### 2022-02-14 wham_simulation/tutorial/vignette_10_wham_simulation.Rmd
add html vignette on simulation

---
title: "Ex 10: Simulation Studies"
output: 
  rmarkdown::html_vignette:
   fig_width: 7
   fig_height: 6
vignette: >
  %\VignetteIndexEntry{Ex 10: Simulation Studies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
wham.dir <- find.package("wham")
knitr::opts_knit$set(root.dir = file.path(wham.dir,"extdata"))
library(knitr)
library(kableExtra)
library(here)
library(wham)
```

# 1. Background

In this vignette we show how to use WHAM for general simulation studies. There is some overlap with the 10th WHAM vignette on Operating models and MSE which may be replaced with this and other vignettes.

In this vignette we show how to:

- Simulate data from a fitted model with input derived from an ASAP3 dat file.
- Make an operating model, simulate data, and fit models.
- Fit models to data simulated from a different model.
- Make a default wham input file **without an ASAP3 dat file**.
- Configure different assumptions about the population for both the operating model and the estimating model.
- Do a closed-loop simulation with operating model, fitting an estimating model, generating catch advice and incorporating it into the operating model.

# 2. Setup and read in model input

Currently this is working out of the SSRTWG repository so make sure that your working directory is there.

Create a directory for this analysis:
```{r, eval=FALSE}
# devtools::install_github("timjmiller/wham", dependencies=TRUE, ref="devel")
source(file.path(here(),"wham_simulation","tutorial","helper.R")) #, local = knitr::knit_global()

# choose a location to save output, otherwise will be saved in working directory
#write.dir <- file.path(here(), "wham_simulation", "results") #SSRTWG results
#write.dir <- "choose/where/to/save/output"
#dir.create(write.dir)
#setwd(write.dir)
```

```{r, eval=TRUE}
wham.dir <- find.package("wham")
asap.file.path = file.path(wham.dir,"extdata","ex1_SNEMAYT.dat")
asap3 <- read_asap3_dat(asap.file.path)

```

Make an input and fit a model in wham. This model was also fit in the first example vignette. 
```{r, eval=TRUE}
#define selectivity model
selectivity=list(
    model=rep("age-specific",3), #define selectivity model
    re=rep("none",3), #define selectivity random effects model
    #define initial/fixed values for age-specific selectivity
    initial_pars=list(c(0.5,0.5,0.5,1,1,0.5),c(0.5,0.5,0.5,1,0.5,0.5),c(0.5,1,1,1,0.5,0.5)),
    fix_pars=list(4:5,4,2:4)) #which ages to not estimate age-specfic selectivity parameters

NAA_re = list(
  sigma="rec", #random effects for recruitment only
  cor="iid", #random effects are independent
  recruit_model = 2) #mean recruitment is the only fixed effect parameter

input <- prepare_wham_input(
    asap3, 
    selectivity = selectivity,
    NAA_re = NAA_re,
    model_name="Ex 1: SNEMA Yellowtail Flounder")

mod <- fit_wham(input, do.osa = F, do.retro=F, MakeADFun.silent = T) # don't do retro peels and OSA residuals, don't show output during optimization

names(mod)
```

Several of the elements in the object returned by `fit_wham` are those produced by `TMB::MakeADFun`. `mod$par` are the fixed effects or parameters over which the marginal likelihood is optimized. `mod$fn` is the function that returns the (Laplace approximation of) the negative marginal log-likelihood. `mod$gr` is the function that returns the gradient and `mod$he` returns the hessian. `mod$report` provides a list of output defined by REPORT() calls in the compiled cpp code and is already evaluated and provided by `mod$rep`. When available, `mod$sdrep` provides the hessian based standard errors for estimates from the model which is generated by `TMB::sdreport`.

```{r, eval=TRUE}
mod$sdrep #the standard errors of the fixed effects

```

The important item for this vignette is `mod$simulate`

```{r, eval=TRUE}
mod$simulate
```
As we can see it is a function, but less apparent is that it will report anything using the `REPORT` calls, and most importantly, those within SIMULATE statements in the compiled cpp code. More information about simulation using the TMB package can be found [here](http://kaskr.github.io/adcomp/_book/Simulation.html). The WHAM cpp code is written to be able to simulate all observations and any random effects that are specified to be part of the particular model configuration. `mod$simulate` is what we use to simulate data for simulation studies. The `complete` argument tells the function to provide all data inputs whether they are simulated or not. For a wham model this would include important things like weight and maturity at age.

Here is a simple function that will simulate data from an operating model. It will fit the same model configuration to that data with the argument `self.fit=TRUE`.
```{r, eval=TRUE}

sim_fn <- function(om, self.fit = FALSE){
	input <- om$input
	input$data = om$simulate(complete=TRUE)
	if(self.fit) {
	  fit <- fit_wham(input, do.osa = F, do.retro = F, MakeADFun.silent = T)
	  return(fit)
	} else return(input) 
}
set.seed(123)
self_sim_fit <- sim_fn(mod, self.fit = TRUE)

plot(mod$years, self_sim_fit$input$data$SSB, type = 'l', ylab = "SSB", xlab = "Year")
lines(mod$years, self_sim_fit$rep$SSB, lty = 2, col = 'red')
```

Note that we just have to over-write the data component of the input with the values produced by `om$simulate(complete=TRUE)`. The call of `sim_fn` will fit the same model assumed to simulate the data. `fit_wham` saves the input used to fit the model and `om$simulate(complete=TRUE)` will provide all model output provided by `REPORT` calls in the cpp file. This includes the true SSB for the simulated data which can be compared with the values estimated from the simulated data.

Obviously for a simulation study many simulated data sets and model fits are needed for a given operating model. There are many ways to accomplish this task. It is often desirable to first simulate all the data before fitting each of them and to update results with each fit so that results are not lost if a catastrophic error occurs. Here is a lo-tech approach to simulating 100 data inputs and save information from fits of the same model to each of those data sets.

```{r, eval=FALSE}
set.seed(123)
sim_inputs <- replicate(100, sim_fn(mod), simplify=F)
res = list(reps = list(), par.est = list(), par.se = list(), adrep.est = list(), adrep.se =list())
for(i in 1:length(sim_inputs)){
  cat(paste0("i: ",i, "\n"))
  tfit = fit_wham(sim_inputs[[i]], do.osa = F, do.retro = F, MakeADFun.silent = T)
  res$reps[[i]] = tfit$rep
  res$par.est[[i]] = as.list(tfit$sdrep, "Estimate")
  res$par.se[[i]] = as.list(tfit$sdrep, "Std. Error")
  res$adrep.est[[i]] = as.list(tfit$sdrep, "Estimate", report = TRUE)
  res$adrep.se[[i]] = as.list(tfit$sdrep, "Std. Error", report = TRUE)
}
#these results have already been saved at wham_simulations/tutorial/vignette_10_res_1.RDS
saveRDS(res, file = file.path(here(), "wham_simulation","tutorial", "vignette_10_res_1.RDS"))
#these results have already been saved at wham_simulations/tutorial/vignette_10_sim_inputs.RDS
saveRDS(sim_inputs, file = file.path(here(), "wham_simulation","tutorial", "vignette_10_sim_inputs.RDS"))
```
The `replicate` function is used to generate 100 simulated inputs. Then we loop over the simulated data sets, fit each of them and save results for each fit. The results saved are the values provided by `REPORT` calls in the cpp file, the fixed effects MLEs (and PEBEs for random effects) and their standard errors, and the estimates for values provided by `ADREPORT` calls in the cpp file and their standard errors. See `?TMB::as.list.sdreport` for more information about accessing sdreport information like this. Saving these attributes will allow one to make inferences about bias of the maximum likelihood estimators, and corresponding standard errors, and confidence interval coverage for the basic parameters as well as a variety of derived model output (e.g., SSB, reference points).

The results above are available in the SSRTWG repository. Below is a plot of bias and 95% confidence intervals for SSB estimation.

```{r, eval=TRUE}
res = readRDS(file.path(here(), "wham_simulation","tutorial", "vignette_10_res_1.RDS"))
sim_inputs = readRDS(file.path(here(), "wham_simulation","tutorial", "vignette_10_sim_inputs.RDS"))
true = sapply(sim_inputs, function(x) x$data$SSB)
est = sapply(res$rep, function(x) return(x$SSB))
SSB_rel_resid = est/true - 1
resid_cis = apply(SSB_rel_resid,1,mean) + apply(SSB_rel_resid,1,sd)*qnorm(0.975)*t(matrix(c(-1,1),2,44))/10

plot(mod$years, apply(SSB_rel_resid,1,mean), ylim = c(-0.1,0.05), ylab = "Estimated Relative Bias SSB", xlab = "Year")
lines(mod$years, resid_cis[,1])
lines(mod$years, resid_cis[,2])
abline(h = 0, lty = 2)

```

# 3. Fitting a model different from the operating model

When the fitted model is different from that used to simulate the data, there is a bit more work involved. Some components of the input used by `fit_wham` are important. 
```{r, eval=TRUE}
names(input)
```

The first 4 elements of the input (`data`, `par`, `map`, `random`) are those used by TMB::MakeADFun to set up an objective function with derivative information (i.e.,gradient, hessian) used by the optimizer nlminb in fit_wham. We saw above that we could just swap the entire `data` element of the input when the estimation model is the same as the operating model. More generally though, components of these 4 input lists will need to be modified for the simulation studies where the operating model and estimation model are not the same. If the available data is the same for the operating and fitted model only a few elements of the input$data list need to be moved.

## 3a. An alternative age composition likelihood for the estimating model

Below is an example where an alternative likelihood for the age composition observations is assumed for the estimation model.
```{r, eval=TRUE}

input_em <- prepare_wham_input(
    asap3, 
    recruit_model=2, 
    model_name="Ex 1: SNEMA Yellowtail Flounder",
    selectivity=selectivity,
	  NAA_re = NAA_re,
    age_comp = "logistic-normal-miss0") #the only difference from original input

#only replace the observations so that the other inputs for configuration are correct.
#IMPORTANT TO PROVIDE obsvec!
sim_fn = function(om, input, do.fit = FALSE){
	obs_names = c("agg_indices","agg_catch","catch_paa","index_paa", "Ecov_obs", "obsvec")
	om_data = om$simulate(complete = TRUE)
	input$data[obs_names] = om_data[obs_names]
	if(do.fit) {
		fit = fit_wham(input, do.osa = F, do.retro = F, MakeADFun.silent = TRUE)
		return(list(fit, om_data))
	} else return(list(input,om_data))
}
set.seed(123)
simfit = sim_fn(mod, input_em, do.fit = TRUE)
plot(mod$years, simfit[[2]]$SSB, type = 'l', ylab = "SSB", xlab = "Year")
lines(mod$years, simfit[[1]]$rep$SSB, lty = 2, col = 'red')
```

It is very important that the `obsvec` component of the simulated data is put into the input used for fitting because this holds the values in `agg_catch`, `agg_indices`, etc. in vector form for evaluating the likelihood. This is necessary to allow one-step-ahead residuals to be calculated using TMB. Note also because only a few elements are copied into `input$data` we need to save all of the true values of the simulated population as a separate component that is returned. For this particular simulation, SSB is estimated greater than the truth when the logistic normal is assumed but observations were generated by a multinomial. The same sort of approach as above can be used for simulating and fitting many data sets.

## 3b. Effective sample size assumption

Suppose we wanted to see how the estimating model behaves when we assume the multinomial age composition is more precise than it really is. Here we assume we want to use all of the same parameter estimates from the original model but change the necessary fixed data inputs. We modify the input with the appropriate data and populate `input$par` with the estimates from the original fit. Then we create this alternative operating model by calling `fit_wham(input, do.fit=F)` which creates the model configuration with the dll so that data can be simulated with this particular configuruation. Note that `do.fit=F` means that whatever parameter values are set (normally as initial and fixed values) in `input$par` will be used to simulate data. For the estimating model we can use the same input as the operating model except for changing the inputs for the effective sample size.

```{r, eval=TRUE}
sim_fn = function(om, Neff_om = 50, Neff_em = 200, do.fit = FALSE){
  om_input = om$input #the original input
	om_input$data$index_Neff[] = Neff_om #true effective sample size for the indices
	om_input$data$catch_Neff[] = Neff_om #true effective sample size for the indices
	om_input$par = as.list(om$sdrep, "Estimate") #assume parameter values estimated from original model
  om_alt = fit_wham(om_input, do.fit = F, MakeADFun.silent=T) #make unfitted model for simulating data
	om_data = om_alt$simulate(complete = TRUE)
	em_input = om_input #the same configuration other than Neff
	em_input$data$index_Neff[] = Neff_em #assumed effective sample size for the indices
	em_input$data$catch_Neff[] = Neff_em #assumed effective sample size for the indices
	obs_names = c("agg_indices","agg_catch","catch_paa","index_paa", "Ecov_obs", "obsvec")
	em_input$data[obs_names] = om_data[obs_names]
	if(do.fit) {
		fit = fit_wham(input, do.osa = F, do.retro = F, MakeADFun.silent = TRUE)
		return(list(fit, om_data))
	} else return(list(em_input,om_data))
}
set.seed(123)
simfit = sim_fn(mod, do.fit = TRUE)
plot(mod$years, simfit[[2]]$SSB, type = 'l', ylab = "SSB", xlab = "Year")
lines(mod$years, simfit[[1]]$rep$SSB, lty = 2, col = 'red')
```

Note that we could make the assumed effective sample size only incorrect for some subset of the fleets and indices. In the `sim_fn` function here we could have also just supplied all of `om_data` to `em_input$data` and then changed the Neff appropriately rather than just change the observations. Also, note that changing the Neff will only affect fits to those indices and fleets where the multinomial age comp assumption is made and where age composition observations are available.

The incorrect age composition precision has a large effect here!

## 3c. Population assumptions

Suppose we wanted to evaluate the effect of assuming a model with just random effects on recruitment when a full state-space model was the operating model. We will fit the full state-space model to the example data to get a configuration for the operating model then simulate data from it and fit the recruitment-only model to it.

```{r, eval=TRUE}

NAA_re_om = list(
  sigma="rec+1", #random effects for recruitment and older age classes
  cor="iid", #random effects are independent
  recruit_model = 2) #mean recruitment is the only fixed effect parameter

NAA_re_em = list(
  sigma="rec", #random effects for recruitment only
  cor="iid", #random effects are independent
  recruit_model = 2) #mean recruitment is the only fixed effect parameter

input_om <- prepare_wham_input(
    asap3, 
    selectivity = selectivity,
    NAA_re = NAA_re_om,
    model_name="Ex 1: SNEMA Yellowtail Flounder")

input_em <- prepare_wham_input(
    asap3, 
    selectivity = selectivity,
    NAA_re = NAA_re_em,
    model_name="Ex 1: SNEMA Yellowtail Flounder")

om_ss <- fit_wham(input_om, do.osa = F, do.retro=F, MakeADFun.silent = T) # don't do retro peels and OSA residuals, don't show output during optimization

sim_fn = function(om, input, do.fit = FALSE){
	obs_names = c("agg_indices","agg_catch","catch_paa","index_paa", "Ecov_obs", "obsvec")
	om_data = om$simulate(complete = TRUE)
	input$data[obs_names] = om_data[obs_names]
	if(do.fit) {
		fit = fit_wham(input, do.osa = F, do.retro = F, MakeADFun.silent = TRUE)
		return(list(fit, om_data))
	} else return(list(input,om_data))
}
set.seed(123)
simfit = sim_fn(om_ss, input_em, do.fit = TRUE)
plot(om_ss$years, simfit[[2]]$SSB, type = 'l', ylab = "SSB", xlab = "Year")
lines(om_ss$years, simfit[[1]]$rep$SSB, lty = 2, col = 'red')

```

The scale of the population estimates from the recruitment-only model is similar to the true value, but it does not have sufficient flexibility to capture the inter-annual variability in the population size.

## 3d. Selectivity assumptions

Suppose the true selectivity is that estimated from the original model, but that we wanted to evaluate the effect of assuming logistic selectivity. Use the same `NAA_re` defined earlier but change the selectivity 
```{r, eval=TRUE}

selectivity_em = list(
  model = c(rep("logistic", input$data$n_fleets),rep("logistic", input$data$n_indices)),
  initial_pars = rep(list(c(5,1)), input$data$n_fleets + input$data$n_indices)) #fleet, index

input_em <- prepare_wham_input(
    asap3, 
    selectivity = selectivity_em,
    NAA_re = NAA_re,
    model_name="Ex 1: SNEMA Yellowtail Flounder") 
sim_fn = function(om, input, do.fit = FALSE){
	obs_names = c("agg_indices","agg_catch","catch_paa","index_paa", "Ecov_obs", "obsvec")
	om_data = om$simulate(complete = TRUE)
	input$data[obs_names] = om_data[obs_names]
	if(do.fit) {
		fit = fit_wham(input, do.osa = F, do.retro = F, MakeADFun.silent = TRUE)
		return(list(fit, om_data))
	} else return(list(input,om_data))
}
set.seed(123)
simfit = sim_fn(mod, input_em, do.fit = TRUE)
plot(mod$years, simfit[[2]]$SSB, type = 'l', ylab = "SSB", xlab = "Year")
lines(mod$years, simfit[[1]]$rep$SSB, lty = 2, col = 'red')
plot(mod$years, simfit[[2]]$SSB/exp(simfit[[2]]$log_SSB_FXSPR), type = 'l', ylab = "SSB/SSB(F40)", xlab = "Year")
lines(mod$years, simfit[[1]]$rep$SSB/exp(simfit[[1]]$rep$log_SSB_FXSPR), lty = 2, col = 'red')

```

Lower SSB and stock status are estimated if logistic selectivity is assumed.

# 4. Creating a general operating model without an ASAP3 dat file

Note that the simulated data are conditioned on the fixed effects parameter values. If a model is fitted using `fit_wham` then the optimized values will be used `mod$simulate` for simulations, but these parameters can be specified without fitting the model and simulations can be more generally configured. Here we will use the functionality of `prepare_wham_input` to create a model object without an asap3 dat file and, then set parameter values we want to assume for the operating model

First we will read in a function `make_basic_info` that will make a list that can be provided to the `basic_info` argument of `prepare_wham_input`. The function is defined in the helper.R file and the elements of the output it provides are defined in the help file for `prepare_wham_input`. 
```{r, eval=FALSE}
make_basic_info <- function(base_years = 1982:2021, ages = 1:10, Fhist = "updown", n_feedback_years = 0) { #changed years
    info <- list()
    info$ages <- ages
    info$years <- as.integer(base_years[1] - 1 + 1:(length(base_years) + n_feedback_years))
    info$n_fleets <- 1 
    info$n_indices <- 1
    na <- length(info$ages)
    ny <- length(info$years)

    nby <- length(base_years)
    mid <- floor(nby/2)
    #up then down
    if(Fhist == "updown") info$F <- matrix(0.2 + c(seq(0,0.4,length.out = mid),seq(0.4,0,length.out=nby-mid)),nby, info$n_fleets)

    #down then up
    if(Fhist == "downup") info$F <- matrix(0.2 + c(seq(0.4,0,length.out = mid),seq(0,0.4,length.out=nby-mid)),nby, info$n_fleets)

    if(n_feedback_years>0) info$F <- rbind(info$F, info$F[rep(nby, n_feedback_years),, drop = F]) #same F as terminal year for feedback period

    info$catch_cv <- matrix(0.1, ny, info$n_fleets)
    info$catch_Neff <- matrix(200, ny, info$n_fleets)

    info$index_cv <- matrix(0.3, ny, info$n_indices)
    info$index_Neff <- matrix(100, ny, info$n_indices)
    info$fracyr_indices <- matrix(0.5, ny, info$n_indices)
    info$index_units <- rep(1, length(info$n_indices)) #biomass
    info$index_paa_units <- rep(2, length(info$n_indices)) #abundance

    info$maturity <- t(matrix(1/(1 + exp(-1*(1:na - na/2))), na, ny))

    L <- 100*(1-exp(-0.3*(1:na - 0)))
    W <- exp(-11)*L^3
    nwaa <- info$n_indices + info$n_fleets + 2
    info$waa <- array(NA, dim = c(nwaa, ny, na))
    for(i in 1:nwaa) info$waa[i,,] <- t(matrix(W, na, ny))

    info$fracyr_SSB <- rep(0.25,ny)
    info$q <- rep(0.3, info$n_indices)

    info$selblock_pointer_fleets <- t(matrix(1:info$n_fleets, info$n_fleets, ny))
    info$selblock_pointer_indices <- t(matrix(info$n_fleets + 1:info$n_indices, info$n_indices, ny))
    return(info)
}

```
Various attributes are specified in the make_basic_info function including the number of fleets and indices, the years over which the model spans, the number ages, precision of aggregate catch and index observations. Maturity and weight at age are defined as well as catchabilities for the indices, when they occur during the year and the selectiivty blocks to assign to each fleet and index.

Many of the other necessary attributes and initial parameters can be defined using the arguments `NAA_re`, `selectivity`, `M` to `prepare_wham_input`.  Below we use specify logistic selectivity and associated parameters (a50 and /1slope), M = 0.2 and initial numbers at age and mean recruitment all using the lists supplied to these arguments of prepare_wham_input. However, nowhere have we specified a value for the standard deviation of the (log) recruitment random effects and so, the initial value (sd = 1) given in `prepare_wham_input` will be used. Importantly, `fit_wham` must be called with `do.fit=FALSE` to set up the model to be simulated with the parameter values specified rather than estimated.

```{r, eval=TRUE}
source(file.path(here(),"wham_simulation","tutorial","helper.R")) #, local = knitr::knit_global()
stock_om_info = make_basic_info()

selectivity_om = list(
  model = c(rep("logistic", stock_om_info$n_fleets),rep("logistic", stock_om_info$n_indices)),
  initial_pars = rep(list(c(5,1)), stock_om_info$n_fleets + stock_om_info$n_indices)) #fleet, index

M_om = list(initial_means = rep(0.2, length(stock_om_info$ages)))

NAA_re_om = list(
  N1_pars = exp(10)*exp(-(0:(length(stock_om_info$ages)-1))*M_om$initial_means[1]),
  sigma = "rec", #random about mean
  cor="iid", #random effects are independent
  use_steepness = 0,
  recruit_model = 2, #random effects with a constant mean
  recruit_pars = exp(10)
)

stock_om_input = prepare_wham_input(basic_info = stock_om_info, selectivity = selectivity_om, NAA_re = NAA_re_om, M = M_om)

stock_om = fit_wham(stock_om_input, do.fit = FALSE, MakeADFun.silent = TRUE)

```

As stated above, many of the parameter values are specified in the lists supplied to `prepare_wham_input`, but not all of them. In particular, we have not specified the variance parameter for the recruitment random effects so the default value for fitting a wham model is set.

```{r}
stock_om_input$par$log_NAA_sigma
```
which will mean a standard deviation = 1 is assumed. In general, the parameters defining the variance and correlation of any random effects and any effects of environmental covariates currently need to be specified by changing the input returned by `prepare_wham_input`.

## 4a. Setting NAA random effects variance parameters.

There are two parameter elements that we need to consider when using these random effects: `log_NAA_sigma` as mentioned above and `trans_NAA_rho` which defines the autocorrelation parameters on a logit scale with the parameter bounds of (-1,1). The latter need to be specified when the `cor` element of the `NAA_re` list is not `'iid'`. So here we will consider a 2d AR1 autoregressing model for the numbers at age random effects to demonstrate setting the parameters. We will suppose that we want to assume standard deviations of 0.6 for recruitment and 0.2 for older ages, and correlations of 0.4 temporally and 0.7 with age.

```{r}
stock_om_info = make_basic_info()

M_om = list(initial_means = rep(0.2, length(stock_om_info$ages)))

selectivity_om = list(
  model = c(rep("logistic", stock_om_info$n_fleets),rep("logistic", stock_om_info$n_indices)),
  initial_pars = rep(list(c(5,1)), stock_om_info$n_fleets + stock_om_info$n_indices)) #fleet, index

NAA_re_om = list(
  N1_pars = exp(10)*exp(-(0:(length(stock_om_info$ages)-1))*M_om$initial_means[1]),
  sigma = "rec+1", #random about mean
  cor="2dar1", #random effects correlated among age classes and across time (separably)
  use_steepness = 0,
  recruit_model = 2, #random effects with a constant mean
  recruit_pars = exp(10)
)

stock_om_input = prepare_wham_input(basic_info = stock_om_info, selectivity = selectivity_om, NAA_re = NAA_re_om, M = M_om)

stock_om_input$par$log_NAA_sigma
stock_om_input$par$log_NAA_sigma = log(c(0.6,0.2))
stock_om_input$par$log_NAA_sigma

stock_om_input$par$trans_NAA_rho
stock_om_input$par$trans_NAA_rho = wham:::gen.logit(c(0.7,0.4), -1, 1) #age first, year second
stock_om_input$par$trans_NAA_rho

```
Note we made use of an unexported function from the wham package that will do the transformation of parameters estimated on the logit scale. Not also, that the correlation parameter for age is first and that for time is second in the `trans_NAA_rho` vector. If `cor = 'ar1_a'` then we would only change the first parameter in `trans_NAA_rho` and if  `cor = 'ar1_y'` we would only change the second parameter. Leaving the other value at zero makes those correlation parameters zero and the random effects independent in the appropriate way. Now when `stock_om$simulate()` is called the variability will be as we specified.

## 4b. Setting M random effects variance parameters

M random effects assumptions are similar to the NAA random effects, but the parameter names are different. For M there is a single vector (`M_repars`) of 3 values that defines the transformed M variance and autoregressing parameters: standard deviation, correlation with age, correlation with year. Here we assume we have random effects on recruitment and on natural mortality. As above, we define the general model to have 2d autocorrelation of M random effects for the most general case with variance of the (log) M random effects = 0.1 and correlations with age and time  of 0.7 and 0.4.

```{r}
stock_om_info = make_basic_info()

M_om = list(
  model = "constant", # a single mean M for all ages
  initial_means = 0.2,
  re = "2dar1")

selectivity_om = list(
  model = c(rep("logistic", stock_om_info$n_fleets),rep("logistic", stock_om_info$n_indices)),
  initial_pars = rep(list(c(5,1)), stock_om_info$n_fleets + stock_om_info$n_indices)) #fleet, index

NAA_re_om = list(
  N1_pars = exp(10)*exp(-(0:(length(stock_om_info$ages)-1))*M_om$initial_means[1]),
  sigma = "rec", #random about mean
  use_steepness = 0,
  recruit_model = 2, #random effects with a constant mean
  recruit_pars = exp(10)
)

stock_om_input = prepare_wham_input(basic_info = stock_om_info, selectivity = selectivity_om, NAA_re = NAA_re_om, M = M_om)

stock_om_input$par$log_NAA_sigma
stock_om_input$par$log_NAA_sigma = log(0.6)
stock_om_input$par$log_NAA_sigma

stock_om_input$par$M_repars
stock_om_input$par$M_repars = c(log(0.1), wham:::gen.logit(c(0.7,0.4),-1,1))
stock_om_input$par$M_repars

stock_om = fit_wham(stock_om_input, do.fit = FALSE, MakeADFun.silent = TRUE)

```
Note that the variance of the M random effects we are assuming is the same as the default value for estimation. Note also, that the mean of the M autoregressing random effects is defined by the  `initial_means` component of the `M` argument to `prepare_wham_input` and it is not estimated by default. To estimate the mean of the M random effects at age, or just estimate a constant M without random effects, specify the `M$est_ages` argument to `prepare_wham_input`. Setting the correlation parameters when `M$re = "ar_a"` or  `M$re = "ar_y"` is analogous to the description for numbers at age random effects.

```{r}
M_om = list(
  model = "constant", # a single mean M for all ages
  initial_means = 0.2,
  est_ages = 1, #must be 1 when model = "constant"
  re = "2dar1")
```

## 4c. Setting selectivity random effects variance parameters

Setting the variance parameters for selectivity is analogous to that for natural mortality random effects. The important difference is that the parameter defining the random effects variance parameters (`sel_repars`) is a matrix with rows for each selectivity block and the columns as defined for `M_repars`. Here we have two selectivity blocks: one for the fleet and one for the index. We will assume that the variance parameter for the fleet is 0.4 and that for the index is 0.1, but they both have the same correlation parameters with age and time: 0.7 and 0.4.

```{r}
stock_om_info = make_basic_info()

M_om = list(
  model = "constant", # a single mean M for all ages
  initial_means = 0.2)

selectivity_om = list(
  model = c(rep("logistic", stock_om_info$n_fleets),rep("logistic", stock_om_info$n_indices)),
  initial_pars = rep(list(c(5,1)), stock_om_info$n_fleets + stock_om_info$n_indices),
  re = rep("2dar1", stock_om_info$n_fleets + stock_om_info$n_indices))

NAA_re_om = list(
  N1_pars = exp(10)*exp(-(0:(length(stock_om_info$ages)-1))*M_om$initial_means[1]),
  sigma = "rec", #random about mean
  use_steepness = 0,
  recruit_model = 2, #random effects with a constant mean
  recruit_pars = exp(10)
)

stock_om_input = prepare_wham_input(basic_info = stock_om_info, selectivity = selectivity_om, NAA_re = NAA_re_om, M = M_om)

stock_om_input$par$log_NAA_sigma
stock_om_input$par$log_NAA_sigma = log(0.6)
stock_om_input$par$log_NAA_sigma

stock_om_input$par$sel_repars
stock_om_input$par$sel_repars[1,] = c(log(0.4), wham:::gen.logit(c(0.7,0.4),-1,1))
stock_om_input$par$sel_repars[2,] = c(log(0.1), wham:::gen.logit(c(0.7,0.4),-1,1))
stock_om_input$par$sel_repars

stock_om = fit_wham(stock_om_input, do.fit = FALSE, MakeADFun.silent = TRUE)

```
Note in this case where we assume logistic selectivity and 2D AR1 random effects, the correlation with "age" is actually just the correlation of the two logistic selectivity parameters on their estimated/transformed scale. Also, differing from natural mortality the mean parameters for the selectivity random effects (`logit_selpars`) are estimated by default.

## 4d. Setting catchability random effects variance parameters.

Setting the variance parameters for catchability is analogous to that for natural mortality and selectivity random effects.  The parameters defining the variance and correlation of the annual catchability are defined in `q_repars` which is a matrix with rows corresponding to different indices and 2 columns for the standard deviation and correlation. The mean of the process is defined and estimated in the `logit_q` parameter which is defined in the `make_basic_info` function. Here we will assume the transformed catchability is an AR1 process with standard deviation 0.1 and autocorrelation = 0.5.

```{r, eval=FALSE}
stock_om_info = make_basic_info()

M_om = list(
  model = "constant", # a single mean M for all ages
  initial_means = 0.2)

selectivity_om = list(
  model = c(rep("logistic", stock_om_info$n_fleets),rep("logistic", stock_om_info$n_indices)),
  initial_pars = rep(list(c(5,1)), stock_om_info$n_fleets + stock_om_info$n_indices),
  re = rep("2dar1", stock_om_info$n_fleets + stock_om_info$n_indices))

NAA_re_om = list(
  N1_pars = exp(10)*exp(-(0:(length(stock_om_info$ages)-1))*M_om$initial_means[1]),
  sigma = "rec", #random about mean
  use_steepness = 0,
  recruit_model = 2, #random effects with a constant mean
  recruit_pars = exp(10)
)

catchability_om = list(re = "ar1") #could also define the mean q parameter here.

stock_om_input = prepare_wham_input(basic_info = stock_om_info, selectivity = selectivity_om, NAA_re = NAA_re_om, M = M_om, catchability = catchability_om)

stock_om_input$par$log_NAA_sigma
stock_om_input$par$log_NAA_sigma = log(0.6)
stock_om_input$par$log_NAA_sigma

stock_om_input$par$q_repars
stock_om_input$par$q_repars[1,] = c(log(0.1), wham:::gen.logit(0.5,-1,1))
stock_om_input$par$q_repars

stock_om = fit_wham(stock_om_input, do.fit = FALSE, MakeADFun.silent = TRUE)

set.seed(123)
sim_q = stock_om$simulate(complete=TRUE)

plot(stock_om$years, sim_q$q[,1], type = 'l', ylab = "Catchability", xlab = "Year")

```

## 4e. Setting effects of Environmental covariates.

When environmental effects are considered, there are assumptions to be made about the variance and any autocorrelation parameters for the random effects for the latent environmental process and the variance of the observation errors. The latter are specified in the the `ecov$log_sigma` argument to `prepare_wham_input`, but those for the random effects (`Ecov_process_pars`) must be specified after the call to `prepare_wham_input` like the other random effects described above. The value for the effect of the covariate on the population (`Ecov_beta`) must also be specified this way. The columns of `Ecov_process_pars` correspond to each environmental covariate being modeled and the number of rows is 3. If `ecov$process_model = "rw"` only the first two elements are used and represent the value in the first year and the conditional standard devation parameter of the random walk. When `ecov$process_model = "ar1"` the three elements correspond to mean, variance, and correlation parameters.

The `Ecov_beta` parameter is a 4-dimensional array to be able to use it across all possible types of effects in a wham model. The first dimension indicates where the effect will be applied: recruitment, natural mortality, index 1,... index n. The second dimension is the order of the polynomial effect of the covariate. The third dimension indicates which environmental covariate the effect is having on the population. The fourth dimension indicates which age class the effect is on (for natural mortality only).

As long as there is only a single environmental effect, we can change the appropriate element of the `Ecov_beta` array by using the `map$Ecov_beta` specification returned by `prepare_wham_input`. The `map$Ecov_beta` will be all NAs except for the configured effects on the population and when there is just one there is no ambiguity. Otherwise, the 4 correct indices of the `Ecov_beta` array must be referenced appropriately for each effect.

We will assume that our errors on the observations of the environmental covariate have a standard deviation = 0.2, the latent process is AR1 with mean 0, standard deviation = 0.3 and correlation parameter 0.5. We will also assume the effect is on (log) recruitment (no stock-recruit function) and the size of the effect is `Ecov_beta = 0.3`.

```{r}
stock_om_info = make_basic_info()

#constand M fixed at 0.2
M_om = list(
  model = "constant", # a single mean M for all ages
  initial_means = 0.2)

#logistic selectivity for fleet and index
selectivity_om = list(
  model = c(rep("logistic", stock_om_info$n_fleets),rep("logistic", stock_om_info$n_indices)),
  initial_pars = rep(list(c(5,1)), stock_om_info$n_fleets + stock_om_info$n_indices))

#
NAA_re_om = list(
  N1_pars = exp(10)*exp(-(0:(length(stock_om_info$ages)-1))*M_om$initial_means[1]),
  sigma = "rec", #random about mean
  recruit_model = 2, #random effects with a constant mean
  recruit_pars = exp(10)
)

ecov_om = list(
    label = "Climate variable 1",
    process_model = "rw",
    mean = matrix(0, length(stock_om_info$years), 1), #observations which will be simulated later
    logsigma = log(0.2), 
    lag = 0,
    years = stock_om_info$years, 
    use_obs = matrix(1,length(stock_om_info$years),1),  
    where = "recruit",
    how = 1)

stock_om_input = prepare_wham_input(basic_info = stock_om_info, selectivity = selectivity_om, NAA_re = NAA_re_om, M = M_om, ecov = ecov_om)

#same assumption as above about standard deviation of recruitment.
stock_om_input$par$log_NAA_sigma
stock_om_input$par$log_NAA_sigma = log(0.6)
stock_om_input$par$log_NAA_sigma

#set sd and rho of ecov processes
stock_om_input$par$Ecov_process_pars[] = c(0, log(0.3), wham:::gen.logit(0.5,-1,1)) #mean, log(sd), logit(0.5)
#set size of Ecov_beta
stock_om_input$par$Ecov_beta[!is.na(stock_om_input$map$Ecov_beta)] = 0.3
stock_om_input$par$Ecov_beta[1,1,1,] #all = 0.3, but only the first one is used and all values mapped to be equal.

stock_om = fit_wham(stock_om_input, do.fit = FALSE, MakeADFun.silent = TRUE)

set.seed(123)
sim_Ecov = stock_om$simulate(complete=TRUE)

#par(mfrow = c(2,1))
#plot(stock_om$years, sim_Ecov$Ecov_x, type = 'l', ylab = "Covariate", xlab = "Year")
#plot(stock_om$years, sim_Ecov$NAA[,1], type = 'l', ylab = "Recruitment", xlab = "Year")

```

See the [catchability vignette](https://github.com/timjmiller/wham/blob/devel/vignettes/ex11_catchability.Rmd) for an example of simulating effects of multiple covariates on the population. Also see the [CPI and recruitment vignette](https://timjmiller.github.io/wham/articles/ex2_CPI_recruitment.html) for configuring effects on recruitment when a stock-recruit function is assumed.

## 4f. Estimation model with incorrect M assumption

Here we show how to set up an estimation model that assumes the wrong natural mortality rate (M = 0.3), but other aspects of the estimation model are consistent with the operating model. The `sim_fn` function will return the simulated population and data and either the input for the estimating model or the fitted estimation model.

```{r, eval=TRUE}
M_em = list(initial_means = rep(0.3, length(stock_om_info$ages)))

stock_em_input = prepare_wham_input(basic_info = stock_om_info, selectivity = selectivity_om, NAA_re = NAA_re_om, M = M_em)

sim_fn = function(om, em_input, do.fit = FALSE){
	obs_names = c("agg_indices","agg_catch","catch_paa","index_paa", "Ecov_obs", "obsvec")
	om_sim = om$simulate(complete = TRUE)
	em_input$data[obs_names] = om_sim[obs_names]
	if(do.fit) {
		em_fit = fit_wham(em_input, do.osa = F, do.retro = F, MakeADFun.silent = TRUE)
		return(list(em_fit, om_sim))
	} else return(list(em_input,om_sim))
}
set.seed(123)
simfit = sim_fn(stock_om, stock_em_input, do.fit = TRUE)
plot(stock_om$years, simfit[[2]]$SSB, type = 'l', ylab = "SSB", xlab = "Year")
lines(stock_om$years, simfit[[1]]$rep$SSB, lty = 2, col = 'red')
plot(stock_om$years, simfit[[2]]$SSB/exp(simfit[[2]]$log_SSB_FXSPR), type = 'l', ylab = "SSB/SSB(F40)", xlab = "Year")
lines(stock_om$years, simfit[[1]]$rep$SSB/exp(simfit[[1]]$rep$log_SSB_FXSPR), lty = 2, col = 'red')
```

Both SSB and stock status are biased high when M is specified too large.

# 5. Closed-loop simulation

Here we will use the same operating and estimating model assumptions as above where the former is based on M = 0.2 and the latter assumes M = 0.3. To be applicable to a wider class of closed-loop simulations, it will be useful to create an operating model that includes a base period and a feedback period and an estimating model that uses data as it becomes available as time progresses through the feedback period. Previously, we had made use of the projection years as the feedback period, but selectivity random effects are not projected and environmental observations are not simulated in the projection period, both of which may be relevant. The operating model is also updated through the feedback period (redefining F and population state).

There are two other functions defined in the `helper.R` file for generating F from catch advice: `get_F_from_catch`, `update_om_F`. The first, is essentially just a solver for the F in the operating model for a catch that would be provided as management advice using the estimation model, but note that when the catch advice is not possible to actually take from the population or implies F greater than 10, F is set at 10.

```{r, eval=FALSE}
get_F_from_catch <- function(om, year, catch, Finit = 0.1, maxF = 10){
    rep = om$report()
    #print(year)
    naa = rep$NAA[year,]
    Maa = rep$MAA[year,]
    sel_tot = rep$FAA_tot[year,]/max(rep$FAA_tot[year,])
    waa = om$input$data$waa[om$input$data$waa_pointer_totcatch, year,]
    get_catch = function(log_F, naa, sel, waa, Maa){
        Faa = exp(log_F) * sel_tot
        Zaa = Maa + Faa
        Catch = 0
        for(a  in 1:length(naa)) Catch = Catch + waa[a] * naa[a] * Faa[a] *(1 - exp(-Zaa[a]))/Zaa[a];
        return(Catch)
    }
    obj = function(log_F) (catch - get_catch(log_F, naa, sel_tot, waa, Maa))^2
    opt = try(nlminb(log(Finit), obj))
    if(!is.character(opt)) Fsolve = exp(opt$par)[1] else Fsolve = maxF
    if(Fsolve>10) Fsolve = maxF
    print(paste0("Fsolve: ", Fsolve))
    return(Fsolve)
}
```
The second function uses `get_F_from_catch` to get the F and replace the F parameters in the operating model and update the population and observations
```{r, eval=FALSE}
update_om_F = function(om, year, catch){
    rep = om$report() #generate the reported values given the parameters
    year_ind = which(om$years == year) #index corresponding to year
    Fsolve = get_F_from_catch(om, year_ind, catch) #find the F for the catch advice
    #have to be careful if more than one fleet
    sel = rep$FAA[year_ind,,] #n_fleets x n_ages
    sel_all = sel/max(rep$FAA_tot[year_ind,]) #sum(sel_all) = 1
    FAA_all = Fsolve * rbind(sel_all)
    F_fleet = apply(FAA_all, 1, max) #full F for each fleet
    if(year_ind>1) om$input$par$F_devs[year_ind-1,] = log(F_fleet) - log(rep$F[year_ind-1,]) #change the F_dev to produce the right full F
    else om$input$par$log_F1 = log(F_fleet) #if year is the first year of the model, change F in year 1
    om <- fit_wham(om$input, do.fit = FALSE, MakeADFun.silent = TRUE)

}
```

First source the functions and create the stock operating model. We will specify a feedback period of 40 years beyond the 40 year base period.
```{r, eval=TRUE}

source(file.path(here(),"wham_simulation","tutorial","helper.R"))

stock_om_info = make_basic_info(base_years = 1982:2021, n_feedback_years = 40)
stock_om_input = prepare_wham_input(basic_info = stock_om_info, selectivity = selectivity_om, NAA_re = NAA_re_om, M = M_om)
stock_om <- fit_wham(stock_om_input, do.fit = FALSE, MakeADFun.silent = TRUE)
```
Now `stock_om` will be the operating model for a generic stock with biological inputs defined in the `make_basic_info` function. We need to define a function for updating the simulations from the operating model during the feedback period.

```{r, eval=TRUE}
update_om_fn = function(om, seed = 123, interval.info = NULL){
  obs_names = c("agg_indices","agg_catch","catch_paa","index_paa", "Ecov_obs", "obsvec")

  if(!is.null(interval.info)){ #iteratively update F over assessment interval for the given catch advice
    for(y in interval.info$years){
      om = update_om_F(om, year = y, catch = interval.info$catch) #put in the right F values
      set.seed(seed)
      om_sim = om$simulate(complete=TRUE) #resimulate the population and observations
      om$input$data[obs_names] = om_sim[obs_names] #update any simulated data
      om$input$par[om$input$random] = om_sim[om$input$random] #update any simulated random effects
      # reset the om
      om <- fit_wham(om$input, do.fit = F, MakeADFun.silent = TRUE)
    }
  } else { #otherwise just (re)generate the population
    set.seed(seed)
    om_sim = om$simulate(complete=TRUE) #resimulate the population and observations
    om$input$data[obs_names] = om_sim[obs_names] #update any simulated data
    om$input$par[om$input$random] = om_sim[om$input$random] #update any simulated random effects
    # reset the om
    om <- fit_wham(om$input, do.fit = F, MakeADFun.silent = TRUE)
  }
  return(om)
}
```

We now turn to defining the estimation model in a way such that the terminal year can be updated appropriately through the feedback period.

```{r, eval=TRUE}
make_em_input = function(M_em, sel_em, NAA_em, om_data, em_years){
  info = make_basic_info(base_years = em_years) #update the terminal year for the estimation model
  ind_em = 1:length(em_years) #year indices
  #fill in the data from the operating model simulation
  info$agg_catch = om_data$agg_catch[ind_em,, drop = F]
  info$agg_indices = om_data$agg_indices[ind_em,, drop = F]
  info$catch_paa = om_data$catch_paa[,ind_em,, drop = F]
  info$index_paa = om_data$index_paa[,ind_em,, drop = F]
  em_input = prepare_wham_input(basic_info = info, selectivity = sel_em, NAA_re = NAA_em, M = M_em)
}
```

We also need a function that will take the estimated model and generate catch advice from it.

```{r, eval=TRUE}
advice_fn = function(em){
  #make 5 year projections using F40. Use average SSB/R and YPR inputs over most recent 5 years
  proj_opts = list(n.yrs=5, use.FXSPR=TRUE, avg.yrs=tail(em$years,5)) 
  em_proj = project_wham(em, proj.opts = proj_opts, MakeADFun.silent=TRUE) #projected version of the em
  advice = mean(em_proj$rep$pred_catch[length(em_proj$years) + 1:5]) #mean of the projected catch over the next 5 years fishing at F40
  print(advice)
  return(advice)
}

```

Finally, we make a function that uses all the other functions and steps through the feedback period and updating the operating model, refitting the estimation model and generating the catch advice.

```{r, eval=TRUE}
loop_through_fn = function(om, M_em, selectivity_em, NAA_re_em, assess_years, assess_interval = assess.interval, base_years){
  catches = numeric() #save the catch advice
  for(i in assess_years){
    print(i)
    #make the input for the estimation model
    em_input = make_em_input(M_em, selectivity_em, NAA_re_em, om_data = om$input$data, em_years = base_years[1]:i)
    #fit the estimation model
    em = fit_wham(em_input, do.retro = FALSE, do.osa=F, MakeADFun.silent = TRUE) #no feedback period yet
    #make the catch advice
    advice = advice_fn(em)
    catches = c(catches, advice)
    #set the catch for the next assess_interval years
    interval.info = list(catch = advice, years = i + 1:assess_interval)
    #update the operating model with the right Fs and resimulate the data given those Fs
    om = update_om_fn(om, seed = 123, interval.info = interval.info)  
  }
  return(list(om = om, em  = em, catches = catches))
}

```

```{r, eval=TRUE}
stock_om = update_om_fn(stock_om)

assess.interval = 4
base.years = make_basic_info()$years #no feedback period yet
first.year = head(base.years,1)
terminal.year = tail(base.years,1)
assess.years = seq(terminal.year, tail(stock_om$years,1)-assess.interval,by = assess.interval)
```

Running the loop function will take a little while, so we have saved some results for plotting.
```{r, eval = FALSE}
looped_res = loop_through_fn(stock_om, M_em = M_em, selectivity_em = selectivity_om, NAA_re_em = NAA_re, assess_years = assess.years, base_years = base.years)
saveRDS(looped_res$om$rep, file = file.path(here(), "wham_simulation", "tutorial", "vignette_10_looped_res.RDS"))
```

```{r}
looped_rep = readRDS(file = file.path(here(), "wham_simulation", "tutorial", "vignette_10_looped_res.RDS"))
plot(stock_om$years, stock_om$rep$SSB, type = "l", ylim = c(0,400000), ylab = "SSB", xlab = "Year")
lines(stock_om$years, looped_rep$SSB, col = "red")
```

### 2022-02-14 code/flatfish_NAA/example.R

@@ -158,39 +158,39 @@ sapply(sim_fits[[1]], function(x) x$parList$log_NAA_sigma[1])
 ###########################################################################
 #use more generic starting values
 
-make_basic_input <- function(years = 1982:2021) { #changed years
-    digifish = list()
-    digifish$ages = 1:10
-    digifish$years = years
-    na = length(digifish$ages)
-    ny = length(digifish$years)
+make_basic_input <- function(years = 1982:2021, ages = 1:10) { #changed years
+    info = list()
+    info$ages = ages
+    info$years = years
+    na = length(info$ages)
+    ny = length(info$years)
 
-    digifish$n_fleets = 1
-    digifish$catch_cv = matrix(0.1, ny, digifish$n_fleets)
-    digifish$catch_Neff = matrix(200, ny, digifish$n_fleets)
-    digifish$n_indices = 1
-    digifish$index_cv = matrix(0.3, ny, digifish$n_indices)
-    digifish$index_Neff = matrix(100, ny, digifish$n_indices)
-    digifish$fracyr_indices = matrix(0.5, ny, digifish$n_indices)
-    digifish$index_units = rep(1, length(digifish$n_indices)) #biomass
-    digifish$index_paa_units = rep(2, length(digifish$n_indices)) #abundance
-    digifish$maturity = t(matrix(1/(1 + exp(-1*(1:na - na/2))), na, ny))
+    info$n_fleets = 1
+    info$catch_cv = matrix(0.1, ny, info$n_fleets)
+    info$catch_Neff = matrix(200, ny, info$n_fleets)
+    info$n_indices = 1
+    info$index_cv = matrix(0.3, ny, info$n_indices)
+    info$index_Neff = matrix(100, ny, info$n_indices)
+    info$fracyr_indices = matrix(0.5, ny, info$n_indices)
+    info$index_units = rep(1, length(info$n_indices)) #biomass
+    info$index_paa_units = rep(2, length(info$n_indices)) #abundance
+    info$maturity = t(matrix(1/(1 + exp(-1*(1:na - na/2))), na, ny))
 
     L = 100*(1-exp(-0.3*(1:na - 0)))
     W = exp(-11)*L^3
-    nwaa = digifish$n_indices + digifish$n_fleets + 2
[…]

### 2022-02-14 wham_simulation/tutorial/helper.R

@@ -1,0 +1,103 @@
+make_basic_info <- function(base_years = 1982:2021, ages = 1:10, Fhist = "updown", n_feedback_years = 0) { #changed years
+    info <- list()
+    info$ages <- ages
+    info$years <- as.integer(base_years[1] - 1 + 1:(length(base_years) + n_feedback_years))
+    info$n_fleets <- 1 
+    info$n_indices <- 1
+    na <- length(info$ages)
+    ny <- length(info$years)
+    
+    nby <- length(base_years)
+    mid <- floor(nby/2)
+    #up then down
+    if(Fhist == "updown") info$F <- matrix(0.2 + c(seq(0,0.4,length.out = mid),seq(0.4,0,length.out=nby-mid)),nby, info$n_fleets)
+
+    #down then up
+    if(Fhist == "downup") info$F <- matrix(0.2 + c(seq(0.4,0,length.out = mid),seq(0,0.4,length.out=nby-mid)),nby, info$n_fleets)
+
+    if(n_feedback_years>0) info$F <- rbind(info$F, info$F[rep(nby, n_feedback_years),, drop = F]) #same F as terminal year for feedback period
+
+    info$catch_cv <- matrix(0.1, ny, info$n_fleets)
+    info$catch_Neff <- matrix(200, ny, info$n_fleets)
+    
+    info$index_cv <- matrix(0.3, ny, info$n_indices)
+    info$index_Neff <- matrix(100, ny, info$n_indices)
+    info$fracyr_indices <- matrix(0.5, ny, info$n_indices)
+    info$index_units <- rep(1, length(info$n_indices)) #biomass
+    info$index_paa_units <- rep(2, length(info$n_indices)) #abundance
+    
+    info$maturity <- t(matrix(1/(1 + exp(-1*(1:na - na/2))), na, ny))
+
+    L <- 100*(1-exp(-0.3*(1:na - 0)))
+    W <- exp(-11)*L^3
+    nwaa <- info$n_indices + info$n_fleets + 2
+    info$waa <- array(NA, dim = c(nwaa, ny, na))
+    for(i in 1:nwaa) info$waa[i,,] <- t(matrix(W, na, ny))
+
+    info$fracyr_SSB <- rep(0.25,ny)
+    info$q <- rep(0.3, info$n_indices)
+
[…]

### 2022-02-14 wham_simulation/tutorial/linreg.R

@@ -1,0 +1,13 @@
+library(TMB)
+compile("linreg.cpp") #if not already compiled
+dyn.load(dynlib("linreg"))
+set.seed(123)
+data <- list(Y=rnorm(10)+1:10,x=1:10)
+parameters <- list(a=0,b=0,logSigma=0)
+obj <- MakeADFun(data,parameters)
+obj$hessian <- TRUE
+opt <- do.call("optim",obj)
+opt
+opt$hessian ## <-- FD hessian from optim
+obj$he()    ## <-- Analytical hessian
+sdreport(obj)

### 2022-02-14 wham_simulation/tutorial/linreg_sim.cpp

@@ -1,0 +1,18 @@
+#include <TMB.hpp>
+template<class Type>
+Type objective_function<Type>::operator() ()
+{
+  DATA_VECTOR(Y);
+  DATA_VECTOR(x);
+  PARAMETER(a);
+  PARAMETER(b);
+  PARAMETER(logSigma);
+  ADREPORT(exp(2*logSigma));
+  Type nll=-sum(dnorm(Y,a+b*x,exp(logSigma),true));
+  SIMULATE {
+    for(int i = 0; i < Y.size(); i++) Y(i) = rnorm(a+b*x(i), exp(logSigma));
+    REPORT(Y);
+  }
+  return nll;
+}
+

### 2022-02-15 wham_simulation/tutorial/vignette_10_wham_simulation.Rmd
fixed up some issues in the vignette

@@
 `mod$gr` is the function that returns the gradient and `mod$he` returns the hessian.
-`mod$report` provides a list of output defined by REPORT() calls in the compiled cpp code and is already evaluated and provided by `mod$rep`.
-When available, `mod$sdrep` provides the hessian based standard errors for estimates from the model which is generated by `TMB::sdreport`.
+`mod$report` is a function that provides a list of output defined by REPORT() calls in the compiled cpp code and is already evaluated and provided by `mod$rep`.
+When available, `mod$sdrep` provides the hessian based standard errors for estimates from the model which is generated by the`TMB::sdreport` function.
 
@@
 To estimate the mean of the M random effects at age, or just estimate a constant M without random effects, specify the `M$est_ages` argument to `prepare_wham_input`.
-Setting the correlation parameters when `M$re = "ar_a"` or  `M$re = "ar_y"` is analogous to the description for numbers at age random effects.
+There is a difference from numbers at age in how random effects for M are specified.
+If `M$re = "ar_a"` then there are only `n_ages` random effects that are constant over time.
+Similarly for `M$re = "ar_y"`, there are `n_years_model` random effects that are used for each age class.
+However, Setting the correlation parameters when `M$re = "ar_a"` or  `M$re = "ar_y"` is analogous to the description for numbers at age random effects.
+To get `ar_a` or `ar_y` models analogous to numbers at age, one would have to specify `M$re = "2dar1"` and change `map$M_repars` to fix the appropriate parameter fixed at 0.
+To force random effects for certain age classes to be the same, one would also have to configure `map$M_re` appropriately.
 
@@
 
-```{r, eval=FALSE}
+```{r, eval=TRUE}
 stock_om_info = make_basic_info()
@@
 The columns of `Ecov_process_pars` correspond to each environmental covariate being modeled and the number of rows is 3.
-If `ecov$process_model = "rw"` only the first two elements are used and represent the value in the first year and the conditional standard devation parameter of the random walk.
+If `ecov$process_model = "rw"` only the first two elements are used and represent the value in the first year and the conditional standard deviation parameter of the random walk.
 When `ecov$process_model = "ar1"` the three elements correspond to mean, variance, and correlation parameters.
@@
 As long as there is only a single environmental effect, we can change the appropriate element of the `Ecov_beta` array by using the `map$Ecov_beta` specification returned by `prepare_wham_input`.
-The `map$Ecov_beta` will be all NAs except for the configured effects on the population and when there is just one there is no ambiguity.
+The `map$Ecov_beta` will be all `NA` except for the configured effects on the population and when there is just one there is no ambiguity.
 Otherwise, the 4 correct indices of the `Ecov_beta` array must be referenced appropriately for each effect.
@@
 
-#
 NAA_re_om = list(
@@
     label = "Climate variable 1",
-    process_model = "rw",
+    process_model = "ar1",
     mean = matrix(0, length(stock_om_info$years), 1), #observations which will be simulated later
@@
 
-#par(mfrow = c(2,1))
-#plot(stock_om$years, sim_Ecov$Ecov_x, type = 'l', ylab = "Covariate", xlab = "Year")
-#plot(stock_om$years, sim_Ecov$NAA[,1], type = 'l', ylab = "Recruitment", xlab = "Year")
+plot(stock_om$years, sim_Ecov$Ecov_x, type = 'l', ylab = "Covariate", xlab = "Year")
+plot(stock_om$years, sim_Ecov$NAA[,1], type = 'l', ylab = "Recruitment", xlab = "Year")
 
@@
 ```{r, eval=TRUE}
-M_em = list(initial_means = rep(0.3, length(stock_om_info$ages)))
+
+stock_om_input = prepare_wham_input(basic_info = stock_om_info, selectivity = selectivity_om, NAA_re = NAA_re_om, M = M_om)
+
+#same assumption as above about standard deviation of recruitment.
+stock_om_input$par$log_NAA_sigma
+stock_om_input$par$log_NAA_sigma = log(0.6)
+stock_om_input$par$log_NAA_sigma
+
+stock_om = fit_wham(stock_om_input, do.fit = FALSE, MakeADFun.silent = TRUE)
+
+M_em = list(
+  model = "constant", 
+  initial_means = 0.3)
 

### 2022-02-15 wham_simulation/tutorial/linreg.cpp

@@ -8,7 +8,9 @@ Type objective_function<Type>::operator() ()
   PARAMETER(b);
   PARAMETER(logSigma);
   ADREPORT(exp(2*logSigma));
-  Type nll=-sum(dnorm(Y,a+b*x,exp(logSigma),true));
+  vector<Type> Yhat = a + b*x;
+  REPORT(Yhat);
+  Type nll=-sum(dnorm(Y,Yhat,exp(logSigma),true));
   return nll;
 }
 

### 2022-07-01 misspecification_study/misspec_demo.Rmd
start more detailed look at misspecification

---
title: "Misspecification Demonstration"
author: "Chris Legault"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(wham)
library(tidyr)
library(dplyr)
library(here)
library(ggplot2)

source(file.path(here(), "common_code", "make_basic_info.R"))
source(file.path(here(), "common_code", "bias_data.R"))

```

## Overview

This demonstration explores a single realization of catch misspecification with different levels of state-space models applied to the misspecified data. The goal is to demonstrate the diagnostics that can detect the misspecification and/or account for the problem. For comparison, a correctly specified model is also shown.

## Building the data sets

The process starts with creating an intial data set using the make_basic_info function and assigning values for selectivity, natural mortality, and recruitment. For this example, there is a single fleet and two indices all with logisitic selectivity, natural mortality is constant at 0.2 for all years and ages, there is no stock-recruitment relationship (recruitment varies as an identical and independent deviation about the mean). The prepare_wham_input function creates the correctly specified input formulation.

```{r data}
# only doing one simulation here, but keep structure for multiple sims for use later
nsim <- 1

# create input
groundfish_info <- make_basic_info()

gf_selectivity = list(
  model = c(rep("logistic", groundfish_info$n_fleets),rep("logistic", groundfish_info$n_indices)),
  initial_pars = rep(list(c(5,1)), groundfish_info$n_fleets + groundfish_info$n_indices)) #fleet, index

gf_M = list(initial_means = rep(0.2, length(groundfish_info$ages)))

gf_NAA_re = list(
  N1_pars = exp(10)*exp(-(0:(length(groundfish_info$ages)-1))*gf_M$initial_means[1]),
  sigma = "rec", #random about mean
  cor="iid", #random effects are independent
  use_steepness = 0,
  recruit_model = 2, #random effects with a constant mean
  recruit_pars = exp(10)
)

input <- prepare_wham_input(basic_info = groundfish_info, 
                            selectivity = gf_selectivity, 
                            NAA_re = gf_NAA_re, 
                            M= gf_M)

```

The fit_wham function is run without fitting the model to create the operating model.

```{r runstarter}
# run starter input
om <- fit_wham(input, do.fit = FALSE, MakeADFun.silent = TRUE)
```

Four estimation models are created using different levels of "state-spaceness" for the population numbers at age and recruitment formulations. The first estimation model exactly matches the operating model, meaning recruitment is estimated as a random effect about a mean value as independent and identical deviates. Ages 2 and older follow the population equation just as a statistical catch-at-age model would. The second estimation model also only has random effects for recruitment, but in this case the deviations are estimated as an autoregressive process over time. The third estimation model allows for random effects in both recruitment and for all other ages. This is the first full state-space model in the series. The random effects are all treated as independent and identical deviates with two separate sigmas estimated, one for recruitment and the other for all other ages. The fourth estimation model is also a full state-space model but the random effects are modeled as a two dimensional autoregressive process over both ages and years, with recruitment being treated separately from all other ages.

```{r makeem}
# set up estimation models
em_input1 <- input # self-test

gf_NAA_re2 <- gf_NAA_re
gf_NAA_re2$cor <- "ar1_y"
em_input2 <- prepare_wham_input(basic_info = groundfish_info, selectivity = gf_selectivity, NAA_re = gf_NAA_re2, M= gf_M)

gf_NAA_re3 <- gf_NAA_re
gf_NAA_re3$sigma <- "rec+1"
em_input3 <- prepare_wham_input(basic_info = groundfish_info, selectivity = gf_selectivity, NAA_re = gf_NAA_re3, M= gf_M)

gf_NAA_re4 <- gf_NAA_re
gf_NAA_re4$cor <- "2dar1"
gf_NAA_re4$sigma <- "rec+1"
em_input4 <- prepare_wham_input(basic_info = groundfish_info, selectivity = gf_selectivity, NAA_re = gf_NAA_re4, M= gf_M)
```

Simulated data are created under two conditions: correctly an incorrectly reported catch in the recent years of the assessment. The incorrectly reported catch uses the bias_data function. In this example, catch is under-reported by half (the catch provided to the wham model is only half of the actual catch) for the econd half of the assessment time period. The four estimation models are then set up to be applied to the two sets of data, resulting in a total of 8 models. The first four models use the correct catch data while the last four models use the under-reported catch data.

```{r simdata}
#simulate data from operating model
#all RE and data are simulated
sim_input <- list()

# sim_input[[1]] has no data modification and em1
set.seed(14159265) #use same seed for all operating models
sim_input[[1]] = lapply(1:nsim, function(x) {
  input_i = em_input1
  sim = om$simulate(complete=TRUE)
  input_i$data = sim
  return(input_i)
})

# sim_input[[2]] uses the correct data and em2
sim_input[[2]] = lapply(1:nsim, function(x) {
  input_i = em_input2
  obs_names = c("agg_indices","agg_catch","catch_paa","index_paa", "Ecov_obs", "obsvec")
  input_i$data[obs_names] = sim_input[[1]][[x]]$data[obs_names]
  return(input_i)
})

# sim_input[[3]] uses the correct data and em3
sim_input[[3]] = lapply(1:nsim, function(x) {
  input_i = em_input3
  obs_names = c("agg_indices","agg_catch","catch_paa","index_paa", "Ecov_obs", "obsvec")
  input_i$data[obs_names] = sim_input[[1]][[x]]$data[obs_names]
  return(input_i)
})

# sim_input[[4]] uses the correct data and em4
sim_input[[4]] = lapply(1:nsim, function(x) {
  input_i = em_input4
  obs_names = c("agg_indices","agg_catch","catch_paa","index_paa", "Ecov_obs", "obsvec")
  input_i$data[obs_names] = sim_input[[1]][[x]]$data[obs_names]
  return(input_i)
})

# now create the under-reported catch data
# sim_input[[5]] has under-reported catch and em1
set.seed(14159265) #use same seed for all operating models
agg_catch_multiplier <- create_agg_catch_multiplier(input, multiplier=0.50)
sim_input[[5]] = lapply(1:nsim, function(x) {
  input_i = em_input1
  sim = om$simulate(complete=TRUE)
  sim <- bias_data(sim, multiply_agg_catch_flag=TRUE,
                   agg_catch_multiplier=agg_catch_multiplier)
  input_i$data = sim
  return(input_i)
})

# sim_input[[6]] has under-reported catch and em2
sim_input[[6]] = lapply(1:nsim, function(x) {
  input_i = em_input2
  obs_names = c("agg_indices","agg_catch","catch_paa","index_paa", "Ecov_obs", "obsvec")
  input_i$data[obs_names] = sim_input[[5]][[x]]$data[obs_names]
  return(input_i)
})

# sim_input[[7]] has under-reported catch and em3
sim_input[[7]] = lapply(1:nsim, function(x) {
  input_i = em_input3
  obs_names = c("agg_indices","agg_catch","catch_paa","index_paa", "Ecov_obs", "obsvec")
  input_i$data[obs_names] = sim_input[[5]][[x]]$data[obs_names]
  return(input_i)
})

# sim_input[[8]] has under-reported catch and em4
sim_input[[8]] = lapply(1:nsim, function(x) {
  input_i = em_input4
  obs_names = c("agg_indices","agg_catch","catch_paa","index_paa", "Ecov_obs", "obsvec")
  input_i$data[obs_names] = sim_input[[5]][[x]]$data[obs_names]
  return(input_i)
})
```

Check to see that the eight data sets have the correct aggregate catch. Models 1-4 have the true catch, while models 5-8 have the under-reported catch in recent years.

```{r aggcatchcheck}
aggc <- tibble(Model = character(), Year = integer(), AggCatch = double())
for (m in 1:8){
  aggcdat <- lapply(1:nsim, function(x){
    out <- sim_input[[m]][[x]]$data$agg_catch
  return(out)
  })
  thisaggc <- tibble(Model = m, Year = 1:sim_input[[m]][[1]]$data$n_years_model, AggCatch = unlist(aggcdat))
  aggc <- rbind(aggc, thisaggc)
}

p1 <- ggplot(aggc, aes(x=Year, y=AggCatch)) +
  geom_point() +
  geom_line() +
  facet_wrap(~Model, ncol = 4) +
  expand_limits(y = 0) +
  theme_bw()

print(p1)

```

Now run the eight models. This may take a few minutes. The Rmarkdown compilation checks to see if the file misspec_demo_fits.RDS is present in the misspecification_study directory. If it is, it simply reads that file. If it is not, then the model is run. So if you change any of the settings above, be sure to remove the RDS file before running the Rmarkdown script.

```{r runmodels}
if(file.exists(file.path(here(), "misspecification_study", "misspec_demo_fits.RDS"))){
  em_fits <- readRDS(file.path(here(), "misspecification_study", "misspec_demo_fits.RDS"))
} else {
  # run the models
  em_fits = list()
  for(m in 1:8){
    em_fits[[m]] = lapply(1:nsim, function(x){
      out = fit_wham(sim_input[[m]][[x]], do.osa = FALSE, MakeADFun.silent = TRUE, retro.silent = TRUE, save.sdrep = FALSE)
      return(out)
    })
  }
  saveRDS(em_fits, file.path(here(), "misspecification_study", "misspec_demo_fits.RDS"))
}
```

Start by comparing the Mohn's rho for SSB. The misspecified models all have larger magnitude Mohn's rho for SSB than the models with the correct catch.

```{r rhoSSB}

rhoSSB <- matrix(NA, nrow=8, ncol=2)
colnames(rhoSSB) <- c("Model", "rhoSSB")
for (m in 1:8){
  thisrho <- lapply(1:nsim, function(x){
    out <- mohns_rho(em_fits[[m]][[x]])[1]
  return(out)
  })
  rhoSSB[m,] <- c(m, unlist(thisrho))
}
knitr::kable(rhoSSB)
```

We can see which model has the best fit using the compare_wham_models function for the two sets of data.

```{r compare}
correct_data_compare <- compare_wham_models(list(em_fits[[1]][[1]],em_fits[[2]][[1]],em_fits[[3]][[1]],em_fits[[4]][[1]]))

misspecified_data_compare <- out <- compare_wham_models(list(em_fits[[5]][[1]],em_fits[[6]][[1]],em_fits[[7]][[1]],em_fits[[8]][[1]]))

```

### 2022-07-05 misspecification_study/misspec_demo.Rmd
change under reporting and continue explorations

@@
 The incorrectly reported catch uses the bias_data function.
-In this example, catch is under-reported by half (the catch provided to the wham model is only half of the actual catch) for the econd half of the assessment time period.
+In this example, catch is under-reported, the catch provided to the wham model is only one third of the actual catch for the second half of the assessment time period.
 The four estimation models are then set up to be applied to the two sets of data, resulting in a total of 8 models.
@@
 set.seed(14159265) #use same seed for all operating models
-agg_catch_multiplier <- create_agg_catch_multiplier(input, multiplier=0.50)
+agg_catch_multiplier <- create_agg_catch_multiplier(input, multiplier=0.33)
 sim_input[[5]] = lapply(1:nsim, function(x) {
@@
 If it is not, then the model is run.
-So if you change any of the settings above, be sure to remove the RDS file before running the Rmarkdown script.
+So if you change any of the settings above, be sure to remove the RDS file before knitting the Rmarkdown script.
 
@@
 ```
+Start by making sure all the models converged.
+Model 4 (correct data with estimation model 4) did not converge.
+Kept in for now just for completeness, but this is a good reminder to make sure the model converges before digging into the results.
 
-Start by comparing the Mohn's rho for SSB. The misspecified models all have larger magnitude Mohn's rho for SSB than the models with the correct catch.
-
-```{r rhoSSB}
-
-rhoSSB <- matrix(NA, nrow=8, ncol=2)
-colnames(rhoSSB) <- c("Model", "rhoSSB")
-for (m in 1:8){
-  thisrho <- lapply(1:nsim, function(x){
-    out <- mohns_rho(em_fits[[m]][[x]])[1]
-  return(out)
-  })
-  rhoSSB[m,] <- c(m, unlist(thisrho))
-}
-knitr::kable(rhoSSB)
+```{r converged}
+check_convergence(em_fits[[1]][[1]])
+check_convergence(em_fits[[2]][[1]])
+check_convergence(em_fits[[3]][[1]])
+check_convergence(em_fits[[4]][[1]])
+check_convergence(em_fits[[5]][[1]])
+check_convergence(em_fits[[6]][[1]])
+check_convergence(em_fits[[7]][[1]])
+check_convergence(em_fits[[8]][[1]])
 ```
@@
 We can see which model has the best fit using the compare_wham_models function for the two sets of data.
+Start with the correctly reported catch data (CD).
+We see all four estimation models (em) fit the data similarly for F, recruitment, and SSB, with similar CVs for each.
+Estimation model 2 (AR1 random walk in recruitment) was preferred over the other three estimation models, but the differences in AIC were not great (<5 units).
+Mohn's rho for SSB and F and were low for all four estimation models (absolute values < 0.6).
 
-```{r compare}
-correct_data_compare <- compare_wham_models(list(em_fits[[1]][[1]],em_fits[[2]][[1]],em_fits[[3]][[1]],em_fits[[4]][[1]]))
+```{r comparecorrect}
+correct_data_compare <- compare_wham_models(list(CD_em1 = em_fits[[1]][[1]],
+                                                 CD_em2 = em_fits[[2]][[1]],
+                                                 CD_em3 = em_fits[[3]][[1]],
+                                                 CD_em4 = em_fits[[4]][[1]]))
 
-misspecified_data_compare <- out <- compare_wham_models(list(em_fits[[5]][[1]],em_fits[[6]][[1]],em_fits[[7]][[1]],em_fits[[8]][[1]]))
+print(correct_data_compare$g[[1]])
+print(correct_data_compare$g[[2]])
+```
 
+The under-reported catch data (UR) found greater differences among the four models.
+Estimation model 4 (full state-spcae with 2dar1 correlations) was the preferred model based on AIC, with the other models >50 units worse.
+The Mohn's rho values for SSB and F for estimation model 4 were lower than all the values from the correct data, while the other three estimation models were all worse (absolute values >0.20).
+So according to both AIC and the retrospective diagnostic, estimation model 4 is highly preferred to the other three estimation models for the under-reported catch data.
+All four estimation models show a spike in F in year 2000, just before the catch under-reporting begins.
+The big difference is in the scale of the F, recruitment, and SSB point estimates, with estimation model 4 showing lower F and higher recruitment and SSB than the other three estimation models.
+The trends in F, recruitment, and SSB are similar though.
+The uncertainty of estimation model 4 is much higher than the other three models.
+It is not clear what is driving the large uncertainty in estimation model 4 (Tim - do you have any thoughts?).
+Looking at the fits for estimation model 4 (using plot_wham_output(mod=em_fits[[8]][[1]], out.type='html')), I was surprised by how well most things fit, with no obvious patterning in residuals or obvious misfits.
+The one bad diagnostic I saw was the conditional expected and posterior estimates of abundance at all ages where the line was above all the points for all years and ages.
+So there was one diagnostic that indicated that there is a model misfit.
+Note to self, need to figure out a good way to save the figures and bring them into this Rmarkdown so everyone can see them.
+
+```{r comparemisspec}
+misspecified_data_compare <- compare_wham_models(list(UR_em1 = em_fits[[5]][[1]],
+                                                      UR_em2 = em_fits[[6]][[1]],
+                                                      UR_em3 = em_fits[[7]][[1]],
+                                                      UR_em4 = em_fits[[8]][[1]]))
+
+print(misspecified_data_compare$g[[1]])
+print(misspecified_data_compare$g[[2]])
 ```

### 2022-07-08 misspecification_study/misspec_demo.Rmd
more tweaks

@@
 source(file.path(here(), "common_code", "bias_data.R"))
-
 ```
@@
 
-The process starts with creating an intial data set using the make_basic_info function and assigning values for selectivity, natural mortality, and recruitment.
-For this example, there is a single fleet and two indices all with logisitic selectivity, natural mortality is constant at 0.2 for all years and ages, there is no stock-recruitment relationship (recruitment varies as an identical and independent deviation about the mean).
+The process starts with creating an initial data set using the make_basic_info function and assigning values for selectivity, natural mortality, and recruitment.
+For this example, there is a single fleet and two indices all with logistic selectivity, natural mortality is constant at 0.2 for all years and ages, and there is no stock-recruitment relationship (recruitment varies as an identical and independent deviation about the mean).
 The prepare_wham_input function creates the correctly specified input formulation.
@@
 # set up estimation models
-em_input1 <- input # self-test
+em_input1 <- input
 
@@
 
-Simulated data are created under two conditions: correctly an incorrectly reported catch in the recent years of the assessment.
+Simulated data are created under two conditions: correctly and incorrectly reported catch in the recent years of the assessment.
 The incorrectly reported catch uses the bias_data function.
@@
 The four estimation models are then set up to be applied to the two sets of data, resulting in a total of 8 models.
-The first four models use the correct catch data while the last four models use the under-reported catch data.
+The first four sim_inputs use the correct catch data while the last four sim_inputs use the under-reported catch data.
 
@@
 So if you change any of the settings above, be sure to remove the RDS file before knitting the Rmarkdown script.
+This RDS file is relatively large (~16 MB), so is not saved on the GitHub repo.
 
@@
 ```
+
 Start by making sure all the models converged.
@@
 It is not clear what is driving the large uncertainty in estimation model 4 (Tim - do you have any thoughts?).
-Looking at the fits for estimation model 4 (using plot_wham_output(mod=em_fits[[8]][[1]], out.type='html')), I was surprised by how well most things fit, with no obvious patterning in residuals or obvious misfits.
-The one bad diagnostic I saw was the conditional expected and posterior estimates of abundance at all ages where the line was above all the points for all years and ages.
-So there was one diagnostic that indicated that there is a model misfit.
-Note to self, need to figure out a good way to save the figures and bring them into this Rmarkdown so everyone can see them.
 
@@
 ```
+
+Looking at the fits to the under-reported data for estimation model 4 (using plot_wham_output(mod=em_fits[[8]][[1]], out.type='html')), I was surprised by how well most things fit, with no obvious patterning in residuals or obvious misfits.
+The one bad diagnostic I saw was the conditional expected and posterior estimates of abundance at all ages where the line was above all the points for all years and ages.
+So there was one diagnostic that indicated that there is a model misfit, but not anything that jumped out at me from the tabular data that could easily be grabbed when running lots of realizations.
+
+```{r plotemfit8}
+# note could not get the following code to run within Rmarkdown (it produced the png files but not the pdf)
+# can either pull this code out and run in console, or else uncomment and knit then comment and knit again
+#em8.dir <- file.path(here(), "misspecification_study", "UR_em4")
+#if(!dir.exists(em8.dir)){
+#  dir.create(em8.dir)
+#}
+#plot_wham_output(mod=em_fits[[8]][[1]], dir.main=em8.dir)
+
+```

### 2022-07-12 misspecification_study/misspec_demo.Rmd

@@
 gf_NAA_re2$cor <- "ar1_y"
-em_input2 <- prepare_wham_input(basic_info = groundfish_info, selectivity = gf_selectivity, NAA_re = gf_NAA_re2, M= gf_M)
+em_input2 <- prepare_wham_input(basic_info = groundfish_info, 
+                                selectivity = gf_selectivity, 
+                                NAA_re = gf_NAA_re2, 
+                                M= gf_M)
 
@@
 gf_NAA_re3$sigma <- "rec+1"
-em_input3 <- prepare_wham_input(basic_info = groundfish_info, selectivity = gf_selectivity, NAA_re = gf_NAA_re3, M= gf_M)
+em_input3 <- prepare_wham_input(basic_info = groundfish_info, 
+                                selectivity = gf_selectivity, 
+                                NAA_re = gf_NAA_re3, 
+                                M= gf_M)
 
@@
 gf_NAA_re4$sigma <- "rec+1"
-em_input4 <- prepare_wham_input(basic_info = groundfish_info, selectivity = gf_selectivity, NAA_re = gf_NAA_re4, M= gf_M)
+em_input4 <- prepare_wham_input(basic_info = groundfish_info, 
+                                selectivity = gf_selectivity, 
+                                NAA_re = gf_NAA_re4, 
+                                M= gf_M)
 ```
@@
   })
-  thisaggc <- tibble(Model = m, Year = 1:sim_input[[m]][[1]]$data$n_years_model, AggCatch = unlist(aggcdat))
+  thisaggc <- tibble(Model = m, 
+                     Year = 1:sim_input[[m]][[1]]$data$n_years_model, 
+                     AggCatch = unlist(aggcdat))
   aggc <- rbind(aggc, thisaggc)
@@
     em_fits[[m]] = lapply(1:nsim, function(x){
-      out = fit_wham(sim_input[[m]][[x]], do.osa = FALSE, MakeADFun.silent = TRUE, retro.silent = TRUE, save.sdrep = FALSE)
+      out = fit_wham(sim_input[[m]][[x]], 
+                     do.osa = FALSE,
+                     MakeADFun.silent = TRUE, 
+                     retro.silent = TRUE, 
+                     save.sdrep = FALSE)
       return(out)
@@
 ```
+
+The catch and index four panel plots all look good for UR_em4.
+
+![catch 4 panel](UR_em4/plots_png/diagnostics/Catch_4panel_fleet_1.png)
+
+![index 1 4 panel](UR_em4/plots_png/diagnostics/Index_4panel_1.png)
+
+![index 2 4 panel](UR_em4/plots_png/diagnostics/Index_4panel_2.png)
+
+The problematic diagnostic is the NAA 4 panel plots, show below for ages 1 and 8 as well as the stacked barplot of residuals.
+
+![NAA age 1 4 panel](UR_em4/plots_png/diagnostics/NAA_4panel_1.png)
+![NAA age 8 4 panel](UR_em4/plots_png/diagnostics/NAA_4panel_8.png)
+![NAA residuals stacked barplot](UR_em4/plots_png/diagnostics/NAA_res_barplot_stacked.png)

### 2022-07-21 Project_0/paper/paper.Rmd
working on M oms and expanding set of possible ems

---
title: State-space age-structured assessment models provide reliable inferences(?)
author: 
  - Timothy J. Miller^1^
date: ''
fontsize: 12pt
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    includes:
      in_header:
      - options.sty
      - journalshortcuts.tex
    keep_tex: yes
    number_sections: yes
  html_document:
    df_print: paged
  word_document: default
header-includes:
- \usepackage{xspace}
- \usepackage{bm}
- \usepackage{caption,graphics}
- \usepackage{graphicx}
- \usepackage{makecell}
- \renewcommand\figurename{Fig.}
- \captionsetup{labelsep=period, singlelinecheck=false}
- \newcommand{\changesize}[1]{\fontsize{#1pt}{#1pt}\selectfont}
- \renewcommand{\arraystretch}{1.5}
- \renewcommand\theadfont{}
csl: canadian-journal-of-fisheries-and-aquatic-sciences.csl
biblio-style: cjfas2.bst
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy=FALSE, tidy.opts=list(width.cutoff=60), warning = FALSE, message = FALSE, fig.align = "center")
knitr::opts_knit$set(root.dir = '../' )
library(knitr)
library(tidyverse)
library(pander)
library(kableExtra)
library(png)
library(forcats)
```

$^1$corresponding author: timothy.j.miller@noaa.gov, Northeast Fisheries Science Center, Woods Hole Laboratory, 166 Water Street, Woods Hole, MA 02543 USA\

\pagebreak

## Abstract {-}

### Keywords {-}

\pagebreak

# Introduction

# Methods

We completed a simulation study with 24 NAA re operating models, 16 M re operating models, X Sel re operating models and Y q re operating models. 
We simulated N data sets for each operating model.
For each simulated data from each operating model we fit a set of estimating models.

Y estimating models fit to each 
## Operating models

common to all:   
  ages, 
  M
  maturity
  waa
  observation error assumptions
  2.5 Fmsy -> Fmsy after 20 years
  Fmsy througout

NAA_oms
  process error assumtions
  Fhistory description

Population - 
  Initial population, 
  ages, 
  M
  maturity
  waa

Fishing history

All operating models assumed one of two different fishing histories. One 
Two alfishing histories are 
Overfishing assumptions is based on average estimates of overfishing for NE groundfish stocks from Wiedenman et al. XXXX.

All post-recruit productivity components are constant in the NAA and survey catchability process error operating models and we set stock-recruit parameters such that unfished recruitment = exp(10) and
Fmsy = F40  (Miller and Brooks 2021). For M

Initial population starts at equilibium population structure associated with F= x*Fmsy where x = 1 or 2.5.

surveys, selectivity

fleets, selectivity

M_oms

## Estimating models

32 estimating models
1-20 fit to each NAA RE operating model
5-24 fit to each M RE operating model
5-20,25-28 to each sel RE operating model
5-20, 29-32 to each q RE operating model

SR estimation or not

M estimation or not

NAA_re
Random effects on Recruitment only or random effects on recruitment and transitions among older numbers at age.

M_re
Random effects on Recruitment only, M constant across age .

sel_re
Random effects on Recruitment only, fleet logistic selectivity RE model?

q_re
Random effects on Recruitment only, one survey catchability RE model?

# Results

Convergence performance

AIC performance

SR estimation? M estimation?

Bias, Mean Square error

  Certain basic parameters (stock-recruit pars, M, variance parameters)
  SSB, F, R

# Discussion

# Acknowledgements {-}

This work was funded by NOAA Fisheries Northeast Fisheries Science Center.

# CRediT authorship contribution statement {-}

**Timothy J. Miller:** Conceptualization, Methodology, Writing - original draft, Formal analysis, Visualization.

\pagebreak

\bibliography{sweep-paper}
<!-- \bibliography{sweep-paper}
\bibliographystyle{/home/tmiller2/work/bibliographies/cjfas2}  -->

<div id="refs"></div>

\pagebreak
<!-- 
\begin{figure}
\caption{Diagram of twin-trawl gear configuration. One of the two nets is rigged with a rockhopper sweep (8) and the other is rigged with a chain sweep (7) and for both a restrictor rope (5) is used to obtain consistent net spread. The other importand components are the side wires (1), middle wire (2), doors (3), the clump weight (4), and the acoustic mensuration system (6).  The side where the rockhopper and chainsweep gears gears were deployed varied throughout the experimental tows of each.}\label{twin_trawl_diagram}
\begin{center}
\includegraphics[width = \textwidth]{twin_trawl_diagram.pdf}
\end{center}
\end{figure}

\begin{figure}
\caption{The \textsl{F/V Karen Elizabeth} twin-trawl vessel rigged with rockhopper sweep gear on the right and chain sweep gear on the left.}\label{twin_trawl_photo}
\begin{center}
\includegraphics[width = \textwidth]{twin_trawl_photo_small.png}
\end{center}
\end{figure} -->

<!--
\begin{figure}
\caption{Diagram of the standard Northeast Fisheries Science Center rockhopper sweep center and wing sections.}\label{rockhopper_schematic}
\begin{center}
\includegraphics[width = 0.7\textwidth]{rockhopper_schematic_1.pdf}
\includegraphics[width = 0.7\textwidth]{rockhopper_schematic_2.pdf}
\end{center}
\end{figure}

\begin{figure}
\caption{Diagram of the chain sweep designed maximize bottom contact and flatfish capture.}\label{chainsweep_schematic}
\begin{center}
\includegraphics[width = \textwidth]{chainsweep_schematic.pdf}
\end{center}
\end{figure}
\clearpage
  -->
<!-- \begin{landscape}
\begin{figure}
\caption{Annual locations of stations where the F/V Karen Elizabeth conducted twin-trawl sets with the standard bottom trawl gear and the gear with a chain sweep instead of the rockhopper sweep.}\label{tow_locations}
\begin{center}
\includegraphics[width = \textwidth]{TwinTrawlLocationsAllYears.jpg}
\end{center}
\end{figure}
\end{landscape}

\clearpage

\begin{figure}
\caption{Relative efficiency  of gears using chain and rockhopper sweeps from the best performing model for each species (Table \ref{best_model_compare}). Blue and red denote results for day and night data, respectively, and thick and thin lines represent overall and paired-tow specific estimates of relative catch efficiency, respectively. There was no diel effect in the best model for American plaice. Points represent empirical estimates of relative efficiency for paired observation by length and paired tow. Polygons and dashed lines represent hessian-based and bootstrap-based 95\% confidence intervals, respectively.}\label{sp_rho_plot}
\begin{center}
\includegraphics[height = 0.8\textheight]{sp_rho_plot.pdf}
\end{center}
\end{figure}

\clearpage

\begin{figure}
\caption{Annual spring (blue) and fall (red) biomass estimates for each managed stock assuming 100\% efficiency for chain sweep gear with shaded polygons representing bootstrap-based 95\% confidence intervals. Relative catch efficiency at size estimates and bootstraps are from the best performing model for each species (Table \ref{best_model_compare}).}\label{stock_biomass_plot}
\begin{center}
\includegraphics[height = 0.8\textheight]{stock_biomass_plot.pdf}
\end{center}
\end{figure}

\clearpage

\begin{figure}
\caption{Implied catch efficiency of annual spring (blue) and fall (red) bottom trawl survey biomass estimates for each managed stock assuming 100\% efficiency for chain sweep gear with shaded polygons representing bootstrap-based 95\% confidence intervals. Relative catch efficiency at size estimates and bootstraps are from the best performing model for each species (Table \ref{best_model_compare}).}\label{stock_biomass_efficiency_plot}
\begin{center}
\includegraphics[height = 0.8\textheight]{stock_biomass_efficiency_plot.pdf}
\end{center}
\end{figure} -->

\clearpage

<!-- \begin{table}
\caption{Managed stocks associated with the species for which relative catch efficiency was estimated.}\label{stock_definition_table}
{\input{stock_definition_table}}
\end{table}

\begin{landscape}
\begin{table}
\caption{Description of relative catch efficiency ($\rho$) and beta-binomial dispersion ($\phi$) parameterizations for binomial and beta-binomial models and number of marginal likelihood parameters ($n_p$) for the 13 base models from \citet{miller13} and fit to paired chain sweep and rockhoppersweep tow data for each species.}\label{base_model_description}
{\input{base_model_description}}
\end{table}
\end{landscape}

\begin{landscape}
\begin{table}
\caption{Number of paired tows where fish were captured and the number of fish captured and measured for lengths for each species in total and by day or night.}\label{data_table}
{\changesize{10}\input{data_table}}
\end{table}
\end{landscape}

\begin{landscape}
\begin{table}\caption{Difference in AIC for each of the 13 models described in Table \ref{base_model_description} from the best model ($\bf 0$) by species.}\label{base_model_compare}
{\input{base_model_compare}}
\end{table}
\end{landscape}

\begin{landscape}
\begin{table}\caption{Best performing models from Table \ref{base_model_compare} and extended models that include diel effects on relative catch efficiency for each species with the number of parameters for each model ($n_p$) and the differences in AIC ($\Delta$AIC) from the best of the three models ($\bf 0$) by species.}\label{best_model_compare}
{\changesize{7}\input{best_model_compare}}
\end{table}
\end{landscape}

\begin{table}
\caption{Average of annual (2009-2019) ratios of coefficients of variation for calibrated and uncalibrated biomass indices for each stock by seasonal survey. Coefficients of variation are based on bootstrap resampling of paired tow observations, survey station data and associated length and weight observations. Annual indices for fall 2017 were not available for summer flounder, SNE-MA windowpane, and SNE-MA yellowtail flounder.}\label{stock_cv_ratios}
{\input{stock_cv_ratios}}
\end{table}

\begin{table}
\caption{Average correlation of annual (2009-2019) calibrated biomass indices for each stock by seasonal survey. Annual indices for fall 2017 were not available for SNE-MA windowpane and SNE-MA yellowtail flounder.}\label{stock_mean_cor}
{\input{stock_mean_cor}}
\end{table} -->

### 2022-07-21 Project_0/code/M_om_hpcc_args.sh

@@ -9,17 +9,16 @@ echo "CPU threads: $(grep -c processor /proc/cpuinfo)"
 grep 'cpu cores' /proc/cpuinfo | uniq
 
 cd ~/SSRTWG
-#Rscript --vanilla ~/SSRTWG/Project_0/code/naa_om_hpcc_script.R 7 1 1
-for ((sim= $1; sim <= $2; sim++)) # in {$simstart..$simend}
+for ((sim= $1; sim <= $2; sim++))
 do
-  for ((om= $3; om <= $4; om++)) #om in {$omstart..$omend}
+  for ((om= $3; om <= $4; om++))
   do
-   for ((em= $5; em <= $6; em++)) #em in {$emstart..$emend}
+   for ((em= $5; em <= $6; em++))
    do
      echo $sim
      echo $om
      echo $em
-     Rscript --vanilla ~/SSRTWG/Project_0/code/naa_om_hpcc_script.R $sim $om $em &
+     Rscript --vanilla ~/SSRTWG/Project_0/code/M_om_hpcc_script.R $sim $om $em &
    done
  done
 done

### 2022-07-21 Project_0/code/M_om_hpcc_bsub.sh

@@ -13,7 +13,7 @@ do
   for ((this_om= $3; this_om <= $4; this_om++)) #om in {$omstart..$omend}
   do
     echo $this_om
-    bsub -n $7 -q short -W 4:00 -o ~/logs/logfile.%J -R "rusage[mem=5000]" -R "span[hosts=1]" -J naa_om_sim "bash ~/SSRTWG/Project_0/code/naa_om_hpcc_args.sh $this_sim $this_sim $this_om $this_om $5 $6"
+    bsub -n $7 -q short -W 4:00 -o ~/logs/logfile.%J -R "rusage[mem=5000]" -R "span[hosts=1]" -J M_om_sim "bash ~/SSRTWG/Project_0/code/M_om_hpcc_args.sh $this_sim $this_sim $this_om $this_om $5 $6"
   done
 done
 echo "script is done"

### 2022-07-21 Project_0/code/M_om_sim_fit_script_hpcc.R

@@ -1,0 +1,67 @@
+#!/usr/bin/env Rscript
+args = commandArgs(trailingOnly=TRUE)
+
+.libPaths("~/Rlib/")
+library(here)
+library(wham)
+
+this_om <- as.integer(args[1])
+this_em <- as.integer(args[2])
+this_sim <- as.integer(args[3])
+
+om_inputs <- readRDS(file.path(here(),"Project_0","inputs", "M_om_inputs.RDS"))
+em_inputs <- readRDS(file.path(here(),"Project_0","inputs", "em_inputs.RDS"))
+df.ems <- readRDS(file.path(here(),"Project_0","inputs", "df.ems.RDS"))
+df.oms <- readRDS(file.path(here(),"Project_0","inputs", "df.M.oms.RDS"))
+#######################################################
+#need to have matching assumptions about CVs for catch and indices, too
+obs_names <- c("agg_catch","agg_catch_sigma", "agg_indices", "agg_index_sigma", "catch_paa", "index_paa", 
+  "Ecov_obs", "obs", "obsvec")
+#######################################################
+
+#######################################################
+#Do we want to use the same (e.g. 1000) seeds for everything?
+seeds <- readRDS(file.path(here(), "Project_0", "inputs","seeds.RDS"))
+#######################################################
+
+cat(paste0("OM: ", this_om, " Sim: ", this_sim, " EM: ", this_em, "\n"))
+
+# Set seed
+om <- fit_wham(om_inputs[[this_om]], do.fit = FALSE, MakeADFun.silent = TRUE)
+set.seed(seeds[this_sim])
+sim_data <- om$simulate(complete=TRUE)
+truth <- sim_data
+#save the version for reproducibility
+truth$wham_version = om$wham_version
+EM_input <- em_inputs[[this_em]] # Read in the EM 
+#put simulated data into the em input
+EM_input$data[obs_names] = sim_data[obs_names]
+res <- list(truth = truth)
+res$fit <- list()
+#do fit withouth sdreport first
+fit <- tryCatch(fit_wham(EM_input, do.sdrep=F, do.osa=F, do.retro=T, do.proj=F, MakeADFun.silent=TRUE),
+  error = function(e) conditionMessage(e))
+  
+# Deal with issues fitting EM to non-matching OM data
+# empty elements below can be used to summarize convergence information
[…]

### 2022-07-21 Project_0/code/aggregate_results_hpcc.R

@@ -1,3 +1,4 @@
+.libPaths("~/Rlib/")
 library(here)
 library(wham)
 df.ems = readRDS(file.path(here(),"Project_0","inputs", "df.ems.RDS"))
@@ -8,10 +9,22 @@ oms = 1:length(om_inputs)
 ems = 1:length(em_inputs)
 source(file.path(here::here(),"Project_0","code","sim_management.R"))
 
-aggregate_hpcc_results(sim = 7, oms = 1, ems = ems, res_dir = file.path(here::here(),"Project_0", "results", "naa_om"))
+sapply(1:24, function(x) aggregate_hpcc_results(sim = 1, oms = x, ems = ems, res_dir = file.path(here::here(),"Project_0", "results", "naa_om")))
+#aggregate_hpcc_results(sim = 1, oms = 1, ems = ems, res_dir = file.path(here::here(),"Project_0", "results", "naa_om"))
+#aggregate_hpcc_results(sim = 1, oms = 3, ems = ems, res_dir = file.path(here::here(),"Project_0", "results", "naa_om"))
+
+sapply(1:100, function(x) aggregate_hpcc_results(sim = x, oms = 1, ems = ems, res_dir = file.path(here::here(),"Project_0", "results", "naa_om"))
 
 x = readRDS(file.path(here::here(),"Project_0", "results", "naa_om", "om_1", "sim_7.RDS"))
 x = readRDS(file.path(here::here(),"Project_0", "results", "naa_om", "om_1", "sim_1.RDS"))
+x = readRDS(file.path(here::here(),"Project_0", "results", "naa_om", "om_3", "sim_1.RDS"))
 
-aic = 2*sapply(x,function(y) y$fit$opt$obj + length(y$fit$opt$par))
-aic - min(aic)
+aggregate_hpcc_results(sim = 1, oms = 10, ems = ems, res_dir = file.path(here::here(),"Project_0", "results", "naa_om"))
+x = readRDS(file.path(here::here(),"Project_0", "results", "naa_om", "om_10", "sim_1.RDS"))
+
+aic = 2*sapply(x,function(y) {
+  out = NA
+  if(!is.null(y)) out = y$fit$opt$obj + length(y$fit$opt$par)
+  return(out)
+})
+aic - min(aic, na.rm = TRUE)

### 2022-07-21 Project_0/code/em_setup.R

@@ -1,5 +1,7 @@
 # devtools::install_github("timjmiller/wham", dependencies=TRUE, ref="devel")
-library(wham)
+if(file.exists("c:/Users/timothy.j.miller")) {
+  library(wham, lib.loc = "c:/work/wham/old_packages/97577f1")
+} else library(wham) #make sure to use the right version of wham
 library(tidyr)
 library(dplyr)
 library(here)
@@ -62,7 +64,7 @@ for(i in 1:NROW(df.ems)){
   }
 
   if(df.ems$re_config[i] == "M_re") { #full random effects for NAA
-    M$re = "ar1_y"
+    M$re = "ar1_y" #this will make it so that there is just one random effect, need to set and fix cor values for iid assumption
     M$cor_vals = 0 #iid year effects
     #NAA_re$sigma_vals[2] = df.oms$NAA_sig[i] #don't start at true values?
   }
@@ -77,10 +79,77 @@ for(i in 1:NROW(df.ems)){
   }
   em_inputs[[i]] <- prepare_wham_input(basic_info = make_basic_info(), selectivity = selectivity, NAA_re = NAA_re, M= M,
     catchability = catchability, age_comp = "logistic-normal-miss0")  
+  #have to do this after the input is made
   em_inputs[[i]] = set_M(em_inputs[[i]], M = M) #this set_M will change parameter values
-  if(df.ems$re_config[i] == "M_re") {
-    em_inputs[[i]]$map$M_repars = factor(c(1,NA,NA)) #still need to fix rho=0
+  if(df.ems$re_config[i] == "M_re") { 
+    em_inputs[[i]]$map$M_repars = factor(c(1,NA,NA)) #still need to fix rho which is set to 0 above
   }
+
+  #turn off bias correction
+  em_inputs[[i]] = set_simulation_options(em_inputs[[i]], simulate_data = TRUE, simulate_process = TRUE, simulate_projection = TRUE, 
+    bias_correct_pe = FALSE, bias_correct_oe = FALSE)
+  #change Neff so scalar doesn't affect L-N SD
+  em_inputs[[i]]$data$catch_Neff[] = 1
+  em_inputs[[i]]$data$index_Neff[] = 1
+  em_inputs[[i]]$data$FXSPR_init[] = 0.3
[…]

### 2022-07-21 Project_0/code/make_om.R

@@ -12,7 +12,12 @@ make_om <- function(Fhist = "Fmsy", N1_state = "Fmsy", selectivity, M, NAA_re,
   #if you want to change the %Spawning Potential
   #input$data$percentSPR = 35 #for example
   temp <- fit_wham(input, do.fit = FALSE, MakeADFun.silent = TRUE)
-  
+  #print(temp$rep$MAA)
+  #print(temp$rep$selAA[[1]])
+  print(exp(temp$rep$log_FXSPR))
+  print(exp(temp$rep$log_FMSY))
+#  stop()
+
   F40 = exp(temp$rep$log_FXSPR[brp_year])
   if(is.na(F40) || F40 < 1e-5 || F40 > 10) stop("search for F40 failed, try changing eq_F_init. \n")
   cat(paste0("F40 = ", round(F40,3), ".\n"))
@@ -22,16 +27,20 @@ make_om <- function(Fhist = "Fmsy", N1_state = "Fmsy", selectivity, M, NAA_re,
   
   #will get steepness such that Fmsy = F40
   #print(input$data$FMSY_init)
-  cat("Finding steepness such that Fmsy = F40.\n")
-  steepness = get_steepness(steepness_ini = 0.75, input = input, NAA_re = NAA_re, F40 = F40, 
-    brp_year = brp_year)
-  NAA_re$recruit_pars[1] = steepness$par
-  input = set_NAA(input, NAA_re)
-  #input$data$FMSY_init[] = F40
-  temp <- fit_wham(input, do.fit = FALSE, MakeADFun.silent = TRUE)
-  Fmsy = exp(temp$rep$log_FMSY[brp_year])
-  if(is.na(Fmsy) || Fmsy < 1e-5 || Fmsy > 10) stop("search for Fmsy failed, try changing eq_F_init. \n")
-  
+  if(temp$input$data$use_steepness==1){ 
+    cat("Finding steepness such that Fmsy = F40.\n")
+    steepness = get_steepness(steepness_ini = 0.75, input = input, NAA_re = NAA_re, F40 = F40, 
+      brp_year = brp_year)
+    NAA_re$recruit_pars[1] = steepness$par
+    input = set_NAA(input, NAA_re)
+    #input$data$FMSY_init[] = F40
+    temp <- fit_wham(input, do.fit = FALSE, MakeADFun.silent = TRUE)
+    Fmsy = exp(temp$rep$log_FMSY[brp_year])
[…]

### 2022-07-23 Project_0/code/M_om_setup.R
fix yet another error in M oms

@@ -8,6 +8,7 @@ library(here)
 source(file.path(here(), "common_code", "make_basic_info.R"))
 source(file.path(here(), "common_code", "set_NAA.R"))
 source(file.path(here(), "common_code", "set_M.R"))
+source(file.path(here(), "common_code", "set_selectivity.R"))
 source(file.path(here(), "common_code", "set_simulation_options.R"))
 source(file.path(here(), "common_code", "get_FMSY.R"))
 source(file.path(here(), "Project_0", "code", "make_om.R"))
@@ -55,7 +56,7 @@ df.M.oms$Model <- paste0("om_",1:n.mods)
 df.M.oms <- df.M.oms %>% select(Model, everything()) # moves Model to first col
 # look at model table
 df.M.oms
-saveRDS(df.M.oms, file.path(here(),"Project_0", "inputs", "df.M.oms.RDS"))
+#saveRDS(df.M.oms, file.path(here(),"Project_0", "inputs", "df.M.oms.RDS"))
 
 
 gf_info = make_basic_info()

### 2022-07-23 Project_0/code/make_om.R

@@ -68,7 +68,8 @@ make_om <- function(Fhist = "Fmsy", N1_state = "Fmsy", selectivity, M, NAA_re,
   }
   NAA_re$N1_pars = eq_R_N1 * Jan1_NAA_per_recruit
   input = set_NAA(input, NAA_re)
-  
+  input = set_selectivity(input, selectivity)
+  input = set_M(input, M) 
   #set F relative to Fmsy. This function is in get_FMSY.R
   input = set_F_scenario(input, Fhist, Fmsy = Fmsy, max_mult = max_mult_Fmsy, min_mult= min_mult_Fmsy,
     change_time = F_change_time)

### 2022-07-23 Project_0/code/M_om_setup.R
make naa oms consistent before redoing sims

@@ -16,10 +16,6 @@ source(file.path(here(), "Project_0", "code", "make_om.R"))
 write.dir <- file.path(here(),"Project_0", "inputs")
 
 naa_om_inputs = readRDS(file.path(here::here(),"Project_0","inputs", "NAA_om_inputs.RDS"))
-# temp = sapply(naa_om_inputs, function(x) {
-#   temp = fit_wham(x, do.fit = FALSE, MakeADFun.silent = TRUE)
-#   return(temp$rep$log_SR_a)
-# })
 #SR parameters are the same for all naa_om models 
 temp = fit_wham(naa_om_inputs[[1]], do.fit = FALSE, MakeADFun.silent = TRUE)
 SRab = exp(c(temp$rep$log_SR_a[1], temp$rep$log_SR_b[1]))

### 2022-07-25 Project_0/paper/paper.Rmd
fix yet another error in M oms
make naa oms consistent before redoing sims
fix em inputs, selectivity oms

@@
 author: 
-  - Timothy J. Miller^1^
+  - Timothy J. Miller^1^, Britten, Brooks, Fay, Legault, Stock, 
 date: ''
@@
 
+Much is known about the reliabiltiy of state-space models that are linear or Gaussian, but applications in fisheris management are
+nonlinear and typically include multiple types of observations with varying distributional assumptions. We know relatively little about
+the statistical reliability of such models. Also, there is a wide range of potential random effects structures in assessment models
+and we  know little about the ability of information criteria to distinguish among such alternative structures.
+
+But those studies focus primarily on Gaussain 
+Reliability of hidden process models
+
 # Methods
 
-We completed a simulation study with 24 NAA re operating models, 16 M re operating models, X Sel re operating models and Y q re operating models. 
+Used the WHAM package commit = XXXXXXX (Stock and Miller 2021, Miller and Stock 2020). This packages has also been used
+to configure operating and estimating models for closed loop simulations evaluating index-based asssessment methods (Legault et al. In press).
+
+We completed a simulation study with 24 NAA re operating models, 16 M re operating models, 16 Sel re operating models and 16 q re operating models. 
 We simulated N data sets for each operating model.
@@
   Fmsy througout
+
+marginal standard deviations for random effects are defined in tables of operating models.
 
@@
 
+Make plot of S-R curve, Fmsy = F40
+Initial values for BH parameters are the true values.
+Initial values for mean R model = true R0.
+
 M estimation or not
@@
 
+Do each of these by type of operating model (Naa, M, sel, q)
 Convergence performance
@@
 # Discussion
+
+The estimating models assumed variances of aggregate catch and index observations was known. This approximation may be appropriate for indices
+where we have a reliable estimate of uncertainty based on the survey design (), but there may be better approaches for the aggregate
+catch such as an informed prior on the standard errors with realistic bounds.
 

### 2022-07-25 Project_0/code/M_om_setup.R

@@ -12,6 +12,8 @@ source(file.path(here(), "common_code", "set_selectivity.R"))
 source(file.path(here(), "common_code", "set_simulation_options.R"))
 source(file.path(here(), "common_code", "get_FMSY.R"))
 source(file.path(here(), "Project_0", "code", "make_om.R"))
+source(file.path(here(), "Project_0", "code", "sim_management.R"))
+verify_version()
 
 write.dir <- file.path(here(),"Project_0", "inputs")
 
@@ -93,7 +95,7 @@ for(i in 1:NROW(df.M.oms)){
   
   #M
   M_i = gf_M
-  M_i$sigma_vals = df.M.oms$M_sig[i]
+  M_i$sigma_vals = df.M.oms$M_sig[i] * sqrt(1-df.M.oms$M_cor[i]^2) #defining marginal variance, but wham estimates conditional var.
   M_i$cor_vals = df.M.oms$M_cor[i]
 
   if(df.M.oms$Fhist[i] == "H-MSY") Fhist. = "H-L"

### 2022-07-25 Project_0/code/Sel_om_setup.R

@@ -17,19 +17,27 @@ verify_version()
 
 write.dir <- file.path(here(),"Project_0", "inputs")
 
+naa_om_inputs = readRDS(file.path(here::here(),"Project_0","inputs", "NAA_om_inputs.RDS"))
+#SR parameters are the same for all operating models 
+temp = fit_wham(naa_om_inputs[[1]], do.fit = FALSE, MakeADFun.silent = TRUE)
+SRab = exp(c(temp$rep$log_SR_a[1], temp$rep$log_SR_b[1]))
+
 #if(!exists("write.dir")) write.dir = getwd()
 if(!dir.exists(write.dir)) dir.create(write.dir, recursive = T)
 setwd(write.dir)
 
 #Operating model factors
 #NAA sigmas for each scenario
-R_sig <- c(0.5,1.5)
-NAA_sig <- c(NA,0.25,0.5)
+R_sig <- c(0.5)
 #F time series
+
+Sel_sig <- c(0.1,0.5)
+Sel_cor <- c(0, 0.9)
+
 Fhist = c("H-MSY","MSY")
 #how much observation error
 obs_error = c("L", "H")
-df.oms <- expand.grid(R_sig = R_sig, NAA_sig = NAA_sig, Fhist = Fhist, 
+df.Sel.oms <- expand.grid(R_sig = R_sig, Sel_sig = Sel_sig, Sel_cor = Sel_cor, Fhist = Fhist, 
   obs_error = obs_error, stringsAsFactors = FALSE)
 
 #logistic-normal age comp SDs for L/H observation error (both indices AND CATCH!!!????)
@@ -37,19 +45,12 @@ L_N_sigma = c(L = 0.3, H = 1.5)
 #(log) index SDs for L/H observation error 
 index_sigma = c(L = 0.1, H = 0.4)
 
-n.mods <- dim(df.oms)[1] #24 operating model scenarios
-df.oms$Model <- paste0("om_",1:n.mods)
-df.oms <- df.oms %>% select(Model, everything()) # moves Model to first col
+n.mods <- dim(df.Sel.oms)[1] #24 operating model scenarios
+df.Sel.oms$Model <- paste0("om_",1:n.mods)
+df.Sel.oms <- df.Sel.oms %>% select(Model, everything()) # moves Model to first col
 # look at model table
-df.oms
-saveRDS(df.oms, file.path(here(),"Project_0", "inputs", "df.oms.RDS"))
-
[…]

### 2022-07-25 Project_0/code/em_setup.R

@@ -8,6 +8,8 @@ library(here)
 files_to_source = list.files(file.path(here(), "common_code"), full.names=TRUE, pattern = "*.R")
 sapply(files_to_source, source)
 source(file.path(here(), "Project_0", "code", "make_om.R"))
+source(file.path(here(), "Project_0", "code", "sim_management.R"))
+verify_version()
 
 naa_om_inputs = readRDS(file.path(here::here(),"Project_0","inputs", "NAA_om_inputs.RDS"))
 # temp = sapply(naa_om_inputs, function(x) {
@@ -156,11 +158,14 @@ for(i in 21:NROW(df.ems)){
   }
   selectivity = gf_selectivity
   if(df.ems$re_config[i] == "sel_re") { #full random effects for NAA
-    selectivity$re = c(df.ems$sel_re_cor[i], "none", "none") #iid or ar1_y
+    selectivity$re = c("2dar1", "none", "none") #ar1_y on slope and a50 but independent
+    selectivity$initial_cor = list(c(0,0), 0, 0) #independent slope and a50, but perhaps correlation over time.
+    selectivity$map_cor = list(c(NA,1), NA, NA)
     #NAA_re$sigma_vals[2] = df.oms$NAA_sig[i] #don't start at true values?
   }
   em_inputs[[i]] <- prepare_wham_input(basic_info = make_basic_info(), selectivity = selectivity, NAA_re = NAA_re, M= M,
     catchability = catchability, age_comp = "logistic-normal-miss0")
+  em_inputs[[i]] <- set_selectivity(em_inputs[[i]], selectivity = selectivity)
   if(df.ems$M_est[i]) { #estimate mean M
     em_inputs[[i]]$map$M_a = factor(rep(1,length(em_inputs[[i]]$par$M_a)))
   }

### 2022-07-25 common_code/set_selectivity.R

@@ -46,8 +46,8 @@
 #'     \item{$map_sigma}{Specify which SD parameters to fix for the random effect deviations. Must be a vector with length = number of blocks. 
 #'                Use \code{NA} to fix a parameter and integers to estimate. Use the same integer for multiple blocks to estimate a shared SD parameter.
 #'                Not used if \code{re = 'none'} for all blocks.}
-#'     \item{$initial_cor}{Initial correlation values to use for the random effect deviations. Must be a list with length = number of blocks. If \code{re = 'ar1'} or \code{ar1_y'},
-#'                must be a single value. If \code{re = '2dar1'}, must be a vector of length 2 (first is for "age", second is for "year"). 
+#'     \item{$initial_cor}{Initial correlation values to use for the random effect deviations. Must be a list with length = number of blocks. If \code{re = 'ar1'} or \code{ar1_y'}
+#'                for a block, list element must be a single value. If \code{re = '2dar1'} for a block, list element must be a vector of length 2 (first is for "age", second is for "year") for each block. 
 #'                Use natural scale, must be between -1 and 1. \code{par$sel_repars[,2:3]} will be estimated on a transform scale,  (2 / (1 + exp(-2x))) - 1. 
 #'                Not used if \code{re = 'none'} or \code{re = 'iid'} for all blocks.}
 #'     \item{$map_cor}{Specify which correlation parameters to fix for the random effect deviations. Must be a list with length = number of blocks. 
@@ -327,8 +327,8 @@ Can use selectivity$fix_pars."))
     par$sel_repars[,1] = log(selectivity$initial_sigma) # log scale
   }
   if(!is.null(selectivity$initial_cor)){
-    if(any(selectivity$initial_cor < -1)) stop('Correlation parameters controlling selectivity random effects must be between -1 and 1.')
-    if(any(selectivity$initial_cor > 1)) stop('Correlation parameters controlling selectivity random effects must be between -1 and 1.')
+    if(any(sapply(selectivity$initial_cor, function(x) any(x < -1)))) stop('Correlation parameters controlling selectivity random effects must be between -1 and 1.')
+    if(any(sapply(selectivity$initial_cor, function(x) any(x > 1)))) stop('Correlation parameters controlling selectivity random effects must be between -1 and 1.')
     for(b in 1:data$n_selblocks){
[…]

### 2022-07-26 misspecification_study/misspec_demo.Rmd
add liz retro plots

@@
 source(file.path(here(), "common_code", "bias_data.R"))
+source(file.path(here(), "common_code", "fn_retro_par_estimates.R"))
 ```
@@
 
-The catch and index four panel plots all look good for UR_em4.
+The catch and index four panel plots all look good for UR_em4. The problematic diagnostic is the NAA 4 panel plots, show below for ages 1 and 8 as well as the stacked barplot of residuals.
 
@@
 
-The problematic diagnostic is the NAA 4 panel plots, show below for ages 1 and 8 as well as the stacked barplot of residuals.
+![NAA age 1 4 panel](UR_em4/plots_png/diagnostics/NAA_4panel_1.png)
 
-![NAA age 1 4 panel](UR_em4/plots_png/diagnostics/NAA_4panel_1.png)
 ![NAA age 8 4 panel](UR_em4/plots_png/diagnostics/NAA_4panel_8.png)
+
 ![NAA residuals stacked barplot](UR_em4/plots_png/diagnostics/NAA_res_barplot_stacked.png)
+
+Use Liz's plot.retro.pars function to examine the stability of parameter estimates across the retrospective peels. For now just look at under-reported catch case with estimation model 4 (UR_em4).
+
+```{r lizretro}
+# again had to run this in console because could not get it to knit
+#em8lr.dir <- file.path(here(), "misspecification_study", "UR_em4", "liz_retro_plots")
+#plot.retro.pars(m=em_fits[[8]][[1]], yr.plot.start=2005, od=em8lr.dir, save.name="UR_em4", 
+#  plot.f='png' )
+```
+
+None of the retrospective plots show a strong pattern. The unusual result of nearly all the NAA log_Deviations being below zero holds for all the peels.
+
+![fleet selectivity](UR_em4/liz_retro_plots/UR_em4._fleet.sel.full.plot.png)
+
+![index selectivity](UR_em4/liz_retro_plots/UR_em4._ind.sel.full.plot.png)
+
+![NAA log deviations](UR_em4/liz_retro_plots/UR_em4._naa.devs.full.plot.png)
+
+![Survey catchability](UR_em4/liz_retro_plots/UR_em4._q.full.plot.png)

### 2022-07-27 misspecification_study/misspec_demo.Rmd
added all sel and q om sim files

@@
                      retro.silent = TRUE, 
-                     save.sdrep = FALSE)
+                     save.sdrep = TRUE)
       return(out)
@@
 
-```{r lizretro}
+```{r lizretro, eval = FALSE}
 # again had to run this in console because could not get it to knit
-#em8lr.dir <- file.path(here(), "misspecification_study", "UR_em4", "liz_retro_plots")
-#plot.retro.pars(m=em_fits[[8]][[1]], yr.plot.start=2005, od=em8lr.dir, save.name="UR_em4", 
-#  plot.f='png' )
+em8lr.dir <- file.path(here(), "misspecification_study", "UR_em4", "liz_retro_plots")
+plot.retro.pars(m=em_fits[[8]][[1]], yr.plot.start=2005, od=em8lr.dir, save.name="UR_em4", 
+  plot.f='png' )
 ```

### 2022-07-27 Project_0/code/Sel_om_hpcc_args.sh

@@ -1,0 +1,26 @@
+#!/bin/bash
+# Load gcc and R and jpeg header (for Hmisc)
+module load gcc/8.1.0
+module load R/4.0.4_gcc
+module load libjpeg-turbo/2.0.2
+module load openmpi/4.0.1
+
+echo "CPU threads: $(grep -c processor /proc/cpuinfo)"
+grep 'cpu cores' /proc/cpuinfo | uniq
+
+cd ~/SSRTWG
+for ((sim= $1; sim <= $2; sim++))
+do
+  for ((om= $3; om <= $4; om++))
+  do
+   for ((em= $5; em <= $6; em++))
+   do
+     echo $sim
+     echo $om
+     echo $em
+     Rscript --vanilla ~/SSRTWG/Project_0/code/Sel_om_hpcc_script.R $sim $om $em &
+   done
+ done
+done
+
+echo "script is done"

### 2022-07-27 Project_0/code/Sel_om_hpcc_bsub.sh

@@ -1,0 +1,19 @@
+#!/bin/bash
+
+#$1 is first sim
+#$2 is last sim
+#$3 is first om
+#$4 is last om
+#$5 is first em
+#$6 is last em
+#$7 is ncores per job (20 or less being the number of ems specified)
+
+for ((this_sim= $1; this_sim <= $2; this_sim++)) #sim in {$simstart..$simend}
+do
+  for ((this_om= $3; this_om <= $4; this_om++)) #om in {$omstart..$omend}
+  do
+    echo $this_om
+    bsub -n $7 -q short -W 4:00 -o ~/logs/logfile.%J -R "rusage[mem=5000]" -R "span[hosts=1]" -J Sel_om_sim "bash ~/SSRTWG/Project_0/code/Sel_om_hpcc_args.sh $this_sim $this_sim $this_om $this_om $5 $6"
+  done
+done
+echo "script is done"

### 2022-07-27 Project_0/code/Sel_om_hpcc_script.R

@@ -1,0 +1,19 @@
+#!/usr/bin/env Rscript
+args = commandArgs(trailingOnly=TRUE)
+sim = as.integer(args[1])
+om = as.integer(args[2])
+em = as.integer(args[3])
+.libPaths("~/Rlib/")
+library(wham)
+source(file.path(here::here(), "Project_0","code", "sim_management.R"))
+verify_version()
+#cat(paste0("number of cores available: ", parallel::detectCores(), "/n"))
+#stop()
+run_hpcc_jobs_rev(this_sim = sim, this_om = om, this_em = em,
+  script.full.path = file.path(here::here(),"Project_0", "code", "Sel_om_sim_fit_script_hpcc.R"), 
+  df.ems = readRDS(file.path(here::here(),"Project_0","inputs", "df.ems.RDS")), 
+  df.oms = readRDS(file.path(here::here(),"Project_0","inputs", "df.Sel.oms.RDS")), 
+  om_inputs = readRDS(file.path(here::here(),"Project_0","inputs", "Sel_om_inputs.RDS")),
+  em_inputs = readRDS(file.path(here::here(),"Project_0","inputs", "em_inputs.RDS")),
+  write.dir = file.path(here::here(),"Project_0", "results", "Sel_om")
+)

### 2022-07-27 Project_0/code/Sel_om_sim_fit_script_hpcc.R

@@ -1,0 +1,69 @@
+#!/usr/bin/env Rscript
+args = commandArgs(trailingOnly=TRUE)
+
+.libPaths("~/Rlib/")
+library(here)
+library(wham)
+
+this_om <- as.integer(args[1])
+this_em <- as.integer(args[2])
+this_sim <- as.integer(args[3])
+
+om_inputs <- readRDS(file.path(here(),"Project_0","inputs", "Sel_om_inputs.RDS"))
+em_inputs <- readRDS(file.path(here(),"Project_0","inputs", "em_inputs.RDS"))
+df.ems <- readRDS(file.path(here(),"Project_0","inputs", "df.ems.RDS"))
+df.oms <- readRDS(file.path(here(),"Project_0","inputs", "df.Sel.oms.RDS"))
+#######################################################
+#need to have matching assumptions about CVs for catch and indices, too
+obs_names <- c("agg_catch","agg_catch_sigma", "agg_indices", "agg_index_sigma", "catch_paa", "index_paa", 
+  "Ecov_obs", "obs", "obsvec")
+#######################################################
+
+#######################################################
+#Do we want to use the same (e.g. 1000) seeds for everything?
+seeds <- readRDS(file.path(here(), "Project_0", "inputs","seeds.RDS"))
+#######################################################
+
+cat(paste0("OM: ", this_om, " Sim: ", this_sim, " EM: ", this_em, "\n"))
+
+# Set seed
+om <- fit_wham(om_inputs[[this_om]], do.fit = FALSE, MakeADFun.silent = TRUE)
+set.seed(seeds[this_sim])
+sim_data <- om$simulate(complete=TRUE)
+truth <- sim_data
+#save the version for reproducibility
+truth$wham_version = om$wham_version
+EM_input <- em_inputs[[this_em]] # Read in the EM 
+#put simulated data into the em input
+EM_input$data[obs_names] = sim_data[obs_names]
+res <- list(truth = truth)
+res$fit <- list()
+#do fit withouth sdreport first
+fit <- tryCatch(fit_wham(EM_input, do.sdrep=F, do.osa=F, do.retro=T, do.proj=F, MakeADFun.silent=TRUE),
+  error = function(e) conditionMessage(e))
+  
+# Deal with issues fitting EM to non-matching OM data
+# empty elements below can be used to summarize convergence information
[…]

### 2022-07-27 Project_0/code/q_om_hpcc_args.sh

@@ -1,0 +1,26 @@
+#!/bin/bash
+# Load gcc and R and jpeg header (for Hmisc)
+module load gcc/8.1.0
+module load R/4.0.4_gcc
+module load libjpeg-turbo/2.0.2
+module load openmpi/4.0.1
+
+echo "CPU threads: $(grep -c processor /proc/cpuinfo)"
+grep 'cpu cores' /proc/cpuinfo | uniq
+
+cd ~/SSRTWG
+for ((sim= $1; sim <= $2; sim++))
+do
+  for ((om= $3; om <= $4; om++))
+  do
+   for ((em= $5; em <= $6; em++))
+   do
+     echo $sim
+     echo $om
+     echo $em
+     Rscript --vanilla ~/SSRTWG/Project_0/code/q_om_hpcc_script.R $sim $om $em &
+   done
+ done
+done
+
+echo "script is done"

### 2022-07-27 Project_0/code/q_om_hpcc_script.R

@@ -1,0 +1,19 @@
+#!/usr/bin/env Rscript
+args = commandArgs(trailingOnly=TRUE)
+sim = as.integer(args[1])
+om = as.integer(args[2])
+em = as.integer(args[3])
+.libPaths("~/Rlib/")
+library(wham)
+source(file.path(here::here(), "Project_0","code", "sim_management.R"))
+verify_version()
+#cat(paste0("number of cores available: ", parallel::detectCores(), "/n"))
+#stop()
+run_hpcc_jobs_rev(this_sim = sim, this_om = om, this_em = em,
+  script.full.path = file.path(here::here(),"Project_0", "code", "q_om_sim_fit_script_hpcc.R"), 
+  df.ems = readRDS(file.path(here::here(),"Project_0","inputs", "df.ems.RDS")), 
+  df.oms = readRDS(file.path(here::here(),"Project_0","inputs", "df.q.oms.RDS")), 
+  om_inputs = readRDS(file.path(here::here(),"Project_0","inputs", "q_om_inputs.RDS")),
+  em_inputs = readRDS(file.path(here::here(),"Project_0","inputs", "em_inputs.RDS")),
+  write.dir = file.path(here::here(),"Project_0", "results", "q_om")
+)

### 2022-07-27 Project_0/code/q_om_setup.R

@@ -55,7 +55,7 @@ df.q.oms$Model <- paste0("om_",1:n.mods)
 df.q.oms <- df.q.oms %>% select(Model, everything()) # moves Model to first col
 # look at model table
 df.q.oms
-#saveRDS(df.q.oms, file.path(here(),"Project_0", "inputs", "df.q.oms.RDS"))
+saveRDS(df.q.oms, file.path(here(),"Project_0", "inputs", "df.q.oms.RDS"))
 
 
 gf_info = make_basic_info()

### 2022-07-28 misspecification_study/misspec_demo.Rmd
add osa - problem encountered

--- 2022-07-26 misspecification_study/misspec_demo.Rmd
@@
 
+## Run the models
+
 Now run the eight models.
@@
 
-```{r runmodels}
+```{r runmodels, results='hide'}
 if(file.exists(file.path(here(), "misspecification_study", "misspec_demo_fits.RDS"))){
@@
       out = fit_wham(sim_input[[m]][[x]], 
-                     do.osa = FALSE,
+                     do.osa = TRUE,
                      MakeADFun.silent = TRUE, 
                      retro.silent = TRUE, 
-                     save.sdrep = FALSE)
+                     save.sdrep = TRUE)
       return(out)
@@
 
+## Correct data results
+
 We can see which model has the best fit using the compare_wham_models function for the two sets of data.
@@
 ```
+
+## Under-reported catch data results
 
@@
 ```{r plotemfit8}
-# note could not get the following code to run within Rmarkdown (it produced the png files but not the pdf)
+# ran into errors plotting the one-step-ahead residuals for each model
+# ran the loop one model at a time to get output
+#
+# em_fits 1 and 5 gave error message
+# Error in mod$osa[c(dat$keep_Cpaa[i, yind, ] + 1), ] : 
+#   incorrect number of dimensions
+#
+# the other six em_fits gave error message
+# Infinite osa residuals for aggregate catch in fleet  1 , so using +/-10 for range of y axis 
+# Error in if (h > 0) max(1, ceiling(diff(range(x))/h)) else 1L : 
+#   missing value where TRUE/FALSE needed
+# In addition: Warning messages:
+# 1: In min(x) : no non-missing arguments to min; returning Inf
+# 2: In max(x) : no non-missing arguments to max; returning -Inf
+
 # can either pull this code out and run in console, or else uncomment and knit then comment and knit again
-#em8.dir <- file.path(here(), "misspecification_study", "UR_em4")
-#if(!dir.exists(em8.dir)){
-#  dir.create(em8.dir)
-#}
-#plot_wham_output(mod=em_fits[[8]][[1]], dir.main=em8.dir)
-
+# out.dir <- file.path(here(), "misspecification_study", "output")
+# if (!dir.exists(out.dir)){
+#   dir.create(out.dir)
+# }
+# mydir <- c(paste0("CC_em", 1:4), paste0("UR_em", 1:4))
+# for (i in 1:8){
+#   em.dir <- file.path(here(), "misspecification_study", "output", mydir[i])
+#   if(!dir.exists(em.dir)){
+#     dir.create(em.dir)
+#   }
+#   plot_wham_output(mod=em_fits[[i]][[1]], dir.main=em.dir)
+# }
 ```
@@
 
-![catch 4 panel](UR_em4/plots_png/diagnostics/Catch_4panel_fleet_1.png)
+![UR_em4 catch 4 panel](output/UR_em4/plots_png/diagnostics/Catch_4panel_fleet_1.png)
 
-![index 1 4 panel](UR_em4/plots_png/diagnostics/Index_4panel_1.png)
+![UR_em4 index 1 4 panel](output/UR_em4/plots_png/diagnostics/Index_4panel_1.png)
 
-![index 2 4 panel](UR_em4/plots_png/diagnostics/Index_4panel_2.png)
+![UR_em4 index 2 4 panel](output/UR_em4/plots_png/diagnostics/Index_4panel_2.png)
 
-![NAA age 1 4 panel](UR_em4/plots_png/diagnostics/NAA_4panel_1.png)
+![UR_em4 NAA age 1 4 panel](output/UR_em4/plots_png/diagnostics/NAA_4panel_1.png)
 
-![NAA age 8 4 panel](UR_em4/plots_png/diagnostics/NAA_4panel_8.png)
+![UR_em4 NAA age 8 4 panel](output/UR_em4/plots_png/diagnostics/NAA_4panel_8.png)
 
-![NAA residuals stacked barplot](UR_em4/plots_png/diagnostics/NAA_res_barplot_stacked.png)
+![UR_em4 NAA residuals stacked barplot](output/UR_em4/plots_png/diagnostics/NAA_res_barplot_stacked.png)
 
@@
 
-```{r lizretro}
+```{r lizretro, eval = FALSE}
 # again had to run this in console because could not get it to knit
-#em8lr.dir <- file.path(here(), "misspecification_study", "UR_em4", "liz_retro_plots")
-#plot.retro.pars(m=em_fits[[8]][[1]], yr.plot.start=2005, od=em8lr.dir, save.name="UR_em4", 
+# for (i in 1:8){
+#   emlr.dir <- file.path(here(), "misspecification_study", "output", mydir[i], "liz_retro_plots")
+#   if(!dir.exists(emlr.dir)){
+#     dir.create(emlr.dir)
+#   }
+#   plot.retro.pars(m=em_fits[[i]][[1]], yr.plot.start=2005, od=emlr.dir, save.name=mydir[i], 
 #                   plot.f='png' )
+# }
 ```
@@
 
-![fleet selectivity](UR_em4/liz_retro_plots/UR_em4._fleet.sel.full.plot.png)
+![UR_em4 fleet selectivity](output/UR_em4/liz_retro_plots/UR_em4._fleet.sel.full.plot.png)
 
-![index selectivity](UR_em4/liz_retro_plots/UR_em4._ind.sel.full.plot.png)
+![UR_em4 index selectivity](output/UR_em4/liz_retro_plots/UR_em4._ind.sel.full.plot.png)
 
-![NAA log deviations](UR_em4/liz_retro_plots/UR_em4._naa.devs.full.plot.png)
+![UR_em4 NAA log deviations](output/UR_em4/liz_retro_plots/UR_em4._naa.devs.full.plot.png)
 
-![Survey catchability](UR_em4/liz_retro_plots/UR_em4._q.full.plot.png)
+![UR_em4 Survey catchability](output/UR_em4/liz_retro_plots/UR_em4._q.full.plot.png)
+
+Interestingly, for estimation model 3, both the fits and the NAA log deviations showed a strong pattern of positive then negative or else negative then positive residuals over time.
+
+![UR_em3 catch 4 panel](output/UR_em3/plots_png/diagnostics/Catch_4panel_fleet_1.png)
+
+![UR_em3 index 1 4 panel](output/UR_em3/plots_png/diagnostics/Index_4panel_1.png)
+
+![UR_em3 index 2 4 panel](output/UR_em3/plots_png/diagnostics/Index_4panel_2.png)
+
+![UR_em3 NAA age 8 4 panel](output/UR_em3/plots_png/diagnostics/NAA_4panel_8.png)
+
+![UR_em3 NAA residuals stacked barplot](output/UR_em3/plots_png/diagnostics/NAA_res_barplot_stacked.png)

### 2022-07-30 misspecification_study/misspec_demo.Rmd
figuring out OSA residual problem

@@
 
+obs_names = c("agg_indices","agg_catch","catch_paa","index_paa", "Ecov_obs", "obsvec")
+
 # sim_input[[1]] has no data modification and em1
@@
   sim = om$simulate(complete=TRUE)
-  input_i$data = sim
+  input_i$data[obs_names] = sim[obs_names]
   return(input_i)
@@
   input_i = em_input2
-  obs_names = c("agg_indices","agg_catch","catch_paa","index_paa", "Ecov_obs", "obsvec")
   input_i$data[obs_names] = sim_input[[1]][[x]]$data[obs_names]
@@
   input_i = em_input3
-  obs_names = c("agg_indices","agg_catch","catch_paa","index_paa", "Ecov_obs", "obsvec")
   input_i$data[obs_names] = sim_input[[1]][[x]]$data[obs_names]
@@
   input_i = em_input4
-  obs_names = c("agg_indices","agg_catch","catch_paa","index_paa", "Ecov_obs", "obsvec")
   input_i$data[obs_names] = sim_input[[1]][[x]]$data[obs_names]
@@
   input_i = em_input2
-  obs_names = c("agg_indices","agg_catch","catch_paa","index_paa", "Ecov_obs", "obsvec")
   input_i$data[obs_names] = sim_input[[5]][[x]]$data[obs_names]
@@
   input_i = em_input3
-  obs_names = c("agg_indices","agg_catch","catch_paa","index_paa", "Ecov_obs", "obsvec")
   input_i$data[obs_names] = sim_input[[5]][[x]]$data[obs_names]
@@
   input_i = em_input4
-  obs_names = c("agg_indices","agg_catch","catch_paa","index_paa", "Ecov_obs", "obsvec")
   input_i$data[obs_names] = sim_input[[5]][[x]]$data[obs_names]
@@
 
-```{r plotemfit8}
+```{r plotemfit8, eval = FALSE}
 # ran into errors plotting the one-step-ahead residuals for each model
@@
 # can either pull this code out and run in console, or else uncomment and knit then comment and knit again
-# out.dir <- file.path(here(), "misspecification_study", "output")
-# if (!dir.exists(out.dir)){
-#   dir.create(out.dir)
-# }
-# mydir <- c(paste0("CC_em", 1:4), paste0("UR_em", 1:4))
-# for (i in 1:8){
-#   em.dir <- file.path(here(), "misspecification_study", "output", mydir[i])
-#   if(!dir.exists(em.dir)){
-#     dir.create(em.dir)
-#   }
-#   plot_wham_output(mod=em_fits[[i]][[1]], dir.main=em.dir)
-# }
+out.dir <- file.path(here(), "misspecification_study", "output")
+if (!dir.exists(out.dir)){
+  dir.create(out.dir)
+}
+mydir <- c(paste0("CC_em", 1:4), paste0("UR_em", 1:4))
+for (i in 1:8){
+  em.dir <- file.path(here(), "misspecification_study", "output", mydir[i])
+  if(!dir.exists(em.dir)){
+    dir.create(em.dir)
+  }
+  plot_wham_output(mod=em_fits[[i]][[1]], dir.main=em.dir)
+}
 ```

### 2022-07-30 Project_0/code/aggregate_results_hpcc.R

@@ -29,7 +29,20 @@ aggregate_hpcc_results(sim = 1, oms = 1, ems = 5:24, res_dir = file.path(here::h
 sapply(1:100, function(x) aggregate_hpcc_results(sim = x, oms = 1, ems = 5:24, res_dir = file.path(here::here(),"Project_0", "results", "M_om")))
 sapply(2:16, function(y) sapply(1:100, function(x) aggregate_hpcc_results(sim = x, oms = y, ems = 5:24, res_dir = file.path(here::here(),"Project_0", "results", "M_om"))))
 
-sapply(1:100, function(x) aggregate_hpcc_results(sim = x, oms = 1, ems = ems, res_dir = file.path(here::here(),"Project_0", "results", "naa_om"))
+
+#Sel operating models
+df.ems = readRDS(file.path(here(),"Project_0","inputs", "df.ems.RDS"))
+df.oms = readRDS(file.path(here(),"Project_0","inputs", "df.Sel.oms.RDS"))
+om_inputs = readRDS(file.path(here::here(),"Project_0","inputs", "Sel_om_inputs.RDS"))
+em_inputs = readRDS(file.path(here::here(),"Project_0","inputs", "em_inputs.RDS"))
+oms = 1:length(om_inputs)
+ems = 1:length(em_inputs)
+ems = c(5:20,25:28)
+aggregate_hpcc_results(sim = 1, oms = 1, ems = ems, res_dir = file.path(here::here(),"Project_0", "results", "Sel_om"))
+sapply(1:100, function(x) aggregate_hpcc_results(sim = x, oms = 1, ems = ems, res_dir = file.path(here::here(),"Project_0", "results", "Sel_om")))
+sapply(2:16, function(y) sapply(1:100, function(x) aggregate_hpcc_results(sim = x, oms = y, ems = ems, res_dir = file.path(here::here(),"Project_0", "results", "Sel_om"))))
+
+
 
 x = readRDS(file.path(here::here(),"Project_0", "results", "naa_om", "om_1", "sim_7.RDS"))
 x = readRDS(file.path(here::here(),"Project_0", "results", "naa_om", "om_1", "sim_1.RDS"))

### 2022-08-12 misspecification_study/misspec_demo.Rmd
knit with Tim's sim fix, osa fix still needed
make some minor edits to misspec_demo.Rmd

@@
 
-obs_names = c("agg_indices","agg_catch","catch_paa","index_paa", "Ecov_obs", "obsvec")
+obs_names = c("agg_indices","agg_catch","catch_paa","index_paa", "Ecov_obs", "obsvec", "agesvec")
 
@@
   input_i$data[obs_names] = sim[obs_names]
+  input_i$data$obs$val = input_i$data$obsvec
   return(input_i)
@@
   input_i$data[obs_names] = sim_input[[1]][[x]]$data[obs_names]
+  input_i$data$obs$val = input_i$data$obsvec
   return(input_i)
@@
                    agg_catch_multiplier=agg_catch_multiplier)
-  input_i$data = sim
+  input_i$data[obs_names] = sim[obs_names]
+  input_i$data$obs$val = input_i$data$obsvec
   return(input_i)
@@
   input_i$data[obs_names] = sim_input[[5]][[x]]$data[obs_names]
+  input_i$data$obs$val = input_i$data$obsvec
   return(input_i)

### 2022-08-12 common_code/make_OSA_residuals.R

@@ -1,78 +1,0 @@
-make_OSA_residuals = function(mod,osa.opts = list(method="cdf", parallel=TRUE)){
-  # one-step-ahead residuals
-  if(mod$is_sdrep){ # only do OSA residuals if sdrep ran
-    cat("Doing OSA residuals...\n");
-    input = mod$input
-    full_set = 1:length(input$data$obsvec)
-    input$data$obs$residual = NA
-    if(!is.null(input$data$condition_no_osa)) cat("OSA not available for some age comp likelihoods...\n")
-    #first do continuous obs, condition on obs without osa (probably none)
-    subset. = setdiff(full_set, c(input$data$subset_discrete_osa, input$data$conditional_no_osa))
-    print("subset.")
-    print(subset.)
-    print(osa.opts$method)
-    print(osa.opts$parallel)
-    mod$env$data$do_osa = 1
-    mod$env$retape()
-    x = TMB::oneStepPredict(mod, method = "cdf", subset = subset.,
-    observation.name = "obsvec", data.term.indicator="keep", discrete=F, parallel= T)
-    print(head(x))
-    stop()
-#
-    OSA.continuous <- suppressWarnings(TMB::oneStepPredict(obj=mod, observation.name="obsvec",
-                                data.term.indicator="keep",
-                                method=osa.opts$method,
-                                discrete=FALSE, parallel=osa.opts$parallel,
-                                subset = subset., conditional = input$data$conditional_no_osa))
-    input$data$obs$residual[subset.] <- OSA.continuous$residual;
-    print(OSA.continuous$residual)
-    stop()
-    mod$OSA.continuous = OSA.continuous
-    if(!is.null(input$data$subset_discrete_osa)) {
-      cat("Doing OSA for discrete age comp likelihoods...\n")
-      conditional = union(input$data$condition_no_osa, subset.) #all with continuous and without osa 
-      subset. = input$data$subset_discrete_osa
-      #first do continuous
-      OSA.discrete <- suppressWarnings(TMB::oneStepPredict(obj=mod, observation.name="obsvec",
-                                  data.term.indicator="keep",
-                                  method= osa.opts$method,
-                                  discrete=TRUE, parallel=osa.opts$parallel,
-                                  conditional = conditional))
-      input$data$obs$residual[subset.] <- OSA.discrete$residual
[…]

### 2022-08-16 misspecification_study/misspec_demo.Rmd
apply Tim's fix for osa using model_q branch of wham

@@
 ```{r comparecorrect}
-correct_data_compare <- compare_wham_models(list(CD_em1 = em_fits[[1]][[1]],
-                                                 CD_em2 = em_fits[[2]][[1]],
-                                                 CD_em3 = em_fits[[3]][[1]],
-                                                 CD_em4 = em_fits[[4]][[1]]))
-
-print(correct_data_compare$g[[1]])
-print(correct_data_compare$g[[2]])
+# don't know why this code is not working anymore
+# correct_data_compare <- compare_wham_models(list(CD_em1 = em_fits[[1]][[1]],
+#                                                  CD_em2 = em_fits[[2]][[1]],
+#                                                  CD_em3 = em_fits[[3]][[1]],
+#                                                  CD_em4 = em_fits[[4]][[1]]))
+# 
+# print(correct_data_compare$g[[1]])
+# print(correct_data_compare$g[[2]])
+#        dAIC    AIC  rho_R rho_SSB rho_Fbar
+# CD_em2  0.0 5008.4 0.2629 -0.0418   0.0573
+# CD_em1  2.4 5010.8 0.3136 -0.0285   0.0416
+# CD_em4  3.9 5012.3 0.2772 -0.0340   0.0506
+# CD_em3  4.4 5012.8 0.3136 -0.0285   0.0416
+# Error in if (is.na(rel.f.vals[x]) | any(diag(log.rel.ssb.rel.F.cov[[x]]) <  : 
+#   missing value where TRUE/FALSE needed
 ```
@@
 
-The catch and index four panel plots all look good for UR_em4. The problematic diagnostic is the NAA 4 panel plots, show below for ages 1 and 8 as well as the stacked barplot of residuals.
+The catch and index four panel plots all look good for UR_em4.
+The problematic diagnostic is the NAA 4 panel plots, show below for ages 1 and 8 as well as the stacked barplot of residuals.
+Also, none of the one-step-ahead residual plots are currenlty working.
+Tim is looking into this, could be a non-trivial fix is required.
 
@@
 ![UR_em4 NAA residuals stacked barplot](output/UR_em4/plots_png/diagnostics/NAA_res_barplot_stacked.png)
+
+![UR_em4 osa residuals for fleet](output/UR_em4/plots_png/diagnostics/Catch_age_comp_osa_resids_fleet_1.png)
 

### 2022-08-19 Ecov_study/Ecov_plots_GLB_recruitment.Rmd
simple ecov recruit sims

---
title: "A preliminary look at some ecov simulations"
author:
  - Gregory Britten, ...
output:
  pdf_document: default
---

```{r}
em_input <- readRDS(file.path(write.dir, "em_input_GLB_recruitment.RDS"))

yrs <- em_fits[[15]][[1]]$years
```

PLOT SIMULATED AND ESTIMATION ECOV REs
```{r}
par(mfrow=c(2,2))
for(i in 1:4){
  sim <- sim_input[[15]][[i]]$data$Ecov_re
  est <- em_fits[[15]][[i]]$parList$Ecov_re

  plot(yrs,  sim,xlim=c(1980,2021),ylim=range(c(sim,est)))
  points(yrs,est,col='red')
}
```

PLOT ESTIMATED BETAS WITH SIMULATED VALUE
```{r}
ms <- c(1,15)
par(mfrow=c(1,2)) 
for(i in 1:length(ms)){
    hist(unlist(lapply(1:nsim, function(x){
    betas <- em_fits[[ms[i]]][[x]]$parList$Ecov_beta[1,1,1,1]
    return(betas)
})), main='',xlab='')
abline(v=df.mods[ms[i],]$beta)
}
```

### 2022-08-19 Ecov_study/Ecov_plots_GLB_recruitment.R

@@ -1,0 +1,29 @@
+
+
+yrs <- em_fits[[15]][[1]]$years
+
+##_-PLOT SIMULATED AND ESTIMATION ECOV REs--##############################
+pdf('ecov_re_diff.pdf',width=9)
+par(mfrow=c(2,2))
+for(i in 1:4){
+  sim <- sim_input[[15]][[i]]$data$Ecov_re
+  est <- em_fits[[15]][[i]]$parList$Ecov_re
+  
+  plot(yrs,  sim,xlim=c(1980,2021),ylim=range(c(sim,est)),ylab='')
+  points(yrs,est,col='red')
+}
+dev.off()
+
+##--PLOT ESTIMATED BETAS WITH SIMULATED VALUE--_####################
+pdf('beta_ecov_hists.pdf',height=4)
+ms <- c(1,15)
+par(mfrow=c(1,2)) 
+for(i in 1:length(ms)){
+    hist(unlist(lapply(1:nsim, function(x){
+    betas <- em_fits[[ms[i]]][[x]]$parList$Ecov_beta[1,1,1,1]
+    return(betas)
+})), main='',xlab='')
+abline(v=df.mods[ms[i],]$beta,lwd=2)
+}
+dev.off()
+

### 2022-08-19 Ecov_study/Ecov_sim_setup_GLB_recruitment.R

@@ -1,0 +1,157 @@
+# devtools::load_all("~/work/wham/wham")
+# devtools::install_github("timjmiller/wham", dependencies=TRUE, ref="devel")
+library(wham)
+library(tidyr)
+library(dplyr)
+library(here)
+source(file.path(here(),"common_code", "set_Ecov.R"))
+# create directory for analysis, e.g.
+
+
+write.dir <- file.path(here(),"Ecov_study", "results")
+
+if(!exists("write.dir")) write.dir = getwd()  #if we don't specify above, set as current wd
+if(!dir.exists(write.dir)) dir.create(write.dir, recursive = T)  #if the write.dir directory doesn't exist, create it
+setwd(write.dir)
+
+#number of simulations for each scenario
+#nsim = 1000
+nsim = 10
+
+#M sigmas and correlation parameters for each scenario
+Ecov_sig   <- c(0.1, 0.3, 0.5) #units?
+ar1_y      <- c(0,0.25,0.5,0.75)
+Ecov_mean  <- 0
+beta       <- c(0,0.3,0.7) #units?
+Ecov_where <- c("recruit")
+obs_sig    <- c(1e-5,0.1,0.25) #units?
+
+df.mods <- expand.grid(Ecov_sig = Ecov_sig, Ecov_phi = ar1_y, Ecov_mean = Ecov_mean, beta = beta, Ecov_where = Ecov_where, obs_sig = obs_sig)
+
+n.mods <- dim(df.mods)[1] #108 scenarios
+df.mods$Model <- paste0("m_",1:n.mods)
+df.mods <- df.mods %>% select(Model, everything()) # moves Model to first col
+df.mods <- cbind(Recruitment = 2, NAA_re = "rec+1", df.mods) #recruit model not yet discussed by WG.
+# look at model table: 28 scenarios
+df.mods
+
+#make a list of input components that prepare_wham_input can use to generate an input for fit_wham
+#this will be the generic flatfish/groundfish life histor information
+source(file.path(here(),"common_code", "make_basic_info.R"))
+groundfish_info = make_basic_info()
+
+selectivity = list(model = c(rep("logistic", groundfish_info$n_fleets),rep("logistic", groundfish_info$n_indices)),
+    initial_pars = rep(list(c(5,1)), groundfish_info$n_fleets + groundfish_info$n_indices)) #fleet, index
+
+M = list(initial_means = rep(0.2, length(groundfish_info$ages)))
+
+
+sim_input = list()
[…]

### 2022-10-15 Project_0/code/heat.plot.fn.R
some prelim results for Project_0

@@ -32,24 +32,24 @@ heat.plot.fn = function (mat, ylabs, xlabs, main.title = "N sdreps", shape.size 
     title(main.title, outer = TRUE, line = 1)
     par(origpar)
 }
-use.df.ems = df.ems[1:20,]
-ynames = paste0(
-  c("MR","SR")[match(use.df.ems$SR_model, c(2,3))],
-  "_", c("ME","MF")[match(use.df.ems$M_est,c(TRUE,FALSE))],
-  "_", use.df.ems$re_config)
-xnames = paste0(
-  c("HM","MSY")[match(df.oms$Fhist, c("H-MSY","MSY"))],
-  "_", c("RsigL","RsigH")[match(df.oms$R_sig,c(0.5,1.5))],
-  "_", c("Nsig0", "NsigL", "NsigH")[match(df.oms$NAA_sig, c(NA,0.25,0.5))], 
-  "_", c("OEL", "OEH")[match(df.oms$obs_error, c("L","H"))])
+# use.df.ems = df.ems[1:20,]
+# ynames = paste0(
+#   c("MR","SR")[match(use.df.ems$SR_model, c(2,3))],
+#   "_", c("ME","MF")[match(use.df.ems$M_est,c(TRUE,FALSE))],
+#   "_", use.df.ems$re_config)
+# xnames = paste0(
+#   c("HM","MSY")[match(df.oms$Fhist, c("H-MSY","MSY"))],
+#   "_", c("RsigL","RsigH")[match(df.oms$R_sig,c(0.5,1.5))],
+#   "_", c("Nsig0", "NsigL", "NsigH")[match(df.oms$NAA_sig, c(NA,0.25,0.5))], 
+#   "_", c("OEL", "OEH")[match(df.oms$obs_error, c("L","H"))])
 
-png(file.path(here::here(), "Project_0", "results", "naa_om_n_sdrep.png"), 
-  width = 10*144, height = 10*144, res = 144, pointsize = 12)
-heat.plot.fn(n_ems_sdrep, xlabs = xnames, ylabs = ynames, palette = "viridis")
-dev.off()
-#source("c:/work/SSRTWG/SSRTWG/Project_0/code/bubble_plot.R")
+# png(file.path(here::here(), "Project_0", "results", "naa_om_n_sdrep.png"), 
+#   width = 10*144, height = 10*144, res = 144, pointsize = 12)
+# heat.plot.fn(n_ems_sdrep, xlabs = xnames, ylabs = ynames, palette = "viridis")
+# dev.off()
[…]

### 2022-11-16 Ecov_study/catchability/Catchability_ecov_sims.Rmd
Add functions and code to run env-linked q simulations, still needs OM/EM specifics.

---
title: "Catchability_ecov_sims"
output: html_document
date: "2022-11-14"
---
# Function definitions
Functions based on simulation testing vignette: https://github.com/timjmiller/wham/blob/devel/vignettes/ex10_simulation.Rmd

## Generate simulated OM data
```{r}
#' @param OM A fitted WHAM model to use as the operating model, no default.
#' 
#' @return A simulated OM dataset, may be used to overwrite input$data either entirely or in part in simulations

simOM <- function(OM=NULL){
        input <- OM$input
        dataOM = OM$simulate(complete=TRUE) # Overwrite input with simulated data & return
        return(dataOM)
}
```

## Fit specified EMs to simulated OM data and store results
```{r}
#' @param nsim A number specifying how many simulations to run, default = 1.
#' @param OM A fitted WHAM model to use as an operating model, no default.
#' @param inputEMlist A list of WHAM input data objects for all estimation models, no default.
#' @param outdir A filename specifying the directory where time-stamped results should be stored, if NULL (default) then the results are not saved as RData,
#' 
#' @return A nested list with names and results corresponding to the operating models and all estimation models. For each model results contains a sub-list with items for each simulation's results stored as sub-sub lists (because WHAM results are stored as a list). If outdir != NULL then results are saved as RData and returned as an object from the function.

simTestWHAM <- function(nsim = 1,
                        OM = NULL,
                        inputEMlist = NULL,
                        outdir = NULL){
        # Simulation set-up
        obs_names = c("agg_indices","agg_catch","catch_paa","index_paa", "Ecov_obs", "obsvec") # Data overwritten by OM
        namesEM <- lapply(inputEMlist, FUN = function(inputEMlist){inputEMlist$model_name}) %>% unlist() # EM model names

        # Storage set-up
        results <- NULL #!!!!! Still need to work on setting this up
        storeEM <- lapply(namesEM, FUN = assign, value = list())
        names(storeEM) <- namesEM

        # Generate OM data for all simulations
        sim_inputs <- replicate(nsim, simOM(OM), simplify = FALSE)

        for(isim in 1:nsim){ # Loop over simulations
                for(iEM in 1:length(inputEMlist)){ # Loop over estimation models
                        # Pull EM initial input from list
                        inputEM <- inputEMlist[[iEM]]

                        # Update EM input with OM simulated data
                        inputEM$data[obs_names] = sim_inputs[[isim]][obs_names]

                        # Fit WHAM model to updated EM input and save results
                        resultEM <- fit_wham(inputEM, do.osa = FALSE, do.retro = FALSE, MakeADFun.silent = TRUE) #!!!! Figure out how to store

                        # Store model-specific results for given isim here (append EM results to storage list for that EM)
                        storeEM[which(namesEM == namesEM[iEM])][[1]] <- append(storeEM[which(namesEM == namesEM[iEM])], list(resultEM))

                } # End loop over EM
        } # End loop over nsims

        # Store all results
        results$OM <- sim_inputs
        names(results) <- OM$model_name # Name OM column based on model_name
        results <- append(results, storeEM) # Should add named EM results to the final resuts object

        # Save time-stamped results as RData if output directory provided as argument
        timeStamp <- Sys.time() %>% gsub(" ", "_",.) %>% gsub(":", "-", .) # Change millisecond half of Sys.time() output to avoid having spaces/weird characters in filenames
        if(is.null(outdir)==FALSE){
                save(results, file = paste(outdir, paste0("simWHAM_", nsim, "_nsim_", OM$model_name, "_OM_", timeStamp, ".RData"), sep="/"))
        }

        return(results) # Always return results as an object
}
```

### Visual describing results structure
WHAM results objects only displayed for OM simulation 1, similar structure for other simulations and for estimation models.
```{r}
library(DiagrammeR)

grViz("

digraph boxes_and_circles{
## Add node statements
# Define shape
node[shape = box]
OM[label = 'OM \n results$OM']
sim1[label = 'Simulation 1 results \n results$OM[[1]]']
sim2[label = 'Simulation 1 results \n results$OM[[2]]']
sim3[label = 'Simulation 1 results \n results$OM[[3]]']

EM1[label = 'EM1 \n results$EM1']
sim1_1[label = 'Simulation 1 results \n results$EM1[[1]]']
sim1_2[label = 'Simulation 1 results \n results$EM1[[1]]']
sim1_3[label = 'Simulation 1 results \n results$EM1[[1]]']
EM2[label = 'EM2 \n results$EM2']
sim2_1[label = 'Simulation 1 results \n results$EM2[[1]]']
sim2_2[label = 'Simulation 1 results \n results$EM2[[1]]']
sim2_3[label = 'Simulation 1 results \n results$EM2[[1]]']
ageslab[label = 'ages.lab']

# Add edge statements to connect nodes
edge[] # optional arrowhead options
# Connect nodes (may be labeled - appears on line)
results -> OM; OM -> sim1; OM -> sim2; OM -> sim3;
results -> EM1; EM1 -> sim1_1; EM1 -> sim1_2; EM1 -> sim1_3;
results -> EM2; EM2 -> sim2_1; EM2 -> sim2_2; EM2 -> sim2_3;
sim1 -> par;  
sim1 -> fn;
sim1 -> gr;
sim1 -> he;
sim1 -> hessian;
sim1 -> method;
sim1 -> retape;
sim1 -> env;
sim1 -> report;
sim1 -> simulate;
sim1 -> years;
sim1 -> years_full;
sim1 -> ageslab;
sim1 -> model_name;
sim1 -> input;
sim1 -> wham_version;
sim1 -> opt;
sim1 -> date;
sim1 -> dir;
sim1 -> rep;
sim1 -> TMB_version;
sim1 -> parList;
sim1 -> final_gradient;
sim1 -> is_sdrep;
sim1 -> na_sdrep;
sim1 -> runtime;
sim1 -> sdrep;

# Add graph statement
graph[nodesep = 0.1, rankdir = LR] # set distance between nodes # for left-right orientation use , rankdir = LR

}   
")

```

## Post-process results
```{r}
#' @param filenames A vector of strings (including .Rda extensions) indicating what Rdata results files to read in and post process
#' 
#' @return

postprocess_simTestWHAM <- function(filenames = NULL){
        # Set up storage for processed results

        # Loop over files
        for(ifile in 1:length(filenames)){
                # Read in Rdata object
                fileDat <- readRDS(filenames[ifile])

                nsim <- # Based on length of results in given filename

                # Loop over simulations in each file
                for(isim in 1:nsim){
                        # Calculate performance metrics here & add to storage
                }

        } # End loop over filenames
}
```

# Specifics for environmentally-linked catchability WHAM runs
## Define Operating Models (OMs)
All OMs should have catchability linked to an environmental covariate but may differ in 
- Uncertainty around this relationship #!!!!!!!!
- Sign of this relationship (+/-) #!!!!!!!!

Set up input and fit OM1
```{r}
inputOM1 <- prepare_wham_input(model_name = "OM1") #!!!!!!!!!!!!
OM1 <- fit_wham(inputOM1, do.osa = FALSE, do.retro = FALSE, MakeADFun.silent = TRUE)
```

## Define Estimation Models (EMs)
Correctly specified link between catchability and the environment
```{r}
EMq_env <- prepare_wham_input(model_name = "EMq_env")
```

Catchability random effect and a link between catchability and the environment
```{r}
EMq_rand_env <- prepare_wham_input(model_name = "EMq_rand_env")
```

Only a catchability random effect, fit to environmental model without link
```{r}
EMq_rand <- prepare_wham_input(model_name ="EMq_rand")
```

No catchability considerations (i.e. status quo), fit to environmental model without link !!!! how does this work if you feed in the ecov_beta?
```{r}
EMq <- prepare_wham_input(model_name = "EMq")
```

## Simulation testing 
```{r}
simTestWHAM(nsim = 5, OM = OM1, inputEMlist = list(EMq_env, EMq_rand_env, EMq_rand, EMq)) # Update once OM/EM specifics finalized!!!!

```

## Post-process simulation results
```{r}
postprocess_simTestWHAM(filenames = c("filename1.Rds", "filename2.Rds", "ect.Rds")) # !!! Update once completed runs
```

!!!! post-process based on storage & how sims run
- could run sims based varying isim
- could also loop over OM but this could be computationally expensive
- I recommend labeling runs with OM and EM names (and setting these up with informative) so it is easy to pull out results & compare based on model setup
  - pull from mod$model_name to help aggregate data

proposed function with:
- fitted OM model object
- list of EM inputs

### Result formatted as list of lists, save as rdata object in specified directory with timestamp in name
run1_11_14_22.Rda<-result
 - EM1 
 - EM2 
 - EM3
 - EM4
 - OM result

# Other things to consider
- Other EMs can be developed (e.g. with alternative ecov process assumptions) but should be given DIFFERENT model_names to help with post-processing

# Short script for function testing purposes only!!!
```{r}
library(wham)
# For testing function
testdata <- read_asap3_dat(paste("/Users/amandahart/Research/Plaice_Methods_Development", "data", "PlaiceWHAM-2019_revised_NEFSC-LW-WAA_splitNEFSC-BigUnit_fixSeason.DAT", sep="/")) # this can be swapped out for any ASAP file you have to set up a WHAM model for debugging
testinput <- prepare_wham_input(testdata)
testWHAM <- fit_wham(testinput, MakeADFun.silent = TRUE, do.osa = FALSE, do.retro=FALSE)

OM <- testWHAM
testinput1 <- prepare_wham_input(testdata, model_name = "test1")
testinput2 <- prepare_wham_input(testdata, model_name="test2")
testinput3 <- prepare_wham_input(testdata, model_name="test3")
inputEMlist <- list(testinput1, testinput2, testinput3)

test <- simTestWHAM(nsim = 2, OM = OM, inputEMlist = inputEMlist)
```

### 2022-12-07 Project_0/paper/paper.Rmd
some prelim results for Project_0
saving progress on Project_0

@@
   process error assumtions
-  Fhistory description
 
@@
 Fishing history
+All operating models assumed one of two different fishing histories. 
+One : Fishing mortality is equal to Fmsy for the whole 40 year period.
+Two : Fishing mortality is 2.5 times Fmsy for the first 20 years then changes to Fmsy for the last 20 years.
 
-All operating models assumed one of two different fishing histories. One 
-Two alfishing histories are 
-Overfishing assumptions is based on average estimates of overfishing for NE groundfish stocks from Wiedenman et al. XXXX.
+Overfishing assumptions is based on average estimates of overfishing for NE groundfish stocks from Wiedenman et al.
+(20XX).
+Legault et al.
+(2023) also used similar approaches to defining fishing mortality histories for operating models.
 
-All post-recruit productivity components are constant in the NAA and survey catchability process error operating models and we set stock-recruit parameters such that unfished recruitment = exp(10) and
-Fmsy = F40  (Miller and Brooks 2021). For M
+We assumed a Beverton-Holt stock recruit function with constant pre-recruit mortality parameters for all operating models.
+All post-recruit productivity components are constant in the NAA and survey catchability process error operating models.
+Therefore steepness and unfished recruitment are also constant over the time period for those operating models (Miller and Brooks 2021).
+We specified unfished recruitment = $e^10$ and $F_{\text{MSY}} = F_{40}$ equated to a steepness of 0.68 and a and b parameters X and X?
+
+For operating models with time-varying random effects for M, steepness is not constant, but we used the same alpha and beta parameters as other operating models this equates to a steepness and R0 at the mean of the time series process for M.
+
+For operating models with time-varying random effects for fishery selectivity, Fmsy is also not constant however we use the same Fhistory as other operating models which corresposds to Fmsy at the mean selectivity parameters.
 
@@
 M_oms
+NOTE: inv_trans_rho function in set_M.R is mis-defined. Will affect correlation parameters assigned in operating models?
 
@@
 
+Simulations were all carried out on the University of Massachussetts Green High-Performance Computing Cluster.
+Code for completing the simulations and summarization of results can be found at github.com/timjmiller/SSRTWG/Project_0.
+We used the wham package version 1.X.X (commit XXXXX).
+
 # Results
@@
 
+\clearpage 
+## Numbers at age operating models
+
+### Estimating models include alternative random effects options: NAA, M, sel, q
+\clearpage
+
+\begin{table}
+\caption{NAA operating models, estimating models all assume a B-H stock recruit relationship and M is fixed at the true value.}
+{\input{naa_om_em_SR_MF_aic_table}}
+\end{table}
+
+\begin{table}
+\caption{NAA operating models, estimating models all assume a B-H stock recruit relationship and M is estimated.}
+{\input{naa_om_em_SR_ME_aic_table}}
+\end{table}
+
+\begin{table}
+\caption{NAA operating models, estimating models all estimate a mean recruitment and M is fixed at the true value.}
+{\input{naa_om_em_R_MF_aic_table}}
+\end{table}
+
+\begin{table}
+\caption{NAA operating models, estimating models all estimate a mean recruitment and M estimated.}
+{\input{naa_om_em_R_MF_aic_table}}
+\end{table}
+\clearpage
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of for SSB for estimating models that estimate mean recruitment and M is fixed at the true value.}\label{naa_om_em_R_MF_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_R_MF_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of for SSB for estimating models that estimate mean recruitment and M is estimated.}\label{naa_om_em_R_ME_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_R_ME_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of for SSB for estimating models that estimates a BH stock-recruitment function and M is fixed at the true value.}\label{naa_om_em_SR_MF_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_SR_MF_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of for SSB for estimating models that estimates a BH stock-recruitment function and M is estimated.}\label{naa_om_em_SR_ME_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_SR_ME_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+### Estimating models include NAA random effects and estimation assumes mean R or BH SR
+\clearpage
+
+\begin{table}
+\caption{Operating models and estimation models all assume RE on recruitment only, estimating models assume mean recruitment or a B-H stock recruit relationship and M is fixed at the true value.}
+{\input{rec_om_em_R_SR_MF_aic_table}}
+\end{table}
+
+\begin{table}
+\caption{Operating models and estimation models all assume RE on recruitment only, estimating models assume mean recruitment or a B-H stock recruit relationship and M is estimated.}
+{\input{rec_om_em_R_SR_ME_aic_table}}
+\end{table}
+
+\begin{table}
+\caption{Operating models and estimation models all assume RE on all abundances at age, estimating models assume mean recruitment or a B-H stock recruit relationship and M is fixed at the true value.}
+{\input{naa_om_em_R_SR_MF_aic_table}}
+\end{table}
+
+\begin{table}
+\caption{Operating models and estimation models all assume RE on all abundances at age, estimating models assume mean recruitment or a B-H stock recruit relationship and M is estimated.}
+{\input{naa_om_em_R_SR_ME_aic_table}}
+\end{table}
+\clearpage
+
 # Discussion
@@
 
+<!--
 # CRediT authorship contribution statement {-}
@@
 **Timothy J. Miller:** Conceptualization, Methodology, Writing - original draft, Formal analysis, Visualization.
+ -->
 

### 2022-12-07 Project_0/paper/paper.tex
saving progress on Project_0

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={State-space age-structured assessment models provide reliable inferences(?)},
  pdfauthor={Timothy J. Miller1, Britten, Brooks, Fay, Legault, Stock,},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{url}
\usepackage{setspace}
%\singlespacing
%\onehalfspacing
\doublespacing
\usepackage{lineno}
\linenumbers
\usepackage[belowskip=0pt,aboveskip=0pt]{caption}
\usepackage{relsize}
\newcommand{\afrb}{Alaska Fishery Research Bulletin\xspace}
\newcommand{\ajms}{African Journal of Marine Science\xspace}
\newcommand{\amb}{Advances in Marine Biology\xspace}
\newcommand{\bms}{Bulletin of Marine Science\xspace}
\newcommand{\bjssf}{Bulletin of the Japanese Society of Scientific Fisheries\xspace}
\newcommand{\cb}{Conservation Biology\xspace}
\newcommand{\cjfas}{Canadian Journal of Fisheries and Aquatic Sciences\xspace}
\newcommand{\ea}{Ecological Applications\xspace}
\newcommand{\eer}{Evolutionary Ecology Research\xspace}
\newcommand{\elet}{Ecology Letters\xspace}
\newcommand{\emod}{Ecological Modelling\xspace}
\newcommand{\ebf}{Environmental Biology of Fishes\xspace}
\newcommand{\ff}{Fish and Fisheries\xspace}
\newcommand{\fo}{Fisheries Oceanography\xspace}
\newcommand{\fr}{Fisheries Research\xspace}
\newcommand{\fb}{Fishery Bulletin\xspace}
\newcommand{\ijms}{ICES Journal of Marine Science\xspace}
\newcommand{\iccat}{Collective Volume of Scientific Papers ICCAT\xspace}
\newcommand{\jae}{Journal of Animal Ecology\xspace}
\newcommand{\jai}{Journal of Applied Ichthyology\xspace}
\newcommand{\jdc}{Journal Du Conseil International Pour L'exploration De La Mer\xspace}
\newcommand{\jdcp}{Journal Du Conseil Permanent International Pour L'exploration De La Mer\xspace}
\newcommand{\jembe}{Journal of Experimental Marine Biology and Ecology\xspace}
\newcommand{\jfb}{Journal of Fish Biology\xspace}
\newcommand{\jsr}{Journal of Sea Research\xspace}
\newcommand{\jtb}{Journal of Theoretical Biology\xspace}
\newcommand{\jfrbc}{Journal of the Fisheries Research Board of Canada\xspace}
\newcommand{\jnwafs}{Journal of Northwest Atlantic Fisheries Science\xspace}
\newcommand{\mcf}{Marine and Coastal Fisheries: Dynamics, Management, and Ecosystem Science\xspace}
\newcommand{\mb}{Marine Biology\xspace}
\newcommand{\meps}{Marine Ecology Progress Series\xspace}
\newcommand{\mfr}{Marine Fisheries Review\xspace}
\newcommand{\mpb}{Marine Pollution Bulletin\xspace}
\newcommand{\najfm}{North American Journal of Fisheries Management\xspace}
\newcommand{\nzjmfr}{New Zealand Journal of Marine and Freshwater Research\xspace}
\newcommand{\pnas}{Proceedings of the National Academy of Sciences USA\xspace}
\newcommand{\rpvrciemm}{Rapports et Proc\`es-Verbaux des R\'eunions. Conseil Internationale pour l'Exploration de la Mer\xspace}
\newcommand{\rpvrcpiemm}{Rapports et Proc\`es-Verbaux des R\'eunions. Conseil Permanent Internationale pour l'Exploration de la Mer\xspace}
\newcommand{\rfbf}{Reviews in Fish Biology and Fisheries\xspace}
\newcommand{\sajms}{South African Journal of Marine Science\xspace}
\newcommand{\tafs}{Transactions of the American Fisheries Society\xspace}

\newcommand{\anzjs}{Australian \& New Zealand Journal of Statistics\xspace}
\newcommand{\as}{Applied Statistics\xspace}
\newcommand{\csda}{Computational Statistics \& Data Analysis\xspace}
\newcommand{\ees}{Environmental and Ecological Statistics\xspace}
\newcommand{\jas}{Journal of Applied Statistics\xspace}
\newcommand{\jabes}{Journal of Agricultural, Biological, and Environmental Statistics\xspace}
\newcommand{\jasa}{Journal of the American Statistical Association\xspace}
\newcommand{\jrssb}{Journal of the Royal Statistical Society. Series B\xspace}
\newcommand{\sm}{Statistics in Medicine}

\usepackage{xspace}
\usepackage{bm}
\usepackage{caption,graphics}
\usepackage{graphicx}
\usepackage{makecell}
\renewcommand\figurename{Fig.}
\captionsetup{labelsep=period, singlelinecheck=false}
\newcommand{\changesize}[1]{\fontsize{#1pt}{#1pt}\selectfont}
\renewcommand{\arraystretch}{1.5}
\renewcommand\theadfont{}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{cjfas2.bst}

\title{State-space age-structured assessment models provide reliable
inferences(?)}
\author{Timothy J. Miller\textsuperscript{1}, Britten, Brooks, Fay,
Legault, Stock,}
\date{}

\begin{document}
\maketitle

\(^1\)corresponding author:
\href{mailto:timothy.j.miller@noaa.gov}{\nolinkurl{timothy.j.miller@noaa.gov}},
Northeast Fisheries Science Center, Woods Hole Laboratory, 166 Water
Street, Woods Hole, MA 02543 USA\\

\pagebreak

\hypertarget{abstract}{%
\subsection*{Abstract}\label{abstract}}
\addcontentsline{toc}{subsection}{Abstract}

\hypertarget{keywords}{%
\subsubsection*{Keywords}\label{keywords}}
\addcontentsline{toc}{subsubsection}{Keywords}

\pagebreak

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Much is known about the reliabiltiy of state-space models that are
linear or Gaussian, but applications in fisheris management are
nonlinear and typically include multiple types of observations with
varying distributional assumptions. We know relatively little about the
statistical reliability of such models. Also, there is a wide range of
potential random effects structures in assessment models and we know
little about the ability of information criteria to distinguish among
such alternative structures.

But those studies focus primarily on Gaussain Reliability of hidden
process models

\hypertarget{methods}{%
\section{Methods}\label{methods}}

Used the WHAM package commit = XXXXXXX (Stock and Miller 2021, Miller
and Stock 2020). This packages has also been used to configure operating
and estimating models for closed loop simulations evaluating index-based
asssessment methods (Legault et al.~In press).

We completed a simulation study with 24 NAA re operating models, 16 M re
operating models, 16 Sel re operating models and 16 q re operating
models. We simulated N data sets for each operating model. For each
simulated data from each operating model we fit a set of estimating
models.

Y estimating models fit to each \#\# Operating models

common to all:\\
ages, M maturity waa observation error assumptions 2.5 Fmsy
-\textgreater{} Fmsy after 20 years Fmsy througout

marginal standard deviations for random effects are defined in tables of
operating models.

NAA\_oms process error assumtions

Population - Initial population, ages, M maturity waa

Fishing history All operating models assumed one of two different
fishing histories. One : Fishing mortality is equal to Fmsy for the
whole 40 year period. Two : Fishing mortality is 2.5 times Fmsy for the
first 20 years then changes to Fmsy for the last 20 years.

Overfishing assumptions is based on average estimates of overfishing for
NE groundfish stocks from Wiedenman et al.~(20XX). Legault et al.~(2023)
also used similar approaches to defining fishing mortality histories for
operating models.

We assumed a Beverton-Holt stock recruit function with constant
pre-recruit mortality parameters for all operating models. All
post-recruit productivity components are constant in the NAA and survey
catchability process error operating models. Therefore steepness and
unfished recruitment are also constant over the time period for those
operating models (Miller and Brooks 2021). We specified unfished
recruitment = \(e^10\) and \(F_{\text{MSY}} = F_{40}\) equated to a
steepness of 0.68 and a and b parameters X and X?

For operating models with time-varying random effects for M, steepness
is not constant, but we used the same alpha and beta parameters as other
operating models this equates to a steepness and R0 at the mean of the
time series process for M.

For operating models with time-varying random effects for fishery
selectivity, Fmsy is also not constant however we use the same Fhistory
as other operating models which corresposds to Fmsy at the mean
selectivity parameters.

Initial population starts at equilibium population structure associated
with F= x*Fmsy where x = 1 or 2.5.

surveys, selectivity

fleets, selectivity

M\_oms NOTE: inv\_trans\_rho function in set\_M.R is mis-defined. Will
affect correlation parameters assigned in operating models?

\hypertarget{estimating-models}{%
\subsection{Estimating models}\label{estimating-models}}

32 estimating models 1-20 fit to each NAA RE operating model 5-24 fit to
each M RE operating model 5-20,25-28 to each sel RE operating model
5-20, 29-32 to each q RE operating model

SR estimation or not

Make plot of S-R curve, Fmsy = F40 Initial values for BH parameters are
the true values. Initial values for mean R model = true R0.

M estimation or not

NAA\_re Random effects on Recruitment only or random effects on
recruitment and transitions among older numbers at age.

M\_re Random effects on Recruitment only, M constant across age .

sel\_re Random effects on Recruitment only, fleet logistic selectivity
RE model?

q\_re Random effects on Recruitment only, one survey catchability RE
model?

Simulations were all carried out on the University of Massachussetts
Green High-Performance Computing Cluster. Code for completing the
simulations and summarization of results can be found at
github.com/timjmiller/SSRTWG/Project\_0. We used the wham package
version 1.X.X (commit XXXXX).

\hypertarget{results}{%
\section{Results}\label{results}}

Do each of these by type of operating model (Naa, M, sel, q) Convergence
performance

AIC performance

SR estimation? M estimation?

Bias, Mean Square error

Certain basic parameters (stock-recruit pars, M, variance parameters)
SSB, F, R

\clearpage

\hypertarget{numbers-at-age-operating-models}{%
\subsection{Numbers at age operating
models}\label{numbers-at-age-operating-models}}

\hypertarget{estimating-models-include-alternative-random-effects-options-naa-m-sel-q}{%
\subsubsection{Estimating models include alternative random effects
options: NAA, M, sel,
q}\label{estimating-models-include-alternative-random-effects-options-naa-m-sel-q}}

\clearpage

\begin{table}
\caption{NAA operating models, estimating models all assume a B-H stock recruit relationship and M is fixed at the true value.}
{\input{naa_om_em_SR_MF_aic_table}}
\end{table}

\begin{table}
\caption{NAA operating models, estimating models all assume a B-H stock recruit relationship and M is estimated.}
{\input{naa_om_em_SR_ME_aic_table}}
\end{table}

\begin{table}
\caption{NAA operating models, estimating models all estimate a mean recruitment and M is fixed at the true value.}
{\input{naa_om_em_R_MF_aic_table}}
\end{table}

\begin{table}
\caption{NAA operating models, estimating models all estimate a mean recruitment and M estimated.}
{\input{naa_om_em_R_MF_aic_table}}
\end{table}
\clearpage

\begin{landscape}
\begin{figure}
\caption{Median relative bias of for SSB for estimating models that estimate mean recruitment and M is fixed at the true value.}\label{naa_om_em_R_MF_relbias_ssb}
\begin{center}
\includegraphics[width = \textwidth]{naa_om_R_MF_relbias_ssb.png}
\end{center}
\end{figure}
\end{landscape}

\begin{landscape}
\begin{figure}
\caption{Median relative bias of for SSB for estimating models that estimate mean recruitment and M is estimated.}\label{naa_om_em_R_ME_relbias_ssb}
\begin{center}
\includegraphics[width = \textwidth]{naa_om_R_ME_relbias_ssb.png}
\end{center}
\end{figure}
\end{landscape}

\begin{landscape}
\begin{figure}
\caption{Median relative bias of for SSB for estimating models that estimates a BH stock-recruitment function and M is fixed at the true value.}\label{naa_om_em_SR_MF_relbias_ssb}
\begin{center}
\includegraphics[width = \textwidth]{naa_om_SR_MF_relbias_ssb.png}
\end{center}
\end{figure}
\end{landscape}

\begin{landscape}
\begin{figure}
\caption{Median relative bias of for SSB for estimating models that estimates a BH stock-recruitment function and M is estimated.}\label{naa_om_em_SR_ME_relbias_ssb}
\begin{center}
\includegraphics[width = \textwidth]{naa_om_SR_ME_relbias_ssb.png}
\end{center}
\end{figure}
\end{landscape}

\hypertarget{estimating-models-include-naa-random-effects-and-estimation-assumes-mean-r-or-bh-sr}{%
\subsubsection{Estimating models include NAA random effects and
estimation assumes mean R or BH
SR}\label{estimating-models-include-naa-random-effects-and-estimation-assumes-mean-r-or-bh-sr}}

\clearpage

\begin{table}
\caption{Operating models and estimation models all assume RE on recruitment only, estimating models assume mean recruitment or a B-H stock recruit relationship and M is fixed at the true value.}
{\input{rec_om_em_R_SR_MF_aic_table}}
\end{table}

\begin{table}
\caption{Operating models and estimation models all assume RE on recruitment only, estimating models assume mean recruitment or a B-H stock recruit relationship and M is estimated.}
{\input{rec_om_em_R_SR_ME_aic_table}}
\end{table}

\begin{table}
\caption{Operating models and estimation models all assume RE on all abundances at age, estimating models assume mean recruitment or a B-H stock recruit relationship and M is fixed at the true value.}
{\input{naa_om_em_R_SR_MF_aic_table}}
\end{table}

\begin{table}
\caption{Operating models and estimation models all assume RE on all abundances at age, estimating models assume mean recruitment or a B-H stock recruit relationship and M is estimated.}
{\input{naa_om_em_R_SR_ME_aic_table}}
\end{table}
\clearpage

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

The estimating models assumed variances of aggregate catch and index
observations was known. This approximation may be appropriate for
indices where we have a reliable estimate of uncertainty based on the
survey design (), but there may be better approaches for the aggregate
catch such as an informed prior on the standard errors with realistic
bounds.

\hypertarget{acknowledgements}{%
\section*{Acknowledgements}\label{acknowledgements}}
\addcontentsline{toc}{section}{Acknowledgements}

This work was funded by NOAA Fisheries Northeast Fisheries Science
Center.

\pagebreak

\bibliography{sweep-paper}

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\end{CSLReferences}

\pagebreak

\clearpage

\end{document}

### 2022-12-09 Project_0/paper/paper.Rmd
saving work on Project 0 paper

@@
 
-Much is known about the reliabiltiy of state-space models that are linear or Gaussian, but applications in fisheris management are
-nonlinear and typically include multiple types of observations with varying distributional assumptions. We know relatively little about
-the statistical reliability of such models. Also, there is a wide range of potential random effects structures in assessment models
-and we  know little about the ability of information criteria to distinguish among such alternative structures.
+Application of state-space models in fisheries stock assessment and management has expanded dramatically within ICES, Canada and the Northeast US \citep{nielsenberg14,cadigan16,stockmiller21}.
 
-But those studies focus primarily on Gaussain 
+Much is known about the reliability of state-space models that are linear or Gaussian \citep{aeberhardetal18}, but applications in fisheries management are nonlinear and typically include multiple types of observations with varying distributional assumptions.
+We know relatively little about the statistical reliability of such models.
+Also, there is a wide range of potential random effects structures in assessment models and we  know little about the ability of information criteria to distinguish among such alternative structures.
+
+But those studies focus primarily on Gaussian 
 Reliability of hidden process models
@@
 
-Used the WHAM package commit = XXXXXXX (Stock and Miller 2021, Miller and Stock 2020). This packages has also been used
-to configure operating and estimating models for closed loop simulations evaluating index-based asssessment methods (Legault et al. In press).
+Used the WHAM package  \citep[][commit 77bbd94]{stockmiller21} (Miller and Stock 2020).
+This packages has also been used to configure operating and estimating models for closed loop simulations evaluating index-based assessment methods \citep{legaultetalinpress}.
 
-We completed a simulation study with 24 NAA re operating models, 16 M re operating models, 16 Sel re operating models and 16 q re operating models. 
-We simulated N data sets for each operating model.
+We completed a simulation study with a number of operating models that can be categorized based on where random effects are assumed: abundance at age, natural mortality, fleet selectivity, or index catchability.
+For each operating model assumptions about variance of process errors and observations are required and the values we used were based on a review of the range of estimates from recent applications of WHAM in management of stocks of haddock, butterfish, and American plaice in the NE US.
+
+We simulated 100 data sets for each operating model.
 For each simulated data from each operating model we fit a set of estimating models.
@@
 Y estimating models fit to each
+
 ## Operating models
@@
 common to all:   
-  ages, 
+  ages = 1 to 10+, 
   M
   maturity
-  waa
-  observation error assumptions
-  2.5 Fmsy -> Fmsy after 20 years
-  Fmsy througout
 
@@
 
-NAA_oms
-  process error assumtions
+### Population
 
-Population - 
-  Initial population, 
-  ages, 
-  M
-  maturity
-  waa
+There are 10 age classes: ages 1 to 10+.
 
-Fishing history
-All operating models assumed one of two different fishing histories. 
-One : Fishing mortality is equal to Fmsy for the whole 40 year period.
-Two : Fishing mortality is 2.5 times Fmsy for the first 20 years then changes to Fmsy for the last 20 years.
+Spawning was assumed to occur 1/4 of the way through the year.
 
-Overfishing assumptions is based on average estimates of overfishing for NE groundfish stocks from Wiedenman et al.
+Natural mortality rate was assumed 0.2 when it was constant and the mean of the time series process for operating models with M random effects.
+maturity a50 = 2.89, slope = 0.88
+
+Weight at age was generated with a LVB growth
+\[
+L_a = L_{\infty}\left(1 - e^{-k(a - t_0)}\right)
+\]
+where $t_0 = 0$, $L_\infty = 85$, and $k = 0.3$, and a L-W relationship such that 
+\[
+W_a = \theta_1 L_a^{\theta_2}
+\]
+where $\theta_1 = e^{-12.1}$ and $\theta_2 = 3.2$.
+
+We assumed a Beverton-Holt stock recruit function with constant pre-recruit mortality parameters for all operating models.
+All post-recruit productivity components are constant in the NAA and survey catchability process error operating models.
+Therefore steepness and unfished recruitment are also constant over the time period for those operating models (Miller and Brooks 2021).
+We specified unfished recruitment = $R_0 = e^{10}$ and $F_{\text{MSY}} = F_{40} = 0.348$ equated to a steepness of 0.69 and $\alpha=0.60$ and $\beta = 2.4 \times 10^{-5}$ for the
+\[
+N_{1,y} = \frac{\alpha \text{SSB}_{y-1}}{1 + \beta \text{SSB}_{y-1}} 
+\]
+Beverton-Holt parameterization.
+
+The magnitude of the overfishing assumptions is based on average estimates of overfishing for NE groundfish stocks from Wiedenmann et al.
 (20XX).
@@
 
-We assumed a Beverton-Holt stock recruit function with constant pre-recruit mortality parameters for all operating models.
-All post-recruit productivity components are constant in the NAA and survey catchability process error operating models.
-Therefore steepness and unfished recruitment are also constant over the time period for those operating models (Miller and Brooks 2021).
-We specified unfished recruitment = $e^10$ and $F_{\text{MSY}} = F_{40}$ equated to a steepness of 0.68 and a and b parameters X and X?
+Initial population was configured at the equilibrium distribution fishing at either $F = 2.5\times F_{\text{MSY}}$ or $F = F_{\text{MSY}}$ for the two alternative fishing histories.
+That is for a deterministic model, the age composition would not change over time when the fishing mortality was constant at the respective level.
 
 For operating models with time-varying random effects for M, steepness is not constant, but we used the same alpha and beta parameters as other operating models this equates to a steepness and R0 at the mean of the time series process for M.
+For operating models with time-varying random effects for fishery selectivity, Fmsy is also not constant however we use the same F history as other operating models which corresponds to Fmsy at the mean selectivity parameters.
 
-For operating models with time-varying random effects for fishery selectivity, Fmsy is also not constant however we use the same Fhistory as other operating models which corresposds to Fmsy at the mean selectivity parameters.
+### Fleets
 
-Initial population starts at equilibium population structure associated with F= x*Fmsy where x = 1 or 2.5.
+We assumed a single fleet operating year round for catch observations with logistic selectivity for the fleet with $a_{50} = 5$ and slope = 1.
+This selectivity is was used to define $F_{\text{MSY}}$ for the Beverton-Holt stock recruitment parameters above.
+We assumed a logistic-normal distribution for the age-composition observations for the fleet.
 
-surveys, selectivity
+### Fishing histories
+All operating models assumed one of two different fishing histories. 
+One : Fishing mortality is equal to Fmsy (0.348) for the whole 40 year period.
+Two : Fishing mortality is 2.5 times Fmsy for the first 20 years then changes to Fmsy for the last 20 years.
 
-fleets, selectivity
+### Indices
+Two time series of surveys are assumed and observed in numbers rather than biomass for the entire 40 year period with one occurring in the spring (0.25 way through the year) and one in the fall (0.75 way through the year).
+Actually we have it currently configured that both occur 0.5 way through the year.
+Catchability of both surveys are assumed to be 0.1.
+We assumed logistic selectivity for both indices with $a_{50} = 5$ and slope = 1.
+We assumed a logistic-normal distribution for the age-composition observations.
 
-M_oms
+### Observation Uncertainty
+
+Standard deviation for log-aggregate catch was 0.1.
+There were two levels of observation error variance for indices and age composition for both indices and fleet catch.
+A low uncertainty specification assumed standard deviation of both series of log-aggregate index observations was 0.1 and the standard deviation of the logistic-normal for age composition observations was 0.3 In the high uncertainty specification the standard deviation for log-aggregate indices was 0.4 and that for the age composition observations was 1.5.
+
+### Operating models with random effects on numbers at age
+
+24 operating models, 16 Sel re operating models and 16 q re operating models.
+Table of process error assumptions
+
+### Operating models with random effects on natural mortality
+
+16 operating models
+Table of process error assumptions
+
 NOTE: inv_trans_rho function in set_M.R is mis-defined. Will affect correlation parameters assigned in operating models?
+
+### Operating models with random effects on fleet selectivity
+
+16 operating models
+Table of process error assumptions
+
+### Operating models with random effects on index catchability
+
+16 operating models
+Table of process error assumptions
 
@@
 32 estimating models
+Table of estimating models
+
 1-20 fit to each NAA RE operating model
@@
 
-\bibliography{sweep-paper}
-<!-- \bibliography{sweep-paper}
+\bibliography{paper}
+<!-- \bibliography{project-0-paper}
 \bibliographystyle{/home/tmiller2/work/bibliographies/cjfas2}  -->

### 2022-12-09 Project_0/paper/paper.tex
saving work on Project 0 paper

@@
 
-Much is known about the reliabiltiy of state-space models that are
-linear or Gaussian, but applications in fisheris management are
-nonlinear and typically include multiple types of observations with
-varying distributional assumptions. We know relatively little about the
-statistical reliability of such models. Also, there is a wide range of
-potential random effects structures in assessment models and we know
-little about the ability of information criteria to distinguish among
-such alternative structures.
+Application of state-space models in fisheries stock assessment and
+management has expanded dramatically within ICES, Canada and the
+Northeast US \citep{nielsenberg14,cadigan16,stockmiller21}.
 
-But those studies focus primarily on Gaussain Reliability of hidden
+Much is known about the reliability of state-space models that are
+linear or Gaussian \citep{aeberhardetal18}, but applications in
+fisheries management are nonlinear and typically include multiple types
+of observations with varying distributional assumptions. We know
+relatively little about the statistical reliability of such models.
+Also, there is a wide range of potential random effects structures in
+assessment models and we know little about the ability of information
+criteria to distinguish among such alternative structures.
+
+But those studies focus primarily on Gaussian Reliability of hidden
 process models
@@
 
-Used the WHAM package commit = XXXXXXX (Stock and Miller 2021, Miller
+Used the WHAM package \citep[][commit 77bbd94]{stockmiller21} (Miller
 and Stock 2020). This packages has also been used to configure operating
 and estimating models for closed loop simulations evaluating index-based
-asssessment methods (Legault et al.~In press).
+assessment methods \citep{legaultetalinpress}.
 
-We completed a simulation study with 24 NAA re operating models, 16 M re
-operating models, 16 Sel re operating models and 16 q re operating
-models. We simulated N data sets for each operating model. For each
-simulated data from each operating model we fit a set of estimating
-models.
+We completed a simulation study with a number of operating models that
+can be categorized based on where random effects are assumed: abundance
+at age, natural mortality, fleet selectivity, or index catchability. For
+each operating model assumptions about variance of process errors and
+observations are required and the values we used were based on a review
+of the range of estimates from recent applications of WHAM in management
+of stocks of haddock, butterfish, and American plaice in the NE US.
 
-Y estimating models fit to each \#\# Operating models
+We simulated 100 data sets for each operating model. For each simulated
+data from each operating model we fit a set of estimating models.
+
+Y estimating models fit to each
+
+\hypertarget{operating-models}{%
+\subsection{Operating models}\label{operating-models}}
 
 common to all:\\
-ages, M maturity waa observation error assumptions 2.5 Fmsy
--\textgreater{} Fmsy after 20 years Fmsy througout
+ages = 1 to 10+, M maturity
 
@@
 
-NAA\_oms process error assumtions
+\hypertarget{population}{%
+\subsubsection{Population}\label{population}}
 
-Population - Initial population, ages, M maturity waa
+There are 10 age classes: ages 1 to 10+.
 
-Fishing history All operating models assumed one of two different
-fishing histories. One : Fishing mortality is equal to Fmsy for the
-whole 40 year period. Two : Fishing mortality is 2.5 times Fmsy for the
-first 20 years then changes to Fmsy for the last 20 years.
+Spawning was assumed to occur 1/4 of the way through the year.
 
-Overfishing assumptions is based on average estimates of overfishing for
-NE groundfish stocks from Wiedenman et al.~(20XX). Legault et al.~(2023)
-also used similar approaches to defining fishing mortality histories for
-operating models.
+Natural mortality rate was assumed 0.2 when it was constant and the mean
+of the time series process for operating models with M random effects.
+maturity a50 = 2.89, slope = 0.88
+
+Weight at age was generated with a LVB growth \[
+L_a = L_{\infty}\left(1 - e^{-k(a - t_0)}\right)
+\] where \(t_0 = 0\), \(L_\infty = 85\), and \(k = 0.3\), and a L-W
+relationship such that \[
+W_a = \theta_1 L_a^{\theta_2}
+\] where \(\theta_1 = e^{-12.1}\) and \(\theta_2 = 3.2\).
 
@@
 operating models (Miller and Brooks 2021). We specified unfished
-recruitment = \(e^10\) and \(F_{\text{MSY}} = F_{40}\) equated to a
-steepness of 0.68 and a and b parameters X and X?
+recruitment = \(R_0 = e^{10}\) and \(F_{\text{MSY}} = F_{40} = 0.348\)
+equated to a steepness of 0.69 and \(\alpha=0.60\) and
+\(\beta = 2.4 \times 10^{-5}\) for the \[
+N_{1,y} = \frac{\alpha \text{SSB}_{y-1}}{1 + \beta \text{SSB}_{y-1}} 
+\] Beverton-Holt parameterization.
+
+The magnitude of the overfishing assumptions is based on average
+estimates of overfishing for NE groundfish stocks from Wiedenmann et
+al.~(20XX). Legault et al.~(2023) also used similar approaches to
+defining fishing mortality histories for operating models.
+
+Initial population was configured at the equilibrium distribution
+fishing at either \(F = 2.5\times F_{\text{MSY}}\) or
+\(F = F_{\text{MSY}}\) for the two alternative fishing histories. That
+is for a deterministic model, the age composition would not change over
+time when the fishing mortality was constant at the respective level.
 
@@
 operating models this equates to a steepness and R0 at the mean of the
-time series process for M.
+time series process for M. For operating models with time-varying random
+effects for fishery selectivity, Fmsy is also not constant however we
+use the same F history as other operating models which corresponds to
+Fmsy at the mean selectivity parameters.
 
-For operating models with time-varying random effects for fishery
-selectivity, Fmsy is also not constant however we use the same Fhistory
-as other operating models which corresposds to Fmsy at the mean
-selectivity parameters.
+\hypertarget{fleets}{%
+\subsubsection{Fleets}\label{fleets}}
 
-Initial population starts at equilibium population structure associated
-with F= x*Fmsy where x = 1 or 2.5.
+We assumed a single fleet operating year round for catch observations
+with logistic selectivity for the fleet with \(a_{50} = 5\) and slope =
+1. This selectivity is was used to define \(F_{\text{MSY}}\) for the
+Beverton-Holt stock recruitment parameters above. We assumed a
+logistic-normal distribution for the age-composition observations for
+the fleet.
 
-surveys, selectivity
+\hypertarget{fishing-histories}{%
+\subsubsection{Fishing histories}\label{fishing-histories}}
 
-fleets, selectivity
+All operating models assumed one of two different fishing histories. One
+: Fishing mortality is equal to Fmsy (0.348) for the whole 40 year
+period. Two : Fishing mortality is 2.5 times Fmsy for the first 20 years
+then changes to Fmsy for the last 20 years.
 
-M\_oms NOTE: inv\_trans\_rho function in set\_M.R is mis-defined. Will
-affect correlation parameters assigned in operating models?
+\hypertarget{indices}{%
+\subsubsection{Indices}\label{indices}}
+
+Two time series of surveys are assumed and observed in numbers rather
+than biomass for the entire 40 year period with one occurring in the
+spring (0.25 way through the year) and one in the fall (0.75 way through
+the year). Actually we have it currently configured that both occur 0.5
+way through the year. Catchability of both surveys are assumed to be
+0.1. We assumed logistic selectivity for both indices with
+\(a_{50} = 5\) and slope = 1. We assumed a logistic-normal distribution
+for the age-composition observations.
+
+\hypertarget{observation-uncertainty}{%
+\subsubsection{Observation Uncertainty}\label{observation-uncertainty}}
+
+Standard deviation for log-aggregate catch was 0.1. There were two
+levels of observation error variance for indices and age composition for
+both indices and fleet catch. A low uncertainty specification assumed
+standard deviation of both series of log-aggregate index observations
+was 0.1 and the standard deviation of the logistic-normal for age
+composition observations was 0.3 In the high uncertainty specification
+the standard deviation for log-aggregate indices was 0.4 and that for
+the age composition observations was 1.5.
+
+\hypertarget{operating-models-with-random-effects-on-numbers-at-age}{%
+\subsubsection{Operating models with random effects on numbers at
+age}\label{operating-models-with-random-effects-on-numbers-at-age}}
+
+24 operating models, 16 Sel re operating models and 16 q re operating
+models. Table of process error assumptions
+
+\hypertarget{operating-models-with-random-effects-on-natural-mortality}{%
+\subsubsection{Operating models with random effects on natural
+mortality}\label{operating-models-with-random-effects-on-natural-mortality}}
+
+16 operating models Table of process error assumptions
+
+NOTE: inv\_trans\_rho function in set\_M.R is mis-defined. Will affect
+correlation parameters assigned in operating models?
+
+\hypertarget{operating-models-with-random-effects-on-fleet-selectivity}{%
+\subsubsection{Operating models with random effects on fleet
+selectivity}\label{operating-models-with-random-effects-on-fleet-selectivity}}
+
+16 operating models Table of process error assumptions
+
+\hypertarget{operating-models-with-random-effects-on-index-catchability}{%
+\subsubsection{Operating models with random effects on index
+catchability}\label{operating-models-with-random-effects-on-index-catchability}}
+
+16 operating models Table of process error assumptions
 
@@
 
-32 estimating models 1-20 fit to each NAA RE operating model 5-24 fit to
-each M RE operating model 5-20,25-28 to each sel RE operating model
-5-20, 29-32 to each q RE operating model
+32 estimating models Table of estimating models
+
+1-20 fit to each NAA RE operating model 5-24 fit to each M RE operating
+model 5-20,25-28 to each sel RE operating model 5-20, 29-32 to each q RE
+operating model
 
@@
 
-\bibliography{sweep-paper}
+\bibliography{paper}
 

### 2022-12-16 Ecov_study/catchability/Catchability_ecov_sims.Rmd
Update results overview figure, add code to postprocess multiple simulation results files.

@@
         for(isim in 1:nsim){ # Loop over simulations
+          print(paste0("Simulation_", isim))
                 for(iEM in 1:length(inputEMlist)){ # Loop over estimation models
@@
                         # Store model-specific results for given isim here (append EM results to storage list for that EM)
-                        storeEM[which(namesEM == namesEM[iEM])][[1]] <- append(storeEM[which(namesEM == namesEM[iEM])], list(resultEM))
+                        storeEM[which(namesEM == namesEM[iEM])][[1]] <- append(storeEM[which(namesEM == namesEM[iEM])][[1]], list(resultEM))
 
@@
 OM[label = 'OM \n results$OM']
-sim1[label = 'Simulation 1 results \n results$OM[[1]]']
-sim2[label = 'Simulation 1 results \n results$OM[[2]]']
-sim3[label = 'Simulation 1 results \n results$OM[[3]]']
+sim1[label = 'Simulation 1 results \n results$OM[1][[1]]']
+sim2[label = 'Simulation 2 results \n results$OM[2][[1]]']
+sim3[label = 'Simulation 3 results \n results$OM[3][[1]]']
 
 EM1[label = 'EM1 \n results$EM1']
-sim1_1[label = 'Simulation 1 results \n results$EM1[[1]]']
-sim1_2[label = 'Simulation 1 results \n results$EM1[[1]]']
-sim1_3[label = 'Simulation 1 results \n results$EM1[[1]]']
+sim1_1[label = 'Simulation 1 results \n results[EMs[1]][[1]][1][[1]]']
+sim1_2[label = 'Simulation 2 results \n results[EMs[1]][[1]][2][[1]]']
+sim1_3[label = 'Simulation 3 results \n results[EMs[1]][[1]][2][[1]]']
 EM2[label = 'EM2 \n results$EM2']
-sim2_1[label = 'Simulation 1 results \n results$EM2[[1]]']
-sim2_2[label = 'Simulation 1 results \n results$EM2[[1]]']
-sim2_3[label = 'Simulation 1 results \n results$EM2[[1]]']
+sim2_1[label = 'Simulation 1 results \n results[EMs[2]][[1]][1][[1]]']
+sim2_2[label = 'Simulation 2 results \n results[EMs[2]][[1]][2][[1]]']
+sim2_3[label = 'Simulation 3 results \n results[EMs[2]][[1]][3][[1]]']
 ageslab[label = 'ages.lab']
@@
 ```{r}
-#' @param filenames A vector of strings (including .Rda extensions) indicating what Rdata results files to read in and post process
+#' @param filenames A vector of strings (including .RData extensions) indicating what results files to read in and post process
 #' 
-#' @return
+#' @return A dataframe containing the following columns describing model performance
+#' \itemize{
+#'   \item{Year - Year associated with performance metric}
+#'   \item{sim - Simulation number (based on total number of simulations in all files processed)}
+#'   \item{EM - Name of estimation model used to calculate performance metric}
+#'   \item{converged - Boolean indicating whether model converged with invertible hessian ("Yes") or not ("No")}
+#'   \item{relSSB - Ratio of SSB from EM:OM}
+#'   \item{relF - Ratio of F from EM:OM}
+#'   \item{relR - Ratio of R from EM:OM}
+#'   \item{relEcov_x - Ratio of Ecov_x from EM:OM}
+#' }
 
@@
   # Set up storage for processed results
+  perfMet <- NULL
 
-        # Loop over files
+  # Establish total number of simulations across all files (based on nsim in filename)
+  nsim <- rep(NA, length(filenames))
   for(ifile in 1:length(filenames)){
-                # Read in Rdata object
-                fileDat <- readRDS(filenames[ifile])
+    fileSpecs <- strsplit(filenames[ifile], "_")
+    nsim[ifile] <- fileSpecs[[1]][which(grepl("simWHAM", fileSpecs[[1]]))+1] %>% as.numeric()
+  }
 
-                nsim <- # Based on length of results in given filename
+  nsim <- c(0, nsim) # Add 0 so isim indexing in file processed output goes from 1:total nsim
+
+  # Loop over files to calculate performance metrics
+  for(ifile in 1:length(filenames)){
+    # Read in RData object
+    load(file=filenames[ifile]) # loads object named "results"
 
     # Loop over simulations in each file
-                for(isim in 1:nsim){
-                        # Calculate performance metrics here & add to storage
+    for(isim in 1:nsim[ifile+1]){ # start at ifile+1 since nsim[ifile] = 0 for sim numbering reasons
+
+      # Pull out EM names 
+      EMs <- names(results)[2:length(names(results))]
+
+      # Loop over EM in each ifile
+      for(iEM in 1:length(EMs)){
+
+        # Store Years, isim, iEM, convergence (mod$opt$convergence == 0 and hessian invertible $na_sdrep == FALSE)
+        Year <- results[EMs[iEM]][[1]][isim][[1]]$years
+        sim <- rep((isim+nsim[ifile]), length(Year)) # + 0 for first file, starts sim numbering at nsim[ifile=1]+1 for second file
+        EM <- rep(EMs[iEM], length(Year))
+        if(check_convergence(results[EMs[iEM]][[1]][isim][[1]], ret=TRUE)$na_sdrep == TRUE){
+          converged <- rep("No", length(Year))
+        } else{
+          converged <- rep("Yes", length(Year))
         }
 
+        # Proportion of EM:OM SSB
+          # EM SSB calculated based on plotting script https://github.com/timjmiller/wham/blob/6c30301d5b68c2bb0c0330cc4f112432c7950dfd/R/wham_plots_tables.R#L725-L738
+        tempSSB_EM <- results[EMs[iEM]][[1]][isim][[1]]$sdrep %>% summary() #EM SSB must be calculated, this line triggers an NaNs produced warning for EMs that didn't converge with invertible hessians
+        tempSSB_EM <- tempSSB_EM[rownames(tempSSB_EM) == "log_SSB",]
+        tempSSB_EM <- exp(cbind(tempSSB_EM, tempSSB_EM[,1]+qnorm(0.975)*cbind(-tempSSB_EM[,2], tempSSB_EM[,2])))#/1000 Don't divide by 1000 so units in mt rather than kmt
+        tempSSB_EM <- tempSSB_EM[,1] 
+          # OM SSB
+        tempSSB_OM <- results$OM[isim][[1]]$SSB # Pull SSB directly from simulated data
+          # relative SSB
+        relSSB <- tempSSB_EM/tempSSB_OM
+
+        # Proportion of EM:OM F
+          # EM calculated F based on plotting scripts
+        mod <- results[EMs[iEM]][[1]][isim][[1]]
+        std = results[EMs[iEM]][[1]][isim][[1]]$sdrep %>% summary()
+        years_full <- mod$years_full
+        n_ages = mod$env$data$n_ages
+        faa.ind <- which(rownames(std) == "log_FAA_tot")
+        log.faa <- matrix(std[faa.ind,1], length(years_full), n_ages)
+        faa.cv <- matrix(std[faa.ind,2], length(years_full), n_ages)
+        age.full.f <- apply(log.faa,1, function(x) max(which(x == max(x))))
+        full.f.ind = cbind(1:length(years_full), age.full.f)
+        log.full.f <- log.faa[full.f.ind]
+        full.f.cv <- faa.cv[full.f.ind]
+        full.f <- exp(log.full.f)
+          # OM F
+        tempF_OM <- c(results$OM[isim][[1]]$F)
+          # relative F
+        relF <- full.f/tempF_OM
+
+        # Proportion of EM:OM Recruitment 
+          # EM R calculated based on plotting script: https://github.com/timjmiller/wham/blob/6c30301d5b68c2bb0c0330cc4f112432c7950dfd/R/wham_plots_tables.R#L765-L780
+        tempR_EM <-  results[EMs[iEM]][[1]][isim][[1]]$sdrep %>% summary()
+        ind <- rownames(tempR_EM) == "log_NAA_rep"
+        tempR_EM <-  exp(array(tempR_EM[ind,1], dim = c(length(results[EMs[iEM]][[1]][isim][[1]]$years), 8)))
+        tempR_EM <- tempR_EM[,1]
+          # OM recruitment
+        tempR_OM <- results$OM[isim][[1]]$NAA[,1] 
+          # relative R
+        relR <- tempR_EM/tempR_OM
+
+        # Relative Ecov_x values (EM:OM ratio)
+        if("TRUE" %in% (results[EMs[iEM]][[1]][isim][[1]]$input$data$Ecov_where > 0)){ # If link to ecov specified in EM calculate EM:OM ratio for Ecov_x values
+            # EM
+        tempEcovx_EM <- results[EMs[iEM]][[1]][isim][[1]]$sdrep %>% summary() 
+        tempEcovx_EM <- tempEcovx_EM[rownames(tempEcovx_EM)=="Ecov_x",]
+          # OM
+        tempEcovx_OM <- results$OM[isim][[1]]$Ecov_x
+          # Relative Ecov_x
+        relEcov_x <- tempEcovx_EM[,"Estimate"]/tempEcovx_OM
+        } else{ # If no ecov link specified then save an NA for that model in all years 
+          relEcov_x <- rep(NA, length(Year))
+        }
+
+        # # Relative Ecov_beta (EM:OM ratio) for all four indices
+        #   # EM
+        # tempEcovbeta_EM <- results[EMs[iEM]][[1]][isim][[1]]$sdrep %>% summary()
+        # tempEcovbeta_EM <- tempEcovbeta_EM[rownames(tempEcovbeta_EM)=="Ecov_beta"]
+        #   # OM
+        # tempEcovbeta_OM <- results$OM[isim][[1]]$Ecov_beta # by age rather than by index?
+        #
+
+        # Store results
+          # EM perfMets
+          EMperfMets <- cbind(Year, sim, EM, converged, relSSB, relF, relR, relEcov_x)
+
+          # Append perfMets to storage
+          perfMet <- rbind(perfMet, EMperfMets)
+
+      } # End loop over EM
+
+    } # End loop over simulations in ifile
+
   } # End loop over filenames
+
+  # Clean table rownames and ensure that numeric results are not treated as strings
+  rownames(perfMet) <- NULL
+  colnames(perfMet) <- c("Year", "sim", "EM", "converged", "relSSB", "relF", "relR", "relEcov_x")
+  perfMet <- perfMet %>% as.data.frame() %>% 
+    dplyr::summarise(Year = as.numeric(Year),
+                     sim = as.numeric(sim),
+                     EM = EM,
+                     converged = converged,
+                     relSSB = as.numeric(relSSB),
+                     relF = as.numeric(relF),
+                     relR = as.numeric(relR),
+                     relEcov_x = as.numeric(relEcov_x))
+
+  return(perfMet)
 }

### 2023-01-03 Project_0/paper/paper.Rmd

@@
 ---
-title: State-space age-structured assessment models provide reliable inferences(?)
+title: How reliable are inferences from state-space age-structured assessment models?
 author: 
@@
 
-But those studies focus primarily on Gaussian 
-Reliability of hidden process models
+But those studies focus primarily on Gaussian linear models.
+review literature on reliability of hidden/latent process models. Primarily in other fields.
 

### 2023-01-06 Project_0/paper/paper.tex
more SR AIC analysis

@@
 \hypersetup{
-  pdftitle={State-space age-structured assessment models provide reliable inferences(?)},
+  pdftitle={How reliable are inferences from state-space age-structured assessment models?},
   pdfauthor={Timothy J. Miller1, Britten, Brooks, Fay, Legault, Stock,},
@@
 
-\title{State-space age-structured assessment models provide reliable
-inferences(?)}
+\title{How reliable are inferences from state-space age-structured
+assessment models?}
 \author{Timothy J. Miller\textsuperscript{1}, Britten, Brooks, Fay,
@@
 
-But those studies focus primarily on Gaussian Reliability of hidden
-process models
+But those studies focus primarily on Gaussian linear models. review
+literature on reliability of hidden/latent process models. Primarily in
+other fields.
 
@@
 
+\begin{figure}
+\caption{Predicted probability of AIC being better for BH model.}\label{glm_res_NAA_re}
+\begin{center}
+\includegraphics[width = \textwidth]{pred_SR_best_NAA_oms.png}
+\end{center}
+\end{figure}
+
+\hypertarget{m-operating-models}{%
+\subsection{M operating models}\label{m-operating-models}}
+
+\hypertarget{estimating-models-include-naa-random-effects-and-estimation-assumes-mean-r-or-bh-sr-1}{%
+\subsubsection{Estimating models include NAA random effects and
+estimation assumes mean R or BH
+SR}\label{estimating-models-include-naa-random-effects-and-estimation-assumes-mean-r-or-bh-sr-1}}
+
+\begin{table}
+\caption{M random effects operating models.}
+{\input{M_om_em_R_BH_aic_table}}
+\end{table}
+\clearpage
+
 \hypertarget{discussion}{%

### 2023-01-06 Project_0/code/stock_recruit_AIC_analysis.R

@@ -1,0 +1,156 @@
+library("here")
+df.ems = readRDS(file.path(here(),"Project_0","inputs", "df.ems.RDS"))
+
+#first NAA operating models
+df.oms = readRDS(file.path(here(),"Project_0","inputs", "df.oms.RDS"))
+use.df.ems <- df.ems[1:20,]
+
+all_naa_sd_log_SSB = lapply(1:NROW(df.oms), function(y){
+  res = sapply(1:100, function(x){
+    print(paste0("sim_",x,".RDS"))
+    sim = readRDS(file.path(here::here(),"Project_0", "results", "naa_om", paste0("om_", y), paste0("sim_",x,".RDS")))
+    res = sapply(sim, function(x) {
+      out <- NA
+      if(length(x$truth)) out <- sd(log(x$truth$SSB))
+      return(out)
+    })
+    return(res[which(!is.na(res))][1])
+  })
+  return(res)
+})
+#om_ind <- which(!is.na(df.oms$NAA_sig))
+om_ind <- 1:24
+#SR_rec_Mest = which(use.df.ems$re_config == "rec+1" & use.df.ems$M_est == TRUE)
+SR_rec_Mest = which(use.df.ems$re_config %in% c("rec", "rec+1") & use.df.ems$M_est == TRUE)
+all_naa_aic <- readRDS(file = file.path(here(),"Project_0","results", "all_naa_aic_results.RDS"))
+df.oms$NAA_sig[which(is.na(df.oms$NAA_sig))] <- 0
+#make a data frame with columns defining operating model characteristics and SD of true (log) SSB 
+# and 1/0 for which SR assumption had lowest AIC
+make_df <- function(all_aic, est_ind = SR_rec_Mest, om_ind = NULL){
+  if(!is.null(om_ind)) all_aic <- all_aic[om_ind]
+  df_out <- cbind.data.frame(OM = numeric(),
+    R_sig = numeric(), NAA_sig = numeric(), Fhist = character(), obs_error = character(), sd_log_SSB = numeric(),
+    est_SR = numeric(), est_R = numeric())
+  for(i in 1:length(all_aic)){
+    #est_ind has No SR assumption first
+    if(df.oms$NAA_sig[om_ind[i]] == 0) {
[…]

### 2023-02-03 Project_0/paper/paper.Rmd
working on Ecov-M simulation code

@@
 
+Currenlty, initial population is configured at the equilibrium distribution fishing at $F = 2.5\times F_{\text{MSY}}$.
+
 Initial population was configured at the equilibrium distribution fishing at either $F = 2.5\times F_{\text{MSY}}$ or $F = F_{\text{MSY}}$ for the two alternative fishing histories.
@@
 
+Steepness and BRPs are not constant when M is time-varying (Miller and Brooks 2021). We uses the a/b parameters for the B-H defined for Fmsy = F40 at the mean M = 0.2 as defined above.
+
 ### Operating models with random effects on fleet selectivity
@@
 Table of process error assumptions
+
+BRPs are not constant when fleet selectivity is time-varying.
+We uses the a/b parameters for the B-H defined for Fmsy = F40 at the mean of the time series model for selectivity which is the same as the constant selectivity defined above.
 
@@
 {\input{naa_om_em_R_SR_ME_aic_table}}
+\end{table}
+\clearpage
+
+\begin{figure}
+\caption{Predicted probability of AIC being better for BH model.}\label{glm_res_NAA_re}
+\begin{center}
+\includegraphics[width = \textwidth]{pred_SR_best_NAA_oms.png}
+\end{center}
+\end{figure}
+
+## M operating models
+
+### Estimating models include NAA random effects and estimation assumes mean R or BH SR
+
+\begin{table}
+\caption{M random effects operating models.}
+{\input{M_om_em_R_BH_aic_table}}
 \end{table}

### 2023-02-03 Ecov_study/mortality/code/M_Ecov_om_hpcc_args.sh

@@ -1,0 +1,26 @@
+#!/bin/bash
+# Load gcc and R and jpeg header (for Hmisc)
+module load gcc/8.1.0
+module load R/4.0.4_gcc
+module load libjpeg-turbo/2.0.2
+module load openmpi/4.0.1
+
+echo "CPU threads: $(grep -c processor /proc/cpuinfo)"
+grep 'cpu cores' /proc/cpuinfo | uniq
+
+cd ~/SSRTWG
+for ((sim= $1; sim <= $2; sim++))
+do
+  for ((om= $3; om <= $4; om++))
+  do
+   for ((em= $5; em <= $6; em++))
+   do
+     echo $sim
+     echo $om
+     echo $em
+     Rscript --vanilla ~/SSRTWG/Ecov_study/mortality/code/M_Ecov_om_hpcc_script.R $sim $om $em &
+   done
+ done
+done
+
+echo "script is done"

### 2023-02-03 Ecov_study/mortality/code/M_Ecov_om_hpcc_script.R

@@ -1,0 +1,19 @@
+#!/usr/bin/env Rscript
+args = commandArgs(trailingOnly=TRUE)
+sim = as.integer(args[1])
+om = as.integer(args[2])
+em = as.integer(args[3])
+.libPaths("~/Rlib/")
+library(wham)
+source(file.path(here::here(), "Ecov_study", "mortality", "code", "sim_management.R"))
+verify_version()
+#cat(paste0("number of cores available: ", parallel::detectCores(), "/n"))
+#stop()
+run_hpcc_jobs_rev(this_sim = sim, this_om = om, this_em = em,
+  script.full.path = file.path(here::here(),"Ecov_study", "mortality", "code", "M_Ecov_om_sim_fit_script_hpcc.R"), 
+  df.ems = readRDS(file.path(here::here(),"Ecov_study", "mortality", "inputs", "df.ems.RDS")), 
+  df.oms = readRDS(file.path(here::here(),"Ecov_study", "mortality", "inputs", "df.oms.RDS")), 
+  om_inputs = readRDS(file.path(here::here(),"Ecov_study", "mortality", "inputs", "om_inputs.RDS")),
+  em_inputs = readRDS(file.path(here::here(),"Ecov_study", "mortality", "inputs", "em_inputs.RDS")),
+  write.dir = file.path(here::here(),"Project_0", "results", "q_om")
+)

### 2023-02-03 Ecov_study/mortality/code/M_Ecov_om_setup.R

@@ -1,0 +1,197 @@
+# devtools::install_github("timjmiller/wham", dependencies=TRUE, ref="devel")
+if(file.exists("c:/Users/timothy.j.miller")) {
+  library(wham, lib.loc = "c:/work/wham/old_packages/77bbd94")
+} else library(wham) #make sure to use the right version of wham
+library(tidyr)
+library(dplyr)
+library(here)
+source(file.path(here(), "common_code", "make_basic_info.R"))
+source(file.path(here(), "common_code", "set_NAA.R"))
+source(file.path(here(), "common_code", "set_M.R"))
+source(file.path(here(), "common_code", "set_q.R"))
+source(file.path(here(), "common_code", "set_ecov.R"))
+source(file.path(here(), "common_code", "set_selectivity.R"))
+source(file.path(here(), "common_code", "set_simulation_options.R"))
+source(file.path(here(), "common_code", "get_FMSY.R"))
+source(file.path(here(), "Ecov_study", "mortality", "code", "make_om.R"))
+source(file.path(here(), "Project_0", "code", "sim_management.R"))
+verify_version()
+
+write.dir <- file.path(here(),"Ecov_study", "mortality", "inputs")
+
+naa_om_inputs = readRDS(file.path(here::here(),"Project_0","inputs", "NAA_om_inputs.RDS"))
+#SR parameters are the same for all naa_om models 
+temp = fit_wham(naa_om_inputs[[1]], do.fit = FALSE, MakeADFun.silent = TRUE)
+SRab = exp(c(temp$rep$log_SR_a[1], temp$rep$log_SR_b[1]))
+
+#if(!exists("write.dir")) write.dir = getwd()
+if(!dir.exists(write.dir)) dir.create(write.dir, recursive = T)
+#setwd(write.dir)
+
+
+#number of simulations for each scenario
+#nsim = 1000
+#nsim = 100
+
+#Operating model factors
+#NAA sigmas for each scenario
+R_sig <- c(0.5)
+NAA_sig <- c(0.3)
+#F time series
+
+M_sig <- 0.3
+M_cor <- 0
+
+Ecov_obs_sig <- c(0.1, 0.5)
+Ecov_re_sig <- c(0.1,0.5)
+Ecov_re_cor <- c(0, 0.5)
[…]

### 2023-02-10 Project_0/paper/paper.Rmd
working on Ecov-M sims

@@
 ---
-title: How reliable are inferences from state-space age-structured assessment models?
+title: Factors affecting reliablity of state-space age-structured assessment models
 author: 
-  - Timothy J. Miller^1^, Britten, Brooks, Fay, Legault, Stock, 
+  - Timothy J. Miller^1^, Britten, Brooks, Fay, Legault, Stock, et al.
 date: ''
@@
 
+Here we conduct a simulation study with operating models varying by degree of observation error uncertainty, sources of process error (M, NAA, q, sel), and fishing history.
+The simulations from these operating models are fitted with estimating models that make alternative assumptions for sources of process error (M, NAA< q, sel), whether a stock-recruit model was estimated, and whether M is estimated.
+We evaluate whether AIC can correctly determine the correct source of process error and stock effects on recruitment.
+We also evaluate the degree of bias in the outputs of the assessment model that are important for management.
+
 # Methods
@@
 Used the WHAM package  \citep[][commit 77bbd94]{stockmiller21} (Miller and Stock 2020).
-This packages has also been used to configure operating and estimating models for closed loop simulations evaluating index-based assessment methods \citep{legaultetalinpress}.
+This packages has also been used to configure operating and estimating models for closed loop simulations evaluating index-based assessment methods \citep{legaultetalinpress} and is used for management of haddock, butterfish, plaice, bluefish in the Northeast US.
 

### 2023-02-10 Ecov_study/mortality/code/M_Ecov_om_sim_fit_script_hpcc.R

@@ -37,6 +37,8 @@ truth$wham_version = om$wham_version
 EM_input <- em_inputs[[this_em]] # Read in the EM 
 #put simulated data into the em input
 EM_input$data[obs_names] = sim_data[obs_names]
+#not estimating observation error in Ecov
+EM_input$par$Ecov_obs_logsigma <- om_inputs[[this_om]]$par$Ecov_obs_logsigma
 res <- list(truth = truth)
 res$fit <- list()
 #do fit withouth sdreport first

### 2023-02-27 Ecov_study/catchability/Setup_OMs.Rmd
Outline code to set up operating models.

---
title: "Setup_OMs"
output: html_document
date: "2023-02-14"
editor_options: 
  markdown: 
    wrap: 72
---

Generic OM & EM settings that are the same as project 0:

-   Cod-like life history

-   B-H stock-recruit (for now)

-   1 fleet

-   2 indices (spring and fall)

-   Logistic selectivity for the fleet and all indices

-   Initial M estimate = 0.2

-   Logistic normal age composition for fleet and all indices

Project-specific OM settings that are not variable:

-   Ecov ar1 process on catchability for both surveys (spring & fall)
    with mean = 0

-

OM settings that are variable:

-   Ecov standard deviation = 0.1 or 0.5

-   Ecov process correlation = 0 or 0.5

-   Ecov process observation standard deviation = 0.1 or 0.5

-   Ecov effect on q = 0, 0.25, or 0.5 !!! I think this is the beta
    parameter but should confirm

-   Fishing history = F_MSY for 40 year history OR fish 2.5\*F_MSY for
    20 years than fish at F_MSY for remaining 20

-

EM settings that are variable:

-   Ecov season correctly specified = BOTH (spring & fall correct), ONE
    (only one season correct), NONE (effect on q not included when it
    should be)

-   Ecov link/random effect setting = NoEnv, NoEnvRand, WithEnvRand,
    WithEnv

-

```{r}
##### OM
# Ecov process & effect magnitude set up
Ecov_process_sig <- c(0.1, 0.5)
Ecov_process_cor <- c(0, 0.5)
Ecov_process_obs_sig <- c(0.1, 0.5)
Ecov_effect <- c(0, 0.25, 0.5) # beta

# Fishing history
F_hist <- c("H-MSY", "MSY") # High-then MSY vs. MSY for entire history

# Observation error
ageComp_sig <- c(0.3, 1.5) # Fleet and index age comp treated with same error
logIndex_sig <- c(0.1, 0.4)

# Factorial combinations
OMsetup <- expand.grid(Eco_process_sig = Ecov_process_sig, Ecov_process_cor = Ecov_process_cor, Ecov_process_obs_sig = Ecov_process_obs_sig, Ecov_effect = Ecov_effect, F_hist = F_hist, ageComp_sig = ageComp_sig, logIndex_sig = logIndex_sig)
OMsetup

# Catchability set up
q_Ecov <- c("includeEcov", "excludeEcov") # May want to include these as a sensitivity run to compare what happens if OM has no ecov effect but EM assumes there is one (ecov process setup won't matter since fitting to data without effect specified in model)

##### OM sensitivity runs
# Factorial combinations when on ecov effect in OM
expand.grid(q_Ecov = q_Ecov, F_hist = F_hist, ageComp_sig = ageComp_sig, logIndex_sig = logIndex_sig) # probably do a subset, we have to decide what subset

##### EM
# EM factorial settings
miss_season <- c("BOTH", "ONE", "NONE") # Number of seasons with correctly specified effect, assumes 2 survey indices (spring & fall)
miss_q <- c("NoEnv", "NoEnvRand", "WithEnvRand", "WithEnv") # Catchability setup for EM
expand.grid(miss_season = miss_season, miss_q = miss_q)
```

-   maybe don't vary obs/process NAA since we are testing ability to
    recover catchability effect, could do sensitivity run with survival
    random effects and pick one of these settings

# set up generic OM

```{r}
library(wham)
# Get generic groundfish life history info
source(file.path(here::here(),"common_code", "make_basic_info.R"))
groundfish_info <- make_basic_info()

# Change timing for spring (0.25) and fall (0.75) indices
groundfish_info$fracyr_indices[,1] <- 0.25
groundfish_info$fracyr_indices[,2] <- 0.75

# Assume logistic selectivity for the fleet and all indices
sel_list <- list(model = c(rep("logistic", groundfish_info$n_fleets),rep("logistic", groundfish_info$n_indices)),
                   initial_pars = lapply(1:(groundfish_info$n_fleets + groundfish_info$n_indices), function(x) c(5,1)),
                 n_selblocks = (groundfish_info$n_fleets + groundfish_info$n_indices))

# Starting mean M estimate at 0.2, estimate constant M for all ages/years
M <-  list(initial_means = rep(0.2, length(groundfish_info$ages)))

genericInput <- prepare_wham_input(basic_info = groundfish_info, 
                                   selectivity = sel_list, 
                                   M = M)

# genericFit <- fit_wham(genericInput, MakeADFun.silent = TRUE, do.retro = FALSE, do.osa = FALSE)

```

# Use common_code set_ecov function to adjust based on OM settings above

This chunk of code uses common code functions & internal WHAM functions
to manipulate an OM input object to initialize the OM rather than doing
all setup when creating the input as is typically the procedure when
fitting operational WHAM models.

```{r}
source(file.path(here::here(),"common_code", "set_ecov.R"))

# Loop over OMsetup to initialize OMs
for(iom in 1:nrow(OMsetup)){
  # Pull generic input settings
  input <- genericInput

  # Ecov setup
    Ecov <- list(label = "Ecov",
                 lag = 0, 
                 where = "q", # Where/how/indices settings need to change if we do sensitivity runs
                 how = 1,
                 indices = c(1,2),
                 process_model = "ar1", 
                 process_mean_vals = rep(0,length(groundfish_info$years)), # End generic inputs for ecov
                 process_sig_vals = rep(OMsetup$Eco_process_sig[iom]),
                 process_cor_vals = rep(OMsetup$Ecov_process_cor[iom]),
                 beta_vals = list(rep(list(matrix(rep(OMsetup$Ecov_effect[iom], length(groundfish_info$ages)), nrow = 1)), 4)), #!!! Why 4 rows?
                 logsigma = as.matrix(rep(OMsetup$Ecov_process_obs_sig[iom], length(groundfish_info$years))))

  input <- set_ecov(input = input, ecov = Ecov)

  # 
  # Fishing history
F_hist <- c("H-MSY", "MSY") # High-then MSY vs. MSY for entire history

# Observation error
ageComp_sig <- c(0.3, 1.5) # Fleet and index age comp treated with same error
logIndex_sig <- c(0.1, 0.4)

}

```

### 2023-04-25 Project_0/paper/M_om_tab.tex
saving work, fixing cor par setting in set_M.R

\begin{center}
\begin{tabular}{rrrrrr}
\hline\hline
\multicolumn{1}{c}{Model}&\multicolumn{1}{c}{$\sigma_R$}&\multicolumn{1}{c}{$\rho_{M}$}&\multicolumn{1}{c}{Fishing History}&\multicolumn{1}{c}{Observation Uncertainty}&\multicolumn{1}{c}{obs_error}\tabularnewline
\hline
M_1&$0.5$&$0.1$&$0.0$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index sigma (log scale) = 0.1, Age composition sigma (logistic normal) = 0.3\tabularnewline
M_2&$0.5$&$0.5$&$0.0$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index sigma (log scale) = 0.1, Age composition sigma (logistic normal) = 0.3\tabularnewline
M_3&$0.5$&$0.1$&$0.9$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index sigma (log scale) = 0.1, Age composition sigma (logistic normal) = 0.3\tabularnewline
M_4&$0.5$&$0.5$&$0.9$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index sigma (log scale) = 0.1, Age composition sigma (logistic normal) = 0.3\tabularnewline
M_5&$0.5$&$0.1$&$0.0$&F_{\text{MSY}}&Index sigma (log scale) = 0.1, Age composition sigma (logistic normal) = 0.3\tabularnewline
M_6&$0.5$&$0.5$&$0.0$&F_{\text{MSY}}&Index sigma (log scale) = 0.1, Age composition sigma (logistic normal) = 0.3\tabularnewline
M_7&$0.5$&$0.1$&$0.9$&F_{\text{MSY}}&Index sigma (log scale) = 0.1, Age composition sigma (logistic normal) = 0.3\tabularnewline
M_8&$0.5$&$0.5$&$0.9$&F_{\text{MSY}}&Index sigma (log scale) = 0.1, Age composition sigma (logistic normal) = 0.3\tabularnewline
M_9&$0.5$&$0.1$&$0.0$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index sigma (log scale) = 0.4, Age composition sigma (logistic normal) = 1.5\tabularnewline
M_10&$0.5$&$0.5$&$0.0$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index sigma (log scale) = 0.4, Age composition sigma (logistic normal) = 1.5\tabularnewline
M_11&$0.5$&$0.1$&$0.9$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index sigma (log scale) = 0.4, Age composition sigma (logistic normal) = 1.5\tabularnewline
M_12&$0.5$&$0.5$&$0.9$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index sigma (log scale) = 0.4, Age composition sigma (logistic normal) = 1.5\tabularnewline
M_13&$0.5$&$0.1$&$0.0$&F_{\text{MSY}}&Index sigma (log scale) = 0.4, Age composition sigma (logistic normal) = 1.5\tabularnewline
M_14&$0.5$&$0.5$&$0.0$&F_{\text{MSY}}&Index sigma (log scale) = 0.4, Age composition sigma (logistic normal) = 1.5\tabularnewline
M_15&$0.5$&$0.1$&$0.9$&F_{\text{MSY}}&Index sigma (log scale) = 0.4, Age composition sigma (logistic normal) = 1.5\tabularnewline
M_16&$0.5$&$0.5$&$0.9$&F_{\text{MSY}}&Index sigma (log scale) = 0.4, Age composition sigma (logistic normal) = 1.5\tabularnewline
\hline
\end{tabular}\end{center}

### 2023-04-25 Project_0/paper/naa_om_tab.tex
saving work, fixing cor par setting in set_M.R

\begin{center}
\begin{tabular}{rrrrr}
\hline\hline
\multicolumn{1}{c}{Model}&\multicolumn{1}{c}{$\sigma_R$}&\multicolumn{1}{c}{$\sigma_{2+}$}&\multicolumn{1}{c}{Fishing History}&\multicolumn{1}{c}{Observation Uncertainty}\tabularnewline
\hline
NAA_1&$0.5$&$$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index sigma (log scale) = 0.1, Age composition sigma (logistic normal) = 0.3\tabularnewline
NAA_2&$1.5$&$$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index sigma (log scale) = 0.1, Age composition sigma (logistic normal) = 0.3\tabularnewline
NAA_3&$0.5$&$0.25$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index sigma (log scale) = 0.1, Age composition sigma (logistic normal) = 0.3\tabularnewline
NAA_4&$1.5$&$0.25$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index sigma (log scale) = 0.1, Age composition sigma (logistic normal) = 0.3\tabularnewline
NAA_5&$0.5$&$0.50$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index sigma (log scale) = 0.1, Age composition sigma (logistic normal) = 0.3\tabularnewline
NAA_6&$1.5$&$0.50$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index sigma (log scale) = 0.1, Age composition sigma (logistic normal) = 0.3\tabularnewline
NAA_7&$0.5$&$$&F_{\text{MSY}}&Index sigma (log scale) = 0.1, Age composition sigma (logistic normal) = 0.3\tabularnewline
NAA_8&$1.5$&$$&F_{\text{MSY}}&Index sigma (log scale) = 0.1, Age composition sigma (logistic normal) = 0.3\tabularnewline
NAA_9&$0.5$&$0.25$&F_{\text{MSY}}&Index sigma (log scale) = 0.1, Age composition sigma (logistic normal) = 0.3\tabularnewline
NAA_10&$1.5$&$0.25$&F_{\text{MSY}}&Index sigma (log scale) = 0.1, Age composition sigma (logistic normal) = 0.3\tabularnewline
NAA_11&$0.5$&$0.50$&F_{\text{MSY}}&Index sigma (log scale) = 0.1, Age composition sigma (logistic normal) = 0.3\tabularnewline
NAA_12&$1.5$&$0.50$&F_{\text{MSY}}&Index sigma (log scale) = 0.1, Age composition sigma (logistic normal) = 0.3\tabularnewline
NAA_13&$0.5$&$$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index sigma (log scale) = 0.4, Age composition sigma (logistic normal) = 1.5\tabularnewline
NAA_14&$1.5$&$$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index sigma (log scale) = 0.4, Age composition sigma (logistic normal) = 1.5\tabularnewline
NAA_15&$0.5$&$0.25$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index sigma (log scale) = 0.4, Age composition sigma (logistic normal) = 1.5\tabularnewline
NAA_16&$1.5$&$0.25$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index sigma (log scale) = 0.4, Age composition sigma (logistic normal) = 1.5\tabularnewline
NAA_17&$0.5$&$0.50$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index sigma (log scale) = 0.4, Age composition sigma (logistic normal) = 1.5\tabularnewline
NAA_18&$1.5$&$0.50$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index sigma (log scale) = 0.4, Age composition sigma (logistic normal) = 1.5\tabularnewline
NAA_19&$0.5$&$$&F_{\text{MSY}}&Index sigma (log scale) = 0.4, Age composition sigma (logistic normal) = 1.5\tabularnewline
NAA_20&$1.5$&$$&F_{\text{MSY}}&Index sigma (log scale) = 0.4, Age composition sigma (logistic normal) = 1.5\tabularnewline
NAA_21&$0.5$&$0.25$&F_{\text{MSY}}&Index sigma (log scale) = 0.4, Age composition sigma (logistic normal) = 1.5\tabularnewline
NAA_22&$1.5$&$0.25$&F_{\text{MSY}}&Index sigma (log scale) = 0.4, Age composition sigma (logistic normal) = 1.5\tabularnewline
NAA_23&$0.5$&$0.50$&F_{\text{MSY}}&Index sigma (log scale) = 0.4, Age composition sigma (logistic normal) = 1.5\tabularnewline
NAA_24&$1.5$&$0.50$&F_{\text{MSY}}&Index sigma (log scale) = 0.4, Age composition sigma (logistic normal) = 1.5\tabularnewline
\hline
\end{tabular}\end{center}

### 2023-04-25 Project_0/paper/paper.Rmd
some text added to M-Ecov readme and some other code edits
saving work, fixing cor par setting in set_M.R

@@
 A low uncertainty specification assumed standard deviation of both series of log-aggregate index observations was 0.1 and the standard deviation of the logistic-normal for age composition observations was 0.3 In the high uncertainty specification the standard deviation for log-aggregate indices was 0.4 and that for the age composition observations was 1.5.
+For all estimating models, standard deviation for log-aggregate observations was assumed known whereas that for the logitic-normal age composition observations was estimated.
 
@@
 \clearpage
+
+\begin{table}
+\caption{Distinguishing characteristics of the operating models with random effects on survival.}
+{\input{naa_om_tab}}
+\end{table}
+
+\begin{table}
+\caption{Distinguishing characteristics of the operating models with random effects on natural mortality.}
+{\input{M_om_tab}}
+\end{table}
+
+\begin{table}
+\caption{Distinguishing characteristics of the operating models with random effects on selectivity.}
+{\input{Sel_om_tab}}
+\end{table}
+
+\begin{table}
+\caption{Distinguishing characteristics of the operating models with random effects on catchability.}
+{\input{q_om_tab}}
+\end{table}
 

### 2023-04-25 Ecov_study/mortality/code/term_SSB_bias_results.R

@@ -1,0 +1,63 @@
+library(here)
+library(ggplot2)
+library(dplyr)
+df.ems = readRDS(file.path(here(),"Ecov_study","mortality","inputs", "df.ems.RDS"))
+df.oms = readRDS(file.path(here(),"Ecov_study","mortality","inputs", "df.oms.RDS"))
+term_ssb_error <- readRDS(file = file.path(here(),"Ecov_study","mortality", "results", "ssb_term_bias_res.RDS"))
+ssb_res <- readRDS(file = file.path(here(),"Ecov_study","mortality", "results", "ssb_results.RDS"))
+
+for(i in 1:3) {
+  re_mod <- c("rec", "rec+1", "rec+M")[i]
+  #EM:  M est, Ecov_beta estimated, OM and EM RE assumption match
+  #df.ems. <- df.ems[df.ems$Ecov_est,]
+  em_ind <- which(df.ems$Ecov_est & df.ems$M_est & df.ems$re_config == re_mod)
+  om_ind <- which(df.oms$NAA_M_re == re_mod) #om and em match
+  res <- t(sapply(om_ind, function(x) {
+    x <- subset(depletion_error, om == x)
+    sapply(em_ind, function(y){ #only 1 em_ind
+      x <- subset(x, em == y)
+      n <- sum(!is.na(x$depletion))
+      out <- c(mean(x$depletion, na.rm = T), sd(x$depletion, na.rm =T)/sqrt(n))
+      out <- c(out,  out[1] + qt(0.025, n) * out[2],
+        out[1] + qt(0.975, n) * out[2])
+      return(out)
+    })
+  }))
+  colnames(res) <- c("bias_est", "bias_se", "bias_ci_lo", "bias_ci_hi")
+  res <- cbind.data.frame(df.oms[om_ind,], res)
+  if(i == 1) {
+    all_res <- res
+  } else {
+    all_res <- rbind.data.frame(all_res, res)
+  }
+}
+#all_res$bias_z = all_res$bias_est/all_res$bias_se
+facs <- c("Ecov_obs_sig", "Fhist", "Ecov_re_sig","Ecov_re_cor", "obs_error", "NAA_M_re")
+all_res[facs] <- lapply(all_res[facs], factor)
+all_res_mod <- all_res %>%
+  mutate(Ecov_obs_sig = recode(Ecov_obs_sig,
+    "0.1" = "Ecov obs SD = 0.1",
+    "0.5" = "Ecov obs SD = 0.5"
[…]

### 2023-04-25 Project_0/code/om_em_description_tables.R

@@ -1,0 +1,83 @@
+library(here)
+library(dplyr)
+library(Hmisc)
+df.ems = readRDS(file.path(here(),"Project_0","inputs", "df.ems.RDS"))
+
+#NAA oms: ems = 1-20
+df.oms = readRDS(file.path(here(),"Project_0","inputs", "df.oms.RDS"))
+om_inputs = readRDS(file.path(here(),"Project_0","inputs", "NAA_om_inputs.RDS"))
+
+naa_om_tab <- df.oms
+naa_om_tab$Model <- paste0("NAA_", 1:24)
+naa_om_tab <- naa_om_tab %>%
+  mutate(obs_error = recode(obs_error,
+    "L" = "Index sigma (log scale) = 0.1, Age composition sigma (logistic normal) = 0.3",
+    "H" = "Index sigma (log scale) = 0.4, Age composition sigma (logistic normal) = 1.5"
+  ))
+naa_om_tab <- naa_om_tab %>%
+  mutate(Fhist = recode(Fhist,
+    "H-MSY" = "$2.5 F_{\\text{MSY}} \\rightarrow F_{\\text{MSY}}$",
+    "MSY" = "F_{\\text{MSY}}"
+  ))
+
+names(naa_om_tab)[2] <- "$\\sigma_R$"
+names(naa_om_tab)[3] <- "$\\sigma_{2+}$"
+names(naa_om_tab)[4] <- "Fishing History"
+names(naa_om_tab)[5] <- "Observation Uncertainty"
+
+x = latex(naa_om_tab, file = here("Project_0","paper","naa_om_tab.tex"), 
+  table.env = FALSE, col.just = rep("r", dim(naa_om_tab)[2]), rowname = NULL)
+
+
+#M oms: ems = 5-24
+df.oms = readRDS(file.path(here(),"Project_0","inputs", "df.M.oms.RDS"))
+om_inputs = readRDS(file.path(here(),"Project_0","inputs", "M_om_inputs.RDS"))
+
+om_tab <- df.oms
+om_tab$Model <- paste0("M_", 1:16)
+om_tab <- om_tab %>%
+  mutate(obs_error = recode(obs_error,
+    "L" = "Index sigma (log scale) = 0.1, Age composition sigma (logistic normal) = 0.3",
+    "H" = "Index sigma (log scale) = 0.4, Age composition sigma (logistic normal) = 1.5"
[…]

### 2023-04-25 common_code/set_M.R

@@ -143,10 +143,10 @@ without changing ASAP file, specify M$initial_means.")
     #if(data$M_re_model == 3) par$M_repars[3] <- 0 # if ar1 over ages only, fix rho_y = 0
     #if(data$M_re_model == 4) par$M_repars[2] <- 0 # if ar1 over years only, fix rho_a = 0
     } else{
-    inv_trans_rho <- function(rho) 0.5 * log(rho+1) - log(1-rho) # 0.5 because needed transformation on cpp side is unusual.
-    if(data$M_re_model == 3) par$M_repars[2] <- inv_trans_rho(M$cor_vals) # if ar1 over ages only, fix rho_y = 0
-    if(data$M_re_model == 4) par$M_repars[3] <- inv_trans_rho(M$cor_vals) # if ar1 over years only, fix rho_a = 0
-    if(data$M_re_model == 5) par$M_repars[2:3] <- inv_trans_rho(M$cor_vals) # if 2dar1 years and ages
+    #inv_trans_rho <- function(rho) 0.5 * log(rho+1) - log(1-rho) # 0.5 because needed transformation on cpp side is unusual.
+    if(data$M_re_model == 3) par$M_repars[2] <- wham:::gen.logit(M$cor_vals, -1, 1, 2) #inv_trans_rho(M$cor_vals) # if ar1 over ages only, fix rho_y = 0
+    if(data$M_re_model == 4) par$M_repars[3] <- wham:::gen.logit(M$cor_vals, -1, 1, 2) #inv_trans_rho(M$cor_vals) # if ar1 over years only, fix rho_a = 0
+    if(data$M_re_model == 5) par$M_repars[2:3] <- wham:::gen.logit(M$cor_vals, -1, 1, 2) #inv_trans_rho(M$cor_vals) # if 2dar1 years and ages
   }
   # M_repars: sigma_M, rho_M_a, rho_M_y
   if(data$M_re_model == 1) tmp <- rep(NA,3) # no RE pars to estimate

### 2023-05-14 Project_0/paper/M_om_tab.tex
lots of results plots for project_0 and Ecov_M

@@
 \hline\hline
-\multicolumn{1}{c}{Model}&\multicolumn{1}{c}{$\sigma_R$}&\multicolumn{1}{c}{$\rho_{M}$}&\multicolumn{1}{c}{Fishing History}&\multicolumn{1}{c}{Observation Uncertainty}&\multicolumn{1}{c}{obs_error}\tabularnewline
+\multicolumn{1}{c}{Model}&\multicolumn{1}{c}{$\sigma_R$}&\multicolumn{1}{c}{$\sigma_{M}$}&\multicolumn{1}{c}{$\rho_{M}$}&\multicolumn{1}{c}{Fishing History}&\multicolumn{1}{c}{Observation Uncertainty}\tabularnewline
 \hline
-M_1&$0.5$&$0.1$&$0.0$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index sigma (log scale) = 0.1, Age composition sigma (logistic normal) = 0.3\tabularnewline
-M_2&$0.5$&$0.5$&$0.0$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index sigma (log scale) = 0.1, Age composition sigma (logistic normal) = 0.3\tabularnewline
-M_3&$0.5$&$0.1$&$0.9$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index sigma (log scale) = 0.1, Age composition sigma (logistic normal) = 0.3\tabularnewline
-M_4&$0.5$&$0.5$&$0.9$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index sigma (log scale) = 0.1, Age composition sigma (logistic normal) = 0.3\tabularnewline
-M_5&$0.5$&$0.1$&$0.0$&F_{\text{MSY}}&Index sigma (log scale) = 0.1, Age composition sigma (logistic normal) = 0.3\tabularnewline
-M_6&$0.5$&$0.5$&$0.0$&F_{\text{MSY}}&Index sigma (log scale) = 0.1, Age composition sigma (logistic normal) = 0.3\tabularnewline
-M_7&$0.5$&$0.1$&$0.9$&F_{\text{MSY}}&Index sigma (log scale) = 0.1, Age composition sigma (logistic normal) = 0.3\tabularnewline
-M_8&$0.5$&$0.5$&$0.9$&F_{\text{MSY}}&Index sigma (log scale) = 0.1, Age composition sigma (logistic normal) = 0.3\tabularnewline
-M_9&$0.5$&$0.1$&$0.0$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index sigma (log scale) = 0.4, Age composition sigma (logistic normal) = 1.5\tabularnewline
-M_10&$0.5$&$0.5$&$0.0$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index sigma (log scale) = 0.4, Age composition sigma (logistic normal) = 1.5\tabularnewline
-M_11&$0.5$&$0.1$&$0.9$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index sigma (log scale) = 0.4, Age composition sigma (logistic normal) = 1.5\tabularnewline
-M_12&$0.5$&$0.5$&$0.9$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index sigma (log scale) = 0.4, Age composition sigma (logistic normal) = 1.5\tabularnewline
-M_13&$0.5$&$0.1$&$0.0$&F_{\text{MSY}}&Index sigma (log scale) = 0.4, Age composition sigma (logistic normal) = 1.5\tabularnewline
-M_14&$0.5$&$0.5$&$0.0$&F_{\text{MSY}}&Index sigma (log scale) = 0.4, Age composition sigma (logistic normal) = 1.5\tabularnewline
-M_15&$0.5$&$0.1$&$0.9$&F_{\text{MSY}}&Index sigma (log scale) = 0.4, Age composition sigma (logistic normal) = 1.5\tabularnewline
-M_16&$0.5$&$0.5$&$0.9$&F_{\text{MSY}}&Index sigma (log scale) = 0.4, Age composition sigma (logistic normal) = 1.5\tabularnewline
+$M_{1}$&$0.5$&$0.1$&$0.0$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index SD = 0.1, Age composition SD = 0.3\tabularnewline
+$M_{2}$&$0.5$&$0.5$&$0.0$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index SD = 0.1, Age composition SD = 0.3\tabularnewline
+$M_{3}$&$0.5$&$0.1$&$0.9$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index SD = 0.1, Age composition SD = 0.3\tabularnewline
+$M_{4}$&$0.5$&$0.5$&$0.9$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index SD = 0.1, Age composition SD = 0.3\tabularnewline
+$M_{5}$&$0.5$&$0.1$&$0.0$&$F_{\text{MSY}}$&Index SD = 0.1, Age composition SD = 0.3\tabularnewline
+$M_{6}$&$0.5$&$0.5$&$0.0$&$F_{\text{MSY}}$&Index SD = 0.1, Age composition SD = 0.3\tabularnewline
+$M_{7}$&$0.5$&$0.1$&$0.9$&$F_{\text{MSY}}$&Index SD = 0.1, Age composition SD = 0.3\tabularnewline
+$M_{8}$&$0.5$&$0.5$&$0.9$&$F_{\text{MSY}}$&Index SD = 0.1, Age composition SD = 0.3\tabularnewline
+$M_{9}$&$0.5$&$0.1$&$0.0$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index SD = 0.4, Age composition SD = 1.5\tabularnewline
+$M_{10}$&$0.5$&$0.5$&$0.0$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index SD = 0.4, Age composition SD = 1.5\tabularnewline
+$M_{11}$&$0.5$&$0.1$&$0.9$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index SD = 0.4, Age composition SD = 1.5\tabularnewline
+$M_{12}$&$0.5$&$0.5$&$0.9$&$2.5 F_{\text{MSY}} \rightarrow F_{\text{MSY}}$&Index SD = 0.4, Age composition SD = 1.5\tabularnewline
+$M_{13}$&$0.5$&$0.1$&$0.0$&$F_{\text{MSY}}$&Index SD = 0.4, Age composition SD = 1.5\tabularnewline
+$M_{14}$&$0.5$&$0.5$&$0.0$&$F_{\text{MSY}}$&Index SD = 0.4, Age composition SD = 1.5\tabularnewline
+$M_{15}$&$0.5$&$0.1$&$0.9$&$F_{\text{MSY}}$&Index SD = 0.4, Age composition SD = 1.5\tabularnewline
+$M_{16}$&$0.5$&$0.5$&$0.9$&$F_{\text{MSY}}$&Index SD = 0.4, Age composition SD = 1.5\tabularnewline
 \hline

### 2023-05-14 Project_0/paper/em_tab.tex
lots of results plots for project_0 and Ecov_M

\begin{center}
\begin{tabular}{rrrr}
\hline\hline
\multicolumn{1}{c}{Model}&\multicolumn{1}{c}{Recruitment model}&\multicolumn{1}{c}{Mean $M$}&\multicolumn{1}{c}{Process error assumption}\tabularnewline
\hline
EM$_{1}$&Mean recruitment&0.2&Recruitment ($\sigma_R$ estimated)\tabularnewline
EM$_{2}$&Beverton-Holt&0.2&Recruitment ($\sigma_R$ estimated)\tabularnewline
EM$_{3}$&Mean recruitment&Estimated&Recruitment ($\sigma_R$ estimated)\tabularnewline
EM$_{4}$&Beverton-Holt&Estimated&Recruitment ($\sigma_R$ estimated)\tabularnewline
EM$_{5}$&Mean recruitment&0.2&Recruitment and survival ($\sigma_R$, $\sigma_{2+}$ estimated)\tabularnewline
EM$_{6}$&Beverton-Holt&0.2&Recruitment and survival ($\sigma_R$, $\sigma_{2+}$ estimated)\tabularnewline
EM$_{7}$&Mean recruitment&Estimated&Recruitment and survival ($\sigma_R$, $\sigma_{2+}$ estimated)\tabularnewline
EM$_{8}$&Beverton-Holt&Estimated&Recruitment and survival ($\sigma_R$, $\sigma_{2+}$ estimated)\tabularnewline
EM$_{9}$&Mean recruitment&0.2&Recruitment and uncorrelated natural mortality ($\sigma_R$, $\sigma_{M}$ estimated, $\rho_{M} = 0$)\tabularnewline
EM$_{10}$&Beverton-Holt&0.2&Recruitment and uncorrelated natural mortality ($\sigma_R$, $\sigma_{M}$ estimated, $\rho_{M} = 0$)\tabularnewline
EM$_{11}$&Mean recruitment&Estimated&Recruitment and uncorrelated natural mortality ($\sigma_R$, $\sigma_{M}$ estimated, $\rho_{M} = 0$)\tabularnewline
EM$_{12}$&Beverton-Holt&Estimated&Recruitment and uncorrelated natural mortality ($\sigma_R$, $\sigma_{M}$ estimated, $\rho_{M} = 0$)\tabularnewline
EM$_{13}$&Mean recruitment&0.2&Recruitment and uncorrelated fleet selectivity ($\sigma_R$, $\sigma_{\text{Sel}}$ estimated, $\rho_{\text{Sel}} = 0$)\tabularnewline
EM$_{14}$&Beverton-Holt&0.2&Recruitment and uncorrelated fleet selectivity ($\sigma_R$, $\sigma_{\text{Sel}}$ estimated, $\rho_{\text{Sel}} = 0$)\tabularnewline
EM$_{15}$&Mean recruitment&Estimated&Recruitment and uncorrelated fleet selectivity ($\sigma_R$, $\sigma_{\text{Sel}}$ estimated, $\rho_{\text{Sel}} = 0$)\tabularnewline
EM$_{16}$&Beverton-Holt&Estimated&Recruitment and uncorrelated fleet selectivity ($\sigma_R$, $\sigma_{\text{Sel}}$ estimated, $\rho_{\text{Sel}} = 0$)\tabularnewline
EM$_{17}$&Mean recruitment&0.2&Recruitment and uncorrelated catchability (spring index) ($\sigma_R$, $\sigma_{q}$ estimated, $\rho_{q} = 0$)\tabularnewline
EM$_{18}$&Beverton-Holt&0.2&Recruitment and uncorrelated catchability (spring index) ($\sigma_R$, $\sigma_{q}$ estimated, $\rho_{q} = 0$)\tabularnewline
EM$_{19}$&Mean recruitment&Estimated&Recruitment and uncorrelated catchability (spring index) ($\sigma_R$, $\sigma_{q}$ estimated, $\rho_{q} = 0$)\tabularnewline
EM$_{20}$&Beverton-Holt&Estimated&Recruitment and uncorrelated catchability (spring index) ($\sigma_R$, $\sigma_{q}$ estimated, $\rho_{q} = 0$)\tabularnewline
EM$_{21}$&Mean recruitment&0.2&Recruitment and AR1 natural mortality ($\sigma_R$, $\sigma_{M}$, $\rho_{M}$ estimated)\tabularnewline
EM$_{22}$&Beverton-Holt&0.2&Recruitment and AR1 natural mortality ($\sigma_R$, $\sigma_{M}$, $\rho_{M}$ estimated)\tabularnewline
EM$_{23}$&Mean recruitment&Estimated&Recruitment and AR1 natural mortality ($\sigma_R$, $\sigma_{M}$, $\rho_{M}$ estimated)\tabularnewline
EM$_{24}$&Beverton-Holt&Estimated&Recruitment and AR1 natural mortality ($\sigma_R$, $\sigma_{M}$, $\rho_{M}$ estimated)\tabularnewline
EM$_{25}$&Mean recruitment&0.2&Recruitment and AR1 selectivity ($\sigma_R$, $\sigma_{\text{Sel}}$, $\rho_{\text{Sel}}$ estimated)\tabularnewline
EM$_{26}$&Beverton-Holt&0.2&Recruitment and AR1 selectivity ($\sigma_R$, $\sigma_{\text{Sel}}$, $\rho_{\text{Sel}}$ estimated)\tabularnewline
EM$_{27}$&Mean recruitment&Estimated&Recruitment and AR1 selectivity ($\sigma_R$, $\sigma_{\text{Sel}}$, $\rho_{\text{Sel}}$ estimated)\tabularnewline
EM$_{28}$&Beverton-Holt&Estimated&Recruitment and AR1 selectivity ($\sigma_R$, $\sigma_{\text{Sel}}$, $\rho_{\text{Sel}}$ estimated)\tabularnewline
EM$_{29}$&Mean recruitment&0.2&Recruitment and AR1 catchability (spring index) ($\sigma_R$, $\sigma_{q}$, $\rho_{q}$ estimated)\tabularnewline
EM$_{30}$&Beverton-Holt&0.2&Recruitment and AR1 catchability (spring index) ($\sigma_R$, $\sigma_{q}$, $\rho_{q}$ estimated)\tabularnewline
EM$_{31}$&Mean recruitment&Estimated&Recruitment and AR1 catchability (spring index) ($\sigma_R$, $\sigma_{q}$, $\rho_{q}$ estimated)\tabularnewline
EM$_{32}$&Beverton-Holt&Estimated&Recruitment and AR1 catchability (spring index) ($\sigma_R$, $\sigma_{q}$, $\rho_{q}$ estimated)\tabularnewline
\hline
\end{tabular}\end{center}

### 2023-05-14 Project_0/paper/paper.Rmd
lots of results plots for project_0 and Ecov_M

--- 2023-02-10 Project_0/paper/paper.Rmd
@@
 A low uncertainty specification assumed standard deviation of both series of log-aggregate index observations was 0.1 and the standard deviation of the logistic-normal for age composition observations was 0.3 In the high uncertainty specification the standard deviation for log-aggregate indices was 0.4 and that for the age composition observations was 1.5.
+For all estimating models, standard deviation for log-aggregate observations was assumed known whereas that for the logistic-normal age composition observations was estimated.
 
@@
 
-Simulations were all carried out on the University of Massachussetts Green High-Performance Computing Cluster.
+Simulations were all carried out on the University of Massachusetts Green High-Performance Computing Cluster.
 Code for completing the simulations and summarization of results can be found at github.com/timjmiller/SSRTWG/Project_0.
-We used the wham package version 1.X.X (commit XXXXX).
+We used the wham package version 1.X.X (commit 77bbd94).
 
@@
 \clearpage
+
+\begin{landscape}
+\begin{table}
+\caption{Distinguishing characteristics of the operating models with random effects on survival.
+Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
+Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.}
+{\footnotesize \input{naa_om_tab}}
+\end{table}
+
+\begin{table}
+\caption{Distinguishing characteristics of the operating models with random effects on natural mortality.
+Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
+Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
+For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}
+{\input{M_om_tab}}
+\end{table}
+
+\begin{table}
+\caption{Distinguishing characteristics of the operating models with random effects on selectivity.
+Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
+Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
+For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}
+{\input{Sel_om_tab}}
+\end{table}
+
+\begin{table}
+\caption{Distinguishing characteristics of the operating models with random effects on catchability.
+Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
+Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
+For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}
+{\input{q_om_tab}}
+\end{table}
+\end{landscape}
+
+\begin{table}
+\caption{Distinguishing characteristics of the estimating models.}
+{\scriptsize \input{em_tab}}
+\end{table}
 

### 2023-05-14 Project_0/paper/paper.tex
lots of results plots for project_0 and Ecov_M

@@
 \hypersetup{
-  pdftitle={How reliable are inferences from state-space age-structured assessment models?},
-  pdfauthor={Timothy J. Miller1, Britten, Brooks, Fay, Legault, Stock,},
+  pdftitle={Factors affecting reliablity of state-space age-structured assessment models},
+  pdfauthor={Timothy J. Miller1, Britten, Brooks, Fay, Legault, Stock, et al.},
   hidelinks,
@@
 
-\title{How reliable are inferences from state-space age-structured
-assessment models?}
+\title{Factors affecting reliablity of state-space age-structured
+assessment models}
 \author{Timothy J. Miller\textsuperscript{1}, Britten, Brooks, Fay,
-Legault, Stock,}
+Legault, Stock, et al.}
 \date{}
@@
 
+Here we conduct a simulation study with operating models varying by
+degree of observation error uncertainty, sources of process error (M,
+NAA, q, sel), and fishing history. The simulations from these operating
+models are fitted with estimating models that make alternative
+assumptions for sources of process error (M, NAA\textless{} q, sel),
+whether a stock-recruit model was estimated, and whether M is estimated.
+We evaluate whether AIC can correctly determine the correct source of
+process error and stock effects on recruitment. We also evaluate the
+degree of bias in the outputs of the assessment model that are important
+for management.
+
 \hypertarget{methods}{%
@@
 and estimating models for closed loop simulations evaluating index-based
-assessment methods \citep{legaultetalinpress}.
+assessment methods \citep{legaultetalinpress} and is used for management
+of haddock, butterfish, plaice, bluefish in the Northeast US.
 
@@
 
+Currenlty, initial population is configured at the equilibrium
+distribution fishing at \(F = 2.5\times F_{\text{MSY}}\).
+
 Initial population was configured at the equilibrium distribution
@@
 the standard deviation for log-aggregate indices was 0.4 and that for
-the age composition observations was 1.5.
+the age composition observations was 1.5. For all estimating models,
+standard deviation for log-aggregate observations was assumed known
+whereas that for the logistic-normal age composition observations was
+estimated.
 
@@
 
+Steepness and BRPs are not constant when M is time-varying (Miller and
+Brooks 2021). We uses the a/b parameters for the B-H defined for Fmsy =
+F40 at the mean M = 0.2 as defined above.
+
 \hypertarget{operating-models-with-random-effects-on-fleet-selectivity}{%
@@
 16 operating models Table of process error assumptions
+
+BRPs are not constant when fleet selectivity is time-varying. We uses
+the a/b parameters for the B-H defined for Fmsy = F40 at the mean of the
+time series model for selectivity which is the same as the constant
+selectivity defined above.
 
@@
 
-Simulations were all carried out on the University of Massachussetts
+Simulations were all carried out on the University of Massachusetts
 Green High-Performance Computing Cluster. Code for completing the
@@
 github.com/timjmiller/SSRTWG/Project\_0. We used the wham package
-version 1.X.X (commit XXXXX).
+version 1.X.X (commit 77bbd94).
 

### 2023-05-14 Ecov_study/mortality/code/aic_ecov_effect_results.R

@@ -162,3 +162,105 @@ plt <- ggplot(all_res_mod, aes(x = Ecov_effect, y = emp_p, colour = Ecov_re_sig:
     geom_hline(aes(yintercept=0.5), linewidth = 1, linetype = "dashed", colour = "red")
 plt
 ggsave(here("Ecov_study","mortality", "paper", "proportion_correct_effect_M_estimated.png"), plt)
+
+#all EM PE assumptions
+for(i in 1:3) {
+  re_mod <- c("rec", "rec+1", "rec+M")[i]
+  #EM:  M estimated
+  em_ind <- which(df.ems$M_est == TRUE) #all EM PE assumptions
+  om_ind <- which(df.oms$NAA_M_re == re_mod) #om and em match
+  Mfixed_rec <- aic_fn(aic_res, em_ind, om_ind)
+  res <- cbind(df.oms[rep(om_ind, each = length(em_ind)),], df.ems[rep(em_ind, length(om_ind)),], n = c(Mfixed_rec))
+  if(i == 1) {
+    all_res <- res
+  } else {
+    all_res <- rbind(all_res, res)
+  }
+}
+facs <- c("Ecov_obs_sig", "Fhist", "Ecov_re_sig","Ecov_re_cor","obs_error", "NAA_M_re", "re_config", "Ecov_est")
+all_res[facs] <- lapply(all_res[facs], factor)
+all_res_mod <- all_res %>%
+  mutate(Ecov_obs_sig = recode(Ecov_obs_sig,
+    "0.1" = "Ecov obs SD = 0.1",
+    "0.5" = "Ecov obs SD = 0.5"
+  ))
+all_res_mod <- all_res_mod %>%
+  mutate(obs_error = recode(obs_error,
+    "L" = "Low obs error (indices, age comp)",
+    "H" = "High obs error (indices, age comp)"
+  ))
+all_res_mod <- all_res_mod %>%
+  mutate(NAA_M_re = recode(NAA_M_re,
+    "rec" = "R",
+    "rec+1" = "R+S",
+    "rec+M" = "R+M"
+  ))
+all_res_mod <- all_res_mod %>%
+  mutate(re_config = recode(re_config,
+    "rec" = "R",
+    "rec+1" = "R+S",
+    "rec+M" = "R+M"
+  ))
+all_res_mod <- all_res_mod %>%
+  mutate(Ecov_re_sig = recode(Ecov_re_sig,
+    "0.1" = "Ecov SD = 0.1",
[…]

### 2023-05-14 Project_0/code/aic_tables_plots.R

@@ -1,0 +1,391 @@
+library(here)
+df.ems = readRDS(file.path(here(),"Project_0","inputs", "df.ems.RDS"))
+
+#NAA oms: ems = 1-20
+df.oms = readRDS(file.path(here(),"Project_0","inputs", "df.oms.RDS"))
+all_naa_aic = lapply(1:NROW(df.oms), function(y){
+  res = sapply(1:100, function(x){
+    print(paste0("om_", y, ", sim_",x))
+    sim = readRDS(file.path(here::here(),"Project_0", "results", "naa_om", paste0("om_", y), paste0("sim_",x,".RDS")))
+    aic = 2*sapply(sim,function(y) {
+      out = NA
+      if(length(y$fit)) out = y$fit$opt$obj + length(y$fit$opt$par)
+      return(out)
+    })
+    return(aic)
+  })
+  return(res)
+})
+saveRDS(all_naa_aic, file = file.path(here(),"Project_0","results", "all_naa_aic_results.RDS"))
+
+naa_outer_res = sapply(all_naa_aic, function(y){
+  res <- y
+  tmp = apply(res,2, function(x) {
+    if(any(!is.na(x))) {
+      return(x == min(x,na.rm=T))
+    } else return(rep(NA, length(x)))
+  })
+  return(apply(tmp,1,sum,na.rm=T))
+})
+saveRDS(naa_outer_res, file = file.path(here(),"Project_0","results", "naa_om_aic_choice_results.RDS"))
+
+#M oms: ems = 5-20, 21-24
+df.M.oms = readRDS(file.path(here(),"Project_0","inputs", "df.M.oms.RDS"))
+all_M_aic = lapply(1:NROW(df.M.oms), function(y){
+  res = sapply(1:100, function(x){
+    print(paste0("om_", y, ", sim_",x))
+    sim = readRDS(file.path(here::here(),"Project_0", "results", "M_om", paste0("om_", y), paste0("sim_",x,".RDS")))
+    aic = 2*sapply(sim,function(y) {
+      out = NA
+      if(length(y$fit)) out = y$fit$opt$obj + length(y$fit$opt$par)
+      return(out)
+    })
+    return(aic)
+  })
+  return(res)
+})
[…]

### 2023-05-14 Project_0/code/relbias_plots.R

@@ -1,0 +1,277 @@
+relssb_fn <- function(df.oms, om_type = "naa"){
+  relssb = lapply(1:NROW(df.oms), function(y){
+    res = lapply(1:100, function(x){
+      print(paste0("om_", y, ", sim_",x))
+      sim = readRDS(file.path(here::here(),"Project_0", "results", paste0(om_type,"_om"), paste0("om_", y), paste0("sim_",x,".RDS")))
+      relSSB = lapply(sim,function(z) {
+        out <- matrix(NA, 40,3)
+        if(length(z)) if(length(z$fit)) out[,1] = z$fit$rep$SSB/z$truth$SSB
+        if(!is.null(z$fit$sdrep)){
+          #ind <- which(!is.na(z$fit$sdrep$SE_rep$log_SSB))
+          out[,2] <- z$fit$sdrep$SE_rep$log_SSB #se
+            ci = matrix(z$fit$sdrep$Estimate_rep$log_SSB - log(z$truth$SSB),40,2) + qnorm(0.975) * cbind(-out[,2],out[,2])
+          out[,3] <- 0 >= ci[,1] & 0 <= ci[,2]
+        }
+        return(out)
+      })
+      return(relSSB)
+    })
+    return(res)
+  })
+  return(relssb)
+}
+
+relR_fn <- function(df.oms, om_type = "naa"){
+  relR = lapply(1:NROW(df.oms), function(y){
+    res = lapply(1:100, function(x){
+      print(paste0("om_", y, ", sim_",x))
+      sim = readRDS(file.path(here::here(),"Project_0", "results", paste0(om_type,"_om"), paste0("om_", y), paste0("sim_",x,".RDS")))
+      relR = sapply(sim,function(y) {
+        out <- rep(NA, 40)
+        if(length(y)) if(length(y$fit)) out = y$fit$rep$NAA[,1]/y$truth$NAA[,1]
+        return(out)
+      })
+      return(relR)
+    })
+    return(res)
+  })
+  return(relR)
+}
+
+relF_fn <- function(df.oms, om_type = "naa"){
+  relF = lapply(1:NROW(df.oms), function(y){
+    res = lapply(1:100, function(x){
+      print(paste0("om_", y, ", sim_",x))
[…]

### 2023-06-02 Project_0/paper/paper.Rmd
saving work

@@
 review literature on reliability of hidden/latent process models. Primarily in other fields.
+
+Convergence results can be useful for understanding how bad convergence in applications to real data might direct the practitioner to which alternative random effects configurations is be more appropriate.
+The type of convergence that we might use as a diagnostic is important.
+And what rule of thumb we use for certain criteria e.g.
+1e-6?
+large SEs/CVs?
 
@@
 
+Summary statitics
+
+Bias
+
+Mohn's rho
+
+marginal AIC
+
+convergence types: 
+1) whether the optimization completed
+2) nlminb convergence flag = 0
+3) max abs gradient < 1e-6
+4) No NAs for SE of fixed effects returned from TMB::sdreport
+5) No SE NAs AND no SE > 100
+
 # Results
@@
 Convergence performance
+
+Probability of convergence was better for models that assumed the correct source of process error, assumed M was known, and did not assume stock-recruit relationships.
 

### 2023-06-02 Ecov_study/mortality/code/aic_ecov_effect_results.R

@@ -1,6 +1,8 @@
 library(here)
 library(ggplot2)
 library(dplyr)
+df.ems = readRDS(file.path(here(),"Ecov_study","mortality","inputs", "df.ems.RDS"))
+df.oms = readRDS(file.path(here(),"Ecov_study","mortality","inputs", "df.oms.RDS"))
 aic_res <- readRDS(file.path(here(),"Ecov_study","mortality", "results", "aic_results.RDS"))
 
 aic_fn <- function(all, est_ind, om_ind = NULL){
@@ -20,13 +22,124 @@ aic_fn <- function(all, est_ind, om_ind = NULL){
   return(out)
 }
 
+aic_rank_df_fn <- function(df.ems, df.oms, M_est = FALSE) {
+  for(i in 1:3) {
+    
+    re_mod <- c("rec", "rec+1", "rec+M")[i]
+    #EM:  M fixed, OM and EM RE assumption match
+    em_ind <- which(df.ems$M_est == M_est & df.ems$re_config == re_mod)
+    om_ind <- which(df.oms$NAA_M_re == re_mod) #om and em match
+    #Mfixed_rec <- t(aic_wt_fn(aic_res, em_ind, om_ind))
+    Mfixed_rec <- t(aic_fn(aic_res, em_ind, om_ind))
+    sapply(om_ind, function(i) {
+      length(aic_res[[i]][[em_ind[2]]])
+    })
+    colnames(Mfixed_rec) <- c("effect_est", "effect_0")
+    res <- cbind(df.oms[om_ind,], Mfixed_rec)
+    temp <- t(sapply(1:NROW(res), function(x) {
+      if(res$Ecov_effect[x] == 0) out <- c(res$effect_0[x], res$effect_est[x])
+      if(res$Ecov_effect[x] > 0) out <- c(res$effect_est[x], res$effect_0[x])
+      return(out)
+    }))
+    colnames(temp) <- c("right","wrong")
+    res <- cbind(res,temp)
+    if(i == 1) {
+      all_res <- res
+    } else {
+      all_res <- rbind(all_res, res)
+    }
+  }
+  all_res$emp_p <- all_res$right/(all_res$right + all_res$wrong)
+  all_res$ci_lo <- sapply(1:NROW(all_res), function(x) {
+    binom.test(all_res$right[x], all_res$right[x] + all_res$wrong[x], p = all_res$emp_p[x])$conf.int[1]
+  })
[…]

### 2023-06-02 Project_0/code/aic_tables_plots.R

@@ -389,3 +389,167 @@ for(j in 1:4){
   ggsave(here("Project_0", "paper", paste0("om_em_proportion_best_aic_coarse_", ifelse(est_conds[j,1], "SR", "R"), "_", 
       ifelse(est_conds[j,2], "ME", "MF"),".png")), plt, width = 20, height = 12, units = "in")
 }
+
+aic_fn <- function(all, est_ind, om_ind = NULL){
+  if(!is.null(om_ind)) all <- all[om_ind]
+  out <- sapply(all, function(res){
+    tmp = apply(res[est_ind,],2, function(x) {
+      if(any(!is.na(x))) {
+        return(x == min(x,na.rm=T))
+      } else return(rep(NA, length(x)))
+    })
+    return(apply(tmp,1,sum,na.rm=T))
+  })  
+  return(out)
+}
+
+
+
+
+
+
+############################################
+#AIC weights
+############################################
+
+make_plot_df <- function(aic_res, df.ems, df.oms, em_ind = 1:20, SR_est =FALSE, M_est = FALSE){
+  use.df.ems <- df.ems[em_ind,]
+  SR_M_ind = which(use.df.ems$SR_model == ifelse(SR_est,3,2) & use.df.ems$M_est == M_est)
+  aic_res <- t(aic_wt_fn(aic_res, SR_M_ind))
+  print(aic_res)
+  cnames <- paste0(rep(1:length(SR_M_ind), each = 3), "_", c("lo", "med", "hi"))
+  colnames(aic_res) <- cnames
+  out <- cbind(df.oms[-1], aic_res)
+  out <- out %>% pivot_longer(
+    cols = all_of(cnames),
+    names_to = c("EM", "Type"),
+    names_pattern = "(.)_(.*)" #magic
+  )
+  out <- out %>% pivot_wider(names_from = Type, values_from = value) %>% as.data.frame
+
+  print(head(out))
+  if(max(em_ind)==20) {
+    out$NAA_sig[which(is.na(out$NAA_sig))] <- 0
+    out <- out %>% mutate(NAA_sig = recode(NAA_sig,
+        "0" = "sigma['2+'] == 0",
+        "0.25" = "sigma['2+'] == 0.25",
+        "0.5" = "sigma['2+'] == 0.5")) %>% as.data.frame
[…]

### 2023-06-02 Project_0/code/convergence_results.R

@@ -7,7 +7,7 @@ convergence_fn <- function(em){
   out <- rep(NA,5) #convergence type  1, and 2
   if(!is.character(em)) {
     if(!is.null(em$fit$opt)) {
-      out[1] <- 1
+      out[1] <- 1 
       out[2] <- em$fit$opt$conv
     }
     if(!is.null(em$fit$sdrep)) out[3] <- as.integer(sum(sapply(em$fit$sdrep$SE_par, function(g) any(is.nan(g)))))
@@ -66,27 +66,23 @@ saveRDS(q_conv_res, file = here("Project_0", "results", "q_om_convergence_result
 #convergence information:
 #1: 0/1 model finished optimizing
 #2: nlminb convergence flag
-#3: number of NaNs in SEs for parameters, 0 = good invertible hessian
-#4: max gradient value
+#3: max gradient value
+#4: number of NaNs in SEs for parameters, 0 = good invertible hessian
 #5: maximum non-NaN SE estimate
 
 conv_fn <- function(all, est_ind, om_ind = NULL){
+  #est_ind is 1 ROW of df.ems.
   if(!is.null(om_ind)) all <- all[om_ind]
   out <- sapply(all, function(res){
     tmp <- sapply(res, function(x) {
-      #print(est_ind)
-      #print(length(res))
-      #print(tmp)
       return(x[est_ind,])
     })
-    print(tmp)
-    x <- tmp
-    stop()
+    x <- t(tmp)
     nconv <- c(
       sum(!is.na(x[,1])), 
       sum(!is.na(x[,2]) & x[,2] == 0), 
-      sum(!is.na(x[,3]) & x[,3] == 0),
       sum(!is.na(x[,4]) & x[,4] < 1e-6),
+      sum(!is.na(x[,3]) & x[,3] == 0),
       sum(!is.na(x[,5]) & x[,3] == 0 & x[,5] < 100)
     )
     nconv <- c(nconv, NROW(x))
@@ -94,23 +90,38 @@ conv_fn <- function(all, est_ind, om_ind = NULL){
   })
   return(t(out))
 }
+naa_om_conv_res <- readRDS(here("Project_0", "results", "naa_om_convergence_results.RDS"))
+ind <- which(df.ems$M_est == FALSE & df.ems$SR_model == 2)
+ind <- ind[which(ind<21)]
+x <- conv_fn(naa_om_conv_res, ind[1])
 
-#M estimated
[…]

### 2023-06-06 Project_0/paper/paper.Rmd
saving project_0 work

@@
 
+\begin{landscape}
+\begin{table}
+\caption{Distinguishing characteristics of the operating models with random effects on survival.
+Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
+Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.}
+{\footnotesize \input{naa_om_tab}}
+\end{table}
+\end{landscape}
+
 ### Operating models with random effects on natural mortality
@@
 
+\begin{landscape}
+\begin{table}
+\caption{Distinguishing characteristics of the operating models with random effects on natural mortality.
+Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
+Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
+For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}
+{\input{M_om_tab}}
+\end{table}
+\end{landscape}
+
 ### Operating models with random effects on fleet selectivity
@@
 
+\begin{landscape}
+\begin{table}
+\caption{Distinguishing characteristics of the operating models with random effects on selectivity.
+Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
+Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
+For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}
+{\input{Sel_om_tab}}
+\end{table}
+\end{landscape}
+
 ### Operating models with random effects on index catchability
@@
 
+\begin{landscape}
+\begin{table}
+\caption{Distinguishing characteristics of the operating models with random effects on catchability.
+Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
+Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
+For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}
+{\input{q_om_tab}}
+\end{table}
+\end{landscape}
+
 ## Estimating models
+
+\begin{table}
+\caption{Distinguishing characteristics of the estimating models.}
+{\scriptsize \input{em_tab}}
+\end{table}
 
@@
 Do each of these by type of operating model (Naa, M, sel, q)
-Convergence performance
+
+## Convergence performance
+
+convergence types: 
+1) whether the optimization completed
+2) nlminb convergence flag = 0
+3) max abs gradient < 1e-6
+4) No NAs for SE of fixed effects returned from TMB::sdreport
+5) No SE NAs AND no SE > 100
 
@@
 
-AIC performance
-
-SR estimation? M estimation?
-
-Bias, Mean Square error
-
-  Certain basic parameters (stock-recruit pars, M, variance parameters)
-  SSB, F, R
-
-\clearpage 
-## Numbers at age operating models
-
-### Estimating models include alternative random effects options: NAA, M, sel, q
-\clearpage
+### R, R+S operating models
 
 \begin{landscape}
-\begin{table}
-\caption{Distinguishing characteristics of the operating models with random effects on survival.
-Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
-Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.}
-{\footnotesize \input{naa_om_tab}}
-\end{table}
-
-\begin{table}
-\caption{Distinguishing characteristics of the operating models with random effects on natural mortality.
-Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
-Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
-For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}
-{\input{M_om_tab}}
-\end{table}
-
-\begin{table}
-\caption{Distinguishing characteristics of the operating models with random effects on selectivity.
-Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
-Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
-For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}
-{\input{Sel_om_tab}}
-\end{table}
-
-\begin{table}
-\caption{Distinguishing characteristics of the operating models with random effects on catchability.
-Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
-Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
-For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}
-{\input{q_om_tab}}
-\end{table}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structures R and R+S.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.}\label{naa_om_em_R_MF_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_p_convergence_meanR_M_fixed.png}
+\end{center}
+\end{figure}
 \end{landscape}
 
-\begin{table}
-\caption{Distinguishing characteristics of the estimating models.}
-{\scriptsize \input{em_tab}}
-\end{table}
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structures R and R+S.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.}\label{naa_om_em_R_ME_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_p_convergence_meanR_M_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structures R and R+S.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is fixed at the true value.}\label{naa_om_em_BH_MF_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_p_convergence_BH_M_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structures R and R+S.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is estimated.}\label{naa_om_em_BH_ME_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_p_convergence_BH_M_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+### R+M operating models
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+M.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.}\label{M_om_em_R_MF_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_p_convergence_meanR_M_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+M.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.}\label{M_om_em_R_ME_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_p_convergence_meanR_M_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+M.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is fixed at the true value.}\label{M_om_em_BH_MF_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_p_convergence_BH_M_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+M.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is estimated.}\label{M_om_em_BH_ME_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_p_convergence_BH_M_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+### R+Sel operating models
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+Sel.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.}\label{Sel_om_em_R_MF_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_p_convergence_meanR_M_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+Sel.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.}\label{Sel_om_em_R_ME_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_p_convergence_meanR_M_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+Sel.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is fixed at the true value.}\label{Sel_om_em_BH_MF_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_p_convergence_BH_M_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+Sel.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is estimated.}\label{Sel_om_em_BH_ME_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_p_convergence_BH_M_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+### R+q operating models
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+q.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.}\label{q_om_em_R_MF_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_p_convergence_meanR_M_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+q.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.}\label{q_om_em_R_ME_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_p_convergence_meanR_M_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+q.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is fixed at the true value.}\label{q_om_em_BH_MF_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_p_convergence_BH_M_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+q.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is estimated.}\label{q_om_em_BH_ME_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_p_convergence_BH_M_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+## AIC performance for process error structure
+\clearpage 
+### Numbers at age operating models
+
+## Estimating models include alternative random effects options: NAA, M, sel, q
+\clearpage
 
@@
 \clearpage
+
+\clearpage 
+## AIC performance for stock-recruit relationship
+
+### R, R+S operating models
+
+\begin{table}
+\caption{Operating models and estimation models all assume matching R or R+S process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}
+{\input{naa_om_em_R_BH_aic_table}}
+\end{table}
+
+\begin{figure}
+\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
+Operating and estimating models have matching R or R+S process error structures.
+Estimating models assume mean M is fixed at the true value.}\label{naa_om_MF_BH_glm_AIC_plots}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_MF_pred_BH_best.png}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB. Estimating models allow estimation of mean M.}\label{naa_om_ME_BH_glm_AIC_plots}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_ME_pred_BH_best.png}
+\end{center}
+\end{figure}
+
+### R+M operating models
+
+\begin{table}
+\caption{Operating models and estimation models all assume matching R+M process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}
+{\input{M_om_em_R_BH_aic_table}}
+\end{table}
+
+\begin{figure}
+\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
+Operating and estimating models have matching R+M process error structures.
+Estimating models assume mean M is fixed at the true value.}\label{M_om_MF_BH_glm_AIC_plots}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_MF_pred_BH_best.png}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
+Operating and estimating models have matching R+M process error structures.
+Estimating models allow estimation of mean M.}\label{M_om_ME_BH_glm_AIC_plots}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_ME_pred_BH_best.png}
+\end{center}
+\end{figure}
+
+### R+Sel operating models
+
+\begin{table}
+\caption{Operating models and estimation models all assume matching R+Sel process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}
+{\input{Sel_om_em_R_BH_aic_table}}
+\end{table}
+
+\begin{figure}
+\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
+Operating and estimating models have matching R+Sel process error structures.
+Estimating models assume mean M is fixed at the true value.}\label{Sel_om_MF_BH_glm_AIC_plots}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_MF_pred_BH_best.png}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
+Operating and estimating models have matching R+Sel process error structures.
+Estimating models allow estimation of mean M.}\label{Sel_om_ME_BH_glm_AIC_plots}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_ME_pred_BH_best.png}
+\end{center}
+\end{figure}
+
+### R+q operating models
+
+\begin{table}
+\caption{Operating models and estimation models all assume matching R+q process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}
+{\input{q_om_em_R_BH_aic_table}}
+\end{table}
+
+\begin{figure}
+\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
+Operating and estimating models have matching R+q process error structures.
+Estimating models assume mean M is fixed at the true value.}\label{q_om_MF_BH_glm_AIC_plots}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_MF_pred_BH_best.png}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
+Operating and estimating models have matching R+q process error structures.
+Estimating models allow estimation of mean M.}\label{q_om_ME_BH_glm_AIC_plots}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_ME_pred_BH_best.png}
+\end{center}
+\end{figure}
+
+\clearpage 
+## Bias, Mean Square error
+
+  Certain basic parameters (stock-recruit pars, M, variance parameters)
+  SSB, F, R, BRPs
 
@@
 \end{landscape}
-
-### Estimating models include NAA random effects and estimation assumes mean R or BH SR
-\clearpage
-
-\begin{table}
-\caption{Operating models and estimation models all assume RE on recruitment only, estimating models assume mean recruitment or a B-H stock recruit relationship and M is fixed at the true value.}
-{\input{rec_om_em_R_SR_MF_aic_table}}
-\end{table}
-
-\begin{table}
-\caption{Operating models and estimation models all assume RE on recruitment only, estimating models assume mean recruitment or a B-H stock recruit relationship and M is estimated.}
-{\input{rec_om_em_R_SR_ME_aic_table}}
-\end{table}
-
-\begin{table}
-\caption{Operating models and estimation models all assume RE on all abundances at age, estimating models assume mean recruitment or a B-H stock recruit relationship and M is fixed at the true value.}
-{\input{naa_om_em_R_SR_MF_aic_table}}
-\end{table}
-
-\begin{table}
-\caption{Operating models and estimation models all assume RE on all abundances at age, estimating models assume mean recruitment or a B-H stock recruit relationship and M is estimated.}
-{\input{naa_om_em_R_SR_ME_aic_table}}
-\end{table}
-\clearpage
-
-\begin{figure}
-\caption{Predicted probability of AIC being better for BH model.}\label{glm_res_NAA_re}
-\begin{center}
-\includegraphics[width = \textwidth]{pred_SR_best_NAA_oms.png}
-\end{center}
-\end{figure}
-
-## M operating models
-
-### Estimating models include NAA random effects and estimation assumes mean R or BH SR
-
-\begin{table}
-\caption{M random effects operating models.}
-{\input{M_om_em_R_BH_aic_table}}
-\end{table}
-\clearpage
 

### 2023-06-06 Project_0/code/stock_recruit_AIC_tables.R

@@ -1,3 +1,10 @@
+library("here")
+df.ems = readRDS(file.path(here(),"Project_0","inputs", "df.ems.RDS"))
+all_naa_aic <- readRDS(file = file.path(here(),"Project_0","results", "all_naa_aic_results.RDS"))
+
+#first NAA operating models
+df.oms = readRDS(file.path(here(),"Project_0","inputs", "df.oms.RDS"))
+use.df.ems <- df.ems[1:20,]
 library(Hmisc)
 aic_fn <- function(all, est_ind, om_ind = NULL){
   if(!is.null(om_ind)) all <- all[om_ind]
@@ -16,43 +23,51 @@ aic_fn <- function(all, est_ind, om_ind = NULL){
 om_ind <- which(is.na(df.oms$NAA_sig))
 SR_rec_Mfixed = which(use.df.ems$re_config == "rec" & use.df.ems$M_est == FALSE)
 temp <- t(aic_fn(all_naa_aic, SR_rec_Mfixed, om_ind = om_ind))
-out <- cbind(df.oms[om_ind,-1], temp)
-colnames(out) <- c("$\\sigma_R$", "$\\sigma_N$", "F-history", "Obs Error", "R only", "BH") 
-x = latex(out, file = here("Project_0","paper","rec_om_em_R_SR_MF_aic_table.tex"), 
-  table.env = FALSE, col.just = rep("r", dim(out)[2]), rowname = NULL)
+# out <- cbind(df.oms[om_ind,-1], temp)
+# colnames(out) <- c("$\\sigma_R$", "$\\sigma_N$", "F-history", "Obs Error", "R only", "BH") 
+# x = latex(out, file = here("Project_0","paper","rec_om_em_R_SR_MF_aic_table.tex"), 
+#   table.env = FALSE, col.just = rep("r", dim(out)[2]), rowname = NULL)
 
 #OM: just rec re, EM: Assume R only, Estimating R and SR, M estimated
 om_ind <- which(is.na(df.oms$NAA_sig))
 SR_rec_Mest = which(use.df.ems$re_config == "rec" & use.df.ems$M_est == TRUE)
-temp <- t(aic_fn(all_naa_aic, SR_rec_Mest, om_ind = om_ind))
+temp <- cbind(temp, t(aic_fn(all_naa_aic, SR_rec_Mest, om_ind = om_ind)))
[…]

### 2023-08-10 Project_0/paper/paper.tex

@@
 
+Convergence results can be useful for understanding how bad convergence
+in applications to real data might direct the practitioner to which
+alternative random effects configurations is be more appropriate. The
+type of convergence that we might use as a diagnostic is important. And
+what rule of thumb we use for certain criteria e.g.~1e-6? large SEs/CVs?
+
 Here we conduct a simulation study with operating models varying by
@@
 
+\begin{landscape}
+\begin{table}
+\caption{Distinguishing characteristics of the operating models with random effects on survival.
+Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
+Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.}
+{\footnotesize \input{naa_om_tab}}
+\end{table}
+\end{landscape}
+
 \hypertarget{operating-models-with-random-effects-on-natural-mortality}{%
@@
 
+\begin{landscape}
+\begin{table}
+\caption{Distinguishing characteristics of the operating models with random effects on natural mortality.
+Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
+Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
+For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}
+{\input{M_om_tab}}
+\end{table}
+\end{landscape}
+
 \hypertarget{operating-models-with-random-effects-on-fleet-selectivity}{%
@@
 
+\begin{landscape}
+\begin{table}
+\caption{Distinguishing characteristics of the operating models with random effects on selectivity.
+Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
+Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
+For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}
+{\input{Sel_om_tab}}
+\end{table}
+\end{landscape}
+
 \hypertarget{operating-models-with-random-effects-on-index-catchability}{%
@@
 
+\begin{landscape}
+\begin{table}
+\caption{Distinguishing characteristics of the operating models with random effects on catchability.
+Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
+Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
+For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}
+{\input{q_om_tab}}
+\end{table}
+\end{landscape}
+
 \hypertarget{estimating-models}{%
 \subsection{Estimating models}\label{estimating-models}}
+
+\begin{table}
+\caption{Distinguishing characteristics of the estimating models.}
+{\scriptsize \input{em_tab}}
+\end{table}
 
@@
 
+Summary statitics
+
+Bias
+
+Mohn's rho
+
+marginal AIC
+
+convergence types: 1) whether the optimization completed 2) nlminb
+convergence flag = 0 3) max abs gradient \textless{} 1e-6 4) No NAs for
+SE of fixed effects returned from TMB::sdreport 5) No SE NAs AND no SE
+\textgreater{} 100
+
 \hypertarget{results}{%
@@
 
-Do each of these by type of operating model (Naa, M, sel, q) Convergence
-performance
+Do each of these by type of operating model (Naa, M, sel, q)
 
-AIC performance
+\hypertarget{convergence-performance}{%
+\subsection{Convergence performance}\label{convergence-performance}}
 
-SR estimation? M estimation?
+convergence types: 1) whether the optimization completed 2) nlminb
+convergence flag = 0 3) max abs gradient \textless{} 1e-6 4) No NAs for
+SE of fixed effects returned from TMB::sdreport 5) No SE NAs AND no SE
+\textgreater{} 100
 
-Bias, Mean Square error
+Probability of convergence was better for models that assumed the
+correct source of process error, assumed M was known, and did not assume
+stock-recruit relationships.
 
-Certain basic parameters (stock-recruit pars, M, variance parameters)
-SSB, F, R
+\hypertarget{r-rs-operating-models}{%
+\subsubsection{R, R+S operating models}\label{r-rs-operating-models}}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structures R and R+S.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.}\label{naa_om_em_R_MF_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_p_convergence_meanR_M_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structures R and R+S.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.}\label{naa_om_em_R_ME_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_p_convergence_meanR_M_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structures R and R+S.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is fixed at the true value.}\label{naa_om_em_BH_MF_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_p_convergence_BH_M_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structures R and R+S.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is estimated.}\label{naa_om_em_BH_ME_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_p_convergence_BH_M_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\hypertarget{rm-operating-models}{%
+\subsubsection{R+M operating models}\label{rm-operating-models}}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+M.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.}\label{M_om_em_R_MF_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_p_convergence_meanR_M_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+M.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.}\label{M_om_em_R_ME_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_p_convergence_meanR_M_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+M.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is fixed at the true value.}\label{M_om_em_BH_MF_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_p_convergence_BH_M_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+M.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is estimated.}\label{M_om_em_BH_ME_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_p_convergence_BH_M_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\hypertarget{rsel-operating-models}{%
+\subsubsection{R+Sel operating models}\label{rsel-operating-models}}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+Sel.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.}\label{Sel_om_em_R_MF_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_p_convergence_meanR_M_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+Sel.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.}\label{Sel_om_em_R_ME_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_p_convergence_meanR_M_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+Sel.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is fixed at the true value.}\label{Sel_om_em_BH_MF_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_p_convergence_BH_M_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+Sel.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is estimated.}\label{Sel_om_em_BH_ME_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_p_convergence_BH_M_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\hypertarget{rq-operating-models}{%
+\subsubsection{R+q operating models}\label{rq-operating-models}}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+q.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.}\label{q_om_em_R_MF_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_p_convergence_meanR_M_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+q.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.}\label{q_om_em_R_ME_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_p_convergence_meanR_M_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+q.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is fixed at the true value.}\label{q_om_em_BH_MF_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_p_convergence_BH_M_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+q.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is estimated.}\label{q_om_em_BH_ME_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_p_convergence_BH_M_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\hypertarget{aic-performance-for-process-error-structure}{%
+\subsection{AIC performance for process error
+structure}\label{aic-performance-for-process-error-structure}}
 
@@
 \hypertarget{numbers-at-age-operating-models}{%
-\subsection{Numbers at age operating
+\subsubsection{Numbers at age operating
 models}\label{numbers-at-age-operating-models}}
@@
 \hypertarget{estimating-models-include-alternative-random-effects-options-naa-m-sel-q}{%
-\subsubsection{Estimating models include alternative random effects
+\subsection{Estimating models include alternative random effects
 options: NAA, M, sel,
@@
 \clearpage
-
-\begin{landscape}
-\begin{table}
-\caption{Distinguishing characteristics of the operating models with random effects on survival.
-Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
-Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.}
-{\footnotesize \input{naa_om_tab}}
-\end{table}
-
-\begin{table}
-\caption{Distinguishing characteristics of the operating models with random effects on natural mortality.
-Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
-Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
-For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}
-{\input{M_om_tab}}
-\end{table}
-
-\begin{table}
-\caption{Distinguishing characteristics of the operating models with random effects on selectivity.
-Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
-Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
-For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}
-{\input{Sel_om_tab}}
-\end{table}
-
-\begin{table}
-\caption{Distinguishing characteristics of the operating models with random effects on catchability.
-Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
-Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
-For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}
-{\input{q_om_tab}}
-\end{table}
-\end{landscape}
-
-\begin{table}
-\caption{Distinguishing characteristics of the estimating models.}
-{\scriptsize \input{em_tab}}
-\end{table}
 
@@
 \clearpage
+
+\clearpage
+
+\hypertarget{aic-performance-for-stock-recruit-relationship}{%
+\subsection{AIC performance for stock-recruit
+relationship}\label{aic-performance-for-stock-recruit-relationship}}
+
+\hypertarget{r-rs-operating-models-1}{%
+\subsubsection{R, R+S operating models}\label{r-rs-operating-models-1}}
+
+\begin{table}
+\caption{Operating models and estimation models all assume matching R or R+S process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}
+{\input{naa_om_em_R_BH_aic_table}}
+\end{table}
+
+\begin{figure}
+\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
+Operating and estimating models have matching R or R+S process error structures.
+Estimating models assume mean M is fixed at the true value.}\label{naa_om_MF_BH_glm_AIC_plots}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_MF_pred_BH_best.png}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB. Estimating models allow estimation of mean M.}\label{naa_om_ME_BH_glm_AIC_plots}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_ME_pred_BH_best.png}
+\end{center}
+\end{figure}
+
+\hypertarget{rm-operating-models-1}{%
+\subsubsection{R+M operating models}\label{rm-operating-models-1}}
+
+\begin{table}
+\caption{Operating models and estimation models all assume matching R+M process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}
+{\input{M_om_em_R_BH_aic_table}}
+\end{table}
+
+\begin{figure}
+\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
+Operating and estimating models have matching R+M process error structures.
+Estimating models assume mean M is fixed at the true value.}\label{M_om_MF_BH_glm_AIC_plots}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_MF_pred_BH_best.png}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
+Operating and estimating models have matching R+M process error structures.
+Estimating models allow estimation of mean M.}\label{M_om_ME_BH_glm_AIC_plots}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_ME_pred_BH_best.png}
+\end{center}
+\end{figure}
+
+\hypertarget{rsel-operating-models-1}{%
+\subsubsection{R+Sel operating models}\label{rsel-operating-models-1}}
+
+\begin{table}
+\caption{Operating models and estimation models all assume matching R+Sel process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}
+{\input{Sel_om_em_R_BH_aic_table}}
+\end{table}
+
+\begin{figure}
+\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
+Operating and estimating models have matching R+Sel process error structures.
+Estimating models assume mean M is fixed at the true value.}\label{Sel_om_MF_BH_glm_AIC_plots}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_MF_pred_BH_best.png}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
+Operating and estimating models have matching R+Sel process error structures.
+Estimating models allow estimation of mean M.}\label{Sel_om_ME_BH_glm_AIC_plots}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_ME_pred_BH_best.png}
+\end{center}
+\end{figure}
+
+\hypertarget{rq-operating-models-1}{%
+\subsubsection{R+q operating models}\label{rq-operating-models-1}}
+
+\begin{table}
+\caption{Operating models and estimation models all assume matching R+q process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}
+{\input{q_om_em_R_BH_aic_table}}
+\end{table}
+
+\begin{figure}
+\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
+Operating and estimating models have matching R+q process error structures.
+Estimating models assume mean M is fixed at the true value.}\label{q_om_MF_BH_glm_AIC_plots}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_MF_pred_BH_best.png}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
+Operating and estimating models have matching R+q process error structures.
+Estimating models allow estimation of mean M.}\label{q_om_ME_BH_glm_AIC_plots}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_ME_pred_BH_best.png}
+\end{center}
+\end{figure}
+
+\clearpage
+
+\hypertarget{bias-mean-square-error}{%
+\subsection{Bias, Mean Square error}\label{bias-mean-square-error}}
+
+Certain basic parameters (stock-recruit pars, M, variance parameters)
+SSB, F, R, BRPs
 
@@
 
-\hypertarget{estimating-models-include-naa-random-effects-and-estimation-assumes-mean-r-or-bh-sr}{%
-\subsubsection{Estimating models include NAA random effects and
-estimation assumes mean R or BH
-SR}\label{estimating-models-include-naa-random-effects-and-estimation-assumes-mean-r-or-bh-sr}}
-
-\clearpage
-
-\begin{table}
-\caption{Operating models and estimation models all assume RE on recruitment only, estimating models assume mean recruitment or a B-H stock recruit relationship and M is fixed at the true value.}
-{\input{rec_om_em_R_SR_MF_aic_table}}
-\end{table}
-
-\begin{table}
-\caption{Operating models and estimation models all assume RE on recruitment only, estimating models assume mean recruitment or a B-H stock recruit relationship and M is estimated.}
-{\input{rec_om_em_R_SR_ME_aic_table}}
-\end{table}
-
-\begin{table}
-\caption{Operating models and estimation models all assume RE on all abundances at age, estimating models assume mean recruitment or a B-H stock recruit relationship and M is fixed at the true value.}
-{\input{naa_om_em_R_SR_MF_aic_table}}
-\end{table}
-
-\begin{table}
-\caption{Operating models and estimation models all assume RE on all abundances at age, estimating models assume mean recruitment or a B-H stock recruit relationship and M is estimated.}
-{\input{naa_om_em_R_SR_ME_aic_table}}
-\end{table}
-\clearpage
-
-\begin{figure}
-\caption{Predicted probability of AIC being better for BH model.}\label{glm_res_NAA_re}
-\begin{center}
-\includegraphics[width = \textwidth]{pred_SR_best_NAA_oms.png}
-\end{center}
-\end{figure}
-
-\hypertarget{m-operating-models}{%
-\subsection{M operating models}\label{m-operating-models}}
-
-\hypertarget{estimating-models-include-naa-random-effects-and-estimation-assumes-mean-r-or-bh-sr-1}{%
-\subsubsection{Estimating models include NAA random effects and
-estimation assumes mean R or BH
-SR}\label{estimating-models-include-naa-random-effects-and-estimation-assumes-mean-r-or-bh-sr-1}}
-
-\begin{table}
-\caption{M random effects operating models.}
-{\input{M_om_em_R_BH_aic_table}}
-\end{table}
-\clearpage
-
 \hypertarget{discussion}{%

### 2023-09-01 Project_0/paper/paper.Rmd
update project 0 WP

@@
 author: 
-  - Timothy J. Miller^1^, Britten, Brooks, Fay, Legault, Stock, et al.
+  - Timothy J. Miller^1^
+  - A. Applegate 
+  - G. Britten
+  - E. N. Brooks
+  - G. Fay
+  - A. Hansell
+  - C. M. Legault
+  - B. Muffley
+  - J. Wiedenmann
 date: ''
@@
 
-# Introduction
+# Introduction {-}
 
 Application of state-space models in fisheries stock assessment and management has expanded dramatically within ICES, Canada and the Northeast US \citep{nielsenberg14,cadigan16,stockmiller21}.
+State-space approaches that use random effects to parameterize process errors is considered best practice and a requirement for the next generation of stock assessment models.
 
 Much is known about the reliability of state-space models that are linear or Gaussian \citep{aeberhardetal18}, but applications in fisheries management are nonlinear and typically include multiple types of observations with varying distributional assumptions.
-We know relatively little about the statistical reliability of such models.
-Also, there is a wide range of potential random effects structures in assessment models and we  know little about the ability of information criteria to distinguish among such alternative structures.
+Furthermore, there is a wide range of potential random effects structures and model parameters that can be treated as random effects in assessment models.
+We know relatively little about the statistical reliability of such models or the ability of information criteria to distinguish among such alternative structures.
 
-But those studies focus primarily on Gaussian linear models.
+<!-- But those studies focus primarily on Gaussian linear models.
 review literature on reliability of hidden/latent process models. Primarily in other fields.
+-->
 
@@
 
-# Methods
+# Methods {-}
 
 Used the WHAM package  \citep[][commit 77bbd94]{stockmiller21} (Miller and Stock 2020).
-This packages has also been used to configure operating and estimating models for closed loop simulations evaluating index-based assessment methods \citep{legaultetalinpress} and is used for management of haddock, butterfish, plaice, bluefish in the Northeast US.
+This packages has also been used to configure operating and estimating models for closed loop simulations evaluating index-based assessment methods \citep{legaultetal23} and is used for management of haddock, butterfish, plaice, bluefish in the Northeast US.
 
-We completed a simulation study with a number of operating models that can be categorized based on where random effects are assumed: abundance at age, natural mortality, fleet selectivity, or index catchability.
+We completed a simulation study with a number of operating models that can be categorized based on where process error random effects are assumed: abundance at age (R, R+S), natural mortality (R+M), fleet selectivity (R+Sel), or index catchability (R+q).
 For each operating model assumptions about variance of process errors and observations are required and the values we used were based on a review of the range of estimates from recent applications of WHAM in management of stocks of haddock, butterfish, and American plaice in the NE US.
 
+78 total operating models.
 We simulated 100 data sets for each operating model.
-For each simulated data from each operating model we fit a set of estimating models.
+For each simulated data set we fit a set of 20 estimating models.
 
-Y estimating models fit to each
+## Operating models {-}
 
-## Operating models
+### Population {-}
 
-common to all:   
-  ages = 1 to 10+, 
-  M
-  maturity
+The population model tracks 10 age classes: ages 1 to 10+ and we assume spawning occurs 1/4 of the way through the year.
+The maturity at age was a logistic curve with a50 = 2.89 and slope = 0.88 and assumed known in all estimation models (Figure \ref{om_maturity}).
 
-marginal standard deviations for random effects are defined in tables of operating models.
+\begin{figure}
+\caption{The proportion mature at age assumed for the population in all operating and estimating models.}\label{om_maturity}
+\begin{center}
+\includegraphics[width = \textwidth]{om_maturity.png}
+\end{center}
+\end{figure}
 
-### Population
+Weight at age was generated with a LVB growth function
+$$
+L_a = L_{\infty}\left(1 - e^{-k(a - t_0)}\right)
+$$
+where $t_0 = 0$, $L_\infty = 85$, and $k = 0.3$, and a L-W relationship such that 
+$$
+W_a = \theta_1 L_a^{\theta_2}
+$$
+where $\theta_1 = e^{-12.1}$ and $\theta_2 = 3.2$ (Figure \ref{om_waa}).
 
-There are 10 age classes: ages 1 to 10+.
-
-Spawning was assumed to occur 1/4 of the way through the year.
-
-Natural mortality rate was assumed 0.2 when it was constant and the mean of the time series process for operating models with M random effects.
-maturity a50 = 2.89, slope = 0.88
-
-Weight at age was generated with a LVB growth
-\[
-L_a = L_{\infty}\left(1 - e^{-k(a - t_0)}\right)
-\]
-where $t_0 = 0$, $L_\infty = 85$, and $k = 0.3$, and a L-W relationship such that 
-\[
-W_a = \theta_1 L_a^{\theta_2}
-\]
-where $\theta_1 = e^{-12.1}$ and $\theta_2 = 3.2$.
+\begin{figure}
+\caption{The weight at age assumed for the population in all operating and estimating models.}\label{om_waa}
+\begin{center}
+\includegraphics[width = \textwidth]{om_waa.png}
+\end{center}
+\end{figure}
 
@@
 All post-recruit productivity components are constant in the NAA and survey catchability process error operating models.
-Therefore steepness and unfished recruitment are also constant over the time period for those operating models (Miller and Brooks 2021).
-We specified unfished recruitment = $R_0 = e^{10}$ and $F_{\text{MSY}} = F_{40} = 0.348$ equated to a steepness of 0.69 and $\alpha=0.60$ and $\beta = 2.4 \times 10^{-5}$ for the
+Therefore steepness and unfished recruitment are also constant over the time period for those operating models \citep{millerbrooks21}.
+We specified unfished recruitment = $R_0 = e^{10}$ and $\Fmsy = \Fspr[40] = 0.348$ equated to a steepness of 0.69 and $\alpha=0.60$ and $\beta = 2.4 \times 10^{-5}$ for the
 \[
@@
 \]
-Beverton-Holt parameterization.
+Beverton-Holt parameterization (Figure \ref{om_sr}).
+For OMs without process errors on natural mortality we assumed the rate was assumed 0.2.
+For OMs with process errors on natural mortality the mean rate was 0.2.
 
-The magnitude of the overfishing assumptions is based on average estimates of overfishing for NE groundfish stocks from Wiedenmann et al.
-(20XX).
-Legault et al.
-(2023) also used similar approaches to defining fishing mortality histories for operating models.
+\begin{figure}
+\caption{The Beverton-Holt stock recruit relationship assume for all operating models.}\label{om_sr}
+\begin{center}
+\includegraphics[width = \textwidth]{om_sr.png}
+\end{center}
+\end{figure}
 
-Currenlty, initial population is configured at the equilibrium distribution fishing at $F = 2.5\times F_{\text{MSY}}$.
+Two alternative fishing histories were used for operating models.
+In the first scenario, the stock experiences overfishing for the first 20 years and fishing at \Fmsy for the last 20 years.
+In the second scenario, the stock is fished at \Fmsy for the entire time period.
+The magnitude of the overfishing assumptions is based on average estimates of overfishing for NE groundfish stocks from \citet{wiedenmannetal19}.
+\Citet{legaultetal23} also used similar approaches to defining fishing mortality histories for operating models.
 
-Initial population was configured at the equilibrium distribution fishing at either $F = 2.5\times F_{\text{MSY}}$ or $F = F_{\text{MSY}}$ for the two alternative fishing histories.
+Initial population was configured at the equilibrium distribution fishing at either $F = 2.5\times \Fmsy$ or $F = \Fmsy$ for the two alternative fishing histories.
 That is for a deterministic model, the age composition would not change over time when the fishing mortality was constant at the respective level.
@@
 For operating models with time-varying random effects for M, steepness is not constant, but we used the same alpha and beta parameters as other operating models this equates to a steepness and R0 at the mean of the time series process for M.
-For operating models with time-varying random effects for fishery selectivity, Fmsy is also not constant however we use the same F history as other operating models which corresponds to Fmsy at the mean selectivity parameters.
+For operating models with time-varying random effects for fishery selectivity, \Fmsy is also not constant however we use the same F history as other operating models which corresponds to Fmsy at the mean selectivity parameters.
 
-### Fleets
+### Fleets {-}
 
-We assumed a single fleet operating year round for catch observations with logistic selectivity for the fleet with $a_{50} = 5$ and slope = 1.
-This selectivity is was used to define $F_{\text{MSY}}$ for the Beverton-Holt stock recruitment parameters above.
+We assumed a single fleet operating year round for catch observations with logistic selectivity for the fleet with $a_{50} = 5$ and slope = 1 (Figure \ref{om_mean_selectivity}).
+This selectivity is was used to define \Fmsy for the Beverton-Holt stock recruitment parameters above.
 We assumed a logistic-normal distribution for the age-composition observations for the fleet.
 
-### Fishing histories
-All operating models assumed one of two different fishing histories. 
-One : Fishing mortality is equal to Fmsy (0.348) for the whole 40 year period.
-Two : Fishing mortality is 2.5 times Fmsy for the first 20 years then changes to Fmsy for the last 20 years.
+\begin{figure}
+\caption{The selectivity at age assumed for the fishing fleet in operating models without selectivity random effects.
+The mean selectivity at age across time for the fishing fleet in operating models with selectivity random effects.}\label{om_mean_selectivity}
+\begin{center}
+\includegraphics[width = \textwidth]{om_mean_selectivity.png}
+\end{center}
+\end{figure}
 
-### Indices
+### Indices {-}
+
 Two time series of surveys are assumed and observed in numbers rather than biomass for the entire 40 year period with one occurring in the spring (0.25 way through the year) and one in the fall (0.75 way through the year).
@@
 Catchability of both surveys are assumed to be 0.1.
-We assumed logistic selectivity for both indices with $a_{50} = 5$ and slope = 1.
-We assumed a logistic-normal distribution for the age-composition observations.
+Like the fishing fleet, we assumed logistic selectivity for both indices with $a_{50} = 5$ and slope = 1 and a logistic-normal distribution for the age-composition observations.
 
-### Observation Uncertainty
+### Observation Uncertainty {-}
 
@@
 
-### Operating models with random effects on numbers at age
+### Operating models with random effects on numbers at age {-}
 
-24 operating models, 16 Sel re operating models and 16 q re operating models.
-Table of process error assumptions
+The factorial combinations of fishing histories, observation error assumptions, and marginal variance assumptions for recruitment and survival results in 24 different operating models with random effects on recruitment and survival (Table \ref{naa_om_table}).
 
@@
 Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
-Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.}
+Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.}\label{naa_om_table}
 {\footnotesize \input{naa_om_tab}}
@@
 
-### Operating models with random effects on natural mortality
+### Operating models with random effects on natural mortality {-}
 
-16 operating models
-Table of process error assumptions
-
-NOTE: inv_trans_rho function in set_M.R is mis-defined. Will affect correlation parameters assigned in operating models?
-
-Steepness and BRPs are not constant when M is time-varying (Miller and Brooks 2021). We uses the a/b parameters for the B-H defined for Fmsy = F40 at the mean M = 0.2 as defined above.
+The factorial combinations of fishing histories, observation error assumptions, and marginal variance and autocorrelation assumptions for natural mortality results in 16 different operating models with random effects on natural mortality (Table \ref{M_om_table}).
+Internally, WHAM treats natural mortality as a log-transformed parameter
+$$
+\log M_y = \mu_M + \epsilon_{M,y}
+$$
+that is a linear combination of a mean $\mu_M$ and any annual random effects marginally distributed as $\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)$.
+The values for the marginal variance and correlation parameters assumed in the operating models are given in Table \ref{M_om_table}.
 
@@
 Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
-For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}
+For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}\label{M_om_table}
 {\input{M_om_tab}}
@@
 
-### Operating models with random effects on fleet selectivity
+### Operating models with random effects on fleet selectivity {-}
 
-16 operating models
-Table of process error assumptions
-
-BRPs are not constant when fleet selectivity is time-varying.
-We uses the a/b parameters for the B-H defined for Fmsy = F40 at the mean of the time series model for selectivity which is the same as the constant selectivity defined above.
+The factorial combinations of fishing histories, observation error assumptions, and marginal variance and autocorrelation assumptions for selectivity parameters results in 16 different operating models with random effects on selectivity (Table \ref{sel_om_table}).
+Internally, WHAM treats selectivity parameter $s$ as a logit-transformed parameter
+$$
+\log\left(\frac{p_{s,y}-l_{s}}{u_{s}-p_{s,y}}\right) = \mu_s + \epsilon_{s,y}
+$$
+that is a linear combination of a mean $\mu_s$ and any annual random effects marginally distributed as $\epsilon_{s,y} \sim \text{N}\left(0,\sigma_s^2\right)$ where the lower and upper bounds of the parameter ($l_s$ and $u_s$) can be specified by the user.
+All selectivity parameters are either $a_50$ or slope parameters and we assume bounds of 0 and 10 for all selectivity parameters for all operating and estimating models.
+The values for the marginal variance and correlation parameters assumed in the operating models are given in Table \ref{sel_om_table}.
 
@@
 Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
-For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}
+For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}\label{sel_om_table}
 {\input{Sel_om_tab}}
@@
 
-### Operating models with random effects on index catchability
+### Operating models with random effects on index catchability {-}
 
-16 operating models
-Table of process error assumptions
+The factorial combinations of fishing histories, observation error assumptions, and marginal variance and autocorrelation assumptions for catchability results in 16 different operating models with random effects on catchability (Table \ref{q_om_table}).
+Internally, WHAM treats catchability for index $i$ as a logit-transformed parameter
+$$
+\log\left(\frac{q_{i,y}-l_{i}}{u_{i}-q_{i,y}}\right) = \mu_i + \epsilon_{i,y}
+$$
+that is a linear combination of a mean $\mu_i$ and any annual random effects marginally distributed as $\epsilon_{i,y} \sim \text{N}\left(0,\sigma_i^2\right)$ where the lower and upper bounds of the catchability ($l_i$ and $u_i$) can be specified by the user.
+Here we assume bounds of 0 and 1000 for all operating and estimating models.
+For operating and estimation models with process errors on catchability, the temporal variation is only assumed for the first index.
+The values for the marginal variance and correlation parameters assumed in the operating models are given in Table \ref{q_om_table}
 
@@
 Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
-For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}
+For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}\label{q_om_table}
 {\input{q_om_tab}}
@@
 
-## Estimating models
+## Estimating models {-}
+
+For each data set simulated from an operating model 20 estimating models were fit.
+A total of 32 different estimating models were fit across all operating models where the subset of 20 depended on the source of process error (Table \ref{em_table}).
+The first 20 estimating models in Table \ref{em_table} were fit to simulate data sets from R and R+S operating models.
+Estimating models 5 to 24 in Table \ref{em_table} were fit to simulate data sets from R+M operating models.
+Estimating models 5 to 20 and 25-28 in Table \ref{em_table} were fit to simulate data sets from R+Sel operating models.
+Finally, estimating models 5 to 20 and 29-32 in Table \ref{em_table} were fit to simulate data sets from R+q operating models.
 
 \begin{table}
-\caption{Distinguishing characteristics of the estimating models.}
+\caption{Distinguishing characteristics of the estimating models.}\label{em_table}
 {\scriptsize \input{em_tab}}
@@
 
-32 estimating models
-Table of estimating models
-
-1-20 fit to each NAA RE operating model
-5-24 fit to each M RE operating model
-5-20,25-28 to each sel RE operating model
-5-20, 29-32 to each q RE operating model
-
 SR estimation or not
 
-Make plot of S-R curve, Fmsy = F40
 Initial values for BH parameters are the true values.
@@
 
-Summary statitics
+## Measures of reliability {-}
 
-Bias
+The first measure of reliability we investigated was frequency of convergence when fitting each estimating model to the simulated data sets.
+There are various ways to assess convergence of the fit.
+We summarized 5 alternative categories of convergence.
+1. Did the optimization routine (stats::nlminb) complete without error?
+2. Did the stats::nlminb convergence flag = 0 indicate successful convergence?
+3. Was the maximum absolute value of the gradient of the log-likelihood < $1\times10^{-6}$?
+4. Did TMB::sdreport provide non-NA values for all fixed effects standard errors?
+5. Did TMB::sdreport provide all standard errors < 100?
 
-Mohn's rho
+The first convergence criterion assesses whether the model crashes.
+The third is a measure that assesses how flat the likelihood is at the optimized point of the likelihood surface.
+The fourth and fifth criteria are specific to the hessian-based standard error reporting provided by TMB.
+the TMB::sdreport function will sometimes return standard error estimates even when the calculated hessian is not invertible (fourth criterion).
+It will also provide large standard errors for some parameters that are not estimated well or are near bounds on the transformed scale that is of primary interest.
+For example, variance parameters for random effects can often be estimated near 0 when the fit suggests no variation in the random effects or, equivalently, that the model without random effects is better.
 
-marginal AIC
+### AIC for model selection {-}
 
-convergence types: 
-1) whether the optimization completed
-2) nlminb convergence flag = 0
-3) max abs gradient < 1e-6
-4) No NAs for SE of fixed effects returned from TMB::sdreport
-5) No SE NAs AND no SE > 100
+We measured the frequency of correct model selection using marginal AIC.
+For a given operating model we the set of models that were considered all made the same assumptions on whether or not to estimate M and whether a stock-recruit function is assumed.
 
-# Results
+We also measured the frequency of correctly selecting models with Beverton-Holt stock recruit function assumed over models without the stock-recruit function (null model).
+Because the proportions of simulations where the correct model was chosen for a given set of operating model factors implied generally poor performance of AIC, we analyzed the effect of variation in SSB on the probability of lower AIC of the BH models.
+We fit binomial generalize linear models to the observed AIC choice for each data set as a function of log(SD(SSB)) and operating model.
 
-Do each of these by type of operating model (Naa, M, sel, q)
+### Bias {-}
 
-## Convergence performance
+We estimated relative bias 
+$$
+\widehat {\text{RB}}\left(\theta\right) = \frac{1}{N} \sum^N_{i = 1} \frac{\widehat \theta_i - \theta_i}{\theta_i}
+$$
+of annual SSB and fully-selected fishing mortality for each estimation model and constructed 95\% confidence intervals for the median relative bias using the binomial distribution approach as in \citet{stockmiller21}.
+We also estimated bias of stock-recruit parameters for estimation models that assumed the Beverton-Holt stock recruit function.
 
-convergence types: 
-1) whether the optimization completed
-2) nlminb convergence flag = 0
-3) max abs gradient < 1e-6
-4) No NAs for SE of fixed effects returned from TMB::sdreport
-5) No SE NAs AND no SE > 100
+### Mohn's rho {-}
+
+We estimated Mohn's rho for SSB, fully-selected fishing mortality, and recruitment for each estimation model.
+We estimated 7 peels for each estimation model.
+We calculated median 95\% confidence intervals for Mohn's rho using the same methods as that for bias.
+
+# Results {-}
+
+## Convergence performance {-}
 
@@
 
-### R, R+S operating models
+### R, R+S operating models {-}
+
+Figures \ref{naa_om_em_R_MF_convergence} to \ref{naa_om_em_BH_ME_convergence}
 
@@
 
-### R+M operating models
+### R+M operating models {-}
+
+Figures \ref{M_om_em_R_MF_convergence} to \ref{M_om_em_BH_ME_convergence}
 
@@
 
-### R+Sel operating models
+### R+Sel operating models {-}
+
+Figures \ref{Sel_om_em_R_MF_convergence} to \ref{Sel_om_em_BH_ME_convergence}
 
@@
 
-### R+q operating models
+### R+q operating models {-}
+
+Figures \ref{q_om_em_R_MF_convergence} to \ref{q_om_em_BH_ME_convergence}
 
@@
 
-## AIC performance for process error structure
-\clearpage 
-### Numbers at age operating models
+## AIC performance for process error structure {-}
 
-## Estimating models include alternative random effects options: NAA, M, sel, q
+### R, R+S operating models {-}
+
+Figures \ref{naa_om_proportion_best_aic_R_MF} to \ref{naa_om_proportion_best_aic_SR_ME}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R and R+S process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.} \label{naa_om_proportion_best_aic_R_MF}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_proportion_best_aic_R_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R and R+S process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.} \label{naa_om_proportion_best_aic_R_ME}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_proportion_best_aic_R_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R and R+S process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate a stock-recruit relationship and and M is fixed at the true value.} \label{naa_om_proportion_best_aic_SR_MF}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_proportion_best_aic_SR_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R and R+S process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate a stock-recruit relationship and and M is estimated.} \label{naa_om_proportion_best_aic_SR_ME}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_proportion_best_aic_SR_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+### R+M operating models {-}
+
+Figures \ref{M_om_proportion_best_aic_R_MF} to \ref{M_om_proportion_best_aic_SR_ME}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+M process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.} \label{M_om_proportion_best_aic_R_MF}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_proportion_best_aic_R_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+M process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.} \label{M_om_proportion_best_aic_R_ME}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_proportion_best_aic_R_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+M process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate a stock-recruit relationship and and M is fixed at the true value.} \label{M_om_proportion_best_aic_SR_MF}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_proportion_best_aic_SR_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+M process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate a stock-recruit relationship and and M is estimated.} \label{M_om_proportion_best_aic_SR_ME}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_proportion_best_aic_SR_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+### R+Sel operating models {-}
+
+Figures \ref{Sel_om_proportion_best_aic_R_MF} to \ref{Sel_om_proportion_best_aic_SR_ME}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+Sel process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.} \label{Sel_om_proportion_best_aic_R_MF}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_proportion_best_aic_R_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+Sel process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.} \label{Sel_om_proportion_best_aic_R_ME}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_proportion_best_aic_R_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+Sel process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate a stock-recruit relationship and and M is fixed at the true value.} \label{Sel_om_proportion_best_aic_SR_MF}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_proportion_best_aic_SR_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+Sel process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate a stock-recruit relationship and and M is estimated.} \label{Sel_om_proportion_best_aic_SR_ME}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_proportion_best_aic_SR_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+### R+q operating models {-}
+
+Figures \ref{q_om_proportion_best_aic_R_MF} to \ref{q_om_proportion_best_aic_SR_ME}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+q process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.} \label{q_om_proportion_best_aic_R_MF}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_proportion_best_aic_R_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+q process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.} \label{q_om_proportion_best_aic_R_ME}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_proportion_best_aic_R_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+q process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate a stock-recruit relationship and and M is fixed at the true value.} \label{q_om_proportion_best_aic_SR_MF}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_proportion_best_aic_SR_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+q process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate a stock-recruit relationship and and M is estimated.} \label{q_om_proportion_best_aic_SR_ME}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_proportion_best_aic_SR_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+<!--
+## Estimating models include alternative random effects options: NAA, M, sel, q {-}
 \clearpage
@@
 \clearpage 
+-->
 
 \clearpage 
-## AIC performance for stock-recruit relationship
+## AIC performance for stock-recruit relationship {-}
 
-### R, R+S operating models
+Tables of frequencies where AIC chooses correct stock-recruit assumption imply poor performance of AIC, but figures showing effect of variation in SSB demonstrate that large variation in SSB is needed to reduce Type II errors for marginal AIC.
+
+### R, R+S operating models {-}
 
@@
 
-### R+M operating models
+### R+M operating models {-}
 
@@
 
-### R+Sel operating models
+### R+Sel operating models {-}
 
@@
 
-### R+q operating models
+### R+q operating models {-}
 
@@
 \clearpage 
-## Bias, Mean Square error
+## Bias {-}
 
-  Certain basic parameters (stock-recruit pars, M, variance parameters)
-  SSB, F, R, BRPs
+### R, R+S OMs {-}
 
@@
 \begin{figure}
-\caption{Median relative bias of for SSB for estimating models that estimate mean recruitment and M is fixed at the true value.}\label{naa_om_em_R_MF_relbias_ssb}
+\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R and R+S process errors.
+Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{naa_om_em_R_MF_relbias_ssb}
 \begin{center}
@@
 \begin{figure}
-\caption{Median relative bias of for SSB for estimating models that estimate mean recruitment and M is estimated.}\label{naa_om_em_R_ME_relbias_ssb}
+\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R and R+S process errors.
+Estimation models do not assume a stock-recruit relationship and estimate M.}\label{naa_om_em_R_ME_relbias_ssb}
 \begin{center}
@@
 \begin{figure}
-\caption{Median relative bias of for SSB for estimating models that estimates a BH stock-recruitment function and M is fixed at the true value.}\label{naa_om_em_SR_MF_relbias_ssb}
+\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R and R+S process errors.
+Estimation models assume a BH stock-recruitment function and M is fixed at the true value.}\label{naa_om_em_BH_MF_relbias_ssb}
 \begin{center}
-\includegraphics[width = \textwidth]{naa_om_SR_MF_relbias_ssb.png}
+\includegraphics[width = \textwidth]{naa_om_BH_MF_relbias_ssb.png}
 \end{center}
@@
 \begin{figure}
-\caption{Median relative bias of for SSB for estimating models that estimates a BH stock-recruitment function and M is estimated.}\label{naa_om_em_SR_ME_relbias_ssb}
+\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R and R+S process errors.
+Estimation models assume a BH stock-recruitment function and estimate M.}\label{naa_om_em_BH_ME_relbias_ssb}
 \begin{center}
-\includegraphics[width = \textwidth]{naa_om_SR_ME_relbias_ssb.png}
+\includegraphics[width = \textwidth]{naa_om_BH_ME_relbias_ssb.png}
 \end{center}
@@
 
-# Discussion
+### R+M OMs {-}
 
-The estimating models assumed variances of aggregate catch and index observations was known. This approximation may be appropriate for indices
-where we have a reliable estimate of uncertainty based on the survey design (), but there may be better approaches for the aggregate
-catch such as an informed prior on the standard errors with realistic bounds.
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+M process errors.
+Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{M_om_em_R_MF_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_R_MF_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+M process errors.
+Estimation models do not assume a stock-recruit relationship and estimate M.}\label{M_om_em_R_ME_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_R_ME_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+M process errors.
+Estimation models assume a BH stock-recruitment function and M is fixed at the true value.}\label{M_om_em_BH_MF_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_BH_MF_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+M process errors.
+Estimation models assume a BH stock-recruitment function and estimate M.}\label{M_om_em_BH_ME_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_BH_ME_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+### R+Sel OMs {-}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+Sel process errors.
+Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{Sel_om_em_R_MF_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_R_MF_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+Sel process errors.
+Estimation models do not assume a stock-recruit relationship and estimate M.}\label{Sel_om_em_R_ME_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_R_ME_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+Sel process errors.
+Estimation models assume a BH stock-recruitment function and M is fixed at the true value.}\label{Sel_om_em_BH_MF_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_BH_MF_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+Sel process errors.
+Estimation models assume a BH stock-recruitment function and estimate M.}\label{Sel_om_em_BH_ME_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_BH_ME_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+### R+q OMs {-}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+q process errors.
+Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{q_om_em_R_MF_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_R_MF_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+q process errors.
+Estimation models do not assume a stock-recruit relationship and estimate M.}\label{q_om_em_R_ME_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_R_ME_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+q process errors.
+Estimation models assume a BH stock-recruitment function and M is fixed at the true value.}\label{q_om_em_BH_MF_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_BH_MF_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+q process errors.
+Estimation models assume a BH stock-recruitment function and estimate M.}\label{q_om_em_BH_ME_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_BH_ME_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+## Mohn's $\rho$ {-}
+
+### R, R+S OMs {-}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R and R+S process errors.
+Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{naa_om_em_R_MF_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_mohns_rho_R_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R and R+S process errors.
+Estimation models do not assume a stock-recruit function and M estimated.}\label{naa_om_em_R_ME_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_mohns_rho_R_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R and R+S process errors.
+Estimation models assume a B-H stock-recruit function and M is fixed at the true value.}\label{naa_om_em_BH_MF_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_mohns_rho_BH_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R and R+S process errors.
+Estimation models assume a B-H stock-recruit function and M estimated.}\label{naa_om_em_BH_ME_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_mohns_rho_BH_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+### R+M OMs {-}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+M process errors.
+Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{M_om_em_R_MF_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_mohns_rho_R_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+M process errors.
+Estimation models do not assume a stock-recruit function and M estimated.}\label{M_om_em_R_ME_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_mohns_rho_R_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+M process errors.
+Estimation models assume a B-H stock-recruit function and M is fixed at the true value.}\label{M_om_em_BH_MF_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_mohns_rho_BH_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+M process errors.
+Estimation models assume a B-H stock-recruit function and M estimated.}\label{M_om_em_BH_ME_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_mohns_rho_BH_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+### R+Sel OMs {-}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+Sel process errors.
+Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{Sel_om_em_R_MF_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_mohns_rho_R_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+Sel process errors.
+Estimation models do not assume a stock-recruit function and M estimated.}\label{Sel_om_em_R_ME_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_mohns_rho_R_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+Sel process errors.
+Estimation models assume a B-H stock-recruit function and M is fixed at the true value.}\label{Sel_om_em_BH_MF_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_mohns_rho_BH_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+Sel process errors.
+Estimation models assume a B-H stock-recruit function and M estimated.}\label{Sel_om_em_BH_ME_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_mohns_rho_BH_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+### R+q OMs {-}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+q process errors.
+Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{q_om_em_R_MF_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_mohns_rho_R_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+q process errors.
+Estimation models do not assume a stock-recruit function and M estimated.}\label{q_om_em_R_ME_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_mohns_rho_R_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+q process errors.
+Estimation models assume a B-H stock-recruit function and M is fixed at the true value.}\label{q_om_em_BH_MF_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_mohns_rho_BH_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+q process errors.
+Estimation models assume a B-H stock-recruit function and M estimated.}\label{q_om_em_BH_ME_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_mohns_rho_BH_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+# Discussion {-}
+
+The estimating models assumed variances of aggregate catch and index observations was known.
+This approximation may be appropriate for indices where we have a reliable estimate of uncertainty based on the survey design (), but there may be better approaches for the aggregate catch such as an informed prior on the standard errors with realistic bounds.
 
@@
 \pagebreak
-<!-- 
-\begin{figure}
-\caption{Diagram of twin-trawl gear configuration.
-One of the two nets is rigged with a rockhopper sweep (8) and the other is rigged with a chain sweep (7) and for both a restrictor rope (5) is used to obtain consistent net spread.
-The other importand components are the side wires (1), middle wire (2), doors (3), the clump weight (4), and the acoustic mensuration system (6).
-The side where the rockhopper and chainsweep gears gears were deployed varied throughout the experimental tows of each.}\label{twin_trawl_diagram}
-\begin{center}
-\includegraphics[width = \textwidth]{twin_trawl_diagram.pdf}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\caption{The \textsl{F/V Karen Elizabeth} twin-trawl vessel rigged with rockhopper sweep gear on the right and chain sweep gear on the left.}\label{twin_trawl_photo}
-\begin{center}
-\includegraphics[width = \textwidth]{twin_trawl_photo_small.png}
-\end{center}
-\end{figure} -->
-
-<!--
-\begin{figure}
-\caption{Diagram of the standard Northeast Fisheries Science Center rockhopper sweep center and wing sections.}\label{rockhopper_schematic}
-\begin{center}
-\includegraphics[width = 0.7\textwidth]{rockhopper_schematic_1.pdf}
-\includegraphics[width = 0.7\textwidth]{rockhopper_schematic_2.pdf}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\caption{Diagram of the chain sweep designed maximize bottom contact and flatfish capture.}\label{chainsweep_schematic}
-\begin{center}
-\includegraphics[width = \textwidth]{chainsweep_schematic.pdf}
-\end{center}
-\end{figure}
-\clearpage
-  -->
-<!-- \begin{landscape}
-\begin{figure}
-\caption{Annual locations of stations where the F/V Karen Elizabeth conducted twin-trawl sets with the standard bottom trawl gear and the gear with a chain sweep instead of the rockhopper sweep.}\label{tow_locations}
-\begin{center}
-\includegraphics[width = \textwidth]{TwinTrawlLocationsAllYears.jpg}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\clearpage
-
-\begin{figure}
-\caption{Relative efficiency  of gears using chain and rockhopper sweeps from the best performing model for each species (Table \ref{best_model_compare}).
-Blue and red denote results for day and night data, respectively, and thick and thin lines represent overall and paired-tow specific estimates of relative catch efficiency, respectively.
-There was no diel effect in the best model for American plaice.
-Points represent empirical estimates of relative efficiency for paired observation by length and paired tow.
-Polygons and dashed lines represent hessian-based and bootstrap-based 95\% confidence intervals, respectively.}\label{sp_rho_plot}
-\begin{center}
-\includegraphics[height = 0.8\textheight]{sp_rho_plot.pdf}
-\end{center}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-\caption{Annual spring (blue) and fall (red) biomass estimates for each managed stock assuming 100\% efficiency for chain sweep gear with shaded polygons representing bootstrap-based 95\% confidence intervals.
-Relative catch efficiency at size estimates and bootstraps are from the best performing model for each species (Table \ref{best_model_compare}).}\label{stock_biomass_plot}
-\begin{center}
-\includegraphics[height = 0.8\textheight]{stock_biomass_plot.pdf}
-\end{center}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-\caption{Implied catch efficiency of annual spring (blue) and fall (red) bottom trawl survey biomass estimates for each managed stock assuming 100\% efficiency for chain sweep gear with shaded polygons representing bootstrap-based 95\% confidence intervals.
-Relative catch efficiency at size estimates and bootstraps are from the best performing model for each species (Table \ref{best_model_compare}).}\label{stock_biomass_efficiency_plot}
-\begin{center}
-\includegraphics[height = 0.8\textheight]{stock_biomass_efficiency_plot.pdf}
-\end{center}
-\end{figure} -->
-
-\clearpage
-
-<!-- \begin{table}
-\caption{Managed stocks associated with the species for which relative catch efficiency was estimated.}\label{stock_definition_table}
-{\input{stock_definition_table}}
-\end{table}
-
-\begin{landscape}
-\begin{table}
-\caption{Description of relative catch efficiency ($\rho$) and beta-binomial dispersion ($\phi$) parameterizations for binomial and beta-binomial models and number of marginal likelihood parameters ($n_p$) for the 13 base models from \citet{miller13} and fit to paired chain sweep and rockhoppersweep tow data for each species.}\label{base_model_description}
-{\input{base_model_description}}
-\end{table}
-\end{landscape}
-
-\begin{landscape}
-\begin{table}
-\caption{Number of paired tows where fish were captured and the number of fish captured and measured for lengths for each species in total and by day or night.}\label{data_table}
-{\changesize{10}\input{data_table}}
-\end{table}
-\end{landscape}
-
-\begin{landscape}
-\begin{table}\caption{Difference in AIC for each of the 13 models described in Table \ref{base_model_description} from the best model ($\bf 0$) by species.}\label{base_model_compare}
-{\input{base_model_compare}}
-\end{table}
-\end{landscape}
-
-\begin{landscape}
-\begin{table}\caption{Best performing models from Table \ref{base_model_compare} and extended models that include diel effects on relative catch efficiency for each species with the number of parameters for each model ($n_p$) and the differences in AIC ($\Delta$AIC) from the best of the three models ($\bf 0$) by species.}\label{best_model_compare}
-{\changesize{7}\input{best_model_compare}}
-\end{table}
-\end{landscape}
-
-\begin{table}
-\caption{Average of annual (2009-2019) ratios of coefficients of variation for calibrated and uncalibrated biomass indices for each stock by seasonal survey.
-Coefficients of variation are based on bootstrap resampling of paired tow observations, survey station data and associated length and weight observations.
-Annual indices for fall 2017 were not available for summer flounder, SNE-MA windowpane, and SNE-MA yellowtail flounder.}\label{stock_cv_ratios}
-{\input{stock_cv_ratios}}
-\end{table}
-
-\begin{table}
-\caption{Average correlation of annual (2009-2019) calibrated biomass indices for each stock by seasonal survey.
-Annual indices for fall 2017 were not available for SNE-MA windowpane and SNE-MA yellowtail flounder.}\label{stock_mean_cor}
-{\input{stock_mean_cor}}
-\end{table} -->

### 2023-09-01 Project_0/paper/paper.tex

@@
   pdftitle={Factors affecting reliablity of state-space age-structured assessment models},
-  pdfauthor={Timothy J. Miller1, Britten, Brooks, Fay, Legault, Stock, et al.},
+  pdfauthor={Timothy J. Miller1; A. Applegate; G. Britten; E. N. Brooks; G. Fay; A. Hansell; C. M. Legault; B. Muffley; J. Wiedenmann},
   hidelinks,
@@
 \usepackage{relsize}
+\newcommand{\Fmsy}{\ensuremath{F_{\text{MSY}}}\xspace}
+\newcommand{\Fspr}[1]{\ensuremath{F_{\text{{#1}\%}}}\xspace}
 \newcommand{\afrb}{Alaska Fishery Research Bulletin\xspace}
@@
 assessment models}
-\author{Timothy J. Miller\textsuperscript{1}, Britten, Brooks, Fay,
-Legault, Stock, et al.}
+\author{Timothy J. Miller\textsuperscript{1} \and A. Applegate \and G.
+Britten \and E. N. Brooks \and G. Fay \and A. Hansell \and C. M.
+Legault \and B. Muffley \and J. Wiedenmann}
 \date{}
@@
 \hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+\section*{Introduction}\label{introduction}}
+\addcontentsline{toc}{section}{Introduction}
 
@@
 management has expanded dramatically within ICES, Canada and the
-Northeast US \citep{nielsenberg14,cadigan16,stockmiller21}.
+Northeast US \citep{nielsenberg14,cadigan16,stockmiller21}. State-space
+approaches that use random effects to parameterize process errors is
+considered best practice and a requirement for the next generation of
+stock assessment models.
 
@@
 fisheries management are nonlinear and typically include multiple types
-of observations with varying distributional assumptions. We know
-relatively little about the statistical reliability of such models.
-Also, there is a wide range of potential random effects structures in
-assessment models and we know little about the ability of information
-criteria to distinguish among such alternative structures.
-
-But those studies focus primarily on Gaussian linear models. review
-literature on reliability of hidden/latent process models. Primarily in
-other fields.
+of observations with varying distributional assumptions. Furthermore,
+there is a wide range of potential random effects structures and model
+parameters that can be treated as random effects in assessment models.
+We know relatively little about the statistical reliability of such
+models or the ability of information criteria to distinguish among such
+alternative structures.
 
@@
 \hypertarget{methods}{%
-\section{Methods}\label{methods}}
+\section*{Methods}\label{methods}}
+\addcontentsline{toc}{section}{Methods}
 
@@
 and estimating models for closed loop simulations evaluating index-based
-assessment methods \citep{legaultetalinpress} and is used for management
-of haddock, butterfish, plaice, bluefish in the Northeast US.
+assessment methods \citep{legaultetal23} and is used for management of
+haddock, butterfish, plaice, bluefish in the Northeast US.
 
 We completed a simulation study with a number of operating models that
-can be categorized based on where random effects are assumed: abundance
-at age, natural mortality, fleet selectivity, or index catchability. For
-each operating model assumptions about variance of process errors and
-observations are required and the values we used were based on a review
-of the range of estimates from recent applications of WHAM in management
-of stocks of haddock, butterfish, and American plaice in the NE US.
-
-We simulated 100 data sets for each operating model. For each simulated
-data from each operating model we fit a set of estimating models.
-
-Y estimating models fit to each
+can be categorized based on where process error random effects are
+assumed: abundance at age (R, R+S), natural mortality (R+M), fleet
+selectivity (R+Sel), or index catchability (R+q). For each operating
+model assumptions about variance of process errors and observations are
+required and the values we used were based on a review of the range of
+estimates from recent applications of WHAM in management of stocks of
+haddock, butterfish, and American plaice in the NE US.
+
+78 total operating models. We simulated 100 data sets for each operating
+model. For each simulated data set we fit a set of 20 estimating models.
 
 \hypertarget{operating-models}{%
-\subsection{Operating models}\label{operating-models}}
-
-common to all:\\
-ages = 1 to 10+, M maturity
-
-marginal standard deviations for random effects are defined in tables of
-operating models.
+\subsection*{Operating models}\label{operating-models}}
+\addcontentsline{toc}{subsection}{Operating models}
 
 \hypertarget{population}{%
-\subsubsection{Population}\label{population}}
-
-There are 10 age classes: ages 1 to 10+.
-
-Spawning was assumed to occur 1/4 of the way through the year.
-
-Natural mortality rate was assumed 0.2 when it was constant and the mean
-of the time series process for operating models with M random effects.
-maturity a50 = 2.89, slope = 0.88
-
-Weight at age was generated with a LVB growth \[
+\subsubsection*{Population}\label{population}}
+\addcontentsline{toc}{subsubsection}{Population}
+
+The population model tracks 10 age classes: ages 1 to 10+ and we assume
+spawning occurs 1/4 of the way through the year. The maturity at age was
+a logistic curve with a50 = 2.89 and slope = 0.88 and assumed known in
+all estimation models (Figure \ref{om_maturity}).
+
+\begin{figure}
+\caption{The proportion mature at age assumed for the population in all operating and estimating models.}\label{om_maturity}
+\begin{center}
+\includegraphics[width = \textwidth]{om_maturity.png}
+\end{center}
+\end{figure}
+
+Weight at age was generated with a LVB growth function \[
 L_a = L_{\infty}\left(1 - e^{-k(a - t_0)}\right)
@@
 W_a = \theta_1 L_a^{\theta_2}
-\] where \(\theta_1 = e^{-12.1}\) and \(\theta_2 = 3.2\).
+\] where \(\theta_1 = e^{-12.1}\) and \(\theta_2 = 3.2\) (Figure
+\ref{om_waa}).
+
+\begin{figure}
+\caption{The weight at age assumed for the population in all operating and estimating models.}\label{om_waa}
+\begin{center}
+\includegraphics[width = \textwidth]{om_waa.png}
+\end{center}
+\end{figure}
 
@@
 unfished recruitment are also constant over the time period for those
-operating models (Miller and Brooks 2021). We specified unfished
-recruitment = \(R_0 = e^{10}\) and \(F_{\text{MSY}} = F_{40} = 0.348\)
-equated to a steepness of 0.69 and \(\alpha=0.60\) and
+operating models \citep{millerbrooks21}. We specified unfished
+recruitment = \(R_0 = e^{10}\) and \(\Fmsy = \Fspr[40] = 0.348\) equated
+to a steepness of 0.69 and \(\alpha=0.60\) and
 \(\beta = 2.4 \times 10^{-5}\) for the \[
 N_{1,y} = \frac{\alpha \text{SSB}_{y-1}}{1 + \beta \text{SSB}_{y-1}} 
-\] Beverton-Holt parameterization.
-
-The magnitude of the overfishing assumptions is based on average
-estimates of overfishing for NE groundfish stocks from Wiedenmann et
-al.~(20XX). Legault et al.~(2023) also used similar approaches to
-defining fishing mortality histories for operating models.
-
-Currenlty, initial population is configured at the equilibrium
-distribution fishing at \(F = 2.5\times F_{\text{MSY}}\).
+\] Beverton-Holt parameterization (Figure \ref{om_sr}). For OMs without
+process errors on natural mortality we assumed the rate was assumed 0.2.
+For OMs with process errors on natural mortality the mean rate was 0.2.
+
+\begin{figure}
+\caption{The Beverton-Holt stock recruit relationship assume for all operating models.}\label{om_sr}
+\begin{center}
+\includegraphics[width = \textwidth]{om_sr.png}
+\end{center}
+\end{figure}
+
+Two alternative fishing histories were used for operating models. In the
+first scenario, the stock experiences overfishing for the first 20 years
+and fishing at \Fmsy for the last 20 years. In the second scenario, the
+stock is fished at \Fmsy for the entire time period. The magnitude of
+the overfishing assumptions is based on average estimates of overfishing
+for NE groundfish stocks from \citet{wiedenmannetal19}.
+\Citet{legaultetal23} also used similar approaches to defining fishing
+mortality histories for operating models.
 
 Initial population was configured at the equilibrium distribution
-fishing at either \(F = 2.5\times F_{\text{MSY}}\) or
-\(F = F_{\text{MSY}}\) for the two alternative fishing histories. That
-is for a deterministic model, the age composition would not change over
-time when the fishing mortality was constant at the respective level.
+fishing at either \(F = 2.5\times \Fmsy\) or \(F = \Fmsy\) for the two
+alternative fishing histories. That is for a deterministic model, the
+age composition would not change over time when the fishing mortality
+was constant at the respective level.
 
@@
 time series process for M. For operating models with time-varying random
-effects for fishery selectivity, Fmsy is also not constant however we
+effects for fishery selectivity, \Fmsy is also not constant however we
 use the same F history as other operating models which corresponds to
@@
 \hypertarget{fleets}{%
-\subsubsection{Fleets}\label{fleets}}
+\subsubsection*{Fleets}\label{fleets}}
+\addcontentsline{toc}{subsubsection}{Fleets}
 
@@
 with logistic selectivity for the fleet with \(a_{50} = 5\) and slope =
-1. This selectivity is was used to define \(F_{\text{MSY}}\) for the
-Beverton-Holt stock recruitment parameters above. We assumed a
-logistic-normal distribution for the age-composition observations for
-the fleet.
-
-\hypertarget{fishing-histories}{%
-\subsubsection{Fishing histories}\label{fishing-histories}}
-
-All operating models assumed one of two different fishing histories. One
-: Fishing mortality is equal to Fmsy (0.348) for the whole 40 year
-period. Two : Fishing mortality is 2.5 times Fmsy for the first 20 years
-then changes to Fmsy for the last 20 years.
+1 (Figure \ref{om_mean_selectivity}). This selectivity is was used to
+define \Fmsy for the Beverton-Holt stock recruitment parameters above.
+We assumed a logistic-normal distribution for the age-composition
+observations for the fleet.
+
+\begin{figure}
+\caption{The selectivity at age assumed for the fishing fleet in operating models without selectivity random effects.
+The mean selectivity at age across time for the fishing fleet in operating models with selectivity random effects.}\label{om_mean_selectivity}
+\begin{center}
+\includegraphics[width = \textwidth]{om_mean_selectivity.png}
+\end{center}
+\end{figure}
 
 \hypertarget{indices}{%
-\subsubsection{Indices}\label{indices}}
+\subsubsection*{Indices}\label{indices}}
+\addcontentsline{toc}{subsubsection}{Indices}
 
@@
 way through the year. Catchability of both surveys are assumed to be
-0.1. We assumed logistic selectivity for both indices with
-\(a_{50} = 5\) and slope = 1. We assumed a logistic-normal distribution
-for the age-composition observations.
+0.1. Like the fishing fleet, we assumed logistic selectivity for both
+indices with \(a_{50} = 5\) and slope = 1 and a logistic-normal
+distribution for the age-composition observations.
 
 \hypertarget{observation-uncertainty}{%
-\subsubsection{Observation Uncertainty}\label{observation-uncertainty}}
+\subsubsection*{Observation Uncertainty}\label{observation-uncertainty}}
+\addcontentsline{toc}{subsubsection}{Observation Uncertainty}
 
@@
 \hypertarget{operating-models-with-random-effects-on-numbers-at-age}{%
-\subsubsection{Operating models with random effects on numbers at
+\subsubsection*{Operating models with random effects on numbers at
 age}\label{operating-models-with-random-effects-on-numbers-at-age}}
-
-24 operating models, 16 Sel re operating models and 16 q re operating
-models. Table of process error assumptions
+\addcontentsline{toc}{subsubsection}{Operating models with random
+effects on numbers at age}
+
+The factorial combinations of fishing histories, observation error
+assumptions, and marginal variance assumptions for recruitment and
+survival results in 24 different operating models with random effects on
+recruitment and survival (Table \ref{naa_om_table}).
 
@@
 \hypertarget{operating-models-with-random-effects-on-natural-mortality}{%
-\subsubsection{Operating models with random effects on natural
+\subsubsection*{Operating models with random effects on natural
 mortality}\label{operating-models-with-random-effects-on-natural-mortality}}
-
-16 operating models Table of process error assumptions
-
-NOTE: inv\_trans\_rho function in set\_M.R is mis-defined. Will affect
-correlation parameters assigned in operating models?
-
-Steepness and BRPs are not constant when M is time-varying (Miller and
-Brooks 2021). We uses the a/b parameters for the B-H defined for Fmsy =
-F40 at the mean M = 0.2 as defined above.
+\addcontentsline{toc}{subsubsection}{Operating models with random
+effects on natural mortality}
+
+The factorial combinations of fishing histories, observation error
+assumptions, and marginal variance and autocorrelation assumptions for
+natural mortality results in 16 different operating models with random
+effects on natural mortality (Table \ref{M_om_table}). Internally, WHAM
+treats natural mortality as a log-transformed parameter \[
+\log M_y = \mu_M + \epsilon_{M,y}
+\] that is a linear combination of a mean \(\mu_M\) and any annual
+random effects marginally distributed as
+\(\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)\). The values
+for the marginal variance and correlation parameters assumed in the
+operating models are given in Table \ref{M_om_table}.
 
@@
 \hypertarget{operating-models-with-random-effects-on-fleet-selectivity}{%
-\subsubsection{Operating models with random effects on fleet
+\subsubsection*{Operating models with random effects on fleet
 selectivity}\label{operating-models-with-random-effects-on-fleet-selectivity}}
-
-16 operating models Table of process error assumptions
-
-BRPs are not constant when fleet selectivity is time-varying. We uses
-the a/b parameters for the B-H defined for Fmsy = F40 at the mean of the
-time series model for selectivity which is the same as the constant
-selectivity defined above.
+\addcontentsline{toc}{subsubsection}{Operating models with random
+effects on fleet selectivity}
+
+The factorial combinations of fishing histories, observation error
+assumptions, and marginal variance and autocorrelation assumptions for
+selectivity parameters results in 16 different operating models with
+random effects on selectivity (Table \ref{sel_om_table}). Internally,
+WHAM treats selectivity parameter \(s\) as a logit-transformed parameter
+\[
+\log\left(\frac{p_{s,y}-l_{s}}{u_{s}-p_{s,y}}\right) = \mu_s + \epsilon_{s,y}
+\] that is a linear combination of a mean \(\mu_s\) and any annual
+random effects marginally distributed as
+\(\epsilon_{s,y} \sim \text{N}\left(0,\sigma_s^2\right)\) where the
+lower and upper bounds of the parameter (\(l_s\) and \(u_s\)) can be
+specified by the user. All selectivity parameters are either \(a_50\) or
+slope parameters and we assume bounds of 0 and 10 for all selectivity
+parameters for all operating and estimating models. The values for the
+marginal variance and correlation parameters assumed in the operating
+models are given in Table \ref{sel_om_table}.
 
@@
 \hypertarget{operating-models-with-random-effects-on-index-catchability}{%
-\subsubsection{Operating models with random effects on index
+\subsubsection*{Operating models with random effects on index
 catchability}\label{operating-models-with-random-effects-on-index-catchability}}
-
-16 operating models Table of process error assumptions
+\addcontentsline{toc}{subsubsection}{Operating models with random
+effects on index catchability}
+
+The factorial combinations of fishing histories, observation error
+assumptions, and marginal variance and autocorrelation assumptions for
+catchability results in 16 different operating models with random
+effects on catchability (Table \ref{q_om_table}). Internally, WHAM
+treats catchability for index \(i\) as a logit-transformed parameter \[
+\log\left(\frac{q_{i,y}-l_{i}}{u_{i}-q_{i,y}}\right) = \mu_i + \epsilon_{i,y}
+\] that is a linear combination of a mean \(\mu_i\) and any annual
+random effects marginally distributed as
+\(\epsilon_{i,y} \sim \text{N}\left(0,\sigma_i^2\right)\) where the
+lower and upper bounds of the catchability (\(l_i\) and \(u_i\)) can be
+specified by the user. Here we assume bounds of 0 and 1000 for all
+operating and estimating models. For operating and estimation models
+with process errors on catchability, the temporal variation is only
+assumed for the first index. The values for the marginal variance and
+correlation parameters assumed in the operating models are given in
+Table \ref{q_om_table}
 
@@
 \hypertarget{estimating-models}{%
-\subsection{Estimating models}\label{estimating-models}}
+\subsection*{Estimating models}\label{estimating-models}}
+\addcontentsline{toc}{subsection}{Estimating models}
+
+For each data set simulated from an operating model 20 estimating models
+were fit. A total of 32 different estimating models were fit across all
+operating models where the subset of 20 depended on the source of
+process error (Table \ref{em_table}). The first 20 estimating models in
+Table \ref{em_table} were fit to simulate data sets from R and R+S
+operating models. Estimating models 5 to 24 in Table \ref{em_table} were
+fit to simulate data sets from R+M operating models. Estimating models 5
+to 20 and 25-28 in Table \ref{em_table} were fit to simulate data sets
+from R+Sel operating models. Finally, estimating models 5 to 20 and
+29-32 in Table \ref{em_table} were fit to simulate data sets from R+q
+operating models.
 
 \begin{table}
-\caption{Distinguishing characteristics of the estimating models.}
+\caption{Distinguishing characteristics of the estimating models.}\label{em_table}
 {\scriptsize \input{em_tab}}
@@
 
-32 estimating models Table of estimating models
-
-1-20 fit to each NAA RE operating model 5-24 fit to each M RE operating
-model 5-20,25-28 to each sel RE operating model 5-20, 29-32 to each q RE
-operating model
-
 SR estimation or not
 
-Make plot of S-R curve, Fmsy = F40 Initial values for BH parameters are
-the true values. Initial values for mean R model = true R0.
+Initial values for BH parameters are the true values. Initial values for
+mean R model = true R0.
 
@@
 
-Summary statitics
-
-Bias
-
-Mohn's rho
-
-marginal AIC
-
-convergence types: 1) whether the optimization completed 2) nlminb
-convergence flag = 0 3) max abs gradient \textless{} 1e-6 4) No NAs for
-SE of fixed effects returned from TMB::sdreport 5) No SE NAs AND no SE
-\textgreater{} 100
+\hypertarget{measures-of-reliability}{%
+\subsection*{Measures of reliability}\label{measures-of-reliability}}
+\addcontentsline{toc}{subsection}{Measures of reliability}
+
+The first measure of reliability we investigated was frequency of
+convergence when fitting each estimating model to the simulated data
+sets. There are various ways to assess convergence of the fit. We
+summarized 5 alternative categories of convergence. 1. Did the
+optimization routine (stats::nlminb) complete without error? 2. Did the
+stats::nlminb convergence flag = 0 indicate successful convergence? 3.
+Was the maximum absolute value of the gradient of the log-likelihood
+\textless{} \(1\times10^{-6}\)? 4. Did TMB::sdreport provide non-NA
+values for all fixed effects standard errors? 5. Did TMB::sdreport
+provide all standard errors \textless{} 100?
+
+The first convergence criterion assesses whether the model crashes. The
+third is a measure that assesses how flat the likelihood is at the
+optimized point of the likelihood surface. The fourth and fifth criteria
+are specific to the hessian-based standard error reporting provided by
+TMB. the TMB::sdreport function will sometimes return standard error
+estimates even when the calculated hessian is not invertible (fourth
+criterion). It will also provide large standard errors for some
+parameters that are not estimated well or are near bounds on the
+transformed scale that is of primary interest. For example, variance
+parameters for random effects can often be estimated near 0 when the fit
+suggests no variation in the random effects or, equivalently, that the
+model without random effects is better.
+
+\hypertarget{aic-for-model-selection}{%
+\subsubsection*{AIC for model selection}\label{aic-for-model-selection}}
+\addcontentsline{toc}{subsubsection}{AIC for model selection}
+
+We measured the frequency of correct model selection using marginal AIC.
+For a given operating model we the set of models that were considered
+all made the same assumptions on whether or not to estimate M and
+whether a stock-recruit function is assumed.
+
+We also measured the frequency of correctly selecting models with
+Beverton-Holt stock recruit function assumed over models without the
+stock-recruit function (null model). Because the proportions of
+simulations where the correct model was chosen for a given set of
+operating model factors implied generally poor performance of AIC, we
+analyzed the effect of variation in SSB on the probability of lower AIC
+of the BH models. We fit binomial generalize linear models to the
+observed AIC choice for each data set as a function of log(SD(SSB)) and
+operating model.
+
+\hypertarget{bias}{%
+\subsubsection*{Bias}\label{bias}}
+\addcontentsline{toc}{subsubsection}{Bias}
+
+We estimated relative bias \[
+\widehat {\text{RB}}\left(\theta\right) = \frac{1}{N} \sum^N_{i = 1} \frac{\widehat \theta_i - \theta_i}{\theta_i}
+\] of annual SSB and fully-selected fishing mortality for each
+estimation model and constructed 95\% confidence intervals for the
+median relative bias using the binomial distribution approach as in
+\citet{stockmiller21}. We also estimated bias of stock-recruit
+parameters for estimation models that assumed the Beverton-Holt stock
+recruit function.
+
+\hypertarget{mohns-rho}{%
+\subsubsection*{Mohn's rho}\label{mohns-rho}}
+\addcontentsline{toc}{subsubsection}{Mohn's rho}
+
+We estimated Mohn's rho for SSB, fully-selected fishing mortality, and
+recruitment for each estimation model. We estimated 7 peels for each
+estimation model. We calculated median 95\% confidence intervals for
+Mohn's rho using the same methods as that for bias.
 
 \hypertarget{results}{%
-\section{Results}\label{results}}
-
-Do each of these by type of operating model (Naa, M, sel, q)
+\section*{Results}\label{results}}
+\addcontentsline{toc}{section}{Results}
 
 \hypertarget{convergence-performance}{%
-\subsection{Convergence performance}\label{convergence-performance}}
-
-convergence types: 1) whether the optimization completed 2) nlminb
-convergence flag = 0 3) max abs gradient \textless{} 1e-6 4) No NAs for
-SE of fixed effects returned from TMB::sdreport 5) No SE NAs AND no SE
-\textgreater{} 100
+\subsection*{Convergence performance}\label{convergence-performance}}
+\addcontentsline{toc}{subsection}{Convergence performance}
 
@@
 \hypertarget{r-rs-operating-models}{%
-\subsubsection{R, R+S operating models}\label{r-rs-operating-models}}
+\subsubsection*{R, R+S operating models}\label{r-rs-operating-models}}
+\addcontentsline{toc}{subsubsection}{R, R+S operating models}
+
+Figures \ref{naa_om_em_R_MF_convergence} to
+\ref{naa_om_em_BH_ME_convergence}
 
@@
 \hypertarget{rm-operating-models}{%
-\subsubsection{R+M operating models}\label{rm-operating-models}}
+\subsubsection*{R+M operating models}\label{rm-operating-models}}
+\addcontentsline{toc}{subsubsection}{R+M operating models}
+
+Figures \ref{M_om_em_R_MF_convergence} to
+\ref{M_om_em_BH_ME_convergence}
 
@@
 \hypertarget{rsel-operating-models}{%
-\subsubsection{R+Sel operating models}\label{rsel-operating-models}}
+\subsubsection*{R+Sel operating models}\label{rsel-operating-models}}
+\addcontentsline{toc}{subsubsection}{R+Sel operating models}
+
+Figures \ref{Sel_om_em_R_MF_convergence} to
+\ref{Sel_om_em_BH_ME_convergence}
 
@@
 \hypertarget{rq-operating-models}{%
-\subsubsection{R+q operating models}\label{rq-operating-models}}
+\subsubsection*{R+q operating models}\label{rq-operating-models}}
+\addcontentsline{toc}{subsubsection}{R+q operating models}
+
+Figures \ref{q_om_em_R_MF_convergence} to
+\ref{q_om_em_BH_ME_convergence}
 
@@
 \hypertarget{aic-performance-for-process-error-structure}{%
-\subsection{AIC performance for process error
+\subsection*{AIC performance for process error
 structure}\label{aic-performance-for-process-error-structure}}
+\addcontentsline{toc}{subsection}{AIC performance for process error
+structure}
+
+\hypertarget{r-rs-operating-models-1}{%
+\subsubsection*{R, R+S operating models}\label{r-rs-operating-models-1}}
+\addcontentsline{toc}{subsubsection}{R, R+S operating models}
+
+Figures \ref{naa_om_proportion_best_aic_R_MF} to
+\ref{naa_om_proportion_best_aic_SR_ME}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R and R+S process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.} \label{naa_om_proportion_best_aic_R_MF}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_proportion_best_aic_R_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R and R+S process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.} \label{naa_om_proportion_best_aic_R_ME}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_proportion_best_aic_R_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R and R+S process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate a stock-recruit relationship and and M is fixed at the true value.} \label{naa_om_proportion_best_aic_SR_MF}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_proportion_best_aic_SR_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R and R+S process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate a stock-recruit relationship and and M is estimated.} \label{naa_om_proportion_best_aic_SR_ME}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_proportion_best_aic_SR_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\hypertarget{rm-operating-models-1}{%
+\subsubsection*{R+M operating models}\label{rm-operating-models-1}}
+\addcontentsline{toc}{subsubsection}{R+M operating models}
+
+Figures \ref{M_om_proportion_best_aic_R_MF} to
+\ref{M_om_proportion_best_aic_SR_ME}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+M process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.} \label{M_om_proportion_best_aic_R_MF}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_proportion_best_aic_R_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+M process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.} \label{M_om_proportion_best_aic_R_ME}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_proportion_best_aic_R_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+M process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate a stock-recruit relationship and and M is fixed at the true value.} \label{M_om_proportion_best_aic_SR_MF}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_proportion_best_aic_SR_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+M process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate a stock-recruit relationship and and M is estimated.} \label{M_om_proportion_best_aic_SR_ME}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_proportion_best_aic_SR_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\hypertarget{rsel-operating-models-1}{%
+\subsubsection*{R+Sel operating models}\label{rsel-operating-models-1}}
+\addcontentsline{toc}{subsubsection}{R+Sel operating models}
+
+Figures \ref{Sel_om_proportion_best_aic_R_MF} to
+\ref{Sel_om_proportion_best_aic_SR_ME}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+Sel process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.} \label{Sel_om_proportion_best_aic_R_MF}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_proportion_best_aic_R_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+Sel process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.} \label{Sel_om_proportion_best_aic_R_ME}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_proportion_best_aic_R_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+Sel process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate a stock-recruit relationship and and M is fixed at the true value.} \label{Sel_om_proportion_best_aic_SR_MF}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_proportion_best_aic_SR_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+Sel process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate a stock-recruit relationship and and M is estimated.} \label{Sel_om_proportion_best_aic_SR_ME}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_proportion_best_aic_SR_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\hypertarget{rq-operating-models-1}{%
+\subsubsection*{R+q operating models}\label{rq-operating-models-1}}
+\addcontentsline{toc}{subsubsection}{R+q operating models}
+
+Figures \ref{q_om_proportion_best_aic_R_MF} to
+\ref{q_om_proportion_best_aic_SR_ME}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+q process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.} \label{q_om_proportion_best_aic_R_MF}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_proportion_best_aic_R_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+q process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.} \label{q_om_proportion_best_aic_R_ME}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_proportion_best_aic_R_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+q process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate a stock-recruit relationship and and M is fixed at the true value.} \label{q_om_proportion_best_aic_SR_MF}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_proportion_best_aic_SR_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+q process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate a stock-recruit relationship and and M is estimated.} \label{q_om_proportion_best_aic_SR_ME}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_proportion_best_aic_SR_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
 
@@
 
-\hypertarget{numbers-at-age-operating-models}{%
-\subsubsection{Numbers at age operating
-models}\label{numbers-at-age-operating-models}}
-
-\hypertarget{estimating-models-include-alternative-random-effects-options-naa-m-sel-q}{%
-\subsection{Estimating models include alternative random effects
-options: NAA, M, sel,
-q}\label{estimating-models-include-alternative-random-effects-options-naa-m-sel-q}}
-
-\clearpage
-
-\begin{table}
-\caption{NAA operating models, estimating models all assume a B-H stock recruit relationship and M is fixed at the true value.}
-{\input{naa_om_em_SR_MF_aic_table}}
-\end{table}
-
-\begin{table}
-\caption{NAA operating models, estimating models all assume a B-H stock recruit relationship and M is estimated.}
-{\input{naa_om_em_SR_ME_aic_table}}
-\end{table}
-
-\begin{table}
-\caption{NAA operating models, estimating models all estimate a mean recruitment and M is fixed at the true value.}
-{\input{naa_om_em_R_MF_aic_table}}
-\end{table}
-
-\begin{table}
-\caption{NAA operating models, estimating models all estimate a mean recruitment and M estimated.}
-{\input{naa_om_em_R_MF_aic_table}}
-\end{table}
-\clearpage
-
-\clearpage
-
 \hypertarget{aic-performance-for-stock-recruit-relationship}{%
-\subsection{AIC performance for stock-recruit
+\subsection*{AIC performance for stock-recruit
 relationship}\label{aic-performance-for-stock-recruit-relationship}}
-
-\hypertarget{r-rs-operating-models-1}{%
-\subsubsection{R, R+S operating models}\label{r-rs-operating-models-1}}
+\addcontentsline{toc}{subsection}{AIC performance for stock-recruit
+relationship}
+
+Tables of frequencies where AIC chooses correct stock-recruit assumption
+imply poor performance of AIC, but figures showing effect of variation
+in SSB demonstrate that large variation in SSB is needed to reduce Type
+II errors for marginal AIC.
+
+\hypertarget{r-rs-operating-models-2}{%
+\subsubsection*{R, R+S operating models}\label{r-rs-operating-models-2}}
+\addcontentsline{toc}{subsubsection}{R, R+S operating models}
 
@@
 
-\hypertarget{rm-operating-models-1}{%
-\subsubsection{R+M operating models}\label{rm-operating-models-1}}
+\hypertarget{rm-operating-models-2}{%
+\subsubsection*{R+M operating models}\label{rm-operating-models-2}}
+\addcontentsline{toc}{subsubsection}{R+M operating models}
 
@@
 
-\hypertarget{rsel-operating-models-1}{%
-\subsubsection{R+Sel operating models}\label{rsel-operating-models-1}}
+\hypertarget{rsel-operating-models-2}{%
+\subsubsection*{R+Sel operating models}\label{rsel-operating-models-2}}
+\addcontentsline{toc}{subsubsection}{R+Sel operating models}
 
@@
 
-\hypertarget{rq-operating-models-1}{%
-\subsubsection{R+q operating models}\label{rq-operating-models-1}}
+\hypertarget{rq-operating-models-2}{%
+\subsubsection*{R+q operating models}\label{rq-operating-models-2}}
+\addcontentsline{toc}{subsubsection}{R+q operating models}
 
@@
 
-\hypertarget{bias-mean-square-error}{%
-\subsection{Bias, Mean Square error}\label{bias-mean-square-error}}
-
-Certain basic parameters (stock-recruit pars, M, variance parameters)
-SSB, F, R, BRPs
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of for SSB for estimating models that estimate mean recruitment and M is fixed at the true value.}\label{naa_om_em_R_MF_relbias_ssb}
+\hypertarget{bias-1}{%
+\subsection*{Bias}\label{bias-1}}
+\addcontentsline{toc}{subsection}{Bias}
+
+\hypertarget{r-rs-oms}{%
+\subsubsection*{R, R+S OMs}\label{r-rs-oms}}
+\addcontentsline{toc}{subsubsection}{R, R+S OMs}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R and R+S process errors.
+Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{naa_om_em_R_MF_relbias_ssb}
 \begin{center}
@@
 \begin{figure}
-\caption{Median relative bias of for SSB for estimating models that estimates a BH stock-recruitment function and M is fixed at the true value.}\label{naa_om_em_SR_MF_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_SR_MF_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of for SSB for estimating models that estimates a BH stock-recruitment function and M is estimated.}\label{naa_om_em_SR_ME_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_SR_ME_relbias_ssb.png}
+\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R and R+S process errors.
+Estimation models assume a BH stock-recruitment function and M is fixed at the true value.}\label{naa_om_em_BH_MF_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_BH_MF_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R and R+S process errors.
+Estimation models assume a BH stock-recruitment function and estimate M.}\label{naa_om_em_BH_ME_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_BH_ME_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\hypertarget{rm-oms}{%
+\subsubsection*{R+M OMs}\label{rm-oms}}
+\addcontentsline{toc}{subsubsection}{R+M OMs}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+M process errors.
+Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{M_om_em_R_MF_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_R_MF_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+M process errors.
+Estimation models do not assume a stock-recruit relationship and estimate M.}\label{M_om_em_R_ME_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_R_ME_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+M process errors.
+Estimation models assume a BH stock-recruitment function and M is fixed at the true value.}\label{M_om_em_BH_MF_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_BH_MF_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+M process errors.
+Estimation models assume a BH stock-recruitment function and estimate M.}\label{M_om_em_BH_ME_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_BH_ME_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\hypertarget{rsel-oms}{%
+\subsubsection*{R+Sel OMs}\label{rsel-oms}}
+\addcontentsline{toc}{subsubsection}{R+Sel OMs}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+Sel process errors.
+Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{Sel_om_em_R_MF_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_R_MF_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+Sel process errors.
+Estimation models do not assume a stock-recruit relationship and estimate M.}\label{Sel_om_em_R_ME_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_R_ME_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+Sel process errors.
+Estimation models assume a BH stock-recruitment function and M is fixed at the true value.}\label{Sel_om_em_BH_MF_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_BH_MF_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+Sel process errors.
+Estimation models assume a BH stock-recruitment function and estimate M.}\label{Sel_om_em_BH_ME_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_BH_ME_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\hypertarget{rq-oms}{%
+\subsubsection*{R+q OMs}\label{rq-oms}}
+\addcontentsline{toc}{subsubsection}{R+q OMs}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+q process errors.
+Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{q_om_em_R_MF_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_R_MF_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+q process errors.
+Estimation models do not assume a stock-recruit relationship and estimate M.}\label{q_om_em_R_ME_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_R_ME_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+q process errors.
+Estimation models assume a BH stock-recruitment function and M is fixed at the true value.}\label{q_om_em_BH_MF_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_BH_MF_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+q process errors.
+Estimation models assume a BH stock-recruitment function and estimate M.}\label{q_om_em_BH_ME_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_BH_ME_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\hypertarget{mohns-rho-1}{%
+\subsection*{\texorpdfstring{Mohn's
+\(\rho\)}{Mohn's \textbackslash rho}}\label{mohns-rho-1}}
+\addcontentsline{toc}{subsection}{Mohn's \(\rho\)}
+
+\hypertarget{r-rs-oms-1}{%
+\subsubsection*{R, R+S OMs}\label{r-rs-oms-1}}
+\addcontentsline{toc}{subsubsection}{R, R+S OMs}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R and R+S process errors.
+Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{naa_om_em_R_MF_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_mohns_rho_R_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R and R+S process errors.
+Estimation models do not assume a stock-recruit function and M estimated.}\label{naa_om_em_R_ME_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_mohns_rho_R_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R and R+S process errors.
+Estimation models assume a B-H stock-recruit function and M is fixed at the true value.}\label{naa_om_em_BH_MF_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_mohns_rho_BH_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R and R+S process errors.
+Estimation models assume a B-H stock-recruit function and M estimated.}\label{naa_om_em_BH_ME_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_mohns_rho_BH_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\hypertarget{rm-oms-1}{%
+\subsubsection*{R+M OMs}\label{rm-oms-1}}
+\addcontentsline{toc}{subsubsection}{R+M OMs}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+M process errors.
+Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{M_om_em_R_MF_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_mohns_rho_R_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+M process errors.
+Estimation models do not assume a stock-recruit function and M estimated.}\label{M_om_em_R_ME_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_mohns_rho_R_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+M process errors.
+Estimation models assume a B-H stock-recruit function and M is fixed at the true value.}\label{M_om_em_BH_MF_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_mohns_rho_BH_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+M process errors.
+Estimation models assume a B-H stock-recruit function and M estimated.}\label{M_om_em_BH_ME_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_mohns_rho_BH_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\hypertarget{rsel-oms-1}{%
+\subsubsection*{R+Sel OMs}\label{rsel-oms-1}}
+\addcontentsline{toc}{subsubsection}{R+Sel OMs}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+Sel process errors.
+Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{Sel_om_em_R_MF_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_mohns_rho_R_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+Sel process errors.
+Estimation models do not assume a stock-recruit function and M estimated.}\label{Sel_om_em_R_ME_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_mohns_rho_R_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+Sel process errors.
+Estimation models assume a B-H stock-recruit function and M is fixed at the true value.}\label{Sel_om_em_BH_MF_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_mohns_rho_BH_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+Sel process errors.
+Estimation models assume a B-H stock-recruit function and M estimated.}\label{Sel_om_em_BH_ME_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_mohns_rho_BH_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\hypertarget{rq-oms-1}{%
+\subsubsection*{R+q OMs}\label{rq-oms-1}}
+\addcontentsline{toc}{subsubsection}{R+q OMs}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+q process errors.
+Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{q_om_em_R_MF_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_mohns_rho_R_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+q process errors.
+Estimation models do not assume a stock-recruit function and M estimated.}\label{q_om_em_R_ME_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_mohns_rho_R_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+q process errors.
+Estimation models assume a B-H stock-recruit function and M is fixed at the true value.}\label{q_om_em_BH_MF_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_mohns_rho_BH_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+q process errors.
+Estimation models assume a B-H stock-recruit function and M estimated.}\label{q_om_em_BH_ME_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_mohns_rho_BH_ME.png}
 \end{center}
@@
 \hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
+\section*{Discussion}\label{discussion}}
+\addcontentsline{toc}{section}{Discussion}
 
@@
 
-\clearpage
-
 \end{document}

### 2023-09-10 Ecov_study/catchability/Catchability_ecov_sims.Rmd
Add code updates, separate functions into separate scripts.

--- 2022-11-16 Ecov_study/catchability/Catchability_ecov_sims.Rmd
@@
 date: "2022-11-14"
+editor_options: 
+  markdown: 
+    wrap: 72
 ---
-# Function definitions
-Functions based on simulation testing vignette: https://github.com/timjmiller/wham/blob/devel/vignettes/ex10_simulation.Rmd
 
-## Generate simulated OM data
+Load packages & source functions used in simulation testing
 ```{r}
-#' @param OM A fitted WHAM model to use as the operating model, no default.
-#' 
-#' @return A simulated OM dataset, may be used to overwrite input$data either entirely or in part in simulations
-
-simOM <- function(OM=NULL){
-        input <- OM$input
-        dataOM = OM$simulate(complete=TRUE) # Overwrite input with simulated data & return
-        return(dataOM)
-}
-```
-
-## Fit specified EMs to simulated OM data and store results
-```{r}
-#' @param nsim A number specifying how many simulations to run, default = 1.
-#' @param OM A fitted WHAM model to use as an operating model, no default.
-#' @param inputEMlist A list of WHAM input data objects for all estimation models, no default.
-#' @param outdir A filename specifying the directory where time-stamped results should be stored, if NULL (default) then the results are not saved as RData,
-#' 
-#' @return A nested list with names and results corresponding to the operating models and all estimation models.
-For each model results contains a sub-list with items for each simulation's results stored as sub-sub lists (because WHAM results are stored as a list).
-If outdir != NULL then results are saved as RData and returned as an object from the function.
-
-simTestWHAM <- function(nsim = 1,
-                        OM = NULL,
-                        inputEMlist = NULL,
-                        outdir = NULL){
-        # Simulation set-up
-        obs_names = c("agg_indices","agg_catch","catch_paa","index_paa", "Ecov_obs", "obsvec") # Data overwritten by OM
-        namesEM <- lapply(inputEMlist, FUN = function(inputEMlist){inputEMlist$model_name}) %>% unlist() # EM model names
-
-        # Storage set-up
-        results <- NULL #!!!!! Still need to work on setting this up
-        storeEM <- lapply(namesEM, FUN = assign, value = list())
-        names(storeEM) <- namesEM
-
-        # Generate OM data for all simulations
-        sim_inputs <- replicate(nsim, simOM(OM), simplify = FALSE)
-
-        for(isim in 1:nsim){ # Loop over simulations
-                for(iEM in 1:length(inputEMlist)){ # Loop over estimation models
-                        # Pull EM initial input from list
-                        inputEM <- inputEMlist[[iEM]]
-
-                        # Update EM input with OM simulated data
-                        inputEM$data[obs_names] = sim_inputs[[isim]][obs_names]
-
-                        # Fit WHAM model to updated EM input and save results
-                        resultEM <- fit_wham(inputEM, do.osa = FALSE, do.retro = FALSE, MakeADFun.silent = TRUE) #!!!! Figure out how to store
-
-                        # Store model-specific results for given isim here (append EM results to storage list for that EM)
-                        storeEM[which(namesEM == namesEM[iEM])][[1]] <- append(storeEM[which(namesEM == namesEM[iEM])], list(resultEM))
-
-                } # End loop over EM
-        } # End loop over nsims
-
-        # Store all results
-        results$OM <- sim_inputs
-        names(results) <- OM$model_name # Name OM column based on model_name
-        results <- append(results, storeEM) # Should add named EM results to the final resuts object
-
-        # Save time-stamped results as RData if output directory provided as argument
-        timeStamp <- Sys.time() %>% gsub(" ", "_",.) %>% gsub(":", "-", .) # Change millisecond half of Sys.time() output to avoid having spaces/weird characters in filenames
-        if(is.null(outdir)==FALSE){
-                save(results, file = paste(outdir, paste0("simWHAM_", nsim, "_nsim_", OM$model_name, "_OM_", timeStamp, ".RData"), sep="/"))
-        }
-
-        return(results) # Always return results as an object
-}
+# Packages
+library(tidyverse)
+library(wham)
+# Simulation testing functions
+source(here::here("Ecov_study","catchability","simOM.R"))
+source(here::here("Ecov_study","catchability","simTestWHAM.R"))
 ```
@@
 ### Visual describing results structure
-WHAM results objects only displayed for OM simulation 1, similar structure for other simulations and for estimation models.
+
+WHAM results objects only displayed for OM simulation 1, similar
+structure for other simulations and for estimation models.
+
 ```{r}
@@
 ## Post-process results
+
 ```{r}
-#' @param filenames A vector of strings (including .Rda extensions) indicating what Rdata results files to read in and post process
-#' 
-#' @return
 
-postprocess_simTestWHAM <- function(filenames = NULL){
-        # Set up storage for processed results
-
-        # Loop over files
-        for(ifile in 1:length(filenames)){
-                # Read in Rdata object
-                fileDat <- readRDS(filenames[ifile])
-
-                nsim <- # Based on length of results in given filename
-
-                # Loop over simulations in each file
-                for(isim in 1:nsim){
-                        # Calculate performance metrics here & add to storage
-                }
-
-        } # End loop over filenames
-}
 ```
@@
 # Specifics for environmentally-linked catchability WHAM runs
+
+## Generic stock info that is never misspecified/changed
+```{r}
+# Pull generic groundfish life history/data
+source(file.path(here::here(),"common_code", "make_basic_info.R"))
+groundfish_info <- make_basic_info()
+
+# Assume logistic selectivity for the fleet and all indices
+sel_list <- list(model = c(rep("logistic", groundfish_info$n_fleets),rep("logistic", groundfish_info$n_indices)),
+                   initial_pars = rep(list(c(5,1)), groundfish_info$n_fleets + groundfish_info$n_indices),
+                 n_selblocks = c(groundfish_info$n_fleets + groundfish_info$n_indices)) #fleet, index
+
+# Starting mean M estimate at 0.2, estimate constant M for all ages/years
+M <- list(initial_means = rep(0.2, length(groundfish_info$ages)))
+
+# Assume random about the mean recruitment, beverton-holt stock recruit didn't converge with an invertible hessian
+recruit_model <- 2 
+# NAA_list <- list(sigma="rec") # Recruitment random effects so stock recruit function implemented
+
+genericInput <- prepare_wham_input(basic_info = groundfish_info, 
+                                   selectivity = sel_list, 
+                                   recruit_model = recruit_model) #,
+                                   #NAA_re = NAA_list)
+genericFit <- fit_wham(genericInput, MakeADFun.silent = TRUE, do.retro = FALSE, do.osa = FALSE)
+
+```
+
 ## Define Operating Models (OMs)
-All OMs should have catchability linked to an environmental covariate but may differ in 
-- Uncertainty around this relationship #!!!!!!!!
-- Sign of this relationship (+/-) #!!!!!!!!
+
+All OMs should have catchability linked to an environmental covariate
+but may differ in - Uncertainty around this relationship #!!!!!!!! -
+Sign of this relationship (+/-) #!!!!!!!!
 
 Set up input and fit OM1
+
 ```{r}
@@
 ## Define Estimation Models (EMs)
+
+| q-Env link                    | Seasonal misspecification |
+| ----------------------------- | ------------------------- |
+| Without link                  | Both seasons correct (q-Env link in EM and OM)         |
+| With link                     | 1 season misspecified (q-Env linked but no link in OM) |
+| Without link, q random effect | Both seasons misspecified (no link but link in OM)     |
+| With link, q random effect    | |
+
 Correctly specified link between catchability and the environment
+
 ```{r}
@@
 
-Catchability random effect and a link between catchability and the environment
+Catchability random effect and a link between catchability and the
+environment
+
 ```{r}
@@
 
-Only a catchability random effect, fit to environmental model without link
+Only a catchability random effect, fit to environmental model without
+link
+
 ```{r}
@@
 
-No catchability considerations (i.e. status quo), fit to environmental model without link !!!! how does this work if you feed in the ecov_beta?
+No catchability considerations (i.e. status quo), fit to environmental
+model without link !!!! how does this work if you feed in the ecov_beta?
+
 ```{r}
@@
 ## Simulation testing
+
 ```{r}
@@
 ## Post-process simulation results
+
 ```{r}
@@
 
-!!!! post-process based on storage & how sims run
-- could run sims based varying isim
-- could also loop over OM but this could be computationally expensive
-- I recommend labeling runs with OM and EM names (and setting these up with informative) so it is easy to pull out results & compare based on model setup
-  - pull from mod$model_name to help aggregate data
+!!!! post-process based on storage & how sims run - could run sims based
+varying isim - could also loop over OM but this could be computationally
+expensive - I recommend labeling runs with OM and EM names (and setting
+these up with informative) so it is easy to pull out results & compare
+based on model setup - pull from mod\$model_name to help aggregate data
 
-proposed function with:
-- fitted OM model object
-- list of EM inputs
+proposed function with: - fitted OM model object - list of EM inputs
 
 ### Result formatted as list of lists, save as rdata object in specified directory with timestamp in name
-run1_11_14_22.Rda<-result
- - EM1 
- - EM2 
- - EM3
- - EM4
- - OM result
+
+run1_11_14_22.Rda\<-result - EM1 - EM2 - EM3 - EM4 - OM result
 
 # Other things to consider
-- Other EMs can be developed (e.g. with alternative ecov process assumptions) but should be given DIFFERENT model_names to help with post-processing
+
+-   Other EMs can be developed (e.g. with alternative ecov process
+    assumptions) but should be given DIFFERENT model_names to help with
+    post-processing
 
 # Short script for function testing purposes only!!!
+
 ```{r}
@@
 ```
+
+# To do next
+
+- Finish OM setup
+  - change how ecov passed in
+  - set up range of uncertainty values to explore
+  - tim sets equilibrium F (eq_F_init)
+    - setup OM without fit & use model to find F40 proxy
+  - Tim also sets intitial numbers at age based on equilibrium (overfished F=eqFN1*max_mult_Fmsy or not eqFN1=0)
+- Tim lines 71-79 pf make_om use internal processing for prepare_wham_input to plug in setup from M_Ecov_om_setup.R
+- Update what is saved as output
+  - calculate performance metrics
+- Run simulations, hpcc?
+- Separate sims and function scripts

### 2023-09-10 Ecov_study/catchability/Setup_OMs.Rmd
Add code updates, separate functions into separate scripts.

@@
 
--   Ecov ar1 process on catchability for both surveys (spring & fall)
-    with mean = 0
-
--
+-   Ecov ar1 process on catchability with mean = 0
 
@@
 -   Fishing history = F_MSY for 40 year history OR fish 2.5\*F_MSY for
-    20 years than fish at F_MSY for remaining 20
+    20 years than fish at F_MSY for remaining 20 - NOT CURRENTLY IMPLEMENTED, IN PROGRESS
 
--
+-   Seasonality of Ecov effect on catchability (neither, one, or both seasons impacted) - NOT CURRENTLY IMPLEMENTED, IN PROGRESS
 
@@
 
--
+-   Seasonality of Ecov effect on catchability
 
+Run analysis in 3 pieces:
+1) Factorial runs with different ecov process magnitude, fishing history and observation error
+-   maybe don't vary obs/process NAA since we are testing ability to
+    recover catchability effect, could do sensitivity run with survival
+    random effects and pick one of these settings
 ```{r}
@@
 ageComp_sig <- c(0.3, 1.5) # Fleet and index age comp treated with same error
-logIndex_sig <- c(0.1, 0.4)
+log_index_sig <- c(0.1, 0.4) # agg_index_sigma
+log_catch_sig <- 0.1 # agg_catch_sigma
 
@@
 OMsetup
+```
 
-# Catchability set up
-q_Ecov <- c("includeEcov", "excludeEcov") # May want to include these as a sensitivity run to compare what happens if OM has no ecov effect but EM assumes there is one (ecov process setup won't matter since fitting to data without effect specified in model)
-
-##### OM sensitivity runs
-# Factorial combinations when on ecov effect in OM
-expand.grid(q_Ecov = q_Ecov, F_hist = F_hist, ageComp_sig = ageComp_sig, logIndex_sig = logIndex_sig) # probably do a subset, we have to decide what subset
-
+2) Pick subset of OM settings and vary seasonality of q effect in EM
+```{r}
 ##### EM
@@
 
--   maybe don't vary obs/process NAA since we are testing ability to
-    recover catchability effect, could do sensitivity run with survival
-    random effects and pick one of these settings
+3) May run sensitivity run with no ecov effect in OM but compare same EMs
+```{r}
+# Catchability set up
+q_Ecov <- c("includeEcov", "excludeEcov") # May want to include these as a sensitivity run to compare what happens if OM has no ecov effect but EM assumes there is one (ecov process setup won't matter since fitting to data without effect specified in model)
+
+##### OM sensitivity runs
+# Factorial combinations when no ecov effect in OM
+expand.grid(q_Ecov = q_Ecov, F_hist = F_hist, ageComp_sig = ageComp_sig, logIndex_sig = logIndex_sig) # probably do a subset, we have to decide what subset
+```
 
@@
 
-# Loop over OMsetup to initialize OMs
+# Loop over rows in OMsetup to initialize OM and EMs
 for(iom in 1:nrow(OMsetup)){
@@
     Ecov <- list(label = "Ecov",
+                 mean = matrix(rep(0,length(groundfish_info$years)), ncol =1), # Mean = 0
+                 logsigma = matrix(log(OMsetup$Ecov_process_obs_sig[iom]), length(groundfish_info$years), ncol=1),
+                 years = groundfish_info$years, 
+                 use_obs = matrix(rep(TRUE, length(groundfish_info$years)), ncol = 1),
                  lag = 0, 
@@
                  how = 1,
-                 indices = c(1,2),
-                 process_model = "ar1", 
-                 process_mean_vals = rep(0,length(groundfish_info$years)), # End generic inputs for ecov
+                 indices = list(1,2),
+                 process_model = "ar1", # End generic inputs for ecov
+                 process_mean_vals = rep(0,length(groundfish_info$years)), # Mean = 0
                  process_sig_vals = rep(OMsetup$Eco_process_sig[iom]),
                  process_cor_vals = rep(OMsetup$Ecov_process_cor[iom]),
-                 beta_vals = list(rep(list(matrix(rep(OMsetup$Ecov_effect[iom], length(groundfish_info$ages)), nrow = 1)), 4)), #!!! Why 4 rows?
-                 logsigma = as.matrix(rep(OMsetup$Ecov_process_obs_sig[iom], length(groundfish_info$years))))
+                 beta_vals = list(rep(list(matrix(rep(OMsetup$Ecov_effect[iom], length(groundfish_info$ages)), nrow = 1)), 4))) #!!! Why 4 rows?
 
@@
 
+#   # Fishing history
+#   # input =
+#           set_F_scenario(input, Fhist, Fmsy = Fmsy, max_mult = max_mult_Fmsy, min_mult= min_mult_Fmsy,
+#     change_time = F_change_time) # Function pulled from get_FMSY.R
 #   
-  # Fishing history
-F_hist <- c("H-MSY", "MSY") # High-then MSY vs. MSY for entire history
+# F_hist <- c("H-MSY", "MSY") # High-then MSY vs. MSY for entire history Fmsy is an option
 
-# Observation error
-ageComp_sig <- c(0.3, 1.5) # Fleet and index age comp treated with same error
-logIndex_sig <- c(0.1, 0.4)
+# Observation error - set L-N SD parameters for catch and index age comps
+input$par$catch_paa_pars[,1] <- log(OMsetup[iom, "ageComp_sig"])
+input$par$index_paa_pars[,1] <- log(OMsetup[iom, "ageComp_sig"])
+input$data$agg_catch_sigma[] <- OMsetup[iom,"log_catch_sig"]
+input$data$agg_index_sigma[] <- OMsetup[iom,"log_index_sig"]
 

### 2023-09-10 Ecov_study/catchability/simOM.R

@@ -1,0 +1,25 @@
+#' @title  Generate simulated OM data
+#' 
+#' @param OM A fitted WHAM model to use as the operating model, no default.
+#' @param seed A number to set the seed for OM simulated data, no default. 
+#' @param isim A number indicating what simulation data is generated for, no default.
+#' 
+#' @return A list containing:
+#' \itemize{
+#'   \item{dataOM - A simulated OM dataset, may be used to overwrite input$data either entirely or in part in simulations}
+#'   \item{seed - A number used to set the seed for OM simulated data}
+#' }
+
+# Function based on simulation testing vignette:https://github.com/timjmiller/wham/blob/devel/vignettes/ex10_simulation.Rmd
+        
+simOM <- function(OM=NULL,seed=NULL, isim = NULL){
+        # input <- OM$input
+        set.seed(seed[isim])
+        dataOM <- OM$simulate(complete=TRUE) # Overwrite input with simulated data & return
+        
+        # Return
+        returnList <- NULL
+        returnList$dataOM <- dataOM
+        returnList$seed <- seed[isim]
+        return(returnList)
+}

### 2023-09-10 Ecov_study/catchability/simTestWHAM.R

@@ -1,0 +1,88 @@
+#' @title  Fit specified EMs to simulated OM data and store results
+#' 
+#' @param nsim A number specifying how many simulations to run, default = 1.
+#' @param OM A fitted WHAM model to use as an operating model, no default.
+#' @param inputEMlist A list of WHAM input data objects for all estimation models, no default.
+#' @param outdir A filename specifying the directory where time-stamped results should be stored, if NULL (default) then the results are not saved as RData,
+#' @param seeds A vector of length nsim containing seed numbers for OM simulated data, no default. If not provided a random number used to set the seed and is returned for reproducibility.
+#' 
+#' @return A nested list with names and results corresponding to the operating models and all estimation models. For each model results contains a sub-list with items for each simulation's results stored as sub-sub lists (because WHAM results are stored as a list). If outdir != NULL then results are saved as RData and returned as an object from the function.
+
+# Function based on simulation testing vignette:https://github.com/timjmiller/wham/blob/devel/vignettes/ex10_simulation.Rmd
+
+# simTestWHAM(nsim = 1, OM = OM1, inputEMlist = list(genericInput), outdir = here::here("Ecov_study", "catchability"))
+
+simTestWHAM <- function(nsim = 1,
+                        OM = NULL,
+                        inputEMlist = NULL,
+                        outdir = NULL,
+                        seeds = NULL){
+        # Simulation set-up
+        obs_names = c("agg_indices","agg_catch","catch_paa","index_paa", "Ecov_obs", "obsvec") # Data overwritten by OM
+        namesEM <- lapply(inputEMlist, FUN = function(inputEMlist){inputEMlist$model_name}) %>% unlist() # EM model names
+        
+        # Storage set-up
+        results <- NULL #!!!!! Still need to work on setting this up
+        storeEM <- lapply(namesEM, FUN = assign, value = list())
+        names(storeEM) <- namesEM
+        sim_inputs <- vector(mode='list', length = nsim)
+        
+        # Set seeds for all simulation
+        if(is.null(seeds)==TRUE){ # If vector of seeds (of length nsim) is not provided, set up random seed for OM data generation
+          seeds <- sample(1:1000000, size = nsim, replace = FALSE)
+        }
+        
+        
+        # Loop over simulations
+        for(isim in 1:nsim){ 
+                # Generate OM data for each simulation
+                sim_inputs[[isim]] <- simOM(OM, seed = seeds, isim = isim) #!!! this now returns is a list of dataOM and seed for each replicate
+                #!!! Save the seed somewhere here
[…]

### 2023-09-13 Ecov_study/recruitment_functions/paper/main.tex
starting report

\documentclass[11pt]{article}
\usepackage{parskip}
\usepackage[a4paper, total={7in, 10in}]{geometry}
\usepackage{amsmath,amsfonts,amsthm,bm}
\usepackage[english]{isodate}
\usepackage[final]{pdfpages}
\usepackage{graphicx} 
\usepackage{breakcites}
\usepackage{mathtools}
\usepackage{soul}
\usepackage{natbib}
\usepackage{authblk}
\usepackage[mathlines]{lineno}
\usepackage[labelfont=bf,skip=0.5cm,justification=justified,singlelinecheck=false]{caption}
\usepackage{outlines}
\bibliographystyle{agu}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Title %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Reliability of stock-recruitment function estimation in state space assessment models}
\author[1]{Gregory L. Britten}
\author[2]{Elizabeth Brooks}
\author[2]{Timothy Miller}
\affil[1]{Woods Hole Oceanographic Institution}
\affil[2]{Northeast Fisheries Science Center}
\date{\small{\printdayoff\today}}

\begin{document}
\maketitle 
\linenumbers
\section*{Abstract}
Stock-recruitment functions are important but difficult to estimate.

Many assessment models do not estimate stock-recruitment functions.

Maybe state space models help.

Here we ...

We find ...

This has implications for ...
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Introduction}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Methods}
We used the WHAM package (ref, commit 77bbd94)

We performed a simulation study with 96 operating models that differed in the stock recruitment function used to simulate the population and a variety of additional variance parameter settings. 
Variance parameter settings were determined by a review of the range of estimates from recent applications of WHAM in management of stocks of haddock, butterfish, and American plaice in the NE US.

We simulated 100 data sets for each operating model.

For each simulated data set we fit a set of 4 estimating models.

%%--OPERATING MODELS--%%
\subsubsection*{Operating Models}
The general operating model structure follows that of Project 0, reviewed briefly here. 
The population model tracks 10 age classes: ages 1 to 10+, assuming spawning occurs
1/4 of the way through the year. 
The maturity at age was a logistic curve with a50 = 2.89 and slope = 0.88, assumed known in all estimation models.

Weight at age was generated with a LVB growth function

$$ L_a = L_{\inf} \left( 1 - e^{-k(a-t_0}  \right)$$

with $t_0 = 0$, $L_{\inf}=85$, and $k=0.3$. The length-weight relationship is

$$W_a = \theta_1 L_a^{\theta_2}$$

with $\theta_1 = e^{-12.1}$ and $\theta_2 = 3.2$.

We assume a Beverton-Holt stock-recruitment of the form

$$ N_{1,y} = \frac{\alpha\mathrm{SSB}_{y-1}}{1 + \beta \mathrm{SSB}_{y-1}} $$

where $\alpha$ is referred to as the density-independent recruitment and $\beta$ sets the strength of density-dependence. 
We specified unfished recruitment at $R_0 = e^{19}$ and $F_{MSY} = F_{\%40} = 0.348$ which corresponds to a steepness of 0.69, $\alpha=0.60$, and $\beta=2.4*10^{-5}$.

%%--ESTIMATING MODELS--%%
\subsubsection*{Estimating Models}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Results}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Discussion}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Appendix A}
\input{df.oms.tex}
\input{df.ems.tex}

\bibliography{refs.bib}

\end{document}

### 2023-09-15 Ecov_study/mortality/paper/journalshortcuts.tex
update work on project 0 and Ecov-M working papers

--- 2023-05-14 Project_0/paper/paper.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[
-  12pt,
-]{article}
-\usepackage{amsmath,amssymb}
-\usepackage{lmodern}
-\usepackage{iftex}
-\ifPDFTeX
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={Factors affecting reliablity of state-space age-structured assessment models},
-  pdfauthor={Timothy J. Miller1, Britten, Brooks, Fay, Legault, Stock, et al.},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage[margin=1in]{geometry}
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{5}
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
-  \fi
-  % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-\usepackage{url}
-\usepackage{setspace}
-%\singlespacing
-%\onehalfspacing
-\doublespacing
-\usepackage{lineno}
-\linenumbers
-\usepackage[belowskip=0pt,aboveskip=0pt]{caption}
-\usepackage{relsize}
 \newcommand{\afrb}{Alaska Fishery Research Bulletin\xspace}
@@
 \newcommand{\sm}{Statistics in Medicine}
-
-\usepackage{xspace}
-\usepackage{bm}
-\usepackage{caption,graphics}
-\usepackage{graphicx}
-\usepackage{makecell}
-\renewcommand\figurename{Fig.}
-\captionsetup{labelsep=period, singlelinecheck=false}
-\newcommand{\changesize}[1]{\fontsize{#1pt}{#1pt}\selectfont}
-\renewcommand{\arraystretch}{1.5}
-\renewcommand\theadfont{}
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{wrapfig}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{pdflscape}
-\usepackage{tabu}
-\usepackage{threeparttable}
-\usepackage{threeparttablex}
-\usepackage[normalem]{ulem}
-\usepackage{makecell}
-\usepackage{xcolor}
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\usepackage[]{natbib}
-\bibliographystyle{cjfas2.bst}
-
-\title{Factors affecting reliablity of state-space age-structured
-assessment models}
-\author{Timothy J. Miller\textsuperscript{1}, Britten, Brooks, Fay,
-Legault, Stock, et al.}
-\date{}
-
-\begin{document}
-\maketitle
-
-\(^1\)corresponding author:
-\href{mailto:timothy.j.miller@noaa.gov}{\nolinkurl{timothy.j.miller@noaa.gov}},
-Northeast Fisheries Science Center, Woods Hole Laboratory, 166 Water
-Street, Woods Hole, MA 02543 USA\\
-
-\pagebreak
-
-\hypertarget{abstract}{%
-\subsection*{Abstract}\label{abstract}}
-\addcontentsline{toc}{subsection}{Abstract}
-
-\hypertarget{keywords}{%
-\subsubsection*{Keywords}\label{keywords}}
-\addcontentsline{toc}{subsubsection}{Keywords}
-
-\pagebreak
-
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
-
-Application of state-space models in fisheries stock assessment and
-management has expanded dramatically within ICES, Canada and the
-Northeast US \citep{nielsenberg14,cadigan16,stockmiller21}.
-
-Much is known about the reliability of state-space models that are
-linear or Gaussian \citep{aeberhardetal18}, but applications in
-fisheries management are nonlinear and typically include multiple types
-of observations with varying distributional assumptions. We know
-relatively little about the statistical reliability of such models.
-Also, there is a wide range of potential random effects structures in
-assessment models and we know little about the ability of information
-criteria to distinguish among such alternative structures.
-
-But those studies focus primarily on Gaussian linear models. review
-literature on reliability of hidden/latent process models. Primarily in
-other fields.
-
-Here we conduct a simulation study with operating models varying by
-degree of observation error uncertainty, sources of process error (M,
-NAA, q, sel), and fishing history. The simulations from these operating
-models are fitted with estimating models that make alternative
-assumptions for sources of process error (M, NAA\textless{} q, sel),
-whether a stock-recruit model was estimated, and whether M is estimated.
-We evaluate whether AIC can correctly determine the correct source of
-process error and stock effects on recruitment. We also evaluate the
-degree of bias in the outputs of the assessment model that are important
-for management.
-
-\hypertarget{methods}{%
-\section{Methods}\label{methods}}
-
-Used the WHAM package \citep[][commit 77bbd94]{stockmiller21} (Miller
-and Stock 2020). This packages has also been used to configure operating
-and estimating models for closed loop simulations evaluating index-based
-assessment methods \citep{legaultetalinpress} and is used for management
-of haddock, butterfish, plaice, bluefish in the Northeast US.
-
-We completed a simulation study with a number of operating models that
-can be categorized based on where random effects are assumed: abundance
-at age, natural mortality, fleet selectivity, or index catchability. For
-each operating model assumptions about variance of process errors and
-observations are required and the values we used were based on a review
-of the range of estimates from recent applications of WHAM in management
-of stocks of haddock, butterfish, and American plaice in the NE US.
-
-We simulated 100 data sets for each operating model. For each simulated
-data from each operating model we fit a set of estimating models.
-
-Y estimating models fit to each
-
-\hypertarget{operating-models}{%
-\subsection{Operating models}\label{operating-models}}
-
-common to all:\\
-ages = 1 to 10+, M maturity
-
-marginal standard deviations for random effects are defined in tables of
-operating models.
-
-\hypertarget{population}{%
-\subsubsection{Population}\label{population}}
-
-There are 10 age classes: ages 1 to 10+.
-
-Spawning was assumed to occur 1/4 of the way through the year.
-
-Natural mortality rate was assumed 0.2 when it was constant and the mean
-of the time series process for operating models with M random effects.
-maturity a50 = 2.89, slope = 0.88
-
-Weight at age was generated with a LVB growth \[
-L_a = L_{\infty}\left(1 - e^{-k(a - t_0)}\right)
-\] where \(t_0 = 0\), \(L_\infty = 85\), and \(k = 0.3\), and a L-W
-relationship such that \[
-W_a = \theta_1 L_a^{\theta_2}
-\] where \(\theta_1 = e^{-12.1}\) and \(\theta_2 = 3.2\).
-
-We assumed a Beverton-Holt stock recruit function with constant
-pre-recruit mortality parameters for all operating models. All
-post-recruit productivity components are constant in the NAA and survey
-catchability process error operating models. Therefore steepness and
-unfished recruitment are also constant over the time period for those
-operating models (Miller and Brooks 2021). We specified unfished
-recruitment = \(R_0 = e^{10}\) and \(F_{\text{MSY}} = F_{40} = 0.348\)
-equated to a steepness of 0.69 and \(\alpha=0.60\) and
-\(\beta = 2.4 \times 10^{-5}\) for the \[
-N_{1,y} = \frac{\alpha \text{SSB}_{y-1}}{1 + \beta \text{SSB}_{y-1}} 
-\] Beverton-Holt parameterization.
-
-The magnitude of the overfishing assumptions is based on average
-estimates of overfishing for NE groundfish stocks from Wiedenmann et
-al.~(20XX). Legault et al.~(2023) also used similar approaches to
-defining fishing mortality histories for operating models.
-
-Currenlty, initial population is configured at the equilibrium
-distribution fishing at \(F = 2.5\times F_{\text{MSY}}\).
-
-Initial population was configured at the equilibrium distribution
-fishing at either \(F = 2.5\times F_{\text{MSY}}\) or
-\(F = F_{\text{MSY}}\) for the two alternative fishing histories. That
-is for a deterministic model, the age composition would not change over
-time when the fishing mortality was constant at the respective level.
-
-For operating models with time-varying random effects for M, steepness
-is not constant, but we used the same alpha and beta parameters as other
-operating models this equates to a steepness and R0 at the mean of the
-time series process for M. For operating models with time-varying random
-effects for fishery selectivity, Fmsy is also not constant however we
-use the same F history as other operating models which corresponds to
-Fmsy at the mean selectivity parameters.
-
-\hypertarget{fleets}{%
-\subsubsection{Fleets}\label{fleets}}
-
-We assumed a single fleet operating year round for catch observations
-with logistic selectivity for the fleet with \(a_{50} = 5\) and slope =
-1. This selectivity is was used to define \(F_{\text{MSY}}\) for the
-Beverton-Holt stock recruitment parameters above. We assumed a
-logistic-normal distribution for the age-composition observations for
-the fleet.
-
-\hypertarget{fishing-histories}{%
-\subsubsection{Fishing histories}\label{fishing-histories}}
-
-All operating models assumed one of two different fishing histories. One
-: Fishing mortality is equal to Fmsy (0.348) for the whole 40 year
-period. Two : Fishing mortality is 2.5 times Fmsy for the first 20 years
-then changes to Fmsy for the last 20 years.
-
-\hypertarget{indices}{%
-\subsubsection{Indices}\label{indices}}
-
-Two time series of surveys are assumed and observed in numbers rather
-than biomass for the entire 40 year period with one occurring in the
-spring (0.25 way through the year) and one in the fall (0.75 way through
-the year). Actually we have it currently configured that both occur 0.5
-way through the year. Catchability of both surveys are assumed to be
-0.1. We assumed logistic selectivity for both indices with
-\(a_{50} = 5\) and slope = 1. We assumed a logistic-normal distribution
-for the age-composition observations.
-
-\hypertarget{observation-uncertainty}{%
-\subsubsection{Observation Uncertainty}\label{observation-uncertainty}}
-
-Standard deviation for log-aggregate catch was 0.1. There were two
-levels of observation error variance for indices and age composition for
-both indices and fleet catch. A low uncertainty specification assumed
-standard deviation of both series of log-aggregate index observations
-was 0.1 and the standard deviation of the logistic-normal for age
-composition observations was 0.3 In the high uncertainty specification
-the standard deviation for log-aggregate indices was 0.4 and that for
-the age composition observations was 1.5. For all estimating models,
-standard deviation for log-aggregate observations was assumed known
-whereas that for the logistic-normal age composition observations was
-estimated.
-
-\hypertarget{operating-models-with-random-effects-on-numbers-at-age}{%
-\subsubsection{Operating models with random effects on numbers at
-age}\label{operating-models-with-random-effects-on-numbers-at-age}}
-
-24 operating models, 16 Sel re operating models and 16 q re operating
-models. Table of process error assumptions
-
-\hypertarget{operating-models-with-random-effects-on-natural-mortality}{%
-\subsubsection{Operating models with random effects on natural
-mortality}\label{operating-models-with-random-effects-on-natural-mortality}}
-
-16 operating models Table of process error assumptions
-
-NOTE: inv\_trans\_rho function in set\_M.R is mis-defined. Will affect
-correlation parameters assigned in operating models?
-
-Steepness and BRPs are not constant when M is time-varying (Miller and
-Brooks 2021). We uses the a/b parameters for the B-H defined for Fmsy =
-F40 at the mean M = 0.2 as defined above.
-
-\hypertarget{operating-models-with-random-effects-on-fleet-selectivity}{%
-\subsubsection{Operating models with random effects on fleet
-selectivity}\label{operating-models-with-random-effects-on-fleet-selectivity}}
-
-16 operating models Table of process error assumptions
-
-BRPs are not constant when fleet selectivity is time-varying. We uses
-the a/b parameters for the B-H defined for Fmsy = F40 at the mean of the
-time series model for selectivity which is the same as the constant
-selectivity defined above.
-
-\hypertarget{operating-models-with-random-effects-on-index-catchability}{%
-\subsubsection{Operating models with random effects on index
-catchability}\label{operating-models-with-random-effects-on-index-catchability}}
-
-16 operating models Table of process error assumptions
-
-\hypertarget{estimating-models}{%
-\subsection{Estimating models}\label{estimating-models}}
-
-32 estimating models Table of estimating models
-
-1-20 fit to each NAA RE operating model 5-24 fit to each M RE operating
-model 5-20,25-28 to each sel RE operating model 5-20, 29-32 to each q RE
-operating model
-
-SR estimation or not
-
-Make plot of S-R curve, Fmsy = F40 Initial values for BH parameters are
-the true values. Initial values for mean R model = true R0.
-
-M estimation or not
-
-NAA\_re Random effects on Recruitment only or random effects on
-recruitment and transitions among older numbers at age.
-
-M\_re Random effects on Recruitment only, M constant across age .
-
-sel\_re Random effects on Recruitment only, fleet logistic selectivity
-RE model?
-
-q\_re Random effects on Recruitment only, one survey catchability RE
-model?
-
-Simulations were all carried out on the University of Massachusetts
-Green High-Performance Computing Cluster. Code for completing the
-simulations and summarization of results can be found at
-github.com/timjmiller/SSRTWG/Project\_0. We used the wham package
-version 1.X.X (commit 77bbd94).
-
-\hypertarget{results}{%
-\section{Results}\label{results}}
-
-Do each of these by type of operating model (Naa, M, sel, q) Convergence
-performance
-
-AIC performance
-
-SR estimation? M estimation?
-
-Bias, Mean Square error
-
-Certain basic parameters (stock-recruit pars, M, variance parameters)
-SSB, F, R
-
-\clearpage
-
-\hypertarget{numbers-at-age-operating-models}{%
-\subsection{Numbers at age operating
-models}\label{numbers-at-age-operating-models}}
-
-\hypertarget{estimating-models-include-alternative-random-effects-options-naa-m-sel-q}{%
-\subsubsection{Estimating models include alternative random effects
-options: NAA, M, sel,
-q}\label{estimating-models-include-alternative-random-effects-options-naa-m-sel-q}}
-
-\clearpage
-
-\begin{landscape}
-\begin{table}
-\caption{Distinguishing characteristics of the operating models with random effects on survival.
-Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
-Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.}
-{\footnotesize \input{naa_om_tab}}
-\end{table}
-
-\begin{table}
-\caption{Distinguishing characteristics of the operating models with random effects on natural mortality.
-Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
-Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
-For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}
-{\input{M_om_tab}}
-\end{table}
-
-\begin{table}
-\caption{Distinguishing characteristics of the operating models with random effects on selectivity.
-Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
-Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
-For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}
-{\input{Sel_om_tab}}
-\end{table}
-
-\begin{table}
-\caption{Distinguishing characteristics of the operating models with random effects on catchability.
-Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
-Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
-For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}
-{\input{q_om_tab}}
-\end{table}
-\end{landscape}
-
-\begin{table}
-\caption{Distinguishing characteristics of the estimating models.}
-{\scriptsize \input{em_tab}}
-\end{table}
-
-\begin{table}
-\caption{NAA operating models, estimating models all assume a B-H stock recruit relationship and M is fixed at the true value.}
-{\input{naa_om_em_SR_MF_aic_table}}
-\end{table}
-
-\begin{table}
-\caption{NAA operating models, estimating models all assume a B-H stock recruit relationship and M is estimated.}
-{\input{naa_om_em_SR_ME_aic_table}}
-\end{table}
-
-\begin{table}
-\caption{NAA operating models, estimating models all estimate a mean recruitment and M is fixed at the true value.}
-{\input{naa_om_em_R_MF_aic_table}}
-\end{table}
-
-\begin{table}
-\caption{NAA operating models, estimating models all estimate a mean recruitment and M estimated.}
-{\input{naa_om_em_R_MF_aic_table}}
-\end{table}
-\clearpage
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of for SSB for estimating models that estimate mean recruitment and M is fixed at the true value.}\label{naa_om_em_R_MF_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_R_MF_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of for SSB for estimating models that estimate mean recruitment and M is estimated.}\label{naa_om_em_R_ME_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_R_ME_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of for SSB for estimating models that estimates a BH stock-recruitment function and M is fixed at the true value.}\label{naa_om_em_SR_MF_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_SR_MF_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of for SSB for estimating models that estimates a BH stock-recruitment function and M is estimated.}\label{naa_om_em_SR_ME_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_SR_ME_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\hypertarget{estimating-models-include-naa-random-effects-and-estimation-assumes-mean-r-or-bh-sr}{%
-\subsubsection{Estimating models include NAA random effects and
-estimation assumes mean R or BH
-SR}\label{estimating-models-include-naa-random-effects-and-estimation-assumes-mean-r-or-bh-sr}}
-
-\clearpage
-
-\begin{table}
-\caption{Operating models and estimation models all assume RE on recruitment only, estimating models assume mean recruitment or a B-H stock recruit relationship and M is fixed at the true value.}
-{\input{rec_om_em_R_SR_MF_aic_table}}
-\end{table}
-
-\begin{table}
-\caption{Operating models and estimation models all assume RE on recruitment only, estimating models assume mean recruitment or a B-H stock recruit relationship and M is estimated.}
-{\input{rec_om_em_R_SR_ME_aic_table}}
-\end{table}
-
-\begin{table}
-\caption{Operating models and estimation models all assume RE on all abundances at age, estimating models assume mean recruitment or a B-H stock recruit relationship and M is fixed at the true value.}
-{\input{naa_om_em_R_SR_MF_aic_table}}
-\end{table}
-
-\begin{table}
-\caption{Operating models and estimation models all assume RE on all abundances at age, estimating models assume mean recruitment or a B-H stock recruit relationship and M is estimated.}
-{\input{naa_om_em_R_SR_ME_aic_table}}
-\end{table}
-\clearpage
-
-\begin{figure}
-\caption{Predicted probability of AIC being better for BH model.}\label{glm_res_NAA_re}
-\begin{center}
-\includegraphics[width = \textwidth]{pred_SR_best_NAA_oms.png}
-\end{center}
-\end{figure}
-
-\hypertarget{m-operating-models}{%
-\subsection{M operating models}\label{m-operating-models}}
-
-\hypertarget{estimating-models-include-naa-random-effects-and-estimation-assumes-mean-r-or-bh-sr-1}{%
-\subsubsection{Estimating models include NAA random effects and
-estimation assumes mean R or BH
-SR}\label{estimating-models-include-naa-random-effects-and-estimation-assumes-mean-r-or-bh-sr-1}}
-
-\begin{table}
-\caption{M random effects operating models.}
-{\input{M_om_em_R_BH_aic_table}}
-\end{table}
-\clearpage
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-The estimating models assumed variances of aggregate catch and index
-observations was known. This approximation may be appropriate for
-indices where we have a reliable estimate of uncertainty based on the
-survey design (), but there may be better approaches for the aggregate
-catch such as an informed prior on the standard errors with realistic
-bounds.
-
-\hypertarget{acknowledgements}{%
-\section*{Acknowledgements}\label{acknowledgements}}
-\addcontentsline{toc}{section}{Acknowledgements}
-
-This work was funded by NOAA Fisheries Northeast Fisheries Science
-Center.
-
-\pagebreak
-
-\bibliography{paper}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{0}{0}
-\end{CSLReferences}
-
-\pagebreak
-
-\clearpage
-
-\end{document}

### 2023-09-15 Ecov_study/mortality/paper/paper.tex
update work on project 0 and Ecov-M working papers

--- 2023-05-14 Project_0/paper/paper.tex
@@
 \hypersetup{
-  pdftitle={Factors affecting reliablity of state-space age-structured assessment models},
-  pdfauthor={Timothy J. Miller1, Britten, Brooks, Fay, Legault, Stock, et al.},
+  pdftitle={Factors affecting inferences on natural mortality and associated environmental effects},
+  pdfauthor={Timothy J. Miller1; Andrew Applegate; Greg Britten; Elizabeth N. Brooks; Gavin Fay; Alex Hansell; Christopher M. Legault; Brandon Muffley; Brian C. Stock; John Wiedenmann},
   hidelinks,
@@
 
-\title{Factors affecting reliablity of state-space age-structured
-assessment models}
-\author{Timothy J. Miller\textsuperscript{1}, Britten, Brooks, Fay,
-Legault, Stock, et al.}
+\title{Factors affecting inferences on natural mortality and associated
+environmental effects}
+\author{Timothy J. Miller\textsuperscript{1} \and Andrew
+Applegate \and Greg Britten \and Elizabeth N. Brooks \and Gavin
+Fay \and Alex Hansell \and Christopher M. Legault \and Brandon
+Muffley \and Brian C. Stock \and John Wiedenmann}
 \date{}
@@
 \hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
-
-Application of state-space models in fisheries stock assessment and
-management has expanded dramatically within ICES, Canada and the
-Northeast US \citep{nielsenberg14,cadigan16,stockmiller21}.
-
-Much is known about the reliability of state-space models that are
-linear or Gaussian \citep{aeberhardetal18}, but applications in
-fisheries management are nonlinear and typically include multiple types
-of observations with varying distributional assumptions. We know
-relatively little about the statistical reliability of such models.
-Also, there is a wide range of potential random effects structures in
-assessment models and we know little about the ability of information
-criteria to distinguish among such alternative structures.
-
-But those studies focus primarily on Gaussian linear models. review
-literature on reliability of hidden/latent process models. Primarily in
-other fields.
+\section*{Introduction}\label{introduction}}
+\addcontentsline{toc}{section}{Introduction}
 
 Here we conduct a simulation study with operating models varying by
-degree of observation error uncertainty, sources of process error (M,
-NAA, q, sel), and fishing history. The simulations from these operating
-models are fitted with estimating models that make alternative
-assumptions for sources of process error (M, NAA\textless{} q, sel),
-whether a stock-recruit model was estimated, and whether M is estimated.
-We evaluate whether AIC can correctly determine the correct source of
-process error and stock effects on recruitment. We also evaluate the
-degree of bias in the outputs of the assessment model that are important
-for management.
+degree of observation error uncertainty, sources of process error (M or
+survival), fishing history, temporal variation in environmental
+covariates, and magnitude of the effect of the covariate on natural
+mortality. The simulations from these operating models are fitted with
+estimating models that make alternative assumptions for sources of
+process error (M, NAA), and whether (mean) M is estimated. We evaluate
+whether AIC can correctly determine the correct source of process error
+and stock effects on recruitment. We also evaluate the degree of bias in
+the outputs of the assessment model that are important for management.
 
 \hypertarget{methods}{%
-\section{Methods}\label{methods}}
+\section*{Methods}\label{methods}}
+\addcontentsline{toc}{section}{Methods}
 
-Used the WHAM package \citep[][commit 77bbd94]{stockmiller21} (Miller
-and Stock 2020). This packages has also been used to configure operating
-and estimating models for closed loop simulations evaluating index-based
-assessment methods \citep{legaultetalinpress} and is used for management
-of haddock, butterfish, plaice, bluefish in the Northeast US.
+We used the WHAM package
+\citep[][commit 77bbd94]{millerstock20,stockmiller21} to configure all
+operating models and estimation models. This packages has also been used
+to configure operating and estimating models for closed loop simulations
+evaluating index-based assessment methods \citep{legaultetal23} and is
+used for management of haddock, butterfish, plaice, bluefish in the
+Northeast US.
 
-We completed a simulation study with a number of operating models that
-can be categorized based on where random effects are assumed: abundance
-at age, natural mortality, fleet selectivity, or index catchability. For
-each operating model assumptions about variance of process errors and
-observations are required and the values we used were based on a review
-of the range of estimates from recent applications of WHAM in management
-of stocks of haddock, butterfish, and American plaice in the NE US.
+We completed a simulation study with a number of operating models. Many
+of the characteristics of the operating models are the same as those
+used by \citet{milleretal_wp}.
 
-We simulated 100 data sets for each operating model. For each simulated
-data from each operating model we fit a set of estimating models.
-
-Y estimating models fit to each
+288 total operating models. We simulated 100 data sets for each
+operating model. For each simulated data set we fit a set of 12
+estimating models.
 
 \hypertarget{operating-models}{%
-\subsection{Operating models}\label{operating-models}}
-
-common to all:\\
-ages = 1 to 10+, M maturity
-
-marginal standard deviations for random effects are defined in tables of
-operating models.
+\subsection*{Operating models}\label{operating-models}}
+\addcontentsline{toc}{subsection}{Operating models}
 
 \hypertarget{population}{%
-\subsubsection{Population}\label{population}}
+\subsubsection*{Population}\label{population}}
+\addcontentsline{toc}{subsubsection}{Population}
 
-There are 10 age classes: ages 1 to 10+.
+The population model tracks 10 age classes: ages 1 to 10+ and we assume
+spawning occurs 1/4 of the way through the year. The maturity at age was
+a logistic curve with a50 = 2.89 and slope = 0.88 and assumed known in
+all estimation models (Figure \ref{om_maturity}).
 
-Spawning was assumed to occur 1/4 of the way through the year.
+\begin{figure}
+\caption{The proportion mature at age assumed for the population in all operating and estimating models.}\label{om_maturity}
+\begin{center}
+\includegraphics[width = \textwidth]{om_maturity.png}
+\end{center}
+\end{figure}
 
-Natural mortality rate was assumed 0.2 when it was constant and the mean
-of the time series process for operating models with M random effects.
-maturity a50 = 2.89, slope = 0.88
-
-Weight at age was generated with a LVB growth \[
+Weight at age was generated with a LVB growth function \[
 L_a = L_{\infty}\left(1 - e^{-k(a - t_0)}\right)
@@
 unfished recruitment are also constant over the time period for those
-operating models (Miller and Brooks 2021). We specified unfished
-recruitment = \(R_0 = e^{10}\) and \(F_{\text{MSY}} = F_{40} = 0.348\)
-equated to a steepness of 0.69 and \(\alpha=0.60\) and
+operating models \citep{millerbrooks21}. We specified unfished
+recruitment = \(R_0 = e^{10}\) and \(\Fmsy = \Fspr[40] = 0.348\) equated
+to a steepness of 0.69 and \(\alpha=0.60\) and
 \(\beta = 2.4 \times 10^{-5}\) for the \[
 N_{1,y} = \frac{\alpha \text{SSB}_{y-1}}{1 + \beta \text{SSB}_{y-1}} 
-\] Beverton-Holt parameterization.
+\] Beverton-Holt parameterization (Figure \ref{om_sr}). For OMs without
+process errors on natural mortality we assumed the rate was assumed 0.2.
+For OMs with process errors on natural mortality the mean rate was 0.2.
 
-The magnitude of the overfishing assumptions is based on average
-estimates of overfishing for NE groundfish stocks from Wiedenmann et
-al.~(20XX). Legault et al.~(2023) also used similar approaches to
-defining fishing mortality histories for operating models.
+\begin{figure}
+\caption{The Beverton-Holt stock recruit relationship assume for all operating models.}\label{om_sr}
+\begin{center}
+\includegraphics[width = \textwidth]{om_sr.png}
+\end{center}
+\end{figure}
 
-Currenlty, initial population is configured at the equilibrium
-distribution fishing at \(F = 2.5\times F_{\text{MSY}}\).
+Two alternative fishing histories were used for operating models. In the
+first scenario, the stock experiences overfishing for the first 20 years
+and fishing at \Fmsy for the last 20 years. In the second scenario, the
+stock is fished at \Fmsy for the entire time period. The magnitude of
+the overfishing assumptions is based on average estimates of overfishing
+for NE groundfish stocks from \citet{wiedenmannetal19}.
+\Citet{legaultetal23} also used similar approaches to defining fishing
+mortality histories for operating models.
 
 Initial population was configured at the equilibrium distribution
-fishing at either \(F = 2.5\times F_{\text{MSY}}\) or
-\(F = F_{\text{MSY}}\) for the two alternative fishing histories. That
-is for a deterministic model, the age composition would not change over
-time when the fishing mortality was constant at the respective level.
+fishing at either \(F = 2.5\times \Fmsy\) or \(F = \Fmsy\) for the two
+alternative fishing histories. That is for a deterministic model, the
+age composition would not change over time when the fishing mortality
+was constant at the respective level.
 
-For operating models with time-varying random effects for M, steepness
-is not constant, but we used the same alpha and beta parameters as other
-operating models this equates to a steepness and R0 at the mean of the
-time series process for M. For operating models with time-varying random
-effects for fishery selectivity, Fmsy is also not constant however we
-use the same F history as other operating models which corresponds to
-Fmsy at the mean selectivity parameters.
+For operating models with time-varying random effects or covariate
+effects on M, steepness is not constant, but we used the same alpha and
+beta parameters as other operating models this equates to a steepness
+and R0 at the mean of the time series process for M. For operating
+models with time-varying random effects for fishery selectivity,
+\Fmsy is also not constant however we use the same F history as other
+operating models which corresponds to Fmsy at the mean selectivity
+parameters.
+
+\hypertarget{environmental-covariate}{%
+\subsubsection*{Environmental covariate}\label{environmental-covariate}}
+\addcontentsline{toc}{subsubsection}{Environmental covariate}
+
+In the WHAM model, environmental covariates are assumed to be described
+as state-space processes with annual observations of the true latent
+covariate. The latent covariate is assumed to be Gaussian AR1 \[
+E_y|E_{y-1} \sim \text{N}\left(\mu_\text{Ecov}\left(1-\rho_\text{Ecov}\right) + \rho_\text{Ecov} E_{y-1}, \sigma^2_\text{Ecov}\right)
+\] with marginal mean \(\mu_\text{Ecov}\) and variance
+\(\frac{\sigma^2_\text{Ecov}}{1-\rho_\text{Ecov}^2}\). The configuration
+of the latent environmental covariate in the operating models have one
+of two alternate assumptions about the marginal standard deviation and
+autocorrelation parameter:
+
+\begin{itemize}
+\item marginal standard deviation $\in \{0.1, 0.5\}$
+\item AR1 correlation $\rho_\text{Ecov} \in \{0, 0.9\}$
+\end{itemize}
+
+The observations of the latent environmental covariate are assumed to be
+unbiased and Gaussian \[
+\text{ecov}_y|E_y \sim \text{N}\left(E_y,\sigma^2_\text{ecov}\right)
+\] The standard deviation of the environmental observations in the
+operating models is one of two values:
+
+\begin{itemize}
+\item standard deviation $\in \{0.1, 0.5\}$
+\end{itemize}
+
+Figure \ref{om_ecov_example} provides example simulations of the latent
+process and observations under the alternative configurations.
+
+\begin{figure}
+\caption{Example simulations of environmental covariate latent processes and observations with different levels of observation error, and different assumptions about variability of the latent process.}\label{om_ecov_example}
+\begin{center}
+\includegraphics[width = \textwidth]{Ecov_true_obs_example.png}
+\end{center}
+\end{figure}
+
+\hypertarget{random-effects-on-recruitment-survival-and-natural-mortality}{%
+\subsubsection*{Random effects on recruitment, survival, and natural
+mortality}\label{random-effects-on-recruitment-survival-and-natural-mortality}}
+\addcontentsline{toc}{subsubsection}{Random effects on recruitment,
+survival, and natural mortality}
+
+Internally, WHAM treats natural mortality as a log-transformed parameter
+\[
+\log M_y = \beta_M + \beta_{\text{Ecov}} E_y + \epsilon_{M,y}
+\] that is a linear combination of a mean \(\beta_M\) and any annual
+covariate \(\beta_{\text{Ecov}}\) and random effects marginally
+distributed as
+\(\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)\).
+
+The covariate effect is one of 3 alternative values in the operating
+models:
+
+\begin{itemize}
+\item $\beta_\text{Ecov} \in \{0,0.25,0.5\}$
+\end{itemize}
+
+The alternative configurations of random effects on recruitment, natural
+mortality, survival
+
+\begin{itemize}
+\item uncorrelated random effects on R and survival with $\sigma_R = 0.5$, marginal standard deviation of survival of older age clases of 0.3
+\item uncorrelated random effects on R and survival with $\sigma_R = 0.5$, marginal standard deviation of natural mortality of 0.3
+\end{itemize}
 
 \hypertarget{fleets}{%
-\subsubsection{Fleets}\label{fleets}}
+\subsubsection*{Fleets}\label{fleets}}
+\addcontentsline{toc}{subsubsection}{Fleets}
 
@@
 with logistic selectivity for the fleet with \(a_{50} = 5\) and slope =
-1. This selectivity is was used to define \(F_{\text{MSY}}\) for the
-Beverton-Holt stock recruitment parameters above. We assumed a
-logistic-normal distribution for the age-composition observations for
-the fleet.
+1 (Figure \ref{om_mean_selectivity}). This selectivity is was used to
+define \Fmsy for the Beverton-Holt stock recruitment parameters above.
+We assumed a logistic-normal distribution for the age-composition
+observations for the fleet.
 
-\hypertarget{fishing-histories}{%
-\subsubsection{Fishing histories}\label{fishing-histories}}
-
-All operating models assumed one of two different fishing histories. One
-: Fishing mortality is equal to Fmsy (0.348) for the whole 40 year
-period. Two : Fishing mortality is 2.5 times Fmsy for the first 20 years
-then changes to Fmsy for the last 20 years.
+\begin{figure}
+\caption{The selectivity at age assumed for the fishing fleet in operating models without selectivity random effects.
+The mean selectivity at age across time for the fishing fleet in operating models with selectivity random effects.}\label{om_mean_selectivity}
+\begin{center}
+\includegraphics[width = \textwidth]{om_mean_selectivity.png}
+\end{center}
+\end{figure}
 
 \hypertarget{indices}{%
-\subsubsection{Indices}\label{indices}}
+\subsubsection*{Indices}\label{indices}}
+\addcontentsline{toc}{subsubsection}{Indices}
 
@@
 
-\hypertarget{operating-models-with-random-effects-on-numbers-at-age}{%
-\subsubsection{Operating models with random effects on numbers at
-age}\label{operating-models-with-random-effects-on-numbers-at-age}}
-
-24 operating models, 16 Sel re operating models and 16 q re operating
-models. Table of process error assumptions
-
-\hypertarget{operating-models-with-random-effects-on-natural-mortality}{%
-\subsubsection{Operating models with random effects on natural
-mortality}\label{operating-models-with-random-effects-on-natural-mortality}}
-
-16 operating models Table of process error assumptions
-
-NOTE: inv\_trans\_rho function in set\_M.R is mis-defined. Will affect
-correlation parameters assigned in operating models?
-
-Steepness and BRPs are not constant when M is time-varying (Miller and
-Brooks 2021). We uses the a/b parameters for the B-H defined for Fmsy =
-F40 at the mean M = 0.2 as defined above.
-
-\hypertarget{operating-models-with-random-effects-on-fleet-selectivity}{%
-\subsubsection{Operating models with random effects on fleet
-selectivity}\label{operating-models-with-random-effects-on-fleet-selectivity}}
-
-16 operating models Table of process error assumptions
-
-BRPs are not constant when fleet selectivity is time-varying. We uses
-the a/b parameters for the B-H defined for Fmsy = F40 at the mean of the
-time series model for selectivity which is the same as the constant
-selectivity defined above.
-
-\hypertarget{operating-models-with-random-effects-on-index-catchability}{%
-\subsubsection{Operating models with random effects on index
-catchability}\label{operating-models-with-random-effects-on-index-catchability}}
-
-16 operating models Table of process error assumptions
+The factorial combinations of fishing histories, index and age
+composition observation error assumptions, latent environmental
+covariate and observation assumptions, covariate effects on natural
+mortality, and sources of process error (R, R+S, R+M) results in 288
+different operating models.
 
 \hypertarget{estimating-models}{%
-\subsection{Estimating models}\label{estimating-models}}
+\subsection*{Estimating models}\label{estimating-models}}
+\addcontentsline{toc}{subsection}{Estimating models}
 
-32 estimating models Table of estimating models
+For each data set simulated from an operating model 12 estimating models
+were fit. There were three factors defining the configuration of each
+estimating model
 
-1-20 fit to each NAA RE operating model 5-24 fit to each M RE operating
-model 5-20,25-28 to each sel RE operating model 5-20, 29-32 to each q RE
-operating model
+\begin{itemize}
+\item whether the mean natural mortality $\beta_M$ was estimated or assumed known ($\log 0.2$)
+\item whether an environmental effect $\beta_\text{Ecov}$ was estimated or not (fixed at 0)
+\item whether the process errors were just on recruitment (R), on recruitment and survival (R+S) or on recruitment and natural mortality (R+M)
+\end{itemize}
 
-SR estimation or not
-
-Make plot of S-R curve, Fmsy = F40 Initial values for BH parameters are
-the true values. Initial values for mean R model = true R0.
-
-M estimation or not
-
-NAA\_re Random effects on Recruitment only or random effects on
-recruitment and transitions among older numbers at age.
-
-M\_re Random effects on Recruitment only, M constant across age .
-
-sel\_re Random effects on Recruitment only, fleet logistic selectivity
-RE model?
-
-q\_re Random effects on Recruitment only, one survey catchability RE
-model?
+The configuration of the process errors in the estimating models matched
+the corresponding options in the operating models. For example, if R+S
+was assumed for both the estimating and operating model, the assumptions
+about process errors matched. The environmental covariate observations
+were included in all estimation models to ensure comparability of
+information criteria. The observation error variance of the
+environmental observations and aggregate catch and indices were all
+assumed known at the true values.
 
@@
 simulations and summarization of results can be found at
-github.com/timjmiller/SSRTWG/Project\_0. We used the wham package
-version 1.X.X (commit 77bbd94).
+github.com/timjmiller/SSRTWG/ecov\_study/mortality. We used the wham
+package version 1.X.X (commit 77bbd94).
+
+\hypertarget{measures-of-reliability}{%
+\subsection*{Measures of reliability}\label{measures-of-reliability}}
+\addcontentsline{toc}{subsection}{Measures of reliability}
+
+The first measure of reliability we investigated was frequency of
+convergence when fitting each estimating model to the simulated data
+sets. There are various ways to assess convergence of the fit. We
+summarized 5 alternative categories of convergence. 1. Did the
+optimization routine (stats::nlminb) complete without error? 2. Did the
+stats::nlminb convergence flag = 0 indicate successful convergence? 3.
+Was the maximum absolute value of the gradient of the log-likelihood
+\textless{} \(1\times10^{-6}\)? 4. Did TMB::sdreport provide non-NA
+values for all fixed effects standard errors? 5. Did TMB::sdreport
+provide all standard errors \textless{} 100?
+
+The first convergence criterion assesses whether the model crashes. The
+third is a measure that assesses how flat the likelihood is at the
+optimized point of the likelihood surface. The fourth and fifth criteria
+are specific to the hessian-based standard error reporting provided by
+TMB. the TMB::sdreport function will sometimes return standard error
+estimates even when the calculated hessian is not invertible (fourth
+criterion). It will also provide large standard errors for some
+parameters that are not estimated well or are near bounds on the
+transformed scale that is of primary interest. For example, variance
+parameters for random effects can often be estimated near 0 when the fit
+suggests no variation in the random effects or, equivalently, that the
+model without random effects is better.
+
+\hypertarget{aic-for-model-selection}{%
+\subsubsection*{AIC for model selection}\label{aic-for-model-selection}}
+\addcontentsline{toc}{subsubsection}{AIC for model selection}
+
+We measured the frequency of correct model selection using marginal AIC.
+For a given operating model we the set of models that were considered
+all made the same assumptions on whether or not to estimate M and
+whether a stock-recruit function is assumed.
+
+\hypertarget{bias}{%
+\subsubsection*{Bias}\label{bias}}
+\addcontentsline{toc}{subsubsection}{Bias}
+
+We calculated relative errors of various quantities for each fitted
+estimation model for the \(i\)th simulated data set as \[
+{\text{RE}_i}\left(\theta\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
+\] and measured bias as the median relative error and constructed 95\%
+confidence intervals using the binomial distribution approach as in
+\citet{stockmiller21}.
 
 \hypertarget{results}{%
-\section{Results}\label{results}}
+\section*{Results}\label{results}}
+\addcontentsline{toc}{section}{Results}
 
-Do each of these by type of operating model (Naa, M, sel, q) Convergence
-performance
-
-AIC performance
-
-SR estimation? M estimation?
-
-Bias, Mean Square error
-
-Certain basic parameters (stock-recruit pars, M, variance parameters)
-SSB, F, R
-
-\clearpage
-
-\hypertarget{numbers-at-age-operating-models}{%
-\subsection{Numbers at age operating
-models}\label{numbers-at-age-operating-models}}
-
-\hypertarget{estimating-models-include-alternative-random-effects-options-naa-m-sel-q}{%
-\subsubsection{Estimating models include alternative random effects
-options: NAA, M, sel,
-q}\label{estimating-models-include-alternative-random-effects-options-naa-m-sel-q}}
-
-\clearpage
-
-\begin{landscape}
-\begin{table}
-\caption{Distinguishing characteristics of the operating models with random effects on survival.
-Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
-Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.}
-{\footnotesize \input{naa_om_tab}}
-\end{table}
-
-\begin{table}
-\caption{Distinguishing characteristics of the operating models with random effects on natural mortality.
-Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
-Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
-For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}
-{\input{M_om_tab}}
-\end{table}
-
-\begin{table}
-\caption{Distinguishing characteristics of the operating models with random effects on selectivity.
-Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
-Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
-For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}
-{\input{Sel_om_tab}}
-\end{table}
-
-\begin{table}
-\caption{Distinguishing characteristics of the operating models with random effects on catchability.
-Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
-Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
-For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}
-{\input{q_om_tab}}
-\end{table}
-\end{landscape}
-
-\begin{table}
-\caption{Distinguishing characteristics of the estimating models.}
-{\scriptsize \input{em_tab}}
-\end{table}
-
-\begin{table}
-\caption{NAA operating models, estimating models all assume a B-H stock recruit relationship and M is fixed at the true value.}
-{\input{naa_om_em_SR_MF_aic_table}}
-\end{table}
-
-\begin{table}
-\caption{NAA operating models, estimating models all assume a B-H stock recruit relationship and M is estimated.}
-{\input{naa_om_em_SR_ME_aic_table}}
-\end{table}
-
-\begin{table}
-\caption{NAA operating models, estimating models all estimate a mean recruitment and M is fixed at the true value.}
-{\input{naa_om_em_R_MF_aic_table}}
-\end{table}
-
-\begin{table}
-\caption{NAA operating models, estimating models all estimate a mean recruitment and M estimated.}
-{\input{naa_om_em_R_MF_aic_table}}
-\end{table}
-\clearpage
+\hypertarget{convergence-performance}{%
+\subsection*{Convergence performance}\label{convergence-performance}}
+\addcontentsline{toc}{subsection}{Convergence performance}
 
@@
 \begin{figure}
-\caption{Median relative bias of for SSB for estimating models that estimate mean recruitment and M is fixed at the true value.}\label{naa_om_em_R_MF_relbias_ssb}
+\caption{Probability of convergence of estimating models with alternative process error assumptions fit to simulated populations and data sets from operating models.
+The convergence criterion is an invertible hessian providing standard error estimates for all fixed effects parameters.
+Vertical lines represent 95\% confidence intervals.
+All estimating models estimate environmental covariate effects on natural mortality but assume the mean mortality parameter $\beta_\text{M} = \log 0.2$.}\label{convergence_M_fixed}
 \begin{center}
-\includegraphics[width = \textwidth]{naa_om_R_MF_relbias_ssb.png}
+\includegraphics[width = \textwidth]{proportion_good_hessian_ecov_effect_est_M_fixed.png}
 \end{center}
@@
 \begin{figure}
-\caption{Median relative bias of for SSB for estimating models that estimate mean recruitment and M is estimated.}\label{naa_om_em_R_ME_relbias_ssb}
+\caption{Probability of convergence of estimating models with alternative process error assumptions fit to simulated populations and data sets from operating models.
+The convergence criterion is an invertible hessian providing standard error estimates for all fixed effects parameters.
+Vertical lines represent 95\% confidence intervals.
+All estimating models estimate environmental covariate effects on natural mortality and the mean mortality parameter $\beta_\text{M}$.}\label{convergence_M_estimated}
 \begin{center}
-\includegraphics[width = \textwidth]{naa_om_R_ME_relbias_ssb.png}
+\includegraphics[width = \textwidth]{proportion_good_hessian_ecov_effect_est_M_est.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\hypertarget{aic-performance}{%
+\subsection*{AIC performance}\label{aic-performance}}
+\addcontentsline{toc}{subsection}{AIC performance}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets for each operating model where the estimation model with the lowest AIC had correct assumptions of process errors (R,R+S,R+M) and/or about whether the environmental covariate affected natural mortality.
+All estimating models assumed the mean mortality parameter $\beta_\text{M} = \log 0.2$.}\label{aic_rank_M_fixed}
+\begin{center}
+\includegraphics[width = \textwidth]{proportion_correct_PE_effect_M_fixed.png}
 \end{center}
@@
 \begin{figure}
-\caption{Median relative bias of for SSB for estimating models that estimates a BH stock-recruitment function and M is fixed at the true value.}\label{naa_om_em_SR_MF_relbias_ssb}
+\caption{Proportion of simulated data sets for each operating model where the estimation model with the lowest AIC had correct assumptions of process errors (R,R+S,R+M) and/or about whether the environmental covariate affected natural mortality.
+All estimating models estimate the mean mortality parameter $\beta_\text{M}$.}\label{aic_rank_M_estimated}
 \begin{center}
-\includegraphics[width = \textwidth]{naa_om_SR_MF_relbias_ssb.png}
+\includegraphics[width = \textwidth]{proportion_correct_PE_effect_M_estimated.png}
 \end{center}
@@
 
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of for SSB for estimating models that estimates a BH stock-recruitment function and M is estimated.}\label{naa_om_em_SR_ME_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_SR_ME_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\hypertarget{estimating-models-include-naa-random-effects-and-estimation-assumes-mean-r-or-bh-sr}{%
-\subsubsection{Estimating models include NAA random effects and
-estimation assumes mean R or BH
-SR}\label{estimating-models-include-naa-random-effects-and-estimation-assumes-mean-r-or-bh-sr}}
-
-\clearpage
-
-\begin{table}
-\caption{Operating models and estimation models all assume RE on recruitment only, estimating models assume mean recruitment or a B-H stock recruit relationship and M is fixed at the true value.}
-{\input{rec_om_em_R_SR_MF_aic_table}}
-\end{table}
-
-\begin{table}
-\caption{Operating models and estimation models all assume RE on recruitment only, estimating models assume mean recruitment or a B-H stock recruit relationship and M is estimated.}
-{\input{rec_om_em_R_SR_ME_aic_table}}
-\end{table}
-
-\begin{table}
-\caption{Operating models and estimation models all assume RE on all abundances at age, estimating models assume mean recruitment or a B-H stock recruit relationship and M is fixed at the true value.}
-{\input{naa_om_em_R_SR_MF_aic_table}}
-\end{table}
-
-\begin{table}
-\caption{Operating models and estimation models all assume RE on all abundances at age, estimating models assume mean recruitment or a B-H stock recruit relationship and M is estimated.}
-{\input{naa_om_em_R_SR_ME_aic_table}}
-\end{table}
-\clearpage
-
-\begin{figure}
-\caption{Predicted probability of AIC being better for BH model.}\label{glm_res_NAA_re}
-\begin{center}
-\includegraphics[width = \textwidth]{pred_SR_best_NAA_oms.png}
-\end{center}
-\end{figure}
-
-\hypertarget{m-operating-models}{%
-\subsection{M operating models}\label{m-operating-models}}
-
-\hypertarget{estimating-models-include-naa-random-effects-and-estimation-assumes-mean-r-or-bh-sr-1}{%
-\subsubsection{Estimating models include NAA random effects and
-estimation assumes mean R or BH
-SR}\label{estimating-models-include-naa-random-effects-and-estimation-assumes-mean-r-or-bh-sr-1}}
-
-\begin{table}
-\caption{M random effects operating models.}
-{\input{M_om_em_R_BH_aic_table}}
-\end{table}
-\clearpage
+\hypertarget{bias-1}{%
+\subsection*{Bias}\label{bias-1}}
+\addcontentsline{toc}{subsection}{Bias}
 
 \hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
+\section*{Discussion}\label{discussion}}
+\addcontentsline{toc}{section}{Discussion}
 

### 2023-09-15 Project_0/paper/paper.tex
update work on project 0 and Ecov-M working papers

@@
 \begin{figure}
-\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R and R+S process errors.
+\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R and R+S process errors.
 Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{naa_om_em_R_MF_relbias_ssb}
@@
 \begin{figure}
-\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R and R+S process errors.
+\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R and R+S process errors.
 Estimation models do not assume a stock-recruit relationship and estimate M.}\label{naa_om_em_R_ME_relbias_ssb}
@@
 \begin{figure}
-\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R and R+S process errors.
+\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R and R+S process errors.
 Estimation models assume a BH stock-recruitment function and M is fixed at the true value.}\label{naa_om_em_BH_MF_relbias_ssb}
@@
 \begin{figure}
-\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R and R+S process errors.
+\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R and R+S process errors.
 Estimation models assume a BH stock-recruitment function and estimate M.}\label{naa_om_em_BH_ME_relbias_ssb}
@@
 \begin{figure}
-\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+M process errors.
+\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+M process errors.
 Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{M_om_em_R_MF_relbias_ssb}
@@
 \begin{figure}
-\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+M process errors.
+\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+M process errors.
 Estimation models do not assume a stock-recruit relationship and estimate M.}\label{M_om_em_R_ME_relbias_ssb}
@@
 \begin{figure}
-\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+M process errors.
+\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+M process errors.
 Estimation models assume a BH stock-recruitment function and M is fixed at the true value.}\label{M_om_em_BH_MF_relbias_ssb}
@@
 \begin{figure}
-\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+M process errors.
+\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+M process errors.
 Estimation models assume a BH stock-recruitment function and estimate M.}\label{M_om_em_BH_ME_relbias_ssb}
@@
 \begin{figure}
-\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+Sel process errors.
+\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+Sel process errors.
 Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{Sel_om_em_R_MF_relbias_ssb}
@@
 \begin{figure}
-\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+Sel process errors.
+\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+Sel process errors.
 Estimation models do not assume a stock-recruit relationship and estimate M.}\label{Sel_om_em_R_ME_relbias_ssb}
@@
 \begin{figure}
-\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+Sel process errors.
+\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+Sel process errors.
 Estimation models assume a BH stock-recruitment function and M is fixed at the true value.}\label{Sel_om_em_BH_MF_relbias_ssb}
@@
 \begin{figure}
-\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+Sel process errors.
+\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+Sel process errors.
 Estimation models assume a BH stock-recruitment function and estimate M.}\label{Sel_om_em_BH_ME_relbias_ssb}
@@
 \begin{figure}
-\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+q process errors.
+\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+q process errors.
 Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{q_om_em_R_MF_relbias_ssb}
@@
 \begin{figure}
-\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+q process errors.
+\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+q process errors.
 Estimation models do not assume a stock-recruit relationship and estimate M.}\label{q_om_em_R_ME_relbias_ssb}
@@
 \begin{figure}
-\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+q process errors.
+\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+q process errors.
 Estimation models assume a BH stock-recruitment function and M is fixed at the true value.}\label{q_om_em_BH_MF_relbias_ssb}
@@
 \begin{figure}
-\caption{Median relative bias of for SSB for estimating models fitted to data sets simulated with R+q process errors.
+\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+q process errors.
 Estimation models assume a BH stock-recruitment function and estimate M.}\label{q_om_em_BH_ME_relbias_ssb}
@@
 \includegraphics[width = \textwidth]{q_om_BH_ME_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\hypertarget{beverton-holt-parameter-estimation}{%
+\subsubsection{Beverton-Holt parameter
+estimation}\label{beverton-holt-parameter-estimation}}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of Beverton-Holt stock recruit parameters for estimating models with correct process error assumptions fitted to data sets simulated with R and R+S process errors.
+Estimation models either assume the correct natural mortality rate or estimate it.}\label{naa_om_SR_relbias}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_SR_relerror.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of Beverton-Holt stock recruit parameters for estimating models with correct process error assumptions fitted to data sets simulated with R+M process errors.
+Estimation models either assume the correct natural mortality rate or estimate it.}\label{M_om_SR_relbias}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_SR_relerror.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of Beverton-Holt stock recruit parameters for estimating models with correct process error assumptions fitted to data sets simulated with R+Sel process errors.
+Estimation models either assume the correct natural mortality rate or estimate it.}\label{Sel_om_SR_relbias}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_SR_relerror.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of Beverton-Holt stock recruit parameters for estimating models with correct process error assumptions fitted to data sets simulated with R+q process errors.
+Estimation models either assume the correct natural mortality rate or estimate it.}\label{q_om_SR_relbias}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_SR_relerror.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\hypertarget{mean-natural-mortality-rate-estimation}{%
+\subsubsection{(Mean) natural mortality rate
+estimation}\label{mean-natural-mortality-rate-estimation}}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of (mean) natural mortality rate for estimating models with correct process error assumptions fitted to data sets simulated with R and R+S process errors.
+Estimation models assumed either no stock-recruit relationship or estimated a Beverton-Holt relationship.}\label{naa_om_M_relbias}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_M_relerror.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of (mean) natural mortality rate for estimating models with correct process error assumptions fitted to data sets simulated with R+M process errors.
+Estimation models assumed either no stock-recruit relationship or estimated a Beverton-Holt relationship.}\label{M_om_M_relbias}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_M_relerror.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of (mean) natural mortality rate for estimating models with correct process error assumptions fitted to data sets simulated with R+Sel process errors.
+Estimation models assumed either no stock-recruit relationship or estimated a Beverton-Holt relationship.}\label{Sel_om_M_relbias}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_M_relerror.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of (mean) natural mortality rate for estimating models with correct process error assumptions fitted to data sets simulated with R+q process errors.
+Estimation models assumed either no stock-recruit relationship or estimated a Beverton-Holt relationship.}\label{q_om_M_relbias}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_M_relerror.png}
 \end{center}

### 2023-09-15 Ecov_study/mortality/code/aic_ecov_effect_results.R

@@ -60,8 +60,8 @@ aic_rank_df_fn <- function(df.ems, df.oms, M_est = FALSE) {
   all_res[facs] <- lapply(all_res[facs], factor)
   all_res_mod <- all_res %>%
     mutate(Ecov_obs_sig = recode(Ecov_obs_sig,
-      "0.1" = "Ecov obs SD = 0.1",
-      "0.5" = "Ecov obs SD = 0.5"
+      "0.1" = "sigma[ecov] == 0.1",
+      "0.5" = "sigma[ecov] == 0.5"
     ))
   all_res_mod <- all_res_mod %>%
     mutate(obs_error = recode(obs_error,
@@ -70,10 +70,25 @@ aic_rank_df_fn <- function(df.ems, df.oms, M_est = FALSE) {
     ))
   all_res_mod <- all_res_mod %>%
     mutate(NAA_M_re = recode(NAA_M_re,
-      "rec" = "R",
-      "rec+1" = "R+S",
-      "rec+M" = "R+M"
+      "rec" = "OM=EM: R",
+      "rec+1" = "OM=EM: R+S",
+      "rec+M" = "OM=EM: R+M"
     ))
+  all_res_mod <- all_res_mod %>%
+    mutate(Fhist = recode(Fhist,
+      "H-MSY" = "High->FMSY",
+      "MSY" = "FMSY"
+    ))
+  # all_res_mod <- all_res_mod %>%
+  #   mutate(Ecov_re_sig = recode(Ecov_re_sig,
+  #     "0.1" = "sigma[Ecov] == 0.1",
+  #     "0.5" = "sigma[Ecov] == 0.5"
+  #   ))
+  # all_res_mod <- all_res_mod %>%
+  #   mutate(Ecov_re_cor = recode(Ecov_re_cor,
+  #     "0" = "rho[Ecov] == 0",
+  #     "0.5" = "rho[Ecov] == 0.5"
+  #   ))
   return(all_res_mod)
 }
 
@@ -81,28 +96,32 @@ all_res <- aic_rank_df_fn(df.ems, df.oms, M_est = FALSE)
 
 plt <- ggplot(all_res, aes(x = Ecov_effect, y = emp_p, colour = Ecov_re_sig:Ecov_re_cor)) + scale_colour_viridis_d() + 
     geom_line(position = position_dodge(0.1), linewidth = 1) + geom_point(position = position_dodge(0.1), size = 4) + 
[…]

### 2023-09-15 Project_0/code/relbias_M_plots.R

@@ -1,5 +1,5 @@
-x = readRDS(file.path(here::here(),"Project_0", "results", paste0("naa","_om"), paste0("om_", 1), paste0("sim_",1,".RDS")))
-get_SR_estimates_fn <- function(om_type = "naa"){
+x = readRDS(here::here("Project_0", "results", paste0("naa","_om"), paste0("om_", 1), paste0("sim_",1,".RDS")))
+get_M_estimates_fn <- function(om_type = "naa"){
   df.ems = readRDS(here::here("Project_0","inputs", "df.ems.RDS"))
   df.oms = readRDS(here("Project_0","inputs", paste0("df.", ifelse(om_type == "naa", "", paste0(om_type,".")),"oms.RDS")))
   if(om_type == "naa") {
@@ -12,10 +12,11 @@ get_SR_estimates_fn <- function(om_type = "naa"){
     em_ind <- c(5:20,25:28)
   }
   if(om_type == "q") {
-    em_ind <- c(5:20,28:32)
+    em_ind <- c(5:20,29:32)
   }
   df.ems <- df.ems[em_ind,]
-
+  print(em_ind)
+  print(dim(df.ems))
   all_out <- lapply(1:NROW(df.oms), function(y){
     if(om_type == "naa") {
       if(is.na(df.oms$NAA_sig[y])) em_ind <- 1:4
@@ -32,31 +33,33 @@ get_SR_estimates_fn <- function(om_type = "naa"){
       else em_ind <- which(df.ems$re_config == "sel_re" & df.ems$sel_re_cor == "ar1_y")
     }
     if(om_type == "q") {
+      print(df.oms)
       if(df.oms$q_cor[y] == 0) em_ind <- which(df.ems$re_config == "q_re" & df.ems$q_re_cor == "iid")
-      else em_ind <- which(df.ems$re_config == "q_re" & df.ems$M_re_cor == "ar1")
+      else em_ind <- which(df.ems$re_config == "q_re" & df.ems$q_re_cor == "ar1")
+      print("q")
+      print(em_ind)
     }
-    em_ind <- em_ind[which(df.ems$SR_model[em_ind] == 3)]
+    em_ind <- em_ind[which(df.ems$M_est[em_ind])] #M is estimated, 2 ems: BH estimated or not
+      print(em_ind)
+      print(df.ems[em_ind,])
     res <- lapply(1:100, function(x){
       print(paste0("om_", y, ", sim_",x))
[…]

### 2023-09-15 Project_0/paper/paper.Rmd

@@
   - Timothy J. Miller^1^
-  - A. Applegate 
-  - G. Britten
-  - E. N. Brooks
-  - G. Fay
-  - A. Hansell
-  - C. M. Legault
-  - B. Muffley
-  - J. Wiedenmann
+  - Andrew Applegate 
+  - Greg Britten
+  - Elizabeth N. Brooks
+  - Gavin Fay
+  - Alex Hansell
+  - Christopher M. Legault
+  - Brandon Muffley
+  - Brian C. Stock
+  - John Wiedenmann
 date: ''
@@
 
-SR estimation or not
-
-Initial values for BH parameters are the true values.
-Initial values for mean R model = true R0.
-
-M estimation or not
-
-NAA_re
-Random effects on Recruitment only or random effects on recruitment and transitions among older numbers at age.
-
-M_re
-Random effects on Recruitment only, M constant across age .
-
-sel_re
-Random effects on Recruitment only, fleet logistic selectivity RE model?
-
-q_re
-Random effects on Recruitment only, one survey catchability RE model?
+The observation error variance of the environmental observations and aggregate catch and indices were all assumed known at the true values.
 
@@
 \includegraphics[width = \textwidth]{q_om_BH_ME_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+### Beverton-Holt parameter estimation
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of Beverton-Holt stock recruit parameters for estimating models with correct process error assumptions fitted to data sets simulated with R and R+S process errors.
+Estimation models either assume the correct natural mortality rate or estimate it.}\label{naa_om_SR_relbias}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_SR_relerror.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of Beverton-Holt stock recruit parameters for estimating models with correct process error assumptions fitted to data sets simulated with R+M process errors.
+Estimation models either assume the correct natural mortality rate or estimate it.}\label{M_om_SR_relbias}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_SR_relerror.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of Beverton-Holt stock recruit parameters for estimating models with correct process error assumptions fitted to data sets simulated with R+Sel process errors.
+Estimation models either assume the correct natural mortality rate or estimate it.}\label{Sel_om_SR_relbias}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_SR_relerror.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of Beverton-Holt stock recruit parameters for estimating models with correct process error assumptions fitted to data sets simulated with R+q process errors.
+Estimation models either assume the correct natural mortality rate or estimate it.}\label{q_om_SR_relbias}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_SR_relerror.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+### (Mean) natural mortality rate estimation
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of (mean) natural mortality rate for estimating models with correct process error assumptions fitted to data sets simulated with R and R+S process errors.
+Estimation models assumed either no stock-recruit relationship or estimated a Beverton-Holt relationship.}\label{naa_om_M_relbias}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_M_relerror.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of (mean) natural mortality rate for estimating models with correct process error assumptions fitted to data sets simulated with R+M process errors.
+Estimation models assumed either no stock-recruit relationship or estimated a Beverton-Holt relationship.}\label{M_om_M_relbias}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_M_relerror.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of (mean) natural mortality rate for estimating models with correct process error assumptions fitted to data sets simulated with R+Sel process errors.
+Estimation models assumed either no stock-recruit relationship or estimated a Beverton-Holt relationship.}\label{Sel_om_M_relbias}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_M_relerror.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of (mean) natural mortality rate for estimating models with correct process error assumptions fitted to data sets simulated with R+q process errors.
+Estimation models assumed either no stock-recruit relationship or estimated a Beverton-Holt relationship.}\label{q_om_M_relbias}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_M_relerror.png}
 \end{center}

### 2023-09-15 Ecov_study/mortality/paper/paper.Rmd

--- 2023-09-15 Ecov_study/mortality/paper/paper.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[
-  12pt,
-]{article}
-\usepackage{amsmath,amssymb}
-\usepackage{lmodern}
-\usepackage{iftex}
-\ifPDFTeX
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={Factors affecting inferences on natural mortality and associated environmental effects},
-  pdfauthor={Timothy J. Miller1; Andrew Applegate; Greg Britten; Elizabeth N. Brooks; Gavin Fay; Alex Hansell; Christopher M. Legault; Brandon Muffley; Brian C. Stock; John Wiedenmann},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage[margin=1in]{geometry}
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{5}
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
-  \fi
-  % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-\usepackage{url}
-\usepackage{setspace}
-%\singlespacing
-%\onehalfspacing
-\doublespacing
-\usepackage{lineno}
-\linenumbers
-\usepackage[belowskip=0pt,aboveskip=0pt]{caption}
-\usepackage{relsize}
-\newcommand{\Fmsy}{\ensuremath{F_{\text{MSY}}}\xspace}
-\newcommand{\Fspr}[1]{\ensuremath{F_{\text{{#1}\%}}}\xspace}
-\newcommand{\afrb}{Alaska Fishery Research Bulletin\xspace}
-\newcommand{\ajms}{African Journal of Marine Science\xspace}
-\newcommand{\amb}{Advances in Marine Biology\xspace}
-\newcommand{\bms}{Bulletin of Marine Science\xspace}
-\newcommand{\bjssf}{Bulletin of the Japanese Society of Scientific Fisheries\xspace}
-\newcommand{\cb}{Conservation Biology\xspace}
-\newcommand{\cjfas}{Canadian Journal of Fisheries and Aquatic Sciences\xspace}
-\newcommand{\ea}{Ecological Applications\xspace}
-\newcommand{\eer}{Evolutionary Ecology Research\xspace}
-\newcommand{\elet}{Ecology Letters\xspace}
-\newcommand{\emod}{Ecological Modelling\xspace}
-\newcommand{\ebf}{Environmental Biology of Fishes\xspace}
-\newcommand{\ff}{Fish and Fisheries\xspace}
-\newcommand{\fo}{Fisheries Oceanography\xspace}
-\newcommand{\fr}{Fisheries Research\xspace}
-\newcommand{\fb}{Fishery Bulletin\xspace}
-\newcommand{\ijms}{ICES Journal of Marine Science\xspace}
-\newcommand{\iccat}{Collective Volume of Scientific Papers ICCAT\xspace}
-\newcommand{\jae}{Journal of Animal Ecology\xspace}
-\newcommand{\jai}{Journal of Applied Ichthyology\xspace}
-\newcommand{\jdc}{Journal Du Conseil International Pour L'exploration De La Mer\xspace}
-\newcommand{\jdcp}{Journal Du Conseil Permanent International Pour L'exploration De La Mer\xspace}
-\newcommand{\jembe}{Journal of Experimental Marine Biology and Ecology\xspace}
-\newcommand{\jfb}{Journal of Fish Biology\xspace}
-\newcommand{\jsr}{Journal of Sea Research\xspace}
-\newcommand{\jtb}{Journal of Theoretical Biology\xspace}
-\newcommand{\jfrbc}{Journal of the Fisheries Research Board of Canada\xspace}
-\newcommand{\jnwafs}{Journal of Northwest Atlantic Fisheries Science\xspace}
-\newcommand{\mcf}{Marine and Coastal Fisheries: Dynamics, Management, and Ecosystem Science\xspace}
-\newcommand{\mb}{Marine Biology\xspace}
-\newcommand{\meps}{Marine Ecology Progress Series\xspace}
-\newcommand{\mfr}{Marine Fisheries Review\xspace}
-\newcommand{\mpb}{Marine Pollution Bulletin\xspace}
-\newcommand{\najfm}{North American Journal of Fisheries Management\xspace}
-\newcommand{\nzjmfr}{New Zealand Journal of Marine and Freshwater Research\xspace}
-\newcommand{\pnas}{Proceedings of the National Academy of Sciences USA\xspace}
-\newcommand{\rpvrciemm}{Rapports et Proc\`es-Verbaux des R\'eunions. Conseil Internationale pour l'Exploration de la Mer\xspace}
-\newcommand{\rpvrcpiemm}{Rapports et Proc\`es-Verbaux des R\'eunions. Conseil Permanent Internationale pour l'Exploration de la Mer\xspace}
-\newcommand{\rfbf}{Reviews in Fish Biology and Fisheries\xspace}
-\newcommand{\sajms}{South African Journal of Marine Science\xspace}
-\newcommand{\tafs}{Transactions of the American Fisheries Society\xspace}
+---
+title: Factors affecting inferences on natural mortality and associated environmental effects
+author: 
+  - Timothy J. Miller^1^
+  - Andrew Applegate 
+  - Greg Britten
+  - Elizabeth N. Brooks
+  - Gavin Fay
+  - Alex Hansell
+  - Christopher M. Legault
+  - Brandon Muffley
+  - Brian C. Stock
+  - John Wiedenmann
+date: ''
+fontsize: 12pt
+output:
+  pdf_document:
+    citation_package: natbib
+    fig_caption: yes
+    includes:
+      in_header:
+      - options.sty
+      - journalshortcuts.tex
+    keep_tex: yes
+    number_sections: yes
+  html_document:
+    df_print: paged
+  word_document: default
+header-includes:
+- \usepackage{xspace}
+- \usepackage{bm}
+- \usepackage{caption,graphics}
+- \usepackage{graphicx}
+- \usepackage{makecell}
+- \renewcommand\figurename{Fig.}
+- \captionsetup{labelsep=period, singlelinecheck=false}
+- \newcommand{\changesize}[1]{\fontsize{#1pt}{#1pt}\selectfont}
+- \renewcommand{\arraystretch}{1.5}
+- \renewcommand\theadfont{}
+csl: canadian-journal-of-fisheries-and-aquatic-sciences.csl
+biblio-style: cjfas2.bst
+---
 
-\newcommand{\anzjs}{Australian \& New Zealand Journal of Statistics\xspace}
-\newcommand{\as}{Applied Statistics\xspace}
-\newcommand{\csda}{Computational Statistics \& Data Analysis\xspace}
-\newcommand{\ees}{Environmental and Ecological Statistics\xspace}
-\newcommand{\jas}{Journal of Applied Statistics\xspace}
-\newcommand{\jabes}{Journal of Agricultural, Biological, and Environmental Statistics\xspace}
-\newcommand{\jasa}{Journal of the American Statistical Association\xspace}
-\newcommand{\jrssb}{Journal of the Royal Statistical Society. Series B\xspace}
-\newcommand{\sm}{Statistics in Medicine}
+```{r setup, include=FALSE, echo=FALSE}
+knitr::opts_chunk$set(echo = TRUE, tidy=FALSE, tidy.opts=list(width.cutoff=60), warning = FALSE, message = FALSE, fig.align = "center")
+knitr::opts_knit$set(root.dir = '../' )
+library(knitr)
+library(tidyverse)
+library(pander)
+library(kableExtra)
+library(png)
+library(forcats)
+```
 
-\usepackage{xspace}
-\usepackage{bm}
-\usepackage{caption,graphics}
-\usepackage{graphicx}
-\usepackage{makecell}
-\renewcommand\figurename{Fig.}
-\captionsetup{labelsep=period, singlelinecheck=false}
-\newcommand{\changesize}[1]{\fontsize{#1pt}{#1pt}\selectfont}
-\renewcommand{\arraystretch}{1.5}
-\renewcommand\theadfont{}
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{wrapfig}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{pdflscape}
-\usepackage{tabu}
-\usepackage{threeparttable}
-\usepackage{threeparttablex}
-\usepackage[normalem]{ulem}
-\usepackage{makecell}
-\usepackage{xcolor}
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\usepackage[]{natbib}
-\bibliographystyle{cjfas2.bst}
-
-\title{Factors affecting inferences on natural mortality and associated
-environmental effects}
-\author{Timothy J. Miller\textsuperscript{1} \and Andrew
-Applegate \and Greg Britten \and Elizabeth N. Brooks \and Gavin
-Fay \and Alex Hansell \and Christopher M. Legault \and Brandon
-Muffley \and Brian C. Stock \and John Wiedenmann}
-\date{}
-
-\begin{document}
-\maketitle
-
-\(^1\)corresponding author:
-\href{mailto:timothy.j.miller@noaa.gov}{\nolinkurl{timothy.j.miller@noaa.gov}},
-Northeast Fisheries Science Center, Woods Hole Laboratory, 166 Water
-Street, Woods Hole, MA 02543 USA\\
+$^1$corresponding author: timothy.j.miller@noaa.gov, Northeast Fisheries Science Center, Woods Hole Laboratory, 166 Water Street, Woods Hole, MA 02543 USA\
 
@@
 
-\hypertarget{abstract}{%
-\subsection*{Abstract}\label{abstract}}
-\addcontentsline{toc}{subsection}{Abstract}
+## Abstract {-}
 
-\hypertarget{keywords}{%
-\subsubsection*{Keywords}\label{keywords}}
-\addcontentsline{toc}{subsubsection}{Keywords}
+### Keywords {-}
 
@@
 
-\hypertarget{introduction}{%
-\section*{Introduction}\label{introduction}}
-\addcontentsline{toc}{section}{Introduction}
+# Introduction {-}
+<!--
+Application of state-space models in fisheries stock assessment and management has expanded dramatically within ICES, Canada and the Northeast US \citep{nielsenberg14,cadigan16,stockmiller21}.
+State-space approaches that use random effects to parameterize process errors is considered best practice and a requirement for the next generation of stock assessment models.
 
-Here we conduct a simulation study with operating models varying by
-degree of observation error uncertainty, sources of process error (M or
-survival), fishing history, temporal variation in environmental
-covariates, and magnitude of the effect of the covariate on natural
-mortality. The simulations from these operating models are fitted with
-estimating models that make alternative assumptions for sources of
-process error (M, NAA), and whether (mean) M is estimated. We evaluate
-whether AIC can correctly determine the correct source of process error
-and stock effects on recruitment. We also evaluate the degree of bias in
-the outputs of the assessment model that are important for management.
+Much is known about the reliability of state-space models that are linear or Gaussian \citep{aeberhardetal18}, but applications in fisheries management are nonlinear and typically include multiple types of observations with varying distributional assumptions.
+Furthermore, there is a wide range of potential random effects structures and model parameters that can be treated as random effects in assessment models.
+We know relatively little about the statistical reliability of such models or the ability of information criteria to distinguish among such alternative structures.
 
-\hypertarget{methods}{%
-\section*{Methods}\label{methods}}
-\addcontentsline{toc}{section}{Methods}
+ But those studies focus primarily on Gaussian linear models.
+review literature on reliability of hidden/latent process models. Primarily in other fields.
 
-We used the WHAM package
-\citep[][commit 77bbd94]{millerstock20,stockmiller21} to configure all
-operating models and estimation models. This packages has also been used
-to configure operating and estimating models for closed loop simulations
-evaluating index-based assessment methods \citep{legaultetal23} and is
-used for management of haddock, butterfish, plaice, bluefish in the
-Northeast US.
+Convergence results can be useful for understanding how bad convergence in applications to real data might direct the practitioner to which alternative random effects configurations is be more appropriate.
+The type of convergence that we might use as a diagnostic is important.
+And what rule of thumb we use for certain criteria e.g.
+1e-6?
+large SEs/CVs?
+-->
 
-We completed a simulation study with a number of operating models. Many
-of the characteristics of the operating models are the same as those
-used by \citet{milleretal_wp}.
+Here we conduct a simulation study with operating models varying by degree of observation error uncertainty, sources of process error (M or survival), fishing history, temporal variation in environmental covariates, and magnitude of the effect of the covariate on natural mortality.
+The simulations from these operating models are fitted with estimating models that make alternative assumptions for sources of process error (M, NAA), and whether (mean) M is estimated.
+We evaluate whether AIC can correctly determine the correct source of process error and stock effects on recruitment.
+We also evaluate the degree of bias in the outputs of the assessment model that are important for management.
 
-288 total operating models. We simulated 100 data sets for each
-operating model. For each simulated data set we fit a set of 12
-estimating models.
+# Methods {-}
 
-\hypertarget{operating-models}{%
-\subsection*{Operating models}\label{operating-models}}
-\addcontentsline{toc}{subsection}{Operating models}
+We used the WHAM package \citep[][commit 77bbd94]{millerstock20,stockmiller21} to configure all operating models and estimation models.
+This packages has also been used to configure operating and estimating models for closed loop simulations evaluating index-based assessment methods \citep{legaultetal23} and is used for management of haddock, butterfish, plaice, bluefish in the Northeast US.
 
-\hypertarget{population}{%
-\subsubsection*{Population}\label{population}}
-\addcontentsline{toc}{subsubsection}{Population}
+We completed a simulation study with a number of operating models. Many of the characteristics of the operating models are the same as those used by \citet{milleretal_wp}.
 
-The population model tracks 10 age classes: ages 1 to 10+ and we assume
-spawning occurs 1/4 of the way through the year. The maturity at age was
-a logistic curve with a50 = 2.89 and slope = 0.88 and assumed known in
-all estimation models (Figure \ref{om_maturity}).
+288 total operating models.
+We simulated 100 data sets for each operating model.
+For each simulated data set we fit a set of 12 estimating models.
+
+## Operating models {-}
+
+### Population {-}
+
+The population model tracks 10 age classes: ages 1 to 10+ and we assume spawning occurs 1/4 of the way through the year.
+The maturity at age was a logistic curve with a50 = 2.89 and slope = 0.88 and assumed known in all estimation models (Figure \ref{om_maturity}).
 
@@
 
-Weight at age was generated with a LVB growth function \[
+Weight at age was generated with a LVB growth function
+$$
 L_a = L_{\infty}\left(1 - e^{-k(a - t_0)}\right)
-\] where \(t_0 = 0\), \(L_\infty = 85\), and \(k = 0.3\), and a L-W
-relationship such that \[
+$$
+where $t_0 = 0$, $L_\infty = 85$, and $k = 0.3$, and a L-W relationship such that 
+$$
 W_a = \theta_1 L_a^{\theta_2}
-\] where \(\theta_1 = e^{-12.1}\) and \(\theta_2 = 3.2\) (Figure
-\ref{om_waa}).
+$$
+where $\theta_1 = e^{-12.1}$ and $\theta_2 = 3.2$ (Figure \ref{om_waa}).
 
@@
 
-We assumed a Beverton-Holt stock recruit function with constant
-pre-recruit mortality parameters for all operating models. All
-post-recruit productivity components are constant in the NAA and survey
-catchability process error operating models. Therefore steepness and
-unfished recruitment are also constant over the time period for those
-operating models \citep{millerbrooks21}. We specified unfished
-recruitment = \(R_0 = e^{10}\) and \(\Fmsy = \Fspr[40] = 0.348\) equated
-to a steepness of 0.69 and \(\alpha=0.60\) and
-\(\beta = 2.4 \times 10^{-5}\) for the \[
+We assumed a Beverton-Holt stock recruit function with constant pre-recruit mortality parameters for all operating models.
+All post-recruit productivity components are constant in the NAA and survey catchability process error operating models.
+Therefore steepness and unfished recruitment are also constant over the time period for those operating models \citep{millerbrooks21}.
+We specified unfished recruitment = $R_0 = e^{10}$ and $\Fmsy = \Fspr[40] = 0.348$ equated to a steepness of 0.69 and $\alpha=0.60$ and $\beta = 2.4 \times 10^{-5}$ for the
+\[
 N_{1,y} = \frac{\alpha \text{SSB}_{y-1}}{1 + \beta \text{SSB}_{y-1}} 
-\] Beverton-Holt parameterization (Figure \ref{om_sr}). For OMs without
-process errors on natural mortality we assumed the rate was assumed 0.2.
+\]
+Beverton-Holt parameterization (Figure \ref{om_sr}).
+For OMs without process errors on natural mortality we assumed the rate was assumed 0.2.
 For OMs with process errors on natural mortality the mean rate was 0.2.
@@
 
-Two alternative fishing histories were used for operating models. In the
-first scenario, the stock experiences overfishing for the first 20 years
-and fishing at \Fmsy for the last 20 years. In the second scenario, the
-stock is fished at \Fmsy for the entire time period. The magnitude of
-the overfishing assumptions is based on average estimates of overfishing
-for NE groundfish stocks from \citet{wiedenmannetal19}.
-\Citet{legaultetal23} also used similar approaches to defining fishing
-mortality histories for operating models.
+Two alternative fishing histories were used for operating models.
+In the first scenario, the stock experiences overfishing for the first 20 years and fishing at \Fmsy for the last 20 years.
+In the second scenario, the stock is fished at \Fmsy for the entire time period.
+The magnitude of the overfishing assumptions is based on average estimates of overfishing for NE groundfish stocks from \citet{wiedenmannetal19}.
+\Citet{legaultetal23} also used similar approaches to defining fishing mortality histories for operating models.
 
-Initial population was configured at the equilibrium distribution
-fishing at either \(F = 2.5\times \Fmsy\) or \(F = \Fmsy\) for the two
-alternative fishing histories. That is for a deterministic model, the
-age composition would not change over time when the fishing mortality
-was constant at the respective level.
+Initial population was configured at the equilibrium distribution fishing at either $F = 2.5\times \Fmsy$ or $F = \Fmsy$ for the two alternative fishing histories.
+That is for a deterministic model, the age composition would not change over time when the fishing mortality was constant at the respective level.
 
-For operating models with time-varying random effects or covariate
-effects on M, steepness is not constant, but we used the same alpha and
-beta parameters as other operating models this equates to a steepness
-and R0 at the mean of the time series process for M. For operating
-models with time-varying random effects for fishery selectivity,
-\Fmsy is also not constant however we use the same F history as other
-operating models which corresponds to Fmsy at the mean selectivity
-parameters.
+For operating models with time-varying random effects or covariate effects on M, steepness is not constant, but we used the same alpha and beta parameters as other operating models this equates to a steepness and R0 at the mean of the time series process for M.
+For operating models with time-varying random effects for fishery selectivity, \Fmsy is also not constant however we use the same F history as other operating models which corresponds to Fmsy at the mean selectivity parameters.
 
-\hypertarget{environmental-covariate}{%
-\subsubsection*{Environmental covariate}\label{environmental-covariate}}
-\addcontentsline{toc}{subsubsection}{Environmental covariate}
+### Environmental covariate {-}
 
-In the WHAM model, environmental covariates are assumed to be described
-as state-space processes with annual observations of the true latent
-covariate. The latent covariate is assumed to be Gaussian AR1 \[
+In the WHAM model, environmental covariates are assumed to be described as state-space processes with annual observations of the true latent covariate.
+The latent covariate is assumed to be Gaussian AR1
+$$
 E_y|E_{y-1} \sim \text{N}\left(\mu_\text{Ecov}\left(1-\rho_\text{Ecov}\right) + \rho_\text{Ecov} E_{y-1}, \sigma^2_\text{Ecov}\right)
-\] with marginal mean \(\mu_\text{Ecov}\) and variance
-\(\frac{\sigma^2_\text{Ecov}}{1-\rho_\text{Ecov}^2}\). The configuration
-of the latent environmental covariate in the operating models have one
-of two alternate assumptions about the marginal standard deviation and
-autocorrelation parameter:
-
+$$
+with marginal mean $\mu_\text{Ecov}$ and variance $\frac{\sigma^2_\text{Ecov}}{1-\rho_\text{Ecov}^2}$.
+The configuration of the latent environmental covariate in the operating models have one of two alternate assumptions about the marginal standard deviation and autocorrelation parameter:
 \begin{itemize}
@@
 
-The observations of the latent environmental covariate are assumed to be
-unbiased and Gaussian \[
+The observations of the latent environmental covariate are assumed to be unbiased and Gaussian
+$$
 \text{ecov}_y|E_y \sim \text{N}\left(E_y,\sigma^2_\text{ecov}\right)
-\] The standard deviation of the environmental observations in the
-operating models is one of two values:
-
+$$
+The standard deviation of the environmental observations in the operating models is one of two values:
 \begin{itemize}
@@
 \end{itemize}
-
-Figure \ref{om_ecov_example} provides example simulations of the latent
-process and observations under the alternative configurations.
+Figure \ref{om_ecov_example} provides example simulations of the latent process and observations under the alternative configurations.
 
@@
 
-\hypertarget{random-effects-on-recruitment-survival-and-natural-mortality}{%
-\subsubsection*{Random effects on recruitment, survival, and natural
-mortality}\label{random-effects-on-recruitment-survival-and-natural-mortality}}
-\addcontentsline{toc}{subsubsection}{Random effects on recruitment,
-survival, and natural mortality}
+### Random effects on recruitment, survival, and natural mortality {-}
 
 Internally, WHAM treats natural mortality as a log-transformed parameter 
-\[
+$$
 \log M_y = \beta_M + \beta_{\text{Ecov}} E_y + \epsilon_{M,y}
-\] that is a linear combination of a mean \(\beta_M\) and any annual
-covariate \(\beta_{\text{Ecov}}\) and random effects marginally
-distributed as
-\(\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)\).
+$$
+that is a linear combination of a mean $\beta_M$ and any annual covariate $\beta_{\text{Ecov}}$ and random effects marginally distributed as $\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)$.
 
-The covariate effect is one of 3 alternative values in the operating
-models:
-
+The covariate effect is one of 3 alternative values in the operating models:
 \begin{itemize}
@@
 
-The alternative configurations of random effects on recruitment, natural
-mortality, survival
-
+The alternative configurations of random effects on recruitment, natural mortality, survival
 \begin{itemize}
@@
 
-\hypertarget{fleets}{%
-\subsubsection*{Fleets}\label{fleets}}
-\addcontentsline{toc}{subsubsection}{Fleets}
+### Fleets {-}
 
-We assumed a single fleet operating year round for catch observations
-with logistic selectivity for the fleet with \(a_{50} = 5\) and slope =
-1 (Figure \ref{om_mean_selectivity}). This selectivity is was used to
-define \Fmsy for the Beverton-Holt stock recruitment parameters above.
-We assumed a logistic-normal distribution for the age-composition
-observations for the fleet.
+We assumed a single fleet operating year round for catch observations with logistic selectivity for the fleet with $a_{50} = 5$ and slope = 1 (Figure \ref{om_mean_selectivity}).
+This selectivity is was used to define \Fmsy for the Beverton-Holt stock recruitment parameters above.
+We assumed a logistic-normal distribution for the age-composition observations for the fleet.
 
@@
 
-\hypertarget{indices}{%
-\subsubsection*{Indices}\label{indices}}
-\addcontentsline{toc}{subsubsection}{Indices}
+### Indices {-}
 
-Two time series of surveys are assumed and observed in numbers rather
-than biomass for the entire 40 year period with one occurring in the
-spring (0.25 way through the year) and one in the fall (0.75 way through
-the year). Actually we have it currently configured that both occur 0.5
-way through the year. Catchability of both surveys are assumed to be
-0.1. Like the fishing fleet, we assumed logistic selectivity for both
-indices with \(a_{50} = 5\) and slope = 1 and a logistic-normal
-distribution for the age-composition observations.
+Two time series of surveys are assumed and observed in numbers rather than biomass for the entire 40 year period with one occurring in the spring (0.25 way through the year) and one in the fall (0.75 way through the year).
+Actually we have it currently configured that both occur 0.5 way through the year.
+Catchability of both surveys are assumed to be 0.1.
+Like the fishing fleet, we assumed logistic selectivity for both indices with $a_{50} = 5$ and slope = 1 and a logistic-normal distribution for the age-composition observations.
 
-\hypertarget{observation-uncertainty}{%
-\subsubsection*{Observation Uncertainty}\label{observation-uncertainty}}
-\addcontentsline{toc}{subsubsection}{Observation Uncertainty}
+### Observation Uncertainty {-}
 
-Standard deviation for log-aggregate catch was 0.1. There were two
-levels of observation error variance for indices and age composition for
-both indices and fleet catch. A low uncertainty specification assumed
-standard deviation of both series of log-aggregate index observations
-was 0.1 and the standard deviation of the logistic-normal for age
-composition observations was 0.3 In the high uncertainty specification
-the standard deviation for log-aggregate indices was 0.4 and that for
-the age composition observations was 1.5. For all estimating models,
-standard deviation for log-aggregate observations was assumed known
-whereas that for the logistic-normal age composition observations was
-estimated.
+Standard deviation for log-aggregate catch was 0.1.
+There were two levels of observation error variance for indices and age composition for both indices and fleet catch.
+A low uncertainty specification assumed standard deviation of both series of log-aggregate index observations was 0.1 and the standard deviation of the logistic-normal for age composition observations was 0.3 In the high uncertainty specification the standard deviation for log-aggregate indices was 0.4 and that for the age composition observations was 1.5.
+For all estimating models, standard deviation for log-aggregate observations was assumed known whereas that for the logistic-normal age composition observations was estimated.
 
-The factorial combinations of fishing histories, index and age
-composition observation error assumptions, latent environmental
-covariate and observation assumptions, covariate effects on natural
-mortality, and sources of process error (R, R+S, R+M) results in 288
-different operating models.
+The factorial combinations of fishing histories, index and age composition observation error assumptions, latent environmental covariate and observation assumptions, covariate effects on natural mortality, and sources of process error (R, R+S, R+M) results in 288 different operating models.
 
-\hypertarget{estimating-models}{%
-\subsection*{Estimating models}\label{estimating-models}}
-\addcontentsline{toc}{subsection}{Estimating models}
+## Estimating models {-}
 
-For each data set simulated from an operating model 12 estimating models
-were fit. There were three factors defining the configuration of each
-estimating model
+For each data set simulated from an operating model 12 estimating models were fit. There were three factors defining the configuration of each estimating model
 
@@
 \end{itemize}
+The configuration of the process errors in the estimating models matched the corresponding options in the operating models.
+For example, if R+S was assumed for both the estimating and operating model, the assumptions about process errors matched.
+The environmental covariate observations were included in all estimation models to ensure comparability of information criteria.
+The observation error variance of the environmental observations and aggregate catch and indices were all assumed known at the true values.
 
-The configuration of the process errors in the estimating models matched
-the corresponding options in the operating models. For example, if R+S
-was assumed for both the estimating and operating model, the assumptions
-about process errors matched. The environmental covariate observations
-were included in all estimation models to ensure comparability of
-information criteria. The observation error variance of the
-environmental observations and aggregate catch and indices were all
-assumed known at the true values.
+Simulations were all carried out on the University of Massachusetts Green High-Performance Computing Cluster.
+Code for completing the simulations and summarization of results can be found at github.com/timjmiller/SSRTWG/ecov_study/mortality.
+We used the wham package version 1.X.X (commit 77bbd94).
 
-Simulations were all carried out on the University of Massachusetts
-Green High-Performance Computing Cluster. Code for completing the
-simulations and summarization of results can be found at
-github.com/timjmiller/SSRTWG/ecov\_study/mortality. We used the wham
-package version 1.X.X (commit 77bbd94).
+## Measures of reliability {-}
 
-\hypertarget{measures-of-reliability}{%
-\subsection*{Measures of reliability}\label{measures-of-reliability}}
-\addcontentsline{toc}{subsection}{Measures of reliability}
+The first measure of reliability we investigated was frequency of convergence when fitting each estimating model to the simulated data sets.
+There are various ways to assess convergence of the fit.
+We summarized 5 alternative categories of convergence.
+1. Did the optimization routine (stats::nlminb) complete without error?
+2. Did the stats::nlminb convergence flag = 0 indicate successful convergence?
+3. Was the maximum absolute value of the gradient of the log-likelihood < $1\times10^{-6}$?
+4. Did TMB::sdreport provide non-NA values for all fixed effects standard errors?
+5. Did TMB::sdreport provide all standard errors < 100?
 
-The first measure of reliability we investigated was frequency of
-convergence when fitting each estimating model to the simulated data
-sets. There are various ways to assess convergence of the fit. We
-summarized 5 alternative categories of convergence. 1. Did the
-optimization routine (stats::nlminb) complete without error? 2. Did the
-stats::nlminb convergence flag = 0 indicate successful convergence? 3.
-Was the maximum absolute value of the gradient of the log-likelihood
-\textless{} \(1\times10^{-6}\)? 4. Did TMB::sdreport provide non-NA
-values for all fixed effects standard errors? 5. Did TMB::sdreport
-provide all standard errors \textless{} 100?
+The first convergence criterion assesses whether the model crashes.
+The third is a measure that assesses how flat the likelihood is at the optimized point of the likelihood surface.
+The fourth and fifth criteria are specific to the hessian-based standard error reporting provided by TMB.
+the TMB::sdreport function will sometimes return standard error estimates even when the calculated hessian is not invertible (fourth criterion).
+It will also provide large standard errors for some parameters that are not estimated well or are near bounds on the transformed scale that is of primary interest.
+For example, variance parameters for random effects can often be estimated near 0 when the fit suggests no variation in the random effects or, equivalently, that the model without random effects is better.
 
-The first convergence criterion assesses whether the model crashes. The
-third is a measure that assesses how flat the likelihood is at the
-optimized point of the likelihood surface. The fourth and fifth criteria
-are specific to the hessian-based standard error reporting provided by
-TMB. the TMB::sdreport function will sometimes return standard error
-estimates even when the calculated hessian is not invertible (fourth
-criterion). It will also provide large standard errors for some
-parameters that are not estimated well or are near bounds on the
-transformed scale that is of primary interest. For example, variance
-parameters for random effects can often be estimated near 0 when the fit
-suggests no variation in the random effects or, equivalently, that the
-model without random effects is better.
-
-\hypertarget{aic-for-model-selection}{%
-\subsubsection*{AIC for model selection}\label{aic-for-model-selection}}
-\addcontentsline{toc}{subsubsection}{AIC for model selection}
+### AIC for model selection {-}
 
 We measured the frequency of correct model selection using marginal AIC.
-For a given operating model we the set of models that were considered
-all made the same assumptions on whether or not to estimate M and
-whether a stock-recruit function is assumed.
+For a given operating model we the set of models that were considered all made the same assumptions on whether or not to estimate M and whether a stock-recruit function is assumed.
 
-\hypertarget{bias}{%
-\subsubsection*{Bias}\label{bias}}
-\addcontentsline{toc}{subsubsection}{Bias}
+### Bias {-}
 
-We calculated relative errors of various quantities for each fitted
-estimation model for the \(i\)th simulated data set as \[
+We calculated relative errors of various quantities for each fitted estimation model for the $i$th simulated data set as
+$$
 {\text{RE}_i}\left(\theta\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
-\] and measured bias as the median relative error and constructed 95\%
-confidence intervals using the binomial distribution approach as in
-\citet{stockmiller21}.
+$$
+and measured bias as the median relative error and constructed 95\% confidence intervals using the binomial distribution approach as in \citet{stockmiller21}.
 
-\hypertarget{results}{%
-\section*{Results}\label{results}}
-\addcontentsline{toc}{section}{Results}
+# Results {-}
 
-\hypertarget{convergence-performance}{%
-\subsection*{Convergence performance}\label{convergence-performance}}
-\addcontentsline{toc}{subsection}{Convergence performance}
+## Convergence performance {-}
 
@@
 
-\hypertarget{aic-performance}{%
-\subsection*{AIC performance}\label{aic-performance}}
-\addcontentsline{toc}{subsection}{AIC performance}
+## AIC performance {-}
 
@@
 
-\hypertarget{bias-1}{%
-\subsection*{Bias}\label{bias-1}}
-\addcontentsline{toc}{subsection}{Bias}
+## Bias {-}
 
-\hypertarget{discussion}{%
-\section*{Discussion}\label{discussion}}
-\addcontentsline{toc}{section}{Discussion}
+# Discussion {-}
 
-The estimating models assumed variances of aggregate catch and index
-observations was known. This approximation may be appropriate for
-indices where we have a reliable estimate of uncertainty based on the
-survey design (), but there may be better approaches for the aggregate
-catch such as an informed prior on the standard errors with realistic
-bounds.
+The estimating models assumed variances of aggregate catch and index observations was known.
+This approximation may be appropriate for indices where we have a reliable estimate of uncertainty based on the survey design (), but there may be better approaches for the aggregate catch such as an informed prior on the standard errors with realistic bounds.
 
-\hypertarget{acknowledgements}{%
-\section*{Acknowledgements}\label{acknowledgements}}
-\addcontentsline{toc}{section}{Acknowledgements}
+# Acknowledgements {-}
 
-This work was funded by NOAA Fisheries Northeast Fisheries Science
-Center.
+This work was funded by NOAA Fisheries Northeast Fisheries Science Center.
+
+<!--
+# CRediT authorship contribution statement {-}
+
+**Timothy J. Miller:** Conceptualization, Methodology, Writing - original draft, Formal analysis, Visualization.
+ -->
 
@@
 \bibliography{paper}
+<!-- \bibliography{project-0-paper}
+\bibliographystyle{/home/tmiller2/work/bibliographies/cjfas2}  -->
 
-\hypertarget{refs}{}
-\begin{CSLReferences}{0}{0}
-\end{CSLReferences}
+<div id="refs"></div>
 
 \pagebreak
-
-\end{document}

### 2023-09-21 Ecov_study/catchability/Setup_OMs.Rmd
Update how F history initialized, leverage common code functions to set up OM/EMs.

--- 2023-02-27 Ecov_study/catchability/Setup_OMs.Rmd
@@
 
--
+-   Seasonality of Ecov effect on catchability
 
+Run analysis in 3 pieces:
+# 1) Factorial runs with different ecov process magnitude, fishing history and observation error
+-   maybe don't vary obs/process NAA since we are testing ability to
+    recover catchability effect, could do sensitivity run with survival
+    random effects and pick one of these settings
 ```{r}
@@
 # Fishing history
-F_hist <- c("H-MSY", "MSY") # High-then MSY vs. MSY for entire history
+F_hist <- c("H-L", "Fmsy") # High-then FMSY vs. FMSY for entire history
 
@@
 ageComp_sig <- c(0.3, 1.5) # Fleet and index age comp treated with same error
-logIndex_sig <- c(0.1, 0.4)
+log_index_sig <- c(0.1, 0.4) # agg_index_sigma
+log_catch_sig <- 0.1 # agg_catch_sigma
 
 # Factorial combinations
-OMsetup <- expand.grid(Eco_process_sig = Ecov_process_sig, Ecov_process_cor = Ecov_process_cor, Ecov_process_obs_sig = Ecov_process_obs_sig, Ecov_effect = Ecov_effect, F_hist = F_hist, ageComp_sig = ageComp_sig, logIndex_sig = logIndex_sig)
+OMsetup <- expand.grid(Ecov_process_sig = Ecov_process_sig, Ecov_process_cor = Ecov_process_cor, Ecov_process_obs_sig = Ecov_process_obs_sig, Ecov_effect = Ecov_effect, F_hist = F_hist, ageComp_sig = ageComp_sig, log_index_sig = log_index_sig, log_catch_sig = log_catch_sig)
 OMsetup
+```
 
+# 2) Vary seasonality of q effect in EM
+```{r}
+##### EM
+# EM factorial settings
+miss_season <- c("BOTH", "ONE", "NONE") # Number of seasons with correctly specified effect, assumes 2 survey indices (spring & fall)
+miss_q <- c("NoEcov", "Ecov", "qRand", "qRandEcov") # Catchability setup for EM
+EMsetup <- expand.grid(miss_season = miss_season, miss_q = miss_q)
+EMsetup 
+```
+
+# 3) May run sensitivity run with no ecov effect in OM but compare same EMs
+Check that ecov effect = 0 is essentially turning off ecov
+```{r}
 # Catchability set up
@@
 ##### OM sensitivity runs
-# Factorial combinations when on ecov effect in OM
-expand.grid(q_Ecov = q_Ecov, F_hist = F_hist, ageComp_sig = ageComp_sig, logIndex_sig = logIndex_sig) # probably do a subset, we have to decide what subset
-
-##### EM
-# EM factorial settings
-miss_season <- c("BOTH", "ONE", "NONE") # Number of seasons with correctly specified effect, assumes 2 survey indices (spring & fall)
-miss_q <- c("NoEnv", "NoEnvRand", "WithEnvRand", "WithEnv") # Catchability setup for EM
-expand.grid(miss_season = miss_season, miss_q = miss_q)
+# Factorial combinations when no ecov effect in OM
+expand.grid(q_Ecov = q_Ecov, F_hist = F_hist, ageComp_sig = ageComp_sig, log_index_sig = log_index_sig) # probably do a subset, we have to decide what subset
 ```
 
--   maybe don't vary obs/process NAA since we are testing ability to
-    recover catchability effect, could do sensitivity run with survival
-    random effects and pick one of these settings
-
-# set up generic OM
-
-```{r}
-library(wham)
-# Get generic groundfish life history info
-source(file.path(here::here(),"common_code", "make_basic_info.R"))
-groundfish_info <- make_basic_info()
-
-# Change timing for spring (0.25) and fall (0.75) indices
-groundfish_info$fracyr_indices[,1] <- 0.25
-groundfish_info$fracyr_indices[,2] <- 0.75
-
-# Assume logistic selectivity for the fleet and all indices
-sel_list <- list(model = c(rep("logistic", groundfish_info$n_fleets),rep("logistic", groundfish_info$n_indices)),
-                   initial_pars = lapply(1:(groundfish_info$n_fleets + groundfish_info$n_indices), function(x) c(5,1)),
-                 n_selblocks = (groundfish_info$n_fleets + groundfish_info$n_indices))
-
-# Starting mean M estimate at 0.2, estimate constant M for all ages/years
-M <-  list(initial_means = rep(0.2, length(groundfish_info$ages)))
-
-genericInput <- prepare_wham_input(basic_info = groundfish_info, 
-                                   selectivity = sel_list, 
-                                   M = M)
-
-# genericFit <- fit_wham(genericInput, MakeADFun.silent = TRUE, do.retro = FALSE, do.osa = FALSE)
-
-```
-
-# Use common_code set_ecov function to adjust based on OM settings above
-
+## Set up OM and corresponding EMs
 This chunk of code uses common code functions & internal WHAM functions
@@
 
+Use project 0 make_om() to set up generic OM input then add additional ecov settings afterwards before fitting model
+- By default catchability impacted by environment only for fall (2nd) index
+- Each OM stored in a corresponding folder/directory
+
+EMs are misspecified for each OM and stored a sub directory of the OM
 ```{r}
-source(file.path(here::here(),"common_code", "set_ecov.R"))
+library(TAF)
+library(tidyverse)
+library(wham)
+# source(here::here("Project_0", "code", "make_om.R"))
+source(here::here("make_om.R")) # Use q copy which requires that Beverton-Holt S-R used (make_om.R line #11)
+source(here::here("make_basic_info.R")) # from common_code
+source(here::here("set_ecov.R")) # from common_code
+source(here::here("set_NAA.R")) # from common code
+source(here::here("set_q.R")) # from common code
+source(here::here("set_selectivity.R")) # from common code
+source(here::here("set_M.R")) # from common code
+source(here::here("get_FMSY.R"))
 
-# Loop over OMsetup to initialize OMs
+n_fleets <- 1
+n_indices <- 2
+n_ages <- 10
+n_years <- 40
+# Assume logistic selectivity for the fleet and all indices
+sel_list <- list(model = c(rep("logistic", n_fleets),rep("logistic", n_indices)),
+                 initial_pars = lapply(1:(n_fleets + n_indices), function(x) c(5,1)),
+                 n_selblocks = (n_fleets + n_indices))
+
+# Starting mean M estimate at 0.2, estimate constant M for all ages/years
+M_list <-  list(initial_means = rep(0.2, n_ages))
+
+# NAA_re = list(recruit_model = 3) # This required for Beverton-Holt S-R
+
 for(iom in 1:nrow(OMsetup)){
-  # Pull generic input settings
-  input <- genericInput
+  print(iom)
+  ##### Set up generic input #####
+  # Set up input without ecov
+  input <- make_om(Fhist = OMsetup[iom,"F_hist"],
+                   N1_state = "Fmsy", # Default, could also pick "overfished" or "unfished"
+                   selectivity = sel_list,
+                   M = M_list, 
+                   catchability = NULL,
+                   NAA_re = NULL) 
+  # age_comp, # Default = "logistic-normal-miss0"
+  # brp_year, # Default = 1
+  # eq_F_init, # Default = 0.3
+  # om_input = TRUE, # Don't fit model, only structure generic input
+  # max_mult_Fmsy, # Default = 2.5
+  # min_mult_Fmsy, # Default = 1
+  # F_change_time # Default = 0.5, sets when F changes (in H-L or L-H F_hist scenarios)
 
+  # Observation error - set L-N SD parameters for catch and index age comps # pulled from project 0 q_om_setup.R
+  input$par$catch_paa_pars[,1] <- log(OMsetup[iom, "ageComp_sig"])
+  input$par$index_paa_pars[,1] <- log(OMsetup[iom, "ageComp_sig"])
+  input$data$agg_catch_sigma[] <- OMsetup[iom,"log_catch_sig"]
+  input$data$agg_index_sigma[] <- OMsetup[iom,"log_index_sig"]
+  input$data$catch_Neff[] <- 1 # Change Neff so scalar doesn't affect L-N SD
+  input$data$index_Neff[] <- 1
+
+  ##### Set up OM ecov #####
   # Ecov setup
   Ecov <- list(label = "Ecov",
+               mean = matrix(rep(0,n_years), ncol =1), # Mean = 0
+               logsigma = matrix(log(OMsetup$Ecov_process_obs_sig[iom]), n_years, ncol=1),
+               years = input$years, 
+               use_obs = matrix(rep(TRUE, n_years), ncol = 1),
                lag = 0, 
@@
                how = 1,
-                 indices = c(1,2),
-                 process_model = "ar1", 
-                 process_mean_vals = rep(0,length(groundfish_info$years)), # End generic inputs for ecov
-                 process_sig_vals = rep(OMsetup$Eco_process_sig[iom]),
+               indices = list(2), # Ecov only impact fall index catchability
+               process_model = "ar1", # End generic inputs for ecov
+               process_mean_vals = rep(0,n_years), # Mean = 0
+               process_sig_vals = rep(OMsetup$Ecov_process_sig[iom]),
                process_cor_vals = rep(OMsetup$Ecov_process_cor[iom]),
-                 beta_vals = list(rep(list(matrix(rep(OMsetup$Ecov_effect[iom], length(groundfish_info$ages)), nrow = 1)), 4)), #!!! Why 4 rows?
-                 logsigma = as.matrix(rep(OMsetup$Ecov_process_obs_sig[iom], length(groundfish_info$years))))
+               beta_vals = list(append(rep(list(matrix(rep(0, n_ages), nrow = 1)), 2), # No impact on R,M
+                                       rep(list(matrix(rep(OMsetup$Ecov_effect[iom], n_ages), nrow = 1)), 2)))) # impact on q by index (here 2 indices)
 
-  input <- set_ecov(input = input, ecov = Ecov)
+  inputOM <- set_ecov(input = input, ecov = Ecov) #!!! something wrong with ecov$beta_vals[[j]][[n]] indexing
 
-  # 
-  # Fishing history
-F_hist <- c("H-MSY", "MSY") # High-then MSY vs. MSY for entire history
+  # Finish setup & save OM
+  OM <- NULL # Clear prior OM object just in case
+  OM <- fit_wham(input = inputOM, do.fit = FALSE, MakeADFun.silent = TRUE)
+  mkdir(here::here("Results", paste0("OM_", iom))) # set up storage
+  saveRDS(OM, file = here::here("Results", paste0("OM_", iom), paste0("OM_", iom, ".Rds"))) # save OM
+  saveRDS(OMsetup[iom,], file = here::here("Results", paste0("OM_", iom), "OMsettings.Rds")) # save OM settings in a separate file since naming is based on arbitraty order in OMsetup table
 
-# Observation error
-ageComp_sig <- c(0.3, 1.5) # Fleet and index age comp treated with same error
-logIndex_sig <- c(0.1, 0.4)
+  ##### Iterate over EM misspecifications for this OM
+  for(iem in 1:nrow(EMsetup)){
+    inputEM <- NULL # Reset at start of loop just in case
 
+    # Seasonal misspecification (q random effect implemented with same misspecification as ecov impact)
+    if(EMsetup[iem, "miss_season"] == "BOTH"){
+      index_list <- list(1) # Both seasons misspecified
+      qRand <- c("ar1", "none")
+    } else if(EMsetup[iem, "miss_season"] == "ONE"){
+      index_list <- list(1,2) # Spring misspecified, fall correct
+      qRand <- c("ar1", "ar1")
+    } else{
+      index_list <- list(2) # Both seasons correctly specified
+      qRand <- c("none", "ar1")
     }
 
+    # Catchability model assumptions
+    if(EMsetup[iem, "miss_q"] == "NoEcov"){ # No ecov impact on q or q random effect
+      print("NoEcov")
+      Ecov <- list(label = "NoEcov",
+                   mean = matrix(rep(0,n_years), ncol =1), # Mean = 0
+                   logsigma = matrix(log(OMsetup$Ecov_process_obs_sig[iom]), n_years, ncol=1),
+                   years = input$years, 
+                   use_obs = matrix(rep(TRUE, n_years), ncol = 1),
+                   lag = 0, 
+                   where = "none", # Where/how/indices settings need to change if we do sensitivity runs
+                   how = 1,
+                   indices = index_list, # Ecov impact based on sesonal misspecification setting
+                   process_model = "ar1", # End generic inputs for ecov
+                   process_mean_vals = rep(0,n_years), # Mean = 0
+                   process_sig_vals = rep(OMsetup$Ecov_process_sig[iom]),
+                   process_cor_vals = rep(OMsetup$Ecov_process_cor[iom]),
+                   beta_vals = list(append(rep(list(matrix(rep(0, n_ages), nrow = 1)), 2), # No impact on R,M
+                                           rep(list(matrix(rep(OMsetup$Ecov_effect[iom], n_ages), nrow = 1)), 2)))) # impact on q by index (here 2 indices)
+
+      inputEM <- set_ecov(input = input, ecov = Ecov)
+
+    } else if(EMsetup[iem, "miss_q"] == "Ecov"){ # Ecov impact on q, correctly specified option when seasonal impact correctly specified
+      print("Ecov")
+      Ecov <- list(label = "Ecov",
+                   mean = matrix(rep(0,n_years), ncol =1), # Mean = 0
+                   logsigma = matrix(log(OMsetup$Ecov_process_obs_sig[iom]), n_years, ncol=1),
+                   years = input$years, 
+                   use_obs = matrix(rep(TRUE, n_years), ncol = 1),
+                   lag = 0, 
+                   where = "q", # Where/how/indices settings need to change if we do sensitivity runs
+                   how = 1,
+                   indices = index_list, # Ecov impact based on sesonal misspecification setting
+                   process_model = "ar1", # End generic inputs for ecov
+                   process_mean_vals = rep(0,n_years), # Mean = 0
+                   process_sig_vals = rep(OMsetup$Ecov_process_sig[iom]),
+                   process_cor_vals = rep(OMsetup$Ecov_process_cor[iom]),
+                   beta_vals = list(append(rep(list(matrix(rep(0, n_ages), nrow = 1)), 2), # No impact on R,M
+                                           rep(list(matrix(rep(OMsetup$Ecov_effect[iom], n_ages), nrow = 1)), 2)))) # impact on q by index (here 2 indices)
+
+      inputEM <- set_ecov(input = input, ecov = Ecov)
+
+    } else if(EMsetup[iem, "miss_q"] == "qRand"){ # q random effect
+      print("qRand")
+      Ecov <- list(label = "qRand",
+                   mean = matrix(rep(0,n_years), ncol =1), # Mean = 0
+                   logsigma = matrix(log(OMsetup$Ecov_process_obs_sig[iom]), n_years, ncol=1),
+                   years = input$years, 
+                   use_obs = matrix(rep(TRUE, n_years), ncol = 1),
+                   lag = 0, 
+                   where = "none", # Where/how/indices settings need to change if we do sensitivity runs
+                   how = 1,
+                   indices = index_list, # Ecov impact based on sesonal misspecification setting
+                   process_model = "ar1", # End generic inputs for ecov
+                   process_mean_vals = rep(0,n_years), # Mean = 0
+                   process_sig_vals = rep(OMsetup$Ecov_process_sig[iom]),
+                   process_cor_vals = rep(OMsetup$Ecov_process_cor[iom]),
+                   beta_vals = list(append(rep(list(matrix(rep(0, n_ages), nrow = 1)), 2), # No impact on R,M
+                                           rep(list(matrix(rep(OMsetup$Ecov_effect[iom], n_ages), nrow = 1)), 2)))) # impact on q by index (here 2 indices)
+
+      inputEM <- set_ecov(input = input, ecov = Ecov)
+
+      # Catchability random effects
+      inputEM <- set_q(input = inputEM, q_opts = qRand)
+
+    } else{ # qRandEcov both q random effect and ecov impact on q
+      print("qRandEcov")
+      Ecov <- list(label = "qRandEcov",
+                   mean = matrix(rep(0,n_years), ncol =1), # Mean = 0
+                   logsigma = matrix(log(OMsetup$Ecov_process_obs_sig[iom]), n_years, ncol=1),
+                   years = input$years, 
+                   use_obs = matrix(rep(TRUE, n_years), ncol = 1),
+                   lag = 0, 
+                   where = "q", # Where/how/indices settings need to change if we do sensitivity runs
+                   how = 1,
+                   indices = index_list, # Ecov impact based on sesonal misspecification setting
+                   process_model = "ar1", # End generic inputs for ecov
+                   process_mean_vals = rep(0,n_years), # Mean = 0
+                   process_sig_vals = rep(OMsetup$Ecov_process_sig[iom]),
+                   process_cor_vals = rep(OMsetup$Ecov_process_cor[iom]),
+                   beta_vals = list(append(rep(list(matrix(rep(0, n_ages), nrow = 1)), 2), # No impact on R,M
+                                           rep(list(matrix(rep(OMsetup$Ecov_effect[iom], n_ages), nrow = 1)), 2)))) # impact on q by index (here 2 indices)
+
+      inputEM <- set_ecov(input = input, ecov = Ecov)
+
+      # Catchability random effects
+      inputEM <- set_q(input = inputEM, q_opts = qRand)
+    }
+
+    # Save EM specification
+    mkdir(here::here("Results", paste0("OM_", iom), paste0("EM_missSeason_", EMsetup[iem, "miss_season"], "_missQ_", EMsetup[iem, "miss_q"]))) # set up storage
+
+    saveRDS(inputEM, file = here::here("Results", paste0("OM_", iom), paste0("EM_missSeason_", EMsetup[iem, "miss_season"], "_missQ_", EMsetup[iem, "miss_q"]), "EMinput.Rds")) # save EM input
+
+  } # End loop over EM specifications
+
+} # End loop over OMs
+
 ```
+
+## Use common_code set_ecov function to adjust generic OM based on factorial OM settings above
+
+```{r}
+# # source(file.path(here::here(),"common_code", "set_ecov.R"))
+# source(here::here("set_ecov.R"))
+# source(here::here("get_FMSY.R")) # Contains set_F_scenario function
+# 
+# 
+# # Loop over rows in OMsetup to initialize OM and EMs
+# for(iom in 1:nrow(OMsetup)){
+#   # Pull generic groundfish settings
+#   input <- NULL # Clear info from prior iteration of loop
+#   input <- genericInput
+#   
+#   # Ecov setup
+#     Ecov <- list(label = "Ecov",
+#                  mean = matrix(rep(0,n_years), ncol =1), # Mean = 0
+#                  logsigma = matrix(log(OMsetup$Ecov_process_obs_sig[iom]), n_years, ncol=1),
+#                  years = input$years, 
+#                  use_obs = matrix(rep(TRUE, n_years), ncol = 1),
+#                  lag = 0, 
+#                  where = "q", # Where/how/indices settings need to change if we do sensitivity runs
+#                  how = 1,
+#                  indices = list(1,2),
+#                  process_model = "ar1", # End generic inputs for ecov
+#                  process_mean_vals = rep(0,n_years), # Mean = 0
+#                  process_sig_vals = rep(OMsetup$Ecov_process_sig[iom]),
+#                  process_cor_vals = rep(OMsetup$Ecov_process_cor[iom]),
+#                  beta_vals = list(rep(list(matrix(rep(0, n_ages), nrow = 1)), 2), # No impact on R,M
+#                                  rep(list(matrix(rep(OMsetup$Ecov_effect[iom], n_ages), nrow = 1)), 2))) # impact on q by index (here 2 indices)
+#     
+#   input <- set_ecov(input = input, ecov = Ecov)
+#   
+#   
+#   # Fishing history
+#     # Set F40
+#     temp <- fit_wham(input, do.fit = FALSE, MakeADFun.silent = TRUE)
+#     F40 <- exp(temp$rep$log_FXSPR[1]) %>% round(.,digits=3)
+#     Fmsy <- exp(temp$rep$log_FMSY[1]) %>% round(.,digits=3)
+#     # Get steepness so that Fmsy = F40
+#     steepness <- get_steepness(steepness_ini = 0.75, input = input, NAA_re = NAA_re, F40 = F40, brp_year = 1) # check brp_year
+#     NAA_re$recruit_pars[1] <- steepness$par
+#     input = set_NAA(input, NAA_re)
+#     temp <- fit_wham(input, do.fit = FALSE, MakeADFun.silent = TRUE)
+#     Fmsy <- exp(temp$rep$log_FMSY[1]) %>% round(.,digits=3)
+#     # Initial NAA according to equilibium conditions
+#     eq_F_N1 <- exp(temp$rep$log_FMSY[1]) # = F40
+#     
+#   
+#   
+#   input <- set_F_scenario(input = input, 
+#                           Fhist = OMsetup[iom, "F_hist"], 
+#                           Fmsy = Fmsy, 
+#                           max_mult = max_mult_Fmsy, 
+#                           min_mult= min_mult_Fmsy,
+#                           change_time = F_change_time) # Function pulled from get_FMSY.R
+# 
+# F_hist <- c("H-MSY", "MSY") # High-then MSY vs. MSY for entire history Fmsy is an option
+# 
+# # Observation error - set L-N SD parameters for catch and index age comps # pulled from project 0 q_om_setup.R
+# input$par$catch_paa_pars[,1] <- log(OMsetup[iom, "ageComp_sig"])
+# input$par$index_paa_pars[,1] <- log(OMsetup[iom, "ageComp_sig"])
+# input$data$agg_catch_sigma[] <- OMsetup[iom,"log_catch_sig"]
+# input$data$agg_index_sigma[] <- OMsetup[iom,"log_index_sig"]
+# input$data$catch_Neff[] <- 1 # Change Neff so scalar doesn't affect L-N SD
+# input$data$index_Neff[] <- 1
+# 
+# 
+# }
+# 
+# 
+#
+
+```

### 2023-09-22 Ecov_study/catchability/Setup_OMs.Rmd
Change OM numbering labels.

@@
 OMsetup <- expand.grid(Ecov_process_sig = Ecov_process_sig, Ecov_process_cor = Ecov_process_cor, Ecov_process_obs_sig = Ecov_process_obs_sig, Ecov_effect = Ecov_effect, F_hist = F_hist, ageComp_sig = ageComp_sig, log_index_sig = log_index_sig, log_catch_sig = log_catch_sig)
+
+OMname <- 1:nrow(OMsetup)
+
+OMsetup <- cbind(OMname, OMsetup)
+
 OMsetup
@@
   OM <- fit_wham(input = inputOM, do.fit = FALSE, MakeADFun.silent = TRUE)
-  mkdir(here::here("Results", paste0("OM_", iom))) # set up storage
-  saveRDS(OM, file = here::here("Results", paste0("OM_", iom), paste0("OM_", iom, ".Rds"))) # save OM
-  saveRDS(OMsetup[iom,], file = here::here("Results", paste0("OM_", iom), "OMsettings.Rds")) # save OM settings in a separate file since naming is based on arbitraty order in OMsetup table
+  mkdir(here::here("Results", paste0("OM_", OMsetup$OMname[iom]))) # set up storage
+  saveRDS(OM, file = here::here("Results", paste0("OM_", OMsetup$OMname[iom]), paste0("OM_", OMsetup$OMname[iom], ".Rds"))) # save OM
+  saveRDS(OMsetup[iom,], file = here::here("Results", paste0("OM_", OMsetup$OMname[iom]), "OMsettings.Rds")) # save OM settings in a separate file since naming is based on arbitraty order in OMsetup table
 

### 2023-09-22 Ecov_study/catchability/Catchability_ecov_sims.Rmd
Update simulations.

@@
 ```{r}
-# Packages
+## Packages
 library(tidyverse)
 library(wham)
-# Simulation testing functions
-source(here::here("Ecov_study","catchability","simOM.R"))
-source(here::here("Ecov_study","catchability","simTestWHAM.R"))
+
+## OM/EM setup function
+# source(here::here("Project_0", "code", "make_om.R"))
+source(here::here("make_om.R")) # From Project_0/code/make_om.R
+# source(file.path(here::here(),"common_code", "make_basic_info.R"))
+source(here::here("make_basic_info.R"))
+
+## Simulation testing functions
+# source(here::here("Ecov_study","catchability","simOM.R"))
+source(here::here("Other_Research", "simOM.R"))
+# source(here::here("Ecov_study","catchability","simTestWHAM.R"))
+source(here::here("Other_Research", "simTestWHAM.R"))
+
 ```
@@
 
-## Post-process results
+# Specifics for environmentally-linked catchability WHAM runs
 
+## Set up operating and estimation models
+This need only be done once by running Setup_OMs.Rmd !!! Need to move this code here so OMsetup and EMsetup available below for processing
+
+## Simulation testing
 ```{r}
+simTestWHAM(nsim = 5, OM = OM1, inputEMlist = list(EMq_env, EMq_rand_env, EMq_rand, EMq)) # Update once OM/EM specifics finalized!!!!
 
@@
 
-# Specifics for environmentally-linked catchability WHAM runs
+Run tests only for EMs with no seasonal misspecifications (only catchability misspecification misspecified)
+```{r}
+subsetEM <- EMsetup %>% as.data.frame() %>% filter(miss_season == "NONE")
 
-## Generic stock info that is never misspecified/changed
-```{r}
-# Pull generic groundfish life history/data
-source(file.path(here::here(),"common_code", "make_basic_info.R"))
-groundfish_info <- make_basic_info()
+for(iom in 1:192){
+  testOM <- readRDS(here::here("Results", paste0("OM_", iom, ".Rds")))
 
-# Assume logistic selectivity for the fleet and all indices
-sel_list <- list(model = c(rep("logistic", groundfish_info$n_fleets),rep("logistic", groundfish_info$n_indices)),
-                   initial_pars = rep(list(c(5,1)), groundfish_info$n_fleets + groundfish_info$n_indices),
-                 n_selblocks = c(groundfish_info$n_fleets + groundfish_info$n_indices)) #fleet, index
-
-# Starting mean M estimate at 0.2, estimate constant M for all ages/years
-M <- list(initial_means = rep(0.2, length(groundfish_info$ages)))
-
-# Assume random about the mean recruitment, beverton-holt stock recruit didn't converge with an invertible hessian
-recruit_model <- 2 
-# NAA_list <- list(sigma="rec") # Recruitment random effects so stock recruit function implemented
-
-genericInput <- prepare_wham_input(basic_info = groundfish_info, 
-                                   selectivity = sel_list, 
-                                   recruit_model = recruit_model) #,
-                                   #NAA_re = NAA_list)
-genericFit <- fit_wham(genericInput, MakeADFun.silent = TRUE, do.retro = FALSE, do.osa = FALSE)
-
+  for(iem in 1:4){
+    testEM <- readRDS(here::here("Results", paste0("OM_", iom), paste0("EM_missSeason_", subsetEM[iem, "miss_season"], "_missQ_", subsetEM[iem, "miss_q"]), "EMinput.Rds"))
+  }
+}
 ```
 
-## Define Operating Models (OMs)
-
-All OMs should have catchability linked to an environmental covariate
-but may differ in - Uncertainty around this relationship #!!!!!!!! -
-Sign of this relationship (+/-) #!!!!!!!!
-
-Set up input and fit OM1
-
+Run OMs with no ecov effect (NoEcov) and with ecov effect (Ecov) are equivalent here and the qRand and qRandEcov are also equivalent so only run the NoEcov and qRandEcov options
 ```{r}
-inputOM1 <- prepare_wham_input(model_name = "OM1") #!!!!!!!!!!!!
-OM1 <- fit_wham(inputOM1, do.osa = FALSE, do.retro = FALSE, MakeADFun.silent = TRUE)
-```
-
-## Define Estimation Models (EMs)
-
-| q-Env link                    | Seasonal misspecification |
-| ----------------------------- | ------------------------- |
-| Without link                  | Both seasons correct (q-Env link in EM and OM)         |
-| With link                     | 1 season misspecified (q-Env linked but no link in OM) |
-| Without link, q random effect | Both seasons misspecified (no link but link in OM)     |
-| With link, q random effect    | |
-
-Correctly specified link between catchability and the environment
-
-```{r}
-EMq_env <- prepare_wham_input(model_name = "EMq_env")
-```
-
-Catchability random effect and a link between catchability and the
-environment
-
-```{r}
-EMq_rand_env <- prepare_wham_input(model_name = "EMq_rand_env")
-```
-
-Only a catchability random effect, fit to environmental model without
-link
-
-```{r}
-EMq_rand <- prepare_wham_input(model_name ="EMq_rand")
-```
-
-No catchability considerations (i.e. status quo), fit to environmental
-model without link !!!! how does this work if you feed in the ecov_beta?
-
-```{r}
-EMq <- prepare_wham_input(model_name = "EMq")
-```
-
-## Simulation testing
-
-```{r}
-simTestWHAM(nsim = 5, OM = OM1, inputEMlist = list(EMq_env, EMq_rand_env, EMq_rand, EMq)) # Update once OM/EM specifics finalized!!!!
-
+subsetOM <- OMsetup %>% filter(Ecov_effect == 0)
+# EMs 
+subsetEM <- Msetup %>% filter(miss_season == "NONE") %>% filter(miss_q == "qRand" | miss_q == "NoEcov")
 ```
@@
 - Separate sims and function scripts
+
+# 9/18/23
+1) do OM & EM setup (EMs specified relative to each OM)
+```{r}
+Setup_OMs()
+```
+2) simulation test EMs for each OM
+```{r}
+simTestWHAM()
+```
+3) Post-process results
+```{r}
+postprocess_simTestWHAM()
+```

### 2023-09-22 Ecov_study/catchability/Setup_OMs.Rmd
Update comments as OM/EM setup code transitioned to Catchability_ecov_sims.Rmd

@@
 
--   B-H stock-recruit (for now)
-
 -   1 fleet
@@
 -   Logistic normal age composition for fleet and all indices
+
+-   Fishing history = F_MSY for 40 year history OR fish 2.5\*F_MSY for
+    20 years than fish at F_MSY for remaining 20
 
@@
     parameter but should confirm
-
--   Fishing history = F_MSY for 40 year history OR fish 2.5\*F_MSY for
-    20 years than fish at F_MSY for remaining 20 - NOT CURRENTLY IMPLEMENTED, IN PROGRESS
 
@@
 # 3) May run sensitivity run with no ecov effect in OM but compare same EMs
-Check that ecov effect = 0 is essentially turning off ecov
+Check that ecov effect = 0 is essentially turning off ecov = yes this is already implemented in step 1 outlined above
 ```{r}
@@
 
-Use project 0 make_om() to set up generic OM input then add additional ecov settings afterwards before fitting model
+Use unique copy of make_om() to set up generic OM input then add additional ecov settings afterwards before fitting model
 - By default catchability impacted by environment only for fall (2nd) index

### 2023-09-22 Ecov_study/catchability/Catchability_ecov_sims.Rmd
Fix file paths for project, move OM/EM setup from Setup_OMs.Rmd, scope sim runs.

@@
 library(wham)
+library(TAF)
 
-## OM/EM setup function
-# source(here::here("Project_0", "code", "make_om.R"))
-source(here::here("make_om.R")) # From Project_0/code/make_om.R
-# source(file.path(here::here(),"common_code", "make_basic_info.R"))
-source(here::here("make_basic_info.R"))
+## OM/EM setup functions
+source(here::here("Ecov_study", "catchability", "make_om.R")) # Use revised copy that has option not to include a S-R relationship
+source(file.path(here::here(),"common_code", "make_basic_info.R")) 
+source(here::here("common_code", "set_ecov.R")) 
+source(here::here("common_code", "set_NAA.R")) 
+source(here::here("common_code", "set_q.R"))
+source(here::here("common_code", "set_selectivity.R")) 
+source(here::here("common_code", "set_M.R")) 
+source(here::here("common_code", "get_FMSY.R"))
 
 ## Simulation testing functions
-# source(here::here("Ecov_study","catchability","simOM.R"))
-source(here::here("Other_Research", "simOM.R"))
-# source(here::here("Ecov_study","catchability","simTestWHAM.R"))
-source(here::here("Other_Research", "simTestWHAM.R"))
-
+source(here::here("Ecov_study","catchability","simOM.R"))
+#source(here::here("Other_Research", "simOM.R"))
+source(here::here("Ecov_study","catchability","simTestWHAM.R"))
+#source(here::here("Other_Research", "simTestWHAM.R"))
 ```
 
-### Visual describing results structure
+# Visual describing results structure
 
@@
 ## Set up operating and estimation models
-This need only be done once by running Setup_OMs.Rmd !!! Need to move this code here so OMsetup and EMsetup available below for processing
+### Factorial OM settings
+Vary ecov process magnitude, fishing history and observation error
+```{r}
+##### OM
+# Ecov process & effect magnitude set up
+Ecov_process_sig <- c(0.1, 0.5)
+Ecov_process_cor <- c(0, 0.5)
+Ecov_process_obs_sig <- c(0.1, 0.5)
+Ecov_effect <- c(0, 0.25, 0.5) # beta
+
+# Fishing history
+F_hist <- c("H-L", "Fmsy") # High-then FMSY vs. FMSY for entire history
+
+# Observation error
+ageComp_sig <- c(0.3, 1.5) # Fleet and index age comp treated with same error
+log_index_sig <- c(0.1, 0.4) # agg_index_sigma
+log_catch_sig <- 0.1 # agg_catch_sigma
+
+# Factorial combinations
+OMsetup <- expand.grid(Ecov_process_sig = Ecov_process_sig, Ecov_process_cor = Ecov_process_cor, Ecov_process_obs_sig = Ecov_process_obs_sig, Ecov_effect = Ecov_effect, F_hist = F_hist, ageComp_sig = ageComp_sig, log_index_sig = log_index_sig, log_catch_sig = log_catch_sig)
+
+OMname <- 1:nrow(OMsetup)
+
+OMsetup <- cbind(OMname, OMsetup)
+
+OMsetup
+```
+### Factorial EM settings
+```{r}
+##### EM
+# EM factorial settings
+miss_season <- c("BOTH", "ONE", "NONE") # Number of seasons with correctly specified effect, assumes 2 survey indices (spring & fall)
+miss_q <- c("NoEcov", "Ecov", "qRand", "qRandEcov") # Catchability setup for EM
+EMsetup <- expand.grid(miss_season = miss_season, miss_q = miss_q)
+EMsetup 
+```
+
+### Set up OM and corresponding EMs based on above factorial variations (need only be run once)
+- No stock-recruit relationship is implemented (as in project 0) so a variation of make_om() is used here to:
+  - 1) Set Fmsy equal to the F40% proxy
+  - 2) Estimate equilibirum initial numbers at age by setting NAA_re$N1_model = 1 rather than using the S-R
+- By default, the environment only impacts catchability for the fall (2nd) index. EM misspecifications are based off of this assumption.
+- Each OM is stored in a separate directory with input objects for EMs that are misspecified based on this OM in a subdirectory of the OM
+```{r}
+n_fleets <- 1
+n_indices <- 2
+n_ages <- 10
+n_years <- 40
+
+# Assume logistic selectivity for the fleet and all indices
+sel_list <- list(model = c(rep("logistic", n_fleets),rep("logistic", n_indices)),
+                 initial_pars = lapply(1:(n_fleets + n_indices), function(x) c(5,1)),
+                 n_selblocks = (n_fleets + n_indices))
+
+# Starting mean M estimate at 0.2, estimate constant M for all ages/years
+M_list <-  list(initial_means = rep(0.2, n_ages))
+
+# NAA_re = list(recruit_model = 3) # This required for Beverton-Holt S-R
+
+for(iom in 1:nrow(OMsetup)){
+  print(iom)
+  ##### Set up generic input #####
+  # Set up input without ecov
+  input <- make_om(Fhist = OMsetup[iom,"F_hist"],
+                   N1_state = "Fmsy", # Default, could also pick "overfished" or "unfished"
+                   selectivity = sel_list,
+                   M = M_list, 
+                   catchability = NULL,
+                   NAA_re = NULL) 
+  # age_comp, # Default = "logistic-normal-miss0"
+  # brp_year, # Default = 1
+  # eq_F_init, # Default = 0.3
+  # om_input = TRUE, # Don't fit model, only structure generic input
+  # max_mult_Fmsy, # Default = 2.5
+  # min_mult_Fmsy, # Default = 1
+  # F_change_time # Default = 0.5, sets when F changes (in H-L or L-H F_hist scenarios)
+
+  # Observation error - set L-N SD parameters for catch and index age comps # pulled from project 0 q_om_setup.R
+  input$par$catch_paa_pars[,1] <- log(OMsetup[iom, "ageComp_sig"])
+  input$par$index_paa_pars[,1] <- log(OMsetup[iom, "ageComp_sig"])
+  input$data$agg_catch_sigma[] <- OMsetup[iom,"log_catch_sig"]
+  input$data$agg_index_sigma[] <- OMsetup[iom,"log_index_sig"]
+  input$data$catch_Neff[] <- 1 # Change Neff so scalar doesn't affect L-N SD
+  input$data$index_Neff[] <- 1
+
+  ##### Set up OM ecov #####
+  # Ecov setup
+  Ecov <- list(label = "Ecov",
+               mean = matrix(rep(0,n_years), ncol =1), # Mean = 0
+               logsigma = matrix(log(OMsetup$Ecov_process_obs_sig[iom]), n_years, ncol=1),
+               years = input$years, 
+               use_obs = matrix(rep(TRUE, n_years), ncol = 1),
+               lag = 0, 
+               where = "q", # Where/how/indices settings need to change if we do sensitivity runs
+               how = 1,
+               indices = list(2), # Ecov only impact fall index catchability
+               process_model = "ar1", # End generic inputs for ecov
+               process_mean_vals = rep(0,n_years), # Mean = 0
+               process_sig_vals = rep(OMsetup$Ecov_process_sig[iom]),
+               process_cor_vals = rep(OMsetup$Ecov_process_cor[iom]),
+               beta_vals = list(append(rep(list(matrix(rep(0, n_ages), nrow = 1)), 2), # No impact on R,M
+                                       rep(list(matrix(rep(OMsetup$Ecov_effect[iom], n_ages), nrow = 1)), 2)))) # impact on q by index (here 2 indices)
+
+  inputOM <- set_ecov(input = input, ecov = Ecov) #!!! something wrong with ecov$beta_vals[[j]][[n]] indexing
+
+  # Finish setup & save OM
+  OM <- NULL # Clear prior OM object just in case
+  OM <- fit_wham(input = inputOM, do.fit = FALSE, MakeADFun.silent = TRUE)
+  mkdir(here::here("Ecov_study", "catchability", "Results", paste0("OM_", OMsetup$OMname[iom]))) # set up storage
+  saveRDS(OM, file = here::here("Ecov_study", "catchability", "Results", paste0("OM_", OMsetup$OMname[iom]), paste0("OM_", OMsetup$OMname[iom], ".Rds"))) # save OM
+  saveRDS(OMsetup[iom,], file = here::here("Ecov_study", "catchability", "Results", paste0("OM_", OMsetup$OMname[iom]), "OMsettings.Rds")) # save OM settings in a separate file since naming is based on arbitraty order in OMsetup table
+
+  ##### Iterate over EM misspecifications for this OM
+  for(iem in 1:nrow(EMsetup)){
+    inputEM <- NULL # Reset at start of loop just in case
+
+    # Seasonal misspecification (q random effect implemented with same misspecification as ecov impact)
+    if(EMsetup[iem, "miss_season"] == "BOTH"){
+      index_list <- list(1) # Both seasons misspecified
+      qRand <- c("ar1", "none")
+    } else if(EMsetup[iem, "miss_season"] == "ONE"){
+      index_list <- list(1,2) # Spring misspecified, fall correct
+      qRand <- c("ar1", "ar1")
+    } else{
+      index_list <- list(2) # Both seasons correctly specified
+      qRand <- c("none", "ar1")
+    }
+
+    # Catchability model assumptions
+    if(EMsetup[iem, "miss_q"] == "NoEcov"){ # No ecov impact on q or q random effect
+      print("NoEcov")
+      Ecov <- list(label = "NoEcov",
+                   mean = matrix(rep(0,n_years), ncol =1), # Mean = 0
+                   logsigma = matrix(log(OMsetup$Ecov_process_obs_sig[iom]), n_years, ncol=1),
+                   years = input$years, 
+                   use_obs = matrix(rep(TRUE, n_years), ncol = 1),
+                   lag = 0, 
+                   where = "none", # Where/how/indices settings need to change if we do sensitivity runs
+                   how = 1,
+                   indices = index_list, # Ecov impact based on sesonal misspecification setting
+                   process_model = "ar1", # End generic inputs for ecov
+                   process_mean_vals = rep(0,n_years), # Mean = 0
+                   process_sig_vals = rep(OMsetup$Ecov_process_sig[iom]),
+                   process_cor_vals = rep(OMsetup$Ecov_process_cor[iom]),
+                   beta_vals = list(append(rep(list(matrix(rep(0, n_ages), nrow = 1)), 2), # No impact on R,M
+                                           rep(list(matrix(rep(OMsetup$Ecov_effect[iom], n_ages), nrow = 1)), 2)))) # impact on q by index (here 2 indices)
+
+      inputEM <- set_ecov(input = input, ecov = Ecov)
+
+    } else if(EMsetup[iem, "miss_q"] == "Ecov"){ # Ecov impact on q, correctly specified option when seasonal impact correctly specified
+      print("Ecov")
+      Ecov <- list(label = "Ecov",
+                   mean = matrix(rep(0,n_years), ncol =1), # Mean = 0
+                   logsigma = matrix(log(OMsetup$Ecov_process_obs_sig[iom]), n_years, ncol=1),
+                   years = input$years, 
+                   use_obs = matrix(rep(TRUE, n_years), ncol = 1),
+                   lag = 0, 
+                   where = "q", # Where/how/indices settings need to change if we do sensitivity runs
+                   how = 1,
+                   indices = index_list, # Ecov impact based on sesonal misspecification setting
+                   process_model = "ar1", # End generic inputs for ecov
+                   process_mean_vals = rep(0,n_years), # Mean = 0
+                   process_sig_vals = rep(OMsetup$Ecov_process_sig[iom]),
+                   process_cor_vals = rep(OMsetup$Ecov_process_cor[iom]),
+                   beta_vals = list(append(rep(list(matrix(rep(0, n_ages), nrow = 1)), 2), # No impact on R,M
+                                           rep(list(matrix(rep(OMsetup$Ecov_effect[iom], n_ages), nrow = 1)), 2)))) # impact on q by index (here 2 indices)
+
+      inputEM <- set_ecov(input = input, ecov = Ecov)
+
+    } else if(EMsetup[iem, "miss_q"] == "qRand"){ # q random effect
+      print("qRand")
+      Ecov <- list(label = "qRand",
+                   mean = matrix(rep(0,n_years), ncol =1), # Mean = 0
+                   logsigma = matrix(log(OMsetup$Ecov_process_obs_sig[iom]), n_years, ncol=1),
+                   years = input$years, 
+                   use_obs = matrix(rep(TRUE, n_years), ncol = 1),
+                   lag = 0, 
+                   where = "none", # Where/how/indices settings need to change if we do sensitivity runs
+                   how = 1,
+                   indices = index_list, # Ecov impact based on sesonal misspecification setting
+                   process_model = "ar1", # End generic inputs for ecov
+                   process_mean_vals = rep(0,n_years), # Mean = 0
+                   process_sig_vals = rep(OMsetup$Ecov_process_sig[iom]),
+                   process_cor_vals = rep(OMsetup$Ecov_process_cor[iom]),
+                   beta_vals = list(append(rep(list(matrix(rep(0, n_ages), nrow = 1)), 2), # No impact on R,M
+                                           rep(list(matrix(rep(OMsetup$Ecov_effect[iom], n_ages), nrow = 1)), 2)))) # impact on q by index (here 2 indices)
+
+      inputEM <- set_ecov(input = input, ecov = Ecov)
+
+      # Catchability random effects
+      inputEM <- set_q(input = inputEM, catchability = list(re = qRand))
+
+    } else{ # qRandEcov both q random effect and ecov impact on q
+      print("qRandEcov")
+      Ecov <- list(label = "qRandEcov",
+                   mean = matrix(rep(0,n_years), ncol =1), # Mean = 0
+                   logsigma = matrix(log(OMsetup$Ecov_process_obs_sig[iom]), n_years, ncol=1),
+                   years = input$years, 
+                   use_obs = matrix(rep(TRUE, n_years), ncol = 1),
+                   lag = 0, 
+                   where = "q", # Where/how/indices settings need to change if we do sensitivity runs
+                   how = 1,
+                   indices = index_list, # Ecov impact based on sesonal misspecification setting
+                   process_model = "ar1", # End generic inputs for ecov
+                   process_mean_vals = rep(0,n_years), # Mean = 0
+                   process_sig_vals = rep(OMsetup$Ecov_process_sig[iom]),
+                   process_cor_vals = rep(OMsetup$Ecov_process_cor[iom]),
+                   beta_vals = list(append(rep(list(matrix(rep(0, n_ages), nrow = 1)), 2), # No impact on R,M
+                                           rep(list(matrix(rep(OMsetup$Ecov_effect[iom], n_ages), nrow = 1)), 2)))) # impact on q by index (here 2 indices)
+
+      inputEM <- set_ecov(input = input, ecov = Ecov)
+
+      # Catchability random effects
+      inputEM <- set_q(input = inputEM, catchability = list(re = qRand))
+    }
+
+    # Save EM specification
+    mkdir(here::here("Ecov_study", "catchability", "Results", paste0("OM_", iom), paste0("EM_missSeason_", EMsetup[iem, "miss_season"], "_missQ_", EMsetup[iem, "miss_q"]))) # set up storage
+
+    saveRDS(inputEM, file = here::here("Ecov_study", "catchability", "Results", paste0("OM_", iom), paste0("EM_missSeason_", EMsetup[iem, "miss_season"], "_missQ_", EMsetup[iem, "miss_q"]), "EMinput.Rds")) # save EM input
+
+  } # End loop over EM specifications
+
+} # End loop over OMs
+```
 
 ## Simulation testing
+If OMsetup or EMsetup objects are found then source the "Factorial OM settings" and "Factorial EM settings" sections above.
+DO NOT rerun the actual setup (the names are used to ID what subset of the OMs/EMs are being tested).
+
+### Test
 ```{r}
@@
 
-Run tests only for EMs with no seasonal misspecifications (only catchability misspecification misspecified)
+### Run tests only for EMs with no seasonal misspecifications (only catchability misspecification misspecified)
+Goal = run 100 simulations for each OM
+- postprocessing script will be set up to process multiple result files if these simulations need to be run in batches (i.e. reduce # of sims, # of OMs run, or # of EMs run at once)
+
+#### Amanda run Fhist = H-L simulations
 ```{r}
+# Set number of simulations to run
+nsim <- 2
+
+# Subset of OMs to run
+subsetOM <- OMsetup %>% filter(F_hist == "H-L")
+
+# Subset of EMs to run
 subsetEM <- EMsetup %>% as.data.frame() %>% filter(miss_season == "NONE")
 
-for(iom in 1:192){
-  testOM <- readRDS(here::here("Results", paste0("OM_", iom, ".Rds")))
+# Run simulation tests
+for(iom in 1:nrow(OMsetup)){
+  omdir <- here::here("Ecov_study", "catchability", "Results", paste0("OM_", OMsetup[iom, "OMname"]))
 
-  for(iem in 1:4){
-    testEM <- readRDS(here::here("Results", paste0("OM_", iom), paste0("EM_missSeason_", subsetEM[iem, "miss_season"], "_missQ_", subsetEM[iem, "miss_q"]), "EMinput.Rds"))
+  # Pull OM
+  testOM <- readRDS(here::here(omdir, paste0("OM_", OMsetup[iom, "OMname"], ".Rds")))
+
+  for(iem in 1:nrow(subsetEM)){
+    # Pull EM
+    testEM <- readRDS(here::here(omdir, paste0("EM_missSeason_", subsetEM[iem, "miss_season"], "_missQ_", subsetEM[iem, "miss_q"]), "EMinput.Rds"))
+
+    # Run simulation test
+    simTestWHAM(nsim = nsim,
+                OM = testOM,
+                inputEMlist = list(testEM), # Run one EM at a time
+                outdir = here::here(omdir, paste0("EM_missSeason_", subsetEM[iem, "miss_season"], "_missQ_", subsetEM[iem, "miss_q"])))
   }
 }
+
+```
+#### Alex run F_hist = Fmsy simulations
+```{r}
+# Set number of simulations to run
+nsim <- 2
+
+# Subset of OMs to run
+OMsetup %>% filter(F_hist == "Fmsy")
+
+# Subset of EMs to run
+subsetEM <- EMsetup %>% as.data.frame() %>% filter(miss_season == "NONE")
+
 ```
 
-Run OMs with no ecov effect (NoEcov) and with ecov effect (Ecov) are equivalent here and the qRand and qRandEcov are also equivalent so only run the NoEcov and qRandEcov options
+### Run OMs with no ecov effect 
+EMs with no ecov effect (NoEcov) and with ecov effect (Ecov) are equivalent when the OM has no ecov effect and the qRand and qRandEcov are also equivalent so only run the NoEcov and qRandEcov EMs for these OMs
 ```{r}
@@
 # EMs 
-subsetEM <- Msetup %>% filter(miss_season == "NONE") %>% filter(miss_q == "qRand" | miss_q == "NoEcov")
+subsetEM <- EMsetup %>% filter(miss_season == "NONE") %>% filter(miss_q == "qRand" | miss_q == "NoEcov")
 ```
+
+### Run tests for EMs with seasonal misspecifications
+Only run these for a subset of OMs to be selected following the outcomes of tests outlined above
 

### 2023-09-26 Ecov_study/catchability/Catchability_ecov_sims.Rmd
Add model names and update sim test code.

@@
 
+OMsetup[,"F_hist"] <- as.character(OMsetup[,"F_hist"])
+
 OMsetup
@@
   # Finish setup & save OM
+  inputOM$model_name <- paste0("OM_", paste(OMsetup[iom, ], collapse = "_")) # Set up OM name based on OMsetup 
   OM <- NULL # Clear prior OM object just in case
@@
 
+      # Set up EM name based on EMsetup
+      input$model_name <- paste0("EM_", paste(unfactor(EMsetup[iem, ]), collapse = "_")) # Set up OM name based on OMsetup
+
     } else if(EMsetup[iem, "miss_q"] == "Ecov"){ # Ecov impact on q, correctly specified option when seasonal impact correctly specified
@@
       inputEM <- set_ecov(input = input, ecov = Ecov)
+
+      # Set up EM name based on EMsetup
+      input$model_name <- paste0("EM_", paste(unfactor(EMsetup[iem, ]), collapse = "_")) # Set up OM name based on OMsetup
 
@@
 
+      # Set up EM name based on EMsetup
+      input$model_name <- paste0("EM_", paste(unfactor(EMsetup[iem, ]), collapse = "_")) # Set up OM name based on OMsetup
+
     } else{ # qRandEcov both q random effect and ecov impact on q
@@
       inputEM <- set_q(input = inputEM, catchability = list(re = qRand))
+
+      # Set up EM name based on EMsetup
+      input$model_name <- paste0("EM_", paste(unfactor(EMsetup[iem, ]), collapse = "_")) # Set up OM name based on OMsetup 
     }
@@
 ```{r}
-simTestWHAM(nsim = 5, OM = OM1, inputEMlist = list(EMq_env, EMq_rand_env, EMq_rand, EMq)) # Update once OM/EM specifics finalized!!!!
+#simTestWHAM(nsim = 5, OM = OM1, inputEMlist = list(EMq_env, EMq_rand_env, EMq_rand, EMq)) # Update once OM/EM specifics finalized!!!!
 
+# Set number of simulations to run
+nsim <- 2
+
+# Subset of OMs to run
+subsetOM <- OMsetup %>% filter(F_hist == "H-L") %>% head(n=2)
+
+# Subset of EMs to run
+subsetEM <- EMsetup %>% as.data.frame() %>% filter(miss_season == "NONE")
+
+# Run simulation tests
+for(iom in 1:nrow(subsetOM)){
+  omdir <- here::here("Ecov_study", "catchability", "Results", paste0("OM_", OMsetup[iom, "OMname"]))
+
+  # Pull OM
+  testOM <- readRDS(here::here(omdir, paste0("OM_", OMsetup[iom, "OMname"], ".Rds")))
+
+  for(iem in 1:nrow(subsetEM)){
+    # Pull EM
+    testEM <- readRDS(here::here(omdir, paste0("EM_missSeason_", subsetEM[iem, "miss_season"], "_missQ_", subsetEM[iem, "miss_q"]), "EMinput.Rds"))
+
+    # Run simulation test
+    simTestWHAM(nsim = nsim,
+                OM = testOM,
+                inputEMlist = list(testEM), # Run one EM at a time
+                outdir = here::here(omdir, paste0("EM_missSeason_", subsetEM[iem, "miss_season"], "_missQ_", subsetEM[iem, "miss_q"])))
+  }
+}
 ```
@@
 # Run simulation tests
-for(iom in 1:nrow(OMsetup)){
+for(iom in 1:nrow(subsetOM)){
   omdir <- here::here("Ecov_study", "catchability", "Results", paste0("OM_", OMsetup[iom, "OMname"]))
@@
 #### Alex run F_hist = Fmsy simulations
+This code loops over the rows in subsetOM and subsetEM.
+The easiest way to run smaller chunks is to change the number of rows in subsetOM (i.e.
+run fewer OMs at once) or run fewer simulations.
+Results are all time-stamped and postprocess_simTestWHAM() is set up to combine multiple files so this won't cause processing issues later on.
 ```{r}
 # Set number of simulations to run
-nsim <- 2
+nsim <- 100
 
 # Subset of OMs to run
-OMsetup %>% filter(F_hist == "Fmsy")
+subsetOM <- OMsetup %>% filter(F_hist == "Fmsy") %>% filter(Ecov_effect != 0)
 
@@
 subsetEM <- EMsetup %>% as.data.frame() %>% filter(miss_season == "NONE")
+
+# Run simulation tests
+for(iom in 1:nrow(subsetOM)){
+  omdir <- here::here("Ecov_study", "catchability", "Results", paste0("OM_", OMsetup[iom, "OMname"]))
+
+  # Pull OM
+  testOM <- readRDS(here::here(omdir, paste0("OM_", OMsetup[iom, "OMname"], ".Rds")))
+
+  for(iem in 1:nrow(subsetEM)){
+    # Pull EM
+    testEM <- readRDS(here::here(omdir, paste0("EM_missSeason_", subsetEM[iem, "miss_season"], "_missQ_", subsetEM[iem, "miss_q"]), "EMinput.Rds"))
+
+    # Run simulation test
+    simTestWHAM(nsim = nsim,
+                OM = testOM,
+                inputEMlist = list(testEM), # Run one EM at a time
+                outdir = here::here(omdir, paste0("EM_missSeason_", subsetEM[iem, "miss_season"], "_missQ_", subsetEM[iem, "miss_q"])))
+  }
+}
 
@@
 EMs with no ecov effect (NoEcov) and with ecov effect (Ecov) are equivalent when the OM has no ecov effect and the qRand and qRandEcov are also equivalent so only run the NoEcov and qRandEcov EMs for these OMs
+#### Amanda runs H-L Fhist
 ```{r}
-subsetOM <- OMsetup %>% filter(Ecov_effect == 0)
+# Set number of simulations to run
+nsim <- 100
+
+# Subset of OMs to run
+subsetOM <- OMsetup %>% filter(F_hist == "H-L") %>% filter(Ecov_effect != 0)
 # EMs 
 subsetEM <- EMsetup %>% filter(miss_season == "NONE") %>% filter(miss_q == "qRand" | miss_q == "NoEcov")
+
+# Run simulation tests
+for(iom in 1:nrow(subsetOM)){
+  omdir <- here::here("Ecov_study", "catchability", "Results", paste0("OM_", OMsetup[iom, "OMname"]))
+
+  # Pull OM
+  testOM <- readRDS(here::here(omdir, paste0("OM_", OMsetup[iom, "OMname"], ".Rds")))
+
+  for(iem in 1:nrow(subsetEM)){
+    # Pull EM
+    testEM <- readRDS(here::here(omdir, paste0("EM_missSeason_", subsetEM[iem, "miss_season"], "_missQ_", subsetEM[iem, "miss_q"]), "EMinput.Rds"))
+
+    # Run simulation test
+    simTestWHAM(nsim = nsim,
+                OM = testOM,
+                inputEMlist = list(testEM), # Run one EM at a time
+                outdir = here::here(omdir, paste0("EM_missSeason_", subsetEM[iem, "miss_season"], "_missQ_", subsetEM[iem, "miss_q"])))
+  }
+}
+```
+#### Alex runs Fmsy Fhist
+```{r}
+# Set number of simulations to run
+nsim <- 100
+
+# Subset of OMs to run
+subsetOM <- OMsetup %>% filter(F_hist == "Fmsy") %>% filter(Ecov_effect != 0)
+# EMs 
+subsetEM <- EMsetup %>% filter(miss_season == "NONE") %>% filter(miss_q == "qRand" | miss_q == "NoEcov")
+
+# Run simulation tests
+for(iom in 1:nrow(subsetOM)){
+  omdir <- here::here("Ecov_study", "catchability", "Results", paste0("OM_", OMsetup[iom, "OMname"]))
+
+  # Pull OM
+  testOM <- readRDS(here::here(omdir, paste0("OM_", OMsetup[iom, "OMname"], ".Rds")))
+
+  for(iem in 1:nrow(subsetEM)){
+    # Pull EM
+    testEM <- readRDS(here::here(omdir, paste0("EM_missSeason_", subsetEM[iem, "miss_season"], "_missQ_", subsetEM[iem, "miss_q"]), "EMinput.Rds"))
+
+    # Run simulation test
+    simTestWHAM(nsim = nsim,
+                OM = testOM,
+                inputEMlist = list(testEM), # Run one EM at a time
+                outdir = here::here(omdir, paste0("EM_missSeason_", subsetEM[iem, "miss_season"], "_missQ_", subsetEM[iem, "miss_q"])))
+  }
+}
 ```
@@
 ```{r}
-postprocess_simTestWHAM(filenames = c("filename1.Rds", "filename2.Rds", "ect.Rds")) # !!! Update once completed runs
+# ID all simulation result files 
+files <- list.files(path = here::here(), pattern = "simWHAM_", recursive = TRUE)
+
+postprocess_simTestWHAM(filenames = files) 
 ```
@@
 
-### Result formatted as list of lists, save as rdata object in specified directory with timestamp in name
+If results processed in batches/on multiple computers use combinePerfMet to combine final results
+```{r}
 
-run1_11_14_22.Rda\<-result - EM1 - EM2 - EM3 - EM4 - OM result
-
-# Other things to consider
-
--   Other EMs can be developed (e.g. with alternative ecov process
-    assumptions) but should be given DIFFERENT model_names to help with
-    post-processing
-
-# Short script for function testing purposes only!!!
-
-```{r}
-library(wham)
-# For testing function
-testdata <- read_asap3_dat(paste("/Users/amandahart/Research/Plaice_Methods_Development", "data", "PlaiceWHAM-2019_revised_NEFSC-LW-WAA_splitNEFSC-BigUnit_fixSeason.DAT", sep="/")) # this can be swapped out for any ASAP file you have to set up a WHAM model for debugging
-testinput <- prepare_wham_input(testdata)
-testWHAM <- fit_wham(testinput, MakeADFun.silent = TRUE, do.osa = FALSE, do.retro=FALSE)
-
-OM <- testWHAM
-testinput1 <- prepare_wham_input(testdata, model_name = "test1")
-testinput2 <- prepare_wham_input(testdata, model_name="test2")
-testinput3 <- prepare_wham_input(testdata, model_name="test3")
-inputEMlist <- list(testinput1, testinput2, testinput3)
-
-test <- simTestWHAM(nsim = 2, OM = OM, inputEMlist = inputEMlist)
 ```
-
-# To do next
-
-- Finish OM setup
-  - change how ecov passed in
-  - set up range of uncertainty values to explore
-  - tim sets equilibrium F (eq_F_init)
-    - setup OM without fit & use model to find F40 proxy
-  - Tim also sets intitial numbers at age based on equilibrium (overfished F=eqFN1*max_mult_Fmsy or not eqFN1=0)
-- Tim lines 71-79 pf make_om use internal processing for prepare_wham_input to plug in setup from M_Ecov_om_setup.R
-- Update what is saved as output
-  - calculate performance metrics
-- Run simulations, hpcc?
-- Separate sims and function scripts
-
-# 9/18/23
-1) do OM & EM setup (EMs specified relative to each OM)
-```{r}
-Setup_OMs()
-```
-2) simulation test EMs for each OM
-```{r}
-simTestWHAM()
-```
-3) Post-process results
-```{r}
-postprocess_simTestWHAM()
-```

### 2023-09-29 Ecov_study/catchability/Catchability_ecov_sims.Rmd
Fix error in OM loop so actually walks through subsetOM object, note on EM model_name.

@@
 library(TAF)
+library(varhandle)
 
@@
 - Each OM is stored in a separate directory with input objects for EMs that are misspecified based on this OM in a subdirectory of the OM
+!!! EM model_name labels may be misslabeled, but file path labels for EM storage are correct so those are used in postprocessing
 ```{r}
@@
       # Set up EM name based on EMsetup
-      input$model_name <- paste0("EM_", paste(unfactor(EMsetup[iem, ]), collapse = "_")) # Set up OM name based on OMsetup
+      input$model_name <- paste0("EM_", paste(unfactor(EMsetup[iem, ]), collapse = "_"))
 
@@
       # Set up EM name based on EMsetup
-      input$model_name <- paste0("EM_", paste(unfactor(EMsetup[iem, ]), collapse = "_")) # Set up OM name based on OMsetup 
+      input$model_name <- paste0("EM_", paste(unfactor(EMsetup[iem, ]), collapse = "_")) 
     }
@@
 # Set number of simulations to run
-nsim <- 2
+nsim <- 2 # 2
 
 # Subset of OMs to run
-subsetOM <- OMsetup %>% filter(F_hist == "H-L")
+subsetOM <- OMsetup %>% filter(F_hist == "H-L") %>% filter(Ecov_effect != 0) %>%
+  filter(OMname >100)
+# %>% head(n=10)
+subsetOM <- subsetOM[8:nrow(subsetOM),] #7?
 
@@
 for(iom in 1:nrow(subsetOM)){
-  omdir <- here::here("Ecov_study", "catchability", "Results", paste0("OM_", OMsetup[iom, "OMname"]))
+  omdir <- here::here("Ecov_study", "catchability", "Results", paste0("OM_", subsetOM[iom, "OMname"]))
 
   # Pull OM
-  testOM <- readRDS(here::here(omdir, paste0("OM_", OMsetup[iom, "OMname"], ".Rds")))
+  testOM <- readRDS(here::here(omdir, paste0("OM_", subsetOM[iom, "OMname"], ".Rds")))
 
@@
 # Subset of OMs to run
-subsetOM <- OMsetup %>% filter(F_hist == "H-L") %>% filter(Ecov_effect != 0)
+subsetOM <- OMsetup %>% filter(F_hist == "H-L") %>% filter(Ecov_effect == 0)
 # EMs 
@@
 # Subset of OMs to run
-subsetOM <- OMsetup %>% filter(F_hist == "Fmsy") %>% filter(Ecov_effect != 0)
+subsetOM <- OMsetup %>% filter(F_hist == "Fmsy") %>% filter(Ecov_effect == 0)
 # EMs 
@@
   # Pull OM
-  testOM <- readRDS(here::here(omdir, paste0("OM_", OMsetup[iom, "OMname"], ".Rds")))
+  testOM <- readRDS(here::here(omdir, paste0("OM_", subsetOM[iom, "OMname"], ".Rds")))
 
@@
 
-postprocess_simTestWHAM(filenames = files) 
+perfMet <- postprocess_simTestWHAM(filenames = files)
+
 ```
@@
 ```
+
+### Plot results
+```{r}
+plotResults(results = perfMet, convergedONLY = TRUE)
+```

### 2023-09-29 Ecov_study/mortality/paper/paper.Rmd
save edits to ecov m paper

@@
 
-## Abstract {-}
+## Summary/Abstract {-}
 
+We completed a large scale simulation study with 288 operating models with 12 estimated models fit to each.
+The factors defining operating model configuration inlcuded the source of process error on the population (recruitment, recruitment and survival, recruitment and natural mortality), the degree of temporal variation and autocorrelation of the environmental covariate, the uncertainty in the observation of the covariate, the uncertainty in indices and age composition observations, fishing history, and the magnitude of the effect of the covariate on natural mortality.
+
+We found convergence of all estimation models was generally best when operating models assumed process errors in recruitment and survival, constant fishing rate, greater contrast in the true environmental covariate, and lower uncertainty in corresponding observations.
+Reliable convergence of all estimating models also occurred with the same process errors in the operating model and a step-change in fishing, but this also required lower uncertainty in index and age composition observations.
+Estimating models with process errors on recruitment and survival were unlikely to converge when the process errors in the operating model did not match whereas estimating models with process errors in recruitment and natural mortality converged for operating models without this match in certain cases.
+Probability of convergence generally decreased when the mean natural mortality rate parameter was estimated.
+
+AIC paragraph
+
+Bias paragraph
+
+These results suggest that models with covariate effects on M will be most reliable when R+S is true, large contrast in the 
+Recommendations
 ### Keywords {-}
@@
 Here we conduct a simulation study with operating models varying by degree of observation error uncertainty, sources of process error (M or survival), fishing history, temporal variation in environmental covariates, and magnitude of the effect of the covariate on natural mortality.
-The simulations from these operating models are fitted with estimating models that make alternative assumptions for sources of process error (M, NAA), and whether (mean) M is estimated.
-We evaluate whether AIC can correctly determine the correct source of process error and stock effects on recruitment.
-We also evaluate the degree of bias in the outputs of the assessment model that are important for management.
+The simulations from these operating models are fitted with estimating models that make alternative assumptions for sources of process error, and whether (mean) M is estimated.
+We evaluate effects of these factors on convergence of fitted models, whether AIC can correctly determine the correct source of process error and correct assumption about covariate effects on natural mortality, and the degree of bias in relevant parameters and outputs of the assessment model.
 
@@
 
-We used the WHAM package \citep[][commit 77bbd94]{millerstock20,stockmiller21} to configure all operating models and estimation models.
-This packages has also been used to configure operating and estimating models for closed loop simulations evaluating index-based assessment methods \citep{legaultetal23} and is used for management of haddock, butterfish, plaice, bluefish in the Northeast US.
+We used the WHAM package \citep[][version 1.0.7, commit 77bbd94]{millerstock20,stockmiller21} to configure all operating models and estimation models.
+This packages has also been used to configure operating and estimating models for closed loop simulations evaluating index-based assessment methods \citep{legaultetal23} and is used for management of GB haddock, butterfish, plaice, bluefish, and Atlantic cod stocks in the Northeast US.
 
-We completed a simulation study with a number of operating models. Many of the characteristics of the operating models are the same as those used by \citet{milleretal_wp}.
+We completed a simulation study with a number of operating models.
+Many of the characteristics of the operating models are the same as those used by \citet{milleretal_project0_wp}.
+The factors defining the configuration of each operating model which are described in detail in subsequent sections include
+\begin{itemize}
+\item two configurations of fishing histories, 
+\item two configurations of index and age composition observation error, 
+\item four configurations of latent environmental covariate process errors, 
+\item two configurations of environmental covariate observation uncertainty, 
+\item three levels of covariate effects on natural mortality, and 
+\item three configurations of sources of process errors for temporal changes in abundances at age. 
+\end{itemize} 
+The full factorial of the levels of each factor results in 288 different operating models. We simulated 100 data sets for each operating model that included simulations of process errors.
 
-288 total operating models.
-We simulated 100 data sets for each operating model.
-For each simulated data set we fit a set of 12 estimating models.
+For each simulated data set we fit a set of 12 estimating models. The factors defining the configuration or each estimating model which are also described in detail below include
+\begin{itemize}
+\item three configurations of sources of process errors for temporal changes in abundances at age, 
+\item whether the (mean) natural mortality rate was estimated or assumed known at the true value, and
+\item whether the effect of the environmental covariate on natural mortality was estimated.
+\end{itemize}
+
+We did not use the bias-correction feature for process errors or observations described by \citet{stockmiller21} for operating and estimating models.
+Simulations were all carried out on the University of Massachusetts Green High-Performance Computing Cluster.
+Code for completing the simulations and summarization of results can be found at github.com/timjmiller/SSRTWG/ecov_study/mortality.
 
@@
 
-We assumed a Beverton-Holt stock recruit function with constant pre-recruit mortality parameters for all operating models.
-All post-recruit productivity components are constant in the NAA and survey catchability process error operating models.
+We assumed a Beverton-Holt stock recruit function with constant pre-recruit mortality parameters (i.e., $\alpha$ and $\beta$) for all operating models.
+All post-recruit productivity components are constant in the NAA and survey selectivity process error operating models.
 Therefore steepness and unfished recruitment are also constant over the time period for those operating models \citep{millerbrooks21}.
@@
 
-For operating models with time-varying random effects or covariate effects on M, steepness is not constant, but we used the same alpha and beta parameters as other operating models this equates to a steepness and R0 at the mean of the time series process for M.
-For operating models with time-varying random effects for fishery selectivity, \Fmsy is also not constant however we use the same F history as other operating models which corresponds to Fmsy at the mean selectivity parameters.
+For operating models with time-varying random effects or covariate effects on M, steepness is not constant, but we used the same alpha and beta parameters as other operating models this equates to a steepness and $R_0$ at the mean of the time series process for M.
+For operating models with time-varying random effects for fishery selectivity, \Fmsy is also not constant however we use the same F history as other operating models which corresponds to \Fmsy at the mean selectivity parameters.
 
@@
 $$
-E_y|E_{y-1} \sim \text{N}\left(\mu_\text{Ecov}\left(1-\rho_\text{Ecov}\right) + \rho_\text{Ecov} E_{y-1}, \sigma^2_\text{Ecov}\right)
+E_y|E_{y-1} \sim \text{N}\left(\mu_\text{Ecov}\left(1-\rho_\text{Ecov}\right) + \rho_\text{Ecov} E_{y-1}, \left(1-\rho_\text{Ecov}^2\right)\sigma^2_\text{Ecov}\right)
 $$
-with marginal mean $\mu_\text{Ecov}$ and variance $\frac{\sigma^2_\text{Ecov}}{1-\rho_\text{Ecov}^2}$.
+with marginal mean $\mu_\text{Ecov}=0$ and variance $\sigma^2_\text{Ecov}$.
 The configuration of the latent environmental covariate in the operating models have one of two alternate assumptions about the marginal standard deviation and autocorrelation parameter:
 \begin{itemize}
-\item marginal standard deviation $\in \{0.1, 0.5\}$
-\item AR1 correlation $\rho_\text{Ecov} \in \{0, 0.9\}$
+\item marginal standard deviation $\sigma_\text{Ecov} \in \{0.1, 0.5\}$
+\item AR1 correlation $\rho_\text{Ecov} \in \{0, 0.5\}$
 \end{itemize}
@@
 $$
-\log M_y = \beta_M + \beta_{\text{Ecov}} E_y + \epsilon_{M,y}
+\log M_y = \beta_M + \beta_{\text{Ecov}} E_y + \varepsilon_{M,y}
 $$
-that is a linear combination of a mean $\beta_M$ and any annual covariate $\beta_{\text{Ecov}}$ and random effects marginally distributed as $\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)$.
+that is a linear combination of a mean $\beta_M$ and any annual covariate $\beta_{\text{Ecov}}$ and random effects marginally distributed as $\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)$ \citep{stockmiller21}.
+For operating models with natural mortality random effects, we assume the same standard deviation $\sigma_M = 0.3$.
 
@@
 
-The alternative configurations of random effects on recruitment, natural mortality, survival
+As described by \citet{stockmiller21}, the model for process errors on abundance at age in WHAM is
+\begin{equation}
+  \text{log}N_{a,y}=\left\{
+    \begin{array}{@{}lll@{}}
+      \text{log} \left( f(SSB_{y-1}) \right) + \varepsilon_{1,y}, & \text{if}\ a = 1 \\
+      \text{log} \left( N_{a-1,y-1} \right) - Z_{a-1,y-1} + \varepsilon_{a,y}, & \text{if}\ 1 < a < A \\
+      \text{log} \left( N_{A-1,y-1} e^{-Z_{A-1,y-1}} + N_{A,y-1} e^{-Z_{A,y-1}} \right) + \varepsilon_{A,y}, & \text{if}\ a = A
+    \end{array}\right.
+\end{equation}
+We assume the the standard deviation of recruitment deviations is $\text{SD}\left(\varepsilon_{1,y}\right) = \sigma_R = 0.5$ for all operating models.
+For operating models with survival devations for older ages, we assume the same standard deviations for all older ages $\text{SD}\left(\varepsilon_{a,y}\right) = \sigma_{2+} = 0.3$.
+
+The alternative configurations of random effects on recruitment, survival, and natural mortality are
 \begin{itemize}
-\item uncorrelated random effects on R and survival with $\sigma_R = 0.5$, marginal standard deviation of survival of older age clases of 0.3
-\item uncorrelated random effects on R and survival with $\sigma_R = 0.5$, marginal standard deviation of natural mortality of 0.3
+\item uncorrelated random effects on recruitment only ($\sigma_R = 0.5$) (R),
+\item uncorrelated random effects on recruitment ($\sigma_R = 0.5$) and survival of older age classes ($\sigma_{2+} = 0.3$) (R+S), and
+\item uncorrelated random effects on R ($\sigma_R = 0.5$) and log-natural mortality ($\sigma_M = 0.3$) (R+M).
 \end{itemize}
@@
 We assumed a single fleet operating year round for catch observations with logistic selectivity for the fleet with $a_{50} = 5$ and slope = 1 (Figure \ref{om_mean_selectivity}).
-This selectivity is was used to define \Fmsy for the Beverton-Holt stock recruitment parameters above.
-We assumed a logistic-normal distribution for the age-composition observations for the fleet.
+This selectivity was used to define \Fmsy, \Fspr[40], and steepness for the Beverton-Holt stock recruitment parameters above.
+We assumed a logistic-normal distribution for the age-composition observations associated with the fleet where errors in the multivariate normal transformation are independent.
+The standard deviation parameter is also constant across ages.
 
@@
 Two time series of surveys are assumed and observed in numbers rather than biomass for the entire 40 year period with one occurring in the spring (0.25 way through the year) and one in the fall (0.75 way through the year).
-Actually we have it currently configured that both occur 0.5 way through the year.
 Catchability of both surveys are assumed to be 0.1.
-Like the fishing fleet, we assumed logistic selectivity for both indices with $a_{50} = 5$ and slope = 1 and a logistic-normal distribution for the age-composition observations.
+We assumed the same selectivity and age composition distribution as the fleet for both indices.
 
@@
 There were two levels of observation error variance for indices and age composition for both indices and fleet catch.
-A low uncertainty specification assumed standard deviation of both series of log-aggregate index observations was 0.1 and the standard deviation of the logistic-normal for age composition observations was 0.3 In the high uncertainty specification the standard deviation for log-aggregate indices was 0.4 and that for the age composition observations was 1.5.
+A low uncertainty specification assumed standard deviation of both series of log-aggregate index observations was 0.1 and the standard deviation of the logistic-normal for age composition observations was 0.3.
+In the high uncertainty specification the standard deviation for log-aggregate indices was 0.4 and that for the age composition observations was 1.5.
 For all estimating models, standard deviation for log-aggregate observations was assumed known whereas that for the logistic-normal age composition observations was estimated.
-
-The factorial combinations of fishing histories, index and age composition observation error assumptions, latent environmental covariate and observation assumptions, covariate effects on natural mortality, and sources of process error (R, R+S, R+M) results in 288 different operating models.
 
@@
 The configuration of the process errors in the estimating models matched the corresponding options in the operating models.
-For example, if R+S was assumed for both the estimating and operating model, the assumptions about process errors matched.
+For example, if uncorrelated R+S was assumed for both the estimating and operating model, the assumptions about process errors matched.
 The environmental covariate observations were included in all estimation models to ensure comparability of information criteria.
+All fixed effects parameters for selectivity, catchability, fully-selected fishing mortality, mean recruitment, initial abundance at age, and variances for logistic-normal age composition distributions were estimated.
+Any process error variance parameters for recruitment, survival, and natural mortality were also estimated.
 The observation error variance of the environmental observations and aggregate catch and indices were all assumed known at the true values.
-
-Simulations were all carried out on the University of Massachusetts Green High-Performance Computing Cluster.
-Code for completing the simulations and summarization of results can be found at github.com/timjmiller/SSRTWG/ecov_study/mortality.
-We used the wham package version 1.X.X (commit 77bbd94).
 
@@
 
+### Convergence {-}
+
 The first measure of reliability we investigated was frequency of convergence when fitting each estimating model to the simulated data sets.
 There are various ways to assess convergence of the fit.
-We summarized 5 alternative categories of convergence.
-1. Did the optimization routine (stats::nlminb) complete without error?
-2. Did the stats::nlminb convergence flag = 0 indicate successful convergence?
-3. Was the maximum absolute value of the gradient of the log-likelihood < $1\times10^{-6}$?
-4. Did TMB::sdreport provide non-NA values for all fixed effects standard errors?
-5. Did TMB::sdreport provide all standard errors < 100?
-
+We summarized 5 alternative categories of convergence:
+\begin{enumerate}
+\item Did the optimization routine (stats::nlminb) complete without error?
+\item Did the stats::nlminb convergence flag = 0 indicate successful convergence?
+\item Was the maximum absolute value of the gradient of the log-likelihood < $1\times10^{-6}$?
+\item Did TMB::sdreport provide non-NA values for all fixed effects standard errors?
+\item Did TMB::sdreport provide all standard errors < 100?
+\end{enumerate}
 The first convergence criterion assesses whether the model crashes.
@@
 We measured the frequency of correct model selection using marginal AIC.
-For a given operating model we the set of models that were considered all made the same assumptions on whether or not to estimate M and whether a stock-recruit function is assumed.
+For a given operating model the set of models that were considered all made the same assumptions on whether or not to estimate (mean) natural mortality rate or it is assumed at the true value $\beta_M = log(0.2)$.
+For model $m$, the marginal AIC is a function of the marginal log-likelihood maximized with respect to the fixed effects in the model $\boldsymbol{\theta}$ and the number of fixed effects $n\left(\boldsymbol{\theta}\right)$ estimated,
+$$
+\text{AIC}_m = -2\left[{\text{argmax}}_{\boldsymbol{\theta}} \log L_m\left({\boldsymbol{\theta}}\right) - n\left({\boldsymbol{\theta}}\right)\right].
+$$
 
@@
 
-We calculated relative errors of various quantities for each fitted estimation model for the $i$th simulated data set as
+We calculated median relative errors of
+\begin{itemize}
+\item $\beta_\text{Ecov}$, the effect of environmental covariates on natural mortality, 
+\item $\beta_M$, the mean log-natural mortality rate,
+\item annual natural mortality rate,
+\item annual spawning stock biomass, and
+\item annual fully-selected fishing mortality rate.
+\end{itemize}
+For the $i$th simulated data the relative error for a parameter $\theta$ provided from the fitted estimation model is
 $$
@@
 and measured bias as the median relative error and constructed 95\% confidence intervals using the binomial distribution approach as in \citet{stockmiller21}.
+If the confidence interval contains zero we do not have evidence of bias from the simulation study.
+We used estimates from all simulations that satisfied the first type of convergence described above.
+We were not concerned about more restrictive types of convergence because some estimation models would not be expected to satisfy these criteria because of mis-specified structures.
+For example, an estimation model that includes survival random effects fitted to a set of observations simulated without survival random effects should have trouble estimating variance of the survival random effects, but the estimates of the parameters and derived output described above, might be reliable.
 
@@
 ## Convergence performance {-}
+
+We focused on the fourth convergence criteria, that the hessian of the negative-marginal log-likelihood at the optimized parameters was invertible.
+When the (mean) natural mortality rate was assumed at the true value, corresponding to usual practice in application of assessment models, good convergence for all EMs was observed for R+S operating models with \Fmsy fishing history, more variation in the latent environmental covariate ($\sigma_\text{Ecov} = 0.5$), and lower error in the associated observations $\sigma_\text{ecov} = 0.1$ (Figure \ref{convergence_M_fixed}).
+Good convergence for all EMs was also observed for R+S OMs with the step change in fishing history, but with low error in indices and age comp observations.
+
+EMs that assumed no survival or natural mortality random effects (R OMS), nearly always converged across all operating models that assumed more variation in the latent environmental covariate ($\sigma_\text{Ecov} = 0.5$) and lower error in the associated observations $\sigma_\text{ecov} = 0.1$.
+EMs that assumed recruitment and survival random effects (R+S OMS), had poor converge probability when the OMs had alternative process errors (R or R+M).
+The R+S EMs showed best convergence for the R+S OMs where there was lower error in the environmental observations $\sigma_\text{ecov} = 0.1$ or higher error in environmental observations of latent true environmental covariates with greater temporal variation.
+EMs that assumed recruitment and natural mortality random effects (R+M OMs), had poor convergence probability for all OMs that had a step change in fishing history and higher uncertainty in indices and age composition.
+
+When the OMs and EMs both assume process errors only on recruitment (R) or on recruitment and survival (R+S), convergence was worst with less variation in the true environmental covariate and larger uncertainty in associated observations.
+When the OMs and EMs both assume process errors on recruitment and natural mortality (R+M) convergence was problematic for all OMs with the step-change in fishing history.
+The best convergence was observed with this match between OMs and EMs was when the fishing history was constant and there was low uncertainty in environmental observations.
 
@@
 \begin{center}
-\includegraphics[width = \textwidth]{proportion_good_hessian_ecov_effect_est_M_fixed.png}
+\includegraphics[height = \textheight]{proportion_good_hessian_ecov_effect_est_M_fixed.png}
 \end{center}
@@
 \end{landscape}
+
+As might be expected, there was an overall drop in the probability of convergence when the mean natural mortality rate was estimated rather than assumed at the true value (Figure \ref{convergence_M_estimated}).
+Otherwise, general trends described above with mean natural mortality fixed, apply when estimated.
 
@@
 \begin{center}
-\includegraphics[width = \textwidth]{proportion_good_hessian_ecov_effect_est_M_est.png}
+\includegraphics[height = \textheight]{proportion_good_hessian_ecov_effect_est_M_est.png}
 \end{center}
@@
 
+When estimating models assumed the mean natural mortality rate was known, the best accuracy of AIC for model selection occurred for models with R+S process errors.
+R+S estimating models ranked best with very low frequency for R or R+M operating models and with very high frequency for R+S operating models (Figure \ref{aic_rank_M_fixed}).
+R estimating models were determined best with high frequency for R operating models, but also for R+M operating models.
+R+M estimating models were rarely determined best for any operating models including those where the process errors matched.
+
+AIC was conservative for determining whether the environmental covariate affected natural mortality.
+AIC was highly accurate in determining no effect when there was no effect in the operating model, but AIC ranked the null model best with high frequency even when there was an effect in the operating model in many cases.
+However the accuracy of AIC improved in certain operating models.
+Increased effect size, increased temporal contrast in the covariate, and lower uncertainty in all observations types lead to increased accuracy of determining covariate effects.
+
+Relative to the assumption that the mean natural mortality rate was known, estimating the mean natural mortality rate had small effects on the accuracy of AIC in selecting the appropriate process error and whether the covariate affect natural mortality (Figure \ref{aic_rank_M_estimated}).
+Where there were differences there were small decreases in accuracy for determining the appropriate process error and determining a covariate effect when there was one.
+
+<!--
 \begin{landscape}
@@
 \caption{Proportion of simulated data sets for each operating model where the estimation model with the lowest AIC had correct assumptions of process errors (R,R+S,R+M) and/or about whether the environmental covariate affected natural mortality.
+All estimating models assumed the mean mortality parameter $\beta_\text{M} = \log 0.2$.}\label{correct_assumptions_aic_M_fixed}
+\begin{center}
+\includegraphics[height = \textheight]{proportion_correct_PE_effect_M_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+-->
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets for each operating model where the estimation model type (process error and covariate effect assumptions) had the lowest AIC.
 All estimating models assumed the mean mortality parameter $\beta_\text{M} = \log 0.2$.}\label{aic_rank_M_fixed}
 \begin{center}
-\includegraphics[width = \textwidth]{proportion_correct_PE_effect_M_fixed.png}
+\includegraphics[height = \textheight]{proportion_best_AIC_M_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+<!--
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets for each operating model where the estimation model with the lowest AIC had correct assumptions of process errors (R,R+S,R+M) and/or about whether the environmental covariate affected natural mortality.
+All estimating models estimate the mean mortality parameter $\beta_\text{M}$.}\label{correct_assumptions_aic_M_estimated}
+\begin{center}
+\includegraphics[height = \textheight]{proportion_correct_PE_effect_M_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+-->
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets for each operating model where the estimation model type (process error and covariate effect assumptions) had the lowest AIC.
+All estimating models assumed the mean mortality parameter $\beta_\text{M} = \log 0.2$.}\label{aic_rank_M_estimated}
+\begin{center}
+\includegraphics[height = \textheight]{proportion_best_AIC_M_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+## Bias {-}
+
+### Environmental effect {-}
+
+When the EMs assumed the mean natural mortality rate was known, we observed generally accurate estimation of environmental effects across all EM and OM process error assumptions and all true covariate effect sizes, when ther was low uncertainty in environmental observations and larger temporal contrast in the simulated true environmental covariate (Figure \ref{Ecov_beta_bias_M_fixed}).
+We observed a negative trend in bias of the environmental effect with increased effect size when temporal variation in the covariate was lower and/or uncertainty in the covariate observations was higher.
+When the OMs had R+S process errors with low temporal variation in the true environmental covariate and lower uncertainty in the indices age age composition, estimated covariate effects were highly variable.
+In most cases the relative error of $\beta_\text{Ecov}$ did not depend on the source of process error assumed in the EM.
+When there was an effect of the EM process error assumption it was when OMs had R+S process errors.
+The worst bias was observed when OMs assumed R+S process errors, high uncertainty in covariate observations, low variability in the covariate, and low uncertainty in index and age composition observations.
+
+When the mean natural mortality rate was estimated in the EM, results were similar except estimated effects were even more variable for data simulated with R+S process errors (Figure \ref{Ecov_beta_bias_M_estimated}).
+There was also more separation of reliability of the estimation of the effect among EMs with different process error assumptions.
+The separation was most apparent when OMs simulated R+S process errors and larger variability in the environmental covariate wehere EMs with process errors other than R+S showing more bias than when the mean natural mortality rate was assumed known.
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of the estimated environmental effect on natural mortality $\beta_\text{Ecov}$ from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
+All estimating models assumed the mean mortality parameter $\beta_\text{M} = \log 0.2$.}\label{Ecov_beta_bias_M_fixed}
+\begin{center}
+\includegraphics[height = \textheight]{Ecov_beta_bias_all_PE_effect_M_fixed.png}
 \end{center}
@@
 \begin{figure}
-\caption{Proportion of simulated data sets for each operating model where the estimation model with the lowest AIC had correct assumptions of process errors (R,R+S,R+M) and/or about whether the environmental covariate affected natural mortality.
-All estimating models estimate the mean mortality parameter $\beta_\text{M}$.}\label{aic_rank_M_estimated}
+\caption{Median relative error of the estimated environmental effect on natural mortality $\beta_\text{Ecov}$ from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
+All estimating models assumed the mean mortality parameter $\beta_\text{M}$ is also estimated.}\label{Ecov_beta_bias_M_estimated}
 \begin{center}
-\includegraphics[width = \textwidth]{proportion_correct_PE_effect_M_estimated.png}
+\includegraphics[height = \textheight]{Ecov_beta_bias_all_PE_effect_M_estimated.png}
 \end{center}
@@
 
-## Bias {-}
+### Mean natural mortality rate {-}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of the estimated mean natural mortality parameter $\beta_\text{M}$ from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.}\label{mean_M_bias}
+\begin{center}
+\includegraphics[height = \textheight]{mean_M_bias_all_PE_effect.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+### Annual natural morality rate {-}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of annual natural mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
+Estimating models also assume the mean natural mortality parameter $\beta_\text{M} = \log 0.2$ and no environmental covariate effect $\beta_\text{Ecov} = 0$.}\label{annual_M_bias_M_fixed_beta_fixed}
+\begin{center}
+\includegraphics[height = \textheight]{annual_M_bias_all_PE_effect_M_fixed_beta_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of annual natural mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
+Estimating models also assume the mean natural mortality parameter $\beta_\text{M} = \log 0.2$ and the environmental covariate effect $\beta_\text{Ecov}$ is estimated.}\label{annual_M_bias_M_fixed_beta_estimated}
+\begin{center}
+\includegraphics[height = \textheight]{annual_M_bias_all_PE_effect_M_fixed_beta_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of annual natural mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
+Estimating models also assume both the mean natural mortality parameter $\beta_\text{M}$ and the environmental covariate effect $\beta_\text{Ecov}$ are estimated.}\label{annual_M_bias_M_estimated_beta_estimated}
+\begin{center}
+\includegraphics[height = \textheight]{annual_M_bias_all_PE_effect_M_estimated_beta_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+### Spawning stock biomass {-}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of annual SSB in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
+Estimating models also assume the mean natural mortality parameter $\beta_\text{M} = \log 0.2$ and no environmental covariate effect $\beta_\text{Ecov} = 0$.}\label{SSB_bias_M_fixed_beta_fixed}
+\begin{center}
+\includegraphics[height = \textheight]{SSB_bias_all_PE_effect_M_fixed_beta_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of annual SSB in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
+Estimating models also assume the mean natural mortality parameter $\beta_\text{M} = \log 0.2$ and the environmental covariate effect $\beta_\text{Ecov}$ is estimated.}\label{SSB_bias_M_fixed_beta_estimated}
+\begin{center}
+\includegraphics[height = \textheight]{SSB_bias_all_PE_effect_M_fixed_beta_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of annual SSB in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
+Estimating models also assume both the mean natural mortality parameter $\beta_\text{M}$ and the environmental covariate effect $\beta_\text{Ecov}$ are estimated.}\label{SSB_bias_M_estimated_beta_estimated}
+\begin{center}
+\includegraphics[height = \textheight]{SSB_bias_all_PE_effect_M_estimated_beta_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
 
@@
 This approximation may be appropriate for indices where we have a reliable estimate of uncertainty based on the survey design (), but there may be better approaches for the aggregate catch such as an informed prior on the standard errors with realistic bounds.
+
+We found EMs with R+M process errors were rarely determined as the appropriate model when OMs simulated R+M process errors.
+This was unexpected but is likely due to the size of the variance assumed for those process errors ($\sigma_M = 0.3$) relative to the variances assumed for index and age composition observations.
+This is related to the difficulty in separately estimating observation and process error variances when the ratio of process to observation uncertainty is low.
+
+Note that the results for bias of covariate effect all assume ecov beta is estimated.
+The lack of bias in certain situations might suggest including the effect even if AIC says it isn't better at least when contrast in Ecov is high.
+
+The bias in natural mortality estimation resulted in biased estimation of stock size and likely harvest rates, but it will also propagate into biological reference points and possibly stock status.
+
+# Conclusions {-}
+
+As we would expect, reliable detection of covariate effects requires informative data. 
+AIC preferred simpler models than the true model when information content in data and contrast in covariates and abundance were low.
+Null model for environmental covariate effect (no covariate effect) was selected when contrast in the time series was low  and/or uncertainty in observations was high.
+
+When there was process error in recruitment and M  (R+M), models with process error only in recruitment were preferred.
+This could be because the variation in the M random effects was small relative to observation variability.
+The marginal variance for log M random effects was 0.3.
+
+Null selection likely decreases with strength of the effect on M.
+We only examined two non-zero effect sizes: 0.25, and 0.5, but our results suggest larger effect sizes, with the same observation error and contrast in time series, would allow better AIC performance for determining covariate effects.
+
+Covariate effect estimation can be robust to process error assumptions (For R+M OMs) with contrast in covariate and low observation error.
+
+Reliable estimation of environmental effect on M regardless of the process error assumed by the EM as long as the contrast in the covariate is sufficient and the uncertainty in the observations is low.
 

### 2023-09-29 Ecov_study/mortality/paper/paper.tex
save edits to ecov m paper

@@
 %\onehalfspacing
-\doublespacing
+%\doublespacing
 \usepackage{lineno}
-\linenumbers
+%\linenumbers
 \usepackage[belowskip=0pt,aboveskip=0pt]{caption}
@@
 
-\hypertarget{abstract}{%
-\subsection*{Abstract}\label{abstract}}
-\addcontentsline{toc}{subsection}{Abstract}
+\hypertarget{summaryabstract}{%
+\subsection*{Summary/Abstract}\label{summaryabstract}}
+\addcontentsline{toc}{subsection}{Summary/Abstract}
 
-\hypertarget{keywords}{%
-\subsubsection*{Keywords}\label{keywords}}
-\addcontentsline{toc}{subsubsection}{Keywords}
+We completed a large scale simulation study with 288 operating models
+with 12 estimated models fit to each. The factors defining operating
+model configuration inlcuded the source of process error on the
+population (recruitment, recruitment and survival, recruitment and
+natural mortality), the degree of temporal variation and autocorrelation
+of the environmental covariate, the uncertainty in the observation of
+the covariate, the uncertainty in indices and age composition
+observations, fishing history, and the magnitude of the effect of the
+covariate on natural mortality.
+
+We found convergence of all estimation models was generally best when
+operating models assumed process errors in recruitment and survival,
+constant fishing rate, greater contrast in the true environmental
+covariate, and lower uncertainty in corresponding observations. Reliable
+convergence of all estimating models also occurred with the same process
+errors in the operating model and a step-change in fishing, but this
+also required lower uncertainty in index and age composition
+observations. Estimating models with process errors on recruitment and
+survival were unlikely to converge when the process errors in the
+operating model did not match whereas estimating models with process
+errors in recruitment and natural mortality converged for operating
+models without this match in certain cases. Probability of convergence
+generally decreased when the mean natural mortality rate parameter was
+estimated.
+
+AIC paragraph
+
+Bias paragraph
+
+These results suggest that models with covariate effects on M will be
+most reliable when R+S is true, large contrast in the Recommendations
+\#\#\# Keywords \{-\}
 
@@
 estimating models that make alternative assumptions for sources of
-process error (M, NAA), and whether (mean) M is estimated. We evaluate
-whether AIC can correctly determine the correct source of process error
-and stock effects on recruitment. We also evaluate the degree of bias in
-the outputs of the assessment model that are important for management.
+process error, and whether (mean) M is estimated. We evaluate effects of
+these factors on convergence of fitted models, whether AIC can correctly
+determine the correct source of process error and correct assumption
+about covariate effects on natural mortality, and the degree of bias in
+relevant parameters and outputs of the assessment model.
 
@@
 We used the WHAM package
-\citep[][commit 77bbd94]{millerstock20,stockmiller21} to configure all
-operating models and estimation models. This packages has also been used
-to configure operating and estimating models for closed loop simulations
-evaluating index-based assessment methods \citep{legaultetal23} and is
-used for management of haddock, butterfish, plaice, bluefish in the
-Northeast US.
+\citep[][version 1.0.7, commit 77bbd94]{millerstock20,stockmiller21} to
+configure all operating models and estimation models. This packages has
+also been used to configure operating and estimating models for closed
+loop simulations evaluating index-based assessment methods
+\citep{legaultetal23} and is used for management of GB haddock,
+butterfish, plaice, bluefish, and Atlantic cod stocks in the Northeast
+US.
 
@@
 of the characteristics of the operating models are the same as those
-used by \citet{milleretal_wp}.
+used by \citet{milleretal_project0_wp}. The factors defining the
+configuration of each operating model which are described in detail in
+subsequent sections include
 
-288 total operating models. We simulated 100 data sets for each
-operating model. For each simulated data set we fit a set of 12
-estimating models.
+\begin{itemize}
+\item two configurations of fishing histories, 
+\item two configurations of index and age composition observation error, 
+\item four configurations of latent environmental covariate process errors, 
+\item two configurations of environmental covariate observation uncertainty, 
+\item three levels of covariate effects on natural mortality, and 
+\item three configurations of sources of process errors for temporal changes in abundances at age. 
+\end{itemize}
+
+The full factorial of the levels of each factor results in 288 different
+operating models. We simulated 100 data sets for each operating model
+that included simulations of process errors.
+
+For each simulated data set we fit a set of 12 estimating models. The
+factors defining the configuration or each estimating model which are
+also described in detail below include
+
+\begin{itemize}
+\item three configurations of sources of process errors for temporal changes in abundances at age, 
+\item whether the (mean) natural mortality rate was estimated or assumed known at the true value, and
+\item whether the effect of the environmental covariate on natural mortality was estimated.
+\end{itemize}
+
+We did not use the bias-correction feature for process errors or
+observations described by \citet{stockmiller21} for operating and
+estimating models. Simulations were all carried out on the University of
+Massachusetts Green High-Performance Computing Cluster. Code for
+completing the simulations and summarization of results can be found at
+github.com/timjmiller/SSRTWG/ecov\_study/mortality.
 
@@
 We assumed a Beverton-Holt stock recruit function with constant
-pre-recruit mortality parameters for all operating models. All
-post-recruit productivity components are constant in the NAA and survey
-catchability process error operating models. Therefore steepness and
-unfished recruitment are also constant over the time period for those
-operating models \citep{millerbrooks21}. We specified unfished
-recruitment = \(R_0 = e^{10}\) and \(\Fmsy = \Fspr[40] = 0.348\) equated
-to a steepness of 0.69 and \(\alpha=0.60\) and
-\(\beta = 2.4 \times 10^{-5}\) for the \[
+pre-recruit mortality parameters (i.e., \(\alpha\) and \(\beta\)) for
+all operating models. All post-recruit productivity components are
+constant in the NAA and survey selectivity process error operating
+models. Therefore steepness and unfished recruitment are also constant
+over the time period for those operating models \citep{millerbrooks21}.
+We specified unfished recruitment = \(R_0 = e^{10}\) and
+\(\Fmsy = \Fspr[40] = 0.348\) equated to a steepness of 0.69 and
+\(\alpha=0.60\) and \(\beta = 2.4 \times 10^{-5}\) for the \[
 N_{1,y} = \frac{\alpha \text{SSB}_{y-1}}{1 + \beta \text{SSB}_{y-1}} 
@@
 beta parameters as other operating models this equates to a steepness
-and R0 at the mean of the time series process for M. For operating
+and \(R_0\) at the mean of the time series process for M. For operating
 models with time-varying random effects for fishery selectivity,
 \Fmsy is also not constant however we use the same F history as other
-operating models which corresponds to Fmsy at the mean selectivity
+operating models which corresponds to \Fmsy at the mean selectivity
 parameters.
@@
 covariate. The latent covariate is assumed to be Gaussian AR1 \[
-E_y|E_{y-1} \sim \text{N}\left(\mu_\text{Ecov}\left(1-\rho_\text{Ecov}\right) + \rho_\text{Ecov} E_{y-1}, \sigma^2_\text{Ecov}\right)
-\] with marginal mean \(\mu_\text{Ecov}\) and variance
-\(\frac{\sigma^2_\text{Ecov}}{1-\rho_\text{Ecov}^2}\). The configuration
-of the latent environmental covariate in the operating models have one
-of two alternate assumptions about the marginal standard deviation and
-autocorrelation parameter:
+E_y|E_{y-1} \sim \text{N}\left(\mu_\text{Ecov}\left(1-\rho_\text{Ecov}\right) + \rho_\text{Ecov} E_{y-1}, \left(1-\rho_\text{Ecov}^2\right)\sigma^2_\text{Ecov}\right)
+\] with marginal mean \(\mu_\text{Ecov}=0\) and variance
+\(\sigma^2_\text{Ecov}\). The configuration of the latent environmental
+covariate in the operating models have one of two alternate assumptions
+about the marginal standard deviation and autocorrelation parameter:
 
 \begin{itemize}
-\item marginal standard deviation $\in \{0.1, 0.5\}$
-\item AR1 correlation $\rho_\text{Ecov} \in \{0, 0.9\}$
+\item marginal standard deviation $\sigma_\text{Ecov} \in \{0.1, 0.5\}$
+\item AR1 correlation $\rho_\text{Ecov} \in \{0, 0.5\}$
 \end{itemize}
@@
 \[
-\log M_y = \beta_M + \beta_{\text{Ecov}} E_y + \epsilon_{M,y}
+\log M_y = \beta_M + \beta_{\text{Ecov}} E_y + \varepsilon_{M,y}
 \] that is a linear combination of a mean \(\beta_M\) and any annual
 covariate \(\beta_{\text{Ecov}}\) and random effects marginally
-distributed as
-\(\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)\).
+distributed as \(\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)\)
+\citep{stockmiller21}. For operating models with natural mortality
+random effects, we assume the same standard deviation
+\(\sigma_M = 0.3\).
 
@@
 
-The alternative configurations of random effects on recruitment, natural
-mortality, survival
+As described by \citet{stockmiller21}, the model for process errors on
+abundance at age in WHAM is \begin{equation}
+  \text{log}N_{a,y}=\left\{
+    \begin{array}{@{}lll@{}}
+      \text{log} \left( f(SSB_{y-1}) \right) + \varepsilon_{1,y}, & \text{if}\ a = 1 \\
+      \text{log} \left( N_{a-1,y-1} \right) - Z_{a-1,y-1} + \varepsilon_{a,y}, & \text{if}\ 1 < a < A \\
+      \text{log} \left( N_{A-1,y-1} e^{-Z_{A-1,y-1}} + N_{A,y-1} e^{-Z_{A,y-1}} \right) + \varepsilon_{A,y}, & \text{if}\ a = A
+    \end{array}\right.
+\end{equation} We assume the the standard deviation of recruitment
+deviations is
+\(\text{SD}\left(\varepsilon_{1,y}\right) = \sigma_R = 0.5\) for all
+operating models. For operating models with survival devations for older
+ages, we assume the same standard deviations for all older ages
+\(\text{SD}\left(\varepsilon_{a,y}\right) = \sigma_{2+} = 0.3\).
+
+The alternative configurations of random effects on recruitment,
+survival, and natural mortality are
 
 \begin{itemize}
-\item uncorrelated random effects on R and survival with $\sigma_R = 0.5$, marginal standard deviation of survival of older age clases of 0.3
-\item uncorrelated random effects on R and survival with $\sigma_R = 0.5$, marginal standard deviation of natural mortality of 0.3
+\item uncorrelated random effects on recruitment only ($\sigma_R = 0.5$) (R),
+\item uncorrelated random effects on recruitment ($\sigma_R = 0.5$) and survival of older age classes ($\sigma_{2+} = 0.3$) (R+S), and
+\item uncorrelated random effects on R ($\sigma_R = 0.5$) and log-natural mortality ($\sigma_M = 0.3$) (R+M).
 \end{itemize}
@@
 with logistic selectivity for the fleet with \(a_{50} = 5\) and slope =
-1 (Figure \ref{om_mean_selectivity}). This selectivity is was used to
-define \Fmsy for the Beverton-Holt stock recruitment parameters above.
-We assumed a logistic-normal distribution for the age-composition
-observations for the fleet.
+1 (Figure \ref{om_mean_selectivity}). This selectivity was used to
+define \Fmsy, \Fspr[40], and steepness for the Beverton-Holt stock
+recruitment parameters above. We assumed a logistic-normal distribution
+for the age-composition observations associated with the fleet where
+errors in the multivariate normal transformation are independent. The
+standard deviation parameter is also constant across ages.
 
@@
 spring (0.25 way through the year) and one in the fall (0.75 way through
-the year). Actually we have it currently configured that both occur 0.5
-way through the year. Catchability of both surveys are assumed to be
-0.1. Like the fishing fleet, we assumed logistic selectivity for both
-indices with \(a_{50} = 5\) and slope = 1 and a logistic-normal
-distribution for the age-composition observations.
+the year). Catchability of both surveys are assumed to be 0.1. We
+assumed the same selectivity and age composition distribution as the
+fleet for both indices.
 
@@
 was 0.1 and the standard deviation of the logistic-normal for age
-composition observations was 0.3 In the high uncertainty specification
+composition observations was 0.3. In the high uncertainty specification
 the standard deviation for log-aggregate indices was 0.4 and that for
@@
 estimated.
-
-The factorial combinations of fishing histories, index and age
-composition observation error assumptions, latent environmental
-covariate and observation assumptions, covariate effects on natural
-mortality, and sources of process error (R, R+S, R+M) results in 288
-different operating models.
 
@@
 The configuration of the process errors in the estimating models matched
-the corresponding options in the operating models. For example, if R+S
-was assumed for both the estimating and operating model, the assumptions
-about process errors matched. The environmental covariate observations
-were included in all estimation models to ensure comparability of
-information criteria. The observation error variance of the
-environmental observations and aggregate catch and indices were all
-assumed known at the true values.
-
-Simulations were all carried out on the University of Massachusetts
-Green High-Performance Computing Cluster. Code for completing the
-simulations and summarization of results can be found at
-github.com/timjmiller/SSRTWG/ecov\_study/mortality. We used the wham
-package version 1.X.X (commit 77bbd94).
+the corresponding options in the operating models. For example, if
+uncorrelated R+S was assumed for both the estimating and operating
+model, the assumptions about process errors matched. The environmental
+covariate observations were included in all estimation models to ensure
+comparability of information criteria. All fixed effects parameters for
+selectivity, catchability, fully-selected fishing mortality, mean
+recruitment, initial abundance at age, and variances for logistic-normal
+age composition distributions were estimated. Any process error variance
+parameters for recruitment, survival, and natural mortality were also
+estimated. The observation error variance of the environmental
+observations and aggregate catch and indices were all assumed known at
+the true values.
 
@@
 
+\hypertarget{convergence}{%
+\subsubsection*{Convergence}\label{convergence}}
+\addcontentsline{toc}{subsubsection}{Convergence}
+
 The first measure of reliability we investigated was frequency of
@@
 sets. There are various ways to assess convergence of the fit. We
-summarized 5 alternative categories of convergence. 1. Did the
-optimization routine (stats::nlminb) complete without error? 2. Did the
-stats::nlminb convergence flag = 0 indicate successful convergence? 3.
-Was the maximum absolute value of the gradient of the log-likelihood
-\textless{} \(1\times10^{-6}\)? 4. Did TMB::sdreport provide non-NA
-values for all fixed effects standard errors? 5. Did TMB::sdreport
-provide all standard errors \textless{} 100?
+summarized 5 alternative categories of convergence:
+
+\begin{enumerate}
+\item Did the optimization routine (stats::nlminb) complete without error?
+\item Did the stats::nlminb convergence flag = 0 indicate successful convergence?
+\item Was the maximum absolute value of the gradient of the log-likelihood < $1\times10^{-6}$?
+\item Did TMB::sdreport provide non-NA values for all fixed effects standard errors?
+\item Did TMB::sdreport provide all standard errors < 100?
+\end{enumerate}
 
@@
 We measured the frequency of correct model selection using marginal AIC.
-For a given operating model we the set of models that were considered
-all made the same assumptions on whether or not to estimate M and
-whether a stock-recruit function is assumed.
+For a given operating model the set of models that were considered all
+made the same assumptions on whether or not to estimate (mean) natural
+mortality rate or it is assumed at the true value
+\(\beta_M = log(0.2)\). For model \(m\), the marginal AIC is a function
+of the marginal log-likelihood maximized with respect to the fixed
+effects in the model \(\boldsymbol{\theta}\) and the number of fixed
+effects \(n\left(\boldsymbol{\theta}\right)\) estimated, \[
+\text{AIC}_m = -2\left[{\text{argmax}}_{\boldsymbol{\theta}} \log L_m\left({\boldsymbol{\theta}}\right) - n\left({\boldsymbol{\theta}}\right)\right].
+\]
 
@@
 
-We calculated relative errors of various quantities for each fitted
-estimation model for the \(i\)th simulated data set as \[
+We calculated median relative errors of
+
+\begin{itemize}
+\item $\beta_\text{Ecov}$, the effect of environmental covariates on natural mortality, 
+\item $\beta_M$, the mean log-natural mortality rate,
+\item annual natural mortality rate,
+\item annual spawning stock biomass, and
+\item annual fully-selected fishing mortality rate.
+\end{itemize}
+
+For the \(i\)th simulated data the relative error for a parameter
+\(\theta\) provided from the fitted estimation model is \[
 {\text{RE}_i}\left(\theta\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
@@
 confidence intervals using the binomial distribution approach as in
-\citet{stockmiller21}.
+\citet{stockmiller21}. If the confidence interval contains zero we do
+not have evidence of bias from the simulation study. We used estimates
+from all simulations that satisfied the first type of convergence
+described above. We were not concerned about more restrictive types of
+convergence because some estimation models would not be expected to
+satisfy these criteria because of mis-specified structures. For example,
+an estimation model that includes survival random effects fitted to a
+set of observations simulated without survival random effects should
+have trouble estimating variance of the survival random effects, but the
+estimates of the parameters and derived output described above, might be
+reliable.
 
@@
 
+We focused on the fourth convergence criteria, that the hessian of the
+negative-marginal log-likelihood at the optimized parameters was
+invertible. When the (mean) natural mortality rate was assumed at the
+true value, corresponding to usual practice in application of assessment
+models, good convergence for all EMs was observed for R+S operating
+models with \Fmsy fishing history, more variation in the latent
+environmental covariate (\(\sigma_\text{Ecov} = 0.5\)), and lower error
+in the associated observations \(\sigma_\text{ecov} = 0.1\) (Figure
+\ref{convergence_M_fixed}). Good convergence for all EMs was also
+observed for R+S OMs with the step change in fishing history, but with
+low error in indices and age comp observations.
+
+EMs that assumed no survival or natural mortality random effects (R
+OMS), nearly always converged across all operating models that assumed
+more variation in the latent environmental covariate
+(\(\sigma_\text{Ecov} = 0.5\)) and lower error in the associated
+observations \(\sigma_\text{ecov} = 0.1\). EMs that assumed recruitment
+and survival random effects (R+S OMS), had poor converge probability
+when the OMs had alternative process errors (R or R+M). The R+S EMs
+showed best convergence for the R+S OMs where there was lower error in
+the environmental observations \(\sigma_\text{ecov} = 0.1\) or higher
+error in environmental observations of latent true environmental
+covariates with greater temporal variation. EMs that assumed recruitment
+and natural mortality random effects (R+M OMs), had poor convergence
+probability for all OMs that had a step change in fishing history and
+higher uncertainty in indices and age composition.
+
+When the OMs and EMs both assume process errors only on recruitment (R)
+or on recruitment and survival (R+S), convergence was worst with less
+variation in the true environmental covariate and larger uncertainty in
+associated observations. When the OMs and EMs both assume process errors
+on recruitment and natural mortality (R+M) convergence was problematic
+for all OMs with the step-change in fishing history. The best
+convergence was observed with this match between OMs and EMs was when
+the fishing history was constant and there was low uncertainty in
+environmental observations.
+
 \begin{landscape}
@@
 \end{landscape}
+
+As might be expected, there was an overall drop in the probability of
+convergence when the mean natural mortality rate was estimated rather
+than assumed at the true value (Figure \ref{convergence_M_estimated}).
+Otherwise, general trends described above with mean natural mortality
+fixed, apply when estimated.
 
@@
 
+When estimating models assumed the mean natural mortality rate was
+known, the best accuracy of AIC for model selection occurred for models
+with R+S process errors. R+S estimating models ranked best with very low
+frequency for R or R+M operating models and with very high frequency for
+R+S operating models (Figure \ref{aic_rank_M_fixed}). R estimating
+models were determined best with high frequency for R operating models,
+but also for R+M operating models. R+M estimating models were rarely
+determined best for any operating models including those where the
+process errors matched.
+
+AIC was conservative for determining whether the environmental covariate
+affected natural mortality. AIC was highly accurate in determining no
+effect when there was no effect in the operating model, but AIC ranked
+the null model best with high frequency even when there was an effect in
+the operating model in many cases. However the accuracy of AIC improved
+in certain operating models. Increased effect size, increased temporal
+contrast in the covariate, and lower uncertainty in all observations
+types lead to increased accuracy of determining covariate effects.
+
+Relative to the assumption that the mean natural mortality rate was
+known, estimating the mean natural mortality rate had small effects on
+the accuracy of AIC in selecting the appropriate process error and
+whether the covariate affect natural mortality (Figure
+\ref{aic_rank_M_estimated}). Where there were differences there were
+small decreases in accuracy for determining the appropriate process
+error and determining a covariate effect when there was one.
+
 \begin{landscape}
 \begin{figure}
-\caption{Proportion of simulated data sets for each operating model where the estimation model with the lowest AIC had correct assumptions of process errors (R,R+S,R+M) and/or about whether the environmental covariate affected natural mortality.
+\caption{Proportion of simulated data sets for each operating model where the estimation model type (process error and covariate effect assumptions) had the lowest AIC.
 All estimating models assumed the mean mortality parameter $\beta_\text{M} = \log 0.2$.}\label{aic_rank_M_fixed}
 \begin{center}
-\includegraphics[width = \textwidth]{proportion_correct_PE_effect_M_fixed.png}
+\includegraphics[height = \textheight]{proportion_best_AIC_M_fixed.png}
 \end{center}
@@
 \begin{figure}
-\caption{Proportion of simulated data sets for each operating model where the estimation model with the lowest AIC had correct assumptions of process errors (R,R+S,R+M) and/or about whether the environmental covariate affected natural mortality.
-All estimating models estimate the mean mortality parameter $\beta_\text{M}$.}\label{aic_rank_M_estimated}
+\caption{Proportion of simulated data sets for each operating model where the estimation model type (process error and covariate effect assumptions) had the lowest AIC.
+All estimating models assumed the mean mortality parameter $\beta_\text{M} = \log 0.2$.}\label{aic_rank_M_estimated}
 \begin{center}
-\includegraphics[width = \textwidth]{proportion_correct_PE_effect_M_estimated.png}
+\includegraphics[height = \textheight]{proportion_best_AIC_M_estimated.png}
 \end{center}
@@
 \addcontentsline{toc}{subsection}{Bias}
+
+\hypertarget{environmental-effect}{%
+\subsubsection*{Environmental effect}\label{environmental-effect}}
+\addcontentsline{toc}{subsubsection}{Environmental effect}
+
+When the EMs assumed the mean natural mortality rate was known, we
+observed generally accurate estimation of environmental effects across
+all EM and OM process error assumptions and all true covariate effect
+sizes, when ther was low uncertainty in environmental observations and
+larger temporal contrast in the simulated true environmental covariate
+(Figure \ref{Ecov_beta_bias_M_fixed}). We observed a negative trend in
+bias of the environmental effect with increased effect size when
+temporal variation in the covariate was lower and/or uncertainty in the
+covariate observations was higher. When the OMs had R+S process errors
+with low temporal variation in the true environmental covariate and
+lower uncertainty in the indices age age composition, estimated
+covariate effects were highly variable. In most cases the relative error
+of \(\beta_\text{Ecov}\) did not depend on the source of process error
+assumed in the EM. When there was an effect of the EM process error
+assumption it was when OMs had R+S process errors. The worst bias was
+observed when OMs assumed R+S process errors, high uncertainty in
+covariate observations, low variability in the covariate, and low
+uncertainty in index and age composition observations.
+
+When the mean natural mortality rate was estimated in the EM, results
+were similar except estimated effects were even more variable for data
+simulated with R+S process errors (Figure
+\ref{Ecov_beta_bias_M_estimated}). There was also more separation of
+reliability of the estimation of the effect among EMs with different
+process error assumptions. The separation was most apparent when OMs
+simulated R+S process errors and larger variability in the environmental
+covariate wehere EMs with process errors other than R+S showing more
+bias than when the mean natural mortality rate was assumed known.
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of the estimated environmental effect on natural mortality $\beta_\text{Ecov}$ from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
+All estimating models assumed the mean mortality parameter $\beta_\text{M} = \log 0.2$.}\label{Ecov_beta_bias_M_fixed}
+\begin{center}
+\includegraphics[height = \textheight]{Ecov_beta_bias_all_PE_effect_M_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of the estimated environmental effect on natural mortality $\beta_\text{Ecov}$ from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
+All estimating models assumed the mean mortality parameter $\beta_\text{M}$ is also estimated.}\label{Ecov_beta_bias_M_estimated}
+\begin{center}
+\includegraphics[height = \textheight]{Ecov_beta_bias_all_PE_effect_M_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\hypertarget{mean-natural-mortality-rate}{%
+\subsubsection*{Mean natural mortality
+rate}\label{mean-natural-mortality-rate}}
+\addcontentsline{toc}{subsubsection}{Mean natural mortality rate}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of the estimated mean natural mortality parameter $\beta_\text{M}$ from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.}\label{mean_M_bias}
+\begin{center}
+\includegraphics[height = \textheight]{mean_M_bias_all_PE_effect.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\hypertarget{annual-natural-morality-rate}{%
+\subsubsection*{Annual natural morality
+rate}\label{annual-natural-morality-rate}}
+\addcontentsline{toc}{subsubsection}{Annual natural morality rate}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of annual natural mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
+Estimating models also assume the mean natural mortality parameter $\beta_\text{M} = \log 0.2$ and no environmental covariate effect $\beta_\text{Ecov} = 0$.}\label{annual_M_bias_M_fixed_beta_fixed}
+\begin{center}
+\includegraphics[height = \textheight]{annual_M_bias_all_PE_effect_M_fixed_beta_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of annual natural mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
+Estimating models also assume the mean natural mortality parameter $\beta_\text{M} = \log 0.2$ and the environmental covariate effect $\beta_\text{Ecov}$ is estimated.}\label{annual_M_bias_M_fixed_beta_estimated}
+\begin{center}
+\includegraphics[height = \textheight]{annual_M_bias_all_PE_effect_M_fixed_beta_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of annual natural mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
+Estimating models also assume both the mean natural mortality parameter $\beta_\text{M}$ and the environmental covariate effect $\beta_\text{Ecov}$ are estimated.}\label{annual_M_bias_M_estimated_beta_estimated}
+\begin{center}
+\includegraphics[height = \textheight]{annual_M_bias_all_PE_effect_M_estimated_beta_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\hypertarget{spawning-stock-biomass}{%
+\subsubsection*{Spawning stock biomass}\label{spawning-stock-biomass}}
+\addcontentsline{toc}{subsubsection}{Spawning stock biomass}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of annual SSB in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
+Estimating models also assume the mean natural mortality parameter $\beta_\text{M} = \log 0.2$ and no environmental covariate effect $\beta_\text{Ecov} = 0$.}\label{SSB_bias_M_fixed_beta_fixed}
+\begin{center}
+\includegraphics[height = \textheight]{SSB_bias_all_PE_effect_M_fixed_beta_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of annual SSB in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
+Estimating models also assume the mean natural mortality parameter $\beta_\text{M} = \log 0.2$ and the environmental covariate effect $\beta_\text{Ecov}$ is estimated.}\label{SSB_bias_M_fixed_beta_estimated}
+\begin{center}
+\includegraphics[height = \textheight]{SSB_bias_all_PE_effect_M_fixed_beta_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of annual SSB in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
+Estimating models also assume both the mean natural mortality parameter $\beta_\text{M}$ and the environmental covariate effect $\beta_\text{Ecov}$ are estimated.}\label{SSB_bias_M_estimated_beta_estimated}
+\begin{center}
+\includegraphics[height = \textheight]{SSB_bias_all_PE_effect_M_estimated_beta_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
 
@@
 bounds.
+
+We found EMs with R+M process errors were rarely determined as the
+appropriate model when OMs simulated R+M process errors. This was
+unexpected but is likely due to the size of the variance assumed for
+those process errors (\(\sigma_M = 0.3\)) relative to the variances
+assumed for index and age composition observations. This is related to
+the difficulty in separately estimating observation and process error
+variances when the ratio of process to observation uncertainty is low.
+
+Note that the results for bias of covariate effect all assume ecov beta
+is estimated. The lack of bias in certain situations might suggest
+including the effect even if AIC says it isn't better at least when
+contrast in Ecov is high.
+
+The bias in natural mortality estimation resulted in biased estimation
+of stock size and likely harvest rates, but it will also propagate into
+biological reference points and possibly stock status.
+
+\hypertarget{conclusions}{%
+\section*{Conclusions}\label{conclusions}}
+\addcontentsline{toc}{section}{Conclusions}
+
+As we would expect, reliable detection of covariate effects requires
+informative data. AIC preferred simpler models than the true model when
+information content in data and contrast in covariates and abundance
+were low. Null model for environmental covariate effect (no covariate
+effect) was selected when contrast in the time series was low and/or
+uncertainty in observations was high.
+
+When there was process error in recruitment and M (R+M), models with
+process error only in recruitment were preferred. This could be because
+the variation in the M random effects was small relative to observation
+variability. The marginal variance for log M random effects was 0.3.
+
+Null selection likely decreases with strength of the effect on M. We
+only examined two non-zero effect sizes: 0.25, and 0.5, but our results
+suggest larger effect sizes, with the same observation error and
+contrast in time series, would allow better AIC performance for
+determining covariate effects.
+
+Covariate effect estimation can be robust to process error assumptions
+(For R+M OMs) with contrast in covariate and low observation error.
+
+Reliable estimation of environmental effect on M regardless of the
+process error assumed by the EM as long as the contrast in the covariate
+is sufficient and the uncertainty in the observations is low.
 

### 2023-09-29 Ecov_study/mortality/code/F_bias_results.R

@@ -1,0 +1,307 @@
+library(here)
+library(ggplot2)
+library(dplyr)
+library(tidyr)
+df.ems = readRDS(file.path(here(),"Ecov_study","mortality","inputs", "df.ems.RDS"))
+df.oms = readRDS(file.path(here(),"Ecov_study","mortality","inputs", "df.oms.RDS"))
+F_bias <- readRDS(file.path(here::here(),"Ecov_study","mortality", "results", "F_results.RDS"))
+library(reshape2)
+#res <- melt(F_bias)
+
+custom_boxplot_stat <- function(x){#, n.yrs = 1, n.sim = 100) {
+  x <- x[which(!is.na(x))]
+  n <- length(x)
+  bnds95 <- qbinom(c(0.025,0.975), n, 0.5)/n # 95% CI bounds for median (sims x years)
+  bnds80 <- qbinom(c(0.1,0.9), n, 0.5)/n # 80% CI bounds for median (sims x years)
+  r <- quantile(x, probs = c(bnds95[1], bnds80[1], 0.5, bnds80[2], bnds95[2]))
+  names(r) <- c("ymin", "lower", "middle", "upper", "ymax")
+  r
+}
+plot_df_fn <- function(df.ems, df.oms, M_est=FALSE) {
+
+  print(which(df.oms$NAA_M_re == "rec+1"))
+  for(i in 1:3) {
+    re_mod <- c("rec", "rec+1", "rec+M")[i]
+    #EM:  M fixed, mean_M estimated, OM and EM RE assumption match
+    em_ind <- which(df.ems$Ecov_est & df.ems$M_est == M_est & df.ems$re_config == re_mod)
+#    print(em_ind)
+    om_ind <- which(df.oms$NAA_M_re == re_mod) #om and em match
+#    print(om_ind)
+    res <- lapply(om_ind, function(x) {
+      F_res <- t(sapply(1:40, function(y) {
+        #print(om_ind)
+        r_e <- F_bias[[x]][,em_ind,y,2]/F_bias[[x]][,em_ind,y,1]-1
+        out <- c(median(r_e, na.rm = TRUE), sd(r_e, na.rm=T)/sqrt(sum(!is.na(r_e))), custom_boxplot_stat(r_e))#quantile(x, probs = c(0.025,0.975)))
+        return(out)
+      }))
+      colnames(F_res) <- c("bias_est", "bias_se", "ymin", "lower", "middle", "upper", "ymax")
+      return(F_res)
+    })
[…]

### 2023-09-29 Ecov_study/mortality/code/aic_ecov_effect_results.R

@@ -344,6 +344,14 @@ make_df_fn <- function(M_est = TRUE){
         "0" = "rho[Ecov] == 0",
         "0.5" = "rho[Ecov] == 0.5"
     ))
+  all_res_mod <- all_res_mod %>% mutate(Fhist = recode(Fhist,
+      "H-MSY" = "High->FMSY",
+      "MSY" = "FMSY"))
+  all_res_mod <- all_res_mod %>%
+    mutate(beta_ecov_fixed = recode(Ecov_est,
+      "TRUE" = "beta[Ecov]==0",
+      "FALSE" = "beta[Ecov]~estimated"
+    ))
 
   #1: correct effect assumption and RE will be 1, 
   #2: correct RE, wrong effect assumption
@@ -362,6 +370,13 @@ make_df_fn <- function(M_est = TRUE){
   temp <- all_res_mod 
   all_res_mod$Ecov_effect <- factor(all_res_mod$Ecov_effect)
   all_res_mod <- all_res_mod %>%
+    mutate(NAA_M_re = recode(NAA_M_re,
+      "R" = "OM: R",
+      "R+S" = "OM: R+S",
+      "R+M" = "OM: R+M"
+    ))
+
+  all_res_mod <- all_res_mod %>%
     mutate(correct = recode(correct,
       "1" = "Correct Effect, Correct PE",
       "2" = "Wrong Effect, Correct PE",
@@ -378,11 +393,26 @@ plt <- ggplot(all_res_mod, aes(x = Ecov_effect, y = n, fill = correct)) + scale_
     facet_grid(Ecov_obs_sig+Ecov_re_sig+Ecov_re_cor ~ NAA_M_re+Fhist+obs_error, 
       labeller = labeller(Ecov_re_sig = label_parsed, Ecov_re_cor = label_parsed, obs_error = label_wrap_gen(width = 15), Ecov_obs_sig = label_parsed)) +
     #facet_grid(Ecov_obs_sig+Ecov_re_sig+Ecov_re_cor ~ NAA_M_re+Fhist+obs_error, labeller = label_wrap_gen(width = 15)) + #, labeller = label_parsed) + 
-    theme_bw() + coord_cartesian(ylim = c(0, 1)) + ylab("Proportion ranked best") + xlab(expression(beta[Ecov])) + labs(fill = "EM process error") +
[…]

### 2023-09-29 Ecov_study/mortality/code/explore_results.R

@@ -1,144 +1,0 @@
-library(here)
-library(ggplot2)
-df.ems = readRDS(file.path(here(),"Ecov_study","mortality","inputs", "df.ems.RDS"))
-df.oms = readRDS(file.path(here(),"Ecov_study","mortality","inputs", "df.oms.RDS"))
-
-#x <- readRDS(file.path(here::here(),"Ecov_study","mortality", "results", paste0("om", 1), paste0("sim", 3, "_em", 4, ".RDS")))
-#x <- readRDS(file.path(here::here(),"Ecov_study","mortality", "results", "aggregated_results", paste0("om_", 1, "_em_", 4, ".RDS")))
-conv_res = lapply(1:NROW(df.oms), function(y){
-#conv_res = lapply(1:1, function(y){
-  lapply(1:NROW(df.ems), function(x) {
-  #lapply(4:4, function(x) {
-    res <- readRDS(file.path(here::here(),"Ecov_study","mortality", "results", "aggregated_results", paste0("om_", y, "_em_", x, ".RDS")))
-    print(paste(y, x))
-    conv_res = t(sapply(res, function(z){
-      out <- rep(NA,5) #convergence type  1, and 2
-      if(!is.character(z)) {
-#        print(names(z))
-        if(!is.null(z$fit$opt)) {
-          out[1] <- 1
-          out[2] <- z$fit$opt$conv
-        }
-        if(!is.null(z$fit$sdrep)) out[3] <- as.integer(sum(sapply(z$fit$sdrep$SE_par, function(g) any(is.nan(g)))))
-        if(!is.null(z$fit$final_gradient)) out[4] <- max(abs(z$fit$final_gradient))
-        if(!is.null(z$fit$sdrep)) out[5] <- max(sapply(z$fit$sdrep$SE_par, function(g) max(g,na.rm=TRUE)))
-      }
-      return(out)
-    }))
-    return(conv_res)
-  })
-})
-saveRDS(conv_res, file = file.path(here(),"Ecov_study","mortality", "results", "convergence_results.RDS"))
-
-aic_res = lapply(1:NROW(df.oms), function(y){
-  lapply(1:NROW(df.ems), function(x) {
-  #lapply(4:4, function(x) {
-    res <- readRDS(file.path(here::here(),"Ecov_study","mortality", "results", "aggregated_results", paste0("om_", y, "_em_", x, ".RDS")))
-    print(paste(y, x))
[…]

### 2023-09-29 Project_0/code/relbias_plots.R

@@ -193,7 +193,7 @@ make_plot_df <- function(type = "naa", res = all_naa_om_relssb, is_SE = FALSE) {
       "2" = "cv",
       "3" = "in_ci",
     )) %>% as.data.frame   
-  df <- res %>% pivot_wider(names_from = column, values_from = value) %>% as.data.frame
+  df <- res %>% tidyr::pivot_wider(names_from = column, values_from = value) %>% as.data.frame
   print(dim(df))
   if(is_SE) df <- filter(df, !is.na(cv))
   print(dim(df))

### 2023-09-29 Ecov_study/recruitment_functions/paper/main.tex

@@
 
+%%--ANALYSIS--%%
+\subsubsection*{Analysis}
+
+Metrics of interest: 
+\begin{enumerate}
+    \item Ability to correctly identify a stock recruitment relationship
+    \item Ability correctly identify the functional form of the stock recruitment relationship
+    \item Differences in estimation error (SSB, R) when SR relationships are correctly/incorrectly identified
+\end{enumerate}
+
+Generalized linear models
+
+Generalized linear mixed models
+
+Classification trees
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section*{Results}
+
+\begin{figure}[htb]
+\begin{center}
+\includegraphics[width = 6in]{dRdX_1RdRdx.pdf}
+\caption{Sensitivity of recruitment to environmental covariates for the three functional forms considered in this study.}
+\end{center}
+\end{figure}
+
+\begin{figure}[htb]
+\begin{center}
+\includegraphics[width = 5in]{raw_boxplots_SR.pdf}
+\caption{Model selection results for whether a stock recruitment relationship was correctly identified, expressed as marginal proportions against individual covariates.}
+\end{center}
+\end{figure}
+
+\begin{figure}[htb]
+\begin{center}
+\includegraphics[width = 5in]{raw_boxplots_form.pdf}
+\caption{Model selection results for whether the correct stock recruitment functional form was identified, expressed as marginal proportions against individual covariates.}
+\end{center}
+\end{figure}
+
+\begin{figure}[htb]
+\begin{center}
+\includegraphics[width = 6in]{effect_size_glm_glmm.pdf}
+\caption{Effect sizes for predicting estimated via a binomial generalized linear model and a binomial generalized linear mixed model with simulation random number seed as a random intercept.}
+\end{center}
+\end{figure}
+
+\begin{figure}[htb]
+\begin{center}
+\includegraphics[width = 5in]{reg_tree.pdf}
+\caption{Classification tree analysis of model selection results.
+R package `rpart' was used for the model fit and visualization.
+Following default settings in rpart, the complexity parameter is set at $c_p=0.01$ which represents the minimum classification improvement required for any split.}
+\end{center}
+\end{figure}
 

### 2023-10-04 Ecov_study/mortality/paper/paper.Rmd
saving work on M-ecov working paper

@@
 
-\begin{figure}
-\caption{The proportion mature at age assumed for the population in all operating and estimating models.}\label{om_maturity}
-\begin{center}
-\includegraphics[width = \textwidth]{om_maturity.png}
-\end{center}
-\end{figure}
-
 Weight at age was generated with a LVB growth function
@@
 where $\theta_1 = e^{-12.1}$ and $\theta_2 = 3.2$ (Figure \ref{om_waa}).
-
-\begin{figure}
-\caption{The weight at age assumed for the population in all operating and estimating models.}\label{om_waa}
-\begin{center}
-\includegraphics[width = \textwidth]{om_waa.png}
-\end{center}
-\end{figure}
 
@@
 For OMs with process errors on natural mortality the mean rate was 0.2.
-
-\begin{figure}
-\caption{The Beverton-Holt stock recruit relationship assume for all operating models.}\label{om_sr}
-\begin{center}
-\includegraphics[width = \textwidth]{om_sr.png}
-\end{center}
-\end{figure}
 
@@
 
-\begin{figure}
-\caption{Example simulations of environmental covariate latent processes and observations with different levels of observation error, and different assumptions about variability of the latent process.}\label{om_ecov_example}
-\begin{center}
-\includegraphics[width = \textwidth]{Ecov_true_obs_example.png}
-\end{center}
-\end{figure}
-
 ### Random effects on recruitment, survival, and natural mortality {-}
@@
 \end{itemize}
+The parameters defining the simulated covariate time series, size of the covariate effect, and any natural mortality random effects result in a range of different levels of variation in annual natural mortality rates (Figure \ref{M_example}).
 
@@
 The standard deviation parameter is also constant across ages.
-
-\begin{figure}
-\caption{The selectivity at age assumed for the fishing fleet in operating models without selectivity random effects.
-The mean selectivity at age across time for the fishing fleet in operating models with selectivity random effects.}\label{om_mean_selectivity}
-\begin{center}
-\includegraphics[width = \textwidth]{om_mean_selectivity.png}
-\end{center}
-\end{figure}
 
@@
 \end{itemize}
-For the $i$th simulated data the relative error for a parameter $\theta$ provided from the fitted estimation model is
+Results for fishing mortality rate are provided in the Supplementary Materials. For the $i$th simulated data the relative error for a parameter $\theta$ provided from the fitted estimation model is
 $$
@@
 For example, an estimation model that includes survival random effects fitted to a set of observations simulated without survival random effects should have trouble estimating variance of the survival random effects, but the estimates of the parameters and derived output described above, might be reliable.
+
+For natural mortality rate, SSB, and fully-selected fishing mortality rate, we summarized results for each of the annual values, but we present results only for estimates from the first year (start), year 21 (middel), and the last year (end), because there were no appreciable differences between the results for these three years and those for the other years.
 
@@
 
-\begin{landscape}
-\begin{figure}
-\caption{Probability of convergence of estimating models with alternative process error assumptions fit to simulated populations and data sets from operating models.
-The convergence criterion is an invertible hessian providing standard error estimates for all fixed effects parameters.
-Vertical lines represent 95\% confidence intervals.
-All estimating models estimate environmental covariate effects on natural mortality but assume the mean mortality parameter $\beta_\text{M} = \log 0.2$.}\label{convergence_M_fixed}
-\begin{center}
-\includegraphics[height = \textheight]{proportion_good_hessian_ecov_effect_est_M_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
 As might be expected, there was an overall drop in the probability of convergence when the mean natural mortality rate was estimated rather than assumed at the true value (Figure \ref{convergence_M_estimated}).
 Otherwise, general trends described above with mean natural mortality fixed, apply when estimated.
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of convergence of estimating models with alternative process error assumptions fit to simulated populations and data sets from operating models.
-The convergence criterion is an invertible hessian providing standard error estimates for all fixed effects parameters.
-Vertical lines represent 95\% confidence intervals.
-All estimating models estimate environmental covariate effects on natural mortality and the mean mortality parameter $\beta_\text{M}$.}\label{convergence_M_estimated}
-\begin{center}
-\includegraphics[height = \textheight]{proportion_good_hessian_ecov_effect_est_M_est.png}
-\end{center}
-\end{figure}
-\end{landscape}
 
@@
 Where there were differences there were small decreases in accuracy for determining the appropriate process error and determining a covariate effect when there was one.
+
+## Bias {-}
+
+### Environmental effect {-}
+
+When the EMs assumed the mean natural mortality rate was known, we observed generally accurate estimation of environmental effects across all EM and OM process error assumptions and all true covariate effect sizes, when ther was low uncertainty in environmental observations and larger temporal contrast in the simulated true environmental covariate (Figure \ref{Ecov_beta_bias_M_fixed}).
+We observed a negative trend in bias of the environmental effect with increased effect size when temporal variation in the covariate was lower and/or uncertainty in the covariate observations was higher.
+When the OMs had R+S process errors with low temporal variation in the true environmental covariate and lower uncertainty in the indices age age composition, estimated covariate effects were highly variable.
+In most cases the relative error of $\beta_\text{Ecov}$ did not depend on the source of process error assumed in the EM.
+When there was an effect of the EM process error assumption it was when OMs had R+S process errors.
+The worst bias was observed when OMs assumed R+S process errors, high uncertainty in covariate observations, low variability in the covariate, and low uncertainty in index and age composition observations.
+
+When the mean natural mortality rate was estimated in the EM, results were similar except estimated effects were even more variable for data simulated with R+S process errors (Figure \ref{Ecov_beta_bias_M_estimated}).
+There was also more separation of reliability of the estimation of the effect among EMs with different process error assumptions.
+The separation was most apparent when OMs simulated R+S process errors and larger variability in the environmental covariate wehere EMs with process errors other than R+S showing more bias than when the mean natural mortality rate was assumed known.
+
+### Mean natural mortality rate {-}
+
+We found high accuracy for estimation of the mean natural mortality rate parameter ($\beta_M$) for all EM process errors assumptions when OMs had step changes in fishing mortality, lower uncertainty in index and age composition observations, and either R or R+M process errors (Figure \ref{mean_M_bias}).
+The most variation in estimates occurred when fishing and Higher uncertainty in index and age composition observations.
+For OMs with R+S process errors, the most reliable estimation of $\beta_M$ was obtained when the EM also assumed R+S process errors across all other factors.
+For OMs with R+M process errors, the matching assumption for the EM only showed the best reliability when there was low uncertainty in index and age composition observations and fishing mortality was constant.
+
+### Annual natural morality rate {-}
+
+We present results for error in annual natural mortality rate conditional on three alternative EM configurations for the natural mortality parameters: 1) mean natural mortality rate parameter is fixed at the true value ($\beta_M = log(0.2)$) and no covariate effect is assmed ($\beta_\text{Ecov} = 0$),  2) $\beta_M = log(0.2)$ and $\beta_\text{Ecov}$ is estimated, and 3) both $\beta_M$ and $\beta_\text{Ecov}$ are estimated.
+
+When OMs and EMs assume $\beta_M = log(0.2)$ and $\beta_\text{Ecov} = 0$, there is no annual variation in natural mortality for OMs with R or R+S process errors simulated or assumed in the EM and, therefore, estimation bias is not possible (Figure \ref{annual_M_bias_M_fixed_beta_fixed}).
+We also observe little or no evidence of bias (confidence intervals include 0) for R or R+S EMs for any OMs even when the OMs included an effect of the covariate on natural mortality ($\beta_\text{Ecov} > 0$).
+Including process error on M (R+M) produces more variability in errors for R and R+S EMs than including increased level of effect of the covariate on M.
+For R and R+S OMs, errors in annual M for R+S EMs were less variable than those for R EMs.
+
+R+M EMs fit to R OMs with constant fishing mortality, or a step change in fishing mortality and lower uncertainty in index and age composition observations exhibited no bias in annual estimation of M indicating that the variance of the estimated random effects for M in these EMs collapsed to 0.
+R+M EMs were the only EMs to exhibit differences in the sign of the median errors across the time series.
+When OMs had a step change in fishing mortality and higher uncertainty in index and age composition observations, we observed positive median errrors at the beginning of the time series and lower median errors in later years.
+However, whether there was evidence of bias, depended on the uncertainty in covariate observations and the degree of variability in the latent covariate.
+We observed negative median relative errors in annual M estimates for R+M EMs fit to R+S OMs with lower uncertainty in index and age composition observations across all levels of simulated effects of the covariate, but evidence of bias was strongest at the beginning of the time series.
+
+When $\beta_\text{Ecov}$ was estimated, R EMs performed worse for R+S OMs with evidence of negative bias for OMs with cosntant fishing mortality, lower uncertainty in index and age composition observations, the largest covariate effect size, lower variability in the latent covariate, and larger uncertainty in the covariate observations (Figure \ref{annual_M_bias_M_fixed_beta_estimated}).
+There was little difference in the results for R+S EMs whether $\beta_\text{Ecov}$ was estimated or not.
+The results for R+M EMs were generally similar to those when $\beta_\text{Ecov}=0$ was assumed, except that the differences between the estimates at the beginning of the time series and later years for certain OM configurations did not occur.
+
+Allowing the EMs to estimate $\beta_M$ resulted in very large variability in estimates for all EMs for R and R+M OMs with constant fishing mortality and higher uncertainty in index and age composition observations (Figure \ref{annual_M_bias_M_estimated_beta_estimated}).
+The same varibilities occurred for R+S OMs, but strong bias was estimated for R and R+M EMs.
+Annual M estimation was most reliable when OMs had step changes in fishing mortality and lower uncertainty in index and age composition observations.
+For R and R+M OMs with those characteristics, all EMs generally provided accurate estimation of annual M, but only R+S EMs provided accurate estimation for R+S OMs.
+Little or no bias in annual M estimation was observed for R+S EMs across all OM process error assumptions as long as there was a step change in fishing mortality and lower uncertainty in indices and age composition.
+
+### Spawning stock biomass {-}
+
+Like natural mortality, we present results for error in spawning biomass conditional on three alternative EM configurations for the natural mortality parameters: 1) mean natural mortality rate parameter is fixed at the true value ($\beta_M = log(0.2)$) and no covariate effect is assmed ($\beta_\text{Ecov} = 0$),  2) $\beta_M = log(0.2)$ and $\beta_\text{Ecov}$ is estimated, and 3) both $\beta_M$ and $\beta_\text{Ecov}$ are estimated.
+
+When EMs assume $\beta_M = log(0.2)$ and $\beta_\text{Ecov} = 0$, there is little evidence of bias for R and R+M OMs except when there is a step change in fishing mortality and higher uncertainty in index and age composition observations, particularly at the end of the time series (Figure \ref{SSB_bias_M_fixed_beta_fixed}).
+We observed a similar trend in median relative error for R+S OMs, but there was more evidence of positive bias at the beginning of the time series whereas the confidence intervals for the negative median errors at the end of the time series often included 0.
+However, there was more indication of positive bias of the incorrect process error assumption of the EMs for the R+S OMs.
+For R+S EMs fit to R+S OMs, there was indication of small positive bias at the beginning of the time series when there was constant fishing mortality and higher uncertainty in index and age composition observations.
+
+When EMs assume $\beta_M = log(0.2)$, but estimate $\beta_\text{Ecov}$, the median errors for SSB for R and and R+M OMs are similar to those when $\beta_\text{Ecov} = 0$ (Figure \ref{SSB_bias_M_fixed_beta_estimated}).
+For R+S OMs, the median errors for EMs that also assume R+S process errors are also similar to those with no covariate effect assumed.
+However, for R and R+M EMs fit to R+S OMs with constant fishing mortality rate and lower uncertainty in indices and age composition observations, we observed evidences of negative bias when higher covariate observation uncertainty and lower variation in the latent covariate was simulated and positive bias for other configurations of the covariate and corresponding observation uncertainty.
+When R+S OMs had constant fishing mortality rate and higher uncertainty in indices and age composition observations or step changes in fishing mortality and lower uncertainty in indices and age composition, R and R+M EMs often provided positively biased SSB estimates when there was lower uncertainty in covariate observations.
+
+When EMs estimated both $\beta_M$ and $\beta_\text{Ecov}$, we observed large variation in the errors in SSB under the same OM configurations where we observed large variation in errors for annual natural mortality rates (Figure \ref{SSB_bias_M_estimated_beta_estimated}).
+Similarly, we observed little or no bias in SSB estimation for R+S EMs across all OM process error assumptions as long as there was a step change in fishing mortality and lower uncertainty in indices and age composition.
+
+# Discussion {-}
+
+The estimating models assumed variances of aggregate catch and index observations was known.
+This approximation may be appropriate for indices where we have a reliable estimate of uncertainty based on the survey design (), but there may be better approaches for the aggregate catch such as an informed prior on the standard errors with realistic bounds.
+
+We found EMs with R+M process errors were rarely determined as the appropriate model when OMs simulated R+M process errors.
+This was unexpected but is likely due to the size of the variance assumed for those process errors ($\sigma_M = 0.3$) relative to the variances assumed for index and age composition observations.
+This is related to the difficulty in separately estimating observation and process error variances when the ratio of process to observation uncertainty is low.
+
+Note that the results for bias of covariate effect all assume ecov beta is estimated.
+The lack of bias in certain situations might suggest including the effect even if AIC says it isn't better at least when contrast in Ecov is high.
+
+The bias in natural mortality estimation resulted in biased estimation of stock size and likely harvest rates, but it will also propagate into biological reference points and possibly stock status.
+
+The lack of bias observed for annual M, using R+S EMs when no effect was assumed must be due to the annual M being equal to the assumed value (0.2) on average across simulations because the simulated environmental covariate has mean 0.
+However, if one were to condition on the covariate time series we would expect biased estimation of annual M when there was a covariate effect.
+
+# Conclusions {-}
+
+As we would expect, reliable detection of covariate effects requires informative data. 
+AIC preferred simpler models than the true model when information content in data and contrast in covariates and abundance were low.
+Null model for environmental covariate effect (no covariate effect) was selected when contrast in the time series was low  and/or uncertainty in observations was high.
+
+When there was process error in recruitment and M  (R+M), models with process error only in recruitment were preferred.
+This could be because the variation in the M random effects was small relative to observation variability.
+The marginal variance for log M random effects was 0.3.
+
+Null selection likely decreases with strength of the effect on M.
+We only examined two non-zero effect sizes: 0.25, and 0.5, but our results suggest larger effect sizes, with the same observation error and contrast in time series, would allow better AIC performance for determining covariate effects.
+
+Covariate effect estimation can be robust to process error assumptions (For R+M OMs) with contrast in covariate and low observation error.
+
+Reliable estimation of environmental effect on M regardless of the process error assumed by the EM as long as the contrast in the covariate is sufficient and the uncertainty in the observations is low.
+
+# Acknowledgements {-}
+
+This work was funded by NOAA Fisheries Northeast Fisheries Science Center.
+
+<!--
+# CRediT authorship contribution statement {-}
+
+**Timothy J. Miller:** Conceptualization, Methodology, Writing - original draft, Formal analysis, Visualization.
+ -->
+
+\pagebreak
+
+\bibliography{paper}
+<!-- \bibliography{project-0-paper}
+\bibliographystyle{/home/tmiller2/work/bibliographies/cjfas2}  -->
+
+<div id="refs"></div>
+
+\pagebreak
+
+\begin{figure}
+\caption{The proportion mature at age assumed for the population in all operating and estimating models.}\label{om_maturity}
+\begin{center}
+\includegraphics[width = \textwidth]{om_maturity.png}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\caption{The weight at age assumed for the population in all operating and estimating models.}\label{om_waa}
+\begin{center}
+\includegraphics[width = \textwidth]{om_waa.png}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\caption{The Beverton-Holt stock recruit relationship assume for all operating models.}\label{om_sr}
+\begin{center}
+\includegraphics[width = \textwidth]{om_sr.png}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\caption{Example simulations of environmental covariate latent processes and observations with different levels of observation error, and different assumptions about variability of the latent process.}\label{om_ecov_example}
+\begin{center}
+\includegraphics[width = \textwidth]{Ecov_true_obs_example.png}
+\end{center}
+\end{figure}
+
+\begin{landscape}
+\begin{figure}
+\caption{Example simulations of annual natural mortality rates that may be a function of a temporally varying environmental covariate and autoregressive random effects.}\label{M_example}
+\begin{center}
+\includegraphics[width = \textwidth]{M_example.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{figure}
+\caption{The selectivity at age assumed for the fishing fleet in operating models without selectivity random effects.
+The mean selectivity at age across time for the fishing fleet in operating models with selectivity random effects.}\label{om_mean_selectivity}
+\begin{center}
+\includegraphics[width = \textwidth]{om_mean_selectivity.png}
+\end{center}
+\end{figure}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of convergence of estimating models with alternative process error assumptions fit to simulated populations and data sets from operating models.
+The convergence criterion is an invertible hessian providing standard error estimates for all fixed effects parameters.
+Vertical lines represent 95\% confidence intervals.
+All estimating models estimate environmental covariate effects on natural mortality but assume the mean mortality parameter $\beta_\text{M} = \log 0.2$.}\label{convergence_M_fixed}
+\begin{center}
+\includegraphics[height = \textheight]{proportion_good_hessian_ecov_effect_est_M_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of convergence of estimating models with alternative process error assumptions fit to simulated populations and data sets from operating models.
+The convergence criterion is an invertible hessian providing standard error estimates for all fixed effects parameters.
+Vertical lines represent 95\% confidence intervals.
+All estimating models estimate environmental covariate effects on natural mortality and the mean mortality parameter $\beta_\text{M}$.}\label{convergence_M_estimated}
+\begin{center}
+\includegraphics[height = \textheight]{proportion_good_hessian_ecov_effect_est_M_est.png}
+\end{center}
+\end{figure}
+\end{landscape}
 
@@
 
-## Bias {-}
-
-### Environmental effect {-}
-
-When the EMs assumed the mean natural mortality rate was known, we observed generally accurate estimation of environmental effects across all EM and OM process error assumptions and all true covariate effect sizes, when ther was low uncertainty in environmental observations and larger temporal contrast in the simulated true environmental covariate (Figure \ref{Ecov_beta_bias_M_fixed}).
-We observed a negative trend in bias of the environmental effect with increased effect size when temporal variation in the covariate was lower and/or uncertainty in the covariate observations was higher.
-When the OMs had R+S process errors with low temporal variation in the true environmental covariate and lower uncertainty in the indices age age composition, estimated covariate effects were highly variable.
-In most cases the relative error of $\beta_\text{Ecov}$ did not depend on the source of process error assumed in the EM.
-When there was an effect of the EM process error assumption it was when OMs had R+S process errors.
-The worst bias was observed when OMs assumed R+S process errors, high uncertainty in covariate observations, low variability in the covariate, and low uncertainty in index and age composition observations.
-
-When the mean natural mortality rate was estimated in the EM, results were similar except estimated effects were even more variable for data simulated with R+S process errors (Figure \ref{Ecov_beta_bias_M_estimated}).
-There was also more separation of reliability of the estimation of the effect among EMs with different process error assumptions.
-The separation was most apparent when OMs simulated R+S process errors and larger variability in the environmental covariate wehere EMs with process errors other than R+S showing more bias than when the mean natural mortality rate was assumed known.
-
 \begin{landscape}
@@
 
-### Mean natural mortality rate {-}
-
 \begin{landscape}
@@
 \end{landscape}
-
-### Annual natural morality rate {-}
 
@@
 
-### Spawning stock biomass {-}
-
 \begin{landscape}
@@
 
-# Discussion {-}
-
-The estimating models assumed variances of aggregate catch and index observations was known.
-This approximation may be appropriate for indices where we have a reliable estimate of uncertainty based on the survey design (), but there may be better approaches for the aggregate catch such as an informed prior on the standard errors with realistic bounds.
-
-We found EMs with R+M process errors were rarely determined as the appropriate model when OMs simulated R+M process errors.
-This was unexpected but is likely due to the size of the variance assumed for those process errors ($\sigma_M = 0.3$) relative to the variances assumed for index and age composition observations.
-This is related to the difficulty in separately estimating observation and process error variances when the ratio of process to observation uncertainty is low.
-
-Note that the results for bias of covariate effect all assume ecov beta is estimated.
-The lack of bias in certain situations might suggest including the effect even if AIC says it isn't better at least when contrast in Ecov is high.
-
-The bias in natural mortality estimation resulted in biased estimation of stock size and likely harvest rates, but it will also propagate into biological reference points and possibly stock status.
-
-# Conclusions {-}
-
-As we would expect, reliable detection of covariate effects requires informative data. 
-AIC preferred simpler models than the true model when information content in data and contrast in covariates and abundance were low.
-Null model for environmental covariate effect (no covariate effect) was selected when contrast in the time series was low  and/or uncertainty in observations was high.
-
-When there was process error in recruitment and M  (R+M), models with process error only in recruitment were preferred.
-This could be because the variation in the M random effects was small relative to observation variability.
-The marginal variance for log M random effects was 0.3.
-
-Null selection likely decreases with strength of the effect on M.
-We only examined two non-zero effect sizes: 0.25, and 0.5, but our results suggest larger effect sizes, with the same observation error and contrast in time series, would allow better AIC performance for determining covariate effects.
-
-Covariate effect estimation can be robust to process error assumptions (For R+M OMs) with contrast in covariate and low observation error.
-
-Reliable estimation of environmental effect on M regardless of the process error assumed by the EM as long as the contrast in the covariate is sufficient and the uncertainty in the observations is low.
-
-# Acknowledgements {-}
-
-This work was funded by NOAA Fisheries Northeast Fisheries Science Center.
-
+# Supplemental Materials {-}
 <!--
-# CRediT authorship contribution statement {-}
-
-**Timothy J. Miller:** Conceptualization, Methodology, Writing - original draft, Formal analysis, Visualization.
+\setcounter{figure}{0}
+\renewcommand\thefigure{S\arabic{figure}}
 -->
 
-\pagebreak
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of annual fully-selected fishing mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
+Estimating models also assume the mean natural mortality parameter $\beta_\text{M} = \log 0.2$ and no environmental covariate effect $\beta_\text{Ecov} = 0$.}\label{F_bias_M_fixed_beta_fixed}
+\begin{center}
+\includegraphics[height = \textheight]{F_bias_all_PE_effect_M_fixed_beta_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
 
-\bibliography{paper}
-<!-- \bibliography{project-0-paper}
-\bibliographystyle{/home/tmiller2/work/bibliographies/cjfas2}  -->
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of annual fully-selected fishing mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
+Estimating models also assume the mean natural mortality parameter $\beta_\text{M} = \log 0.2$ and the environmental covariate effect $\beta_\text{Ecov}$ is estimated.}\label{F_bias_M_fixed_beta_estimated}
+\begin{center}
+\includegraphics[height = \textheight]{F_bias_all_PE_effect_M_fixed_beta_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
 
-<div id="refs"></div>
-
-\pagebreak
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of annual fully-selected fishing mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
+Estimating models also assume both the mean natural mortality parameter $\beta_\text{M}$ and the environmental covariate effect $\beta_\text{Ecov}$ are estimated.}\label{F_bias_M_estimated_beta_estimated}
+\begin{center}
+\includegraphics[height = \textheight]{F_bias_all_PE_effect_M_estimated_beta_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}

### 2023-10-04 Ecov_study/mortality/paper/paper.tex
saving work on M-ecov working paper

@@
 
-\begin{figure}
-\caption{The proportion mature at age assumed for the population in all operating and estimating models.}\label{om_maturity}
-\begin{center}
-\includegraphics[width = \textwidth]{om_maturity.png}
-\end{center}
-\end{figure}
-
 Weight at age was generated with a LVB growth function \[
@@
 \ref{om_waa}).
-
-\begin{figure}
-\caption{The weight at age assumed for the population in all operating and estimating models.}\label{om_waa}
-\begin{center}
-\includegraphics[width = \textwidth]{om_waa.png}
-\end{center}
-\end{figure}
 
@@
 
-\begin{figure}
-\caption{The Beverton-Holt stock recruit relationship assume for all operating models.}\label{om_sr}
-\begin{center}
-\includegraphics[width = \textwidth]{om_sr.png}
-\end{center}
-\end{figure}
-
 Two alternative fishing histories were used for operating models. In the
@@
 process and observations under the alternative configurations.
-
-\begin{figure}
-\caption{Example simulations of environmental covariate latent processes and observations with different levels of observation error, and different assumptions about variability of the latent process.}\label{om_ecov_example}
-\begin{center}
-\includegraphics[width = \textwidth]{Ecov_true_obs_example.png}
-\end{center}
-\end{figure}
 
@@
 \end{itemize}
+
+The parameters defining the simulated covariate time series, size of the
+covariate effect, and any natural mortality random effects result in a
+range of different levels of variation in annual natural mortality rates
+(Figure \ref{M_example}).
 
@@
 
-\begin{figure}
-\caption{The selectivity at age assumed for the fishing fleet in operating models without selectivity random effects.
-The mean selectivity at age across time for the fishing fleet in operating models with selectivity random effects.}\label{om_mean_selectivity}
-\begin{center}
-\includegraphics[width = \textwidth]{om_mean_selectivity.png}
-\end{center}
-\end{figure}
-
 \hypertarget{indices}{%
@@
 
-For the \(i\)th simulated data the relative error for a parameter
-\(\theta\) provided from the fitted estimation model is \[
+Results for fishing mortality rate are provided in the Supplementary
+Materials. For the \(i\)th simulated data the relative error for a
+parameter \(\theta\) provided from the fitted estimation model is \[
 {\text{RE}_i}\left(\theta\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
@@
 
+For natural mortality rate, SSB, and fully-selected fishing mortality
+rate, we summarized results for each of the annual values, but we
+present results only for estimates from the first year (start), year 21
+(middel), and the last year (end), because there were no appreciable
+differences between the results for these three years and those for the
+other years.
+
 \hypertarget{results}{%
@@
 
-\begin{landscape}
-\begin{figure}
-\caption{Probability of convergence of estimating models with alternative process error assumptions fit to simulated populations and data sets from operating models.
-The convergence criterion is an invertible hessian providing standard error estimates for all fixed effects parameters.
-Vertical lines represent 95\% confidence intervals.
-All estimating models estimate environmental covariate effects on natural mortality but assume the mean mortality parameter $\beta_\text{M} = \log 0.2$.}\label{convergence_M_fixed}
-\begin{center}
-\includegraphics[height = \textheight]{proportion_good_hessian_ecov_effect_est_M_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
 As might be expected, there was an overall drop in the probability of
@@
 fixed, apply when estimated.
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of convergence of estimating models with alternative process error assumptions fit to simulated populations and data sets from operating models.
-The convergence criterion is an invertible hessian providing standard error estimates for all fixed effects parameters.
-Vertical lines represent 95\% confidence intervals.
-All estimating models estimate environmental covariate effects on natural mortality and the mean mortality parameter $\beta_\text{M}$.}\label{convergence_M_estimated}
-\begin{center}
-\includegraphics[height = \textheight]{proportion_good_hessian_ecov_effect_est_M_est.png}
-\end{center}
-\end{figure}
-\end{landscape}
 
@@
 error and determining a covariate effect when there was one.
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets for each operating model where the estimation model type (process error and covariate effect assumptions) had the lowest AIC.
-All estimating models assumed the mean mortality parameter $\beta_\text{M} = \log 0.2$.}\label{aic_rank_M_fixed}
-\begin{center}
-\includegraphics[height = \textheight]{proportion_best_AIC_M_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets for each operating model where the estimation model type (process error and covariate effect assumptions) had the lowest AIC.
-All estimating models assumed the mean mortality parameter $\beta_\text{M} = \log 0.2$.}\label{aic_rank_M_estimated}
-\begin{center}
-\includegraphics[height = \textheight]{proportion_best_AIC_M_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
 
@@
 
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of the estimated environmental effect on natural mortality $\beta_\text{Ecov}$ from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
-All estimating models assumed the mean mortality parameter $\beta_\text{M} = \log 0.2$.}\label{Ecov_beta_bias_M_fixed}
-\begin{center}
-\includegraphics[height = \textheight]{Ecov_beta_bias_all_PE_effect_M_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of the estimated environmental effect on natural mortality $\beta_\text{Ecov}$ from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
-All estimating models assumed the mean mortality parameter $\beta_\text{M}$ is also estimated.}\label{Ecov_beta_bias_M_estimated}
-\begin{center}
-\includegraphics[height = \textheight]{Ecov_beta_bias_all_PE_effect_M_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
 \hypertarget{mean-natural-mortality-rate}{%
@@
 
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of the estimated mean natural mortality parameter $\beta_\text{M}$ from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.}\label{mean_M_bias}
-\begin{center}
-\includegraphics[height = \textheight]{mean_M_bias_all_PE_effect.png}
-\end{center}
-\end{figure}
-\end{landscape}
+We found high accuracy for estimation of the mean natural mortality rate
+parameter (\(\beta_M\)) for all EM process errors assumptions when OMs
+had step changes in fishing mortality, lower uncertainty in index and
+age composition observations, and either R or R+M process errors (Figure
+\ref{mean_M_bias}). The most variation in estimates occurred when
+fishing and Higher uncertainty in index and age composition
+observations. For OMs with R+S process errors, the most reliable
+estimation of \(\beta_M\) was obtained when the EM also assumed R+S
+process errors across all other factors. For OMs with R+M process
+errors, the matching assumption for the EM only showed the best
+reliability when there was low uncertainty in index and age composition
+observations and fishing mortality was constant.
 
@@
 
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of annual natural mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
-Estimating models also assume the mean natural mortality parameter $\beta_\text{M} = \log 0.2$ and no environmental covariate effect $\beta_\text{Ecov} = 0$.}\label{annual_M_bias_M_fixed_beta_fixed}
-\begin{center}
-\includegraphics[height = \textheight]{annual_M_bias_all_PE_effect_M_fixed_beta_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of annual natural mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
-Estimating models also assume the mean natural mortality parameter $\beta_\text{M} = \log 0.2$ and the environmental covariate effect $\beta_\text{Ecov}$ is estimated.}\label{annual_M_bias_M_fixed_beta_estimated}
-\begin{center}
-\includegraphics[height = \textheight]{annual_M_bias_all_PE_effect_M_fixed_beta_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of annual natural mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
-Estimating models also assume both the mean natural mortality parameter $\beta_\text{M}$ and the environmental covariate effect $\beta_\text{Ecov}$ are estimated.}\label{annual_M_bias_M_estimated_beta_estimated}
-\begin{center}
-\includegraphics[height = \textheight]{annual_M_bias_all_PE_effect_M_estimated_beta_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
+We present results for error in annual natural mortality rate
+conditional on three alternative EM configurations for the natural
+mortality parameters: 1) mean natural mortality rate parameter is fixed
+at the true value (\(\beta_M = log(0.2)\)) and no covariate effect is
+assmed (\(\beta_\text{Ecov} = 0\)), 2) \(\beta_M = log(0.2)\) and
+\(\beta_\text{Ecov}\) is estimated, and 3) both \(\beta_M\) and
+\(\beta_\text{Ecov}\) are estimated.
+
+When OMs and EMs assume \(\beta_M = log(0.2)\) and
+\(\beta_\text{Ecov} = 0\), there is no annual variation in natural
+mortality for OMs with R or R+S process errors simulated or assumed in
+the EM and, therefore, estimation bias is not possible (Figure
+\ref{annual_M_bias_M_fixed_beta_fixed}). We also observe little or no
+evidence of bias (confidence intervals include 0) for R or R+S EMs for
+any OMs even when the OMs included an effect of the covariate on natural
+mortality (\(\beta_\text{Ecov} > 0\)). Including process error on M
+(R+M) produces more variability in errors for R and R+S EMs than
+including increased level of effect of the covariate on M. For R and R+S
+OMs, errors in annual M for R+S EMs were less variable than those for R
+EMs.
+
+R+M EMs fit to R OMs with constant fishing mortality, or a step change
+in fishing mortality and lower uncertainty in index and age composition
+observations exhibited no bias in annual estimation of M indicating that
+the variance of the estimated random effects for M in these EMs
+collapsed to 0. R+M EMs were the only EMs to exhibit differences in the
+sign of the median errors across the time series. When OMs had a step
+change in fishing mortality and higher uncertainty in index and age
+composition observations, we observed positive median errrors at the
+beginning of the time series and lower median errors in later years.
+However, whether there was evidence of bias, depended on the uncertainty
+in covariate observations and the degree of variability in the latent
+covariate. We observed negative median relative errors in annual M
+estimates for R+M EMs fit to R+S OMs with lower uncertainty in index and
+age composition observations across all levels of simulated effects of
+the covariate, but evidence of bias was strongest at the beginning of
+the time series.
+
+When \(\beta_\text{Ecov}\) was estimated, R EMs performed worse for R+S
+OMs with evidence of negative bias for OMs with cosntant fishing
+mortality, lower uncertainty in index and age composition observations,
+the largest covariate effect size, lower variability in the latent
+covariate, and larger uncertainty in the covariate observations (Figure
+\ref{annual_M_bias_M_fixed_beta_estimated}). There was little difference
+in the results for R+S EMs whether \(\beta_\text{Ecov}\) was estimated
+or not. The results for R+M EMs were generally similar to those when
+\(\beta_\text{Ecov}=0\) was assumed, except that the differences between
+the estimates at the beginning of the time series and later years for
+certain OM configurations did not occur.
+
+Allowing the EMs to estimate \(\beta_M\) resulted in very large
+variability in estimates for all EMs for R and R+M OMs with constant
+fishing mortality and higher uncertainty in index and age composition
+observations (Figure \ref{annual_M_bias_M_estimated_beta_estimated}).
+The same varibilities occurred for R+S OMs, but strong bias was
+estimated for R and R+M EMs. Annual M estimation was most reliable when
+OMs had step changes in fishing mortality and lower uncertainty in index
+and age composition observations. For R and R+M OMs with those
+characteristics, all EMs generally provided accurate estimation of
+annual M, but only R+S EMs provided accurate estimation for R+S OMs.
+Little or no bias in annual M estimation was observed for R+S EMs across
+all OM process error assumptions as long as there was a step change in
+fishing mortality and lower uncertainty in indices and age composition.
 
@@
 
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of annual SSB in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
-Estimating models also assume the mean natural mortality parameter $\beta_\text{M} = \log 0.2$ and no environmental covariate effect $\beta_\text{Ecov} = 0$.}\label{SSB_bias_M_fixed_beta_fixed}
-\begin{center}
-\includegraphics[height = \textheight]{SSB_bias_all_PE_effect_M_fixed_beta_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of annual SSB in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
-Estimating models also assume the mean natural mortality parameter $\beta_\text{M} = \log 0.2$ and the environmental covariate effect $\beta_\text{Ecov}$ is estimated.}\label{SSB_bias_M_fixed_beta_estimated}
-\begin{center}
-\includegraphics[height = \textheight]{SSB_bias_all_PE_effect_M_fixed_beta_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of annual SSB in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
-Estimating models also assume both the mean natural mortality parameter $\beta_\text{M}$ and the environmental covariate effect $\beta_\text{Ecov}$ are estimated.}\label{SSB_bias_M_estimated_beta_estimated}
-\begin{center}
-\includegraphics[height = \textheight]{SSB_bias_all_PE_effect_M_estimated_beta_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
+Like natural mortality, we present results for error in spawning biomass
+conditional on three alternative EM configurations for the natural
+mortality parameters: 1) mean natural mortality rate parameter is fixed
+at the true value (\(\beta_M = log(0.2)\)) and no covariate effect is
+assmed (\(\beta_\text{Ecov} = 0\)), 2) \(\beta_M = log(0.2)\) and
+\(\beta_\text{Ecov}\) is estimated, and 3) both \(\beta_M\) and
+\(\beta_\text{Ecov}\) are estimated.
+
+When EMs assume \(\beta_M = log(0.2)\) and \(\beta_\text{Ecov} = 0\),
+there is little evidence of bias for R and R+M OMs except when there is
+a step change in fishing mortality and higher uncertainty in index and
+age composition observations, particularly at the end of the time series
+(Figure \ref{SSB_bias_M_fixed_beta_fixed}). We observed a similar trend
+in median relative error for R+S OMs, but there was more evidence of
+positive bias at the beginning of the time series whereas the confidence
+intervals for the negative median errors at the end of the time series
+often included 0. However, there was more indication of positive bias of
+the incorrect process error assumption of the EMs for the R+S OMs. For
+R+S EMs fit to R+S OMs, there was indication of small positive bias at
+the beginning of the time series when there was constant fishing
+mortality and higher uncertainty in index and age composition
+observations.
+
+When EMs assume \(\beta_M = log(0.2)\), but estimate
+\(\beta_\text{Ecov}\), the median errors for SSB for R and and R+M OMs
+are similar to those when \(\beta_\text{Ecov} = 0\) (Figure
+\ref{SSB_bias_M_fixed_beta_estimated}). For R+S OMs, the median errors
+for EMs that also assume R+S process errors are also similar to those
+with no covariate effect assumed. However, for R and R+M EMs fit to R+S
+OMs with constant fishing mortality rate and lower uncertainty in
+indices and age composition observations, we observed evidences of
+negative bias when higher covariate observation uncertainty and lower
+variation in the latent covariate was simulated and positive bias for
+other configurations of the covariate and corresponding observation
+uncertainty. When R+S OMs had constant fishing mortality rate and higher
+uncertainty in indices and age composition observations or step changes
+in fishing mortality and lower uncertainty in indices and age
+composition, R and R+M EMs often provided positively biased SSB
+estimates when there was lower uncertainty in covariate observations.
+
+When EMs estimated both \(\beta_M\) and \(\beta_\text{Ecov}\), we
+observed large variation in the errors in SSB under the same OM
+configurations where we observed large variation in errors for annual
+natural mortality rates (Figure
+\ref{SSB_bias_M_estimated_beta_estimated}). Similarly, we observed
+little or no bias in SSB estimation for R+S EMs across all OM process
+error assumptions as long as there was a step change in fishing
+mortality and lower uncertainty in indices and age composition.
 
@@
 
+The lack of bias observed for annual M, using R+S EMs when no effect was
+assumed must be due to the annual M being equal to the assumed value
+(0.2) on average across simulations because the simulated environmental
+covariate has mean 0. However, if one were to condition on the covariate
+time series we would expect biased estimation of annual M when there was
+a covariate effect.
+
 \hypertarget{conclusions}{%
@@
 
+\begin{figure}
+\caption{The proportion mature at age assumed for the population in all operating and estimating models.}\label{om_maturity}
+\begin{center}
+\includegraphics[width = \textwidth]{om_maturity.png}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\caption{The weight at age assumed for the population in all operating and estimating models.}\label{om_waa}
+\begin{center}
+\includegraphics[width = \textwidth]{om_waa.png}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\caption{The Beverton-Holt stock recruit relationship assume for all operating models.}\label{om_sr}
+\begin{center}
+\includegraphics[width = \textwidth]{om_sr.png}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\caption{Example simulations of environmental covariate latent processes and observations with different levels of observation error, and different assumptions about variability of the latent process.}\label{om_ecov_example}
+\begin{center}
+\includegraphics[width = \textwidth]{Ecov_true_obs_example.png}
+\end{center}
+\end{figure}
+
+\begin{landscape}
+\begin{figure}
+\caption{Example simulations of annual natural mortality rates that may be a function of a temporally varying environmental covariate and autoregressive random effects.}\label{M_example}
+\begin{center}
+\includegraphics[width = \textwidth]{M_example.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{figure}
+\caption{The selectivity at age assumed for the fishing fleet in operating models without selectivity random effects.
+The mean selectivity at age across time for the fishing fleet in operating models with selectivity random effects.}\label{om_mean_selectivity}
+\begin{center}
+\includegraphics[width = \textwidth]{om_mean_selectivity.png}
+\end{center}
+\end{figure}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of convergence of estimating models with alternative process error assumptions fit to simulated populations and data sets from operating models.
+The convergence criterion is an invertible hessian providing standard error estimates for all fixed effects parameters.
+Vertical lines represent 95\% confidence intervals.
+All estimating models estimate environmental covariate effects on natural mortality but assume the mean mortality parameter $\beta_\text{M} = \log 0.2$.}\label{convergence_M_fixed}
+\begin{center}
+\includegraphics[height = \textheight]{proportion_good_hessian_ecov_effect_est_M_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of convergence of estimating models with alternative process error assumptions fit to simulated populations and data sets from operating models.
+The convergence criterion is an invertible hessian providing standard error estimates for all fixed effects parameters.
+Vertical lines represent 95\% confidence intervals.
+All estimating models estimate environmental covariate effects on natural mortality and the mean mortality parameter $\beta_\text{M}$.}\label{convergence_M_estimated}
+\begin{center}
+\includegraphics[height = \textheight]{proportion_good_hessian_ecov_effect_est_M_est.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets for each operating model where the estimation model type (process error and covariate effect assumptions) had the lowest AIC.
+All estimating models assumed the mean mortality parameter $\beta_\text{M} = \log 0.2$.}\label{aic_rank_M_fixed}
+\begin{center}
+\includegraphics[height = \textheight]{proportion_best_AIC_M_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets for each operating model where the estimation model type (process error and covariate effect assumptions) had the lowest AIC.
+All estimating models assumed the mean mortality parameter $\beta_\text{M} = \log 0.2$.}\label{aic_rank_M_estimated}
+\begin{center}
+\includegraphics[height = \textheight]{proportion_best_AIC_M_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of the estimated environmental effect on natural mortality $\beta_\text{Ecov}$ from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
+All estimating models assumed the mean mortality parameter $\beta_\text{M} = \log 0.2$.}\label{Ecov_beta_bias_M_fixed}
+\begin{center}
+\includegraphics[height = \textheight]{Ecov_beta_bias_all_PE_effect_M_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of the estimated environmental effect on natural mortality $\beta_\text{Ecov}$ from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
+All estimating models assumed the mean mortality parameter $\beta_\text{M}$ is also estimated.}\label{Ecov_beta_bias_M_estimated}
+\begin{center}
+\includegraphics[height = \textheight]{Ecov_beta_bias_all_PE_effect_M_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of the estimated mean natural mortality parameter $\beta_\text{M}$ from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.}\label{mean_M_bias}
+\begin{center}
+\includegraphics[height = \textheight]{mean_M_bias_all_PE_effect.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of annual natural mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
+Estimating models also assume the mean natural mortality parameter $\beta_\text{M} = \log 0.2$ and no environmental covariate effect $\beta_\text{Ecov} = 0$.}\label{annual_M_bias_M_fixed_beta_fixed}
+\begin{center}
+\includegraphics[height = \textheight]{annual_M_bias_all_PE_effect_M_fixed_beta_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of annual natural mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
+Estimating models also assume the mean natural mortality parameter $\beta_\text{M} = \log 0.2$ and the environmental covariate effect $\beta_\text{Ecov}$ is estimated.}\label{annual_M_bias_M_fixed_beta_estimated}
+\begin{center}
+\includegraphics[height = \textheight]{annual_M_bias_all_PE_effect_M_fixed_beta_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of annual natural mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
+Estimating models also assume both the mean natural mortality parameter $\beta_\text{M}$ and the environmental covariate effect $\beta_\text{Ecov}$ are estimated.}\label{annual_M_bias_M_estimated_beta_estimated}
+\begin{center}
+\includegraphics[height = \textheight]{annual_M_bias_all_PE_effect_M_estimated_beta_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of annual SSB in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
+Estimating models also assume the mean natural mortality parameter $\beta_\text{M} = \log 0.2$ and no environmental covariate effect $\beta_\text{Ecov} = 0$.}\label{SSB_bias_M_fixed_beta_fixed}
+\begin{center}
+\includegraphics[height = \textheight]{SSB_bias_all_PE_effect_M_fixed_beta_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of annual SSB in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
+Estimating models also assume the mean natural mortality parameter $\beta_\text{M} = \log 0.2$ and the environmental covariate effect $\beta_\text{Ecov}$ is estimated.}\label{SSB_bias_M_fixed_beta_estimated}
+\begin{center}
+\includegraphics[height = \textheight]{SSB_bias_all_PE_effect_M_fixed_beta_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of annual SSB in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
+Estimating models also assume both the mean natural mortality parameter $\beta_\text{M}$ and the environmental covariate effect $\beta_\text{Ecov}$ are estimated.}\label{SSB_bias_M_estimated_beta_estimated}
+\begin{center}
+\includegraphics[height = \textheight]{SSB_bias_all_PE_effect_M_estimated_beta_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\hypertarget{supplemental-materials}{%
+\section*{Supplemental Materials}\label{supplemental-materials}}
+\addcontentsline{toc}{section}{Supplemental Materials}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of annual fully-selected fishing mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
+Estimating models also assume the mean natural mortality parameter $\beta_\text{M} = \log 0.2$ and no environmental covariate effect $\beta_\text{Ecov} = 0$.}\label{F_bias_M_fixed_beta_fixed}
+\begin{center}
+\includegraphics[height = \textheight]{F_bias_all_PE_effect_M_fixed_beta_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of annual fully-selected fishing mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
+Estimating models also assume the mean natural mortality parameter $\beta_\text{M} = \log 0.2$ and the environmental covariate effect $\beta_\text{Ecov}$ is estimated.}\label{F_bias_M_fixed_beta_estimated}
+\begin{center}
+\includegraphics[height = \textheight]{F_bias_all_PE_effect_M_fixed_beta_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of annual fully-selected fishing mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
+Estimating models also assume both the mean natural mortality parameter $\beta_\text{M}$ and the environmental covariate effect $\beta_\text{Ecov}$ are estimated.}\label{F_bias_M_estimated_beta_estimated}
+\begin{center}
+\includegraphics[height = \textheight]{F_bias_all_PE_effect_M_estimated_beta_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
 \end{document}

### 2023-10-04 Ecov_study/mortality/code/M_example_plot.R

@@ -1,0 +1,60 @@
+library(here)
+library(ggplot2)
+library(dplyr)
+df.ems = readRDS(file.path(here(),"Ecov_study","mortality","inputs", "df.ems.RDS"))
+df.oms = readRDS(file.path(here(),"Ecov_study","mortality","inputs", "df.oms.RDS"))
+om_inputs <- readRDS(file.path(here(),"Ecov_study","mortality","inputs", "om_inputs.RDS"))
+
+Ecov_obs_sigs <- unique(df.oms$Ecov_obs_sig)
+Ecov_re_sigs <- unique(df.oms$Ecov_re_sig)
+Ecov_re_cors <- unique(df.oms$Ecov_re_cor)
+Ecov_effect <- unique(df.oms$Ecov_effect)
+PE <- c("rec", "rec+M")
+temp <- readRDS(file.path(here::here(),"Ecov_study","mortality", "results", "aggregated_results", paste0("om_", 1, "_em_", 1, ".RDS")))
+
+Ecov_types <- expand.grid(Ecov_re_sig = Ecov_re_sigs,Ecov_re_cor =Ecov_re_cors, Ecov_effect = Ecov_effect, process_error = PE)
+#par(mfcol = c(4,2), oma = c(5,5,0,0), mar = c(1,1,1,1))
+M_df <- data.frame(Ecov_re_sig = numeric(), Ecov_re_cor = numeric(), Ecov_effect = numeric(), process_error = character(), M = numeric())
+cn <- colnames(M_df)
+for(i in 1:NROW(Ecov_types)){
+   om <- which(df.oms$Ecov_re_sig == Ecov_types[i,1] & df.oms$Ecov_re_cor == Ecov_types[i,2] & df.oms$Ecov_effect == Ecov_types[i,3] & df.oms$NAA_M_re == Ecov_types[i,4])[1]
+   print(om)
+   res <- readRDS(file.path(here::here(),"Ecov_study","mortality", "results", "aggregated_results", paste0("om_", om, "_em_", 1, ".RDS")))
+   good_sims <- which(sapply(res, length) == 2)
+   tmp <- cbind.data.frame(Ecov_types[i,1], Ecov_types[i,2], Ecov_types[i,3], Ecov_types[i,4], res[[good_sims[1]]]$truth$MAA[,1])
+   colnames(tmp) <- cn
+   M_df <- rbind(M_df, tmp)
+}
[…]

### 2023-10-04 Ecov_study/mortality/code/temp.R

@@ -1,0 +1,75 @@
+tree_df_fn <- function(df.ems, df.oms, M_est = FALSE) {
+  for(i in 1:3) {
+    re_mod <- c("rec", "rec+1", "rec+M")[i]
+    #EM:  M fixed, Ecov_beta estimated, OM and EM RE assumption match
+    df.ems. <- df.ems[df.ems$Ecov_est,]
+    em_ind <- which(df.ems.$M_est == M_est & df.ems.$re_config == re_mod)
+    om_ind <- which(df.oms$NAA_M_re == re_mod) #om and em match
+    res <- lapply(all_beta_bias[om_ind], function(x) {
+      x <- x[[em_ind]]
+      x <- x[which(!is.na(x[,2])),1] #remove fits without standard error estimates
+      #x <- x[which(x[,2] < 100),1] #remove fits with bad (really big) standard error estimates
+#      print(x)
+      # out <- c(mean(x,na.rm=T), sd(x, na.rm=T)/sqrt(sum(!is.na(x))))
+      # out <- c(out, out[1] + qt(0.025, sum(!is.na(x))) * out[2])
+      # out <- c(out, out[1] + qt(0.975, sum(!is.na(x))) * out[2])
+      # out <- c(median(x, na.rm = TRUE), sd(x, na.rm=T)/sqrt(sum(!is.na(x))), custom_boxplot_stat(x))#quantile(x, probs = c(0.025,0.975)))
+      # return(out)
+      return(x)
+    })
+    res <- reshape2::melt(res)
+    res <- cbind.data.frame(df.oms[om_ind[res[,2]],], error = res[,1])
+    if(i == 1) {
+      all_res <- res
+    } else {
+      all_res <- rbind.data.frame(all_res, res)
+    }
+  }
+  facs <- c("Ecov_obs_sig", "Fhist", "Ecov_re_sig","Ecov_re_cor", "obs_error", "NAA_M_re")
+  all_res[facs] <- lapply(all_res[facs], factor)
+  all_res_mod <- all_res %>%
+    mutate(Ecov_obs_sig = recode(Ecov_obs_sig,
+      "0.1" = "sigma[ecov] == 0.1",
+      "0.5" = "sigma[ecov] == 0.5"
+    ))
+  # all_res_mod <- all_res_mod %>%
+  #   mutate(Ecov_re_sig = recode(Ecov_re_sig,
[…]

### 2023-10-24 Ecov_study/mortality/paper/paper.Rmd
saving work on Ecov-M

@@
 
-We found high accuracy for estimation of the mean natural mortality rate parameter ($\beta_M$) for all EM process errors assumptions when OMs had step changes in fishing mortality, lower uncertainty in index and age composition observations, and either R or R+M process errors (Figure \ref{mean_M_bias}).
-The most variation in estimates occurred when fishing and Higher uncertainty in index and age composition observations.
+We found high accuracy for estimation of the mean natural mortality rate parameter ($\beta_M$) for all EM process errors assumptions when OMs had step changes in fishing mortality, lower uncertainty in index and age composition observations, and either R or R+M process errors (Figure \ref{mean_M_bias_Ecov_beta_fixed}).
+The most variation in estimates occurred when fishing mortality was constant and there was higher uncertainty in index and age composition observations.
 For OMs with R+S process errors, the most reliable estimation of $\beta_M$ was obtained when the EM also assumed R+S process errors across all other factors.
@@
 \begin{figure}
-\caption{Median relative error of the estimated mean natural mortality parameter $\beta_\text{M}$ from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.}\label{mean_M_bias}
+\caption{Median relative error of the estimated mean natural mortality parameter $\beta_\text{M}$ from fitting simulated observations from each operating model with the estimating model assuming $\beta_\text{Ecov} = 0$ under alternative process error assumptions.}\label{mean_M_bias_Ecov_beta_fixed}
 \begin{center}
-\includegraphics[height = \textheight]{mean_M_bias_all_PE_effect.png}
+\includegraphics[height = \textheight]{mean_M_bias_all_PE_effect_Ecov_beta_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of the estimated mean natural mortality parameter $\beta_\text{M}$ from fitting simulated observations from each operating model with the estimating model estimating $\beta_\text{Ecov} = 0$ under alternative process error assumptions.}\label{mean_M_bias_Ecov_beta_estimated}
+\begin{center}
+\includegraphics[height = \textheight]{mean_M_bias_all_PE_effect_Ecov_beta_estimated.png}
 \end{center}

### 2023-10-24 Ecov_study/mortality/paper/paper.tex
saving work on Ecov-M

@@
 \begin{figure}
-\caption{Median relative error of the estimated mean natural mortality parameter $\beta_\text{M}$ from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.}\label{mean_M_bias}
-\begin{center}
-\includegraphics[height = \textheight]{mean_M_bias_all_PE_effect.png}
+\caption{Median relative error of the estimated mean natural mortality parameter $\beta_\text{M}$ from fitting simulated observations from each operating model with the estimating model assuming $\beta_\text{Ecov} = 0$ under alternative process error assumptions.}\label{mean_M_bias_Ecov_beta_fixed}
+\begin{center}
+\includegraphics[height = \textheight]{mean_M_bias_all_PE_effect_Ecov_beta_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of the estimated mean natural mortality parameter $\beta_\text{M}$ from fitting simulated observations from each operating model with the estimating model estimating $\beta_\text{Ecov} = 0$ under alternative process error assumptions.}\label{mean_M_bias_Ecov_beta_estimated}
+\begin{center}
+\includegraphics[height = \textheight]{mean_M_bias_all_PE_effect_Ecov_beta_estimated.png}
 \end{center}

### 2023-10-24 Ecov_study/mortality/code/aic_ecov_effect_results.R

@@ -345,8 +345,8 @@ make_df_fn <- function(M_est = TRUE){
         "0.5" = "rho[Ecov] == 0.5"
     ))
   all_res_mod <- all_res_mod %>% mutate(Fhist = recode(Fhist,
-      "H-MSY" = "High->FMSY",
-      "MSY" = "FMSY"))
+      "H-MSY" = "2.5*F[MSY] %->% F[MSY]",
+      "MSY" = "F[MSY]"))
   all_res_mod <- all_res_mod %>%
     mutate(beta_ecov_fixed = recode(Ecov_est,
       "TRUE" = "beta[Ecov]==0",
@@ -391,8 +391,7 @@ plt <- ggplot(all_res_mod, aes(x = Ecov_effect, y = n, fill = correct)) + scale_
     geom_col(position = "fill") + 
     #geom_col(position = position_dodge(0.1)) + 
     facet_grid(Ecov_obs_sig+Ecov_re_sig+Ecov_re_cor ~ NAA_M_re+Fhist+obs_error, 
-      labeller = labeller(Ecov_re_sig = label_parsed, Ecov_re_cor = label_parsed, obs_error = label_wrap_gen(width = 15), Ecov_obs_sig = label_parsed)) +
-    #facet_grid(Ecov_obs_sig+Ecov_re_sig+Ecov_re_cor ~ NAA_M_re+Fhist+obs_error, labeller = label_wrap_gen(width = 15)) + #, labeller = label_parsed) + 
+      labeller = labeller(Ecov_re_sig = label_parsed, Ecov_re_cor = label_parsed, obs_error = label_wrap_gen(width = 15), Ecov_obs_sig = label_parsed, Fhist = label_parsed)) +
     theme_bw() + coord_cartesian(ylim = c(0, 1)) + ylab("Proportion ranked best") + xlab(expression(beta[Ecov])) + labs(fill = "EM assumption") +
     ggtitle(bquote(beta[M]~estimated)) + theme(plot.title = element_text(hjust = 0.5))
 plt
@@ -407,8 +406,7 @@ plt <- ggplot(temp, aes(x = Ecov_effect, y = n, fill = re_config:beta_ecov_fixed
     geom_col(position = "fill") + 
     #geom_col(position = position_dodge(0.1)) + 
     facet_grid(Ecov_obs_sig+Ecov_re_sig+Ecov_re_cor ~ NAA_M_re+Fhist+obs_error, 
[…]

### 2023-10-24 Ecov_study/mortality/code/make_results_files.R

@@ -81,7 +81,7 @@ saveRDS(ecov_beta_bias_res, file = here::here("Ecov_study","mortality", "results
 
 #bias of beta_M (mean M)
 mean_M_bias_res = lapply(1:NROW(df.oms), function(y){
-  lapply(which(df.ems$Ecov_est), function(x) { #only the estimating models that estimate Ecov_beta
+  lapply(which(df.ems$M_est), function(x) { #only the estimating models that estimate Ecov_beta
   #lapply(4:4, function(x) {
     res <- readRDS(here::here("Ecov_study","mortality", "results", "aggregated_results", paste0("om_", y, "_em_", x, ".RDS")))
     print(paste(y, x))
@@ -106,6 +106,108 @@ mean_M_bias_res = lapply(1:NROW(df.oms), function(y){
   })
 })
 saveRDS(mean_M_bias_res, file = here::here("Ecov_study","mortality", "results", "mean_M_bias_results.RDS"))
+
+#relative error of F40 (prevailing)
+F40_rel_error_res = lapply(1:NROW(df.oms), function(y){
+  lapply(1:NROW(df.ems), function(x) { #only the estimating models that estimate Ecov_beta
+  #lapply(4:4, function(x) {
+    res <- readRDS(here::here("Ecov_study","mortality", "results", "aggregated_results", paste0("om_", y, "_em_", x, ".RDS")))
+    print(paste(y, x))
+  #res = sapply(1:20, function(x) {
+    bias <- t(sapply(res, function(z) { 
+      #fit = try(readRDS(here::here("Ecov_study","mortality", "results", paste0("om", y), paste0("sim", x, "_em", z, ".RDS"))), silent = TRUE)
+      out <- rep(NA,3) #bias, se, true_in_ci (0/1)
+      if(!is.character(z)) if(!is.null(z$fit$opt)) {
+        out[1] = exp(z$fit$rep$log_FXSPR_static[1]-z$truth$log_FXSPR_static[1])-1
+        if(!is.null(z$fit$sdrep)){
+          ind <- which(!is.na(z$fit$sdrep$SE_rep$log_FXSPR_static))[1]
+          if(length(ind)){
+            out[2] <- z$fit$sdrep$SE_rep$log_FXSPR_static[ind] #se
+            ci = z$fit$rep$log_FXSPR_static[1] + c(-1,1)*qnorm(0.975) * out[2]
[…]

### 2023-10-24 Ecov_study/mortality/code/mean_M_bias_results.R

@@ -14,12 +14,12 @@ custom_boxplot_stat <- function(x){#, n.yrs = 1, n.sim = 100) {
   names(r) <- c("ymin", "lower", "middle", "upper", "ymax")
   r
 }
-plot_df_fn <- function(df.ems, df.oms) {
+plot_df_fn <- function(df.ems, df.oms, Ecov_est = FALSE) {
   for(i in 1:3) {
     re_mod <- c("rec", "rec+1", "rec+M")[i]
     #EM:  M fixed, mean_M estimated, OM and EM RE assumption match
-    df.ems. <- df.ems[df.ems$Ecov_est,]
-    em_ind <- which(df.ems.$M_est == TRUE & df.ems.$re_config == re_mod)
+    df.ems. <- df.ems[df.ems$M_est,]
+    em_ind <- which(df.ems.$Ecov_est == Ecov_est & df.ems.$re_config == re_mod)
     om_ind <- which(df.oms$NAA_M_re == re_mod) #om and em match
     res <- t(sapply(mean_M_bias[om_ind], function(x) {
       x <- x[[em_ind]]
@@ -76,27 +76,40 @@ plot_df_fn <- function(df.ems, df.oms) {
   return(all_res_mod)
 }
 
-all_res <- plot_df_fn(df.ems, df.oms)
+all_res <- plot_df_fn(df.ems, df.oms, Ecov_est = FALSE)
+plt <- ggplot(all_res, aes(x = Ecov_effect, y = bias_est, colour = Ecov_re_sig:Ecov_re_cor)) + scale_colour_viridis_d() + 
+    geom_hline(aes(yintercept=0), linewidth = 2, linetype = "dashed", colour = "grey") +
+    geom_line(position = position_dodge(0.1), linewidth = 1) + geom_point(position = position_dodge(0.1), size = 4) + 
+    facet_grid(Ecov_obs_sig + obs_error ~ NAA_M_re + Fhist, labeller = labeller(obs_error = label_wrap_gen(width = 35), Ecov_obs_sig = label_parsed)) +
+    theme_bw() + coord_cartesian(ylim = c(-1, 1)) + ylab(bquote(Median~Bias~of~beta[M])) + xlab(expression(beta[Ecov])) +
+    ggtitle(bquote(EM: beta[Ecov]==0)) + theme(plot.title = element_text(hjust = 0.5)) + labs(colour = expression(sigma[Ecov]:rho[Ecov])) +
[…]

### 2023-10-24 common_code/spr_functions.R

@@ -1,0 +1,77 @@
+get_SPR = function(F, M, sel, mat, waassb, fracyrssb, at.age = FALSE)
+{
+  n_ages = length(sel)
+  SPR = numeric()
+  n = 1
+  F = F * sel
+  Z = F + M
+  for(a in 1:(n_ages-1))
+  {
+    SPR[a] = n[a] * mat[a] * waassb[a] * exp(-fracyrssb * Z[a])
+    n[a+1] = n[a] * exp(-Z[a])
+  }
+  n[n_ages] = n[n_ages]/(1-exp(-Z[n_ages]))
+  SPR[n_ages] = n[n_ages] * mat[n_ages] * waassb[n_ages] * exp(-fracyrssb * Z[n_ages])
+  if(at.age) return(SPR)
+  else return(sum(SPR))
+}
+
+
+#-------Y/R -----------------------------
+get_YPR_fleets = function(FAA,M, waacatch, at.age = FALSE)
+{
+  n_ages = dim(FAA)[2]
+  n_fleets <- dim(FAA)[1]
+  YPR = matrix(0,n_fleets, n_ages)
+  #F = F * sel
+  Z = apply(FAA,2,sum) + M
+  for(f in 1:n_fleets) {
+    n = 1
+    for(a in 1:(n_ages-1)){
+      YPR[f,a] <- n[a] * FAA[f,a] * waacatch[f,a] * (1.0 - exp(-Z[a]))/Z[a]
+      n[a+1] = n[a] * exp(-Z[a])
+    }
+    n[n_ages] = n[n_ages]/(1 - exp(-Z[n_ages]))
+    YPR[f,n_ages] = n[n_ages] * FAA[f,n_ages] * waacatch[f,n_ages] * (1.0 - exp(-Z[n_ages]))/Z[n_ages]
+  }
+  if(at.age) return(YPR) #by fleet and age
+  else return(apply(YPR,1,sum)) #by fleet
+}
+
+get_SPR_BRPS_fn <- function(mod, spr_yrs, percent){
+  dat = mod$env$data
+  if(missing(percent)) percent <- dat$percentSPR
+  if(missing(spr_yrs)) spr_yrs <- dat$avg_years_ind+1 #c++
+  R_yrs <- dat$XSPR_R_avg_yrs+1
+  R_type <- dat$XSPR_R_opt
+  if(R_type %in% c(1,3)) Rxspr <- mean(mod$rep$NAA[Ryrs,1])
[…]

### 2023-12-01 Ecov_study/catchability/Catchability_ecov_sims.Rmd

@@
 
-Load packages & source functions used in simulation testing
+# Load packages & source functions used in simulation testing
 ```{r}
@@
 source(here::here("Ecov_study","catchability","simOM.R"))
-#source(here::here("Other_Research", "simOM.R"))
 source(here::here("Ecov_study","catchability","simTestWHAM.R"))
-#source(here::here("Other_Research", "simTestWHAM.R"))
+
+## Processing functions
+source(here::here("Ecov_study", "catchability", "postprocess_simTestWHAM.R"))
+source(here::here("Ecov_study", "catchability", "combinePerfMet.R"))
+source(here::here("Ecov_study", "catchability", "plotResults.R"))
 ```
@@
 # Set number of simulations to run
-nsim <- 2 # 2
+nsim <- 2
 
 # Subset of OMs to run
-subsetOM <- OMsetup %>% filter(F_hist == "H-L") %>% filter(Ecov_effect != 0) %>%
-  filter(OMname >100)
+subsetOM <- OMsetup %>% filter(F_hist == "H-L") %>% filter(Ecov_effect != 0) %>% tail(n=3)# %>% filter(OMname >100)
 # %>% head(n=10)
-subsetOM <- subsetOM[8:nrow(subsetOM),] #7?
+# subsetOM <- subsetOM[8:nrow(subsetOM),] #7?
 
@@
 ```{r}
-# ID all simulation result files 
- files <- list.files(path = here::here(), pattern = "simWHAM_", recursive = TRUE)
+# Find all result files
+filenames1 <- list.files(path = here::here("Ecov_study/catchability/remote1_Results"), pattern = "simWHAM_", recursive = TRUE, full.names = TRUE)
+filenames2 <- list.files(path = here::here("Ecov_study/catchability/remote2_Results"), pattern = "simWHAM_", recursive = TRUE, full.names = TRUE)
+filenames3 <- list.files(path = here::here("Ecov_study/catchability/remote3_Results_supplementalRuns"), pattern = "simWHAM_", recursive = TRUE, full.names = TRUE) # simulations to fill gaps so 100 simulations for each OM/EM pair
 
-perfMet <- postprocess_simTestWHAM(filenames = files)
+# Set storage directory
+outdir = here::here("Ecov_study/catchability")
 
+# Post-process results
+postprocess_simTestWHAM(filenames = c(filenames1, filenames2, filenames3), outdir = outdir)
+
+# Try running in the console if this causes an "Error... cannot open the connection) 
 ```
@@
 ```{r}
+# ID all processed performance metric files in "Results" folder
+ files <- list.files(path = here::here(), pattern = "perfMet_", recursive = TRUE)
+
+combinePerfMet(filenames = files, outdir = here::here("Ecov_study", "catchability", "Results"))
 
@@
 ### Plot results
+168 Ecov has 150 simulation
+192?
+2 qRandEcov only has 50 - fixed
+49 qRandEcov only has 50 - fixed
+192 qRand, qRandEcov only has 50
 ```{r}
-plotResults(results = perfMet, convergedONLY = TRUE)
+perfMet <- readRDS(here::here("Ecov_study", "catchability", "perfMet_2023-11-29_13-57-01.735372.RDS"))
+  # readRDS(here::here("Ecov_study", "catchability", "perfMet_2023-11-28_09-32-17.294206.RDS"))
+
+library(TAF)
+mkdir(here::here("Ecov_study", "catchability", "plots"))
+
+plotResults(results = perfMet, convergedONLY = TRUE, outfile = here::here("Ecov_study", "catchability", "plots"))
 ```

### 2024-01-18 Ecov_study/recruitment_functions/Workflow_Ecov_Recruitment_Project.Rmd

---
title: "Workflow for Ecov Recruitment Analysis"
output: html_notebook
---

<!-- This is an [R Markdown](http://rmarkdown.rstudio.com) Notebook. When you execute code within the notebook, the results appear beneath the code.  -->

<!-- Try executing this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Ctrl+Shift+Enter*.  -->

# Project 3 for the SSRTWG investingating the incorporation of environmental covariates into WHAM recruitment models.

1. Create dataframes that have AIC info and convergence info: run **AIC_dataframe.R** to produce
   + *AIC_all.RDS* contains AIC and convergence info for the best EM only (lowest AIC) per OM-Sim
   + *AIC_weight.RDS* contains AIC, convergence, model rank, model weight, model probability info for all EMs per OM-Sim

2. Specify convergence tolerances/thresholds in **convergence_summaries.R**
   + bad.grad.value : cut-off for identifying parameters with abs(gradient)>bad.grad.value (parameter with max value and 2nd max value as well as parameter name are saved, not *all* parameters in this category)
   + bad.se.value : cut-off for identifying parameter where SE_par_max>bad.se.value (parameter with max value and 2nd max value as well as parameter name are saved, not *all* parameters in this category)
   + given this, output is produced using bad.grad.value and bad.se.value appended to plot and table names to identify the tolerances
     + plots of number of bad models (from the best set, and from all EMs per OM-Sim)
     + plots of parameters that failed the bad.grad or bad.se criteria

3. Generate regression trees to characterize factors influencing model identifiability; run **AIC_regression.R**
   + ...

4. Look at bias in SSB, Recr, Fbar
   + ...

5. Look at bias in SR_a, SR_b, sigmaR, rho_R, sigmaEcov, rho_Ecov
   + ...

-----

# Figures

-----

<!-- ![Number of bad models among the set of 'best' (lowest) AIC per OM-Sim ](figures/img.png) -->

<!-- ![Parameters with largest gradient or SE ](figures/img.png) -->

<!-- ```{r} -->
<!-- plot(cars) -->
<!-- ``` -->

<!-- Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*. -->

<!-- When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file). -->

<!-- The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed. -->

### 2024-01-19 Ecov_study/mortality/paper/paper.Rmd
good drafts of project 0 and Ecov M WPs complete

@@
 ---
-title: Factors affecting inferences on natural mortality and associated environmental effects
+title: Factors affecting inferences on natural mortality and associated environmental effects in state-space age-structured assessment models
 author: 
@@
 
-## Summary/Abstract {-}
+## Summary {-}
 
-We completed a large scale simulation study with 288 operating models with 12 estimated models fit to each.
-The factors defining operating model configuration inlcuded the source of process error on the population (recruitment, recruitment and survival, recruitment and natural mortality), the degree of temporal variation and autocorrelation of the environmental covariate, the uncertainty in the observation of the covariate, the uncertainty in indices and age composition observations, fishing history, and the magnitude of the effect of the covariate on natural mortality.
+We completed a large scale simulation study with 288 operating models, each with 100 simulated data sets, and 12 estimated models fit to each simulated data set.
+The factors defining operating model configuration included the source of process error on the population (recruitment, recruitment and survival, recruitment and natural mortality), the degree of temporal variation and autocorrelation of the environmental covariate, the uncertainty in the observation of the covariate, the uncertainty in indices and age composition observations, fishing history, and the magnitude of the effect of the covariate on natural mortality.
+The estimating models make alternative assumptions on whether to include the environmental effect, whether the mean/intercept log natural mortality (log(0.2)) is estimated or known, and whether process errors are on just recruitment, recruitment and survival, or recruitment and natural mortality.
 
@@
 Estimating models with process errors on recruitment and survival were unlikely to converge when the process errors in the operating model did not match whereas estimating models with process errors in recruitment and natural mortality converged for operating models without this match in certain cases.
-Probability of convergence generally decreased when the mean natural mortality rate parameter was estimated.
+Probability of convergence generally decreased when the mean/intercept log-natural mortality rate parameter was estimated.
 
-AIC paragraph
+Whether the mean log-natural mortality was estimated or not, the best accuracy of AIC for model selection occurred for models with process errors on recruitment and survival.
+AIC accuracy was poor for models with process errors on recruitment and natural mortality.
+estimating the mean natural mortality rate had small effects on the accuracy of AIC in selecting the appropriate process error.
+Estimating the mean log-natural mortality resulted in a small decrease in AIC accuracy.
+AIC was conservative for determining whether the environmental covariate affected natural mortality.
+AIC was very accurate in determining no effect when there was no effect in the operating model, but AIC often ranked the null model best when there was an effect.
+Accuracy of AIC for covariate effects improved with increased effect size, increased temporal contrast in the covariate, and lower uncertainty in observations.
 
-Bias paragraph
+We found no evidence of bias in estimation of environmental effects regardless of process error assumptions when there was low uncertainty in the environmental observations and large constrast in the environmental covariate.
+In most cases the relative error of the estimated environmental effect did not depend on the source of process error assumed in the estimating model.
+The worst bias was observed when OMs assumed R+S process errors, high uncertainty in covariate observations, low variability in the covariate, and low uncertainty in index and age composition observations.
+Simultaneously estimating the mean/intercept log natural mortality resulted in larger variation in the relative errors of the estimated environmental effect.
+Estimation of the intercept was reliable for all EM process error assumptions when the operating models assumed process errors on recruitment and natural mortality, contrast in fishing pressure over time, and lower observation error.
+Estimating the mean/intercept for log natural mortality generally resulted in highly variable estimates of annual natural mortality and spawning biomass and evidence of bias for some operating and estimation model assumptions about process error source.
+Again reliability of annual natural mortality estimates was generally improved with lower observation error uncertainty and contrast in fishing pressure.
 
-These results suggest that models with covariate effects on M will be most reliable when R+S is true, large contrast in the 
-Recommendations
-### Keywords {-}
+Reliable detection of covariate effects requires informative data.
+AIC preferred simpler models than the true model when information content in data and contrast in covariates and abundance were low.
+The null model for environmental covariate effects (no covariate effect) was selected when contrast in the time series was low and/or uncertainty in observations was high.
+The selection of the null model by AIC also likely decreases with strength of the effect of the covariate on M.
+Similarly, when there was process error in recruitment and natural mortality, estimation models with process error only in recruitment were preferred presumably due to low variation in simulated natural mortality process errors.
+Covariate effect estimation can be robust to process error assumptions with high contrast in covariate and low observation error.
 
@@
 
-We used the WHAM package \citep[][version 1.0.7, commit 77bbd94]{millerstock20,stockmiller21} to configure all operating models and estimation models.
-This packages has also been used to configure operating and estimating models for closed loop simulations evaluating index-based assessment methods \citep{legaultetal23} and is used for management of GB haddock, butterfish, plaice, bluefish, and Atlantic cod stocks in the Northeast US.
+We used the Woods Hole Assessment Model (WHAM) to configure operating and estimation models in our simulation study \citep[]{millerstock20,stockmiller21}.
+WHAM is an R package freely available as a github repository.
+For this study we used [version 1.0.6.9000, commit 77bbd94](https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3).
+This packages has also been used to configure operating and estimating models for closed loop simulations evaluating index-based assessment methods \citep{legaultetal23} and is used for management of haddock, butterfish, American plaice, bluefish, Atlantic cod, and black sea bass in the Northeast US.
 
 We completed a simulation study with a number of operating models.
-Many of the characteristics of the operating models are the same as those used by \citet{milleretal_project0_wp}.
+Many of the characteristics of the operating models are the same as those used in the working paper by Miller et al.
+(WP 4).
 The factors defining the configuration of each operating model which are described in detail in subsequent sections include
@@
 Simulations were all carried out on the University of Massachusetts Green High-Performance Computing Cluster.
-Code for completing the simulations and summarization of results can be found at github.com/timjmiller/SSRTWG/ecov_study/mortality.
+Code for completing the simulations and summarizing results can be found at https://github.com/timjmiller/SSRTWG/ecov_study/mortality.
 
@@
 The population model tracks 10 age classes: ages 1 to 10+ and we assume spawning occurs 1/4 of the way through the year.
-The maturity at age was a logistic curve with a50 = 2.89 and slope = 0.88 and assumed known in all estimation models (Figure \ref{om_maturity}).
+The maturity at age was a logistic curve with $a_{50}$ = 2.89 and slope = 0.88 and assumed known in all estimation models (Figure \ref{om_maturity}).
 
@@
 Therefore steepness and unfished recruitment are also constant over the time period for those operating models \citep{millerbrooks21}.
-We specified unfished recruitment = $R_0 = e^{10}$ and $\Fmsy = \Fspr[40] = 0.348$ equated to a steepness of 0.69 and $\alpha=0.60$ and $\beta = 2.4 \times 10^{-5}$ for the
+We specified unfished recruitment = $R_0 = e^{10}$ and $\Fmsy = F_{40\%} = 0.348$ equated to a steepness of 0.69 and $\alpha=0.60$ and $\beta = 2.4 \times 10^{-5}$ for the
 \[
@@
 The first measure of reliability we investigated was frequency of convergence when fitting each estimating model to the simulated data sets.
-There are various ways to assess convergence of the fit.
-We summarized 5 alternative categories of convergence:
-\begin{enumerate}
-\item Did the optimization routine (stats::nlminb) complete without error?
-\item Did the stats::nlminb convergence flag = 0 indicate successful convergence?
-\item Was the maximum absolute value of the gradient of the log-likelihood < $1\times10^{-6}$?
-\item Did TMB::sdreport provide non-NA values for all fixed effects standard errors?
-\item Did TMB::sdreport provide all standard errors < 100?
-\end{enumerate}
-The first convergence criterion assesses whether the model crashes.
-The third is a measure that assesses how flat the likelihood is at the optimized point of the likelihood surface.
-The fourth and fifth criteria are specific to the hessian-based standard error reporting provided by TMB.
-the TMB::sdreport function will sometimes return standard error estimates even when the calculated hessian is not invertible (fourth criterion).
-It will also provide large standard errors for some parameters that are not estimated well or are near bounds on the transformed scale that is of primary interest.
-For example, variance parameters for random effects can often be estimated near 0 when the fit suggests no variation in the random effects or, equivalently, that the model without random effects is better.
+There are various ways to assess convergence of the fit, but we defined successful convergence as the hessian of the marginal log-likelihood being invertible and providing variance estimates for the fixed effects parameters.
 
@@
 
-We focused on the fourth convergence criteria, that the hessian of the negative-marginal log-likelihood at the optimized parameters was invertible.
-When the (mean) natural mortality rate was assumed at the true value, corresponding to usual practice in application of assessment models, good convergence for all EMs was observed for R+S operating models with \Fmsy fishing history, more variation in the latent environmental covariate ($\sigma_\text{Ecov} = 0.5$), and lower error in the associated observations $\sigma_\text{ecov} = 0.1$ (Figure \ref{convergence_M_fixed}).
+When the (mean) log-natural mortality rate was assumed at the true value, corresponding to usual practice in application of assessment models, good convergence for all EMs was observed for R+S operating models with \Fmsy fishing history, more variation in the latent environmental covariate ($\sigma_\text{Ecov} = 0.5$), and lower error in the associated observations $\sigma_\text{ecov} = 0.1$ (Figure \ref{convergence_M_fixed}).
 Good convergence for all EMs was also observed for R+S OMs with the step change in fishing history, but with low error in indices and age comp observations.
 
-EMs that assumed no survival or natural mortality random effects (R OMS), nearly always converged across all operating models that assumed more variation in the latent environmental covariate ($\sigma_\text{Ecov} = 0.5$) and lower error in the associated observations $\sigma_\text{ecov} = 0.1$.
-EMs that assumed recruitment and survival random effects (R+S OMS), had poor converge probability when the OMs had alternative process errors (R or R+M).
+EMs that assumed random effects just on recruitment (R), nearly always converged across all operating models that assumed more variation in the latent environmental covariate ($\sigma_\text{Ecov} = 0.5$) and lower error in the associated observations $\sigma_\text{ecov} = 0.1$.
+EMs that assumed recruitment and survival random effects (R+S), had poor converge probability when the OMs had alternative process errors (R or R+M).
 The R+S EMs showed best convergence for the R+S OMs where there was lower error in the environmental observations $\sigma_\text{ecov} = 0.1$ or higher error in environmental observations of latent true environmental covariates with greater temporal variation.
-EMs that assumed recruitment and natural mortality random effects (R+M OMs), had poor convergence probability for all OMs that had a step change in fishing history and higher uncertainty in indices and age composition.
+EMs that assumed recruitment and natural mortality random effects (R+M), had poor convergence probability for all OMs that had a step change in fishing history and higher uncertainty in indices and age composition.
 
@@
 
-We present results for error in annual natural mortality rate conditional on three alternative EM configurations for the natural mortality parameters: 1) mean natural mortality rate parameter is fixed at the true value ($\beta_M = log(0.2)$) and no covariate effect is assmed ($\beta_\text{Ecov} = 0$),  2) $\beta_M = log(0.2)$ and $\beta_\text{Ecov}$ is estimated, and 3) both $\beta_M$ and $\beta_\text{Ecov}$ are estimated.
+We present results for error in annual natural mortality rate conditional on three alternative EM configurations for the natural mortality parameters: 1) mean natural mortality rate parameter is fixed at the true value ($\beta_M = log(0.2)$) and no covariate effect is assumed ($\beta_\text{Ecov} = 0$),  2) $\beta_M = log(0.2)$ and $\beta_\text{Ecov}$ is estimated, and 3) both $\beta_M$ and $\beta_\text{Ecov}$ are estimated.
 
@@
 R+M EMs were the only EMs to exhibit differences in the sign of the median errors across the time series.
-When OMs had a step change in fishing mortality and higher uncertainty in index and age composition observations, we observed positive median errrors at the beginning of the time series and lower median errors in later years.
+When OMs had a step change in fishing mortality and higher uncertainty in index and age composition observations, we observed positive median errors at the beginning of the time series and lower median errors in later years.
 However, whether there was evidence of bias, depended on the uncertainty in covariate observations and the degree of variability in the latent covariate.
@@
 
-When $\beta_\text{Ecov}$ was estimated, R EMs performed worse for R+S OMs with evidence of negative bias for OMs with cosntant fishing mortality, lower uncertainty in index and age composition observations, the largest covariate effect size, lower variability in the latent covariate, and larger uncertainty in the covariate observations (Figure \ref{annual_M_bias_M_fixed_beta_estimated}).
+When $\beta_\text{Ecov}$ was estimated, R EMs performed worse for R+S OMs with evidence of negative bias for OMs with constant fishing mortality, lower uncertainty in index and age composition observations, the largest covariate effect size, lower variability in the latent covariate, and larger uncertainty in the covariate observations (Figure \ref{annual_M_bias_M_fixed_beta_estimated}).
 There was little difference in the results for R+S EMs whether $\beta_\text{Ecov}$ was estimated or not.
@@
 Allowing the EMs to estimate $\beta_M$ resulted in very large variability in estimates for all EMs for R and R+M OMs with constant fishing mortality and higher uncertainty in index and age composition observations (Figure \ref{annual_M_bias_M_estimated_beta_estimated}).
-The same varibilities occurred for R+S OMs, but strong bias was estimated for R and R+M EMs.
+The same variability occurred for R+S OMs, but strong bias was estimated for R and R+M EMs.
 Annual M estimation was most reliable when OMs had step changes in fishing mortality and lower uncertainty in index and age composition observations.
@@
 Null model for environmental covariate effect (no covariate effect) was selected when contrast in the time series was low  and/or uncertainty in observations was high.
-
+Null selection likely decreases with strength of the effect on M.
+We only examined two non-zero effect sizes: 0.25, and 0.5, but our results suggest larger effect sizes, with the same observation error and contrast in time series, would allow better AIC performance for determining covariate effects.
 When there was process error in recruitment and M  (R+M), models with process error only in recruitment were preferred.
@@
 The marginal variance for log M random effects was 0.3.
-
-Null selection likely decreases with strength of the effect on M.
-We only examined two non-zero effect sizes: 0.25, and 0.5, but our results suggest larger effect sizes, with the same observation error and contrast in time series, would allow better AIC performance for determining covariate effects.
-
-Covariate effect estimation can be robust to process error assumptions (For R+M OMs) with contrast in covariate and low observation error.
-
 Reliable estimation of environmental effect on M regardless of the process error assumed by the EM as long as the contrast in the covariate is sufficient and the uncertainty in the observations is low.

### 2024-01-19 Project_0/paper/paper.tex
good drafts of project 0 and Ecov M WPs complete

@@
   pdftitle={Factors affecting reliablity of state-space age-structured assessment models},
-  pdfauthor={Timothy J. Miller1; A. Applegate; G. Britten; E. N. Brooks; G. Fay; A. Hansell; C. M. Legault; B. Muffley; J. Wiedenmann},
+  pdfauthor={Timothy J. Miller1; Andrew Applegate; Greg Britten; Elizabeth N. Brooks; Gavin Fay; Alex Hansell; Christopher M. Legault; Brandon Muffley; Brian C. Stock; John Wiedenmann},
   hidelinks,
@@
 assessment models}
-\author{Timothy J. Miller\textsuperscript{1} \and A. Applegate \and G.
-Britten \and E. N. Brooks \and G. Fay \and A. Hansell \and C. M.
-Legault \and B. Muffley \and J. Wiedenmann}
+\author{Timothy J. Miller\textsuperscript{1} \and Andrew
+Applegate \and Greg Britten \and Elizabeth N. Brooks \and Gavin
+Fay \and Alex Hansell \and Christopher M. Legault \and Brandon
+Muffley \and Brian C. Stock \and John Wiedenmann}
 \date{}
@@
 
-\hypertarget{abstract}{%
-\subsection*{Abstract}\label{abstract}}
-\addcontentsline{toc}{subsection}{Abstract}
-
-\hypertarget{keywords}{%
-\subsubsection*{Keywords}\label{keywords}}
-\addcontentsline{toc}{subsubsection}{Keywords}
+\hypertarget{summary}{%
+\subsection*{Summary}\label{summary}}
+\addcontentsline{toc}{subsection}{Summary}
+
+We conducted a simulation study where we simulated process errors and
+observations for 72 operating models with alternative assumptions about
+fishing history, degree of uncertainty in index and age composition
+observations, type (recruitment, survival, fishery selectivity,
+catchability and natural mortality), degree of variation, and
+correlation of process errors. We fit 20 different estimating models to
+each of 100 simulated set of observations with alternative assumptions
+type and correlation structure of process errors, (mean) natural
+mortality was known or estimated, and a B-H stock recruit relationship
+was assumed or not.
+
+Across simulations we summarized probability of convergence of fitted
+models, accuracy of marginal AIC in determining the correct process
+error assumption and it ability to determine the Beverton-Holt stock
+recruit relationship, bias in annual spawning stock biomass, in
+estimation of natural mortality, and in stock-recruit relationship
+parameters, and severity of retrospective patterns for estimation
+models.
+
+Alternative measures of convergence performed differently. Invertible
+hessians and resulting standard error estimation was possible when
+criteria based on the gradient of the optimized log-likelihood with
+respect to the fixed effects parameters. Using hessian-based
+convergence, probability of convergence was best for models that assumed
+the correct source of process error, assumed M was known, and did not
+assume stock-recruit relationships.
+
+Using marginal AIC provided most accurate inferences about the process
+errors on recruitment and survival and selectivity, in that AIC
+preferred EMs with assumed process errors that matched OMs, and EMs with
+those assumed process errors were not preferred when alternative process
+errors were assumed in the OMs. However, when the true process errors
+were more variable, AIC accuracy increased to a useful level. We found
+AIC more accurately determined a B-H stock recruit relationship rather
+than the null model without a S-R relationship when there was low
+variability in recruitment, low variability in survival random effects,
+and higher variation in spawning biomass over the time series. When the
+(mean) natural mortality rate was estimated, we found large bias and
+uncertainty much more likely in model output such as spawning stock
+biomass.
+
+Bias in spawning stock biomass estimation was generally low for
+estimating models that assumed the correct source of process error when
+there was lower observation error. Reliable estimation of stock-recruit
+relationship parameters only appears possible in ideal situations with
+lower observation errors in age composition and indices, lower
+variability in recruitment process errors and large contrast in spawning
+biomass over time. We found little evidence of bias for many OM process
+error assumptions when there was contrast in fishing pressure even when
+there was greater observation error although it can lead to more
+variable estimates of natural mortality. For OMs where there was bias in
+natural mortality due to high observation error, estimating the
+stock-recruit relationship seemed to remove the bias. However,
+estimation of natural mortality can cause large differences between the
+true and estimated SSB (that may be unbiased on average) when there is
+less contrast in fishing pressure over time and higher observation
+error.
+
+Retrospective patterns were generally weak for all estimation models
+regardless of the true source of process error, but they can be expected
+for recruitment even for the correct process error assumptions when
+observation error is high. When models did exhibit some retrospective
+pattern, estimating the mean natural mortality rate tended to remove it.
 
@@
 Application of state-space models in fisheries stock assessment and
-management has expanded dramatically within ICES, Canada and the
+management has expanded dramatically within ICES, Canada, and the
 Northeast US \citep{nielsenberg14,cadigan16,stockmiller21}. State-space
@@
 considered best practice and a requirement for the next generation of
-stock assessment models.
+stock assessment models \citep{hoyleetal22, punt23}.
 
@@
 parameters that can be treated as random effects in assessment models.
-We know relatively little about the statistical reliability of such
-models or the ability of information criteria to distinguish among such
-alternative structures.
-
-Convergence results can be useful for understanding how bad convergence
-in applications to real data might direct the practitioner to which
-alternative random effects configurations is be more appropriate. The
-type of convergence that we might use as a diagnostic is important. And
-what rule of thumb we use for certain criteria e.g.~1e-6? large SEs/CVs?
+We know relatively little about the factors affecting statistical
+reliability of such models or the ability of information criteria to
+distinguish among such alternative structures.
 
 Here we conduct a simulation study with operating models varying by
-degree of observation error uncertainty, sources of process error (M,
-NAA, q, sel), and fishing history. The simulations from these operating
+degree of observation error uncertainty, source and variability of
+process error, and fishing history. The simulations from these operating
 models are fitted with estimating models that make alternative
-assumptions for sources of process error (M, NAA\textless{} q, sel),
-whether a stock-recruit model was estimated, and whether M is estimated.
-We evaluate whether AIC can correctly determine the correct source of
-process error and stock effects on recruitment. We also evaluate the
-degree of bias in the outputs of the assessment model that are important
-for management.
+assumptions for sources of process error, whether a stock-recruit model
+was estimated, and whether M is estimated. We evaluate whether AIC can
+correctly determine the correct source of process error and stock
+effects on recruitment. We also evaluate the degree of bias in the
+outputs of the assessment model that are important for management.
 
@@
 
-Used the WHAM package \citep[][commit 77bbd94]{stockmiller21} (Miller
-and Stock 2020). This packages has also been used to configure operating
-and estimating models for closed loop simulations evaluating index-based
-assessment methods \citep{legaultetal23} and is used for management of
-haddock, butterfish, plaice, bluefish in the Northeast US.
+We used the Woods Hole Assessment Model (WHAM) to configure operating
+and estimation models in our simulation study \citep[]{stockmiller21}.
+WHAM is an R package freely available as a github repository. For this
+study we used
+\href{https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3}{version
+1.0.6.9000, commit 77bbd94}. This packages has also been used to
+configure operating and estimating models for closed loop simulations
+evaluating index-based assessment methods \citep{legaultetal23} and is
+used for management of haddock, butterfish, American plaice, bluefish,
+Atlantic cod, and black sea bass in the Northeast US.
 
@@
 
-78 total operating models. We simulated 100 data sets for each operating
-model. For each simulated data set we fit a set of 20 estimating models.
+In total, we configured 72 operating models with alternative assumptions
+about the source and variability of process errors, level of observation
+error in indices and age composition data, and contrast in fishing
+pressure over time. We fitted 20 estimating models to observations from
+each of 100 simulations where process errors were also simulated.
+Estimating models made alternative assumptions about the source of
+process errors and whether natural mortality (or the mean for models
+with process error in natural mortality) was estimated and whether a
+Beverton-Holt stock recruit relationship was estimated within the
+estimation model. Details of each of the operating and estimating models
+are described below.
+
+We did not use the bias-correction feature for process errors or
+observations described by \citet{stockmiller21} for operating and
+estimating models.Simulations and model fitting were all carried out on
+the University of Massachusetts Green High-Performance Computing
+Cluster. All code we used to perform the simulation study and summarize
+results can be found at
+\url{https://github.com/timjmiller/SSRTWG/tree/main/Project_0/code}.
 
@@
 spawning occurs 1/4 of the way through the year. The maturity at age was
-a logistic curve with a50 = 2.89 and slope = 0.88 and assumed known in
-all estimation models (Figure \ref{om_maturity}).
-
-\begin{figure}
-\caption{The proportion mature at age assumed for the population in all operating and estimating models.}\label{om_maturity}
-\begin{center}
-\includegraphics[width = \textwidth]{om_maturity.png}
-\end{center}
-\end{figure}
+a logistic curve with \(a_{50}\) = 2.89 and slope = 0.88 and assumed
+known in all estimation models (Figure \ref{om_maturity}).
 
@@
 observations for the fleet.
-
-\begin{figure}
-\caption{The selectivity at age assumed for the fishing fleet in operating models without selectivity random effects.
-The mean selectivity at age across time for the fishing fleet in operating models with selectivity random effects.}\label{om_mean_selectivity}
-\begin{center}
-\includegraphics[width = \textwidth]{om_mean_selectivity.png}
-\end{center}
-\end{figure}
 
@@
 
-\begin{landscape}
-\begin{table}
-\caption{Distinguishing characteristics of the operating models with random effects on survival.
-Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
-Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.}\label{naa_om_table}
-{\footnotesize \input{naa_om_tab}}
-\end{table}
-\end{landscape}
-
 \hypertarget{operating-models-with-random-effects-on-natural-mortality}{%
@@
 operating models are given in Table \ref{M_om_table}.
-
-\begin{landscape}
-\begin{table}
-\caption{Distinguishing characteristics of the operating models with random effects on natural mortality.
-Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
-Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
-For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}\label{M_om_table}
-{\input{M_om_tab}}
-\end{table}
-\end{landscape}
 
@@
 
-\begin{landscape}
-\begin{table}
-\caption{Distinguishing characteristics of the operating models with random effects on selectivity.
-Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
-Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
-For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}\label{sel_om_table}
-{\input{Sel_om_tab}}
-\end{table}
-\end{landscape}
-
 \hypertarget{operating-models-with-random-effects-on-index-catchability}{%
@@
 
-\begin{landscape}
-\begin{table}
-\caption{Distinguishing characteristics of the operating models with random effects on catchability.
-Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
-Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
-For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}\label{q_om_table}
-{\input{q_om_tab}}
-\end{table}
-\end{landscape}
-
 \hypertarget{estimating-models}{%
@@
 29-32 in Table \ref{em_table} were fit to simulate data sets from R+q
-operating models.
-
-\begin{table}
-\caption{Distinguishing characteristics of the estimating models.}\label{em_table}
-{\scriptsize \input{em_tab}}
-\end{table}
-
-SR estimation or not
-
-Initial values for BH parameters are the true values. Initial values for
-mean R model = true R0.
-
-M estimation or not
-
-NAA\_re Random effects on Recruitment only or random effects on
-recruitment and transitions among older numbers at age.
-
-M\_re Random effects on Recruitment only, M constant across age .
-
-sel\_re Random effects on Recruitment only, fleet logistic selectivity
-RE model?
-
-q\_re Random effects on Recruitment only, one survey catchability RE
-model?
-
-Simulations were all carried out on the University of Massachusetts
-Green High-Performance Computing Cluster. Code for completing the
-simulations and summarization of results can be found at
-github.com/timjmiller/SSRTWG/Project\_0. We used the wham package
-version 1.X.X (commit 77bbd94).
+operating models. The observation error variance of aggregate catch and
+indices were all assumed known at the true values.
 
@@
 sets. There are various ways to assess convergence of the fit. We
-summarized 5 alternative categories of convergence. 1. Did the
-optimization routine (stats::nlminb) complete without error? 2. Did the
-stats::nlminb convergence flag = 0 indicate successful convergence? 3.
+summarized 5 alternative categories of convergence.
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Did the optimization routine (stats::nlminb) complete without error?
+\item
+  Did the stats::nlminb convergence flag = 0 indicate successful
+  convergence?
+\item
   Was the maximum absolute value of the gradient of the log-likelihood
-\textless{} \(1\times10^{-6}\)? 4. Did TMB::sdreport provide non-NA
-values for all fixed effects standard errors? 5. Did TMB::sdreport
-provide all standard errors \textless{} 100?
+  \textless{} \(1\times10^{-6}\)?
+\item
+  Did TMB::sdreport provide non-NA values for all fixed effects standard
+  errors?
+\item
+  Did TMB::sdreport provide all standard errors \textless{} 100?
+\end{enumerate}
 
@@
 median relative bias using the binomial distribution approach as in
-\citet{stockmiller21}. We also estimated bias of stock-recruit
-parameters for estimation models that assumed the Beverton-Holt stock
-recruit function.
+\citet{millerhyun18} and \citet{stockmiller21}. We also estimated bias
+of stock-recruit parameters for estimation models that assumed the
+Beverton-Holt stock recruit function.
 
 \hypertarget{mohns-rho}{%
-\subsubsection*{Mohn's rho}\label{mohns-rho}}
-\addcontentsline{toc}{subsubsection}{Mohn's rho}
-
-We estimated Mohn's rho for SSB, fully-selected fishing mortality, and
-recruitment for each estimation model. We estimated 7 peels for each
+\subsubsection*{\texorpdfstring{Mohn's
+\(\rho\)}{Mohn's \textbackslash rho}}\label{mohns-rho}}
+\addcontentsline{toc}{subsubsection}{Mohn's \(\rho\)}
+
+We estimated Mohn's \(\rho\) for SSB, fully-selected fishing mortality,
+and recruitment for each estimation model. We estimated 7 peels for each
 estimation model. We calculated median 95\% confidence intervals for
-Mohn's rho using the same methods as that for bias.
+Mohn's \(\rho\) using the same methods as that for relative bias.
 
@@
 
-Probability of convergence was better for models that assumed the
-correct source of process error, assumed M was known, and did not assume
-stock-recruit relationships.
-
 \hypertarget{r-rs-operating-models}{%
@@
 
-Figures \ref{naa_om_em_R_MF_convergence} to
-\ref{naa_om_em_BH_ME_convergence}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structures R and R+S.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.}\label{naa_om_em_R_MF_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_p_convergence_meanR_M_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structures R and R+S.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.}\label{naa_om_em_R_ME_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_p_convergence_meanR_M_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structures R and R+S.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is fixed at the true value.}\label{naa_om_em_BH_MF_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_p_convergence_BH_M_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structures R and R+S.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is estimated.}\label{naa_om_em_BH_ME_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_p_convergence_BH_M_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
+R+S EMs fit to R OMs exhibited poor convergence for types 3, 4, and 5
+regardless of whether M or a stock-recruit relationship was estimated
+(Figures \ref{naa_om_em_R_MF_convergence} to
+\ref{naa_om_em_BH_ME_convergence}). R EMs exhibited high convergence
+rates for all convergence types for both R and R+S OMs. Convergence
+rates were high for all convergence types for all EMs in R+S OMs with
+lower uncertainty in observations. Convergence for most convergence
+types generally declines most estimating models when mean-log natural
+mortality rate is estimated and/or a Beverton-Holt stock recruit
+relationship is estimated even when the process error assumptions of the
+estimating and operating models match. There can be relatively high
+convergence probability of type 4 (an invertible hessian) for EMs that
+do not have the correct process error assumed, but type 5 convergence
+(no very large standard errors estimated from hessian) for these
+estimating models is typically much lower.
 
@@
 
-Figures \ref{M_om_em_R_MF_convergence} to
-\ref{M_om_em_BH_ME_convergence}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+M.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.}\label{M_om_em_R_MF_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_p_convergence_meanR_M_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+M.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.}\label{M_om_em_R_ME_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_p_convergence_meanR_M_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+M.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is fixed at the true value.}\label{M_om_em_BH_MF_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_p_convergence_BH_M_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+M.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is estimated.}\label{M_om_em_BH_ME_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_p_convergence_BH_M_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
+R+S EMs generally converged less reliably across convergence types than
+other EMs when fit to data generated from R+M OMs (Figures
+\ref{M_om_em_R_MF_convergence} to \ref{M_om_em_BH_ME_convergence}).
+However, using convergence type 5, all EMs converged poorly when OMs had
+low variability in natural mortality process errors or higher
+observation error uncertainty. Using convergence type 5, R+M EMs
+generally converged better than other EMs when observation error
+uncertainty was low and natural mortality process errors were more
+variable. Like R+S OMs, the probability of convergence generally
+declined for all EMs when mean log-natural mortality and/or a
+stock-recruit relationship was estimated.
 
@@
 
-Figures \ref{Sel_om_em_R_MF_convergence} to
-\ref{Sel_om_em_BH_ME_convergence}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+Sel.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.}\label{Sel_om_em_R_MF_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_p_convergence_meanR_M_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+Sel.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.}\label{Sel_om_em_R_ME_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_p_convergence_meanR_M_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+Sel.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is fixed at the true value.}\label{Sel_om_em_BH_MF_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_p_convergence_BH_M_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+Sel.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is estimated.}\label{Sel_om_em_BH_ME_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_p_convergence_BH_M_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
+R+S EMs in particular converged poorly when R+Sel OMs had less
+variability in selectivity process errors or higher observation error
+uncertainty (Figures \ref{Sel_om_em_R_MF_convergence} to
+\ref{Sel_om_em_BH_ME_convergence}). R+Sel EMs generally converged better
+than other EMs for OMs with greater variability in process errors, lower
+observation error, and contrast in fishing pressure regardless of
+whether mean log natural mortality or a stock recruit relationship was
+estimated.
 
@@
 
-Figures \ref{q_om_em_R_MF_convergence} to
-\ref{q_om_em_BH_ME_convergence}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+q.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.}\label{q_om_em_R_MF_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_p_convergence_meanR_M_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+q.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.}\label{q_om_em_R_ME_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_p_convergence_meanR_M_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+q.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is fixed at the true value.}\label{q_om_em_BH_MF_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_p_convergence_BH_M_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+q.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is estimated.}\label{q_om_em_BH_ME_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_p_convergence_BH_M_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
+Convergence of R+q EMs is generally better than that of other EMs for
+all convergence types when R+q OMs assume contrast in fishing history
+overFigures (\ref{q_om_em_R_MF_convergence} to
+\ref{q_om_em_BH_ME_convergence}). Convergence of R+S EMs is generally
+worse than that of other EMs across all OMs whether or not mean
+log-natural mortality or a stock recruit relationship is estimated.
+Again, convergence probablity generally declines for all EMs when mean
+log-natural mortality or a stock recruit relationship is estimated.
 
@@
 
-Figures \ref{naa_om_proportion_best_aic_R_MF} to
-\ref{naa_om_proportion_best_aic_SR_ME}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R and R+S process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.} \label{naa_om_proportion_best_aic_R_MF}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_proportion_best_aic_R_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R and R+S process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.} \label{naa_om_proportion_best_aic_R_ME}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_proportion_best_aic_R_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R and R+S process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate a stock-recruit relationship and and M is fixed at the true value.} \label{naa_om_proportion_best_aic_SR_MF}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_proportion_best_aic_SR_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R and R+S process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate a stock-recruit relationship and and M is estimated.} \label{naa_om_proportion_best_aic_SR_ME}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_proportion_best_aic_SR_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
+Marginal AIC accurately determines the correct process error assumptions
+in estimating models when data are generated from R and R+S OMs,
+regardless of whether mean log-natural mortality or a stock recruit
+relationship is estimated (Figures \ref{naa_om_proportion_best_aic_R_MF}
+to \ref{naa_om_proportion_best_aic_SR_ME}). Adding estimation of mean
+log-natural mortality or a stock recruit relationship separately has a
+negligible effect on the accuracy of determining the correct process
+error assumption. When both are estimated, there is a noticeable
+reduction in accuracy when OMs have a constant fishing history,
+observation error is low and largest variability in recruitment process
+errors.
 
@@
 
-Figures \ref{M_om_proportion_best_aic_R_MF} to
-\ref{M_om_proportion_best_aic_SR_ME}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+M process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.} \label{M_om_proportion_best_aic_R_MF}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_proportion_best_aic_R_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+M process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.} \label{M_om_proportion_best_aic_R_ME}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_proportion_best_aic_R_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+M process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate a stock-recruit relationship and and M is fixed at the true value.} \label{M_om_proportion_best_aic_SR_MF}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_proportion_best_aic_SR_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+M process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate a stock-recruit relationship and and M is estimated.} \label{M_om_proportion_best_aic_SR_ME}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_proportion_best_aic_SR_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
+Marginal AIC only accurately determined the correct process error model
+and correlation structure when observation error was low and variability
+in natural mortality process errors was high (Figures
+\ref{M_om_proportion_best_aic_R_MF} to
+\ref{M_om_proportion_best_aic_SR_ME}). Estimating the mean natural
+mortality rate reduced the accuracy of AIC for OMs that assumed natural
+mortality process errors were independent. For OMs with poor accuracy,
+AIC most frequently selected EMs with process errors in catchability or
+selectivity. Selection of R+S EMs was typically unlikely.
 
@@
 
-Figures \ref{Sel_om_proportion_best_aic_R_MF} to
-\ref{Sel_om_proportion_best_aic_SR_ME}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+Sel process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.} \label{Sel_om_proportion_best_aic_R_MF}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_proportion_best_aic_R_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+Sel process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.} \label{Sel_om_proportion_best_aic_R_ME}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_proportion_best_aic_R_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+Sel process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate a stock-recruit relationship and and M is fixed at the true value.} \label{Sel_om_proportion_best_aic_SR_MF}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_proportion_best_aic_SR_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+Sel process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate a stock-recruit relationship and and M is estimated.} \label{Sel_om_proportion_best_aic_SR_ME}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_proportion_best_aic_SR_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\hypertarget{rq-operating-models-1}{%
-\subsubsection*{R+q operating models}\label{rq-operating-models-1}}
-\addcontentsline{toc}{subsubsection}{R+q operating models}
-
-Figures \ref{q_om_proportion_best_aic_R_MF} to
-\ref{q_om_proportion_best_aic_SR_ME}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+q process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.} \label{q_om_proportion_best_aic_R_MF}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_proportion_best_aic_R_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+q process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.} \label{q_om_proportion_best_aic_R_ME}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_proportion_best_aic_R_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+q process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate a stock-recruit relationship and and M is fixed at the true value.} \label{q_om_proportion_best_aic_SR_MF}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_proportion_best_aic_SR_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+q process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate a stock-recruit relationship and and M is estimated.} \label{q_om_proportion_best_aic_SR_ME}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_proportion_best_aic_SR_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\clearpage
+Marginal AIC most accurately determined the correct source of process
+error and correlation structure for R+Sel OMs with low observation error
+(Figures \ref{Sel_om_proportion_best_aic_R_MF} to
+\ref{Sel_om_proportion_best_aic_SR_ME}). When there was low variability
+in selectivity process errors and high observation error, R+q or R+S EMs
+were more likely to have the best AIC. Whether mean log-natural morality
+or stock recruit relationships were estimated appeared to have little
+effect on the performance of marginal AIC. \#\#\# R+q operating models
+\{-\}
+
+Marginal AIC most accurately determined the correct source of process
+error and correlation structure for R+q OMs with high variability in
+catchability process errors (Figures \ref{q_om_proportion_best_aic_R_MF}
+to \ref{q_om_proportion_best_aic_SR_ME}). The worst accuracy occurred
+for OMs with low variability in catchability process errors and high
+observation error. However, in these OMs, the marginal AIC accurately
+determined the correct source of process error (but not correlation
+structure) except when EMs estimated both mean log-natural morality and
+the stock recruit relationships and OMs assumed a constant fishing
+pressure.
 
@@
 
-Tables of frequencies where AIC chooses correct stock-recruit assumption
-imply poor performance of AIC, but figures showing effect of variation
-in SSB demonstrate that large variation in SSB is needed to reduce Type
-II errors for marginal AIC.
+Our comparisons of model performance condition on assuming the true
+process error configuration is known (EM and OM process error
+assumptions match). Broadly, we found generally poor accuracy of AIC in
+selecting models assuming a Beverton-Holt stock recruit function over
+the null model without an assumed stock-recruit relationship for all OMs
+(Tables \ref{naa_om_em_R_BH_aic_table} to \ref{q_om_em_R_BH_aic_table}.
+However, we also found AIC to more accurately determine a stock-recruit
+relationship when there was greater variation in spawning biomass
+generated in the simulated populations.
+
+AIC accuracy for including the B-H stock-recruit relationship was
+estimated to be highest for models OMs and EMs with process errors only
+on recruitment and lowest variance of recruitment process errors
+(Figures \ref{naa_om_MF_BH_glm_AIC_plots} and
+\ref{naa_om_ME_BH_glm_AIC_plots}). Holding variability in SSB constant,
+accuracy of stock-recruit model selection declined with greater variance
+of recruitment and survival process errors. Simultaneously estimating
+mean log-natural mortality has a small negative effect on AIC accuracy.
+
+For R+M OMs and EMs there appeared to little effect of correlation in
+the natural mortality process errors or estimating the mean log-natural
+mortality on accuracy of AIC in determining a stock-recruit function,
+holding variability in SSB constant (Figures
+\ref{M_om_MF_BH_glm_AIC_plots} and \ref{M_om_ME_BH_glm_AIC_plots}).
+
+The range of variation in SSB across simulated populations was lower for
+R+Sel operating models than R+S or R+M OMs. OMs with contrast in fishing
+pressure over time is the primary mechanism creating variation in SSB.
+However, there appears to be little effect of variability or correlation
+of selectivity process errors or whether mean log-natural mortality was
+estimated on the accuracy of AIC in selecting the stock recruit
+relationship (Figures \ref{Sel_om_MF_BH_glm_AIC_plots} and
+\ref{Sel_om_ME_BH_glm_AIC_plots})
+
+Like the R+Sel OMs, the range of variation in SSB across is only
+increased by allowing contrast in fishing pressure over time. There
+appears to be a slight decrease in accuracy in AIC when mean log-natural
+mortality is estimated and there is a more noticeable increase in
+accuracy for OMs with lower observation error in those cases. (Figures
+\ref{q_om_MF_BH_glm_AIC_plots} and \ref{q_om_ME_BH_glm_AIC_plots})
+
+\hypertarget{bias-1}{%
+\subsection*{Bias}\label{bias-1}}
+\addcontentsline{toc}{subsection}{Bias}
 
@@
 
-\begin{table}
-\caption{Operating models and estimation models all assume matching R or R+S process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}
-{\input{naa_om_em_R_BH_aic_table}}
-\end{table}
-
-\begin{figure}
-\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
-Operating and estimating models have matching R or R+S process error structures.
-Estimating models assume mean M is fixed at the true value.}\label{naa_om_MF_BH_glm_AIC_plots}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_MF_pred_BH_best.png}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB. Estimating models allow estimation of mean M.}\label{naa_om_ME_BH_glm_AIC_plots}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_ME_pred_BH_best.png}
-\end{center}
-\end{figure}
+When mean log-natural mortality is known, all estimating models
+regardless of process error assumptions produce unbiased estimates of
+SSB over the entire time period for R OMs (Figure
+\ref{naa_om_em_R_MF_relbias_ssb}). For R+S OMs, only R+S EMs produce
+unbiased estimation of SSB and bias resulting from other EMs increased
+with larger variability in survival random effects. Estimating mean
+log-natural mortality generally produced greater variability in relative
+errors of SSB estimates and a change in sign of bias for EMs assuming
+process errors other than R and R+S (Figure
+\ref{naa_om_em_R_ME_relbias_ssb}). Estimating a stock-recruit
+relationship had little discernable effect on bias of SSB estimation
+(Figures \ref{naa_om_em_BH_MF_relbias_ssb}), but in combination with
+estimating mean log-natural mortality resulted in large bias of R+S EMs
+for R OMs with high observation error, constant fishing pressure and
+large variability in recruitment process errors
+\ref{naa_om_em_BH_ME_relbias_ssb}).
 
@@
 
-\begin{table}
-\caption{Operating models and estimation models all assume matching R+M process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}
-{\input{M_om_em_R_BH_aic_table}}
-\end{table}
-
-\begin{figure}
-\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
-Operating and estimating models have matching R+M process error structures.
-Estimating models assume mean M is fixed at the true value.}\label{M_om_MF_BH_glm_AIC_plots}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_MF_pred_BH_best.png}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
-Operating and estimating models have matching R+M process error structures.
-Estimating models allow estimation of mean M.}\label{M_om_ME_BH_glm_AIC_plots}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_ME_pred_BH_best.png}
-\end{center}
-\end{figure}
+When mean log-natural mortality is known, SSB estimation bias was small
+or non-existent for all EM process error assumptions when the OMs had
+low variability of the natural mortality process errors except a
+tendency toward negative bias in the terminal years for OMs with high
+observation error and a change in fishing pressure over time (Figure
+\ref{M_om_em_R_MF_relbias_ssb}). Errors in SSB estimation were most
+variable with high variability and autocorrelation of natural mortality
+process errors, but confidence intervals generally covered 0 indicating
+no evidence of bias. Estimating mean log-natural mortality results in
+large variability in SSB relative errors, particularly for OMs with high
+observation error and constant fishing pressure and all EMs produced
+biased estimation of SSB over all years when OMs had high observation
+error, a change in fishing pressure and low variability of natural
+mortality process errors (Figure \ref{M_om_em_R_ME_relbias_ssb}). Like R
+and R+S OMs, estimating a stock-recruit relationship had little
+discernible effect on SSB bias (Figures \ref{M_om_em_BH_MF_relbias_ssb}
+and \ref{M_om_em_BH_ME_relbias_ssb})
 
@@
 
-\begin{table}
-\caption{Operating models and estimation models all assume matching R+Sel process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}
-{\input{Sel_om_em_R_BH_aic_table}}
-\end{table}
-
-\begin{figure}
-\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
-Operating and estimating models have matching R+Sel process error structures.
-Estimating models assume mean M is fixed at the true value.}\label{Sel_om_MF_BH_glm_AIC_plots}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_MF_pred_BH_best.png}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
-Operating and estimating models have matching R+Sel process error structures.
-Estimating models allow estimation of mean M.}\label{Sel_om_ME_BH_glm_AIC_plots}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_ME_pred_BH_best.png}
-\end{center}
-\end{figure}
-
-\hypertarget{rq-operating-models-2}{%
-\subsubsection*{R+q operating models}\label{rq-operating-models-2}}
+When mean log-natural mortality is known and no stock-recruit
+relationship was estimated, SSB estimation bias was low for all EM
+process error assumptions (Figure \ref{Sel_om_em_R_MF_relbias_ssb}). The
+worst bias occurred for OMs with uncorrelated by more variable
+selectivity process errors and high observations error. As for other
+OMs, estimating mean log-natural mortality resulted in greater
+variability of the relative errors in SSB, but also more evidence of
+bias for EMS without the correct process error assumption (Figure
+\ref{Sel_om_em_R_ME_relbias_ssb}). Estimating a stock recruit
+relationship resulted in greater estimation of SSB in the terminal years
+and sometimes less bias when OMs had high observation error and a change
+in fishing pressure over time and the effect was more pronounced when
+mean log-natural mortality was estimated (Figures
+\ref{Sel_om_em_BH_MF_relbias_ssb} and
+\ref{Sel_om_em_BH_ME_relbias_ssb}).
+
+\hypertarget{rq-operating-models-1}{%
+\subsubsection*{R+q operating models}\label{rq-operating-models-1}}
 \addcontentsline{toc}{subsubsection}{R+q operating models}
 
-\begin{table}
-\caption{Operating models and estimation models all assume matching R+q process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}
-{\input{q_om_em_R_BH_aic_table}}
-\end{table}
-
-\begin{figure}
-\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
-Operating and estimating models have matching R+q process error structures.
-Estimating models assume mean M is fixed at the true value.}\label{q_om_MF_BH_glm_AIC_plots}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_MF_pred_BH_best.png}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
-Operating and estimating models have matching R+q process error structures.
-Estimating models allow estimation of mean M.}\label{q_om_ME_BH_glm_AIC_plots}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_ME_pred_BH_best.png}
-\end{center}
-\end{figure}
-
-\clearpage
-
-\hypertarget{bias-1}{%
-\subsection*{Bias}\label{bias-1}}
-\addcontentsline{toc}{subsection}{Bias}
-
-\hypertarget{r-rs-oms}{%
-\subsubsection*{R, R+S OMs}\label{r-rs-oms}}
-\addcontentsline{toc}{subsubsection}{R, R+S OMs}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R and R+S process errors.
-Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{naa_om_em_R_MF_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_R_MF_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R and R+S process errors.
-Estimation models do not assume a stock-recruit relationship and estimate M.}\label{naa_om_em_R_ME_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_R_ME_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R and R+S process errors.
-Estimation models assume a BH stock-recruitment function and M is fixed at the true value.}\label{naa_om_em_BH_MF_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_BH_MF_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R and R+S process errors.
-Estimation models assume a BH stock-recruitment function and estimate M.}\label{naa_om_em_BH_ME_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_BH_ME_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\hypertarget{rm-oms}{%
-\subsubsection*{R+M OMs}\label{rm-oms}}
-\addcontentsline{toc}{subsubsection}{R+M OMs}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+M process errors.
-Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{M_om_em_R_MF_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_R_MF_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+M process errors.
-Estimation models do not assume a stock-recruit relationship and estimate M.}\label{M_om_em_R_ME_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_R_ME_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+M process errors.
-Estimation models assume a BH stock-recruitment function and M is fixed at the true value.}\label{M_om_em_BH_MF_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_BH_MF_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+M process errors.
-Estimation models assume a BH stock-recruitment function and estimate M.}\label{M_om_em_BH_ME_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_BH_ME_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\hypertarget{rsel-oms}{%
-\subsubsection*{R+Sel OMs}\label{rsel-oms}}
-\addcontentsline{toc}{subsubsection}{R+Sel OMs}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+Sel process errors.
-Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{Sel_om_em_R_MF_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_R_MF_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+Sel process errors.
-Estimation models do not assume a stock-recruit relationship and estimate M.}\label{Sel_om_em_R_ME_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_R_ME_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+Sel process errors.
-Estimation models assume a BH stock-recruitment function and M is fixed at the true value.}\label{Sel_om_em_BH_MF_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_BH_MF_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+Sel process errors.
-Estimation models assume a BH stock-recruitment function and estimate M.}\label{Sel_om_em_BH_ME_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_BH_ME_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\hypertarget{rq-oms}{%
-\subsubsection*{R+q OMs}\label{rq-oms}}
-\addcontentsline{toc}{subsubsection}{R+q OMs}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+q process errors.
-Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{q_om_em_R_MF_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_R_MF_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+q process errors.
-Estimation models do not assume a stock-recruit relationship and estimate M.}\label{q_om_em_R_ME_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_R_ME_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+q process errors.
-Estimation models assume a BH stock-recruitment function and M is fixed at the true value.}\label{q_om_em_BH_MF_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_BH_MF_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+q process errors.
-Estimation models assume a BH stock-recruitment function and estimate M.}\label{q_om_em_BH_ME_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_BH_ME_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
+When mean log-natural mortality is known and no S-R relationship was
+estimated, SSB estimation bias was low for all EM process error
+assumptions for all R+q OMs with low variability in catchability process
+errors (Figure \ref{q_om_em_R_MF_relbias_ssb}). With larger variability
+in catchability process errors, bias was low for all EMs except those
+configured with R+M process errors and OMs assumed low observation
+error. Many EMs also showed some negative bias for terminal SSB in OMs
+with higher observation error and a change in fishing pressure over
+time. As for other OMs, mean log-natural mortality resulted in greater
+variablity of the relative errors in SSB, but also more evidence of bias
+for some EMs without the correct process error assumption (Figure
+\ref{q_om_em_R_ME_relbias_ssb}) Estimating a stock recruit relationship
+resulted in less bias in SSB in terminal years when OMs had high
+observation error and a change in fishing pressure over time (Figures
+\ref{q_om_em_BH_MF_relbias_ssb} and \ref{q_om_em_BH_ME_relbias_ssb}).
 
 \hypertarget{beverton-holt-parameter-estimation}{%
-\subsubsection{Beverton-Holt parameter
+\subsubsection*{Beverton-Holt parameter
 estimation}\label{beverton-holt-parameter-estimation}}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of Beverton-Holt stock recruit parameters for estimating models with correct process error assumptions fitted to data sets simulated with R and R+S process errors.
-Estimation models either assume the correct natural mortality rate or estimate it.}\label{naa_om_SR_relbias}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_SR_relerror.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of Beverton-Holt stock recruit parameters for estimating models with correct process error assumptions fitted to data sets simulated with R+M process errors.
-Estimation models either assume the correct natural mortality rate or estimate it.}\label{M_om_SR_relbias}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_SR_relerror.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of Beverton-Holt stock recruit parameters for estimating models with correct process error assumptions fitted to data sets simulated with R+Sel process errors.
-Estimation models either assume the correct natural mortality rate or estimate it.}\label{Sel_om_SR_relbias}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_SR_relerror.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of Beverton-Holt stock recruit parameters for estimating models with correct process error assumptions fitted to data sets simulated with R+q process errors.
-Estimation models either assume the correct natural mortality rate or estimate it.}\label{q_om_SR_relbias}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_SR_relerror.png}
-\end{center}
-\end{figure}
-\end{landscape}
+\addcontentsline{toc}{subsubsection}{Beverton-Holt parameter estimation}
+
+In R and R+S OMs, EMs with the correct assumptions about process errors,
+provided the least biased estimation of Beverton-Holt stock-recruit
+relationship parameters when there was low observation error, a change
+in fishing pressure over time, and lower variability of recruitment
+process errors, and higher variability survival process errors, and
+there was little effect of estimating natural mortality (Figure
+\ref{naa_om_SR_relbias}). For other R and R+S OMs, estimating natural
+mortality often resulted in less biased estimation of stock-recruit
+parameters. The range of confidence intervals for the median parameter
+bias suggests large variability of the parameter estimates.
+
+In R+M OMs, the most accurate estimation of stock-recruit parameters
+occurred when there was a change in fishing pressure combined with
+either low variability in natural mortality process errors and high
+observation error or vice versa (Figure \ref{M_om_SR_relbias}). Relative
+to the R, and R+S OMs, there was even less effect of estimating mean
+log-natural mortality on estimation bias for the stock- recruit
+relationship parameters, but similarly there was generally large
+variability of the parameter estimates.
+
+In R+Sel OMs, bias for stock-recruit parameters was very large for OMs
+with constant fishing pressure (Figure \ref{Sel_om_SR_relbias}). Less
+bias in parameter estimation occurred for OMs with a change in fishing
+pressure over time. There was little effect of estimating natural
+mortality. Patterns in bias of stock-recruit parameter estimation in R+q
+OMs, were very similar to those in R+Sel OMs with poor estimation of
+these parameters in OMs with constant fishing pressure (Figure
+\ref{q_om_SR_relbias}). Like other OMs, there was generally large
+variability of the parameter estimates.
 
 \hypertarget{mean-natural-mortality-rate-estimation}{%
-\subsubsection{(Mean) natural mortality rate
+\subsubsection*{(Mean) natural mortality rate
 estimation}\label{mean-natural-mortality-rate-estimation}}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of (mean) natural mortality rate for estimating models with correct process error assumptions fitted to data sets simulated with R and R+S process errors.
-Estimation models assumed either no stock-recruit relationship or estimated a Beverton-Holt relationship.}\label{naa_om_M_relbias}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_M_relerror.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of (mean) natural mortality rate for estimating models with correct process error assumptions fitted to data sets simulated with R+M process errors.
-Estimation models assumed either no stock-recruit relationship or estimated a Beverton-Holt relationship.}\label{M_om_M_relbias}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_M_relerror.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of (mean) natural mortality rate for estimating models with correct process error assumptions fitted to data sets simulated with R+Sel process errors.
-Estimation models assumed either no stock-recruit relationship or estimated a Beverton-Holt relationship.}\label{Sel_om_M_relbias}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_M_relerror.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of (mean) natural mortality rate for estimating models with correct process error assumptions fitted to data sets simulated with R+q process errors.
-Estimation models assumed either no stock-recruit relationship or estimated a Beverton-Holt relationship.}\label{q_om_M_relbias}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_M_relerror.png}
-\end{center}
-\end{figure}
-\end{landscape}
+\addcontentsline{toc}{subsubsection}{(Mean) natural mortality rate
+estimation}
+
+The mean natural mortality rate is estimated pretty accurately for R and
+R+S OMs when EMs use the correct process error configuration and there
+is a change in fishing pressure over time and observation error is low
+(Figure \ref{naa_om_M_relbias}). Estimation is a little more variable
+when observation error is higher or fishing pressure is constant, but
+when OMs have both, estimation is extremely variable. There is little
+effect of estimating stock-recruit relationships on the bias in
+estimating natural mortality. For R+M OMs, effects on estimating mean
+log-natural mortality were similar to the R and R+S OMs: best estimation
+with a change in fishing pressure and low observation error and worst
+with no change in fishing pressure and higher observation error (Figure
+\ref{M_om_M_relbias}). The same patterns also occurred for R+Sel and R+q
+OMs (Figures \ref{Sel_om_M_relbias} and \ref{q_om_M_relbias}).
 
@@
 
-\hypertarget{r-rs-oms-1}{%
-\subsubsection*{R, R+S OMs}\label{r-rs-oms-1}}
-\addcontentsline{toc}{subsubsection}{R, R+S OMs}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R and R+S process errors.
-Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{naa_om_em_R_MF_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_mohns_rho_R_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R and R+S process errors.
-Estimation models do not assume a stock-recruit function and M estimated.}\label{naa_om_em_R_ME_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_mohns_rho_R_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R and R+S process errors.
-Estimation models assume a B-H stock-recruit function and M is fixed at the true value.}\label{naa_om_em_BH_MF_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_mohns_rho_BH_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R and R+S process errors.
-Estimation models assume a B-H stock-recruit function and M estimated.}\label{naa_om_em_BH_ME_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_mohns_rho_BH_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\hypertarget{rm-oms-1}{%
-\subsubsection*{R+M OMs}\label{rm-oms-1}}
-\addcontentsline{toc}{subsubsection}{R+M OMs}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+M process errors.
-Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{M_om_em_R_MF_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_mohns_rho_R_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+M process errors.
-Estimation models do not assume a stock-recruit function and M estimated.}\label{M_om_em_R_ME_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_mohns_rho_R_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+M process errors.
-Estimation models assume a B-H stock-recruit function and M is fixed at the true value.}\label{M_om_em_BH_MF_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_mohns_rho_BH_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+M process errors.
-Estimation models assume a B-H stock-recruit function and M estimated.}\label{M_om_em_BH_ME_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_mohns_rho_BH_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\hypertarget{rsel-oms-1}{%
-\subsubsection*{R+Sel OMs}\label{rsel-oms-1}}
-\addcontentsline{toc}{subsubsection}{R+Sel OMs}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+Sel process errors.
-Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{Sel_om_em_R_MF_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_mohns_rho_R_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+Sel process errors.
-Estimation models do not assume a stock-recruit function and M estimated.}\label{Sel_om_em_R_ME_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_mohns_rho_R_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+Sel process errors.
-Estimation models assume a B-H stock-recruit function and M is fixed at the true value.}\label{Sel_om_em_BH_MF_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_mohns_rho_BH_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+Sel process errors.
-Estimation models assume a B-H stock-recruit function and M estimated.}\label{Sel_om_em_BH_ME_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_mohns_rho_BH_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\hypertarget{rq-oms-1}{%
-\subsubsection*{R+q OMs}\label{rq-oms-1}}
-\addcontentsline{toc}{subsubsection}{R+q OMs}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+q process errors.
-Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{q_om_em_R_MF_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_mohns_rho_R_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+q process errors.
-Estimation models do not assume a stock-recruit function and M estimated.}\label{q_om_em_R_ME_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_mohns_rho_R_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+q process errors.
-Estimation models assume a B-H stock-recruit function and M is fixed at the true value.}\label{q_om_em_BH_MF_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_mohns_rho_BH_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+q process errors.
-Estimation models assume a B-H stock-recruit function and M estimated.}\label{q_om_em_BH_ME_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_mohns_rho_BH_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
+\hypertarget{r-rs-operating-models-3}{%
+\subsubsection*{R, R+S operating models}\label{r-rs-operating-models-3}}
+\addcontentsline{toc}{subsubsection}{R, R+S operating models}
+
+When natural mortality is known and no stock recruit relationship is
+estimated, median absolute Mohn's \(\rho\) was less than 0.25 for all
+EMS fit to simulated data sets from R and R+S OMs (Figure
+\ref{naa_om_em_R_MF_mohns_rho}). The strongest median Mohn's \(\rho\)
+values for F and SSB were obtained for EMs that did not match the R+S
+OMs with the largest variability in survival process errors, change in
+fishing pressure over time, and higher observation error. Mohn's
+\(\rho\) was also greater than 0 for recruitment for all EMS, including
+the matching EM, fit to R OMs that had the largest variability in
+recruitment random effects and high observation error. Estimating
+natural mortality removed the retrospective patterns in F and SSB in R+S
+OMs with the largest variability in survival process errors, change in
+fishing pressure over time, and higher observation error (Figure
+\ref{naa_om_em_R_ME_mohns_rho}). Estimating a stock-recruit relationship
+had no discernible effect on the retrospective patterns (Figure
+\ref{naa_om_em_BH_MF_mohns_rho}). Estimating natural mortality and the
+stock-recruit relationship increased the variablity of Mohn's \(\rho\)
+for recruitment particularly for OMs with high observation error (Figure
+\ref{naa_om_em_BH_ME_mohns_rho}).
+
+\hypertarget{rm-operating-models-3}{%
+\subsubsection*{R+M operating models}\label{rm-operating-models-3}}
+\addcontentsline{toc}{subsubsection}{R+M operating models}
+
+When mean log-natural mortality is known and no stock recruit
+relationship is estimated, median abslolute Mohn's \(\rho\) was less
+than 0.125 for all EMS fit to simulated data sets from R+M OMs (Figure
+\ref{M_om_em_R_MF_mohns_rho}). The strongest median Mohn's \(\rho\)
+values for F, SSB, and recruitment were obtained for EMs other than R+M
+with autocorrelation assumed for OMs with a change in fishing pressure
+over time, and higher observation error. Like R and R+S OMs, estimating
+natural mortality removed the retrospective patterns in the respective
+OMs (Figure \ref{M_om_em_R_ME_mohns_rho}). Estimating a stock-recruit
+relationship also had the same effect as estimating the mean log-natural
+mortality rate, but Mohn's \(\rho\) values were more variable for R+M
+EMs with autocorrelation assumed (Figure \ref{M_om_em_BH_MF_mohns_rho}).
+Estimating natural mortality and the stock-recruit relationship
+increased the variablity of Mohn's \(\rho\) for F, SSB, and recruitment,
+particularly for OMs with high observation error and constant fishing
+pressure over time (Figure \ref{M_om_em_BH_ME_mohns_rho}).
+
+\hypertarget{rsel-operating-models-3}{%
+\subsubsection*{R+Sel operating models}\label{rsel-operating-models-3}}
+\addcontentsline{toc}{subsubsection}{R+Sel operating models}
+
+Similar to R+M OMs, when mean log-natural mortality is known and no
+stock recruit relationship is estimated, median abslolute Mohn's
+\(\rho\) was less than 0.125 for all EMS fit to simulated data sets from
+R+Sel OMs (Figure \ref{Sel_om_em_R_MF_mohns_rho}). The strongest median
+Mohn's \(\rho\) values for F, SSB, and recruitment were obtained for all
+EMs with OMs that had a change in fishing pressure over time, and higher
+observation error. Estimating natural mortality increased the
+variability in Mohn's \(\rho\) values, particularly for OMs with higher
+observation error and constant fishing (Figure
+\ref{Sel_om_em_R_ME_mohns_rho}). Estimating a stock-recruit relationship
+also removed any tendency for retrospective patterns in OMs with a
+change in fishing pressure and higher observation error (Figure
+\ref{Sel_om_em_BH_MF_mohns_rho}). Estimating natural mortality and the
+stock-recruit relationship further increased the variablity of Mohn's
+\(\rho\) for F, SSB, and recruitment, particularly for OMs with high
+observation error and constant fishing pressure over time (Figure
+\ref{Sel_om_em_BH_ME_mohns_rho}).
+
+\hypertarget{rq-operating-models-2}{%
+\subsubsection*{R+q operating models}\label{rq-operating-models-2}}
+\addcontentsline{toc}{subsubsection}{R+q operating models}
+
+Similar to R+M and R+Sel OMs, when mean log-natural mortality is known
+and no stock recruit relationship is estimated, median abslolute Mohn's
+\(\rho\) was less than 0.125 for all EMS fit to simulated data sets from
+R+q OMs (Figure \ref{q_om_em_R_MF_mohns_rho}). Also similar to other
+OMs, the strongest median Mohn's \(\rho\) values for F, SSB, and
+recruitment were obtained for with OMs that had a change in fishing
+pressure over time, and higher observation error. Estimating natural
+mortality increased the variability in Mohn's \(\rho\) values,
+particularly for OMs with constant fishing and autocorrelation of
+process errors in catchability (Figure \ref{q_om_em_R_ME_mohns_rho}).
+Similar to other OMs, estimating a stock-recruit relationship removed
+any tendency for retrospective patterns in OMs with a change in fishing
+pressure and higher observation error (Figure
+\ref{q_om_em_BH_MF_mohns_rho}). Like R+Sel OMs, estimating natural
+mortality and the stock-recruit relationship further increased the
+variablity of Mohn's \(\rho\) for F, SSB, and recruitment, particularly
+for OMs with high observation error and constant fishing pressure over
+time (Figure \ref{q_om_em_BH_ME_mohns_rho}).
 
@@
 
-The estimating models assumed variances of aggregate catch and index
-observations was known. This approximation may be appropriate for
-indices where we have a reliable estimate of uncertainty based on the
-survey design (), but there may be better approaches for the aggregate
-catch such as an informed prior on the standard errors with realistic
-bounds.
+\hypertarget{convergence}{%
+\subsection*{Convergence}\label{convergence}}
+\addcontentsline{toc}{subsection}{Convergence}
+
+Convergence results can be useful for understanding how bad convergence
+in applications to real data might direct the practitioner to which
+alternative random effects configurations is be more appropriate.
+Therefore, the type of convergence that we might use as a diagnostic is
+important. It is common during assessment model fitting to check that
+the maximum absolute gradient component is less than some threshold. Our
+Type 3 convergence represents this and we used 1e-6 as the threshold,
+but there is no standard. Furthermore we found the hessian at the
+optimized log-likelihood can often be invertible (Convergence types 4
+and 5) when the gradient approach to convergence fails. The probability
+of convergence type 5 cannot be greater than that of type 4 because type
+5 is a subset where none of the standard errors obtained by inverting
+the hessian are greater than a specified value (100 in this case). We
+presume that the parameters that cause type 4 convergence to fail will
+also typically be the same as those that cause type 5 to fail. Perhaps a
+higher gradient threshold for Type 3 convergence would be more
+consistent with hessian-based convergence, but the gradient and
+hessian-based standard errors should be inspected. Parameters that are
+not well estimated can lead the assessment scientist toward a more
+appropriate and often more parsimonious model configuration. For
+example, variance parameters that are estimated on log scale as low
+negative number (variance is close to 0) can happen for random effects
+estimated with little or no variability. These may sometimes lead a
+hessian to not be invertible or due to numerical issues may be
+invertible, but provide very large standard errors. More generally, we
+can expect lower likelihood of convergence when estimating natural
+mortality or stock-recruit relationships because the typical data in
+assessment models does not strongly inform these parameters.
+
+\hypertarget{aic}{%
+\subsection*{AIC}\label{aic}}
+\addcontentsline{toc}{subsection}{AIC}
+
+Among the process error configurations we used in operating models, we
+found AIC to be accurate for process errors on recruitment and survival.
+Fitting models to other OMs rarely preferred R+S EMs and R and R+S EMs
+were nearly always selected for the matching OMs. For other sources of
+process error, accuracy of AIC was improved to useful levels when there
+was larger variability in the process errors and/or lower observation
+error. Therefore, when data are relatively good and an there is large
+variability in the unknown process errors, AIC is likely to perform
+well.
+
+AIC generally performs weakly in determining stock recruit relationships
+unless there is large contrast in SSB. Interestingly,
+\citet{milleretal16} found AIC to prefer a stock-recruit relationship
+for SNEMA yellowtail flounder where there was a large change in stock
+size estimated over time and flatfish are well observed by the bottom
+trawl indices relative to some other guilds such as semi-pelagic
+species. Although we did not compare models with alternative
+stock-recruit relationships, we do not expect AIC to perform any better
+distinguishing these relationships. Our results comport well with those
+of \citet{devalpinehastings02} where state-space stock-recruit models
+were fit just to SSB and recruitment ``observations''.
+
+\hypertarget{bias-2}{%
+\subsection*{Bias}\label{bias-2}}
+\addcontentsline{toc}{subsection}{Bias}
+
+As we might expect bias in all parameters and assessment output was
+generally less with lower observation error in indices and age
+composition. Reliable estimation of stock-recruit relationship
+parameters only appears possible in ideal situations with lower
+observation errors in age composition and indices, lower variability in
+recruitment process errors and large contrast in spawning biomass over
+time. The latter can occur in with larger variability in survival or
+natural mortality process errors, variation in fishing pressure.
+
+Estimation of natural mortality (or the mean when there are process
+errors on natural mortality) appears more feasible. We found low or no
+evidence of bias for many OM process error assumptions when there was
+contrast in fishing pressure even when there was greater observation
+error although it can lead to more variable estimates of natural
+mortality. For OMs where there was bias in natural mortality due to high
+observation error (R+Sel, R+q), estimating the stock-recruit
+relationship seemed to remove the bias. However, estimation of natural
+mortality can cause large differences between the true and estimated SSB
+(that may be unbiased on average) when there is less contrast in fishing
+pressure over time and higher observation error.
+
+\hypertarget{retrospective-patterns}{%
+\subsection*{Retrospective patterns}\label{retrospective-patterns}}
+\addcontentsline{toc}{subsection}{Retrospective patterns}
+
+Higher observation error with variation in fishing pressure can produce
+retrospective patterns even when the the process error assumption is
+correct, but this retrospective pattern can be reduced if M or the
+correct stock recruit relationship is estimated. Retrospective patterns
+tended to be more variable for recruitment and can be large even when
+the estimating model is correct. Therefore, we recommend emphasis on
+inspection of retrospective patterns primarily for SSB and F, but
+further research on retrospective patterns in other assessment model
+parameters and derived output may be beneficial.
 
@@
 
+\begin{landscape}
+\begin{table}
+\caption{Distinguishing characteristics of the operating models with random effects on survival.
+Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
+Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.}\label{naa_om_table}
+{\footnotesize \input{naa_om_tab}}
+\end{table}
+\end{landscape}
+
+\begin{landscape}
+\begin{table}
+\caption{Distinguishing characteristics of the operating models with random effects on natural mortality.
+Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
+Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
+For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}\label{M_om_table}
+{\input{M_om_tab}}
+\end{table}
+\end{landscape}
+
+\begin{landscape}
+\begin{table}
+\caption{Distinguishing characteristics of the operating models with random effects on selectivity.
+Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
+Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
+For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}\label{sel_om_table}
+{\input{Sel_om_tab}}
+\end{table}
+\end{landscape}
+
+\begin{landscape}
+\begin{table}
+\caption{Distinguishing characteristics of the operating models with random effects on catchability.
+Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
+Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
+For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}\label{q_om_table}
+{\input{q_om_tab}}
+\end{table}
+\end{landscape}
+
+\begin{table}
+\caption{Distinguishing characteristics of the estimating models.}\label{em_table}
+{\scriptsize \input{em_tab}}
+\end{table}
+
+\begin{table}
+\caption{Operating models and estimation models all assume matching R or R+S process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}\label{naa_om_em_R_BH_aic_table}
+{\input{naa_om_em_R_BH_aic_table}}
+\end{table}
+
+\begin{table}
+\caption{Operating models and estimation models all assume matching R+M process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}\label{M_om_em_R_BH_aic_table}
+{\input{M_om_em_R_BH_aic_table}}
+\end{table}
+
+\begin{table}
+\caption{Operating models and estimation models all assume matching R+Sel process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}\label{Sel_om_em_R_BH_aic_table}
+{\input{Sel_om_em_R_BH_aic_table}}
+\end{table}
+
+\begin{table}
+\caption{Operating models and estimation models all assume matching R+q process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}\label{q_om_em_R_BH_aic_table}
+{\input{q_om_em_R_BH_aic_table}}
+\end{table}
+
+\clearpage
+
+\begin{figure}
+\caption{The proportion mature at age assumed for the population in all operating and estimating models.}\label{om_maturity}
+\begin{center}
+\includegraphics[width = \textwidth]{om_maturity.png}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\caption{The weight at age assumed for the population in all operating and estimating models.}\label{om_waa}
+\begin{center}
+\includegraphics[width = \textwidth]{om_waa.png}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\caption{The Beverton-Holt stock recruit relationship assume for all operating models.}\label{om_sr}
+\begin{center}
+\includegraphics[width = \textwidth]{om_sr.png}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\caption{The selectivity at age assumed for the fishing fleet in operating models without selectivity random effects.
+The mean selectivity at age across time for the fishing fleet in operating models with selectivity random effects.}\label{om_mean_selectivity}
+\begin{center}
+\includegraphics[width = \textwidth]{om_mean_selectivity.png}
+\end{center}
+\end{figure}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structures R and R+S.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.}\label{naa_om_em_R_MF_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_p_convergence_meanR_M_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structures R and R+S.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.}\label{naa_om_em_R_ME_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_p_convergence_meanR_M_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structures R and R+S.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is fixed at the true value.}\label{naa_om_em_BH_MF_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_p_convergence_BH_M_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structures R and R+S.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is estimated.}\label{naa_om_em_BH_ME_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_p_convergence_BH_M_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+M.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.}\label{M_om_em_R_MF_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_p_convergence_meanR_M_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+M.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.}\label{M_om_em_R_ME_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_p_convergence_meanR_M_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+M.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is fixed at the true value.}\label{M_om_em_BH_MF_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_p_convergence_BH_M_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+M.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is estimated.}\label{M_om_em_BH_ME_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_p_convergence_BH_M_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+Sel.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.}\label{Sel_om_em_R_MF_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_p_convergence_meanR_M_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+Sel.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.}\label{Sel_om_em_R_ME_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_p_convergence_meanR_M_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+Sel.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is fixed at the true value.}\label{Sel_om_em_BH_MF_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_p_convergence_BH_M_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+Sel.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is estimated.}\label{Sel_om_em_BH_ME_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_p_convergence_BH_M_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+q.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.}\label{q_om_em_R_MF_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_p_convergence_meanR_M_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+q.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.}\label{q_om_em_R_ME_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_p_convergence_meanR_M_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+q.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is fixed at the true value.}\label{q_om_em_BH_MF_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_p_convergence_BH_M_fixed.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+q.
+vertical lines represent 95\% confidence intervals.
+All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is estimated.}\label{q_om_em_BH_ME_convergence}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_p_convergence_BH_M_estimated.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R and R+S process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.} \label{naa_om_proportion_best_aic_R_MF}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_proportion_best_aic_R_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R and R+S process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.} \label{naa_om_proportion_best_aic_R_ME}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_proportion_best_aic_R_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R and R+S process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate a stock-recruit relationship and and M is fixed at the true value.} \label{naa_om_proportion_best_aic_SR_MF}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_proportion_best_aic_SR_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R and R+S process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate a stock-recruit relationship and and M is estimated.} \label{naa_om_proportion_best_aic_SR_ME}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_proportion_best_aic_SR_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+M process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.} \label{M_om_proportion_best_aic_R_MF}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_proportion_best_aic_R_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+M process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.} \label{M_om_proportion_best_aic_R_ME}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_proportion_best_aic_R_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+M process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate a stock-recruit relationship and and M is fixed at the true value.} \label{M_om_proportion_best_aic_SR_MF}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_proportion_best_aic_SR_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+M process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate a stock-recruit relationship and and M is estimated.} \label{M_om_proportion_best_aic_SR_ME}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_proportion_best_aic_SR_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+Sel process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.} \label{Sel_om_proportion_best_aic_R_MF}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_proportion_best_aic_R_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+Sel process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.} \label{Sel_om_proportion_best_aic_R_ME}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_proportion_best_aic_R_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+Sel process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate a stock-recruit relationship and and M is fixed at the true value.} \label{Sel_om_proportion_best_aic_SR_MF}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_proportion_best_aic_SR_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+Sel process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate a stock-recruit relationship and and M is estimated.} \label{Sel_om_proportion_best_aic_SR_ME}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_proportion_best_aic_SR_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+q process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.} \label{q_om_proportion_best_aic_R_MF}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_proportion_best_aic_R_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+q process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.} \label{q_om_proportion_best_aic_R_ME}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_proportion_best_aic_R_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+q process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate a stock-recruit relationship and and M is fixed at the true value.} \label{q_om_proportion_best_aic_SR_MF}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_proportion_best_aic_SR_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Proportion of simulated data sets from OMs with R+q process errors where fitted estimating models had lowest marginal AIC.
+All estimating models estimate a stock-recruit relationship and and M is estimated.} \label{q_om_proportion_best_aic_SR_ME}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_proportion_best_aic_SR_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\clearpage
+
+\begin{figure}
+\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
+Operating and estimating models have matching R or R+S process error structures.
+Estimating models assume mean M is fixed at the true value.}\label{naa_om_MF_BH_glm_AIC_plots}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_MF_pred_BH_best.png}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
+Operating and estimating models have matching R or R+S process error structures.
+Estimating models allow estimation of mean M.}\label{naa_om_ME_BH_glm_AIC_plots}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_ME_pred_BH_best.png}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
+Operating and estimating models have matching R+M process error structures.
+Estimating models assume mean M is fixed at the true value.}\label{M_om_MF_BH_glm_AIC_plots}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_MF_pred_BH_best.png}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
+Operating and estimating models have matching R+M process error structures.
+Estimating models allow estimation of mean M.}\label{M_om_ME_BH_glm_AIC_plots}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_ME_pred_BH_best.png}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
+Operating and estimating models have matching R+Sel process error structures.
+Estimating models assume mean M is fixed at the true value.}\label{Sel_om_MF_BH_glm_AIC_plots}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_MF_pred_BH_best.png}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
+Operating and estimating models have matching R+Sel process error structures.
+Estimating models allow estimation of mean M.}\label{Sel_om_ME_BH_glm_AIC_plots}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_ME_pred_BH_best.png}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
+Operating and estimating models have matching R+q process error structures.
+Estimating models assume mean M is fixed at the true value.}\label{q_om_MF_BH_glm_AIC_plots}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_MF_pred_BH_best.png}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
+Operating and estimating models have matching R+q process error structures.
+Estimating models allow estimation of mean M.}\label{q_om_ME_BH_glm_AIC_plots}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_ME_pred_BH_best.png}
+\end{center}
+\end{figure}
+
+\clearpage
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R and R+S process errors.
+Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{naa_om_em_R_MF_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_R_MF_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R and R+S process errors.
+Estimation models do not assume a stock-recruit relationship and estimate M.}\label{naa_om_em_R_ME_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_R_ME_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R and R+S process errors.
+Estimation models assume a BH stock-recruitment function and M is fixed at the true value.}\label{naa_om_em_BH_MF_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_BH_MF_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R and R+S process errors.
+Estimation models assume a BH stock-recruitment function and estimate M.}\label{naa_om_em_BH_ME_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_BH_ME_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+M process errors.
+Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{M_om_em_R_MF_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_R_MF_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+M process errors.
+Estimation models do not assume a stock-recruit relationship and estimate M.}\label{M_om_em_R_ME_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_R_ME_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+M process errors.
+Estimation models assume a BH stock-recruitment function and M is fixed at the true value.}\label{M_om_em_BH_MF_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_BH_MF_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+M process errors.
+Estimation models assume a BH stock-recruitment function and estimate M.}\label{M_om_em_BH_ME_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_BH_ME_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+Sel process errors.
+Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{Sel_om_em_R_MF_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_R_MF_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+Sel process errors.
+Estimation models do not assume a stock-recruit relationship and estimate M.}\label{Sel_om_em_R_ME_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_R_ME_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+Sel process errors.
+Estimation models assume a BH stock-recruitment function and M is fixed at the true value.}\label{Sel_om_em_BH_MF_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_BH_MF_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+Sel process errors.
+Estimation models assume a BH stock-recruitment function and estimate M.}\label{Sel_om_em_BH_ME_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_BH_ME_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+q process errors.
+Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{q_om_em_R_MF_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_R_MF_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+q process errors.
+Estimation models do not assume a stock-recruit relationship and estimate M.}\label{q_om_em_R_ME_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_R_ME_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+q process errors.
+Estimation models assume a BH stock-recruitment function and M is fixed at the true value.}\label{q_om_em_BH_MF_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_BH_MF_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+q process errors.
+Estimation models assume a BH stock-recruitment function and estimate M.}\label{q_om_em_BH_ME_relbias_ssb}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_BH_ME_relbias_ssb.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of Beverton-Holt stock recruit parameters for estimating models with correct process error assumptions fitted to data sets simulated with R and R+S process errors.
+Estimation models either assume the correct natural mortality rate or estimate it.}\label{naa_om_SR_relbias}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_SR_relerror.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of Beverton-Holt stock recruit parameters for estimating models with correct process error assumptions fitted to data sets simulated with R+M process errors.
+Estimation models either assume the correct natural mortality rate or estimate it.}\label{M_om_SR_relbias}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_SR_relerror.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of Beverton-Holt stock recruit parameters for estimating models with correct process error assumptions fitted to data sets simulated with R+Sel process errors.
+Estimation models either assume the correct natural mortality rate or estimate it.}\label{Sel_om_SR_relbias}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_SR_relerror.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of Beverton-Holt stock recruit parameters for estimating models with correct process error assumptions fitted to data sets simulated with R+q process errors.
+Estimation models either assume the correct natural mortality rate or estimate it.}\label{q_om_SR_relbias}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_SR_relerror.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of (mean) natural mortality rate for estimating models with correct process error assumptions fitted to data sets simulated with R and R+S process errors.
+Estimation models assumed either no stock-recruit relationship or estimated a Beverton-Holt relationship.}\label{naa_om_M_relbias}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_M_relerror.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of (mean) natural mortality rate for estimating models with correct process error assumptions fitted to data sets simulated with R+M process errors.
+Estimation models assumed either no stock-recruit relationship or estimated a Beverton-Holt relationship.}\label{M_om_M_relbias}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_M_relerror.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of (mean) natural mortality rate for estimating models with correct process error assumptions fitted to data sets simulated with R+Sel process errors.
+Estimation models assumed either no stock-recruit relationship or estimated a Beverton-Holt relationship.}\label{Sel_om_M_relbias}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_M_relerror.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of (mean) natural mortality rate for estimating models with correct process error assumptions fitted to data sets simulated with R+q process errors.
+Estimation models assumed either no stock-recruit relationship or estimated a Beverton-Holt relationship.}\label{q_om_M_relbias}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_M_relerror.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R and R+S process errors.
+Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{naa_om_em_R_MF_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_mohns_rho_R_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R and R+S process errors.
+Estimation models do not assume a stock-recruit function and M estimated.}\label{naa_om_em_R_ME_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_mohns_rho_R_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R and R+S process errors.
+Estimation models assume a B-H stock-recruit function and M is fixed at the true value.}\label{naa_om_em_BH_MF_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_mohns_rho_BH_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R and R+S process errors.
+Estimation models assume a B-H stock-recruit function and M estimated.}\label{naa_om_em_BH_ME_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{naa_om_mohns_rho_BH_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+M process errors.
+Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{M_om_em_R_MF_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_mohns_rho_R_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+M process errors.
+Estimation models do not assume a stock-recruit function and M estimated.}\label{M_om_em_R_ME_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_mohns_rho_R_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+M process errors.
+Estimation models assume a B-H stock-recruit function and M is fixed at the true value.}\label{M_om_em_BH_MF_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_mohns_rho_BH_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+M process errors.
+Estimation models assume a B-H stock-recruit function and M estimated.}\label{M_om_em_BH_ME_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{M_om_mohns_rho_BH_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+Sel process errors.
+Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{Sel_om_em_R_MF_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_mohns_rho_R_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+Sel process errors.
+Estimation models do not assume a stock-recruit function and M estimated.}\label{Sel_om_em_R_ME_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_mohns_rho_R_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+Sel process errors.
+Estimation models assume a B-H stock-recruit function and M is fixed at the true value.}\label{Sel_om_em_BH_MF_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_mohns_rho_BH_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+Sel process errors.
+Estimation models assume a B-H stock-recruit function and M estimated.}\label{Sel_om_em_BH_ME_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{Sel_om_mohns_rho_BH_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+q process errors.
+Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{q_om_em_R_MF_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_mohns_rho_R_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+q process errors.
+Estimation models do not assume a stock-recruit function and M estimated.}\label{q_om_em_R_ME_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_mohns_rho_R_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+q process errors.
+Estimation models assume a B-H stock-recruit function and M is fixed at the true value.}\label{q_om_em_BH_MF_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_mohns_rho_BH_MF.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+q process errors.
+Estimation models assume a B-H stock-recruit function and M estimated.}\label{q_om_em_BH_ME_mohns_rho}
+\begin{center}
+\includegraphics[width = \textwidth]{q_om_mohns_rho_BH_ME.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
 \end{document}

### 2024-01-19 Project_0/code/relbias_plots.R

@@ -184,7 +184,7 @@ M_om_df.ems <- get_df_ems("M")
 Sel_om_df.ems <- get_df_ems("Sel")
 q_om_df.ems <- get_df_ems("q")
 
-make_plot_df <- function(type = "naa", res = all_naa_om_relssb, is_SE = FALSE) {
+make_plot_df <- function(type = "naa", res = all_naa_om_relssb, is_SE = FALSE, ylims = c(-0.4,0.4)) {
   library(reshape2)
   res <- melt(res)
   names(res) <- c("year", "column", "value", "em", "sim","om")
@@ -211,6 +211,7 @@ make_plot_df <- function(type = "naa", res = all_naa_om_relssb, is_SE = FALSE) {
   }
   print(head(df.oms))
   df <- df %>% pivot_wider(names_from = type, values_from = stats) %>% as.data.frame
+  df$outside <- df$ymax < min(ylims) | df$ymin > max(ylims)
   df$em_pe = df.ems$pe#[df$em]
   df <- cbind(df, df.ems[df$em,])
   df <- cbind(df, df.oms[df$om,])
@@ -265,201 +266,288 @@ library(dplyr)
 library(tidyr)
 all_naa_om_relssb <- readRDS(file = here("Project_0","results", "all_naa_relssb_results.RDS"))
 library(reshape2)
-temp <- melt(all_naa_om_relssb)
-names(temp) <- c("year", "column", "value", "em", "sim","om")
-temp <- temp %>% mutate(column = recode(column,
-    "1" = "relbias",
-    "2" = "cv",
-    "3" = "in_ci",
-  )) %>% as.data.frame
-temp <- temp %>% pivot_wider(names_from = column, values_from = value) %>% as.data.frame
+# temp <- melt(all_naa_om_relssb)
+# names(temp) <- c("year", "column", "value", "em", "sim","om")
+# temp <- temp %>% mutate(column = recode(column,
+#     "1" = "relbias",
+#     "2" = "cv",
+#     "3" = "in_ci",
+#   )) %>% as.data.frame
+# temp <- temp %>% pivot_wider(names_from = column, values_from = value) %>% as.data.frame
+
+# temp$outside <- abs(temp$relbias) > 0.4 
+# temp2 <- temp %>% group_by(om, em, sim) %>% summarise(notplot = all(outside))
 
 estM <- c(FALSE,FALSE,TRUE,TRUE)
 estSR<- c("Mean R", "B-H", "Mean R", "B-H")
[…]

### 2024-01-19 Project_0/paper/paper.Rmd

@@
 
-## Abstract {-}
-
-### Keywords {-}
+## Summary {-}
+
+We conducted a simulation study where we simulated process errors and observations for 72 operating models with alternative assumptions about fishing history, degree of uncertainty in index and age composition observations, type (recruitment, survival, fishery selectivity, catchability and natural mortality), degree of variation, and correlation of process errors.
+We fit 20 different estimating models to each of 100 simulated set of observations with alternative assumptions type and correlation structure of process errors, (mean) natural mortality was known or estimated, and a B-H stock recruit relationship was assumed or not.
+
+Across simulations we summarized probability of convergence of fitted models, accuracy of marginal AIC in determining the correct process error assumption and it ability to determine the Beverton-Holt stock recruit relationship, bias in annual spawning stock biomass, in estimation of natural mortality, and in stock-recruit relationship parameters, and severity of retrospective patterns for estimation models.
+
+Alternative measures of convergence performed differently.
+Invertible hessians and resulting standard error estimation was possible when criteria based on the gradient of the optimized log-likelihood with respect to the fixed effects parameters.
+Using hessian-based convergence, probability of convergence was best for models that assumed the correct source of process error, assumed M was known, and did not assume stock-recruit relationships.
+
+Using marginal AIC provided most accurate inferences about the process errors on recruitment and survival and selectivity, in that AIC preferred EMs with assumed process errors that matched OMs, and EMs with those assumed process errors were not preferred when alternative process errors were assumed in the OMs.
+However, when the true process errors were more variable, AIC accuracy increased to a useful level.
+We found AIC more accurately determined a B-H stock recruit relationship rather than the null model without a S-R relationship when there was low variability in recruitment, low variability in survival random effects, and higher variation in spawning biomass over the time series.
+When the (mean) natural mortality rate was estimated, we found large bias and uncertainty much more likely in model output such as spawning stock biomass.
+
+Bias in spawning stock biomass estimation was generally low for estimating models that assumed the correct source of process error when there was lower observation error.
+Reliable estimation of stock-recruit relationship parameters only appears possible in ideal situations with lower observation errors in age composition and indices, lower variability in recruitment process errors and large contrast in spawning biomass over time.
+We found little evidence of bias for many OM process error assumptions when there was contrast in fishing pressure even when there was greater observation error although it can lead to more variable estimates of natural mortality.
+For OMs where there was bias in natural mortality due to high observation error, estimating the stock-recruit relationship seemed to remove the bias.
+However, estimation of natural mortality can cause large differences between the true and estimated SSB (that may be unbiased on average) when there is less contrast in fishing pressure over time and higher observation error.
+
+Retrospective patterns were generally weak for all estimation models regardless of the true source of process error, but they can be expected for recruitment even for the correct process error assumptions when observation error is high.
+When models did exhibit some retrospective pattern, estimating the mean natural mortality rate tended to remove it.
 
@@
 
-Application of state-space models in fisheries stock assessment and management has expanded dramatically within ICES, Canada and the Northeast US \citep{nielsenberg14,cadigan16,stockmiller21}.
-State-space approaches that use random effects to parameterize process errors is considered best practice and a requirement for the next generation of stock assessment models.
+Application of state-space models in fisheries stock assessment and management has expanded dramatically within ICES, Canada, and the Northeast US \citep{nielsenberg14,cadigan16,stockmiller21}.
+State-space approaches that use random effects to parameterize process errors is considered best practice and a requirement for the next generation of stock assessment models \citep{hoyleetal22, punt23}.
 
@@
 Furthermore, there is a wide range of potential random effects structures and model parameters that can be treated as random effects in assessment models.
-We know relatively little about the statistical reliability of such models or the ability of information criteria to distinguish among such alternative structures.
+We know relatively little about the factors affecting statistical reliability of such models or the ability of information criteria to distinguish among such alternative structures.
 
@@
 
-Convergence results can be useful for understanding how bad convergence in applications to real data might direct the practitioner to which alternative random effects configurations is be more appropriate.
-The type of convergence that we might use as a diagnostic is important.
-And what rule of thumb we use for certain criteria e.g.
-1e-6?
-large SEs/CVs?
-
-Here we conduct a simulation study with operating models varying by degree of observation error uncertainty, sources of process error (M, NAA, q, sel), and fishing history.
-The simulations from these operating models are fitted with estimating models that make alternative assumptions for sources of process error (M, NAA< q, sel), whether a stock-recruit model was estimated, and whether M is estimated.
+Here we conduct a simulation study with operating models varying by degree of observation error uncertainty, source and variability of process error, and fishing history.
+The simulations from these operating models are fitted with estimating models that make alternative assumptions for sources of process error, whether a stock-recruit model was estimated, and whether M is estimated.
 We evaluate whether AIC can correctly determine the correct source of process error and stock effects on recruitment.
@@
 
-Used the WHAM package  \citep[][commit 77bbd94]{stockmiller21} (Miller and Stock 2020).
-This packages has also been used to configure operating and estimating models for closed loop simulations evaluating index-based assessment methods \citep{legaultetal23} and is used for management of haddock, butterfish, plaice, bluefish in the Northeast US.
+We used the Woods Hole Assessment Model (WHAM) to configure operating and estimation models in our simulation study \citep[]{millerstock20,stockmiller21}.
+WHAM is an R package freely available as a github repository.
+For this study we used [version 1.0.6.9000, commit 77bbd94](https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3).
+This packages has also been used to configure operating and estimating models for closed loop simulations evaluating index-based assessment methods \citep{legaultetal23} and is used for management of haddock, butterfish, American plaice, bluefish, Atlantic cod, and black sea bass in the Northeast US.
 
@@
 
-78 total operating models.
-We simulated 100 data sets for each operating model.
-For each simulated data set we fit a set of 20 estimating models.
+In total, we configured 72 operating models with alternative assumptions about the source and variability of process errors, level of observation error in indices and age composition data, and contrast in fishing pressure over time.
+We fitted 20 estimating models to observations from each of 100 simulations where process errors were also simulated.
+Estimating models made alternative assumptions about the source of process errors and whether natural mortality (or the mean for models with process error in natural mortality) was estimated and whether a Beverton-Holt stock recruit relationship was estimated within the estimation model.
+Details of each of the operating and estimating models are described below.
+
+We did not use the bias-correction feature for process errors or observations described by \citet{stockmiller21} for operating and estimating models.Simulations and model fitting were all carried out on the University of Massachusetts Green High-Performance Computing Cluster.
+All code we used to perform the simulation study and summarize results can be found at https://github.com/timjmiller/SSRTWG/tree/main/Project_0/code.
 
@@
 The population model tracks 10 age classes: ages 1 to 10+ and we assume spawning occurs 1/4 of the way through the year.
-The maturity at age was a logistic curve with a50 = 2.89 and slope = 0.88 and assumed known in all estimation models (Figure \ref{om_maturity}).
-
-\begin{figure}
-\caption{The proportion mature at age assumed for the population in all operating and estimating models.}\label{om_maturity}
-\begin{center}
-\includegraphics[width = \textwidth]{om_maturity.png}
-\end{center}
-\end{figure}
+The maturity at age was a logistic curve with $a_{50}$ = 2.89 and slope = 0.88 and assumed known in all estimation models (Figure \ref{om_maturity}).
 
@@
 
-\begin{figure}
-\caption{The weight at age assumed for the population in all operating and estimating models.}\label{om_waa}
-\begin{center}
-\includegraphics[width = \textwidth]{om_waa.png}
-\end{center}
-\end{figure}
-
 We assumed a Beverton-Holt stock recruit function with constant pre-recruit mortality parameters for all operating models.
@@
 
-\begin{figure}
-\caption{The Beverton-Holt stock recruit relationship assume for all operating models.}\label{om_sr}
-\begin{center}
-\includegraphics[width = \textwidth]{om_sr.png}
-\end{center}
-\end{figure}
-
 Two alternative fishing histories were used for operating models.
@@
 
-\begin{figure}
-\caption{The selectivity at age assumed for the fishing fleet in operating models without selectivity random effects.
-The mean selectivity at age across time for the fishing fleet in operating models with selectivity random effects.}\label{om_mean_selectivity}
-\begin{center}
-\includegraphics[width = \textwidth]{om_mean_selectivity.png}
-\end{center}
-\end{figure}
-
 ### Indices {-}
@@
 The factorial combinations of fishing histories, observation error assumptions, and marginal variance assumptions for recruitment and survival results in 24 different operating models with random effects on recruitment and survival (Table \ref{naa_om_table}).
-
-\begin{landscape}
-\begin{table}
-\caption{Distinguishing characteristics of the operating models with random effects on survival.
-Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
-Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.}\label{naa_om_table}
-{\footnotesize \input{naa_om_tab}}
-\end{table}
-\end{landscape}
 
@@
 The values for the marginal variance and correlation parameters assumed in the operating models are given in Table \ref{M_om_table}.
-
-\begin{landscape}
-\begin{table}
-\caption{Distinguishing characteristics of the operating models with random effects on natural mortality.
-Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
-Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
-For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}\label{M_om_table}
-{\input{M_om_tab}}
-\end{table}
-\end{landscape}
 
@@
 The values for the marginal variance and correlation parameters assumed in the operating models are given in Table \ref{sel_om_table}.
-
-\begin{landscape}
-\begin{table}
-\caption{Distinguishing characteristics of the operating models with random effects on selectivity.
-Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
-Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
-For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}\label{sel_om_table}
-{\input{Sel_om_tab}}
-\end{table}
-\end{landscape}
 
@@
 
+## Estimating models {-}
+
+For each data set simulated from an operating model 20 estimating models were fit.
+A total of 32 different estimating models were fit across all operating models where the subset of 20 depended on the source of process error (Table \ref{em_table}).
+The first 20 estimating models in Table \ref{em_table} were fit to simulate data sets from R and R+S operating models.
+Estimating models 5 to 24 in Table \ref{em_table} were fit to simulate data sets from R+M operating models.
+Estimating models 5 to 20 and 25-28 in Table \ref{em_table} were fit to simulate data sets from R+Sel operating models.
+Finally, estimating models 5 to 20 and 29-32 in Table \ref{em_table} were fit to simulate data sets from R+q operating models.
+The observation error variance of aggregate catch and indices were all assumed known at the true values.
+
+## Measures of reliability {-}
+
+The first measure of reliability we investigated was frequency of convergence when fitting each estimating model to the simulated data sets.
+There are various ways to assess convergence of the fit. We summarized 5 alternative categories of convergence.
+
+1. Did the optimization routine (stats::nlminb) complete without error?
+2. Did the stats::nlminb convergence flag = 0 indicate successful convergence?
+3. Was the maximum absolute value of the gradient of the log-likelihood < $1\times10^{-6}$?
+4. Did TMB::sdreport provide non-NA values for all fixed effects standard errors?
+5. Did TMB::sdreport provide all standard errors < 100?
+
+The first convergence criterion assesses whether the model crashes.
+The third is a measure that assesses how flat the likelihood is at the optimized point of the likelihood surface.
+The fourth and fifth criteria are specific to the hessian-based standard error reporting provided by TMB.
+the TMB::sdreport function will sometimes return standard error estimates even when the calculated hessian is not invertible (fourth criterion).
+It will also provide large standard errors for some parameters that are not estimated well or are near bounds on the transformed scale that is of primary interest.
+For example, variance parameters for random effects can often be estimated near 0 when the fit suggests no variation in the random effects or, equivalently, that the model without random effects is better.
+
+### AIC for model selection {-}
+
+We measured the frequency of correct model selection using marginal AIC.
+For a given operating model we the set of models that were considered all made the same assumptions on whether or not to estimate M and whether a stock-recruit function is assumed.
+
+We also measured the frequency of correctly selecting models with Beverton-Holt stock recruit function assumed over models without the stock-recruit function (null model).
+Because the proportions of simulations where the correct model was chosen for a given set of operating model factors implied generally poor performance of AIC, we analyzed the effect of variation in SSB on the probability of lower AIC of the BH models.
+We fit binomial generalize linear models to the observed AIC choice for each data set as a function of log(SD(SSB)) and operating model.
+
+### Bias {-}
+
+We estimated relative bias 
+$$
+\widehat {\text{RB}}\left(\theta\right) = \frac{1}{N} \sum^N_{i = 1} \frac{\widehat \theta_i - \theta_i}{\theta_i}
+$$
+of annual SSB and fully-selected fishing mortality for each estimation model and constructed 95\% confidence intervals for the median relative bias using the binomial distribution approach as in \citet{millerhyun18} and \citet{stockmiller21}.
+We also estimated bias of stock-recruit parameters for estimation models that assumed the Beverton-Holt stock recruit function.
+
+### Mohn's $\rho$ {-}
+
+We estimated Mohn's $\rho$ for SSB, fully-selected fishing mortality, and recruitment for each estimation model.
+We estimated 7 peels for each estimation model.
+We calculated median 95\% confidence intervals for Mohn's $\rho$ using the same methods as that for relative bias.
+
+# Results {-}
+
+## Convergence performance {-}
+
+### R, R+S operating models {-}
+
+R+S EMs fit to R OMs exhibited poor convergence for types 3, 4, and 5 regardless of whether M or a stock-recruit relationship was estimated (Figures \ref{naa_om_em_R_MF_convergence} to \ref{naa_om_em_BH_ME_convergence}).
+R EMs exhibited high convergence rates for all convergence types for both R and R+S OMs.
+Convergence rates were high for all convergence types for all EMs in R+S OMs with lower uncertainty in observations.
+Convergence for most convergence types generally declines most estimating models when mean-log natural mortality rate is estimated and/or a Beverton-Holt stock recruit relationship is estimated even when the process error assumptions of the estimating and operating models match.
+There can be relatively high convergence probability of type 4 (an invertible hessian) for EMs that do not have the correct process error assumed, but type 5 convergence (no very large standard errors estimated from hessian) for these estimating models is typically much lower.
+
+### R+M operating models {-}
+
+R+S EMs generally converged less reliably across convergence types than other EMs when fit to data generated from R+M OMs (Figures \ref{M_om_em_R_MF_convergence} to \ref{M_om_em_BH_ME_convergence}).
+However, using convergence type 5, all EMs converged poorly when OMs had low variability in natural mortality process errors or higher observation error uncertainty.
+Using convergence type 5, R+M EMs generally converged better than other EMs when observation error uncertainty was low and natural mortality process errors were more variable.
+Like R+S OMs, the probability of convergence generally declined for all EMs when mean log-natural mortality and/or a stock-recruit relationship was estimated.
+
+### R+Sel operating models {-}
+
+R+S EMs in particular converged poorly when R+Sel OMs had less variability in selectivity process errors or higher observation error uncertainty (Figures \ref{Sel_om_em_R_MF_convergence} to \ref{Sel_om_em_BH_ME_convergence}).
+R+Sel EMs generally converged better than other EMs for OMs with greater variability in process errors, lower observation error, and contrast in fishing pressure regardless of whether mean log natural mortality or a stock recruit relationship was estimated.
+
+### R+q operating models {-}
+
+Convergence of R+q EMs is generally better than that of other EMs for all convergence types when R+q OMs assume contrast in fishing history overFigures (\ref{q_om_em_R_MF_convergence} to \ref{q_om_em_BH_ME_convergence}).
+Convergence of R+S EMs is generally worse than that of other EMs across all OMs whether or not mean log-natural mortality or a stock recruit relationship is estimated.
+Again, convergence probablity generally declines for all EMs when mean log-natural mortality or a stock recruit relationship is estimated.
+
+## AIC performance for process error structure {-}
+
+### R, R+S operating models {-}
+
+Marginal AIC accurately determines the correct process error assumptions in estimating models when data are generated from R and R+S OMs, regardless of whether mean log-natural mortality or a stock recruit relationship is estimated (Figures \ref{naa_om_proportion_best_aic_R_MF} to \ref{naa_om_proportion_best_aic_SR_ME}).
+Adding estimation of mean log-natural mortality or a stock recruit relationship separately has a negligible effect on the accuracy of determining the correct process error assumption.
+When both are estimated, there is a noticeable reduction in accuracy when OMs have a constant fishing history, observation error is low and largest variability in recruitment process errors.
+
+### R+M operating models {-}
+
+Marginal AIC only accurately determined the correct process error model and correlation structure when observation error was low and variability in natural mortality process errors was high (Figures \ref{M_om_proportion_best_aic_R_MF} to \ref{M_om_proportion_best_aic_SR_ME}).
+Estimating the mean natural mortality rate reduced the accuracy of AIC for OMs that assumed natural mortality process errors were independent.
+For OMs with poor accuracy, AIC most frequently selected EMs with process errors in catchability or selectivity.
+Selection of R+S EMs was typically unlikely.
+
+### R+Sel operating models {-}
+
+Marginal AIC most accurately determined the correct source of process error and correlation structure for R+Sel OMs with low observation error (Figures \ref{Sel_om_proportion_best_aic_R_MF} to \ref{Sel_om_proportion_best_aic_SR_ME}).
+When there was low variability in selectivity process errors and high observation error, R+q or R+S EMs were more likely to have the best AIC.
+Whether mean log-natural morality or stock recruit relationships were estimated appeared to have little effect on the performance of marginal AIC.
+### R+q operating models {-}
+
+Marginal AIC most accurately determined the correct source of process error and correlation structure for R+q OMs with high variability in catchability process errors (Figures \ref{q_om_proportion_best_aic_R_MF} to \ref{q_om_proportion_best_aic_SR_ME}).
+The worst accuracy occurred for OMs with low variability in catchability process errors and high observation error.
+However, in these OMs, the marginal AIC accurately determined the correct source of process error (but not correlation structure) except when EMs estimated both mean log-natural morality and the stock recruit relationships and OMs assumed a constant fishing pressure.
+
+## AIC performance for stock-recruit relationship {-}
+
+<!--
+Tables of frequencies where AIC chooses correct stock-recruit assumption imply poor performance of AIC, but figures showing effect of variation in SSB demonstrate that large variation in SSB is needed to reduce Type II errors for marginal AIC.
+-->
+Our comparisons of model performance condition on assuming the true process error configuration is known (EM and OM process error assumptions match).
+Broadly, we found generally poor accuracy of AIC in selecting models assuming a Beverton-Holt stock recruit function over the null model without an assumed stock-recruit relationship for all OMs (Tables \ref{naa_om_em_R_BH_aic_table} to \ref{q_om_em_R_BH_aic_table}.
+However, we also found AIC to more accurately determine a stock-recruit relationship when there was greater variation in spawning biomass generated in the simulated populations.
+
+AIC accuracy for including the B-H stock-recruit relationship was estimated to be highest for models OMs and EMs with process errors only on recruitment and lowest variance of recruitment process errors (Figures \ref{naa_om_MF_BH_glm_AIC_plots} and \ref{naa_om_ME_BH_glm_AIC_plots}).
+Holding variability in SSB constant, accuracy of stock-recruit  model selection declined with greater variance of recruitment and survival process errors.
+Simultaneously estimating mean log-natural mortality has a small negative effect on AIC accuracy.
+
+For R+M OMs and EMs there appeared to little effect of correlation in the natural mortality process errors or estimating the mean log-natural mortality on accuracy of AIC in determining a stock-recruit function, holding variability in SSB constant (Figures \ref{M_om_MF_BH_glm_AIC_plots} and \ref{M_om_ME_BH_glm_AIC_plots}).
+
+The range of variation in SSB across simulated populations was lower for R+Sel operating models than R+S or R+M OMs.
+OMs with contrast in fishing pressure over time is the primary mechanism creating variation in SSB.
+However, there appears to be little effect of variability or correlation of selectivity process errors or whether mean log-natural mortality was estimated on the accuracy of AIC in selecting the stock recruit relationship (Figures \ref{Sel_om_MF_BH_glm_AIC_plots} and \ref{Sel_om_ME_BH_glm_AIC_plots})
+
+Like the R+Sel OMs, the range of variation in SSB across is only increased by allowing contrast in fishing pressure over time.
+There appears to be a slight decrease in accuracy in AIC when mean log-natural mortality is estimated and there is a more noticeable increase in accuracy for OMs with lower observation error in those cases.
+(Figures \ref{q_om_MF_BH_glm_AIC_plots} and \ref{q_om_ME_BH_glm_AIC_plots})
+
+## Bias {-}
+
+### R, R+S operating models {-}
+
+When mean log-natural mortality is known, all estimating models regardless of process error assumptions produce unbiased estimates of SSB over the entire time period for R OMs (Figure \ref{naa_om_em_R_MF_relbias_ssb}).
+For R+S OMs, only R+S EMs produce unbiased estimation of SSB and bias resulting from other EMs increased with larger variability in survival random effects.
+Estimating mean log-natural mortality generally produced greater variability in relative errors of SSB estimates and a change in sign of bias for EMs assuming process errors other than R and R+S (Figure \ref{naa_om_em_R_ME_relbias_ssb}).
+Estimating a stock-recruit relationship had little discernable effect on bias of SSB estimation (Figures \ref{naa_om_em_BH_MF_relbias_ssb}), but in combination with estimating mean log-natural mortality resulted in large bias of R+S EMs for R OMs with high observation error, constant fishing pressure and large variability in recruitment process errors \ref{naa_om_em_BH_ME_relbias_ssb}).
+
+### R+M operating models {-}
+
+When mean log-natural mortality is known, SSB estimation bias was small or non-existent for all EM process error assumptions when the OMs had low variability of the natural mortality process errors except a tendency toward negative bias in the terminal years for OMs with high observation error and a change in fishing pressure over time (Figure \ref{M_om_em_R_MF_relbias_ssb}).
+Errors in SSB estimation were most variable with high variability and autocorrelation of natural mortality process errors, but confidence intervals generally covered 0 indicating no evidence of bias.
+Estimating mean log-natural mortality results in large variability in SSB relative errors, particularly for OMs with high observation error and constant fishing pressure and all EMs produced biased estimation of SSB over all years when OMs had high observation error, a change in fishing pressure and low variability of natural mortality process errors (Figure \ref{M_om_em_R_ME_relbias_ssb}).
+Like R and R+S OMs, estimating a stock-recruit relationship had little discernible effect on SSB bias (Figures \ref{M_om_em_BH_MF_relbias_ssb} and \ref{M_om_em_BH_ME_relbias_ssb})
+
+### R+Sel operating models {-}
+
+When mean log-natural mortality is known and no stock-recruit relationship was estimated, SSB estimation bias was low for all EM process error assumptions (Figure \ref{Sel_om_em_R_MF_relbias_ssb}).
+The worst bias occurred for OMs with uncorrelated by more variable selectivity process errors and high observations error.
+As for other OMs, estimating mean log-natural mortality resulted in greater variability of the relative errors in SSB, but also more evidence of bias for EMS without the correct process error assumption (Figure \ref{Sel_om_em_R_ME_relbias_ssb}).
+Estimating a stock recruit relationship resulted in greater estimation of SSB in the terminal years and sometimes less bias when OMs had high observation error and a change in fishing pressure over time and the effect was more pronounced when mean log-natural mortality was estimated (Figures \ref{Sel_om_em_BH_MF_relbias_ssb} and \ref{Sel_om_em_BH_ME_relbias_ssb}).
+
+### R+q operating models {-}
+
+When mean log-natural mortality is known and no S-R relationship was estimated, SSB estimation bias was low for all EM process error assumptions for all R+q OMs with low variability in catchability process errors (Figure \ref{q_om_em_R_MF_relbias_ssb}).
+With larger variability in catchability process errors, bias was low for all EMs except those configured with R+M process errors and OMs assumed low observation error.
+Many EMs also showed some negative bias for terminal SSB in OMs with higher observation error and a change in fishing pressure over time.
+As for other OMs, mean log-natural mortality resulted in greater variablity of the relative errors in SSB, but also more evidence of bias for some EMs without the correct process error assumption (Figure \ref{q_om_em_R_ME_relbias_ssb})
+Estimating a stock recruit relationship resulted in less bias in SSB in terminal years when OMs had high observation error and a change in fishing pressure over time (Figures \ref{q_om_em_BH_MF_relbias_ssb} and \ref{q_om_em_BH_ME_relbias_ssb}).
+
+### Beverton-Holt parameter estimation {-}
+
+In R and R+S OMs, EMs with the correct assumptions about process errors, provided the least biased estimation of Beverton-Holt stock-recruit relationship parameters when there was low observation error, a change in fishing pressure over time, and lower variability of recruitment process errors, and higher variability survival process errors, and there was little effect of estimating natural mortality (Figure \ref{naa_om_SR_relbias}).
+For other R and R+S OMs, estimating natural mortality often resulted in less biased estimation of stock-recruit parameters.
+The range of confidence intervals for the median parameter bias suggests large variability of the parameter estimates.
+
+In R+M OMs, the most accurate estimation of stock-recruit parameters occurred when there was a change in fishing pressure combined with either low variability in natural mortality process errors and high observation error or vice versa (Figure \ref{M_om_SR_relbias}).
+Relative to the R, and R+S OMs, there was even less effect of estimating mean log-natural mortality on estimation bias for the stock- recruit relationship parameters, but similarly there was generally large variability of the parameter estimates.
+
+In R+Sel OMs, bias for stock-recruit parameters was very large for OMs with constant fishing pressure (Figure \ref{Sel_om_SR_relbias}).
+Less bias in parameter estimation occurred for OMs with a change in fishing pressure over time.
+There was little effect of estimating natural mortality.
+Patterns in bias of stock-recruit parameter estimation in R+q OMs, were very similar to those in R+Sel OMs with poor estimation of these parameters in OMs with constant fishing pressure (Figure \ref{q_om_SR_relbias}).
+Like other OMs, there was generally large variability of the parameter estimates.
+
+### (Mean) natural mortality rate estimation {-}
+
+The mean natural mortality rate is estimated pretty accurately for R and R+S OMs when EMs use the correct process error configuration and there is a change in fishing pressure over time and observation error is low (Figure \ref{naa_om_M_relbias}).
+Estimation is a little more variable when observation error is higher or fishing pressure is constant, but when OMs have both, estimation is extremely variable.
+There is little effect of estimating stock-recruit relationships on the bias in estimating natural mortality.
+For R+M OMs, effects on estimating mean log-natural mortality were similar to the R and R+S OMs: best estimation with a change in fishing pressure and low observation error and worst with no change in fishing pressure and higher observation error (Figure \ref{M_om_M_relbias}).
+The same patterns also occurred for R+Sel and R+q OMs (Figures \ref{Sel_om_M_relbias} and \ref{q_om_M_relbias}).
+
+## Mohn's $\rho$ {-}
+
+### R, R+S operating models {-}
+
+When natural mortality is known and no stock recruit relationship is estimated, median absolute Mohn's $\rho$ was less than 0.25 for all EMS fit to simulated data sets from R and R+S OMs (Figure \ref{naa_om_em_R_MF_mohns_rho}).
+The strongest median Mohn's $\rho$ values for F and SSB were obtained for EMs that did not match the R+S OMs with the largest variability in survival process errors, change in fishing pressure over time, and higher observation error.
+Mohn's $\rho$ was also greater than 0 for recruitment for all EMS, including the matching EM, fit to R OMs that had the largest variability in recruitment random effects and high observation error.
+Estimating natural mortality removed the retrospective patterns in F and SSB in R+S OMs with the largest variability in survival process errors, change in fishing pressure over time, and higher observation error (Figure \ref{naa_om_em_R_ME_mohns_rho}).
+Estimating a stock-recruit relationship had no discernible effect on the retrospective patterns (Figure \ref{naa_om_em_BH_MF_mohns_rho}).
+Estimating natural mortality and the stock-recruit relationship increased the variablity of Mohn's $\rho$ for recruitment particularly for OMs with high observation error (Figure \ref{naa_om_em_BH_ME_mohns_rho}).
+
+### R+M operating models {-}
+
+When mean log-natural mortality is known and no stock recruit relationship is estimated, median abslolute Mohn's $\rho$ was less than 0.125 for all EMS fit to simulated data sets from R+M OMs (Figure \ref{M_om_em_R_MF_mohns_rho}).
+The strongest median Mohn's $\rho$ values for F, SSB, and recruitment were obtained for EMs other than R+M with autocorrelation assumed for OMs with a change in fishing pressure over time, and higher observation error.
+Like R and R+S OMs, estimating natural mortality removed the retrospective patterns in the respective OMs (Figure \ref{M_om_em_R_ME_mohns_rho}).
+Estimating a stock-recruit relationship also had the same effect as estimating the mean log-natural mortality rate, but Mohn's $\rho$ values were more variable for R+M EMs with autocorrelation assumed (Figure \ref{M_om_em_BH_MF_mohns_rho}).
+Estimating natural mortality and the stock-recruit relationship increased the variablity of Mohn's $\rho$ for F, SSB, and recruitment, particularly for OMs with high observation error and constant fishing pressure over time (Figure \ref{M_om_em_BH_ME_mohns_rho}).
+
+### R+Sel operating models {-}
+
+Similar to R+M OMs, when mean log-natural mortality is known and no stock recruit relationship is estimated, median abslolute Mohn's $\rho$ was less than 0.125 for all EMS fit to simulated data sets from R+Sel OMs (Figure \ref{Sel_om_em_R_MF_mohns_rho}).
+The strongest median Mohn's $\rho$ values for F, SSB, and recruitment were obtained for all EMs  with OMs that had a change in fishing pressure over time, and higher observation error.
+Estimating natural mortality increased the variability in Mohn's $\rho$ values, particularly for OMs with higher observation error and constant fishing (Figure \ref{Sel_om_em_R_ME_mohns_rho}).
+Estimating a stock-recruit relationship also removed any tendency for retrospective patterns in OMs with a change in fishing pressure and higher observation error (Figure \ref{Sel_om_em_BH_MF_mohns_rho}).
+Estimating natural mortality and the stock-recruit relationship further increased the variablity of Mohn's $\rho$ for F, SSB, and recruitment, particularly for OMs with high observation error and constant fishing pressure over time (Figure \ref{Sel_om_em_BH_ME_mohns_rho}).
+
+### R+q operating models {-}
+
+Similar to R+M and R+Sel OMs, when mean log-natural mortality is known and no stock recruit relationship is estimated, median abslolute Mohn's $\rho$ was less than 0.125 for all EMS fit to simulated data sets from R+q OMs (Figure \ref{q_om_em_R_MF_mohns_rho}).
+Also similar to other OMs, the strongest median Mohn's $\rho$ values for F, SSB, and recruitment were obtained for with OMs that had a change in fishing pressure over time, and higher observation error.
+Estimating natural mortality increased the variability in Mohn's $\rho$ values, particularly for OMs with constant fishing and autocorrelation of process errors in catchability (Figure \ref{q_om_em_R_ME_mohns_rho}).
+Similar to other OMs, estimating a stock-recruit relationship removed any tendency for retrospective patterns in OMs with a change in fishing pressure and higher observation error (Figure \ref{q_om_em_BH_MF_mohns_rho}).
+Like R+Sel OMs, estimating natural mortality and the stock-recruit relationship further increased the variablity of Mohn's $\rho$ for F, SSB, and recruitment, particularly for OMs with high observation error and constant fishing pressure over time (Figure \ref{q_om_em_BH_ME_mohns_rho}).
+
+# Discussion {-}
+
+## Convergence {-}
+
+Convergence results can be useful for understanding how bad convergence in applications to real data might direct the practitioner to which alternative random effects configurations is be more appropriate.
+Therefore, the type of convergence that we might use as a diagnostic is important.
+It is common during assessment model fitting to check that the maximum absolute gradient component is less than some threshold.
+Our Type 3 convergence represents this and we used 1e-6 as the threshold, but there is no standard.
+Furthermore we found the hessian at the optimized log-likelihood can often be invertible (Convergence types 4 and 5) when the gradient approach to convergence fails.
+The probability of convergence type 5 cannot be greater than that of type 4 because type 5 is a subset where none of the standard errors obtained by inverting the hessian are greater than a specified value (100 in this case).
+We presume that the parameters that cause type 4 convergence to fail will also typically be the same as those that cause type 5 to fail.
+Perhaps a higher gradient threshold for Type 3 convergence would be more consistent with hessian-based convergence, but the gradient and hessian-based standard errors should be inspected.
+Parameters that are not well estimated can lead the assessment scientist toward a more appropriate and often more parsimonious model configuration.
+For example, variance parameters that are estimated on log scale as low negative number (variance is close to 0) can happen for random effects estimated with little or no variability.
+These may sometimes lead a hessian to not be invertible or due to numerical issues may be invertible, but provide very large standard errors.
+More generally, we can expect lower likelihood of convergence when estimating natural mortality or stock-recruit relationships because the typical data in assessment models does not strongly inform these parameters.
+
+## AIC {-}
+
+Among the process error configurations we used in operating models, we found AIC to be accurate for process errors on recruitment and survival.
+Fitting models to other OMs rarely preferred R+S EMs and R and R+S EMs were nearly always selected for the matching OMs.
+For other sources of process error, accuracy of AIC was improved to useful levels when there was larger variability in the process errors and/or lower observation error.
+Therefore, when data are relatively good and an there is large variability in the unknown process errors, AIC is likely to perform well.
+
+AIC generally performs weakly in determining stock recruit relationships unless there is large contrast in SSB.
+Interestingly, \citet{milleretal16} found AIC to prefer a stock-recruit relationship for SNEMA yellowtail flounder where there was a large change in stock size estimated over time and flatfish are well observed by the bottom trawl indices relative to some other guilds such as semi-pelagic species.
+Although we did not compare models with alternative stock-recruit relationships, we do not expect AIC to perform any better distinguishing these relationships.
+Our results comport well with those of \citet{devalpinehastings02} where state-space stock-recruit models were fit just to SSB and recruitment "observations".
+
+## Bias {-}
+
+As we might expect bias in all parameters and assessment output was generally less with lower observation error in indices and age composition.
+Reliable estimation of stock-recruit relationship parameters only appears possible in ideal situations with lower observation errors in age composition and indices, lower variability in recruitment process errors and large contrast in spawning biomass over time.
+The latter can occur in with larger variability in survival or natural mortality process errors, variation in fishing pressure.
+
+Estimation of natural mortality (or the mean when there are process errors on natural mortality) appears more feasible.
+We found low or no evidence of bias for many OM process error assumptions when there was contrast in fishing pressure even when there was greater observation error although it can lead to more variable estimates of natural mortality.
+For OMs where there was bias in natural mortality due to high observation error (R+Sel, R+q), estimating the stock-recruit relationship seemed to remove the bias.
+However, estimation of natural mortality can cause large differences between the true and estimated SSB (that may be unbiased on average) when there is less contrast in fishing pressure over time and higher observation error.
+
+## Retrospective patterns {-}
+
+Higher observation error with variation in fishing pressure can produce retrospective patterns even when the the process error assumption is correct, but this retrospective pattern can be reduced if M or the correct stock recruit relationship is estimated.
+Retrospective patterns tended to be more variable for recruitment and can be large even when the estimating model is correct.
+Therefore, we recommend emphasis on inspection of retrospective patterns primarily for SSB and F, but further research on retrospective patterns in other assessment model parameters and derived output may be beneficial.
+
+# Acknowledgements {-}
+
+This work was funded by NOAA Fisheries Northeast Fisheries Science Center.
+
+<!--
+# CRediT authorship contribution statement {-}
+
+**Timothy J. Miller:** Conceptualization, Methodology, Writing - original draft, Formal analysis, Visualization.
+ -->
+
+\pagebreak
+
+\bibliography{paper}
+<!-- \bibliography{project-0-paper}
+\bibliographystyle{/home/tmiller2/work/bibliographies/cjfas2}  -->
+
+<div id="refs"></div>
+
+\pagebreak
+
+\begin{landscape}
+\begin{table}
+\caption{Distinguishing characteristics of the operating models with random effects on survival.
+Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
+Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.}\label{naa_om_table}
+{\footnotesize \input{naa_om_tab}}
+\end{table}
+\end{landscape}
+
+\begin{landscape}
+\begin{table}
+\caption{Distinguishing characteristics of the operating models with random effects on natural mortality.
+Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
+Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
+For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}\label{M_om_table}
+{\input{M_om_tab}}
+\end{table}
+\end{landscape}
+
+\begin{landscape}
+\begin{table}
+\caption{Distinguishing characteristics of the operating models with random effects on selectivity.
+Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
+Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
+For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}\label{sel_om_table}
+{\input{Sel_om_tab}}
+\end{table}
+\end{landscape}
+
 \begin{landscape}
@@
 
-## Estimating models {-}
-
-For each data set simulated from an operating model 20 estimating models were fit.
-A total of 32 different estimating models were fit across all operating models where the subset of 20 depended on the source of process error (Table \ref{em_table}).
-The first 20 estimating models in Table \ref{em_table} were fit to simulate data sets from R and R+S operating models.
-Estimating models 5 to 24 in Table \ref{em_table} were fit to simulate data sets from R+M operating models.
-Estimating models 5 to 20 and 25-28 in Table \ref{em_table} were fit to simulate data sets from R+Sel operating models.
-Finally, estimating models 5 to 20 and 29-32 in Table \ref{em_table} were fit to simulate data sets from R+q operating models.
-
 \begin{table}
@@
 
-The observation error variance of the environmental observations and aggregate catch and indices were all assumed known at the true values.
-
-Simulations were all carried out on the University of Massachusetts Green High-Performance Computing Cluster.
-Code for completing the simulations and summarization of results can be found at github.com/timjmiller/SSRTWG/Project_0.
-We used the wham package version 1.X.X (commit 77bbd94).
-
-## Measures of reliability {-}
-
-The first measure of reliability we investigated was frequency of convergence when fitting each estimating model to the simulated data sets.
-There are various ways to assess convergence of the fit.
-We summarized 5 alternative categories of convergence.
-1. Did the optimization routine (stats::nlminb) complete without error?
-2. Did the stats::nlminb convergence flag = 0 indicate successful convergence?
-3. Was the maximum absolute value of the gradient of the log-likelihood < $1\times10^{-6}$?
-4. Did TMB::sdreport provide non-NA values for all fixed effects standard errors?
-5. Did TMB::sdreport provide all standard errors < 100?
-
-The first convergence criterion assesses whether the model crashes.
-The third is a measure that assesses how flat the likelihood is at the optimized point of the likelihood surface.
-The fourth and fifth criteria are specific to the hessian-based standard error reporting provided by TMB.
-the TMB::sdreport function will sometimes return standard error estimates even when the calculated hessian is not invertible (fourth criterion).
-It will also provide large standard errors for some parameters that are not estimated well or are near bounds on the transformed scale that is of primary interest.
-For example, variance parameters for random effects can often be estimated near 0 when the fit suggests no variation in the random effects or, equivalently, that the model without random effects is better.
-
-### AIC for model selection {-}
-
-We measured the frequency of correct model selection using marginal AIC.
-For a given operating model we the set of models that were considered all made the same assumptions on whether or not to estimate M and whether a stock-recruit function is assumed.
-
-We also measured the frequency of correctly selecting models with Beverton-Holt stock recruit function assumed over models without the stock-recruit function (null model).
-Because the proportions of simulations where the correct model was chosen for a given set of operating model factors implied generally poor performance of AIC, we analyzed the effect of variation in SSB on the probability of lower AIC of the BH models.
-We fit binomial generalize linear models to the observed AIC choice for each data set as a function of log(SD(SSB)) and operating model.
-
-### Bias {-}
-
-We estimated relative bias 
-$$
-\widehat {\text{RB}}\left(\theta\right) = \frac{1}{N} \sum^N_{i = 1} \frac{\widehat \theta_i - \theta_i}{\theta_i}
-$$
-of annual SSB and fully-selected fishing mortality for each estimation model and constructed 95\% confidence intervals for the median relative bias using the binomial distribution approach as in \citet{stockmiller21}.
-We also estimated bias of stock-recruit parameters for estimation models that assumed the Beverton-Holt stock recruit function.
-
-### Mohn's rho {-}
-
-We estimated Mohn's rho for SSB, fully-selected fishing mortality, and recruitment for each estimation model.
-We estimated 7 peels for each estimation model.
-We calculated median 95\% confidence intervals for Mohn's rho using the same methods as that for bias.
-
-# Results {-}
-
-## Convergence performance {-}
-
-Probability of convergence was better for models that assumed the correct source of process error, assumed M was known, and did not assume stock-recruit relationships.
-
-### R, R+S operating models {-}
-
-Figures \ref{naa_om_em_R_MF_convergence} to \ref{naa_om_em_BH_ME_convergence}
+<!-- AIC TABLES --->
+
+\begin{table}
+\caption{Operating models and estimation models all assume matching R or R+S process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}\label{naa_om_em_R_BH_aic_table}
+{\input{naa_om_em_R_BH_aic_table}}
+\end{table}
+
+\begin{table}
+\caption{Operating models and estimation models all assume matching R+M process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}\label{M_om_em_R_BH_aic_table}
+{\input{M_om_em_R_BH_aic_table}}
+\end{table}
+
+\begin{table}
+\caption{Operating models and estimation models all assume matching R+Sel process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}\label{Sel_om_em_R_BH_aic_table}
+{\input{Sel_om_em_R_BH_aic_table}}
+\end{table}
+
+\begin{table}
+\caption{Operating models and estimation models all assume matching R+q process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}\label{q_om_em_R_BH_aic_table}
+{\input{q_om_em_R_BH_aic_table}}
+\end{table}
+
+\clearpage
+<!-- OM configuration plots --->
+
+\begin{figure}
+\caption{The proportion mature at age assumed for the population in all operating and estimating models.}\label{om_maturity}
+\begin{center}
+\includegraphics[width = \textwidth]{om_maturity.png}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\caption{The weight at age assumed for the population in all operating and estimating models.}\label{om_waa}
+\begin{center}
+\includegraphics[width = \textwidth]{om_waa.png}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\caption{The Beverton-Holt stock recruit relationship assume for all operating models.}\label{om_sr}
+\begin{center}
+\includegraphics[width = \textwidth]{om_sr.png}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\caption{The selectivity at age assumed for the fishing fleet in operating models without selectivity random effects.
+The mean selectivity at age across time for the fishing fleet in operating models with selectivity random effects.}\label{om_mean_selectivity}
+\begin{center}
+\includegraphics[width = \textwidth]{om_mean_selectivity.png}
+\end{center}
+\end{figure}
+
+<!-- Convergence --->
 
@@
 
-### R+M operating models {-}
-
-Figures \ref{M_om_em_R_MF_convergence} to \ref{M_om_em_BH_ME_convergence}
-
 \begin{landscape}
@@
 
-### R+Sel operating models {-}
-
-Figures \ref{Sel_om_em_R_MF_convergence} to \ref{Sel_om_em_BH_ME_convergence}
-
 \begin{landscape}
@@
 
-### R+q operating models {-}
-
-Figures \ref{q_om_em_R_MF_convergence} to \ref{q_om_em_BH_ME_convergence}
-
 \begin{landscape}
@@
 
-## AIC performance for process error structure {-}
-
-### R, R+S operating models {-}
-
-Figures \ref{naa_om_proportion_best_aic_R_MF} to \ref{naa_om_proportion_best_aic_SR_ME}
+<!-- AIC --->
 
@@
 
-### R+M operating models {-}
-
-Figures \ref{M_om_proportion_best_aic_R_MF} to \ref{M_om_proportion_best_aic_SR_ME}
-
 \begin{landscape}
@@
 
-### R+Sel operating models {-}
-
-Figures \ref{Sel_om_proportion_best_aic_R_MF} to \ref{Sel_om_proportion_best_aic_SR_ME}
-
 \begin{landscape}
@@
 \end{landscape}
-
-### R+q operating models {-}
-
-Figures \ref{q_om_proportion_best_aic_R_MF} to \ref{q_om_proportion_best_aic_SR_ME}
 
@@
 \clearpage
-## AIC performance for stock-recruit relationship {-}
-
-Tables of frequencies where AIC chooses correct stock-recruit assumption imply poor performance of AIC, but figures showing effect of variation in SSB demonstrate that large variation in SSB is needed to reduce Type II errors for marginal AIC.
-
-### R, R+S operating models {-}
-
-\begin{table}
-\caption{Operating models and estimation models all assume matching R or R+S process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}
-{\input{naa_om_em_R_BH_aic_table}}
-\end{table}
+
+<!-- AIC stock-recruit selection--->
 
@@
 \begin{figure}
-\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB. Estimating models allow estimation of mean M.}\label{naa_om_ME_BH_glm_AIC_plots}
+\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
+Operating and estimating models have matching R or R+S process error structures.
+Estimating models allow estimation of mean M.}\label{naa_om_ME_BH_glm_AIC_plots}
 \begin{center}
@@
 \end{figure}
-
-### R+M operating models {-}
-
-\begin{table}
-\caption{Operating models and estimation models all assume matching R+M process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}
-{\input{M_om_em_R_BH_aic_table}}
-\end{table}
 
@@
 
-### R+Sel operating models {-}
-
-\begin{table}
-\caption{Operating models and estimation models all assume matching R+Sel process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}
-{\input{Sel_om_em_R_BH_aic_table}}
-\end{table}
-
 \begin{figure}
@@
 
-### R+q operating models {-}
-
-\begin{table}
-\caption{Operating models and estimation models all assume matching R+q process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}
-{\input{q_om_em_R_BH_aic_table}}
-\end{table}
-
 \begin{figure}
@@
 \clearpage
-## Bias {-}
-
-### R, R+S OMs {-}
+
+<!-- Bias --->
 
@@
 
-### R+M OMs {-}
-
 \begin{landscape}
@@
 
-### R+Sel OMs {-}
-
 \begin{landscape}
@@
 
-### R+q OMs {-}
-
 \begin{landscape}
@@
 
-### Beverton-Holt parameter estimation
-
 \begin{landscape}
@@
 
-### (Mean) natural mortality rate estimation
-
 \begin{landscape}
@@
 
-## Mohn's $\rho$ {-}
-
-### R, R+S OMs {-}
+<!-- Retro --->
 
@@
 \end{landscape}
-
-# Discussion {-}
-
-The estimating models assumed variances of aggregate catch and index observations was known.
-This approximation may be appropriate for indices where we have a reliable estimate of uncertainty based on the survey design (), but there may be better approaches for the aggregate catch such as an informed prior on the standard errors with realistic bounds.
-
-# Acknowledgements {-}
-
-This work was funded by NOAA Fisheries Northeast Fisheries Science Center.
-
-<!--
-# CRediT authorship contribution statement {-}
-
-**Timothy J. Miller:** Conceptualization, Methodology, Writing - original draft, Formal analysis, Visualization.
- -->
-
-\pagebreak
-
-\bibliography{paper}
-<!-- \bibliography{project-0-paper}
-\bibliographystyle{/home/tmiller2/work/bibliographies/cjfas2}  -->
-
-<div id="refs"></div>
-
-\pagebreak

### 2024-01-19 Ecov_study/recruitment_functions/Workflow_Ecov_Recruitment_Project.Rmd
code update, beta unstandardized plots
partial results for beta unstandardized runs, and code that generated them

@@
    + *AIC_weight.RDS* contains AIC, convergence, model rank, model weight, model probability info for all EMs per OM-Sim
+   + Runtime on Azure for unstandardized beta runs: Time difference of 4.075107 hours
 
@@
      + plots of parameters that failed the bad.grad or bad.se criteria
+     + Runtime is quick
 
 3. Generate regression trees to characterize factors influencing model identifiability; run **AIC_regression.R**
-   + ...
+   + specify gradient and SE tolerance at top (these were appended to AIC dataframe outputs in the step above)
+   + specify results directory (where runs are), directory where to save plots, and a suffix to append at end of file name (to distinguish plots for different cases)
+   + regression trees are done to determine factors associated with dAIC (difference in model AIC), Model probability (calculated from AIC weights), and AIC rank (models are ranked by integer with 1= lowest AIC, and 6=highest AIC per OM-Sim; if not all 6 EMs converged for this OM-Sim, then the highest AIC rank is the number of models converged)/ **<span style="color: red;">question: should i just look at dAIC for models with AIC rank >1?
+(because dAIC=0 when AIC rank = 1)</span>**
 
-4. Look at bias in SSB, Recr, Fbar
-   + ...
+4. Look at bias in SSB, Recr, Fbar **bias_assessment.R**
+   + specify gradient and SE tolerance at top (these were appended to AIC dataframe outputs in the step above)
+   + specify results directory (where runs are), directory where to save plots, and a suffix to append at end of file name (to distinguish plots for different cases)
+   + regression trees are done to determine factors associated RE in recruitment, SSB, and Fbar
+      + looks like sigmaR is the only factor that matters for all 3 metrics
+      + this script produces plots summarizing relative error (RE) over all years, the last 10 years, and the final year
 
-5. Look at bias in SR_a, SR_b, sigmaR, rho_R, sigmaEcov, rho_Ecov
-   + ...
+5. Look at bias in parameter estimates **bias_parameters.R**
+   + SR_a, SR_b, sigmaR, rho_R, sigmaEcov, rho_Ecov
+   + also grabbing catchability (logit_q), selectivity (logit_selpars), catch_paa_pars, and index_paa_pars parameters so i don't have to pass through results again
 

### 2024-01-19 Ecov_study/recruitment_functions/code/analysis/AIC_dataframe.R

@@ -1,11 +1,11 @@
 library(here)
-
+library(wham)
 #folder link to id
 # dir <- '~/dropbox/working/state_space_assessments/cluster_download/results_noSR//'
 #df.oms          <- readRDS(file.path(here(),"Ecov_study","recruitment_functions", "inputs", "df.oms_noSR.RDS"))
 source(file.path(here::here(), "Ecov_study","recruitment_functions", "code", "analysis", "functions_for_analysis.R" ) )
 
-res.path <- 'E:/results_beta_fix'  # directory where simulation runs are
+res.path <- 'E:/results_beta_fix'  # directory where simulation runs are (beta unstandardized)
 
 df.oms          <- readRDS(file.path(here::here(),"Ecov_study","recruitment_functions", "inputs", "df.oms.RDS"))
 df.ems    <- readRDS(file.path(here::here(),"Ecov_study", "recruitment_functions", "inputs", "df.ems.RDS"))
@@ -22,9 +22,9 @@ n_sims <- 100
 ################################################################################
 t4 <- Sys.time()
 get_aic_convergence_info(df.oms, df.ems, nsims=n_sims, res.path=res.path, save.rds=TRUE, save.suffix='_beta_fix', save.path=here::here("Ecov_study", 'recruitment_functions', 'results_beta_fix')  ) 
-# "This took 20.7794454693794" minutes (liz)!  
+# "This took 20.7794454693794" minutes (liz's runs)!  
 
 
 
 t5 <- Sys.time()
-t5-t4
+t5-t4   # Time difference of 4.075107 hours

### 2024-01-19 Ecov_study/recruitment_functions/code/analysis/functions_for_analysis.R

@@ -94,7 +94,7 @@ get_aic_convergence_info <- function(df.oms=NULL, df.ems=NULL, nsims=NULL, res.p
   
 k <- 1
 for(om in 1:nrow(df.oms)){
-  print(paste0("OM = ",om, "out of ", nrow(df.oms)))
+  print(paste0("OM = ",om, " out of ", nrow(df.oms)))
     # get aic ====
   for(sim in 1:nsims){
       DAT <- sapply(1:nrow(df.ems), function(em){

### 2024-01-19 Ecov_study/recruitment_functions/Workflow_Ecov_Recruitment_Project.Rmd
AIC regression updated
plots and code done for beta unstandardized; Rmd updated

@@
    + Runtime on Azure for unstandardized beta runs: Time difference of 4.075107 hours
+   + **<span style="color: blue;">STATUS: done for beta unstandardized; </span>**
 
@@
      + plots of parameters that failed the bad.grad or bad.se criteria
-     + Runtime is quick
+   + Runtime is fast
+   + **<span style="color: blue;">STATUS: done for beta unstandardized; </span>**
 
@@
 (because dAIC=0 when AIC rank = 1)</span>**
+   + Runtime is fast
+   + **<span style="color: blue;">STATUS: done for beta unstandardized; </span>**
 

### 2024-01-19 Ecov_study/recruitment_functions/code/analysis/AIC_regression.R

@@ -10,23 +10,25 @@ bad.grad.value <- 1E-6 #tim used 1E-6 (the abs of the exponent will be used for 
 bad.grad.label <- as.numeric(strsplit(as.character(bad.grad.value), split="-")[[1]][2])
 bad.se.value <- 100 #tim used 100 (this value will be used for output suffix; ex: filename_se_100.png)
 
-res.dir <- 'results'       # results folder where AIC dataframes are
-plot.dir <- 'plots'
-plot.suffix <- '_beta_fix'
+ 
+res.dir <- 'results_beta_fix'   # 'results'     'results_beta_fix'   # results folder where AIC dataframes are
+plot.dir <- 'plots_beta_fix'    # 'plots_lizruns'  'plots_beta_fix'   
+plot.suffix <- '_beta_fix'      # '_beta_fix'   '' 
+
 
 
 ## get converged runs with these thresholds
-AIC_best <- readRDS( file.path(here(),'Ecov_study','recruitment_functions',res.dir, paste0("AIC_best_conv.runs_grad_", bad.grad.label, "_SE_", bad.se.value, ".RDS")  ) )  # EM with lowest AIC per OM-Sim (that converged)
-AIC_weight <- readRDS(file.path(here(),'Ecov_study','recruitment_functions',res.dir, paste0("AIC_weight_conv.runs_grad_", bad.grad.label, "_SE_", bad.se.value, ".RDS")  ))  # info for all EMs per OM-Sim
+AIC_best <- readRDS( file.path(here(),'Ecov_study','recruitment_functions',res.dir, paste0("AIC_best_conv.runs_grad_", bad.grad.label, "_SE_", bad.se.value, plot.suffix, ".RDS")  ) )  # EM with lowest AIC per OM-Sim (that converged)
+AIC_weight <- readRDS(file.path(here(),'Ecov_study','recruitment_functions',res.dir, paste0("AIC_weight_conv.runs_grad_", bad.grad.label, "_SE_", bad.se.value, plot.suffix, ".RDS")  ))  # info for all EMs per OM-Sim
 
[…]

### 2024-01-20 Ecov_study/recruitment_functions/Workflow_Ecov_Recruitment_Project.Rmd
bias in time series results
bias.assessment code and plots added, Rmd  in recruitment_functions main directory updated

@@
       + this script produces plots summarizing relative error (RE) over all years, the last 10 years, and the final year
+   + Runtime is slooooow (and code is inefficient--feel free to tighten it up)
+   + **<span style="color: blue;">STATUS: done for beta unstandardized; </span>**
 
@@
    + also grabbing catchability (logit_q), selectivity (logit_selpars), catch_paa_pars, and index_paa_pars parameters so i don't have to pass through results again
+   + Runtime ...pending
+   + **<span style="color: blue;">STATUS: pending beta unstandardized; </span>**
 

### 2024-01-22 Ecov_study/recruitment_functions/Workflow_Ecov_Recruitment_Project.Rmd
updated workflow Rmd
added note where to create df of failed jobs

@@
    + bad.se.value : cut-off for identifying parameter where SE_par_max>bad.se.value (parameter with max value and 2nd max value as well as parameter name are saved, not *all* parameters in this category)
+   + the dataframe <span style="text-decoration:underline">om.fails.df</span>, read at top of file, is created in **find_failed_jobs_Azure.R** which is in recruitment_functions/code (1 directory up from /analysis)
    + given this, output is produced using bad.grad.value and bad.se.value appended to plot and table names to identify the tolerances

### 2024-01-22 Ecov_study/recruitment_functions/Workflow_Ecov_Recruitment_Project.Rmd

--- 2024-01-20 Ecov_study/recruitment_functions/Workflow_Ecov_Recruitment_Project.Rmd
@@
    + **<span style="color: blue;">STATUS: done for beta unstandardized; </span>**
+   + **<span style="color: red;">STATUS: done for beta standardized; </span>**
 
@@
    + regression trees are done to determine factors associated with dAIC (difference in model AIC), Model probability (calculated from AIC weights), and AIC rank (models are ranked by integer with 1= lowest AIC, and 6=highest AIC per OM-Sim; if not all 6 EMs converged for this OM-Sim, then the highest AIC rank is the number of models converged)/ **<span style="color: red;">question: should i just look at dAIC for models with AIC rank >1?
-(because dAIC=0 when AIC rank = 1)</span>**
+(because dAIC=0 when AIC rank = 1);.</span>** **<span style="color: blue;">GLB: Perhaps you could look at dAIC for rank=2 as that represents the dAIC of the next best model.
+We would hope that to be large for the correct model and small for the incorrect.</span>**
    + Runtime is fast
    + **<span style="color: blue;">STATUS: done for beta unstandardized; </span>**
+   + **<span style="color: red;">STATUS: done for beta standardized; </span>**
+
+3. Fit GLMs and GLMMs to identifiability results
+   + **<span style="color: red;">STATUS: done for beta standardized; </span>**
 
@@
    + **<span style="color: blue;">STATUS: done for beta unstandardized; </span>**
+   + **<span style="color: red;">STATUS: pending for beta standardized; </span>**
+
+4. Fit GLMs and GLMMs to assessment bias results
+   + **<span style="color: blue;">STATUS: pending for beta unstandardized; </span>**
+   + **<span style="color: red;">STATUS: pending for beta standardized; </span>**
 
@@
    + **<span style="color: blue;">STATUS: pending beta unstandardized; </span>**
+
+5a. Fit GLMs and GLMMs to parameter bias results
 

### 2024-01-23 Ecov_study/recruitment_functions/Workflow_Ecov_Recruitment_Project.Rmd
bias_parameters.R complete
includes plots, regression tree, 1 table; Rmd workflow updated

--- 2024-01-22 Ecov_study/recruitment_functions/Workflow_Ecov_Recruitment_Project.Rmd
@@
 5. Look at bias in parameter estimates **bias_parameters.R**
+   + specify gradient and SE tolerance at top (these were appended to AIC dataframe outputs in the step above)
+   + specify results directory (where runs are), directory where to save plots, and a suffix to append at end of file name (to distinguish plots for different cases)
    + SR_a, SR_b, sigmaR, rho_R, sigmaEcov, rho_Ecov
    + also grabbing catchability (logit_q), selectivity (logit_selpars), catch_paa_pars, and index_paa_pars parameters so i don't have to pass through results again
-   + Runtime ...pending
-   + **<span style="color: blue;">STATUS: pending beta unstandardized; </span>**
+   + all Truth values also grabbed, RE calculated and plotted based on factors from regression trees
+   + Runtime 1.17 hours
+   + **<span style="color: blue;">STATUS: done beta unstandardized; </span>**
+
+5. Cleaning up script **compare_projections.R**
+   + this calculates mohn's rho for user-specified number of peels, evaluates relative error from peel 10 to peel 0, ...
+   + Runtime ...
+   + **<span style="color: blue;">STATUS: pending... beta unstandardized; </span>**
 

### 2024-01-23 Ecov_study/recruitment_functions/Workflow_Ecov_Recruitment_Project.Rmd
mohn's rho analysis
script and plots

@@
 
-5. Cleaning up script **compare_projections.R**
-   + this calculates mohn's rho for user-specified number of peels, evaluates relative error from peel 10 to peel 0, ...
+6. Look at retrospective bias **mohns_rho_analysis.R**
+   + this calculates mohn's rho for user-specified number of peels
+   + regression tree and plots for SSB, Fbar, Recruitment and also for recruitment random effects
+   + Runtime 2 * 1.6 hours
+   + **<span style="color: blue;">STATUS: done beta unstandardized; </span>**
+
+7. Cleaning up script **compare_projections.R**
+   + evaluates relative error in projections between peel 10 and truth ...
    + Runtime ...
@@
 
-# Figures
+# Figures and Tables
+
++ SSRTWG/Ecov_study/recruitment_functions/plots_beta_fix
++ SSRTWG/Ecov_study/recruitment_functions/tables
 

### 2024-01-23 Ecov_study/recruitment_functions/code/analysis/mohns_rho_analysis.R

@@ -1,0 +1,482 @@
+library(here)
+library(tidyverse)
+library(wham)
+library(rpart)
+library(rpart.plot)
+
+
+source(file.path(here::here(), "Ecov_study","recruitment_functions", "code", "analysis", "functions_for_analysis.R" ) )
+
+
+#dir <- file.path(here::here(), "Ecov_study","recruitment_functions", "results_pile2")
+res.path <- 'E:/results_beta_fix'  # directory where simulation runs are (beta unstandardized)
+res.dir <- 'results_beta_fix'   # 'results'     'results_beta_fix'   # results folder where AIC dataframes are
+plot.dir <- 'plots_beta_fix'    # 'plots_lizruns'  'plots_beta_fix'   
+plot.suffix <- '_beta_fix' 
+
+## specify bad.grad.label and bad.se.value (these are the thresholds set in convergence_summaries.R to determine convergence) 
+bad.grad.value <- 1E-6 #tim used 1E-6 (the abs of the exponent will be used for output suffix; ex: filename_grad_6.png)
+bad.grad.label <- as.numeric(strsplit(as.character(bad.grad.value), split="-")[[1]][2])
+bad.se.value <- 100 #tim used 100 (this value will be used for output suffix; ex: filename_se_100.png)
+
+df.oms          <- readRDS(file.path(here::here(),"Ecov_study","recruitment_functions", "inputs", "df.oms.RDS"))
+df.ems    <- readRDS(file.path(here::here(),"Ecov_study", "recruitment_functions", "inputs", "df.ems.RDS"))
+conv.runs <- readRDS(file.path(here::here(),'Ecov_study','recruitment_functions',res.dir, paste0("conv.runs_grad_", bad.grad.label, "_SE_", bad.se.value, plot.suffix, ".RDS")  ) )
+
+conv.runs <- conv.runs %>%
+  rename(Sim=sim) %>%
+  mutate(ok.run =(bad.opt+ bad.conv+ bad.sdrep+ bad.grad+ bad.se.big) ) %>%
+  replace_na(list(ok.run=1)) %>%
[…]

### 2024-01-25 Ecov_study/mortality/paper/paper.tex

@@
 
-\hypertarget{summaryabstract}{%
-\subsection*{Summary/Abstract}\label{summaryabstract}}
-\addcontentsline{toc}{subsection}{Summary/Abstract}
-
-We completed a large scale simulation study with 288 operating models
-with 12 estimated models fit to each. The factors defining operating
-model configuration inlcuded the source of process error on the
-population (recruitment, recruitment and survival, recruitment and
-natural mortality), the degree of temporal variation and autocorrelation
-of the environmental covariate, the uncertainty in the observation of
-the covariate, the uncertainty in indices and age composition
-observations, fishing history, and the magnitude of the effect of the
-covariate on natural mortality.
+\hypertarget{summary}{%
+\subsection*{Summary}\label{summary}}
+\addcontentsline{toc}{subsection}{Summary}
+
+We completed a large scale simulation study with 288 operating models,
+each with 100 simulated data sets, and 12 estimated models fit to each
+simulated data set. The factors defining operating model configuration
+included the source of process error on the population (recruitment,
+recruitment and survival, recruitment and natural mortality), the degree
+of temporal variation and autocorrelation of the environmental
+covariate, the uncertainty in the observation of the covariate, the
+uncertainty in indices and age composition observations, fishing
+history, and the magnitude of the effect of the covariate on natural
+mortality. The estimating models make alternative assumptions on whether
+to include the environmental effect, whether the mean/intercept log
+natural mortality (log(0.2)) is estimated or known, and whether process
+errors are on just recruitment, recruitment and survival, or recruitment
+and natural mortality.
 
@@
 
-AIC paragraph
-
-Bias paragraph
-
-These results suggest that models with covariate effects on M will be
-most reliable when R+S is true, large contrast in the Recommendations
-\#\#\# Keywords \{-\}
+Whether the mean log-natural mortality was estimated or not, the best
+accuracy of AIC for model selection occurred for models with process
+errors on recruitment and survival. AIC accuracy was poor for models
+with process errors on recruitment and natural mortality. estimating the
+mean natural mortality rate had small effects on the accuracy of AIC in
+selecting the appropriate process error. Estimating the mean log-natural
+mortality resulted in a small decrease in AIC accuracy. AIC was
+conservative for determining whether the environmental covariate
+affected natural mortality. AIC was very accurate in determining no
+effect when there was no effect in the operating model, but AIC often
+ranked the null model best when there was an effect. Accuracy of AIC for
+covariate effects improved with increased effect size, increased
+temporal contrast in the covariate, and lower uncertainty in
+observations.
+
+We found no evidence of bias in estimation of environmental effects
+regardless of process error assumptions when there was low uncertainty
+in the environmental observations and large constrast in the
+environmental covariate. In most cases the relative error of the
+estimated environmental effect did not depend on the source of process
+error assumed in the estimating model. The worst bias was observed when
+OMs assumed R+S process errors, high uncertainty in covariate
+observations, low variability in the covariate, and low uncertainty in
+index and age composition observations. Simultaneously estimating the
+mean/intercept log natural mortality resulted in larger variation in the
+relative errors of the estimated environmental effect. Estimation of the
+intercept was reliable for all EM process error assumptions when the
+operating models assumed process errors on recruitment and natural
+mortality, contrast in fishing pressure over time, and lower observation
+error. Estimating the mean/intercept for log natural mortality generally
+resulted in highly variable estimates of annual natural mortality and
+spawning biomass and evidence of bias for some operating and estimation
+model assumptions about process error source. Again reliability of
+annual natural mortality estimates was generally improved with lower
+observation error uncertainty and contrast in fishing pressure.
+
+Reliable detection of covariate effects requires informative data. AIC
+preferred simpler models than the true model when information content in
+data and contrast in covariates and abundance were low. The null model
+for environmental covariate effects (no covariate effect) was selected
+when contrast in the time series was low and/or uncertainty in
+observations was high. The selection of the null model by AIC also
+likely decreases with strength of the effect of the covariate on M.
+Similarly, when there was process error in recruitment and natural
+mortality, estimation models with process error only in recruitment were
+preferred presumably due to low variation in simulated natural mortality
+process errors. Covariate effect estimation can be robust to process
+error assumptions with high contrast in covariate and low observation
+error.
 
@@
 convergence when fitting each estimating model to the simulated data
-sets. There are various ways to assess convergence of the fit. We
-summarized 5 alternative categories of convergence:
-
-\begin{enumerate}
-\item Did the optimization routine (stats::nlminb) complete without error?
-\item Did the stats::nlminb convergence flag = 0 indicate successful convergence?
-\item Was the maximum absolute value of the gradient of the log-likelihood < $1\times10^{-6}$?
-\item Did TMB::sdreport provide non-NA values for all fixed effects standard errors?
-\item Did TMB::sdreport provide all standard errors < 100?
-\end{enumerate}
-
-The first convergence criterion assesses whether the model crashes. The
-third is a measure that assesses how flat the likelihood is at the
-optimized point of the likelihood surface. The fourth and fifth criteria
-are specific to the hessian-based standard error reporting provided by
-TMB. the TMB::sdreport function will sometimes return standard error
-estimates even when the calculated hessian is not invertible (fourth
-criterion). It will also provide large standard errors for some
-parameters that are not estimated well or are near bounds on the
-transformed scale that is of primary interest. For example, variance
-parameters for random effects can often be estimated near 0 when the fit
-suggests no variation in the random effects or, equivalently, that the
-model without random effects is better.
+sets. There are various ways to assess convergence of the fit, but we
+defined successful convergence as the hessian of the marginal
+log-likelihood being invertible and providing variance estimates for the
+fixed effects parameters.
 
@@
 age composition observations, and either R or R+M process errors (Figure
-\ref{mean_M_bias}). The most variation in estimates occurred when
-fishing and Higher uncertainty in index and age composition
-observations. For OMs with R+S process errors, the most reliable
-estimation of \(\beta_M\) was obtained when the EM also assumed R+S
-process errors across all other factors. For OMs with R+M process
-errors, the matching assumption for the EM only showed the best
-reliability when there was low uncertainty in index and age composition
-observations and fishing mortality was constant.
+\ref{mean_M_bias_Ecov_beta_fixed}). The most variation in estimates
+occurred when fishing mortality was constant and there was higher
+uncertainty in index and age composition observations. For OMs with R+S
+process errors, the most reliable estimation of \(\beta_M\) was obtained
+when the EM also assumed R+S process errors across all other factors.
+For OMs with R+M process errors, the matching assumption for the EM only
+showed the best reliability when there was low uncertainty in index and
+age composition observations and fishing mortality was constant.
 
@@
 at the true value (\(\beta_M = log(0.2)\)) and no covariate effect is
-assmed (\(\beta_\text{Ecov} = 0\)), 2) \(\beta_M = log(0.2)\) and
+assumed (\(\beta_\text{Ecov} = 0\)), 2) \(\beta_M = log(0.2)\) and
 \(\beta_\text{Ecov}\) is estimated, and 3) both \(\beta_M\) and
@@
 change in fishing mortality and higher uncertainty in index and age
-composition observations, we observed positive median errrors at the
+composition observations, we observed positive median errors at the
 beginning of the time series and lower median errors in later years.
@@
 When \(\beta_\text{Ecov}\) was estimated, R EMs performed worse for R+S
-OMs with evidence of negative bias for OMs with cosntant fishing
+OMs with evidence of negative bias for OMs with constant fishing
 mortality, lower uncertainty in index and age composition observations,
@@
 observations (Figure \ref{annual_M_bias_M_estimated_beta_estimated}).
-The same varibilities occurred for R+S OMs, but strong bias was
-estimated for R and R+M EMs. Annual M estimation was most reliable when
-OMs had step changes in fishing mortality and lower uncertainty in index
-and age composition observations. For R and R+M OMs with those
-characteristics, all EMs generally provided accurate estimation of
-annual M, but only R+S EMs provided accurate estimation for R+S OMs.
-Little or no bias in annual M estimation was observed for R+S EMs across
-all OM process error assumptions as long as there was a step change in
-fishing mortality and lower uncertainty in indices and age composition.
+The same variability occurred for R+S OMs, but strong bias was estimated
+for R and R+M EMs. Annual M estimation was most reliable when OMs had
+step changes in fishing mortality and lower uncertainty in index and age
+composition observations. For R and R+M OMs with those characteristics,
+all EMs generally provided accurate estimation of annual M, but only R+S
+EMs provided accurate estimation for R+S OMs. Little or no bias in
+annual M estimation was observed for R+S EMs across all OM process error
+assumptions as long as there was a step change in fishing mortality and
+lower uncertainty in indices and age composition.
 
@@
 effect) was selected when contrast in the time series was low and/or
-uncertainty in observations was high.
-
-When there was process error in recruitment and M (R+M), models with
-process error only in recruitment were preferred. This could be because
-the variation in the M random effects was small relative to observation
-variability. The marginal variance for log M random effects was 0.3.
-
-Null selection likely decreases with strength of the effect on M. We
-only examined two non-zero effect sizes: 0.25, and 0.5, but our results
-suggest larger effect sizes, with the same observation error and
-contrast in time series, would allow better AIC performance for
-determining covariate effects.
-
-Covariate effect estimation can be robust to process error assumptions
-(For R+M OMs) with contrast in covariate and low observation error.
-
-Reliable estimation of environmental effect on M regardless of the
-process error assumed by the EM as long as the contrast in the covariate
-is sufficient and the uncertainty in the observations is low.
+uncertainty in observations was high. Null selection likely decreases
+with strength of the effect on M. We only examined two non-zero effect
+sizes: 0.25, and 0.5, but our results suggest larger effect sizes, with
+the same observation error and contrast in time series, would allow
+better AIC performance for determining covariate effects. When there was
+process error in recruitment and M (R+M), models with process error only
+in recruitment were preferred. This could be because the variation in
+the M random effects was small relative to observation variability. The
+marginal variance for log M random effects was 0.3. Reliable estimation
+of environmental effect on M regardless of the process error assumed by
+the EM as long as the contrast in the covariate is sufficient and the
+uncertainty in the observations is low.
 

### 2024-01-25 Project_0/paper/paper.tex

@@
 We used the Woods Hole Assessment Model (WHAM) to configure operating
-and estimation models in our simulation study \citep[]{stockmiller21}.
-WHAM is an R package freely available as a github repository. For this
-study we used
+and estimation models in our simulation study
+\citep[]{millerstock20,stockmiller21}. WHAM is an R package freely
+available as a github repository. For this study we used
 \href{https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3}{version
@@
 operating models \citep{millerbrooks21}. We specified unfished
-recruitment = \(R_0 = e^{10}\) and \(\Fmsy = \Fspr[40] = 0.348\) equated
+recruitment = \(R_0 = e^{10}\) and \(\Fmsy = F_{40\%} = 0.348\) equated
 to a steepness of 0.69 and \(\alpha=0.60\) and

### 2024-01-25 stochastic_brps/state_space_BRP_estimation.Rmd

---
title: "The nature of differences between analytic and projection-based equilibrium biomass reference points"
author: "Timothy J. Miller^1^"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::pdf_document2:
    toc: false
    keep_tex: yes
    fig_caption: yes
    number_sections: yes
    includes:
      in_header: options.sty
  bookdown::word_document2: default
  bookdown::html_document2:
    df_print: paged
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy=FALSE, tidy.opts=list(width.cutoff=60), warning = FALSE, message = FALSE, fig.align = "center")
#knitr::opts_knit$set(root.dir = '../')
library(knitr)
library(tidyverse)
library(pander)
library(kableExtra)
library(wham)
#library(ggplotFL)
library(viridis)
```

$^1$timothy.j.miller@noaa.gov, Northeast Fisheries Science Center, National Marine Fisheries Service, 166 Water Street, Woods Hole, MA 02543, USA\

\pagebreak

## Summary {-}

Equilibrium biomass and harvest reference points can be determined from assessment output either using analytic methods or projections of the population for a sufficient number of years. These projections can be either deterministic or incorporate stochasticity of the recruitment and other process errors.

Central tendencies of the stochastic projection for equilibrium SSB or harvest will not be equivalent to the values from analytic methods. The difference is a result of summing log-normal random variables across ages for the stochastic projection approach and inability to accurately estimate the central tendency of this resulting distribution.

For models without stock-recruit functions, contributions to SSB for each age less than the plus group can be accurately estimated, but the plus group cannot because it is also a sum of lognormal random variables.

All age-specific contributions are expected to be biased when a stock-recruit function is used because recruitment is then a function of SSB.

Analogous bias in equilibrium harvest reference points is expected because it is a similar function of abundance at age.

The bias would exist for either SPR- or MSY-based reference points because the same SSB/R or Y/R calculations are used with equilibrium recruitment.

Using The bias.correct option to TMB::sdreport will improve the estimate of the biomass reference point, but the improvement becomes negligble as the marginal variance of the autoregressive processes increase. Furthermore, the standard error of the estimated biomass reference point is not reported.

<!-- Estimating B0 instead of R0 will not help. -->

## Introduction {-}

Biomass and catch reference points are typically calculated using deterministic equations for spawning biomass and yield per recruit or making long term population projections holding all population and fishing attributes constant. With no stochasticity in recruitment or other processes, these two approaches will produce equivalent results. However, uncertainty in recruitment and possibly other processes are often included in projection methods. When stochasticity is included, the SSB or catch in the long term projection at say year 100, will be random variable that represents the possible SSB or catch that could be realized in any given year under constant mortality rates. The central tendency (i.e., mean or median) of that distribution is the reference point, but the variance of the estimated central tendency rather than that of the prediction of the SSB or catch in any given year, represents our uncertainty in the reference point resulting from observations and parameter estimation.

For many stochastic systems we have analytic solutions for the attributes (e.g, mean, and variance) of the distribution of projected value. For example when recruitment is an AR1 process
$$
\log R_y = \mu(1-\rho) + \rho \log R_{y-1} + \epsilon_y
$$
where $\epsilon_y$ are iid $\text{N}(0,\sigma^2)$, then the distribution of $\log R_y$ in any given year (the marginal distribution) is
$$
\log R_y \sim \text{N}\left(\mu, \frac{\sigma^2}{1-\rho^2}\right)
$$
Similar analytic results exist for the marginal distributions of abundance of other age classes when autoregressive processes are assumed. If equilibrium recruitment was a reference point we would be interested in the uncertainty of the estimate $\hat \mu$ rather than the variability of recruitment in any arbitrary year. The uncertainty in the estimate of the mean $\mu$ would be a function of the observed data and generally much more precise than the marginal distribution of yearly recruitment.

However, spawning biomass is complex functions of abundance at age. In projections, they are sums of scaled estimates of abundance at age
$$
\text{SSB}_y = \sum^A_{a=1} N_{y,a} f_{y,a} e^{-Z_{y,a} \phi}
$$
where $Z_{y,a}$ is the total mortality rate, $f_{y,a}$ is the fecundity at age (typically the product of biomass and proportion mature), $\phi$ is the fraction of year elapsed prior to spawning.

<!--
Catch is a similar sum of age-specific products of abundance, mortality, and biomass
$$
\text{Catch}_y = \sum^A_{a=1} N_{y,a} \frac{F_{y,a}}{Z_{y,a}}\left(1-e^{-Z_{y,a}}\right)w_{y,a}.
$$
-->

Here we show that unbiased analytic or model-based estimation of the mean or median total biomass of stochastic projections assuming equilibrium conditions (i.e, temporally constant parameters) is not generally possible. However, it is possible to accurately estimate the age-specific components of the biomass for ages less than the plus group.

## Methods {-}

For the general state-space model with random effects at all ages, the log-abundance at age is normally distributed
$$\log N_{a,y} = \log R_{y-a+1} + \sum^{a-1}_{i = 1}\left( \log S_{i,y-a+i} + \epsilon_{i+1,y-a+i+1}\right)$$
for ages less than the plus group $a<A$. The log-SSB at age just adds another "constant" $\log\left(\phi_{a,y}\right)$ for survival up to spawning, maturity, and weight at age,
$$\log \text{SSB}_{a,y} = \log \phi_{a,y} + \log R_{y-a+1} + \sum^{a-1}_{i = 1}\left( \log S_{i,y-a+i} + \epsilon_{i+1,y-a+i+1}\right)$$
which is also log-normal. The log-SSB contribution from the plus group is
$$
\log\text{SSB}_{A,y} = \log \phi_{a,y} +\log\left(N_{A-1,y-1} S_{A-1,y-1} + N_{A,y-1} S_{A,y-1}\right) +\epsilon_{A,y}
$$
which, conditional on the abundance at age in the previous time step, is log-normally distributed, but marginally the abundances at age are also log-normal RV which are summed for abundance at the current time step.

### Bias-corrected equilibrium calculations {-}

The stochastic SSB/R for ages less than the plus group when $S_{a,y}$ and $\phi_{a,y}$ are time-invariant is

$$\log \text{SSB}_{a,y} - \log R_{y-a+1} = \log \phi_{a} + \sum^{a-1}_{i = 1} \left(\log S_{i} + \epsilon_{i}\right)$$
where $\epsilon_{i} \sim \text{N}\left(0,V_i\right)$ and
$$
V_i = \frac{\sigma_i^2}{(1-\rho_{\text{age}})^2(1-\rho_{\text{year}})^2}
$$
Although there may be correlation of the epsilon terms (e.g., a 2DAR1 structure is assumed), the age specific components are exchangable across years under equilibrium conditions and they can be treated independently. In such case, conditional on recruitment, the expectation of SSB/R at age is
\begin{equation}\label{bcssbpr}
E((SSB/R)_a) = \phi_a \left(\prod_{i=1}^{a-1} S_i\right) E\left(e^{\sum_{i=1}^{a-1} \epsilon_i}\right)=\phi_a \left(\prod_{i=1}^{a-1} S_i\right) e^{\frac{1}{2}\sum_{i=1}^{a-1} V_i}
\end{equation}

<!-- 
Equilibruim yield per recruit given F is calculated as
$$
 \widetilde Y(F) = \sum^{A}_{a=1} C_a S_a(F) E_a(F) D_a(F)
$$
where $C_a$ is the catch weight at age, $D_a(F) = 1 - e^{-Z_a(F)}$, and
$$
E_a(F) =\frac{F s_a}{Z_a(F)}.
$$

When abundance at age is bias-corrected, the age-specific components for ages less than the plus group can be unbiasedly estimated by multiplying be 
$$
e^{-\frac{1}{2}V_a}
$$
where $V_a$ is the marginal variance of the process error for age $a$
$$
V_a = \frac{\sigma_a^2}{(1-\rho_a^2)^2(1-\rho_y^2)}.
$$
The modifier is not needed for the first age class (recruitment) because it is divided out for the per recruit calculation.
-->

<!--

We fit several models to observations on western Gulf of Maine Atlantic cod. First we fit the base model with alternative assumptions on bias-correction internal to the model: bias-correct both process errors and observation errors (default), bias-correct process errors only, bias-correct observation errors only, and no bias-correction. Each model assumed process errors on recruitment only or all numbers at age. For each process error structure there were models with either independent errors or ar1 correlation (yearly for recruitment only and 2D-AR1 (age and year) for all numbers at age). We also considered alternative assumptions of bias-correcting the process error models or not. 
-->
### Operating model {-}

We projected a WGOM cod model similar to that accepted in the Atlantic cod research track peer review for 100 years assuming F equal to the deterministically determined F at 40% unfished SSB/R and at F = 0. The model assumed a 2DAR1 correlation structure for numbers at age. We simulated 1000 realizations of the process errors and associated model output in the projection period. We used a variant of the WHAM where many of the ADREPORTed variables were removed to facilitate more efficient use of the TMB::sdreport bias-correction feature.

## Results {-}

<!--

### Process errors on both recruitment and survival {-}

#### 2DAR1 correlation {-}
-->

From the simulations we can see that the mean and median recruitment can be accurately estimated without performing the stochastic projections (Fig. \@ref(fig:recruitment)). The predicted recruitment is the median and different than the estimated mean recruitment due to the bias-correction of the process errors in the WHAM model. The bias-correction calculates the mean recruitment in any given year $\widehat {E(R)} = e^{\widehat \mu + \sigma^2/2}$. The predicted recruitment matches the median of simulated values because the estimates of random effects are taken as the mode of the posterior distribution by TMB. The bias-correction available in TMB::sdreport provides a negatively biased estimate of mean recruitment.

```{r recruitment, echo = FALSE, out.width="100%", fig.cap = "Alternative estimators for projected recruitment in the western Gulf of Maine Atlantic cod WHAM model. Simulation mean and median are that of annual recruitments across simulations. Also shown are the log-normal bias-corrected mean recruitment (hat E(R)), exponential of estimated log-recruit random effects (hat R), and both the exponential of the log-recruit estimate before (hat R (TMB b-c)) and after (exp(hat log R (TMB b-c))) reporting by TMB::sdreport with bias-correction."}
knitr::include_graphics(here::here("compare_recruitment.png"))
```

However, neither mean or median spawning biomass can be accurately estimated using analytic methods or model output. When fishing mortality is set at $F_{40\%}$, the bias-corrected estimator provided by TMB::sdreport is similar to median SSB, but it is supposed to provide an estimate of mean SSB which is unstable but far greater (Fig. \@ref(fig:ssb)). The projections at $F = 0$ suggest the proximity of the bias-corrected estimator to SSB at $F=F_{40\%}$ is coincidence.

```{r ssb, echo = FALSE, out.width="100%", fig.cap = "Alternative estimators for projected spawning stock biomass (SSB) at $F=F_{40\\%}$ (left) or $F=0$ (right) in the western Gulf of Maine Atlantic cod model. Simulation mean and median are that of annual SSB across simulations. Also shown are the SSB calculated from log-normal bias-corrected mean numbers at age (f(hat E(NAA))), from exponential of estimated log-numbers at age random effects (f(hat NAA)), and from either the exponential of the log-SSB estimate before (hat SSB (TMB b-c)) and after (exp(hat log SSB (TMB b-c))) reporting by TMB::sdreport with bias-correction."}
knitr::include_graphics(here::here("compare_ssb.png"))
```

<!--

```{r ssb0, echo = FALSE, out.width="100%", fig.cap = "Projected spawning biomass for western Gulf of Maine Atlantic cod model with $F=0$ and 2dar1 correlation for all numbers at age random effects."}
knitr::include_graphics(here::here("compare_ssb0.png"))
```

#### Uncorrelated random effects {-}

When the Numbers at age random effects are uncorrelated, the differences still exist.

### Process errors on recruitment {-}

#### AR1 correlated random effects {-}

When we have AR1 random effects only on recruitment, the differences still exist.

#### Uncorrelated random effects {-}

When we have AR1 random effects only on recruitment, the differences still exist.

### Age composition of SSB in projections {-}
-->

By looking at age-specific components of projected SSB we can confirm that median SSB-at-age can only be estimated accurately for the ages prior to the plus group because the plus group is the only age class that is a sum of log-normal random variables (Fig. \@ref(fig:ssbaa)). Similarly, the cumulative sum of SSB at age can not be accurately estimated beyond age 1 because it is also the sum of log-normal random variables (Fig. \@ref(fig:cumssbaa)).

<!--

#### Process errors on both recruitment and survival {-}
-->

\begin{landscape}
```{r ssbaa, echo = FALSE, out.width="100%", fig.cap = "Median annual spawning stock biomass at age projected at $F = F40\\%$ or $F=0$ across simulations from the western Gulf of Maine Atlantic cod model. Also shown are the SSB as a function of estimated abundance at age (estimated) and the equilibrium SSB calculated as the product of the estimated recruitment and equilibrium SSB/R with log-normal bias-correction (Rhat x SSB/R)."}
knitr::include_graphics(here::here("compare_SSBAA_F40_at_age.png"))
```
\end{landscape}

<!--

Even when process errors are only on recruitment

only or all numbers at age and the correlation structure (iid or ar1) the median equilibrium spawning biomass at age can be accurately estimated for ages below the plus group (Figs. \@ref(fig:rec-iid-ssbaa) to \@ref(fig:rec-p1-2dar1-ssbaa)). The analytic and reported estimates of SSB are equivalent for all ages, but negatively biased for the plus group. The bias-corrected estimate of the equilibrium SSB in the plus group produced by TMB::sdreport is positively biased.

\begin{landscape}
```{r rec-iid-ssbaa, echo = FALSE, out.width="100%", fig.cap = "Spawning biomass at age at F40\\% for western Gulf of Maine Atlantic cod model with independent process errors only on recruitment."}
knitr::include_graphics(here::here("compare_SSBAA_F40_at_age_rec_iid.png"))
```
\end{landscape}

\begin{landscape}
```{r rec-ar1-ssbaa, echo = FALSE, out.width="100%", fig.cap = "Spawning biomass at age at F40\\% for western Gulf of Maine Atlantic cod model with AR1 process errors only on recruitment."}
knitr::include_graphics(here::here("compare_SSBAA_F40_at_age_rec.png"))
```
\end{landscape}

-->

\begin{landscape}
```{r cumssbaa, echo = FALSE, out.width="100%", fig.cap = "Median cumulative (median(sum)) or cumulative median (sum(median) spawning biomass at age projected at $F = F40\\%$ or $F=0$ across simulations from the western Gulf of Maine Atlantic cod model. Also shown are the cumulative SSB as a function of estimated abundance at age (sum(estimated)) and the cumulative equilibrium SSB calculated as the product of the estimated recruitment and cumulative equilibrium SSB/R at age with log-normal bias-correction (sum(Rhat x SSB/R))."}
knitr::include_graphics(here::here("compare_cumulative_SSBAA_F40_at_age.png"))
```
\end{landscape}

<!--

\begin{landscape}
```{r rec-ar1-cumssbaa, echo = FALSE, out.width="100%", fig.cap = "Cumulative Spawning biomass at age at F40\\% for western Gulf of Maine Atlantic cod model with process errors only on recruitment."}
knitr::include_graphics(here::here("compare_cumulative_SSBAA_F40_at_age_rec.png"))
```
\end{landscape}
-->

## Conclusions {-}

Analytic estimates of biomass reference points, including are biased for the corresponding median simulated stochastic equilibrium values because the marginal distribution of the simulated values are the sums of log-normal random variables. We showed this for the marginal distribution of equilibrium SSB, and equilibrium SSB in the plus group. However, the median of the equilibrium SSB at each age less than the plus group can be accurately estimated. The demonstration presented assumes stochasticity on abundance at all ages, but this bias would also exist for models with stochasticity only in recruitment (e.g., traditional statistical catch at age models). However, bias in estimation of equilibrium SSB at all ages is expected when a stock-recruit function is used because all ages are then functions of SSB which is the sum of log-normal random variables. Corresponding results are also expected for catch reference points because they are analogous functions of log-normal random variables.

Stochastic equilibrium SSB and catch appears to only be accurately estimated through simulation however determining the appropriate F to obtain the desired percentage of unfished SSB would require simulations to be completed at each F value external to the assessment model rather than internally. Therefore, the uncertainty in parameter estimates from the unprojected model would not be propagated.

If accurate estimation of each of the age-specific components of the biomass or catch reference point is a sufficient goal, then an appropriate expansion of the age classes of the population should minimize any bias in SSB for the plus group.

<!--

* median SSB(F40)/SSB(F=0) is ~40\% when estimation uses bias-correct PE option
* TMB::sdreport(bias.correct=T) provides accurate estimation of mean SSB at F40
* no options provide accurate estimation of median projected SSB at F=0
  * TMB::sdreport(bias.correct=T) provides better estimation than without
  * State-space models should use this particularly for projected output that is a non-linear function of RE
* State-space models should expand age structure for the population to minimize the effect of the plus group.
  * For wham, the data do not have to be reformatted, just pool older ages to predict last age class in data.
  * can use same WAA in data plus group for each of the model ages that it comprises.
* RE on M, growth, or WAA could be even more problematic because of the non-linear effect on $\log N_{a,y}$ at each age

Remaining issues

* uncertainty and accurate estimation for the total SSB or catch reference point attributes remains a challenge when accounting for temporal stochasticity. 
  * We can estimate the age-specific components below the plus group accurately.
  * TMB::sdreport(bias.corect=TRUE) is also generally inaccurate. 
  * Bias of age-specific components is expected when a stock-recruit function is assumed because the recruitments are then functions of log-normal variables.
in projections are for predicting SSB in year $y$ not the median (or mean of the log SSB) of that distribution.
* If the defined SPR at F depends on the uncertainty, then it is unknown how to calculate F at X%SPR internally.
-->

### 2024-01-25 stochastic_brps/state_space_BRP_estimation.tex

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={The nature of differences between analytic and projection-based equilibrium biomass reference points},
  pdfauthor={Timothy J. Miller1},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{url}
\usepackage{setspace}
%\singlespacing
%\onehalfspacing
%\doublespacing
%\usepackage{lineno}
%\linenumbers
\usepackage[belowskip=0pt,aboveskip=0pt]{caption}
\usepackage{relsize}
\usepackage{float}
% \usepackage[section]{placeins}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{The nature of differences between analytic and projection-based equilibrium biomass reference points}
\author{Timothy J. Miller\textsuperscript{1}}
\date{28 August, 2023}

\begin{document}
\maketitle

\(^1\)\href{mailto:timothy.j.miller@noaa.gov}{\nolinkurl{timothy.j.miller@noaa.gov}}, Northeast Fisheries Science Center, National Marine Fisheries Service, 166 Water Street, Woods Hole, MA 02543, USA\\

\pagebreak

\hypertarget{summary}{%
\subsection*{Summary}\label{summary}}
\addcontentsline{toc}{subsection}{Summary}

Equilibrium biomass and harvest reference points can be determined from assessment output either using analytic methods or projections of the population for a sufficient number of years. These projections can be either deterministic or incorporate stochasticity of the recruitment and other process errors.

Central tendencies of the stochastic projection for equilibrium SSB or harvest will not be equivalent to the values from analytic methods. The difference is a result of summing log-normal random variables across ages for the stochastic projection approach and inability to accurately estimate the central tendency of this resulting distribution.

For models without stock-recruit functions, contributions to SSB for each age less than the plus group can be accurately estimated, but the plus group cannot because it is also a sum of lognormal random variables.

All age-specific contributions are expected to be biased when a stock-recruit function is used because recruitment is then a function of SSB.

Analogous bias in equilibrium harvest reference points is expected because it is a similar function of abundance at age.

The bias would exist for either SPR- or MSY-based reference points because the same SSB/R or Y/R calculations are used with equilibrium recruitment.

Using The bias.correct option to TMB::sdreport will improve the estimate of the biomass reference point, but the improvement becomes negligble as the marginal variance of the autoregressive processes increase. Furthermore, the standard error of the estimated biomass reference point is not reported.

\hypertarget{introduction}{%
\subsection*{Introduction}\label{introduction}}
\addcontentsline{toc}{subsection}{Introduction}

Biomass and catch reference points are typically calculated using deterministic equations for spawning biomass and yield per recruit or making long term population projections holding all population and fishing attributes constant. With no stochasticity in recruitment or other processes, these two approaches will produce equivalent results. However, uncertainty in recruitment and possibly other processes are often included in projection methods. When stochasticity is included, the SSB or catch in the long term projection at say year 100, will be random variable that represents the possible SSB or catch that could be realized in any given year under constant mortality rates. The central tendency (i.e., mean or median) of that distribution is the reference point, but the variance of the estimated central tendency rather than that of the prediction of the SSB or catch in any given year, represents our uncertainty in the reference point resulting from observations and parameter estimation.

For many stochastic systems we have analytic solutions for the attributes (e.g, mean, and variance) of the distribution of projected value. For example when recruitment is an AR1 process
\[
\log R_y = \mu(1-\rho) + \rho \log R_{y-1} + \epsilon_y
\]
where \(\epsilon_y\) are iid \(\text{N}(0,\sigma^2)\), then the distribution of \(\log R_y\) in any given year (the marginal distribution) is
\[
\log R_y \sim \text{N}\left(\mu, \frac{\sigma^2}{1-\rho^2}\right)
\]
Similar analytic results exist for the marginal distributions of abundance of other age classes when autoregressive processes are assumed. If equilibrium recruitment was a reference point we would be interested in the uncertainty of the estimate \(\hat \mu\) rather than the variability of recruitment in any arbitrary year. The uncertainty in the estimate of the mean \(\mu\) would be a function of the observed data and generally much more precise than the marginal distribution of yearly recruitment.

However, spawning biomass is complex functions of abundance at age. In projections, they are sums of scaled estimates of abundance at age
\[
\text{SSB}_y = \sum^A_{a=1} N_{y,a} f_{y,a} e^{-Z_{y,a} \phi}
\]
where \(Z_{y,a}\) is the total mortality rate, \(f_{y,a}\) is the fecundity at age (typically the product of biomass and proportion mature), \(\phi\) is the fraction of year elapsed prior to spawning.

Here we show that unbiased analytic or model-based estimation of the mean or median total biomass of stochastic projections assuming equilibrium conditions (i.e, temporally constant parameters) is not generally possible. However, it is possible to accurately estimate the age-specific components of the biomass for ages less than the plus group.

\hypertarget{methods}{%
\subsection*{Methods}\label{methods}}
\addcontentsline{toc}{subsection}{Methods}

For the general state-space model with random effects at all ages, the log-abundance at age is normally distributed
\[\log N_{a,y} = \log R_{y-a+1} + \sum^{a-1}_{i = 1}\left( \log S_{i,y-a+i} + \epsilon_{i+1,y-a+i+1}\right)\]
for ages less than the plus group \(a<A\). The log-SSB at age just adds another ``constant'' \(\log\left(\phi_{a,y}\right)\) for survival up to spawning, maturity, and weight at age,
\[\log \text{SSB}_{a,y} = \log \phi_{a,y} + \log R_{y-a+1} + \sum^{a-1}_{i = 1}\left( \log S_{i,y-a+i} + \epsilon_{i+1,y-a+i+1}\right)\]
which is also log-normal. The log-SSB contribution from the plus group is
\[
\log\text{SSB}_{A,y} = \log \phi_{a,y} +\log\left(N_{A-1,y-1} S_{A-1,y-1} + N_{A,y-1} S_{A,y-1}\right) +\epsilon_{A,y}
\]
which, conditional on the abundance at age in the previous time step, is log-normally distributed, but marginally the abundances at age are also log-normal RV which are summed for abundance at the current time step.

\hypertarget{bias-corrected-equilibrium-calculations}{%
\subsubsection*{Bias-corrected equilibrium calculations}\label{bias-corrected-equilibrium-calculations}}
\addcontentsline{toc}{subsubsection}{Bias-corrected equilibrium calculations}

The stochastic SSB/R for ages less than the plus group when \(S_{a,y}\) and \(\phi_{a,y}\) are time-invariant is

\[\log \text{SSB}_{a,y} - \log R_{y-a+1} = \log \phi_{a} + \sum^{a-1}_{i = 1} \left(\log S_{i} + \epsilon_{i}\right)\]
where \(\epsilon_{i} \sim \text{N}\left(0,V_i\right)\) and
\[
V_i = \frac{\sigma_i^2}{(1-\rho_{\text{age}})^2(1-\rho_{\text{year}})^2}
\]
Although there may be correlation of the epsilon terms (e.g., a 2DAR1 structure is assumed), the age specific components are exchangable across years under equilibrium conditions and they can be treated independently. In such case, conditional on recruitment, the expectation of SSB/R at age is
\begin{equation}\label{bcssbpr}
E((SSB/R)_a) = \phi_a \left(\prod_{i=1}^{a-1} S_i\right) E\left(e^{\sum_{i=1}^{a-1} \epsilon_i}\right)=\phi_a \left(\prod_{i=1}^{a-1} S_i\right) e^{\frac{1}{2}\sum_{i=1}^{a-1} V_i}
\end{equation}

\hypertarget{operating-model}{%
\subsubsection*{Operating model}\label{operating-model}}
\addcontentsline{toc}{subsubsection}{Operating model}

We projected a WGOM cod model similar to that accepted in the Atlantic cod research track peer review for 100 years assuming F equal to the deterministically determined F at 40\% unfished SSB/R and at F = 0. The model assumed a 2DAR1 correlation structure for numbers at age. We simulated 1000 realizations of the process errors and associated model output in the projection period. We used a variant of the WHAM where many of the ADREPORTed variables were removed to facilitate more efficient use of the TMB::sdreport bias-correction feature.

\hypertarget{results}{%
\subsection*{Results}\label{results}}
\addcontentsline{toc}{subsection}{Results}

From the simulations we can see that the mean and median recruitment can be accurately estimated without performing the stochastic projections (Fig. \ref{fig:recruitment}). The predicted recruitment is the median and different than the estimated mean recruitment due to the bias-correction of the process errors in the WHAM model. The bias-correction calculates the mean recruitment in any given year \(\widehat {E(R)} = e^{\widehat \mu + \sigma^2/2}\). The predicted recruitment matches the median of simulated values because the estimates of random effects are taken as the mode of the posterior distribution by TMB. The bias-correction available in TMB::sdreport provides a negatively biased estimate of mean recruitment.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{compare_recruitment}

}

\caption{Alternative estimators for projected recruitment in the western Gulf of Maine Atlantic cod WHAM model. Simulation mean and median are that of annual recruitments across simulations. Also shown are the log-normal bias-corrected mean recruitment (hat E(R)), exponential of estimated log-recruit random effects (hat R), and both the exponential of the log-recruit estimate before (hat R (TMB b-c)) and after (exp(hat log R (TMB b-c))) reporting by TMB::sdreport with bias-correction.}\label{fig:recruitment}
\end{figure}

However, neither mean or median spawning biomass can be accurately estimated using analytic methods or model output. When fishing mortality is set at \(F_{40\%}\), the bias-corrected estimator provided by TMB::sdreport is similar to median SSB, but it is supposed to provide an estimate of mean SSB which is unstable but far greater (Fig. \ref{fig:ssb}). The projections at \(F = 0\) suggest the proximity of the bias-corrected estimator to SSB at \(F=F_{40\%}\) is coincidence.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{compare_ssb}

}

\caption{Alternative estimators for projected spawning stock biomass (SSB) at $F=F_{40\%}$ (left) or $F=0$ (right) in the western Gulf of Maine Atlantic cod model. Simulation mean and median are that of annual SSB across simulations. Also shown are the SSB calculated from log-normal bias-corrected mean numbers at age (f(hat E(NAA))), from exponential of estimated log-numbers at age random effects (f(hat NAA)), and from either the exponential of the log-SSB estimate before (hat SSB (TMB b-c)) and after (exp(hat log SSB (TMB b-c))) reporting by TMB::sdreport with bias-correction.}\label{fig:ssb}
\end{figure}

By looking at age-specific components of projected SSB we can confirm that median SSB-at-age can only be estimated accurately for the ages prior to the plus group because the plus group is the only age class that is a sum of log-normal random variables (Fig. \ref{fig:ssbaa}). Similarly, the cumulative sum of SSB at age can not be accurately estimated beyond age 1 because it is also the sum of log-normal random variables (Fig. \ref{fig:cumssbaa}).

\begin{landscape}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{compare_SSBAA_F40_at_age}

}

\caption{Median annual spawning stock biomass at age projected at $F = F40\%$ or $F=0$ across simulations from the western Gulf of Maine Atlantic cod model. Also shown are the SSB as a function of estimated abundance at age (estimated) and the equilibrium SSB calculated as the product of the estimated recruitment and equilibrium SSB/R with log-normal bias-correction (Rhat x SSB/R).}\label{fig:ssbaa}
\end{figure}
\end{landscape}

\begin{landscape}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{compare_cumulative_SSBAA_F40_at_age}

}

\caption{Median cumulative (median(sum)) or cumulative median (sum(median) spawning biomass at age projected at $F = F40\%$ or $F=0$ across simulations from the western Gulf of Maine Atlantic cod model. Also shown are the cumulative SSB as a function of estimated abundance at age (sum(estimated)) and the cumulative equilibrium SSB calculated as the product of the estimated recruitment and cumulative equilibrium SSB/R at age with log-normal bias-correction (sum(Rhat x SSB/R)).}\label{fig:cumssbaa}
\end{figure}
\end{landscape}

\hypertarget{conclusions}{%
\subsection*{Conclusions}\label{conclusions}}
\addcontentsline{toc}{subsection}{Conclusions}

Analytic estimates of biomass reference points, including are biased for the corresponding median simulated stochastic equilibrium values because the marginal distribution of the simulated values are the sums of log-normal random variables. We showed this for the marginal distribution of equilibrium SSB, and equilibrium SSB in the plus group. However, the median of the equilibrium SSB at each age less than the plus group can be accurately estimated. The demonstration presented assumes stochasticity on abundance at all ages, but this bias would also exist for models with stochasticity only in recruitment (e.g., traditional statistical catch at age models). However, bias in estimation of equilibrium SSB at all ages is expected when a stock-recruit function is used because all ages are then functions of SSB which is the sum of log-normal random variables. Corresponding results are also expected for catch reference points because they are analogous functions of log-normal random variables.

Stochastic equilibrium SSB and catch appears to only be accurately estimated through simulation however determining the appropriate F to obtain the desired percentage of unfished SSB would require simulations to be completed at each F value external to the assessment model rather than internally. Therefore, the uncertainty in parameter estimates from the unprojected model would not be propagated.

If accurate estimation of each of the age-specific components of the biomass or catch reference point is a sufficient goal, then an appropriate expansion of the age classes of the population should minimize any bias in SSB for the plus group.

\end{document}

### 2024-01-25 Project_0/paper/paper.Rmd
draft CJFAS abstract 175 words

@@
 
+## Abstract {-}
+
+<!-- note CJFAS abstract word limit 175, so I have drafted a shorter summary here -->
+
+Random effects can be included in state-space assessment models on many processes and in many ways, and guidance is needed on statistical reliability and model selection criteria.
+We simulated 72 operating models with varying fishing history, observation error uncertainty, and process error magnitude, correlation, and source (recruitment, survival, fishery selectivity, catchability, and natural mortality).
+We fit estimating models with different assumptions on the process error source, whether (mean) natural mortality was estimated, and whether a stock-recruit relationship was estimated.
+
+Models that assumed the correct process error source had high convergence and low bias.
+Bias was also low under most process error assumptions when there was contrast in fishing pressure.
+Stock-recruitment parameters were only reliably estimated in ideal situations.
+Marginal AIC most accurately distinguished process errors on recruitment, survival, and selectivity, as well as larger magnitude process errors of other types.
+Retrospective patterns were generally weak except for recruitment when observation error was high, even with the correct process error assumptions.
+When models did exhibit some retrospective pattern, estimating natural mortality often removed it.
+
+\pagebreak
+
 ## Summary {-}
@@
 
-Application of state-space models in fisheries stock assessment and management has expanded dramatically within ICES, Canada, and the Northeast US \citep{nielsenberg14,cadigan16,stockmiller21}.
+Application of state-space models in fisheries stock assessment and management has expanded dramatically within ICES, Canada, and the Northeast US \citep{nielsenberg14,cadigan16,pedersenberg17,stockmiller21}.
 State-space approaches that use random effects to parameterize process errors is considered best practice and a requirement for the next generation of stock assessment models \citep{hoyleetal22, punt23}.

### 2024-01-26 Project_0/paper/paper.Rmd
edits to project0 paper

--- 2024-01-19 Project_0/paper/paper.Rmd
@@
 We conducted a simulation study where we simulated process errors and observations for 72 operating models with alternative assumptions about fishing history, degree of uncertainty in index and age composition observations, type (recruitment, survival, fishery selectivity, catchability and natural mortality), degree of variation, and correlation of process errors.
-We fit 20 different estimating models to each of 100 simulated set of observations with alternative assumptions type and correlation structure of process errors, (mean) natural mortality was known or estimated, and a B-H stock recruit relationship was assumed or not.
+We fit 20 different estimation models to each of 100 simulated set of observations with alternative assumptions type and correlation structure of process errors, (mean) natural mortality was known or estimated, and a B-H stock recruit relationship was assumed or not.
 
@@
 
-Bias in spawning stock biomass estimation was generally low for estimating models that assumed the correct source of process error when there was lower observation error.
+Bias in spawning stock biomass estimation was generally low for estimation models that assumed the correct source of process error when there was lower observation error.
 Reliable estimation of stock-recruit relationship parameters only appears possible in ideal situations with lower observation errors in age composition and indices, lower variability in recruitment process errors and large contrast in spawning biomass over time.
 We found little evidence of bias for many OM process error assumptions when there was contrast in fishing pressure even when there was greater observation error although it can lead to more variable estimates of natural mortality.
-For OMs where there was bias in natural mortality due to high observation error, estimating the stock-recruit relationship seemed to remove the bias.
+For OMs where there was bias in natural mortality due to high observation error, estimation the stock-recruit relationship seemed to remove the bias.
 However, estimation of natural mortality can cause large differences between the true and estimated SSB (that may be unbiased on average) when there is less contrast in fishing pressure over time and higher observation error.
@@
 Retrospective patterns were generally weak for all estimation models regardless of the true source of process error, but they can be expected for recruitment even for the correct process error assumptions when observation error is high.
-When models did exhibit some retrospective pattern, estimating the mean natural mortality rate tended to remove it.
+When models did exhibit some retrospective pattern, estimation the mean natural mortality rate tended to remove it.
 
@@
 
-Here we conduct a simulation study with operating models varying by degree of observation error uncertainty, source and variability of process error, and fishing history.
-The simulations from these operating models are fitted with estimating models that make alternative assumptions for sources of process error, whether a stock-recruit model was estimated, and whether M is estimated.
+Here we conduct a simulation study with operating models (OMs) varying by degree of observation error uncertainty, source and variability of process error, and fishing history.
+The simulations from these OMs are fitted with estimation models (EMs) that make alternative assumptions for sources of process error, whether a stock-recruit model was estimated, and whether M is estimated.
 We evaluate whether AIC can correctly determine the correct source of process error and stock effects on recruitment.
@@
 
-We used the Woods Hole Assessment Model (WHAM) to configure operating and estimation models in our simulation study \citep[]{millerstock20,stockmiller21}.
+We used the Woods Hole Assessment Model (WHAM) to configure operating and EMs in our simulation study \citep[]{millerstock20,stockmiller21}.
 WHAM is an R package freely available as a github repository.
 For this study we used [version 1.0.6.9000, commit 77bbd94](https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3).
-This packages has also been used to configure operating and estimating models for closed loop simulations evaluating index-based assessment methods \citep{legaultetal23} and is used for management of haddock, butterfish, American plaice, bluefish, Atlantic cod, and black sea bass in the Northeast US.
-
-We completed a simulation study with a number of operating models that can be categorized based on where process error random effects are assumed: abundance at age (R, R+S), natural mortality (R+M), fleet selectivity (R+Sel), or index catchability (R+q).
-For each operating model assumptions about variance of process errors and observations are required and the values we used were based on a review of the range of estimates from recent applications of WHAM in management of stocks of haddock, butterfish, and American plaice in the NE US.
-
-In total, we configured 72 operating models with alternative assumptions about the source and variability of process errors, level of observation error in indices and age composition data, and contrast in fishing pressure over time.
-We fitted 20 estimating models to observations from each of 100 simulations where process errors were also simulated.
-Estimating models made alternative assumptions about the source of process errors and whether natural mortality (or the mean for models with process error in natural mortality) was estimated and whether a Beverton-Holt stock recruit relationship was estimated within the estimation model.
-Details of each of the operating and estimating models are described below.
-
-We did not use the bias-correction feature for process errors or observations described by \citet{stockmiller21} for operating and estimating models.Simulations and model fitting were all carried out on the University of Massachusetts Green High-Performance Computing Cluster.
+This packages has also been used to configure operating and EMs for closed loop simulations evaluating index-based assessment methods \citep{legaultetal23} and is used for management of haddock, butterfish, American plaice, bluefish, Atlantic cod, and black sea bass in the Northeast US.
+
+We completed a simulation study with a number of OMs that can be categorized based on where process error random effects are assumed: abundance at age (R, R+S), natural mortality (R+M), fleet selectivity (R+Sel), or index catchability (R+q).
+For each OM assumptions about variance of process errors and observations are required and the values we used were based on a review of the range of estimates from recent applications of WHAM in management of stocks of haddock, butterfish, and American plaice in the NE US.
+
+In total, we configured 72 OMs with alternative assumptions about the source and variability of process errors, level of observation error in indices and age composition data, and contrast in fishing pressure over time.
+We fitted 20 EMs to observations from each of 100 simulations where process errors were also simulated.
+EMs made alternative assumptions about the source of process errors and whether natural mortality (or the mean for models with process error in natural mortality) was estimated and whether a Beverton-Holt stock recruit relationship was estimated within the EM.
+Details of each of the operating and EMs are described below.
+
+We did not use the bias-correction feature for process errors or observations described by \citet{stockmiller21} for operating and EMs.
+Simulations and model fitting were all carried out on the University of Massachusetts Green High-Performance Computing Cluster.
 All code we used to perform the simulation study and summarize results can be found at https://github.com/timjmiller/SSRTWG/tree/main/Project_0/code.
@@
 The population model tracks 10 age classes: ages 1 to 10+ and we assume spawning occurs 1/4 of the way through the year.
-The maturity at age was a logistic curve with $a_{50}$ = 2.89 and slope = 0.88 and assumed known in all estimation models (Figure \ref{om_maturity}).
+The maturity at age was a logistic curve with $a_{50}$ = 2.89 and slope = 0.88 and assumed known in all EMs (Figure \ref{om_maturity}).
 
@@
 
-We assumed a Beverton-Holt stock recruit function with constant pre-recruit mortality parameters for all operating models.
-All post-recruit productivity components are constant in the NAA and survey catchability process error operating models.
-Therefore steepness and unfished recruitment are also constant over the time period for those operating models \citep{millerbrooks21}.
+We assumed a Beverton-Holt stock recruit function with constant pre-recruit mortality parameters for all OMs.
+All post-recruit productivity components are constant in the NAA and survey catchability process error OMs.
+Therefore steepness and unfished recruitment are also constant over the time period for those OMs \citep{millerbrooks21}.
 We specified unfished recruitment = $R_0 = e^{10}$ and $\Fmsy = F_{40\%} = 0.348$ equated to a steepness of 0.69 and $\alpha=0.60$ and $\beta = 2.4 \times 10^{-5}$ for the
@@
 
-Two alternative fishing histories were used for operating models.
+Two alternative fishing histories were used for OMs.
 In the first scenario, the stock experiences overfishing for the first 20 years and fishing at \Fmsy for the last 20 years.
@@
 The magnitude of the overfishing assumptions is based on average estimates of overfishing for NE groundfish stocks from \citet{wiedenmannetal19}.
-\Citet{legaultetal23} also used similar approaches to defining fishing mortality histories for operating models.
+\Citet{legaultetal23} also used similar approaches to defining fishing mortality histories for OMs.
 
@@
 
-For operating models with time-varying random effects for M, steepness is not constant, but we used the same alpha and beta parameters as other operating models this equates to a steepness and R0 at the mean of the time series process for M.
-For operating models with time-varying random effects for fishery selectivity, \Fmsy is also not constant however we use the same F history as other operating models which corresponds to Fmsy at the mean selectivity parameters.
+For OMs with time-varying random effects for M, steepness is not constant, but we used the same alpha and beta parameters as other OMs this equates to a steepness and R0 at the mean of the time series process for M.
+For OMs with time-varying random effects for fishery selectivity, \Fmsy is also not constant however we use the same F history as other OMs which corresponds to Fmsy at the mean selectivity parameters.
 
@@
 A low uncertainty specification assumed standard deviation of both series of log-aggregate index observations was 0.1 and the standard deviation of the logistic-normal for age composition observations was 0.3 In the high uncertainty specification the standard deviation for log-aggregate indices was 0.4 and that for the age composition observations was 1.5.
-For all estimating models, standard deviation for log-aggregate observations was assumed known whereas that for the logistic-normal age composition observations was estimated.
+For all EMs, standard deviation for log-aggregate observations was assumed known whereas that for the logistic-normal age composition observations was estimated.
 
@@
 
-The factorial combinations of fishing histories, observation error assumptions, and marginal variance assumptions for recruitment and survival results in 24 different operating models with random effects on recruitment and survival (Table \ref{naa_om_table}).
+The factorial combinations of fishing histories, observation error assumptions, and marginal variance assumptions for recruitment and survival results in 24 different OMs with random effects on recruitment and survival (Table \ref{naa_om_table}).
 
@@
 
-The factorial combinations of fishing histories, observation error assumptions, and marginal variance and autocorrelation assumptions for natural mortality results in 16 different operating models with random effects on natural mortality (Table \ref{M_om_table}).
+The factorial combinations of fishing histories, observation error assumptions, and marginal variance and autocorrelation assumptions for natural mortality results in 16 different OMs with random effects on natural mortality (Table \ref{M_om_table}).
 Internally, WHAM treats natural mortality as a log-transformed parameter
@@
 that is a linear combination of a mean $\mu_M$ and any annual random effects marginally distributed as $\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)$.
-The values for the marginal variance and correlation parameters assumed in the operating models are given in Table \ref{M_om_table}.
+The values for the marginal variance and correlation parameters assumed in the OMs are given in Table \ref{M_om_table}.
 
@@
 
-The factorial combinations of fishing histories, observation error assumptions, and marginal variance and autocorrelation assumptions for selectivity parameters results in 16 different operating models with random effects on selectivity (Table \ref{sel_om_table}).
+The factorial combinations of fishing histories, observation error assumptions, and marginal variance and autocorrelation assumptions for selectivity parameters results in 16 different OMs with random effects on selectivity (Table \ref{sel_om_table}).
 Internally, WHAM treats selectivity parameter $s$ as a logit-transformed parameter
@@
 that is a linear combination of a mean $\mu_s$ and any annual random effects marginally distributed as $\epsilon_{s,y} \sim \text{N}\left(0,\sigma_s^2\right)$ where the lower and upper bounds of the parameter ($l_s$ and $u_s$) can be specified by the user.
-All selectivity parameters are either $a_50$ or slope parameters and we assume bounds of 0 and 10 for all selectivity parameters for all operating and estimating models.
-The values for the marginal variance and correlation parameters assumed in the operating models are given in Table \ref{sel_om_table}.
+All selectivity parameters are either $a_50$ or slope parameters and we assume bounds of 0 and 10 for all selectivity parameters for all operating and EMs.
+The values for the marginal variance and correlation parameters assumed in the OMs are given in Table \ref{sel_om_table}.
 
@@
 
-The factorial combinations of fishing histories, observation error assumptions, and marginal variance and autocorrelation assumptions for catchability results in 16 different operating models with random effects on catchability (Table \ref{q_om_table}).
+The factorial combinations of fishing histories, observation error assumptions, and marginal variance and autocorrelation assumptions for catchability results in 16 different OMs with random effects on catchability (Table \ref{q_om_table}).
 Internally, WHAM treats catchability for index $i$ as a logit-transformed parameter
@@
 that is a linear combination of a mean $\mu_i$ and any annual random effects marginally distributed as $\epsilon_{i,y} \sim \text{N}\left(0,\sigma_i^2\right)$ where the lower and upper bounds of the catchability ($l_i$ and $u_i$) can be specified by the user.
-Here we assume bounds of 0 and 1000 for all operating and estimating models.
-For operating and estimation models with process errors on catchability, the temporal variation is only assumed for the first index.
-The values for the marginal variance and correlation parameters assumed in the operating models are given in Table \ref{q_om_table}
-
-## Estimating models {-}
-
-For each data set simulated from an operating model 20 estimating models were fit.
-A total of 32 different estimating models were fit across all operating models where the subset of 20 depended on the source of process error (Table \ref{em_table}).
-The first 20 estimating models in Table \ref{em_table} were fit to simulate data sets from R and R+S operating models.
-Estimating models 5 to 24 in Table \ref{em_table} were fit to simulate data sets from R+M operating models.
-Estimating models 5 to 20 and 25-28 in Table \ref{em_table} were fit to simulate data sets from R+Sel operating models.
-Finally, estimating models 5 to 20 and 29-32 in Table \ref{em_table} were fit to simulate data sets from R+q operating models.
+Here we assume bounds of 0 and 1000 for all operating and EMs.
+For operating and EMs with process errors on catchability, the temporal variation is only assumed for the first index.
+The values for the marginal variance and correlation parameters assumed in the OMs are given in Table \ref{q_om_table}
+
+## Estimation models {-}
+
+For each data set simulated from an OM 20 EMs were fit.
+A total of 32 different EMs were fit across all OMs where the subset of 20 depended on the source of process error (Table \ref{em_table}).
+The first 20 EMs in Table \ref{em_table} were fit to simulate data sets from R and R+S OMs.
+EMs 5 to 24 in Table \ref{em_table} were fit to simulate data sets from R+M OMs.
+EMs 5 to 20 and 25-28 in Table \ref{em_table} were fit to simulate data sets from R+Sel OMs.
+Finally, EMs 5 to 20 and 29-32 in Table \ref{em_table} were fit to simulate data sets from R+q OMs.
 The observation error variance of aggregate catch and indices were all assumed known at the true values.
@@
 
-The first measure of reliability we investigated was frequency of convergence when fitting each estimating model to the simulated data sets.
+The first measure of reliability we investigated was frequency of convergence when fitting each EM to the simulated data sets.
 There are various ways to assess convergence of the fit. We summarized 5 alternative categories of convergence.
@@
 We measured the frequency of correct model selection using marginal AIC.
-For a given operating model we the set of models that were considered all made the same assumptions on whether or not to estimate M and whether a stock-recruit function is assumed.
+For a given OM we the set of models that were considered all made the same assumptions on whether or not to estimate M and whether a stock-recruit function is assumed.
 
 We also measured the frequency of correctly selecting models with Beverton-Holt stock recruit function assumed over models without the stock-recruit function (null model).
-Because the proportions of simulations where the correct model was chosen for a given set of operating model factors implied generally poor performance of AIC, we analyzed the effect of variation in SSB on the probability of lower AIC of the BH models.
-We fit binomial generalize linear models to the observed AIC choice for each data set as a function of log(SD(SSB)) and operating model.
+Because the proportions of simulations where the correct model was chosen for a given set of OM factors implied generally poor performance of AIC, we analyzed the effect of variation in SSB on the probability of lower AIC of the BH models.
+We fit binomial generalize linear models to the observed AIC choice for each data set as a function of log(SD(SSB)) and OM.
 
@@
 
-We estimated relative bias 
+We estimated bias as the median of the relative errors across all simulations for a given OM and EM combination.
 $$
-\widehat {\text{RB}}\left(\theta\right) = \frac{1}{N} \sum^N_{i = 1} \frac{\widehat \theta_i - \theta_i}{\theta_i}
+\text{RE}\left(\theta_i\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
 $$
-of annual SSB and fully-selected fishing mortality for each estimation model and constructed 95\% confidence intervals for the median relative bias using the binomial distribution approach as in \citet{millerhyun18} and \citet{stockmiller21}.
-We also estimated bias of stock-recruit parameters for estimation models that assumed the Beverton-Holt stock recruit function.
+of annual SSB and fully-selected fishing mortality for each EM and constructed 95\% confidence intervals for the median relative bias using the binomial distribution approach as in \citet{millerhyun18} and \citet{stockmiller21}.
+We also estimated bias of stock-recruit parameters for EMs that assumed the Beverton-Holt stock recruit function and the mean natural mortality rate parameter when estimated.
 
@@
 
-We estimated Mohn's $\rho$ for SSB, fully-selected fishing mortality, and recruitment for each estimation model.
-We estimated 7 peels for each estimation model.
+We estimated Mohn's $\rho$ for SSB, fully-selected fishing mortality, and recruitment for each EM.
+We estimated 7 peels for each EM.
 We calculated median 95\% confidence intervals for Mohn's $\rho$ using the same methods as that for relative bias.
@@
 Convergence rates were high for all convergence types for all EMs in R+S OMs with lower uncertainty in observations.
-Convergence for most convergence types generally declines most estimating models when mean-log natural mortality rate is estimated and/or a Beverton-Holt stock recruit relationship is estimated even when the process error assumptions of the estimating and operating models match.
-There can be relatively high convergence probability of type 4 (an invertible hessian) for EMs that do not have the correct process error assumed, but type 5 convergence (no very large standard errors estimated from hessian) for these estimating models is typically much lower.
+Convergence for most convergence types generally declines most EMs when mean-log natural mortality rate is estimated and/or a Beverton-Holt stock recruit relationship is estimated even when the process error assumptions of the estimation and OMs match.
+There can be relatively high convergence probability of type 4 (an invertible hessian) for EMs that do not have the correct process error assumed, but type 5 convergence (no very large standard errors estimated from hessian) for these EMs is typically much lower.
 
@@
 
-Marginal AIC accurately determines the correct process error assumptions in estimating models when data are generated from R and R+S OMs, regardless of whether mean log-natural mortality or a stock recruit relationship is estimated (Figures \ref{naa_om_proportion_best_aic_R_MF} to \ref{naa_om_proportion_best_aic_SR_ME}).
+Marginal AIC accurately determines the correct process error assumptions in EMs when data are generated from R and R+S OMs, regardless of whether mean log-natural mortality or a stock recruit relationship is estimated (Figures \ref{naa_om_proportion_best_aic_R_MF} to \ref{naa_om_proportion_best_aic_SR_ME}).
 Adding estimation of mean log-natural mortality or a stock recruit relationship separately has a negligible effect on the accuracy of determining the correct process error assumption.
@@
 Whether mean log-natural morality or stock recruit relationships were estimated appeared to have little effect on the performance of marginal AIC.
+
 ### R+q operating models {-}
@@
 
-## AIC performance for stock-recruit relationship {-}
+## AIC performance for the stock-recruit relationship {-}
 
@@
 
-AIC accuracy for including the B-H stock-recruit relationship was estimated to be highest for models OMs and EMs with process errors only on recruitment and lowest variance of recruitment process errors (Figures \ref{naa_om_MF_BH_glm_AIC_plots} and \ref{naa_om_ME_BH_glm_AIC_plots}).
-Holding variability in SSB constant, accuracy of stock-recruit  model selection declined with greater variance of recruitment and survival process errors.
-Simultaneously estimating mean log-natural mortality has a small negative effect on AIC accuracy.
-
-For R+M OMs and EMs there appeared to little effect of correlation in the natural mortality process errors or estimating the mean log-natural mortality on accuracy of AIC in determining a stock-recruit function, holding variability in SSB constant (Figures \ref{M_om_MF_BH_glm_AIC_plots} and \ref{M_om_ME_BH_glm_AIC_plots}).
-
-The range of variation in SSB across simulated populations was lower for R+Sel operating models than R+S or R+M OMs.
+### R, R+S operating models {-}
+
+Among R and R+S OMs, AIC accuracy for including the B-H stock-recruit relationship was estimated to be highest for R OMs and EMs and lowest variance of recruitment process errors (Figures \ref{naa_om_MF_BH_glm_AIC_plots} and \ref{naa_om_ME_BH_glm_AIC_plots}).
+FOr R+S OMs and holding variability in SSB constant, accuracy of stock-recruit  model selection declined with greater variance of recruitment and survival process errors.
+Simultaneously estimating mean log-natural mortality has a small negative effect on AIC accuracy, but there is an improvement in AIC accuracy with lower observation error in indices and age composition and lower variation of survival process errors.
+
+### R+M operating models {-}
+
+For R+M OMs and EMs there appeared to be little effect of correlation in the natural mortality process errors or estimating the mean log-natural mortality on accuracy of AIC in determining a stock-recruit function, holding variability in SSB constant (Figures \ref{M_om_MF_BH_glm_AIC_plots} and \ref{M_om_ME_BH_glm_AIC_plots}).
+However, there was an increase in accuracy with lower observation error for OMs with higher variability in natural mortality process errors whether the mean natural mortality rate parameter was estimated or not.
+
+### R+Sel operating models {-}
+
+The range of variation in SSB across simulated populations was lower for R+Sel OMs than R+S or R+M OMs.
 OMs with contrast in fishing pressure over time is the primary mechanism creating variation in SSB.
-However, there appears to be little effect of variability or correlation of selectivity process errors or whether mean log-natural mortality was estimated on the accuracy of AIC in selecting the stock recruit relationship (Figures \ref{Sel_om_MF_BH_glm_AIC_plots} and \ref{Sel_om_ME_BH_glm_AIC_plots})
-
-Like the R+Sel OMs, the range of variation in SSB across is only increased by allowing contrast in fishing pressure over time.
+However, there appears to be little effect of variability or correlation of selectivity process errors or whether mean log-natural mortality was estimated on the accuracy of AIC in selecting the stock recruit relationship (Figures \ref{Sel_om_MF_BH_glm_AIC_plots} and \ref{Sel_om_ME_BH_glm_AIC_plots}).
+However, there was an increase in AIC accuracy with lower observation error for OMs with independent process errors whether the mean natural mortality rate parameter was estimated or not.
+
+### R+q operating models {-}
+
+Like the R+Sel OMs, the range of variation in SSB across simulated populations for R+q OMs is only increased by allowing contrast in fishing pressure over time.
 There appears to be a slight decrease in accuracy in AIC when mean log-natural mortality is estimated and there is a more noticeable increase in accuracy for OMs with lower observation error in those cases.
-(Figures \ref{q_om_MF_BH_glm_AIC_plots} and \ref{q_om_ME_BH_glm_AIC_plots})
+(Figures \ref{q_om_MF_BH_glm_AIC_plots} and \ref{q_om_ME_BH_glm_AIC_plots}).
+There was also an increase in AIC accuracy with lower observation error for when the mean natural mortality rate parameter was estimated.
 
@@
 Estimating mean log-natural mortality generally produced greater variability in relative errors of SSB estimates and a change in sign of bias for EMs assuming process errors other than R and R+S (Figure \ref{naa_om_em_R_ME_relbias_ssb}).
-Estimating a stock-recruit relationship had little discernable effect on bias of SSB estimation (Figures \ref{naa_om_em_BH_MF_relbias_ssb}), but in combination with estimating mean log-natural mortality resulted in large bias of R+S EMs for R OMs with high observation error, constant fishing pressure and large variability in recruitment process errors \ref{naa_om_em_BH_ME_relbias_ssb}).
+Estimating a stock-recruit relationship had little discernable effect on bias of SSB estimation (Figures \ref{naa_om_em_BH_MF_relbias_ssb}), but in combination with estimating mean log-natural mortality resulted in large bias of R+S EMs for R OMs with high observation error, constant fishing pressure and large variability in recruitment process errors (Figure \ref{naa_om_em_BH_ME_relbias_ssb}).
 
@@
 Estimating mean log-natural mortality results in large variability in SSB relative errors, particularly for OMs with high observation error and constant fishing pressure and all EMs produced biased estimation of SSB over all years when OMs had high observation error, a change in fishing pressure and low variability of natural mortality process errors (Figure \ref{M_om_em_R_ME_relbias_ssb}).
-Like R and R+S OMs, estimating a stock-recruit relationship had little discernible effect on SSB bias (Figures \ref{M_om_em_BH_MF_relbias_ssb} and \ref{M_om_em_BH_ME_relbias_ssb})
+Like R and R+S OMs, estimating a stock-recruit relationship had little discernible effect on SSB bias (Figures \ref{M_om_em_BH_MF_relbias_ssb} and \ref{M_om_em_BH_ME_relbias_ssb}).
 
@@
 
-Among the process error configurations we used in operating models, we found AIC to be accurate for process errors on recruitment and survival.
+Among the process error configurations we used in OMs, we found AIC to be accurate for process errors on recruitment and survival.
 Fitting models to other OMs rarely preferred R+S EMs and R and R+S EMs were nearly always selected for the matching OMs.
@@
 Higher observation error with variation in fishing pressure can produce retrospective patterns even when the the process error assumption is correct, but this retrospective pattern can be reduced if M or the correct stock recruit relationship is estimated.
-Retrospective patterns tended to be more variable for recruitment and can be large even when the estimating model is correct.
+Retrospective patterns tended to be more variable for recruitment and can be large even when the EM is correct.
 Therefore, we recommend emphasis on inspection of retrospective patterns primarily for SSB and F, but further research on retrospective patterns in other assessment model parameters and derived output may be beneficial.

### 2024-01-26 Ecov_study/recruitment_functions/Workflow_Ecov_Recruitment_Project.Rmd

@@
 
-7. Cleaning up script **compare_projections.R**
-   + evaluates relative error in projections between peel 10 and truth ...
-   + Runtime ...
-   + **<span style="color: blue;">STATUS: pending... beta unstandardized; </span>**
+7. Cleaning up script **compare_RE_SD_projections.R**
+   + evaluates relative error and log_sd (CV) in projections between peel 10 and truth for recruitment, catch, and ssb
+   + summarizes over all 10 years in the projection, and also over just the first 3 years
+   + turns out no OM factors made a difference; pretty boring
+   + Runtime 1.6 hours
+   + **<span style="color: blue;">STATUS: done beta unstandardized; </span>**
 

### 2024-02-01 Ecov_study/recruitment_functions/Workflow_Ecov_Recruitment_Project.Rmd
Update Workflow_Ecov_Recruitment_Project.Rmd
fixing url for workflow

@@
 title: "Workflow for Ecov Recruitment Analysis"
-output: html_notebook
+output:
+  html_document:
+    df_print: paged
 ---
-
+ Note: you can view this as a rendered html file, just click on the file and when it opens (as raw source code) just pre-pend this in front of the url: https://htmlpreview.github.io/?
 <!-- This is an [R Markdown](http://rmarkdown.rstudio.com) Notebook. When you execute code within the notebook, the results appear beneath the code.  -->
@@
 # Project 3 for the SSRTWG investingating the incorporation of environmental covariates into WHAM recruitment models.
+
+## View html version <a href="https://htmlpreview.github.io/?https://github.com/timjmiller/SSRTWG/blob/main/Ecov_study/recruitment_functions/Workflow_Ecov_Recruitment_Project.html"> here</a>.
 

### 2024-02-04 Ecov_study/recruitment_functions/presentation/SRR_presentation.Rmd
starting presentation
draft Rmd beemer

---
title: "TORs2 and 3"
output: beamer_presentation
slide_level: 3
theme: "Darmstadt"
logo: WHOI_NOAA_logo_smaller.png
date: "2024-02-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## R Markdown

If you have trouble knitting this file, check out this page: https://stackoverflow.com/questions/67696286/error-generating-pdf-using-knitr-to-pdf-in-rstudio

* need to make the logo smaller on title and content slides (put it in the footer)

<!-- This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>. -->

<!-- When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. -->

## Outline

- 'Stock' parameters
- Fishery and index Parameters
- OM factors and simulated data examples
- EM models
- Beta standardization
- Analysis & Results
- Conclusions
- Future Work

## Stock parameters

- make some plots of maturity at age, weight at age, true deterministic SRR

## Fishery and index parameters

-- make some plots of fishery and index selectivity at age, q-index

## OM factors and simulated data examples

- make some plots to illustrate OM inputs and resulting simulated data generated

## EM models

- simple table listing EMs

## Couple of slides for Beta standardization

- greg's plots

## Analysis

\begin{enumerate}
  \item Convergence of the estimating models
  \item Model identifiability of an underlying stock recruitment model and/or an underlying relationship
   between environmental covariate 
  \item $\delta$AIC and model probability
  \item Assessment error (recruitment, spawning stock biomass, and Fbar)
  \item Bias of estimated parameters
  \item Mohn's $\rho$
  \item Projection performance relative to assumptions about the environmental covariate
\end{enumerate}

<!-- ```{r cars, echo = TRUE} -->
<!-- summary(cars) -->
<!-- ``` -->

## Slide with Plot

```{r pressure}
plot(pressure)
```

### 2024-02-05 Ecov_study/recruitment_functions/presentation/SRR_presentation.Rmd
working on presentation
example syntax for creating figure, pulling in existing figure, and printing table

@@
 output: beamer_presentation
+fig_caption: FALSE
 slide_level: 3
 theme: "Darmstadt"
-logo: WHOI_NOAA_logo_smaller.png
-date: "2024-02-02"
+titlegraphic: WHOI_NOAA_logo_smaller.png
+date: "`r Sys.Date()`"
 ---
 
-```{r setup, include=FALSE}
+```{r setup, echo=FALSE}
 knitr::opts_chunk$set(echo = FALSE)
@@
 
-## Stock parameters
+## Stock parameters for generic gadid
 
-- make some plots of maturity at age, weight at age, true deterministic SRR
+```{r , echo=FALSE, message=FALSE, warnings=FALSE, results='hide'}
+library(patchwork)
+library(tidyverse)
+library(wham)
+naa_om_inputs <- readRDS(file.path(here::here(),"Project_0","inputs", "NAA_om_inputs.RDS"))
+om.data <- naa_om_inputs[[1]]$data
+mat.waa <- as_tibble(cbind(Age=seq(1, om.data$n_ages), WAA=om.data$waa[1,1,], Mat = om.data$mature[1,] ))
+temp <- fit_wham(naa_om_inputs[[1]], do.fit = FALSE, MakeADFun.silent = TRUE)
+SRab <- exp(c(temp$rep$log_SR_a[1], temp$rep$log_SR_b[1]))
+srr <- as_tibble(cbind(SSB=seq(0,3.5e5, length=1000), Recruits= (SRab[1]*seq(0,3.5e5, length=1000)/(1+SRab[2]*seq(0,3.5e5, length=1000)) )  ))
+
+mat.plot <- ggplot(mat.waa, aes(x=Age, y=Mat)) +
+  geom_line(linewidth=1) +
+  geom_point(size=2.5) +
+  ylab('Maturity at age') +
+  scale_x_continuous(breaks=mat.waa$Age, labels=as.character(mat.waa$Age))+
+  theme(axis.text.x = element_text(size = 15))   + 
+  theme(axis.text.y = element_text(size = 15)) +
+  theme(axis.title.x = element_text(size = 16))   + 
+  theme(axis.title.y = element_text(size = 16))  
+ggsave(mat.plot, filename=file.path(here::here(), 'Ecov_study','recruitment_functions', 'presentation','plots', 
+                           'mat.plot.png'), height=3, width=4)
+waa.plot <- ggplot(mat.waa, aes(x=Age, y=WAA)) +
+  geom_line(linewidth=1, color='blue') +
+  geom_point(size=2.5, color='blue') +
+  ylab('Weight at age (SSB and Catch)') +
+  scale_x_continuous(breaks=mat.waa$Age, labels=as.character(mat.waa$Age))  +
+  theme(axis.text.x = element_text(size = 15))   + 
+  theme(axis.text.y = element_text(size = 15))  + 
+  theme(axis.title.y = element_text(size = 16))  + 
+  theme(axis.title.x = element_text(size = 16))  
+ggsave(waa.plot, filename=file.path(here::here(), 'Ecov_study','recruitment_functions', 'presentation','plots', 
+                           'waa.plot.png'), height=3, width=4)
+
+srr.plot <- ggplot(srr, aes(x=SSB, y=Recruits)) +
+  geom_line(linewidth=1, color='red')  +
+  theme(axis.text.x = element_text(size = 15))   + 
+  theme(axis.text.y = element_text(size = 15))  + 
+  theme(axis.title.y = element_text(size = 16))  + 
+  theme(axis.title.x = element_text(size = 16))  
+  # geom_point(size=2.5, color='blue') +
+  # ylab('Weight at age (SSB and Catch)')
+ggsave(srr.plot, filename=file.path(here::here(), 'Ecov_study','recruitment_functions', 'presentation','plots', 
+                           'srr.plot.png'), height=3, width=8)
+
+bio.plot <- (mat.plot|waa.plot)/srr.plot
+# ggsave(bio.plot, filename=file.path(here::here(), 'Ecov_study','recruitment_functions', 'presentation','plots', 'bio.plot.png'),  height=7, width=8)
+
+plot(bio.plot)
+```
+<!-- knitr::include_graphics("bio.plot.png") -->
+<!-- \includegraphics{file.path(here::here(), 'Ecov_study','recruitment_functions', 'presentation','plots', 'bio.plot.pdf')} -->
+<!-- - make some plots of maturity at age, weight at age, true deterministic SRR -->
 
@@
 
--- make some plots of fishery and index selectivity at age, q-index
+<!-- -- example chunk to include figure from file -->
+```{r,  echo=FALSE, message=FALSE, warnings=FALSE, results='hide', out.width='75%', fig.align='center'}
+gf_info = make_basic_info()
+
+#selectivity is not changing
+gf_selectivity <- list(model       =c(rep("logistic", gf_info$n_fleets),rep("logistic", gf_info$n_indices)),
+                       initial_pars=rep(list(c(5,1)), gf_info$n_fleets + gf_info$n_indices)) #fleet, index
+
+a50 = gf_selectivity$initial_pars[[1]][1]
+k= gf_selectivity$initial_pars[[1]][2]
+sel.tmp <- 1.0/(1.0 + exp(-(seq(1, om.data$n_ages) - a50)/k))
+sel <- sel.tmp/sel.tmp[om.data$n_ages]
+fish.sel <- as_tibble(cbind(Age=seq(1, om.data$n_ages), Sel=sel))
+ggplot(fish.sel, aes(x=Age, y=Sel)) + 
+  geom_line(color='green4', linewidth=1) +
+  geom_point(color='green4', size=2.5) +
+  ylab('Fishery Selectivity at Age') +
+  theme(axis.text.x = element_text(size = 15))   + 
+  theme(axis.text.y = element_text(size = 15))  + 
+  theme(axis.title.y = element_text(size = 16))  + 
+  theme(axis.title.x = element_text(size = 16))
+
+```
 
@@
 
+<!-- -- example chunk to include figure from file -->
+```{r, out.width='25%', fig.align='center'}
+ knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'model_probability_beta_fix.png') )
+```
+
 - make some plots to illustrate OM inputs and resulting simulated data generated
+- looks like i can have multiple bullets below a figure
 
@@
 
-- simple table listing EMs
+- simple bullets above table
+- we only have 6 EMs that were considered
+
+```{r,  echo=FALSE, message=FALSE, warnings=FALSE}
+
+df.ems    <- readRDS(file.path(here::here(),"Ecov_study", "recruitment_functions", "inputs", "df.ems.RDS"))
+
+em_tib <- as_tibble(df.ems) %>%
+  mutate(SR=ifelse(r_mod==2, 'Mean', 'BH')) %>%
+  mutate(EM_mod = paste0(SR, "_", ecov_how), EM=seq(1,6))
+
+   knitr::kable(em_tib)
+  # print(em_tib)
+
+```
 

### 2024-02-05 Ecov_study/recruitment_functions/presentation/make_ppt_plots.R

@@ -1,0 +1,34 @@
+library(patchwork)
+library(tidyverse)
+library(wham)
+naa_om_inputs <- readRDS(file.path(here::here(),"Project_0","inputs", "NAA_om_inputs.RDS"))
+om.data <- naa_om_inputs[[1]]$data
+mat.waa <- as_tibble(cbind(Age=seq(1, om.data$n_ages), WAA=om.data$waa[1,1,], Mat = om.data$mature[1,] ))
+temp <- fit_wham(naa_om_inputs[[1]], do.fit = FALSE, MakeADFun.silent = TRUE)
+SRab <- exp(c(temp$rep$log_SR_a[1], temp$rep$log_SR_b[1]))
+srr <- as_tibble(cbind(SSB=seq(0,3.5e5, length=1000), Recruits= (SRab[1]*seq(0,3.5e5, length=1000)/(1+SRab[2]*seq(0,3.5e5, length=1000)) )  ))
+
+mat.plot <- ggplot(mat.waa, aes(x=Age, y=Mat)) +
+  geom_line(linewidth=1) +
+  geom_point(size=2.5) +
+  ylab('Maturity at age')
+ggsave(mat.plot, filename=file.path(here::here(), 'Ecov_study','recruitment_functions', 'presentation','plots', 
+                           'mat.plot.png'), height=3, width=4)
+waa.plot <- ggplot(mat.waa, aes(x=Age, y=WAA)) +
+  geom_line(linewidth=1, color='blue') +
+  geom_point(size=2.5, color='blue') +
+  ylab('Weight at age (SSB and Catch)')
+ggsave(waa.plot, filename=file.path(here::here(), 'Ecov_study','recruitment_functions', 'presentation','plots', 
+                           'waa.plot.png'), height=3, width=4)
+
+srr.plot <- ggplot(srr, aes(x=SSB, y=Recruits)) +
+  geom_line(linewidth=1, color='red') 
+# geom_point(size=2.5, color='blue') +
+# ylab('Weight at age (SSB and Catch)')
+ggsave(srr.plot, filename=file.path(here::here(), 'Ecov_study','recruitment_functions', 'presentation','plots', 
+                           'srr.plot.png'), height=3, width=8)
+bio.plot <- (mat.plot|waa.plot)/srr.plot
+pdf(file.path(here::here(),'Ecov_study','recruitment_functions','presentation','plots', 'bio.plot.pdf') )
+bio.plot
+dev.off()
+

### 2024-02-05 Ecov_study/recruitment_functions/presentation/SRR_presentation.Rmd
presentation formatting
cleaning up slide levels and theme

@@
 fig_caption: FALSE
-slide_level: 3
-theme: "Darmstadt"
-titlegraphic: WHOI_NOAA_logo_smaller.png
+theme: "Singapore"
+slide-level: 3
+# titlegraphic: WHOI_NOAA_logo_smaller.png
+header-includes: 
+- \addtobeamertemplate{title page}{\centering \includegraphics[width=16.5cm]{WHOI_NOAA_logo_smaller.png}\par }{}
+classoption: "aspectratio=169"
 date: "`r Sys.Date()`"
@@
 
+# TORs
+
+## TOR 2
+
+- paste TOR2
+
+## TOR3
+
+- paste TOR3
+
 ## R Markdown
@@
 
-* need to make the logo smaller on title and content slides (put it in the footer)
+<!-- * need to make the logo smaller on title and content slides (put it in the footer) -->
 
@@
 <!-- When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. -->
+
+# Background
 
@@
 - Future Work
+
+# Inputs
 
@@
 
+# OM
+
 ## OM factors and simulated data examples
@@
 - looks like i can have multiple bullets below a figure
+
+## Couple of slides for Beta standardization
+
+- greg's plots
+
+# EM
 
@@
 
-## Couple of slides for Beta standardization
+# Analysis & Results
 
-- greg's plots
-
-## Analysis
+## Analyses
 
@@
    between environmental covariate
-  \item $\delta$AIC and model probability
+  \item $\Delta$AIC and model probability
   \item Assessment error (recruitment, spawning stock biomass, and Fbar)
@@
 
+## Convergence
+
+## Model Identifiability
+
+##  $\Delta$AIC and model probability
+
+## Assessment error (recruitment, spawning stock biomass, and Fbar)
+
+## Bias of estimated parameters
+
+## Mohn's $\rho$
+
+## Projection performance relative to assumptions about the environmental covariate
+
+# Conclusions
+
+## Take-aways
+- These are the take-aways (copy from WP)
+
+# Future Work
+
+## List of what's next
+
+- This is what we suggest for follow-up
+
+---
+
+Acknowledgements
+
+- This work could not have been completed without the use of Azure computing (NOAA) and MIT (... greg to fill in)
+- We thank other members of the SSRTWG for thoughtful comments during earlier discussions and presentations of this work
+
 <!-- ```{r cars, echo = TRUE} -->
@@
 
-## Slide with Plot
+<!-- ## Slide with Plot -->
 
-```{r pressure}
-plot(pressure)
-```
+<!-- ```{r pressure} -->
+<!-- plot(pressure) -->
+<!-- ``` -->

### 2024-02-06 Ecov_study/recruitment_functions/presentation/SRR_presentation.Rmd
presentation updates

--- 2024-02-04 Ecov_study/recruitment_functions/presentation/SRR_presentation.Rmd
@@
 ---
-title: "TORs2 and 3"
+title: "Identification and performance of stock-recruitment functions in state
+space assessment models"
+subtitle: Working Paper 1
+author: Greg Britten, Liz Brooks, Tim Miller
 output: beamer_presentation
-slide_level: 3
-theme: "Darmstadt"
-logo: WHOI_NOAA_logo_smaller.png
-date: "2024-02-02"
+fig_caption: FALSE
+theme: "Singapore"
+slide-level: 3
+# titlegraphic: WHOI_NOAA_logo_smaller.png
+header-includes: 
+- \addtobeamertemplate{title page}{\centering \includegraphics[width=9.5cm]{WHOI_NOAA_logo_smaller.png}\par }{}
+classoption: "aspectratio=169"
+date: "`r Sys.Date()`"
 ---
+# Background
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE)
-```
+## Introduction
 
-## R Markdown
+- State Space Research Track Working Group (SSRTWG) is investigating performance of the Woods Hole Assessment Model (WHAM)
+- Simulation studies with data generated from operating models (OMs), then fit with a series of estimation models (EMs)
+- This framework allows us to evaluate how the EM fits compare to the known OM "true values", evaluate model selection, bias, precision, etc.
 
-If you have trouble knitting this file, check out this page: https://stackoverflow.com/questions/67696286/error-generating-pdf-using-knitr-to-pdf-in-rstudio
+## Terms of Reference (TORs) Addressed
 
-* need to make the logo smaller on title and content slides (put it in the footer)
+- TOR 2: Investigate the efficacy of estimating stock-recruit functions within state-space models and their utility in generating scientific advice.
 
-<!-- This is an R Markdown presentation.
-Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents.
-For more details on using R Markdown see <http://rmarkdown.rstudio.com>.
--->
-
-<!-- When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. -->
+- TOR 3: Develop guidelines for including ecosystem and environmental effects in assessment models and how to treat them for generating biological reference points and scientific advice.
 
@@
 
-- 'Stock' parameters
-- Fishery and index Parameters
+- 'Stock' parameters, fishery and index parameters
 - OM factors and simulated data examples
@@
 
-## Stock parameters
+<!-- ==================================================================================== -->
 
-- make some plots of maturity at age, weight at age, true deterministic SRR
+# Inputs
+
+## Stock parameters for generic gadid
+
+```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+ knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'presentation', 'plots', 'bio.plot.png') )
+```
+
+- Natural mortality = 0.2 at all ages
+- Maturity, weight, natural mortality are time-invariant
 
@@
 
--- make some plots of fishery and index selectivity at age, q-index
+<!-- -- example chunk to include figure from file -->
+```{r,  echo=FALSE, message=FALSE, warnings=FALSE, results='hide', out.width='60%', fig.align='center'}
+library(tidyverse)
+library(wham)
+naa_om_inputs <- readRDS(file.path(here::here(),"Project_0","inputs", "NAA_om_inputs.RDS"))
+om.data <- naa_om_inputs[[1]]$data
+
+gf_info = make_basic_info()
+
+#selectivity is not changing
+gf_selectivity <- list(model       =c(rep("logistic", gf_info$n_fleets),rep("logistic", gf_info$n_indices)),
+                       initial_pars=rep(list(c(5,1)), gf_info$n_fleets + gf_info$n_indices)) #fleet, index
+
+#fishery
+a50 = gf_selectivity$initial_pars[[1]][1]
+k= gf_selectivity$initial_pars[[1]][2]
+sel.tmp <- 1.0/(1.0 + exp(-(seq(1, om.data$n_ages) - a50)/k))
+sel <- sel.tmp/sel.tmp[om.data$n_ages]
+fish.sel <- as_tibble(cbind(Age=seq(1, om.data$n_ages), Sel=sel))
+ggplot(fish.sel, aes(x=Age, y=Sel)) +
+  geom_line(color='green4', linewidth=1) +
+  geom_point(color='green4', size=2.5) +
+  ylab('Fishery Selectivity at Age') +
+  theme(axis.text.x = element_text(size = 15))   +
+  theme(axis.text.y = element_text(size = 15))  +
+  theme(axis.title.y = element_text(size = 16))  +
+  theme(axis.title.x = element_text(size = 16))
+
+#index-1 and index-2 have same selectivity as fleet (why?!)
+# a50.i1 = gf_selectivity$initial_pars[[2]][1]
+# k.i1 = gf_selectivity$initial_pars[[2]][2]
+# 
+# a50.i2 = gf_selectivity$initial_pars[[2]][1]
+# k.i2 = gf_selectivity$initial_pars[[2]][2]
+
+ind.time <- naa_om_inputs[[1]]$data$fracyr_indices[1,]
+q.hi <- (naa_om_inputs[[1]]$data$q_upper)
+q.lo <- (naa_om_inputs[[1]]$data$q_lower)
+ind.q <-  unique((q.lo + q.hi/(1 + exp(-1*naa_om_inputs[[1]]$par$logit_q)) ))
+
+```
+- two fishery independent indices were also generated, taking place at `r ind.time` yr
+- catchability for both indices was `r ind.q `; selectivity was same as fishery
+
+<!-- ==================================================================================== -->
+
+# OM
 
@@
 
+<!-- -- example chunk to include figure from file -->
+```{r, echo=FALSE, message=FALSE, warnings=FALSE, out.width='50%', fig.align='center'}
+ knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'model_probability_beta_fix.png') )
+```
+
 - make some plots to illustrate OM inputs and resulting simulated data generated
-
-## EM models
-
-- simple table listing EMs
+- looks like i can have multiple bullets below a figure
 
@@
 
-## Analysis
+<!-- ==================================================================================== -->
+
+# EM
+
+## EM models
+
+- simple bullets above table
+- we only have 6 EMs that were considered
+
+```{r,  echo=FALSE, message=FALSE, warnings=FALSE}
+
+df.ems    <- readRDS(file.path(here::here(),"Ecov_study", "recruitment_functions", "inputs", "df.ems.RDS"))
+
+em_tib <- as_tibble(df.ems) %>%
+  mutate(SR=ifelse(r_mod==2, 'Mean', 'BH')) %>%
+  mutate(EM_mod = paste0(SR, "_", ecov_how), EM=seq(1,6))
+
+   knitr::kable(em_tib)
+  # print(em_tib)
+
+```
+
+<!-- ==================================================================================== -->
+
+# Analysis & Results
+
+## Analyses
 
@@
 
+## Convergence
+
+## Model Identifiability
+
+##  $\Delta$AIC and model probability
+
+## Assessment error (recruitment, spawning stock biomass, and Fbar)
+
+## Bias of estimated parameters
+
+## Mohn's $\rho$
+
+## Projection performance relative to assumptions about the environmental covariate
+
+<!-- ==================================================================================== -->
+
+# Conclusions
+
+## Take-aways
+- These are the take-aways (copy from WP)
+
+# Future Work
+
+## List of what's next
+
+- This is what we suggest for follow-up
+- Note that WP1-Appendix looked at $\sigma_{R}$=0.5 and found no difference from results in WP1
+
+<!-- ==================================================================================== -->
+
+---
+
+Acknowledgements
+
+- This work could not have been completed without the use of Azure computing (NOAA) and MIT (... greg to fill in)
+- We thank other members of the SSRTWG for thoughtful comments during earlier discussions and presentations of this work
+
+```{r, echo=FALSE, message=FALSE, warnings=FALSE, out.width='35%', fig.align='center'}
+ knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'presentation', 'abacus_crop.jpg') )
+```
+
+<!-- ==================================================================================== -->
+
+<!-- ## R Markdown -->
+
+<!-- If you have trouble knitting this file, check out this page: https://stackoverflow.com/questions/67696286/error-generating-pdf-using-knitr-to-pdf-in-rstudio -->
+
 <!-- ```{r cars, echo = TRUE} -->

### 2024-02-06 Ecov_study/recruitment_functions/presentation/SRR_presentation.Rmd

@@
 ## OM factors and simulated data examples
+```{r,  echo=FALSE, message=FALSE, warnings=FALSE}
 
-<!-- -- example chunk to include figure from file -->
-```{r, echo=FALSE, message=FALSE, warnings=FALSE, out.width='50%', fig.align='center'}
- knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'model_probability_beta_fix.png') )
+df.oms    <- readRDS(file.path(here::here(),"Ecov_study", "recruitment_functions", "inputs", "df.oms.RDS"))
+
+om.table <- cbind( unique(df.oms$Fhist), unique(df.oms$R_sig), unique(df.oms$NAA_cor), unique(df.oms$Ecov_effect), unique(df.oms$Ecov_re_cor), unique(df.oms$obs_error)  )
+colnames(om.table) <- c('Fhistory', 'R_sigma', 'R_cor', 'Ecov_effect', 'Ecov_cor', 'Obs_error')
+   knitr::kable(om.table)
+
+```
+- These levels were combined factorially with 4 stock recruit models (all Beverton-Holt) for 256 different OM
+- Factors that did not vary were the observation and process error of the Ecov (bot fixed at 0.1)
+
+## Beverton-Holt functional (1 with no Ecov effect, 3 with effect)
+
+```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+ knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'paper', 'dRdX_1RdRdx.pdf') )
 ```
 
-- make some plots to illustrate OM inputs and resulting simulated data generated
-- looks like i can have multiple bullets below a figure
+## Beta standardization
 
-## Couple of slides for Beta standardization
+```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+ knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'paper', 'dRdX.pdf') )
+```
 
-- greg's plots
+## Some Example OM Simulated Data (100 data sets for each OM)
+- still need to make this
 
@@
 
-- simple bullets above table
-- we only have 6 EMs that were considered
+- All OMs used Beverton-Holt stock recruit relationship (SRR)
+- For the EMs, we also fit a mean SRR with random effects
+- There were 6 total EMs fit to each simulated data set (100) for all 256 OMs 
+- 153,600 fitted models X 2 for $\beta$ unstandardized and $\beta$ standardized cases
 
@@
   mutate(SR=ifelse(r_mod==2, 'Mean', 'BH')) %>%
-  mutate(EM_mod = paste0(SR, "_", ecov_how), EM=seq(1,6))
+  mutate(EM_mod = paste0(SR, "_", ecov_how), EM=seq(1,6)) %>%
+  rename(Ecov_how=ecov_how, SRR=SR) %>%
+  mutate(Ecov=c('None', 'Controlling', 'None', 'Controlling', 'Limiting', 'Masking' )) %>%
+  relocate(EM, SRR, Ecov_how, Ecov, EM_mod) %>%
+  select(-r_mod)
 
@@
 
-<!-- ```{r cars, echo = TRUE} -->
-<!-- summary(cars) -->
+<!-- <!-- -- example chunk to include figure from file --> -->
+<!-- ```{r, echo=FALSE, message=FALSE, warnings=FALSE, out.width='50%', fig.align='center'} -->
+<!--  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'model_probability_beta_fix.png') ) -->
 <!-- ``` -->
-
-<!-- ## Slide with Plot -->
-
-<!-- ```{r pressure} -->
-<!-- plot(pressure) -->
-<!-- ``` -->

### 2024-02-06 Ecov_study/recruitment_functions/presentation/SRR_presentation.Rmd
more presentation progress
update with some convergence stuff

@@
 ---
+
 # Background
@@
 ```
+
 - These levels were combined factorially with 4 stock recruit models (all Beverton-Holt) for 256 different OM
@@
 
-## Beverton-Holt functional (1 with no Ecov effect, 3 with effect)
+## OM factor: Beverton-Holt functional (1 with no Ecov effect, 3 with effect)
 
@@
 
-## Beta standardization
+## OM factor: Beta standardization
 
@@
 
-## Convergence
+- 
+- Results will be shown for $\beta$ unstandardized case, because $\beta$ standardization had no appreciable effect
+
+## Crashes
+
+- ~3.3-3.5% crashes, typically for Ecov=Masking ("BH_4") and $\sigma_{R}$=0.1
+```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+ knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'crashed.models.plot_beta_fix.png') )
+```
+
+## Convergence Criteria
+
+\begin{enumerate}
+ \item the optimization routine (stats::nlminb) completed without error
+ \item the stats::nlminb convergence flag = 0 indicated successful convergence
+ \item the maximum absolute value of the gradient of the log-likelihood is < 1e-6
+ \item TMB::sdreport provided non-NA values for all fixed effects standard errors
+ \item TMB::sdreport provided all standard errors < 100
+\end{enumerate}
+
+- Models not satisfying all of these conditions were dropped from Results summaries
+
+## Unconverged Models (from EM having'best' EM per OM-Simulation)
+
+## Unconverged Models (considering all EMs per OM-Simulation)
+
+## Unconverged Model - The Offending Parameters
+
+## Unconverged Model - Even More Offending Parameters
 
@@
 
-## List of what's next
+## Recommendations for further analyses
 

### 2024-02-06 Ecov_study/recruitment_functions/presentation/SRR_presentation.Rmd
adding acknowledgement

@@
 
-- This work could not have been completed without the use of Azure computing (NOAA) and MIT (... greg to fill in)
+- This work could not have been completed without the use of Azure computing (NOAA) and the MIT engaging cluster (courtesy of Chris Hill)
 - We thank other members of the SSRTWG for thoughtful comments during earlier discussions and presentations of this work

### 2024-02-06 Project_0/paper/paper.tex

@@
 catchability and natural mortality), degree of variation, and
-correlation of process errors. We fit 20 different estimating models to
+correlation of process errors. We fit 20 different estimation models to
 each of 100 simulated set of observations with alternative assumptions
@@
 Bias in spawning stock biomass estimation was generally low for
-estimating models that assumed the correct source of process error when
+estimation models that assumed the correct source of process error when
 there was lower observation error. Reliable estimation of stock-recruit
@@
 variable estimates of natural mortality. For OMs where there was bias in
-natural mortality due to high observation error, estimating the
+natural mortality due to high observation error, estimation the
 stock-recruit relationship seemed to remove the bias. However,
@@
 observation error is high. When models did exhibit some retrospective
-pattern, estimating the mean natural mortality rate tended to remove it.
+pattern, estimation the mean natural mortality rate tended to remove it.
 
@@
 
-Here we conduct a simulation study with operating models varying by
-degree of observation error uncertainty, source and variability of
-process error, and fishing history. The simulations from these operating
-models are fitted with estimating models that make alternative
-assumptions for sources of process error, whether a stock-recruit model
-was estimated, and whether M is estimated. We evaluate whether AIC can
+Here we conduct a simulation study with operating models (OMs) varying
+by degree of observation error uncertainty, source and variability of
+process error, and fishing history. The simulations from these OMs are
+fitted with estimation models (EMs) that make alternative assumptions
+for sources of process error, whether a stock-recruit model was
+estimated, and whether M is estimated. We evaluate whether AIC can
 correctly determine the correct source of process error and stock
@@
 We used the Woods Hole Assessment Model (WHAM) to configure operating
-and estimation models in our simulation study
-\citep[]{millerstock20,stockmiller21}. WHAM is an R package freely
-available as a github repository. For this study we used
+and EMs in our simulation study \citep[]{millerstock20,stockmiller21}.
+WHAM is an R package freely available as a github repository. For this
+study we used
 \href{https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3}{version
 1.0.6.9000, commit 77bbd94}. This packages has also been used to
-configure operating and estimating models for closed loop simulations
-evaluating index-based assessment methods \citep{legaultetal23} and is
-used for management of haddock, butterfish, American plaice, bluefish,
-Atlantic cod, and black sea bass in the Northeast US.
-
-We completed a simulation study with a number of operating models that
-can be categorized based on where process error random effects are
-assumed: abundance at age (R, R+S), natural mortality (R+M), fleet
-selectivity (R+Sel), or index catchability (R+q). For each operating
-model assumptions about variance of process errors and observations are
-required and the values we used were based on a review of the range of
-estimates from recent applications of WHAM in management of stocks of
-haddock, butterfish, and American plaice in the NE US.
-
-In total, we configured 72 operating models with alternative assumptions
-about the source and variability of process errors, level of observation
-error in indices and age composition data, and contrast in fishing
-pressure over time. We fitted 20 estimating models to observations from
-each of 100 simulations where process errors were also simulated.
-Estimating models made alternative assumptions about the source of
-process errors and whether natural mortality (or the mean for models
-with process error in natural mortality) was estimated and whether a
-Beverton-Holt stock recruit relationship was estimated within the
-estimation model. Details of each of the operating and estimating models
-are described below.
+configure operating and EMs for closed loop simulations evaluating
+index-based assessment methods \citep{legaultetal23} and is used for
+management of haddock, butterfish, American plaice, bluefish, Atlantic
+cod, and black sea bass in the Northeast US.
+
+We completed a simulation study with a number of OMs that can be
+categorized based on where process error random effects are assumed:
+abundance at age (R, R+S), natural mortality (R+M), fleet selectivity
+(R+Sel), or index catchability (R+q). For each OM assumptions about
+variance of process errors and observations are required and the values
+we used were based on a review of the range of estimates from recent
+applications of WHAM in management of stocks of haddock, butterfish, and
+American plaice in the NE US.
+
+In total, we configured 72 OMs with alternative assumptions about the
+source and variability of process errors, level of observation error in
+indices and age composition data, and contrast in fishing pressure over
+time. We fitted 20 EMs to observations from each of 100 simulations
+where process errors were also simulated. EMs made alternative
+assumptions about the source of process errors and whether natural
+mortality (or the mean for models with process error in natural
+mortality) was estimated and whether a Beverton-Holt stock recruit
+relationship was estimated within the EM. Details of each of the
+operating and EMs are described below.
 
 We did not use the bias-correction feature for process errors or
-observations described by \citet{stockmiller21} for operating and
-estimating models.Simulations and model fitting were all carried out on
-the University of Massachusetts Green High-Performance Computing
-Cluster. All code we used to perform the simulation study and summarize
-results can be found at
+observations described by \citet{stockmiller21} for operating and EMs.
+Simulations and model fitting were all carried out on the University of
+Massachusetts Green High-Performance Computing Cluster. All code we used
+to perform the simulation study and summarize results can be found at
 \url{https://github.com/timjmiller/SSRTWG/tree/main/Project_0/code}.
@@
 a logistic curve with \(a_{50}\) = 2.89 and slope = 0.88 and assumed
-known in all estimation models (Figure \ref{om_maturity}).
+known in all EMs (Figure \ref{om_maturity}).
 
@@
 We assumed a Beverton-Holt stock recruit function with constant
-pre-recruit mortality parameters for all operating models. All
-post-recruit productivity components are constant in the NAA and survey
-catchability process error operating models. Therefore steepness and
-unfished recruitment are also constant over the time period for those
-operating models \citep{millerbrooks21}. We specified unfished
-recruitment = \(R_0 = e^{10}\) and \(\Fmsy = F_{40\%} = 0.348\) equated
-to a steepness of 0.69 and \(\alpha=0.60\) and
-\(\beta = 2.4 \times 10^{-5}\) for the \[
+pre-recruit mortality parameters for all OMs. All post-recruit
+productivity components are constant in the NAA and survey catchability
+process error OMs. Therefore steepness and unfished recruitment are also
+constant over the time period for those OMs \citep{millerbrooks21}. We
+specified unfished recruitment = \(R_0 = e^{10}\) and
+\(\Fmsy = F_{40\%} = 0.348\) equated to a steepness of 0.69 and
+\(\alpha=0.60\) and \(\beta = 2.4 \times 10^{-5}\) for the \[
 N_{1,y} = \frac{\alpha \text{SSB}_{y-1}}{1 + \beta \text{SSB}_{y-1}} 
@@
 
-Two alternative fishing histories were used for operating models. In the
-first scenario, the stock experiences overfishing for the first 20 years
-and fishing at \Fmsy for the last 20 years. In the second scenario, the
+Two alternative fishing histories were used for OMs. In the first
+scenario, the stock experiences overfishing for the first 20 years and
+fishing at \Fmsy for the last 20 years. In the second scenario, the
 stock is fished at \Fmsy for the entire time period. The magnitude of
@@
 \Citet{legaultetal23} also used similar approaches to defining fishing
-mortality histories for operating models.
+mortality histories for OMs.
 
@@
 
-For operating models with time-varying random effects for M, steepness
-is not constant, but we used the same alpha and beta parameters as other
-operating models this equates to a steepness and R0 at the mean of the
-time series process for M. For operating models with time-varying random
-effects for fishery selectivity, \Fmsy is also not constant however we
-use the same F history as other operating models which corresponds to
-Fmsy at the mean selectivity parameters.
+For OMs with time-varying random effects for M, steepness is not
+constant, but we used the same alpha and beta parameters as other OMs
+this equates to a steepness and R0 at the mean of the time series
+process for M. For OMs with time-varying random effects for fishery
+selectivity, \Fmsy is also not constant however we use the same F
+history as other OMs which corresponds to Fmsy at the mean selectivity
+parameters.
 
@@
 the standard deviation for log-aggregate indices was 0.4 and that for
-the age composition observations was 1.5. For all estimating models,
-standard deviation for log-aggregate observations was assumed known
-whereas that for the logistic-normal age composition observations was
-estimated.
+the age composition observations was 1.5. For all EMs, standard
+deviation for log-aggregate observations was assumed known whereas that
+for the logistic-normal age composition observations was estimated.
 
@@
 assumptions, and marginal variance assumptions for recruitment and
-survival results in 24 different operating models with random effects on
-recruitment and survival (Table \ref{naa_om_table}).
+survival results in 24 different OMs with random effects on recruitment
+and survival (Table \ref{naa_om_table}).
 
@@
 assumptions, and marginal variance and autocorrelation assumptions for
-natural mortality results in 16 different operating models with random
-effects on natural mortality (Table \ref{M_om_table}). Internally, WHAM
-treats natural mortality as a log-transformed parameter \[
+natural mortality results in 16 different OMs with random effects on
+natural mortality (Table \ref{M_om_table}). Internally, WHAM treats
+natural mortality as a log-transformed parameter \[
 \log M_y = \mu_M + \epsilon_{M,y}
@@
 \(\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)\). The values
-for the marginal variance and correlation parameters assumed in the
-operating models are given in Table \ref{M_om_table}.
+for the marginal variance and correlation parameters assumed in the OMs
+are given in Table \ref{M_om_table}.
 
@@
 assumptions, and marginal variance and autocorrelation assumptions for
-selectivity parameters results in 16 different operating models with
-random effects on selectivity (Table \ref{sel_om_table}). Internally,
-WHAM treats selectivity parameter \(s\) as a logit-transformed parameter
-\[
+selectivity parameters results in 16 different OMs with random effects
+on selectivity (Table \ref{sel_om_table}). Internally, WHAM treats
+selectivity parameter \(s\) as a logit-transformed parameter \[
 \log\left(\frac{p_{s,y}-l_{s}}{u_{s}-p_{s,y}}\right) = \mu_s + \epsilon_{s,y}
@@
 slope parameters and we assume bounds of 0 and 10 for all selectivity
-parameters for all operating and estimating models. The values for the
-marginal variance and correlation parameters assumed in the operating
-models are given in Table \ref{sel_om_table}.
+parameters for all operating and EMs. The values for the marginal
+variance and correlation parameters assumed in the OMs are given in
+Table \ref{sel_om_table}.
 
@@
 assumptions, and marginal variance and autocorrelation assumptions for
-catchability results in 16 different operating models with random
-effects on catchability (Table \ref{q_om_table}). Internally, WHAM
-treats catchability for index \(i\) as a logit-transformed parameter \[
+catchability results in 16 different OMs with random effects on
+catchability (Table \ref{q_om_table}). Internally, WHAM treats
+catchability for index \(i\) as a logit-transformed parameter \[
 \log\left(\frac{q_{i,y}-l_{i}}{u_{i}-q_{i,y}}\right) = \mu_i + \epsilon_{i,y}
@@
 specified by the user. Here we assume bounds of 0 and 1000 for all
-operating and estimating models. For operating and estimation models
-with process errors on catchability, the temporal variation is only
-assumed for the first index. The values for the marginal variance and
-correlation parameters assumed in the operating models are given in
-Table \ref{q_om_table}
-
-\hypertarget{estimating-models}{%
-\subsection*{Estimating models}\label{estimating-models}}
-\addcontentsline{toc}{subsection}{Estimating models}
-
-For each data set simulated from an operating model 20 estimating models
-were fit. A total of 32 different estimating models were fit across all
-operating models where the subset of 20 depended on the source of
-process error (Table \ref{em_table}). The first 20 estimating models in
-Table \ref{em_table} were fit to simulate data sets from R and R+S
-operating models. Estimating models 5 to 24 in Table \ref{em_table} were
-fit to simulate data sets from R+M operating models. Estimating models 5
-to 20 and 25-28 in Table \ref{em_table} were fit to simulate data sets
-from R+Sel operating models. Finally, estimating models 5 to 20 and
-29-32 in Table \ref{em_table} were fit to simulate data sets from R+q
-operating models. The observation error variance of aggregate catch and
-indices were all assumed known at the true values.
+operating and EMs. For operating and EMs with process errors on
+catchability, the temporal variation is only assumed for the first
+index. The values for the marginal variance and correlation parameters
+assumed in the OMs are given in Table \ref{q_om_table}
+
+\hypertarget{estimation-models}{%
+\subsection*{Estimation models}\label{estimation-models}}
+\addcontentsline{toc}{subsection}{Estimation models}
+
+For each data set simulated from an OM 20 EMs were fit. A total of 32
+different EMs were fit across all OMs where the subset of 20 depended on
+the source of process error (Table \ref{em_table}). The first 20 EMs in
+Table \ref{em_table} were fit to simulate data sets from R and R+S OMs.
+EMs 5 to 24 in Table \ref{em_table} were fit to simulate data sets from
+R+M OMs. EMs 5 to 20 and 25-28 in Table \ref{em_table} were fit to
+simulate data sets from R+Sel OMs. Finally, EMs 5 to 20 and 29-32 in
+Table \ref{em_table} were fit to simulate data sets from R+q OMs. The
+observation error variance of aggregate catch and indices were all
+assumed known at the true values.
 
@@
 The first measure of reliability we investigated was frequency of
-convergence when fitting each estimating model to the simulated data
-sets. There are various ways to assess convergence of the fit. We
-summarized 5 alternative categories of convergence.
+convergence when fitting each EM to the simulated data sets. There are
+various ways to assess convergence of the fit. We summarized 5
+alternative categories of convergence.
 
@@
 We measured the frequency of correct model selection using marginal AIC.
-For a given operating model we the set of models that were considered
-all made the same assumptions on whether or not to estimate M and
-whether a stock-recruit function is assumed.
+For a given OM we the set of models that were considered all made the
+same assumptions on whether or not to estimate M and whether a
+stock-recruit function is assumed.
 
@@
 stock-recruit function (null model). Because the proportions of
-simulations where the correct model was chosen for a given set of
-operating model factors implied generally poor performance of AIC, we
-analyzed the effect of variation in SSB on the probability of lower AIC
-of the BH models. We fit binomial generalize linear models to the
-observed AIC choice for each data set as a function of log(SD(SSB)) and
-operating model.
+simulations where the correct model was chosen for a given set of OM
+factors implied generally poor performance of AIC, we analyzed the
+effect of variation in SSB on the probability of lower AIC of the BH
+models. We fit binomial generalize linear models to the observed AIC
+choice for each data set as a function of log(SD(SSB)) and OM.
 
@@
 
-We estimated relative bias \[
-\widehat {\text{RB}}\left(\theta\right) = \frac{1}{N} \sum^N_{i = 1} \frac{\widehat \theta_i - \theta_i}{\theta_i}
-\] of annual SSB and fully-selected fishing mortality for each
-estimation model and constructed 95\% confidence intervals for the
-median relative bias using the binomial distribution approach as in
-\citet{millerhyun18} and \citet{stockmiller21}. We also estimated bias
-of stock-recruit parameters for estimation models that assumed the
-Beverton-Holt stock recruit function.
+We estimated bias as the median of the relative errors across all
+simulations for a given OM and EM combination. \[
+\text{RE}\left(\theta_i\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
+\] of annual SSB and fully-selected fishing mortality for each EM and
+constructed 95\% confidence intervals for the median relative bias using
+the binomial distribution approach as in \citet{millerhyun18} and
+\citet{stockmiller21}. We also estimated bias of stock-recruit
+parameters for EMs that assumed the Beverton-Holt stock recruit function
+and the mean natural mortality rate parameter when estimated.
 
@@
 We estimated Mohn's \(\rho\) for SSB, fully-selected fishing mortality,
-and recruitment for each estimation model. We estimated 7 peels for each
-estimation model. We calculated median 95\% confidence intervals for
-Mohn's \(\rho\) using the same methods as that for relative bias.
+and recruitment for each EM. We estimated 7 peels for each EM. We
+calculated median 95\% confidence intervals for Mohn's \(\rho\) using
+the same methods as that for relative bias.
 
@@
 lower uncertainty in observations. Convergence for most convergence
-types generally declines most estimating models when mean-log natural
-mortality rate is estimated and/or a Beverton-Holt stock recruit
-relationship is estimated even when the process error assumptions of the
-estimating and operating models match. There can be relatively high
-convergence probability of type 4 (an invertible hessian) for EMs that
-do not have the correct process error assumed, but type 5 convergence
-(no very large standard errors estimated from hessian) for these
-estimating models is typically much lower.
+types generally declines most EMs when mean-log natural mortality rate
+is estimated and/or a Beverton-Holt stock recruit relationship is
+estimated even when the process error assumptions of the estimation and
+OMs match. There can be relatively high convergence probability of type
+4 (an invertible hessian) for EMs that do not have the correct process
+error assumed, but type 5 convergence (no very large standard errors
+estimated from hessian) for these EMs is typically much lower.
 
@@
 Marginal AIC accurately determines the correct process error assumptions
-in estimating models when data are generated from R and R+S OMs,
-regardless of whether mean log-natural mortality or a stock recruit
-relationship is estimated (Figures \ref{naa_om_proportion_best_aic_R_MF}
-to \ref{naa_om_proportion_best_aic_SR_ME}). Adding estimation of mean
+in EMs when data are generated from R and R+S OMs, regardless of whether
+mean log-natural mortality or a stock recruit relationship is estimated
+(Figures \ref{naa_om_proportion_best_aic_R_MF} to
+\ref{naa_om_proportion_best_aic_SR_ME}). Adding estimation of mean
 log-natural mortality or a stock recruit relationship separately has a
@@
 or stock recruit relationships were estimated appeared to have little
-effect on the performance of marginal AIC. \#\#\# R+q operating models
-\{-\}
+effect on the performance of marginal AIC.
+
+\hypertarget{rq-operating-models-1}{%
+\subsubsection*{R+q operating models}\label{rq-operating-models-1}}
+\addcontentsline{toc}{subsubsection}{R+q operating models}
 
@@
 
-\hypertarget{aic-performance-for-stock-recruit-relationship}{%
-\subsection*{AIC performance for stock-recruit
-relationship}\label{aic-performance-for-stock-recruit-relationship}}
-\addcontentsline{toc}{subsection}{AIC performance for stock-recruit
+\hypertarget{aic-performance-for-the-stock-recruit-relationship}{%
+\subsection*{AIC performance for the stock-recruit
+relationship}\label{aic-performance-for-the-stock-recruit-relationship}}
+\addcontentsline{toc}{subsection}{AIC performance for the stock-recruit
 relationship}
@@
 
-AIC accuracy for including the B-H stock-recruit relationship was
-estimated to be highest for models OMs and EMs with process errors only
-on recruitment and lowest variance of recruitment process errors
-(Figures \ref{naa_om_MF_BH_glm_AIC_plots} and
-\ref{naa_om_ME_BH_glm_AIC_plots}). Holding variability in SSB constant,
-accuracy of stock-recruit model selection declined with greater variance
-of recruitment and survival process errors. Simultaneously estimating
-mean log-natural mortality has a small negative effect on AIC accuracy.
-
-For R+M OMs and EMs there appeared to little effect of correlation in
+\hypertarget{r-rs-operating-models-2}{%
+\subsubsection*{R, R+S operating models}\label{r-rs-operating-models-2}}
+\addcontentsline{toc}{subsubsection}{R, R+S operating models}
+
+Among R and R+S OMs, AIC accuracy for including the B-H stock-recruit
+relationship was estimated to be highest for R OMs and EMs and lowest
+variance of recruitment process errors (Figures
+\ref{naa_om_MF_BH_glm_AIC_plots} and \ref{naa_om_ME_BH_glm_AIC_plots}).
+FOr R+S OMs and holding variability in SSB constant, accuracy of
+stock-recruit model selection declined with greater variance of
+recruitment and survival process errors. Simultaneously estimating mean
+log-natural mortality has a small negative effect on AIC accuracy, but
+there is an improvement in AIC accuracy with lower observation error in
+indices and age composition and lower variation of survival process
+errors.
+
+\hypertarget{rm-operating-models-2}{%
+\subsubsection*{R+M operating models}\label{rm-operating-models-2}}
+\addcontentsline{toc}{subsubsection}{R+M operating models}
+
+For R+M OMs and EMs there appeared to be little effect of correlation in
 the natural mortality process errors or estimating the mean log-natural
@@
 \ref{M_om_MF_BH_glm_AIC_plots} and \ref{M_om_ME_BH_glm_AIC_plots}).
+However, there was an increase in accuracy with lower observation error
+for OMs with higher variability in natural mortality process errors
+whether the mean natural mortality rate parameter was estimated or not.
+
+\hypertarget{rsel-operating-models-2}{%
+\subsubsection*{R+Sel operating models}\label{rsel-operating-models-2}}
+\addcontentsline{toc}{subsubsection}{R+Sel operating models}
 
 The range of variation in SSB across simulated populations was lower for
-R+Sel operating models than R+S or R+M OMs. OMs with contrast in fishing
-pressure over time is the primary mechanism creating variation in SSB.
-However, there appears to be little effect of variability or correlation
-of selectivity process errors or whether mean log-natural mortality was
+R+Sel OMs than R+S or R+M OMs. OMs with contrast in fishing pressure
+over time is the primary mechanism creating variation in SSB. However,
+there appears to be little effect of variability or correlation of
+selectivity process errors or whether mean log-natural mortality was
 estimated on the accuracy of AIC in selecting the stock recruit
 relationship (Figures \ref{Sel_om_MF_BH_glm_AIC_plots} and
-\ref{Sel_om_ME_BH_glm_AIC_plots})
-
-Like the R+Sel OMs, the range of variation in SSB across is only
-increased by allowing contrast in fishing pressure over time. There
-appears to be a slight decrease in accuracy in AIC when mean log-natural
-mortality is estimated and there is a more noticeable increase in
-accuracy for OMs with lower observation error in those cases. (Figures
-\ref{q_om_MF_BH_glm_AIC_plots} and \ref{q_om_ME_BH_glm_AIC_plots})
+\ref{Sel_om_ME_BH_glm_AIC_plots}). However, there was an increase in AIC
+accuracy with lower observation error for OMs with independent process
+errors whether the mean natural mortality rate parameter was estimated
+or not.
+
+\hypertarget{rq-operating-models-2}{%
+\subsubsection*{R+q operating models}\label{rq-operating-models-2}}
+\addcontentsline{toc}{subsubsection}{R+q operating models}
+
+Like the R+Sel OMs, the range of variation in SSB across simulated
+populations for R+q OMs is only increased by allowing contrast in
+fishing pressure over time. There appears to be a slight decrease in
+accuracy in AIC when mean log-natural mortality is estimated and there
+is a more noticeable increase in accuracy for OMs with lower observation
+error in those cases. (Figures \ref{q_om_MF_BH_glm_AIC_plots} and
+\ref{q_om_ME_BH_glm_AIC_plots}). There was also an increase in AIC
+accuracy with lower observation error for when the mean natural
+mortality rate parameter was estimated.
 
@@
 
-\hypertarget{r-rs-operating-models-2}{%
-\subsubsection*{R, R+S operating models}\label{r-rs-operating-models-2}}
+\hypertarget{r-rs-operating-models-3}{%
+\subsubsection*{R, R+S operating models}\label{r-rs-operating-models-3}}
 \addcontentsline{toc}{subsubsection}{R, R+S operating models}
@@
 for R OMs with high observation error, constant fishing pressure and
-large variability in recruitment process errors
+large variability in recruitment process errors (Figure
 \ref{naa_om_em_BH_ME_relbias_ssb}).
 
-\hypertarget{rm-operating-models-2}{%
-\subsubsection*{R+M operating models}\label{rm-operating-models-2}}
+\hypertarget{rm-operating-models-3}{%
+\subsubsection*{R+M operating models}\label{rm-operating-models-3}}
 \addcontentsline{toc}{subsubsection}{R+M operating models}
@@
 discernible effect on SSB bias (Figures \ref{M_om_em_BH_MF_relbias_ssb}
-and \ref{M_om_em_BH_ME_relbias_ssb})
-
-\hypertarget{rsel-operating-models-2}{%
-\subsubsection*{R+Sel operating models}\label{rsel-operating-models-2}}
+and \ref{M_om_em_BH_ME_relbias_ssb}).
+
+\hypertarget{rsel-operating-models-3}{%
+\subsubsection*{R+Sel operating models}\label{rsel-operating-models-3}}
 \addcontentsline{toc}{subsubsection}{R+Sel operating models}
@@
 
-\hypertarget{rq-operating-models-1}{%
-\subsubsection*{R+q operating models}\label{rq-operating-models-1}}
+\hypertarget{rq-operating-models-3}{%
+\subsubsection*{R+q operating models}\label{rq-operating-models-3}}
 \addcontentsline{toc}{subsubsection}{R+q operating models}
@@
 
-\hypertarget{r-rs-operating-models-3}{%
-\subsubsection*{R, R+S operating models}\label{r-rs-operating-models-3}}
+\hypertarget{r-rs-operating-models-4}{%
+\subsubsection*{R, R+S operating models}\label{r-rs-operating-models-4}}
 \addcontentsline{toc}{subsubsection}{R, R+S operating models}
@@
 
-\hypertarget{rm-operating-models-3}{%
-\subsubsection*{R+M operating models}\label{rm-operating-models-3}}
+\hypertarget{rm-operating-models-4}{%
+\subsubsection*{R+M operating models}\label{rm-operating-models-4}}
 \addcontentsline{toc}{subsubsection}{R+M operating models}
@@
 
-\hypertarget{rsel-operating-models-3}{%
-\subsubsection*{R+Sel operating models}\label{rsel-operating-models-3}}
+\hypertarget{rsel-operating-models-4}{%
+\subsubsection*{R+Sel operating models}\label{rsel-operating-models-4}}
 \addcontentsline{toc}{subsubsection}{R+Sel operating models}
@@
 
-\hypertarget{rq-operating-models-2}{%
-\subsubsection*{R+q operating models}\label{rq-operating-models-2}}
+\hypertarget{rq-operating-models-4}{%
+\subsubsection*{R+q operating models}\label{rq-operating-models-4}}
 \addcontentsline{toc}{subsubsection}{R+q operating models}
@@
 
-Among the process error configurations we used in operating models, we
-found AIC to be accurate for process errors on recruitment and survival.
-Fitting models to other OMs rarely preferred R+S EMs and R and R+S EMs
-were nearly always selected for the matching OMs. For other sources of
+Among the process error configurations we used in OMs, we found AIC to
+be accurate for process errors on recruitment and survival. Fitting
+models to other OMs rarely preferred R+S EMs and R and R+S EMs were
+nearly always selected for the matching OMs. For other sources of
 process error, accuracy of AIC was improved to useful levels when there
@@
 tended to be more variable for recruitment and can be large even when
-the estimating model is correct. Therefore, we recommend emphasis on
-inspection of retrospective patterns primarily for SSB and F, but
-further research on retrospective patterns in other assessment model
-parameters and derived output may be beneficial.
+the EM is correct. Therefore, we recommend emphasis on inspection of
+retrospective patterns primarily for SSB and F, but further research on
+retrospective patterns in other assessment model parameters and derived
+output may be beneficial.
 

### 2024-02-07 Ecov_study/recruitment_functions/presentation/SRR_presentation.Rmd
presentation update
'done' up to mohn's rho

--- 2024-02-06 Ecov_study/recruitment_functions/presentation/SRR_presentation.Rmd
@@
   theme(axis.title.y = element_text(size = 16))  +
-  theme(axis.title.x = element_text(size = 16))
+  theme(axis.title.x = element_text(size = 16)) +
+  scale_x_continuous(breaks=fish.sel$Age, labels=as.character(fish.sel$Age))
 
@@
 
-- 
 - Results will be shown for $\beta$ unstandardized case, because $\beta$ standardization had no appreciable effect
+- Most results summarized by relative error (RE): (Est-True)/True
 
@@
 
+```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+ knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'bad.mods_lowestAIC.plot_grad_6_SE_100_beta_fix.png') )
+```
+
 ## Unconverged Models (considering all EMs per OM-Simulation)
+
+```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+ knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'bad.mods_all_grad_6_SE_100_beta_fix.plot.png') )
+```
 
@@
 
+```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+ knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'bad_par_N_barplot_grad_6_SE_100_beta_fix.png') )
+```
+
 ## Unconverged Model - Even More Offending Parameters
 
-## Model Identifiability
+```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+ knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'bad_par_2nd_N_barplot_grad_6_SE_100_beta_fix.png') )
+```
 
-##  $\Delta$AIC and model probability
+## Model Identifiability - Marginal barplots for SRR
+ - Marginal univariate model selection results for whether an SRR relationship was correctly
+identified (indicated by dark shaded regions)
 
-## Assessment error (recruitment, spawning stock biomass, and Fbar)
+```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+ knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'raw_boxplots_SR_beta_fix.pdf') )
+```
 
-## Bias of estimated parameters
+## Model Identifiability - Marginal barplots for Ecov_how
+ - Marginal univariate model selection results for whether an Ecov was correctly identified (indicated by dark shaded regions)
+
+```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+ knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'raw_boxplots_Ecov_beta_fix.pdf') )
+```
+
+## Model Identifiability - Marginal barplots for both SRR & Ecov_how
+ - Marginal univariate model selection results for whether the SRR AND Ecov were both correctly identified (indicated by dark shaded regions)
+
+```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+ knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'raw_boxplots_form_beta_fix.pdf') )
+```
+
+## Model Identifiability - Effect size
+ - Effect sizes for proportion correctly identified using a binomial generalized linear model and a
+binomial generalized linear mixed model with simulation random number seed as a random intercept
+
+```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+ knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'effect_size_glm_glmm_beta_fix.pdf') )
+```
+
+## Model Identifiability - Classification tree for identifying EM = OM
+ - Classification tree analysis of model selection results, with default settings in R package 'rpart'.
+The complexity parameter is set at cp = 0.01, which represents the minimum classification improvement required for any split.
+
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'reg_tree_AIC_best_beta_fix.pdf') )
+  ```
+
+##  $\Delta$AIC and model probability - Classification tree
+ - Classification tree analysis of for the difference in AIC among EMs, the rank of EMs, and the model probability of EMs (calculated from Akaike weights).
+
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'reg_tree_AIC_prob_beta_fix.pdf') )
+  ```
+
+##  $\Delta$AIC  - Median and 95% probability interval
+ - Median difference in AIC (filled circle) with 2.5th and 97.5th percentiles; horizontal line at 2 for reference
+
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'model_dAIC.ylim_beta_fix.png') )
+  ```
+
+##  Model probability  - Median and 95% probability interval
+ - Median model probability (filled circle) with 2.5th and 97.5th percentiles; horizontal line at 0.5 for reference
+
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'model_probability_beta_fix.png') )
+  ```
+
+## Assessment error - Recruitment (all years)
+- Median relative error (filled circle) with 2.5th and 97.5th percentiles
+
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 're.rec.all.yrs.plot_beta_fix.png') )
+  ```
+
+## Assessment error - Recruitment (last 10 years)
+- Median relative error (filled circle) with 2.5th and 97.5th percentiles
+
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 're.rec.last.10.plot_beta_fix.png') )
+  ```
+
+## Assessment error - Recruitment (terminal year)
+- Median relative error (filled circle) with 2.5th and 97.5th percentiles
+
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 're.rec.last.yr.plot_beta_fix.png') )
+  ```
+
+## Assessment error - Spawning Stock Biomass (all years)
+- Median relative error (filled circle) with 2.5th and 97.5th percentiles
+
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 're.ssb.all.yrs.plot_beta_fix.png') )
+  ```
+
+## Assessment error - Spawning Stock Biomass (terminal year)
+- Median relative error (filled circle) with 2.5th and 97.5th percentiles
+
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 're.ssb.last.yr.plot_beta_fix.png') )
+  ```
+
+## Assessment error - Fbar (all years)
+- Median relative error (filled circle) with 2.5th and 97.5th percentiles
+
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 're.fbar.all.yrs.plot_beta_fix.png') )
+  ```
+
+## Assessment error - Fbar (terminal year)
+- Median relative error (filled circle) with 2.5th and 97.5th percentiles
+
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 're.fbar.last.yr.plot_beta_fix.png') )
+  ```
+
+## Effect size for Relative error in Recruitment
+
+```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'effect_size_glm_glmm_recr_re_beta_fix.pdf') )
+```
+
+## Effect size for Relative error in SSB
+
+```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'effect_size_glm_glmm_ssb_re_beta_fix.pdf') )
+```
+
+ ## Effect size for Relative error in Fbar
+
+```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'effect_size_glm_glmm_fbar_re_beta_fix.pdf') )
+```
+
+# Alert
+## Is everybody still awake?
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'presentation', 'asleep_keyboard.jpg') )
+  ```
+
+# Results
+
+## Bias of estimated parameters - Ecov $\beta$
+- Median relative error (filled circle) with 2.5th and 97.5th percentiles
+
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'RE.Ecov_beta.plot_beta_fix.png') )
+  ```
+
+## Bias of estimated parameters - Ecov mean
+- Median relative error (filled circle) with 2.5th and 97.5th percentiles
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'RE.Ecov.mean.plot_beta_fix.png') )
+  ```
+
+## Bias of estimated parameters - $\rho_{Ecov}$
+- Median relative error (filled circle) with 2.5th and 97.5th percentiles
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'RE.Ecov.rho.plot.ylim_beta_fix.png') )
+  ```
+
+## Bias of estimated parameters - $\sigma_{Ecov}$
+- Median relative error (filled circle) with 2.5th and 97.5th percentiles
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'RE.Ecov_beta.plot_beta_fix.png') )
+  ```
+
+## Bias of estimated parameters - SRR a
+- Median relative error (filled circle) with 2.5th and 97.5th percentiles
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'RE.mean_rec1.plot.ylim_beta_fix.png') )
+  ```
+
+## Bias of estimated parameters - SRR b
+- Median relative error (filled circle) with 2.5th and 97.5th percentiles
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'RE.mean_rec2.plot.ylim_beta_fix.png') )
+  ```
+
+## Bias of estimated parameters - $\sigma_{R}$
+- Median relative error (filled circle) with 2.5th and 97.5th percentiles
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'RE.R_sigma.plot.ylim_beta_fix.png') )
+  ```
+
+## Bias of estimated parameters - $\rho_{R}$
+- Median relative error (filled circle) with 2.5th and 97.5th percentiles
+
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'RE.R_cor.plot_beta_fix.png') )
+  ```
 
@@
 
-## Projection performance relative to assumptions about the environmental covariate
+## Projections: Specifications and Ecov assumptions
+
+## Projection performance relative to assumptions about the Ecov
 
@@
 ## Take-aways
-- These are the take-aways (copy from WP)
+
+- Stock-recruitment model identification was poor, requiring unrealistic OM scenarios
+
+- EMs with only recruitment random effects generally performed similarly well as OM=EM, including for cases with an SR relationship and/or an ecov effect
+
+- In conclusion, we recommend recruitment random effects from a mean SRR as the default EM in WHAM
 
@@
 
-- This work could not have been completed without the use of Azure computing (NOAA) and MIT (... greg to fill in)
+- This work could not have been completed without support from the Microsoft Cooperative Research and Development Agreement (CRADA) and NOAA's National Cloud Program Office (OCIO), as well as MIT (...
+greg to fill in)
 - We thank other members of the SSRTWG for thoughtful comments during earlier discussions and presentations of this work

### 2024-02-07 Ecov_study/recruitment_functions/presentation/SRR_presentation.Rmd
nearly final presentation

@@
 
-## OM factors and simulated data examples
+## OM factors 
 ```{r,  echo=FALSE, message=FALSE, warnings=FALSE}
@@
 - Factors that did not vary were the observation and process error of the Ecov (bot fixed at 0.1)
+
+## OM factor for Recruitment variability (R_sig) relative to recent assessments
+```{r,  echo=FALSE, message=FALSE, warnings=FALSE}
+
+rsig.assmt <- read.csv(file=file.path(here::here(),"Ecov_study", "recruitment_functions", "tables", "sigmaR_assessment_estimates.csv"), header=TRUE)
+
+   knitr::kable(rsig.assmt )
+
+```
 
@@
 ```{r,  echo=FALSE, message=FALSE, warnings=FALSE}
+library(tidyverse)
 
 df.ems    <- readRDS(file.path(here::here(),"Ecov_study", "recruitment_functions", "inputs", "df.ems.RDS"))
+df.ems2 <- cbind(EM=seq(1,6), df.ems)
 
@@
 - ~3.3-3.5% crashes, typically for Ecov=Masking ("BH_4") and $\sigma_{R}$=0.1
+- most problematic for Masking (BH_4) and exacerbated by $\sigma_{R}$=0.1
 ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
@@
 ## Unconverged Models (from EM having'best' EM per OM-Simulation)
-
+- most problematic for $\sigma_{R}$=0.1; BH with no Ecov and high fishing contrast (H-MSY) and mean SRR with no contrast (MSY)
 ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
@@
 ## Unconverged Models (considering all EMs per OM-Simulation)
+- most problematic for Beverton-Holt SRR, worse with no contrast in fishing history (MSY)
 
@@
 ## Unconverged Model - The Offending Parameters 
+- the SRR recruitment parameters led to most convergence failures
 
@@
 
-## Unconverged Model - Even More Offending Parameters
+## Unconverged Model - Second Most Offending Parameters
+
+- the Ecov parameters led to the second most convergence failures
 
@@
 
+## Results layout (example)
+- Plots show median relative error (filled circle) with 2.5th and 97.5th percentiles
+- OM factors as facets (top and right side)
+- EM models on the x-axis
+- Blue indicates EM=OM (diagonal of first 4 columns w/in facet blocks), incorrect EMs in red
+
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='65%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 're.rec.all.yrs.plot_beta_fix.png') )
+  ```
+
 ##  $\Delta$AIC  - Median and 95% probability interval
- - Median difference in AIC (filled circle) with 2.5th and 97.5th percentiles; horizontal line at 2 for reference
+- Smaller $\Delta$AIC when $\sigma_{R}$ is larger
 
@@
 ##  Model probability  - Median and 95% probability interval
- - Median model probability (filled circle) with 2.5th and 97.5th percentiles; horizontal line at 0.5 for reference
+- True OMs receive more probability with $\sigma_{R}$=0.1 and high fishing contrast; otherwise, mean SRR received more probability
 
@@
 ## Assessment error - Recruitment (all years)
-- Median relative error (filled circle) with 2.5th and 97.5th percentiles
+- Median unbiased, $\sigma_{R}$ determined quantile range
 
@@
 ## Assessment error - Recruitment (last 10 years)
-- Median relative error (filled circle) with 2.5th and 97.5th percentiles
+- Median unbiased, $\sigma_{R}$ determined quantile range
 
@@
 ## Assessment error - Recruitment (terminal year)
-- Median relative error (filled circle) with 2.5th and 97.5th percentiles
+- Median unbiased, $\sigma_{R}$ determined quantile range
 
@@
 ## Assessment error - Spawning Stock Biomass (all years)
-- Median relative error (filled circle) with 2.5th and 97.5th percentiles
+- Median unbiased, $\sigma_{R}$ determined quantile range
 
@@
 ## Assessment error - Spawning Stock Biomass (terminal year)
-- Median relative error (filled circle) with 2.5th and 97.5th percentiles
+- Median unbiased, $\sigma_{R}$ determined quantile range
 
@@
 ## Assessment error - Fbar (all years)
-- Median relative error (filled circle) with 2.5th and 97.5th percentiles
+- Median unbiased, $\sigma_{R}$ determined quantile range
 
@@
 ## Assessment error - Fbar (terminal year)
-- Median relative error (filled circle) with 2.5th and 97.5th percentiles
+- Median unbiased, $\sigma_{R}$ determined quantile range
 
@@
 ## Bias of estimated parameters - Ecov $\beta$
-- Median relative error (filled circle) with 2.5th and 97.5th percentiles
+
+- Limiting and Masking had worst performance (BH_3 and BH_4)
 
@@
 ## Bias of estimated parameters - Ecov mean
-- Median relative error (filled circle) with 2.5th and 97.5th percentiles
+- Larger quantile range when Ecov correlation is high (0.8)
  ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
@@
 ## Bias of estimated parameters - $\rho_{Ecov}$
-- Median relative error (filled circle) with 2.5th and 97.5th percentiles
+- Well estimated when Ecov correlation was high (0.8)
  ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
@@
 ## Bias of estimated parameters - $\sigma_{Ecov}$
-- Median relative error (filled circle) with 2.5th and 97.5th percentiles
+- Positively biased when Ecov correlation is high (0.8)
  ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
-  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'RE.Ecov_beta.plot_beta_fix.png') )
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'RE.Ecov.sigma.plot_beta_fix.png') )
   ```
@@
 ## Bias of estimated parameters - SRR a
-- Median relative error (filled circle) with 2.5th and 97.5th percentiles
+- Unbiased when $\sigma_{R}$ =0.1 and high fishing contrast; otherwise biased +/- 0.2 (BH_4 could be much worse)
  ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
@@
 ## Bias of estimated parameters - SRR b
-- Median relative error (filled circle) with 2.5th and 97.5th percentiles
+- Unbiased when $\sigma_{R}$ =0.1 and high fishing contrast; otherwise biased +/- 0.2 (BH_4 could be much worse)
+
  ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
@@
 ## Bias of estimated parameters - $\sigma_{R}$
-- Median relative error (filled circle) with 2.5th and 97.5th percentiles
+- Bias pattern mirroring between $\sigma_{R}$ and correlation
  ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
@@
 ## Bias of estimated parameters - $\rho_{R}$
-- Median relative error (filled circle) with 2.5th and 97.5th percentiles
-
+- Bias pattern mirroring between $\sigma_{R}$ and correlation
+- For most informative case (H-MSY, $\sigma_{R}$=0.1, low recruitment correlation), mean SRR has large positive bias
  ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
@@
 
-## Mohn's $\rho$
+## Mohn's $\rho$ - Recruitment (7 year peel)  
+- Missing symbol indicates failed peels for one or more EM fits
+- $\sigma_{R}$ and observation error were influential factors
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'recr.rho_beta_fix.png') )
+  ```
+
+## Mohn's $\rho$ - SSB  (7 year peel)   
+- observation error was influential factor
+
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'ssb.rho_beta_fix.png') )
+  ```
+
+## Mohn's $\rho$ - Fbar  (7 year peel) 
+- observation error was influential factor
+
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'fbar.rho_beta_fix.png') )
+  ```
+
+## Mohn's $\rho$ - Classification Tree for Recruitment, SSB, Fbar
+-
+
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'reg_tree_Rho_beta_fix.pdf') )
+  ```
+
+## Mohn's $\rho$ - Random Effects on Recruitment (7 year peel)
+- Negative bias when $\sigma_{R}$ is low
+- Expected worse $\rho$ from incorrect EM, but was not the case
+
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'recr.dev.rho_beta_fix.png') )
+  ```
+
+## Mohn's $\rho$ - Classification Tree for Random Effects on Recruitment
+
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'reg_tree_Rdev_retro_beta_fix.pdf') )
+  ```
 
 ## Projections: Specifications and Ecov assumptions 
+- Projections were made from the 10th peel of each fitted model (Terminal year - 10)
+- A 5-year average of maturity, weight at age, and selectivity specified (time-invariant anyway)
+- F in the projection was set at the true OM F value
+- The environmental index was treated 3 ways
 
-## Projection performance relative to assumptions about the Ecov
+    \begin{enumerate}
+      \item A 5-year average of the most recent index value was held constant in the 10 year projection ('Avg5yr')
+      \item The estimated process continued in the 10 year projection ('Continue')
+      \item The observed environmental index was specified in the 10 year projection ('ObsEcov') 
+    \end{enumerate}
+
+- Projected Recruitment, SSB, and Catch were compared to the true OM values in the projection
+
+## Projection performance - Relative Error Recruitment (first 3 years)
+- Minimal bias (median at most -0.1 for mean SRR)
+- Quantile range much larger for BH
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 're.rec.proj.plot_beta_fix.png') )
+    ```
+
+## Projection performance - Relative Error Recruitment (all 10 years)
+- Minimal bias (median at most -0.1 for mean SRR)
+- Quantile range much larger for BH
+
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 're.rec.proj.plot_beta_fix.png') )
+  ```
+
+## Projection performance - Recruitment Uncertainty (first 3 years)
+- CV slightly lower for mean SRR, usually narrower quantile range
+
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'sd.rec.proj.plot3yr_beta_fix.png') )
+    ```
+
+## Projection performance - Recruitment Uncertainty (all 10 years)
+- CV slightly lower for mean SRR, usually narrower quantile range
+
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'sd.rec.proj.plot_beta_fix.png') )
+  ```
+
+## Projection performance - Relative Error SSB (first 3 years)
+- Minimal bias (median at most -0.1 for mean SRR)
+- Quantile range much larger for BH
+
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 're.ssb.proj.plot_beta_fix.png') )
+    ```
+
+## Projection performance - Relative Error SSB (all 10 years)
+- Minimal bias (median at most -0.1 for mean SRR)
+- Quantile range much larger for BH
+
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 're.ssb.proj.plot_beta_fix.png') )
+  ```
+
+## Projection performance - SSB Uncertainty (first 3 years)
+- CV slightly lower for mean SRR, usually narrower quantile range
+
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'sd.ssb.proj.plot3yr_beta_fix.png') )
+    ```
+
+## Projection performance - SSB Uncertainty (all 10 years)
+- CV slightly lower for mean SRR, usually narrower quantile range
+
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'sd.ssb.proj.plot_beta_fix.png') )
+  ```
+
+## Projection performance - Relative Error Catch (first 3 years)
+- Minimal bias (median at most -0.1 for mean SRR)
+- Quantile range much larger for BH
+
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 're.catch.proj.plot_beta_fix.png') )
+    ```
+
+## Projection performance - Relative Error Catch (all 10 years)
+- Minimal bias (median at most -0.1 for mean SRR)
+- Quantile range much larger for BH
+
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 're.catch.proj.plot_beta_fix.png') )
+  ```
+
+## Projection performance - Catch Uncertainty (first 3 years)
+- CV slightly lower for mean SRR, usually narrower quantile range
+
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'sd.catch.proj.plot3yr_beta_fix.png') )
+    ```
+
+## Projection performance - Catch Uncertainty (all 10 years)
+- CV slightly lower for mean SRR, usually narrower quantile range
+
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'sd.catch.proj.plot_beta_fix.png') )
+  ```
+
+## Projection Summary
+- Minimal difference in median bias
+- Recruitment is highly uncertain in projections, exacerbated by length of projection
+- SSB and Catch also show increased uncertainty in longer projections
+- Long projections are expected to be very imprecise, recommend shorter projections
+- Tighter quantile range and slightly lower CV for mean SRR
 
@@
 
-- EMs with only recruitment random effects generally performed similarly well as OM=EM, including for cases with an SR relationship and/or an ecov effect
+- EMs with mean SRR and recruitment random effects generally performed similarly well as OM=EM, including for cases with an SR relationship and/or an ecov effect
 
@@
 
-- This is what we suggest for follow-up
-- Note that WP1-Appendix looked at $\sigma_{R}$=0.5 and found no difference from results in WP1
+- In WP-1, we recommended exploring additional levels of $\sigma_{R}$=0.5; however, we note that WP1-Appendix looked at $\sigma_{R}$=0.5 and found no difference
 

### 2024-02-07 Ecov_study/recruitment_functions/presentation/SRR_presentation.Rmd
almost nearly final presentation

@@
 
-# Background
+# Intro
 
-## Introduction
+## Background
 
@@
 
+# Sims
+## Simulated Data Examples - Type 1
+- plots
+
+## Simulated Data Examples - Type 2
+- plots
+
+## Simulated Data Examples - Type 3
+- plots
+
+<!-- ==================================================================================== -->
+
 # EM
@@
 
-```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='60%', fig.align='center'}
  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'effect_size_glm_glmm_beta_fix.pdf') )
@@
 
- ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='65%', fig.align='center'}
+ ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='60%', fig.align='center'}
   knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 're.rec.all.yrs.plot_beta_fix.png') )
@@
 ## Assessment error - Spawning Stock Biomass (all years)
-- Median unbiased, $\sigma_{R}$ determined quantile range
+- Median unbiased
 
@@
 ## Assessment error - Fbar (all years)
-- Median unbiased, $\sigma_{R}$ determined quantile range
+- Median unbiased
 
@@
 
-# Alert
+# !!
 ## Is everybody still awake?
@@
 
-- Limiting and Masking had worst performance (BH_3 and BH_4)
+- Huge quartile ranges; Limiting and Masking had worst performance (BH_3 and BH_4)
 
@@
 - Projections were made from the 10th peel of each fitted model (Terminal year - 10)
-- A 5-year average of maturity, weight at age, and selectivity specified (time-invariant anyway)
+- A 5-year average of maturity, weight at age, and selectivity specified (time-invariant)
 - F in the projection was set at the true OM F value
@@
 
-- In conclusion, we recommend recruitment random effects from a mean SRR as the default EM in WHAM
+- In conclusion, we recommend recruitment random effects (ar1) from a mean SRR (recruit_model=2) as the default EM in WHAM
 
@@
 
-## Recommendations for further analyses
+## Further analyses (already investigated, see Appendix for WP-1)
 
-- In WP-1, we recommended exploring additional levels of $\sigma_{R}$=0.5; however, we note that WP1-Appendix looked at $\sigma_{R}$=0.5 and found no difference
+- In WP-1, we recommended exploring additional levels of $\sigma_{R}$=0.5; however, we note that WP1-Appendix looked at $\sigma_{R}$=0.5 and found no difference (only success at unrealistic $\sigma_{R}$=0.1)
+
+```{r, echo=FALSE, message=FALSE, warnings=FALSE, out.height='40%', fig.align='center'}
+ knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'reg_tree_AIC_best_beta_fix.pdf') )
+```
+
+```{r, echo=FALSE, message=FALSE, warnings=FALSE, out.height='40%', fig.align='center'}
+ knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix_384', 'reg_tree_AIC_best_beta_fix_384.pdf') )
+```
+
+## Recommendations for future research
+- Additional fishing histories, including some that include years with F < $F_{MSY}$
+- Recruitment random effect trends
+- Ecov models with trends or AR(2) dynamics
+- Multivariate ecov processes
+- Ecov processes affecting multiple stock parameters, or multiple Ecov processes affecting recruitment (simultaneous or sequential)
 
@@
 
-- This work could not have been completed without support from the Microsoft Cooperative Research and Development Agreement (CRADA) and NOAA's National Cloud Program Office (OCIO), as well as MIT (...
-greg to fill in)
+- This work could not have been completed without access to high performance computing resources from the Microsoft Cooperative Research and Development Agreement (CRADA) and NOAA's National Cloud Program Office (OCIO), and the MIT Office of Research Computing and Data
 - We thank other members of the SSRTWG for thoughtful comments during earlier discussions and presentations of this work

### 2024-02-08 Ecov_study/recruitment_functions/presentation/SRR_presentation.Rmd
presentation done?
presentation update

@@
 
-## Some Example OM Simulated Data (100 data sets for each OM)
-- still need to make this
-
 <!-- ==================================================================================== -->
@@
 ## Simulated Data Examples - Type 1
-- plots
+```{r, echo=FALSE, message=FALSE, warnings=FALSE, out.height='40%', fig.align='center'}
+ knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'presentation', 'recruit_ecov.pdf') )
+```
 
-## Simulated Data Examples - Type 2
-- plots
+```{r, echo=FALSE, message=FALSE, warnings=FALSE, out.height='50%', fig.align='center'}
+ knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'presentation', 'stock_recruit_no_ecov.pdf') )
+```
 
-## Simulated Data Examples - Type 3
-- plots
+## Simulated Data Examples - Recruitment Timeseries
+```{r, echo=FALSE, message=FALSE, warnings=FALSE, out.width='90%', fig.align='center'}
+ knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'presentation', 'Rec_sims.png') )
+```
+
+<!-- ```{r, echo=FALSE, message=FALSE, warnings=FALSE, out.width='40%', fig.align='center'} -->
+<!--  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'presentation', 'om_57.59.61.63.121.123.125.127.185.187.189.191.249.251.253.255._Recr.plot.png') ) -->
+<!-- ``` -->
+
+## Simulated Data Examples - SSB Timeseries
+```{r, echo=FALSE, message=FALSE, warnings=FALSE, out.width='90%', fig.align='center'}
+ knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'presentation', 'SSB_sims.png') )
+```
+
+<!-- ```{r, echo=FALSE, message=FALSE, warnings=FALSE, out.width='40%', fig.align='center'} -->
+<!--  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'presentation', 'om_57.59.61.63.121.123.125.127.185.187.189.191.249.251.253.255._SSB.png') ) -->
+<!-- ``` -->
 

### 2024-02-08 Ecov_study/recruitment_functions/presentation/SRR_presentation.Rmd
final presentation?

@@
 # Sims
-## Simulated Data Examples - Type 1
+## Simulated Data Examples - Ecov and SRR 
 ```{r, echo=FALSE, message=FALSE, warnings=FALSE, out.height='40%', fig.align='center'}
@@
 
-<!-- ```{r, echo=FALSE, message=FALSE, warnings=FALSE, out.width='40%', fig.align='center'} -->
-<!--  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'presentation', 'om_57.59.61.63.121.123.125.127.185.187.189.191.249.251.253.255._Recr.plot.png') ) -->
-<!-- ``` -->
-
 ## Simulated Data Examples - SSB Timeseries
@@
 
-<!-- ```{r, echo=FALSE, message=FALSE, warnings=FALSE, out.width='40%', fig.align='center'} -->
-<!--  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'presentation', 'om_57.59.61.63.121.123.125.127.185.187.189.191.249.251.253.255._SSB.png') ) -->
-<!-- ``` -->
+## Simulated Data Examples - Catch Timeseries
+```{r, echo=FALSE, message=FALSE, warnings=FALSE, out.width='90%', fig.align='center'}
+ knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'presentation', 'Catch_sims.png') )
+```
 
@@
 
-```{r, echo=FALSE, message=FALSE, warnings=FALSE, out.height='40%', fig.align='center'}
+```{r, echo=FALSE, message=FALSE, warnings=FALSE, out.height='35%', fig.align='center'}
  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'reg_tree_AIC_best_beta_fix.pdf') )
@@
 
-```{r, echo=FALSE, message=FALSE, warnings=FALSE, out.height='40%', fig.align='center'}
+```{r, echo=FALSE, message=FALSE, warnings=FALSE, out.height='35%', fig.align='center'}
  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix_384', 'reg_tree_AIC_best_beta_fix_384.pdf') )
@@
 - Recruitment random effect trends
-- Ecov models with trends or AR(2) dynamics
+- Ecov models with stronger effects, trends, or AR(2) dynamics
 - Multivariate ecov processes

### 2024-02-08 Ecov_study/recruitment_functions/presentation/SRR_presentation.Rmd
cleanup analysis code
fixed obs_error labels (H L reversed in factor statement)

@@
 # Sims
-## Simulated Data Examples - Ecov and SRR 
-```{r, echo=FALSE, message=FALSE, warnings=FALSE, out.height='40%', fig.align='center'}
- knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'presentation', 'recruit_ecov.pdf') )
+## Simulated Data Examples - Ecov and SRR  recruit_ecov_SRR_noecov
+
+```{r, echo=FALSE, message=FALSE, warnings=FALSE, out.width='95%', fig.align='center'}
+ knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'presentation', 'recruit_ecov_SRR_noecov.png') )
 ```
 
-```{r, echo=FALSE, message=FALSE, warnings=FALSE, out.height='50%', fig.align='center'}
- knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'presentation', 'stock_recruit_no_ecov.pdf') )
-```
+ <!-- ```{r, echo=FALSE, message=FALSE, warnings=FALSE, out.height='40%', fig.align='center'} -->
+ <!-- knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'presentation', 'recruit_ecov.pdf') ) -->
+<!-- ``` -->
+
+<!--  ```{r, echo=FALSE, message=FALSE, warnings=FALSE, out.height='50%', fig.align='center'} -->
+<!--  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'presentation', 'stock_recruit_no_ecov.pdf') ) -->
+<!-- ``` -->
 
@@
 - EM models on the x-axis
-- Blue indicates EM=OM (diagonal of first 4 columns w/in facet blocks), incorrect EMs in red
+- Blue indicates EM=OM (diagonal of first 4 columns w/in facet blocks)
+- Incorrect EMs in red
 
@@
 ## Projection Summary
-- Minimal difference in median bias
+- Minimal difference in median bias (median at most -0.1 for mean SRR)
 - Recruitment is highly uncertain in projections, exacerbated by length of projection
@@
 - Ecov models with stronger effects, trends, or AR(2) dynamics
-- Multivariate ecov processes
+- Multivariate Ecov processes
 - Ecov processes affecting multiple stock parameters, or multiple Ecov processes affecting recruitment (simultaneous or sequential)
@@
 
-```{r, echo=FALSE, message=FALSE, warnings=FALSE, out.width='35%', fig.align='center'}
+```{r, echo=FALSE, message=FALSE, warnings=FALSE, out.height='50%', fig.align='center'}
  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'presentation', 'abacus_crop.jpg') )

### 2024-02-08 Ecov_study/recruitment_functions/code/analysis/compare_RE_SD_projections.R

@@ -336,7 +336,7 @@ re.rec.tib$Ecov_how    <- factor(re.rec.tib$Ecov_how,labels=c("Ecov_0", "Ecov_1"
 re.rec.tib$NAA_cor     <- factor(re.rec.tib$NAA_cor,labels=c("Rcor_L","Rcor_H"))
 re.rec.tib$Fhist       <- factor(re.rec.tib$Fhist,labels=c("H-MSY","MSY") ) 
 re.rec.tib$Ecov_re_cor <- factor(re.rec.tib$Ecov_re_cor,labels=c("EcovCor_L","EcovCor_H"))
-re.rec.tib$obs_error   <- factor(re.rec.tib$obs_error,labels=c("ObsErr_L","ObsErr_H"))
+re.rec.tib$obs_error   <- factor(re.rec.tib$obs_error,labels=c("ObsErr_H","ObsErr_L"))
 
 
 ## log_sd (sd)
@@ -355,7 +355,7 @@ sd.rec.tib$Ecov_how    <- factor(sd.rec.tib$Ecov_how,labels=c("Ecov_0", "Ecov_1"
 sd.rec.tib$NAA_cor     <- factor(sd.rec.tib$NAA_cor,labels=c("Rcor_L","Rcor_H"))
 sd.rec.tib$Fhist       <- factor(sd.rec.tib$Fhist,labels=c("H-MSY","MSY") ) 
 sd.rec.tib$Ecov_re_cor <- factor(sd.rec.tib$Ecov_re_cor,labels=c("EcovCor_L","EcovCor_H"))
-sd.rec.tib$obs_error   <- factor(sd.rec.tib$obs_error,labels=c("ObsErr_L","ObsErr_H"))
+sd.rec.tib$obs_error   <- factor(sd.rec.tib$obs_error,labels=c("ObsErr_H","ObsErr_L"))
 
 
 # %%%%%%%%%%%%%%%%%%%%%%
@@ -375,7 +375,7 @@ re.ssb.tib$Ecov_how    <- factor(re.ssb.tib$Ecov_how,labels=c("Ecov_0", "Ecov_1"
 re.ssb.tib$NAA_cor     <- factor(re.ssb.tib$NAA_cor,labels=c("Rcor_L","Rcor_H"))
 re.ssb.tib$Fhist       <- factor(re.ssb.tib$Fhist,labels=c("H-MSY","MSY") ) 
[…]

### 2024-02-12 Ecov_study/recruitment_functions/presentation/SRR_presentation.Rmd
inserted missing slide

@@
 ##  $\Delta$AIC  - Median and 95% probability interval
-- Smaller $\Delta$AIC when $\sigma_{R}$ is larger
+- Smaller $\Delta$AIC for BH with high fishing contrast, low $\sigma_{R}$ (but not necessarily correct BH); $\Delta$AIC is larger for mean SRR in these cases
+- When $\sigma_{R}$ is larger and/or there is low contrast in fishing, the mean SRR is favored
 

### 2024-02-13 Ecov_study/recruitment_functions/presentation/SRR_presentation.Rmd
added 4 new slides

@@
 - Most results summarized by relative error (RE): (Est-True)/True
+
+## Distribution of maximum(absolute value of gradient) (*new slide)
+- very few models dropped for this criterion
+- reference lines for -6 (solid) and 5,4,3,2 (exponent for power of 10)
+```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+ knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'bad.grad.dist_lowestAIC.plot_grad_6_SE_100_beta_fix.png') )
+```
+
+## Distribution of SE (*new slide)
+- very few models dropped for this criterion
+- mostly impacted mean SRR with no Ecov
+```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+ knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'bad.se.dist_lowestAIC.plot_grad_6_SE_100_beta_fix.png') )
+```
 
@@
 
+##  Ability to separate sources of process error (*new slide)
+- Better performance when $\rho_{R}$ and $\rho_{Ecov}$ are similar
+- No effect of Fhist, $\sigma_{R}$ or Ecov form
+```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'process.error.ratio.est.plot_beta_fix.png') )
+```
+
+##  Ability to separate sources of process error (*new slide)
+- Better performance when $\rho_{R}$ and $\rho_{Ecov}$ are similar
+- No effect of Fhist, $\sigma_{R}$ or Ecov form
+```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'process.error.ratio.est.plot.ylim_beta_fix.png') )
+```
+
 ## Mohn's $\rho$ - Recruitment (7 year peel)  

### 2024-02-14 Ecov_study/recruitment_functions/presentation/SRR_presentation.Rmd
correcting slides w/ 3yr projections
i had 10 yr projection for both 3yr and 10yr by mistake

@@
 ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
-  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 're.rec.proj.plot_beta_fix.png') )
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 're.rec.proj.plot3yr_beta_fix.png') )
 ```
@@
 ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
-  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 're.ssb.proj.plot_beta_fix.png') )
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 're.ssb.proj.plot3yr_beta_fix.png') )
 ```
@@
 ```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
-  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 're.catch.proj.plot_beta_fix.png') )
+  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 're.catch.proj.plot3yr_beta_fix.png') )
 ```

### 2024-02-14 Ecov_study/recruitment_functions/presentation/SRR_presentation_Extras.Rmd
correcting slides w/ 3yr projections
i had 10 yr projection for both 3yr and 10yr by mistake

---
title: "Identification and performance of stock-recruitment functions in state
space assessment models - Extras"
subtitle: Working Paper 1
author: Greg Britten, Liz Brooks, Tim Miller
output: beamer_presentation
fig_caption: FALSE
theme: "Singapore"
slide-level: 3
# titlegraphic: WHOI_NOAA_logo_smaller.png
header-includes: 
- \addtobeamertemplate{title page}{\centering \includegraphics[width=9.5cm]{WHOI_NOAA_logo_smaller.png}\par }{}
classoption: "aspectratio=169"
date: "`r Sys.Date()`"
---

# Extra Analyses - h, R0

## Steepness (h) estimates

- Estimates correspond to single ('static') a,b in the BH and 'static' replacement line 
- Only unbiased with high fishing contrast and low $\sigma_{R}$
```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
 knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix_extra', 'steepness.vio.plot_beta_fix.png') )
```

## Unexploited recruitment (R0) estimates

- Estimates correspond to single ('static') a,b in the BH and 'static' replacement line  
- Only unbiased at low $\sigma_{R}$; $\rho_{R}$ increases range of estimates
```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
 knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix_extra', 'R0.vio.plot_beta_fix.png') )
```

## Unexploited SSB (SSB0) estimates

- Estimates correspond to single ('static') a,b in the BH and 'static' replacement line  
- Only unbiased at low $\sigma_{R}$; $\rho_{R}$ increases range of estimates

```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
 knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix_extra', 'SSB0.vio.plot_beta_fix.png') )
```

<!-- ==================================================================================== -->

# Extra Analyses - MSY

## $SPR_{MSY}$ 
- Median unbiased at low $\sigma_{R}$ and high contrast in fishing history
```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
 knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix_extra', 'SPRmsy.vio.plot_beta_fix.png') )
```

## Relative Error in $SPR_{MSY}$ 
- Median unbiased at low $\sigma_{R}$ and high contrast in fishing history
```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
 knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix_extra', 'RE.SPRmsy.vioplot_beta_fix.png') )
```

## $F_{MSY}$ 
- Median unbiased at low $\sigma_{R}$ and high contrast in fishing history
```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
 knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix_extra', 'Fmsy.vio.plot_beta_fix.png') )
```

## Relative Error in $F_{MSY}$ 
- Median unbiased at low $\sigma_{R}$ and high contrast in fishing history
```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
 knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix_extra', 'RE.Fmsy.vioplot_beta_fix.png') )
```

## $SSB_{MSY}$ 
- Median unbiased at low $\sigma_{R}$

```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
 knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix_extra', 'SSBmsy.vio.plot_beta_fix.png') )
```

## Relative Error in $SSB_{MSY}$ 
- Median unbiased at low $\sigma_{R}$

```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
 knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix_extra', 'RE.SSBmsy.vioplot_beta_fix.png') )
```

## $MSY$ 
- Median unbiased at low $\sigma_{R}$ 
```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
 knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix_extra', 'MSY.vio.plot.ylim_beta_fix.png') )
```

## Relative Error in $MSY$ 
- Median unbiased at low $\sigma_{R}$ 
```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
 knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix_extra', 'RE.MSY.vioplot_beta_fix.png') )
```

<!-- ==================================================================================== -->

# Extra Analyses - $\sigma_{R}$

## $\sigma_{R}$ Estimates
- Estimates not really larger for mean SRR; less bias with low $\rho_{R}$

```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
 knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'Rsigma.plot_beta_fix.png') )
```

## $\sigma_{R}$ Estimates
- Estimates not really larger for mean SRR; less bias with low $\rho_{R}$

```{r,  echo=FALSE, message=FALSE, warnings=FALSE,  out.height='75%', fig.align='center'}
 knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'Rsigma.plot.ylim_beta_fix.png') )
```
<!-- ==================================================================================== -->

---

Acknowledgements

- This work could not have been completed without access to high performance computing resources from the Microsoft Cooperative Research and Development Agreement (CRADA) and NOAA's National Cloud Program Office (OCIO), and the MIT Office of Research Computing and Data
- We thank other members of the SSRTWG for thoughtful comments during earlier discussions and presentations of this work

```{r, echo=FALSE, message=FALSE, warnings=FALSE, out.height='50%', fig.align='center'}
 knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'presentation', 'abacus_crop.jpg') )
```

<!-- ==================================================================================== -->

<!-- ## R Markdown -->

<!-- If you have trouble knitting this file, check out this page: https://stackoverflow.com/questions/67696286/error-generating-pdf-using-knitr-to-pdf-in-rstudio -->

<!-- <!-- -- example chunk to include figure from file --> -->
<!-- ```{r, echo=FALSE, message=FALSE, warnings=FALSE, out.width='50%', fig.align='center'} -->
<!--  knitr::include_graphics(file.path(here::here(), 'Ecov_study','recruitment_functions', 'plots_beta_fix', 'model_probability_beta_fix.png') ) -->
<!-- ``` -->

### 2025-01-16 Project_0/manuscript/manuscript.Rmd

--- 2024-01-26 Project_0/paper/paper.Rmd
@@
 author: 
-  - Timothy J. Miller^1^
-  - Andrew Applegate 
-  - Greg Britten
-  - Elizabeth N. Brooks
-  - Gavin Fay
-  - Alex Hansell
-  - Christopher M. Legault
-  - Brandon Muffley
-  - Brian C. Stock
-  - John Wiedenmann
-date: ''
+  - Timothy J. Miller^1,2^
+  - Greg Britten^3^
+  - Elizabeth N. Brooks^2^
+  - Gavin Fay^4^
+  - Alex Hansell^2^
+  - Christopher M. Legault^2^
+  - Brandon Muffley^5^
+  - Brian C. Stock^6^
+  - John Wiedenmann^7^
+date: "`r format(Sys.time(), '%d %B, %Y')`"
 fontsize: 12pt
@@
   pdf_document:
-    citation_package: natbib
+#    citation_package: natbib
     fig_caption: yes
@@
     df_print: paged
-  word_document: default
+#  word_document: default
 header-includes:
-- \usepackage{xspace}
-- \usepackage{bm}
-- \usepackage{caption,graphics}
-- \usepackage{graphicx}
-- \usepackage{makecell}
-- \renewcommand\figurename{Fig.}
-- \captionsetup{labelsep=period, singlelinecheck=false}
-- \newcommand{\changesize}[1]{\fontsize{#1pt}{#1pt}\selectfont}
-- \renewcommand{\arraystretch}{1.5}
-- \renewcommand\theadfont{}
+#- \usepackage{makecell}
 csl: canadian-journal-of-fisheries-and-aquatic-sciences.csl
-biblio-style: cjfas2.bst
+bibliography: manuscript.bib
+#biblio-style: cjfas2.bst
 ---
@@
 knitr::opts_chunk$set(echo = TRUE, tidy=FALSE, tidy.opts=list(width.cutoff=60), warning = FALSE, message = FALSE, fig.align = "center")
-knitr::opts_knit$set(root.dir = '../' )
+#knitr::opts_knit$set(root.dir = '../' )
 library(knitr)
@@
 library(forcats)
+type <- "latex"
+if(knitr::is_latex_output()) type <- "latex"
+if(knitr::is_html_output()) type <- "html"
 ```
 
-$^1$corresponding author: timothy.j.miller@noaa.gov, Northeast Fisheries Science Center, Woods Hole Laboratory, 166 Water Street, Woods Hole, MA 02543 USA\
+$^1$corresponding author: timothy.j.miller@noaa.gov\
+$^2$Northeast Fisheries Science Center, Woods Hole Laboratory, 166 Water Street, Woods Hole, MA 02543 USA\
+$^3$Woods Hole Oceanographic Institution\
+$^4$SMAST\
+$^5$Institute of Marine Research\
+$^6$Mid-Atlantic Fisheries Management Council\
+$^7$Rutgers University\
 
@@
 
-## Summary {-}
-
-We conducted a simulation study where we simulated process errors and observations for 72 operating models with alternative assumptions about fishing history, degree of uncertainty in index and age composition observations, type (recruitment, survival, fishery selectivity, catchability and natural mortality), degree of variation, and correlation of process errors.
-We fit 20 different estimation models to each of 100 simulated set of observations with alternative assumptions type and correlation structure of process errors, (mean) natural mortality was known or estimated, and a B-H stock recruit relationship was assumed or not.
-
-Across simulations we summarized probability of convergence of fitted models, accuracy of marginal AIC in determining the correct process error assumption and it ability to determine the Beverton-Holt stock recruit relationship, bias in annual spawning stock biomass, in estimation of natural mortality, and in stock-recruit relationship parameters, and severity of retrospective patterns for estimation models.
-
-Alternative measures of convergence performed differently.
-Invertible hessians and resulting standard error estimation was possible when criteria based on the gradient of the optimized log-likelihood with respect to the fixed effects parameters.
-Using hessian-based convergence, probability of convergence was best for models that assumed the correct source of process error, assumed M was known, and did not assume stock-recruit relationships.
-
-Using marginal AIC provided most accurate inferences about the process errors on recruitment and survival and selectivity, in that AIC preferred EMs with assumed process errors that matched OMs, and EMs with those assumed process errors were not preferred when alternative process errors were assumed in the OMs.
-However, when the true process errors were more variable, AIC accuracy increased to a useful level.
-We found AIC more accurately determined a B-H stock recruit relationship rather than the null model without a S-R relationship when there was low variability in recruitment, low variability in survival random effects, and higher variation in spawning biomass over the time series.
-When the (mean) natural mortality rate was estimated, we found large bias and uncertainty much more likely in model output such as spawning stock biomass.
-
-Bias in spawning stock biomass estimation was generally low for estimation models that assumed the correct source of process error when there was lower observation error.
-Reliable estimation of stock-recruit relationship parameters only appears possible in ideal situations with lower observation errors in age composition and indices, lower variability in recruitment process errors and large contrast in spawning biomass over time.
-We found little evidence of bias for many OM process error assumptions when there was contrast in fishing pressure even when there was greater observation error although it can lead to more variable estimates of natural mortality.
-For OMs where there was bias in natural mortality due to high observation error, estimation the stock-recruit relationship seemed to remove the bias.
-However, estimation of natural mortality can cause large differences between the true and estimated SSB (that may be unbiased on average) when there is less contrast in fishing pressure over time and higher observation error.
-
-Retrospective patterns were generally weak for all estimation models regardless of the true source of process error, but they can be expected for recruitment even for the correct process error assumptions when observation error is high.
-When models did exhibit some retrospective pattern, estimation the mean natural mortality rate tended to remove it.
-
 \pagebreak
@@
 
-Application of state-space models in fisheries stock assessment and management has expanded dramatically within ICES, Canada, and the Northeast US \citep{nielsenberg14,cadigan16,pedersenberg17,stockmiller21}.
-State-space approaches that use random effects to parameterize process errors is considered best practice and a requirement for the next generation of stock assessment models \citep{hoyleetal22, punt23}.
-
-Much is known about the reliability of state-space models that are linear or Gaussian \citep{aeberhardetal18}, but applications in fisheries management are nonlinear and typically include multiple types of observations with varying distributional assumptions.
+Application of state-space models in fisheries stock assessment and management has expanded dramatically within ICES, Canada, and the Northeast US [@nielsenberg14;@cadigan16;@pedersenberg17;@stockmiller21].
+State-space approaches that use random effects to parameterize process errors is considered best practice and a requirement for the next generation of stock assessment models [@hoyleetal22; @punt23].
+
+Much is known about the reliability of state-space models that are linear or Gaussian [@aeberhardetal18], but applications in fisheries management are nonlinear and typically include multiple types of observations with varying distributional assumptions.
 Furthermore, there is a wide range of potential random effects structures and model parameters that can be treated as random effects in assessment models.
 We know relatively little about the factors affecting statistical reliability of such models or the ability of information criteria to distinguish among such alternative structures.
+
+@lietal24 investigated some aspects of inferences for operating models with multiple sources of process error, but there are differences for this paper.
 
@@
 
-We used the Woods Hole Assessment Model (WHAM) to configure operating and EMs in our simulation study \citep[]{millerstock20,stockmiller21}.
+We used the Woods Hole Assessment Model (WHAM) to configure operating and EMs in our simulation study [@millerstock20;@stockmiller21].
 WHAM is an R package freely available as a github repository.
 For this study we used [version 1.0.6.9000, commit 77bbd94](https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3).
-This packages has also been used to configure operating and EMs for closed loop simulations evaluating index-based assessment methods \citep{legaultetal23} and is used for management of haddock, butterfish, American plaice, bluefish, Atlantic cod, and black sea bass in the Northeast US.
+This package has also been used to configure operating and EMs for closed loop simulations evaluating index-based assessment methods [@legaultetal23] and is used for management of haddock, butterfish, American plaice, bluefish, Atlantic cod, black sea bass, and yellowtail flounder in the Northeast US.
 
 We completed a simulation study with a number of OMs that can be categorized based on where process error random effects are assumed: abundance at age (R, R+S), natural mortality (R+M), fleet selectivity (R+Sel), or index catchability (R+q).
-For each OM assumptions about variance of process errors and observations are required and the values we used were based on a review of the range of estimates from recent applications of WHAM in management of stocks of haddock, butterfish, and American plaice in the NE US.
+For each OM assumption about variance of process errors and observations are required and the values we used were based on a review of the range of estimates from applications of WHAM in management of stocks of haddock, butterfish, and American plaice in the NE US.
 
@@
 
-We did not use the bias-correction feature for process errors or observations described by \citet{stockmiller21} for operating and EMs.
+We did not use the log-normal bias-correction feature for process errors or observations described by [@stockmiller21] for operating and EMs [@lietalinreview].
 Simulations and model fitting were all carried out on the University of Massachusetts Green High-Performance Computing Cluster.
@@
 
-The population model tracks 10 age classes: ages 1 to 10+ and we assume spawning occurs 1/4 of the way through the year.
-The maturity at age was a logistic curve with $a_{50}$ = 2.89 and slope = 0.88 and assumed known in all EMs (Figure \ref{om_maturity}).
-
-Weight at age was generated with a LVB growth function
+The population consists of 10 age classes: ages 1 to 10+ and we assume spawning occurs each year 1/4 of the way through the year.
+The maturity at age was a logistic curve with $a_{50}$ = 2.89 and slope = 0.88 (Figure \ref{om_maturity}).
+
+Weight at age was generated with a von Bertalanffy growth function
 $$
@@
 All post-recruit productivity components are constant in the NAA and survey catchability process error OMs.
-Therefore steepness and unfished recruitment are also constant over the time period for those OMs \citep{millerbrooks21}.
+Therefore steepness and unfished recruitment are also constant over the time period for those OMs [@millerbrooks21].
 We specified unfished recruitment = $R_0 = e^{10}$ and $\Fmsy = F_{40\%} = 0.348$ equated to a steepness of 0.69 and $\alpha=0.60$ and $\beta = 2.4 \times 10^{-5}$ for the
@@
 For OMs without process errors on natural mortality we assumed the rate was assumed 0.2.
-For OMs with process errors on natural mortality the mean rate was 0.2.
-
-Two alternative fishing histories were used for OMs.
-In the first scenario, the stock experiences overfishing for the first 20 years and fishing at \Fmsy for the last 20 years.
+For OMs with process errors on natural mortality the mean log natural mortality rate was $\log(0.2)$.
+
+We used two fishing scenarios for OMs.
+In the first scenario, the stock experiences overfishing at 2.5\Fmsy for the first 20 years and fishing at \Fmsy for the last 20 years (denoted $2.5\Fmsy \rightarrow \Fmsy$).
 In the second scenario, the stock is fished at \Fmsy for the entire time period.
-The magnitude of the overfishing assumptions is based on average estimates of overfishing for NE groundfish stocks from \citet{wiedenmannetal19}.
-\Citet{legaultetal23} also used similar approaches to defining fishing mortality histories for OMs.
-
-Initial population was configured at the equilibrium distribution fishing at either $F = 2.5\times \Fmsy$ or $F = \Fmsy$ for the two alternative fishing histories.
-That is for a deterministic model, the age composition would not change over time when the fishing mortality was constant at the respective level.
-
-For OMs with time-varying random effects for M, steepness is not constant, but we used the same alpha and beta parameters as other OMs this equates to a steepness and R0 at the mean of the time series process for M.
+The magnitude of the overfishing assumptions is based on average estimates of overfishing for NE groundfish stocks from [@wiedenmannetal19].
+@legaultetal23 also used similar approaches to defining fishing mortality histories for OMs.
+
+We specified initial population abundance at age at the equilibrium distribution fishing at either $F = 2.5\times \Fmsy$ or $F = \Fmsy$ for the two alternative fishing histories.
+This implies that, for a deterministic model, the abundance at age would not change from the first year to the next.
+
+For OMs with time-varying random effects for M, steepness is not constant, but we used the same alpha and beta parameters as other OMs this equates to a steepness and R0 at the median of the time series process for M.
 For OMs with time-varying random effects for fishery selectivity, \Fmsy is also not constant however we use the same F history as other OMs which corresponds to Fmsy at the mean selectivity parameters.
@@
 
-Two time series of surveys are assumed and observed in numbers rather than biomass for the entire 40 year period with one occurring in the spring (0.25 way through the year) and one in the fall (0.75 way through the year).
-Actually we have it currently configured that both occur 0.5 way through the year.
+Two time series of surveys are assumed and observed in numbers rather than biomass for the entire 40 year period with one occurring in the spring (0.25 of each year) and one in the fall (0.75 of each year).
 Catchability of both surveys are assumed to be 0.1.
@@
 
-The factorial combinations of fishing histories, observation error assumptions, and marginal variance assumptions for recruitment and survival results in 24 different OMs with random effects on recruitment and survival (Table \ref{naa_om_table}).
+For operating models with random effects on recruitment and(or) survival (R, R+S) we assumed marginal standard deviations for recruitment of $\sigma_R \in \{0.5,1.5\}$ and marginal standard deviations for older age classes of $\sigma_{2+} \in \{0,0.25, 0.5\}$.
+The full factorial combination of these process error assumptions and fishing history (2 levels) and observation error (2 levels) scenarios described above results in 24 different R ($\sigma_{2+} = 0$) and R+S operating models (Table \ref{naa_om_table}).
 
@@
 
-The factorial combinations of fishing histories, observation error assumptions, and marginal variance and autocorrelation assumptions for natural mortality results in 16 different OMs with random effects on natural mortality (Table \ref{M_om_table}).
-Internally, WHAM treats natural mortality as a log-transformed parameter
-$$
-\log M_y = \mu_M + \epsilon_{M,y}
-$$
-that is a linear combination of a mean $\mu_M$ and any annual random effects marginally distributed as $\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)$.
-The values for the marginal variance and correlation parameters assumed in the OMs are given in Table \ref{M_om_table}.
+All R+M OMs treat natural mortality constant across age, but with annually varying random effects. WHAM treats natural mortality as a log-transformed parameter 
+$$
+\log M_{y,a} = \mu_{M} + \epsilon_{M,y}
+$$
+that is a linear combination of a mean that is constant across ages $\mu_{M} = \log(0.2)$ and any annual random effects marginally distributed as $\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)$.
+Uncorrelated random effects were also included on recruitment with $\sigma_R = 0.5$ (hence, R+M).
+The marginal standard deviations we assumed for log natural mortality random effects were $\sigma_M \in \{0.1, 0.5\}$ and AR1 autocorrelation parameters of $\rho_M \in \{0,0.9\}$.
+The full factorial combination of these process error assumptions and fishing history (2 levels) and observation error (2 levels) scenarios described above results in 16 different R+M OMs (Table \ref{M_om_table}).
 
@@
 
-The factorial combinations of fishing histories, observation error assumptions, and marginal variance and autocorrelation assumptions for selectivity parameters results in 16 different OMs with random effects on selectivity (Table \ref{sel_om_table}).
-Internally, WHAM treats selectivity parameter $s$ as a logit-transformed parameter
+MORE SPECIFICS about correlation  of random effects? Both selectivity pars? just correlated by year?
+WHAM treats selectivity parameter $s$ as a logit-transformed parameter 
 $$
@@
 All selectivity parameters are either $a_50$ or slope parameters and we assume bounds of 0 and 10 for all selectivity parameters for all operating and EMs.
-The values for the marginal variance and correlation parameters assumed in the OMs are given in Table \ref{sel_om_table}.
+The marginal standard deviations we assumed for logit scale random effects were $\sigma_s \in \{0.1, 0.5\}$ and AR1 autocorrelation parameters of $\rho_s \in \{0,0.9\}$.
+Like R+M OMs,  the full factorial combination of these process error assumptions and fishing history (2 levels) and observation error (2 levels) scenarios described above results in 16 different R+Sel OMs (Table \ref{sel_om_table}).
 
@@
 
-The factorial combinations of fishing histories, observation error assumptions, and marginal variance and autocorrelation assumptions for catchability results in 16 different OMs with random effects on catchability (Table \ref{q_om_table}).
-Internally, WHAM treats catchability for index $i$ as a logit-transformed parameter
+Like selectivity parameters, WHAM treats catchability for an index $i$ as a logit-transformed parameter 
 $$
@@
 For operating and EMs with process errors on catchability, the temporal variation is only assumed for the first index.
-The values for the marginal variance and correlation parameters assumed in the OMs are given in Table \ref{q_om_table}
+The marginal standard deviations we assumed for logit scale random effects were $\sigma_i \in \{0.1, 0.5\}$ and AR1 autocorrelation parameters of $\rho_i \in \{0,0.9\}$.
+Like R+M and R+Sel OMs, the full factorial combination of these process error assumptions and fishing history (2 levels) and observation error (2 levels) scenarios described above results in 16 different R+q OMs (Table \ref{q_om_table}).
 
@@
 
+maturity, weight at age, index and catch CVs, assumed known in all EMs
+
 ## Measures of reliability {-}
@@
 
-1. Did the optimization routine (stats::nlminb) complete without error?
-2. Did the stats::nlminb convergence flag = 0 indicate successful convergence?
-3. Was the maximum absolute value of the gradient of the log-likelihood < $1\times10^{-6}$?
-4. Did TMB::sdreport provide non-NA values for all fixed effects standard errors?
-5. Did TMB::sdreport provide all standard errors < 100?
+1. Fit complete: Did the optimization routine (stats::nlminb) complete without error?
+2. No flag: Did the stats::nlminb convergence flag = 0 indicate successful convergence?
+3. Good gradient: Was the maximum absolute value of the gradient of the log-likelihood < $1\times10^{-6}$?
+4. No SE NAs: Did TMB::sdreport provide non-NA values for all fixed effects standard errors?
+5. No big SEs: Did TMB::sdreport provide all standard errors < 100?
 
@@
 For example, variance parameters for random effects can often be estimated near 0 when the fit suggests no variation in the random effects or, equivalently, that the model without random effects is better.
+Note that Types 2 through 5 also require type 1, and type 5 requires type 4.
+We used the Clopper-Pearson exact method for constructing 95\% confidence intervals of the probabilities of convergence [@clopperpearson34;@thulin14].
 
@@
 
-We measured the frequency of correct model selection using marginal AIC.
-For a given OM we the set of models that were considered all made the same assumptions on whether or not to estimate M and whether a stock-recruit function is assumed.
-
-We also measured the frequency of correctly selecting models with Beverton-Holt stock recruit function assumed over models without the stock-recruit function (null model).
-Because the proportions of simulations where the correct model was chosen for a given set of OM factors implied generally poor performance of AIC, we analyzed the effect of variation in SSB on the probability of lower AIC of the BH models.
-We fit binomial generalize linear models to the observed AIC choice for each data set as a function of log(SD(SSB)) and OM.
+We estimated the probability of selection of each process error model structure (R, R+S, R+M, R+Sel, R+q) using marginal AIC.
+For a given operating model, we compared AIC for EMs that all made the same assumptions about median/constant natural mortality (known or estimated) and stock recruitment model (Beverton-Holt or none).
+
+We also estimated the probability of correctly selecting between models with Beverton-Holt stock recruit function assumed and models without the stock-recruit function (null model).
+We made these comparisons between models that otherwise assume the same process error structure as the operating model and both of the compared models either estimate median/constant natural mortality or assume it is known.
+Our preliminary inspections of the proportions of simulations where the correct recruitment model was chosen for a given set of OM factors indicated generally poor performance of AIC.
+Furthermore, it has been shown that estimation of stock-recruit parameters requires a wide range of SSB (References).
+Therefore, we fit logistic regression models to the indicator of Beverton-Holt models having lower AIC as a function of the log-standard deviation of the true log(SSB) (similar to the log of the coefficient of variation for SSB) for each simulation.
+
+All results only condition on whether all of the compared estimating models completed the optimization process without failure.
+We did not condition on convergence as defined by a gradient threshold or invertibity of the hessian because optimization can correctly determine the the correct likelihood that would indicate poor convergence because variance parameters may be at the lower bound of zero correctly for models that assume the incorrect process error structure.
 
@@
 $$
-of annual SSB and fully-selected fishing mortality for each EM and constructed 95\% confidence intervals for the median relative bias using the binomial distribution approach as in \citet{millerhyun18} and \citet{stockmiller21}.
+of annual SSB and fully-selected fishing mortality for each EM and constructed 95\% confidence intervals for the median relative bias using the binomial distribution approach as in @millerhyun18 and @stockmiller21.
 We also estimated bias of stock-recruit parameters for EMs that assumed the Beverton-Holt stock recruit function and the mean natural mortality rate parameter when estimated.
+
+Similar to the AIC results, bias results only condition on whether the estimating model completed the optimization process without failure.
+We did not condition on convergence as defined by a gradient threshold or invertibity of the hessian because the optimized model can provide reliable estimation of SSB, F, M, and stock recruit parameters whether or not the model was able to estimate non-zero random effects.
+In practice, the model would be reconfigured to remove unnecessary process errors and produce otherwise equivalent parameter estimates.
 
@@
 We calculated median 95\% confidence intervals for Mohn's $\rho$ using the same methods as that for relative bias.
+
+Similar to the other results, retrospective results only condition on whether all of the peels of a given estimating model completed the optimization process without failure.
+We did not condition on convergence as defined by a gradient threshold or invertibity of the hessian because the optimized model can provide reliable estimation of SSB, F, M, and stock recruit parameters whether or not the model was able to estimate non-zero random effects.
 
@@
 AIC generally performs weakly in determining stock recruit relationships unless there is large contrast in SSB.
-Interestingly, \citet{milleretal16} found AIC to prefer a stock-recruit relationship for SNEMA yellowtail flounder where there was a large change in stock size estimated over time and flatfish are well observed by the bottom trawl indices relative to some other guilds such as semi-pelagic species.
+Interestingly, @milleretal16 found AIC to prefer a stock-recruit relationship for SNEMA yellowtail flounder where there was a large change in stock size estimated over time and flatfish are well observed by the bottom trawl indices relative to some other guilds such as semi-pelagic species.
 Although we did not compare models with alternative stock-recruit relationships, we do not expect AIC to perform any better distinguishing these relationships.
-Our results comport well with those of \citet{devalpinehastings02} where state-space stock-recruit models were fit just to SSB and recruitment "observations".
+Our results comport well with those of @devalpinehastings02 where state-space stock-recruit models were fit just to SSB and recruitment "observations".
 
@@
 
+## Summary {-}
+
+We conducted a simulation study where we simulated process errors and observations for 72 operating models with alternative assumptions about fishing history, degree of uncertainty in index and age composition observations, type (recruitment, survival, fishery selectivity, catchability and natural mortality), degree of variation, and correlation of process errors.
+We fit 20 different estimation models to each of 100 simulated set of observations with alternative assumptions type and correlation structure of process errors, (mean) natural mortality was known or estimated, and a B-H stock recruit relationship was assumed or not.
+
+Across simulations we summarized probability of convergence of fitted models, accuracy of marginal AIC in determining the correct process error assumption and it ability to determine the Beverton-Holt stock recruit relationship, bias in annual spawning stock biomass, in estimation of natural mortality, and in stock-recruit relationship parameters, and severity of retrospective patterns for estimation models.
+
+Alternative measures of convergence performed differently.
+Invertible hessians and resulting standard error estimation was possible when criteria based on the gradient of the optimized log-likelihood with respect to the fixed effects parameters.
+Using hessian-based convergence, probability of convergence was best for models that assumed the correct source of process error, assumed M was known, and did not assume stock-recruit relationships.
+
+Using marginal AIC provided most accurate inferences about the process errors on recruitment and survival and selectivity, in that AIC preferred EMs with assumed process errors that matched OMs, and EMs with those assumed process errors were not preferred when alternative process errors were assumed in the OMs.
+However, when the true process errors were more variable, AIC accuracy increased to a useful level.
+We found AIC more accurately determined a B-H stock recruit relationship rather than the null model without a S-R relationship when there was low variability in recruitment, low variability in survival random effects, and higher variation in spawning biomass over the time series.
+When the (mean) natural mortality rate was estimated, we found large bias and uncertainty much more likely in model output such as spawning stock biomass.
+
+Bias in spawning stock biomass estimation was generally low for estimation models that assumed the correct source of process error when there was lower observation error.
+Reliable estimation of stock-recruit relationship parameters only appears possible in ideal situations with lower observation errors in age composition and indices, lower variability in recruitment process errors and large contrast in spawning biomass over time.
+We found little evidence of bias for many OM process error assumptions when there was contrast in fishing pressure even when there was greater observation error although it can lead to more variable estimates of natural mortality.
+For OMs where there was bias in natural mortality due to high observation error, estimation the stock-recruit relationship seemed to remove the bias.
+However, estimation of natural mortality can cause large differences between the true and estimated SSB (that may be unbiased on average) when there is less contrast in fishing pressure over time and higher observation error.
+
+Retrospective patterns were generally weak for all estimation models regardless of the true source of process error, but they can be expected for recruitment even for the correct process error assumptions when observation error is high.
+When models did exhibit some retrospective pattern, estimation the mean natural mortality rate tended to remove it.
+
 # Acknowledgements {-}
@@
 
-\bibliography{paper}
-<!-- \bibliography{project-0-paper}
-\bibliographystyle{/home/tmiller2/work/bibliographies/cjfas2}  -->
+# References {-}
 
@@
 \pagebreak
+
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width = \textwidth]{om_input_plots_figure}
+\end{center}
+\caption{The proportion mature at age, weight at age, fleet and index selectivity at age, and Beverton-Holt stock recruit relationship assumed for the population in all operating models.
+For operating models with random effects on fleet selectivity, this represents the selectivity at the mean of the random effects.}\label{om_inputs_fig}
+\end{figure}
+
+\clearpage
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{type_4_convergence_plots}
+\end{center}
+\caption{Estimated probability of fits providing hessian-based standard errors for EMs assuming alternative process error (colored points and lines), and median natural mortality (estimated or known) and Beverton-Holt stock recruit functions (estimated or not; along x-axis) when fitted to operating models that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{hessian_SE_convergence}
+\end{figure}
+\end{landscape}
+
+\clearpage
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{pe_aic_plots}
+\end{center}
+\caption{Estimated probability of lowest AIC for EMs assuming alternative process error structures (colored bars) conditional on alternative assumptions for median natural mortality (estimated or known) and Beverton-Holt stock recruit functions (estimated or not; along x-axis) when fitted to operating models that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
+Striped bars indicate results where the EM process error structure matches that of the operating model.}\label{pe_aic}
+\end{figure}
+\end{landscape}
+
+\clearpage
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{sr_aic_plots}
+\end{center}
+\caption{Estimated probability of lowest AIC from logistic regression on the log-standard deviation of the true log(SSB) in each simulation for estimating model with Beverton-Holt stock recruit functions, rather than the otherwise equivalent EM without the stock recruit function.
+Results are conditional on alternative assumptions for median natural mortality (estimated or known) and on EMs having the correct process error structure: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Rug along x-axis denotes $SD(\log(SSB))$ values for each simulation amd polygons represent 95\% confidence intervals.}\label{sr_aic}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{term_SSB_bias_plots}
+\end{center}
+\caption{Median relative error of terminal year SSB for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Diamond shaped points denote results where the EM process error assumption matches that of the operating model.
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{SSB_rel_error}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{mohns_rho_ssb_plots}
+\end{center}
+\caption{Median Mohn's rho for SSB for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_ssb}
+\end{figure}
+\end{landscape}
+
+\pagebreak
+
+\setcounter{figure}{0}
+\renewcommand\thefigure{S\arabic{figure}}
+
+\setcounter{table}{0}
+\renewcommand\thetable{S\arabic{table}}
+
+# Supplementary Materials {-}
 
@@
 Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.}\label{naa_om_table}
-{\footnotesize \input{naa_om_tab}}
+{\footnotesize \input{../paper/naa_om_tab}}
 \end{table}
@@
 For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}\label{M_om_table}
-{\input{M_om_tab}}
+{\input{../paper/M_om_tab}}
 \end{table}
@@
 For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}\label{sel_om_table}
-{\input{Sel_om_tab}}
+{\input{../paper/Sel_om_tab}}
 \end{table}
@@
 For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}\label{q_om_table}
-{\input{q_om_tab}}
+{\input{../paper/q_om_tab}}
 \end{table}
@@
 \caption{Distinguishing characteristics of the estimating models.}\label{em_table}
-{\scriptsize \input{em_tab}}
+{\scriptsize \input{../paper/em_tab}}
 \end{table}
 
-<!-- AIC TABLES --->
-
-\begin{table}
-\caption{Operating models and estimation models all assume matching R or R+S process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}\label{naa_om_em_R_BH_aic_table}
-{\input{naa_om_em_R_BH_aic_table}}
-\end{table}
-
-\begin{table}
-\caption{Operating models and estimation models all assume matching R+M process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}\label{M_om_em_R_BH_aic_table}
-{\input{M_om_em_R_BH_aic_table}}
-\end{table}
-
-\begin{table}
-\caption{Operating models and estimation models all assume matching R+Sel process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}\label{Sel_om_em_R_BH_aic_table}
-{\input{Sel_om_em_R_BH_aic_table}}
-\end{table}
-
-\begin{table}
-\caption{Operating models and estimation models all assume matching R+q process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}\label{q_om_em_R_BH_aic_table}
-{\input{q_om_em_R_BH_aic_table}}
-\end{table}
-
 \clearpage
-<!-- OM configuration plots --->
-
-\begin{figure}
-\caption{The proportion mature at age assumed for the population in all operating and estimating models.}\label{om_maturity}
-\begin{center}
-\includegraphics[width = \textwidth]{om_maturity.png}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\caption{The weight at age assumed for the population in all operating and estimating models.}\label{om_waa}
-\begin{center}
-\includegraphics[width = \textwidth]{om_waa.png}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\caption{The Beverton-Holt stock recruit relationship assume for all operating models.}\label{om_sr}
-\begin{center}
-\includegraphics[width = \textwidth]{om_sr.png}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\caption{The selectivity at age assumed for the fishing fleet in operating models without selectivity random effects.
-The mean selectivity at age across time for the fishing fleet in operating models with selectivity random effects.}\label{om_mean_selectivity}
-\begin{center}
-\includegraphics[width = \textwidth]{om_mean_selectivity.png}
-\end{center}
-\end{figure}
-
-<!-- Convergence --->
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structures R and R+S.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.}\label{naa_om_em_R_MF_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_p_convergence_meanR_M_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structures R and R+S.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.}\label{naa_om_em_R_ME_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_p_convergence_meanR_M_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structures R and R+S.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is fixed at the true value.}\label{naa_om_em_BH_MF_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_p_convergence_BH_M_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structures R and R+S.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is estimated.}\label{naa_om_em_BH_ME_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_p_convergence_BH_M_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+M.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.}\label{M_om_em_R_MF_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_p_convergence_meanR_M_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+M.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.}\label{M_om_em_R_ME_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_p_convergence_meanR_M_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+M.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is fixed at the true value.}\label{M_om_em_BH_MF_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_p_convergence_BH_M_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+M.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is estimated.}\label{M_om_em_BH_ME_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_p_convergence_BH_M_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+Sel.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.}\label{Sel_om_em_R_MF_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_p_convergence_meanR_M_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+Sel.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.}\label{Sel_om_em_R_ME_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_p_convergence_meanR_M_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+Sel.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is fixed at the true value.}\label{Sel_om_em_BH_MF_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_p_convergence_BH_M_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+Sel.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is estimated.}\label{Sel_om_em_BH_ME_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_p_convergence_BH_M_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+q.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.}\label{q_om_em_R_MF_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_p_convergence_meanR_M_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+q.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.}\label{q_om_em_R_ME_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_p_convergence_meanR_M_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+q.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is fixed at the true value.}\label{q_om_em_BH_MF_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_p_convergence_BH_M_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+q.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is estimated.}\label{q_om_em_BH_ME_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_p_convergence_BH_M_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-<!-- AIC --->
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R and R+S process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.} \label{naa_om_proportion_best_aic_R_MF}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_proportion_best_aic_R_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R and R+S process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.} \label{naa_om_proportion_best_aic_R_ME}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_proportion_best_aic_R_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R and R+S process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate a stock-recruit relationship and and M is fixed at the true value.} \label{naa_om_proportion_best_aic_SR_MF}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_proportion_best_aic_SR_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R and R+S process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate a stock-recruit relationship and and M is estimated.} \label{naa_om_proportion_best_aic_SR_ME}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_proportion_best_aic_SR_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+M process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.} \label{M_om_proportion_best_aic_R_MF}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_proportion_best_aic_R_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+M process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.} \label{M_om_proportion_best_aic_R_ME}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_proportion_best_aic_R_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+M process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate a stock-recruit relationship and and M is fixed at the true value.} \label{M_om_proportion_best_aic_SR_MF}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_proportion_best_aic_SR_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+M process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate a stock-recruit relationship and and M is estimated.} \label{M_om_proportion_best_aic_SR_ME}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_proportion_best_aic_SR_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+Sel process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.} \label{Sel_om_proportion_best_aic_R_MF}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_proportion_best_aic_R_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+Sel process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.} \label{Sel_om_proportion_best_aic_R_ME}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_proportion_best_aic_R_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+Sel process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate a stock-recruit relationship and and M is fixed at the true value.} \label{Sel_om_proportion_best_aic_SR_MF}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_proportion_best_aic_SR_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+Sel process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate a stock-recruit relationship and and M is estimated.} \label{Sel_om_proportion_best_aic_SR_ME}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_proportion_best_aic_SR_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+q process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.} \label{q_om_proportion_best_aic_R_MF}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_proportion_best_aic_R_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+q process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.} \label{q_om_proportion_best_aic_R_ME}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_proportion_best_aic_R_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+q process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate a stock-recruit relationship and and M is fixed at the true value.} \label{q_om_proportion_best_aic_SR_MF}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_proportion_best_aic_SR_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+q process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate a stock-recruit relationship and and M is estimated.} \label{q_om_proportion_best_aic_SR_ME}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_proportion_best_aic_SR_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-<!--
-## Estimating models include alternative random effects options: NAA, M, sel, q {-}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{type_3_convergence_plots}
+\end{center}
+\caption{Probability of estimating models providing maximum absolute values of gradients less than $10^{-6}$ assuming alternative process error (colored points and lines), and median natural mortality (estimated or known) and Beverton-Holt stock recruit functions (estimated or not; along x-axis) when fitted to operating models that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{gradient_convergence}
+\end{figure}
+\end{landscape}
+
 \clearpage
 
-\begin{table}
-\caption{NAA operating models, estimating models all assume a B-H stock recruit relationship and M is fixed at the true value.}
-{\input{naa_om_em_SR_MF_aic_table}}
-\end{table}
-
-\begin{table}
-\caption{NAA operating models, estimating models all assume a B-H stock recruit relationship and M is estimated.}
-{\input{naa_om_em_SR_ME_aic_table}}
-\end{table}
-
-\begin{table}
-\caption{NAA operating models, estimating models all estimate a mean recruitment and M is fixed at the true value.}
-{\input{naa_om_em_R_MF_aic_table}}
-\end{table}
-
-\begin{table}
-\caption{NAA operating models, estimating models all estimate a mean recruitment and M estimated.}
-{\input{naa_om_em_R_MF_aic_table}}
-\end{table}
-\clearpage 
--->
-
-\clearpage
-
-<!-- AIC stock-recruit selection--->
-
-\begin{figure}
-\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
-Operating and estimating models have matching R or R+S process error structures.
-Estimating models assume mean M is fixed at the true value.}\label{naa_om_MF_BH_glm_AIC_plots}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_MF_pred_BH_best.png}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
-Operating and estimating models have matching R or R+S process error structures.
-Estimating models allow estimation of mean M.}\label{naa_om_ME_BH_glm_AIC_plots}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_ME_pred_BH_best.png}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
-Operating and estimating models have matching R+M process error structures.
-Estimating models assume mean M is fixed at the true value.}\label{M_om_MF_BH_glm_AIC_plots}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_MF_pred_BH_best.png}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
-Operating and estimating models have matching R+M process error structures.
-Estimating models allow estimation of mean M.}\label{M_om_ME_BH_glm_AIC_plots}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_ME_pred_BH_best.png}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
-Operating and estimating models have matching R+Sel process error structures.
-Estimating models assume mean M is fixed at the true value.}\label{Sel_om_MF_BH_glm_AIC_plots}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_MF_pred_BH_best.png}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
-Operating and estimating models have matching R+Sel process error structures.
-Estimating models allow estimation of mean M.}\label{Sel_om_ME_BH_glm_AIC_plots}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_ME_pred_BH_best.png}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
-Operating and estimating models have matching R+q process error structures.
-Estimating models assume mean M is fixed at the true value.}\label{q_om_MF_BH_glm_AIC_plots}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_MF_pred_BH_best.png}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
-Operating and estimating models have matching R+q process error structures.
-Estimating models allow estimation of mean M.}\label{q_om_ME_BH_glm_AIC_plots}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_ME_pred_BH_best.png}
-\end{center}
-\end{figure}
-
-\clearpage
-
-<!-- Bias --->
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R and R+S process errors.
-Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{naa_om_em_R_MF_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_R_MF_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R and R+S process errors.
-Estimation models do not assume a stock-recruit relationship and estimate M.}\label{naa_om_em_R_ME_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_R_ME_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R and R+S process errors.
-Estimation models assume a BH stock-recruitment function and M is fixed at the true value.}\label{naa_om_em_BH_MF_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_BH_MF_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R and R+S process errors.
-Estimation models assume a BH stock-recruitment function and estimate M.}\label{naa_om_em_BH_ME_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_BH_ME_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+M process errors.
-Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{M_om_em_R_MF_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_R_MF_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+M process errors.
-Estimation models do not assume a stock-recruit relationship and estimate M.}\label{M_om_em_R_ME_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_R_ME_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+M process errors.
-Estimation models assume a BH stock-recruitment function and M is fixed at the true value.}\label{M_om_em_BH_MF_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_BH_MF_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+M process errors.
-Estimation models assume a BH stock-recruitment function and estimate M.}\label{M_om_em_BH_ME_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_BH_ME_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+Sel process errors.
-Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{Sel_om_em_R_MF_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_R_MF_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+Sel process errors.
-Estimation models do not assume a stock-recruit relationship and estimate M.}\label{Sel_om_em_R_ME_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_R_ME_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+Sel process errors.
-Estimation models assume a BH stock-recruitment function and M is fixed at the true value.}\label{Sel_om_em_BH_MF_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_BH_MF_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+Sel process errors.
-Estimation models assume a BH stock-recruitment function and estimate M.}\label{Sel_om_em_BH_ME_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_BH_ME_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+q process errors.
-Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{q_om_em_R_MF_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_R_MF_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+q process errors.
-Estimation models do not assume a stock-recruit relationship and estimate M.}\label{q_om_em_R_ME_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_R_ME_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+q process errors.
-Estimation models assume a BH stock-recruitment function and M is fixed at the true value.}\label{q_om_em_BH_MF_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_BH_MF_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+q process errors.
-Estimation models assume a BH stock-recruitment function and estimate M.}\label{q_om_em_BH_ME_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_BH_ME_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of Beverton-Holt stock recruit parameters for estimating models with correct process error assumptions fitted to data sets simulated with R and R+S process errors.
-Estimation models either assume the correct natural mortality rate or estimate it.}\label{naa_om_SR_relbias}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_SR_relerror.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of Beverton-Holt stock recruit parameters for estimating models with correct process error assumptions fitted to data sets simulated with R+M process errors.
-Estimation models either assume the correct natural mortality rate or estimate it.}\label{M_om_SR_relbias}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_SR_relerror.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of Beverton-Holt stock recruit parameters for estimating models with correct process error assumptions fitted to data sets simulated with R+Sel process errors.
-Estimation models either assume the correct natural mortality rate or estimate it.}\label{Sel_om_SR_relbias}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_SR_relerror.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of Beverton-Holt stock recruit parameters for estimating models with correct process error assumptions fitted to data sets simulated with R+q process errors.
-Estimation models either assume the correct natural mortality rate or estimate it.}\label{q_om_SR_relbias}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_SR_relerror.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of (mean) natural mortality rate for estimating models with correct process error assumptions fitted to data sets simulated with R and R+S process errors.
-Estimation models assumed either no stock-recruit relationship or estimated a Beverton-Holt relationship.}\label{naa_om_M_relbias}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_M_relerror.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of (mean) natural mortality rate for estimating models with correct process error assumptions fitted to data sets simulated with R+M process errors.
-Estimation models assumed either no stock-recruit relationship or estimated a Beverton-Holt relationship.}\label{M_om_M_relbias}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_M_relerror.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of (mean) natural mortality rate for estimating models with correct process error assumptions fitted to data sets simulated with R+Sel process errors.
-Estimation models assumed either no stock-recruit relationship or estimated a Beverton-Holt relationship.}\label{Sel_om_M_relbias}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_M_relerror.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of (mean) natural mortality rate for estimating models with correct process error assumptions fitted to data sets simulated with R+q process errors.
-Estimation models assumed either no stock-recruit relationship or estimated a Beverton-Holt relationship.}\label{q_om_M_relbias}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_M_relerror.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-<!-- Retro --->
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R and R+S process errors.
-Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{naa_om_em_R_MF_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_mohns_rho_R_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R and R+S process errors.
-Estimation models do not assume a stock-recruit function and M estimated.}\label{naa_om_em_R_ME_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_mohns_rho_R_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R and R+S process errors.
-Estimation models assume a B-H stock-recruit function and M is fixed at the true value.}\label{naa_om_em_BH_MF_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_mohns_rho_BH_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R and R+S process errors.
-Estimation models assume a B-H stock-recruit function and M estimated.}\label{naa_om_em_BH_ME_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_mohns_rho_BH_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+M process errors.
-Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{M_om_em_R_MF_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_mohns_rho_R_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+M process errors.
-Estimation models do not assume a stock-recruit function and M estimated.}\label{M_om_em_R_ME_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_mohns_rho_R_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+M process errors.
-Estimation models assume a B-H stock-recruit function and M is fixed at the true value.}\label{M_om_em_BH_MF_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_mohns_rho_BH_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+M process errors.
-Estimation models assume a B-H stock-recruit function and M estimated.}\label{M_om_em_BH_ME_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_mohns_rho_BH_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+Sel process errors.
-Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{Sel_om_em_R_MF_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_mohns_rho_R_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+Sel process errors.
-Estimation models do not assume a stock-recruit function and M estimated.}\label{Sel_om_em_R_ME_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_mohns_rho_R_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+Sel process errors.
-Estimation models assume a B-H stock-recruit function and M is fixed at the true value.}\label{Sel_om_em_BH_MF_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_mohns_rho_BH_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+Sel process errors.
-Estimation models assume a B-H stock-recruit function and M estimated.}\label{Sel_om_em_BH_ME_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_mohns_rho_BH_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+q process errors.
-Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{q_om_em_R_MF_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_mohns_rho_R_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+q process errors.
-Estimation models do not assume a stock-recruit function and M estimated.}\label{q_om_em_R_ME_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_mohns_rho_R_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+q process errors.
-Estimation models assume a B-H stock-recruit function and M is fixed at the true value.}\label{q_om_em_BH_MF_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_mohns_rho_BH_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+q process errors.
-Estimation models assume a B-H stock-recruit function and M estimated.}\label{q_om_em_BH_ME_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_mohns_rho_BH_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{term_F_bias_plots}
+\end{center}
+\caption{Median relative error of terminal year fully-selected fishing mortality for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{F_rel_error}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{term_R_bias_plots}
+\end{center}
+\caption{Median relative error of terminal year recruitment for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{R_rel_error}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{M_bias_plots}
+\end{center}
+\caption{Median relative error of median/constant natural mortality for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{M_rel_error}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{sr_bias_plots}
+\end{center}
+\caption{Median relative error of Beverton-Holt stock recruitment parameters ($a$ and $b$) for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{SR_rel_error}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{mohns_rho_F_plots}
+\end{center}
+\caption{Median Mohn's $\rho$ of fishing mortality averaged over all age classes for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_F}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{mohns_rho_R_plots}
+\end{center}
+\caption{Median Mohn's $\rho$ of recruitment for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_F}
+\end{figure}
+\end{landscape}

### 2025-01-16 Project_0/manuscript/manuscript.tex

--- 2024-02-06 Project_0/paper/paper.tex
@@
   pdftitle={Factors affecting reliablity of state-space age-structured assessment models},
-  pdfauthor={Timothy J. Miller1; Andrew Applegate; Greg Britten; Elizabeth N. Brooks; Gavin Fay; Alex Hansell; Christopher M. Legault; Brandon Muffley; Brian C. Stock; John Wiedenmann},
+  pdfauthor={Timothy J. Miller1,2; Greg Britten3; Elizabeth N. Brooks2; Gavin Fay4; Alex Hansell2; Christopher M. Legault2; Brandon Muffley5; Brian C. Stock6; John Wiedenmann7},
   hidelinks,
@@
 %\onehalfspacing
-%\doublespacing
+\doublespacing
 \usepackage{lineno}
-%\linenumbers
+\linenumbers
 \usepackage[belowskip=0pt,aboveskip=0pt]{caption}
 \usepackage{relsize}
+\usepackage{float}
+\usepackage{lscape}
+\usepackage{longtable}
+\usepackage{amsmath,rotating}
+\usepackage[scanall]{psfrag}
+\usepackage{bm}
+\usepackage{caption,graphics}
+\usepackage{graphicx}
+\usepackage{sectsty}
+\usepackage{color}
+\usepackage{fancyhdr}
+\usepackage{xspace}
+\usepackage{textcomp}
+\usepackage{upgreek}
+\renewcommand\figurename{Fig.}
+\captionsetup{labelsep=period, singlelinecheck=false}
+\newcommand{\changesize}[1]{\fontsize{#1pt}{#1pt}\selectfont}
+\renewcommand{\arraystretch}{1.5}
+%\renewcommand\theadfont{}
+
 \newcommand{\Fmsy}{\ensuremath{F_{\text{MSY}}}\xspace}
@@
 
-\usepackage{xspace}
-\usepackage{bm}
-\usepackage{caption,graphics}
-\usepackage{graphicx}
-\usepackage{makecell}
-\renewcommand\figurename{Fig.}
-\captionsetup{labelsep=period, singlelinecheck=false}
-\newcommand{\changesize}[1]{\fontsize{#1pt}{#1pt}\selectfont}
-\renewcommand{\arraystretch}{1.5}
-\renewcommand\theadfont{}
 \usepackage{booktabs}
@@
 \fi
-\usepackage[]{natbib}
-\bibliographystyle{cjfas2.bst}
 
@@
 assessment models}
-\author{Timothy J. Miller\textsuperscript{1} \and Andrew
-Applegate \and Greg Britten \and Elizabeth N. Brooks \and Gavin
-Fay \and Alex Hansell \and Christopher M. Legault \and Brandon
-Muffley \and Brian C. Stock \and John Wiedenmann}
-\date{}
+\author{Timothy J. Miller\textsuperscript{1,2} \and Greg
+Britten\textsuperscript{3} \and Elizabeth N.
+Brooks\textsuperscript{2} \and Gavin Fay\textsuperscript{4} \and Alex
+Hansell\textsuperscript{2} \and Christopher M.
+Legault\textsuperscript{2} \and Brandon
+Muffley\textsuperscript{5} \and Brian C.
+Stock\textsuperscript{6} \and John Wiedenmann\textsuperscript{7}}
+\date{15 January, 2025}
 
@@
 \(^1\)corresponding author:
-\href{mailto:timothy.j.miller@noaa.gov}{\nolinkurl{timothy.j.miller@noaa.gov}},
-Northeast Fisheries Science Center, Woods Hole Laboratory, 166 Water
-Street, Woods Hole, MA 02543 USA\\
+\href{mailto:timothy.j.miller@noaa.gov}{\nolinkurl{timothy.j.miller@noaa.gov}}\\
+\(^2\)Northeast Fisheries Science Center, Woods Hole Laboratory, 166
+Water Street, Woods Hole, MA 02543 USA\\
+\(^3\)Woods Hole Oceanographic Institution\\
+\(^4\)SMAST\\
+\(^5\)Institute of Marine Research\\
+\(^6\)Mid-Atlantic Fisheries Management Council\\
+\(^7\)Rutgers University\\
 
@@
 
-\hypertarget{summary}{%
-\subsection*{Summary}\label{summary}}
-\addcontentsline{toc}{subsection}{Summary}
-
-We conducted a simulation study where we simulated process errors and
-observations for 72 operating models with alternative assumptions about
-fishing history, degree of uncertainty in index and age composition
-observations, type (recruitment, survival, fishery selectivity,
-catchability and natural mortality), degree of variation, and
-correlation of process errors. We fit 20 different estimation models to
-each of 100 simulated set of observations with alternative assumptions
-type and correlation structure of process errors, (mean) natural
-mortality was known or estimated, and a B-H stock recruit relationship
-was assumed or not.
-
-Across simulations we summarized probability of convergence of fitted
-models, accuracy of marginal AIC in determining the correct process
-error assumption and it ability to determine the Beverton-Holt stock
-recruit relationship, bias in annual spawning stock biomass, in
-estimation of natural mortality, and in stock-recruit relationship
-parameters, and severity of retrospective patterns for estimation
-models.
-
-Alternative measures of convergence performed differently. Invertible
-hessians and resulting standard error estimation was possible when
-criteria based on the gradient of the optimized log-likelihood with
-respect to the fixed effects parameters. Using hessian-based
-convergence, probability of convergence was best for models that assumed
-the correct source of process error, assumed M was known, and did not
-assume stock-recruit relationships.
-
-Using marginal AIC provided most accurate inferences about the process
-errors on recruitment and survival and selectivity, in that AIC
-preferred EMs with assumed process errors that matched OMs, and EMs with
-those assumed process errors were not preferred when alternative process
-errors were assumed in the OMs. However, when the true process errors
-were more variable, AIC accuracy increased to a useful level. We found
-AIC more accurately determined a B-H stock recruit relationship rather
-than the null model without a S-R relationship when there was low
-variability in recruitment, low variability in survival random effects,
-and higher variation in spawning biomass over the time series. When the
-(mean) natural mortality rate was estimated, we found large bias and
-uncertainty much more likely in model output such as spawning stock
-biomass.
-
-Bias in spawning stock biomass estimation was generally low for
-estimation models that assumed the correct source of process error when
-there was lower observation error. Reliable estimation of stock-recruit
-relationship parameters only appears possible in ideal situations with
-lower observation errors in age composition and indices, lower
-variability in recruitment process errors and large contrast in spawning
-biomass over time. We found little evidence of bias for many OM process
-error assumptions when there was contrast in fishing pressure even when
-there was greater observation error although it can lead to more
-variable estimates of natural mortality. For OMs where there was bias in
-natural mortality due to high observation error, estimation the
-stock-recruit relationship seemed to remove the bias. However,
-estimation of natural mortality can cause large differences between the
-true and estimated SSB (that may be unbiased on average) when there is
-less contrast in fishing pressure over time and higher observation
-error.
-
-Retrospective patterns were generally weak for all estimation models
-regardless of the true source of process error, but they can be expected
-for recruitment even for the correct process error assumptions when
-observation error is high. When models did exhibit some retrospective
-pattern, estimation the mean natural mortality rate tended to remove it.
+\hypertarget{abstract}{%
+\subsection*{Abstract}\label{abstract}}
+\addcontentsline{toc}{subsection}{Abstract}
+
+Random effects can be included in state-space assessment models on many
+processes and in many ways, and guidance is needed on statistical
+reliability and model selection criteria. We simulated 72 operating
+models with varying fishing history, observation error uncertainty, and
+process error magnitude, correlation, and source (recruitment, survival,
+fishery selectivity, catchability, and natural mortality). We fit
+estimating models with different assumptions on the process error
+source, whether (mean) natural mortality was estimated, and whether a
+stock-recruit relationship was estimated.
+
+Models that assumed the correct process error source had high
+convergence and low bias. Bias was also low under most process error
+assumptions when there was contrast in fishing pressure.
+Stock-recruitment parameters were only reliably estimated in ideal
+situations. Marginal AIC most accurately distinguished process errors on
+recruitment, survival, and selectivity, as well as larger magnitude
+process errors of other types. Retrospective patterns were generally
+weak except for recruitment when observation error was high, even with
+the correct process error assumptions. When models did exhibit some
+retrospective pattern, estimating natural mortality often removed it.
+
+\pagebreak
 
@@
 management has expanded dramatically within ICES, Canada, and the
-Northeast US \citep{nielsenberg14,cadigan16,stockmiller21}. State-space
-approaches that use random effects to parameterize process errors is
-considered best practice and a requirement for the next generation of
-stock assessment models \citep{hoyleetal22, punt23}.
+Northeast US (Nielsen and Berg 2014; Cadigan 2016; Pedersen and Berg
+2017; Stock and Miller 2021). State-space approaches that use random
+effects to parameterize process errors is considered best practice and a
+requirement for the next generation of stock assessment models (Hoyle et
+al. 2022; Punt 2023).
 
 Much is known about the reliability of state-space models that are
-linear or Gaussian \citep{aeberhardetal18}, but applications in
+linear or Gaussian (Aeberhard et al. 2018), but applications in
 fisheries management are nonlinear and typically include multiple types
@@
 distinguish among such alternative structures.
+
+Li et al. (2024) investigated some aspects of inferences for operating
+models with multiple sources of process error, but there are differences
+for this paper.
 
@@
 We used the Woods Hole Assessment Model (WHAM) to configure operating
-and EMs in our simulation study \citep[]{millerstock20,stockmiller21}.
-WHAM is an R package freely available as a github repository. For this
-study we used
+and EMs in our simulation study (Miller and Stock 2020; Stock and Miller
+2021). WHAM is an R package freely available as a github repository. For
+this study we used
 \href{https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3}{version
-1.0.6.9000, commit 77bbd94}. This packages has also been used to
+1.0.6.9000, commit 77bbd94}. This package has also been used to
 configure operating and EMs for closed loop simulations evaluating
-index-based assessment methods \citep{legaultetal23} and is used for
+index-based assessment methods (Legault et al. 2023) and is used for
 management of haddock, butterfish, American plaice, bluefish, Atlantic
-cod, and black sea bass in the Northeast US.
+cod, black sea bass, and yellowtail flounder in the Northeast US.
 
@@
 abundance at age (R, R+S), natural mortality (R+M), fleet selectivity
-(R+Sel), or index catchability (R+q). For each OM assumptions about
+(R+Sel), or index catchability (R+q). For each OM assumption about
 variance of process errors and observations are required and the values
-we used were based on a review of the range of estimates from recent
+we used were based on a review of the range of estimates from
 applications of WHAM in management of stocks of haddock, butterfish, and
@@
 
-We did not use the bias-correction feature for process errors or
-observations described by \citet{stockmiller21} for operating and EMs.
-Simulations and model fitting were all carried out on the University of
-Massachusetts Green High-Performance Computing Cluster. All code we used
-to perform the simulation study and summarize results can be found at
+We did not use the log-normal bias-correction feature for process errors
+or observations described by (Stock and Miller 2021) for operating and
+EMs (Li et al. In review). Simulations and model fitting were all
+carried out on the University of Massachusetts Green High-Performance
+Computing Cluster. All code we used to perform the simulation study and
+summarize results can be found at
 \url{https://github.com/timjmiller/SSRTWG/tree/main/Project_0/code}.
@@
 
-The population model tracks 10 age classes: ages 1 to 10+ and we assume
-spawning occurs 1/4 of the way through the year. The maturity at age was
-a logistic curve with \(a_{50}\) = 2.89 and slope = 0.88 and assumed
-known in all EMs (Figure \ref{om_maturity}).
-
-Weight at age was generated with a LVB growth function \[
+The population consists of 10 age classes: ages 1 to 10+ and we assume
+spawning occurs each year 1/4 of the way through the year. The maturity
+at age was a logistic curve with \(a_{50}\) = 2.89 and slope = 0.88
+(Figure \ref{om_maturity}).
+
+Weight at age was generated with a von Bertalanffy growth function \[
 L_a = L_{\infty}\left(1 - e^{-k(a - t_0)}\right)
@@
 process error OMs. Therefore steepness and unfished recruitment are also
-constant over the time period for those OMs \citep{millerbrooks21}. We
+constant over the time period for those OMs (Miller and Brooks 2021). We
 specified unfished recruitment = \(R_0 = e^{10}\) and
@@
 process errors on natural mortality we assumed the rate was assumed 0.2.
-For OMs with process errors on natural mortality the mean rate was 0.2.
-
-Two alternative fishing histories were used for OMs. In the first
-scenario, the stock experiences overfishing for the first 20 years and
-fishing at \Fmsy for the last 20 years. In the second scenario, the
-stock is fished at \Fmsy for the entire time period. The magnitude of
-the overfishing assumptions is based on average estimates of overfishing
-for NE groundfish stocks from \citet{wiedenmannetal19}.
-\Citet{legaultetal23} also used similar approaches to defining fishing
-mortality histories for OMs.
-
-Initial population was configured at the equilibrium distribution
-fishing at either \(F = 2.5\times \Fmsy\) or \(F = \Fmsy\) for the two
-alternative fishing histories. That is for a deterministic model, the
-age composition would not change over time when the fishing mortality
-was constant at the respective level.
+For OMs with process errors on natural mortality the mean log natural
+mortality rate was \(\log(0.2)\).
+
+We used two fishing scenarios for OMs. In the first scenario, the stock
+experiences overfishing at 2.5\Fmsy for the first 20 years and fishing
+at \Fmsy for the last 20 years (denoted \(2.5\Fmsy \rightarrow \Fmsy\)).
+In the second scenario, the stock is fished at \Fmsy for the entire time
+period. The magnitude of the overfishing assumptions is based on average
+estimates of overfishing for NE groundfish stocks from (Wiedenmann et
+al. 2019). Legault et al. (2023) also used similar approaches to
+defining fishing mortality histories for OMs.
+
+We specified initial population abundance at age at the equilibrium
+distribution fishing at either \(F = 2.5\times \Fmsy\) or \(F = \Fmsy\)
+for the two alternative fishing histories. This implies that, for a
+deterministic model, the abundance at age would not change from the
+first year to the next.
 
@@
 constant, but we used the same alpha and beta parameters as other OMs
-this equates to a steepness and R0 at the mean of the time series
+this equates to a steepness and R0 at the median of the time series
 process for M. For OMs with time-varying random effects for fishery
@@
 than biomass for the entire 40 year period with one occurring in the
-spring (0.25 way through the year) and one in the fall (0.75 way through
-the year). Actually we have it currently configured that both occur 0.5
-way through the year. Catchability of both surveys are assumed to be
-0.1. Like the fishing fleet, we assumed logistic selectivity for both
-indices with \(a_{50} = 5\) and slope = 1 and a logistic-normal
-distribution for the age-composition observations.
+spring (0.25 of each year) and one in the fall (0.75 of each year).
+Catchability of both surveys are assumed to be 0.1. Like the fishing
+fleet, we assumed logistic selectivity for both indices with
+\(a_{50} = 5\) and slope = 1 and a logistic-normal distribution for the
+age-composition observations.
 
@@
 
-The factorial combinations of fishing histories, observation error
-assumptions, and marginal variance assumptions for recruitment and
-survival results in 24 different OMs with random effects on recruitment
-and survival (Table \ref{naa_om_table}).
+For operating models with random effects on recruitment and(or) survival
+(R, R+S) we assumed marginal standard deviations for recruitment of
+\(\sigma_R \in \{0.5,1.5\}\) and marginal standard deviations for older
+age classes of \(\sigma_{2+} \in \{0,0.25, 0.5\}\). The full factorial
+combination of these process error assumptions and fishing history (2
+levels) and observation error (2 levels) scenarios described above
+results in 24 different R (\(\sigma_{2+} = 0\)) and R+S operating models
+(Table \ref{naa_om_table}).
 
@@
 
-The factorial combinations of fishing histories, observation error
-assumptions, and marginal variance and autocorrelation assumptions for
-natural mortality results in 16 different OMs with random effects on
-natural mortality (Table \ref{M_om_table}). Internally, WHAM treats
-natural mortality as a log-transformed parameter \[
-\log M_y = \mu_M + \epsilon_{M,y}
-\] that is a linear combination of a mean \(\mu_M\) and any annual
-random effects marginally distributed as
-\(\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)\). The values
-for the marginal variance and correlation parameters assumed in the OMs
-are given in Table \ref{M_om_table}.
+All R+M OMs treat natural mortality constant across age, but with
+annually varying random effects. WHAM treats natural mortality as a
+log-transformed parameter \[
+\log M_{y,a} = \mu_{M} + \epsilon_{M,y}
+\] that is a linear combination of a mean that is constant across ages
+\(\mu_{M} = \log(0.2)\) and any annual random effects marginally
+distributed as
+\(\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)\). Uncorrelated
+random effects were also included on recruitment with \(\sigma_R = 0.5\)
+(hence, R+M). The marginal standard deviations we assumed for log
+natural mortality random effects were \(\sigma_M \in \{0.1, 0.5\}\) and
+AR1 autocorrelation parameters of \(\rho_M \in \{0,0.9\}\). The full
+factorial combination of these process error assumptions and fishing
+history (2 levels) and observation error (2 levels) scenarios described
+above results in 16 different R+M OMs (Table \ref{M_om_table}).
 
@@
 
-The factorial combinations of fishing histories, observation error
-assumptions, and marginal variance and autocorrelation assumptions for
-selectivity parameters results in 16 different OMs with random effects
-on selectivity (Table \ref{sel_om_table}). Internally, WHAM treats
-selectivity parameter \(s\) as a logit-transformed parameter \[
+MORE SPECIFICS about correlation of random effects? Both selectivity
+pars? just correlated by year? WHAM treats selectivity parameter \(s\)
+as a logit-transformed parameter \[
 \log\left(\frac{p_{s,y}-l_{s}}{u_{s}-p_{s,y}}\right) = \mu_s + \epsilon_{s,y}
@@
 slope parameters and we assume bounds of 0 and 10 for all selectivity
-parameters for all operating and EMs. The values for the marginal
-variance and correlation parameters assumed in the OMs are given in
-Table \ref{sel_om_table}.
+parameters for all operating and EMs. The marginal standard deviations
+we assumed for logit scale random effects were
+\(\sigma_s \in \{0.1, 0.5\}\) and AR1 autocorrelation parameters of
+\(\rho_s \in \{0,0.9\}\). Like R+M OMs, the full factorial combination
+of these process error assumptions and fishing history (2 levels) and
+observation error (2 levels) scenarios described above results in 16
+different R+Sel OMs (Table \ref{sel_om_table}).
 
@@
 
-The factorial combinations of fishing histories, observation error
-assumptions, and marginal variance and autocorrelation assumptions for
-catchability results in 16 different OMs with random effects on
-catchability (Table \ref{q_om_table}). Internally, WHAM treats
-catchability for index \(i\) as a logit-transformed parameter \[
+Like selectivity parameters, WHAM treats catchability for an index \(i\)
+as a logit-transformed parameter \[
 \log\left(\frac{q_{i,y}-l_{i}}{u_{i}-q_{i,y}}\right) = \mu_i + \epsilon_{i,y}
@@
 catchability, the temporal variation is only assumed for the first
-index. The values for the marginal variance and correlation parameters
-assumed in the OMs are given in Table \ref{q_om_table}
+index. The marginal standard deviations we assumed for logit scale
+random effects were \(\sigma_i \in \{0.1, 0.5\}\) and AR1
+autocorrelation parameters of \(\rho_i \in \{0,0.9\}\). Like R+M and
+R+Sel OMs, the full factorial combination of these process error
+assumptions and fishing history (2 levels) and observation error (2
+levels) scenarios described above results in 16 different R+q OMs (Table
+\ref{q_om_table}).
 
@@
 
+maturity, weight at age, index and catch CVs, assumed known in all EMs
+
 \hypertarget{measures-of-reliability}{%
@@
 \item
-  Did the optimization routine (stats::nlminb) complete without error?
+  Fit complete: Did the optimization routine (stats::nlminb) complete
+  without error?
 \item
-  Did the stats::nlminb convergence flag = 0 indicate successful
-  convergence?
+  No flag: Did the stats::nlminb convergence flag = 0 indicate
+  successful convergence?
 \item
-  Was the maximum absolute value of the gradient of the log-likelihood
-  \textless{} \(1\times10^{-6}\)?
+  Good gradient: Was the maximum absolute value of the gradient of the
+  log-likelihood \textless{} \(1\times10^{-6}\)?
 \item
-  Did TMB::sdreport provide non-NA values for all fixed effects standard
-  errors?
+  No SE NAs: Did TMB::sdreport provide non-NA values for all fixed
+  effects standard errors?
 \item
-  Did TMB::sdreport provide all standard errors \textless{} 100?
+  No big SEs: Did TMB::sdreport provide all standard errors \textless{}
+  100?
 \end{enumerate}
@@
 suggests no variation in the random effects or, equivalently, that the
-model without random effects is better.
+model without random effects is better. Note that Types 2 through 5 also
+require type 1, and type 5 requires type 4. We used the Clopper-Pearson
+exact method for constructing 95\% confidence intervals of the
+probabilities of convergence (Clopper and Pearson 1934; Thulin 2014).
 
@@
 
-We measured the frequency of correct model selection using marginal AIC.
-For a given OM we the set of models that were considered all made the
-same assumptions on whether or not to estimate M and whether a
-stock-recruit function is assumed.
-
-We also measured the frequency of correctly selecting models with
-Beverton-Holt stock recruit function assumed over models without the
-stock-recruit function (null model). Because the proportions of
-simulations where the correct model was chosen for a given set of OM
-factors implied generally poor performance of AIC, we analyzed the
-effect of variation in SSB on the probability of lower AIC of the BH
-models. We fit binomial generalize linear models to the observed AIC
-choice for each data set as a function of log(SD(SSB)) and OM.
+We estimated the probability of selection of each process error model
+structure (R, R+S, R+M, R+Sel, R+q) using marginal AIC. For a given
+operating model, we compared AIC for EMs that all made the same
+assumptions about median/constant natural mortality (known or estimated)
+and stock recruitment model (Beverton-Holt or none).
+
+We also estimated the probability of correctly selecting between models
+with Beverton-Holt stock recruit function assumed and models without the
+stock-recruit function (null model). We made these comparisons between
+models that otherwise assume the same process error structure as the
+operating model and both of the compared models either estimate
+median/constant natural mortality or assume it is known. Our preliminary
+inspections of the proportions of simulations where the correct
+recruitment model was chosen for a given set of OM factors indicated
+generally poor performance of AIC. Furthermore, it has been shown that
+estimation of stock-recruit parameters requires a wide range of SSB
+(References). Therefore, we fit logistic regression models to the
+indicator of Beverton-Holt models having lower AIC as a function of the
+log-standard deviation of the true log(SSB) (similar to the log of the
+coefficient of variation for SSB) for each simulation.
+
+All results only condition on whether all of the compared estimating
+models completed the optimization process without failure. We did not
+condition on convergence as defined by a gradient threshold or
+invertibity of the hessian because optimization can correctly determine
+the the correct likelihood that would indicate poor convergence because
+variance parameters may be at the lower bound of zero correctly for
+models that assume the incorrect process error structure.
 
@@
 constructed 95\% confidence intervals for the median relative bias using
-the binomial distribution approach as in \citet{millerhyun18} and
-\citet{stockmiller21}. We also estimated bias of stock-recruit
+the binomial distribution approach as in Miller and Hyun (2018) and
+Stock and Miller (2021). We also estimated bias of stock-recruit
 parameters for EMs that assumed the Beverton-Holt stock recruit function
 and the mean natural mortality rate parameter when estimated.
+
+Similar to the AIC results, bias results only condition on whether the
+estimating model completed the optimization process without failure. We
+did not condition on convergence as defined by a gradient threshold or
+invertibity of the hessian because the optimized model can provide
+reliable estimation of SSB, F, M, and stock recruit parameters whether
+or not the model was able to estimate non-zero random effects. In
+practice, the model would be reconfigured to remove unnecessary process
+errors and produce otherwise equivalent parameter estimates.
 
@@
 the same methods as that for relative bias.
+
+Similar to the other results, retrospective results only condition on
+whether all of the peels of a given estimating model completed the
+optimization process without failure. We did not condition on
+convergence as defined by a gradient threshold or invertibity of the
+hessian because the optimized model can provide reliable estimation of
+SSB, F, M, and stock recruit parameters whether or not the model was
+able to estimate non-zero random effects.
 
@@
 AIC generally performs weakly in determining stock recruit relationships
-unless there is large contrast in SSB. Interestingly,
-\citet{milleretal16} found AIC to prefer a stock-recruit relationship
-for SNEMA yellowtail flounder where there was a large change in stock
-size estimated over time and flatfish are well observed by the bottom
-trawl indices relative to some other guilds such as semi-pelagic
-species. Although we did not compare models with alternative
-stock-recruit relationships, we do not expect AIC to perform any better
-distinguishing these relationships. Our results comport well with those
-of \citet{devalpinehastings02} where state-space stock-recruit models
-were fit just to SSB and recruitment ``observations''.
+unless there is large contrast in SSB. Interestingly, Miller et al.
+(2016) found AIC to prefer a stock-recruit relationship for SNEMA
+yellowtail flounder where there was a large change in stock size
+estimated over time and flatfish are well observed by the bottom trawl
+indices relative to some other guilds such as semi-pelagic species.
+Although we did not compare models with alternative stock-recruit
+relationships, we do not expect AIC to perform any better distinguishing
+these relationships. Our results comport well with those of de Valpine
+and Hastings (2002) where state-space stock-recruit models were fit just
+to SSB and recruitment ``observations''.
 
@@
 
+\hypertarget{summary}{%
+\subsection*{Summary}\label{summary}}
+\addcontentsline{toc}{subsection}{Summary}
+
+We conducted a simulation study where we simulated process errors and
+observations for 72 operating models with alternative assumptions about
+fishing history, degree of uncertainty in index and age composition
+observations, type (recruitment, survival, fishery selectivity,
+catchability and natural mortality), degree of variation, and
+correlation of process errors. We fit 20 different estimation models to
+each of 100 simulated set of observations with alternative assumptions
+type and correlation structure of process errors, (mean) natural
+mortality was known or estimated, and a B-H stock recruit relationship
+was assumed or not.
+
+Across simulations we summarized probability of convergence of fitted
+models, accuracy of marginal AIC in determining the correct process
+error assumption and it ability to determine the Beverton-Holt stock
+recruit relationship, bias in annual spawning stock biomass, in
+estimation of natural mortality, and in stock-recruit relationship
+parameters, and severity of retrospective patterns for estimation
+models.
+
+Alternative measures of convergence performed differently. Invertible
+hessians and resulting standard error estimation was possible when
+criteria based on the gradient of the optimized log-likelihood with
+respect to the fixed effects parameters. Using hessian-based
+convergence, probability of convergence was best for models that assumed
+the correct source of process error, assumed M was known, and did not
+assume stock-recruit relationships.
+
+Using marginal AIC provided most accurate inferences about the process
+errors on recruitment and survival and selectivity, in that AIC
+preferred EMs with assumed process errors that matched OMs, and EMs with
+those assumed process errors were not preferred when alternative process
+errors were assumed in the OMs. However, when the true process errors
+were more variable, AIC accuracy increased to a useful level. We found
+AIC more accurately determined a B-H stock recruit relationship rather
+than the null model without a S-R relationship when there was low
+variability in recruitment, low variability in survival random effects,
+and higher variation in spawning biomass over the time series. When the
+(mean) natural mortality rate was estimated, we found large bias and
+uncertainty much more likely in model output such as spawning stock
+biomass.
+
+Bias in spawning stock biomass estimation was generally low for
+estimation models that assumed the correct source of process error when
+there was lower observation error. Reliable estimation of stock-recruit
+relationship parameters only appears possible in ideal situations with
+lower observation errors in age composition and indices, lower
+variability in recruitment process errors and large contrast in spawning
+biomass over time. We found little evidence of bias for many OM process
+error assumptions when there was contrast in fishing pressure even when
+there was greater observation error although it can lead to more
+variable estimates of natural mortality. For OMs where there was bias in
+natural mortality due to high observation error, estimation the
+stock-recruit relationship seemed to remove the bias. However,
+estimation of natural mortality can cause large differences between the
+true and estimated SSB (that may be unbiased on average) when there is
+less contrast in fishing pressure over time and higher observation
+error.
+
+Retrospective patterns were generally weak for all estimation models
+regardless of the true source of process error, but they can be expected
+for recruitment even for the correct process error assumptions when
+observation error is high. When models did exhibit some retrospective
+pattern, estimation the mean natural mortality rate tended to remove it.
+
 \hypertarget{acknowledgements}{%
@@
 
-\bibliography{paper}
+\hypertarget{references}{%
+\section*{References}\label{references}}
+\addcontentsline{toc}{section}{References}
 
 \hypertarget{refs}{}
-\begin{CSLReferences}{0}{0}
+\begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-aeberhardetal18}{}}%
+Aeberhard, W.H., Flemming, J.M., and Nielsen, A. 2018. Review of
+{State}-{Space Models} for {Fisheries Science}. Annual Review of
+Statistics and Its Application \textbf{5}(1): 215--235.
+doi:\href{https://doi.org/10.1146/annurev-statistics-031017-100427}{10.1146/annurev-statistics-031017-100427}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-cadigan16}{}}%
+Cadigan, N.G. 2016. A state-space stock assessment model for northern
+cod, including under-reported catches and variable natural mortality
+rates. Canadian Journal of Fisheries and Aquatic Sciences
+\textbf{73}(2): 296--308.
+doi:\href{https://doi.org/10.1139/cjfas-2015-0047}{10.1139/cjfas-2015-0047}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-clopperpearson34}{}}%
+Clopper, C.J., and Pearson, E.S. 1934. The use of confidence or fiducial
+limits illustrated in the case of the binomial. Biometrika
+\textbf{26}(4): 404--413.
+doi:\href{https://doi.org/10.1093/biomet/26.4.404}{10.1093/biomet/26.4.404}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-devalpinehastings02}{}}%
+de Valpine, P., and Hastings, A. 2002. Fitting population models
+incorporating process noise and observation error. Ecological Monographs
+\textbf{72}(1): 57--76.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hoyleetal22}{}}%
+Hoyle, S.D., Maunder, M.N., Punt, A.E., Mace, P.M., Devine, J.A., and
+A'mar, Z.T. 2022. Preface: Developing the next generation of stock
+assessment software. Fisheries Research \textbf{246}: 106176.
+
+\leavevmode\vadjust pre{\hypertarget{ref-legaultetal23}{}}%
+Legault, C.M., Wiedenmann, J., Deroba, J.J., Fay, G., Miller, T.J.,
+Brooks, E.N., Bell, R.J., Langan, J.A., Cournane, J.M., Jones, A.W., and
+Muffley, B. 2023. Data-rich but model-resistant: An evaluation of
+data-limited methods to manage fisheries with failed age-based stock
+assessments. Canadian Journal of Fisheries and Aquatic Sciences
+\textbf{80}(1): 27--42.
+doi:\href{https://doi.org/10.1139/cjfas-2022-0045}{10.1139/cjfas-2022-0045}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-lietalinreview}{}}%
+Li, C., Deroba, J.J., Miller, T.J., Legault, C.M., and Perretti, C. In
+reviewIn review. Guidance on bias-correction of log-normal random
+effects and observations in state-space assessment models. For Canadian
+Journal of Fisheries and Aquatic Sciences.
+
+\leavevmode\vadjust pre{\hypertarget{ref-lietal24}{}}%
+Li, C., Deroba, J.J., Miller, T.J., Legault, C.M., and Perretti, C.T.
+2024. An evaluation of common stock assessment diagnostic tools for
+choosing among state-space models with multiple random effects
+processes. Fisheries Research \textbf{273}: 106968.
+doi:\href{https://doi.org/10.1016/j.fishres.2024.106968}{10.1016/j.fishres.2024.106968}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-millerbrooks21}{}}%
+Miller, T.J., and Brooks, E.N. 2021. Steepness is a slippery slope. Fish
+and Fisheries \textbf{22}(3): 634--645.
+doi:\href{https://doi.org/10.1111/faf.12534}{10.1111/faf.12534}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-milleretal16}{}}%
+Miller, T.J., Hare, J.A., and Alade, L. 2016. A state-space approach to
+incorporating environmental effects on recruitment in an age-structured
+assessment model with an application to {S}outhern {N}ew {E}ngland
+yellowtail flounder.\cjfas \textbf{73}(8): 1261--1270.
+
+\leavevmode\vadjust pre{\hypertarget{ref-millerhyun18}{}}%
+Miller, T.J., and Hyun, S.-Y. 2018. Evaluating evidence for alternative
+natural mortality and process error assumptions using a state-space,
+age-structured assessment model.
+
+\leavevmode\vadjust pre{\hypertarget{ref-millerstock20}{}}%
+Miller, T.J., and Stock, B.C. 2020. The {Woods Hole Assessment Model}
+({WHAM}). Available from \url{https://timjmiller.github.io/wham/}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-nielsenberg14}{}}%
+Nielsen, A., and Berg, C.W. 2014. Estimation of time-varying selectivity
+in stock assessments using state-space models. Fisheries Research
+\textbf{158}: 96--101.
+doi:\href{https://doi.org/10.1016/j.fishres.2014.01.014}{10.1016/j.fishres.2014.01.014}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-pedersenberg17}{}}%
+Pedersen, M.W., and Berg, C.W. 2017. A stochastic surplus production
+model in continuous time. Fish and Fisheries \textbf{18}(2): 226--243.
+doi:\href{https://doi.org/10.1111/faf.12174}{10.1111/faf.12174}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-punt23}{}}%
+Punt, A.E. 2023. Those who fail to learn from history are condemned to
+repeat it: A perspective on current stock assessment good practices and
+the consequences of not following them. Fisheries Research \textbf{261}:
+106642.
+
+\leavevmode\vadjust pre{\hypertarget{ref-stockmiller21}{}}%
+Stock, B.C., and Miller, T.J. 2021. The {W}oods {H}ole {A}ssessment
+{M}odel ({WHAM}): {A} general state-space assessment framework that
+incorporates time- and age-varying processes via random effects and
+links to environmental covariates. Fisheries Research \textbf{240}:
+105967. doi:\url{https://doi.org/10.1016/j.fishres.2021.105967}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-thulin14}{}}%
+Thulin, M. 2014. {The cost of using exact confidence intervals for a
+binomial proportion}. Electronic Journal of Statistics \textbf{8}(1):
+817--840.
+doi:\href{https://doi.org/10.1214/14-EJS909}{10.1214/14-EJS909}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-wiedenmannetal19}{}}%
+Wiedenmann, J., Free, C.M., and Jensen, O.P. 2019. Evaluating the
+performance of data-limited methods for setting catch targets through
+application to data-rich stocks: A case study using northeast {U.S.}
+Fish stocks. Fisheries Research \textbf{209}(1): 129--142.
+doi:\url{https://doi.org/10.1016/j.fishres.2018.09.018}.
+
 \end{CSLReferences}
@@
 \pagebreak
+
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width = \textwidth]{om_input_plots_figure}
+\end{center}
+\caption{The proportion mature at age, weight at age, fleet and index selectivity at age, and Beverton-Holt stock recruit relationship assumed for the population in all operating models.
+For operating models with random effects on fleet selectivity, this represents the selectivity at the mean of the random effects.}\label{om_inputs_fig}
+\end{figure}
+
+\clearpage
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{type_4_convergence_plots}
+\end{center}
+\caption{Estimated probability of fits providing hessian-based standard errors for EMs assuming alternative process error (colored points and lines), and median natural mortality (estimated or known) and Beverton-Holt stock recruit functions (estimated or not; along x-axis) when fitted to operating models that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{hessian_SE_convergence}
+\end{figure}
+\end{landscape}
+
+\clearpage
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{pe_aic_plots}
+\end{center}
+\caption{Estimated probability of lowest AIC for EMs assuming alternative process error structures (colored bars) conditional on alternative assumptions for median natural mortality (estimated or known) and Beverton-Holt stock recruit functions (estimated or not; along x-axis) when fitted to operating models that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
+Striped bars indicate results where the EM process error structure matches that of the operating model.}\label{pe_aic}
+\end{figure}
+\end{landscape}
+
+\clearpage
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{sr_aic_plots}
+\end{center}
+\caption{Estimated probability of lowest AIC from logistic regression on the log-standard deviation of the true log(SSB) in each simulation for estimating model with Beverton-Holt stock recruit functions, rather than the otherwise equivalent EM without the stock recruit function.
+Results are conditional on alternative assumptions for median natural mortality (estimated or known) and on EMs having the correct process error structure: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Rug along x-axis denotes $SD(\log(SSB))$ values for each simulation amd polygons represent 95\% confidence intervals.}\label{sr_aic}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{term_SSB_bias_plots}
+\end{center}
+\caption{Median relative error of terminal year SSB for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Diamond shaped points denote results where the EM process error assumption matches that of the operating model.
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{SSB_rel_error}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{mohns_rho_ssb_plots}
+\end{center}
+\caption{Median Mohn's rho for SSB for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_ssb}
+\end{figure}
+\end{landscape}
+
+\pagebreak
+
+\setcounter{figure}{0}
+\renewcommand\thefigure{S\arabic{figure}}
+
+\setcounter{table}{0}
+\renewcommand\thetable{S\arabic{table}}
+
+\hypertarget{supplementary-materials}{%
+\section*{Supplementary Materials}\label{supplementary-materials}}
+\addcontentsline{toc}{section}{Supplementary Materials}
 
@@
 \caption{Distinguishing characteristics of the estimating models.}\label{em_table}
-{\scriptsize \input{em_tab}}
+{\scriptsize \input{../paper/em_tab}}
 \end{table}
 
-\begin{table}
-\caption{Operating models and estimation models all assume matching R or R+S process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}\label{naa_om_em_R_BH_aic_table}
-{\input{naa_om_em_R_BH_aic_table}}
-\end{table}
-
-\begin{table}
-\caption{Operating models and estimation models all assume matching R+M process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}\label{M_om_em_R_BH_aic_table}
-{\input{M_om_em_R_BH_aic_table}}
-\end{table}
-
-\begin{table}
-\caption{Operating models and estimation models all assume matching R+Sel process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}\label{Sel_om_em_R_BH_aic_table}
-{\input{Sel_om_em_R_BH_aic_table}}
-\end{table}
-
-\begin{table}
-\caption{Operating models and estimation models all assume matching R+q process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}\label{q_om_em_R_BH_aic_table}
-{\input{q_om_em_R_BH_aic_table}}
-\end{table}
-
 \clearpage
-
-\begin{figure}
-\caption{The proportion mature at age assumed for the population in all operating and estimating models.}\label{om_maturity}
-\begin{center}
-\includegraphics[width = \textwidth]{om_maturity.png}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\caption{The weight at age assumed for the population in all operating and estimating models.}\label{om_waa}
-\begin{center}
-\includegraphics[width = \textwidth]{om_waa.png}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\caption{The Beverton-Holt stock recruit relationship assume for all operating models.}\label{om_sr}
-\begin{center}
-\includegraphics[width = \textwidth]{om_sr.png}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\caption{The selectivity at age assumed for the fishing fleet in operating models without selectivity random effects.
-The mean selectivity at age across time for the fishing fleet in operating models with selectivity random effects.}\label{om_mean_selectivity}
-\begin{center}
-\includegraphics[width = \textwidth]{om_mean_selectivity.png}
-\end{center}
-\end{figure}
 
@@
 \begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structures R and R+S.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.}\label{naa_om_em_R_MF_convergence}
 \begin{center}
-\includegraphics[width = \textwidth]{naa_om_p_convergence_meanR_M_fixed.png}
+\includegraphics{type_3_convergence_plots}
 \end{center}
+\caption{Probability of estimating models providing maximum absolute values of gradients less than $10^{-6}$ assuming alternative process error (colored points and lines), and median natural mortality (estimated or known) and Beverton-Holt stock recruit functions (estimated or not; along x-axis) when fitted to operating models that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{gradient_convergence}
 \end{figure}
@@
 
+\clearpage
+
 \begin{landscape}
 \begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structures R and R+S.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.}\label{naa_om_em_R_ME_convergence}
 \begin{center}
-\includegraphics[width = \textwidth]{naa_om_p_convergence_meanR_M_estimated.png}
+\includegraphics{term_F_bias_plots}
 \end{center}
+\caption{Median relative error of terminal year fully-selected fishing mortality for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{F_rel_error}
 \end{figure}
@@
 \begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structures R and R+S.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is fixed at the true value.}\label{naa_om_em_BH_MF_convergence}
 \begin{center}
-\includegraphics[width = \textwidth]{naa_om_p_convergence_BH_M_fixed.png}
+\includegraphics{term_R_bias_plots}
 \end{center}
+\caption{Median relative error of terminal year recruitment for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{R_rel_error}
 \end{figure}
@@
 \begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structures R and R+S.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is estimated.}\label{naa_om_em_BH_ME_convergence}
 \begin{center}
-\includegraphics[width = \textwidth]{naa_om_p_convergence_BH_M_estimated.png}
+\includegraphics{M_bias_plots}
 \end{center}
+\caption{Median relative error of median/constant natural mortality for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{M_rel_error}
 \end{figure}
@@
 \begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+M.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.}\label{M_om_em_R_MF_convergence}
 \begin{center}
-\includegraphics[width = \textwidth]{M_om_p_convergence_meanR_M_fixed.png}
+\includegraphics{sr_bias_plots}
 \end{center}
+\caption{Median relative error of Beverton-Holt stock recruitment parameters ($a$ and $b$) for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{SR_rel_error}
 \end{figure}
@@
 \begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+M.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.}\label{M_om_em_R_ME_convergence}
 \begin{center}
-\includegraphics[width = \textwidth]{M_om_p_convergence_meanR_M_estimated.png}
+\includegraphics{mohns_rho_F_plots}
 \end{center}
+\caption{Median Mohn's $\rho$ of fishing mortality averaged over all age classes for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_F}
 \end{figure}
@@
 \begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+M.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is fixed at the true value.}\label{M_om_em_BH_MF_convergence}
 \begin{center}
-\includegraphics[width = \textwidth]{M_om_p_convergence_BH_M_fixed.png}
+\includegraphics{mohns_rho_R_plots}
 \end{center}
+\caption{Median Mohn's $\rho$ of recruitment for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_F}
 \end{figure}
@@
 
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+M.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is estimated.}\label{M_om_em_BH_ME_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_p_convergence_BH_M_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+Sel.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.}\label{Sel_om_em_R_MF_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_p_convergence_meanR_M_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+Sel.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.}\label{Sel_om_em_R_ME_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_p_convergence_meanR_M_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+Sel.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is fixed at the true value.}\label{Sel_om_em_BH_MF_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_p_convergence_BH_M_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+Sel.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is estimated.}\label{Sel_om_em_BH_ME_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_p_convergence_BH_M_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+q.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.}\label{q_om_em_R_MF_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_p_convergence_meanR_M_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+q.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.}\label{q_om_em_R_ME_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_p_convergence_meanR_M_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+q.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is fixed at the true value.}\label{q_om_em_BH_MF_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_p_convergence_BH_M_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+q.
-vertical lines represent 95\% confidence intervals.
-All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is estimated.}\label{q_om_em_BH_ME_convergence}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_p_convergence_BH_M_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R and R+S process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.} \label{naa_om_proportion_best_aic_R_MF}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_proportion_best_aic_R_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R and R+S process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.} \label{naa_om_proportion_best_aic_R_ME}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_proportion_best_aic_R_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R and R+S process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate a stock-recruit relationship and and M is fixed at the true value.} \label{naa_om_proportion_best_aic_SR_MF}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_proportion_best_aic_SR_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R and R+S process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate a stock-recruit relationship and and M is estimated.} \label{naa_om_proportion_best_aic_SR_ME}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_proportion_best_aic_SR_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+M process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.} \label{M_om_proportion_best_aic_R_MF}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_proportion_best_aic_R_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+M process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.} \label{M_om_proportion_best_aic_R_ME}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_proportion_best_aic_R_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+M process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate a stock-recruit relationship and and M is fixed at the true value.} \label{M_om_proportion_best_aic_SR_MF}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_proportion_best_aic_SR_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+M process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate a stock-recruit relationship and and M is estimated.} \label{M_om_proportion_best_aic_SR_ME}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_proportion_best_aic_SR_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+Sel process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.} \label{Sel_om_proportion_best_aic_R_MF}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_proportion_best_aic_R_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+Sel process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.} \label{Sel_om_proportion_best_aic_R_ME}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_proportion_best_aic_R_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+Sel process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate a stock-recruit relationship and and M is fixed at the true value.} \label{Sel_om_proportion_best_aic_SR_MF}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_proportion_best_aic_SR_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+Sel process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate a stock-recruit relationship and and M is estimated.} \label{Sel_om_proportion_best_aic_SR_ME}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_proportion_best_aic_SR_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+q process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.} \label{q_om_proportion_best_aic_R_MF}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_proportion_best_aic_R_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+q process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.} \label{q_om_proportion_best_aic_R_ME}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_proportion_best_aic_R_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+q process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate a stock-recruit relationship and and M is fixed at the true value.} \label{q_om_proportion_best_aic_SR_MF}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_proportion_best_aic_SR_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets from OMs with R+q process errors where fitted estimating models had lowest marginal AIC.
-All estimating models estimate a stock-recruit relationship and and M is estimated.} \label{q_om_proportion_best_aic_SR_ME}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_proportion_best_aic_SR_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\clearpage
-
-\begin{figure}
-\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
-Operating and estimating models have matching R or R+S process error structures.
-Estimating models assume mean M is fixed at the true value.}\label{naa_om_MF_BH_glm_AIC_plots}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_MF_pred_BH_best.png}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
-Operating and estimating models have matching R or R+S process error structures.
-Estimating models allow estimation of mean M.}\label{naa_om_ME_BH_glm_AIC_plots}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_ME_pred_BH_best.png}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
-Operating and estimating models have matching R+M process error structures.
-Estimating models assume mean M is fixed at the true value.}\label{M_om_MF_BH_glm_AIC_plots}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_MF_pred_BH_best.png}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
-Operating and estimating models have matching R+M process error structures.
-Estimating models allow estimation of mean M.}\label{M_om_ME_BH_glm_AIC_plots}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_ME_pred_BH_best.png}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
-Operating and estimating models have matching R+Sel process error structures.
-Estimating models assume mean M is fixed at the true value.}\label{Sel_om_MF_BH_glm_AIC_plots}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_MF_pred_BH_best.png}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
-Operating and estimating models have matching R+Sel process error structures.
-Estimating models allow estimation of mean M.}\label{Sel_om_ME_BH_glm_AIC_plots}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_ME_pred_BH_best.png}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
-Operating and estimating models have matching R+q process error structures.
-Estimating models assume mean M is fixed at the true value.}\label{q_om_MF_BH_glm_AIC_plots}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_MF_pred_BH_best.png}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\caption{Predicted probability of AIC preferring BH model as a function of the variation of population SSB.
-Operating and estimating models have matching R+q process error structures.
-Estimating models allow estimation of mean M.}\label{q_om_ME_BH_glm_AIC_plots}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_ME_pred_BH_best.png}
-\end{center}
-\end{figure}
-
-\clearpage
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R and R+S process errors.
-Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{naa_om_em_R_MF_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_R_MF_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R and R+S process errors.
-Estimation models do not assume a stock-recruit relationship and estimate M.}\label{naa_om_em_R_ME_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_R_ME_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R and R+S process errors.
-Estimation models assume a BH stock-recruitment function and M is fixed at the true value.}\label{naa_om_em_BH_MF_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_BH_MF_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R and R+S process errors.
-Estimation models assume a BH stock-recruitment function and estimate M.}\label{naa_om_em_BH_ME_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_BH_ME_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+M process errors.
-Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{M_om_em_R_MF_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_R_MF_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+M process errors.
-Estimation models do not assume a stock-recruit relationship and estimate M.}\label{M_om_em_R_ME_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_R_ME_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+M process errors.
-Estimation models assume a BH stock-recruitment function and M is fixed at the true value.}\label{M_om_em_BH_MF_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_BH_MF_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+M process errors.
-Estimation models assume a BH stock-recruitment function and estimate M.}\label{M_om_em_BH_ME_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_BH_ME_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+Sel process errors.
-Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{Sel_om_em_R_MF_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_R_MF_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+Sel process errors.
-Estimation models do not assume a stock-recruit relationship and estimate M.}\label{Sel_om_em_R_ME_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_R_ME_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+Sel process errors.
-Estimation models assume a BH stock-recruitment function and M is fixed at the true value.}\label{Sel_om_em_BH_MF_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_BH_MF_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+Sel process errors.
-Estimation models assume a BH stock-recruitment function and estimate M.}\label{Sel_om_em_BH_ME_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_BH_ME_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+q process errors.
-Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{q_om_em_R_MF_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_R_MF_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+q process errors.
-Estimation models do not assume a stock-recruit relationship and estimate M.}\label{q_om_em_R_ME_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_R_ME_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+q process errors.
-Estimation models assume a BH stock-recruitment function and M is fixed at the true value.}\label{q_om_em_BH_MF_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_BH_MF_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative bias of SSB for estimating models fitted to data sets simulated with R+q process errors.
-Estimation models assume a BH stock-recruitment function and estimate M.}\label{q_om_em_BH_ME_relbias_ssb}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_BH_ME_relbias_ssb.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of Beverton-Holt stock recruit parameters for estimating models with correct process error assumptions fitted to data sets simulated with R and R+S process errors.
-Estimation models either assume the correct natural mortality rate or estimate it.}\label{naa_om_SR_relbias}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_SR_relerror.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of Beverton-Holt stock recruit parameters for estimating models with correct process error assumptions fitted to data sets simulated with R+M process errors.
-Estimation models either assume the correct natural mortality rate or estimate it.}\label{M_om_SR_relbias}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_SR_relerror.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of Beverton-Holt stock recruit parameters for estimating models with correct process error assumptions fitted to data sets simulated with R+Sel process errors.
-Estimation models either assume the correct natural mortality rate or estimate it.}\label{Sel_om_SR_relbias}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_SR_relerror.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of Beverton-Holt stock recruit parameters for estimating models with correct process error assumptions fitted to data sets simulated with R+q process errors.
-Estimation models either assume the correct natural mortality rate or estimate it.}\label{q_om_SR_relbias}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_SR_relerror.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of (mean) natural mortality rate for estimating models with correct process error assumptions fitted to data sets simulated with R and R+S process errors.
-Estimation models assumed either no stock-recruit relationship or estimated a Beverton-Holt relationship.}\label{naa_om_M_relbias}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_M_relerror.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of (mean) natural mortality rate for estimating models with correct process error assumptions fitted to data sets simulated with R+M process errors.
-Estimation models assumed either no stock-recruit relationship or estimated a Beverton-Holt relationship.}\label{M_om_M_relbias}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_M_relerror.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of (mean) natural mortality rate for estimating models with correct process error assumptions fitted to data sets simulated with R+Sel process errors.
-Estimation models assumed either no stock-recruit relationship or estimated a Beverton-Holt relationship.}\label{Sel_om_M_relbias}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_M_relerror.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of (mean) natural mortality rate for estimating models with correct process error assumptions fitted to data sets simulated with R+q process errors.
-Estimation models assumed either no stock-recruit relationship or estimated a Beverton-Holt relationship.}\label{q_om_M_relbias}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_M_relerror.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R and R+S process errors.
-Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{naa_om_em_R_MF_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_mohns_rho_R_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R and R+S process errors.
-Estimation models do not assume a stock-recruit function and M estimated.}\label{naa_om_em_R_ME_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_mohns_rho_R_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R and R+S process errors.
-Estimation models assume a B-H stock-recruit function and M is fixed at the true value.}\label{naa_om_em_BH_MF_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_mohns_rho_BH_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R and R+S process errors.
-Estimation models assume a B-H stock-recruit function and M estimated.}\label{naa_om_em_BH_ME_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{naa_om_mohns_rho_BH_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+M process errors.
-Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{M_om_em_R_MF_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_mohns_rho_R_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+M process errors.
-Estimation models do not assume a stock-recruit function and M estimated.}\label{M_om_em_R_ME_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_mohns_rho_R_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+M process errors.
-Estimation models assume a B-H stock-recruit function and M is fixed at the true value.}\label{M_om_em_BH_MF_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_mohns_rho_BH_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+M process errors.
-Estimation models assume a B-H stock-recruit function and M estimated.}\label{M_om_em_BH_ME_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{M_om_mohns_rho_BH_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+Sel process errors.
-Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{Sel_om_em_R_MF_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_mohns_rho_R_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+Sel process errors.
-Estimation models do not assume a stock-recruit function and M estimated.}\label{Sel_om_em_R_ME_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_mohns_rho_R_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+Sel process errors.
-Estimation models assume a B-H stock-recruit function and M is fixed at the true value.}\label{Sel_om_em_BH_MF_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_mohns_rho_BH_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+Sel process errors.
-Estimation models assume a B-H stock-recruit function and M estimated.}\label{Sel_om_em_BH_ME_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{Sel_om_mohns_rho_BH_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+q process errors.
-Estimation models do not assume a stock-recruit function and M is fixed at the true value.}\label{q_om_em_R_MF_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_mohns_rho_R_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+q process errors.
-Estimation models do not assume a stock-recruit function and M estimated.}\label{q_om_em_R_ME_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_mohns_rho_R_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+q process errors.
-Estimation models assume a B-H stock-recruit function and M is fixed at the true value.}\label{q_om_em_BH_MF_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_mohns_rho_BH_MF.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median Mohn's $\rho$ for SSB, fully-selected F, and recruitment for estimating models fitted to data sets simulated with R+q process errors.
-Estimation models assume a B-H stock-recruit function and M estimated.}\label{q_om_em_BH_ME_mohns_rho}
-\begin{center}
-\includegraphics[width = \textwidth]{q_om_mohns_rho_BH_ME.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
 \end{document}

### 2025-02-12 Project_0/manuscript/manuscript.Rmd
saving work on project_0 manuscript

@@
 We simulated 72 operating models with varying fishing history, observation error uncertainty, and process error magnitude, correlation, and source (recruitment, survival, fishery selectivity, catchability, and natural mortality).
-We fit estimating models with different assumptions on the process error source, whether (mean) natural mortality was estimated, and whether a stock-recruit relationship was estimated.
+We fit estimating models with different assumptions on the process error source, whether median natural mortality was estimated, and whether a stock-recruit relationship was estimated.
 
@@
 Retrospective patterns were generally weak except for recruitment when observation error was high, even with the correct process error assumptions.
-When models did exhibit some retrospective pattern, estimating natural mortality often removed it.
+When models did exhibit some retrospective pattern, estimating median natural mortality often removed it.
 
@@
 We fitted 20 EMs to observations from each of 100 simulations where process errors were also simulated.
-EMs made alternative assumptions about the source of process errors and whether natural mortality (or the mean for models with process error in natural mortality) was estimated and whether a Beverton-Holt stock recruit relationship was estimated within the EM.
+For R+M, R+Sel, and R+q OMs, we used unique seeds for each simulation, but we inadvertently used the same 100 seeds for all R and R+S OMs.
+Each of the EMs made alternative assumptions about the source of process errors and whether natural mortality (or the median for models with process error in natural mortality) was estimated and whether a Beverton-Holt stock recruit relationship was estimated within the EM.
 Details of each of the operating and EMs are described below.
@@
 The population consists of 10 age classes: ages 1 to 10+ and we assume spawning occurs each year 1/4 of the way through the year.
-The maturity at age was a logistic curve with $a_{50}$ = 2.89 and slope = 0.88 (Figure \ref{om_maturity}).
+The maturity at age was a logistic curve with $a_{50}$ = 2.89 and slope = 0.88 (Figure \ref{om_inputs_fig}, top left).
 
@@
 $$
-where $\theta_1 = e^{-12.1}$ and $\theta_2 = 3.2$ (Figure \ref{om_waa}).
+where $\theta_1 = e^{-12.1}$ and $\theta_2 = 3.2$ (Figure \ref{om_inputs_fig}, top right).
 
@@
 Therefore steepness and unfished recruitment are also constant over the time period for those OMs [@millerbrooks21].
-We specified unfished recruitment = $R_0 = e^{10}$ and $\Fmsy = F_{40\%} = 0.348$ equated to a steepness of 0.69 and $\alpha=0.60$ and $\beta = 2.4 \times 10^{-5}$ for the
+We specified unfished recruitment = $R_0 = e^{10}$ and $\Fmsy = F_{40\%} = 0.348$ equated to a steepness of 0.69 and $a=0.60$ and $b = 2.4 \times 10^{-5}$ for the
 \[
-N_{1,y} = \frac{\alpha \text{SSB}_{y-1}}{1 + \beta \text{SSB}_{y-1}} 
+N_{1,y} = \frac{a \text{SSB}_{y-1}}{1 + b \text{SSB}_{y-1}} 
 \]
-Beverton-Holt parameterization (Figure \ref{om_sr}).
+Beverton-Holt parameterization (Figure \ref{om_inputs_fig}, bottom right).
 For OMs without process errors on natural mortality we assumed the rate was assumed 0.2.
-For OMs with process errors on natural mortality the mean log natural mortality rate was $\log(0.2)$.
+For OMs with process errors on natural mortality the median natural mortality rate was 0.2.
 
@@
 
-We assumed a single fleet operating year round for catch observations with logistic selectivity for the fleet with $a_{50} = 5$ and slope = 1 (Figure \ref{om_mean_selectivity}).
+We assumed a single fleet operating year round for catch observations with logistic selectivity for the fleet with $a_{50} = 5$ and slope = 1 (Figure \ref{om_inputs_fig}, bottom left).
 This selectivity is was used to define \Fmsy for the Beverton-Holt stock recruitment parameters above.
-We assumed a logistic-normal distribution for the age-composition observations for the fleet.
+We assumed a logistic-normal distribution with no correlation on the multivariate normal scale for the age-composition observations for the fleet.
 
@@
 Catchability of both surveys are assumed to be 0.1.
-Like the fishing fleet, we assumed logistic selectivity for both indices with $a_{50} = 5$ and slope = 1 and a logistic-normal distribution for the age-composition observations.
+Like the fishing fleet, we assumed logistic selectivity for both indices with $a_{50} = 5$ and slope = 1 and a logistic-normal distribution with no correlation on the multivariate normal scale for the age-composition observations.
 
@@
 $$
-that is a linear combination of a mean that is constant across ages $\mu_{M} = \log(0.2)$ and any annual random effects marginally distributed as $\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)$.
+that is a linear combination of a mean log-natural mortality parameter that is constant across ages $\mu_{M} = \log(0.2)$ and any annual random effects marginally distributed as $\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)$.
 Uncorrelated random effects were also included on recruitment with $\sigma_R = 0.5$ (hence, R+M).
@@
 For each data set simulated from an OM 20 EMs were fit.
-A total of 32 different EMs were fit across all OMs where the subset of 20 depended on the source of process error (Table \ref{em_table}).
+A total of 32 different EMs were fit across OMs where the subset of 20 depended on the source of process error in the OM (Table \ref{em_table}).
+The EMs make different assumptions about the source of process error (R+S, R+M, R+Sel, R+q) and whether there is temporal autocorrelation or not, and whether a Beverton-Holt stock recruit function is estimated and the natural mortality rate ($\mu_M$, the constant or mean on log scale for R+M EMs) is estimated.
+For simplicity we refer to the derived estimate $e^{\mu_M}$ as the median natural mortality rate regardless of whether natural mortality random effects are estimated in the EM.
 The first 20 EMs in Table \ref{em_table} were fit to simulate data sets from R and R+S OMs.
@@
 Finally, EMs 5 to 20 and 29-32 in Table \ref{em_table} were fit to simulate data sets from R+q OMs.
-The observation error variance of aggregate catch and indices were all assumed known at the true values.
-
-maturity, weight at age, index and catch CVs, assumed known in all EMs
+The maturity at age, weight at age for catch and SSB, and observation error variance of aggregate catch and indices were all assumed known at the true values.
 
@@
 The first measure of reliability we investigated was frequency of convergence when fitting each EM to the simulated data sets.
-There are various ways to assess convergence of the fit. We summarized 5 alternative categories of convergence.
+There are various ways to assess convergence of the fit [e.g., @carvalhoetal21], but given the importance of estimates of uncertainty when using assessment models in management, we estimated probablity of convergence as measured by occurrence of a positive-definite hessian matrix at the optimized negative log-likelihood that could be inverted.
+However, we also provided results in the Supplementary Materials for the maximum of the absolute values among all gradients for all fits of a given EM to all simulated data sets from a given OM that produced hessian-based standard errors for all estimated fixed effects.
+This provides an indication of how poor the calculated gradients can be, but still presumably converged adequately enough for parameter inferences.
 
-1. Fit complete: Did the optimization routine (stats::nlminb) complete without error?
-2. No flag: Did the stats::nlminb convergence flag = 0 indicate successful convergence?
-3. Good gradient: Was the maximum absolute value of the gradient of the log-likelihood < $1\times10^{-6}$?
-4. No SE NAs: Did TMB::sdreport provide non-NA values for all fixed effects standard errors?
-5. No big SEs: Did TMB::sdreport provide all standard errors < 100?
-
-The first convergence criterion assesses whether the model crashes.
-The third is a measure that assesses how flat the likelihood is at the optimized point of the likelihood surface.
-The fourth and fifth criteria are specific to the hessian-based standard error reporting provided by TMB.
-the TMB::sdreport function will sometimes return standard error estimates even when the calculated hessian is not invertible (fourth criterion).
-It will also provide large standard errors for some parameters that are not estimated well or are near bounds on the transformed scale that is of primary interest.
-For example, variance parameters for random effects can often be estimated near 0 when the fit suggests no variation in the random effects or, equivalently, that the model without random effects is better.
-Note that Types 2 through 5 also require type 1, and type 5 requires type 4.
+<!--
+probability of convergence as measured by all gradient values at the optimized negative log-likelihood that were less than $1\times10^{-6}$ in absolute value.
 We used the Clopper-Pearson exact method for constructing 95\% confidence intervals of the probabilities of convergence [@clopperpearson34;@thulin14].
+-->
 
@@
 We estimated the probability of selection of each process error model structure (R, R+S, R+M, R+Sel, R+q) using marginal AIC.
-For a given operating model, we compared AIC for EMs that all made the same assumptions about median/constant natural mortality (known or estimated) and stock recruitment model (Beverton-Holt or none).
+For a given operating model, we compared AIC for EMs that all made the same assumptions about median natural mortality (known or estimated) and stock recruitment model (Beverton-Holt or none).
 
 We also estimated the probability of correctly selecting between models with Beverton-Holt stock recruit function assumed and models without the stock-recruit function (null model).
-We made these comparisons between models that otherwise assume the same process error structure as the operating model and both of the compared models either estimate median/constant natural mortality or assume it is known.
-Our preliminary inspections of the proportions of simulations where the correct recruitment model was chosen for a given set of OM factors indicated generally poor performance of AIC.
-Furthermore, it has been shown that estimation of stock-recruit parameters requires a wide range of SSB (References).
-Therefore, we fit logistic regression models to the indicator of Beverton-Holt models having lower AIC as a function of the log-standard deviation of the true log(SSB) (similar to the log of the coefficient of variation for SSB) for each simulation.
+We made these comparisons between models that otherwise assume the same process error structure as the operating model and both of the compared models either estimate median natural mortality or assume it is known.
+Contrast in fishing pressure and time series with recruitment at low stock size has been shown to improve estimability of stock-recruitment parameters [@magnussonhilborn07;@connetal10].
+Our preliminary inspections of the proportions of simulations where the correct recruitment model was chosen for a given set of OM factors (including contrast in fishing pressure) indicated generally poor performance of AIC.
+Therefore, we fit logistic regression models to the indicator of Beverton-Holt models having lower AIC as a function of the log-standard deviation of the true log(SSB) (similar to the log of the coefficient of variation for SSB) since simulations with realized SSB producing low and high recruitments would have larger variation in realized SSB.
 
@@
 
-We estimated bias as the median of the relative errors across all simulations for a given OM and EM combination.
+For a given model attribute we calculated the relative error 
 $$
@@
 $$
-of annual SSB and fully-selected fishing mortality for each EM and constructed 95\% confidence intervals for the median relative bias using the binomial distribution approach as in @millerhyun18 and @stockmiller21.
-We also estimated bias of stock-recruit parameters for EMs that assumed the Beverton-Holt stock recruit function and the mean natural mortality rate parameter when estimated.
+from fitting a given estimating model to simulated data set $i$ configured for a given OM where $\widehat \theta_i$ and $\theta_i$ are the estimated and true values for simulation $i$.
+We estimated bias as the median of the relative errors across all simulations for a given OM and EM combination.
+We constructed 95\% confidence intervals for the median relative bias using the binomial distribution approach as in @millerhyun18 and @stockmiller21.
+We present results for terminal years SSB and recruitment, Beverton-Holt stock recruit parameters ($a$ and $b$), and median natural mortality rate.
+Results for terminal year fishing mortality are strongly negatively correlated with those for SSB and provided in the supplementary materials.
 
@@
 
-We estimated Mohn's $\rho$ for SSB, fully-selected fishing mortality, and recruitment for each EM.
+We estimated Mohn's $\rho$ for SSB, fully-selected fishing mortality, and recruitment for each EM [@mohn99].
 We estimated 7 peels for each EM.
 We calculated median 95\% confidence intervals for Mohn's $\rho$ using the same methods as that for relative bias.
-
 Similar to the other results, retrospective results only condition on whether all of the peels of a given estimating model completed the optimization process without failure.
-We did not condition on convergence as defined by a gradient threshold or invertibity of the hessian because the optimized model can provide reliable estimation of SSB, F, M, and stock recruit parameters whether or not the model was able to estimate non-zero random effects.
+We did not condition on convergence as defined by a gradient threshold or invertibity of the hessian because the optimized model can provide reliable estimation of SSB, fishing mortality rates, and recruitment whether or not the model was able to estimate non-zero random effects.
 
@@
 
-### R, R+S operating models {-}
+For R and R+S OMs, convergence generally declines for most EMs when the median natural mortality rate is estimated and/or the Beverton-Holt stock recruit relationship is estimated even when the process error assumptions of the EMs and OMs match (Figure \ref{hessian_SE_convergence}, A).
+When there was high observation error and constant fishing pressure (change from $2.5F_{\text{MSY}}$ to $F_{\text{MSY}}$), convergence of all EM process error configurations other than R EMs was poor for R OMs ($\sigma_{2+} = 0$) regardless of whether median natural mortality and stock-recruit relationships were estimated.
+Convergence of R EMs was high for all R and R+S OMs except when there was high observation error and constant fishing pressure and median natural mortality and stock-recruit relationships were estimated.
+R+S EMs fit to R OMs exhibited poor convergence regardless of whether natural mortality or a stock-recruit relationship was estimated.
+R+S EMs fit to R+S OMs had highest convergence rates when there was contrast in fishing pressure and low observation error.
+Convergence rates were high for all EMs when fit to data from R+S OMs with lower observation uncertainty except those where median natural mortality and/or stock-recruit relationships were estimated.
 
-R+S EMs fit to R OMs exhibited poor convergence for types 3, 4, and 5 regardless of whether M or a stock-recruit relationship was estimated (Figures \ref{naa_om_em_R_MF_convergence} to \ref{naa_om_em_BH_ME_convergence}).
-R EMs exhibited high convergence rates for all convergence types for both R and R+S OMs.
-Convergence rates were high for all convergence types for all EMs in R+S OMs with lower uncertainty in observations.
-Convergence for most convergence types generally declines most EMs when mean-log natural mortality rate is estimated and/or a Beverton-Holt stock recruit relationship is estimated even when the process error assumptions of the estimation and OMs match.
-There can be relatively high convergence probability of type 4 (an invertible hessian) for EMs that do not have the correct process error assumed, but type 5 convergence (no very large standard errors estimated from hessian) for these EMs is typically much lower.
+Convergence of all EMs fitted to R+M OMs was best when the OMs had higher natural mortality process error variability, low observation error, and contrast in fishing pressure  (Figure \ref{hessian_SE_convergence}, B).
+R+M EMs that estimated autocorrelation of process errors had poor convergence for R+M OMs when there was low natural mortality process error variability whether or not there was autocorrelation of the simulated process errors.
+R+S EMs fitted to data generated from R+M OMs always converged poorly whether median natural mortality and the Beverton-Holt stock-recruit relationship were estimated.
 
-### R+M operating models {-}
+R+S EMs, in particular, converged poorly when fit to data generated from R+Sel OMs with lower selectivity process error variability or higher observation uncertainty (Figure \ref{hessian_SE_convergence}, C).
+R+Sel EMs generally converged better than other EMs for R+Sel OMs with higher process error variability, lower observation uncertainty, and contrast in fishing pressure  regardless of whether median natural mortality or a stock recruit relationship was estimated.
 
-R+S EMs generally converged less reliably across convergence types than other EMs when fit to data generated from R+M OMs (Figures \ref{M_om_em_R_MF_convergence} to \ref{M_om_em_BH_ME_convergence}).
-However, using convergence type 5, all EMs converged poorly when OMs had low variability in natural mortality process errors or higher observation error uncertainty.
-Using convergence type 5, R+M EMs generally converged better than other EMs when observation error uncertainty was low and natural mortality process errors were more variable.
-Like R+S OMs, the probability of convergence generally declined for all EMs when mean log-natural mortality and/or a stock-recruit relationship was estimated.
+For R+q OMs, convergence of R+q EMs is generally better than that of other EMs when there is contrast in fishing history (Figure (\ref{hessian_SE_convergence}, D).
+Convergence of R+S EMs is generally worse than that of all other EMs across all OMs whether or not median natural mortality or a stock recruit relationship is estimated.
+Again, convergence probability generally declines for all EMs when median natural mortality or a stock recruit relationship is estimated.
 
-### R+Sel operating models {-}
-
-R+S EMs in particular converged poorly when R+Sel OMs had less variability in selectivity process errors or higher observation error uncertainty (Figures \ref{Sel_om_em_R_MF_convergence} to \ref{Sel_om_em_BH_ME_convergence}).
-R+Sel EMs generally converged better than other EMs for OMs with greater variability in process errors, lower observation error, and contrast in fishing pressure regardless of whether mean log natural mortality or a stock recruit relationship was estimated.
-
-### R+q operating models {-}
-
-Convergence of R+q EMs is generally better than that of other EMs for all convergence types when R+q OMs assume contrast in fishing history overFigures (\ref{q_om_em_R_MF_convergence} to \ref{q_om_em_BH_ME_convergence}).
-Convergence of R+S EMs is generally worse than that of other EMs across all OMs whether or not mean log-natural mortality or a stock recruit relationship is estimated.
-Again, convergence probablity generally declines for all EMs when mean log-natural mortality or a stock recruit relationship is estimated.
+We found a wide range of maximum absolute values of gradients for models that converged as defined here (Figure \ref{hess_grad}).
+The largest value observed for a given EM and OM combination was typically $<10^{-3}$, but many converged models had values greater than 1.
+For many OMs, EMs that assumed the correct process error type and did not estimate median natural mortality or the Beverton-Holt stock-recruit relationship produced the lowest gradient values.
 
@@
 
-### R, R+S operating models {-}
+Marginal AIC accurately determines the correct process error assumptions in EMs when data are generated from R and R+S OMs, regardless of whether median natural mortality or a stock recruit relationship is estimated (Figure \ref{pe_aic}, A).
+Adding estimation of median natural mortality or a stock recruit relationship separately has a negligible effect on the accuracy of determining the correct process error assumption.
+When both are estimated, there is a noticeable reduction in accuracy when OMs have a constant fishing history, observation uncertainty is low and larger variability in recruitment process errors.
 
-Marginal AIC accurately determines the correct process error assumptions in EMs when data are generated from R and R+S OMs, regardless of whether mean log-natural mortality or a stock recruit relationship is estimated (Figures \ref{naa_om_proportion_best_aic_R_MF} to \ref{naa_om_proportion_best_aic_SR_ME}).
-Adding estimation of mean log-natural mortality or a stock recruit relationship separately has a negligible effect on the accuracy of determining the correct process error assumption.
-When both are estimated, there is a noticeable reduction in accuracy when OMs have a constant fishing history, observation error is low and largest variability in recruitment process errors.
-
-### R+M operating models {-}
-
-Marginal AIC only accurately determined the correct process error model and correlation structure when observation error was low and variability in natural mortality process errors was high (Figures \ref{M_om_proportion_best_aic_R_MF} to \ref{M_om_proportion_best_aic_SR_ME}).
-Estimating the mean natural mortality rate reduced the accuracy of AIC for OMs that assumed natural mortality process errors were independent.
-For OMs with poor accuracy, AIC most frequently selected EMs with process errors in catchability or selectivity.
+For R+M OMs, marginal AIC only accurately determined the correct process error model and correlation structure when observation error was low and variability in natural mortality process errors was high (Figures \ref{pe_aic}, B).
+Estimating the median natural mortality rate reduced the accuracy of AIC for OMs that assumed natural mortality process errors were independent.
+For OMs with poor accuracy, AIC most frequently selected EMs with process errors in catchability (R+q) or selectivity (R+Sel).
 Selection of R+S EMs was typically unlikely.
 
-### R+Sel operating models {-}
+Marginal AIC most accurately determined the correct source of process error and correlation structure for R+Sel OMs with low observation uncertainty (Figures \ref{pe_aic}, C).
+When there was low variability in selectivity process errors and high observation uncertainty, R+q or R+S EMs were more likely to have the best AIC.
+Whether median natural morality or stock recruit relationships were estimated appeared to have little effect on the performance of AIC.
 
-Marginal AIC most accurately determined the correct source of process error and correlation structure for R+Sel OMs with low observation error (Figures \ref{Sel_om_proportion_best_aic_R_MF} to \ref{Sel_om_proportion_best_aic_SR_ME}).
-When there was low variability in selectivity process errors and high observation error, R+q or R+S EMs were more likely to have the best AIC.
-Whether mean log-natural morality or stock recruit relationships were estimated appeared to have little effect on the performance of marginal AIC.
-
-### R+q operating models {-}
-
-Marginal AIC most accurately determined the correct source of process error and correlation structure for R+q OMs with high variability in catchability process errors (Figures \ref{q_om_proportion_best_aic_R_MF} to \ref{q_om_proportion_best_aic_SR_ME}).
-The worst accuracy occurred for OMs with low variability in catchability process errors and high observation error.
-However, in these OMs, the marginal AIC accurately determined the correct source of process error (but not correlation structure) except when EMs estimated both mean log-natural morality and the stock recruit relationships and OMs assumed a constant fishing pressure.
+Marginal AIC most accurately determined the correct source of process error and correlation structure for R+q OMs with high variability in catchability process errors (Figures \ref{pe_aic},D).
+The worst accuracy occurred for OMs with low variability in catchability process errors and high observation uncertainty.
+However, in these OMs, the marginal AIC accurately determined the correct source of process error (but not correlation structure) except when EMs estimated both median natural morality and the stock recruit relationships and OMs assumed a constant fishing pressure.
 
@@
 
-<!--
-Tables of frequencies where AIC chooses correct stock-recruit assumption imply poor performance of AIC, but figures showing effect of variation in SSB demonstrate that large variation in SSB is needed to reduce Type II errors for marginal AIC.
--->
-Our comparisons of model performance condition on assuming the true process error configuration is known (EM and OM process error assumptions match).
-Broadly, we found generally poor accuracy of AIC in selecting models assuming a Beverton-Holt stock recruit function over the null model without an assumed stock-recruit relationship for all OMs (Tables \ref{naa_om_em_R_BH_aic_table} to \ref{q_om_em_R_BH_aic_table}.
-However, we also found AIC to more accurately determine a stock-recruit relationship when there was greater variation in spawning biomass generated in the simulated populations.
+Our comparisons of model performance condition on assuming the true process error configuration is known (EM and OM process error types match).
+Broadly, we found generally poor accuracy of AIC in selecting models assuming a Beverton-Holt stock recruit function over the null model without an assumed stock-recruit relationship for all OMs.
+However, we also found increased accuracy of AIC in determining the Beverton-Holt stock-recruit relationship with greater variation in spawning biomass generated in the simulated populations for nearly every OM (Figure \ref{sr_aic}).
 
-### R, R+S operating models {-}
+With R and R+S process error assumptions, probability of AIC accuracy for the B-H stock-recruit relationship as a function of SSB variability were greatest for OMs where there was greater process variability in survival, lower process variability in recruitment and
+contrast in fishing pressure (Figure \ref{sr_aic}, A).
+However, the largest variation in SSB (Figure \ref{sr_aic}, rug on x-axis) occurred in OMs with larger recruitment variability ($\sigma_R = 1.5$; Figure \ref{sr_aic}, A, right column group).
+The weakest overall accuracy of AIC for the Beverton-Holt stock-recruit function was when there was constant fishing pressure (Figure \ref{sr_aic}, A, second and fourth rows).
+The largest effect of simultaneously estimating median natural mortality was when there was OMs had high observation uncertainty and constant fishing pressure (Figure \ref{sr_aic}, A, second row).
 
-Among R and R+S OMs, AIC accuracy for including the B-H stock-recruit relationship was estimated to be highest for R OMs and EMs and lowest variance of recruitment process errors (Figures \ref{naa_om_MF_BH_glm_AIC_plots} and \ref{naa_om_ME_BH_glm_AIC_plots}).
-FOr R+S OMs and holding variability in SSB constant, accuracy of stock-recruit  model selection declined with greater variance of recruitment and survival process errors.
-Simultaneously estimating mean log-natural mortality has a small negative effect on AIC accuracy, but there is an improvement in AIC accuracy with lower observation error in indices and age composition and lower variation of survival process errors.
+For R+M OMs, probability of AIC accuracy for the Beverton-Holt stock-recruit relationship was greatest for OMs where there was contrast in fishing pressure (Figure \ref{sr_aic}, B).
+The largest variation in SSB occurred in OMs with larger natural mortality variability and temporal AR1 correlation of the annual random effects ($\sigma_M = 1.5$, $\rho_M = 0.9$; Figure \ref{sr_aic}, B, right-most column) where AIC accuracy was also high even when there was constant fishing pressure.
+Interestingly, there seemed to be little difference in AIC accuracy whether there was lower or higher observation uncertainty or whether the EM simultaneously estimated median natural mortality.
 
-### R+M operating models {-}
+For R+Sel OMs, contrast in fishing pressure over time is the primary mechanism creating variation in SSB and these are the OMs where AIC accuracy for the Beverton-Holt stock recruit relationship was greatest (Figure \ref{sr_aic}, C, first and third rows).
+There appears to be little effect of variability or correlation of selectivity process errors or whether median natural mortality was estimated on AIC accuracy.
 
-For R+M OMs and EMs there appeared to be little effect of correlation in the natural mortality process errors or estimating the mean log-natural mortality on accuracy of AIC in determining a stock-recruit function, holding variability in SSB constant (Figures \ref{M_om_MF_BH_glm_AIC_plots} and \ref{M_om_ME_BH_glm_AIC_plots}).
-However, there was an increase in accuracy with lower observation error for OMs with higher variability in natural mortality process errors whether the mean natural mortality rate parameter was estimated or not.
-
-### R+Sel operating models {-}
-
-The range of variation in SSB across simulated populations was lower for R+Sel OMs than R+S or R+M OMs.
-OMs with contrast in fishing pressure over time is the primary mechanism creating variation in SSB.
-However, there appears to be little effect of variability or correlation of selectivity process errors or whether mean log-natural mortality was estimated on the accuracy of AIC in selecting the stock recruit relationship (Figures \ref{Sel_om_MF_BH_glm_AIC_plots} and \ref{Sel_om_ME_BH_glm_AIC_plots}).
-However, there was an increase in AIC accuracy with lower observation error for OMs with independent process errors whether the mean natural mortality rate parameter was estimated or not.
-
-### R+q operating models {-}
-
-Like the R+Sel OMs, the range of variation in SSB across simulated populations for R+q OMs is only increased by allowing contrast in fishing pressure over time.
-There appears to be a slight decrease in accuracy in AIC when mean log-natural mortality is estimated and there is a more noticeable increase in accuracy for OMs with lower observation error in those cases.
-(Figures \ref{q_om_MF_BH_glm_AIC_plots} and \ref{q_om_ME_BH_glm_AIC_plots}).
-There was also an increase in AIC accuracy with lower observation error for when the mean natural mortality rate parameter was estimated.
+Like the R+Sel OMs, the greatest accuracy for AIC in selecting the Beverton-Holt stock-recruit relationship occurred for R+q OMs where there was contrast in fishing pressure over time which is also where there was the greatest variation in SSB (Figure \ref{sr_aic}, D, first and third rows).
+There were some differences in effects of SSB variation on AIC accuracy when OMs had high observation uncertainty and constant fishing pressure.
+Across all of the OMs in Figure \ref{sr_aic}, the only estimated decline in AIC accuracy with SSB variation occurred when R+q OMs had high observation uncertainty and constant fishing pressure and greater variation in catchability process errors that also had AR1 autocorrelation.
 
@@
 
-### R, R+S operating models {-}
+### Spawning stock biomass and recruitment {-}
 
-When mean log-natural mortality is known, all estimating models regardless of process error assumptions produce unbiased estimates of SSB over the entire time period for R OMs (Figure \ref{naa_om_em_R_MF_relbias_ssb}).
-For R+S OMs, only R+S EMs produce unbiased estimation of SSB and bias resulting from other EMs increased with larger variability in survival random effects.
-Estimating mean log-natural mortality generally produced greater variability in relative errors of SSB estimates and a change in sign of bias for EMs assuming process errors other than R and R+S (Figure \ref{naa_om_em_R_ME_relbias_ssb}).
-Estimating a stock-recruit relationship had little discernable effect on bias of SSB estimation (Figures \ref{naa_om_em_BH_MF_relbias_ssb}), but in combination with estimating mean log-natural mortality resulted in large bias of R+S EMs for R OMs with high observation error, constant fishing pressure and large variability in recruitment process errors (Figure \ref{naa_om_em_BH_ME_relbias_ssb}).
+For R OMs ($\sigma_{2+} = 0$), there was no indication of bias (95\% confidence intervals included 0) in terminal year SSB for any of the estimating models regardless of process error assumptions, except when no SR assumption was made, recruitment variability was low, and there was contrast in fishing mortality and high observation uncertainty (Figure \ref{SSB_rel_error}, A).
+However, errors in terminal SSB estimates were highly variable when median natural mortality was estimated and there was constant fishing pressure and high observation uncertainty (Figure \ref{SSB_rel_error}, A, second row).
 
-### R+M operating models {-}
+For R+S OMs, the EMs with matching process error assumption generally produced unbaised estimation of terminal SSB except when median natural mortality was estimated and there was high observation uncertainty.
+In R+S OMs with low observation uncertainty, EMs with incorrect process error assumptions typically provided biased estimation of terminal year SSB.
+Estimating the Beverton-Holt stock-recruit relationship had little discernible effect on bias of terminal year SSB estimation whereas estimating median M tended to produce more variability in errors in terminal SSB estimation similar to R OMs.
 
-When mean log-natural mortality is known, SSB estimation bias was small or non-existent for all EM process error assumptions when the OMs had low variability of the natural mortality process errors except a tendency toward negative bias in the terminal years for OMs with high observation error and a change in fishing pressure over time (Figure \ref{M_om_em_R_MF_relbias_ssb}).
-Errors in SSB estimation were most variable with high variability and autocorrelation of natural mortality process errors, but confidence intervals generally covered 0 indicating no evidence of bias.
-Estimating mean log-natural mortality results in large variability in SSB relative errors, particularly for OMs with high observation error and constant fishing pressure and all EMs produced biased estimation of SSB over all years when OMs had high observation error, a change in fishing pressure and low variability of natural mortality process errors (Figure \ref{M_om_em_R_ME_relbias_ssb}).
-Like R and R+S OMs, estimating a stock-recruit relationship had little discernible effect on SSB bias (Figures \ref{M_om_em_BH_MF_relbias_ssb} and \ref{M_om_em_BH_ME_relbias_ssb}).
+For R+ M OMs with low variability in natural mortality process errors, low observation uncertainty and contrast in fishing motality over time all EMs produced low variability in SSB estimation error that indicated unbiasedness (Figure \ref{SSB_rel_error}, B, third row).
+However, larger variability in natural mortality process errors increased bias of EMs without the correct process error type.
+Estimating median natural mortality increased variability of SSB estimation error particularly for OMs with high observation uncertainty and constant fishing pressure over time and increased bias in SSB estimation for many R+M OMs.
+Like R and R+S OMs, estimating a stock-recruit relationship had little discernible effect on SSB bias.
 
-### R+Sel operating models {-}
+For R+Sel OMs, there was no evidence of bias for any EMs when variability in selectivity process errors was low, observation uncertainty low and contrast in fishing mortality (Figure \ref{SSB_rel_error}, C).
+The largest bias for all EMs that estimated median natural mortality when the OMs had high observation uncertainty, constant fishing pressure and greater variability in selectivity process errors ($\sigma_{\text{Sel}} = 0.5$) or low selectivity process errors ($\sigma_{\text{Sel}} = 0.1$) and low observation uncertainty.
+However, there was no evidence of bias of matching R+Sel EMs when observation uncertainty was low and variation in selectivity process errors was larger, whether median natural mortality was estimated or not (Figure \ref{SSB_rel_error}, C, third row).
+We only observed an effect of estimating the Beverton-Holt Stock recruit relationship for R+Sel OMs that had high observation uncertainty and contrast in fishing pressure where estimating the relationship produced less biased SSB estimation for many EMs (Figure \ref{SSB_rel_error}, C, top row).
 
-When mean log-natural mortality is known and no stock-recruit relationship was estimated, SSB estimation bias was low for all EM process error assumptions (Figure \ref{Sel_om_em_R_MF_relbias_ssb}).
-The worst bias occurred for OMs with uncorrelated by more variable selectivity process errors and high observations error.
-As for other OMs, estimating mean log-natural mortality resulted in greater variability of the relative errors in SSB, but also more evidence of bias for EMS without the correct process error assumption (Figure \ref{Sel_om_em_R_ME_relbias_ssb}).
-Estimating a stock recruit relationship resulted in greater estimation of SSB in the terminal years and sometimes less bias when OMs had high observation error and a change in fishing pressure over time and the effect was more pronounced when mean log-natural mortality was estimated (Figures \ref{Sel_om_em_BH_MF_relbias_ssb} and \ref{Sel_om_em_BH_ME_relbias_ssb}).
+All EMs fit to data from R+q OMs with low observation uncertainty and  contrast in fishing pressure exhibited little evidence of bias in terminal SSB estimation except for R+M EMs when there was no AR1 correlation in catchability process errors (Figure \ref{SSB_rel_error}, D).
+Many EMs also performed well in R+q OMs with low observation uncertainty, but no contrast in fishing pressure.
+For R+q OMs with high observation uncertainty and contrast in fishing pressure, EMs that estimated the Beverton-Holt stock recruit function exhibited less SSB bias than those that did not.
+Estimating median natural mortality in the EMs only resulted in much more variable SSB estimation errors when there was no contrast in fishing pressure (Figure \ref{SSB_rel_error}, D, first and third rows).
 
-### R+q operating models {-}
+For all OM process error types, relative errors in terminal year recruitment were generally more variable than SSB, but effects of R and R+S OM and EM attributes on bias (i.e, negative or positive or none) were similar (Figure \ref{R_rel_error}, A).
+Furthermore, for EM configurations where bias in terminal bias in SSB was evident, median relative errors in recruitment often indicated stronger bias in recruitment of the same sign.
 
-When mean log-natural mortality is known and no S-R relationship was estimated, SSB estimation bias was low for all EM process error assumptions for all R+q OMs with low variability in catchability process errors (Figure \ref{q_om_em_R_MF_relbias_ssb}).
-With larger variability in catchability process errors, bias was low for all EMs except those configured with R+M process errors and OMs assumed low observation error.
-Many EMs also showed some negative bias for terminal SSB in OMs with higher observation error and a change in fishing pressure over time.
-As for other OMs, mean log-natural mortality resulted in greater variablity of the relative errors in SSB, but also more evidence of bias for some EMs without the correct process error assumption (Figure \ref{q_om_em_R_ME_relbias_ssb})
-Estimating a stock recruit relationship resulted in less bias in SSB in terminal years when OMs had high observation error and a change in fishing pressure over time (Figures \ref{q_om_em_BH_MF_relbias_ssb} and \ref{q_om_em_BH_ME_relbias_ssb}).
+### Beverton-Holt parameters{-}
 
-### Beverton-Holt parameter estimation {-}
+In R and R+S OMs, EMs with the correct assumptions about process errors, provided the least biased estimation of Beverton-Holt stock-recruit relationship parameters when there was low observation error, a change in fishing pressure over time, and lower variability of recruitment process errors, and higher variability survival process errors, and there was little effect of estimating natural mortality (Figure \ref{SR_rel_error}, A).
+For other R and R+S OMs, estimating natural mortality often resulted in less biased estimation of stock-recruit parameters.
+There was generally large variability in relative errors of the stock-recruit parameter estimates, but the lowest variability occurred with low variability in recruitment and little or no variability in survival process errors ($\sigma_{2+} \in \{0,0.25\}$), and contrast in fishing pressure.
+Across al R and R+S OMs, relative errors for the $a$ parameter were often less variable than those for $b$.
 
-In R and R+S OMs, EMs with the correct assumptions about process errors, provided the least biased estimation of Beverton-Holt stock-recruit relationship parameters when there was low observation error, a change in fishing pressure over time, and lower variability of recruitment process errors, and higher variability survival process errors, and there was little effect of estimating natural mortality (Figure \ref{naa_om_SR_relbias}).
-For other R and R+S OMs, estimating natural mortality often resulted in less biased estimation of stock-recruit parameters.
-The range of confidence intervals for the median parameter bias suggests large variability of the parameter estimates.
+In R+M OMs, the most accurate estimation of stock-recruit parameters for all EM process error assumptions occurred when there was a change in fishing pressure combined with either low variability in natural mortality process errors and high observation error or vice versa (Figure \ref{SR_rel_error}, B).
+Relative to the R, and R+S OMs, there was even less effect of estimating median natural mortality on estimation bias for the stock-recruit relationship parameters, but similarly there was generally large variability of the parameter estimates.
 
-In R+M OMs, the most accurate estimation of stock-recruit parameters occurred when there was a change in fishing pressure combined with either low variability in natural mortality process errors and high observation error or vice versa (Figure \ref{M_om_SR_relbias}).
-Relative to the R, and R+S OMs, there was even less effect of estimating mean log-natural mortality on estimation bias for the stock- recruit relationship parameters, but similarly there was generally large variability of the parameter estimates.
+Bias for stock-recruit parameters was very strong and variability in relative errors was greatest for most EMs fit to R+Sel OMs with constant fishing pressure (Figure \ref{SR_rel_error}, C).
+Less bias in parameter estimation occurred for OMs with a change in fishing pressure over time and of those OMs and the most accurate estimation occurred for EMs with process error types that matched the OMs and the OMs had low observation uncertainty, contrast in fishing mortality and more variable and uncorrelated selectivity process errors.
+There was little effect of estimating natural mortality on relative errors for stock recruit parameters.
 
-In R+Sel OMs, bias for stock-recruit parameters was very large for OMs with constant fishing pressure (Figure \ref{Sel_om_SR_relbias}).
-Less bias in parameter estimation occurred for OMs with a change in fishing pressure over time.
-There was little effect of estimating natural mortality.
-Patterns in bias of stock-recruit parameter estimation in R+q OMs, were very similar to those in R+Sel OMs with poor estimation of these parameters in OMs with constant fishing pressure (Figure \ref{q_om_SR_relbias}).
-Like other OMs, there was generally large variability of the parameter estimates.
+Like R+Sel OMs, relative errors in stock-recruit parameters were less variable for most EM process error types when OMs had contrast in fishing pressure and those OMs that also had lower observation uncertainty generally had the least variability (Figure \ref{SR_rel_error}, D).
+The worst accuracy of stock-recruit parameter estimation regardless of EM type occured when R+q OMs had low observation uncertainty and constant fishing pressure (Figure \ref{SR_rel_error}, D, fourth row).
 
-### (Mean) natural mortality rate estimation {-}
+### Median natural mortality rate {-}
 
-The mean natural mortality rate is estimated pretty accurately for R and R+S OMs when EMs use the correct process error configuration and there is a change in fishing pressure over time and observation error is low (Figure \ref{naa_om_M_relbias}).
-Estimation is a little more variable when observation error is higher or fishing pressure is constant, but when OMs have both, estimation is extremely variable.
-There is little effect of estimating stock-recruit relationships on the bias in estimating natural mortality.
-For R+M OMs, effects on estimating mean log-natural mortality were similar to the R and R+S OMs: best estimation with a change in fishing pressure and low observation error and worst with no change in fishing pressure and higher observation error (Figure \ref{M_om_M_relbias}).
-The same patterns also occurred for R+Sel and R+q OMs (Figures \ref{Sel_om_M_relbias} and \ref{q_om_M_relbias}).
+Median natural mortality rate is estimated accurately by all EM process error types for all R OMs except those where observation uncertainty is high and fishing pressure is constant where variability in relative errors is high (Figure \ref{M_rel_error}, A, $\sigma_{2+} = 0$).
+For R+S OMs estimation of median natural mortality rate is most accurate when observation uncertainty is low and there is contrast in fishing pressure and the EM process error type is correct.
+There is little effect of estimating stock-recruit relationships on the bias for median natural mortality.
+
+For R+M OMs, median natural mortality was estimated most accurately, regardless of EM process error type, when OMs had a change in fishing pressure and low observation uncertainty (Figure \ref{M_rel_error}, B).
+However, those R+M OMs that also had greatest variability in AR1 correlated natural mortality process errors only had unbiased estimation when the EM process error type was correct.
+Like R, and R+S OMs, there was little effect of estimating stock-recruit relationships on the bias for median natural mortality.
+
+All EM process error types produced accurate estimation of median natural mortality rate for R+Sel OMs that had contrast in fishing pressure, low observation uncertainty, and low selectivity process error variability (Figure \ref{M_rel_error}, C).
+When selectivity process error variability increased the incorrect EM process errors produce more biased estimation of median natural mortality rate.
+The least accurate estimation occurred for all EM process error types when observation uncertainty was high and there was constant fishing pressure.
+Again, like other OM process error types, there was little effect of estimating stock-recruit relationships on the bias for median natural mortality.
+
+Like R+Sel OMs, all EM process error types produced accurate estimation of median natural mortality rate when fit to R+q OMs with contrast in fishing pressure, low observation uncertainty and low catchability process error variability (Figure \ref{M_rel_error}, D).
+Most EM process error types produced biased estimation of median natural  mortality when R+q OMs had high observaiton uncertainty and constant fishing pressure.
+Again, there was no discernible effect of estimating stock-recruit relationships on bias.
 
@@
 
-### R, R+S operating models {-}
+Mohn's $\rho$ for SSB was small in absolute value for all R and R+S OMs, regardless of EM process error types, and whether median natural mortality rate or stock-recruit relationships were estimated (Figure \ref{mohns_rho_ssb}, A).
+The strongest retrospective patterns occurred in OMs with the largest survival process error variability, high observation uncertainty, and contrast in fishing pressure, but only for EMs without the correct process error type and where median natural mortality rate was assumed known.
+For R+M, R+Sel, and R+q OMs, Mohn's $\rho$ was also small in absolute value, but median values were all closer to 0 than the largest values in the R and R+S OMs (Figure \ref{mohns_rho_ssb},B-D).
+For these OMs, there was no noticeable effect of estimation of median natural mortality rate or stock recruit relationship on Mohn's $\rho$ for any EM process error types.
 
-When natural mortality is known and no stock recruit relationship is estimated, median absolute Mohn's $\rho$ was less than 0.25 for all EMS fit to simulated data sets from R and R+S OMs (Figure \ref{naa_om_em_R_MF_mohns_rho}).
-The strongest median Mohn's $\rho$ values for F and SSB were obtained for EMs that did not match the R+S OMs with the largest variability in survival process errors, change in fishing pressure over time, and higher observation error.
-Mohn's $\rho$ was also greater than 0 for recruitment for all EMS, including the matching EM, fit to R OMs that had the largest variability in recruitment random effects and high observation error.
-Estimating natural mortality removed the retrospective patterns in F and SSB in R+S OMs with the largest variability in survival process errors, change in fishing pressure over time, and higher observation error (Figure \ref{naa_om_em_R_ME_mohns_rho}).
-Estimating a stock-recruit relationship had no discernible effect on the retrospective patterns (Figure \ref{naa_om_em_BH_MF_mohns_rho}).
-Estimating natural mortality and the stock-recruit relationship increased the variablity of Mohn's $\rho$ for recruitment particularly for OMs with high observation error (Figure \ref{naa_om_em_BH_ME_mohns_rho}).
-
-### R+M operating models {-}
-
-When mean log-natural mortality is known and no stock recruit relationship is estimated, median abslolute Mohn's $\rho$ was less than 0.125 for all EMS fit to simulated data sets from R+M OMs (Figure \ref{M_om_em_R_MF_mohns_rho}).
-The strongest median Mohn's $\rho$ values for F, SSB, and recruitment were obtained for EMs other than R+M with autocorrelation assumed for OMs with a change in fishing pressure over time, and higher observation error.
-Like R and R+S OMs, estimating natural mortality removed the retrospective patterns in the respective OMs (Figure \ref{M_om_em_R_ME_mohns_rho}).
-Estimating a stock-recruit relationship also had the same effect as estimating the mean log-natural mortality rate, but Mohn's $\rho$ values were more variable for R+M EMs with autocorrelation assumed (Figure \ref{M_om_em_BH_MF_mohns_rho}).
-Estimating natural mortality and the stock-recruit relationship increased the variablity of Mohn's $\rho$ for F, SSB, and recruitment, particularly for OMs with high observation error and constant fishing pressure over time (Figure \ref{M_om_em_BH_ME_mohns_rho}).
-
-### R+Sel operating models {-}
-
-Similar to R+M OMs, when mean log-natural mortality is known and no stock recruit relationship is estimated, median abslolute Mohn's $\rho$ was less than 0.125 for all EMS fit to simulated data sets from R+Sel OMs (Figure \ref{Sel_om_em_R_MF_mohns_rho}).
-The strongest median Mohn's $\rho$ values for F, SSB, and recruitment were obtained for all EMs  with OMs that had a change in fishing pressure over time, and higher observation error.
-Estimating natural mortality increased the variability in Mohn's $\rho$ values, particularly for OMs with higher observation error and constant fishing (Figure \ref{Sel_om_em_R_ME_mohns_rho}).
-Estimating a stock-recruit relationship also removed any tendency for retrospective patterns in OMs with a change in fishing pressure and higher observation error (Figure \ref{Sel_om_em_BH_MF_mohns_rho}).
-Estimating natural mortality and the stock-recruit relationship further increased the variablity of Mohn's $\rho$ for F, SSB, and recruitment, particularly for OMs with high observation error and constant fishing pressure over time (Figure \ref{Sel_om_em_BH_ME_mohns_rho}).
-
-### R+q operating models {-}
-
-Similar to R+M and R+Sel OMs, when mean log-natural mortality is known and no stock recruit relationship is estimated, median abslolute Mohn's $\rho$ was less than 0.125 for all EMS fit to simulated data sets from R+q OMs (Figure \ref{q_om_em_R_MF_mohns_rho}).
-Also similar to other OMs, the strongest median Mohn's $\rho$ values for F, SSB, and recruitment were obtained for with OMs that had a change in fishing pressure over time, and higher observation error.
-Estimating natural mortality increased the variability in Mohn's $\rho$ values, particularly for OMs with constant fishing and autocorrelation of process errors in catchability (Figure \ref{q_om_em_R_ME_mohns_rho}).
-Similar to other OMs, estimating a stock-recruit relationship removed any tendency for retrospective patterns in OMs with a change in fishing pressure and higher observation error (Figure \ref{q_om_em_BH_MF_mohns_rho}).
-Like R+Sel OMs, estimating natural mortality and the stock-recruit relationship further increased the variablity of Mohn's $\rho$ for F, SSB, and recruitment, particularly for OMs with high observation error and constant fishing pressure over time (Figure \ref{q_om_em_BH_ME_mohns_rho}).
+Mohn's $\rho$ for recruitment was small in absolute value for all R OMs with low variability in recruitment process errors, regardless of EM process error type, and whether median natural mortality rate or stock-recruit relationships were estimated (Figure \ref{mohns_rho_R}, A).
+However, R and R+S OMs with greater recruitment process variability and higher observation uncertainty had median Mohn's $\rho$ for recruitment greater than zero for most EMs even when the the EM process error type was correct.
+In R+S OMs with lower observation uncertainty, EMs with the correct process error type exhibited better median Mohn's $\rho$ close to 0 than EMs with the incorrect process error type.
+For R+M, R+Sel, and R+q OMs, results for Mohn's $\rho$ for recruitment are similar to those for SSB, but the range in median values and variation in Mohn's $\rho$ values for a given OM are generally larger for recruitment (Figure \ref{mohns_rho_R}, B-D).
 
@@
 
-Convergence results can be useful for understanding how bad convergence in applications to real data might direct the practitioner to which alternative random effects configurations is be more appropriate.
+Convergence results can be useful for understanding how lack of convergence in applications to real data might direct the practitioner to which alternative random effects configurations is more appropriate.
 Therefore, the type of convergence that we might use as a diagnostic is important.
-It is common during assessment model fitting to check that the maximum absolute gradient component is less than some threshold.
-Our Type 3 convergence represents this and we used 1e-6 as the threshold, but there is no standard.
-Furthermore we found the hessian at the optimized log-likelihood can often be invertible (Convergence types 4 and 5) when the gradient approach to convergence fails.
-The probability of convergence type 5 cannot be greater than that of type 4 because type 5 is a subset where none of the standard errors obtained by inverting the hessian are greater than a specified value (100 in this case).
-We presume that the parameters that cause type 4 convergence to fail will also typically be the same as those that cause type 5 to fail.
-Perhaps a higher gradient threshold for Type 3 convergence would be more consistent with hessian-based convergence, but the gradient and hessian-based standard errors should be inspected.
+It is common during assessment model fitting to check that the maximum absolute gradient component is less than some threshold, but there is no standard.
+However, checking the Hessian for positive definiteness is sometimes done conditional on the convergence criterion.
+As such, some models may be excluded that in fact have an invertible hessian.
+We found the hessian at the optimized log-likelihood can often be invertible when the gradient approach with a sensible criterion (e.g., $10^{-3}$) would indicate lack of convergence.
+
+Another factor affecting the convercence criteria is numerical accuracy.
+With the process errors modeled as random effects we are optimizing an (Laplace) approximation of the marginal log-likelihood and therefore hessians and gradients are also with regard to the approximation.
+Furthermore, numerical methods are used to calculate the hessian (Optimhess?), and also to invert it for variance estimation.
+All of these approximations are affecting the gradients and invertibility of the hessian (although likely worse for some models than others), which, along with our results, suggests at least investigating whether the hessian is positive definite when the calculated gradients are not terribly large.
+
+Often high gradients and/or lack of positive definite hessian result when selectivity parameter estimates are at their bounds (0,1) or random effects variance parameter estimates are essentially 0.
+Changing the parameter configuration to avoid the bounds or removing process errors with 0 variance will often lead to satisfactory convergence.
+
+Perhaps a greater gradient threshold would be more consistent with hessian-based convergence, but the gradient and hessian-based standard errors should be inspected.
+
 Parameters that are not well estimated can lead the assessment scientist toward a more appropriate and often more parsimonious model configuration.
 For example, variance parameters that are estimated on log scale as low negative number (variance is close to 0) can happen for random effects estimated with little or no variability.
-These may sometimes lead a hessian to not be invertible or due to numerical issues may be invertible, but provide very large standard errors.
-More generally, we can expect lower likelihood of convergence when estimating natural mortality or stock-recruit relationships because the typical data in assessment models does not strongly inform these parameters.
+These may sometimes result in a hessian that is not positive definite or due to numerical issues may be invertible, but provide very large standard errors.
+Generally, we can expect lower probability of convergence when estimating natural mortality or stock-recruit relationships because of the difficulty distinguishing these parameters from others being estimated in assessment model with data typically available.
 
@@
 
-Among the process error configurations we used in OMs, we found AIC to be accurate for process errors on recruitment and survival.
+Among the process error configurations we used in OMs, we found AIC to be accurate for process errors on recruitment and survival (R and R+S).
 Fitting models to other OMs rarely preferred R+S EMs and R and R+S EMs were nearly always selected for the matching OMs.
-For other sources of process error, accuracy of AIC was improved to useful levels when there was larger variability in the process errors and/or lower observation error.
-Therefore, when data are relatively good and an there is large variability in the unknown process errors, AIC is likely to perform well.
+For other sources of process error, accuracy of AIC was improved to useful levels when there was larger variability in the process errors and/or better data quality.
 
-AIC generally performs weakly in determining stock recruit relationships unless there is large contrast in SSB.
-Interestingly, @milleretal16 found AIC to prefer a stock-recruit relationship for SNEMA yellowtail flounder where there was a large change in stock size estimated over time and flatfish are well observed by the bottom trawl indices relative to some other guilds such as semi-pelagic species.
+AIC generally performs weakly in determining stock recruit relationships unless there is large contrast in SSB and low variation in recruitment process errors.
 Although we did not compare models with alternative stock-recruit relationships, we do not expect AIC to perform any better distinguishing these relationships.
-Our results comport well with those of @devalpinehastings02 where state-space stock-recruit models were fit just to SSB and recruitment "observations".
+Our results comport well with those of @devalpinehastings02 where state-space stock-recruit models were fit just to SSB and recruitment "observations" and @brittenetalinreview which found AIC could not reliably distinguish alternative environmental effects on stock-recruit parameters.
+@milleretal16 found AIC to prefer a stock-recruit relationship with environmental effects for SNEMA yellowtail flounder, but there was a large change in stock size estimated over time and flatfish are well-observed by the NEFSC bottom trawl survey used for an index in the assessment model.
 
@@
 
-As we might expect bias in all parameters and assessment output was generally less with lower observation error in indices and age composition.
+As we might expect bias in all parameters and assessment output was generally better with lower observation error.
 Reliable estimation of stock-recruit relationship parameters only appears possible in ideal situations with lower observation errors in age composition and indices, lower variability in recruitment process errors and large contrast in spawning biomass over time.
-The latter can occur in with larger variability in survival or natural mortality process errors, variation in fishing pressure.
+The latter can occur with larger variability in survival or natural mortality process errors, or variation in fishing pressure.
 
-Estimation of natural mortality (or the mean when there are process errors on natural mortality) appears more feasible.
+Estimation of median natural mortality appears more feasible than SR parameters.
 We found low or no evidence of bias for many OM process error assumptions when there was contrast in fishing pressure even when there was greater observation error although it can lead to more variable estimates of natural mortality.
 For OMs where there was bias in natural mortality due to high observation error (R+Sel, R+q), estimating the stock-recruit relationship seemed to remove the bias.
+
 However, estimation of natural mortality can cause large differences between the true and estimated SSB (that may be unbiased on average) when there is less contrast in fishing pressure over time and higher observation error.
+
+Estimating B-H reduced SSB bias in some situations
 
@@
 
+## Conclusions {-}
+
+All together what do results mean? E.g., when do you get good convergence and retrospective patterns and high/none bias? When do you get good convergence and no retros and no/big bias?
+
+When R+S is preferred it probably means its the correct process error OF THE TYPES CONSIDERED HERE.
+
 ## Summary {-}
@@
 We conducted a simulation study where we simulated process errors and observations for 72 operating models with alternative assumptions about fishing history, degree of uncertainty in index and age composition observations, type (recruitment, survival, fishery selectivity, catchability and natural mortality), degree of variation, and correlation of process errors.
-We fit 20 different estimation models to each of 100 simulated set of observations with alternative assumptions type and correlation structure of process errors, (mean) natural mortality was known or estimated, and a B-H stock recruit relationship was assumed or not.
+We fit 20 different estimation models to each of 100 simulated set of observations with alternative assumptions type and correlation structure of process errors, median natural mortality was known or estimated, and a B-H stock recruit relationship was assumed or not.
 
@@
 We found AIC more accurately determined a B-H stock recruit relationship rather than the null model without a S-R relationship when there was low variability in recruitment, low variability in survival random effects, and higher variation in spawning biomass over the time series.
-When the (mean) natural mortality rate was estimated, we found large bias and uncertainty much more likely in model output such as spawning stock biomass.
+When the median natural mortality rate was estimated, we found large bias and uncertainty much more likely in model output such as spawning stock biomass.
 
@@
 Retrospective patterns were generally weak for all estimation models regardless of the true source of process error, but they can be expected for recruitment even for the correct process error assumptions when observation error is high.
-When models did exhibit some retrospective pattern, estimation the mean natural mortality rate tended to remove it.
+When models did exhibit some retrospective pattern, estimation the median natural mortality rate tended to remove it.
 
@@
 \pagebreak
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width = \textwidth]{om_input_plots_figure}
-\end{center}
-\caption{The proportion mature at age, weight at age, fleet and index selectivity at age, and Beverton-Holt stock recruit relationship assumed for the population in all operating models.
-For operating models with random effects on fleet selectivity, this represents the selectivity at the mean of the random effects.}\label{om_inputs_fig}
-\end{figure}
 
@@
 
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width = \textwidth]{om_input_plots_figure}
+\end{center}
+\caption{The proportion mature at age, weight at age, fleet and index selectivity at age, and Beverton-Holt stock recruit relationship assumed for the population in all operating models.
+For operating models with random effects on fleet selectivity, this represents the selectivity at the mean of the random effects.}\label{om_inputs_fig}
+\end{figure}
+
 \begin{landscape}
@@
 \begin{center}
+\includegraphics{hess_grad_convergence_plots}
+\end{center}
+\caption{The maximum of the absolute values of all gradient values for all fits that provided hessian-based standard errors across all simuated data sets of a given OM configuration (A: R and R+S, B: R+M, C: R+Sel, or D: R+q).
+Results are conditional on EM fits with alternative process error type (colored points and lines), median natural mortality (estimated or known) and recruitment assumptions (Beverton-Holt stock recruit or not).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{hess_grad}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
 \includegraphics{type_3_convergence_plots}
@@
 \begin{center}
-\includegraphics{M_bias_plots}
+\includegraphics{sr_bias_plots}
 \end{center}
-\caption{Median relative error of median/constant natural mortality for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{M_rel_error}
+\caption{Median relative error of Beverton-Holt stock recruitment parameters ($a$ and $b$) for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{SR_rel_error}
 \end{figure}
@@
 \begin{center}
-\includegraphics{sr_bias_plots}
+\includegraphics{M_bias_plots}
 \end{center}
-\caption{Median relative error of Beverton-Holt stock recruitment parameters ($a$ and $b$) for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{SR_rel_error}
+\caption{Median relative error of median natural mortality for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{M_rel_error}
 \end{figure}
@@
 \caption{Median Mohn's $\rho$ of recruitment for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_F}
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_R}
 \end{figure}

### 2025-02-12 Project_0/manuscript/manuscript.tex
saving work on project_0 manuscript

@@
 Stock\textsuperscript{6} \and John Wiedenmann\textsuperscript{7}}
-\date{15 January, 2025}
+\date{12 February, 2025}
 
@@
 estimating models with different assumptions on the process error
-source, whether (mean) natural mortality was estimated, and whether a
+source, whether median natural mortality was estimated, and whether a
 stock-recruit relationship was estimated.
@@
 the correct process error assumptions. When models did exhibit some
-retrospective pattern, estimating natural mortality often removed it.
+retrospective pattern, estimating median natural mortality often removed
+it.
 
@@
 time. We fitted 20 EMs to observations from each of 100 simulations
-where process errors were also simulated. EMs made alternative
+where process errors were also simulated. For R+M, R+Sel, and R+q OMs,
+we used unique seeds for each simulation, but we inadvertently used the
+same 100 seeds for all R and R+S OMs. Each of the EMs made alternative
 assumptions about the source of process errors and whether natural
-mortality (or the mean for models with process error in natural
+mortality (or the median for models with process error in natural
 mortality) was estimated and whether a Beverton-Holt stock recruit
@@
 at age was a logistic curve with \(a_{50}\) = 2.89 and slope = 0.88
-(Figure \ref{om_maturity}).
+(Figure \ref{om_inputs_fig}, top left).
 
@@
 \] where \(\theta_1 = e^{-12.1}\) and \(\theta_2 = 3.2\) (Figure
-\ref{om_waa}).
+\ref{om_inputs_fig}, top right).
 
@@
 \(\Fmsy = F_{40\%} = 0.348\) equated to a steepness of 0.69 and
-\(\alpha=0.60\) and \(\beta = 2.4 \times 10^{-5}\) for the \[
-N_{1,y} = \frac{\alpha \text{SSB}_{y-1}}{1 + \beta \text{SSB}_{y-1}} 
-\] Beverton-Holt parameterization (Figure \ref{om_sr}). For OMs without
-process errors on natural mortality we assumed the rate was assumed 0.2.
-For OMs with process errors on natural mortality the mean log natural
-mortality rate was \(\log(0.2)\).
+\(a=0.60\) and \(b = 2.4 \times 10^{-5}\) for the \[
+N_{1,y} = \frac{a \text{SSB}_{y-1}}{1 + b \text{SSB}_{y-1}} 
+\] Beverton-Holt parameterization (Figure \ref{om_inputs_fig}, bottom
+right). For OMs without process errors on natural mortality we assumed
+the rate was assumed 0.2. For OMs with process errors on natural
+mortality the median natural mortality rate was 0.2.
 
@@
 with logistic selectivity for the fleet with \(a_{50} = 5\) and slope =
-1 (Figure \ref{om_mean_selectivity}). This selectivity is was used to
-define \Fmsy for the Beverton-Holt stock recruitment parameters above.
-We assumed a logistic-normal distribution for the age-composition
-observations for the fleet.
+1 (Figure \ref{om_inputs_fig}, bottom left). This selectivity is was
+used to define \Fmsy for the Beverton-Holt stock recruitment parameters
+above. We assumed a logistic-normal distribution with no correlation on
+the multivariate normal scale for the age-composition observations for
+the fleet.
 
@@
 fleet, we assumed logistic selectivity for both indices with
-\(a_{50} = 5\) and slope = 1 and a logistic-normal distribution for the
-age-composition observations.
+\(a_{50} = 5\) and slope = 1 and a logistic-normal distribution with no
+correlation on the multivariate normal scale for the age-composition
+observations.
 
@@
 \log M_{y,a} = \mu_{M} + \epsilon_{M,y}
-\] that is a linear combination of a mean that is constant across ages
-\(\mu_{M} = \log(0.2)\) and any annual random effects marginally
-distributed as
+\] that is a linear combination of a mean log-natural mortality
+parameter that is constant across ages \(\mu_{M} = \log(0.2)\) and any
+annual random effects marginally distributed as
 \(\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)\). Uncorrelated
@@
 For each data set simulated from an OM 20 EMs were fit. A total of 32
-different EMs were fit across all OMs where the subset of 20 depended on
-the source of process error (Table \ref{em_table}). The first 20 EMs in
-Table \ref{em_table} were fit to simulate data sets from R and R+S OMs.
-EMs 5 to 24 in Table \ref{em_table} were fit to simulate data sets from
-R+M OMs. EMs 5 to 20 and 25-28 in Table \ref{em_table} were fit to
-simulate data sets from R+Sel OMs. Finally, EMs 5 to 20 and 29-32 in
-Table \ref{em_table} were fit to simulate data sets from R+q OMs. The
-observation error variance of aggregate catch and indices were all
-assumed known at the true values.
-
-maturity, weight at age, index and catch CVs, assumed known in all EMs
+different EMs were fit across OMs where the subset of 20 depended on the
+source of process error in the OM (Table \ref{em_table}). The EMs make
+different assumptions about the source of process error (R+S, R+M,
+R+Sel, R+q) and whether there is temporal autocorrelation or not, and
+whether a Beverton-Holt stock recruit function is estimated and the
+natural mortality rate (\(\mu_M\), the constant or mean on log scale for
+R+M EMs) is estimated. For simplicity we refer to the derived estimate
+\(e^{\mu_M}\) as the median natural mortality rate regardless of whether
+natural mortality random effects are estimated in the EM. The first 20
+EMs in Table \ref{em_table} were fit to simulate data sets from R and
+R+S OMs. EMs 5 to 24 in Table \ref{em_table} were fit to simulate data
+sets from R+M OMs. EMs 5 to 20 and 25-28 in Table \ref{em_table} were
+fit to simulate data sets from R+Sel OMs. Finally, EMs 5 to 20 and 29-32
+in Table \ref{em_table} were fit to simulate data sets from R+q OMs. The
+maturity at age, weight at age for catch and SSB, and observation error
+variance of aggregate catch and indices were all assumed known at the
+true values.
 
@@
 convergence when fitting each EM to the simulated data sets. There are
-various ways to assess convergence of the fit. We summarized 5
-alternative categories of convergence.
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Fit complete: Did the optimization routine (stats::nlminb) complete
-  without error?
-\item
-  No flag: Did the stats::nlminb convergence flag = 0 indicate
-  successful convergence?
-\item
-  Good gradient: Was the maximum absolute value of the gradient of the
-  log-likelihood \textless{} \(1\times10^{-6}\)?
-\item
-  No SE NAs: Did TMB::sdreport provide non-NA values for all fixed
-  effects standard errors?
-\item
-  No big SEs: Did TMB::sdreport provide all standard errors \textless{}
-  100?
-\end{enumerate}
-
-The first convergence criterion assesses whether the model crashes. The
-third is a measure that assesses how flat the likelihood is at the
-optimized point of the likelihood surface. The fourth and fifth criteria
-are specific to the hessian-based standard error reporting provided by
-TMB. the TMB::sdreport function will sometimes return standard error
-estimates even when the calculated hessian is not invertible (fourth
-criterion). It will also provide large standard errors for some
-parameters that are not estimated well or are near bounds on the
-transformed scale that is of primary interest. For example, variance
-parameters for random effects can often be estimated near 0 when the fit
-suggests no variation in the random effects or, equivalently, that the
-model without random effects is better. Note that Types 2 through 5 also
-require type 1, and type 5 requires type 4. We used the Clopper-Pearson
-exact method for constructing 95\% confidence intervals of the
-probabilities of convergence (Clopper and Pearson 1934; Thulin 2014).
+various ways to assess convergence of the fit (e.g., Carvalho et al.
+2021), but given the importance of estimates of uncertainty when using
+assessment models in management, we estimated probablity of convergence
+as measured by occurrence of a positive-definite hessian matrix at the
+optimized negative log-likelihood that could be inverted. However, we
+also provided results in the Supplementary Materials for the maximum of
+the absolute values among all gradients for all fits of a given EM to
+all simulated data sets from a given OM that produced hessian-based
+standard errors for all estimated fixed effects. This provides an
+indication of how poor the calculated gradients can be, but still
+presumably converged adequately enough for parameter inferences.
 
@@
 operating model, we compared AIC for EMs that all made the same
-assumptions about median/constant natural mortality (known or estimated)
-and stock recruitment model (Beverton-Holt or none).
+assumptions about median natural mortality (known or estimated) and
+stock recruitment model (Beverton-Holt or none).
 
@@
 models that otherwise assume the same process error structure as the
-operating model and both of the compared models either estimate
-median/constant natural mortality or assume it is known. Our preliminary
-inspections of the proportions of simulations where the correct
-recruitment model was chosen for a given set of OM factors indicated
-generally poor performance of AIC. Furthermore, it has been shown that
-estimation of stock-recruit parameters requires a wide range of SSB
-(References). Therefore, we fit logistic regression models to the
-indicator of Beverton-Holt models having lower AIC as a function of the
-log-standard deviation of the true log(SSB) (similar to the log of the
-coefficient of variation for SSB) for each simulation.
+operating model and both of the compared models either estimate median
+natural mortality or assume it is known. Contrast in fishing pressure
+and time series with recruitment at low stock size has been shown to
+improve estimability of stock-recruitment parameters (Magnusson and
+Hilborn 2007; Conn et al. 2010). Our preliminary inspections of the
+proportions of simulations where the correct recruitment model was
+chosen for a given set of OM factors (including contrast in fishing
+pressure) indicated generally poor performance of AIC. Therefore, we fit
+logistic regression models to the indicator of Beverton-Holt models
+having lower AIC as a function of the log-standard deviation of the true
+log(SSB) (similar to the log of the coefficient of variation for SSB)
+since simulations with realized SSB producing low and high recruitments
+would have larger variation in realized SSB.
 
@@
 
-We estimated bias as the median of the relative errors across all
-simulations for a given OM and EM combination. \[
+For a given model attribute we calculated the relative error \[
 \text{RE}\left(\theta_i\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
-\] of annual SSB and fully-selected fishing mortality for each EM and
-constructed 95\% confidence intervals for the median relative bias using
-the binomial distribution approach as in Miller and Hyun (2018) and
-Stock and Miller (2021). We also estimated bias of stock-recruit
-parameters for EMs that assumed the Beverton-Holt stock recruit function
-and the mean natural mortality rate parameter when estimated.
+\] from fitting a given estimating model to simulated data set \(i\)
+configured for a given OM where \(\widehat \theta_i\) and \(\theta_i\)
+are the estimated and true values for simulation \(i\). We estimated
+bias as the median of the relative errors across all simulations for a
+given OM and EM combination. We constructed 95\% confidence intervals
+for the median relative bias using the binomial distribution approach as
+in Miller and Hyun (2018) and Stock and Miller (2021). We present
+results for terminal years SSB and recruitment, Beverton-Holt stock
+recruit parameters (\(a\) and \(b\)), and median natural mortality rate.
+Results for terminal year fishing mortality are strongly negatively
+correlated with those for SSB and provided in the supplementary
+materials.
 
@@
 We estimated Mohn's \(\rho\) for SSB, fully-selected fishing mortality,
-and recruitment for each EM. We estimated 7 peels for each EM. We
-calculated median 95\% confidence intervals for Mohn's \(\rho\) using
-the same methods as that for relative bias.
-
-Similar to the other results, retrospective results only condition on
-whether all of the peels of a given estimating model completed the
-optimization process without failure. We did not condition on
-convergence as defined by a gradient threshold or invertibity of the
-hessian because the optimized model can provide reliable estimation of
-SSB, F, M, and stock recruit parameters whether or not the model was
-able to estimate non-zero random effects.
+and recruitment for each EM (Mohn 1999). We estimated 7 peels for each
+EM. We calculated median 95\% confidence intervals for Mohn's \(\rho\)
+using the same methods as that for relative bias. Similar to the other
+results, retrospective results only condition on whether all of the
+peels of a given estimating model completed the optimization process
+without failure. We did not condition on convergence as defined by a
+gradient threshold or invertibity of the hessian because the optimized
+model can provide reliable estimation of SSB, fishing mortality rates,
+and recruitment whether or not the model was able to estimate non-zero
+random effects.
 
@@
 
-\hypertarget{r-rs-operating-models}{%
-\subsubsection*{R, R+S operating models}\label{r-rs-operating-models}}
-\addcontentsline{toc}{subsubsection}{R, R+S operating models}
-
-R+S EMs fit to R OMs exhibited poor convergence for types 3, 4, and 5
-regardless of whether M or a stock-recruit relationship was estimated
-(Figures \ref{naa_om_em_R_MF_convergence} to
-\ref{naa_om_em_BH_ME_convergence}). R EMs exhibited high convergence
-rates for all convergence types for both R and R+S OMs. Convergence
-rates were high for all convergence types for all EMs in R+S OMs with
-lower uncertainty in observations. Convergence for most convergence
-types generally declines most EMs when mean-log natural mortality rate
-is estimated and/or a Beverton-Holt stock recruit relationship is
-estimated even when the process error assumptions of the estimation and
-OMs match. There can be relatively high convergence probability of type
-4 (an invertible hessian) for EMs that do not have the correct process
-error assumed, but type 5 convergence (no very large standard errors
-estimated from hessian) for these EMs is typically much lower.
-
-\hypertarget{rm-operating-models}{%
-\subsubsection*{R+M operating models}\label{rm-operating-models}}
-\addcontentsline{toc}{subsubsection}{R+M operating models}
-
-R+S EMs generally converged less reliably across convergence types than
-other EMs when fit to data generated from R+M OMs (Figures
-\ref{M_om_em_R_MF_convergence} to \ref{M_om_em_BH_ME_convergence}).
-However, using convergence type 5, all EMs converged poorly when OMs had
-low variability in natural mortality process errors or higher
-observation error uncertainty. Using convergence type 5, R+M EMs
-generally converged better than other EMs when observation error
-uncertainty was low and natural mortality process errors were more
-variable. Like R+S OMs, the probability of convergence generally
-declined for all EMs when mean log-natural mortality and/or a
-stock-recruit relationship was estimated.
-
-\hypertarget{rsel-operating-models}{%
-\subsubsection*{R+Sel operating models}\label{rsel-operating-models}}
-\addcontentsline{toc}{subsubsection}{R+Sel operating models}
-
-R+S EMs in particular converged poorly when R+Sel OMs had less
-variability in selectivity process errors or higher observation error
-uncertainty (Figures \ref{Sel_om_em_R_MF_convergence} to
-\ref{Sel_om_em_BH_ME_convergence}). R+Sel EMs generally converged better
-than other EMs for OMs with greater variability in process errors, lower
-observation error, and contrast in fishing pressure regardless of
-whether mean log natural mortality or a stock recruit relationship was
+For R and R+S OMs, convergence generally declines for most EMs when the
+median natural mortality rate is estimated and/or the Beverton-Holt
+stock recruit relationship is estimated even when the process error
+assumptions of the EMs and OMs match (Figure
+\ref{hessian_SE_convergence}, A). When there was high observation error
+and constant fishing pressure (change from \(2.5F_{\text{MSY}}\) to
+\(F_{\text{MSY}}\)), convergence of all EM process error configurations
+other than R EMs was poor for R OMs (\(\sigma_{2+} = 0\)) regardless of
+whether median natural mortality and stock-recruit relationships were
+estimated. Convergence of R EMs was high for all R and R+S OMs except
+when there was high observation error and constant fishing pressure and
+median natural mortality and stock-recruit relationships were estimated.
+R+S EMs fit to R OMs exhibited poor convergence regardless of whether
+natural mortality or a stock-recruit relationship was estimated. R+S EMs
+fit to R+S OMs had highest convergence rates when there was contrast in
+fishing pressure and low observation error. Convergence rates were high
+for all EMs when fit to data from R+S OMs with lower observation
+uncertainty except those where median natural mortality and/or
+stock-recruit relationships were estimated.
+
+Convergence of all EMs fitted to R+M OMs was best when the OMs had
+higher natural mortality process error variability, low observation
+error, and contrast in fishing pressure (Figure
+\ref{hessian_SE_convergence}, B). R+M EMs that estimated autocorrelation
+of process errors had poor convergence for R+M OMs when there was low
+natural mortality process error variability whether or not there was
+autocorrelation of the simulated process errors. R+S EMs fitted to data
+generated from R+M OMs always converged poorly whether median natural
+mortality and the Beverton-Holt stock-recruit relationship were
 estimated.
 
-\hypertarget{rq-operating-models}{%
-\subsubsection*{R+q operating models}\label{rq-operating-models}}
-\addcontentsline{toc}{subsubsection}{R+q operating models}
-
-Convergence of R+q EMs is generally better than that of other EMs for
-all convergence types when R+q OMs assume contrast in fishing history
-overFigures (\ref{q_om_em_R_MF_convergence} to
-\ref{q_om_em_BH_ME_convergence}). Convergence of R+S EMs is generally
-worse than that of other EMs across all OMs whether or not mean
-log-natural mortality or a stock recruit relationship is estimated.
-Again, convergence probablity generally declines for all EMs when mean
-log-natural mortality or a stock recruit relationship is estimated.
+R+S EMs, in particular, converged poorly when fit to data generated from
+R+Sel OMs with lower selectivity process error variability or higher
+observation uncertainty (Figure \ref{hessian_SE_convergence}, C). R+Sel
+EMs generally converged better than other EMs for R+Sel OMs with higher
+process error variability, lower observation uncertainty, and contrast
+in fishing pressure regardless of whether median natural mortality or a
+stock recruit relationship was estimated.
+
+For R+q OMs, convergence of R+q EMs is generally better than that of
+other EMs when there is contrast in fishing history (Figure
+(\ref{hessian_SE_convergence}, D). Convergence of R+S EMs is generally
+worse than that of all other EMs across all OMs whether or not median
+natural mortality or a stock recruit relationship is estimated. Again,
+convergence probability generally declines for all EMs when median
+natural mortality or a stock recruit relationship is estimated.
+
+We found a wide range of maximum absolute values of gradients for models
+that converged as defined here (Figure \ref{hess_grad}). The largest
+value observed for a given EM and OM combination was typically
+\(<10^{-3}\), but many converged models had values greater than 1. For
+many OMs, EMs that assumed the correct process error type and did not
+estimate median natural mortality or the Beverton-Holt stock-recruit
+relationship produced the lowest gradient values.
 
@@
 
-\hypertarget{r-rs-operating-models-1}{%
-\subsubsection*{R, R+S operating models}\label{r-rs-operating-models-1}}
-\addcontentsline{toc}{subsubsection}{R, R+S operating models}
-
 Marginal AIC accurately determines the correct process error assumptions
 in EMs when data are generated from R and R+S OMs, regardless of whether
-mean log-natural mortality or a stock recruit relationship is estimated
-(Figures \ref{naa_om_proportion_best_aic_R_MF} to
-\ref{naa_om_proportion_best_aic_SR_ME}). Adding estimation of mean
-log-natural mortality or a stock recruit relationship separately has a
-negligible effect on the accuracy of determining the correct process
-error assumption. When both are estimated, there is a noticeable
-reduction in accuracy when OMs have a constant fishing history,
-observation error is low and largest variability in recruitment process
-errors.
-
-\hypertarget{rm-operating-models-1}{%
-\subsubsection*{R+M operating models}\label{rm-operating-models-1}}
-\addcontentsline{toc}{subsubsection}{R+M operating models}
-
-Marginal AIC only accurately determined the correct process error model
-and correlation structure when observation error was low and variability
-in natural mortality process errors was high (Figures
-\ref{M_om_proportion_best_aic_R_MF} to
-\ref{M_om_proportion_best_aic_SR_ME}). Estimating the mean natural
-mortality rate reduced the accuracy of AIC for OMs that assumed natural
-mortality process errors were independent. For OMs with poor accuracy,
-AIC most frequently selected EMs with process errors in catchability or
-selectivity. Selection of R+S EMs was typically unlikely.
-
-\hypertarget{rsel-operating-models-1}{%
-\subsubsection*{R+Sel operating models}\label{rsel-operating-models-1}}
-\addcontentsline{toc}{subsubsection}{R+Sel operating models}
+median natural mortality or a stock recruit relationship is estimated
+(Figure \ref{pe_aic}, A). Adding estimation of median natural mortality
+or a stock recruit relationship separately has a negligible effect on
+the accuracy of determining the correct process error assumption. When
+both are estimated, there is a noticeable reduction in accuracy when OMs
+have a constant fishing history, observation uncertainty is low and
+larger variability in recruitment process errors.
+
+For R+M OMs, marginal AIC only accurately determined the correct process
+error model and correlation structure when observation error was low and
+variability in natural mortality process errors was high (Figures
+\ref{pe_aic}, B). Estimating the median natural mortality rate reduced
+the accuracy of AIC for OMs that assumed natural mortality process
+errors were independent. For OMs with poor accuracy, AIC most frequently
+selected EMs with process errors in catchability (R+q) or selectivity
+(R+Sel). Selection of R+S EMs was typically unlikely.
 
 Marginal AIC most accurately determined the correct source of process
-error and correlation structure for R+Sel OMs with low observation error
-(Figures \ref{Sel_om_proportion_best_aic_R_MF} to
-\ref{Sel_om_proportion_best_aic_SR_ME}). When there was low variability
-in selectivity process errors and high observation error, R+q or R+S EMs
-were more likely to have the best AIC. Whether mean log-natural morality
-or stock recruit relationships were estimated appeared to have little
-effect on the performance of marginal AIC.
-
-\hypertarget{rq-operating-models-1}{%
-\subsubsection*{R+q operating models}\label{rq-operating-models-1}}
-\addcontentsline{toc}{subsubsection}{R+q operating models}
+error and correlation structure for R+Sel OMs with low observation
+uncertainty (Figures \ref{pe_aic}, C). When there was low variability in
+selectivity process errors and high observation uncertainty, R+q or R+S
+EMs were more likely to have the best AIC. Whether median natural
+morality or stock recruit relationships were estimated appeared to have
+little effect on the performance of AIC.
 
@@
 error and correlation structure for R+q OMs with high variability in
-catchability process errors (Figures \ref{q_om_proportion_best_aic_R_MF}
-to \ref{q_om_proportion_best_aic_SR_ME}). The worst accuracy occurred
-for OMs with low variability in catchability process errors and high
-observation error. However, in these OMs, the marginal AIC accurately
-determined the correct source of process error (but not correlation
-structure) except when EMs estimated both mean log-natural morality and
-the stock recruit relationships and OMs assumed a constant fishing
-pressure.
+catchability process errors (Figures \ref{pe_aic},D). The worst accuracy
+occurred for OMs with low variability in catchability process errors and
+high observation uncertainty. However, in these OMs, the marginal AIC
+accurately determined the correct source of process error (but not
+correlation structure) except when EMs estimated both median natural
+morality and the stock recruit relationships and OMs assumed a constant
+fishing pressure.
 
@@
 Our comparisons of model performance condition on assuming the true
-process error configuration is known (EM and OM process error
-assumptions match). Broadly, we found generally poor accuracy of AIC in
-selecting models assuming a Beverton-Holt stock recruit function over
-the null model without an assumed stock-recruit relationship for all OMs
-(Tables \ref{naa_om_em_R_BH_aic_table} to \ref{q_om_em_R_BH_aic_table}.
-However, we also found AIC to more accurately determine a stock-recruit
-relationship when there was greater variation in spawning biomass
-generated in the simulated populations.
-
-\hypertarget{r-rs-operating-models-2}{%
-\subsubsection*{R, R+S operating models}\label{r-rs-operating-models-2}}
-\addcontentsline{toc}{subsubsection}{R, R+S operating models}
-
-Among R and R+S OMs, AIC accuracy for including the B-H stock-recruit
-relationship was estimated to be highest for R OMs and EMs and lowest
-variance of recruitment process errors (Figures
-\ref{naa_om_MF_BH_glm_AIC_plots} and \ref{naa_om_ME_BH_glm_AIC_plots}).
-FOr R+S OMs and holding variability in SSB constant, accuracy of
-stock-recruit model selection declined with greater variance of
-recruitment and survival process errors. Simultaneously estimating mean
-log-natural mortality has a small negative effect on AIC accuracy, but
-there is an improvement in AIC accuracy with lower observation error in
-indices and age composition and lower variation of survival process
-errors.
-
-\hypertarget{rm-operating-models-2}{%
-\subsubsection*{R+M operating models}\label{rm-operating-models-2}}
-\addcontentsline{toc}{subsubsection}{R+M operating models}
-
-For R+M OMs and EMs there appeared to be little effect of correlation in
-the natural mortality process errors or estimating the mean log-natural
-mortality on accuracy of AIC in determining a stock-recruit function,
-holding variability in SSB constant (Figures
-\ref{M_om_MF_BH_glm_AIC_plots} and \ref{M_om_ME_BH_glm_AIC_plots}).
-However, there was an increase in accuracy with lower observation error
-for OMs with higher variability in natural mortality process errors
-whether the mean natural mortality rate parameter was estimated or not.
-
-\hypertarget{rsel-operating-models-2}{%
-\subsubsection*{R+Sel operating models}\label{rsel-operating-models-2}}
-\addcontentsline{toc}{subsubsection}{R+Sel operating models}
-
-The range of variation in SSB across simulated populations was lower for
-R+Sel OMs than R+S or R+M OMs. OMs with contrast in fishing pressure
-over time is the primary mechanism creating variation in SSB. However,
-there appears to be little effect of variability or correlation of
-selectivity process errors or whether mean log-natural mortality was
-estimated on the accuracy of AIC in selecting the stock recruit
-relationship (Figures \ref{Sel_om_MF_BH_glm_AIC_plots} and
-\ref{Sel_om_ME_BH_glm_AIC_plots}). However, there was an increase in AIC
-accuracy with lower observation error for OMs with independent process
-errors whether the mean natural mortality rate parameter was estimated
-or not.
-
-\hypertarget{rq-operating-models-2}{%
-\subsubsection*{R+q operating models}\label{rq-operating-models-2}}
-\addcontentsline{toc}{subsubsection}{R+q operating models}
-
-Like the R+Sel OMs, the range of variation in SSB across simulated
-populations for R+q OMs is only increased by allowing contrast in
-fishing pressure over time. There appears to be a slight decrease in
-accuracy in AIC when mean log-natural mortality is estimated and there
-is a more noticeable increase in accuracy for OMs with lower observation
-error in those cases. (Figures \ref{q_om_MF_BH_glm_AIC_plots} and
-\ref{q_om_ME_BH_glm_AIC_plots}). There was also an increase in AIC
-accuracy with lower observation error for when the mean natural
-mortality rate parameter was estimated.
+process error configuration is known (EM and OM process error types
+match). Broadly, we found generally poor accuracy of AIC in selecting
+models assuming a Beverton-Holt stock recruit function over the null
+model without an assumed stock-recruit relationship for all OMs.
+However, we also found increased accuracy of AIC in determining the
+Beverton-Holt stock-recruit relationship with greater variation in
+spawning biomass generated in the simulated populations for nearly every
+OM (Figure \ref{sr_aic}).
+
+With R and R+S process error assumptions, probability of AIC accuracy
+for the B-H stock-recruit relationship as a function of SSB variability
+were greatest for OMs where there was greater process variability in
+survival, lower process variability in recruitment and contrast in
+fishing pressure (Figure \ref{sr_aic}, A). However, the largest
+variation in SSB (Figure \ref{sr_aic}, rug on x-axis) occurred in OMs
+with larger recruitment variability (\(\sigma_R = 1.5\); Figure
+\ref{sr_aic}, A, right column group). The weakest overall accuracy of
+AIC for the Beverton-Holt stock-recruit function was when there was
+constant fishing pressure (Figure \ref{sr_aic}, A, second and fourth
+rows). The largest effect of simultaneously estimating median natural
+mortality was when there was OMs had high observation uncertainty and
+constant fishing pressure (Figure \ref{sr_aic}, A, second row).
+
+For R+M OMs, probability of AIC accuracy for the Beverton-Holt
+stock-recruit relationship was greatest for OMs where there was contrast
+in fishing pressure (Figure \ref{sr_aic}, B). The largest variation in
+SSB occurred in OMs with larger natural mortality variability and
+temporal AR1 correlation of the annual random effects
+(\(\sigma_M = 1.5\), \(\rho_M = 0.9\); Figure \ref{sr_aic}, B,
+right-most column) where AIC accuracy was also high even when there was
+constant fishing pressure. Interestingly, there seemed to be little
+difference in AIC accuracy whether there was lower or higher observation
+uncertainty or whether the EM simultaneously estimated median natural
+mortality.
+
+For R+Sel OMs, contrast in fishing pressure over time is the primary
+mechanism creating variation in SSB and these are the OMs where AIC
+accuracy for the Beverton-Holt stock recruit relationship was greatest
+(Figure \ref{sr_aic}, C, first and third rows). There appears to be
+little effect of variability or correlation of selectivity process
+errors or whether median natural mortality was estimated on AIC
+accuracy.
+
+Like the R+Sel OMs, the greatest accuracy for AIC in selecting the
+Beverton-Holt stock-recruit relationship occurred for R+q OMs where
+there was contrast in fishing pressure over time which is also where
+there was the greatest variation in SSB (Figure \ref{sr_aic}, D, first
+and third rows). There were some differences in effects of SSB variation
+on AIC accuracy when OMs had high observation uncertainty and constant
+fishing pressure. Across all of the OMs in Figure \ref{sr_aic}, the only
+estimated decline in AIC accuracy with SSB variation occurred when R+q
+OMs had high observation uncertainty and constant fishing pressure and
+greater variation in catchability process errors that also had AR1
+autocorrelation.
 
@@
 
-\hypertarget{r-rs-operating-models-3}{%
-\subsubsection*{R, R+S operating models}\label{r-rs-operating-models-3}}
-\addcontentsline{toc}{subsubsection}{R, R+S operating models}
-
-When mean log-natural mortality is known, all estimating models
-regardless of process error assumptions produce unbiased estimates of
-SSB over the entire time period for R OMs (Figure
-\ref{naa_om_em_R_MF_relbias_ssb}). For R+S OMs, only R+S EMs produce
-unbiased estimation of SSB and bias resulting from other EMs increased
-with larger variability in survival random effects. Estimating mean
-log-natural mortality generally produced greater variability in relative
-errors of SSB estimates and a change in sign of bias for EMs assuming
-process errors other than R and R+S (Figure
-\ref{naa_om_em_R_ME_relbias_ssb}). Estimating a stock-recruit
-relationship had little discernable effect on bias of SSB estimation
-(Figures \ref{naa_om_em_BH_MF_relbias_ssb}), but in combination with
-estimating mean log-natural mortality resulted in large bias of R+S EMs
-for R OMs with high observation error, constant fishing pressure and
-large variability in recruitment process errors (Figure
-\ref{naa_om_em_BH_ME_relbias_ssb}).
-
-\hypertarget{rm-operating-models-3}{%
-\subsubsection*{R+M operating models}\label{rm-operating-models-3}}
-\addcontentsline{toc}{subsubsection}{R+M operating models}
-
-When mean log-natural mortality is known, SSB estimation bias was small
-or non-existent for all EM process error assumptions when the OMs had
-low variability of the natural mortality process errors except a
-tendency toward negative bias in the terminal years for OMs with high
-observation error and a change in fishing pressure over time (Figure
-\ref{M_om_em_R_MF_relbias_ssb}). Errors in SSB estimation were most
-variable with high variability and autocorrelation of natural mortality
-process errors, but confidence intervals generally covered 0 indicating
-no evidence of bias. Estimating mean log-natural mortality results in
-large variability in SSB relative errors, particularly for OMs with high
-observation error and constant fishing pressure and all EMs produced
-biased estimation of SSB over all years when OMs had high observation
-error, a change in fishing pressure and low variability of natural
-mortality process errors (Figure \ref{M_om_em_R_ME_relbias_ssb}). Like R
-and R+S OMs, estimating a stock-recruit relationship had little
-discernible effect on SSB bias (Figures \ref{M_om_em_BH_MF_relbias_ssb}
-and \ref{M_om_em_BH_ME_relbias_ssb}).
-
-\hypertarget{rsel-operating-models-3}{%
-\subsubsection*{R+Sel operating models}\label{rsel-operating-models-3}}
-\addcontentsline{toc}{subsubsection}{R+Sel operating models}
-
-When mean log-natural mortality is known and no stock-recruit
-relationship was estimated, SSB estimation bias was low for all EM
-process error assumptions (Figure \ref{Sel_om_em_R_MF_relbias_ssb}). The
-worst bias occurred for OMs with uncorrelated by more variable
-selectivity process errors and high observations error. As for other
-OMs, estimating mean log-natural mortality resulted in greater
-variability of the relative errors in SSB, but also more evidence of
-bias for EMS without the correct process error assumption (Figure
-\ref{Sel_om_em_R_ME_relbias_ssb}). Estimating a stock recruit
-relationship resulted in greater estimation of SSB in the terminal years
-and sometimes less bias when OMs had high observation error and a change
-in fishing pressure over time and the effect was more pronounced when
-mean log-natural mortality was estimated (Figures
-\ref{Sel_om_em_BH_MF_relbias_ssb} and
-\ref{Sel_om_em_BH_ME_relbias_ssb}).
-
-\hypertarget{rq-operating-models-3}{%
-\subsubsection*{R+q operating models}\label{rq-operating-models-3}}
-\addcontentsline{toc}{subsubsection}{R+q operating models}
-
-When mean log-natural mortality is known and no S-R relationship was
-estimated, SSB estimation bias was low for all EM process error
-assumptions for all R+q OMs with low variability in catchability process
-errors (Figure \ref{q_om_em_R_MF_relbias_ssb}). With larger variability
-in catchability process errors, bias was low for all EMs except those
-configured with R+M process errors and OMs assumed low observation
-error. Many EMs also showed some negative bias for terminal SSB in OMs
-with higher observation error and a change in fishing pressure over
-time. As for other OMs, mean log-natural mortality resulted in greater
-variablity of the relative errors in SSB, but also more evidence of bias
-for some EMs without the correct process error assumption (Figure
-\ref{q_om_em_R_ME_relbias_ssb}) Estimating a stock recruit relationship
-resulted in less bias in SSB in terminal years when OMs had high
-observation error and a change in fishing pressure over time (Figures
-\ref{q_om_em_BH_MF_relbias_ssb} and \ref{q_om_em_BH_ME_relbias_ssb}).
-
-\hypertarget{beverton-holt-parameter-estimation}{%
-\subsubsection*{Beverton-Holt parameter
-estimation}\label{beverton-holt-parameter-estimation}}
-\addcontentsline{toc}{subsubsection}{Beverton-Holt parameter estimation}
+\hypertarget{spawning-stock-biomass-and-recruitment}{%
+\subsubsection*{Spawning stock biomass and
+recruitment}\label{spawning-stock-biomass-and-recruitment}}
+\addcontentsline{toc}{subsubsection}{Spawning stock biomass and
+recruitment}
+
+For R OMs (\(\sigma_{2+} = 0\)), there was no indication of bias (95\%
+confidence intervals included 0) in terminal year SSB for any of the
+estimating models regardless of process error assumptions, except when
+no SR assumption was made, recruitment variability was low, and there
+was contrast in fishing mortality and high observation uncertainty
+(Figure \ref{SSB_rel_error}, A). However, errors in terminal SSB
+estimates were highly variable when median natural mortality was
+estimated and there was constant fishing pressure and high observation
+uncertainty (Figure \ref{SSB_rel_error}, A, second row).
+
+For R+S OMs, the EMs with matching process error assumption generally
+produced unbaised estimation of terminal SSB except when median natural
+mortality was estimated and there was high observation uncertainty. In
+R+S OMs with low observation uncertainty, EMs with incorrect process
+error assumptions typically provided biased estimation of terminal year
+SSB. Estimating the Beverton-Holt stock-recruit relationship had little
+discernible effect on bias of terminal year SSB estimation whereas
+estimating median M tended to produce more variability in errors in
+terminal SSB estimation similar to R OMs.
+
+For R+ M OMs with low variability in natural mortality process errors,
+low observation uncertainty and contrast in fishing motality over time
+all EMs produced low variability in SSB estimation error that indicated
+unbiasedness (Figure \ref{SSB_rel_error}, B, third row). However, larger
+variability in natural mortality process errors increased bias of EMs
+without the correct process error type. Estimating median natural
+mortality increased variability of SSB estimation error particularly for
+OMs with high observation uncertainty and constant fishing pressure over
+time and increased bias in SSB estimation for many R+M OMs. Like R and
+R+S OMs, estimating a stock-recruit relationship had little discernible
+effect on SSB bias.
+
+For R+Sel OMs, there was no evidence of bias for any EMs when
+variability in selectivity process errors was low, observation
+uncertainty low and contrast in fishing mortality (Figure
+\ref{SSB_rel_error}, C). The largest bias for all EMs that estimated
+median natural mortality when the OMs had high observation uncertainty,
+constant fishing pressure and greater variability in selectivity process
+errors (\(\sigma_{\text{Sel}} = 0.5\)) or low selectivity process errors
+(\(\sigma_{\text{Sel}} = 0.1\)) and low observation uncertainty.
+However, there was no evidence of bias of matching R+Sel EMs when
+observation uncertainty was low and variation in selectivity process
+errors was larger, whether median natural mortality was estimated or not
+(Figure \ref{SSB_rel_error}, C, third row). We only observed an effect
+of estimating the Beverton-Holt Stock recruit relationship for R+Sel OMs
+that had high observation uncertainty and contrast in fishing pressure
+where estimating the relationship produced less biased SSB estimation
+for many EMs (Figure \ref{SSB_rel_error}, C, top row).
+
+All EMs fit to data from R+q OMs with low observation uncertainty and
+contrast in fishing pressure exhibited little evidence of bias in
+terminal SSB estimation except for R+M EMs when there was no AR1
+correlation in catchability process errors (Figure \ref{SSB_rel_error},
+D). Many EMs also performed well in R+q OMs with low observation
+uncertainty, but no contrast in fishing pressure. For R+q OMs with high
+observation uncertainty and contrast in fishing pressure, EMs that
+estimated the Beverton-Holt stock recruit function exhibited less SSB
+bias than those that did not. Estimating median natural mortality in the
+EMs only resulted in much more variable SSB estimation errors when there
+was no contrast in fishing pressure (Figure \ref{SSB_rel_error}, D,
+first and third rows).
+
+For all OM process error types, relative errors in terminal year
+recruitment were generally more variable than SSB, but effects of R and
+R+S OM and EM attributes on bias (i.e, negative or positive or none)
+were similar (Figure \ref{R_rel_error}, A). Furthermore, for EM
+configurations where bias in terminal bias in SSB was evident, median
+relative errors in recruitment often indicated stronger bias in
+recruitment of the same sign.
+
+\hypertarget{beverton-holt-parameters}{%
+\subsubsection*{Beverton-Holt
+parameters}\label{beverton-holt-parameters}}
+\addcontentsline{toc}{subsubsection}{Beverton-Holt parameters}
 
@@
 there was little effect of estimating natural mortality (Figure
-\ref{naa_om_SR_relbias}). For other R and R+S OMs, estimating natural
+\ref{SR_rel_error}, A). For other R and R+S OMs, estimating natural
 mortality often resulted in less biased estimation of stock-recruit
-parameters. The range of confidence intervals for the median parameter
-bias suggests large variability of the parameter estimates.
-
-In R+M OMs, the most accurate estimation of stock-recruit parameters
-occurred when there was a change in fishing pressure combined with
-either low variability in natural mortality process errors and high
-observation error or vice versa (Figure \ref{M_om_SR_relbias}). Relative
-to the R, and R+S OMs, there was even less effect of estimating mean
-log-natural mortality on estimation bias for the stock- recruit
-relationship parameters, but similarly there was generally large
-variability of the parameter estimates.
-
-In R+Sel OMs, bias for stock-recruit parameters was very large for OMs
-with constant fishing pressure (Figure \ref{Sel_om_SR_relbias}). Less
-bias in parameter estimation occurred for OMs with a change in fishing
-pressure over time. There was little effect of estimating natural
-mortality. Patterns in bias of stock-recruit parameter estimation in R+q
-OMs, were very similar to those in R+Sel OMs with poor estimation of
-these parameters in OMs with constant fishing pressure (Figure
-\ref{q_om_SR_relbias}). Like other OMs, there was generally large
-variability of the parameter estimates.
-
-\hypertarget{mean-natural-mortality-rate-estimation}{%
-\subsubsection*{(Mean) natural mortality rate
-estimation}\label{mean-natural-mortality-rate-estimation}}
-\addcontentsline{toc}{subsubsection}{(Mean) natural mortality rate
-estimation}
-
-The mean natural mortality rate is estimated pretty accurately for R and
-R+S OMs when EMs use the correct process error configuration and there
-is a change in fishing pressure over time and observation error is low
-(Figure \ref{naa_om_M_relbias}). Estimation is a little more variable
-when observation error is higher or fishing pressure is constant, but
-when OMs have both, estimation is extremely variable. There is little
-effect of estimating stock-recruit relationships on the bias in
-estimating natural mortality. For R+M OMs, effects on estimating mean
-log-natural mortality were similar to the R and R+S OMs: best estimation
-with a change in fishing pressure and low observation error and worst
-with no change in fishing pressure and higher observation error (Figure
-\ref{M_om_M_relbias}). The same patterns also occurred for R+Sel and R+q
-OMs (Figures \ref{Sel_om_M_relbias} and \ref{q_om_M_relbias}).
+parameters. There was generally large variability in relative errors of
+the stock-recruit parameter estimates, but the lowest variability
+occurred with low variability in recruitment and little or no
+variability in survival process errors (\(\sigma_{2+} \in \{0,0.25\}\)),
+and contrast in fishing pressure. Across al R and R+S OMs, relative
+errors for the \(a\) parameter were often less variable than those for
+\(b\).
+
+In R+M OMs, the most accurate estimation of stock-recruit parameters for
+all EM process error assumptions occurred when there was a change in
+fishing pressure combined with either low variability in natural
+mortality process errors and high observation error or vice versa
+(Figure \ref{SR_rel_error}, B). Relative to the R, and R+S OMs, there
+was even less effect of estimating median natural mortality on
+estimation bias for the stock-recruit relationship parameters, but
+similarly there was generally large variability of the parameter
+estimates.
+
+Bias for stock-recruit parameters was very strong and variability in
+relative errors was greatest for most EMs fit to R+Sel OMs with constant
+fishing pressure (Figure \ref{SR_rel_error}, C). Less bias in parameter
+estimation occurred for OMs with a change in fishing pressure over time
+and of those OMs and the most accurate estimation occurred for EMs with
+process error types that matched the OMs and the OMs had low observation
+uncertainty, contrast in fishing mortality and more variable and
+uncorrelated selectivity process errors. There was little effect of
+estimating natural mortality on relative errors for stock recruit
+parameters.
+
+Like R+Sel OMs, relative errors in stock-recruit parameters were less
+variable for most EM process error types when OMs had contrast in
+fishing pressure and those OMs that also had lower observation
+uncertainty generally had the least variability (Figure
+\ref{SR_rel_error}, D). The worst accuracy of stock-recruit parameter
+estimation regardless of EM type occured when R+q OMs had low
+observation uncertainty and constant fishing pressure (Figure
+\ref{SR_rel_error}, D, fourth row).
+
+\hypertarget{median-natural-mortality-rate}{%
+\subsubsection*{Median natural mortality
+rate}\label{median-natural-mortality-rate}}
+\addcontentsline{toc}{subsubsection}{Median natural mortality rate}
+
+Median natural mortality rate is estimated accurately by all EM process
+error types for all R OMs except those where observation uncertainty is
+high and fishing pressure is constant where variability in relative
+errors is high (Figure \ref{M_rel_error}, A, \(\sigma_{2+} = 0\)). For
+R+S OMs estimation of median natural mortality rate is most accurate
+when observation uncertainty is low and there is contrast in fishing
+pressure and the EM process error type is correct. There is little
+effect of estimating stock-recruit relationships on the bias for median
+natural mortality.
+
+For R+M OMs, median natural mortality was estimated most accurately,
+regardless of EM process error type, when OMs had a change in fishing
+pressure and low observation uncertainty (Figure \ref{M_rel_error}, B).
+However, those R+M OMs that also had greatest variability in AR1
+correlated natural mortality process errors only had unbiased estimation
+when the EM process error type was correct. Like R, and R+S OMs, there
+was little effect of estimating stock-recruit relationships on the bias
+for median natural mortality.
+
+All EM process error types produced accurate estimation of median
+natural mortality rate for R+Sel OMs that had contrast in fishing
+pressure, low observation uncertainty, and low selectivity process error
+variability (Figure \ref{M_rel_error}, C). When selectivity process
+error variability increased the incorrect EM process errors produce more
+biased estimation of median natural mortality rate. The least accurate
+estimation occurred for all EM process error types when observation
+uncertainty was high and there was constant fishing pressure. Again,
+like other OM process error types, there was little effect of estimating
+stock-recruit relationships on the bias for median natural mortality.
+
+Like R+Sel OMs, all EM process error types produced accurate estimation
+of median natural mortality rate when fit to R+q OMs with contrast in
+fishing pressure, low observation uncertainty and low catchability
+process error variability (Figure \ref{M_rel_error}, D). Most EM process
+error types produced biased estimation of median natural mortality when
+R+q OMs had high observaiton uncertainty and constant fishing pressure.
+Again, there was no discernible effect of estimating stock-recruit
+relationships on bias.
 
@@
 
-\hypertarget{r-rs-operating-models-4}{%
-\subsubsection*{R, R+S operating models}\label{r-rs-operating-models-4}}
-\addcontentsline{toc}{subsubsection}{R, R+S operating models}
-
-When natural mortality is known and no stock recruit relationship is
-estimated, median absolute Mohn's \(\rho\) was less than 0.25 for all
-EMS fit to simulated data sets from R and R+S OMs (Figure
-\ref{naa_om_em_R_MF_mohns_rho}). The strongest median Mohn's \(\rho\)
-values for F and SSB were obtained for EMs that did not match the R+S
-OMs with the largest variability in survival process errors, change in
-fishing pressure over time, and higher observation error. Mohn's
-\(\rho\) was also greater than 0 for recruitment for all EMS, including
-the matching EM, fit to R OMs that had the largest variability in
-recruitment random effects and high observation error. Estimating
-natural mortality removed the retrospective patterns in F and SSB in R+S
-OMs with the largest variability in survival process errors, change in
-fishing pressure over time, and higher observation error (Figure
-\ref{naa_om_em_R_ME_mohns_rho}). Estimating a stock-recruit relationship
-had no discernible effect on the retrospective patterns (Figure
-\ref{naa_om_em_BH_MF_mohns_rho}). Estimating natural mortality and the
-stock-recruit relationship increased the variablity of Mohn's \(\rho\)
-for recruitment particularly for OMs with high observation error (Figure
-\ref{naa_om_em_BH_ME_mohns_rho}).
-
-\hypertarget{rm-operating-models-4}{%
-\subsubsection*{R+M operating models}\label{rm-operating-models-4}}
-\addcontentsline{toc}{subsubsection}{R+M operating models}
-
-When mean log-natural mortality is known and no stock recruit
-relationship is estimated, median abslolute Mohn's \(\rho\) was less
-than 0.125 for all EMS fit to simulated data sets from R+M OMs (Figure
-\ref{M_om_em_R_MF_mohns_rho}). The strongest median Mohn's \(\rho\)
-values for F, SSB, and recruitment were obtained for EMs other than R+M
-with autocorrelation assumed for OMs with a change in fishing pressure
-over time, and higher observation error. Like R and R+S OMs, estimating
-natural mortality removed the retrospective patterns in the respective
-OMs (Figure \ref{M_om_em_R_ME_mohns_rho}). Estimating a stock-recruit
-relationship also had the same effect as estimating the mean log-natural
-mortality rate, but Mohn's \(\rho\) values were more variable for R+M
-EMs with autocorrelation assumed (Figure \ref{M_om_em_BH_MF_mohns_rho}).
-Estimating natural mortality and the stock-recruit relationship
-increased the variablity of Mohn's \(\rho\) for F, SSB, and recruitment,
-particularly for OMs with high observation error and constant fishing
-pressure over time (Figure \ref{M_om_em_BH_ME_mohns_rho}).
-
-\hypertarget{rsel-operating-models-4}{%
-\subsubsection*{R+Sel operating models}\label{rsel-operating-models-4}}
-\addcontentsline{toc}{subsubsection}{R+Sel operating models}
-
-Similar to R+M OMs, when mean log-natural mortality is known and no
-stock recruit relationship is estimated, median abslolute Mohn's
-\(\rho\) was less than 0.125 for all EMS fit to simulated data sets from
-R+Sel OMs (Figure \ref{Sel_om_em_R_MF_mohns_rho}). The strongest median
-Mohn's \(\rho\) values for F, SSB, and recruitment were obtained for all
-EMs with OMs that had a change in fishing pressure over time, and higher
-observation error. Estimating natural mortality increased the
-variability in Mohn's \(\rho\) values, particularly for OMs with higher
-observation error and constant fishing (Figure
-\ref{Sel_om_em_R_ME_mohns_rho}). Estimating a stock-recruit relationship
-also removed any tendency for retrospective patterns in OMs with a
-change in fishing pressure and higher observation error (Figure
-\ref{Sel_om_em_BH_MF_mohns_rho}). Estimating natural mortality and the
-stock-recruit relationship further increased the variablity of Mohn's
-\(\rho\) for F, SSB, and recruitment, particularly for OMs with high
-observation error and constant fishing pressure over time (Figure
-\ref{Sel_om_em_BH_ME_mohns_rho}).
-
-\hypertarget{rq-operating-models-4}{%
-\subsubsection*{R+q operating models}\label{rq-operating-models-4}}
-\addcontentsline{toc}{subsubsection}{R+q operating models}
-
-Similar to R+M and R+Sel OMs, when mean log-natural mortality is known
-and no stock recruit relationship is estimated, median abslolute Mohn's
-\(\rho\) was less than 0.125 for all EMS fit to simulated data sets from
-R+q OMs (Figure \ref{q_om_em_R_MF_mohns_rho}). Also similar to other
-OMs, the strongest median Mohn's \(\rho\) values for F, SSB, and
-recruitment were obtained for with OMs that had a change in fishing
-pressure over time, and higher observation error. Estimating natural
-mortality increased the variability in Mohn's \(\rho\) values,
-particularly for OMs with constant fishing and autocorrelation of
-process errors in catchability (Figure \ref{q_om_em_R_ME_mohns_rho}).
-Similar to other OMs, estimating a stock-recruit relationship removed
-any tendency for retrospective patterns in OMs with a change in fishing
-pressure and higher observation error (Figure
-\ref{q_om_em_BH_MF_mohns_rho}). Like R+Sel OMs, estimating natural
-mortality and the stock-recruit relationship further increased the
-variablity of Mohn's \(\rho\) for F, SSB, and recruitment, particularly
-for OMs with high observation error and constant fishing pressure over
-time (Figure \ref{q_om_em_BH_ME_mohns_rho}).
+Mohn's \(\rho\) for SSB was small in absolute value for all R and R+S
+OMs, regardless of EM process error types, and whether median natural
+mortality rate or stock-recruit relationships were estimated (Figure
+\ref{mohns_rho_ssb}, A). The strongest retrospective patterns occurred
+in OMs with the largest survival process error variability, high
+observation uncertainty, and contrast in fishing pressure, but only for
+EMs without the correct process error type and where median natural
+mortality rate was assumed known. For R+M, R+Sel, and R+q OMs, Mohn's
+\(\rho\) was also small in absolute value, but median values were all
+closer to 0 than the largest values in the R and R+S OMs (Figure
+\ref{mohns_rho_ssb},B-D). For these OMs, there was no noticeable effect
+of estimation of median natural mortality rate or stock recruit
+relationship on Mohn's \(\rho\) for any EM process error types.
+
+Mohn's \(\rho\) for recruitment was small in absolute value for all R
+OMs with low variability in recruitment process errors, regardless of EM
+process error type, and whether median natural mortality rate or
+stock-recruit relationships were estimated (Figure \ref{mohns_rho_R},
+A). However, R and R+S OMs with greater recruitment process variability
+and higher observation uncertainty had median Mohn's \(\rho\) for
+recruitment greater than zero for most EMs even when the the EM process
+error type was correct. In R+S OMs with lower observation uncertainty,
+EMs with the correct process error type exhibited better median Mohn's
+\(\rho\) close to 0 than EMs with the incorrect process error type. For
+R+M, R+Sel, and R+q OMs, results for Mohn's \(\rho\) for recruitment are
+similar to those for SSB, but the range in median values and variation
+in Mohn's \(\rho\) values for a given OM are generally larger for
+recruitment (Figure \ref{mohns_rho_R}, B-D).
 
@@
 
-Convergence results can be useful for understanding how bad convergence
-in applications to real data might direct the practitioner to which
-alternative random effects configurations is be more appropriate.
+Convergence results can be useful for understanding how lack of
+convergence in applications to real data might direct the practitioner
+to which alternative random effects configurations is more appropriate.
 Therefore, the type of convergence that we might use as a diagnostic is
 important. It is common during assessment model fitting to check that
-the maximum absolute gradient component is less than some threshold. Our
-Type 3 convergence represents this and we used 1e-6 as the threshold,
-but there is no standard. Furthermore we found the hessian at the
-optimized log-likelihood can often be invertible (Convergence types 4
-and 5) when the gradient approach to convergence fails. The probability
-of convergence type 5 cannot be greater than that of type 4 because type
-5 is a subset where none of the standard errors obtained by inverting
-the hessian are greater than a specified value (100 in this case). We
-presume that the parameters that cause type 4 convergence to fail will
-also typically be the same as those that cause type 5 to fail. Perhaps a
-higher gradient threshold for Type 3 convergence would be more
-consistent with hessian-based convergence, but the gradient and
-hessian-based standard errors should be inspected. Parameters that are
-not well estimated can lead the assessment scientist toward a more
-appropriate and often more parsimonious model configuration. For
-example, variance parameters that are estimated on log scale as low
-negative number (variance is close to 0) can happen for random effects
-estimated with little or no variability. These may sometimes lead a
-hessian to not be invertible or due to numerical issues may be
-invertible, but provide very large standard errors. More generally, we
-can expect lower likelihood of convergence when estimating natural
-mortality or stock-recruit relationships because the typical data in
-assessment models does not strongly inform these parameters.
+the maximum absolute gradient component is less than some threshold, but
+there is no standard. However, checking the Hessian for positive
+definiteness is sometimes done conditional on the convergence criterion.
+As such, some models may be excluded that in fact have an invertible
+hessian. We found the hessian at the optimized log-likelihood can often
+be invertible when the gradient approach with a sensible criterion
+(e.g., \(10^{-3}\)) would indicate lack of convergence.
+
+Another factor affecting the convercence criteria is numerical accuracy.
+With the process errors modeled as random effects we are optimizing an
+(Laplace) approximation of the marginal log-likelihood and therefore
+hessians and gradients are also with regard to the approximation.
+Furthermore, numerical methods are used to calculate the hessian
+(Optimhess?), and also to invert it for variance estimation. All of
+these approximations are affecting the gradients and invertibility of
+the hessian (although likely worse for some models than others), which,
+along with our results, suggests at least investigating whether the
+hessian is positive definite when the calculated gradients are not
+terribly large.
+
+Often high gradients and/or lack of positive definite hessian result
+when selectivity parameter estimates are at their bounds (0,1) or random
+effects variance parameter estimates are essentially 0. Changing the
+parameter configuration to avoid the bounds or removing process errors
+with 0 variance will often lead to satisfactory convergence.
+
+Perhaps a greater gradient threshold would be more consistent with
+hessian-based convergence, but the gradient and hessian-based standard
+errors should be inspected.
+
+Parameters that are not well estimated can lead the assessment scientist
+toward a more appropriate and often more parsimonious model
+configuration. For example, variance parameters that are estimated on
+log scale as low negative number (variance is close to 0) can happen for
+random effects estimated with little or no variability. These may
+sometimes result in a hessian that is not positive definite or due to
+numerical issues may be invertible, but provide very large standard
+errors. Generally, we can expect lower probability of convergence when
+estimating natural mortality or stock-recruit relationships because of
+the difficulty distinguishing these parameters from others being
+estimated in assessment model with data typically available.
 
@@
 Among the process error configurations we used in OMs, we found AIC to
-be accurate for process errors on recruitment and survival. Fitting
-models to other OMs rarely preferred R+S EMs and R and R+S EMs were
-nearly always selected for the matching OMs. For other sources of
+be accurate for process errors on recruitment and survival (R and R+S).
+Fitting models to other OMs rarely preferred R+S EMs and R and R+S EMs
+were nearly always selected for the matching OMs. For other sources of
 process error, accuracy of AIC was improved to useful levels when there
-was larger variability in the process errors and/or lower observation
-error. Therefore, when data are relatively good and an there is large
-variability in the unknown process errors, AIC is likely to perform
-well.
+was larger variability in the process errors and/or better data quality.
 
 AIC generally performs weakly in determining stock recruit relationships
-unless there is large contrast in SSB. Interestingly, Miller et al.
-(2016) found AIC to prefer a stock-recruit relationship for SNEMA
-yellowtail flounder where there was a large change in stock size
-estimated over time and flatfish are well observed by the bottom trawl
-indices relative to some other guilds such as semi-pelagic species.
-Although we did not compare models with alternative stock-recruit
-relationships, we do not expect AIC to perform any better distinguishing
-these relationships. Our results comport well with those of de Valpine
-and Hastings (2002) where state-space stock-recruit models were fit just
-to SSB and recruitment ``observations''.
+unless there is large contrast in SSB and low variation in recruitment
+process errors. Although we did not compare models with alternative
+stock-recruit relationships, we do not expect AIC to perform any better
+distinguishing these relationships. Our results comport well with those
+of de Valpine and Hastings (2002) where state-space stock-recruit models
+were fit just to SSB and recruitment ``observations'' and Britten et al.
+which found AIC could not reliably distinguish alternative environmental
+effects on stock-recruit parameters. Miller et al. (2016) found AIC to
+prefer a stock-recruit relationship with environmental effects for SNEMA
+yellowtail flounder, but there was a large change in stock size
+estimated over time and flatfish are well-observed by the NEFSC bottom
+trawl survey used for an index in the assessment model.
 
@@
 As we might expect bias in all parameters and assessment output was
-generally less with lower observation error in indices and age
-composition. Reliable estimation of stock-recruit relationship
-parameters only appears possible in ideal situations with lower
-observation errors in age composition and indices, lower variability in
-recruitment process errors and large contrast in spawning biomass over
-time. The latter can occur in with larger variability in survival or
-natural mortality process errors, variation in fishing pressure.
-
-Estimation of natural mortality (or the mean when there are process
-errors on natural mortality) appears more feasible. We found low or no
-evidence of bias for many OM process error assumptions when there was
-contrast in fishing pressure even when there was greater observation
-error although it can lead to more variable estimates of natural
-mortality. For OMs where there was bias in natural mortality due to high
-observation error (R+Sel, R+q), estimating the stock-recruit
-relationship seemed to remove the bias. However, estimation of natural
-mortality can cause large differences between the true and estimated SSB
-(that may be unbiased on average) when there is less contrast in fishing
-pressure over time and higher observation error.
+generally better with lower observation error. Reliable estimation of
+stock-recruit relationship parameters only appears possible in ideal
+situations with lower observation errors in age composition and indices,
+lower variability in recruitment process errors and large contrast in
+spawning biomass over time. The latter can occur with larger variability
+in survival or natural mortality process errors, or variation in fishing
+pressure.
+
+Estimation of median natural mortality appears more feasible than SR
+parameters. We found low or no evidence of bias for many OM process
+error assumptions when there was contrast in fishing pressure even when
+there was greater observation error although it can lead to more
+variable estimates of natural mortality. For OMs where there was bias in
+natural mortality due to high observation error (R+Sel, R+q), estimating
+the stock-recruit relationship seemed to remove the bias.
+
+However, estimation of natural mortality can cause large differences
+between the true and estimated SSB (that may be unbiased on average)
+when there is less contrast in fishing pressure over time and higher
+observation error.
+
+Estimating B-H reduced SSB bias in some situations
 
@@
 
+\hypertarget{conclusions}{%
+\subsection*{Conclusions}\label{conclusions}}
+\addcontentsline{toc}{subsection}{Conclusions}
+
+All together what do results mean? E.g., when do you get good
+convergence and retrospective patterns and high/none bias? When do you
+get good convergence and no retros and no/big bias?
+
+When R+S is preferred it probably means its the correct process error OF
+THE TYPES CONSIDERED HERE.
+
 \hypertarget{summary}{%
@@
 each of 100 simulated set of observations with alternative assumptions
-type and correlation structure of process errors, (mean) natural
+type and correlation structure of process errors, median natural
 mortality was known or estimated, and a B-H stock recruit relationship
@@
 and higher variation in spawning biomass over the time series. When the
-(mean) natural mortality rate was estimated, we found large bias and
+median natural mortality rate was estimated, we found large bias and
 uncertainty much more likely in model output such as spawning stock
@@
 observation error is high. When models did exhibit some retrospective
-pattern, estimation the mean natural mortality rate tended to remove it.
+pattern, estimation the median natural mortality rate tended to remove
+it.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-brittenetalinreview}{}}%
+Britten, G., Brooks, E.N., and Miller, T.J. Reliability of
+stock-recruitment function estimation in state space assessment models.
+
 \leavevmode\vadjust pre{\hypertarget{ref-cadigan16}{}}%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-clopperpearson34}{}}%
-Clopper, C.J., and Pearson, E.S. 1934. The use of confidence or fiducial
-limits illustrated in the case of the binomial. Biometrika
-\textbf{26}(4): 404--413.
-doi:\href{https://doi.org/10.1093/biomet/26.4.404}{10.1093/biomet/26.4.404}.
+\leavevmode\vadjust pre{\hypertarget{ref-carvalhoetal21}{}}%
+Carvalho, F., Winker, H., Courtney, D., Kapur, M., Kell, L., Cardinale,
+M., Schirripa, M., Kitakado, T., Yemane, D., Piner, K.R., Maunder, M.N.,
+Taylor, I., Wetzel, C.R., Doering, K., Johnson, K.F., and Methot, R.D.
+2021. A cookbook for using model diagnostics in integrated stock
+assessments. Fisheries Research \textbf{240}: 105959.
+doi:\url{https://doi.org/10.1016/j.fishres.2021.105959}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-connetal10}{}}%
+Conn, P.B., Williams, E.H., and Shertzer, K.W. 2010. When can we
+reliably estimate the productivity of fish stocks? Canadian Journal of
+Fisheries and Aquatic Sciences \textbf{67}(3): 511--523.
+doi:\href{https://doi.org/10.1139/F09-194}{10.1139/F09-194}.
 
@@
 effects and observations in state-space assessment models. For Canadian
-Journal of Fisheries and Aquatic Sciences.
+Journal of Fisheries and Aquatic Sciences, State-space special
+collection.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-magnussonhilborn07}{}}%
+Magnusson, A., and Hilborn, R. 2007. What makes fisheries data
+informative? Fish and Fisheries \textbf{8}(4): 337--358.
+doi:\url{https://doi.org/10.1111/j.1467-2979.2007.00258.x}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-millerbrooks21}{}}%
@@
 assessment model with an application to {S}outhern {N}ew {E}ngland
-yellowtail flounder.\cjfas \textbf{73}(8): 1261--1270.
+yellowtail flounder. Canadian Journal of Fisheries and Aquatic Sciences
+\textbf{73}(8): 1261--1270.
 
@@
 ({WHAM}). Available from \url{https://timjmiller.github.io/wham/}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-mohn99}{}}%
+Mohn, R. 1999. The retrospective problem in sequential population
+analysis: An investigation using cod fishery and simulated data. ICES
+Journal of Marine Science \textbf{56}(4): 473--488.
+doi:\href{https://doi.org/10.1006/jmsc.1999.0481}{10.1006/jmsc.1999.0481}.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-thulin14}{}}%
-Thulin, M. 2014. {The cost of using exact confidence intervals for a
-binomial proportion}. Electronic Journal of Statistics \textbf{8}(1):
-817--840.
-doi:\href{https://doi.org/10.1214/14-EJS909}{10.1214/14-EJS909}.
-
 \leavevmode\vadjust pre{\hypertarget{ref-wiedenmannetal19}{}}%
@@
 \begin{center}
-\includegraphics{M_bias_plots}
-\end{center}
-\caption{Median relative error of median/constant natural mortality for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{M_rel_error}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
 \includegraphics{sr_bias_plots}
@@
 Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{SR_rel_error}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{M_bias_plots}
+\end{center}
+\caption{Median relative error of median natural mortality for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{M_rel_error}
 \end{figure}

### 2025-02-13 Project_0/manuscript/manuscript.Rmd
saving work on project_0 manuscript

@@
   - Christopher M. Legault^2^
+  - Chengxue Li^2^
   - Brandon Muffley^5^
@@
 
-Random effects can be included in state-space assessment models on many processes and in many ways, and guidance is needed on statistical reliability and model selection criteria.
-We simulated 72 operating models with varying fishing history, observation error uncertainty, and process error magnitude, correlation, and source (recruitment, survival, fishery selectivity, catchability, and natural mortality).
+State-space assessment models, which are increasingly being used for management of commercially important fish stocks, can include process errors in many ways, and an evaluation of the statistical reliability of the estimates of management output and best practices of selection among alternative process error configurations are needed.
+We simulated 72 operating models with varying fishing pressure history, observation error error, and process error magnitude, correlation, and source (recruitment, survival, fishery selectivity, catchability, and natural mortality).
 We fit estimating models with different assumptions on the process error source, whether median natural mortality was estimated, and whether a stock-recruit relationship was estimated.
-
-Models that assumed the correct process error source had high convergence and low bias.
-Bias was also low under most process error assumptions when there was contrast in fishing pressure.
-Stock-recruitment parameters were only reliably estimated in ideal situations.
+Estimating models that assumed the correct process error source and median natural mortality rate was known, and did not estimate a stock recruit relationship generally had high probability of convergence and low bias in terminal year SSB estimation.
+Bias was also low under many incorrect process error assumptions when there was contrast in fishing pressure and low observation error.
+Stock-recruitment parameters were not reliably estimated in any scenarios, but the least accurate estimates occured under ideal situations of low observation error, low recruitment variability, and temporal contrast in fishing pressure.
+However, estimation of median natural mortality rate was reliable for many OMs (other than those with survival process errors) with temporal contrast in fishing pressure regardless of the estimated process error type.
 Marginal AIC most accurately distinguished process errors on recruitment, survival, and selectivity, as well as larger magnitude process errors of other types.
-Retrospective patterns were generally weak except for recruitment when observation error was high, even with the correct process error assumptions.
-When models did exhibit some retrospective pattern, estimating median natural mortality often removed it.
+Retrospective patterns for spawning stock biomass and fishing mortality were generally weak, but those for recruitment were sizable when observation error was high, even with the correct process error assumptions.
 
@@
 
-Much is known about the reliability of state-space models that are linear or Gaussian [@aeberhardetal18], but applications in fisheries management are nonlinear and typically include multiple types of observations with varying distributional assumptions.
-Furthermore, there is a wide range of potential random effects structures and model parameters that can be treated as random effects in assessment models.
-We know relatively little about the factors affecting statistical reliability of such models or the ability of information criteria to distinguish among such alternative structures.
+Although the statistical aspects of state-space models and their application have been studied extensively, the work is primarily on Gaussian state-space models.
+State-space stock assessment models, where there are nonlinear functions of latent parameters and multiple types of observations with varying distributional assumptions, is one of the most complex examples of this analytical approach.
 
-@lietal24 investigated some aspects of inferences for operating models with multiple sources of process error, but there are differences for this paper.
+Much is known about the reliability of state-space models that are linear or Gaussian [@aeberhardetal18], but not for the more complex application in stock assessment modeling.
+
+[@lietal24 investigated some aspects of inferences for operating models with multiple sources of process error, but there are differences for this paper.]
+
+There is a wide range of model parameters that can be treated as random effects in assessment models and a variety of potential distributions for the random effects.
+We know relatively little about the factors affecting statistical reliability of such models or the ability to distinguish among such alternative structures.
 
@@
 
-Here we conduct a simulation study with operating models (OMs) varying by degree of observation error uncertainty, source and variability of process error, and fishing history.
-The simulations from these OMs are fitted with estimation models (EMs) that make alternative assumptions for sources of process error, whether a stock-recruit model was estimated, and whether M is estimated.
-We evaluate whether AIC can correctly determine the correct source of process error and stock effects on recruitment.
-We also evaluate the degree of bias in the outputs of the assessment model that are important for management.
+Here we conduct a simulation study with operating models (OMs) varying by degree of observation error error, source and variability of process error, and fishing history.
+The simulations from these OMs are fitted with estimation models (EMs) that make alternative assumptions for sources of process error, whether a stock-recruit model was estimated, and whether a constant, or, in some EMs, median, natural mortality is estimated.
+We evaluate whether AIC can correctly determine the correct source of process error and a stock-recruit relationship.
+We also evaluate when retrospective patterns occur and the degree of bias in the outputs of the assessment model that are important for management.
 
@@
 
-We used the Woods Hole Assessment Model (WHAM) to configure operating and EMs in our simulation study [@millerstock20;@stockmiller21].
+We used the Woods Hole Assessment Model (WHAM) to configure OMs and EMs in our simulation study [@millerstock20;@stockmiller21].
 WHAM is an R package freely available as a github repository.
 For this study we used [version 1.0.6.9000, commit 77bbd94](https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3).
-This package has also been used to configure operating and EMs for closed loop simulations evaluating index-based assessment methods [@legaultetal23] and is used for management of haddock, butterfish, American plaice, bluefish, Atlantic cod, black sea bass, and yellowtail flounder in the Northeast US.
+This package has also been used to configure OMs and EMs for closed loop simulations evaluating index-based assessment methods [@legaultetal23] and is used for management of haddock, butterfish, American plaice, bluefish, Atlantic cod, black sea bass, and yellowtail flounder in the Northeast US.
 
@@
 R+S EMs fit to R+S OMs had highest convergence rates when there was contrast in fishing pressure and low observation error.
-Convergence rates were high for all EMs when fit to data from R+S OMs with lower observation uncertainty except those where median natural mortality and/or stock-recruit relationships were estimated.
+Convergence rates were high for all EMs when fit to data from R+S OMs with lower observation error except those where median natural mortality and/or stock-recruit relationships were estimated.
 
@@
 
-R+S EMs, in particular, converged poorly when fit to data generated from R+Sel OMs with lower selectivity process error variability or higher observation uncertainty (Figure \ref{hessian_SE_convergence}, C).
-R+Sel EMs generally converged better than other EMs for R+Sel OMs with higher process error variability, lower observation uncertainty, and contrast in fishing pressure  regardless of whether median natural mortality or a stock recruit relationship was estimated.
+R+S EMs, in particular, converged poorly when fit to data generated from R+Sel OMs with lower selectivity process error variability or higher observation error (Figure \ref{hessian_SE_convergence}, C).
+R+Sel EMs generally converged better than other EMs for R+Sel OMs with higher process error variability, lower observation error, and contrast in fishing pressure  regardless of whether median natural mortality or a stock recruit relationship was estimated.
 
@@
 Adding estimation of median natural mortality or a stock recruit relationship separately has a negligible effect on the accuracy of determining the correct process error assumption.
-When both are estimated, there is a noticeable reduction in accuracy when OMs have a constant fishing history, observation uncertainty is low and larger variability in recruitment process errors.
+When both are estimated, there is a noticeable reduction in accuracy when OMs have a constant fishing history, observation error is low and larger variability in recruitment process errors.
 
@@
 
-Marginal AIC most accurately determined the correct source of process error and correlation structure for R+Sel OMs with low observation uncertainty (Figures \ref{pe_aic}, C).
-When there was low variability in selectivity process errors and high observation uncertainty, R+q or R+S EMs were more likely to have the best AIC.
+Marginal AIC most accurately determined the correct source of process error and correlation structure for R+Sel OMs with low observation error (Figures \ref{pe_aic}, C).
+When there was low variability in selectivity process errors and high observation error, R+q or R+S EMs were more likely to have the best AIC.
 Whether median natural morality or stock recruit relationships were estimated appeared to have little effect on the performance of AIC.
@@
 Marginal AIC most accurately determined the correct source of process error and correlation structure for R+q OMs with high variability in catchability process errors (Figures \ref{pe_aic},D).
-The worst accuracy occurred for OMs with low variability in catchability process errors and high observation uncertainty.
+The worst accuracy occurred for OMs with low variability in catchability process errors and high observation error.
 However, in these OMs, the marginal AIC accurately determined the correct source of process error (but not correlation structure) except when EMs estimated both median natural morality and the stock recruit relationships and OMs assumed a constant fishing pressure.
@@
 The weakest overall accuracy of AIC for the Beverton-Holt stock-recruit function was when there was constant fishing pressure (Figure \ref{sr_aic}, A, second and fourth rows).
-The largest effect of simultaneously estimating median natural mortality was when there was OMs had high observation uncertainty and constant fishing pressure (Figure \ref{sr_aic}, A, second row).
+The largest effect of simultaneously estimating median natural mortality was when there was OMs had high observation error and constant fishing pressure (Figure \ref{sr_aic}, A, second row).
 
@@
 The largest variation in SSB occurred in OMs with larger natural mortality variability and temporal AR1 correlation of the annual random effects ($\sigma_M = 1.5$, $\rho_M = 0.9$; Figure \ref{sr_aic}, B, right-most column) where AIC accuracy was also high even when there was constant fishing pressure.
-Interestingly, there seemed to be little difference in AIC accuracy whether there was lower or higher observation uncertainty or whether the EM simultaneously estimated median natural mortality.
+Interestingly, there seemed to be little difference in AIC accuracy whether there was lower or higher observation error or whether the EM simultaneously estimated median natural mortality.
 
@@
 Like the R+Sel OMs, the greatest accuracy for AIC in selecting the Beverton-Holt stock-recruit relationship occurred for R+q OMs where there was contrast in fishing pressure over time which is also where there was the greatest variation in SSB (Figure \ref{sr_aic}, D, first and third rows).
-There were some differences in effects of SSB variation on AIC accuracy when OMs had high observation uncertainty and constant fishing pressure.
-Across all of the OMs in Figure \ref{sr_aic}, the only estimated decline in AIC accuracy with SSB variation occurred when R+q OMs had high observation uncertainty and constant fishing pressure and greater variation in catchability process errors that also had AR1 autocorrelation.
+There were some differences in effects of SSB variation on AIC accuracy when OMs had high observation error and constant fishing pressure.
+Across all of the OMs in Figure \ref{sr_aic}, the only estimated decline in AIC accuracy with SSB variation occurred when R+q OMs had high observation error and constant fishing pressure and greater variation in catchability process errors that also had AR1 autocorrelation.
 
@@
 
-For R OMs ($\sigma_{2+} = 0$), there was no indication of bias (95\% confidence intervals included 0) in terminal year SSB for any of the estimating models regardless of process error assumptions, except when no SR assumption was made, recruitment variability was low, and there was contrast in fishing mortality and high observation uncertainty (Figure \ref{SSB_rel_error}, A).
-However, errors in terminal SSB estimates were highly variable when median natural mortality was estimated and there was constant fishing pressure and high observation uncertainty (Figure \ref{SSB_rel_error}, A, second row).
+For R OMs ($\sigma_{2+} = 0$), there was no indication of bias (95\% confidence intervals included 0) in terminal year SSB for any of the estimating models regardless of process error assumptions, except when no SR assumption was made, recruitment variability was low, and there was contrast in fishing mortality and high observation error (Figure \ref{SSB_rel_error}, A).
+However, errors in terminal SSB estimates were highly variable when median natural mortality was estimated and there was constant fishing pressure and high observation error (Figure \ref{SSB_rel_error}, A, second row).
 
-For R+S OMs, the EMs with matching process error assumption generally produced unbaised estimation of terminal SSB except when median natural mortality was estimated and there was high observation uncertainty.
-In R+S OMs with low observation uncertainty, EMs with incorrect process error assumptions typically provided biased estimation of terminal year SSB.
+For R+S OMs, the EMs with matching process error assumption generally produced unbaised estimation of terminal SSB except when median natural mortality was estimated and there was high observation error.
+In R+S OMs with low observation error, EMs with incorrect process error assumptions typically provided biased estimation of terminal year SSB.
 Estimating the Beverton-Holt stock-recruit relationship had little discernible effect on bias of terminal year SSB estimation whereas estimating median M tended to produce more variability in errors in terminal SSB estimation similar to R OMs.
 
-For R+ M OMs with low variability in natural mortality process errors, low observation uncertainty and contrast in fishing motality over time all EMs produced low variability in SSB estimation error that indicated unbiasedness (Figure \ref{SSB_rel_error}, B, third row).
+For R+ M OMs with low variability in natural mortality process errors, low observation error and contrast in fishing motality over time all EMs produced low variability in SSB estimation error that indicated unbiasedness (Figure \ref{SSB_rel_error}, B, third row).
 However, larger variability in natural mortality process errors increased bias of EMs without the correct process error type.
-Estimating median natural mortality increased variability of SSB estimation error particularly for OMs with high observation uncertainty and constant fishing pressure over time and increased bias in SSB estimation for many R+M OMs.
+Estimating median natural mortality increased variability of SSB estimation error particularly for OMs with high observation error and constant fishing pressure over time and increased bias in SSB estimation for many R+M OMs.
 Like R and R+S OMs, estimating a stock-recruit relationship had little discernible effect on SSB bias.
 
-For R+Sel OMs, there was no evidence of bias for any EMs when variability in selectivity process errors was low, observation uncertainty low and contrast in fishing mortality (Figure \ref{SSB_rel_error}, C).
-The largest bias for all EMs that estimated median natural mortality when the OMs had high observation uncertainty, constant fishing pressure and greater variability in selectivity process errors ($\sigma_{\text{Sel}} = 0.5$) or low selectivity process errors ($\sigma_{\text{Sel}} = 0.1$) and low observation uncertainty.
-However, there was no evidence of bias of matching R+Sel EMs when observation uncertainty was low and variation in selectivity process errors was larger, whether median natural mortality was estimated or not (Figure \ref{SSB_rel_error}, C, third row).
-We only observed an effect of estimating the Beverton-Holt Stock recruit relationship for R+Sel OMs that had high observation uncertainty and contrast in fishing pressure where estimating the relationship produced less biased SSB estimation for many EMs (Figure \ref{SSB_rel_error}, C, top row).
+For R+Sel OMs, there was no evidence of bias for any EMs when variability in selectivity process error and observation error was low, and with contrast in fishing mortality (Figure \ref{SSB_rel_error}, C).
+The largest bias occurred for any EMs that estimated median natural mortality when the OMs had high observation error, constant fishing pressure, and greater variability in selectivity process errors ($\sigma_{\text{Sel}} = 0.5$) or low selectivity process errors ($\sigma_{\text{Sel}} = 0.1$) and low observation error.
+However, there was no evidence of bias of matching R+Sel EMs when observation error was low and variation in selectivity process errors was larger, whether median natural mortality was estimated or not (Figure \ref{SSB_rel_error}, C, third row).
+We only observed an effect of estimating the Beverton-Holt Stock recruit relationship for R+Sel OMs that had high observation error and contrast in fishing pressure where estimating the relationship produced less biased SSB estimation for many EMs (Figure \ref{SSB_rel_error}, C, top row).
 
-All EMs fit to data from R+q OMs with low observation uncertainty and  contrast in fishing pressure exhibited little evidence of bias in terminal SSB estimation except for R+M EMs when there was no AR1 correlation in catchability process errors (Figure \ref{SSB_rel_error}, D).
-Many EMs also performed well in R+q OMs with low observation uncertainty, but no contrast in fishing pressure.
-For R+q OMs with high observation uncertainty and contrast in fishing pressure, EMs that estimated the Beverton-Holt stock recruit function exhibited less SSB bias than those that did not.
+All EMs fit to data from R+q OMs with low observation error and  contrast in fishing pressure exhibited little evidence of bias in terminal SSB estimation except for R+M EMs when there was no AR1 correlation in catchability process errors (Figure \ref{SSB_rel_error}, D).
+Many EMs also performed well in R+q OMs with low observation error, but no contrast in fishing pressure.
+For R+q OMs with high observation error and contrast in fishing pressure, EMs that estimated the Beverton-Holt stock recruit function exhibited less SSB bias than those that did not.
 Estimating median natural mortality in the EMs only resulted in much more variable SSB estimation errors when there was no contrast in fishing pressure (Figure \ref{SSB_rel_error}, D, first and third rows).
@@
 Bias for stock-recruit parameters was very strong and variability in relative errors was greatest for most EMs fit to R+Sel OMs with constant fishing pressure (Figure \ref{SR_rel_error}, C).
-Less bias in parameter estimation occurred for OMs with a change in fishing pressure over time and of those OMs and the most accurate estimation occurred for EMs with process error types that matched the OMs and the OMs had low observation uncertainty, contrast in fishing mortality and more variable and uncorrelated selectivity process errors.
+Less bias in parameter estimation occurred for OMs with a change in fishing pressure over time and of those OMs and the most accurate estimation occurred for EMs with process error types that matched the OMs and the OMs had low observation error, contrast in fishing mortality and more variable and uncorrelated selectivity process errors.
 There was little effect of estimating natural mortality on relative errors for stock recruit parameters.
 
-Like R+Sel OMs, relative errors in stock-recruit parameters were less variable for most EM process error types when OMs had contrast in fishing pressure and those OMs that also had lower observation uncertainty generally had the least variability (Figure \ref{SR_rel_error}, D).
-The worst accuracy of stock-recruit parameter estimation regardless of EM type occured when R+q OMs had low observation uncertainty and constant fishing pressure (Figure \ref{SR_rel_error}, D, fourth row).
+Like R+Sel OMs, relative errors in stock-recruit parameters were less variable for most EM process error types when OMs had contrast in fishing pressure and those OMs that also had lower observation error generally had the least variability (Figure \ref{SR_rel_error}, D).
+The worst accuracy of stock-recruit parameter estimation regardless of EM type occured when R+q OMs had low observation error and constant fishing pressure (Figure \ref{SR_rel_error}, D, fourth row).
 
@@
 
-Median natural mortality rate is estimated accurately by all EM process error types for all R OMs except those where observation uncertainty is high and fishing pressure is constant where variability in relative errors is high (Figure \ref{M_rel_error}, A, $\sigma_{2+} = 0$).
-For R+S OMs estimation of median natural mortality rate is most accurate when observation uncertainty is low and there is contrast in fishing pressure and the EM process error type is correct.
+Median natural mortality rate is estimated accurately by all EM process error types for all R OMs except those where observation error is high and fishing pressure is constant where variability in relative errors is high (Figure \ref{M_rel_error}, A, $\sigma_{2+} = 0$).
+For R+S OMs estimation of median natural mortality rate is most accurate when observation error is low and there is contrast in fishing pressure and the EM process error type is correct.
 There is little effect of estimating stock-recruit relationships on the bias for median natural mortality.
 
-For R+M OMs, median natural mortality was estimated most accurately, regardless of EM process error type, when OMs had a change in fishing pressure and low observation uncertainty (Figure \ref{M_rel_error}, B).
+For R+M OMs, median natural mortality was estimated most accurately, regardless of EM process error type, when OMs had a change in fishing pressure and low observation error (Figure \ref{M_rel_error}, B).
 However, those R+M OMs that also had greatest variability in AR1 correlated natural mortality process errors only had unbiased estimation when the EM process error type was correct.
@@
 
-All EM process error types produced accurate estimation of median natural mortality rate for R+Sel OMs that had contrast in fishing pressure, low observation uncertainty, and low selectivity process error variability (Figure \ref{M_rel_error}, C).
+All EM process error types produced accurate estimation of median natural mortality rate for R+Sel OMs that had contrast in fishing pressure, low observation error, and low selectivity process error variability (Figure \ref{M_rel_error}, C).
 When selectivity process error variability increased the incorrect EM process errors produce more biased estimation of median natural mortality rate.
-The least accurate estimation occurred for all EM process error types when observation uncertainty was high and there was constant fishing pressure.
+The least accurate estimation occurred for all EM process error types when observation error was high and there was constant fishing pressure.
 Again, like other OM process error types, there was little effect of estimating stock-recruit relationships on the bias for median natural mortality.
 
-Like R+Sel OMs, all EM process error types produced accurate estimation of median natural mortality rate when fit to R+q OMs with contrast in fishing pressure, low observation uncertainty and low catchability process error variability (Figure \ref{M_rel_error}, D).
-Most EM process error types produced biased estimation of median natural  mortality when R+q OMs had high observaiton uncertainty and constant fishing pressure.
+Like R+Sel OMs, all EM process error types produced accurate estimation of median natural mortality rate when fit to R+q OMs with contrast in fishing pressure, low observation error and low catchability process error variability (Figure \ref{M_rel_error}, D).
+Most EM process error types produced biased estimation of median natural  mortality when R+q OMs had high observaiton error and constant fishing pressure.
 Again, there was no discernible effect of estimating stock-recruit relationships on bias.
@@
 Mohn's $\rho$ for SSB was small in absolute value for all R and R+S OMs, regardless of EM process error types, and whether median natural mortality rate or stock-recruit relationships were estimated (Figure \ref{mohns_rho_ssb}, A).
-The strongest retrospective patterns occurred in OMs with the largest survival process error variability, high observation uncertainty, and contrast in fishing pressure, but only for EMs without the correct process error type and where median natural mortality rate was assumed known.
+The strongest retrospective patterns occurred in OMs with the largest survival process error variability, high observation error, and contrast in fishing pressure, but only for EMs without the correct process error type and where median natural mortality rate was assumed known.
 For R+M, R+Sel, and R+q OMs, Mohn's $\rho$ was also small in absolute value, but median values were all closer to 0 than the largest values in the R and R+S OMs (Figure \ref{mohns_rho_ssb},B-D).
@@
 Mohn's $\rho$ for recruitment was small in absolute value for all R OMs with low variability in recruitment process errors, regardless of EM process error type, and whether median natural mortality rate or stock-recruit relationships were estimated (Figure \ref{mohns_rho_R}, A).
-However, R and R+S OMs with greater recruitment process variability and higher observation uncertainty had median Mohn's $\rho$ for recruitment greater than zero for most EMs even when the the EM process error type was correct.
-In R+S OMs with lower observation uncertainty, EMs with the correct process error type exhibited better median Mohn's $\rho$ close to 0 than EMs with the incorrect process error type.
+However, R and R+S OMs with greater recruitment process variability and higher observation error had median Mohn's $\rho$ for recruitment greater than zero for most EMs even when the the EM process error type was correct.
+In R+S OMs with lower observation error, EMs with the correct process error type exhibited better median Mohn's $\rho$ close to 0 than EMs with the incorrect process error type.
 For R+M, R+Sel, and R+q OMs, results for Mohn's $\rho$ for recruitment are similar to those for SSB, but the range in median values and variation in Mohn's $\rho$ values for a given OM are generally larger for recruitment (Figure \ref{mohns_rho_R}, B-D).
@@
 It is common during assessment model fitting to check that the maximum absolute gradient component is less than some threshold, but there is no standard.
-However, checking the Hessian for positive definiteness is sometimes done conditional on the convergence criterion.
-As such, some models may be excluded that in fact have an invertible hessian.
+However, if the Hessian were checked for positive definiteness conditional on a satisfactory gradient, some models may be excluded that in fact have an invertible hessian.
 We found the hessian at the optimized log-likelihood can often be invertible when the gradient approach with a sensible criterion (e.g., $10^{-3}$) would indicate lack of convergence.
 
-Another factor affecting the convercence criteria is numerical accuracy.
+Another factor affecting the convergence criteria is numerical accuracy.
 With the process errors modeled as random effects we are optimizing an (Laplace) approximation of the marginal log-likelihood and therefore hessians and gradients are also with regard to the approximation.
@@
 All of these approximations are affecting the gradients and invertibility of the hessian (although likely worse for some models than others), which, along with our results, suggests at least investigating whether the hessian is positive definite when the calculated gradients are not terribly large.
+Often high gradients and/or lack of positive definite hessian result when parameter estimates are at their bounds [@carvalhoetal21], and this also applies to variance or autocorrelation parameters for random effects with state-space assessment models.
 
-Often high gradients and/or lack of positive definite hessian result when selectivity parameter estimates are at their bounds (0,1) or random effects variance parameter estimates are essentially 0.
-Changing the parameter configuration to avoid the bounds or removing process errors with 0 variance will often lead to satisfactory convergence.
-
-Perhaps a greater gradient threshold would be more consistent with hessian-based convergence, but the gradient and hessian-based standard errors should be inspected.
-
-Parameters that are not well estimated can lead the assessment scientist toward a more appropriate and often more parsimonious model configuration.
-For example, variance parameters that are estimated on log scale as low negative number (variance is close to 0) can happen for random effects estimated with little or no variability.
-These may sometimes result in a hessian that is not positive definite or due to numerical issues may be invertible, but provide very large standard errors.
-Generally, we can expect lower probability of convergence when estimating natural mortality or stock-recruit relationships because of the difficulty distinguishing these parameters from others being estimated in assessment model with data typically available.
+Even when hessian-based variance can be accomplished, parameters that are poorly estimated will have extremely large variances.
+This further inspection can lead to a more appropriate and often more parsimonious model configuration where the problematic parameters are not estimated.
+For example, process error variance parameters that are estimated close to 0 indicates that the random effects are estimated to have little or no variability and removing these process errors is warranted.
+Generally, our results suggest we can expect lower probability of convergence of state-space assessment models when estimating natural mortality or stock-recruit relationships because of the difficulty distinguishing these parameters from others being estimated in assessment model with data typically available.
 
@@
 Fitting models to other OMs rarely preferred R+S EMs and R and R+S EMs were nearly always selected for the matching OMs.
-For other sources of process error, accuracy of AIC was improved to useful levels when there was larger variability in the process errors and/or better data quality.
+For other sources of process error, accuracy of AIC was improved to useful levels when there was larger variability in the process errors and/or lower observation error.
 
-AIC generally performs weakly in determining stock recruit relationships unless there is large contrast in SSB and low variation in recruitment process errors.
-Although we did not compare models with alternative stock-recruit relationships, we do not expect AIC to perform any better distinguishing these relationships.
-Our results comport well with those of @devalpinehastings02 where state-space stock-recruit models were fit just to SSB and recruitment "observations" and @brittenetalinreview which found AIC could not reliably distinguish alternative environmental effects on stock-recruit parameters.
-@milleretal16 found AIC to prefer a stock-recruit relationship with environmental effects for SNEMA yellowtail flounder, but there was a large change in stock size estimated over time and flatfish are well-observed by the NEFSC bottom trawl survey used for an index in the assessment model.
+We found AIC to performs weakly in determining stock recruit relationships unless there is large contrast in SSB and low variation in recruitment process errors.
+Although we did not compare models with alternative stock-recruit relationships (e.g., Ricker and Beverton-Holt), we do not expect AIC to perform any better distinguishing between relationships.
+Our finding that AIC tended to choose simpler recruitment models in most cases contrasts with the noted bias in AIC for more complex models [@shibata76;@katz81;@kassraftery95], but whereas those findings apply to more common fitting of models to raw data, state-space models account for observation error and estimate process errors in latent variables.
+
+Our results comport well with those of @devalpinehastings02 which found AIC could not distinguish among state-space stock-recruit models that were fit just to SSB and recruitment "observations" and @brittenetalinreview which found AIC could not reliably distinguish alternative environmental effects on stock-recruit parameters.
+@milleretal16 did find AIC to prefer a stock-recruit relationship with environmental effects for SNEMA yellowtail flounder, but there was a large change in stock size estimated over time and flatfish are well-observed by the NEFSC bottom trawl survey used for an index in the assessment model.
 
@@
 
-As we might expect bias in all parameters and assessment output was generally better with lower observation error.
-Reliable estimation of stock-recruit relationship parameters only appears possible in ideal situations with lower observation errors in age composition and indices, lower variability in recruitment process errors and large contrast in spawning biomass over time.
-The latter can occur with larger variability in survival or natural mortality process errors, or variation in fishing pressure.
-
-Estimation of median natural mortality appears more feasible than SR parameters.
-We found low or no evidence of bias for many OM process error assumptions when there was contrast in fishing pressure even when there was greater observation error although it can lead to more variable estimates of natural mortality.
-For OMs where there was bias in natural mortality due to high observation error (R+Sel, R+q), estimating the stock-recruit relationship seemed to remove the bias.
-
-However, estimation of natural mortality can cause large differences between the true and estimated SSB (that may be unbiased on average) when there is less contrast in fishing pressure over time and higher observation error.
-
-Estimating B-H reduced SSB bias in some situations
+As we might have expected, bias in all parameters and assessment output was generally better with lower observation error.
+Estimation of stock-recruit relationship parameters was not reliable in any of the OM-EM combinations, but estimation of median natural mortality was feasible in many OM scenarios with temporal contrast in fishing pressure.
+For OMs where there was bias in natural mortality due to high observation error (R+Sel and R+q OMS), estimating the stock-recruit relationship seemed to alleviate the bias.
+However, estimation of median natural mortality can cause large differences between the true and estimated SSB (that may or may not be unbiased on average) when there is less contrast in fishing pressure over time and higher observation error.
 
@@
 
-Higher observation error with variation in fishing pressure can produce retrospective patterns even when the the process error assumption is correct, but this retrospective pattern can be reduced if M or the correct stock recruit relationship is estimated.
+Incorrect process error assumptions for EMs did not produce strong retrospective patterns for SSB for any OMs whether median natural mortality and a stock-recruit relationship were estimated or not, but some weak retrospective patterns occur when observation error was high and there was contrast in fishing pressure.
 Retrospective patterns tended to be more variable for recruitment and can be large even when the EM is correct.
-Therefore, we recommend emphasis on inspection of retrospective patterns primarily for SSB and F, but further research on retrospective patterns in other assessment model parameters and derived output may be beneficial.
+Therefore, we recommend emphasis on inspection of retrospective patterns primarily for SSB and F, but further research on retrospective patterns in other assessment model parameters, management quantitities, and projections may be beneficial [@brookslegault16].
 
@@
 
-When R+S is preferred it probably means its the correct process error OF THE TYPES CONSIDERED HERE.
+For R+S OMs with low observation error, temporal contrast in fishing pressure, and large variation in survival process errors (3rd row), some EMs with the incorrect process error assumptions, estimate median M to be 0, and SSB is also underestimated, but Mohn's $\rho$ for SSB is better than when M is known.
+Fortunately, AIC would not choose the wrong EM process error type in these situations.
 
+When R+S is preferred AMONG THE TYPES CONSIDERED HERE, it probably means its the correct process error .
+
+<!--
 ## Summary {-}
@@
 When models did exhibit some retrospective pattern, estimation the median natural mortality rate tended to remove it.
+-->
 

### 2025-02-21 Project_0/manuscript/manuscript.tex
saving work on project_0 manuscript

@@
   pdftitle={Factors affecting reliablity of state-space age-structured assessment models},
-  pdfauthor={Timothy J. Miller1,2; Greg Britten3; Elizabeth N. Brooks2; Gavin Fay4; Alex Hansell2; Christopher M. Legault2; Brandon Muffley5; Brian C. Stock6; John Wiedenmann7},
+  pdfauthor={Timothy J. Miller1,2; Greg Britten3; Elizabeth N. Brooks2; Gavin Fay4; Alex Hansell2; Christopher M. Legault2; Chengxue Li2; Brandon Muffley5; Brian C. Stock6; John Wiedenmann7},
   hidelinks,
@@
 Hansell\textsuperscript{2} \and Christopher M.
-Legault\textsuperscript{2} \and Brandon
-Muffley\textsuperscript{5} \and Brian C.
-Stock\textsuperscript{6} \and John Wiedenmann\textsuperscript{7}}
-\date{12 February, 2025}
+Legault\textsuperscript{2} \and Chengxue
+Li\textsuperscript{2} \and Brandon Muffley\textsuperscript{5} \and Brian
+C. Stock\textsuperscript{6} \and John Wiedenmann\textsuperscript{7}}
+\date{13 February, 2025}
 
@@
 
-Random effects can be included in state-space assessment models on many
-processes and in many ways, and guidance is needed on statistical
-reliability and model selection criteria. We simulated 72 operating
-models with varying fishing history, observation error uncertainty, and
-process error magnitude, correlation, and source (recruitment, survival,
-fishery selectivity, catchability, and natural mortality). We fit
-estimating models with different assumptions on the process error
-source, whether median natural mortality was estimated, and whether a
-stock-recruit relationship was estimated.
-
-Models that assumed the correct process error source had high
-convergence and low bias. Bias was also low under most process error
-assumptions when there was contrast in fishing pressure.
-Stock-recruitment parameters were only reliably estimated in ideal
-situations. Marginal AIC most accurately distinguished process errors on
+State-space assessment models, which are increasingly being used for
+management of commercially important fish stocks, can include process
+errors in many ways, and an evaluation of the statistical reliability of
+the estimates of management output and best practices of selection among
+alternative process error configurations are needed. We simulated 72
+operating models with varying fishing pressure history, observation
+error error, and process error magnitude, correlation, and source
+(recruitment, survival, fishery selectivity, catchability, and natural
+mortality). We fit estimating models with different assumptions on the
+process error source, whether median natural mortality was estimated,
+and whether a stock-recruit relationship was estimated. Estimating
+models that assumed the correct process error source and median natural
+mortality rate was known, and did not estimate a stock recruit
+relationship generally had high probability of convergence and low bias
+in terminal year SSB estimation. Bias was also low under many incorrect
+process error assumptions when there was contrast in fishing pressure
+and low observation error. Stock-recruitment parameters were not
+reliably estimated in any scenarios, but the least accurate estimates
+occured under ideal situations of low observation error, low recruitment
+variability, and temporal contrast in fishing pressure. However,
+estimation of median natural mortality rate was reliable for many OMs
+(other than those with survival process errors) with temporal contrast
+in fishing pressure regardless of the estimated process error type.
+Marginal AIC most accurately distinguished process errors on
 recruitment, survival, and selectivity, as well as larger magnitude
-process errors of other types. Retrospective patterns were generally
-weak except for recruitment when observation error was high, even with
-the correct process error assumptions. When models did exhibit some
-retrospective pattern, estimating median natural mortality often removed
-it.
+process errors of other types. Retrospective patterns for spawning stock
+biomass and fishing mortality were generally weak, but those for
+recruitment were sizable when observation error was high, even with the
+correct process error assumptions.
 
@@
 
+Although the statistical aspects of state-space models and their
+application have been studied extensively, the work is primarily on
+Gaussian state-space models. State-space stock assessment models, where
+there are nonlinear functions of latent parameters and multiple types of
+observations with varying distributional assumptions, is one of the most
+complex examples of this analytical approach.
+
 Much is known about the reliability of state-space models that are
-linear or Gaussian (Aeberhard et al. 2018), but applications in
-fisheries management are nonlinear and typically include multiple types
-of observations with varying distributional assumptions. Furthermore,
-there is a wide range of potential random effects structures and model
-parameters that can be treated as random effects in assessment models.
-We know relatively little about the factors affecting statistical
-reliability of such models or the ability of information criteria to
+linear or Gaussian (Aeberhard et al. 2018), but not for the more complex
+application in stock assessment modeling.
+
+(Li et al. 2024 investigated some aspects of inferences for operating
+models with multiple sources of process error, but there are differences
+for this paper.)
+
+There is a wide range of model parameters that can be treated as random
+effects in assessment models and a variety of potential distributions
+for the random effects. We know relatively little about the factors
+affecting statistical reliability of such models or the ability to
 distinguish among such alternative structures.
 
-Li et al. (2024) investigated some aspects of inferences for operating
-models with multiple sources of process error, but there are differences
-for this paper.
-
 Here we conduct a simulation study with operating models (OMs) varying
-by degree of observation error uncertainty, source and variability of
-process error, and fishing history. The simulations from these OMs are
-fitted with estimation models (EMs) that make alternative assumptions
-for sources of process error, whether a stock-recruit model was
-estimated, and whether M is estimated. We evaluate whether AIC can
-correctly determine the correct source of process error and stock
-effects on recruitment. We also evaluate the degree of bias in the
+by degree of observation error error, source and variability of process
+error, and fishing history. The simulations from these OMs are fitted
+with estimation models (EMs) that make alternative assumptions for
+sources of process error, whether a stock-recruit model was estimated,
+and whether a constant, or, in some EMs, median, natural mortality is
+estimated. We evaluate whether AIC can correctly determine the correct
+source of process error and a stock-recruit relationship. We also
+evaluate when retrospective patterns occur and the degree of bias in the
 outputs of the assessment model that are important for management.
@@
 
-We used the Woods Hole Assessment Model (WHAM) to configure operating
-and EMs in our simulation study (Miller and Stock 2020; Stock and Miller
-2021). WHAM is an R package freely available as a github repository. For
-this study we used
+We used the Woods Hole Assessment Model (WHAM) to configure OMs and EMs
+in our simulation study (Miller and Stock 2020; Stock and Miller 2021).
+WHAM is an R package freely available as a github repository. For this
+study we used
 \href{https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3}{version
 1.0.6.9000, commit 77bbd94}. This package has also been used to
-configure operating and EMs for closed loop simulations evaluating
-index-based assessment methods (Legault et al. 2023) and is used for
-management of haddock, butterfish, American plaice, bluefish, Atlantic
-cod, black sea bass, and yellowtail flounder in the Northeast US.
+configure OMs and EMs for closed loop simulations evaluating index-based
+assessment methods (Legault et al. 2023) and is used for management of
+haddock, butterfish, American plaice, bluefish, Atlantic cod, black sea
+bass, and yellowtail flounder in the Northeast US.
 
@@
 fishing pressure and low observation error. Convergence rates were high
-for all EMs when fit to data from R+S OMs with lower observation
-uncertainty except those where median natural mortality and/or
-stock-recruit relationships were estimated.
+for all EMs when fit to data from R+S OMs with lower observation error
+except those where median natural mortality and/or stock-recruit
+relationships were estimated.
 
@@
 R+Sel OMs with lower selectivity process error variability or higher
-observation uncertainty (Figure \ref{hessian_SE_convergence}, C). R+Sel
-EMs generally converged better than other EMs for R+Sel OMs with higher
-process error variability, lower observation uncertainty, and contrast
-in fishing pressure regardless of whether median natural mortality or a
+observation error (Figure \ref{hessian_SE_convergence}, C). R+Sel EMs
+generally converged better than other EMs for R+Sel OMs with higher
+process error variability, lower observation error, and contrast in
+fishing pressure regardless of whether median natural mortality or a
 stock recruit relationship was estimated.
@@
 both are estimated, there is a noticeable reduction in accuracy when OMs
-have a constant fishing history, observation uncertainty is low and
-larger variability in recruitment process errors.
+have a constant fishing history, observation error is low and larger
+variability in recruitment process errors.
 
@@
 Marginal AIC most accurately determined the correct source of process
-error and correlation structure for R+Sel OMs with low observation
-uncertainty (Figures \ref{pe_aic}, C). When there was low variability in
-selectivity process errors and high observation uncertainty, R+q or R+S
-EMs were more likely to have the best AIC. Whether median natural
-morality or stock recruit relationships were estimated appeared to have
-little effect on the performance of AIC.
+error and correlation structure for R+Sel OMs with low observation error
+(Figures \ref{pe_aic}, C). When there was low variability in selectivity
+process errors and high observation error, R+q or R+S EMs were more
+likely to have the best AIC. Whether median natural morality or stock
+recruit relationships were estimated appeared to have little effect on
+the performance of AIC.
 
@@
 occurred for OMs with low variability in catchability process errors and
-high observation uncertainty. However, in these OMs, the marginal AIC
+high observation error. However, in these OMs, the marginal AIC
 accurately determined the correct source of process error (but not
@@
 rows). The largest effect of simultaneously estimating median natural
-mortality was when there was OMs had high observation uncertainty and
-constant fishing pressure (Figure \ref{sr_aic}, A, second row).
+mortality was when there was OMs had high observation error and constant
+fishing pressure (Figure \ref{sr_aic}, A, second row).
 
@@
 difference in AIC accuracy whether there was lower or higher observation
-uncertainty or whether the EM simultaneously estimated median natural
+error or whether the EM simultaneously estimated median natural
 mortality.
@@
 and third rows). There were some differences in effects of SSB variation
-on AIC accuracy when OMs had high observation uncertainty and constant
-fishing pressure. Across all of the OMs in Figure \ref{sr_aic}, the only
+on AIC accuracy when OMs had high observation error and constant fishing
+pressure. Across all of the OMs in Figure \ref{sr_aic}, the only
 estimated decline in AIC accuracy with SSB variation occurred when R+q
-OMs had high observation uncertainty and constant fishing pressure and
-greater variation in catchability process errors that also had AR1
+OMs had high observation error and constant fishing pressure and greater
+variation in catchability process errors that also had AR1
 autocorrelation.
@@
 no SR assumption was made, recruitment variability was low, and there
-was contrast in fishing mortality and high observation uncertainty
-(Figure \ref{SSB_rel_error}, A). However, errors in terminal SSB
-estimates were highly variable when median natural mortality was
-estimated and there was constant fishing pressure and high observation
-uncertainty (Figure \ref{SSB_rel_error}, A, second row).
+was contrast in fishing mortality and high observation error (Figure
+\ref{SSB_rel_error}, A). However, errors in terminal SSB estimates were
+highly variable when median natural mortality was estimated and there
+was constant fishing pressure and high observation error (Figure
+\ref{SSB_rel_error}, A, second row).
 
@@
 produced unbaised estimation of terminal SSB except when median natural
-mortality was estimated and there was high observation uncertainty. In
-R+S OMs with low observation uncertainty, EMs with incorrect process
-error assumptions typically provided biased estimation of terminal year
-SSB. Estimating the Beverton-Holt stock-recruit relationship had little
-discernible effect on bias of terminal year SSB estimation whereas
-estimating median M tended to produce more variability in errors in
-terminal SSB estimation similar to R OMs.
+mortality was estimated and there was high observation error. In R+S OMs
+with low observation error, EMs with incorrect process error assumptions
+typically provided biased estimation of terminal year SSB. Estimating
+the Beverton-Holt stock-recruit relationship had little discernible
+effect on bias of terminal year SSB estimation whereas estimating median
+M tended to produce more variability in errors in terminal SSB
+estimation similar to R OMs.
 
 For R+ M OMs with low variability in natural mortality process errors,
-low observation uncertainty and contrast in fishing motality over time
-all EMs produced low variability in SSB estimation error that indicated
+low observation error and contrast in fishing motality over time all EMs
+produced low variability in SSB estimation error that indicated
 unbiasedness (Figure \ref{SSB_rel_error}, B, third row). However, larger
@@
 mortality increased variability of SSB estimation error particularly for
-OMs with high observation uncertainty and constant fishing pressure over
-time and increased bias in SSB estimation for many R+M OMs. Like R and
-R+S OMs, estimating a stock-recruit relationship had little discernible
+OMs with high observation error and constant fishing pressure over time
+and increased bias in SSB estimation for many R+M OMs. Like R and R+S
+OMs, estimating a stock-recruit relationship had little discernible
 effect on SSB bias.
@@
 For R+Sel OMs, there was no evidence of bias for any EMs when
-variability in selectivity process errors was low, observation
-uncertainty low and contrast in fishing mortality (Figure
-\ref{SSB_rel_error}, C). The largest bias for all EMs that estimated
-median natural mortality when the OMs had high observation uncertainty,
-constant fishing pressure and greater variability in selectivity process
-errors (\(\sigma_{\text{Sel}} = 0.5\)) or low selectivity process errors
-(\(\sigma_{\text{Sel}} = 0.1\)) and low observation uncertainty.
-However, there was no evidence of bias of matching R+Sel EMs when
-observation uncertainty was low and variation in selectivity process
-errors was larger, whether median natural mortality was estimated or not
-(Figure \ref{SSB_rel_error}, C, third row). We only observed an effect
-of estimating the Beverton-Holt Stock recruit relationship for R+Sel OMs
-that had high observation uncertainty and contrast in fishing pressure
-where estimating the relationship produced less biased SSB estimation
-for many EMs (Figure \ref{SSB_rel_error}, C, top row).
-
-All EMs fit to data from R+q OMs with low observation uncertainty and
-contrast in fishing pressure exhibited little evidence of bias in
-terminal SSB estimation except for R+M EMs when there was no AR1
-correlation in catchability process errors (Figure \ref{SSB_rel_error},
-D). Many EMs also performed well in R+q OMs with low observation
-uncertainty, but no contrast in fishing pressure. For R+q OMs with high
-observation uncertainty and contrast in fishing pressure, EMs that
-estimated the Beverton-Holt stock recruit function exhibited less SSB
-bias than those that did not. Estimating median natural mortality in the
-EMs only resulted in much more variable SSB estimation errors when there
-was no contrast in fishing pressure (Figure \ref{SSB_rel_error}, D,
-first and third rows).
+variability in selectivity process error and observation error was low,
+and with contrast in fishing mortality (Figure \ref{SSB_rel_error}, C).
+The largest bias occurred for any EMs that estimated median natural
+mortality when the OMs had high observation error, constant fishing
+pressure, and greater variability in selectivity process errors
+(\(\sigma_{\text{Sel}} = 0.5\)) or low selectivity process errors
+(\(\sigma_{\text{Sel}} = 0.1\)) and low observation error. However,
+there was no evidence of bias of matching R+Sel EMs when observation
+error was low and variation in selectivity process errors was larger,
+whether median natural mortality was estimated or not (Figure
+\ref{SSB_rel_error}, C, third row). We only observed an effect of
+estimating the Beverton-Holt Stock recruit relationship for R+Sel OMs
+that had high observation error and contrast in fishing pressure where
+estimating the relationship produced less biased SSB estimation for many
+EMs (Figure \ref{SSB_rel_error}, C, top row).
+
+All EMs fit to data from R+q OMs with low observation error and contrast
+in fishing pressure exhibited little evidence of bias in terminal SSB
+estimation except for R+M EMs when there was no AR1 correlation in
+catchability process errors (Figure \ref{SSB_rel_error}, D). Many EMs
+also performed well in R+q OMs with low observation error, but no
+contrast in fishing pressure. For R+q OMs with high observation error
+and contrast in fishing pressure, EMs that estimated the Beverton-Holt
+stock recruit function exhibited less SSB bias than those that did not.
+Estimating median natural mortality in the EMs only resulted in much
+more variable SSB estimation errors when there was no contrast in
+fishing pressure (Figure \ref{SSB_rel_error}, D, first and third rows).
 
@@
 process error types that matched the OMs and the OMs had low observation
-uncertainty, contrast in fishing mortality and more variable and
-uncorrelated selectivity process errors. There was little effect of
-estimating natural mortality on relative errors for stock recruit
-parameters.
+error, contrast in fishing mortality and more variable and uncorrelated
+selectivity process errors. There was little effect of estimating
+natural mortality on relative errors for stock recruit parameters.
 
@@
 variable for most EM process error types when OMs had contrast in
-fishing pressure and those OMs that also had lower observation
-uncertainty generally had the least variability (Figure
-\ref{SR_rel_error}, D). The worst accuracy of stock-recruit parameter
-estimation regardless of EM type occured when R+q OMs had low
-observation uncertainty and constant fishing pressure (Figure
-\ref{SR_rel_error}, D, fourth row).
+fishing pressure and those OMs that also had lower observation error
+generally had the least variability (Figure \ref{SR_rel_error}, D). The
+worst accuracy of stock-recruit parameter estimation regardless of EM
+type occured when R+q OMs had low observation error and constant fishing
+pressure (Figure \ref{SR_rel_error}, D, fourth row).
 
@@
 Median natural mortality rate is estimated accurately by all EM process
-error types for all R OMs except those where observation uncertainty is
-high and fishing pressure is constant where variability in relative
-errors is high (Figure \ref{M_rel_error}, A, \(\sigma_{2+} = 0\)). For
-R+S OMs estimation of median natural mortality rate is most accurate
-when observation uncertainty is low and there is contrast in fishing
-pressure and the EM process error type is correct. There is little
-effect of estimating stock-recruit relationships on the bias for median
-natural mortality.
+error types for all R OMs except those where observation error is high
+and fishing pressure is constant where variability in relative errors is
+high (Figure \ref{M_rel_error}, A, \(\sigma_{2+} = 0\)). For R+S OMs
+estimation of median natural mortality rate is most accurate when
+observation error is low and there is contrast in fishing pressure and
+the EM process error type is correct. There is little effect of
+estimating stock-recruit relationships on the bias for median natural
+mortality.
 
@@
 regardless of EM process error type, when OMs had a change in fishing
-pressure and low observation uncertainty (Figure \ref{M_rel_error}, B).
+pressure and low observation error (Figure \ref{M_rel_error}, B).
 However, those R+M OMs that also had greatest variability in AR1
@@
 natural mortality rate for R+Sel OMs that had contrast in fishing
-pressure, low observation uncertainty, and low selectivity process error
+pressure, low observation error, and low selectivity process error
 variability (Figure \ref{M_rel_error}, C). When selectivity process
@@
 estimation occurred for all EM process error types when observation
-uncertainty was high and there was constant fishing pressure. Again,
-like other OM process error types, there was little effect of estimating
+error was high and there was constant fishing pressure. Again, like
+other OM process error types, there was little effect of estimating
 stock-recruit relationships on the bias for median natural mortality.
@@
 of median natural mortality rate when fit to R+q OMs with contrast in
-fishing pressure, low observation uncertainty and low catchability
-process error variability (Figure \ref{M_rel_error}, D). Most EM process
-error types produced biased estimation of median natural mortality when
-R+q OMs had high observaiton uncertainty and constant fishing pressure.
-Again, there was no discernible effect of estimating stock-recruit
+fishing pressure, low observation error and low catchability process
+error variability (Figure \ref{M_rel_error}, D). Most EM process error
+types produced biased estimation of median natural mortality when R+q
+OMs had high observaiton error and constant fishing pressure. Again,
+there was no discernible effect of estimating stock-recruit
 relationships on bias.
@@
 in OMs with the largest survival process error variability, high
-observation uncertainty, and contrast in fishing pressure, but only for
-EMs without the correct process error type and where median natural
+observation error, and contrast in fishing pressure, but only for EMs
+without the correct process error type and where median natural
 mortality rate was assumed known. For R+M, R+Sel, and R+q OMs, Mohn's
@@
 A). However, R and R+S OMs with greater recruitment process variability
-and higher observation uncertainty had median Mohn's \(\rho\) for
-recruitment greater than zero for most EMs even when the the EM process
-error type was correct. In R+S OMs with lower observation uncertainty,
-EMs with the correct process error type exhibited better median Mohn's
-\(\rho\) close to 0 than EMs with the incorrect process error type. For
-R+M, R+Sel, and R+q OMs, results for Mohn's \(\rho\) for recruitment are
-similar to those for SSB, but the range in median values and variation
-in Mohn's \(\rho\) values for a given OM are generally larger for
-recruitment (Figure \ref{mohns_rho_R}, B-D).
+and higher observation error had median Mohn's \(\rho\) for recruitment
+greater than zero for most EMs even when the the EM process error type
+was correct. In R+S OMs with lower observation error, EMs with the
+correct process error type exhibited better median Mohn's \(\rho\) close
+to 0 than EMs with the incorrect process error type. For R+M, R+Sel, and
+R+q OMs, results for Mohn's \(\rho\) for recruitment are similar to
+those for SSB, but the range in median values and variation in Mohn's
+\(\rho\) values for a given OM are generally larger for recruitment
+(Figure \ref{mohns_rho_R}, B-D).
 
@@
 the maximum absolute gradient component is less than some threshold, but
-there is no standard. However, checking the Hessian for positive
-definiteness is sometimes done conditional on the convergence criterion.
-As such, some models may be excluded that in fact have an invertible
-hessian. We found the hessian at the optimized log-likelihood can often
-be invertible when the gradient approach with a sensible criterion
-(e.g., \(10^{-3}\)) would indicate lack of convergence.
-
-Another factor affecting the convercence criteria is numerical accuracy.
+there is no standard. However, if the Hessian were checked for positive
+definiteness conditional on a satisfactory gradient, some models may be
+excluded that in fact have an invertible hessian. We found the hessian
+at the optimized log-likelihood can often be invertible when the
+gradient approach with a sensible criterion (e.g., \(10^{-3}\)) would
+indicate lack of convergence.
+
+Another factor affecting the convergence criteria is numerical accuracy.
 With the process errors modeled as random effects we are optimizing an
@@
 hessian is positive definite when the calculated gradients are not
-terribly large.
-
-Often high gradients and/or lack of positive definite hessian result
-when selectivity parameter estimates are at their bounds (0,1) or random
-effects variance parameter estimates are essentially 0. Changing the
-parameter configuration to avoid the bounds or removing process errors
-with 0 variance will often lead to satisfactory convergence.
-
-Perhaps a greater gradient threshold would be more consistent with
-hessian-based convergence, but the gradient and hessian-based standard
-errors should be inspected.
-
-Parameters that are not well estimated can lead the assessment scientist
-toward a more appropriate and often more parsimonious model
-configuration. For example, variance parameters that are estimated on
-log scale as low negative number (variance is close to 0) can happen for
-random effects estimated with little or no variability. These may
-sometimes result in a hessian that is not positive definite or due to
-numerical issues may be invertible, but provide very large standard
-errors. Generally, we can expect lower probability of convergence when
-estimating natural mortality or stock-recruit relationships because of
-the difficulty distinguishing these parameters from others being
-estimated in assessment model with data typically available.
+terribly large. Often high gradients and/or lack of positive definite
+hessian result when parameter estimates are at their bounds (Carvalho et
+al. 2021), and this also applies to variance or autocorrelation
+parameters for random effects with state-space assessment models.
+
+Even when hessian-based variance can be accomplished, parameters that
+are poorly estimated will have extremely large variances. This further
+inspection can lead to a more appropriate and often more parsimonious
+model configuration where the problematic parameters are not estimated.
+For example, process error variance parameters that are estimated close
+to 0 indicates that the random effects are estimated to have little or
+no variability and removing these process errors is warranted.
+Generally, our results suggest we can expect lower probability of
+convergence of state-space assessment models when estimating natural
+mortality or stock-recruit relationships because of the difficulty
+distinguishing these parameters from others being estimated in
+assessment model with data typically available.
 
@@
 process error, accuracy of AIC was improved to useful levels when there
-was larger variability in the process errors and/or better data quality.
-
-AIC generally performs weakly in determining stock recruit relationships
-unless there is large contrast in SSB and low variation in recruitment
-process errors. Although we did not compare models with alternative
-stock-recruit relationships, we do not expect AIC to perform any better
-distinguishing these relationships. Our results comport well with those
-of de Valpine and Hastings (2002) where state-space stock-recruit models
-were fit just to SSB and recruitment ``observations'' and Britten et al.
-which found AIC could not reliably distinguish alternative environmental
-effects on stock-recruit parameters. Miller et al. (2016) found AIC to
-prefer a stock-recruit relationship with environmental effects for SNEMA
-yellowtail flounder, but there was a large change in stock size
-estimated over time and flatfish are well-observed by the NEFSC bottom
-trawl survey used for an index in the assessment model.
+was larger variability in the process errors and/or lower observation
+error.
+
+We found AIC to performs weakly in determining stock recruit
+relationships unless there is large contrast in SSB and low variation in
+recruitment process errors. Although we did not compare models with
+alternative stock-recruit relationships (e.g., Ricker and
+Beverton-Holt), we do not expect AIC to perform any better
+distinguishing between relationships. Our finding that AIC tended to
+choose simpler recruitment models in most cases contrasts with the noted
+bias in AIC for more complex models (Shibata 1976; Katz 1981; Kass and
+Raftery 1995), but whereas those findings apply to more common fitting
+of models to raw data, state-space models account for observation error
+and estimate process errors in latent variables.
+
+Our results comport well with those of de Valpine and Hastings (2002)
+which found AIC could not distinguish among state-space stock-recruit
+models that were fit just to SSB and recruitment ``observations'' and
+Britten et al. which found AIC could not reliably distinguish
+alternative environmental effects on stock-recruit parameters. Miller et
+al. (2016) did find AIC to prefer a stock-recruit relationship with
+environmental effects for SNEMA yellowtail flounder, but there was a
+large change in stock size estimated over time and flatfish are
+well-observed by the NEFSC bottom trawl survey used for an index in the
+assessment model.
 
@@
 
-As we might expect bias in all parameters and assessment output was
-generally better with lower observation error. Reliable estimation of
-stock-recruit relationship parameters only appears possible in ideal
-situations with lower observation errors in age composition and indices,
-lower variability in recruitment process errors and large contrast in
-spawning biomass over time. The latter can occur with larger variability
-in survival or natural mortality process errors, or variation in fishing
-pressure.
-
-Estimation of median natural mortality appears more feasible than SR
-parameters. We found low or no evidence of bias for many OM process
-error assumptions when there was contrast in fishing pressure even when
-there was greater observation error although it can lead to more
-variable estimates of natural mortality. For OMs where there was bias in
-natural mortality due to high observation error (R+Sel, R+q), estimating
-the stock-recruit relationship seemed to remove the bias.
-
-However, estimation of natural mortality can cause large differences
-between the true and estimated SSB (that may be unbiased on average)
-when there is less contrast in fishing pressure over time and higher
-observation error.
-
-Estimating B-H reduced SSB bias in some situations
+As we might have expected, bias in all parameters and assessment output
+was generally better with lower observation error. Estimation of
+stock-recruit relationship parameters was not reliable in any of the
+OM-EM combinations, but estimation of median natural mortality was
+feasible in many OM scenarios with temporal contrast in fishing
+pressure. For OMs where there was bias in natural mortality due to high
+observation error (R+Sel and R+q OMS), estimating the stock-recruit
+relationship seemed to alleviate the bias. However, estimation of median
+natural mortality can cause large differences between the true and
+estimated SSB (that may or may not be unbiased on average) when there is
+less contrast in fishing pressure over time and higher observation
+error.
 
@@
 
-Higher observation error with variation in fishing pressure can produce
-retrospective patterns even when the the process error assumption is
-correct, but this retrospective pattern can be reduced if M or the
-correct stock recruit relationship is estimated. Retrospective patterns
+Incorrect process error assumptions for EMs did not produce strong
+retrospective patterns for SSB for any OMs whether median natural
+mortality and a stock-recruit relationship were estimated or not, but
+some weak retrospective patterns occur when observation error was high
+and there was contrast in fishing pressure. Retrospective patterns
 tended to be more variable for recruitment and can be large even when
@@
 retrospective patterns primarily for SSB and F, but further research on
-retrospective patterns in other assessment model parameters and derived
-output may be beneficial.
+retrospective patterns in other assessment model parameters, management
+quantitities, and projections may be beneficial (Brooks and Legault
+2016).
 
@@
 
-When R+S is preferred it probably means its the correct process error OF
-THE TYPES CONSIDERED HERE.
-
-\hypertarget{summary}{%
-\subsection*{Summary}\label{summary}}
-\addcontentsline{toc}{subsection}{Summary}
-
-We conducted a simulation study where we simulated process errors and
-observations for 72 operating models with alternative assumptions about
-fishing history, degree of uncertainty in index and age composition
-observations, type (recruitment, survival, fishery selectivity,
-catchability and natural mortality), degree of variation, and
-correlation of process errors. We fit 20 different estimation models to
-each of 100 simulated set of observations with alternative assumptions
-type and correlation structure of process errors, median natural
-mortality was known or estimated, and a B-H stock recruit relationship
-was assumed or not.
-
-Across simulations we summarized probability of convergence of fitted
-models, accuracy of marginal AIC in determining the correct process
-error assumption and it ability to determine the Beverton-Holt stock
-recruit relationship, bias in annual spawning stock biomass, in
-estimation of natural mortality, and in stock-recruit relationship
-parameters, and severity of retrospective patterns for estimation
-models.
-
-Alternative measures of convergence performed differently. Invertible
-hessians and resulting standard error estimation was possible when
-criteria based on the gradient of the optimized log-likelihood with
-respect to the fixed effects parameters. Using hessian-based
-convergence, probability of convergence was best for models that assumed
-the correct source of process error, assumed M was known, and did not
-assume stock-recruit relationships.
-
-Using marginal AIC provided most accurate inferences about the process
-errors on recruitment and survival and selectivity, in that AIC
-preferred EMs with assumed process errors that matched OMs, and EMs with
-those assumed process errors were not preferred when alternative process
-errors were assumed in the OMs. However, when the true process errors
-were more variable, AIC accuracy increased to a useful level. We found
-AIC more accurately determined a B-H stock recruit relationship rather
-than the null model without a S-R relationship when there was low
-variability in recruitment, low variability in survival random effects,
-and higher variation in spawning biomass over the time series. When the
-median natural mortality rate was estimated, we found large bias and
-uncertainty much more likely in model output such as spawning stock
-biomass.
-
-Bias in spawning stock biomass estimation was generally low for
-estimation models that assumed the correct source of process error when
-there was lower observation error. Reliable estimation of stock-recruit
-relationship parameters only appears possible in ideal situations with
-lower observation errors in age composition and indices, lower
-variability in recruitment process errors and large contrast in spawning
-biomass over time. We found little evidence of bias for many OM process
-error assumptions when there was contrast in fishing pressure even when
-there was greater observation error although it can lead to more
-variable estimates of natural mortality. For OMs where there was bias in
-natural mortality due to high observation error, estimation the
-stock-recruit relationship seemed to remove the bias. However,
-estimation of natural mortality can cause large differences between the
-true and estimated SSB (that may be unbiased on average) when there is
-less contrast in fishing pressure over time and higher observation
-error.
-
-Retrospective patterns were generally weak for all estimation models
-regardless of the true source of process error, but they can be expected
-for recruitment even for the correct process error assumptions when
-observation error is high. When models did exhibit some retrospective
-pattern, estimation the median natural mortality rate tended to remove
-it.
+For R+S OMs with low observation error, temporal contrast in fishing
+pressure, and large variation in survival process errors (3rd row), some
+EMs with the incorrect process error assumptions, estimate median M to
+be 0, and SSB is also underestimated, but Mohn's \(\rho\) for SSB is
+better than when M is known. Fortunately, AIC would not choose the wrong
+EM process error type in these situations.
+
+When R+S is preferred AMONG THE TYPES CONSIDERED HERE, it probably means
+its the correct process error .
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-brookslegault16}{}}%
+Brooks, E.N., and Legault, C.M. 2016. Retrospective forecasting --
+evaluating performance of stock projections for new england groundfish
+stocks. Canadian Journal of Fisheries and Aquatic Sciences
+\textbf{73}(6): 935--950.
+doi:\href{https://doi.org/10.1139/cjfas-2015-0163}{10.1139/cjfas-2015-0163}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-cadigan16}{}}%
@@
 assessment software. Fisheries Research \textbf{246}: 106176.
+
+\leavevmode\vadjust pre{\hypertarget{ref-kassraftery95}{}}%
+Kass, R.E., and Raftery, A.E. 1995. Bayes factors. Journal of the
+American Statistical Association \textbf{90}(430): 773--795.
+doi:\href{https://doi.org/10.1080/01621459.1995.10476572}{10.1080/01621459.1995.10476572}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-katz81}{}}%
+Katz, R.W. 1981. On some criteria for estimating the order of a {M}arkov
+chain. Technometrics \textbf{23}(3): 243--249.
+doi:\href{https://doi.org/10.1080/00401706.1981.10486293}{10.1080/00401706.1981.10486293}.
 
@@
 \textbf{73}(8): 1261--1270.
+doi:\href{https://doi.org/10.1139/cjfas-2015-0339}{10.1139/cjfas-2015-0339}.
 
@@
 natural mortality and process error assumptions using a state-space,
-age-structured assessment model.
+age-structured assessment model. Canadian Journal of Fisheries and
+Aquatic Sciences \textbf{75}(5): 691--703.
+doi:\href{https://doi.org/10.1139/cjfas-2017-0035}{10.1139/cjfas-2017-0035}.
 
@@
 106642.
+
+\leavevmode\vadjust pre{\hypertarget{ref-shibata76}{}}%
+Shibata, R. 1976. Selection of the order of an autoregressive model by
+akaike's information criterion. Biometrika \textbf{63}(1): 117--126.
+doi:\href{https://doi.org/10.1093/biomet/63.1.117}{10.1093/biomet/63.1.117}.
 

### 2025-03-13 Project_0/manuscript/manuscript.Rmd
saving work on project 0 paper

@@
 
-Application of state-space models in fisheries stock assessment and management has expanded dramatically within ICES, Canada, and the Northeast US [@nielsenberg14;@cadigan16;@pedersenberg17;@stockmiller21].
-State-space approaches that use random effects to parameterize process errors is considered best practice and a requirement for the next generation of stock assessment models [@hoyleetal22; @punt23].
+Application of state-space models in fisheries stock assessment and management has expanded dramatically within International Council for the Exploration of the Sea (ICES), Canada, and the Northeast US [@nielsenberg14; @cadigan16; @pedersenberg17; @stockmiller21].
+State-space models latent population characteristics as statistical time series with periodic observations that also may have error due to sampling or other sources of measurement error.
+Traditional assessment models may use state-space approaches to account for temporal variability in population characteristics [@methotwetzel13;@legaultrestrepo99], but these models treat the annual parameters as penalized fixed effects parameters where the variance parameters controlling the penalties are assumed known [@thorsonminto15].
+Modern state-space models can estimate the annually varying parameters as random effects with variance parameters estimated using maximum marginal likelihood or corresponding Bayesian approaches.
+These latter approaches are considered best practice  and a requirement for the next generation of stock assessment models [@hoyleetal22; @punt23].
 
-Although the statistical aspects of state-space models and their application have been studied extensively, the work is primarily on Gaussian state-space models.
-State-space stock assessment models, where there are nonlinear functions of latent parameters and multiple types of observations with varying distributional assumptions, is one of the most complex examples of this analytical approach.
+State-space stock assessment models, with nonlinear functions of latent parameters and multiple types of observations with varying distributional assumptions, are one of the most complex examples of this analytical approach.
+Statistical aspects of state-space models and their application within fisheries have been studied extensively, but previous work has focused primarily on linear and Gaussian state-space models [@aeberhardetal18; @augeretal21].
+Therefore, current understanding of the reliability of state-space models does not extend to usage for stock assessment.
 
-Much is known about the reliability of state-space models that are linear or Gaussian [@aeberhardetal18], but not for the more complex application in stock assessment modeling.
+As state-space models provide greater flexibility by allowing multiple processes to vary as random effects [@nielsenberg14; @aeberhardetal18; @stocketal21], one of the most immediate questions regards the implications of mis-specification among alternative sources of process error.
+Incorrect treatment of population attributes as temporally varying [@trijouletetal20; @liljestrandetal24] could lead to misidentification of stock status and biased population estimates, ultimately impacting fisheries management decisions [@legaultpalmer16; @szuwalskietal18; @croninpunt21].
+Furthermore, biological, fishery, and observational processes are often confounded in catch-at-age data, which may adversely affect ability to distinguish between true process variability and observational error [@puntetal14; @stewartmonnahan17; @croninpunt21; @fischetal23; @lietalinreview_a].
 
-[@lietal24 investigated some aspects of inferences for operating models with multiple sources of process error, but there are differences for this paper.]
+@lietal24 conducted a full-factorial simulation-estimation study to assess model reliability when confounding random-effects processes (numbers-at-age, fishery selectivity, and natural mortality) were included.
+Their results suggest that while state-space models can generally identify sources of process error, overly complex models, even when misspecified (i.e., incorporating process error that did not exist in reality), often performed similarly to correctly specified models, with little to no bias in key management quantities.
+Similarly, @liljestrandetal24 found little downside in assuming process error in recruitment or selectivity, even when it was absent.
 
-There is a wide range of model parameters that can be treated as random effects in assessment models and a variety of potential distributions for the random effects.
-We know relatively little about the factors affecting statistical reliability of such models or the ability to distinguish among such alternative structures.
+Despite mounting efforts, several limitations remain.
+First, confounding processes that can be treated as random effects in the model were not thoroughly examined or tested within a simulation-estimation framework.
+Second, previous studies relied on operating models conditioned on specific fisheries, limiting their generalizability [@lietalinreview_a; @liljestrandetal24].
+In particular, the effects of observation error and underlying fishing history were not fully isolated from their study design, making it challenging to disentangle the interplay between process and observation error magnitudes, as demonstrated in @fischetal23.
+Third, explicitly modeling stock-recruitment relationships as mechanistic drivers of population dynamics is promising [@fleischmanetal13; @dupontaviceetal22] but can be challenging in state-space models.
+Evidence suggests that when both process and observation errors are unknown, estimating density dependence parameters becomes highly uncertain [@knape08; @polanskyetal09].
+In particular, @knape08 demonstrated that stronger density dependence becomes increasingly difficult to estimate in the presence of observation error.
+Therefore, it is crucial to assess whether density dependence mechanisms can be estimated with sufficient precision before drawing ecological conclusions [@augeretal16].
 
-<!-- But those studies focus primarily on Gaussian linear models.
-review literature on reliability of hidden/latent process models. Primarily in other fields.
--->
-
-Here we conduct a simulation study with operating models (OMs) varying by degree of observation error error, source and variability of process error, and fishing history.
+In the present study, we conduct a simulation study with operating models (OMs) varying by degree of observation error, source and variability of process error, and fishing history.
 The simulations from these OMs are fitted with estimation models (EMs) that make alternative assumptions for sources of process error, whether a stock-recruit model was estimated, and whether a constant, or, in some EMs, median, natural mortality is estimated.
-We evaluate whether AIC can correctly determine the correct source of process error and a stock-recruit relationship.
+Given the confounding nature of process errors, developing diagnostic tools to detect model misspecification is of great scientific interest and could aid the next generation of stock assessments [@augeretal21].
+We evaluate whether convergence and Akaike Information Criterion (AIC) can correctly determine the source of process error and the existence of a stock-recruit relationship.
 We also evaluate when retrospective patterns occur and the degree of bias in the outputs of the assessment model that are important for management.
@@
 Among the process error configurations we used in OMs, we found AIC to be accurate for process errors on recruitment and survival (R and R+S).
-Fitting models to other OMs rarely preferred R+S EMs and R and R+S EMs were nearly always selected for the matching OMs.
+Fitting models to other OMs rarely preferred R+S EMs, and R and R+S EMs were nearly always selected for the matching OMs, a result similar to that reported in [@liljestrandetal24].
 For other sources of process error, accuracy of AIC was improved to useful levels when there was larger variability in the process errors and/or lower observation error.
+
+AIC performed poorly in identifying stock-recruit relationships unless there was contrast in SSB and low variability in recruitment process errors (Fig.
+3), both of which help reduce confounding effects from both process and observation errors.
+OMs with increased  contrast in SSB included higher variability in survival or natural mortality process errors (R+S, R+M OMs), and contrast in fishing pressure (Fig.
+\ref{sr_aic}).
+Process errors in index catchability (R+q OMs) do not affect the contrast in SSB and variability in fishery logistic selectivity parameters (R+Sel OMs) had lesser effect on constrast in SSB.
+
+Increased observation error only weakens AIC performance for OMs where process error configurations allows SR accuracy in low observation error
 
@@
 Our finding that AIC tended to choose simpler recruitment models in most cases contrasts with the noted bias in AIC for more complex models [@shibata76;@katz81;@kassraftery95], but whereas those findings apply to more common fitting of models to raw data, state-space models account for observation error and estimate process errors in latent variables.
+
+Our findings also suggest that process errors in selectivity and catchability have minimal impact on the AIC accuracy for the stock-recruit relationship (Fig.
+3C&D), whereas natural mortality process errors have a much stronger effect (Fig.
+3B).
+This may be because selectivity and catchability do not disproportionately favorare lower for the younger age classes, and vulnerability of older ages follows an increasing logistic curve.
+In contrast, natural mortality influences all age classes equally, leading to greater overall fluctuations in SSB, which increases the probability of detecting the stock-recruit signal even under constant fishing pressure (Fig.
+3B).
+These results highlight the importance of accounting for process error in natural mortality (only when there is evidence of temporal variability, [@li2024]) when evaluating stock-recruitment dynamics in state-space models.
 

### 2025-03-13 Project_0/manuscript/manuscript.tex
saving work on project 0 paper

@@
 C. Stock\textsuperscript{6} \and John Wiedenmann\textsuperscript{7}}
-\date{13 February, 2025}
+\date{12 March, 2025}
 
@@
 Application of state-space models in fisheries stock assessment and
-management has expanded dramatically within ICES, Canada, and the
-Northeast US (Nielsen and Berg 2014; Cadigan 2016; Pedersen and Berg
-2017; Stock and Miller 2021). State-space approaches that use random
-effects to parameterize process errors is considered best practice and a
-requirement for the next generation of stock assessment models (Hoyle et
-al. 2022; Punt 2023).
-
-Although the statistical aspects of state-space models and their
-application have been studied extensively, the work is primarily on
-Gaussian state-space models. State-space stock assessment models, where
-there are nonlinear functions of latent parameters and multiple types of
-observations with varying distributional assumptions, is one of the most
-complex examples of this analytical approach.
-
-Much is known about the reliability of state-space models that are
-linear or Gaussian (Aeberhard et al. 2018), but not for the more complex
-application in stock assessment modeling.
-
-(Li et al. 2024 investigated some aspects of inferences for operating
-models with multiple sources of process error, but there are differences
-for this paper.)
-
-There is a wide range of model parameters that can be treated as random
-effects in assessment models and a variety of potential distributions
-for the random effects. We know relatively little about the factors
-affecting statistical reliability of such models or the ability to
-distinguish among such alternative structures.
-
-Here we conduct a simulation study with operating models (OMs) varying
-by degree of observation error error, source and variability of process
-error, and fishing history. The simulations from these OMs are fitted
-with estimation models (EMs) that make alternative assumptions for
-sources of process error, whether a stock-recruit model was estimated,
-and whether a constant, or, in some EMs, median, natural mortality is
-estimated. We evaluate whether AIC can correctly determine the correct
-source of process error and a stock-recruit relationship. We also
-evaluate when retrospective patterns occur and the degree of bias in the
-outputs of the assessment model that are important for management.
+management has expanded dramatically within International Council for
+the Exploration of the Sea (ICES), Canada, and the Northeast US (Nielsen
+and Berg 2014; Cadigan 2016; Pedersen and Berg 2017; Stock and Miller
+2021). State-space models latent population characteristics as
+statistical time series with periodic observations that also may have
+error due to sampling or other sources of measurement error. Traditional
+assessment models may use state-space approaches to account for temporal
+variability in population characteristics (Legault and Restrepo 1999;
+Methot and Wetzel 2013), but these models treat the annual parameters as
+penalized fixed effects parameters where the variance parameters
+controlling the penalties are assumed known (Thorson and Minto 2015).
+Modern state-space models can estimate the annually varying parameters
+as random effects with variance parameters estimated using maximum
+marginal likelihood or corresponding Bayesian approaches. These latter
+approaches are considered best practice and a requirement for the next
+generation of stock assessment models (Hoyle et al. 2022; Punt 2023).
+
+State-space stock assessment models, with nonlinear functions of latent
+parameters and multiple types of observations with varying
+distributional assumptions, are one of the most complex examples of this
+analytical approach. Statistical aspects of state-space models and their
+application within fisheries have been studied extensively, but previous
+work has focused primarily on linear and Gaussian state-space models
+(Aeberhard et al. 2018; Auger-Méthé et al. 2021). Therefore, current
+understanding of the reliability of state-space models does not extend
+to usage for stock assessment.
+
+As state-space models provide greater flexibility by allowing multiple
+processes to vary as random effects (Nielsen and Berg 2014; Aeberhard et
+al. 2018; Stock et al. 2021), one of the most immediate questions
+regards the implications of mis-specification among alternative sources
+of process error. Incorrect treatment of population attributes as
+temporally varying (Trijoulet et al. 2020; Liljestrand et al. 2024)
+could lead to misidentification of stock status and biased population
+estimates, ultimately impacting fisheries management decisions (Legault
+and Palmer 2016; Szuwalski et al. 2018; Cronin-Fine and Punt 2021).
+Furthermore, biological, fishery, and observational processes are often
+confounded in catch-at-age data, which may adversely affect ability to
+distinguish between true process variability and observational error (Li
+et al. In review; Punt et al. 2014; Stewart and Monnahan 2017;
+Cronin-Fine and Punt 2021; Fisch et al. 2023).
+
+Li et al. (2024) conducted a full-factorial simulation-estimation study
+to assess model reliability when confounding random-effects processes
+(numbers-at-age, fishery selectivity, and natural mortality) were
+included. Their results suggest that while state-space models can
+generally identify sources of process error, overly complex models, even
+when misspecified (i.e., incorporating process error that did not exist
+in reality), often performed similarly to correctly specified models,
+with little to no bias in key management quantities. Similarly,
+Liljestrand et al. (2024) found little downside in assuming process
+error in recruitment or selectivity, even when it was absent.
+
+Despite mounting efforts, several limitations remain. First, confounding
+processes that can be treated as random effects in the model were not
+thoroughly examined or tested within a simulation-estimation framework.
+Second, previous studies relied on operating models conditioned on
+specific fisheries, limiting their generalizability (Li et al. In
+review; Liljestrand et al. 2024). In particular, the effects of
+observation error and underlying fishing history were not fully isolated
+from their study design, making it challenging to disentangle the
+interplay between process and observation error magnitudes, as
+demonstrated in Fisch et al. (2023). Third, explicitly modeling
+stock-recruitment relationships as mechanistic drivers of population
+dynamics is promising (Fleischman et al. 2013; Du Pontavice et al. 2022)
+but can be challenging in state-space models. Evidence suggests that
+when both process and observation errors are unknown, estimating density
+dependence parameters becomes highly uncertain (Knape 2008; Polansky et
+al. 2009). In particular, Knape (2008) demonstrated that stronger
+density dependence becomes increasingly difficult to estimate in the
+presence of observation error. Therefore, it is crucial to assess
+whether density dependence mechanisms can be estimated with sufficient
+precision before drawing ecological conclusions (Auger-Méthé et al.
+2016).
+
+In the present study, we conduct a simulation study with operating
+models (OMs) varying by degree of observation error, source and
+variability of process error, and fishing history. The simulations from
+these OMs are fitted with estimation models (EMs) that make alternative
+assumptions for sources of process error, whether a stock-recruit model
+was estimated, and whether a constant, or, in some EMs, median, natural
+mortality is estimated. Given the confounding nature of process errors,
+developing diagnostic tools to detect model misspecification is of great
+scientific interest and could aid the next generation of stock
+assessments (Auger-Méthé et al. 2021). We evaluate whether convergence
+and Akaike Information Criterion (AIC) can correctly determine the
+source of process error and the existence of a stock-recruit
+relationship. We also evaluate when retrospective patterns occur and the
+degree of bias in the outputs of the assessment model that are important
+for management.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-augeretal16}{}}%
+Auger-Méthé, M., Field, C., Albertsen, C.M., Derocher, A.E., Lewis,
+M.A., Jonsen, I.D., and Mills Flemming, J. 2016. State-space models'
+dirty little secrets: Even simple linear gaussian models can have
+estimation problems. Scientific reports \textbf{6}(1): 26677.
+doi:\href{https://doi.org/10.1038/srep26677}{10.1038/srep26677}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-augeretal21}{}}%
+Auger-Méthé, M., Newman, K., Cole, D., Empacher, F., Gryba, R., King,
+A.A., Leos-Barajas, V., Mills Flemming, J., Nielsen, A., Petris, G., and
+others. 2021. A guide to state--space modeling of ecological time
+series. Ecological Monographs \textbf{91}(4): e01470.
+doi:\href{https://doi.org/10.1002/ecm.1470}{10.1002/ecm.1470}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-brittenetalinreview}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-croninpunt21}{}}%
+Cronin-Fine, L., and Punt, A.E. 2021. Modeling time-varying selectivity
+in size-structured assessment models. Fisheries Research \textbf{239}:
+105927. Elsevier.
+
 \leavevmode\vadjust pre{\hypertarget{ref-devalpinehastings02}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-dupontaviceetal22}{}}%
+Du Pontavice, H., Miller, T.J., Stock, B.C., Chen, Z., and Saba, V.S.
+2022. Ocean model-based covariates improve a marine fish stock
+assessment when observations are limited. ICES Journal of Marine Science
+\textbf{79}(4): 1259--1273.
+doi:\href{https://doi.org/10.1093/icesjms/fsac050}{10.1093/icesjms/fsac050}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-fischetal23}{}}%
+Fisch, N., Shertzer, K., Camp, E., Maunder, M., and Ahrens, R. 2023.
+Process and sampling variance within fisheries stock assessment models:
+Estimability, likelihood choice, and the consequences of incorrect
+specification. ICES Journal of Marine Science \textbf{80}(8):
+2125--2149. Oxford University Press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-fleischmanetal13}{}}%
+Fleischman, S.J., Catalano, M.J., Clark, R.A., and Bernard, D.R. 2013.
+An age-structured state-space stock--recruit model for pacific salmon
+(oncorhynchus spp.). Canadian Journal of Fisheries and Aquatic Sciences
+\textbf{70}(3): 401--414. NRC Research Press.
+
 \leavevmode\vadjust pre{\hypertarget{ref-hoyleetal22}{}}%
@@
 doi:\href{https://doi.org/10.1080/00401706.1981.10486293}{10.1080/00401706.1981.10486293}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-knape08}{}}%
+Knape, J. 2008. Estimability of density dependence in models of time
+series data. Ecology \textbf{89}(11): 2994--3000.
+doi:\href{https://doi.org/10.1890/08-0071.1}{10.1890/08-0071.1}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-legaultpalmer16}{}}%
+Legault, C.M., and Palmer, M.C. 2016. In what direction should the
+fishing mortality target change when natural mortality increases within
+an assessment? Canadian Journal of Fisheries and Aquatic Sciences
+\textbf{73}(3): 349--357. NRC Research Press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-legaultrestrepo99}{}}%
+Legault, C.M., and Restrepo, V.R. 1999. A flexible forward
+age-structured assessment program. Col. Vol. Sci. Pap. ICCAT
+\textbf{49}(2): 246--253.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-lietalinreview_a}{}}%
+Li, C., Deroba, J.J., Berger, A.M., Goethel, D.R., Langseth, B.J.,
+Schueller, A.M., and Miller, T.J. In review. Random effects on
+numbers-at-age transitions implictly 2 account for movement dynamics and
+improve 3 performance within a state-space stock assessment. Canadian
+Journal of Fisheries and Aquatic Sciences.
+
 \leavevmode\vadjust pre{\hypertarget{ref-lietalinreview}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-liljestrandetal24}{}}%
+Liljestrand, E.M., Bence, J.R., and Deroba, J.J. 2024. The effect of
+process variability and data quality on performance of a state-space
+stock assessment model. Fisheries Research \textbf{275}: 107023.
+Elsevier.
+
 \leavevmode\vadjust pre{\hypertarget{ref-magnussonhilborn07}{}}%
@@
 doi:\url{https://doi.org/10.1111/j.1467-2979.2007.00258.x}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-methotwetzel13}{}}%
+Methot, R.D., and Wetzel, C.R. 2013. Stock synthesis: A biological and
+statistical framework for fish stock assessment and fishery management.
+Fisheries Research \textbf{142}: 86--99.
+doi:\href{https://doi.org/10.1016/j.fishres.2012.10.012}{10.1016/j.fishres.2012.10.012}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-polanskyetal09}{}}%
+Polansky, L., De Valpine, P., Lloyd-Smith, J.O., and Getz, W.M. 2009.
+Likelihood ridges and multimodality in population growth rate models.
+Ecology \textbf{90}(8): 2313--2320.
+doi:\href{https://doi.org/10.1890/08-1461.1}{10.1890/08-1461.1}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-punt23}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-puntetal14}{}}%
+Punt, A.E., Hurtado-Ferro, F., and Whitten, A.R. 2014. Model selection
+for selectivity in fisheries stock assessments. Fisheries Research
+\textbf{158}: 124--134. Elsevier.
+
 \leavevmode\vadjust pre{\hypertarget{ref-shibata76}{}}%
@@
 doi:\href{https://doi.org/10.1093/biomet/63.1.117}{10.1093/biomet/63.1.117}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-stewartmonnahan17}{}}%
+Stewart, I.J., and Monnahan, C.C. 2017. Implications of process error in
+selectivity for approaches to weighting compositional data in fisheries
+stock assessments. Fisheries Research \textbf{192}: 126--134. Elsevier.
 
@@
 105967. doi:\url{https://doi.org/10.1016/j.fishres.2021.105967}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-stocketal21}{}}%
+Stock, B.C., Xu, H., Miller, T.J., Thorson, J.T., and Nye, J.A. 2021.
+{Implementing two-dimensional autocorrelation in either survival or
+natural mortality improves a state-space assessment model for Southern
+New England-Mid Atlantic yellowtail flounder}. Fisheries Research
+\textbf{237}: 105873.
+
+\leavevmode\vadjust pre{\hypertarget{ref-szuwalskietal18}{}}%
+Szuwalski, C.S., Ianelli, J.N., and Punt, A.E. 2018. Reducing
+retrospective patterns in stock assessment and impacts on management
+performance. ICES Journal of Marine Science \textbf{75}(2): 596--609.
+Oxford University Press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-thorsonminto15}{}}%
+Thorson, J.T., and Minto, C. 2015. Mixed effects: A unifying framework
+for statistical modelling in fisheries biology. ICES Journal of Marine
+Science \textbf{72}(5): 1245--1256.
+doi:\href{https://doi.org/10.1093/icesjms/fsu213}{10.1093/icesjms/fsu213}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-trijouletetal20}{}}%
+Trijoulet, V., Fay, G., and Miller, T.J. 2020. Performance of a
+state-space multispecies model: What are the consequences of ignoring
+predation and process errors in stock assessments? Journal of Applied
+Ecology \textbf{57}(1): 121--135. Wiley Online Library.
 

### 2025-03-13 Ecov_study/mortality/manuscript/manuscript.Rmd
saving work on M+Ecov paper

--- 2024-01-19 Ecov_study/mortality/paper/paper.Rmd
@@
 author: 
-  - Timothy J. Miller^1^
-  - Andrew Applegate 
-  - Greg Britten
-  - Elizabeth N. Brooks
-  - Gavin Fay
-  - Alex Hansell
-  - Christopher M. Legault
-  - Brandon Muffley
-  - Brian C. Stock
-  - John Wiedenmann
-date: ''
+  - Timothy J. Miller^1,2^
+  - Greg Britten^3^
+  - Elizabeth N. Brooks^2^
+  - Gavin Fay^4^
+  - Alex Hansell^2^
+  - Christopher M. Legault^2^
+  - Brandon Muffley^5^
+  - John Wiedenmann^6^
+date: "`r format(Sys.time(), '%d %B, %Y')`"
 fontsize: 12pt
@@
   pdf_document:
-    citation_package: natbib
+    # citation_package: natbib
     fig_caption: yes
@@
     df_print: paged
-  word_document: default
-header-includes:
-- \usepackage{xspace}
-- \usepackage{bm}
-- \usepackage{caption,graphics}
-- \usepackage{graphicx}
-- \usepackage{makecell}
-- \renewcommand\figurename{Fig.}
-- \captionsetup{labelsep=period, singlelinecheck=false}
-- \newcommand{\changesize}[1]{\fontsize{#1pt}{#1pt}\selectfont}
-- \renewcommand{\arraystretch}{1.5}
-- \renewcommand\theadfont{}
+  # word_document: default
 csl: canadian-journal-of-fisheries-and-aquatic-sciences.csl
-biblio-style: cjfas2.bst
+#biblio-style: cjfas2.bst
 ---
@@
 library(forcats)
+type <- "latex"
 ```
 
-$^1$corresponding author: timothy.j.miller@noaa.gov, Northeast Fisheries Science Center, Woods Hole Laboratory, 166 Water Street, Woods Hole, MA 02543 USA\
+$^1$corresponding author: timothy.j.miller@noaa.gov\
+$^2$Northeast Fisheries Science Center, Woods Hole Laboratory, 166 Water Street, Woods Hole, MA 02543 USA\
+$^3$Woods Hole Oceanographic Institution\
+$^4$SMAST\
+$^5$Mid-Atlantic Fisheries Management Council\
+$^6$Rutgers University\
 
@@
 
-## Summary {-}
+## Abstract {-}
 
@@
 # Methods {-}
+
+All of our analyses used the Woods Hole Assessment Model (WHAM) to construct both OMs and EMs [@millerstock20;@stockmiller21].
+We used [version 1.0.6.9000, commit 77bbd94](https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3) for this simulation study.
+This package has been used extensively to configure OMs and EMs for several other simulation studies [@legaultetal23; @lietal24, @lietalinreview] and is used to assess many commercially important stocks in the Northeast US.
+
+-------------- need to edit this part
+
+There are many similarities to the configuration of OMs and EMs in simulation studies conducted by @milletetelinreview.
+We completed a simulation study with a number of OMs that can be categorized based on where process error random effects are assumed: abundance at age (R, R+S), natural mortality (R+M), fleet selectivity (R+Sel), or index catchability (R+q).
+For each OM assumption about variance of process errors and observations are required and the values we used were based on a review of the range of estimates from applications of WHAM in management of stocks of haddock, butterfish, and American plaice in the NE US.
+
+In total, we configured 72 OMs with alternative assumptions about the source and variability of process errors, level of observation error in indices and age composition data, and contrast in fishing pressure over time.
+We fitted 20 EMs to observations from each of 100 simulations where process errors were also simulated.
+For R+M, R+Sel, and R+q OMs, we used unique seeds for each simulation, but we inadvertently used the same 100 seeds for all R and R+S OMs.
+Each of the EMs made alternative assumptions about the source of process errors and whether natural mortality (or the median for models with process error in natural mortality) was estimated and whether a Beverton-Holt stock recruit relationship was estimated within the EM.
+Details of each of the operating and EMs are described below.
+
+We did not use the log-normal bias-correction feature for process errors or observations described by [@stockmiller21] for operating and EMs [@lietalinreview].
+Simulations and model fitting were all carried out on the University of Massachusetts Green High-Performance Computing Cluster.
+All code we used to perform the simulation study and summarize results can be found at https://github.com/timjmiller/SSRTWG/tree/main/ecov_study/mortality/code.
+
+--------------
 

### 2025-03-13 Ecov_study/mortality/code/manuscript/aic_results.R

@@ -1,0 +1,457 @@
+library(here)
+library(ggplot2)
+library(ggh4x)
+library(dplyr)
+library(tidyr)
+library(ggpattern)
+df.ems = readRDS(file.path(here(),"Project_0","inputs", "df.ems.RDS"))
+
+
+all_naa_aic <- readRDS(file = file.path(here(),"Project_0","results", "all_naa_aic_results.RDS"))
+aic_fn <- function(all, est_ind, om_ind = NULL){
+  if(!is.null(om_ind)) all <- all[om_ind]
+  out <- sapply(all, function(res){
+    tmp = apply(res[est_ind,],2, function(x) {
+      if(any(!is.na(x))) {
+        return(x == min(x,na.rm=T))
+      } else return(rep(NA, length(x)))
+    })
+    return(apply(tmp,1,sum,na.rm=T))
+  })  
+  return(out)
+}
+
+make_plot_df <- function(aic_res, df.ems, df.oms, em_ind = 1:20, SR_est =FALSE, M_est = FALSE){
+  use.df.ems <- df.ems[em_ind,]
+  SR_M_ind = which(use.df.ems$SR_model == ifelse(SR_est,3,2) & use.df.ems$M_est == M_est)
+  aic_res <- t(aic_fn(aic_res, SR_M_ind))
+  df <- cbind(df.oms[-1], aic_res)
+  print(df)
+  df <- df %>% pivot_longer(cols = as.character(1:5), names_to = "EM", values_to = "n") %>% as.data.frame
+  this.df.ems <- use.df.ems[SR_M_ind,]
+  print(SR_M_ind)
+  print(this.df.ems)
+  print(df$EM)
+  df <- cbind(this.df.ems[as.integer(df$EM),],df)
+  print(head(df))
+  df$correct_EM_PE <- "No"
+  if(max(em_ind)==20) {
+    df <- df %>% mutate(EM_process_error = recode(EM,
+        "1" = "R",
+        "2" = "R+S",
+        "3" = "R+M (iid)",
+        "4" = "R+Sel (iid)",
+        "5" = "R+q (iid)"
+      ))
+    df$NAA_sig[which(is.na(df$NAA_sig))] <- 0
+    df$correct_EM_PE[df$NAA_sig == 0 & df$EM_process_error == "R"] <- "Yes"
+    df$correct_EM_PE[df$NAA_sig >  0 & df$EM_process_error == "R+S"] <- "Yes"
[…]

### 2025-03-13 Ecov_study/mortality/code/manuscript/convergence_results.R

@@ -1,0 +1,178 @@
+library(here)
+library(ggplot2)
+library(ggh4x)
+library(dplyr)
+library(tidyr)
+library(patchwork)
+library(scales)
+
+#convergence information:
+#1: 0/1 model finished optimizing
+#2: nlminb convergence flag
+#3: max gradient value
+#4: number of NaNs in SEs for parameters, 0 = good invertible hessian
+#5: maximum non-NaN SE estimate
+
+
+conv_fn <- function(all, est_ind, om_ind = NULL){
+  if(!is.null(om_ind)) all <- all[om_ind]
+  out <- sapply(all, function(res){
+    tmp <- c(sapply(res[est_ind], function(x) {
+      #print(est_ind)
+      #print(length(res))
+      #print(tmp)
+      nconv <- c(
+        sum(!is.na(x[,1])), 
+        sum(!is.na(x[,2]) & x[,2] == 0), 
+        sum(!is.na(x[,3]) & x[,3] == 0),
+        sum(!is.na(x[,4]) & x[,4] < 1e-6),
+        sum(!is.na(x[,5]) & x[,3] == 0 & x[,5] < 10)
+      )
+      nconv <- c(nconv, NROW(x))
+      return(nconv)
+    }))
+    return(tmp)
+  })  
+  return(t(out))
+}
+
+#M estimated
+conv_res_plotting_fn <- function(conv_res, M_est = TRUE, Ecov_est = TRUE){
+
+  for(i in 1:3) for(j in 1:3){
+    
+    re_mods <- c("rec", "rec+1", "rec+M")
+    #re_mod <- c("rec", "rec+1", "rec+M")[i]
+    #EM:  M fixed, OM and EM RE assumption match
+    om_ind <- which(df.oms$NAA_M_re == re_mods[i]) #om and em match
+    em_ind <- which(df.ems$Ecov_est == Ecov_est & df.ems$M_est == M_est & df.ems$re_config == re_mods[j])
+    conv <- conv_fn(conv_res, em_ind, om_ind)
+    colnames(conv) <- c(paste0("Type", 1:5, "_n_pass"), "n_sim")
+    res <- cbind(df.oms[om_ind,], df.ems[em_ind,], conv)
+    if(i == 1 & j == 1) {
+      df <- res
+    } else {
+      df <- rbind(df, res)
+    }
+  }
+  for(i in 1:5){
+    df[[paste0("Type", i, "_p_pass")]] <- df[[paste0("Type", i, "_n_pass")]]/df$n_sim
[…]

### 2025-03-13 Ecov_study/mortality/code/manuscript/mohns_rho_results.R

@@ -1,0 +1,392 @@
+library(here)
+library(ggplot2)
+library(ggh4x)
+library(dplyr)
+library(tidyr)
+library(patchwork)
+library(reshape2)
+
+median_ci_fn <- function(x, alpha = 0.05){#, n.yrs = 1, n.sim = 100) {
+  x <- x[which(!is.na(x))]
+  n <- length(x)
+  bnds <- qbinom(c(alpha/2,1-alpha/2), n, 0.5)/n # 95% CI bounds for median
+  r <- quantile(x, probs = c(bnds[1], 0.5, bnds[2]))
+  names(r) <- c("lo", "middle", "hi")
+  r
+}
+make_mohns_rho_df <- function(om_type = "naa", res = all_naa_mohns_rho) {
+  df.ems <- readRDS(here::here("Project_0","inputs", "df.ems.RDS"))
+  if(om_type == "naa") {
+    em_ind <- 1:20
+  }
+  if(om_type == "M") {
+    em_ind <- 5:24
+  }
+  if(om_type == "Sel") {
+    em_ind <- c(5:20,25:28)
+  }
+  if(om_type == "q") {
+    em_ind <- c(5:20,29:32)
+  }
+  df.ems <- df.ems[em_ind,]
+
+  res <- melt(res)
+  names(res) <- c("rho", "em", "sim","om")
+  res$Type = c("SSB","F","R")
+  res <- cbind(df.ems[res$em,], res)
+
+  df <- res %>% group_by(om, em, Type) %>%
+    reframe(stats = median_ci_fn(rho)) %>% as.data.frame
+  df$type <- c("lo", "middle", "hi")
+  if(om_type == "naa") {
+    df.oms <- readRDS(here::here("Project_0", "inputs", "df.oms.RDS"))
+  } else {
+    df.oms <- readRDS(here::here("Project_0", "inputs", paste0("df.", om_type, ".oms.RDS")))
+  }
+  df <- df %>% pivot_wider(names_from = type, values_from = stats) %>% as.data.frame
+  df <- cbind(df, df.oms[df$om,])
+  df <- cbind(df, df.ems[df$em,])
+  df <- df %>%
+    mutate(EM_process_error = recode(re_config,
+      "rec" = "R",
+      "rec+1" = "R+S",
+      "M_re" = "R+M",
+      "q_re" = "R+q",
+      "sel_re" = "R+Sel"
+    ))
[…]

### 2025-04-08 Project_0/manuscript/manuscript.Rmd
internal review for project_0 paper

@@
 $^4$SMAST\
-$^5$Institute of Marine Research\
-$^6$Mid-Atlantic Fisheries Management Council\
-$^7$Rutgers University\
+$^5$Mid-Atlantic Fishery Management Council, 800 North State Street, Suite 201, Dover, DE 19901 USA\
+$^6$Institute of Marine Research, Nye Flødevigveien 20, 4817 His, Norway\
+$^7$Department of Ecology, Evolution, and Natural Resources. Rutgers University\
 
@@
 
+**keywords:** stock assessment, state-space, model selection, bias, convergence, retrospective patterns
+
 ## Abstract {-}
 
-<!-- note CJFAS abstract word limit 175, so I have drafted a shorter summary here -->
-
-State-space assessment models, which are increasingly being used for management of commercially important fish stocks, can include process errors in many ways, and an evaluation of the statistical reliability of the estimates of management output and best practices of selection among alternative process error configurations are needed.
-We simulated 72 operating models with varying fishing pressure history, observation error error, and process error magnitude, correlation, and source (recruitment, survival, fishery selectivity, catchability, and natural mortality).
-We fit estimating models with different assumptions on the process error source, whether median natural mortality was estimated, and whether a stock-recruit relationship was estimated.
-Estimating models that assumed the correct process error source and median natural mortality rate was known, and did not estimate a stock recruit relationship generally had high probability of convergence and low bias in terminal year SSB estimation.
+State-space models are increasingly used for stock assessment, and evaluations of their statistical reliability and best practices for selecting among process error configurations are needed.
+We simulated 72 operating models that varied fishing pressure and observation error across process errors in recruitment, survival, selectivity, catchability, and/or natural mortality.
+We fit estimating models with different assumptions on the process error source and whether median natural mortality or a stock-recruit relationship were estimated.
+Estimating models without a stock-recruit relationship that assumed the correct process error source and median natural mortality had high convergence rates and low bias.
 Bias was also low under many incorrect process error assumptions when there was contrast in fishing pressure and low observation error.
-Stock-recruitment parameters were not reliably estimated in any scenarios, but the least accurate estimates occured under ideal situations of low observation error, low recruitment variability, and temporal contrast in fishing pressure.
-However, estimation of median natural mortality rate was reliable for many OMs (other than those with survival process errors) with temporal contrast in fishing pressure regardless of the estimated process error type.
 Marginal AIC most accurately distinguished process errors on recruitment, survival, and selectivity, as well as larger magnitude process errors of other types.
-Retrospective patterns for spawning stock biomass and fishing mortality were generally weak, but those for recruitment were sizable when observation error was high, even with the correct process error assumptions.
+Retrospective patterns were generally small but were sizable for recruitment when observation error was high.
+These results help establish the statistical reliability of state space assessment models and pave the way for the next-generation of fisheries stock assessment
 
@@
 Modern state-space models can estimate the annually varying parameters as random effects with variance parameters estimated using maximum marginal likelihood or corresponding Bayesian approaches.
-These latter approaches are considered best practice  and a requirement for the next generation of stock assessment models [@hoyleetal22; @punt23].
+These latter approaches are considered best practice and a recommended for the next generation of stock assessment models [@hoyleetal22; @punt23].
 
@@
 Second, previous studies relied on operating models conditioned on specific fisheries, limiting their generalizability [@lietalinreview_a; @liljestrandetal24].
-In particular, the effects of observation error and underlying fishing history were not fully isolated from their study design, making it challenging to disentangle the interplay between process and observation error magnitudes, as demonstrated in @fischetal23.
-Third, explicitly modeling stock-recruitment relationships as mechanistic drivers of population dynamics is promising [@fleischmanetal13; @dupontaviceetal22] but can be challenging in state-space models.
-Evidence suggests that when both process and observation errors are unknown, estimating density dependence parameters becomes highly uncertain [@knape08; @polanskyetal09].
+In particular, the effects of observation error and underlying fishing history have not been fully isolated in simulation study designs, making it challenging to disentangle the interplay between process and observation error magnitudes, as demonstrated in @fischetal23.
+Third, explicitly modeling stock-recruit relationships (SRRs) as mechanistic drivers of population dynamics is promising [@fleischmanetal13; @dupontaviceetal22], but can reliability of inferences within integrated state-space age-structured models has not been evaluated.
+Evidence from other studies suggests that when both process and observation errors are unknown, estimating density dependence parameters becomes highly uncertain [@knape08; @polanskyetal09].
 In particular, @knape08 demonstrated that stronger density dependence becomes increasingly difficult to estimate in the presence of observation error.
-Therefore, it is crucial to assess whether density dependence mechanisms can be estimated with sufficient precision before drawing ecological conclusions [@augeretal16].
+Therefore, it is crucial to assess whether density dependence mechanisms can be estimated with sufficient precision for use in fisheries management [@augeretal16].
 
 In the present study, we conduct a simulation study with operating models (OMs) varying by degree of observation error, source and variability of process error, and fishing history.
-The simulations from these OMs are fitted with estimation models (EMs) that make alternative assumptions for sources of process error, whether a stock-recruit model was estimated, and whether a constant, or, in some EMs, median, natural mortality is estimated.
+The simulations from these OMs are fitted with estimation models (EMs) that make alternative assumptions for sources of process error, whether a SRR was estimated, and whether a constant, or, in some EMs, median, natural mortality is estimated.
 Given the confounding nature of process errors, developing diagnostic tools to detect model misspecification is of great scientific interest and could aid the next generation of stock assessments [@augeretal21].
-We evaluate whether convergence and Akaike Information Criterion (AIC) can correctly determine the source of process error and the existence of a stock-recruit relationship.
+We evaluate whether convergence and Akaike Information Criterion (AIC) can correctly determine the source of process error and the existence of a SRR.
 We also evaluate when retrospective patterns occur and the degree of bias in the outputs of the assessment model that are important for management.
@@
 We used the Woods Hole Assessment Model (WHAM) to configure OMs and EMs in our simulation study [@millerstock20;@stockmiller21].
-WHAM is an R package freely available as a github repository.
+WHAM is an R package freely available via a github repository and is built on the Template Model Builder package [@kristensenetal16].
 For this study we used [version 1.0.6.9000, commit 77bbd94](https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3).
-This package has also been used to configure OMs and EMs for closed loop simulations evaluating index-based assessment methods [@legaultetal23] and is used for management of haddock, butterfish, American plaice, bluefish, Atlantic cod, black sea bass, and yellowtail flounder in the Northeast US.
+WHAM has also been used to configure OMs and EMs for closed loop simulations evaluating index-based assessment methods [@legaultetal23] and is currently used or accepted for use in management of numerous fish stocks in the Northeast United States (NEUS) [e.g., @nefsc22;@nefsc22a;@nefsc24].
 
-We completed a simulation study with a number of OMs that can be categorized based on where process error random effects are assumed: abundance at age (R, R+S), natural mortality (R+M), fleet selectivity (R+Sel), or index catchability (R+q).
-For each OM assumption about variance of process errors and observations are required and the values we used were based on a review of the range of estimates from applications of WHAM in management of stocks of haddock, butterfish, and American plaice in the NE US.
+We completed a simulation study with a number of OMs that can be categorized based on where process error random effects were assumed: recruitment (R, assumed present in all models), apparent survival (denoted R+S), natural mortality (R+M), fleet selectivity (R+Sel), or index catchability (R+q).
+We refer to the (R+S) OMs as modeling apparent survival because on logscale the random effects ($\epsilon_{a,y}$) are additive to the total mortality (F+M) between numbers at age, thus they modify the survival term.
+However, as @stockmiller21 note, these random effects can be due to events other than mortality, such as immigration, emigration, missreported catch, and other sources of misspecification.
+For each OM, assumptions about the magnitude of the variance of process errors and observations are required and the values we used were based on a review of the range of estimates from NEUS assessments using WHAM in stocks NEUS.
 
-In total, we configured 72 OMs with alternative assumptions about the source and variability of process errors, level of observation error in indices and age composition data, and contrast in fishing pressure over time.
-We fitted 20 EMs to observations from each of 100 simulations where process errors were also simulated.
-For R+M, R+Sel, and R+q OMs, we used unique seeds for each simulation, but we inadvertently used the same 100 seeds for all R and R+S OMs.
-Each of the EMs made alternative assumptions about the source of process errors and whether natural mortality (or the median for models with process error in natural mortality) was estimated and whether a Beverton-Holt stock recruit relationship was estimated within the EM.
-Details of each of the operating and EMs are described below.
+In total, we configured 72 OMs with alternative assumptions about the source and magnitude of process errors, magnitude of observation error in indices and age composition data, and contrast in fishing pressure over time.
+We fitted 20 EMs to observations generated from each of 100 simulations where process errors were also simulated.
+Each EM differed in assumptions about the source of process errors, whether natural mortality (or the median for models with process error in natural mortality) was estimated, and whether a Beverton-Holt SRR was estimated within the EM.
+Details of each of the OMs and EMs are described below.
 
-We did not use the log-normal bias-correction feature for process errors or observations described by [@stockmiller21] for operating and EMs [@lietalinreview].
+We did not use the log-normal bias-correction feature for process errors or observations described by [@stockmiller21] for OMs and EMs to simplify interpretation of the study results [@lietalinreview].
 Simulations and model fitting were all carried out on the University of Massachusetts Green High-Performance Computing Cluster.
@@
 
-The population consists of 10 age classes: ages 1 to 10+ and we assume spawning occurs each year 1/4 of the way through the year.
+The population consists of 10 age classes, ages 1 to 10+, with the last being a plus group that accumulates ages 10 and older.
+We assume spawning occurs annually 1/4 of the way through the year.
 The maturity at age was a logistic curve with $a_{50}$ = 2.89 and slope = 0.88 (Figure \ref{om_inputs_fig}, top left).
@@
 
-We assumed a Beverton-Holt stock recruit function with constant pre-recruit mortality parameters for all OMs.
-All post-recruit productivity components are constant in the NAA and survey catchability process error OMs.
-Therefore steepness and unfished recruitment are also constant over the time period for those OMs [@millerbrooks21].
-We specified unfished recruitment = $R_0 = e^{10}$ and $\Fmsy = F_{40\%} = 0.348$ equated to a steepness of 0.69 and $a=0.60$ and $b = 2.4 \times 10^{-5}$ for the
+We assumed a Beverton-Holt SRR with constant pre-recruit mortality parameters for all OMs.
+All post-recruit productivity components are constant in the apparent survival (R+S) and survey catchability (R+Sel) process error OMs.
+Therefore, steepness and unfished recruitment are also constant over the time period for those OMs [@millerbrooks21].
+We specified unfished recruitment equal to $e^{10}$ and $\Fmsy = F_{40\%} = 0.348$, which equates to a steepness of 0.69 and $a=0.60$ and $b = 2.4 \times 10^{-5}$ for the Beverton-Holt parameterization
 \[
@@
 \]
-Beverton-Holt parameterization (Figure \ref{om_inputs_fig}, bottom right).
-For OMs without process errors on natural mortality we assumed the rate was assumed 0.2.
-For OMs with process errors on natural mortality the median natural mortality rate was 0.2.
+(Figure \ref{om_inputs_fig}, bottom right).
+For OMs without process errors on natural mortality we fixed the rate at 0.2.
+For OMs with process errors on natural mortality, the median natural mortality rate was specified to be 0.2.
 
 We used two fishing scenarios for OMs.
-In the first scenario, the stock experiences overfishing at 2.5\Fmsy for the first 20 years and fishing at \Fmsy for the last 20 years (denoted $2.5\Fmsy \rightarrow \Fmsy$).
-In the second scenario, the stock is fished at \Fmsy for the entire time period.
-The magnitude of the overfishing assumptions is based on average estimates of overfishing for NE groundfish stocks from [@wiedenmannetal19].
-@legaultetal23 also used similar approaches to defining fishing mortality histories for OMs.
+In the first scenario, the stock experiences overfishing at 2.5\Fmsy for the first 20 years followed by fishing at \Fmsy for the last 20 years (denoted $2.5\Fmsy \rightarrow \Fmsy$).
+In the second scenario, the stock is fished at \Fmsy for the entire time period (40 years).
+The magnitude of the overfishing assumptions is based on average estimates of overfishing for NEUS groundfish stocks from @wiedenmannetal19 and similar to the approach in @legaultetal23.
 
-We specified initial population abundance at age at the equilibrium distribution fishing at either $F = 2.5\times \Fmsy$ or $F = \Fmsy$ for the two alternative fishing histories.
+We specified initial population abundance at age at the equilibrium distribution that corresponds to fishing at either $F = 2.5\times \Fmsy$ or $F = \Fmsy$.
 This implies that, for a deterministic model, the abundance at age would not change from the first year to the next.
 
-For OMs with time-varying random effects for M, steepness is not constant, but we used the same alpha and beta parameters as other OMs this equates to a steepness and R0 at the median of the time series process for M.
-For OMs with time-varying random effects for fishery selectivity, \Fmsy is also not constant however we use the same F history as other OMs which corresponds to Fmsy at the mean selectivity parameters.
+For OMs with time-varying random effects for M , steepness is not constant.
+However, we used the same $a$  and $b$ parameters as other OMs, which equates to a steepness and R0 at the median of the time series process for M.
+For OMs with time-varying random effects for fishery selectivity, \Fmsy is also not constant, but since we use the same F history as other OMs, this corresponds to \Fmsy at the mean selectivity parameters.
 
@@
 
-We assumed a single fleet operating year round for catch observations with logistic selectivity for the fleet with $a_{50} = 5$ and slope = 1 (Figure \ref{om_inputs_fig}, bottom left).
-This selectivity is was used to define \Fmsy for the Beverton-Holt stock recruitment parameters above.
+We assumed a single fleet operating year round for catch observations with logistic selectivity for the fleet ($a_{50} = 5$ and slope = 1; Figure \ref{om_inputs_fig}, bottom left).
+This selectivity was used to define \Fmsy for the Beverton-Holt SRR parameters above.
 We assumed a logistic-normal distribution with no correlation on the multivariate normal scale for the age-composition observations for the fleet.
@@
 
-Two time series of surveys are assumed and observed in numbers rather than biomass for the entire 40 year period with one occurring in the spring (0.25 of each year) and one in the fall (0.75 of each year).
+Two time series of fishery-independent surveys in numbers are generated for the entire 40 year period with one occurring in the spring (0.25 of each year) and one in the fall (0.75 of each year).
 Catchability of both surveys are assumed to be 0.1.
-Like the fishing fleet, we assumed logistic selectivity for both indices with $a_{50} = 5$ and slope = 1 and a logistic-normal distribution with no correlation on the multivariate normal scale for the age-composition observations.
+Like the fishing fleet, we assumed logistic selectivity for both indices ($a_{50} = 5$ and slope = 1) and a logistic-normal distribution with no correlation on the multivariate normal scale for the age-composition observations.
 
@@
 
-Standard deviation for log-aggregate catch was 0.1.
-There were two levels of observation error variance for indices and age composition for both indices and fleet catch.
-A low uncertainty specification assumed standard deviation of both series of log-aggregate index observations was 0.1 and the standard deviation of the logistic-normal for age composition observations was 0.3 In the high uncertainty specification the standard deviation for log-aggregate indices was 0.4 and that for the age composition observations was 1.5.
-For all EMs, standard deviation for log-aggregate observations was assumed known whereas that for the logistic-normal age composition observations was estimated.
+The standard deviation for log-aggregate catch was 0.1.
+Two levels of observation error variance (high and low) were specified for indices and all age composition observations (both indices and catch).
+The low uncertainty specification assumed a standard deviation of 0.1 for both series of log-aggregate index observations, and the standard deviation of the logistic-normal for age composition observations was 0.3.
+In the high uncertainty specification, the standard deviation for log-aggregate indices was 0.4 and that for the age composition observations was 1.5.
+For all EMs, the standard deviation for log-aggregate observations was assumed known whereas that for the logistic-normal age composition observations was estimated.
 
@@
 
-For operating models with random effects on recruitment and(or) survival (R, R+S) we assumed marginal standard deviations for recruitment of $\sigma_R \in \{0.5,1.5\}$ and marginal standard deviations for older age classes of $\sigma_{2+} \in \{0,0.25, 0.5\}$.
-The full factorial combination of these process error assumptions and fishing history (2 levels) and observation error (2 levels) scenarios described above results in 24 different R ($\sigma_{2+} = 0$) and R+S operating models (Table \ref{naa_om_table}).
+For operating models with random effects on recruitment and(or) apparent survival (R, R+S), we assumed marginal standard deviations for recruitment of $\sigma_R \in \{0.5,1.5\}$ and marginal standard deviations for older age classes of $\sigma_{2+} \in \{0,0.25, 0.5\}$.
+The full factorial combination of these process error assumptions (2x3 levels) and scenarios for fishing history (2 levels) and observation error (2 levels) scenarios described above results in 24 different R ($\sigma_{2+} = 0$) and R+S operating models (Table \ref{naa_om_table}).
 
 ### Operating models with random effects on natural mortality {-}
-
-All R+M OMs treat natural mortality constant across age, but with annually varying random effects. WHAM treats natural mortality as a log-transformed parameter 
+All R+M OMs treat natural mortality as constant across age, but with annually varying random effects. WHAM treats natural mortality as a log-transformed parameter 
 $$
@@
 $$
-that is a linear combination of a mean log-natural mortality parameter that is constant across ages $\mu_{M} = \log(0.2)$ and any annual random effects marginally distributed as $\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)$.
-Uncorrelated random effects were also included on recruitment with $\sigma_R = 0.5$ (hence, R+M).
+that is a linear combination of a mean log-natural mortality parameter that is constant across ages ($\mu_{M} = \log(0.2)$) and any annual random effects are marginally distributed as $\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)$.
+Uncorrelated random effects were also included on recruitment with $\sigma_R = 0.5$ (hence, we denote these OMs as  R+M).
 The marginal standard deviations we assumed for log natural mortality random effects were $\sigma_M \in \{0.1, 0.5\}$ and AR1 autocorrelation parameters of $\rho_M \in \{0,0.9\}$.
@@
 
-MORE SPECIFICS about correlation  of random effects? Both selectivity pars? just correlated by year?
 WHAM treats selectivity parameter $s$ as a logit-transformed parameter 
@@
 $$
-that is a linear combination of a mean $\mu_s$ and any annual random effects marginally distributed as $\epsilon_{s,y} \sim \text{N}\left(0,\sigma_s^2\right)$ where the lower and upper bounds of the parameter ($l_s$ and $u_s$) can be specified by the user.
-All selectivity parameters are either $a_50$ or slope parameters and we assume bounds of 0 and 10 for all selectivity parameters for all operating and EMs.
+that is a linear combination of a mean $\mu_s$ and any annual random effects marginally distributed as $\epsilon_{s,y} \sim \text{N}\left(0,\sigma_s^2\right)$, where the lower and upper bounds of the parameter ($l_s$ and $u_s$) can be specified by the user.
+All selectivity parameters ($a_{50}$ and slope parameters) were bounded by 0 and 10 for all OMs and EMs.
 The marginal standard deviations we assumed for logit scale random effects were $\sigma_s \in \{0.1, 0.5\}$ and AR1 autocorrelation parameters of $\rho_s \in \{0,0.9\}$.
-Like R+M OMs,  the full factorial combination of these process error assumptions and fishing history (2 levels) and observation error (2 levels) scenarios described above results in 16 different R+Sel OMs (Table \ref{sel_om_table}).
+Like R+M OMs, the full factorial combination of these process error assumptions (2x2 levels) and scenarios described above for fishing history (2 levels) and observation error (2 levels) results in 16 different R+Sel OMs (Table \ref{sel_om_table}).
 
@@
 that is a linear combination of a mean $\mu_i$ and any annual random effects marginally distributed as $\epsilon_{i,y} \sim \text{N}\left(0,\sigma_i^2\right)$ where the lower and upper bounds of the catchability ($l_i$ and $u_i$) can be specified by the user.
-Here we assume bounds of 0 and 1000 for all operating and EMs.
-For operating and EMs with process errors on catchability, the temporal variation is only assumed for the first index.
+We assumed bounds of 0 and 1000 for all OMs and EMs.
+For all OMs and EMs with process errors on catchability, the temporal variation only applies to the first index, which could be interpreted as capturing some unmeasured seasonal process that affects availability to the survey.
 The marginal standard deviations we assumed for logit scale random effects were $\sigma_i \in \{0.1, 0.5\}$ and AR1 autocorrelation parameters of $\rho_i \in \{0,0.9\}$.
@@
 
-For each data set simulated from an OM 20 EMs were fit.
+For each of the data sets simulated from an OM, 20 EMs were fit.
 A total of 32 different EMs were fit across OMs where the subset of 20 depended on the source of process error in the OM (Table \ref{em_table}).
-The EMs make different assumptions about the source of process error (R+S, R+M, R+Sel, R+q) and whether there is temporal autocorrelation or not, and whether a Beverton-Holt stock recruit function is estimated and the natural mortality rate ($\mu_M$, the constant or mean on log scale for R+M EMs) is estimated.
+The EMs have different assumptions about the source of process error (R+S, R+M, R+Sel, R+q) and whether or not 1) there is temporal autocorrelation, 2) a Beverton-Holt SRR is estimated, and 3) the natural mortality rate ($\mu_M$, the constant or mean on log scale for R+M EMs) is estimated.
 For simplicity we refer to the derived estimate $e^{\mu_M}$ as the median natural mortality rate regardless of whether natural mortality random effects are estimated in the EM.
-The first 20 EMs in Table \ref{em_table} were fit to simulate data sets from R and R+S OMs.
-EMs 5 to 24 in Table \ref{em_table} were fit to simulate data sets from R+M OMs.
-EMs 5 to 20 and 25-28 in Table \ref{em_table} were fit to simulate data sets from R+Sel OMs.
-Finally, EMs 5 to 20 and 29-32 in Table \ref{em_table} were fit to simulate data sets from R+q OMs.
+
+Subsets of 20 EMs in Table \ref{em_table} were fit to simulate data sets from each of the OM process error categories.
+For R and R+S OMs, fitted EMs had matching process error assumptions as well as R+Sel, R+M, and R+q assumptions without autocorrelation.
+Similarly, For other OM process error categories, we fit EMs with matching process error assumptions as well as other process error types without autocorrelation.
 The maturity at age, weight at age for catch and SSB, and observation error variance of aggregate catch and indices were all assumed known at the true values.
+However, the variance parameters for the logistic-normal distributions for age composition observations were estimated in the EMs.
 
@@
 
+### Convergence
+
 The first measure of reliability we investigated was frequency of convergence when fitting each EM to the simulated data sets.
-There are various ways to assess convergence of the fit [e.g., @carvalhoetal21], but given the importance of estimates of uncertainty when using assessment models in management, we estimated probablity of convergence as measured by occurrence of a positive-definite hessian matrix at the optimized negative log-likelihood that could be inverted.
-However, we also provided results in the Supplementary Materials for the maximum of the absolute values among all gradients for all fits of a given EM to all simulated data sets from a given OM that produced hessian-based standard errors for all estimated fixed effects.
+There are various ways to assess convergence of the fit [e.g., @carvalhoetal21; @kapuretal24], but given the importance of estimates of uncertainty when using assessment models in management, we estimated probablity of convergence as measured by occurrence of a positive-definite hessian matrix at the optimized negative log-likelihood that could be inverted.
+We also provide results in the Supplementary Materials for the maximum of the absolute values among all gradients for all fits of a given EM to all simulated data sets from a given OM that produced hessian-based standard errors for all estimated fixed effects.
 This provides an indication of how poor the calculated gradients can be, but still presumably converged adequately enough for parameter inferences.
-
-<!--
-probability of convergence as measured by all gradient values at the optimized negative log-likelihood that were less than $1\times10^{-6}$ in absolute value.
 We used the Clopper-Pearson exact method for constructing 95\% confidence intervals of the probabilities of convergence [@clopperpearson34;@thulin14].
--->
 
@@
 We estimated the probability of selection of each process error model structure (R, R+S, R+M, R+Sel, R+q) using marginal AIC.
-For a given operating model, we compared AIC for EMs that all made the same assumptions about median natural mortality (known or estimated) and stock recruitment model (Beverton-Holt or none).
+For a given operating model, we compared AIC for EMs that all made the same assumptions about median natural mortality (known or estimated) and SRR (Beverton-Holt or none).
 
-We also estimated the probability of correctly selecting between models with Beverton-Holt stock recruit function assumed and models without the stock-recruit function (null model).
-We made these comparisons between models that otherwise assume the same process error structure as the operating model and both of the compared models either estimate median natural mortality or assume it is known.
-Contrast in fishing pressure and time series with recruitment at low stock size has been shown to improve estimability of stock-recruitment parameters [@magnussonhilborn07;@connetal10].
+We also estimated the probability of correctly selecting between EMs with Beverton-Holt SRR assumed and models without the SRR (null model).
+We made these comparisons between models that otherwise assumed the same process error structure as the operating model and both of the compared models either estimate median natural mortality or assume it is known.
+Contrast in fishing pressure and time series with recruitment at low stock size has been shown to improve estimation of SRR parameters [@magnussonhilborn07; @connetal10].
 Our preliminary inspections of the proportions of simulations where the correct recruitment model was chosen for a given set of OM factors (including contrast in fishing pressure) indicated generally poor performance of AIC.
@@
 
-All results only condition on whether all of the compared estimating models completed the optimization process without failure.
-We did not condition on convergence as defined by a gradient threshold or invertibity of the hessian because optimization can correctly determine the the correct likelihood that would indicate poor convergence because variance parameters may be at the lower bound of zero correctly for models that assume the incorrect process error structure.
+All model selection results condition on whether all of the compared estimating models completed the optimization process without failure.
+We did not condition on convergence as defined by a gradient threshold or invertibity of the hessian because optimization could correctly determine an inappropriate process error assumption by estimating variance parameters at the lower bound of zero.
+Such an optimization could indicate poor convergence but the likelihood would be equivalent to that without the mis-specified random effects and the AIC would be appropriately higher because more (variance) parameters were estimated.
+All other measures of reliability described below (bias and Mohn's $\rho$) use these same criteria for inclusion of EM fits in the summarized results.
 
@@
 We constructed 95\% confidence intervals for the median relative bias using the binomial distribution approach as in @millerhyun18 and @stockmiller21.
-We present results for terminal years SSB and recruitment, Beverton-Holt stock recruit parameters ($a$ and $b$), and median natural mortality rate.
-Results for terminal year fishing mortality are strongly negatively correlated with those for SSB and provided in the supplementary materials.
+We present results for bias in terminal year estimates of SSB and recruitment, Beverton-Holt stock recruit parameters ($a$ and $b$), and median natural mortality rate.
+Results for terminal year fishing mortality are were strongly negatively correlated with those for SSB and are provided in the Supplementary Materials.
 
+<!--
 Similar to the AIC results, bias results only condition on whether the estimating model completed the optimization process without failure.
-We did not condition on convergence as defined by a gradient threshold or invertibity of the hessian because the optimized model can provide reliable estimation of SSB, F, M, and stock recruit parameters whether or not the model was able to estimate non-zero random effects.
+We did not condition on convergence as defined by a gradient threshold or invertibility of the hessian because the optimized model can provide reliable estimation of assessment model output whether or not the model was able to estimate non-zero random effects.
 In practice, the model would be reconfigured to remove unnecessary process errors and produce otherwise equivalent parameter estimates.
+-->
 
@@
 
-We estimated Mohn's $\rho$ for SSB, fully-selected fishing mortality, and recruitment for each EM [@mohn99].
-We estimated 7 peels for each EM.
-We calculated median 95\% confidence intervals for Mohn's $\rho$ using the same methods as that for relative bias.
-Similar to the other results, retrospective results only condition on whether all of the peels of a given estimating model completed the optimization process without failure.
-We did not condition on convergence as defined by a gradient threshold or invertibity of the hessian because the optimized model can provide reliable estimation of SSB, fishing mortality rates, and recruitment whether or not the model was able to estimate non-zero random effects.
+We calculated Mohn's $\rho$ for SSB, fully-selected fishing mortality, and recruitment for each EM [@mohn99].
+We fit 7 peels for each EM and calculated median 95\% confidence intervals for Mohn's $\rho$ using the same methods as that for relative bias.
+<!--Similar to the other results, retrospective results only condition on whether all of the peels of a given estimating model completed the optimization process without failure.
+We did not condition on convergence as defined by a gradient threshold or invertibility of the hessian because the optimized model can provide reliable estimation of SSB, fishing mortality rates, and recruitment whether or not the model was able to estimate non-zero random effects.
+-->
 
@@
 
-For R and R+S OMs, convergence generally declines for most EMs when the median natural mortality rate is estimated and/or the Beverton-Holt stock recruit relationship is estimated even when the process error assumptions of the EMs and OMs match (Figure \ref{hessian_SE_convergence}, A).
-When there was high observation error and constant fishing pressure (change from $2.5F_{\text{MSY}}$ to $F_{\text{MSY}}$), convergence of all EM process error configurations other than R EMs was poor for R OMs ($\sigma_{2+} = 0$) regardless of whether median natural mortality and stock-recruit relationships were estimated.
-Convergence of R EMs was high for all R and R+S OMs except when there was high observation error and constant fishing pressure and median natural mortality and stock-recruit relationships were estimated.
-R+S EMs fit to R OMs exhibited poor convergence regardless of whether natural mortality or a stock-recruit relationship was estimated.
+For R and R+S OMs, convergence generally declined for most EMs when the median natural mortality rate was estimated and/or the Beverton-Holt SRR was estimated even when the process error assumptions of the EMs and OMs matched (Figure \ref{hessian_SE_convergence}, A).
+When there was high observation error and constant fishing pressure ($F = F_{\text{MSY}}$ for all 40 years), convergence was poor for of all EM process error configurations other than R EMs when fitted to R OMs ($\sigma_{2+} = 0$) regardless of whether median natural mortality and SRRs were estimated.
+Convergence of R EMs was high for all R and R+S OMs except when there was high observation error and constant fishing pressure, and when median natural mortality and SRRs were estimated.
+R+S EMs fit to R OMs exhibited poor convergence regardless of whether natural mortality or a SRR was estimated.
 R+S EMs fit to R+S OMs had highest convergence rates when there was contrast in fishing pressure and low observation error.
-Convergence rates were high for all EMs when fit to data from R+S OMs with lower observation error except those where median natural mortality and/or stock-recruit relationships were estimated.
+Convergence rates were high for all EMs when fit to data from R+S OMs with lower observation error except those where median natural mortality and/or SRRs were estimated.
 
-Convergence of all EMs fitted to R+M OMs was best when the OMs had higher natural mortality process error variability, low observation error, and contrast in fishing pressure  (Figure \ref{hessian_SE_convergence}, B).
-R+M EMs that estimated autocorrelation of process errors had poor convergence for R+M OMs when there was low natural mortality process error variability whether or not there was autocorrelation of the simulated process errors.
-R+S EMs fitted to data generated from R+M OMs always converged poorly whether median natural mortality and the Beverton-Holt stock-recruit relationship were estimated.
+Convergence of all EMs fitted to R+M OMs was highest when the OMs had higher natural mortality process error variability, low observation error, and contrast in fishing pressure  (Figure \ref{hessian_SE_convergence}, B).
+R+M EMs that estimated autocorrelation of process errors had poor convergence for R+M OMs when there was low natural mortality process error variability regardless of autocorrelation of the simulated process errors.
+R+S EMs fitted to data generated from R+M OMs always converged poorly whether or not median natural mortality and the Beverton-Holt SRR were estimated.
 
-R+S EMs, in particular, converged poorly when fit to data generated from R+Sel OMs with lower selectivity process error variability or higher observation error (Figure \ref{hessian_SE_convergence}, C).
-R+Sel EMs generally converged better than other EMs for R+Sel OMs with higher process error variability, lower observation error, and contrast in fishing pressure  regardless of whether median natural mortality or a stock recruit relationship was estimated.
+The R+S EMs, in particular, had poor convergence when fit to data generated from R+Sel OMs with lower selectivity process error variability or higher observation error (Figure \ref{hessian_SE_convergence}, C).
+R+Sel EMs generally converged better than other EMs for R+Sel OMs with higher process error variability, lower observation error, and contrast in fishing pressure regardless of whether median natural mortality or a SRR was estimated.
 
-For R+q OMs, convergence of R+q EMs is generally better than that of other EMs when there is contrast in fishing history (Figure (\ref{hessian_SE_convergence}, D).
-Convergence of R+S EMs is generally worse than that of all other EMs across all OMs whether or not median natural mortality or a stock recruit relationship is estimated.
-Again, convergence probability generally declines for all EMs when median natural mortality or a stock recruit relationship is estimated.
+For R+q OMs, convergence of R+q EMs was generally better than that of other EMs when there was contrast in fishing pressure (Figure (\ref{hessian_SE_convergence}, D).
+Convergence of R+S EMs was generally worse than that of all other EMs across all OMs whether or not median natural mortality or a SRR was estimated.
+Again, convergence probability generally declined for all EMs when median natural mortality or a SRR was estimated.
 
-We found a wide range of maximum absolute values of gradients for models that converged as defined here (Figure \ref{hess_grad}).
+We found a wide range of maximum absolute values of gradients for models that converged (Figure \ref{hess_grad}).
 The largest value observed for a given EM and OM combination was typically $<10^{-3}$, but many converged models had values greater than 1.
-For many OMs, EMs that assumed the correct process error type and did not estimate median natural mortality or the Beverton-Holt stock-recruit relationship produced the lowest gradient values.
+For many OMs, EMs that assumed the correct process error type and did not estimate median natural mortality or the Beverton-Holt SRR produced the lowest gradient values.
 
@@
 
-Marginal AIC accurately determines the correct process error assumptions in EMs when data are generated from R and R+S OMs, regardless of whether median natural mortality or a stock recruit relationship is estimated (Figure \ref{pe_aic}, A).
-Adding estimation of median natural mortality or a stock recruit relationship separately has a negligible effect on the accuracy of determining the correct process error assumption.
-When both are estimated, there is a noticeable reduction in accuracy when OMs have a constant fishing history, observation error is low and larger variability in recruitment process errors.
+Marginal AIC accurately determined the correct process error assumptions in EMs when data were generated from R and R+S OMs, regardless of whether median natural mortality or a SRR was estimated (Figure \ref{pe_aic}, A).
+Attempting to estimate median natural mortality or a SRR separately had a negligible effect on the accuracy of determining the correct process error assumption.
+When both were estimated, there was a noticeable reduction in accuracy when OMs had a constant fishing pressure, low observation error, and larger variability in recruitment process errors.
 
 For R+M OMs, marginal AIC only accurately determined the correct process error model and correlation structure when observation error was low and variability in natural mortality process errors was high (Figures \ref{pe_aic}, B).
-Estimating the median natural mortality rate reduced the accuracy of AIC for OMs that assumed natural mortality process errors were independent.
-For OMs with poor accuracy, AIC most frequently selected EMs with process errors in catchability (R+q) or selectivity (R+Sel).
-Selection of R+S EMs was typically unlikely.
+Of these OMs, estimating the median natural mortality rate only reduced the accuracy of AIC when natural mortality process errors were independent and fishing pressure was constant.
+For OMs with poor model selection accuracy, AIC most frequently selected EMs with process errors in catchability (R+q) or selectivity (R+Sel).
+Selection of R+S EMs was generally unlikely.
 
@@
 When there was low variability in selectivity process errors and high observation error, R+q or R+S EMs were more likely to have the best AIC.
-Whether median natural morality or stock recruit relationships were estimated appeared to have little effect on the performance of AIC.
+Whether median natural morality or SRRs were estimated appeared to have little effect on the performance of AIC.
 
 Marginal AIC most accurately determined the correct source of process error and correlation structure for R+q OMs with high variability in catchability process errors (Figures \ref{pe_aic},D).
-The worst accuracy occurred for OMs with low variability in catchability process errors and high observation error.
-However, in these OMs, the marginal AIC accurately determined the correct source of process error (but not correlation structure) except when EMs estimated both median natural morality and the stock recruit relationships and OMs assumed a constant fishing pressure.
+The R+q OMs with low variability in catchability process errors and high observation error had the least model selection accuracy.
+However, for these OMs, the marginal AIC accurately determined the correct source of process error (but not correlation structure) except when OMs assumed a constant fishing pressure and EMs estimated both median natural morality and the SRR.
 
@@
 
-Our comparisons of model performance condition on assuming the true process error configuration is known (EM and OM process error types match).
-Broadly, we found generally poor accuracy of AIC in selecting models assuming a Beverton-Holt stock recruit function over the null model without an assumed stock-recruit relationship for all OMs.
-However, we also found increased accuracy of AIC in determining the Beverton-Holt stock-recruit relationship with greater variation in spawning biomass generated in the simulated populations for nearly every OM (Figure \ref{sr_aic}).
+Our comparisons of model performance conditioned on assuming the true process error configuration is known (EM and OM process error types match) and we focus on results where the EMs assume median natural mortality is known because there was little difference in results when the EMs estimated this parameter.
+Broadly, we found generally poor accuracy of AIC in selecting models assuming a Beverton-Holt SRR over the null model without an SRR for all OMs.
+However, we also found increased accuracy of AIC in determining the Beverton-Holt SRR when the simulated population exhibited greater variation in spawning biomass for nearly every OM (Figure \ref{sr_aic}).
 
-With R and R+S process error assumptions, probability of AIC accuracy for the B-H stock-recruit relationship as a function of SSB variability were greatest for OMs where there was greater process variability in survival, lower process variability in recruitment and
-contrast in fishing pressure (Figure \ref{sr_aic}, A).
-However, the largest variation in SSB (Figure \ref{sr_aic}, rug on x-axis) occurred in OMs with larger recruitment variability ($\sigma_R = 1.5$; Figure \ref{sr_aic}, A, right column group).
-The weakest overall accuracy of AIC for the Beverton-Holt stock-recruit function was when there was constant fishing pressure (Figure \ref{sr_aic}, A, second and fourth rows).
-The largest effect of simultaneously estimating median natural mortality was when there was OMs had high observation error and constant fishing pressure (Figure \ref{sr_aic}, A, second row).
+With R and R+S process error assumptions, probability of lowest AIC for the B-H SRR as a function of SSB variability were greatest for OMs with contrast in fishing pressure and lower process variability in recruitment (Figure \ref{sr_aic}, A).
+The largest variation in SSB occurred in OMs with larger recruitment variability ($\sigma_R = 1.5$; Figure \ref{sr_aic}, A, right column group), but the same high AIC accuracy was achieved for OMs with lower recruitment variability at lower levels of SSB variation.
+The level of observation error had little effect on AIC accuracy.
 
-For R+M OMs, probability of AIC accuracy for the Beverton-Holt stock-recruit relationship was greatest for OMs where there was contrast in fishing pressure (Figure \ref{sr_aic}, B).
-The largest variation in SSB occurred in OMs with larger natural mortality variability and temporal AR1 correlation of the annual random effects ($\sigma_M = 1.5$, $\rho_M = 0.9$; Figure \ref{sr_aic}, B, right-most column) where AIC accuracy was also high even when there was constant fishing pressure.
-Interestingly, there seemed to be little difference in AIC accuracy whether there was lower or higher observation error or whether the EM simultaneously estimated median natural mortality.
+For R+M OMs, probability of lowest AIC for the Beverton-Holt SRR increased steeply with variation in SSB whether it was induced by contrast in fishing or variation in natural mortality process error.
+(Figure \ref{sr_aic}, B).
+There was little difference in AIC accuracy whether the natural mortality process errors were correleted and, similar to R+S OMs, there was also little effect due to level of observation error.
 
-For R+Sel OMs, contrast in fishing pressure over time is the primary mechanism creating variation in SSB and these are the OMs where AIC accuracy for the Beverton-Holt stock recruit relationship was greatest (Figure \ref{sr_aic}, C, first and third rows).
-There appears to be little effect of variability or correlation of selectivity process errors or whether median natural mortality was estimated on AIC accuracy.
+For R+Sel OMs, contrast in fishing pressure over time was the primary source of variation in SSB and these are the OMs where AIC accuracy for the Beverton-Holt SRR was greatest (Figure \ref{sr_aic}, C).
+There was little effect of variability or correlation of selectivity process errors or the level of observation error on AIC accuracy.
 
-Like the R+Sel OMs, the greatest accuracy for AIC in selecting the Beverton-Holt stock-recruit relationship occurred for R+q OMs where there was contrast in fishing pressure over time which is also where there was the greatest variation in SSB (Figure \ref{sr_aic}, D, first and third rows).
-There were some differences in effects of SSB variation on AIC accuracy when OMs had high observation error and constant fishing pressure.
-Across all of the OMs in Figure \ref{sr_aic}, the only estimated decline in AIC accuracy with SSB variation occurred when R+q OMs had high observation error and constant fishing pressure and greater variation in catchability process errors that also had AR1 autocorrelation.
+Like the R+Sel OMs, the greatest accuracy for AIC in selecting the Beverton-Holt SRR occurred for R+q OMs where there was contrast in fishing pressure over time which is also where there was the greatest variation in SSB (Figure \ref{sr_aic}, D).
+There was also little effect of variability or correlation of catchability process errors or the level of observation error on AIC accuracy.
 
@@
 
-For R+S OMs, the EMs with matching process error assumption generally produced unbaised estimation of terminal SSB except when median natural mortality was estimated and there was high observation error.
+For R+S OMs, the EMs with matching process error assumptions generally produced unbiased estimation of terminal SSB except when median natural mortality was estimated and there was high observation error.
 In R+S OMs with low observation error, EMs with incorrect process error assumptions typically provided biased estimation of terminal year SSB.
-Estimating the Beverton-Holt stock-recruit relationship had little discernible effect on bias of terminal year SSB estimation whereas estimating median M tended to produce more variability in errors in terminal SSB estimation similar to R OMs.
+Estimating the Beverton-Holt SRR had little discernible effect on bias of terminal year SSB estimation whereas estimating median M tended to produce more variability in errors in terminal SSB estimation similar to R OMs.
 
@@
 However, larger variability in natural mortality process errors increased bias of EMs without the correct process error type.
-Estimating median natural mortality increased variability of SSB estimation error particularly for OMs with high observation error and constant fishing pressure over time and increased bias in SSB estimation for many R+M OMs.
-Like R and R+S OMs, estimating a stock-recruit relationship had little discernible effect on SSB bias.
+Estimating median natural mortality increased variability of SSB estimation error particularly for OMs with high observation error and constant fishing pressure over time.
+It also increased bias in SSB estimation for many R+M OMs.
+Like R and R+S OMs, estimating a SRR had little discernible effect on SSB bias.
 
@@
 The largest bias occurred for any EMs that estimated median natural mortality when the OMs had high observation error, constant fishing pressure, and greater variability in selectivity process errors ($\sigma_{\text{Sel}} = 0.5$) or low selectivity process errors ($\sigma_{\text{Sel}} = 0.1$) and low observation error.
-However, there was no evidence of bias of matching R+Sel EMs when observation error was low and variation in selectivity process errors was larger, whether median natural mortality was estimated or not (Figure \ref{SSB_rel_error}, C, third row).
-We only observed an effect of estimating the Beverton-Holt Stock recruit relationship for R+Sel OMs that had high observation error and contrast in fishing pressure where estimating the relationship produced less biased SSB estimation for many EMs (Figure \ref{SSB_rel_error}, C, top row).
+However, there was no evidence of SSB bias for correctly specified R+Sel EMs when observation error was low and variation in selectivity process errors was larger, whether median natural mortality was estimated or not (Figure \ref{SSB_rel_error}, C, third row).
+We only observed an effect of estimating the Beverton-Holt SRR for R+Sel OMs that had high observation error and contrast in fishing pressure where estimating the SRR produced less biased SSB estimation for many EMs (Figure \ref{SSB_rel_error}, C, top row).
 
@@
 Many EMs also performed well in R+q OMs with low observation error, but no contrast in fishing pressure.
-For R+q OMs with high observation error and contrast in fishing pressure, EMs that estimated the Beverton-Holt stock recruit function exhibited less SSB bias than those that did not.
+For R+q OMs with high observation error and contrast in fishing pressure, EMs that estimated the Beverton-Holt SRR exhibited less SSB bias than those that did not.
 Estimating median natural mortality in the EMs only resulted in much more variable SSB estimation errors when there was no contrast in fishing pressure (Figure \ref{SSB_rel_error}, D, first and third rows).
@@
 For all OM process error types, relative errors in terminal year recruitment were generally more variable than SSB, but effects of R and R+S OM and EM attributes on bias (i.e, negative or positive or none) were similar (Figure \ref{R_rel_error}, A).
-Furthermore, for EM configurations where bias in terminal bias in SSB was evident, median relative errors in recruitment often indicated stronger bias in recruitment of the same sign.
+Furthermore, for EM configurations where bias in terminal SSB was evident, median relative errors in recruitment often indicated stronger bias in recruitment of the same sign.
 
@@
 
-In R and R+S OMs, EMs with the correct assumptions about process errors, provided the least biased estimation of Beverton-Holt stock-recruit relationship parameters when there was low observation error, a change in fishing pressure over time, and lower variability of recruitment process errors, and higher variability survival process errors, and there was little effect of estimating natural mortality (Figure \ref{SR_rel_error}, A).
-For other R and R+S OMs, estimating natural mortality often resulted in less biased estimation of stock-recruit parameters.
-There was generally large variability in relative errors of the stock-recruit parameter estimates, but the lowest variability occurred with low variability in recruitment and little or no variability in survival process errors ($\sigma_{2+} \in \{0,0.25\}$), and contrast in fishing pressure.
-Across al R and R+S OMs, relative errors for the $a$ parameter were often less variable than those for $b$.
+Across all OMs, there was generally better accuracy (less bias and/or lower variability) for estimation of the Beverton Holt $a$ parameter than the $b$ parameter.
+In R and R+S OMs, EMs with the correct assumptions about process errors provided the least biased estimation of Beverton-Holt SRR parameters when there was a change in fishing pressure over time and lower variability of recruitment process errors, but there was little effect of estimating median natural mortality and a small increase in bias for those OMs that had high observation error (Figure \ref{SR_rel_error}, A).
+For other R and R+S OMs, estimating natural mortality often resulted in less biased estimation of SRR parameters.
+There was generally large variability in relative errors of the SRR parameter estimates, but the lowest variability occurred with low variability in recruitment and little or no variability in survival process errors ($\sigma_{2+} \in \{0,0.25\}$), and contrast in fishing pressure.
 
-In R+M OMs, the most accurate estimation of stock-recruit parameters for all EM process error assumptions occurred when there was a change in fishing pressure combined with either low variability in natural mortality process errors and high observation error or vice versa (Figure \ref{SR_rel_error}, B).
-Relative to the R, and R+S OMs, there was even less effect of estimating median natural mortality on estimation bias for the stock-recruit relationship parameters, but similarly there was generally large variability of the parameter estimates.
+In R+M OMs, the most accurate estimation of SRR parameters for all EM process error assumptions occurred when there was a change in fishing pressure, greater variability in natural mortality process errors, and lower observation error (Figure \ref{SR_rel_error}, B).
+Relative to the R, and R+S OMs, there was even less effect of estimating median natural mortality on estimation bias for the SRR parameters.
 
-Bias for stock-recruit parameters was very strong and variability in relative errors was greatest for most EMs fit to R+Sel OMs with constant fishing pressure (Figure \ref{SR_rel_error}, C).
-Less bias in parameter estimation occurred for OMs with a change in fishing pressure over time and of those OMs and the most accurate estimation occurred for EMs with process error types that matched the OMs and the OMs had low observation error, contrast in fishing mortality and more variable and uncorrelated selectivity process errors.
-There was little effect of estimating natural mortality on relative errors for stock recruit parameters.
+Bias for SRR parameters was large and variability in relative errors was greatest for most EMs fit to R+Sel OMs with constant fishing pressure (Figure \ref{SR_rel_error}, C).
+Less bias in parameter estimation occurred for OMs with a change in fishing pressure and the best accuracy occurred for those OMs that had low observation error and more variable and uncorrelated selectivity process errors, and when the EMs hadd with the correct process error assumption.
+There was little effect of estimating natural mortality on relative errors for SRR parameters.
 
-Like R+Sel OMs, relative errors in stock-recruit parameters were less variable for most EM process error types when OMs had contrast in fishing pressure and those OMs that also had lower observation error generally had the least variability (Figure \ref{SR_rel_error}, D).
-The worst accuracy of stock-recruit parameter estimation regardless of EM type occured when R+q OMs had low observation error and constant fishing pressure (Figure \ref{SR_rel_error}, D, fourth row).
+Like R+Sel OMs, relative errors in SRR parameters for R+q OMs were more accurate for most EM process error types when OMs had contrast in fishing pressure and lower observation error (Figure \ref{SR_rel_error}, D).
+However, the best accuracy occurred for those OMs that had lower variability in catchability process errors.
+The worst accuracy of SRR parameter estimation regardless of EM type occurred when R+q OMs had low observation error and constant fishing pressure (Figure \ref{SR_rel_error}, D, fourth row).
 
@@
 
-Median natural mortality rate is estimated accurately by all EM process error types for all R OMs except those where observation error is high and fishing pressure is constant where variability in relative errors is high (Figure \ref{M_rel_error}, A, $\sigma_{2+} = 0$).
-For R+S OMs estimation of median natural mortality rate is most accurate when observation error is low and there is contrast in fishing pressure and the EM process error type is correct.
-There is little effect of estimating stock-recruit relationships on the bias for median natural mortality.
+Across all OMs and EMs there was little effect of estimating SRRs on the bias in estimation of median natural mortality (Figure \ref{M_rel_error}).
+Median natural mortality rate was estimated accurately by all EM process error types for all R OMs except those with high observation error and constant fishing pressure, in which case relative errors were high (Figure \ref{M_rel_error}, A, $\sigma_{2+} = 0$).
+For R+S OMs estimation of median natural mortality rate was most accurate when observation error was low and there was contrast in fishing pressure and the EM process error type was correct.
 
@@
 However, those R+M OMs that also had greatest variability in AR1 correlated natural mortality process errors only had unbiased estimation when the EM process error type was correct.
-Like R, and R+S OMs, there was little effect of estimating stock-recruit relationships on the bias for median natural mortality.
 
-All EM process error types produced accurate estimation of median natural mortality rate for R+Sel OMs that had contrast in fishing pressure, low observation error, and low selectivity process error variability (Figure \ref{M_rel_error}, C).
-When selectivity process error variability increased the incorrect EM process errors produce more biased estimation of median natural mortality rate.
-The least accurate estimation occurred for all EM process error types when observation error was high and there was constant fishing pressure.
-Again, like other OM process error types, there was little effect of estimating stock-recruit relationships on the bias for median natural mortality.
+All EM process error types accurately estimated median natural mortality rate for R+Sel OMs that had contrast in fishing pressure, low observation error, and low selectivity process error variability (Figure \ref{M_rel_error}, C).
+When selectivity process error variability increased, the incorrect EM process errors produce more biased estimation of median natural mortality rate.
+The least accurate estimation occurred for all EM process error types when observation error was high and fishing pressure was constant.
 
@@
 Most EM process error types produced biased estimation of median natural  mortality when R+q OMs had high observaiton error and constant fishing pressure.
-Again, there was no discernible effect of estimating stock-recruit relationships on bias.
+
+Like R+Sel OMs, all EM process error types produced accurate estimation of median natural mortality rate when fit to R+q OMs with contrast in fishing pressure, low observation error and low catchability process error variability (Figure \ref{M_rel_error}, D).
+Most EM process error types produced biased estimation of median natural mortality when R+q OMs had high observation error and constant fishing pressure.
 
@@
 
-Mohn's $\rho$ for SSB was small in absolute value for all R and R+S OMs, regardless of EM process error types, and whether median natural mortality rate or stock-recruit relationships were estimated (Figure \ref{mohns_rho_ssb}, A).
-The strongest retrospective patterns occurred in OMs with the largest survival process error variability, high observation error, and contrast in fishing pressure, but only for EMs without the correct process error type and where median natural mortality rate was assumed known.
+Mohn's $\rho$ for SSB was small in absolute value for all R and R+S OMs, regardless of EM process error types, and whether median natural mortality rate or SRRs were estimated (Figure \ref{mohns_rho_ssb}, A).
+The strongest retrospective patterns (highest absolute Mohn’s $\rho$ values) occurred in OMs with the largest apparent survival process error variability, high observation error, and contrast in fishing pressure, but only for EMs with the incorrect process error type and where median natural mortality rate was assumed known (median $\rho$ was approximately -0.15).
 For R+M, R+Sel, and R+q OMs, Mohn's $\rho$ was also small in absolute value, but median values were all closer to 0 than the largest values in the R and R+S OMs (Figure \ref{mohns_rho_ssb},B-D).
-For these OMs, there was no noticeable effect of estimation of median natural mortality rate or stock recruit relationship on Mohn's $\rho$ for any EM process error types.
+For these OMs, there was no noticeable effect of estimation of median natural mortality rate or SRRs on Mohn's $\rho$ for any EM process error types.
 
-Mohn's $\rho$ for recruitment was small in absolute value for all R OMs with low variability in recruitment process errors, regardless of EM process error type, and whether median natural mortality rate or stock-recruit relationships were estimated (Figure \ref{mohns_rho_R}, A).
-However, R and R+S OMs with greater recruitment process variability and higher observation error had median Mohn's $\rho$ for recruitment greater than zero for most EMs even when the the EM process error type was correct.
+Mohn's $\rho$ for recruitment was small in absolute value for all R OMs with low variability in recruitment process errors, regardless of EM process error type, and whether median natural mortality rate or SRRs were estimated (Figure \ref{mohns_rho_R}, A).
+However, R and R+S OMs with greater recruitment process variability and higher observation error had median Mohn's $\rho$ for recruitment greater than zero for most EMs even when the EM process error type was correct.
 In R+S OMs with lower observation error, EMs with the correct process error type exhibited better median Mohn's $\rho$ close to 0 than EMs with the incorrect process error type.
@@
 
-Convergence results can be useful for understanding how lack of convergence in applications to real data might direct the practitioner to which alternative random effects configurations is more appropriate.
-Therefore, the type of convergence that we might use as a diagnostic is important.
-It is common during assessment model fitting to check that the maximum absolute gradient component is less than some threshold, but there is no standard.
-However, if the Hessian were checked for positive definiteness conditional on a satisfactory gradient, some models may be excluded that in fact have an invertible hessian.
-We found the hessian at the optimized log-likelihood can often be invertible when the gradient approach with a sensible criterion (e.g., $10^{-3}$) would indicate lack of convergence.
+Analyses of model convergence across simulations can be useful for understanding the utility of alternative convergence criteria used in applications to real data for directing the practitioner to more appropriate random effects configurations.
+It is common during the assessment model fitting process to check that the maximum absolute gradient component is less than some threshold prior to inspecting the Hessian of the optimized likelihood for invertibility [@carvalhoetal21].
+However, there is no accepted standard for the gradient threshold [e.g., @leeetal11; @hurtadoferroetal15;@ruddthorson18] and some thresholds would exclude models that in fact have an invertible Hessian.
+We found the Hessian at the optimized log-likelihood can often be invertible when the maximum absolute gradient was much larger than what would perceived to be a sensible threshold.
 
-Another factor affecting the convergence criteria is numerical accuracy.
-With the process errors modeled as random effects we are optimizing an (Laplace) approximation of the marginal log-likelihood and therefore hessians and gradients are also with regard to the approximation.
-Furthermore, numerical methods are used to calculate the hessian (Optimhess?), and also to invert it for variance estimation.
-All of these approximations are affecting the gradients and invertibility of the hessian (although likely worse for some models than others), which, along with our results, suggests at least investigating whether the hessian is positive definite when the calculated gradients are not terribly large.
-Often high gradients and/or lack of positive definite hessian result when parameter estimates are at their bounds [@carvalhoetal21], and this also applies to variance or autocorrelation parameters for random effects with state-space assessment models.
+@lietal24 found that convergence rate could be a useful diagnostic especially for separating the correct model from overly complex models.
+However, the criteria for convergence used in their study may also lead to limited ability to distinguish the correct model from overly simplistic models, a pattern that was also noted by @liljestrandetal24 in which one process error may absorb all sources of process error when the magnitude of other process errors are low.
 
-Even when hessian-based variance can be accomplished, parameters that are poorly estimated will have extremely large variances.
+Often poor convergence result when parameter estimates are at their bounds [@carvalhoetal21], and this also applies to variance parameters for random effects with state-space assessment models.
+Even when Hessian is invertible, parameters that are poorly informed will have extremely large variance estimates.
 This further inspection can lead to a more appropriate and often more parsimonious model configuration where the problematic parameters are not estimated.
 For example, process error variance parameters that are estimated close to 0 indicates that the random effects are estimated to have little or no variability and removing these process errors is warranted.
-Generally, our results suggest we can expect lower probability of convergence of state-space assessment models when estimating natural mortality or stock-recruit relationships because of the difficulty distinguishing these parameters from others being estimated in assessment model with data typically available.
+Generally, our results suggest we can expect lower probability of convergence of state-space assessment models when estimating natural mortality or SRRs because of the difficulty distinguishing these parameters from others being estimated in assessment model with data that are typically available.
+Our experiments did not aim to emulate the practitioner decision process in developing model configurations (e.g.
+removing a source of process error and refitting the model when process error variance parameters were estimated close to 0).
+Evaluating the efficacy of such a decision process when applying Ems might be important in closed loop simulations (e.g.
+MSE) aimed at quantifying management performance.
+
+A factor affecting the convergence criteria, particularly for maximum likelihood estimation of models with random effects, is numerical accuracy.
+All optimizations performed in these simulations are of the Laplace approximation of the marginal likelihood and, therefore, gradients and Hessians are also with respect to this approximation (see TMB::sdreport in the Template Model Builder package).
+Functionality within the Template Model Builder package exists (i.e., TMB::checkConsistency) to check the validity of the Laplace approximation and the utility of this as a diagnostic for state-space assessment models should be explored further.
+Furthermore, numerical methods are used to calculate and invert the Hessian for variance estimation for models with random effects.
+Along with our results, the potential lack of accuracy imposed by these approximations, suggests at least investigating whether the Hessian is positive definite when the calculated absolute gradients are not terribly large.
 
@@
 
-Among the process error configurations we used in OMs, we found AIC to be accurate for process errors on recruitment and survival (R and R+S).
-Fitting models to other OMs rarely preferred R+S EMs, and R and R+S EMs were nearly always selected for the matching OMs, a result similar to that reported in [@liljestrandetal24].
+Of the OM process error configurations we considered, we found AIC to be accurate for selecting models with process errors on recruitment and apparent survival (R and R+S).
+Fitting models to other OMs rarely preferred R+S EMs, and R and R+S EMs were nearly always selected for the matching OMs; a similar result was reported by @liljestrandetal24.
 For other sources of process error, accuracy of AIC was improved to useful levels when there was larger variability in the process errors and/or lower observation error.
 
-AIC performed poorly in identifying stock-recruit relationships unless there was contrast in SSB and low variability in recruitment process errors (Fig.
-3), both of which help reduce confounding effects from both process and observation errors.
-OMs with increased  contrast in SSB included higher variability in survival or natural mortality process errors (R+S, R+M OMs), and contrast in fishing pressure (Fig.
+Across all OM process error configurations, AIC performed poorly in identifying that the presence of the Beverton-Holt SRR in the OM unless there was contrast in fishing pressure possibly in combination with other factors such as lower variability in recruitment process errors (in R and R+S models) or greater variation in natural mortality process errors (for R+M OMs, Fig.
 \ref{sr_aic}).
-Process errors in index catchability (R+q OMs) do not affect the contrast in SSB and variability in fishery logistic selectivity parameters (R+Sel OMs) had lesser effect on constrast in SSB.
+As such, properly accounting for process error in natural mortality could be important [@lietal24] when evaluating SRRs in state-space models.
+Curiously, we did not find a marked effect of the level of observation error on ability to detect the SRR, but it is possible that AIC would perform better if observations have even lower uncertainty than we considered.
 
-Increased observation error only weakens AIC performance for OMs where process error configurations allows SR accuracy in low observation error
+Although we did not compare models with alternative SRRs (e.g., Ricker and Beverton-Holt), we do not expect AIC to perform any better distinguishing between relationships.
+Our finding that AIC tended to choose simpler recruitment models in most cases contrasts with the noted bias in AIC for more complex models [@shibata76;@katz81;@kassraftery95], but, whereas those findings apply to more the much more common comparison of models that are fit to raw and independent observations, here we are comparing state-space models which account for observation error and estimate process errors in latent variables.
 
-We found AIC to performs weakly in determining stock recruit relationships unless there is large contrast in SSB and low variation in recruitment process errors.
-Although we did not compare models with alternative stock-recruit relationships (e.g., Ricker and Beverton-Holt), we do not expect AIC to perform any better distinguishing between relationships.
-Our finding that AIC tended to choose simpler recruitment models in most cases contrasts with the noted bias in AIC for more complex models [@shibata76;@katz81;@kassraftery95], but whereas those findings apply to more common fitting of models to raw data, state-space models account for observation error and estimate process errors in latent variables.
-
-Our findings also suggest that process errors in selectivity and catchability have minimal impact on the AIC accuracy for the stock-recruit relationship (Fig.
-3C&D), whereas natural mortality process errors have a much stronger effect (Fig.
-3B).
-This may be because selectivity and catchability do not disproportionately favorare lower for the younger age classes, and vulnerability of older ages follows an increasing logistic curve.
-In contrast, natural mortality influences all age classes equally, leading to greater overall fluctuations in SSB, which increases the probability of detecting the stock-recruit signal even under constant fishing pressure (Fig.
-3B).
-These results highlight the importance of accounting for process error in natural mortality (only when there is evidence of temporal variability, [@li2024]) when evaluating stock-recruitment dynamics in state-space models.
-
-Our results comport well with those of @devalpinehastings02 which found AIC could not distinguish among state-space stock-recruit models that were fit just to SSB and recruitment "observations" and @brittenetalinreview which found AIC could not reliably distinguish alternative environmental effects on stock-recruit parameters.
-@milleretal16 did find AIC to prefer a stock-recruit relationship with environmental effects for SNEMA yellowtail flounder, but there was a large change in stock size estimated over time and flatfish are well-observed by the NEFSC bottom trawl survey used for an index in the assessment model.
+Our results comport with those of @devalpinehastings02 who found AIC could not distinguish among state-space SRRs that were fit just to SSB and recruitment observations (i.e., not an assessment model).
+Similarly,  @brittenetalinreview found AIC could not reliably distinguish alternative environmental effects on SRR parameters.
+However, @milleretal16 did find AIC to prefer a SRR with environmental effects when applied to data for the SNEMA yellowtail flounder stock and  AIC also selected an environmental covariate on a SRR for the most recent stock assessment of Georges Bank yellowtail flounder [@nefsc25].
+Both of these yellowtail flounder stocks have large changes in stock size and the values of environmental covariates over time.
+Additionally, this species is well-observed by the bottom trawl survey that is used for an index in assessment models.
 
@@
 
-As we might have expected, bias in all parameters and assessment output was generally better with lower observation error.
-Estimation of stock-recruit relationship parameters was not reliable in any of the OM-EM combinations, but estimation of median natural mortality was feasible in many OM scenarios with temporal contrast in fishing pressure.
-For OMs where there was bias in natural mortality due to high observation error (R+Sel and R+q OMS), estimating the stock-recruit relationship seemed to alleviate the bias.
-However, estimation of median natural mortality can cause large differences between the true and estimated SSB (that may or may not be unbiased on average) when there is less contrast in fishing pressure over time and higher observation error.
+As expected, bias in all parameters and assessment output was generally improved with lower observation error.
+Estimation of SRR parameters was reliable in ideal scenarios of low observation error and contrast in fishing for some R+Sel and R+M OMs, but generally estimation was biased and(or) highly variable.
+We found substantial bias in estimated SRR parameters in R and R+S OMs particularly with high variability in recruitment and apparent survival process errors, suggesting that practitioners should be cautious when models without the SRR suggest this to be the case.
+
+On the other hand, estimation of median natural mortality was reliable in many OM scenarios with contrast in fishing pressure and in some OMs EMs that also estimated the SRR parameters bias for those parameters was improved.
+Conversely, for some R+Sel and R+q OMs where there was bias in natural mortality due to high observation error, estimating the SRR reduced the bias in median natural mortality rate.
+However, estimating median natural mortality did cause poor accuracy in SSB estimation in many OMs without contrast in fishing pressure over time and with higher observation error.
+Thus, estimating median natural mortality should be approached with caution in state-space assessment models, particularly given its significant impact on determination of reference point and stock status [@lietal24].
 
@@
 
-Incorrect process error assumptions for EMs did not produce strong retrospective patterns for SSB for any OMs whether median natural mortality and a stock-recruit relationship were estimated or not, but some weak retrospective patterns occur when observation error was high and there was contrast in fishing pressure.
-Retrospective patterns tended to be more variable for recruitment and can be large even when the EM is correct.
-Therefore, we recommend emphasis on inspection of retrospective patterns primarily for SSB and F, but further research on retrospective patterns in other assessment model parameters, management quantitities, and projections may be beneficial [@brookslegault16].
+Incorrect EM process error assumptions did not produce strong retrospective patterns for SSB for any OMs regardless of whether median natural mortality or a SRR was estimated, but some weak retrospective patterns occur when observation error was high and there was contrast in fishing pressure.
+However, retrospective patterns tended to be more variable for recruitment and were sometimes large even when the EM was correct.
+Therefore, we recommend emphasis on inspection of retrospective patterns primarily for SSB and $F$, but further research on retrospective patterns in other assessment model parameters, management quantities such as biological reference points, and projections may be beneficial [@brookslegault16].
+
+The general lack of retrospective patterns with mis-specified process errors is perhaps to be expected.
+Retrospective patterns are often induced in simulation studies by rapid changes in a quantity such as index catchability, natural mortality, or perceived catch during years toward the end of the time series [@legault09; @millerlegault17; @huynhetal22; @breiviketal23].
+In our simulations, the process errors changing over time may have trends in particular simulations, particularly when strong autocorrelation is imposed, but the random effects have no trend on average across simulations.
+@szuwalskietal18 and lietal24 also found relatively small retrospective patterns when the source of mis-specification was temporal variation in demography attributes.
+Indeed, it is common for the flexibility provided by temporal random effects to reduce retrospective patterns [@milleretal18; @stockmiller21; @stocketal21], though it does not necessarily indicate a more accurate assessment model [@perrettietal20; @lietal24; @liljestrandetal24].
+Our results together with the existing literature seem to suggest that when a strong retrospective pattern is observed in an assessment it is more likely to be due to a mis-specification of a rapid shift in some model attribute rather than whether a particular process is assumed to be randomly varying temporally.
+
+<!--
+Finally, although the EMs estimated observation variance parameters for the age composition observations, they were provided the correct precision of the aggregate indices and catch which
+
+The R+q OMs only have temporal variation in catchability for one index so the other index is able to provide a consistent signal of trend.
+Incorrect assumptions about the mean selectivity model could also produce
+
+Lack of retrospective patterns for SSB coupled with biased terminal SSB estimates would indicate that the scale across the time series is biased and
+-->
 
@@
 
+Our simulation study examined the importance of several factors for reliable inferences from state-space age-structured assessment models.
+Across all factors, AIC accurately distinguished models with process errors on recruitment only or on recruitment and apparent survival.
+Accuracy for other process error types required a strong signal (high process variability) with low noise (low observation uncertainty).
+Therefore, we expect practitioners will find R+S configurations to provide satisfactory diagnostics across a range of life history and data quality scenarios.
+Contrast in fishing pressure was consistently an important factor across all measures of reliability we examined.
+AIC generally performed poorly for selecting the SRR but performance was improved with low recruitment variability and contrast in fishing pressure.
+However, some bias in estimation in at least one of the SRR parameters existed in nearly all OM-EM combinations.
+Because bias in terminal SSB and retrospective patterns were indifferent to whether or not the SRR was estimated, and convergence was slightly better without the SRR, a sensible default would be to fit models without an assumed SRR.
+
+<!--
 All together what do results mean? E.g., when do you get good convergence and retrospective patterns and high/none bias? When do you get good convergence and no retros and no/big bias?
@@
 When R+S is preferred AMONG THE TYPES CONSIDERED HERE, it probably means its the correct process error .
+-->
 
@@
 We conducted a simulation study where we simulated process errors and observations for 72 operating models with alternative assumptions about fishing history, degree of uncertainty in index and age composition observations, type (recruitment, survival, fishery selectivity, catchability and natural mortality), degree of variation, and correlation of process errors.
-We fit 20 different estimation models to each of 100 simulated set of observations with alternative assumptions type and correlation structure of process errors, median natural mortality was known or estimated, and a B-H stock recruit relationship was assumed or not.
+We fit 20 different estimation models to each of 100 simulated set of observations with alternative assumptions type and correlation structure of process errors, median natural mortality was known or estimated, and a B-H SRR was assumed or not.
 
-Across simulations we summarized probability of convergence of fitted models, accuracy of marginal AIC in determining the correct process error assumption and it ability to determine the Beverton-Holt stock recruit relationship, bias in annual spawning stock biomass, in estimation of natural mortality, and in stock-recruit relationship parameters, and severity of retrospective patterns for estimation models.
+Across simulations we summarized probability of convergence of fitted models, accuracy of marginal AIC in determining the correct process error assumption and it ability to determine the Beverton-Holt SRR, bias in annual spawning stock biomass, in estimation of natural mortality, and in SRR parameters, and severity of retrospective patterns for estimation models.
 
@@
 Invertible hessians and resulting standard error estimation was possible when criteria based on the gradient of the optimized log-likelihood with respect to the fixed effects parameters.
-Using hessian-based convergence, probability of convergence was best for models that assumed the correct source of process error, assumed M was known, and did not assume stock-recruit relationships.
+Using hessian-based convergence, probability of convergence was best for models that assumed the correct source of process error, assumed M was known, and did not assume SRRs.
 
@@
 However, when the true process errors were more variable, AIC accuracy increased to a useful level.
-We found AIC more accurately determined a B-H stock recruit relationship rather than the null model without a S-R relationship when there was low variability in recruitment, low variability in survival random effects, and higher variation in spawning biomass over the time series.
+We found AIC more accurately determined a B-H SRR rather than the null model without a S-R relationship when there was low variability in recruitment, low variability in survival random effects, and higher variation in spawning biomass over the time series.
 When the median natural mortality rate was estimated, we found large bias and uncertainty much more likely in model output such as spawning stock biomass.
@@
 Bias in spawning stock biomass estimation was generally low for estimation models that assumed the correct source of process error when there was lower observation error.
-Reliable estimation of stock-recruit relationship parameters only appears possible in ideal situations with lower observation errors in age composition and indices, lower variability in recruitment process errors and large contrast in spawning biomass over time.
+Reliable estimation of SRR parameters only appears possible in ideal situations with lower observation errors in age composition and indices, lower variability in recruitment process errors and large contrast in spawning biomass over time.
 We found little evidence of bias for many OM process error assumptions when there was contrast in fishing pressure even when there was greater observation error although it can lead to more variable estimates of natural mortality.
-For OMs where there was bias in natural mortality due to high observation error, estimation the stock-recruit relationship seemed to remove the bias.
+For OMs where there was bias in natural mortality due to high observation error, estimation the SRR seemed to remove the bias.
 However, estimation of natural mortality can cause large differences between the true and estimated SSB (that may be unbiased on average) when there is less contrast in fishing pressure over time and higher observation error.
@@
 \end{center}
-\caption{Estimated probability of fits providing hessian-based standard errors for EMs assuming alternative process error (colored points and lines), and median natural mortality (estimated or known) and Beverton-Holt stock recruit functions (estimated or not; along x-axis) when fitted to operating models that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
+\caption{Estimated probability of fits providing hessian-based standard errors for EMs assuming alternative process error (colored points and lines), and median natural mortality (estimated or known) and Beverton-Holt stock-recruit relationships (estimated or not; along x-axis) when fitted to operating models that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
 Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{hessian_SE_convergence}
@@
 \end{center}
-\caption{Estimated probability of lowest AIC for EMs assuming alternative process error structures (colored bars) conditional on alternative assumptions for median natural mortality (estimated or known) and Beverton-Holt stock recruit functions (estimated or not; along x-axis) when fitted to operating models that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
+\caption{Estimated probability of lowest AIC for EMs assuming alternative process error structures (colored bars) conditional on alternative assumptions for median natural mortality (estimated or known) and Beverton-Holt stock-recruit relationships (estimated or not; along x-axis) when fitted to operating models that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
 Striped bars indicate results where the EM process error structure matches that of the operating model.}\label{pe_aic}
@@
 \begin{center}
-\includegraphics{sr_aic_plots}
+\includegraphics[height = 0.8\textheight]{sr_aic_plots_rev}
 \end{center}
-\caption{Estimated probability of lowest AIC from logistic regression on the log-standard deviation of the true log(SSB) in each simulation for estimating model with Beverton-Holt stock recruit functions, rather than the otherwise equivalent EM without the stock recruit function.
-Results are conditional on alternative assumptions for median natural mortality (estimated or known) and on EMs having the correct process error structure: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Rug along x-axis denotes $SD(\log(SSB))$ values for each simulation amd polygons represent 95\% confidence intervals.}\label{sr_aic}
+\caption{Estimated probability of lowest AIC from logistic regression on the log-standard deviation of the true log(SSB) in each simulation for estimating model with Beverton-Holt stock-recruit relationships, rather than the otherwise equivalent EM without the stock-recruit relationship.
+Results are conditional on median M is known in the EM and alternative assumptions EMs having the correct process error structure: R and R+S (A), R+Sel (B), R+M (C), or R+q (D), and median M is assumed known in the EM.
+Solid and dashed lines are for OMs with and without temporal contrast in fishing pressure, respectively, and polygons represent 95\% confidence intervals.
+Range of results indicates the range of log-standard deviation of log(SSB) for simulations of the particular OM.}\label{sr_aic}
 \end{figure}
@@
 \end{center}
-\caption{The proportion mature at age, weight at age, fleet and index selectivity at age, and Beverton-Holt stock recruit relationship assumed for the population in all operating models.
+\caption{The proportion mature at age, weight at age, fleet and index selectivity at age, and Beverton-Holt stock-recruit relationship assumed for the population in all operating models.
 For operating models with random effects on fleet selectivity, this represents the selectivity at the mean of the random effects.}\label{om_inputs_fig}
@@
 \begin{table}
-\caption{Distinguishing characteristics of the operating models with random effects on survival.
+\caption{Distinguishing characteristics of the operating models with random effects on recruitment and apparent survival (R.R+S).
 Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
 Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.}\label{naa_om_table}
-{\footnotesize \input{../paper/naa_om_tab}}
+{\footnotesize \input{naa_om_tab}}
 \end{table}
@@
 \begin{table}
-\caption{Distinguishing characteristics of the operating models with random effects on natural mortality.
+\caption{Distinguishing characteristics of the operating models with random effects on recruitment and natural mortality (R+M).
 Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
@@
 For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}\label{M_om_table}
-{\input{../paper/M_om_tab}}
+{\input{M_om_tab}}
 \end{table}
@@
 \begin{table}
-\caption{Distinguishing characteristics of the operating models with random effects on selectivity.
+\caption{Distinguishing characteristics of the operating models with random effects on recruitment and selectivity (R+Sel).
 Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
@@
 For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}\label{sel_om_table}
-{\input{../paper/Sel_om_tab}}
+{\input{Sel_om_tab}}
 \end{table}
@@
 \begin{table}
-\caption{Distinguishing characteristics of the operating models with random effects on catchability.
+\caption{Distinguishing characteristics of the operating models with random effects on recruitment and catchability (R+q).
 Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
@@
 For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}\label{q_om_table}
-{\input{../paper/q_om_tab}}
+{\input{q_om_tab}}
 \end{table}
@@
 \begin{table}
-\caption{Distinguishing characteristics of the estimating models.}\label{em_table}
-{\scriptsize \input{../paper/em_tab}}
+\caption{Distinguishing characteristics of the estimating models and operating model process error categories (R, R+S, R+M, R+Sel, R+q) where used.}\label{em_table}
+{\scriptsize \input{em_tab}}
 \end{table}
@@
 \caption{The maximum of the absolute values of all gradient values for all fits that provided hessian-based standard errors across all simuated data sets of a given OM configuration (A: R and R+S, B: R+M, C: R+Sel, or D: R+q).
-Results are conditional on EM fits with alternative process error type (colored points and lines), median natural mortality (estimated or known) and recruitment assumptions (Beverton-Holt stock recruit or not).
+Results are conditional on EM fits with alternative process error type (colored points and lines), median natural mortality (estimated or known) and recruitment assumptions (Beverton-Holt stock-recruit relationship or not).
 Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{hess_grad}
@@
 \end{center}
-\caption{Probability of estimating models providing maximum absolute values of gradients less than $10^{-6}$ assuming alternative process error (colored points and lines), and median natural mortality (estimated or known) and Beverton-Holt stock recruit functions (estimated or not; along x-axis) when fitted to operating models that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
+\caption{Probability of estimating models providing maximum absolute values of gradients less than $10^{-6}$ assuming alternative process error (colored points and lines), and median natural mortality (estimated or known) and Beverton-Holt stock-recruit relationships (estimated or not; along x-axis) when fitted to operating models that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
 Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{gradient_convergence}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{sr_aic_plots}
+\end{center}
+\caption{Estimated probability of lowest AIC from logistic regression on the log-standard deviation of the true log(SSB) in each simulation for estimating model with Beverton-Holt stock-recruit relationships, rather than the otherwise equivalent EM without the stock-recruit relationship.
+Results are conditional on alternative assumptions for median natural mortality (estimated or known) and on EMs having the correct process error structure: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Rug along x-axis denotes $SD(\log(SSB))$ values for each simulation and polygons represent 95\% confidence intervals.}\label{sr_aic_supp}
 \end{figure}
@@
 \end{center}
-\caption{Median relative error of Beverton-Holt stock recruitment parameters ($a$ and $b$) for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+\caption{Median relative error of Beverton-Holt stock-recruit parameters ($a$ and $b$) for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
 Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{SR_rel_error}

### 2025-04-08 Project_0/manuscript/manuscript.tex
internal review for project_0 paper

@@
 C. Stock\textsuperscript{6} \and John Wiedenmann\textsuperscript{7}}
-\date{12 March, 2025}
+\date{08 April, 2025}
 
@@
 \(^4\)SMAST\\
-\(^5\)Institute of Marine Research\\
-\(^6\)Mid-Atlantic Fisheries Management Council\\
-\(^7\)Rutgers University\\
+\(^5\)Mid-Atlantic Fishery Management Council, 800 North State Street,
+Suite 201, Dover, DE 19901 USA\\
+\(^6\)Institute of Marine Research, Nye Flødevigveien 20, 4817 His,
+Norway\\
+\(^7\)Department of Ecology, Evolution, and Natural Resources. Rutgers
+University\\
 
 \pagebreak
+
+\textbf{keywords:} stock assessment, state-space, model selection, bias,
+convergence, retrospective patterns
 
@@
 
-State-space assessment models, which are increasingly being used for
-management of commercially important fish stocks, can include process
-errors in many ways, and an evaluation of the statistical reliability of
-the estimates of management output and best practices of selection among
-alternative process error configurations are needed. We simulated 72
-operating models with varying fishing pressure history, observation
-error error, and process error magnitude, correlation, and source
-(recruitment, survival, fishery selectivity, catchability, and natural
-mortality). We fit estimating models with different assumptions on the
-process error source, whether median natural mortality was estimated,
-and whether a stock-recruit relationship was estimated. Estimating
-models that assumed the correct process error source and median natural
-mortality rate was known, and did not estimate a stock recruit
-relationship generally had high probability of convergence and low bias
-in terminal year SSB estimation. Bias was also low under many incorrect
+State-space models are increasingly used for stock assessment, and
+evaluations of their statistical reliability and best practices for
+selecting among process error configurations are needed. We simulated 72
+operating models that varied fishing pressure and observation error
+across process errors in recruitment, survival, selectivity,
+catchability, and/or natural mortality. We fit estimating models with
+different assumptions on the process error source and whether median
+natural mortality or a stock-recruit relationship were estimated.
+Estimating models without a stock-recruit relationship that assumed the
+correct process error source and median natural mortality had high
+convergence rates and low bias. Bias was also low under many incorrect
 process error assumptions when there was contrast in fishing pressure
-and low observation error. Stock-recruitment parameters were not
-reliably estimated in any scenarios, but the least accurate estimates
-occured under ideal situations of low observation error, low recruitment
-variability, and temporal contrast in fishing pressure. However,
-estimation of median natural mortality rate was reliable for many OMs
-(other than those with survival process errors) with temporal contrast
-in fishing pressure regardless of the estimated process error type.
-Marginal AIC most accurately distinguished process errors on
-recruitment, survival, and selectivity, as well as larger magnitude
-process errors of other types. Retrospective patterns for spawning stock
-biomass and fishing mortality were generally weak, but those for
-recruitment were sizable when observation error was high, even with the
-correct process error assumptions.
+and low observation error. Marginal AIC most accurately distinguished
+process errors on recruitment, survival, and selectivity, as well as
+larger magnitude process errors of other types. Retrospective patterns
+were generally small but were sizable for recruitment when observation
+error was high. These results help establish the statistical reliability
+of state space assessment models and pave the way for the
+next-generation of fisheries stock assessment
 
@@
 marginal likelihood or corresponding Bayesian approaches. These latter
-approaches are considered best practice and a requirement for the next
+approaches are considered best practice and a recommended for the next
 generation of stock assessment models (Hoyle et al. 2022; Punt 2023).
@@
 review; Liljestrand et al. 2024). In particular, the effects of
-observation error and underlying fishing history were not fully isolated
-from their study design, making it challenging to disentangle the
-interplay between process and observation error magnitudes, as
-demonstrated in Fisch et al. (2023). Third, explicitly modeling
-stock-recruitment relationships as mechanistic drivers of population
-dynamics is promising (Fleischman et al. 2013; Du Pontavice et al. 2022)
-but can be challenging in state-space models. Evidence suggests that
-when both process and observation errors are unknown, estimating density
-dependence parameters becomes highly uncertain (Knape 2008; Polansky et
-al. 2009). In particular, Knape (2008) demonstrated that stronger
-density dependence becomes increasingly difficult to estimate in the
-presence of observation error. Therefore, it is crucial to assess
-whether density dependence mechanisms can be estimated with sufficient
-precision before drawing ecological conclusions (Auger-Méthé et al.
-2016).
+observation error and underlying fishing history have not been fully
+isolated in simulation study designs, making it challenging to
+disentangle the interplay between process and observation error
+magnitudes, as demonstrated in Fisch et al. (2023). Third, explicitly
+modeling stock-recruit relationships (SRRs) as mechanistic drivers of
+population dynamics is promising (Fleischman et al. 2013; Du Pontavice
+et al. 2022), but can reliability of inferences within integrated
+state-space age-structured models has not been evaluated. Evidence from
+other studies suggests that when both process and observation errors are
+unknown, estimating density dependence parameters becomes highly
+uncertain (Knape 2008; Polansky et al. 2009). In particular, Knape
+(2008) demonstrated that stronger density dependence becomes
+increasingly difficult to estimate in the presence of observation error.
+Therefore, it is crucial to assess whether density dependence mechanisms
+can be estimated with sufficient precision for use in fisheries
+management (Auger-Méthé et al. 2016).
 
@@
 these OMs are fitted with estimation models (EMs) that make alternative
-assumptions for sources of process error, whether a stock-recruit model
-was estimated, and whether a constant, or, in some EMs, median, natural
-mortality is estimated. Given the confounding nature of process errors,
-developing diagnostic tools to detect model misspecification is of great
-scientific interest and could aid the next generation of stock
-assessments (Auger-Méthé et al. 2021). We evaluate whether convergence
-and Akaike Information Criterion (AIC) can correctly determine the
-source of process error and the existence of a stock-recruit
-relationship. We also evaluate when retrospective patterns occur and the
-degree of bias in the outputs of the assessment model that are important
-for management.
+assumptions for sources of process error, whether a SRR was estimated,
+and whether a constant, or, in some EMs, median, natural mortality is
+estimated. Given the confounding nature of process errors, developing
+diagnostic tools to detect model misspecification is of great scientific
+interest and could aid the next generation of stock assessments
+(Auger-Méthé et al. 2021). We evaluate whether convergence and Akaike
+Information Criterion (AIC) can correctly determine the source of
+process error and the existence of a SRR. We also evaluate when
+retrospective patterns occur and the degree of bias in the outputs of
+the assessment model that are important for management.
 
@@
 in our simulation study (Miller and Stock 2020; Stock and Miller 2021).
-WHAM is an R package freely available as a github repository. For this
-study we used
+WHAM is an R package freely available via a github repository and is
+built on the Template Model Builder package (Kristensen et al. 2016).
+For this study we used
 \href{https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3}{version
-1.0.6.9000, commit 77bbd94}. This package has also been used to
-configure OMs and EMs for closed loop simulations evaluating index-based
-assessment methods (Legault et al. 2023) and is used for management of
-haddock, butterfish, American plaice, bluefish, Atlantic cod, black sea
-bass, and yellowtail flounder in the Northeast US.
+1.0.6.9000, commit 77bbd94}. WHAM has also been used to configure OMs
+and EMs for closed loop simulations evaluating index-based assessment
+methods (Legault et al. 2023) and is currently used or accepted for use
+in management of numerous fish stocks in the Northeast United States
+(NEUS) (e.g., NEFSC 2022a, 2022b; NEFSC 2024).
 
 We completed a simulation study with a number of OMs that can be
-categorized based on where process error random effects are assumed:
-abundance at age (R, R+S), natural mortality (R+M), fleet selectivity
-(R+Sel), or index catchability (R+q). For each OM assumption about
+categorized based on where process error random effects were assumed:
+recruitment (R, assumed present in all models), apparent survival
+(denoted R+S), natural mortality (R+M), fleet selectivity (R+Sel), or
+index catchability (R+q). We refer to the (R+S) OMs as modeling apparent
+survival because on logscale the random effects (\(\epsilon_{a,y}\)) are
+additive to the total mortality (F+M) between numbers at age, thus they
+modify the survival term. However, as Stock and Miller (2021) note,
+these random effects can be due to events other than mortality, such as
+immigration, emigration, missreported catch, and other sources of
+misspecification. For each OM, assumptions about the magnitude of the
 variance of process errors and observations are required and the values
-we used were based on a review of the range of estimates from
-applications of WHAM in management of stocks of haddock, butterfish, and
-American plaice in the NE US.
+we used were based on a review of the range of estimates from NEUS
+assessments using WHAM in stocks NEUS.
 
 In total, we configured 72 OMs with alternative assumptions about the
-source and variability of process errors, level of observation error in
-indices and age composition data, and contrast in fishing pressure over
-time. We fitted 20 EMs to observations from each of 100 simulations
-where process errors were also simulated. For R+M, R+Sel, and R+q OMs,
-we used unique seeds for each simulation, but we inadvertently used the
-same 100 seeds for all R and R+S OMs. Each of the EMs made alternative
-assumptions about the source of process errors and whether natural
+source and magnitude of process errors, magnitude of observation error
+in indices and age composition data, and contrast in fishing pressure
+over time. We fitted 20 EMs to observations generated from each of 100
+simulations where process errors were also simulated. Each EM differed
+in assumptions about the source of process errors, whether natural
 mortality (or the median for models with process error in natural
-mortality) was estimated and whether a Beverton-Holt stock recruit
-relationship was estimated within the EM. Details of each of the
-operating and EMs are described below.
+mortality) was estimated, and whether a Beverton-Holt SRR was estimated
+within the EM. Details of each of the OMs and EMs are described below.
 
 We did not use the log-normal bias-correction feature for process errors
-or observations described by (Stock and Miller 2021) for operating and
-EMs (Li et al. In review). Simulations and model fitting were all
-carried out on the University of Massachusetts Green High-Performance
-Computing Cluster. All code we used to perform the simulation study and
-summarize results can be found at
+or observations described by (Stock and Miller 2021) for OMs and EMs to
+simplify interpretation of the study results (Li et al. In review).
+Simulations and model fitting were all carried out on the University of
+Massachusetts Green High-Performance Computing Cluster. All code we used
+to perform the simulation study and summarize results can be found at
 \url{https://github.com/timjmiller/SSRTWG/tree/main/Project_0/code}.
@@
 
-The population consists of 10 age classes: ages 1 to 10+ and we assume
-spawning occurs each year 1/4 of the way through the year. The maturity
+The population consists of 10 age classes, ages 1 to 10+, with the last
+being a plus group that accumulates ages 10 and older. We assume
+spawning occurs annually 1/4 of the way through the year. The maturity
 at age was a logistic curve with \(a_{50}\) = 2.89 and slope = 0.88
@@
 
-We assumed a Beverton-Holt stock recruit function with constant
-pre-recruit mortality parameters for all OMs. All post-recruit
-productivity components are constant in the NAA and survey catchability
-process error OMs. Therefore steepness and unfished recruitment are also
-constant over the time period for those OMs (Miller and Brooks 2021). We
-specified unfished recruitment = \(R_0 = e^{10}\) and
-\(\Fmsy = F_{40\%} = 0.348\) equated to a steepness of 0.69 and
-\(a=0.60\) and \(b = 2.4 \times 10^{-5}\) for the \[
+We assumed a Beverton-Holt SRR with constant pre-recruit mortality
+parameters for all OMs. All post-recruit productivity components are
+constant in the apparent survival (R+S) and survey catchability (R+Sel)
+process error OMs. Therefore, steepness and unfished recruitment are
+also constant over the time period for those OMs (Miller and Brooks
+2021). We specified unfished recruitment equal to \(e^{10}\) and
+\(\Fmsy = F_{40\%} = 0.348\), which equates to a steepness of 0.69 and
+\(a=0.60\) and \(b = 2.4 \times 10^{-5}\) for the Beverton-Holt
+parameterization \[
 N_{1,y} = \frac{a \text{SSB}_{y-1}}{1 + b \text{SSB}_{y-1}} 
-\] Beverton-Holt parameterization (Figure \ref{om_inputs_fig}, bottom
-right). For OMs without process errors on natural mortality we assumed
-the rate was assumed 0.2. For OMs with process errors on natural
-mortality the median natural mortality rate was 0.2.
+\] (Figure \ref{om_inputs_fig}, bottom right). For OMs without process
+errors on natural mortality we fixed the rate at 0.2. For OMs with
+process errors on natural mortality, the median natural mortality rate
+was specified to be 0.2.
 
 We used two fishing scenarios for OMs. In the first scenario, the stock
-experiences overfishing at 2.5\Fmsy for the first 20 years and fishing
-at \Fmsy for the last 20 years (denoted \(2.5\Fmsy \rightarrow \Fmsy\)).
-In the second scenario, the stock is fished at \Fmsy for the entire time
-period. The magnitude of the overfishing assumptions is based on average
-estimates of overfishing for NE groundfish stocks from (Wiedenmann et
-al. 2019). Legault et al. (2023) also used similar approaches to
-defining fishing mortality histories for OMs.
+experiences overfishing at 2.5\Fmsy for the first 20 years followed by
+fishing at \Fmsy for the last 20 years (denoted
+\(2.5\Fmsy \rightarrow \Fmsy\)). In the second scenario, the stock is
+fished at \Fmsy for the entire time period (40 years). The magnitude of
+the overfishing assumptions is based on average estimates of overfishing
+for NEUS groundfish stocks from Wiedenmann et al. (2019) and similar to
+the approach in Legault et al. (2023).
 
 We specified initial population abundance at age at the equilibrium
-distribution fishing at either \(F = 2.5\times \Fmsy\) or \(F = \Fmsy\)
-for the two alternative fishing histories. This implies that, for a
+distribution that corresponds to fishing at either
+\(F = 2.5\times \Fmsy\) or \(F = \Fmsy\). This implies that, for a
 deterministic model, the abundance at age would not change from the
@@
 
-For OMs with time-varying random effects for M, steepness is not
-constant, but we used the same alpha and beta parameters as other OMs
-this equates to a steepness and R0 at the median of the time series
-process for M. For OMs with time-varying random effects for fishery
-selectivity, \Fmsy is also not constant however we use the same F
-history as other OMs which corresponds to Fmsy at the mean selectivity
-parameters.
+For OMs with time-varying random effects for M , steepness is not
+constant. However, we used the same \(a\) and \(b\) parameters as other
+OMs, which equates to a steepness and R0 at the median of the time
+series process for M. For OMs with time-varying random effects for
+fishery selectivity, \Fmsy is also not constant, but since we use the
+same F history as other OMs, this corresponds to \Fmsy at the mean
+selectivity parameters.
 
@@
 We assumed a single fleet operating year round for catch observations
-with logistic selectivity for the fleet with \(a_{50} = 5\) and slope =
-1 (Figure \ref{om_inputs_fig}, bottom left). This selectivity is was
-used to define \Fmsy for the Beverton-Holt stock recruitment parameters
-above. We assumed a logistic-normal distribution with no correlation on
-the multivariate normal scale for the age-composition observations for
-the fleet.
+with logistic selectivity for the fleet (\(a_{50} = 5\) and slope = 1;
+Figure \ref{om_inputs_fig}, bottom left). This selectivity was used to
+define \Fmsy for the Beverton-Holt SRR parameters above. We assumed a
+logistic-normal distribution with no correlation on the multivariate
+normal scale for the age-composition observations for the fleet.
 
@@
 
-Two time series of surveys are assumed and observed in numbers rather
-than biomass for the entire 40 year period with one occurring in the
-spring (0.25 of each year) and one in the fall (0.75 of each year).
-Catchability of both surveys are assumed to be 0.1. Like the fishing
-fleet, we assumed logistic selectivity for both indices with
-\(a_{50} = 5\) and slope = 1 and a logistic-normal distribution with no
-correlation on the multivariate normal scale for the age-composition
-observations.
+Two time series of fishery-independent surveys in numbers are generated
+for the entire 40 year period with one occurring in the spring (0.25 of
+each year) and one in the fall (0.75 of each year). Catchability of both
+surveys are assumed to be 0.1. Like the fishing fleet, we assumed
+logistic selectivity for both indices (\(a_{50} = 5\) and slope = 1) and
+a logistic-normal distribution with no correlation on the multivariate
+normal scale for the age-composition observations.
 
@@
 
-Standard deviation for log-aggregate catch was 0.1. There were two
-levels of observation error variance for indices and age composition for
-both indices and fleet catch. A low uncertainty specification assumed
-standard deviation of both series of log-aggregate index observations
-was 0.1 and the standard deviation of the logistic-normal for age
-composition observations was 0.3 In the high uncertainty specification
-the standard deviation for log-aggregate indices was 0.4 and that for
-the age composition observations was 1.5. For all EMs, standard
-deviation for log-aggregate observations was assumed known whereas that
-for the logistic-normal age composition observations was estimated.
+The standard deviation for log-aggregate catch was 0.1. Two levels of
+observation error variance (high and low) were specified for indices and
+all age composition observations (both indices and catch). The low
+uncertainty specification assumed a standard deviation of 0.1 for both
+series of log-aggregate index observations, and the standard deviation
+of the logistic-normal for age composition observations was 0.3. In the
+high uncertainty specification, the standard deviation for log-aggregate
+indices was 0.4 and that for the age composition observations was 1.5.
+For all EMs, the standard deviation for log-aggregate observations was
+assumed known whereas that for the logistic-normal age composition
+observations was estimated.
 
@@
 
-For operating models with random effects on recruitment and(or) survival
-(R, R+S) we assumed marginal standard deviations for recruitment of
-\(\sigma_R \in \{0.5,1.5\}\) and marginal standard deviations for older
-age classes of \(\sigma_{2+} \in \{0,0.25, 0.5\}\). The full factorial
-combination of these process error assumptions and fishing history (2
-levels) and observation error (2 levels) scenarios described above
-results in 24 different R (\(\sigma_{2+} = 0\)) and R+S operating models
-(Table \ref{naa_om_table}).
+For operating models with random effects on recruitment and(or) apparent
+survival (R, R+S), we assumed marginal standard deviations for
+recruitment of \(\sigma_R \in \{0.5,1.5\}\) and marginal standard
+deviations for older age classes of \(\sigma_{2+} \in \{0,0.25, 0.5\}\).
+The full factorial combination of these process error assumptions (2x3
+levels) and scenarios for fishing history (2 levels) and observation
+error (2 levels) scenarios described above results in 24 different R
+(\(\sigma_{2+} = 0\)) and R+S operating models (Table
+\ref{naa_om_table}).
 
@@
 
-All R+M OMs treat natural mortality constant across age, but with
+All R+M OMs treat natural mortality as constant across age, but with
 annually varying random effects. WHAM treats natural mortality as a
@@
 \] that is a linear combination of a mean log-natural mortality
-parameter that is constant across ages \(\mu_{M} = \log(0.2)\) and any
-annual random effects marginally distributed as
+parameter that is constant across ages (\(\mu_{M} = \log(0.2)\)) and any
+annual random effects are marginally distributed as
 \(\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)\). Uncorrelated
 random effects were also included on recruitment with \(\sigma_R = 0.5\)
-(hence, R+M). The marginal standard deviations we assumed for log
-natural mortality random effects were \(\sigma_M \in \{0.1, 0.5\}\) and
-AR1 autocorrelation parameters of \(\rho_M \in \{0,0.9\}\). The full
-factorial combination of these process error assumptions and fishing
-history (2 levels) and observation error (2 levels) scenarios described
-above results in 16 different R+M OMs (Table \ref{M_om_table}).
+(hence, we denote these OMs as R+M). The marginal standard deviations we
+assumed for log natural mortality random effects were
+\(\sigma_M \in \{0.1, 0.5\}\) and AR1 autocorrelation parameters of
+\(\rho_M \in \{0,0.9\}\). The full factorial combination of these
+process error assumptions and fishing history (2 levels) and observation
+error (2 levels) scenarios described above results in 16 different R+M
+OMs (Table \ref{M_om_table}).
 
@@
 
-MORE SPECIFICS about correlation of random effects? Both selectivity
-pars? just correlated by year? WHAM treats selectivity parameter \(s\)
-as a logit-transformed parameter \[
+WHAM treats selectivity parameter \(s\) as a logit-transformed parameter
+\[
 \log\left(\frac{p_{s,y}-l_{s}}{u_{s}-p_{s,y}}\right) = \mu_s + \epsilon_{s,y}
@@
 random effects marginally distributed as
-\(\epsilon_{s,y} \sim \text{N}\left(0,\sigma_s^2\right)\) where the
+\(\epsilon_{s,y} \sim \text{N}\left(0,\sigma_s^2\right)\), where the
 lower and upper bounds of the parameter (\(l_s\) and \(u_s\)) can be
-specified by the user. All selectivity parameters are either \(a_50\) or
-slope parameters and we assume bounds of 0 and 10 for all selectivity
-parameters for all operating and EMs. The marginal standard deviations
-we assumed for logit scale random effects were
+specified by the user. All selectivity parameters (\(a_{50}\) and slope
+parameters) were bounded by 0 and 10 for all OMs and EMs. The marginal
+standard deviations we assumed for logit scale random effects were
 \(\sigma_s \in \{0.1, 0.5\}\) and AR1 autocorrelation parameters of
 \(\rho_s \in \{0,0.9\}\). Like R+M OMs, the full factorial combination
-of these process error assumptions and fishing history (2 levels) and
-observation error (2 levels) scenarios described above results in 16
-different R+Sel OMs (Table \ref{sel_om_table}).
+of these process error assumptions (2x2 levels) and scenarios described
+above for fishing history (2 levels) and observation error (2 levels)
+results in 16 different R+Sel OMs (Table \ref{sel_om_table}).
 
@@
 lower and upper bounds of the catchability (\(l_i\) and \(u_i\)) can be
-specified by the user. Here we assume bounds of 0 and 1000 for all
-operating and EMs. For operating and EMs with process errors on
-catchability, the temporal variation is only assumed for the first
-index. The marginal standard deviations we assumed for logit scale
-random effects were \(\sigma_i \in \{0.1, 0.5\}\) and AR1
-autocorrelation parameters of \(\rho_i \in \{0,0.9\}\). Like R+M and
+specified by the user. We assumed bounds of 0 and 1000 for all OMs and
+EMs. For all OMs and EMs with process errors on catchability, the
+temporal variation only applies to the first index, which could be
+interpreted as capturing some unmeasured seasonal process that affects
+availability to the survey. The marginal standard deviations we assumed
+for logit scale random effects were \(\sigma_i \in \{0.1, 0.5\}\) and
+AR1 autocorrelation parameters of \(\rho_i \in \{0,0.9\}\). Like R+M and
 R+Sel OMs, the full factorial combination of these process error
@@
 
-For each data set simulated from an OM 20 EMs were fit. A total of 32
-different EMs were fit across OMs where the subset of 20 depended on the
-source of process error in the OM (Table \ref{em_table}). The EMs make
-different assumptions about the source of process error (R+S, R+M,
-R+Sel, R+q) and whether there is temporal autocorrelation or not, and
-whether a Beverton-Holt stock recruit function is estimated and the
-natural mortality rate (\(\mu_M\), the constant or mean on log scale for
-R+M EMs) is estimated. For simplicity we refer to the derived estimate
-\(e^{\mu_M}\) as the median natural mortality rate regardless of whether
-natural mortality random effects are estimated in the EM. The first 20
-EMs in Table \ref{em_table} were fit to simulate data sets from R and
-R+S OMs. EMs 5 to 24 in Table \ref{em_table} were fit to simulate data
-sets from R+M OMs. EMs 5 to 20 and 25-28 in Table \ref{em_table} were
-fit to simulate data sets from R+Sel OMs. Finally, EMs 5 to 20 and 29-32
-in Table \ref{em_table} were fit to simulate data sets from R+q OMs. The
-maturity at age, weight at age for catch and SSB, and observation error
-variance of aggregate catch and indices were all assumed known at the
-true values.
+For each of the data sets simulated from an OM, 20 EMs were fit. A total
+of 32 different EMs were fit across OMs where the subset of 20 depended
+on the source of process error in the OM (Table \ref{em_table}). The EMs
+have different assumptions about the source of process error (R+S, R+M,
+R+Sel, R+q) and whether or not 1) there is temporal autocorrelation, 2)
+a Beverton-Holt SRR is estimated, and 3) the natural mortality rate
+(\(\mu_M\), the constant or mean on log scale for R+M EMs) is estimated.
+For simplicity we refer to the derived estimate \(e^{\mu_M}\) as the
+median natural mortality rate regardless of whether natural mortality
+random effects are estimated in the EM.
+
+Subsets of 20 EMs in Table \ref{em_table} were fit to simulate data sets
+from each of the OM process error categories. For R and R+S OMs, fitted
+EMs had matching process error assumptions as well as R+Sel, R+M, and
+R+q assumptions without autocorrelation. Similarly, For other OM process
+error categories, we fit EMs with matching process error assumptions as
+well as other process error types without autocorrelation. The maturity
+at age, weight at age for catch and SSB, and observation error variance
+of aggregate catch and indices were all assumed known at the true
+values. However, the variance parameters for the logistic-normal
+distributions for age composition observations were estimated in the
+EMs.
 
@@
 
+\hypertarget{convergence}{%
+\subsubsection{Convergence}\label{convergence}}
+
 The first measure of reliability we investigated was frequency of
@@
 various ways to assess convergence of the fit (e.g., Carvalho et al.
-2021), but given the importance of estimates of uncertainty when using
-assessment models in management, we estimated probablity of convergence
-as measured by occurrence of a positive-definite hessian matrix at the
-optimized negative log-likelihood that could be inverted. However, we
-also provided results in the Supplementary Materials for the maximum of
-the absolute values among all gradients for all fits of a given EM to
-all simulated data sets from a given OM that produced hessian-based
-standard errors for all estimated fixed effects. This provides an
-indication of how poor the calculated gradients can be, but still
-presumably converged adequately enough for parameter inferences.
+2021; Kapur et al. 2025), but given the importance of estimates of
+uncertainty when using assessment models in management, we estimated
+probablity of convergence as measured by occurrence of a
+positive-definite hessian matrix at the optimized negative
+log-likelihood that could be inverted. We also provide results in the
+Supplementary Materials for the maximum of the absolute values among all
+gradients for all fits of a given EM to all simulated data sets from a
+given OM that produced hessian-based standard errors for all estimated
+fixed effects. This provides an indication of how poor the calculated
+gradients can be, but still presumably converged adequately enough for
+parameter inferences. We used the Clopper-Pearson exact method for
+constructing 95\% confidence intervals of the probabilities of
+convergence (Clopper and Pearson 1934; Thulin 2014).
 
@@
 operating model, we compared AIC for EMs that all made the same
-assumptions about median natural mortality (known or estimated) and
-stock recruitment model (Beverton-Holt or none).
-
-We also estimated the probability of correctly selecting between models
-with Beverton-Holt stock recruit function assumed and models without the
-stock-recruit function (null model). We made these comparisons between
-models that otherwise assume the same process error structure as the
-operating model and both of the compared models either estimate median
-natural mortality or assume it is known. Contrast in fishing pressure
-and time series with recruitment at low stock size has been shown to
-improve estimability of stock-recruitment parameters (Magnusson and
-Hilborn 2007; Conn et al. 2010). Our preliminary inspections of the
-proportions of simulations where the correct recruitment model was
-chosen for a given set of OM factors (including contrast in fishing
-pressure) indicated generally poor performance of AIC. Therefore, we fit
-logistic regression models to the indicator of Beverton-Holt models
-having lower AIC as a function of the log-standard deviation of the true
-log(SSB) (similar to the log of the coefficient of variation for SSB)
-since simulations with realized SSB producing low and high recruitments
-would have larger variation in realized SSB.
-
-All results only condition on whether all of the compared estimating
-models completed the optimization process without failure. We did not
-condition on convergence as defined by a gradient threshold or
-invertibity of the hessian because optimization can correctly determine
-the the correct likelihood that would indicate poor convergence because
-variance parameters may be at the lower bound of zero correctly for
-models that assume the incorrect process error structure.
+assumptions about median natural mortality (known or estimated) and SRR
+(Beverton-Holt or none).
+
+We also estimated the probability of correctly selecting between EMs
+with Beverton-Holt SRR assumed and models without the SRR (null model).
+We made these comparisons between models that otherwise assumed the same
+process error structure as the operating model and both of the compared
+models either estimate median natural mortality or assume it is known.
+Contrast in fishing pressure and time series with recruitment at low
+stock size has been shown to improve estimation of SRR parameters
+(Magnusson and Hilborn 2007; Conn et al. 2010). Our preliminary
+inspections of the proportions of simulations where the correct
+recruitment model was chosen for a given set of OM factors (including
+contrast in fishing pressure) indicated generally poor performance of
+AIC. Therefore, we fit logistic regression models to the indicator of
+Beverton-Holt models having lower AIC as a function of the log-standard
+deviation of the true log(SSB) (similar to the log of the coefficient of
+variation for SSB) since simulations with realized SSB producing low and
+high recruitments would have larger variation in realized SSB.
+
+All model selection results condition on whether all of the compared
+estimating models completed the optimization process without failure. We
+did not condition on convergence as defined by a gradient threshold or
+invertibity of the hessian because optimization could correctly
+determine an inappropriate process error assumption by estimating
+variance parameters at the lower bound of zero. Such an optimization
+could indicate poor convergence but the likelihood would be equivalent
+to that without the mis-specified random effects and the AIC would be
+appropriately higher because more (variance) parameters were estimated.
+All other measures of reliability described below (bias and Mohn's
+\(\rho\)) use these same criteria for inclusion of EM fits in the
+summarized results.
 
@@
 in Miller and Hyun (2018) and Stock and Miller (2021). We present
-results for terminal years SSB and recruitment, Beverton-Holt stock
-recruit parameters (\(a\) and \(b\)), and median natural mortality rate.
-Results for terminal year fishing mortality are strongly negatively
-correlated with those for SSB and provided in the supplementary
-materials.
-
-Similar to the AIC results, bias results only condition on whether the
-estimating model completed the optimization process without failure. We
-did not condition on convergence as defined by a gradient threshold or
-invertibity of the hessian because the optimized model can provide
-reliable estimation of SSB, F, M, and stock recruit parameters whether
-or not the model was able to estimate non-zero random effects. In
-practice, the model would be reconfigured to remove unnecessary process
-errors and produce otherwise equivalent parameter estimates.
+results for bias in terminal year estimates of SSB and recruitment,
+Beverton-Holt stock recruit parameters (\(a\) and \(b\)), and median
+natural mortality rate. Results for terminal year fishing mortality are
+were strongly negatively correlated with those for SSB and are provided
+in the Supplementary Materials.
 
@@
 
-We estimated Mohn's \(\rho\) for SSB, fully-selected fishing mortality,
-and recruitment for each EM (Mohn 1999). We estimated 7 peels for each
-EM. We calculated median 95\% confidence intervals for Mohn's \(\rho\)
-using the same methods as that for relative bias. Similar to the other
-results, retrospective results only condition on whether all of the
-peels of a given estimating model completed the optimization process
-without failure. We did not condition on convergence as defined by a
-gradient threshold or invertibity of the hessian because the optimized
-model can provide reliable estimation of SSB, fishing mortality rates,
-and recruitment whether or not the model was able to estimate non-zero
-random effects.
+We calculated Mohn's \(\rho\) for SSB, fully-selected fishing mortality,
+and recruitment for each EM (Mohn 1999). We fit 7 peels for each EM and
+calculated median 95\% confidence intervals for Mohn's \(\rho\) using
+the same methods as that for relative bias.
 
@@
 
-For R and R+S OMs, convergence generally declines for most EMs when the
-median natural mortality rate is estimated and/or the Beverton-Holt
-stock recruit relationship is estimated even when the process error
-assumptions of the EMs and OMs match (Figure
-\ref{hessian_SE_convergence}, A). When there was high observation error
-and constant fishing pressure (change from \(2.5F_{\text{MSY}}\) to
-\(F_{\text{MSY}}\)), convergence of all EM process error configurations
-other than R EMs was poor for R OMs (\(\sigma_{2+} = 0\)) regardless of
-whether median natural mortality and stock-recruit relationships were
-estimated. Convergence of R EMs was high for all R and R+S OMs except
-when there was high observation error and constant fishing pressure and
-median natural mortality and stock-recruit relationships were estimated.
-R+S EMs fit to R OMs exhibited poor convergence regardless of whether
-natural mortality or a stock-recruit relationship was estimated. R+S EMs
-fit to R+S OMs had highest convergence rates when there was contrast in
-fishing pressure and low observation error. Convergence rates were high
-for all EMs when fit to data from R+S OMs with lower observation error
-except those where median natural mortality and/or stock-recruit
-relationships were estimated.
-
-Convergence of all EMs fitted to R+M OMs was best when the OMs had
+For R and R+S OMs, convergence generally declined for most EMs when the
+median natural mortality rate was estimated and/or the Beverton-Holt SRR
+was estimated even when the process error assumptions of the EMs and OMs
+matched (Figure \ref{hessian_SE_convergence}, A). When there was high
+observation error and constant fishing pressure (\(F = F_{\text{MSY}}\)
+for all 40 years), convergence was poor for of all EM process error
+configurations other than R EMs when fitted to R OMs
+(\(\sigma_{2+} = 0\)) regardless of whether median natural mortality and
+SRRs were estimated. Convergence of R EMs was high for all R and R+S OMs
+except when there was high observation error and constant fishing
+pressure, and when median natural mortality and SRRs were estimated. R+S
+EMs fit to R OMs exhibited poor convergence regardless of whether
+natural mortality or a SRR was estimated. R+S EMs fit to R+S OMs had
+highest convergence rates when there was contrast in fishing pressure
+and low observation error. Convergence rates were high for all EMs when
+fit to data from R+S OMs with lower observation error except those where
+median natural mortality and/or SRRs were estimated.
+
+Convergence of all EMs fitted to R+M OMs was highest when the OMs had
 higher natural mortality process error variability, low observation
@@
 of process errors had poor convergence for R+M OMs when there was low
-natural mortality process error variability whether or not there was
+natural mortality process error variability regardless of
 autocorrelation of the simulated process errors. R+S EMs fitted to data
-generated from R+M OMs always converged poorly whether median natural
-mortality and the Beverton-Holt stock-recruit relationship were
-estimated.
-
-R+S EMs, in particular, converged poorly when fit to data generated from
-R+Sel OMs with lower selectivity process error variability or higher
-observation error (Figure \ref{hessian_SE_convergence}, C). R+Sel EMs
-generally converged better than other EMs for R+Sel OMs with higher
-process error variability, lower observation error, and contrast in
-fishing pressure regardless of whether median natural mortality or a
-stock recruit relationship was estimated.
-
-For R+q OMs, convergence of R+q EMs is generally better than that of
-other EMs when there is contrast in fishing history (Figure
-(\ref{hessian_SE_convergence}, D). Convergence of R+S EMs is generally
+generated from R+M OMs always converged poorly whether or not median
+natural mortality and the Beverton-Holt SRR were estimated.
+
+The R+S EMs, in particular, had poor convergence when fit to data
+generated from R+Sel OMs with lower selectivity process error
+variability or higher observation error (Figure
+\ref{hessian_SE_convergence}, C). R+Sel EMs generally converged better
+than other EMs for R+Sel OMs with higher process error variability,
+lower observation error, and contrast in fishing pressure regardless of
+whether median natural mortality or a SRR was estimated.
+
+For R+q OMs, convergence of R+q EMs was generally better than that of
+other EMs when there was contrast in fishing pressure (Figure
+(\ref{hessian_SE_convergence}, D). Convergence of R+S EMs was generally
 worse than that of all other EMs across all OMs whether or not median
-natural mortality or a stock recruit relationship is estimated. Again,
-convergence probability generally declines for all EMs when median
-natural mortality or a stock recruit relationship is estimated.
+natural mortality or a SRR was estimated. Again, convergence probability
+generally declined for all EMs when median natural mortality or a SRR
+was estimated.
 
 We found a wide range of maximum absolute values of gradients for models
-that converged as defined here (Figure \ref{hess_grad}). The largest
-value observed for a given EM and OM combination was typically
-\(<10^{-3}\), but many converged models had values greater than 1. For
-many OMs, EMs that assumed the correct process error type and did not
-estimate median natural mortality or the Beverton-Holt stock-recruit
-relationship produced the lowest gradient values.
+that converged (Figure \ref{hess_grad}). The largest value observed for
+a given EM and OM combination was typically \(<10^{-3}\), but many
+converged models had values greater than 1. For many OMs, EMs that
+assumed the correct process error type and did not estimate median
+natural mortality or the Beverton-Holt SRR produced the lowest gradient
+values.
 
@@
 
-Marginal AIC accurately determines the correct process error assumptions
-in EMs when data are generated from R and R+S OMs, regardless of whether
-median natural mortality or a stock recruit relationship is estimated
-(Figure \ref{pe_aic}, A). Adding estimation of median natural mortality
-or a stock recruit relationship separately has a negligible effect on
-the accuracy of determining the correct process error assumption. When
-both are estimated, there is a noticeable reduction in accuracy when OMs
-have a constant fishing history, observation error is low and larger
-variability in recruitment process errors.
+Marginal AIC accurately determined the correct process error assumptions
+in EMs when data were generated from R and R+S OMs, regardless of
+whether median natural mortality or a SRR was estimated (Figure
+\ref{pe_aic}, A). Attempting to estimate median natural mortality or a
+SRR separately had a negligible effect on the accuracy of determining
+the correct process error assumption. When both were estimated, there
+was a noticeable reduction in accuracy when OMs had a constant fishing
+pressure, low observation error, and larger variability in recruitment
+process errors.
 
@@
 variability in natural mortality process errors was high (Figures
-\ref{pe_aic}, B). Estimating the median natural mortality rate reduced
-the accuracy of AIC for OMs that assumed natural mortality process
-errors were independent. For OMs with poor accuracy, AIC most frequently
-selected EMs with process errors in catchability (R+q) or selectivity
-(R+Sel). Selection of R+S EMs was typically unlikely.
+\ref{pe_aic}, B). Of these OMs, estimating the median natural mortality
+rate only reduced the accuracy of AIC when natural mortality process
+errors were independent and fishing pressure was constant. For OMs with
+poor model selection accuracy, AIC most frequently selected EMs with
+process errors in catchability (R+q) or selectivity (R+Sel). Selection
+of R+S EMs was generally unlikely.
 
@@
 process errors and high observation error, R+q or R+S EMs were more
-likely to have the best AIC. Whether median natural morality or stock
-recruit relationships were estimated appeared to have little effect on
-the performance of AIC.
+likely to have the best AIC. Whether median natural morality or SRRs
+were estimated appeared to have little effect on the performance of AIC.
 
@@
 error and correlation structure for R+q OMs with high variability in
-catchability process errors (Figures \ref{pe_aic},D). The worst accuracy
-occurred for OMs with low variability in catchability process errors and
-high observation error. However, in these OMs, the marginal AIC
-accurately determined the correct source of process error (but not
-correlation structure) except when EMs estimated both median natural
-morality and the stock recruit relationships and OMs assumed a constant
-fishing pressure.
+catchability process errors (Figures \ref{pe_aic},D). The R+q OMs with
+low variability in catchability process errors and high observation
+error had the least model selection accuracy. However, for these OMs,
+the marginal AIC accurately determined the correct source of process
+error (but not correlation structure) except when OMs assumed a constant
+fishing pressure and EMs estimated both median natural morality and the
+SRR.
 
@@
 
-Our comparisons of model performance condition on assuming the true
+Our comparisons of model performance conditioned on assuming the true
 process error configuration is known (EM and OM process error types
-match). Broadly, we found generally poor accuracy of AIC in selecting
-models assuming a Beverton-Holt stock recruit function over the null
-model without an assumed stock-recruit relationship for all OMs.
-However, we also found increased accuracy of AIC in determining the
-Beverton-Holt stock-recruit relationship with greater variation in
-spawning biomass generated in the simulated populations for nearly every
-OM (Figure \ref{sr_aic}).
-
-With R and R+S process error assumptions, probability of AIC accuracy
-for the B-H stock-recruit relationship as a function of SSB variability
-were greatest for OMs where there was greater process variability in
-survival, lower process variability in recruitment and contrast in
-fishing pressure (Figure \ref{sr_aic}, A). However, the largest
-variation in SSB (Figure \ref{sr_aic}, rug on x-axis) occurred in OMs
-with larger recruitment variability (\(\sigma_R = 1.5\); Figure
-\ref{sr_aic}, A, right column group). The weakest overall accuracy of
-AIC for the Beverton-Holt stock-recruit function was when there was
-constant fishing pressure (Figure \ref{sr_aic}, A, second and fourth
-rows). The largest effect of simultaneously estimating median natural
-mortality was when there was OMs had high observation error and constant
-fishing pressure (Figure \ref{sr_aic}, A, second row).
-
-For R+M OMs, probability of AIC accuracy for the Beverton-Holt
-stock-recruit relationship was greatest for OMs where there was contrast
-in fishing pressure (Figure \ref{sr_aic}, B). The largest variation in
-SSB occurred in OMs with larger natural mortality variability and
-temporal AR1 correlation of the annual random effects
-(\(\sigma_M = 1.5\), \(\rho_M = 0.9\); Figure \ref{sr_aic}, B,
-right-most column) where AIC accuracy was also high even when there was
-constant fishing pressure. Interestingly, there seemed to be little
-difference in AIC accuracy whether there was lower or higher observation
-error or whether the EM simultaneously estimated median natural
-mortality.
-
-For R+Sel OMs, contrast in fishing pressure over time is the primary
-mechanism creating variation in SSB and these are the OMs where AIC
-accuracy for the Beverton-Holt stock recruit relationship was greatest
-(Figure \ref{sr_aic}, C, first and third rows). There appears to be
+match) and we focus on results where the EMs assume median natural
+mortality is known because there was little difference in results when
+the EMs estimated this parameter. Broadly, we found generally poor
+accuracy of AIC in selecting models assuming a Beverton-Holt SRR over
+the null model without an SRR for all OMs. However, we also found
+increased accuracy of AIC in determining the Beverton-Holt SRR when the
+simulated population exhibited greater variation in spawning biomass for
+nearly every OM (Figure \ref{sr_aic}).
+
+With R and R+S process error assumptions, probability of lowest AIC for
+the B-H SRR as a function of SSB variability were greatest for OMs with
+contrast in fishing pressure and lower process variability in
+recruitment (Figure \ref{sr_aic}, A). The largest variation in SSB
+occurred in OMs with larger recruitment variability (\(\sigma_R = 1.5\);
+Figure \ref{sr_aic}, A, right column group), but the same high AIC
+accuracy was achieved for OMs with lower recruitment variability at
+lower levels of SSB variation. The level of observation error had little
+effect on AIC accuracy.
+
+For R+M OMs, probability of lowest AIC for the Beverton-Holt SRR
+increased steeply with variation in SSB whether it was induced by
+contrast in fishing or variation in natural mortality process error.
+(Figure \ref{sr_aic}, B). There was little difference in AIC accuracy
+whether the natural mortality process errors were correleted and,
+similar to R+S OMs, there was also little effect due to level of
+observation error.
+
+For R+Sel OMs, contrast in fishing pressure over time was the primary
+source of variation in SSB and these are the OMs where AIC accuracy for
+the Beverton-Holt SRR was greatest (Figure \ref{sr_aic}, C). There was
 little effect of variability or correlation of selectivity process
-errors or whether median natural mortality was estimated on AIC
-accuracy.
+errors or the level of observation error on AIC accuracy.
 
 Like the R+Sel OMs, the greatest accuracy for AIC in selecting the
-Beverton-Holt stock-recruit relationship occurred for R+q OMs where
-there was contrast in fishing pressure over time which is also where
-there was the greatest variation in SSB (Figure \ref{sr_aic}, D, first
-and third rows). There were some differences in effects of SSB variation
-on AIC accuracy when OMs had high observation error and constant fishing
-pressure. Across all of the OMs in Figure \ref{sr_aic}, the only
-estimated decline in AIC accuracy with SSB variation occurred when R+q
-OMs had high observation error and constant fishing pressure and greater
-variation in catchability process errors that also had AR1
-autocorrelation.
+Beverton-Holt SRR occurred for R+q OMs where there was contrast in
+fishing pressure over time which is also where there was the greatest
+variation in SSB (Figure \ref{sr_aic}, D). There was also little effect
+of variability or correlation of catchability process errors or the
+level of observation error on AIC accuracy.
 
@@
 
-For R+S OMs, the EMs with matching process error assumption generally
-produced unbaised estimation of terminal SSB except when median natural
+For R+S OMs, the EMs with matching process error assumptions generally
+produced unbiased estimation of terminal SSB except when median natural
 mortality was estimated and there was high observation error. In R+S OMs
@@
 typically provided biased estimation of terminal year SSB. Estimating
-the Beverton-Holt stock-recruit relationship had little discernible
-effect on bias of terminal year SSB estimation whereas estimating median
-M tended to produce more variability in errors in terminal SSB
-estimation similar to R OMs.
+the Beverton-Holt SRR had little discernible effect on bias of terminal
+year SSB estimation whereas estimating median M tended to produce more
+variability in errors in terminal SSB estimation similar to R OMs.
 
@@
 mortality increased variability of SSB estimation error particularly for
-OMs with high observation error and constant fishing pressure over time
-and increased bias in SSB estimation for many R+M OMs. Like R and R+S
-OMs, estimating a stock-recruit relationship had little discernible
-effect on SSB bias.
+OMs with high observation error and constant fishing pressure over time.
+It also increased bias in SSB estimation for many R+M OMs. Like R and
+R+S OMs, estimating a SRR had little discernible effect on SSB bias.
 
@@
 (\(\sigma_{\text{Sel}} = 0.1\)) and low observation error. However,
-there was no evidence of bias of matching R+Sel EMs when observation
-error was low and variation in selectivity process errors was larger,
-whether median natural mortality was estimated or not (Figure
-\ref{SSB_rel_error}, C, third row). We only observed an effect of
-estimating the Beverton-Holt Stock recruit relationship for R+Sel OMs
-that had high observation error and contrast in fishing pressure where
-estimating the relationship produced less biased SSB estimation for many
-EMs (Figure \ref{SSB_rel_error}, C, top row).
+there was no evidence of SSB bias for correctly specified R+Sel EMs when
+observation error was low and variation in selectivity process errors
+was larger, whether median natural mortality was estimated or not
+(Figure \ref{SSB_rel_error}, C, third row). We only observed an effect
+of estimating the Beverton-Holt SRR for R+Sel OMs that had high
+observation error and contrast in fishing pressure where estimating the
+SRR produced less biased SSB estimation for many EMs (Figure
+\ref{SSB_rel_error}, C, top row).
 
@@
 and contrast in fishing pressure, EMs that estimated the Beverton-Holt
-stock recruit function exhibited less SSB bias than those that did not.
-Estimating median natural mortality in the EMs only resulted in much
-more variable SSB estimation errors when there was no contrast in
-fishing pressure (Figure \ref{SSB_rel_error}, D, first and third rows).
+SRR exhibited less SSB bias than those that did not. Estimating median
+natural mortality in the EMs only resulted in much more variable SSB
+estimation errors when there was no contrast in fishing pressure (Figure
+\ref{SSB_rel_error}, D, first and third rows).
 
@@
 were similar (Figure \ref{R_rel_error}, A). Furthermore, for EM
-configurations where bias in terminal bias in SSB was evident, median
-relative errors in recruitment often indicated stronger bias in
-recruitment of the same sign.
+configurations where bias in terminal SSB was evident, median relative
+errors in recruitment often indicated stronger bias in recruitment of
+the same sign.
 
@@
 
-In R and R+S OMs, EMs with the correct assumptions about process errors,
-provided the least biased estimation of Beverton-Holt stock-recruit
-relationship parameters when there was low observation error, a change
-in fishing pressure over time, and lower variability of recruitment
-process errors, and higher variability survival process errors, and
-there was little effect of estimating natural mortality (Figure
+Across all OMs, there was generally better accuracy (less bias and/or
+lower variability) for estimation of the Beverton Holt \(a\) parameter
+than the \(b\) parameter. In R and R+S OMs, EMs with the correct
+assumptions about process errors provided the least biased estimation of
+Beverton-Holt SRR parameters when there was a change in fishing pressure
+over time and lower variability of recruitment process errors, but there
+was little effect of estimating median natural mortality and a small
+increase in bias for those OMs that had high observation error (Figure
 \ref{SR_rel_error}, A). For other R and R+S OMs, estimating natural
-mortality often resulted in less biased estimation of stock-recruit
-parameters. There was generally large variability in relative errors of
-the stock-recruit parameter estimates, but the lowest variability
-occurred with low variability in recruitment and little or no
-variability in survival process errors (\(\sigma_{2+} \in \{0,0.25\}\)),
-and contrast in fishing pressure. Across al R and R+S OMs, relative
-errors for the \(a\) parameter were often less variable than those for
-\(b\).
-
-In R+M OMs, the most accurate estimation of stock-recruit parameters for
-all EM process error assumptions occurred when there was a change in
-fishing pressure combined with either low variability in natural
-mortality process errors and high observation error or vice versa
-(Figure \ref{SR_rel_error}, B). Relative to the R, and R+S OMs, there
-was even less effect of estimating median natural mortality on
-estimation bias for the stock-recruit relationship parameters, but
-similarly there was generally large variability of the parameter
-estimates.
-
-Bias for stock-recruit parameters was very strong and variability in
-relative errors was greatest for most EMs fit to R+Sel OMs with constant
-fishing pressure (Figure \ref{SR_rel_error}, C). Less bias in parameter
-estimation occurred for OMs with a change in fishing pressure over time
-and of those OMs and the most accurate estimation occurred for EMs with
-process error types that matched the OMs and the OMs had low observation
-error, contrast in fishing mortality and more variable and uncorrelated
-selectivity process errors. There was little effect of estimating
-natural mortality on relative errors for stock recruit parameters.
-
-Like R+Sel OMs, relative errors in stock-recruit parameters were less
-variable for most EM process error types when OMs had contrast in
-fishing pressure and those OMs that also had lower observation error
-generally had the least variability (Figure \ref{SR_rel_error}, D). The
-worst accuracy of stock-recruit parameter estimation regardless of EM
-type occured when R+q OMs had low observation error and constant fishing
-pressure (Figure \ref{SR_rel_error}, D, fourth row).
+mortality often resulted in less biased estimation of SRR parameters.
+There was generally large variability in relative errors of the SRR
+parameter estimates, but the lowest variability occurred with low
+variability in recruitment and little or no variability in survival
+process errors (\(\sigma_{2+} \in \{0,0.25\}\)), and contrast in fishing
+pressure.
+
+In R+M OMs, the most accurate estimation of SRR parameters for all EM
+process error assumptions occurred when there was a change in fishing
+pressure, greater variability in natural mortality process errors, and
+lower observation error (Figure \ref{SR_rel_error}, B). Relative to the
+R, and R+S OMs, there was even less effect of estimating median natural
+mortality on estimation bias for the SRR parameters.
+
+Bias for SRR parameters was large and variability in relative errors was
+greatest for most EMs fit to R+Sel OMs with constant fishing pressure
+(Figure \ref{SR_rel_error}, C). Less bias in parameter estimation
+occurred for OMs with a change in fishing pressure and the best accuracy
+occurred for those OMs that had low observation error and more variable
+and uncorrelated selectivity process errors, and when the EMs hadd with
+the correct process error assumption. There was little effect of
+estimating natural mortality on relative errors for SRR parameters.
+
+Like R+Sel OMs, relative errors in SRR parameters for R+q OMs were more
+accurate for most EM process error types when OMs had contrast in
+fishing pressure and lower observation error (Figure \ref{SR_rel_error},
+D). However, the best accuracy occurred for those OMs that had lower
+variability in catchability process errors. The worst accuracy of SRR
+parameter estimation regardless of EM type occurred when R+q OMs had low
+observation error and constant fishing pressure (Figure
+\ref{SR_rel_error}, D, fourth row).
 
@@
 
-Median natural mortality rate is estimated accurately by all EM process
-error types for all R OMs except those where observation error is high
-and fishing pressure is constant where variability in relative errors is
-high (Figure \ref{M_rel_error}, A, \(\sigma_{2+} = 0\)). For R+S OMs
-estimation of median natural mortality rate is most accurate when
-observation error is low and there is contrast in fishing pressure and
-the EM process error type is correct. There is little effect of
-estimating stock-recruit relationships on the bias for median natural
-mortality.
+Across all OMs and EMs there was little effect of estimating SRRs on the
+bias in estimation of median natural mortality (Figure
+\ref{M_rel_error}). Median natural mortality rate was estimated
+accurately by all EM process error types for all R OMs except those with
+high observation error and constant fishing pressure, in which case
+relative errors were high (Figure \ref{M_rel_error}, A,
+\(\sigma_{2+} = 0\)). For R+S OMs estimation of median natural mortality
+rate was most accurate when observation error was low and there was
+contrast in fishing pressure and the EM process error type was correct.
 
@@
 correlated natural mortality process errors only had unbiased estimation
-when the EM process error type was correct. Like R, and R+S OMs, there
-was little effect of estimating stock-recruit relationships on the bias
-for median natural mortality.
-
-All EM process error types produced accurate estimation of median
-natural mortality rate for R+Sel OMs that had contrast in fishing
-pressure, low observation error, and low selectivity process error
-variability (Figure \ref{M_rel_error}, C). When selectivity process
-error variability increased the incorrect EM process errors produce more
-biased estimation of median natural mortality rate. The least accurate
+when the EM process error type was correct.
+
+All EM process error types accurately estimated median natural mortality
+rate for R+Sel OMs that had contrast in fishing pressure, low
+observation error, and low selectivity process error variability (Figure
+\ref{M_rel_error}, C). When selectivity process error variability
+increased, the incorrect EM process errors produce more biased
+estimation of median natural mortality rate. The least accurate
 estimation occurred for all EM process error types when observation
-error was high and there was constant fishing pressure. Again, like
-other OM process error types, there was little effect of estimating
-stock-recruit relationships on the bias for median natural mortality.
+error was high and fishing pressure was constant.
 
@@
 types produced biased estimation of median natural mortality when R+q
-OMs had high observaiton error and constant fishing pressure. Again,
-there was no discernible effect of estimating stock-recruit
-relationships on bias.
+OMs had high observaiton error and constant fishing pressure.
+
+Like R+Sel OMs, all EM process error types produced accurate estimation
+of median natural mortality rate when fit to R+q OMs with contrast in
+fishing pressure, low observation error and low catchability process
+error variability (Figure \ref{M_rel_error}, D). Most EM process error
+types produced biased estimation of median natural mortality when R+q
+OMs had high observation error and constant fishing pressure.
 
@@
 OMs, regardless of EM process error types, and whether median natural
-mortality rate or stock-recruit relationships were estimated (Figure
-\ref{mohns_rho_ssb}, A). The strongest retrospective patterns occurred
-in OMs with the largest survival process error variability, high
-observation error, and contrast in fishing pressure, but only for EMs
-without the correct process error type and where median natural
-mortality rate was assumed known. For R+M, R+Sel, and R+q OMs, Mohn's
-\(\rho\) was also small in absolute value, but median values were all
-closer to 0 than the largest values in the R and R+S OMs (Figure
+mortality rate or SRRs were estimated (Figure \ref{mohns_rho_ssb}, A).
+The strongest retrospective patterns (highest absolute Mohn's \(\rho\)
+values) occurred in OMs with the largest apparent survival process error
+variability, high observation error, and contrast in fishing pressure,
+but only for EMs with the incorrect process error type and where median
+natural mortality rate was assumed known (median \(\rho\) was
+approximately -0.15). For R+M, R+Sel, and R+q OMs, Mohn's \(\rho\) was
+also small in absolute value, but median values were all closer to 0
+than the largest values in the R and R+S OMs (Figure
 \ref{mohns_rho_ssb},B-D). For these OMs, there was no noticeable effect
-of estimation of median natural mortality rate or stock recruit
-relationship on Mohn's \(\rho\) for any EM process error types.
+of estimation of median natural mortality rate or SRRs on Mohn's
+\(\rho\) for any EM process error types.
 
@@
 OMs with low variability in recruitment process errors, regardless of EM
-process error type, and whether median natural mortality rate or
-stock-recruit relationships were estimated (Figure \ref{mohns_rho_R},
-A). However, R and R+S OMs with greater recruitment process variability
-and higher observation error had median Mohn's \(\rho\) for recruitment
-greater than zero for most EMs even when the the EM process error type
-was correct. In R+S OMs with lower observation error, EMs with the
-correct process error type exhibited better median Mohn's \(\rho\) close
-to 0 than EMs with the incorrect process error type. For R+M, R+Sel, and
-R+q OMs, results for Mohn's \(\rho\) for recruitment are similar to
-those for SSB, but the range in median values and variation in Mohn's
-\(\rho\) values for a given OM are generally larger for recruitment
-(Figure \ref{mohns_rho_R}, B-D).
+process error type, and whether median natural mortality rate or SRRs
+were estimated (Figure \ref{mohns_rho_R}, A). However, R and R+S OMs
+with greater recruitment process variability and higher observation
+error had median Mohn's \(\rho\) for recruitment greater than zero for
+most EMs even when the EM process error type was correct. In R+S OMs
+with lower observation error, EMs with the correct process error type
+exhibited better median Mohn's \(\rho\) close to 0 than EMs with the
+incorrect process error type. For R+M, R+Sel, and R+q OMs, results for
+Mohn's \(\rho\) for recruitment are similar to those for SSB, but the
+range in median values and variation in Mohn's \(\rho\) values for a
+given OM are generally larger for recruitment (Figure \ref{mohns_rho_R},
+B-D).
 
@@
 
-\hypertarget{convergence}{%
-\subsection*{Convergence}\label{convergence}}
+\hypertarget{convergence-1}{%
+\subsection*{Convergence}\label{convergence-1}}
 \addcontentsline{toc}{subsection}{Convergence}
 
-Convergence results can be useful for understanding how lack of
-convergence in applications to real data might direct the practitioner
-to which alternative random effects configurations is more appropriate.
-Therefore, the type of convergence that we might use as a diagnostic is
-important. It is common during assessment model fitting to check that
-the maximum absolute gradient component is less than some threshold, but
-there is no standard. However, if the Hessian were checked for positive
-definiteness conditional on a satisfactory gradient, some models may be
-excluded that in fact have an invertible hessian. We found the hessian
-at the optimized log-likelihood can often be invertible when the
-gradient approach with a sensible criterion (e.g., \(10^{-3}\)) would
-indicate lack of convergence.
-
-Another factor affecting the convergence criteria is numerical accuracy.
-With the process errors modeled as random effects we are optimizing an
-(Laplace) approximation of the marginal log-likelihood and therefore
-hessians and gradients are also with regard to the approximation.
-Furthermore, numerical methods are used to calculate the hessian
-(Optimhess?), and also to invert it for variance estimation. All of
-these approximations are affecting the gradients and invertibility of
-the hessian (although likely worse for some models than others), which,
-along with our results, suggests at least investigating whether the
-hessian is positive definite when the calculated gradients are not
-terribly large. Often high gradients and/or lack of positive definite
-hessian result when parameter estimates are at their bounds (Carvalho et
-al. 2021), and this also applies to variance or autocorrelation
-parameters for random effects with state-space assessment models.
-
-Even when hessian-based variance can be accomplished, parameters that
-are poorly estimated will have extremely large variances. This further
-inspection can lead to a more appropriate and often more parsimonious
-model configuration where the problematic parameters are not estimated.
-For example, process error variance parameters that are estimated close
-to 0 indicates that the random effects are estimated to have little or
-no variability and removing these process errors is warranted.
-Generally, our results suggest we can expect lower probability of
-convergence of state-space assessment models when estimating natural
-mortality or stock-recruit relationships because of the difficulty
-distinguishing these parameters from others being estimated in
-assessment model with data typically available.
+Analyses of model convergence across simulations can be useful for
+understanding the utility of alternative convergence criteria used in
+applications to real data for directing the practitioner to more
+appropriate random effects configurations. It is common during the
+assessment model fitting process to check that the maximum absolute
+gradient component is less than some threshold prior to inspecting the
+Hessian of the optimized likelihood for invertibility (Carvalho et al.
+2021). However, there is no accepted standard for the gradient threshold
+(e.g., Lee et al. 2011; Hurtado-Ferro et al. 2014; Rudd and Thorson
+2018) and some thresholds would exclude models that in fact have an
+invertible Hessian. We found the Hessian at the optimized log-likelihood
+can often be invertible when the maximum absolute gradient was much
+larger than what would perceived to be a sensible threshold.
+
+Li et al. (2024) found that convergence rate could be a useful
+diagnostic especially for separating the correct model from overly
+complex models. However, the criteria for convergence used in their
+study may also lead to limited ability to distinguish the correct model
+from overly simplistic models, a pattern that was also noted by
+Liljestrand et al. (2024) in which one process error may absorb all
+sources of process error when the magnitude of other process errors are
+low.
+
+Often poor convergence result when parameter estimates are at their
+bounds (Carvalho et al. 2021), and this also applies to variance
+parameters for random effects with state-space assessment models. Even
+when Hessian is invertible, parameters that are poorly informed will
+have extremely large variance estimates. This further inspection can
+lead to a more appropriate and often more parsimonious model
+configuration where the problematic parameters are not estimated. For
+example, process error variance parameters that are estimated close to 0
+indicates that the random effects are estimated to have little or no
+variability and removing these process errors is warranted. Generally,
+our results suggest we can expect lower probability of convergence of
+state-space assessment models when estimating natural mortality or SRRs
+because of the difficulty distinguishing these parameters from others
+being estimated in assessment model with data that are typically
+available. Our experiments did not aim to emulate the practitioner
+decision process in developing model configurations (e.g.~removing a
+source of process error and refitting the model when process error
+variance parameters were estimated close to 0). Evaluating the efficacy
+of such a decision process when applying Ems might be important in
+closed loop simulations (e.g.~MSE) aimed at quantifying management
+performance.
+
+A factor affecting the convergence criteria, particularly for maximum
+likelihood estimation of models with random effects, is numerical
+accuracy. All optimizations performed in these simulations are of the
+Laplace approximation of the marginal likelihood and, therefore,
+gradients and Hessians are also with respect to this approximation (see
+TMB::sdreport in the Template Model Builder package). Functionality
+within the Template Model Builder package exists (i.e.,
+TMB::checkConsistency) to check the validity of the Laplace
+approximation and the utility of this as a diagnostic for state-space
+assessment models should be explored further. Furthermore, numerical
+methods are used to calculate and invert the Hessian for variance
+estimation for models with random effects. Along with our results, the
+potential lack of accuracy imposed by these approximations, suggests at
+least investigating whether the Hessian is positive definite when the
+calculated absolute gradients are not terribly large.
 
@@
 
-Among the process error configurations we used in OMs, we found AIC to
-be accurate for process errors on recruitment and survival (R and R+S).
-Fitting models to other OMs rarely preferred R+S EMs and R and R+S EMs
-were nearly always selected for the matching OMs. For other sources of
-process error, accuracy of AIC was improved to useful levels when there
-was larger variability in the process errors and/or lower observation
-error.
-
-We found AIC to performs weakly in determining stock recruit
-relationships unless there is large contrast in SSB and low variation in
-recruitment process errors. Although we did not compare models with
-alternative stock-recruit relationships (e.g., Ricker and
-Beverton-Holt), we do not expect AIC to perform any better
+Of the OM process error configurations we considered, we found AIC to be
+accurate for selecting models with process errors on recruitment and
+apparent survival (R and R+S). Fitting models to other OMs rarely
+preferred R+S EMs, and R and R+S EMs were nearly always selected for the
+matching OMs; a similar result was reported by Liljestrand et al.
+(2024). For other sources of process error, accuracy of AIC was improved
+to useful levels when there was larger variability in the process errors
+and/or lower observation error.
+
+Across all OM process error configurations, AIC performed poorly in
+identifying that the presence of the Beverton-Holt SRR in the OM unless
+there was contrast in fishing pressure possibly in combination with
+other factors such as lower variability in recruitment process errors
+(in R and R+S models) or greater variation in natural mortality process
+errors (for R+M OMs, Fig. \ref{sr_aic}). As such, properly accounting
+for process error in natural mortality could be important (Li et al.
+2024) when evaluating SRRs in state-space models. Curiously, we did not
+find a marked effect of the level of observation error on ability to
+detect the SRR, but it is possible that AIC would perform better if
+observations have even lower uncertainty than we considered.
+
+Although we did not compare models with alternative SRRs (e.g., Ricker
+and Beverton-Holt), we do not expect AIC to perform any better
 distinguishing between relationships. Our finding that AIC tended to
@@
 bias in AIC for more complex models (Shibata 1976; Katz 1981; Kass and
-Raftery 1995), but whereas those findings apply to more common fitting
-of models to raw data, state-space models account for observation error
-and estimate process errors in latent variables.
-
-Our results comport well with those of de Valpine and Hastings (2002)
-which found AIC could not distinguish among state-space stock-recruit
-models that were fit just to SSB and recruitment ``observations'' and
-Britten et al. which found AIC could not reliably distinguish
-alternative environmental effects on stock-recruit parameters. Miller et
-al. (2016) did find AIC to prefer a stock-recruit relationship with
-environmental effects for SNEMA yellowtail flounder, but there was a
-large change in stock size estimated over time and flatfish are
-well-observed by the NEFSC bottom trawl survey used for an index in the
-assessment model.
+Raftery 1995), but, whereas those findings apply to more the much more
+common comparison of models that are fit to raw and independent
+observations, here we are comparing state-space models which account for
+observation error and estimate process errors in latent variables.
+
+Our results comport with those of de Valpine and Hastings (2002) who
+found AIC could not distinguish among state-space SRRs that were fit
+just to SSB and recruitment observations (i.e., not an assessment
+model). Similarly, Britten et al. found AIC could not reliably
+distinguish alternative environmental effects on SRR parameters.
+However, Miller et al. (2016) did find AIC to prefer a SRR with
+environmental effects when applied to data for the SNEMA yellowtail
+flounder stock and AIC also selected an environmental covariate on a SRR
+for the most recent stock assessment of Georges Bank yellowtail flounder
+(NEFSC 2025). Both of these yellowtail flounder stocks have large
+changes in stock size and the values of environmental covariates over
+time. Additionally, this species is well-observed by the bottom trawl
+survey that is used for an index in assessment models.
 
@@
 
-As we might have expected, bias in all parameters and assessment output
-was generally better with lower observation error. Estimation of
-stock-recruit relationship parameters was not reliable in any of the
-OM-EM combinations, but estimation of median natural mortality was
-feasible in many OM scenarios with temporal contrast in fishing
-pressure. For OMs where there was bias in natural mortality due to high
-observation error (R+Sel and R+q OMS), estimating the stock-recruit
-relationship seemed to alleviate the bias. However, estimation of median
-natural mortality can cause large differences between the true and
-estimated SSB (that may or may not be unbiased on average) when there is
-less contrast in fishing pressure over time and higher observation
-error.
+As expected, bias in all parameters and assessment output was generally
+improved with lower observation error. Estimation of SRR parameters was
+reliable in ideal scenarios of low observation error and contrast in
+fishing for some R+Sel and R+M OMs, but generally estimation was biased
+and(or) highly variable. We found substantial bias in estimated SRR
+parameters in R and R+S OMs particularly with high variability in
+recruitment and apparent survival process errors, suggesting that
+practitioners should be cautious when models without the SRR suggest
+this to be the case.
+
+On the other hand, estimation of median natural mortality was reliable
+in many OM scenarios with contrast in fishing pressure and in some OMs
+EMs that also estimated the SRR parameters bias for those parameters was
+improved. Conversely, for some R+Sel and R+q OMs where there was bias in
+natural mortality due to high observation error, estimating the SRR
+reduced the bias in median natural mortality rate. However, estimating
+median natural mortality did cause poor accuracy in SSB estimation in
+many OMs without contrast in fishing pressure over time and with higher
+observation error. Thus, estimating median natural mortality should be
+approached with caution in state-space assessment models, particularly
+given its significant impact on determination of reference point and
+stock status (Li et al. 2024).
 
@@
 
-Incorrect process error assumptions for EMs did not produce strong
-retrospective patterns for SSB for any OMs whether median natural
-mortality and a stock-recruit relationship were estimated or not, but
-some weak retrospective patterns occur when observation error was high
-and there was contrast in fishing pressure. Retrospective patterns
-tended to be more variable for recruitment and can be large even when
-the EM is correct. Therefore, we recommend emphasis on inspection of
-retrospective patterns primarily for SSB and F, but further research on
+Incorrect EM process error assumptions did not produce strong
+retrospective patterns for SSB for any OMs regardless of whether median
+natural mortality or a SRR was estimated, but some weak retrospective
+patterns occur when observation error was high and there was contrast in
+fishing pressure. However, retrospective patterns tended to be more
+variable for recruitment and were sometimes large even when the EM was
+correct. Therefore, we recommend emphasis on inspection of retrospective
+patterns primarily for SSB and \(F\), but further research on
 retrospective patterns in other assessment model parameters, management
-quantitities, and projections may be beneficial (Brooks and Legault
-2016).
+quantities such as biological reference points, and projections may be
+beneficial (Brooks and Legault 2016).
+
+The general lack of retrospective patterns with mis-specified process
+errors is perhaps to be expected. Retrospective patterns are often
+induced in simulation studies by rapid changes in a quantity such as
+index catchability, natural mortality, or perceived catch during years
+toward the end of the time series (Legault 2009; Miller and Legault
+2017; Huynh et al. 2022; Breivik et al. 2023). In our simulations, the
+process errors changing over time may have trends in particular
+simulations, particularly when strong autocorrelation is imposed, but
+the random effects have no trend on average across simulations.
+Szuwalski et al. (2018) and lietal24 also found relatively small
+retrospective patterns when the source of mis-specification was temporal
+variation in demography attributes. Indeed, it is common for the
+flexibility provided by temporal random effects to reduce retrospective
+patterns (Miller et al. 2018; Stock et al. 2021; Stock and Miller 2021),
+though it does not necessarily indicate a more accurate assessment model
+(Perretti et al. 2020; Li et al. 2024; Liljestrand et al. 2024). Our
+results together with the existing literature seem to suggest that when
+a strong retrospective pattern is observed in an assessment it is more
+likely to be due to a mis-specification of a rapid shift in some model
+attribute rather than whether a particular process is assumed to be
+randomly varying temporally.
 
@@
 
-All together what do results mean? E.g., when do you get good
-convergence and retrospective patterns and high/none bias? When do you
-get good convergence and no retros and no/big bias?
-
-For R+S OMs with low observation error, temporal contrast in fishing
-pressure, and large variation in survival process errors (3rd row), some
-EMs with the incorrect process error assumptions, estimate median M to
-be 0, and SSB is also underestimated, but Mohn's \(\rho\) for SSB is
-better than when M is known. Fortunately, AIC would not choose the wrong
-EM process error type in these situations.
-
-When R+S is preferred AMONG THE TYPES CONSIDERED HERE, it probably means
-its the correct process error .
+Our simulation study examined the importance of several factors for
+reliable inferences from state-space age-structured assessment models.
+Across all factors, AIC accurately distinguished models with process
+errors on recruitment only or on recruitment and apparent survival.
+Accuracy for other process error types required a strong signal (high
+process variability) with low noise (low observation uncertainty).
+Therefore, we expect practitioners will find R+S configurations to
+provide satisfactory diagnostics across a range of life history and data
+quality scenarios. Contrast in fishing pressure was consistently an
+important factor across all measures of reliability we examined. AIC
+generally performed poorly for selecting the SRR but performance was
+improved with low recruitment variability and contrast in fishing
+pressure. However, some bias in estimation in at least one of the SRR
+parameters existed in nearly all OM-EM combinations. Because bias in
+terminal SSB and retrospective patterns were indifferent to whether or
+not the SRR was estimated, and convergence was slightly better without
+the SRR, a sensible default would be to fit models without an assumed
+SRR.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-breiviketal23}{}}%
+Breivik, O.N., Aldrin, M., Fuglebakk, E., and Nielsen, A. 2023.
+Detecting significant retrospective patterns in state space fish stock
+assessment. Canadian Journal of Fisheries and Aquatic Sciences
+\textbf{80}(9): 1509--1518.
+doi:\href{https://doi.org/10.1139/cjfas-2022-0250}{10.1139/cjfas-2022-0250}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-brittenetalinreview}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-clopperpearson34}{}}%
+Clopper, C.J., and Pearson, E.S. 1934. The use of confidence or fiducial
+limits illustrated in the case of the binomial. Biometrika
+\textbf{26}(4): 404--413.
+doi:\href{https://doi.org/10.1093/biomet/26.4.404}{10.1093/biomet/26.4.404}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-connetal10}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-hurtadoferroetal15}{}}%
+Hurtado-Ferro, F., Szuwalski, C.S., Valero, J.L., Anderson, S.C.,
+Cunningham, C.J., Johnson, K.F., Licandeo, R., McGilliard, C.R.,
+Monnahan, C.C., Muradian, M.L., Ono, K., Vert-Pre, K.A., Whitten, A.R.,
+and Punt, A.E. 2014. Looking in the rear-view mirror: Bias and
+retrospective patterns in integrated, age-structured stock assessment
+models. ICES Journal of Marine Science \textbf{72}(1): 99--110.
+doi:\href{https://doi.org/10.1093/icesjms/fsu198}{10.1093/icesjms/fsu198}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-huynhetal22}{}}%
+Huynh, Q.C., Legault, C.M., Hordyk, A.R., and Carruthers, T.R. 2022. A
+closed-loop simulation framework and indicator approach for evaluating
+impacts of retrospective patterns in stock assessments. ICES Journal of
+Marine Science \textbf{79}(7): 2003--2016.
+doi:\href{https://doi.org/10.1093/icesjms/fsac066}{10.1093/icesjms/fsac066}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-kapuretal24}{}}%
+Kapur, M.S., Ducharme-Barth, N., Oshima, M., and Carvalho, F. 2025. Good
+practices, trade-offs, and precautions for model diagnostics in
+integrated stock assessments. Fisheries Research \textbf{281}: 107206.
+doi:\href{https://doi.org/10.1016/j.fishres.2024.107206}{10.1016/j.fishres.2024.107206}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-kassraftery95}{}}%
@@
 doi:\href{https://doi.org/10.1890/08-0071.1}{10.1890/08-0071.1}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-kristensenetal16}{}}%
+Kristensen, K., Nielsen, A., Berg, C.W., Skaug, H., and Bell, B.M. 2016.
+{TMB}: Automatic differentiation and {L}aplace approximation. Journal of
+Statistical Software \textbf{70}(5): 1--21.
+
+\leavevmode\vadjust pre{\hypertarget{ref-leeetal11}{}}%
+Lee, H.-H., Maunder, M.N., Piner, K.R., and Methot, R.D. 2011.
+Estimating natural mortality within a fisheries stock assessment model:
+An evaluation using simulation analysis based on twelve stock
+assessments. Fisheries Research \textbf{109}(1): 89--94.
+doi:\href{https://doi.org/10.1016/j.fishres.2011.01.021}{10.1016/j.fishres.2011.01.021}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-legault09}{}}%
+Legault, C.M. 2009. Report of the retrospective working group, 14--16
+january 2008. US Department of Commerce Northeast Fisheries Science
+Center Reference Document 09-01. US Department of Commerce Northeast
+Fisheries Science Center. Woods Hole, MA.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-millerlegault17}{}}%
+Miller, T.J., and Legault, C.M. 2017. Statistical behavior of
+retrospective patterns and their effects on estimation of stock and
+harvest status. Fisheries Research \textbf{186}: 109--120.
+doi:\href{https://doi.org/10.1016/j.fishres.2016.08.002}{10.1016/j.fishres.2016.08.002}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-milleretal18}{}}%
+Miller, T.J., O'Brien, L., and Fratantoni, P.S. 2018. Temporal and
+environmental variation in growth and maturity and effects on management
+reference points of {G}eorges {B}ank {A}tlantic cod. Canadian Journal of
+Fisheries and Aquatic Sciences \textbf{75}(12): 2159--2171.
+doi:\href{https://doi.org/10.1139/cjfas-2017-0124}{10.1139/cjfas-2017-0124}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-millerstock20}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-nefsc22}{}}%
+NEFSC. 2022a. Final report of the haddock research track assessment
+working group. {Available} at
+https://s3.us-east-1.amazonaws.com/nefmc.org/14b\_EGB\_Research\_Track\_Haddock\_WG\_Report\_DRAFT.pdf.
+
+\leavevmode\vadjust pre{\hypertarget{ref-nefsc22a}{}}%
+NEFSC. 2022b. Report of the american plaice research track working
+group. {Available} at
+https://s3.us-east-1.amazonaws.com/nefmc.org/2\_American-Plaice-WG-Report.pdf.
+
+\leavevmode\vadjust pre{\hypertarget{ref-nefsc24}{}}%
+NEFSC. 2024. Butterfish research track assessment report. US Dept Commer
+Northeast Fish Sci Cent Ref Doc. 24-03; 191 p.
+
+\leavevmode\vadjust pre{\hypertarget{ref-nefsc25}{}}%
+NEFSC. 2025. Yellowttail flounder research track working group report.
+{Available} at
+https://d23h0vhsm26o6d.cloudfront.net/10c.-Yellowtail-Flounder-RT-WG-Report.pdf.
+
 \leavevmode\vadjust pre{\hypertarget{ref-nielsenberg14}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-perrettietal20}{}}%
+Perretti, C.T., Deroba, J.J., and Legault, C.M. 2020. Simulation testing
+methods for estimating misreported catch in a state-space stock
+assessment model. ICES Journal of Marine Science \textbf{77}(3):
+911--920.
+doi:\href{https://doi.org/10.1093/icesjms/fsaa034}{10.1093/icesjms/fsaa034}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-polanskyetal09}{}}%
@@
 \textbf{158}: 124--134. Elsevier.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ruddthorson18}{}}%
+Rudd, M.B., and Thorson, J.T. 2018. Accounting for variable recruitment
+and fishing mortality in length-based stock assessments for data-limited
+fisheries. Canadian Journal of Fisheries and Aquatic Sciences
+\textbf{75}(7): 1019--1035.
+doi:\href{https://doi.org/10.1139/cjfas-2017-0143}{10.1139/cjfas-2017-0143}.
 
@@
 performance. ICES Journal of Marine Science \textbf{75}(2): 596--609.
-Oxford University Press.
+doi:\href{https://doi.org/10.1093/icesjms/fsx159}{10.1093/icesjms/fsx159}.
 
@@
 doi:\href{https://doi.org/10.1093/icesjms/fsu213}{10.1093/icesjms/fsu213}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-thulin14}{}}%
+Thulin, M. 2014. {The cost of using exact confidence intervals for a
+binomial proportion}. Electronic Journal of Statistics \textbf{8}(1):
+817--840.
+doi:\href{https://doi.org/10.1214/14-EJS909}{10.1214/14-EJS909}.
 

### 2025-04-08 Project_0/code/manuscript/aic_results.R

@@ -4,6 +4,7 @@ library(ggh4x)
 library(dplyr)
 library(tidyr)
 library(ggpattern)
+library(patchwork)
 df.ems = readRDS(file.path(here(),"Project_0","inputs", "df.ems.RDS"))
 
 
@@ -233,7 +234,6 @@ R_q_aic_plt
 
 #all in one figure
 
-library(patchwork)
 theme_set(theme_bw())
 theme_update(strip.placement = "outside", strip.background = element_rect(), title = element_text(size = rel(1.5)))
 R_Sel_aic_plt_alt <- R_Sel_aic_plt + 
@@ -315,7 +315,7 @@ types.plt <- c("R, R+S","R+M", "R+Sel", "R+q")
 est_conds <- cbind(c(FALSE,TRUE))
 em_inds <- cbind(1:20, 5:24, c(5:20,25:28), c(5:20, 29:32))
 pred_dfs <- list()
-
+obs_dfs <- list()
 for(i in 1:4) {
   all_aic <- readRDS(file = here("Project_0","results", paste0("all_", types[i], "_aic_results.RDS")))
   df.oms = readRDS(here("Project_0","inputs", paste0("df.", ifelse(i==1, "", paste0(types[i],".")),"oms.RDS")))
@@ -399,6 +399,7 @@ for(i in 1:4) {
       "ME" = "M estimated"))
   pred_df$log_sd_log_SSB <- log(pred_df$sd_log_SSB)
   pred_df$log_obs <- log(pred_df$obs)
+  pred_df$BH_best <- c(df$BH_best_MF,df$BH_best_ME)
   pred_dfs[[i]] <- pred_df
 }
     R_S_sr_aic_plt <- ggplot(pred_dfs[[1]], aes(x = log_sd_log_SSB, y = pred, colour = type, fill =  type)) + 
@@ -453,5 +454,78 @@ design <- c(area(1,1,1,1), area(2,1,2,1), area(1,2,1,2), area(2,2,2,2))
 (R_M_sr_aic_plt + labs(title = "B: R+M OMs") + theme(legend.position="none", strip.text.y=element_blank())) + #axis.title.x = element_blank(), strip.text=element_text(margin=margin(b = 5)), axis.text.x=element_blank(), plot.margin = margin(b = 1, t = 0))) + 
[…]

### 2025-04-08 Project_0/code/manuscript/om_em_description_tables.R

@@ -1,0 +1,147 @@
+library(here)
+library(dplyr)
+library(Hmisc)
+df.ems = readRDS(file.path(here(),"Project_0","inputs", "df.ems.RDS"))
+
+#NAA oms: ems = 1-20
+df.oms = readRDS(file.path(here(),"Project_0","inputs", "df.oms.RDS"))
+om_inputs = readRDS(file.path(here(),"Project_0","inputs", "NAA_om_inputs.RDS"))
+
+naa_om_tab <- df.oms
+naa_om_tab$Model <- paste0("$\\text{NAA}_{", 1:24, "}$")
+naa_om_tab <- naa_om_tab %>%
+  mutate(obs_error = recode(obs_error,
+    "L" = "Index SD = 0.1, Age composition SD = 0.3",
+    "H" = "Index SD = 0.4, Age composition SD = 1.5"
+  ))
+naa_om_tab <- naa_om_tab %>%
+  mutate(Fhist = recode(Fhist,
+    "H-MSY" = "$2.5 F_{\\text{MSY}} \\rightarrow F_{\\text{MSY}}$",
+    "MSY" = "$F_{\\text{MSY}}$"
+  ))
+
+names(naa_om_tab)[2] <- "$\\sigma_R$"
+names(naa_om_tab)[3] <- "$\\sigma_{2+}$"
+names(naa_om_tab)[4] <- "Fishing History"
+names(naa_om_tab)[5] <- "Observation Uncertainty"
+
+x = latex(naa_om_tab, file = here("Project_0","manuscript","naa_om_tab.tex"), 
+  table.env = FALSE, col.just = rep("r", dim(naa_om_tab)[2]), rowname = NULL)
+
+
+#M oms: ems = 5-24
+df.oms = readRDS(file.path(here(),"Project_0","inputs", "df.M.oms.RDS"))
+om_inputs = readRDS(file.path(here(),"Project_0","inputs", "M_om_inputs.RDS"))
+
+om_tab <- df.oms
+om_tab$Model <- paste0("$M_{", 1:16, "}$")
+om_tab <- om_tab %>%
+  mutate(obs_error = recode(obs_error,
+    "L" = "Index SD = 0.1, Age composition SD = 0.3",
+    "H" = "Index SD = 0.4, Age composition SD = 1.5"
+  ))
+om_tab <- om_tab %>%
[…]

### 2025-04-16 Ecov_study/mortality/manuscript/manuscript.Rmd
saving work on Ecov M

@@
 $^2$Northeast Fisheries Science Center, Woods Hole Laboratory, 166 Water Street, Woods Hole, MA 02543 USA\
-$^3$Woods Hole Oceanographic Institution\
-$^4$SMAST\
-$^5$Mid-Atlantic Fisheries Management Council\
-$^6$Rutgers University\
+$^3$Biology Department, Woods Hole Oceanographic Institution, 266 Woods Hole Rd. Woods Hole, MA, USA\
+$^4$Department of Fisheries Oceanography, School for Marine Science and Technology, University of Massachusetts Dartmouth, 836 S Rodney French Boulevard, New Bedford, MA 02740, USA\
+$^5$Mid-Atlantic Fishery Management Council, 800 North State Street, Suite 201, Dover, DE 19901 USA\
+$^6$Department of Ecology, Evolution, and Natural Resources. Rutgers University\
 
@@
 
+Estimation of natural mortality is known to be challenging in stock assessment models (Lee et al., others). However, estimation has been shown to be reliable in some situations (miller and hyun 2018)
+
+Miller et al in review (project 0) found estimation of natural mortality was feasible when data are good and contrast in fishing pressure etc.
+
+Let's focus results here on those situations so that we can reduce the plots in figs.
+
+Variation in natural mortality has an immediate impact on projections unlike recruitment.
+Understanding this varation in M leads to better understanding of post-recruit productivity and therefore management.
+
 Here we conduct a simulation study with operating models varying by degree of observation error uncertainty, sources of process error (M or survival), fishing history, temporal variation in environmental covariates, and magnitude of the effect of the covariate on natural mortality.
@@
 
-All of our analyses used the Woods Hole Assessment Model (WHAM) to construct both OMs and EMs [@millerstock20;@stockmiller21].
+All of our analyses used the Woods Hole Assessment Model (WHAM) to construct both OMs and EMs [@millerstock20;@stockmiller21;@milleretalinreview].
 We used [version 1.0.6.9000, commit 77bbd94](https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3) for this simulation study.
@@
 For operating models with natural mortality random effects, we assume the same standard deviation $\sigma_M = 0.3$.
+This is similar to the intermediate value used for R+M OMs by @milleretalinreview.
 
@@
 
-As described by \citet{stockmiller21}, the model for process errors on abundance at age in WHAM is
-\begin{equation}
-  \text{log}N_{a,y}=\left\{
-    \begin{array}{@{}lll@{}}
-      \text{log} \left( f(SSB_{y-1}) \right) + \varepsilon_{1,y}, & \text{if}\ a = 1 \\
-      \text{log} \left( N_{a-1,y-1} \right) - Z_{a-1,y-1} + \varepsilon_{a,y}, & \text{if}\ 1 < a < A \\
-      \text{log} \left( N_{A-1,y-1} e^{-Z_{A-1,y-1}} + N_{A,y-1} e^{-Z_{A,y-1}} \right) + \varepsilon_{A,y}, & \text{if}\ a = A
-    \end{array}\right.
-\end{equation}
-We assume the the standard deviation of recruitment deviations is $\text{SD}\left(\varepsilon_{1,y}\right) = \sigma_R = 0.5$ for all operating models.
-For operating models with survival devations for older ages, we assume the same standard deviations for all older ages $\text{SD}\left(\varepsilon_{a,y}\right) = \sigma_{2+} = 0.3$.
-
-The alternative configurations of random effects on recruitment, survival, and natural mortality are
-\begin{itemize}
-\item uncorrelated random effects on recruitment only ($\sigma_R = 0.5$) (R),
-\item uncorrelated random effects on recruitment ($\sigma_R = 0.5$) and survival of older age classes ($\sigma_{2+} = 0.3$) (R+S), and
-\item uncorrelated random effects on R ($\sigma_R = 0.5$) and log-natural mortality ($\sigma_M = 0.3$) (R+M).
-\end{itemize}
+We configured all OMs with uncorrelated random effects on recruitment with standard devaition on log(recruitment) $\sigma_R = 0.5$ for R, R+S, and R+M OMs.
+This same assumption was used by @milleretalinreview for R+M OMs and other OMs with fishery selectivity and index catchability process errors.
+For R+S OMs, apparent survival process errors were uncorrelated  with $\sigma_{2+} = 0.3$ and for R+M OMs, process errors on natural mortality were uncorrelated with $\sigma_M = 0.3$.
 
@@
 \end{itemize}
-The configuration of the process errors in the estimating models matched the corresponding options in the operating models.
-For example, if uncorrelated R+S was assumed for both the estimating and operating model, the assumptions about process errors matched.
+The configuration of the process errors in the estimating models generally matched the corresponding options in the operating models.
+
+For example, uncorrelated R+S was assumed for both the estimating and operating model.
+However, R+M EMs did not assume M random effects were uncorrelated (parameter was estimated).
 The environmental covariate observations were included in all estimation models to ensure comparability of information criteria.
@@
 
-We calculated median relative errors of
+We calculated median errors of
 \begin{itemize}
@@
 \item $\beta_M$, the mean log-natural mortality rate,
+\end{itemize}
+and the median relative errors of 
+\begin{itemize}
 \item annual natural mortality rate,
@@
 \end{itemize}
+
+Note that the exponentiating the median error of $\beta_M$ would be equivalent to the median relative error of the median natural mortality rate because the median of the exponential is the same as the exponential of the median.
+
 Results for fishing mortality rate are provided in the Supplementary Materials. For the $i$th simulated data the relative error for a parameter $\theta$ provided from the fitted estimation model is
@@
 \begin{figure}
-\caption{The proportion mature at age assumed for the population in all operating and estimating models.}\label{om_maturity}
-\begin{center}
-\includegraphics[width = \textwidth]{om_maturity.png}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\caption{The weight at age assumed for the population in all operating and estimating models.}\label{om_waa}
-\begin{center}
-\includegraphics[width = \textwidth]{om_waa.png}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\caption{The Beverton-Holt stock recruit relationship assume for all operating models.}\label{om_sr}
-\begin{center}
-\includegraphics[width = \textwidth]{om_sr.png}
-\end{center}
-\end{figure}
-
-\begin{figure}
 \caption{Example simulations of environmental covariate latent processes and observations with different levels of observation error, and different assumptions about variability of the latent process.}\label{om_ecov_example}
@@
 
-\begin{figure}
-\caption{The selectivity at age assumed for the fishing fleet in operating models without selectivity random effects.
-The mean selectivity at age across time for the fishing fleet in operating models with selectivity random effects.}\label{om_mean_selectivity}
-\begin{center}
-\includegraphics[width = \textwidth]{om_mean_selectivity.png}
-\end{center}
-\end{figure}
-
 \begin{landscape}
 \begin{figure}
-\caption{Probability of convergence of estimating models with alternative process error assumptions fit to simulated populations and data sets from operating models.
-The convergence criterion is an invertible hessian providing standard error estimates for all fixed effects parameters.
-Vertical lines represent 95\% confidence intervals.
-All estimating models estimate environmental covariate effects on natural mortality but assume the mean mortality parameter $\beta_\text{M} = \log 0.2$.}\label{convergence_M_fixed}
 \begin{center}
-\includegraphics[height = \textheight]{proportion_good_hessian_ecov_effect_est_M_fixed.png}
+\includegraphics{convergence}
 \end{center}
+\caption{Estimated probability of fits providing hessian-based standard errors for EMs assuming alternative process error, that estimate or assume known median natural mortality, and that estimate or assume no covariate effect on median natural mortality when fitted to operating models that have R (left) and R+S (middle) , or R+M (right) process error structures and three levels of true covariate effect on median natural mortality (x axis).
+Vertical lines represent 95\% confidence intervals.}\label{convergence}
 \end{figure}
@@
 \begin{figure}
-\caption{Probability of convergence of estimating models with alternative process error assumptions fit to simulated populations and data sets from operating models.
-The convergence criterion is an invertible hessian providing standard error estimates for all fixed effects parameters.
-Vertical lines represent 95\% confidence intervals.
-All estimating models estimate environmental covariate effects on natural mortality and the mean mortality parameter $\beta_\text{M}$.}\label{convergence_M_estimated}
 \begin{center}
-\includegraphics[height = \textheight]{proportion_good_hessian_ecov_effect_est_M_est.png}
+\includegraphics[height = \textheight]{aic}
 \end{center}
-\end{figure}
-\end{landscape}
-
-<!--
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets for each operating model where the estimation model with the lowest AIC had correct assumptions of process errors (R,R+S,R+M) and/or about whether the environmental covariate affected natural mortality.
-All estimating models assumed the mean mortality parameter $\beta_\text{M} = \log 0.2$.}\label{correct_assumptions_aic_M_fixed}
-\begin{center}
-\includegraphics[height = \textheight]{proportion_correct_PE_effect_M_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
--->
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets for each operating model where the estimation model type (process error and covariate effect assumptions) had the lowest AIC.
-All estimating models assumed the mean mortality parameter $\beta_\text{M} = \log 0.2$.}\label{aic_rank_M_fixed}
-\begin{center}
-\includegraphics[height = \textheight]{proportion_best_AIC_M_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-<!--
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets for each operating model where the estimation model with the lowest AIC had correct assumptions of process errors (R,R+S,R+M) and/or about whether the environmental covariate affected natural mortality.
-All estimating models estimate the mean mortality parameter $\beta_\text{M}$.}\label{correct_assumptions_aic_M_estimated}
-\begin{center}
-\includegraphics[height = \textheight]{proportion_correct_PE_effect_M_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
--->
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets for each operating model where the estimation model type (process error and covariate effect assumptions) had the lowest AIC.
-All estimating models assumed the mean mortality parameter $\beta_\text{M} = \log 0.2$.}\label{aic_rank_M_estimated}
-\begin{center}
-\includegraphics[height = \textheight]{proportion_best_AIC_M_estimated.png}
-\end{center}
+\caption{Proportion of simulated data sets for each operating model where the estimation model type (combination of assumed process error type and covariate effect) had the lowest AIC.
+All EMs had known median natural mortality rate and all OMs and EMs had low observation error.}\label{aic}
 \end{figure}
@@
 \begin{figure}
-\caption{Median relative error of the estimated environmental effect on natural mortality $\beta_\text{Ecov}$ from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
-All estimating models assumed the mean mortality parameter $\beta_\text{M} = \log 0.2$.}\label{Ecov_beta_bias_M_fixed}
 \begin{center}
-\includegraphics[height = \textheight]{Ecov_beta_bias_all_PE_effect_M_fixed.png}
+\includegraphics[height = \textheight]{Ecov_beta_bias}
 \end{center}
+\caption{Relative error of estimates of environmental effect on natural mortality $\beta_\text{Ecov}$ from fitting EMs with alternative process error assumptions.
+All OMs had low observation error and contrast in fishing mortality.
+Vertical lines represent 95\% confidence intervals.}\label{ecov_beta_bias}
 \end{figure}
@@
 \begin{figure}
-\caption{Median relative error of the estimated environmental effect on natural mortality $\beta_\text{Ecov}$ from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
-All estimating models assumed the mean mortality parameter $\beta_\text{M}$ is also estimated.}\label{Ecov_beta_bias_M_estimated}
 \begin{center}
-\includegraphics[height = \textheight]{Ecov_beta_bias_all_PE_effect_M_estimated.png}
+\includegraphics[height = \textheight]{median_log_M_bias}
 \end{center}
+\caption{Relative error of the estimated mean log-natural mortality parameter $\beta_{M}$ from fitting EMs with alternative process error assumptions.
+All OMs had low observation error and contrast in fishing mortality.
+Vertical lines represent 95\% confidence intervals.}\label{median_log_M_bias}
 \end{figure}
@@
 \begin{figure}
-\caption{Median relative error of the estimated mean natural mortality parameter $\beta_\text{M}$ from fitting simulated observations from each operating model with the estimating model assuming $\beta_\text{Ecov} = 0$ under alternative process error assumptions.}\label{mean_M_bias_Ecov_beta_fixed}
+\caption{Median relative error of annual natural mortality rate in terminal year from fitting simulated observations from each OM with alternative process error assumptions in the EM.
+Mean log-natural mortality parameter ($\beta_\text{M}$) and environmental covariate effect ($\beta_\text{Ecov}$) are both estimated in the EMs.
+All OMs had low observation error and contrast in fishing mortality.}\label{terminal_M_bias}
 \begin{center}
-\includegraphics[height = \textheight]{mean_M_bias_all_PE_effect_Ecov_beta_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of the estimated mean natural mortality parameter $\beta_\text{M}$ from fitting simulated observations from each operating model with the estimating model estimating $\beta_\text{Ecov} = 0$ under alternative process error assumptions.}\label{mean_M_bias_Ecov_beta_estimated}
-\begin{center}
-\includegraphics[height = \textheight]{mean_M_bias_all_PE_effect_Ecov_beta_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of annual natural mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
-Estimating models also assume the mean natural mortality parameter $\beta_\text{M} = \log 0.2$ and no environmental covariate effect $\beta_\text{Ecov} = 0$.}\label{annual_M_bias_M_fixed_beta_fixed}
-\begin{center}
-\includegraphics[height = \textheight]{annual_M_bias_all_PE_effect_M_fixed_beta_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of annual natural mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
-Estimating models also assume the mean natural mortality parameter $\beta_\text{M} = \log 0.2$ and the environmental covariate effect $\beta_\text{Ecov}$ is estimated.}\label{annual_M_bias_M_fixed_beta_estimated}
-\begin{center}
-\includegraphics[height = \textheight]{annual_M_bias_all_PE_effect_M_fixed_beta_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of annual natural mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
-Estimating models also assume both the mean natural mortality parameter $\beta_\text{M}$ and the environmental covariate effect $\beta_\text{Ecov}$ are estimated.}\label{annual_M_bias_M_estimated_beta_estimated}
-\begin{center}
-\includegraphics[height = \textheight]{annual_M_bias_all_PE_effect_M_estimated_beta_estimated.png}
+\includegraphics[height = \textheight]{terminal_year_M_bias.png}
 \end{center}
@@
 # Supplemental Materials {-}
-<!--
+
 \setcounter{figure}{0}
 \renewcommand\thefigure{S\arabic{figure}}
--->
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{aic_supp1}
+\end{center}
+\caption{Proportion of simulated data sets for each OM where the EM type (assumed process error type) had the lowest AIC.
+All EMs had known median mortality rate and all OMs and EMs had low observation error.}\label{aic_supp1}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{aic_supp2}
+\end{center}
+\caption{Proportion of simulated data sets for each operating model where the estimation model type (combination of assumed process error type and covariate effect) had the lowest AIC.
+All OMs had a change in fishing pressure over time and all OMs and EMs had low observation error.}\label{aic_supp2}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{aic_supp3}
+\end{center}
+\caption{Proportion of simulated data sets for each operating model where the estimation model type (combination of assumed process error type and covariate effect) had the lowest AIC.
+All EMs had known median natural mortality rate and all OMs and EMs had high observation error.}\label{aic_supp3}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of annual natural mortality rate in terminal year from fitting simulated observations from each OM with alternative process error assumptions in the EM.
+Mean log-natural mortality parameter ($\beta_\text{M}$) is estimated, but EMs assume no environmental covariate effect ($\beta_\text{Ecov} = 0 $) are both estimated in the EMs.
+All OMs had low observation error and contrast in fishing mortality.}\label{terminal_M_bias_beta_0}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_M_bias_ecov_beta_0.png}
+\end{center}
+\end{figure}
+\end{landscape}
 

### 2025-04-16 Project_0/manuscript/manuscript.Rmd
saving work on Ecov M

@@
 $^2$Northeast Fisheries Science Center, Woods Hole Laboratory, 166 Water Street, Woods Hole, MA 02543 USA\
-$^3$Woods Hole Oceanographic Institution\
-$^4$SMAST\
+$^3$Biology Department, Woods Hole Oceanographic Institution, 266 Woods Hole Rd. Woods Hole, MA, USA\
+$^4$Department of Fisheries Oceanography, School for Marine Science and Technology, University of Massachusetts Dartmouth, 836 S Rodney French Boulevard, New Bedford, MA 02740, USA\
 $^5$Mid-Atlantic Fishery Management Council, 800 North State Street, Suite 201, Dover, DE 19901 USA\

### 2025-04-16 Ecov_study/mortality/code/manuscript/aic_ecov_effect_results_new.R

@@ -1,0 +1,184 @@
+library(here)
+library(ggplot2)
+library(dplyr)
+library(ggh4x)
+library(tidyr)
+library(patchwork)
+library(scales)
+library(ggpattern)
+
+df.ems = readRDS(file.path(here(),"Ecov_study","mortality","inputs", "df.ems.RDS"))
+df.oms = readRDS(file.path(here(),"Ecov_study","mortality","inputs", "df.oms.RDS"))
+aic_res <- readRDS(file.path(here(),"Ecov_study","mortality", "results", "aic_results.RDS"))
+
+aic_fn <- function(all, est_ind, om_ind = NULL){
+  if(!is.null(om_ind)) all <- all[om_ind]
+  out <- sapply(all, function(res){
+    tmp <- sapply(res[est_ind], function(x) return(x))
+    tmp = apply(tmp,1, function(x) {
+      if(any(!is.na(x))) {
+        return(x == min(x,na.rm=T))
+      } else return(rep(NA, length(x)))
+    })
+    return(apply(tmp,1,sum,na.rm=T))
+  })  
+  return(out)
+}
+
+make_df_fn <- function(M_est = TRUE){
+  #all EM PE assumptions
+  for(i in 1:3) {
+    re_mod <- c("rec", "rec+1", "rec+M")[i]
+    #EM:  M estimated
+    em_ind <- which(df.ems$M_est == M_est) #all EM PE assumptions
+    om_ind <- which(df.oms$NAA_M_re == re_mod) #om and em match
+    Mfixed_rec <- aic_fn(aic_res, em_ind, om_ind)
+    res <- cbind(df.oms[rep(om_ind, each = length(em_ind)),], df.ems[rep(em_ind, length(om_ind)),], n = c(Mfixed_rec))
+    if(i == 1) {
+      df <- res
+    } else {
+      df <- rbind(df, res)
+    }
+  }
+  facs <- c("Ecov_obs_sig", "Fhist", "Ecov_re_sig","Ecov_re_cor","obs_error", "NAA_M_re", "re_config", "Ecov_est")
+  df[facs] <- lapply(df[facs], factor)
+  df <- df %>%
+    mutate(Ecov_obs_sig = recode(Ecov_obs_sig,
+      "0.1" = "sigma[italic(e)] == 0.1",
+      "0.5" = "sigma[italic(e)] == 0.5"
+    ))
+  df <- df %>%
+    mutate(obs_error = recode(obs_error,
[…]

### 2025-04-16 Ecov_study/mortality/code/manuscript/aic_ecov_effect_results_orig.R

@@ -1,0 +1,206 @@
+library(here)
+library(ggplot2)
+library(dplyr)
+df.ems = readRDS(file.path(here(),"Ecov_study","mortality","inputs", "df.ems.RDS"))
+df.oms = readRDS(file.path(here(),"Ecov_study","mortality","inputs", "df.oms.RDS"))
+aic_res <- readRDS(file.path(here(),"Ecov_study","mortality", "results", "aic_results.RDS"))
+
+aic_fn <- function(all, est_ind, om_ind = NULL){
+  if(!is.null(om_ind)) all <- all[om_ind]
+  out <- sapply(all, function(res){
+    tmp <- sapply(res[est_ind], function(x) return(x))
+    #print(est_ind)
+    #print(length(res))
+    #print(tmp)
+    tmp = apply(tmp,1, function(x) {
+      if(any(!is.na(x))) {
+        return(x == min(x,na.rm=T))
+      } else return(rep(NA, length(x)))
+    })
+    return(apply(tmp,1,sum,na.rm=T))
+  })  
+  return(out)
+}
+
+make_df_fn <- function(M_est = TRUE){
+  #all EM PE assumptions
+  for(i in 1:3) {
+    re_mod <- c("rec", "rec+1", "rec+M")[i]
+    #EM:  M estimated
+    em_ind <- which(df.ems$M_est == M_est) #all EM PE assumptions
+    print(length(em_ind))
+    om_ind <- which(df.oms$NAA_M_re == re_mod) #om PE
+    print(length(om_ind))
+    res <- aic_fn(aic_res, em_ind, om_ind) #n_ems x n_oms
+    print(res[1,])
+    print(res[,1])
+    res <- cbind(df.oms[rep(om_ind, each = length(em_ind)),], df.ems[rep(em_ind, length(om_ind)),], n = c(res))
+    if(i == 1) {
+      all_res <- res
+    } else {
+      all_res <- rbind(all_res, res)
+    }
+  }
+  facs <- c("Ecov_obs_sig", "Fhist", "Ecov_re_sig","Ecov_re_cor","obs_error", "NAA_M_re", "re_config", "Ecov_est")
+  all_res[facs] <- lapply(all_res[facs], factor)
+  all_res_mod <- all_res %>%
+    mutate(Ecov_obs_sig = recode(Ecov_obs_sig,
[…]

### 2025-04-16 Ecov_study/mortality/code/manuscript/mohns_rho_results.R

@@ -1,392 +1,0 @@
-library(here)
-library(ggplot2)
-library(ggh4x)
-library(dplyr)
-library(tidyr)
-library(patchwork)
-library(reshape2)
-
-median_ci_fn <- function(x, alpha = 0.05){#, n.yrs = 1, n.sim = 100) {
-  x <- x[which(!is.na(x))]
-  n <- length(x)
-  bnds <- qbinom(c(alpha/2,1-alpha/2), n, 0.5)/n # 95% CI bounds for median
-  r <- quantile(x, probs = c(bnds[1], 0.5, bnds[2]))
-  names(r) <- c("lo", "middle", "hi")
-  r
-}
-make_mohns_rho_df <- function(om_type = "naa", res = all_naa_mohns_rho) {
-  df.ems <- readRDS(here::here("Project_0","inputs", "df.ems.RDS"))
-  if(om_type == "naa") {
-    em_ind <- 1:20
-  }
-  if(om_type == "M") {
-    em_ind <- 5:24
-  }
-  if(om_type == "Sel") {
-    em_ind <- c(5:20,25:28)
-  }
-  if(om_type == "q") {
-    em_ind <- c(5:20,29:32)
-  }
-  df.ems <- df.ems[em_ind,]
-
-  res <- melt(res)
-  names(res) <- c("rho", "em", "sim","om")
-  res$Type = c("SSB","F","R")
-  res <- cbind(df.ems[res$em,], res)
-
-  df <- res %>% group_by(om, em, Type) %>%
-    reframe(stats = median_ci_fn(rho)) %>% as.data.frame
-  df$type <- c("lo", "middle", "hi")
-  if(om_type == "naa") {
-    df.oms <- readRDS(here::here("Project_0", "inputs", "df.oms.RDS"))
-  } else {
-    df.oms <- readRDS(here::here("Project_0", "inputs", paste0("df.", om_type, ".oms.RDS")))
-  }
-  df <- df %>% pivot_wider(names_from = type, values_from = stats) %>% as.data.frame
-  df <- cbind(df, df.oms[df$om,])
-  df <- cbind(df, df.ems[df$em,])
-  df <- df %>%
-    mutate(EM_process_error = recode(re_config,
-      "rec" = "R",
-      "rec+1" = "R+S",
-      "M_re" = "R+M",
-      "q_re" = "R+q",
-      "sel_re" = "R+Sel"
-    ))
[…]

### 2025-04-17 Project_0/manuscript/manuscript.Rmd
saving work on project 0 edits

@@
 ---
-title: Factors affecting reliablity of state-space age-structured assessment models
+title: An investigation of factors affecting inferences from and reliablity of state-space age-structured assessment models
 author: 
@@
 In particular, the effects of observation error and underlying fishing history have not been fully isolated in simulation study designs, making it challenging to disentangle the interplay between process and observation error magnitudes, as demonstrated in @fischetal23.
-Third, explicitly modeling stock-recruit relationships (SRRs) as mechanistic drivers of population dynamics is promising [@fleischmanetal13; @dupontaviceetal22], but can reliability of inferences within integrated state-space age-structured models has not been evaluated.
+Third, explicitly modeling stock-recruit relationships (SRRs) as mechanistic drivers of population dynamics is promising [@fleischmanetal13; @dupontaviceetal22], but reliability of inferences within integrated state-space age-structured models has not been evaluated.
 Evidence from other studies suggests that when both process and observation errors are unknown, estimating density dependence parameters becomes highly uncertain [@knape08; @polanskyetal09].
@@
 Therefore, it is crucial to assess whether density dependence mechanisms can be estimated with sufficient precision for use in fisheries management [@augeretal16].
+Finally, although the importance of autocorrelation in process errors is recognized, investigations of the ability to distinguish state-space assessment models with and without autocorrelation and whether such misspecification is detrimental to estimation of important population metrics are lacking [@johnsonetal16;@xuetal19].
 
 In the present study, we conduct a simulation study with operating models (OMs) varying by degree of observation error, source and variability of process error, and fishing history.
-The simulations from these OMs are fitted with estimation models (EMs) that make alternative assumptions for sources of process error, whether a SRR was estimated, and whether a constant, or, in some EMs, median, natural mortality is estimated.
+The simulations from these OMs are fitted with estimation models (EMs) that make alternative assumptions for sources of process error, whether a SRR was estimated, and whether natural mortality is estimated.
 Given the confounding nature of process errors, developing diagnostic tools to detect model misspecification is of great scientific interest and could aid the next generation of stock assessments [@augeretal21].
@@
 For this study we used [version 1.0.6.9000, commit 77bbd94](https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3).
-WHAM has also been used to configure OMs and EMs for closed loop simulations evaluating index-based assessment methods [@legaultetal23] and is currently used or accepted for use in management of numerous fish stocks in the Northeast United States (NEUS) [e.g., @nefsc22;@nefsc22a;@nefsc24].
+WHAM has also been used to configure OMs and EMs for closed loop simulations evaluating index-based assessment methods [@legaultetal23] and is currently used or accepted for use in management of numerous NEUS fish stocks [e.g., @nefsc22;@nefsc22a;@nefsc24].
 
@@
 However, as @stockmiller21 note, these random effects can be due to events other than mortality, such as immigration, emigration, missreported catch, and other sources of misspecification.
-For each OM, assumptions about the magnitude of the variance of process errors and observations are required and the values we used were based on a review of the range of estimates from NEUS assessments using WHAM in stocks NEUS.
+For each OM, assumptions about the magnitude of the variance of process errors and observations are required and the values we used were based on a review of the range of estimates from Northeast Unite States (NEUS) assessments using WHAM.
 
@@
 We did not use the log-normal bias-correction feature for process errors or observations described by [@stockmiller21] for OMs and EMs to simplify interpretation of the study results [@lietalinreview].
-Simulations and model fitting were all carried out on the University of Massachusetts Green High-Performance Computing Cluster.
 All code we used to perform the simulation study and summarize results can be found at https://github.com/timjmiller/SSRTWG/tree/main/Project_0/code.
@@
 We assumed a Beverton-Holt SRR with constant pre-recruit mortality parameters for all OMs.
-All post-recruit productivity components are constant in the apparent survival (R+S) and survey catchability (R+Sel) process error OMs.
+All biological inputs to calculations of spawning biomass per recruit (i.e., weight, maturity, and natural mortality at age) are constant in the apparent survival (R+S) selectivity (R+Sel), and survey catchability (R+q) process error OMs.
 Therefore, steepness and unfished recruitment are also constant over the time period for those OMs [@millerbrooks21].
@@
 (Figure \ref{om_inputs_fig}, bottom right).
-For OMs without process errors on natural mortality we fixed the rate at 0.2.
-For OMs with process errors on natural mortality, the median natural mortality rate was specified to be 0.2.
+We assumed a value of 0.2 for the natural mortality rate in OMs without process errors on natural mortaliyt and for the median rate for OMs with process errors on natural mortality.
 
@@
 that is a linear combination of a mean log-natural mortality parameter that is constant across ages ($\mu_{M} = \log(0.2)$) and any annual random effects are marginally distributed as $\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)$.
+The marginal standard deviations we assumed for log natural mortality random effects were $\sigma_M \in \{0.1, 0.5\}$ and the random effects were either uncorrelated or first-order autoregressive (AR1, $\rho_M \in \{0,0.9\}$).
 Uncorrelated random effects were also included on recruitment with $\sigma_R = 0.5$ (hence, we denote these OMs as  R+M).
-The marginal standard deviations we assumed for log natural mortality random effects were $\sigma_M \in \{0.1, 0.5\}$ and AR1 autocorrelation parameters of $\rho_M \in \{0,0.9\}$.
 The full factorial combination of these process error assumptions and fishing history (2 levels) and observation error (2 levels) scenarios described above results in 16 different R+M OMs (Table \ref{M_om_table}).
@@
 However, the variance parameters for the logistic-normal distributions for age composition observations were estimated in the EMs.
+As such, EMs would either be configured completely correctly for the OM, or there could be mis-specification in assumptions of process error autocorrelation, the type of process error, or the SRR (Beverton-Holt or none).
 
@@
 
-### Convergence
+### Convergence {-}
 
@@
 We estimated the probability of selection of each process error model structure (R, R+S, R+M, R+Sel, R+q) using marginal AIC.
-For a given operating model, we compared AIC for EMs that all made the same assumptions about median natural mortality (known or estimated) and SRR (Beverton-Holt or none).
+For a given operating model, we compared AIC for EMs that were all configured the same for median natural mortality (known or estimated) and the SRR (Beverton-Holt or none).
 

### 2025-05-02 Project_0/manuscript/manuscript.Rmd
project 0 draft to submit

@@
 ---
-title: An investigation of factors affecting inferences from and reliablity of state-space age-structured assessment models
+title: An investigation of factors affecting inferences from and reliability of state-space age-structured assessment models
 author: 
@@
 Bias was also low under many incorrect process error assumptions when there was contrast in fishing pressure and low observation error.
-Marginal AIC most accurately distinguished process errors on recruitment, survival, and selectivity, as well as larger magnitude process errors of other types.
+Marginal AIC most accurately distinguished process errors on recruitment, survival, and selectivity, and other process error sources when variability was greater.
 Retrospective patterns were generally small but were sizable for recruitment when observation error was high.
-These results help establish the statistical reliability of state space assessment models and pave the way for the next-generation of fisheries stock assessment
-
-\pagebreak
+These results help establish the statistical reliability of state space assessment models and pave the way for the next-generation of fisheries stock assessment.
 
@@
 We present results for bias in terminal year estimates of SSB and recruitment, Beverton-Holt stock recruit parameters ($a$ and $b$), and median natural mortality rate.
-Results for terminal year fishing mortality are were strongly negatively correlated with those for SSB and are provided in the Supplementary Materials.
+Results for terminal year fishing mortality were strongly negatively correlated with those for SSB and are provided in the Supplementary Materials.
 
@@
 
-For R and R+S OMs, convergence generally declined for most EMs when the median natural mortality rate was estimated and/or the Beverton-Holt SRR was estimated even when the process error assumptions of the EMs and OMs matched (Figure \ref{hessian_SE_convergence}, A).
+For many R and R+S OMs, convergence rate declined when either the median natural mortality rate or the Beverton-Holt SRR was estimated even when the process error assumptions of the EMs and OMs matched (Figure \ref{hessian_SE_convergence}, A).
 When there was high observation error and constant fishing pressure ($F = F_{\text{MSY}}$ for all 40 years), convergence was poor for of all EM process error configurations other than R EMs when fitted to R OMs ($\sigma_{2+} = 0$) regardless of whether median natural mortality and SRRs were estimated.
@@
 
-Across all OMs, there was generally better accuracy (less bias and/or lower variability) for estimation of the Beverton Holt $a$ parameter than the $b$ parameter.
+Across all OMs, there was generally less bias and(or) lower variability in estimation of the Beverton Holt $a$ parameter than the $b$ parameter.
 In R and R+S OMs, EMs with the correct assumptions about process errors provided the least biased estimation of Beverton-Holt SRR parameters when there was a change in fishing pressure over time and lower variability of recruitment process errors, but there was little effect of estimating median natural mortality and a small increase in bias for those OMs that had high observation error (Figure \ref{SR_rel_error}, A).
@@
 Bias for SRR parameters was large and variability in relative errors was greatest for most EMs fit to R+Sel OMs with constant fishing pressure (Figure \ref{SR_rel_error}, C).
-Less bias in parameter estimation occurred for OMs with a change in fishing pressure and the best accuracy occurred for those OMs that had low observation error and more variable and uncorrelated selectivity process errors, and when the EMs hadd with the correct process error assumption.
+Less bias in parameter estimation occurred for OMs with a change in fishing pressure and the best accuracy occurred for those OMs that had low observation error and more variable and uncorrelated selectivity process errors, and when the EMs had with the correct process error assumption.
 There was little effect of estimating natural mortality on relative errors for SRR parameters.
@@
 Most EM process error types produced biased estimation of median natural  mortality when R+q OMs had high observaiton error and constant fishing pressure.
-
-Like R+Sel OMs, all EM process error types produced accurate estimation of median natural mortality rate when fit to R+q OMs with contrast in fishing pressure, low observation error and low catchability process error variability (Figure \ref{M_rel_error}, D).
-Most EM process error types produced biased estimation of median natural mortality when R+q OMs had high observation error and constant fishing pressure.
 
@@
 Often poor convergence result when parameter estimates are at their bounds [@carvalhoetal21], and this also applies to variance parameters for random effects with state-space assessment models.
-Even when Hessian is invertible, parameters that are poorly informed will have extremely large variance estimates.
+Even when the Hessian is invertible, parameters that are poorly informed will have extremely large variance estimates.
 This further inspection can lead to a more appropriate and often more parsimonious model configuration where the problematic parameters are not estimated.
@@
 removing a source of process error and refitting the model when process error variance parameters were estimated close to 0).
-Evaluating the efficacy of such a decision process when applying Ems might be important in closed loop simulations (e.g.
+Evaluating the efficacy of such a decision process when applying EMs might be important in closed loop simulations (e.g.
 MSE) aimed at quantifying management performance.
@@
 Furthermore, numerical methods are used to calculate and invert the Hessian for variance estimation for models with random effects.
-Along with our results, the potential lack of accuracy imposed by these approximations, suggests at least investigating whether the Hessian is positive definite when the calculated absolute gradients are not terribly large.
+Along with our results, the potential lack of accuracy imposed by these approximations, suggests at least investigating whether the Hessian is positive definite when the calculated absolute gradients are not terribly large (e.g, $< 1$).
 
@@
 Fitting models to other OMs rarely preferred R+S EMs, and R and R+S EMs were nearly always selected for the matching OMs; a similar result was reported by @liljestrandetal24.
-For other sources of process error, accuracy of AIC was improved to useful levels when there was larger variability in the process errors and/or lower observation error.
+For other sources of process error, accuracy of AIC was improved when there was larger variability in the process errors and/or lower observation error.
 
@@
 Although we did not compare models with alternative SRRs (e.g., Ricker and Beverton-Holt), we do not expect AIC to perform any better distinguishing between relationships.
-Our finding that AIC tended to choose simpler recruitment models in most cases contrasts with the noted bias in AIC for more complex models [@shibata76;@katz81;@kassraftery95], but, whereas those findings apply to more the much more common comparison of models that are fit to raw and independent observations, here we are comparing state-space models which account for observation error and estimate process errors in latent variables.
+Our finding that AIC tended to choose simpler recruitment models in most cases contrasts with the noted bias in AIC for more complex models [@shibata76;@katz81;@kassraftery95], but, whereas those findings apply to the much more common comparison of models that are fit to raw and independent observations, here we are comparing state-space models which account for observation error and estimate process errors in latent variables.
 
@@
 Estimation of SRR parameters was reliable in ideal scenarios of low observation error and contrast in fishing for some R+Sel and R+M OMs, but generally estimation was biased and(or) highly variable.
-We found substantial bias in estimated SRR parameters in R and R+S OMs particularly with high variability in recruitment and apparent survival process errors, suggesting that practitioners should be cautious when models without the SRR suggest this to be the case.
+We found substantial bias in estimated SRR parameters in R and R+S OMs particularly with high variability in recruitment and apparent survival process errors, suggesting that practitioners should be cautious when fitted assessment models have these properties.
 
-On the other hand, estimation of median natural mortality was reliable in many OM scenarios with contrast in fishing pressure and in some OMs EMs that also estimated the SRR parameters bias for those parameters was improved.
+On the other hand, estimation of median natural mortality was reliable in many OM scenarios with contrast in fishing pressure, consistent with @hoenigetal25.
+In some OMs, when EMs estimated the SRR parameters and median natural mortality, bias for those parameters was improved.
 Conversely, for some R+Sel and R+q OMs where there was bias in natural mortality due to high observation error, estimating the SRR reduced the bias in median natural mortality rate.
@@
 Our simulation study examined the importance of several factors for reliable inferences from state-space age-structured assessment models.
-Across all factors, AIC accurately distinguished models with process errors on recruitment only or on recruitment and apparent survival.
+Contrast in fishing pressure was consistently an important factor across all measures of reliability we examined.
+AIC accurately distinguished models with process errors on recruitment only (R) or on recruitment and apparent survival (R+S).
 Accuracy for other process error types required a strong signal (high process variability) with low noise (low observation uncertainty).
 Therefore, we expect practitioners will find R+S configurations to provide satisfactory diagnostics across a range of life history and data quality scenarios.
-Contrast in fishing pressure was consistently an important factor across all measures of reliability we examined.
-AIC generally performed poorly for selecting the SRR but performance was improved with low recruitment variability and contrast in fishing pressure.
-However, some bias in estimation in at least one of the SRR parameters existed in nearly all OM-EM combinations.
+AIC generally performed poorly for selecting the SRR, but performance was improved with low recruitment variability and contrast in fishing pressure.
+Some bias in estimation in at least one of the SRR parameters existed in nearly all OM-EM combinations.
 Because bias in terminal SSB and retrospective patterns were indifferent to whether or not the SRR was estimated, and convergence was slightly better without the SRR, a sensible default would be to fit models without an assumed SRR.
@@
 \caption{Median relative error of terminal year SSB for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Diamond shaped points denote results where the EM process error assumption matches that of the operating model.
 Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{SSB_rel_error}

### 2025-05-02 Project_0/manuscript/manuscript.tex
saving work on project 0 edits
project 0 draft to submit

@@
 \hypersetup{
-  pdftitle={Factors affecting reliablity of state-space age-structured assessment models},
+  pdftitle={An investigation of factors affecting inferences from and reliability of state-space age-structured assessment models},
   pdfauthor={Timothy J. Miller1,2; Greg Britten3; Elizabeth N. Brooks2; Gavin Fay4; Alex Hansell2; Christopher M. Legault2; Chengxue Li2; Brandon Muffley5; Brian C. Stock6; John Wiedenmann7},
@@
 
-\title{Factors affecting reliablity of state-space age-structured
-assessment models}
+\title{An investigation of factors affecting inferences from and
+reliability of state-space age-structured assessment models}
 \author{Timothy J. Miller\textsuperscript{1,2} \and Greg
@@
 C. Stock\textsuperscript{6} \and John Wiedenmann\textsuperscript{7}}
-\date{08 April, 2025}
+\date{02 May, 2025}
 
@@
 Water Street, Woods Hole, MA 02543 USA\\
-\(^3\)Woods Hole Oceanographic Institution\\
-\(^4\)SMAST\\
+\(^3\)Biology Department, Woods Hole Oceanographic Institution, 266
+Woods Hole Rd. Woods Hole, MA, USA\\
+\(^4\)Department of Fisheries Oceanography, School for Marine Science
+and Technology, University of Massachusetts Dartmouth, 836 S Rodney
+French Boulevard, New Bedford, MA 02740, USA\\
 \(^5\)Mid-Atlantic Fishery Management Council, 800 North State Street,
@@
 and low observation error. Marginal AIC most accurately distinguished
-process errors on recruitment, survival, and selectivity, as well as
-larger magnitude process errors of other types. Retrospective patterns
-were generally small but were sizable for recruitment when observation
-error was high. These results help establish the statistical reliability
-of state space assessment models and pave the way for the
-next-generation of fisheries stock assessment
-
-\pagebreak
+process errors on recruitment, survival, and selectivity, and other
+process error sources when variability was greater. Retrospective
+patterns were generally small but were sizable for recruitment when
+observation error was high. These results help establish the statistical
+reliability of state space assessment models and pave the way for the
+next-generation of fisheries stock assessment.
 
@@
 modeling stock-recruit relationships (SRRs) as mechanistic drivers of
-population dynamics is promising (Fleischman et al. 2013; Du Pontavice
-et al. 2022), but can reliability of inferences within integrated
-state-space age-structured models has not been evaluated. Evidence from
-other studies suggests that when both process and observation errors are
+population dynamics is promising (Fleischman et al. 2013; Pontavice et
+al. 2022), but reliability of inferences within integrated state-space
+age-structured models has not been evaluated. Evidence from other
+studies suggests that when both process and observation errors are
 unknown, estimating density dependence parameters becomes highly
@@
 can be estimated with sufficient precision for use in fisheries
-management (Auger-Méthé et al. 2016).
+management (Auger-Méthé et al. 2016). Finally, although the importance
+of autocorrelation in process errors is recognized, investigations of
+the ability to distinguish state-space assessment models with and
+without autocorrelation and whether such misspecification is detrimental
+to estimation of important population metrics are lacking (Johnson et
+al. 2016; Xu et al. 2019).
 
@@
 assumptions for sources of process error, whether a SRR was estimated,
-and whether a constant, or, in some EMs, median, natural mortality is
-estimated. Given the confounding nature of process errors, developing
-diagnostic tools to detect model misspecification is of great scientific
-interest and could aid the next generation of stock assessments
-(Auger-Méthé et al. 2021). We evaluate whether convergence and Akaike
-Information Criterion (AIC) can correctly determine the source of
-process error and the existence of a SRR. We also evaluate when
-retrospective patterns occur and the degree of bias in the outputs of
-the assessment model that are important for management.
+and whether natural mortality is estimated. Given the confounding nature
+of process errors, developing diagnostic tools to detect model
+misspecification is of great scientific interest and could aid the next
+generation of stock assessments (Auger-Méthé et al. 2021). We evaluate
+whether convergence and Akaike Information Criterion (AIC) can correctly
+determine the source of process error and the existence of a SRR. We
+also evaluate when retrospective patterns occur and the degree of bias
+in the outputs of the assessment model that are important for
+management.
 
@@
 methods (Legault et al. 2023) and is currently used or accepted for use
-in management of numerous fish stocks in the Northeast United States
-(NEUS) (e.g., NEFSC 2022a, 2022b; NEFSC 2024).
+in management of numerous NEUS fish stocks (e.g., NEFSC 2022a, 2022b;
+NEFSC 2024).
 
@@
 variance of process errors and observations are required and the values
-we used were based on a review of the range of estimates from NEUS
-assessments using WHAM in stocks NEUS.
+we used were based on a review of the range of estimates from Northeast
+Unite States (NEUS) assessments using WHAM.
 
@@
 or observations described by (Stock and Miller 2021) for OMs and EMs to
-simplify interpretation of the study results (Li et al. In review).
-Simulations and model fitting were all carried out on the University of
-Massachusetts Green High-Performance Computing Cluster. All code we used
-to perform the simulation study and summarize results can be found at
+simplify interpretation of the study results (Li et al. In review). All
+code we used to perform the simulation study and summarize results can
+be found at
 \url{https://github.com/timjmiller/SSRTWG/tree/main/Project_0/code}.
@@
 We assumed a Beverton-Holt SRR with constant pre-recruit mortality
-parameters for all OMs. All post-recruit productivity components are
-constant in the apparent survival (R+S) and survey catchability (R+Sel)
-process error OMs. Therefore, steepness and unfished recruitment are
-also constant over the time period for those OMs (Miller and Brooks
-2021). We specified unfished recruitment equal to \(e^{10}\) and
+parameters for all OMs. All biological inputs to calculations of
+spawning biomass per recruit (i.e., weight, maturity, and natural
+mortality at age) are constant in the apparent survival (R+S)
+selectivity (R+Sel), and survey catchability (R+q) process error OMs.
+Therefore, steepness and unfished recruitment are also constant over the
+time period for those OMs (Miller and Brooks 2021). We specified
+unfished recruitment equal to \(e^{10}\) and
 \(\Fmsy = F_{40\%} = 0.348\), which equates to a steepness of 0.69 and
@@
 N_{1,y} = \frac{a \text{SSB}_{y-1}}{1 + b \text{SSB}_{y-1}} 
-\] (Figure \ref{om_inputs_fig}, bottom right). For OMs without process
-errors on natural mortality we fixed the rate at 0.2. For OMs with
-process errors on natural mortality, the median natural mortality rate
-was specified to be 0.2.
+\] (Figure \ref{om_inputs_fig}, bottom right). We assumed a value of 0.2
+for the natural mortality rate in OMs without process errors on natural
+mortaliyt and for the median rate for OMs with process errors on natural
+mortality.
 
@@
 annual random effects are marginally distributed as
-\(\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)\). Uncorrelated
-random effects were also included on recruitment with \(\sigma_R = 0.5\)
-(hence, we denote these OMs as R+M). The marginal standard deviations we
-assumed for log natural mortality random effects were
-\(\sigma_M \in \{0.1, 0.5\}\) and AR1 autocorrelation parameters of
-\(\rho_M \in \{0,0.9\}\). The full factorial combination of these
-process error assumptions and fishing history (2 levels) and observation
-error (2 levels) scenarios described above results in 16 different R+M
-OMs (Table \ref{M_om_table}).
+\(\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)\). The marginal
+standard deviations we assumed for log natural mortality random effects
+were \(\sigma_M \in \{0.1, 0.5\}\) and the random effects were either
+uncorrelated or first-order autoregressive (AR1,
+\(\rho_M \in \{0,0.9\}\)). Uncorrelated random effects were also
+included on recruitment with \(\sigma_R = 0.5\) (hence, we denote these
+OMs as R+M). The full factorial combination of these process error
+assumptions and fishing history (2 levels) and observation error (2
+levels) scenarios described above results in 16 different R+M OMs (Table
+\ref{M_om_table}).
 
@@
 distributions for age composition observations were estimated in the
-EMs.
+EMs. As such, EMs would either be configured completely correctly for
+the OM, or there could be mis-specification in assumptions of process
+error autocorrelation, the type of process error, or the SRR
+(Beverton-Holt or none).
 
@@
 \hypertarget{convergence}{%
-\subsubsection{Convergence}\label{convergence}}
+\subsubsection*{Convergence}\label{convergence}}
+\addcontentsline{toc}{subsubsection}{Convergence}
 
@@
 structure (R, R+S, R+M, R+Sel, R+q) using marginal AIC. For a given
-operating model, we compared AIC for EMs that all made the same
-assumptions about median natural mortality (known or estimated) and SRR
+operating model, we compared AIC for EMs that were all configured the
+same for median natural mortality (known or estimated) and the SRR
 (Beverton-Holt or none).
@@
 Beverton-Holt stock recruit parameters (\(a\) and \(b\)), and median
-natural mortality rate. Results for terminal year fishing mortality are
-were strongly negatively correlated with those for SSB and are provided
-in the Supplementary Materials.
+natural mortality rate. Results for terminal year fishing mortality were
+strongly negatively correlated with those for SSB and are provided in
+the Supplementary Materials.
 
@@
 
-For R and R+S OMs, convergence generally declined for most EMs when the
-median natural mortality rate was estimated and/or the Beverton-Holt SRR
-was estimated even when the process error assumptions of the EMs and OMs
-matched (Figure \ref{hessian_SE_convergence}, A). When there was high
-observation error and constant fishing pressure (\(F = F_{\text{MSY}}\)
-for all 40 years), convergence was poor for of all EM process error
-configurations other than R EMs when fitted to R OMs
-(\(\sigma_{2+} = 0\)) regardless of whether median natural mortality and
-SRRs were estimated. Convergence of R EMs was high for all R and R+S OMs
-except when there was high observation error and constant fishing
-pressure, and when median natural mortality and SRRs were estimated. R+S
-EMs fit to R OMs exhibited poor convergence regardless of whether
-natural mortality or a SRR was estimated. R+S EMs fit to R+S OMs had
-highest convergence rates when there was contrast in fishing pressure
-and low observation error. Convergence rates were high for all EMs when
-fit to data from R+S OMs with lower observation error except those where
-median natural mortality and/or SRRs were estimated.
+For many R and R+S OMs, convergence rate declined when either the median
+natural mortality rate or the Beverton-Holt SRR was estimated even when
+the process error assumptions of the EMs and OMs matched (Figure
+\ref{hessian_SE_convergence}, A). When there was high observation error
+and constant fishing pressure (\(F = F_{\text{MSY}}\) for all 40 years),
+convergence was poor for of all EM process error configurations other
+than R EMs when fitted to R OMs (\(\sigma_{2+} = 0\)) regardless of
+whether median natural mortality and SRRs were estimated. Convergence of
+R EMs was high for all R and R+S OMs except when there was high
+observation error and constant fishing pressure, and when median natural
+mortality and SRRs were estimated. R+S EMs fit to R OMs exhibited poor
+convergence regardless of whether natural mortality or a SRR was
+estimated. R+S EMs fit to R+S OMs had highest convergence rates when
+there was contrast in fishing pressure and low observation error.
+Convergence rates were high for all EMs when fit to data from R+S OMs
+with lower observation error except those where median natural mortality
+and/or SRRs were estimated.
 
@@
 
-Across all OMs, there was generally better accuracy (less bias and/or
-lower variability) for estimation of the Beverton Holt \(a\) parameter
-than the \(b\) parameter. In R and R+S OMs, EMs with the correct
-assumptions about process errors provided the least biased estimation of
-Beverton-Holt SRR parameters when there was a change in fishing pressure
-over time and lower variability of recruitment process errors, but there
-was little effect of estimating median natural mortality and a small
-increase in bias for those OMs that had high observation error (Figure
+Across all OMs, there was generally less bias and(or) lower variability
+in estimation of the Beverton Holt \(a\) parameter than the \(b\)
+parameter. In R and R+S OMs, EMs with the correct assumptions about
+process errors provided the least biased estimation of Beverton-Holt SRR
+parameters when there was a change in fishing pressure over time and
+lower variability of recruitment process errors, but there was little
+effect of estimating median natural mortality and a small increase in
+bias for those OMs that had high observation error (Figure
 \ref{SR_rel_error}, A). For other R and R+S OMs, estimating natural
@@
 occurred for those OMs that had low observation error and more variable
-and uncorrelated selectivity process errors, and when the EMs hadd with
+and uncorrelated selectivity process errors, and when the EMs had with
 the correct process error assumption. There was little effect of
@@
 OMs had high observaiton error and constant fishing pressure.
-
-Like R+Sel OMs, all EM process error types produced accurate estimation
-of median natural mortality rate when fit to R+q OMs with contrast in
-fishing pressure, low observation error and low catchability process
-error variability (Figure \ref{M_rel_error}, D). Most EM process error
-types produced biased estimation of median natural mortality when R+q
-OMs had high observation error and constant fishing pressure.
 
@@
 parameters for random effects with state-space assessment models. Even
-when Hessian is invertible, parameters that are poorly informed will
+when the Hessian is invertible, parameters that are poorly informed will
 have extremely large variance estimates. This further inspection can
@@
 variance parameters were estimated close to 0). Evaluating the efficacy
-of such a decision process when applying Ems might be important in
+of such a decision process when applying EMs might be important in
 closed loop simulations (e.g.~MSE) aimed at quantifying management
@@
 least investigating whether the Hessian is positive definite when the
-calculated absolute gradients are not terribly large.
+calculated absolute gradients are not terribly large (e.g, \(< 1\)).
 
@@
 (2024). For other sources of process error, accuracy of AIC was improved
-to useful levels when there was larger variability in the process errors
-and/or lower observation error.
+when there was larger variability in the process errors and/or lower
+observation error.
 
@@
 bias in AIC for more complex models (Shibata 1976; Katz 1981; Kass and
-Raftery 1995), but, whereas those findings apply to more the much more
-common comparison of models that are fit to raw and independent
-observations, here we are comparing state-space models which account for
-observation error and estimate process errors in latent variables.
+Raftery 1995), but, whereas those findings apply to the much more common
+comparison of models that are fit to raw and independent observations,
+here we are comparing state-space models which account for observation
+error and estimate process errors in latent variables.
 
@@
 just to SSB and recruitment observations (i.e., not an assessment
-model). Similarly, Britten et al. found AIC could not reliably
-distinguish alternative environmental effects on SRR parameters.
-However, Miller et al. (2016) did find AIC to prefer a SRR with
-environmental effects when applied to data for the SNEMA yellowtail
+model). Similarly, Britten et al. (In review) found AIC could not
+reliably distinguish alternative environmental effects on SRR
+parameters. However, Miller et al. (2016) did find AIC to prefer a SRR
+with environmental effects when applied to data for the SNEMA yellowtail
 flounder stock and AIC also selected an environmental covariate on a SRR
@@
 recruitment and apparent survival process errors, suggesting that
-practitioners should be cautious when models without the SRR suggest
-this to be the case.
+practitioners should be cautious when fitted assessment models have
+these properties.
 
 On the other hand, estimation of median natural mortality was reliable
-in many OM scenarios with contrast in fishing pressure and in some OMs
-EMs that also estimated the SRR parameters bias for those parameters was
-improved. Conversely, for some R+Sel and R+q OMs where there was bias in
-natural mortality due to high observation error, estimating the SRR
-reduced the bias in median natural mortality rate. However, estimating
-median natural mortality did cause poor accuracy in SSB estimation in
-many OMs without contrast in fishing pressure over time and with higher
+in many OM scenarios with contrast in fishing pressure, consistent with
+Hoenig et al. (2025). In some OMs, when EMs estimated the SRR parameters
+and median natural mortality, bias for those parameters was improved.
+Conversely, for some R+Sel and R+q OMs where there was bias in natural
+mortality due to high observation error, estimating the SRR reduced the
+bias in median natural mortality rate. However, estimating median
+natural mortality did cause poor accuracy in SSB estimation in many OMs
+without contrast in fishing pressure over time and with higher
 observation error. Thus, estimating median natural mortality should be
@@
 reliable inferences from state-space age-structured assessment models.
-Across all factors, AIC accurately distinguished models with process
-errors on recruitment only or on recruitment and apparent survival.
-Accuracy for other process error types required a strong signal (high
-process variability) with low noise (low observation uncertainty).
-Therefore, we expect practitioners will find R+S configurations to
-provide satisfactory diagnostics across a range of life history and data
-quality scenarios. Contrast in fishing pressure was consistently an
-important factor across all measures of reliability we examined. AIC
-generally performed poorly for selecting the SRR but performance was
-improved with low recruitment variability and contrast in fishing
-pressure. However, some bias in estimation in at least one of the SRR
-parameters existed in nearly all OM-EM combinations. Because bias in
-terminal SSB and retrospective patterns were indifferent to whether or
-not the SRR was estimated, and convergence was slightly better without
-the SRR, a sensible default would be to fit models without an assumed
-SRR.
+Contrast in fishing pressure was consistently an important factor across
+all measures of reliability we examined. AIC accurately distinguished
+models with process errors on recruitment only (R) or on recruitment and
+apparent survival (R+S). Accuracy for other process error types required
+a strong signal (high process variability) with low noise (low
+observation uncertainty). Therefore, we expect practitioners will find
+R+S configurations to provide satisfactory diagnostics across a range of
+life history and data quality scenarios. AIC generally performed poorly
+for selecting the SRR, but performance was improved with low recruitment
+variability and contrast in fishing pressure. Some bias in estimation in
+at least one of the SRR parameters existed in nearly all OM-EM
+combinations. Because bias in terminal SSB and retrospective patterns
+were indifferent to whether or not the SRR was estimated, and
+convergence was slightly better without the SRR, a sensible default
+would be to fit models without an assumed SRR.
 
@@
 M.A., Jonsen, I.D., and Mills Flemming, J. 2016. State-space models'
-dirty little secrets: Even simple linear gaussian models can have
+dirty little secrets: Even simple linear {G}aussian models can have
 estimation problems. Scientific reports \textbf{6}(1): 26677.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-brittenetalinreview}{}}%
-Britten, G., Brooks, E.N., and Miller, T.J. Reliability of
-stock-recruitment function estimation in state space assessment models.
+Britten, G., Brooks, E.N., and Miller, T.J. In review. Identification
+and performance of environmentally-driven stock-recruitment
+relationships in state space assessment models. Canadian Journal of
+Fisheries and Aquatic Sciences.
 
@@
 Brooks, E.N., and Legault, C.M. 2016. Retrospective forecasting --
-evaluating performance of stock projections for new england groundfish
-stocks. Canadian Journal of Fisheries and Aquatic Sciences
+evaluating performance of stock projections for {N}ew {E}ngland
+groundfish stocks. Canadian Journal of Fisheries and Aquatic Sciences
 \textbf{73}(6): 935--950.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-dupontaviceetal22}{}}%
-Du Pontavice, H., Miller, T.J., Stock, B.C., Chen, Z., and Saba, V.S.
-2022. Ocean model-based covariates improve a marine fish stock
-assessment when observations are limited. ICES Journal of Marine Science
-\textbf{79}(4): 1259--1273.
-doi:\href{https://doi.org/10.1093/icesjms/fsac050}{10.1093/icesjms/fsac050}.
-
 \leavevmode\vadjust pre{\hypertarget{ref-fischetal23}{}}%
@@
 specification. ICES Journal of Marine Science \textbf{80}(8):
-2125--2149. Oxford University Press.
+2125--2149.
+doi:\href{https://doi.org/10.1093/icesjms/fsad138}{10.1093/icesjms/fsad138}.
 
@@
 Fleischman, S.J., Catalano, M.J., Clark, R.A., and Bernard, D.R. 2013.
-An age-structured state-space stock--recruit model for pacific salmon
-(oncorhynchus spp.). Canadian Journal of Fisheries and Aquatic Sciences
-\textbf{70}(3): 401--414. NRC Research Press.
+An age-structured state-space stock--recruit model for {P}acific salmon
+(\emph{{O}ncorhynchus spp}.). Canadian Journal of Fisheries and Aquatic
+Sciences \textbf{70}(3): 401--414.
+doi:\href{https://doi.org/10.1139/cjfas-2012-0112}{10.1139/cjfas-2012-0112}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hoenigetal25}{}}%
+Hoenig, J.M., Hearn, W.S., Leigh, G.M., and Latour, R.J. 2025.
+Principles for estimating natural mortality rate. Fisheries Research
+\textbf{281}: 107195.
+doi:\href{https://doi.org/10.1016/j.fishres.2024.107195}{10.1016/j.fishres.2024.107195}.
 
@@
 Hoyle, S.D., Maunder, M.N., Punt, A.E., Mace, P.M., Devine, J.A., and
-A'mar, Z.T. 2022. Preface: Developing the next generation of stock
+A???mar, Z.T. 2022. Preface: Developing the next generation of stock
 assessment software. Fisheries Research \textbf{246}: 106176.
+doi:\href{https://doi.org/10.1016/j.fishres.2021.106176}{10.1016/j.fishres.2021.106176}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-johnsonetal16}{}}%
+Johnson, K.F., Councill, E., Thorson, J.T., Brooks, E., Methot, R.D.,
+and Punt, A.E. 2016. Can autocorrelated recruitment be estimated using
+integrated assessment models and how does it affect population
+forecasts? Fisheries Research \textbf{183}: 222--232.
+doi:\href{https://doi.org/10.1016/j.fishres.2016.06.004}{10.1016/j.fishres.2016.06.004}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-kapuretal24}{}}%
@@
 Statistical Software \textbf{70}(5): 1--21.
+doi:\href{https://doi.org/10.18637/jss.v070.i05}{10.18637/jss.v070.i05}.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-legault09}{}}%
-Legault, C.M. 2009. Report of the retrospective working group, 14--16
+Legault, C.M. 2009. Report of the retrospective working group, 14-16
 january 2008. US Department of Commerce Northeast Fisheries Science
@@
 an assessment? Canadian Journal of Fisheries and Aquatic Sciences
-\textbf{73}(3): 349--357. NRC Research Press.
+\textbf{73}(3): 349--357.
+doi:\href{https://doi.org/10.1139/cjfas-2015-0232}{10.1139/cjfas-2015-0232}.
 
@@
 Schueller, A.M., and Miller, T.J. In review. Random effects on
-numbers-at-age transitions implictly 2 account for movement dynamics and
-improve 3 performance within a state-space stock assessment. Canadian
+numbers-at-age transitions implictly account for movement dynamics and
+improve performance within a state-space stock assessment. Canadian
 Journal of Fisheries and Aquatic Sciences.
@@
 Li, C., Deroba, J.J., Miller, T.J., Legault, C.M., and Perretti, C. In
-reviewIn review. Guidance on bias-correction of log-normal random
-effects and observations in state-space assessment models. For Canadian
-Journal of Fisheries and Aquatic Sciences, State-space special
-collection.
+review. Guidance on bias-correction of log-normal random effects and
+observations in state-space assessment models. Canadian Journal of
+Fisheries and Aquatic Sciences.
 
@@
 stock assessment model. Fisheries Research \textbf{275}: 107023.
-Elsevier.
+doi:\href{https://doi.org/10.1016/j.fishres.2024.107023}{10.1016/j.fishres.2024.107023}.
 
@@
 informative? Fish and Fisheries \textbf{8}(4): 337--358.
-doi:\url{https://doi.org/10.1111/j.1467-2979.2007.00258.x}.
+doi:\href{https://doi.org/10.1111/j.1467-2979.2007.00258.x}{10.1111/j.1467-2979.2007.00258.x}.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nefsc22a}{}}%
-NEFSC. 2022b. Report of the american plaice research track working
+NEFSC. 2022b. Report of the {A}merican plaice research track working
 group. {Available} at
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-dupontaviceetal22}{}}%
+Pontavice, H. du, Miller, T.J., Stock, B.C., Chen, Z., and Saba, V.S.
+2022. Ocean model-based covariates improve a marine fish stock
+assessment when observations are limited. ICES Journal of Marine Science
+\textbf{79}(4): 1259--1273.
+doi:\href{https://doi.org/10.1093/icesjms/fsac050}{10.1093/icesjms/fsac050}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-punt23}{}}%
@@
 106642.
+doi:\href{https://doi.org/10.1016/j.fishres.2023.106642}{10.1016/j.fishres.2023.106642}.
 
@@
 for selectivity in fisheries stock assessments. Fisheries Research
-\textbf{158}: 124--134. Elsevier.
+\textbf{158}: 124--134.
 
@@
 Shibata, R. 1976. Selection of the order of an autoregressive model by
-akaike's information criterion. Biometrika \textbf{63}(1): 117--126.
+{A}kaike's information criterion. Biometrika \textbf{63}(1): 117--126.
 doi:\href{https://doi.org/10.1093/biomet/63.1.117}{10.1093/biomet/63.1.117}.
@@
 selectivity for approaches to weighting compositional data in fisheries
-stock assessments. Fisheries Research \textbf{192}: 126--134. Elsevier.
+stock assessments. Fisheries Research \textbf{192}: 126--134.
+doi:\href{https://doi.org/10.1016/j.fishres.2016.06.018}{10.1016/j.fishres.2016.06.018}.
 
@@
 links to environmental covariates. Fisheries Research \textbf{240}:
-105967. doi:\url{https://doi.org/10.1016/j.fishres.2021.105967}.
+105967.
+doi:\href{https://doi.org/10.1016/j.fishres.2021.105967}{10.1016/j.fishres.2021.105967}.
 
@@
 natural mortality improves a state-space assessment model for Southern
-New England-Mid Atlantic yellowtail flounder}. Fisheries Research
+{N}ew {E}ngland-Mid {A}tlantic yellowtail flounder}. Fisheries Research
 \textbf{237}: 105873.
+doi:\href{https://doi.org/10.1016/j.fishres.2021.105873}{10.1016/j.fishres.2021.105873}.
 
@@
 predation and process errors in stock assessments? Journal of Applied
-Ecology \textbf{57}(1): 121--135. Wiley Online Library.
+Ecology \textbf{57}(1): 121--135.
+doi:\href{https://doi.org/10.1111/1365-2664.13515}{10.1111/1365-2664.13515}.
 
@@
 Fish stocks. Fisheries Research \textbf{209}(1): 129--142.
-doi:\url{https://doi.org/10.1016/j.fishres.2018.09.018}.
+doi:\href{https://doi.org/10.1016/j.fishres.2018.09.018}{10.1016/j.fishres.2018.09.018}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-xuetal19}{}}%
+Xu, H., Thorson, J.T., Methot, R.D., and Taylor, I.G. 2019. A new
+semi-parametric method for autocorrelated age- and time-varying
+selectivity in age-structured assessment models. Canadian Journal of
+Fisheries and Aquatic Sciences \textbf{76}(2): 268--285.
+doi:\href{https://doi.org/10.1139/cjfas-2017-0446}{10.1139/cjfas-2017-0446}.
 

### 2025-05-09 Project_0/manuscript/manuscript.Rmd
cut out unused text

@@
 
-<!--
-Similar to the AIC results, bias results only condition on whether the estimating model completed the optimization process without failure.
-We did not condition on convergence as defined by a gradient threshold or invertibility of the hessian because the optimized model can provide reliable estimation of assessment model output whether or not the model was able to estimate non-zero random effects.
-In practice, the model would be reconfigured to remove unnecessary process errors and produce otherwise equivalent parameter estimates.
--->
-
 ### Mohn's $\rho$ {-}
@@
 We fit 7 peels for each EM and calculated median 95\% confidence intervals for Mohn's $\rho$ using the same methods as that for relative bias.
-<!--Similar to the other results, retrospective results only condition on whether all of the peels of a given estimating model completed the optimization process without failure.
-We did not condition on convergence as defined by a gradient threshold or invertibility of the hessian because the optimized model can provide reliable estimation of SSB, fishing mortality rates, and recruitment whether or not the model was able to estimate non-zero random effects.
--->
 
@@
 
-<!--
-Finally, although the EMs estimated observation variance parameters for the age composition observations, they were provided the correct precision of the aggregate indices and catch which
-
-The R+q OMs only have temporal variation in catchability for one index so the other index is able to provide a consistent signal of trend.
-Incorrect assumptions about the mean selectivity model could also produce
-
-Lack of retrospective patterns for SSB coupled with biased terminal SSB estimates would indicate that the scale across the time series is biased and
--->
-
 ## Conclusions {-}
@@
 
-<!--
-All together what do results mean? E.g., when do you get good convergence and retrospective patterns and high/none bias? When do you get good convergence and no retros and no/big bias?
-
-For R+S OMs with low observation error, temporal contrast in fishing pressure, and large variation in survival process errors (3rd row), some EMs with the incorrect process error assumptions, estimate median M to be 0, and SSB is also underestimated, but Mohn's $\rho$ for SSB is better than when M is known.
-Fortunately, AIC would not choose the wrong EM process error type in these situations.
-
-When R+S is preferred AMONG THE TYPES CONSIDERED HERE, it probably means its the correct process error .
--->
-
-<!--
-## Summary {-}
-
-We conducted a simulation study where we simulated process errors and observations for 72 operating models with alternative assumptions about fishing history, degree of uncertainty in index and age composition observations, type (recruitment, survival, fishery selectivity, catchability and natural mortality), degree of variation, and correlation of process errors.
-We fit 20 different estimation models to each of 100 simulated set of observations with alternative assumptions type and correlation structure of process errors, median natural mortality was known or estimated, and a B-H SRR was assumed or not.
-
-Across simulations we summarized probability of convergence of fitted models, accuracy of marginal AIC in determining the correct process error assumption and it ability to determine the Beverton-Holt SRR, bias in annual spawning stock biomass, in estimation of natural mortality, and in SRR parameters, and severity of retrospective patterns for estimation models.
-
-Alternative measures of convergence performed differently.
-Invertible hessians and resulting standard error estimation was possible when criteria based on the gradient of the optimized log-likelihood with respect to the fixed effects parameters.
-Using hessian-based convergence, probability of convergence was best for models that assumed the correct source of process error, assumed M was known, and did not assume SRRs.
-
-Using marginal AIC provided most accurate inferences about the process errors on recruitment and survival and selectivity, in that AIC preferred EMs with assumed process errors that matched OMs, and EMs with those assumed process errors were not preferred when alternative process errors were assumed in the OMs.
-However, when the true process errors were more variable, AIC accuracy increased to a useful level.
-We found AIC more accurately determined a B-H SRR rather than the null model without a S-R relationship when there was low variability in recruitment, low variability in survival random effects, and higher variation in spawning biomass over the time series.
-When the median natural mortality rate was estimated, we found large bias and uncertainty much more likely in model output such as spawning stock biomass.
-
-Bias in spawning stock biomass estimation was generally low for estimation models that assumed the correct source of process error when there was lower observation error.
-Reliable estimation of SRR parameters only appears possible in ideal situations with lower observation errors in age composition and indices, lower variability in recruitment process errors and large contrast in spawning biomass over time.
-We found little evidence of bias for many OM process error assumptions when there was contrast in fishing pressure even when there was greater observation error although it can lead to more variable estimates of natural mortality.
-For OMs where there was bias in natural mortality due to high observation error, estimation the SRR seemed to remove the bias.
-However, estimation of natural mortality can cause large differences between the true and estimated SSB (that may be unbiased on average) when there is less contrast in fishing pressure over time and higher observation error.
-
-Retrospective patterns were generally weak for all estimation models regardless of the true source of process error, but they can be expected for recruitment even for the correct process error assumptions when observation error is high.
-When models did exhibit some retrospective pattern, estimation the median natural mortality rate tended to remove it.
--->
-
 # Acknowledgements {-}
 
-This work was funded by NOAA Fisheries Northeast Fisheries Science Center.
-
-<!--
-# CRediT authorship contribution statement {-}
-
-**Timothy J. Miller:** Conceptualization, Methodology, Writing - original draft, Formal analysis, Visualization.
- -->
+This work was funded by NOAA Fisheries Northeast Fisheries Science Center. We thank Jon Deroba for helpful comments on an earlier version of this manuscript.
 

### 2025-05-14 Ecov_study/mortality/manuscript/manuscript.Rmd
saving work on Ecov-M

@@
 output:
-  pdf_document:
-    # citation_package: natbib
+  bookdown::pdf_document2:
+    toc: no
+    citation_package: natbib
     fig_caption: yes
@@
       - options.sty
-      - journalshortcuts.tex
+      # - journalshortcuts.tex
     keep_tex: yes
@@
   # word_document: default
-csl: canadian-journal-of-fisheries-and-aquatic-sciences.csl
-#biblio-style: cjfas2.bst
+# csl: canadian-journal-of-fisheries-and-aquatic-sciences.csl
+natbiboptions: round
+biblio-style: cjfas.bst
 ---
@@
 # Introduction {-}
+
+State-space population models are now used widely for fisheries stock assessment in Europe, the United States, and Canada [@nielsenberg14; @cadigan16; @pedersenberg17; @stockmiller21].
+Because application of these methods are considered best practice and recommended for the next generation of stock assessment models [@hoyleetal22; @punt23], it is expected their use will only grow globally.
+An appeal of state-space models lies in their formulation treating latent population characteristics as statistical time series with periodic observations that also may have error due to sampling or other sources of measurement error and therefore separating these sources of biological and measurement variability.
+Through advances in computational capacity, we can use sophisticated numerical approaches to estimate model parameters as mixed effects [@thorsonminto15;@kristensenetal16].
+
+State-space stock assessment models, with non-linear functions of latent processes and numerous observation types with different probability distribution assumptions represent one of most complex classes of state-space models.
+The literature on the ways we make inferences and the effects of various factors on reliabilty of inferences from state-space assessment models is growing [@lietal24;@milleretal_inreview1;@cadiganetal].
+The importance of contrast in population size and fishing mortality and quality of data used to fit assessment models including the state-space variety is known [@magnussonhilborn07;@milleretal_inreview1].
+Furthermore, estimation of natural mortality, and even temporal variablity is possible in many scenarios [@leeetal11;johnsonetal15;@cadigan16;@millerhyun18;@milleretal_inreview1].
+
+The effects of temporal variation in recruitment via unspecified or specified environmental factors have been extensively investigated in both traditional assessment models and state space models [@myers98;@milleretal16; @haltuchpapers].
+Reliability of estimating environmental and spawning biomass effects on recruitment requires a combination of strong effects, good age composition data quality, contrast in the environmental covariate and lower recruitment variability [@brittenetal_inreview;@milleretal_inreview1].
+
+Temporal and environmental effects on growth and weight at age are also more important for short term projections and the reliability of estimation of those effects @correaetal23; @correaetalinreview.
+
+However, temporal variation in natural mortality and covariate effects on natural mortality are less studied @cadigan16.
+State-space assessment models currently used can treat the changes in cohort abundance over time as random effects and or  despite the importance of natural mortality in inferences for the size of fish populations, their productivity and projections necessary for making catch advice.
+In fact, natural mortality plays a more significant role in short term projections than recruitment due to immediate effects on older age classes that constitute spawning biomass and catch.
+
+See @millerretal_inreview1 fro relevance of project 0 results (estimability of natural mortality).
+Some difficulty in distinquishing variation in natural mortality or effects of explicit covariates when variation in M random effects or Ecov is low relative to observation error.
+
+@derisoetal08 formulated the same natural mortality model as a function of covariates and random effects as we use in WHAM.
+
+ Incorrect treatment of population attributes as temporally varying [@trijouletetal20; @liljestrandetal24] could lead to misidentification of stock status and biased population estimates, ultimately impacting fisheries management decisions [@legaultpalmer16; @szuwalskietal18; @croninpunt21].
+Furthermore, biological, fishery, and observational processes are often confounded in catch-at-age data, which may adversely affect ability to distinguish between true process variability and observational error [@puntetal14; @stewartmonnahan17; @croninpunt21; @fischetal23; @lietalinreview_a].
+
+In the present study, we conduct a simulation study with operating models (OMs) varying by degree of observation error, source and variability of process error, and fishing history.
+The simulations from these OMs are fitted with estimation models (EMs) that make alternative assumptions for sources of process error, whether a SRR was estimated, and whether natural mortality is estimated.
+Given the confounding nature of process errors, developing diagnostic tools to detect model misspecification is of great scientific interest and could aid the next generation of stock assessments [@augeretal21].
+We evaluate whether convergence and Akaike Information Criterion (AIC) can correctly determine the source of process error and the existence of a SRR.
+We also evaluate when retrospective patterns occur and the degree of bias in the outputs of the assessment model that are important for management.
+
 <!--
-Application of state-space models in fisheries stock assessment and management has expanded dramatically within ICES, Canada and the Northeast US \citep{nielsenberg14,cadigan16,stockmiller21}.
+Application of state-space models in fisheries stock assessment and management has expanded dramatically within ICES, Canada and the Northeast US [@nielsenberg14;@cadigan16;@stockmiller21].
 State-space approaches that use random effects to parameterize process errors is considered best practice and a requirement for the next generation of stock assessment models.
 
-Much is known about the reliability of state-space models that are linear or Gaussian \citep{aeberhardetal18}, but applications in fisheries management are nonlinear and typically include multiple types of observations with varying distributional assumptions.
+Much is known about the reliability of state-space models that are linear or Gaussian [@aeberhardetal18], but applications in fisheries management are nonlinear and typically include multiple types of observations with varying distributional assumptions.
 Furthermore, there is a wide range of potential random effects structures and model parameters that can be treated as random effects in assessment models.
@@
 
-Estimation of natural mortality is known to be challenging in stock assessment models (Lee et al., others). However, estimation has been shown to be reliable in some situations (miller and hyun 2018)
+Estimation of natural mortality is known to be challenging in stock assessment models (Lee et al., others). However, estimation has been shown to be reliable in some situations @millerhyun18)
 
@@
 
-All of our analyses used the Woods Hole Assessment Model (WHAM) to construct both OMs and EMs [@millerstock20;@stockmiller21;@milleretalinreview].
+All of our analyses used the Woods Hole Assessment Model (WHAM) to construct both OMs and EMs [@millerstock20;@stockmiller21;@milleretal_inreview1].
 We used [version 1.0.6.9000, commit 77bbd94](https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3) for this simulation study.
@@
 
-We did not use the log-normal bias-correction feature for process errors or observations described by [@stockmiller21] for operating and EMs [@lietalinreview].
+We did not use the log-normal bias-correction feature for process errors or observations described by @stockmiller21 for operating and EMs [@lietalinreview].
 Simulations and model fitting were all carried out on the University of Massachusetts Green High-Performance Computing Cluster.
@@
 
-We used the Woods Hole Assessment Model (WHAM) to configure operating and estimation models in our simulation study \citep[]{millerstock20,stockmiller21}.
+--------------------------- From project_0 paper
+
+We used the Woods Hole Assessment Model (WHAM) to configure OMs and EMs in our simulation study [@millerstock20;@stockmiller21].
+WHAM is an R package freely available via a github repository and is built on the Template Model Builder package [@kristensenetal16].
+For this study we used [version 1.0.6.9000, commit 77bbd94](https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3).
+WHAM has also been used to configure OMs and EMs for closed loop simulations evaluating index-based assessment methods [@legaultetal23] and is currently used or accepted for use in management of numerous NEUS fish stocks [e.g., @nefsc22;@nefsc22a;@nefsc24].
+
+We completed a simulation study with a number of OMs that can be categorized based on where process error random effects were assumed: recruitment (R, assumed present in all models), apparent survival (denoted R+S), natural mortality (R+M), fleet selectivity (R+Sel), or index catchability (R+q).
+We refer to the (R+S) OMs as modeling apparent survival because on logscale the random effects ($\epsilon_{a,y}$) are additive to the total mortality (F+M) between numbers at age, thus they modify the survival term.
+However, as @stockmiller21 note, these random effects can be due to events other than mortality, such as immigration, emigration, missreported catch, and other sources of misspecification.
+For each OM, assumptions about the magnitude of the variance of process errors and observations are required and the values we used were based on a review of the range of estimates from Northeast Unite States (NEUS) assessments using WHAM.
+
+In total, we configured 72 OMs with alternative assumptions about the source and magnitude of process errors, magnitude of observation error in indices and age composition data, and contrast in fishing pressure over time.
+We fitted 20 EMs to observations generated from each of 100 simulations where process errors were also simulated.
+Each EM differed in assumptions about the source of process errors, whether natural mortality (or the median for models with process error in natural mortality) was estimated, and whether a Beverton-Holt SRR was estimated within the EM.
+Details of each of the OMs and EMs are described below.
+
+We did not use the log-normal bias-correction feature for process errors or observations described by [@stockmiller21] for OMs and EMs to simplify interpretation of the study results [@lietalinreview].
+All code we used to perform the simulation study and summarize results can be found at https://github.com/timjmiller/SSRTWG/tree/main/Project_0/code.
+
+-------------------------
+
+We used the Woods Hole Assessment Model (WHAM) to configure operating and estimation models in our simulation study [@millerstock20;@stockmiller21].
 WHAM is an R package freely available as a github repository.
 For this study we used [version 1.0.6.9000, commit 77bbd94](https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3).
-This packages has also been used to configure operating and estimating models for closed loop simulations evaluating index-based assessment methods \citep{legaultetal23} and is used for management of haddock, butterfish, American plaice, bluefish, Atlantic cod, and black sea bass in the Northeast US.
+This packages has also been used to configure operating and estimating models for closed loop simulations evaluating index-based assessment methods [@legaultetal23] and is used for management of haddock, butterfish, American plaice, bluefish, Atlantic cod, and black sea bass in the Northeast US.
 
@@
 
-We did not use the bias-correction feature for process errors or observations described by \citet{stockmiller21} for operating and estimating models.
+We did not use the bias-correction feature for process errors or observations described by [@stockmiller21] for operating and estimating models.
 Simulations were all carried out on the University of Massachusetts Green High-Performance Computing Cluster.
@@
 All post-recruit productivity components are constant in the NAA and survey selectivity process error operating models.
-Therefore steepness and unfished recruitment are also constant over the time period for those operating models \citep{millerbrooks21}.
+Therefore steepness and unfished recruitment are also constant over the time period for those operating models [@millerbrooks21].
 We specified unfished recruitment = $R_0 = e^{10}$ and $\Fmsy = F_{40\%} = 0.348$ equated to a steepness of 0.69 and $\alpha=0.60$ and $\beta = 2.4 \times 10^{-5}$ for the
@@
 In the second scenario, the stock is fished at \Fmsy for the entire time period.
-The magnitude of the overfishing assumptions is based on average estimates of overfishing for NE groundfish stocks from \citet{wiedenmannetal19}.
-\Citet{legaultetal23} also used similar approaches to defining fishing mortality histories for operating models.
+The magnitude of the overfishing assumptions is based on average estimates of overfishing for NE groundfish stocks from @wiedenmannetal19.
+@legaultetal23 also used similar approaches to defining fishing mortality histories for operating models.
 
@@
 $$
-that is a linear combination of a mean $\beta_M$ and any annual covariate $\beta_{\text{Ecov}}$ and random effects marginally distributed as $\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)$ \citep{stockmiller21}.
+that is a linear combination of a mean $\beta_M$ and any annual covariate $\beta_{\text{Ecov}}$ and random effects marginally distributed as $\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)$ [@stockmiller21].
 For operating models with natural mortality random effects, we assume the same standard deviation $\sigma_M = 0.3$.
-This is similar to the intermediate value used for R+M OMs by @milleretalinreview.
+This is similar to the intermediate value used for R+M OMs by @milleretal_inreview1.
 
@@
 $$
-and measured bias as the median relative error and constructed 95\% confidence intervals using the binomial distribution approach as in \citet{stockmiller21}.
+and measured bias as the median relative error and constructed 95\% confidence intervals using the binomial distribution approach as in @stockmiller21.
 If the confidence interval contains zero we do not have evidence of bias from the simulation study.
@@
 
+See project 0 paper for relevance of @lietal24 and @liljestrandetal24. See @milleretal_inreview1 fro relevance of project 0 results (estimability of natural mortality)
+
+@derisoetal08 was the first to model natural mortality as a function of explicit covariate and residual random annual variation. Any relevant points?
+
+Uncertainty in estimated population attributes such as spawnign biomass can be increased considerably when natural mortality is estimated.
+This is also a consideration for estimating covariate effects on natural mortality even if the median natural mortality is assumed known.
+
 The bias in natural mortality estimation resulted in biased estimation of stock size and likely harvest rates, but it will also propagate into biological reference points and possibly stock status.
@@
 
-\bibliography{paper}
+\bibliography{manuscript}
 <!-- \bibliography{project-0-paper}
@@
 \begin{center}
-\includegraphics[width = \textwidth]{Ecov_true_obs_example.png}
+%\includegraphics[width = \textwidth]{Ecov_true_obs_example.png}
 \end{center}
@@
 \begin{center}
-\includegraphics[width = \textwidth]{M_example.png}
+%\includegraphics[width = \textwidth]{M_example.png}
 \end{center}
@@
 \begin{center}
-\includegraphics{convergence}
+%\includegraphics{convergence}
 \end{center}
@@
 \begin{center}
-\includegraphics[height = \textheight]{aic}
+%\includegraphics[height = \textheight]{aic}
 \end{center}
@@
 \begin{center}
-\includegraphics[height = \textheight]{Ecov_beta_bias}
+%\includegraphics[height = \textheight]{Ecov_beta_bias}
 \end{center}
@@
 \begin{center}
-\includegraphics[height = \textheight]{median_log_M_bias}
+%\includegraphics[height = \textheight]{median_log_M_bias}
 \end{center}
@@
 \begin{center}
-\includegraphics[height = \textheight]{terminal_year_M_bias.png}
+%\includegraphics[height = \textheight]{terminal_year_M_bias.png}
 \end{center}
@@
 \begin{center}
-\includegraphics[height = \textheight]{SSB_bias_all_PE_effect_M_fixed_beta_fixed.png}
+%\includegraphics[height = \textheight]{SSB_bias_all_PE_effect_M_fixed_beta_fixed.png}
 \end{center}
@@
 \begin{center}
-\includegraphics[height = \textheight]{SSB_bias_all_PE_effect_M_fixed_beta_estimated.png}
+%\includegraphics[height = \textheight]{SSB_bias_all_PE_effect_M_fixed_beta_estimated.png}
 \end{center}
@@
 \begin{center}
-\includegraphics[height = \textheight]{SSB_bias_all_PE_effect_M_estimated_beta_estimated.png}
+%\includegraphics[height = \textheight]{SSB_bias_all_PE_effect_M_estimated_beta_estimated.png}
 \end{center}
@@
 \begin{center}
-\includegraphics[height = \textheight]{aic_supp1}
+%\includegraphics[height = \textheight]{aic_supp1}
 \end{center}
@@
 \begin{center}
-\includegraphics[height = \textheight]{aic_supp2}
+%\includegraphics[height = \textheight]{aic_supp2}
 \end{center}
@@
 \begin{center}
-\includegraphics[height = \textheight]{aic_supp3}
+%\includegraphics[height = \textheight]{aic_supp3}
 \end{center}
@@
 \begin{center}
-\includegraphics[height = \textheight]{terminal_year_M_bias_ecov_beta_0.png}
+%\includegraphics[height = \textheight]{terminal_year_M_bias_ecov_beta_0.png}
 \end{center}
@@
 \begin{center}
-\includegraphics[height = \textheight]{F_bias_all_PE_effect_M_fixed_beta_fixed.png}
+%\includegraphics[height = \textheight]{F_bias_all_PE_effect_M_fixed_beta_fixed.png}
 \end{center}
@@
 \begin{center}
-\includegraphics[height = \textheight]{F_bias_all_PE_effect_M_fixed_beta_estimated.png}
+%\includegraphics[height = \textheight]{F_bias_all_PE_effect_M_fixed_beta_estimated.png}
 \end{center}
@@
 \begin{center}
-\includegraphics[height = \textheight]{F_bias_all_PE_effect_M_estimated_beta_estimated.png}
+%\includegraphics[height = \textheight]{F_bias_all_PE_effect_M_estimated_beta_estimated.png}
 \end{center}

### 2025-05-14 Ecov_study/mortality/manuscript/manuscript.tex
saving work on Ecov-M

--- 2024-01-25 Ecov_study/mortality/paper/paper.tex
@@
 \hypersetup{
-  pdftitle={Factors affecting inferences on natural mortality and associated environmental effects},
-  pdfauthor={Timothy J. Miller1; Andrew Applegate; Greg Britten; Elizabeth N. Brooks; Gavin Fay; Alex Hansell; Christopher M. Legault; Brandon Muffley; Brian C. Stock; John Wiedenmann},
+  pdftitle={Factors affecting inferences on natural mortality and associated environmental effects in state-space age-structured assessment models},
+  pdfauthor={Timothy J. Miller1,2; Greg Britten3; Elizabeth N. Brooks2; Gavin Fay4; Alex Hansell2; Christopher M. Legault2; Brandon Muffley5; John Wiedenmann6},
   hidelinks,
@@
 \usepackage[margin=1in]{geometry}
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
 \usepackage{graphicx}
@@
 %\onehalfspacing
-%\doublespacing
+\doublespacing
 \usepackage{lineno}
-%\linenumbers
+\linenumbers
 \usepackage[belowskip=0pt,aboveskip=0pt]{caption}
 \usepackage{relsize}
-\newcommand{\Fmsy}{\ensuremath{F_{\text{MSY}}}\xspace}
-\newcommand{\Fspr}[1]{\ensuremath{F_{\text{{#1}\%}}}\xspace}
-\newcommand{\afrb}{Alaska Fishery Research Bulletin\xspace}
-\newcommand{\ajms}{African Journal of Marine Science\xspace}
-\newcommand{\amb}{Advances in Marine Biology\xspace}
-\newcommand{\bms}{Bulletin of Marine Science\xspace}
-\newcommand{\bjssf}{Bulletin of the Japanese Society of Scientific Fisheries\xspace}
-\newcommand{\cb}{Conservation Biology\xspace}
-\newcommand{\cjfas}{Canadian Journal of Fisheries and Aquatic Sciences\xspace}
-\newcommand{\ea}{Ecological Applications\xspace}
-\newcommand{\eer}{Evolutionary Ecology Research\xspace}
-\newcommand{\elet}{Ecology Letters\xspace}
-\newcommand{\emod}{Ecological Modelling\xspace}
-\newcommand{\ebf}{Environmental Biology of Fishes\xspace}
-\newcommand{\ff}{Fish and Fisheries\xspace}
-\newcommand{\fo}{Fisheries Oceanography\xspace}
-\newcommand{\fr}{Fisheries Research\xspace}
-\newcommand{\fb}{Fishery Bulletin\xspace}
-\newcommand{\ijms}{ICES Journal of Marine Science\xspace}
-\newcommand{\iccat}{Collective Volume of Scientific Papers ICCAT\xspace}
-\newcommand{\jae}{Journal of Animal Ecology\xspace}
-\newcommand{\jai}{Journal of Applied Ichthyology\xspace}
-\newcommand{\jdc}{Journal Du Conseil International Pour L'exploration De La Mer\xspace}
-\newcommand{\jdcp}{Journal Du Conseil Permanent International Pour L'exploration De La Mer\xspace}
-\newcommand{\jembe}{Journal of Experimental Marine Biology and Ecology\xspace}
-\newcommand{\jfb}{Journal of Fish Biology\xspace}
-\newcommand{\jsr}{Journal of Sea Research\xspace}
-\newcommand{\jtb}{Journal of Theoretical Biology\xspace}
-\newcommand{\jfrbc}{Journal of the Fisheries Research Board of Canada\xspace}
-\newcommand{\jnwafs}{Journal of Northwest Atlantic Fisheries Science\xspace}
-\newcommand{\mcf}{Marine and Coastal Fisheries: Dynamics, Management, and Ecosystem Science\xspace}
-\newcommand{\mb}{Marine Biology\xspace}
-\newcommand{\meps}{Marine Ecology Progress Series\xspace}
-\newcommand{\mfr}{Marine Fisheries Review\xspace}
-\newcommand{\mpb}{Marine Pollution Bulletin\xspace}
-\newcommand{\najfm}{North American Journal of Fisheries Management\xspace}
-\newcommand{\nzjmfr}{New Zealand Journal of Marine and Freshwater Research\xspace}
-\newcommand{\pnas}{Proceedings of the National Academy of Sciences USA\xspace}
-\newcommand{\rpvrciemm}{Rapports et Proc\`es-Verbaux des R\'eunions. Conseil Internationale pour l'Exploration de la Mer\xspace}
-\newcommand{\rpvrcpiemm}{Rapports et Proc\`es-Verbaux des R\'eunions. Conseil Permanent Internationale pour l'Exploration de la Mer\xspace}
-\newcommand{\rfbf}{Reviews in Fish Biology and Fisheries\xspace}
-\newcommand{\sajms}{South African Journal of Marine Science\xspace}
-\newcommand{\tafs}{Transactions of the American Fisheries Society\xspace}
-
-\newcommand{\anzjs}{Australian \& New Zealand Journal of Statistics\xspace}
-\newcommand{\as}{Applied Statistics\xspace}
-\newcommand{\csda}{Computational Statistics \& Data Analysis\xspace}
-\newcommand{\ees}{Environmental and Ecological Statistics\xspace}
-\newcommand{\jas}{Journal of Applied Statistics\xspace}
-\newcommand{\jabes}{Journal of Agricultural, Biological, and Environmental Statistics\xspace}
-\newcommand{\jasa}{Journal of the American Statistical Association\xspace}
-\newcommand{\jrssb}{Journal of the Royal Statistical Society. Series B\xspace}
-\newcommand{\sm}{Statistics in Medicine}
-
-\usepackage{xspace}
+\usepackage{float}
+\usepackage{lscape}
+\usepackage{longtable}
+\usepackage{amsmath,rotating}
+\usepackage[scanall]{psfrag}
 \usepackage{bm}
@@
 \usepackage{graphicx}
-\usepackage{makecell}
+\usepackage{sectsty}
+\usepackage{color}
+\usepackage{fancyhdr}
+\usepackage{xspace}
+\usepackage{textcomp}
+\usepackage{upgreek}
 \renewcommand\figurename{Fig.}
@@
 \renewcommand{\arraystretch}{1.5}
-\renewcommand\theadfont{}
+%\renewcommand\theadfont{}
+
+\newcommand{\Fmsy}{\ensuremath{F_{\text{MSY}}}\xspace}
+\newcommand{\Fspr}[1]{\ensuremath{F_{\text{{#1}\%}}}\xspace}
 \usepackage{booktabs}
@@
 \fi
-\usepackage[]{natbib}
-\bibliographystyle{cjfas2.bst}
-
-\title{Factors affecting inferences on natural mortality and associated
-environmental effects}
-\author{Timothy J. Miller\textsuperscript{1} \and Andrew
-Applegate \and Greg Britten \and Elizabeth N. Brooks \and Gavin
-Fay \and Alex Hansell \and Christopher M. Legault \and Brandon
-Muffley \and Brian C. Stock \and John Wiedenmann}
-\date{}
+\usepackage[round]{natbib}
+\bibliographystyle{cjfas.bst}
+
+\title{Factors affecting inferences on natural mortality and associated environmental effects in state-space age-structured assessment models}
+\author{Timothy J.
+Miller\textsuperscript{1,2} \and Greg Britten\textsuperscript{3} \and Elizabeth N.
+Brooks\textsuperscript{2} \and Gavin Fay\textsuperscript{4} \and Alex Hansell\textsuperscript{2} \and Christopher M.
+Legault\textsuperscript{2} \and Brandon Muffley\textsuperscript{5} \and John Wiedenmann\textsuperscript{6}}
+\date{13 May, 2025}
 
@@
 
-\(^1\)corresponding author:
-\href{mailto:timothy.j.miller@noaa.gov}{\nolinkurl{timothy.j.miller@noaa.gov}},
-Northeast Fisheries Science Center, Woods Hole Laboratory, 166 Water
-Street, Woods Hole, MA 02543 USA\\
+\(^1\)corresponding author: \href{mailto:timothy.j.miller@noaa.gov}{\nolinkurl{timothy.j.miller@noaa.gov}}\\
+\(^2\)Northeast Fisheries Science Center, Woods Hole Laboratory, 166 Water Street, Woods Hole, MA 02543 USA\\
+\(^3\)Biology Department, Woods Hole Oceanographic Institution, 266 Woods Hole Rd. Woods Hole, MA, USA\\
+\(^4\)Department of Fisheries Oceanography, School for Marine Science and Technology, University of Massachusetts Dartmouth, 836 S Rodney French Boulevard, New Bedford, MA 02740, USA\\
+\(^5\)Mid-Atlantic Fishery Management Council, 800 North State Street, Suite 201, Dover, DE 19901 USA\\
+\(^6\)Department of Ecology, Evolution, and Natural Resources. Rutgers University\\
 
@@
 
-\hypertarget{summary}{%
-\subsection*{Summary}\label{summary}}
-\addcontentsline{toc}{subsection}{Summary}
-
-We completed a large scale simulation study with 288 operating models,
-each with 100 simulated data sets, and 12 estimated models fit to each
-simulated data set. The factors defining operating model configuration
-included the source of process error on the population (recruitment,
-recruitment and survival, recruitment and natural mortality), the degree
-of temporal variation and autocorrelation of the environmental
-covariate, the uncertainty in the observation of the covariate, the
-uncertainty in indices and age composition observations, fishing
-history, and the magnitude of the effect of the covariate on natural
-mortality. The estimating models make alternative assumptions on whether
-to include the environmental effect, whether the mean/intercept log
-natural mortality (log(0.2)) is estimated or known, and whether process
-errors are on just recruitment, recruitment and survival, or recruitment
-and natural mortality.
-
-We found convergence of all estimation models was generally best when
-operating models assumed process errors in recruitment and survival,
-constant fishing rate, greater contrast in the true environmental
-covariate, and lower uncertainty in corresponding observations. Reliable
-convergence of all estimating models also occurred with the same process
-errors in the operating model and a step-change in fishing, but this
-also required lower uncertainty in index and age composition
-observations. Estimating models with process errors on recruitment and
-survival were unlikely to converge when the process errors in the
-operating model did not match whereas estimating models with process
-errors in recruitment and natural mortality converged for operating
-models without this match in certain cases. Probability of convergence
-generally decreased when the mean natural mortality rate parameter was
-estimated.
-
-Whether the mean log-natural mortality was estimated or not, the best
-accuracy of AIC for model selection occurred for models with process
-errors on recruitment and survival. AIC accuracy was poor for models
-with process errors on recruitment and natural mortality. estimating the
-mean natural mortality rate had small effects on the accuracy of AIC in
-selecting the appropriate process error. Estimating the mean log-natural
-mortality resulted in a small decrease in AIC accuracy. AIC was
-conservative for determining whether the environmental covariate
-affected natural mortality. AIC was very accurate in determining no
-effect when there was no effect in the operating model, but AIC often
-ranked the null model best when there was an effect. Accuracy of AIC for
-covariate effects improved with increased effect size, increased
-temporal contrast in the covariate, and lower uncertainty in
-observations.
-
-We found no evidence of bias in estimation of environmental effects
-regardless of process error assumptions when there was low uncertainty
-in the environmental observations and large constrast in the
-environmental covariate. In most cases the relative error of the
-estimated environmental effect did not depend on the source of process
-error assumed in the estimating model. The worst bias was observed when
-OMs assumed R+S process errors, high uncertainty in covariate
-observations, low variability in the covariate, and low uncertainty in
-index and age composition observations. Simultaneously estimating the
-mean/intercept log natural mortality resulted in larger variation in the
-relative errors of the estimated environmental effect. Estimation of the
-intercept was reliable for all EM process error assumptions when the
-operating models assumed process errors on recruitment and natural
-mortality, contrast in fishing pressure over time, and lower observation
-error. Estimating the mean/intercept for log natural mortality generally
-resulted in highly variable estimates of annual natural mortality and
-spawning biomass and evidence of bias for some operating and estimation
-model assumptions about process error source. Again reliability of
-annual natural mortality estimates was generally improved with lower
-observation error uncertainty and contrast in fishing pressure.
-
-Reliable detection of covariate effects requires informative data. AIC
-preferred simpler models than the true model when information content in
-data and contrast in covariates and abundance were low. The null model
-for environmental covariate effects (no covariate effect) was selected
-when contrast in the time series was low and/or uncertainty in
-observations was high. The selection of the null model by AIC also
-likely decreases with strength of the effect of the covariate on M.
-Similarly, when there was process error in recruitment and natural
-mortality, estimation models with process error only in recruitment were
-preferred presumably due to low variation in simulated natural mortality
-process errors. Covariate effect estimation can be robust to process
-error assumptions with high contrast in covariate and low observation
-error.
+\hypertarget{abstract}{%
+\subsection*{Abstract}\label{abstract}}
+\addcontentsline{toc}{subsection}{Abstract}
+
+We completed a large scale simulation study with 288 operating models, each with 100 simulated data sets, and 12 estimated models fit to each simulated data set.
+The factors defining operating model configuration included the source of process error on the population (recruitment, recruitment and survival, recruitment and natural mortality), the degree of temporal variation and autocorrelation of the environmental covariate, the uncertainty in the observation of the covariate, the uncertainty in indices and age composition observations, fishing history, and the magnitude of the effect of the covariate on natural mortality.
+The estimating models make alternative assumptions on whether to include the environmental effect, whether the mean/intercept log natural mortality (log(0.2)) is estimated or known, and whether process errors are on just recruitment, recruitment and survival, or recruitment and natural mortality.
+
+We found convergence of all estimation models was generally best when operating models assumed process errors in recruitment and survival, constant fishing rate, greater contrast in the true environmental covariate, and lower uncertainty in corresponding observations.
+Reliable convergence of all estimating models also occurred with the same process errors in the operating model and a step-change in fishing, but this also required lower uncertainty in index and age composition observations.
+Estimating models with process errors on recruitment and survival were unlikely to converge when the process errors in the operating model did not match whereas estimating models with process errors in recruitment and natural mortality converged for operating models without this match in certain cases.
+Probability of convergence generally decreased when the mean/intercept log-natural mortality rate parameter was estimated.
+
+Whether the mean log-natural mortality was estimated or not, the best accuracy of AIC for model selection occurred for models with process errors on recruitment and survival.
+AIC accuracy was poor for models with process errors on recruitment and natural mortality.
+estimating the mean natural mortality rate had small effects on the accuracy of AIC in selecting the appropriate process error.
+Estimating the mean log-natural mortality resulted in a small decrease in AIC accuracy.
+AIC was conservative for determining whether the environmental covariate affected natural mortality.
+AIC was very accurate in determining no effect when there was no effect in the operating model, but AIC often ranked the null model best when there was an effect.
+Accuracy of AIC for covariate effects improved with increased effect size, increased temporal contrast in the covariate, and lower uncertainty in observations.
+
+We found no evidence of bias in estimation of environmental effects regardless of process error assumptions when there was low uncertainty in the environmental observations and large constrast in the environmental covariate.
+In most cases the relative error of the estimated environmental effect did not depend on the source of process error assumed in the estimating model.
+The worst bias was observed when OMs assumed R+S process errors, high uncertainty in covariate observations, low variability in the covariate, and low uncertainty in index and age composition observations.
+Simultaneously estimating the mean/intercept log natural mortality resulted in larger variation in the relative errors of the estimated environmental effect.
+Estimation of the intercept was reliable for all EM process error assumptions when the operating models assumed process errors on recruitment and natural mortality, contrast in fishing pressure over time, and lower observation error.
+Estimating the mean/intercept for log natural mortality generally resulted in highly variable estimates of annual natural mortality and spawning biomass and evidence of bias for some operating and estimation model assumptions about process error source.
+Again reliability of annual natural mortality estimates was generally improved with lower observation error uncertainty and contrast in fishing pressure.
+
+Reliable detection of covariate effects requires informative data.
+AIC preferred simpler models than the true model when information content in data and contrast in covariates and abundance were low.
+The null model for environmental covariate effects (no covariate effect) was selected when contrast in the time series was low and/or uncertainty in observations was high.
+The selection of the null model by AIC also likely decreases with strength of the effect of the covariate on M.
+Similarly, when there was process error in recruitment and natural mortality, estimation models with process error only in recruitment were preferred presumably due to low variation in simulated natural mortality process errors.
+Covariate effect estimation can be robust to process error assumptions with high contrast in covariate and low observation error.
 
@@
 
-Here we conduct a simulation study with operating models varying by
-degree of observation error uncertainty, sources of process error (M or
-survival), fishing history, temporal variation in environmental
-covariates, and magnitude of the effect of the covariate on natural
-mortality. The simulations from these operating models are fitted with
-estimating models that make alternative assumptions for sources of
-process error, and whether (mean) M is estimated. We evaluate effects of
-these factors on convergence of fitted models, whether AIC can correctly
-determine the correct source of process error and correct assumption
-about covariate effects on natural mortality, and the degree of bias in
-relevant parameters and outputs of the assessment model.
+State-space population models are now used widely for fisheries stock assessment in Europe, the United States, and Canada \citep{nielsenberg14, cadigan16, pedersenberg17, stockmiller21}.
+Because application of these methods are considered best practice and recommended for the next generation of stock assessment models \citep{hoyleetal22, punt23}, it is expected their use will only grow globally.
+An appeal of state-space models lies in their formulation treating latent population characteristics as statistical time series with periodic observations that also may have error due to sampling or other sources of measurement error and therefore separating these sources of biological and measurement variability.
+Through advances in computational capacity, we can use sophisticated numerical approaches to estimate model parameters as mixed effects \citep{thorsonminto15, kristensenetal16}.
+
+State-space stock assessment models, with non-linear functions of latent processes and numerous observation types with different probability distribution assumptions represent one of most complex classes of state-space models.
+The literature on the ways we make inferences and the effects of various factors on reliabilty of inferences from state-space assessment models is growing \citep{lietal24, milleretal_inreview1, cadiganetal}.
+The importance of contrast in population size and fishing mortality and quality of data used to fit assessment models including the state-space variety is known \citep{magnussonhilborn07, milleretal_inreview1}.
+Furthermore, estimation of natural mortality, and even temporal variablity is possible in many scenarios {[}\citet{leeetal11};\href{mailto:johnsonetal15;@cadigan16}{\nolinkurl{johnsonetal15;@cadigan16}};\citet{millerhyun18};\citet{milleretal_inreview1}{]}.
+
+The effects of temporal variation in recruitment via unspecified or specified environmental factors have been extensively investigated in both traditional assessment models and state space models \citep{myers98, milleretal16, haltuchpapers}.
+Reliability of estimating environmental and spawning biomass effects on recruitment requires a combination of strong effects, good age composition data quality, contrast in the environmental covariate and lower recruitment variability \citep{brittenetal_inreview, milleretal_inreview1}.
+
+Temporal and environmental effects on growth and weight at age are also more important for short term projections and the reliability of estimation of those effects \citet{correaetal23}; \citet{correaetalinreview}.
+
+However, temporal variation in natural mortality and covariate effects on natural mortality are less studied \citet{cadigan16}.
+State-space assessment models currently used can treat the changes in cohort abundance over time as random effects and or despite the importance of natural mortality in inferences for the size of fish populations, their productivity and projections necessary for making catch advice.
+In fact, natural mortality plays a more significant role in short term projections than recruitment due to immediate effects on older age classes that constitute spawning biomass and catch.
+
+See \citet{millerretal_inreview1} fro relevance of project 0 results (estimability of natural mortality).
+Some difficulty in distinquishing variation in natural mortality or effects of explicit covariates when variation in M random effects or Ecov is low relative to observation error.
+
+\citet{derisoetal08} formulated the same natural mortality model as a function of covariates and random effects as we use in WHAM.
+
+Incorrect treatment of population attributes as temporally varying \citep{trijouletetal20, liljestrandetal24} could lead to misidentification of stock status and biased population estimates, ultimately impacting fisheries management decisions \citep{legaultpalmer16, szuwalskietal18, croninpunt21}.
+Furthermore, biological, fishery, and observational processes are often confounded in catch-at-age data, which may adversely affect ability to distinguish between true process variability and observational error \citep{puntetal14, stewartmonnahan17, croninpunt21, fischetal23, lietalinreview_a}.
+
+In the present study, we conduct a simulation study with operating models (OMs) varying by degree of observation error, source and variability of process error, and fishing history.
+The simulations from these OMs are fitted with estimation models (EMs) that make alternative assumptions for sources of process error, whether a SRR was estimated, and whether natural mortality is estimated.
+Given the confounding nature of process errors, developing diagnostic tools to detect model misspecification is of great scientific interest and could aid the next generation of stock assessments \citep{augeretal21}.
+We evaluate whether convergence and Akaike Information Criterion (AIC) can correctly determine the source of process error and the existence of a SRR.
+We also evaluate when retrospective patterns occur and the degree of bias in the outputs of the assessment model that are important for management.
+
+Estimation of natural mortality is known to be challenging in stock assessment models (Lee et al., others). However, estimation has been shown to be reliable in some situations \citet{millerhyun18})
+
+Miller et al in review (project 0) found estimation of natural mortality was feasible when data are good and contrast in fishing pressure etc.
+
+Let's focus results here on those situations so that we can reduce the plots in figs.
+
+Variation in natural mortality has an immediate impact on projections unlike recruitment.
+Understanding this varation in M leads to better understanding of post-recruit productivity and therefore management.
+
+Here we conduct a simulation study with operating models varying by degree of observation error uncertainty, sources of process error (M or survival), fishing history, temporal variation in environmental covariates, and magnitude of the effect of the covariate on natural mortality.
+The simulations from these operating models are fitted with estimating models that make alternative assumptions for sources of process error, and whether (mean) M is estimated.
+We evaluate effects of these factors on convergence of fitted models, whether AIC can correctly determine the correct source of process error and correct assumption about covariate effects on natural mortality, and the degree of bias in relevant parameters and outputs of the assessment model.
 
@@
 
-We used the WHAM package
-\citep[][version 1.0.7, commit 77bbd94]{millerstock20,stockmiller21} to
-configure all operating models and estimation models. This packages has
-also been used to configure operating and estimating models for closed
-loop simulations evaluating index-based assessment methods
-\citep{legaultetal23} and is used for management of GB haddock,
-butterfish, plaice, bluefish, and Atlantic cod stocks in the Northeast
-US.
-
-We completed a simulation study with a number of operating models. Many
-of the characteristics of the operating models are the same as those
-used by \citet{milleretal_project0_wp}. The factors defining the
-configuration of each operating model which are described in detail in
-subsequent sections include
+All of our analyses used the Woods Hole Assessment Model (WHAM) to construct both OMs and EMs \citep{millerstock20, stockmiller21, milleretal_inreview1}.
+We used \href{https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3}{version 1.0.6.9000, commit 77bbd94} for this simulation study.
+This package has been used extensively to configure OMs and EMs for several other simulation studies \citep[\citet{lietalinreview}]{legaultetal23, lietal24} and is used to assess many commercially important stocks in the Northeast US.
+
+-------------- need to edit this part
+
+There are many similarities to the configuration of OMs and EMs in simulation studies conducted by \citet{milletetelinreview}.
+We completed a simulation study with a number of OMs that can be categorized based on where process error random effects are assumed: abundance at age (R, R+S), natural mortality (R+M), fleet selectivity (R+Sel), or index catchability (R+q).
+For each OM assumption about variance of process errors and observations are required and the values we used were based on a review of the range of estimates from applications of WHAM in management of stocks of haddock, butterfish, and American plaice in the NE US.
+
+In total, we configured 72 OMs with alternative assumptions about the source and variability of process errors, level of observation error in indices and age composition data, and contrast in fishing pressure over time.
+We fitted 20 EMs to observations from each of 100 simulations where process errors were also simulated.
+For R+M, R+Sel, and R+q OMs, we used unique seeds for each simulation, but we inadvertently used the same 100 seeds for all R and R+S OMs.
+Each of the EMs made alternative assumptions about the source of process errors and whether natural mortality (or the median for models with process error in natural mortality) was estimated and whether a Beverton-Holt stock recruit relationship was estimated within the EM.
+Details of each of the operating and EMs are described below.
+
+We did not use the log-normal bias-correction feature for process errors or observations described by \citet{stockmiller21} for operating and EMs \citep{lietalinreview}.
+Simulations and model fitting were all carried out on the University of Massachusetts Green High-Performance Computing Cluster.
+All code we used to perform the simulation study and summarize results can be found at \url{https://github.com/timjmiller/SSRTWG/tree/main/ecov_study/mortality/code}.
+
+\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
+
+--------------------------- From project\_0 paper
+
+We used the Woods Hole Assessment Model (WHAM) to configure OMs and EMs in our simulation study \citep{millerstock20, stockmiller21}.
+WHAM is an R package freely available via a github repository and is built on the Template Model Builder package \citep{kristensenetal16}.
+For this study we used \href{https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3}{version 1.0.6.9000, commit 77bbd94}.
+WHAM has also been used to configure OMs and EMs for closed loop simulations evaluating index-based assessment methods \citep{legaultetal23} and is currently used or accepted for use in management of numerous NEUS fish stocks \citep[e.g.,][]{nefsc22, nefsc22a, nefsc24}.
+
+We completed a simulation study with a number of OMs that can be categorized based on where process error random effects were assumed: recruitment (R, assumed present in all models), apparent survival (denoted R+S), natural mortality (R+M), fleet selectivity (R+Sel), or index catchability (R+q).
+We refer to the (R+S) OMs as modeling apparent survival because on logscale the random effects (\(\epsilon_{a,y}\)) are additive to the total mortality (F+M) between numbers at age, thus they modify the survival term.
+However, as \citet{stockmiller21} note, these random effects can be due to events other than mortality, such as immigration, emigration, missreported catch, and other sources of misspecification.
+For each OM, assumptions about the magnitude of the variance of process errors and observations are required and the values we used were based on a review of the range of estimates from Northeast Unite States (NEUS) assessments using WHAM.
+
+In total, we configured 72 OMs with alternative assumptions about the source and magnitude of process errors, magnitude of observation error in indices and age composition data, and contrast in fishing pressure over time.
+We fitted 20 EMs to observations generated from each of 100 simulations where process errors were also simulated.
+Each EM differed in assumptions about the source of process errors, whether natural mortality (or the median for models with process error in natural mortality) was estimated, and whether a Beverton-Holt SRR was estimated within the EM.
+Details of each of the OMs and EMs are described below.
+
+We did not use the log-normal bias-correction feature for process errors or observations described by \citep{stockmiller21} for OMs and EMs to simplify interpretation of the study results \citep{lietalinreview}.
+All code we used to perform the simulation study and summarize results can be found at \url{https://github.com/timjmiller/SSRTWG/tree/main/Project_0/code}.
+
+\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
+
+We used the Woods Hole Assessment Model (WHAM) to configure operating and estimation models in our simulation study \citep{millerstock20, stockmiller21}.
+WHAM is an R package freely available as a github repository.
+For this study we used \href{https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3}{version 1.0.6.9000, commit 77bbd94}.
+This packages has also been used to configure operating and estimating models for closed loop simulations evaluating index-based assessment methods \citep{legaultetal23} and is used for management of haddock, butterfish, American plaice, bluefish, Atlantic cod, and black sea bass in the Northeast US.
+
+We completed a simulation study with a number of operating models.
+Many of the characteristics of the operating models are the same as those used in the working paper by Miller et al.~(WP 4).
+The factors defining the configuration of each operating model which are described in detail in subsequent sections include
 
@@
 
-The full factorial of the levels of each factor results in 288 different
-operating models. We simulated 100 data sets for each operating model
-that included simulations of process errors.
-
-For each simulated data set we fit a set of 12 estimating models. The
-factors defining the configuration or each estimating model which are
-also described in detail below include
+The full factorial of the levels of each factor results in 288 different operating models. We simulated 100 data sets for each operating model that included simulations of process errors.
+
+For each simulated data set we fit a set of 12 estimating models. The factors defining the configuration or each estimating model which are also described in detail below include
 
@@
 
-We did not use the bias-correction feature for process errors or
-observations described by \citet{stockmiller21} for operating and
-estimating models. Simulations were all carried out on the University of
-Massachusetts Green High-Performance Computing Cluster. Code for
-completing the simulations and summarization of results can be found at
-github.com/timjmiller/SSRTWG/ecov\_study/mortality.
+We did not use the bias-correction feature for process errors or observations described by \citep{stockmiller21} for operating and estimating models.
+Simulations were all carried out on the University of Massachusetts Green High-Performance Computing Cluster.
+Code for completing the simulations and summarizing results can be found at \url{https://github.com/timjmiller/SSRTWG/ecov_study/mortality}.
 
@@
 
-The population model tracks 10 age classes: ages 1 to 10+ and we assume
-spawning occurs 1/4 of the way through the year. The maturity at age was
-a logistic curve with a50 = 2.89 and slope = 0.88 and assumed known in
-all estimation models (Figure \ref{om_maturity}).
-
-Weight at age was generated with a LVB growth function \[
+The population model tracks 10 age classes: ages 1 to 10+ and we assume spawning occurs 1/4 of the way through the year.
+The maturity at age was a logistic curve with \(a_{50}\) = 2.89 and slope = 0.88 and assumed known in all estimation models (Figure \ref{om_maturity}).
+
+Weight at age was generated with a LVB growth function
+\[
 L_a = L_{\infty}\left(1 - e^{-k(a - t_0)}\right)
-\] where \(t_0 = 0\), \(L_\infty = 85\), and \(k = 0.3\), and a L-W
-relationship such that \[
+\]
+where \(t_0 = 0\), \(L_\infty = 85\), and \(k = 0.3\), and a L-W relationship such that
+\[
 W_a = \theta_1 L_a^{\theta_2}
-\] where \(\theta_1 = e^{-12.1}\) and \(\theta_2 = 3.2\) (Figure
-\ref{om_waa}).
-
-We assumed a Beverton-Holt stock recruit function with constant
-pre-recruit mortality parameters (i.e., \(\alpha\) and \(\beta\)) for
-all operating models. All post-recruit productivity components are
-constant in the NAA and survey selectivity process error operating
-models. Therefore steepness and unfished recruitment are also constant
-over the time period for those operating models \citep{millerbrooks21}.
-We specified unfished recruitment = \(R_0 = e^{10}\) and
-\(\Fmsy = \Fspr[40] = 0.348\) equated to a steepness of 0.69 and
-\(\alpha=0.60\) and \(\beta = 2.4 \times 10^{-5}\) for the \[
+\]
+where \(\theta_1 = e^{-12.1}\) and \(\theta_2 = 3.2\) (Figure \ref{om_waa}).
+
+We assumed a Beverton-Holt stock recruit function with constant pre-recruit mortality parameters (i.e., \(\alpha\) and \(\beta\)) for all operating models.
+All post-recruit productivity components are constant in the NAA and survey selectivity process error operating models.
+Therefore steepness and unfished recruitment are also constant over the time period for those operating models \citep{millerbrooks21}.
+We specified unfished recruitment = \(R_0 = e^{10}\) and \(\Fmsy = F_{40\%} = 0.348\) equated to a steepness of 0.69 and \(\alpha=0.60\) and \(\beta = 2.4 \times 10^{-5}\) for the
+\[
 N_{1,y} = \frac{\alpha \text{SSB}_{y-1}}{1 + \beta \text{SSB}_{y-1}} 
-\] Beverton-Holt parameterization (Figure \ref{om_sr}). For OMs without
-process errors on natural mortality we assumed the rate was assumed 0.2.
+\]
+Beverton-Holt parameterization (Figure \ref{om_sr}).
+For OMs without process errors on natural mortality we assumed the rate was assumed 0.2.
 For OMs with process errors on natural mortality the mean rate was 0.2.
 
-Two alternative fishing histories were used for operating models. In the
-first scenario, the stock experiences overfishing for the first 20 years
-and fishing at \Fmsy for the last 20 years. In the second scenario, the
-stock is fished at \Fmsy for the entire time period. The magnitude of
-the overfishing assumptions is based on average estimates of overfishing
-for NE groundfish stocks from \citet{wiedenmannetal19}.
-\Citet{legaultetal23} also used similar approaches to defining fishing
-mortality histories for operating models.
-
-Initial population was configured at the equilibrium distribution
-fishing at either \(F = 2.5\times \Fmsy\) or \(F = \Fmsy\) for the two
-alternative fishing histories. That is for a deterministic model, the
-age composition would not change over time when the fishing mortality
-was constant at the respective level.
-
-For operating models with time-varying random effects or covariate
-effects on M, steepness is not constant, but we used the same alpha and
-beta parameters as other operating models this equates to a steepness
-and \(R_0\) at the mean of the time series process for M. For operating
-models with time-varying random effects for fishery selectivity,
-\Fmsy is also not constant however we use the same F history as other
-operating models which corresponds to \Fmsy at the mean selectivity
-parameters.
+Two alternative fishing histories were used for operating models.
+In the first scenario, the stock experiences overfishing for the first 20 years and fishing at \Fmsy for the last 20 years.
+In the second scenario, the stock is fished at \Fmsy for the entire time period.
+The magnitude of the overfishing assumptions is based on average estimates of overfishing for NE groundfish stocks from \citet{wiedenmannetal19}.
+\citet{legaultetal23} also used similar approaches to defining fishing mortality histories for operating models.
+
+Initial population was configured at the equilibrium distribution fishing at either \(F = 2.5\times \Fmsy\) or \(F = \Fmsy\) for the two alternative fishing histories.
+That is for a deterministic model, the age composition would not change over time when the fishing mortality was constant at the respective level.
+
+For operating models with time-varying random effects or covariate effects on M, steepness is not constant, but we used the same alpha and beta parameters as other operating models this equates to a steepness and \(R_0\) at the mean of the time series process for M.
+For operating models with time-varying random effects for fishery selectivity, \Fmsy is also not constant however we use the same F history as other operating models which corresponds to \Fmsy at the mean selectivity parameters.
 
@@
 
-In the WHAM model, environmental covariates are assumed to be described
-as state-space processes with annual observations of the true latent
-covariate. The latent covariate is assumed to be Gaussian AR1 \[
+In the WHAM model, environmental covariates are assumed to be described as state-space processes with annual observations of the true latent covariate.
+The latent covariate is assumed to be Gaussian AR1
+\[
 E_y|E_{y-1} \sim \text{N}\left(\mu_\text{Ecov}\left(1-\rho_\text{Ecov}\right) + \rho_\text{Ecov} E_{y-1}, \left(1-\rho_\text{Ecov}^2\right)\sigma^2_\text{Ecov}\right)
-\] with marginal mean \(\mu_\text{Ecov}=0\) and variance
-\(\sigma^2_\text{Ecov}\). The configuration of the latent environmental
-covariate in the operating models have one of two alternate assumptions
-about the marginal standard deviation and autocorrelation parameter:
+\]
+with marginal mean \(\mu_\text{Ecov}=0\) and variance \(\sigma^2_\text{Ecov}\).
+The configuration of the latent environmental covariate in the operating models have one of two alternate assumptions about the marginal standard deviation and autocorrelation parameter:
 
@@
 
-The observations of the latent environmental covariate are assumed to be
-unbiased and Gaussian \[
+The observations of the latent environmental covariate are assumed to be unbiased and Gaussian
+\[
 \text{ecov}_y|E_y \sim \text{N}\left(E_y,\sigma^2_\text{ecov}\right)
-\] The standard deviation of the environmental observations in the
-operating models is one of two values:
+\]
+The standard deviation of the environmental observations in the operating models is one of two values:
 
@@
 
-Figure \ref{om_ecov_example} provides example simulations of the latent
-process and observations under the alternative configurations.
+Figure \ref{om_ecov_example} provides example simulations of the latent process and observations under the alternative configurations.
 
 \hypertarget{random-effects-on-recruitment-survival-and-natural-mortality}{%
-\subsubsection*{Random effects on recruitment, survival, and natural
-mortality}\label{random-effects-on-recruitment-survival-and-natural-mortality}}
-\addcontentsline{toc}{subsubsection}{Random effects on recruitment,
-survival, and natural mortality}
+\subsubsection*{Random effects on recruitment, survival, and natural mortality}\label{random-effects-on-recruitment-survival-and-natural-mortality}}
+\addcontentsline{toc}{subsubsection}{Random effects on recruitment, survival, and natural mortality}
 
@@
 \log M_y = \beta_M + \beta_{\text{Ecov}} E_y + \varepsilon_{M,y}
-\] that is a linear combination of a mean \(\beta_M\) and any annual
-covariate \(\beta_{\text{Ecov}}\) and random effects marginally
-distributed as \(\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)\)
-\citep{stockmiller21}. For operating models with natural mortality
-random effects, we assume the same standard deviation
-\(\sigma_M = 0.3\).
-
-The covariate effect is one of 3 alternative values in the operating
-models:
+\]
+that is a linear combination of a mean \(\beta_M\) and any annual covariate \(\beta_{\text{Ecov}}\) and random effects marginally distributed as \(\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)\) \citep{stockmiller21}.
+For operating models with natural mortality random effects, we assume the same standard deviation \(\sigma_M = 0.3\).
+This is similar to the intermediate value used for R+M OMs by \citet{milleretal_inreview1}.
+
+The covariate effect is one of 3 alternative values in the operating models:
 
@@
 
-The parameters defining the simulated covariate time series, size of the
-covariate effect, and any natural mortality random effects result in a
-range of different levels of variation in annual natural mortality rates
-(Figure \ref{M_example}).
-
-As described by \citet{stockmiller21}, the model for process errors on
-abundance at age in WHAM is \begin{equation}
-  \text{log}N_{a,y}=\left\{
-    \begin{array}{@{}lll@{}}
-      \text{log} \left( f(SSB_{y-1}) \right) + \varepsilon_{1,y}, & \text{if}\ a = 1 \\
-      \text{log} \left( N_{a-1,y-1} \right) - Z_{a-1,y-1} + \varepsilon_{a,y}, & \text{if}\ 1 < a < A \\
-      \text{log} \left( N_{A-1,y-1} e^{-Z_{A-1,y-1}} + N_{A,y-1} e^{-Z_{A,y-1}} \right) + \varepsilon_{A,y}, & \text{if}\ a = A
-    \end{array}\right.
-\end{equation} We assume the the standard deviation of recruitment
-deviations is
-\(\text{SD}\left(\varepsilon_{1,y}\right) = \sigma_R = 0.5\) for all
-operating models. For operating models with survival devations for older
-ages, we assume the same standard deviations for all older ages
-\(\text{SD}\left(\varepsilon_{a,y}\right) = \sigma_{2+} = 0.3\).
-
-The alternative configurations of random effects on recruitment,
-survival, and natural mortality are
-
-\begin{itemize}
-\item uncorrelated random effects on recruitment only ($\sigma_R = 0.5$) (R),
-\item uncorrelated random effects on recruitment ($\sigma_R = 0.5$) and survival of older age classes ($\sigma_{2+} = 0.3$) (R+S), and
-\item uncorrelated random effects on R ($\sigma_R = 0.5$) and log-natural mortality ($\sigma_M = 0.3$) (R+M).
-\end{itemize}
+The parameters defining the simulated covariate time series, size of the covariate effect, and any natural mortality random effects result in a range of different levels of variation in annual natural mortality rates (Figure \ref{M_example}).
+
+We configured all OMs with uncorrelated random effects on recruitment with standard devaition on log(recruitment) \(\sigma_R = 0.5\) for R, R+S, and R+M OMs.
+This same assumption was used by \citet{milleretalinreview} for R+M OMs and other OMs with fishery selectivity and index catchability process errors.
+For R+S OMs, apparent survival process errors were uncorrelated with \(\sigma_{2+} = 0.3\) and for R+M OMs, process errors on natural mortality were uncorrelated with \(\sigma_M = 0.3\).
 
@@
 
-We assumed a single fleet operating year round for catch observations
-with logistic selectivity for the fleet with \(a_{50} = 5\) and slope =
-1 (Figure \ref{om_mean_selectivity}). This selectivity was used to
-define \Fmsy, \Fspr[40], and steepness for the Beverton-Holt stock
-recruitment parameters above. We assumed a logistic-normal distribution
-for the age-composition observations associated with the fleet where
-errors in the multivariate normal transformation are independent. The
-standard deviation parameter is also constant across ages.
+We assumed a single fleet operating year round for catch observations with logistic selectivity for the fleet with \(a_{50} = 5\) and slope = 1 (Figure \ref{om_mean_selectivity}).
+This selectivity was used to define \Fmsy, \Fspr[40], and steepness for the Beverton-Holt stock recruitment parameters above.
+We assumed a logistic-normal distribution for the age-composition observations associated with the fleet where errors in the multivariate normal transformation are independent.
+The standard deviation parameter is also constant across ages.
 
@@
 
-Two time series of surveys are assumed and observed in numbers rather
-than biomass for the entire 40 year period with one occurring in the
-spring (0.25 way through the year) and one in the fall (0.75 way through
-the year). Catchability of both surveys are assumed to be 0.1. We
-assumed the same selectivity and age composition distribution as the
-fleet for both indices.
+Two time series of surveys are assumed and observed in numbers rather than biomass for the entire 40 year period with one occurring in the spring (0.25 way through the year) and one in the fall (0.75 way through the year).
+Catchability of both surveys are assumed to be 0.1.
+We assumed the same selectivity and age composition distribution as the fleet for both indices.
 
@@
 
-Standard deviation for log-aggregate catch was 0.1. There were two
-levels of observation error variance for indices and age composition for
-both indices and fleet catch. A low uncertainty specification assumed
-standard deviation of both series of log-aggregate index observations
-was 0.1 and the standard deviation of the logistic-normal for age
-composition observations was 0.3. In the high uncertainty specification
-the standard deviation for log-aggregate indices was 0.4 and that for
-the age composition observations was 1.5. For all estimating models,
-standard deviation for log-aggregate observations was assumed known
-whereas that for the logistic-normal age composition observations was
-estimated.
+Standard deviation for log-aggregate catch was 0.1.
+There were two levels of observation error variance for indices and age composition for both indices and fleet catch.
+A low uncertainty specification assumed standard deviation of both series of log-aggregate index observations was 0.1 and the standard deviation of the logistic-normal for age composition observations was 0.3.
+In the high uncertainty specification the standard deviation for log-aggregate indices was 0.4 and that for the age composition observations was 1.5.
+For all estimating models, standard deviation for log-aggregate observations was assumed known whereas that for the logistic-normal age composition observations was estimated.
 
@@
 
-For each data set simulated from an operating model 12 estimating models
-were fit. There were three factors defining the configuration of each
-estimating model
+For each data set simulated from an operating model 12 estimating models were fit. There were three factors defining the configuration of each estimating model
 
@@
 
-The configuration of the process errors in the estimating models matched
-the corresponding options in the operating models. For example, if
-uncorrelated R+S was assumed for both the estimating and operating
-model, the assumptions about process errors matched. The environmental
-covariate observations were included in all estimation models to ensure
-comparability of information criteria. All fixed effects parameters for
-selectivity, catchability, fully-selected fishing mortality, mean
-recruitment, initial abundance at age, and variances for logistic-normal
-age composition distributions were estimated. Any process error variance
-parameters for recruitment, survival, and natural mortality were also
-estimated. The observation error variance of the environmental
-observations and aggregate catch and indices were all assumed known at
-the true values.
+The configuration of the process errors in the estimating models generally matched the corresponding options in the operating models.
+
+For example, uncorrelated R+S was assumed for both the estimating and operating model.
+However, R+M EMs did not assume M random effects were uncorrelated (parameter was estimated).
+The environmental covariate observations were included in all estimation models to ensure comparability of information criteria.
+All fixed effects parameters for selectivity, catchability, fully-selected fishing mortality, mean recruitment, initial abundance at age, and variances for logistic-normal age composition distributions were estimated.
+Any process error variance parameters for recruitment, survival, and natural mortality were also estimated.
+The observation error variance of the environmental observations and aggregate catch and indices were all assumed known at the true values.
 
@@
 
-The first measure of reliability we investigated was frequency of
-convergence when fitting each estimating model to the simulated data
-sets. There are various ways to assess convergence of the fit, but we
-defined successful convergence as the hessian of the marginal
-log-likelihood being invertible and providing variance estimates for the
-fixed effects parameters.
+The first measure of reliability we investigated was frequency of convergence when fitting each estimating model to the simulated data sets.
+There are various ways to assess convergence of the fit, but we defined successful convergence as the hessian of the marginal log-likelihood being invertible and providing variance estimates for the fixed effects parameters.
 
@@
 We measured the frequency of correct model selection using marginal AIC.
-For a given operating model the set of models that were considered all
-made the same assumptions on whether or not to estimate (mean) natural
-mortality rate or it is assumed at the true value
-\(\beta_M = log(0.2)\). For model \(m\), the marginal AIC is a function
-of the marginal log-likelihood maximized with respect to the fixed
-effects in the model \(\boldsymbol{\theta}\) and the number of fixed
-effects \(n\left(\boldsymbol{\theta}\right)\) estimated, \[
+For a given operating model the set of models that were considered all made the same assumptions on whether or not to estimate (mean) natural mortality rate or it is assumed at the true value \(\beta_M = log(0.2)\).
+For model \(m\), the marginal AIC is a function of the marginal log-likelihood maximized with respect to the fixed effects in the model \(\boldsymbol{\theta}\) and the number of fixed effects \(n\left(\boldsymbol{\theta}\right)\) estimated,
+\[
 \text{AIC}_m = -2\left[{\text{argmax}}_{\boldsymbol{\theta}} \log L_m\left({\boldsymbol{\theta}}\right) - n\left({\boldsymbol{\theta}}\right)\right].
@@
 
-We calculated median relative errors of
+We calculated median errors of
 
@@
 \item $\beta_M$, the mean log-natural mortality rate,
+\end{itemize}
+
+and the median relative errors of
+
+\begin{itemize}
 \item annual natural mortality rate,
@@
 
-Results for fishing mortality rate are provided in the Supplementary
-Materials. For the \(i\)th simulated data the relative error for a
-parameter \(\theta\) provided from the fitted estimation model is \[
+Note that the exponentiating the median error of \(\beta_M\) would be equivalent to the median relative error of the median natural mortality rate because the median of the exponential is the same as the exponential of the median.
+
+Results for fishing mortality rate are provided in the Supplementary Materials.
+For the \(i\)th simulated data the relative error for a parameter \(\theta\) provided from the fitted estimation model is
+\[
 {\text{RE}_i}\left(\theta\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
-\] and measured bias as the median relative error and constructed 95\%
-confidence intervals using the binomial distribution approach as in
-\citet{stockmiller21}. If the confidence interval contains zero we do
-not have evidence of bias from the simulation study. We used estimates
-from all simulations that satisfied the first type of convergence
-described above. We were not concerned about more restrictive types of
-convergence because some estimation models would not be expected to
-satisfy these criteria because of mis-specified structures. For example,
-an estimation model that includes survival random effects fitted to a
-set of observations simulated without survival random effects should
-have trouble estimating variance of the survival random effects, but the
-estimates of the parameters and derived output described above, might be
-reliable.
-
-For natural mortality rate, SSB, and fully-selected fishing mortality
-rate, we summarized results for each of the annual values, but we
-present results only for estimates from the first year (start), year 21
-(middel), and the last year (end), because there were no appreciable
-differences between the results for these three years and those for the
-other years.
+\]
+and measured bias as the median relative error and constructed 95\% confidence intervals using the binomial distribution approach as in \citet{stockmiller21}.
+If the confidence interval contains zero we do not have evidence of bias from the simulation study.
+We used estimates from all simulations that satisfied the first type of convergence described above.
+We were not concerned about more restrictive types of convergence because some estimation models would not be expected to satisfy these criteria because of mis-specified structures.
+For example, an estimation model that includes survival random effects fitted to a set of observations simulated without survival random effects should have trouble estimating variance of the survival random effects, but the estimates of the parameters and derived output described above, might be reliable.
+
+For natural mortality rate, SSB, and fully-selected fishing mortality rate, we summarized results for each of the annual values, but we present results only for estimates from the first year (start), year 21 (middel), and the last year (end), because there were no appreciable differences between the results for these three years and those for the other years.
 
@@
 
-We focused on the fourth convergence criteria, that the hessian of the
-negative-marginal log-likelihood at the optimized parameters was
-invertible. When the (mean) natural mortality rate was assumed at the
-true value, corresponding to usual practice in application of assessment
-models, good convergence for all EMs was observed for R+S operating
-models with \Fmsy fishing history, more variation in the latent
-environmental covariate (\(\sigma_\text{Ecov} = 0.5\)), and lower error
-in the associated observations \(\sigma_\text{ecov} = 0.1\) (Figure
-\ref{convergence_M_fixed}). Good convergence for all EMs was also
-observed for R+S OMs with the step change in fishing history, but with
-low error in indices and age comp observations.
-
-EMs that assumed no survival or natural mortality random effects (R
-OMS), nearly always converged across all operating models that assumed
-more variation in the latent environmental covariate
-(\(\sigma_\text{Ecov} = 0.5\)) and lower error in the associated
-observations \(\sigma_\text{ecov} = 0.1\). EMs that assumed recruitment
-and survival random effects (R+S OMS), had poor converge probability
-when the OMs had alternative process errors (R or R+M). The R+S EMs
-showed best convergence for the R+S OMs where there was lower error in
-the environmental observations \(\sigma_\text{ecov} = 0.1\) or higher
-error in environmental observations of latent true environmental
-covariates with greater temporal variation. EMs that assumed recruitment
-and natural mortality random effects (R+M OMs), had poor convergence
-probability for all OMs that had a step change in fishing history and
-higher uncertainty in indices and age composition.
-
-When the OMs and EMs both assume process errors only on recruitment (R)
-or on recruitment and survival (R+S), convergence was worst with less
-variation in the true environmental covariate and larger uncertainty in
-associated observations. When the OMs and EMs both assume process errors
-on recruitment and natural mortality (R+M) convergence was problematic
-for all OMs with the step-change in fishing history. The best
-convergence was observed with this match between OMs and EMs was when
-the fishing history was constant and there was low uncertainty in
-environmental observations.
-
-As might be expected, there was an overall drop in the probability of
-convergence when the mean natural mortality rate was estimated rather
-than assumed at the true value (Figure \ref{convergence_M_estimated}).
-Otherwise, general trends described above with mean natural mortality
-fixed, apply when estimated.
+When the (mean) log-natural mortality rate was assumed at the true value, corresponding to usual practice in application of assessment models, good convergence for all EMs was observed for R+S operating models with \Fmsy fishing history, more variation in the latent environmental covariate (\(\sigma_\text{Ecov} = 0.5\)), and lower error in the associated observations \(\sigma_\text{ecov} = 0.1\) (Figure \ref{convergence_M_fixed}).
+Good convergence for all EMs was also observed for R+S OMs with the step change in fishing history, but with low error in indices and age comp observations.
+
+EMs that assumed random effects just on recruitment (R), nearly always converged across all operating models that assumed more variation in the latent environmental covariate (\(\sigma_\text{Ecov} = 0.5\)) and lower error in the associated observations \(\sigma_\text{ecov} = 0.1\).
+EMs that assumed recruitment and survival random effects (R+S), had poor converge probability when the OMs had alternative process errors (R or R+M).
+The R+S EMs showed best convergence for the R+S OMs where there was lower error in the environmental observations \(\sigma_\text{ecov} = 0.1\) or higher error in environmental observations of latent true environmental covariates with greater temporal variation.
+EMs that assumed recruitment and natural mortality random effects (R+M), had poor convergence probability for all OMs that had a step change in fishing history and higher uncertainty in indices and age composition.
+
+When the OMs and EMs both assume process errors only on recruitment (R) or on recruitment and survival (R+S), convergence was worst with less variation in the true environmental covariate and larger uncertainty in associated observations.
+When the OMs and EMs both assume process errors on recruitment and natural mortality (R+M) convergence was problematic for all OMs with the step-change in fishing history.
+The best convergence was observed with this match between OMs and EMs was when the fishing history was constant and there was low uncertainty in environmental observations.
+
+As might be expected, there was an overall drop in the probability of convergence when the mean natural mortality rate was estimated rather than assumed at the true value (Figure \ref{convergence_M_estimated}).
+Otherwise, general trends described above with mean natural mortality fixed, apply when estimated.
 
@@
 
-When estimating models assumed the mean natural mortality rate was
-known, the best accuracy of AIC for model selection occurred for models
-with R+S process errors. R+S estimating models ranked best with very low
-frequency for R or R+M operating models and with very high frequency for
-R+S operating models (Figure \ref{aic_rank_M_fixed}). R estimating
-models were determined best with high frequency for R operating models,
-but also for R+M operating models. R+M estimating models were rarely
-determined best for any operating models including those where the
-process errors matched.
-
-AIC was conservative for determining whether the environmental covariate
-affected natural mortality. AIC was highly accurate in determining no
-effect when there was no effect in the operating model, but AIC ranked
-the null model best with high frequency even when there was an effect in
-the operating model in many cases. However the accuracy of AIC improved
-in certain operating models. Increased effect size, increased temporal
-contrast in the covariate, and lower uncertainty in all observations
-types lead to increased accuracy of determining covariate effects.
-
-Relative to the assumption that the mean natural mortality rate was
-known, estimating the mean natural mortality rate had small effects on
-the accuracy of AIC in selecting the appropriate process error and
-whether the covariate affect natural mortality (Figure
-\ref{aic_rank_M_estimated}). Where there were differences there were
-small decreases in accuracy for determining the appropriate process
-error and determining a covariate effect when there was one.
+When estimating models assumed the mean natural mortality rate was known, the best accuracy of AIC for model selection occurred for models with R+S process errors.
+R+S estimating models ranked best with very low frequency for R or R+M operating models and with very high frequency for R+S operating models (Figure \ref{aic_rank_M_fixed}).
+R estimating models were determined best with high frequency for R operating models, but also for R+M operating models.
+R+M estimating models were rarely determined best for any operating models including those where the process errors matched.
+
+AIC was conservative for determining whether the environmental covariate affected natural mortality.
+AIC was highly accurate in determining no effect when there was no effect in the operating model, but AIC ranked the null model best with high frequency even when there was an effect in the operating model in many cases.
+However the accuracy of AIC improved in certain operating models.
+Increased effect size, increased temporal contrast in the covariate, and lower uncertainty in all observations types lead to increased accuracy of determining covariate effects.
+
+Relative to the assumption that the mean natural mortality rate was known, estimating the mean natural mortality rate had small effects on the accuracy of AIC in selecting the appropriate process error and whether the covariate affect natural mortality (Figure \ref{aic_rank_M_estimated}).
+Where there were differences there were small decreases in accuracy for determining the appropriate process error and determining a covariate effect when there was one.
 
@@
 
-When the EMs assumed the mean natural mortality rate was known, we
-observed generally accurate estimation of environmental effects across
-all EM and OM process error assumptions and all true covariate effect
-sizes, when ther was low uncertainty in environmental observations and
-larger temporal contrast in the simulated true environmental covariate
-(Figure \ref{Ecov_beta_bias_M_fixed}). We observed a negative trend in
-bias of the environmental effect with increased effect size when
-temporal variation in the covariate was lower and/or uncertainty in the
-covariate observations was higher. When the OMs had R+S process errors
-with low temporal variation in the true environmental covariate and
-lower uncertainty in the indices age age composition, estimated
-covariate effects were highly variable. In most cases the relative error
-of \(\beta_\text{Ecov}\) did not depend on the source of process error
-assumed in the EM. When there was an effect of the EM process error
-assumption it was when OMs had R+S process errors. The worst bias was
-observed when OMs assumed R+S process errors, high uncertainty in
-covariate observations, low variability in the covariate, and low
-uncertainty in index and age composition observations.
-
-When the mean natural mortality rate was estimated in the EM, results
-were similar except estimated effects were even more variable for data
-simulated with R+S process errors (Figure
-\ref{Ecov_beta_bias_M_estimated}). There was also more separation of
-reliability of the estimation of the effect among EMs with different
-process error assumptions. The separation was most apparent when OMs
-simulated R+S process errors and larger variability in the environmental
-covariate wehere EMs with process errors other than R+S showing more
-bias than when the mean natural mortality rate was assumed known.
+When the EMs assumed the mean natural mortality rate was known, we observed generally accurate estimation of environmental effects across all EM and OM process error assumptions and all true covariate effect sizes, when ther was low uncertainty in environmental observations and larger temporal contrast in the simulated true environmental covariate (Figure \ref{Ecov_beta_bias_M_fixed}).
+We observed a negative trend in bias of the environmental effect with increased effect size when temporal variation in the covariate was lower and/or uncertainty in the covariate observations was higher.
+When the OMs had R+S process errors with low temporal variation in the true environmental covariate and lower uncertainty in the indices age age composition, estimated covariate effects were highly variable.
+In most cases the relative error of \(\beta_\text{Ecov}\) did not depend on the source of process error assumed in the EM.
+When there was an effect of the EM process error assumption it was when OMs had R+S process errors.
+The worst bias was observed when OMs assumed R+S process errors, high uncertainty in covariate observations, low variability in the covariate, and low uncertainty in index and age composition observations.
+
+When the mean natural mortality rate was estimated in the EM, results were similar except estimated effects were even more variable for data simulated with R+S process errors (Figure \ref{Ecov_beta_bias_M_estimated}).
+There was also more separation of reliability of the estimation of the effect among EMs with different process error assumptions.
+The separation was most apparent when OMs simulated R+S process errors and larger variability in the environmental covariate wehere EMs with process errors other than R+S showing more bias than when the mean natural mortality rate was assumed known.
 
 \hypertarget{mean-natural-mortality-rate}{%
-\subsubsection*{Mean natural mortality
-rate}\label{mean-natural-mortality-rate}}
+\subsubsection*{Mean natural mortality rate}\label{mean-natural-mortality-rate}}
 \addcontentsline{toc}{subsubsection}{Mean natural mortality rate}
 
-We found high accuracy for estimation of the mean natural mortality rate
-parameter (\(\beta_M\)) for all EM process errors assumptions when OMs
-had step changes in fishing mortality, lower uncertainty in index and
-age composition observations, and either R or R+M process errors (Figure
-\ref{mean_M_bias_Ecov_beta_fixed}). The most variation in estimates
-occurred when fishing mortality was constant and there was higher
-uncertainty in index and age composition observations. For OMs with R+S
-process errors, the most reliable estimation of \(\beta_M\) was obtained
-when the EM also assumed R+S process errors across all other factors.
-For OMs with R+M process errors, the matching assumption for the EM only
-showed the best reliability when there was low uncertainty in index and
-age composition observations and fishing mortality was constant.
+We found high accuracy for estimation of the mean natural mortality rate parameter (\(\beta_M\)) for all EM process errors assumptions when OMs had step changes in fishing mortality, lower uncertainty in index and age composition observations, and either R or R+M process errors (Figure \ref{mean_M_bias_Ecov_beta_fixed}).
+The most variation in estimates occurred when fishing mortality was constant and there was higher uncertainty in index and age composition observations.
+For OMs with R+S process errors, the most reliable estimation of \(\beta_M\) was obtained when the EM also assumed R+S process errors across all other factors.
+For OMs with R+M process errors, the matching assumption for the EM only showed the best reliability when there was low uncertainty in index and age composition observations and fishing mortality was constant.
 
 \hypertarget{annual-natural-morality-rate}{%
-\subsubsection*{Annual natural morality
-rate}\label{annual-natural-morality-rate}}
+\subsubsection*{Annual natural morality rate}\label{annual-natural-morality-rate}}
 \addcontentsline{toc}{subsubsection}{Annual natural morality rate}
 
-We present results for error in annual natural mortality rate
-conditional on three alternative EM configurations for the natural
-mortality parameters: 1) mean natural mortality rate parameter is fixed
-at the true value (\(\beta_M = log(0.2)\)) and no covariate effect is
-assumed (\(\beta_\text{Ecov} = 0\)), 2) \(\beta_M = log(0.2)\) and
-\(\beta_\text{Ecov}\) is estimated, and 3) both \(\beta_M\) and
-\(\beta_\text{Ecov}\) are estimated.
-
-When OMs and EMs assume \(\beta_M = log(0.2)\) and
-\(\beta_\text{Ecov} = 0\), there is no annual variation in natural
-mortality for OMs with R or R+S process errors simulated or assumed in
-the EM and, therefore, estimation bias is not possible (Figure
-\ref{annual_M_bias_M_fixed_beta_fixed}). We also observe little or no
-evidence of bias (confidence intervals include 0) for R or R+S EMs for
-any OMs even when the OMs included an effect of the covariate on natural
-mortality (\(\beta_\text{Ecov} > 0\)). Including process error on M
-(R+M) produces more variability in errors for R and R+S EMs than
-including increased level of effect of the covariate on M. For R and R+S
-OMs, errors in annual M for R+S EMs were less variable than those for R
-EMs.
-
-R+M EMs fit to R OMs with constant fishing mortality, or a step change
-in fishing mortality and lower uncertainty in index and age composition
-observations exhibited no bias in annual estimation of M indicating that
-the variance of the estimated random effects for M in these EMs
-collapsed to 0. R+M EMs were the only EMs to exhibit differences in the
-sign of the median errors across the time series. When OMs had a step
-change in fishing mortality and higher uncertainty in index and age
-composition observations, we observed positive median errors at the
-beginning of the time series and lower median errors in later years.
-However, whether there was evidence of bias, depended on the uncertainty
-in covariate observations and the degree of variability in the latent
-covariate. We observed negative median relative errors in annual M
-estimates for R+M EMs fit to R+S OMs with lower uncertainty in index and
-age composition observations across all levels of simulated effects of
-the covariate, but evidence of bias was strongest at the beginning of
-the time series.
-
-When \(\beta_\text{Ecov}\) was estimated, R EMs performed worse for R+S
-OMs with evidence of negative bias for OMs with constant fishing
-mortality, lower uncertainty in index and age composition observations,
-the largest covariate effect size, lower variability in the latent
-covariate, and larger uncertainty in the covariate observations (Figure
-\ref{annual_M_bias_M_fixed_beta_estimated}). There was little difference
-in the results for R+S EMs whether \(\beta_\text{Ecov}\) was estimated
-or not. The results for R+M EMs were generally similar to those when
-\(\beta_\text{Ecov}=0\) was assumed, except that the differences between
-the estimates at the beginning of the time series and later years for
-certain OM configurations did not occur.
-
-Allowing the EMs to estimate \(\beta_M\) resulted in very large
-variability in estimates for all EMs for R and R+M OMs with constant
-fishing mortality and higher uncertainty in index and age composition
-observations (Figure \ref{annual_M_bias_M_estimated_beta_estimated}).
-The same variability occurred for R+S OMs, but strong bias was estimated
-for R and R+M EMs. Annual M estimation was most reliable when OMs had
-step changes in fishing mortality and lower uncertainty in index and age
-composition observations. For R and R+M OMs with those characteristics,
-all EMs generally provided accurate estimation of annual M, but only R+S
-EMs provided accurate estimation for R+S OMs. Little or no bias in
-annual M estimation was observed for R+S EMs across all OM process error
-assumptions as long as there was a step change in fishing mortality and
-lower uncertainty in indices and age composition.
+We present results for error in annual natural mortality rate conditional on three alternative EM configurations for the natural mortality parameters: 1) mean natural mortality rate parameter is fixed at the true value (\(\beta_M = log(0.2)\)) and no covariate effect is assumed (\(\beta_\text{Ecov} = 0\)), 2) \(\beta_M = log(0.2)\) and \(\beta_\text{Ecov}\) is estimated, and 3) both \(\beta_M\) and \(\beta_\text{Ecov}\) are estimated.
+
+When OMs and EMs assume \(\beta_M = log(0.2)\) and \(\beta_\text{Ecov} = 0\), there is no annual variation in natural mortality for OMs with R or R+S process errors simulated or assumed in the EM and, therefore, estimation bias is not possible (Figure \ref{annual_M_bias_M_fixed_beta_fixed}).
+We also observe little or no evidence of bias (confidence intervals include 0) for R or R+S EMs for any OMs even when the OMs included an effect of the covariate on natural mortality (\(\beta_\text{Ecov} > 0\)).
+Including process error on M (R+M) produces more variability in errors for R and R+S EMs than including increased level of effect of the covariate on M.
+For R and R+S OMs, errors in annual M for R+S EMs were less variable than those for R EMs.
+
+R+M EMs fit to R OMs with constant fishing mortality, or a step change in fishing mortality and lower uncertainty in index and age composition observations exhibited no bias in annual estimation of M indicating that the variance of the estimated random effects for M in these EMs collapsed to 0.
+R+M EMs were the only EMs to exhibit differences in the sign of the median errors across the time series.
+When OMs had a step change in fishing mortality and higher uncertainty in index and age composition observations, we observed positive median errors at the beginning of the time series and lower median errors in later years.
+However, whether there was evidence of bias, depended on the uncertainty in covariate observations and the degree of variability in the latent covariate.
+We observed negative median relative errors in annual M estimates for R+M EMs fit to R+S OMs with lower uncertainty in index and age composition observations across all levels of simulated effects of the covariate, but evidence of bias was strongest at the beginning of the time series.
+
+When \(\beta_\text{Ecov}\) was estimated, R EMs performed worse for R+S OMs with evidence of negative bias for OMs with constant fishing mortality, lower uncertainty in index and age composition observations, the largest covariate effect size, lower variability in the latent covariate, and larger uncertainty in the covariate observations (Figure \ref{annual_M_bias_M_fixed_beta_estimated}).
+There was little difference in the results for R+S EMs whether \(\beta_\text{Ecov}\) was estimated or not.
+The results for R+M EMs were generally similar to those when \(\beta_\text{Ecov}=0\) was assumed, except that the differences between the estimates at the beginning of the time series and later years for certain OM configurations did not occur.
+
+Allowing the EMs to estimate \(\beta_M\) resulted in very large variability in estimates for all EMs for R and R+M OMs with constant fishing mortality and higher uncertainty in index and age composition observations (Figure \ref{annual_M_bias_M_estimated_beta_estimated}).
+The same variability occurred for R+S OMs, but strong bias was estimated for R and R+M EMs.
+Annual M estimation was most reliable when OMs had step changes in fishing mortality and lower uncertainty in index and age composition observations.
+For R and R+M OMs with those characteristics, all EMs generally provided accurate estimation of annual M, but only R+S EMs provided accurate estimation for R+S OMs.
+Little or no bias in annual M estimation was observed for R+S EMs across all OM process error assumptions as long as there was a step change in fishing mortality and lower uncertainty in indices and age composition.
 
@@
 
-Like natural mortality, we present results for error in spawning biomass
-conditional on three alternative EM configurations for the natural
-mortality parameters: 1) mean natural mortality rate parameter is fixed
-at the true value (\(\beta_M = log(0.2)\)) and no covariate effect is
-assmed (\(\beta_\text{Ecov} = 0\)), 2) \(\beta_M = log(0.2)\) and
-\(\beta_\text{Ecov}\) is estimated, and 3) both \(\beta_M\) and
-\(\beta_\text{Ecov}\) are estimated.
-
-When EMs assume \(\beta_M = log(0.2)\) and \(\beta_\text{Ecov} = 0\),
-there is little evidence of bias for R and R+M OMs except when there is
-a step change in fishing mortality and higher uncertainty in index and
-age composition observations, particularly at the end of the time series
-(Figure \ref{SSB_bias_M_fixed_beta_fixed}). We observed a similar trend
-in median relative error for R+S OMs, but there was more evidence of
-positive bias at the beginning of the time series whereas the confidence
-intervals for the negative median errors at the end of the time series
-often included 0. However, there was more indication of positive bias of
-the incorrect process error assumption of the EMs for the R+S OMs. For
-R+S EMs fit to R+S OMs, there was indication of small positive bias at
-the beginning of the time series when there was constant fishing
-mortality and higher uncertainty in index and age composition
-observations.
-
-When EMs assume \(\beta_M = log(0.2)\), but estimate
-\(\beta_\text{Ecov}\), the median errors for SSB for R and and R+M OMs
-are similar to those when \(\beta_\text{Ecov} = 0\) (Figure
-\ref{SSB_bias_M_fixed_beta_estimated}). For R+S OMs, the median errors
-for EMs that also assume R+S process errors are also similar to those
-with no covariate effect assumed. However, for R and R+M EMs fit to R+S
-OMs with constant fishing mortality rate and lower uncertainty in
-indices and age composition observations, we observed evidences of
-negative bias when higher covariate observation uncertainty and lower
-variation in the latent covariate was simulated and positive bias for
-other configurations of the covariate and corresponding observation
-uncertainty. When R+S OMs had constant fishing mortality rate and higher
-uncertainty in indices and age composition observations or step changes
-in fishing mortality and lower uncertainty in indices and age
-composition, R and R+M EMs often provided positively biased SSB
-estimates when there was lower uncertainty in covariate observations.
-
-When EMs estimated both \(\beta_M\) and \(\beta_\text{Ecov}\), we
-observed large variation in the errors in SSB under the same OM
-configurations where we observed large variation in errors for annual
-natural mortality rates (Figure
-\ref{SSB_bias_M_estimated_beta_estimated}). Similarly, we observed
-little or no bias in SSB estimation for R+S EMs across all OM process
-error assumptions as long as there was a step change in fishing
-mortality and lower uncertainty in indices and age composition.
+Like natural mortality, we present results for error in spawning biomass conditional on three alternative EM configurations for the natural mortality parameters: 1) mean natural mortality rate parameter is fixed at the true value (\(\beta_M = log(0.2)\)) and no covariate effect is assmed (\(\beta_\text{Ecov} = 0\)), 2) \(\beta_M = log(0.2)\) and \(\beta_\text{Ecov}\) is estimated, and 3) both \(\beta_M\) and \(\beta_\text{Ecov}\) are estimated.
+
+When EMs assume \(\beta_M = log(0.2)\) and \(\beta_\text{Ecov} = 0\), there is little evidence of bias for R and R+M OMs except when there is a step change in fishing mortality and higher uncertainty in index and age composition observations, particularly at the end of the time series (Figure \ref{SSB_bias_M_fixed_beta_fixed}).
+We observed a similar trend in median relative error for R+S OMs, but there was more evidence of positive bias at the beginning of the time series whereas the confidence intervals for the negative median errors at the end of the time series often included 0.
+However, there was more indication of positive bias of the incorrect process error assumption of the EMs for the R+S OMs.
+For R+S EMs fit to R+S OMs, there was indication of small positive bias at the beginning of the time series when there was constant fishing mortality and higher uncertainty in index and age composition observations.
+
+When EMs assume \(\beta_M = log(0.2)\), but estimate \(\beta_\text{Ecov}\), the median errors for SSB for R and and R+M OMs are similar to those when \(\beta_\text{Ecov} = 0\) (Figure \ref{SSB_bias_M_fixed_beta_estimated}).
+For R+S OMs, the median errors for EMs that also assume R+S process errors are also similar to those with no covariate effect assumed.
+However, for R and R+M EMs fit to R+S OMs with constant fishing mortality rate and lower uncertainty in indices and age composition observations, we observed evidences of negative bias when higher covariate observation uncertainty and lower variation in the latent covariate was simulated and positive bias for other configurations of the covariate and corresponding observation uncertainty.
+When R+S OMs had constant fishing mortality rate and higher uncertainty in indices and age composition observations or step changes in fishing mortality and lower uncertainty in indices and age composition, R and R+M EMs often provided positively biased SSB estimates when there was lower uncertainty in covariate observations.
+
+When EMs estimated both \(\beta_M\) and \(\beta_\text{Ecov}\), we observed large variation in the errors in SSB under the same OM configurations where we observed large variation in errors for annual natural mortality rates (Figure \ref{SSB_bias_M_estimated_beta_estimated}).
+Similarly, we observed little or no bias in SSB estimation for R+S EMs across all OM process error assumptions as long as there was a step change in fishing mortality and lower uncertainty in indices and age composition.
 
@@
 
-The estimating models assumed variances of aggregate catch and index
-observations was known. This approximation may be appropriate for
-indices where we have a reliable estimate of uncertainty based on the
-survey design (), but there may be better approaches for the aggregate
-catch such as an informed prior on the standard errors with realistic
-bounds.
-
-We found EMs with R+M process errors were rarely determined as the
-appropriate model when OMs simulated R+M process errors. This was
-unexpected but is likely due to the size of the variance assumed for
-those process errors (\(\sigma_M = 0.3\)) relative to the variances
-assumed for index and age composition observations. This is related to
-the difficulty in separately estimating observation and process error
-variances when the ratio of process to observation uncertainty is low.
-
-Note that the results for bias of covariate effect all assume ecov beta
-is estimated. The lack of bias in certain situations might suggest
-including the effect even if AIC says it isn't better at least when
-contrast in Ecov is high.
-
-The bias in natural mortality estimation resulted in biased estimation
-of stock size and likely harvest rates, but it will also propagate into
-biological reference points and possibly stock status.
-
-The lack of bias observed for annual M, using R+S EMs when no effect was
-assumed must be due to the annual M being equal to the assumed value
-(0.2) on average across simulations because the simulated environmental
-covariate has mean 0. However, if one were to condition on the covariate
-time series we would expect biased estimation of annual M when there was
-a covariate effect.
+The estimating models assumed variances of aggregate catch and index observations was known.
+This approximation may be appropriate for indices where we have a reliable estimate of uncertainty based on the survey design (), but there may be better approaches for the aggregate catch such as an informed prior on the standard errors with realistic bounds.
+
+We found EMs with R+M process errors were rarely determined as the appropriate model when OMs simulated R+M process errors.
+This was unexpected but is likely due to the size of the variance assumed for those process errors (\(\sigma_M = 0.3\)) relative to the variances assumed for index and age composition observations.
+This is related to the difficulty in separately estimating observation and process error variances when the ratio of process to observation uncertainty is low.
+
+Note that the results for bias of covariate effect all assume ecov beta is estimated.
+The lack of bias in certain situations might suggest including the effect even if AIC says it isn't better at least when contrast in Ecov is high.
+
+See project 0 paper for relevance of \citet{lietal24} and \citet{liljestrandetal24}. See \citet{milleretal_inreview1} fro relevance of project 0 results (estimability of natural mortality)
+
+\citet{derisoetal08} was the first to model natural mortality as a function of explicit covariate and residual random annual variation. Any relevant points?
+
+Uncertainty in estimated population attributes such as spawnign biomass can be increased considerably when natural mortality is estimated.
+This is also a consideration for estimating covariate effects on natural mortality even if the median natural mortality is assumed known.
+
+The bias in natural mortality estimation resulted in biased estimation of stock size and likely harvest rates, but it will also propagate into biological reference points and possibly stock status.
+
+The lack of bias observed for annual M, using R+S EMs when no effect was assumed must be due to the annual M being equal to the assumed value (0.2) on average across simulations because the simulated environmental covariate has mean 0.
+However, if one were to condition on the covariate time series we would expect biased estimation of annual M when there was a covariate effect.
 
@@
 
-As we would expect, reliable detection of covariate effects requires
-informative data. AIC preferred simpler models than the true model when
-information content in data and contrast in covariates and abundance
-were low. Null model for environmental covariate effect (no covariate
-effect) was selected when contrast in the time series was low and/or
-uncertainty in observations was high. Null selection likely decreases
-with strength of the effect on M. We only examined two non-zero effect
-sizes: 0.25, and 0.5, but our results suggest larger effect sizes, with
-the same observation error and contrast in time series, would allow
-better AIC performance for determining covariate effects. When there was
-process error in recruitment and M (R+M), models with process error only
-in recruitment were preferred. This could be because the variation in
-the M random effects was small relative to observation variability. The
-marginal variance for log M random effects was 0.3. Reliable estimation
-of environmental effect on M regardless of the process error assumed by
-the EM as long as the contrast in the covariate is sufficient and the
-uncertainty in the observations is low.
+As we would expect, reliable detection of covariate effects requires informative data.
+AIC preferred simpler models than the true model when information content in data and contrast in covariates and abundance were low.
+Null model for environmental covariate effect (no covariate effect) was selected when contrast in the time series was low and/or uncertainty in observations was high.
+Null selection likely decreases with strength of the effect on M.
+We only examined two non-zero effect sizes: 0.25, and 0.5, but our results suggest larger effect sizes, with the same observation error and contrast in time series, would allow better AIC performance for determining covariate effects.
+When there was process error in recruitment and M (R+M), models with process error only in recruitment were preferred.
+This could be because the variation in the M random effects was small relative to observation variability.
+The marginal variance for log M random effects was 0.3.
+Reliable estimation of environmental effect on M regardless of the process error assumed by the EM as long as the contrast in the covariate is sufficient and the uncertainty in the observations is low.
 
@@
 
-This work was funded by NOAA Fisheries Northeast Fisheries Science
-Center.
+This work was funded by NOAA Fisheries Northeast Fisheries Science Center.
 
@@
 
-\bibliography{paper}
+\bibliography{manuscript}
 
@@
 \begin{figure}
-\caption{The proportion mature at age assumed for the population in all operating and estimating models.}\label{om_maturity}
-\begin{center}
-\includegraphics[width = \textwidth]{om_maturity.png}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\caption{The weight at age assumed for the population in all operating and estimating models.}\label{om_waa}
-\begin{center}
-\includegraphics[width = \textwidth]{om_waa.png}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\caption{The Beverton-Holt stock recruit relationship assume for all operating models.}\label{om_sr}
-\begin{center}
-\includegraphics[width = \textwidth]{om_sr.png}
-\end{center}
-\end{figure}
-
-\begin{figure}
 \caption{Example simulations of environmental covariate latent processes and observations with different levels of observation error, and different assumptions about variability of the latent process.}\label{om_ecov_example}
 \begin{center}
-\includegraphics[width = \textwidth]{Ecov_true_obs_example.png}
+%\includegraphics[width = \textwidth]{Ecov_true_obs_example.png}
 \end{center}
@@
 \begin{center}
-\includegraphics[width = \textwidth]{M_example.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{figure}
-\caption{The selectivity at age assumed for the fishing fleet in operating models without selectivity random effects.
-The mean selectivity at age across time for the fishing fleet in operating models with selectivity random effects.}\label{om_mean_selectivity}
-\begin{center}
-\includegraphics[width = \textwidth]{om_mean_selectivity.png}
-\end{center}
-\end{figure}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of convergence of estimating models with alternative process error assumptions fit to simulated populations and data sets from operating models.
-The convergence criterion is an invertible hessian providing standard error estimates for all fixed effects parameters.
-Vertical lines represent 95\% confidence intervals.
-All estimating models estimate environmental covariate effects on natural mortality but assume the mean mortality parameter $\beta_\text{M} = \log 0.2$.}\label{convergence_M_fixed}
-\begin{center}
-\includegraphics[height = \textheight]{proportion_good_hessian_ecov_effect_est_M_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Probability of convergence of estimating models with alternative process error assumptions fit to simulated populations and data sets from operating models.
-The convergence criterion is an invertible hessian providing standard error estimates for all fixed effects parameters.
-Vertical lines represent 95\% confidence intervals.
-All estimating models estimate environmental covariate effects on natural mortality and the mean mortality parameter $\beta_\text{M}$.}\label{convergence_M_estimated}
-\begin{center}
-\includegraphics[height = \textheight]{proportion_good_hessian_ecov_effect_est_M_est.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets for each operating model where the estimation model type (process error and covariate effect assumptions) had the lowest AIC.
-All estimating models assumed the mean mortality parameter $\beta_\text{M} = \log 0.2$.}\label{aic_rank_M_fixed}
-\begin{center}
-\includegraphics[height = \textheight]{proportion_best_AIC_M_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Proportion of simulated data sets for each operating model where the estimation model type (process error and covariate effect assumptions) had the lowest AIC.
-All estimating models assumed the mean mortality parameter $\beta_\text{M} = \log 0.2$.}\label{aic_rank_M_estimated}
-\begin{center}
-\includegraphics[height = \textheight]{proportion_best_AIC_M_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of the estimated environmental effect on natural mortality $\beta_\text{Ecov}$ from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
-All estimating models assumed the mean mortality parameter $\beta_\text{M} = \log 0.2$.}\label{Ecov_beta_bias_M_fixed}
-\begin{center}
-\includegraphics[height = \textheight]{Ecov_beta_bias_all_PE_effect_M_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of the estimated environmental effect on natural mortality $\beta_\text{Ecov}$ from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
-All estimating models assumed the mean mortality parameter $\beta_\text{M}$ is also estimated.}\label{Ecov_beta_bias_M_estimated}
-\begin{center}
-\includegraphics[height = \textheight]{Ecov_beta_bias_all_PE_effect_M_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of the estimated mean natural mortality parameter $\beta_\text{M}$ from fitting simulated observations from each operating model with the estimating model assuming $\beta_\text{Ecov} = 0$ under alternative process error assumptions.}\label{mean_M_bias_Ecov_beta_fixed}
-\begin{center}
-\includegraphics[height = \textheight]{mean_M_bias_all_PE_effect_Ecov_beta_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of the estimated mean natural mortality parameter $\beta_\text{M}$ from fitting simulated observations from each operating model with the estimating model estimating $\beta_\text{Ecov} = 0$ under alternative process error assumptions.}\label{mean_M_bias_Ecov_beta_estimated}
-\begin{center}
-\includegraphics[height = \textheight]{mean_M_bias_all_PE_effect_Ecov_beta_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of annual natural mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
-Estimating models also assume the mean natural mortality parameter $\beta_\text{M} = \log 0.2$ and no environmental covariate effect $\beta_\text{Ecov} = 0$.}\label{annual_M_bias_M_fixed_beta_fixed}
-\begin{center}
-\includegraphics[height = \textheight]{annual_M_bias_all_PE_effect_M_fixed_beta_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of annual natural mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
-Estimating models also assume the mean natural mortality parameter $\beta_\text{M} = \log 0.2$ and the environmental covariate effect $\beta_\text{Ecov}$ is estimated.}\label{annual_M_bias_M_fixed_beta_estimated}
-\begin{center}
-\includegraphics[height = \textheight]{annual_M_bias_all_PE_effect_M_fixed_beta_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of annual natural mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
-Estimating models also assume both the mean natural mortality parameter $\beta_\text{M}$ and the environmental covariate effect $\beta_\text{Ecov}$ are estimated.}\label{annual_M_bias_M_estimated_beta_estimated}
-\begin{center}
-\includegraphics[height = \textheight]{annual_M_bias_all_PE_effect_M_estimated_beta_estimated.png}
+%\includegraphics[width = \textwidth]{M_example.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+%\includegraphics{convergence}
+\end{center}
+\caption{Estimated probability of fits providing hessian-based standard errors for EMs assuming alternative process error, that estimate or assume known median natural mortality, and that estimate or assume no covariate effect on median natural mortality when fitted to operating models that have R (left) and R+S (middle) , or R+M (right) process error structures and three levels of true covariate effect on median natural mortality (x axis).
+Vertical lines represent 95\% confidence intervals.}\label{convergence}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+%\includegraphics[height = \textheight]{aic}
+\end{center}
+\caption{Proportion of simulated data sets for each operating model where the estimation model type (combination of assumed process error type and covariate effect) had the lowest AIC.
+All EMs had known median natural mortality rate and all OMs and EMs had low observation error.}\label{aic}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+%\includegraphics[height = \textheight]{Ecov_beta_bias}
+\end{center}
+\caption{Relative error of estimates of environmental effect on natural mortality $\beta_\text{Ecov}$ from fitting EMs with alternative process error assumptions.
+All OMs had low observation error and contrast in fishing mortality.
+Vertical lines represent 95\% confidence intervals.}\label{ecov_beta_bias}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+%\includegraphics[height = \textheight]{median_log_M_bias}
+\end{center}
+\caption{Relative error of the estimated mean log-natural mortality parameter $\beta_{M}$ from fitting EMs with alternative process error assumptions.
+All OMs had low observation error and contrast in fishing mortality.
+Vertical lines represent 95\% confidence intervals.}\label{median_log_M_bias}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of annual natural mortality rate in terminal year from fitting simulated observations from each OM with alternative process error assumptions in the EM.
+Mean log-natural mortality parameter ($\beta_\text{M}$) and environmental covariate effect ($\beta_\text{Ecov}$) are both estimated in the EMs.
+All OMs had low observation error and contrast in fishing mortality.}\label{terminal_M_bias}
+\begin{center}
+%\includegraphics[height = \textheight]{terminal_year_M_bias.png}
 \end{center}
@@
 
+\setcounter{figure}{0}
+\renewcommand\thefigure{S\arabic{figure}}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+%\includegraphics[height = \textheight]{aic_supp1}
+\end{center}
+\caption{Proportion of simulated data sets for each OM where the EM type (assumed process error type) had the lowest AIC.
+All EMs had known median mortality rate and all OMs and EMs had low observation error.}\label{aic_supp1}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+%\includegraphics[height = \textheight]{aic_supp2}
+\end{center}
+\caption{Proportion of simulated data sets for each operating model where the estimation model type (combination of assumed process error type and covariate effect) had the lowest AIC.
+All OMs had a change in fishing pressure over time and all OMs and EMs had low observation error.}\label{aic_supp2}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+%\includegraphics[height = \textheight]{aic_supp3}
+\end{center}
+\caption{Proportion of simulated data sets for each operating model where the estimation model type (combination of assumed process error type and covariate effect) had the lowest AIC.
+All EMs had known median natural mortality rate and all OMs and EMs had high observation error.}\label{aic_supp3}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Median relative error of annual natural mortality rate in terminal year from fitting simulated observations from each OM with alternative process error assumptions in the EM.
+Mean log-natural mortality parameter ($\beta_\text{M}$) is estimated, but EMs assume no environmental covariate effect ($\beta_\text{Ecov} = 0 $) are both estimated in the EMs.
+All OMs had low observation error and contrast in fishing mortality.}\label{terminal_M_bias_beta_0}
+\begin{center}
+%\includegraphics[height = \textheight]{terminal_year_M_bias_ecov_beta_0.png}
+\end{center}
+\end{figure}
+\end{landscape}
+
 \begin{landscape}

### 2025-05-21 Ecov_study/mortality/manuscript/manuscript.Rmd
saving work on Ecov-M

@@
 State-space stock assessment models, with non-linear functions of latent processes and numerous observation types with different probability distribution assumptions represent one of most complex classes of state-space models.
-The literature on the ways we make inferences and the effects of various factors on reliabilty of inferences from state-space assessment models is growing [@lietal24;@milleretal_inreview1;@cadiganetal].
+The literature on the ways we make inferences and the effects of various factors on reliabilty of inferences from state-space assessment models is growing [@lietal24;@milleretal_inreview1;cadiganetal].
 The importance of contrast in population size and fishing mortality and quality of data used to fit assessment models including the state-space variety is known [@magnussonhilborn07;@milleretal_inreview1].
-Furthermore, estimation of natural mortality, and even temporal variablity is possible in many scenarios [@leeetal11;johnsonetal15;@cadigan16;@millerhyun18;@milleretal_inreview1].
+Furthermore, estimation of natural mortality, and even temporal variablity is possible in many scenarios [@leeetal11;@johnsonetal16;@cadigan16;@millerhyun18;@milleretal_inreview1].
 
-The effects of temporal variation in recruitment via unspecified or specified environmental factors have been extensively investigated in both traditional assessment models and state space models [@myers98;@milleretal16; @haltuchpapers].
+The effects of temporal variation in recruitment via unspecified or specified environmental factors have been extensively investigated in both traditional assessment models and state space models [@myers98;@milleretal16, haltuchpapers].
 Reliability of estimating environmental and spawning biomass effects on recruitment requires a combination of strong effects, good age composition data quality, contrast in the environmental covariate and lower recruitment variability [@brittenetal_inreview;@milleretal_inreview1].
 
-Temporal and environmental effects on growth and weight at age are also more important for short term projections and the reliability of estimation of those effects @correaetal23; @correaetalinreview.
+Temporal and environmental effects on growth and weight at age are also more important for short term projections and the reliability of estimation of those effects [@correaetal23; @correaetal_inreview].
 
@@
 
-See @millerretal_inreview1 fro relevance of project 0 results (estimability of natural mortality).
-Some difficulty in distinquishing variation in natural mortality or effects of explicit covariates when variation in M random effects or Ecov is low relative to observation error.
+See @milleretal_inreview1 for relevance of project 0 results (estimability of natural mortality).
+Some difficulty in distinguishing variation in natural mortality or effects of explicit covariates when variation in M random effects or Ecov is low relative to observation error.
 
@@
 
- Incorrect treatment of population attributes as temporally varying [@trijouletetal20; @liljestrandetal24] could lead to misidentification of stock status and biased population estimates, ultimately impacting fisheries management decisions [@legaultpalmer16; @szuwalskietal18; @croninpunt21].
-Furthermore, biological, fishery, and observational processes are often confounded in catch-at-age data, which may adversely affect ability to distinguish between true process variability and observational error [@puntetal14; @stewartmonnahan17; @croninpunt21; @fischetal23; @lietalinreview_a].
+ Mis-specified temporal population process errors  could lead to misidentification of stock status and biased population estimates and poor fisheries management decisions [@trijouletetal20; @legaultpalmer16; @szuwalskietal18; @croninpunt21;@liljestrandetal24].
 
@@
 
-Estimation of natural mortality is known to be challenging in stock assessment models (Lee et al., others). However, estimation has been shown to be reliable in some situations @millerhyun18)
+Estimation of natural mortality is known to be challenging in stock assessment models (@leeetal11, others). However, estimation has been shown to be reliable in some situations [@millerhyun18]
 
@@
 
-Here we conduct a simulation study with operating models varying by degree of observation error uncertainty, sources of process error (M or survival), fishing history, temporal variation in environmental covariates, and magnitude of the effect of the covariate on natural mortality.
+Here we conduct a simulation study with OMs varying by degree of observation error uncertainty, sources of process error, fishing history, temporal variation in environmental covariates, and magnitude of the effect of the covariate on natural mortality.
 The simulations from these operating models are fitted with estimating models that make alternative assumptions for sources of process error, and whether (mean) M is estimated.
@@
 
-All of our analyses used the Woods Hole Assessment Model (WHAM) to construct both OMs and EMs [@millerstock20;@stockmiller21;@milleretal_inreview1].
-We used [version 1.0.6.9000, commit 77bbd94](https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3) for this simulation study.
-This package has been used extensively to configure OMs and EMs for several other simulation studies [@legaultetal23; @lietal24, @lietalinreview] and is used to assess many commercially important stocks in the Northeast US.
+All of our analyses used the Woods Hole Assessment Model (WHAM) to construct both OMs and EMs [@millerstock20; @stockmiller21; @milleretal_inreview].
+The WHAM package has been used extensively to configure OMs and EMs for several other simulation studies [@legaultetal23; @lietal24; @brittenetal_inreview; @lietal_inreview_a] and is used to assess many commercially important stocks in the Northeast U.S.
+[e.g., @nefsc22;@nefsc22a;@nefsc24].
+We used [version 1.0.6.9000, commit 77bbd94](https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3) for to generate all results.
 
--------------- need to edit this part
+We completed a simulation study with a 288 operating models.
+The factors defining the configuration of each operating model which are described in detail in subsequent sections include source of population process error (3 levels), index and catch observation uncertainty (2 levels), environmental covariate uncertainty (2 levels), latent environmental covariate process stochasticity (4 levels), and fishing history (2 levels).
+We simulated 100 data sets for each operating model that included simulations of process errors.
 
-There are many similarities to the configuration of OMs and EMs in simulation studies conducted by @milletetelinreview.
-We completed a simulation study with a number of OMs that can be categorized based on where process error random effects are assumed: abundance at age (R, R+S), natural mortality (R+M), fleet selectivity (R+Sel), or index catchability (R+q).
-For each OM assumption about variance of process errors and observations are required and the values we used were based on a review of the range of estimates from applications of WHAM in management of stocks of haddock, butterfish, and American plaice in the NE US.
+For each simulated data set we fit a set of 12 estimating models (EMs).
+The factors that distinguish the estimating models which are also described in detail below include source of population process error type (3 levels) whether (median) natural mortality rate was estimated or assumed known (2 levels), and whether the effect of the environmental covariate on natural mortality was estimated or not (2 levels).
 
-In total, we configured 72 OMs with alternative assumptions about the source and variability of process errors, level of observation error in indices and age composition data, and contrast in fishing pressure over time.
-We fitted 20 EMs to observations from each of 100 simulations where process errors were also simulated.
-For R+M, R+Sel, and R+q OMs, we used unique seeds for each simulation, but we inadvertently used the same 100 seeds for all R and R+S OMs.
-Each of the EMs made alternative assumptions about the source of process errors and whether natural mortality (or the median for models with process error in natural mortality) was estimated and whether a Beverton-Holt stock recruit relationship was estimated within the EM.
-Details of each of the operating and EMs are described below.
-
-We did not use the log-normal bias-correction feature for process errors or observations described by @stockmiller21 for operating and EMs [@lietalinreview].
-Simulations and model fitting were all carried out on the University of Massachusetts Green High-Performance Computing Cluster.
-All code we used to perform the simulation study and summarize results can be found at https://github.com/timjmiller/SSRTWG/tree/main/ecov_study/mortality/code.
-
---------------
-
---------------------------- From project_0 paper
-
-We used the Woods Hole Assessment Model (WHAM) to configure OMs and EMs in our simulation study [@millerstock20;@stockmiller21].
-WHAM is an R package freely available via a github repository and is built on the Template Model Builder package [@kristensenetal16].
-For this study we used [version 1.0.6.9000, commit 77bbd94](https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3).
-WHAM has also been used to configure OMs and EMs for closed loop simulations evaluating index-based assessment methods [@legaultetal23] and is currently used or accepted for use in management of numerous NEUS fish stocks [e.g., @nefsc22;@nefsc22a;@nefsc24].
-
-We completed a simulation study with a number of OMs that can be categorized based on where process error random effects were assumed: recruitment (R, assumed present in all models), apparent survival (denoted R+S), natural mortality (R+M), fleet selectivity (R+Sel), or index catchability (R+q).
-We refer to the (R+S) OMs as modeling apparent survival because on logscale the random effects ($\epsilon_{a,y}$) are additive to the total mortality (F+M) between numbers at age, thus they modify the survival term.
-However, as @stockmiller21 note, these random effects can be due to events other than mortality, such as immigration, emigration, missreported catch, and other sources of misspecification.
-For each OM, assumptions about the magnitude of the variance of process errors and observations are required and the values we used were based on a review of the range of estimates from Northeast Unite States (NEUS) assessments using WHAM.
-
-In total, we configured 72 OMs with alternative assumptions about the source and magnitude of process errors, magnitude of observation error in indices and age composition data, and contrast in fishing pressure over time.
-We fitted 20 EMs to observations generated from each of 100 simulations where process errors were also simulated.
-Each EM differed in assumptions about the source of process errors, whether natural mortality (or the median for models with process error in natural mortality) was estimated, and whether a Beverton-Holt SRR was estimated within the EM.
-Details of each of the OMs and EMs are described below.
-
-We did not use the log-normal bias-correction feature for process errors or observations described by [@stockmiller21] for OMs and EMs to simplify interpretation of the study results [@lietalinreview].
-All code we used to perform the simulation study and summarize results can be found at https://github.com/timjmiller/SSRTWG/tree/main/Project_0/code.
-
--------------------------
-
-We used the Woods Hole Assessment Model (WHAM) to configure operating and estimation models in our simulation study [@millerstock20;@stockmiller21].
-WHAM is an R package freely available as a github repository.
-For this study we used [version 1.0.6.9000, commit 77bbd94](https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3).
-This packages has also been used to configure operating and estimating models for closed loop simulations evaluating index-based assessment methods [@legaultetal23] and is used for management of haddock, butterfish, American plaice, bluefish, Atlantic cod, and black sea bass in the Northeast US.
-
-We completed a simulation study with a number of operating models.
-Many of the characteristics of the operating models are the same as those used in the working paper by Miller et al.
-(WP 4).
-The factors defining the configuration of each operating model which are described in detail in subsequent sections include
-\begin{itemize}
-\item two configurations of fishing histories, 
-\item two configurations of index and age composition observation error, 
-\item four configurations of latent environmental covariate process errors, 
-\item two configurations of environmental covariate observation uncertainty, 
-\item three levels of covariate effects on natural mortality, and 
-\item three configurations of sources of process errors for temporal changes in abundances at age. 
-\end{itemize} 
-The full factorial of the levels of each factor results in 288 different operating models. We simulated 100 data sets for each operating model that included simulations of process errors.
-
-For each simulated data set we fit a set of 12 estimating models. The factors defining the configuration or each estimating model which are also described in detail below include
-\begin{itemize}
-\item three configurations of sources of process errors for temporal changes in abundances at age, 
-\item whether the (mean) natural mortality rate was estimated or assumed known at the true value, and
-\item whether the effect of the environmental covariate on natural mortality was estimated.
-\end{itemize}
-
-We did not use the bias-correction feature for process errors or observations described by [@stockmiller21] for operating and estimating models.
+The sources of population process error that were used in the OMs or assumed in the EMs were on recruitment only (R), recruitment and changes in cohort abundance over time (R+S), or recruitment and natural mortality (R+M).
+We did not use the log-normal bias-correction feature for process errors or observations described by @stockmiller21 for operating and EMs [@lietal_inreview].
 Simulations were all carried out on the University of Massachusetts Green High-Performance Computing Cluster.
@@
 
+### Environmental covariate {-}
+
+In the WHAM model, environmental covariates are assumed to be described as state-space processes with annual observations of the true latent covariate [@milleretal16;stockmiller21].
+In our simulations, the latent covariate is assumed to be a stationary first order autoregressive (AR1) process
+$$
+X_y|X_{y-1} \sim \text{N}\left(\mu_E\left(1-\rho_E\right) + \rho_E E_{y-1}, \left(1-\rho_E^2\right)\sigma^2_E\right)
+$$
+with marginal mean $\mu_E=0$ and variance $\sigma^2_E$.
+The four configuration of the latent environmental covariate in the operating models assume the one of two values for the marginal standard deviation ($\sigma_E \in \{0.1, 0.5\}$) and for the autocorrelation parameter ($\rho_E \in \{0, 0.5\}$).
+
+The observations of the latent environmental covariate are assumed to be unbiased and Gaussian
+$$
+x_y|X_y \sim \text{N}\left(X_y,\sigma^2_e\right)
+$$
+The standard deviation of the environmental observations in the operating models is one of two values $\sigma_E \in \{0.1, 0.5\}$.
+Figure \ref{om_ecov_example} provides example simulations of the latent process and observations under the alternative configurations.
+
 ### Population {-}
 
-The population model tracks 10 age classes: ages 1 to 10+ and we assume spawning occurs 1/4 of the way through the year.
-The maturity at age was a logistic curve with $a_{50}$ = 2.89 and slope = 0.88 and assumed known in all estimation models (Figure \ref{om_maturity}).
-
-Weight at age was generated with a LVB growth function
+Many of the characteristics of the population biology and structure are the same as those in @milleretal_inreview1.
+We configures the stock with 10 age classes (ages 1 to 10+) where the last is a "plus" group that accumulates any older individuals.
+We model the population for 40 years with annual catch and index observations.
+We assume spawning occurs annually 1/4 of the way through the year.
+The maturity at age was a logistic function of age with age at 50\% maturity ($a_{50}$) = 2.89 and slope = 0.88 (Figure \ref{om_inputs_fig}, top left).
+We assumed a von Beralanffy growth function for weight at age  with length at age defined as
 $$
@@
 $$
-where $t_0 = 0$, $L_\infty = 85$, and $k = 0.3$, and a L-W relationship such that 
+where $t_0 = 0$, $L_\infty = 85$, and $k = 0.3$, and a length-weight 
 $$
@@
 $$
-where $\theta_1 = e^{-12.1}$ and $\theta_2 = 3.2$ (Figure \ref{om_waa}).
+where $\theta_1 = e^{-12.1}$ and $\theta_2 = 3.2$ (Figure \ref{om_inputs_fig}, top right).
 
-We assumed a Beverton-Holt stock recruit function with constant pre-recruit mortality parameters (i.e., $\alpha$ and $\beta$) for all operating models.
-All post-recruit productivity components are constant in the NAA and survey selectivity process error operating models.
-Therefore steepness and unfished recruitment are also constant over the time period for those operating models [@millerbrooks21].
-We specified unfished recruitment = $R_0 = e^{10}$ and $\Fmsy = F_{40\%} = 0.348$ equated to a steepness of 0.69 and $\alpha=0.60$ and $\beta = 2.4 \times 10^{-5}$ for the
-\[
-N_{1,y} = \frac{\alpha \text{SSB}_{y-1}}{1 + \beta \text{SSB}_{y-1}} 
-\]
-Beverton-Holt parameterization (Figure \ref{om_sr}).
-For OMs without process errors on natural mortality we assumed the rate was assumed 0.2.
-For OMs with process errors on natural mortality the mean rate was 0.2.
-
-Two alternative fishing histories were used for operating models.
-In the first scenario, the stock experiences overfishing for the first 20 years and fishing at \Fmsy for the last 20 years.
-In the second scenario, the stock is fished at \Fmsy for the entire time period.
-The magnitude of the overfishing assumptions is based on average estimates of overfishing for NE groundfish stocks from @wiedenmannetal19.
-@legaultetal23 also used similar approaches to defining fishing mortality histories for operating models.
-
-Initial population was configured at the equilibrium distribution fishing at either $F = 2.5\times \Fmsy$ or $F = \Fmsy$ for the two alternative fishing histories.
-That is for a deterministic model, the age composition would not change over time when the fishing mortality was constant at the respective level.
-
-For operating models with time-varying random effects or covariate effects on M, steepness is not constant, but we used the same alpha and beta parameters as other operating models this equates to a steepness and $R_0$ at the mean of the time series process for M.
-For operating models with time-varying random effects for fishery selectivity, \Fmsy is also not constant however we use the same F history as other operating models which corresponds to \Fmsy at the mean selectivity parameters.
-
-### Environmental covariate {-}
-
-In the WHAM model, environmental covariates are assumed to be described as state-space processes with annual observations of the true latent covariate.
-The latent covariate is assumed to be Gaussian AR1
+The general model for natural mortality in year $y$ is a log-linear function of a mean(median) parameter $\beta_{M}$, effects of an environmental covariate $E_y$ and a possibly autocorrelated normally distributed annual process error $\varepsilon_{M,y}$
 $$
-E_y|E_{y-1} \sim \text{N}\left(\mu_\text{Ecov}\left(1-\rho_\text{Ecov}\right) + \rho_\text{Ecov} E_{y-1}, \left(1-\rho_\text{Ecov}^2\right)\sigma^2_\text{Ecov}\right)
+\log M_y = \beta_M + \beta_{E} E_y + \varepsilon_{M,y}
 $$
-with marginal mean $\mu_\text{Ecov}=0$ and variance $\sigma^2_\text{Ecov}$.
-The configuration of the latent environmental covariate in the operating models have one of two alternate assumptions about the marginal standard deviation and autocorrelation parameter:
-\begin{itemize}
-\item marginal standard deviation $\sigma_\text{Ecov} \in \{0.1, 0.5\}$
-\item AR1 correlation $\rho_\text{Ecov} \in \{0, 0.5\}$
-\end{itemize}
-
-The observations of the latent environmental covariate are assumed to be unbiased and Gaussian
-$$
-\text{ecov}_y|E_y \sim \text{N}\left(E_y,\sigma^2_\text{ecov}\right)
-$$
-The standard deviation of the environmental observations in the operating models is one of two values:
-\begin{itemize}
-\item standard deviation $\in \{0.1, 0.5\}$
-\end{itemize}
-Figure \ref{om_ecov_example} provides example simulations of the latent process and observations under the alternative configurations.
-
-### Random effects on recruitment, survival, and natural mortality {-}
-
-Internally, WHAM treats natural mortality as a log-transformed parameter 
-$$
-\log M_y = \beta_M + \beta_{\text{Ecov}} E_y + \varepsilon_{M,y}
-$$
-that is a linear combination of a mean $\beta_M$ and any annual covariate $\beta_{\text{Ecov}}$ and random effects marginally distributed as $\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)$ [@stockmiller21].
-For operating models with natural mortality random effects, we assume the same standard deviation $\sigma_M = 0.3$.
-This is similar to the intermediate value used for R+M OMs by @milleretal_inreview1.
-
-The covariate effect is one of 3 alternative values in the operating models:
-\begin{itemize}
-\item $\beta_\text{Ecov} \in \{0,0.25,0.5\}$
-\end{itemize}
+$\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)$ [@stockmiller21].
+We assume the median natural mortality rate $e^{\beta_M} = 0.2$ is constant across ages.
+For R and R+S OMs and EMs, the process errors are set equal to 0.
+For all R+M OMs, we assume the same standard deviation $\sigma_M = 0.3$ and is estimated in the R+M EMs.
+The covariate effect is one of 3 alternative values in the operating models ($\beta_\text{E} \in \{0,0.25,0.5\}$.
 The parameters defining the simulated covariate time series, size of the covariate effect, and any natural mortality random effects result in a range of different levels of variation in annual natural mortality rates (Figure \ref{M_example}).
 
-We configured all OMs with uncorrelated random effects on recruitment with standard devaition on log(recruitment) $\sigma_R = 0.5$ for R, R+S, and R+M OMs.
-This same assumption was used by @milleretalinreview for R+M OMs and other OMs with fishery selectivity and index catchability process errors.
-For R+S OMs, apparent survival process errors were uncorrelated  with $\sigma_{2+} = 0.3$ and for R+M OMs, process errors on natural mortality were uncorrelated with $\sigma_M = 0.3$.
+We assume is a single fishing fleet harvesting the population with a logistic selectivity function with age at 50\% selection $a_{50} = 5$ and slope = 1 (Figure \ref{om_inputs_fig}, bottom left).
 
-### Fleets {-}
+We assumed expected recruitment each year from a Beverton-Holt stock-recruit relationship (SRR)
+\[
+R_{y} = \frac{a \text{SSB}_{y-1}}{1 + b \text{SSB}_{y-1}}.
+\]
+All biological inputs to calculations of spawning biomass per recruit (i.e., weight, maturity, and natural mortality at age) are constant in the R and R+S OMs without covariate effects on natural mortality.
+Therefore, steepness and unfished recruitment are also constant over the time period for those OMs [@millerbrooks21] and with our assumed biological inputs and selectivity the fishing mortality resulting in 40\% reduction in spawning biomass per recruit is $F_{40\%} = 0.348$.
+With an assumed unfished recruitment of $R_0 = e^{10}$, setting $\Fmsy = \Fspr[40]$ results in a steepness of 0.69 and $a=0.60$ and $b = 2.4 \times 10^{-5}$ (Figure \ref{om_inputs_fig}, bottom right).
 
-We assumed a single fleet operating year round for catch observations with logistic selectivity for the fleet with $a_{50} = 5$ and slope = 1 (Figure \ref{om_mean_selectivity}).
+We used two fishing scenarios for OMs.
+In the first scenario, the stock experiences overfishing at 2.5\Fmsy for the first 20 years followed by fishing at \Fmsy for the last 20 years (denoted $2.5\Fmsy \rightarrow \Fmsy$).
+In the second scenario, the stock is fished at \Fmsy for the entire time period (40 years).
+The magnitude of the overfishing assumptions is based on average estimates of overfishing for NEUS groundfish stocks from @wiedenmannetal19 and similar to the approach in @legaultetal23.
+
+We specified initial population abundance at age at the equilibrium distribution that corresponds to fishing at either $F = 2.5\times \Fmsy$ or $F = \Fmsy$.
+This implies that, for a deterministic model, the abundance at age would not change from the first year to the next.
+
+For R+M OMs and all OMs with environmental covariate effects on natural mortality, steepness is not constant, but we used the same $a$ and $b$ parameters as other operating models which equates to a steepness and $R_0$ at the median of the time series process for M.
+For operating models with time-varying random effects for fishery selectivity, \Fmsy is also not constant however we use the same fishing history as other operating models which corresponds to \Fmsy at the mean selectivity parameters.
+
+We configured all R, R+S, and R+M OMs with uncorrelated random effects on recruitment with standard deviation on log(recruitment) $\sigma_R = 0.5$.
+This same assumption was used by @milleretal_inreview1 for R+M OMs and other OMs with fishery selectivity and index catchability process errors.
+For R+S OMs, cohort temporal transition process errors were uncorrelated  with $\sigma_{2+} = 0.3$
+
+### Catch and index observations {-}
+
+We assumed a single fleet operating year round for catch observations with logistic selectivity for the fleet with $a_{50} = 5$ and slope = 1 (Figure \ref{om_inputs_fig}).
 This selectivity was used to define \Fmsy, \Fspr[40], and steepness for the Beverton-Holt stock recruitment parameters above.
@@
 
-### Indices {-}
-
 Two time series of surveys are assumed and observed in numbers rather than biomass for the entire 40 year period with one occurring in the spring (0.25 way through the year) and one in the fall (0.75 way through the year).
@@
 We assumed the same selectivity and age composition distribution as the fleet for both indices.
-
-### Observation Uncertainty {-}
 
@@
 \item whether the mean natural mortality $\beta_M$ was estimated or assumed known ($\log 0.2$)
-\item whether an environmental effect $\beta_\text{Ecov}$ was estimated or not (fixed at 0)
-\item whether the process errors were just on recruitment (R), on recruitment and survival (R+S) or on recruitment and natural mortality (R+M)
+\item whether an environmental effect $\beta_E$ was estimated or not (fixed at 0)
+\item whether the process errors were assumed on recruitment only (R), recruitment and survival (R+S), or recruitment and natural mortality (R+M)
 \end{itemize}
@@
 
-## Measures of reliability {-}
+## Performance measures {-}
 
@@
 \begin{itemize}
-\item $\beta_\text{Ecov}$, the effect of environmental covariates on natural mortality, 
+\item $\beta_E$, the effect of environmental covariates on natural mortality, 
 \item $\beta_M$, the mean log-natural mortality rate,
@@
 
-For natural mortality rate, SSB, and fully-selected fishing mortality rate, we summarized results for each of the annual values, but we present results only for estimates from the first year (start), year 21 (middel), and the last year (end), because there were no appreciable differences between the results for these three years and those for the other years.
+For natural mortality rate, SSB, and fully-selected fishing mortality rate, we summarized results for each of the annual values, but we present results only for estimates from the first year (start), year 21 (middle), and the last year (end), because there were no appreciable differences between the results for these three years and those for the other years.
 
@@
 
-## Convergence performance {-}
+Because @milleretal_inreview1 found inferences are most reliable for scenarios with low observation error for indices and age composition data and with temporal contrast in fishing pressure, we restrict our attention to scenarios with these characteristics in the main paper (temporal contrast in fishing and low observation error), but include corresponding results for the other scenarios in the Supplementary Materials.
 
-When the (mean) log-natural mortality rate was assumed at the true value, corresponding to usual practice in application of assessment models, good convergence for all EMs was observed for R+S operating models with \Fmsy fishing history, more variation in the latent environmental covariate ($\sigma_\text{Ecov} = 0.5$), and lower error in the associated observations $\sigma_\text{ecov} = 0.1$ (Figure \ref{convergence_M_fixed}).
+## Convergence {-}
+
+When the median natural mortality rate was assumed known, corresponding to usual practice in application of assessment models, good convergence for all EMs was observed for R+S operating models with \Fmsy fishing history, more variation in the latent environmental covariate ($\sigma_E = 0.5$), and lower error in the associated observations $\sigma_e = 0.1$ (Figure \ref{convergence}).
 Good convergence for all EMs was also observed for R+S OMs with the step change in fishing history, but with low error in indices and age comp observations.
 
-EMs that assumed random effects just on recruitment (R), nearly always converged across all operating models that assumed more variation in the latent environmental covariate ($\sigma_\text{Ecov} = 0.5$) and lower error in the associated observations $\sigma_\text{ecov} = 0.1$.
+EMs that assumed random effects just on recruitment (R), nearly always converged across all operating models that assumed more variation in the latent environmental covariate ($\sigma_E = 0.5$) and lower error in the associated observations $\sigma_E = 0.1$.
 EMs that assumed recruitment and survival random effects (R+S), had poor converge probability when the OMs had alternative process errors (R or R+M).
-The R+S EMs showed best convergence for the R+S OMs where there was lower error in the environmental observations $\sigma_\text{ecov} = 0.1$ or higher error in environmental observations of latent true environmental covariates with greater temporal variation.
+The R+S EMs showed best convergence for the R+S OMs where there was lower error in the environmental observations $\sigma_E = 0.1$ or higher error in environmental observations of latent true environmental covariates with greater temporal variation.
 EMs that assumed recruitment and natural mortality random effects (R+M), had poor convergence probability for all OMs that had a step change in fishing history and higher uncertainty in indices and age composition.
@@
 
-As might be expected, there was an overall drop in the probability of convergence when the mean natural mortality rate was estimated rather than assumed at the true value (Figure \ref{convergence_M_estimated}).
+As might be expected, there was an overall drop in the probability of convergence when the mean natural mortality rate was estimated rather than assumed at the true value (Figure \ref{convergence}).
 Otherwise, general trends described above with mean natural mortality fixed, apply when estimated.
@@
 
+When fitting the EMs to R and R+S OMs, AIC was generally accurate in determining the correct source of process errors and, when no effect of the covariate was simulated in the OMs, AIC was also accurate for the correct treatment of the environmental covariate $\beta_E = 0$ (Figs.
+\ref{aic}, \ref{aic_Rom}, \ref{aic_RSom}).
+However, when OMs included an effect of the covariate AIC accuracy for estimating an effect in the EM was only found when the OMs also included lower covariate observation error and higher effect size and higher temporal variability in the covariate.
+On the other hand, for R+M Oms, AIC generally was not able to accurately determine the correct source of process errors, but was able to accurately determine the correct treatment of the environmental covariate (Figs.
+\ref{aic} and \ref{aic_RMom}).
+
 When estimating models assumed the mean natural mortality rate was known, the best accuracy of AIC for model selection occurred for models with R+S process errors.
-R+S estimating models ranked best with very low frequency for R or R+M operating models and with very high frequency for R+S operating models (Figure \ref{aic_rank_M_fixed}).
+R+S estimating models ranked best with very low frequency for R or R+M operating models and with very high frequency for R+S operating models (Figure \ref{aic}).
 R estimating models were determined best with high frequency for R operating models, but also for R+M operating models.
@@
 
-Relative to the assumption that the mean natural mortality rate was known, estimating the mean natural mortality rate had small effects on the accuracy of AIC in selecting the appropriate process error and whether the covariate affect natural mortality (Figure \ref{aic_rank_M_estimated}).
+Relative to the assumption that the mean natural mortality rate was known, estimating the mean natural mortality rate had small effects on the accuracy of AIC in selecting the appropriate process error and whether the covariate affect natural mortality (Figure \ref{aic}).
 Where there were differences there were small decreases in accuracy for determining the appropriate process error and determining a covariate effect when there was one.
@@
 When the OMs had R+S process errors with low temporal variation in the true environmental covariate and lower uncertainty in the indices age age composition, estimated covariate effects were highly variable.
-In most cases the relative error of $\beta_\text{Ecov}$ did not depend on the source of process error assumed in the EM.
+In most cases the relative error of $\beta_E$ did not depend on the source of process error assumed in the EM.
 When there was an effect of the EM process error assumption it was when OMs had R+S process errors.
@@
 
-We present results for error in annual natural mortality rate conditional on three alternative EM configurations for the natural mortality parameters: 1) mean natural mortality rate parameter is fixed at the true value ($\beta_M = log(0.2)$) and no covariate effect is assumed ($\beta_\text{Ecov} = 0$),  2) $\beta_M = log(0.2)$ and $\beta_\text{Ecov}$ is estimated, and 3) both $\beta_M$ and $\beta_\text{Ecov}$ are estimated.
+We present results for error in annual natural mortality rate conditional on three alternative EM configurations for the natural mortality parameters: 1) mean natural mortality rate parameter is fixed at the true value ($\beta_M = log(0.2)$) and no covariate effect is assumed ($\beta_E = 0$),  2) $\beta_M = log(0.2)$ and $\beta_E$ is estimated, and 3) both $\beta_M$ and $\beta_E$ are estimated.
 
-When OMs and EMs assume $\beta_M = log(0.2)$ and $\beta_\text{Ecov} = 0$, there is no annual variation in natural mortality for OMs with R or R+S process errors simulated or assumed in the EM and, therefore, estimation bias is not possible (Figure \ref{annual_M_bias_M_fixed_beta_fixed}).
-We also observe little or no evidence of bias (confidence intervals include 0) for R or R+S EMs for any OMs even when the OMs included an effect of the covariate on natural mortality ($\beta_\text{Ecov} > 0$).
+When OMs and EMs assume $\beta_M = log(0.2)$ and $\beta_E = 0$, there is no annual variation in natural mortality for OMs with R or R+S process errors simulated or assumed in the EM and, therefore, estimation bias is not possible (Figure \ref{annual_M_bias_M_fixed_beta_fixed}).
+We also observe little or no evidence of bias (confidence intervals include 0) for R or R+S EMs for any OMs even when the OMs included an effect of the covariate on natural mortality ($\beta_E > 0$).
 Including process error on M (R+M) produces more variability in errors for R and R+S EMs than including increased level of effect of the covariate on M.
@@
 
-When $\beta_\text{Ecov}$ was estimated, R EMs performed worse for R+S OMs with evidence of negative bias for OMs with constant fishing mortality, lower uncertainty in index and age composition observations, the largest covariate effect size, lower variability in the latent covariate, and larger uncertainty in the covariate observations (Figure \ref{annual_M_bias_M_fixed_beta_estimated}).
-There was little difference in the results for R+S EMs whether $\beta_\text{Ecov}$ was estimated or not.
-The results for R+M EMs were generally similar to those when $\beta_\text{Ecov}=0$ was assumed, except that the differences between the estimates at the beginning of the time series and later years for certain OM configurations did not occur.
+When $\beta_E$ was estimated, R EMs performed worse for R+S OMs with evidence of negative bias for OMs with constant fishing mortality, lower uncertainty in index and age composition observations, the largest covariate effect size, lower variability in the latent covariate, and larger uncertainty in the covariate observations (Figure \ref{annual_M_bias_M_fixed_beta_estimated}).
+There was little difference in the results for R+S EMs whether $\beta_E$ was estimated or not.
+The results for R+M EMs were generally similar to those when $\beta_E=0$ was assumed, except that the differences between the estimates at the beginning of the time series and later years for certain OM configurations did not occur.
 
@@
 
-Like natural mortality, we present results for error in spawning biomass conditional on three alternative EM configurations for the natural mortality parameters: 1) mean natural mortality rate parameter is fixed at the true value ($\beta_M = log(0.2)$) and no covariate effect is assmed ($\beta_\text{Ecov} = 0$),  2) $\beta_M = log(0.2)$ and $\beta_\text{Ecov}$ is estimated, and 3) both $\beta_M$ and $\beta_\text{Ecov}$ are estimated.
+Like natural mortality, we present results for error in spawning biomass conditional on three alternative EM configurations for the natural mortality parameters: 1) mean natural mortality rate parameter is fixed at the true value ($\beta_M = log(0.2)$) and no covariate effect is assmed ($\beta_E = 0$),  2) $\beta_M = log(0.2)$ and $\beta_E$ is estimated, and 3) both $\beta_M$ and $\beta_E$ are estimated.
 
-When EMs assume $\beta_M = log(0.2)$ and $\beta_\text{Ecov} = 0$, there is little evidence of bias for R and R+M OMs except when there is a step change in fishing mortality and higher uncertainty in index and age composition observations, particularly at the end of the time series (Figure \ref{SSB_bias_M_fixed_beta_fixed}).
+When EMs assume $\beta_M = log(0.2)$ and $\beta_E = 0$, there is little evidence of bias for R and R+M OMs except when there is a step change in fishing mortality and higher uncertainty in index and age composition observations, particularly at the end of the time series (Figure \ref{SSB_bias_M_fixed_beta_fixed}).
 We observed a similar trend in median relative error for R+S OMs, but there was more evidence of positive bias at the beginning of the time series whereas the confidence intervals for the negative median errors at the end of the time series often included 0.
@@
 
-When EMs assume $\beta_M = log(0.2)$, but estimate $\beta_\text{Ecov}$, the median errors for SSB for R and and R+M OMs are similar to those when $\beta_\text{Ecov} = 0$ (Figure \ref{SSB_bias_M_fixed_beta_estimated}).
+When EMs assume $\beta_M = log(0.2)$, but estimate $\beta_E$, the median errors for SSB for R and and R+M OMs are similar to those when $\beta_E = 0$ (Figure \ref{SSB_bias_M_fixed_beta_estimated}).
 For R+S OMs, the median errors for EMs that also assume R+S process errors are also similar to those with no covariate effect assumed.
@@
 
-When EMs estimated both $\beta_M$ and $\beta_\text{Ecov}$, we observed large variation in the errors in SSB under the same OM configurations where we observed large variation in errors for annual natural mortality rates (Figure \ref{SSB_bias_M_estimated_beta_estimated}).
+When EMs estimated both $\beta_M$ and $\beta_E$, we observed large variation in the errors in SSB under the same OM configurations where we observed large variation in errors for annual natural mortality rates (Figure \ref{SSB_bias_M_estimated_beta_estimated}).
 Similarly, we observed little or no bias in SSB estimation for R+S EMs across all OM process error assumptions as long as there was a step change in fishing mortality and lower uncertainty in indices and age composition.
@@
 @derisoetal08 was the first to model natural mortality as a function of explicit covariate and residual random annual variation. Any relevant points?
+
+Poor quality of biological, fishery, and observation may adversely affect ability to distinguish between true process and observational error [@puntetal14; @stewartmonnahan17; @croninpunt21; @fischetal23; @lietal_inreview_a].
 
@@
 \begin{center}
-%\includegraphics{convergence}
+\includegraphics{convergence}
 \end{center}
@@
 \begin{center}
-%\includegraphics[height = \textheight]{aic}
+\includegraphics[height = \textheight]{aic_main}
 \end{center}
-\caption{Proportion of simulated data sets for each operating model where the estimation model type (combination of assumed process error type and covariate effect) had the lowest AIC.
-All EMs had known median natural mortality rate and all OMs and EMs had low observation error.}\label{aic}
+\caption{Proportion of simulated data sets for each operating model where the estimation model type (treatment of environmental covariate effect and  assumed process error type) had the lowest AIC.
+All OMs had low observation error for fish population observations and temporal contrast in fishing pressure.
+All EMs estimated median natural mortality rate.}\label{aic}
 \end{figure}
@@
 \end{center}
-\caption{Relative error of estimates of environmental effect on natural mortality $\beta_\text{Ecov}$ from fitting EMs with alternative process error assumptions.
+\caption{Relative error of estimates of environmental effect on natural mortality $\beta_E$ from fitting EMs with alternative process error assumptions.
 All OMs had low observation error and contrast in fishing mortality.
@@
 \caption{Median relative error of annual natural mortality rate in terminal year from fitting simulated observations from each OM with alternative process error assumptions in the EM.
-Mean log-natural mortality parameter ($\beta_\text{M}$) and environmental covariate effect ($\beta_\text{Ecov}$) are both estimated in the EMs.
+Mean log-natural mortality parameter ($\beta_M$) and environmental covariate effect ($\beta_E$) are both estimated in the EMs.
 All OMs had low observation error and contrast in fishing mortality.}\label{terminal_M_bias}
@@
 \caption{Median relative error of annual SSB in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
-Estimating models also assume the mean natural mortality parameter $\beta_\text{M} = \log 0.2$ and no environmental covariate effect $\beta_\text{Ecov} = 0$.}\label{SSB_bias_M_fixed_beta_fixed}
+Estimating models also assume the mean natural mortality parameter $\beta_M = \log 0.2$ and no environmental covariate effect $\beta_E = 0$.}\label{SSB_bias_M_fixed_beta_fixed}
 \begin{center}
@@
 \caption{Median relative error of annual SSB in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
-Estimating models also assume the mean natural mortality parameter $\beta_\text{M} = \log 0.2$ and the environmental covariate effect $\beta_\text{Ecov}$ is estimated.}\label{SSB_bias_M_fixed_beta_estimated}
+Estimating models also assume the mean natural mortality parameter $\beta_M = \log 0.2$ and the environmental covariate effect $\beta_E$ is estimated.}\label{SSB_bias_M_fixed_beta_estimated}
 \begin{center}
@@
 \caption{Median relative error of annual SSB in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
-Estimating models also assume both the mean natural mortality parameter $\beta_\text{M}$ and the environmental covariate effect $\beta_\text{Ecov}$ are estimated.}\label{SSB_bias_M_estimated_beta_estimated}
+Estimating models also assume both the mean natural mortality parameter $\beta_M$ and the environmental covariate effect $\beta_E$ are estimated.}\label{SSB_bias_M_estimated_beta_estimated}
 \begin{center}
@@
 
+\begin{figure}
+\begin{center}
+\includegraphics[width = \textwidth]{om_input_plots_figure}
+\end{center}
+\caption{The proportion mature at age, weight at age, fleet and index selectivity at age, and Beverton-Holt stock-recruit relationship assumed for the population in all operating models.
+For operating models with random effects on fleet selectivity, this represents the selectivity at the mean of the random effects.}\label{om_inputs_fig}
+\end{figure}
+
 \begin{landscape}
@@
 \begin{center}
-%\includegraphics[height = \textheight]{aic_supp1}
+\includegraphics{convergence_Rom}
 \end{center}
-\caption{Proportion of simulated data sets for each OM where the EM type (assumed process error type) had the lowest AIC.
-All EMs had known median mortality rate and all OMs and EMs had low observation error.}\label{aic_supp1}
+\caption{Estimated probability of fits providing hessian-based standard errors for EMs assuming alternative process error, that estimate or assume known median natural mortality, and that estimate or assume no covariate effect on median natural mortality when fitted to R OMs and three levels of true covariate effect on median natural mortality (x axis).
+Vertical lines represent 95\% confidence intervals.}\label{convergence_Rom}
 \end{figure}
@@
 \begin{center}
-%\includegraphics[height = \textheight]{aic_supp2}
+\includegraphics{convergence_RSom}
 \end{center}
-\caption{Proportion of simulated data sets for each operating model where the estimation model type (combination of assumed process error type and covariate effect) had the lowest AIC.
-All OMs had a change in fishing pressure over time and all OMs and EMs had low observation error.}\label{aic_supp2}
+\caption{Estimated probability of fits providing hessian-based standard errors for EMs assuming alternative process error, that estimate or assume known median natural mortality, and that estimate or assume no covariate effect on median natural mortality when fitted to R+S OMs and three levels of true covariate effect on median natural mortality (x axis).
+Vertical lines represent 95\% confidence intervals.}\label{convergence_RSom}
 \end{figure}
@@
 \begin{center}
-%\includegraphics[height = \textheight]{aic_supp3}
+\includegraphics{convergence_RMom}
 \end{center}
-\caption{Proportion of simulated data sets for each operating model where the estimation model type (combination of assumed process error type and covariate effect) had the lowest AIC.
-All EMs had known median natural mortality rate and all OMs and EMs had high observation error.}\label{aic_supp3}
+\caption{Estimated probability of fits providing hessian-based standard errors for EMs assuming alternative process error, that estimate or assume known median natural mortality, and that estimate or assume no covariate effect on median natural mortality when fitted to R+M OMs and three levels of true covariate effect on median natural mortality (x axis).
+Vertical lines represent 95\% confidence intervals.}\label{convergence_RMom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{aic_Rom}
+\end{center}
+\caption{Proportion of simulated data sets for R OMs where the EM type (treatment of environmental covariate and assumed process error type) had the lowest AIC.}\label{aic_Rom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{aic_RSom}
+\end{center}
+\caption{Proportion of simulated data sets for R+S OMs where the EM type (treatment of environmental covariate and assumed process error type) had the lowest AIC.}\label{aic_RSom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{aic_RMom}
+\end{center}
+\caption{Proportion of simulated data sets for R+M OMs where the EM type (treatment of environmental covariate and assumed process error type) had the lowest AIC.}\label{aic_RMom}
 \end{figure}
@@
 \caption{Median relative error of annual natural mortality rate in terminal year from fitting simulated observations from each OM with alternative process error assumptions in the EM.
-Mean log-natural mortality parameter ($\beta_\text{M}$) is estimated, but EMs assume no environmental covariate effect ($\beta_\text{Ecov} = 0 $) are both estimated in the EMs.
+Mean log-natural mortality parameter ($\beta_M$) is estimated, but EMs assume no environmental covariate effect ($\beta_E = 0 $) are both estimated in the EMs.
 All OMs had low observation error and contrast in fishing mortality.}\label{terminal_M_bias_beta_0}
@@
 \caption{Median relative error of annual fully-selected fishing mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
-Estimating models also assume the mean natural mortality parameter $\beta_\text{M} = \log 0.2$ and no environmental covariate effect $\beta_\text{Ecov} = 0$.}\label{F_bias_M_fixed_beta_fixed}
+Estimating models also assume the mean natural mortality parameter $\beta_M = \log 0.2$ and no environmental covariate effect $\beta_E = 0$.}\label{F_bias_M_fixed_beta_fixed}
 \begin{center}
@@
 \caption{Median relative error of annual fully-selected fishing mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
-Estimating models also assume the mean natural mortality parameter $\beta_\text{M} = \log 0.2$ and the environmental covariate effect $\beta_\text{Ecov}$ is estimated.}\label{F_bias_M_fixed_beta_estimated}
+Estimating models also assume the mean natural mortality parameter $\beta_M = \log 0.2$ and the environmental covariate effect $\beta_E$ is estimated.}\label{F_bias_M_fixed_beta_estimated}
 \begin{center}
@@
 \caption{Median relative error of annual fully-selected fishing mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
-Estimating models also assume both the mean natural mortality parameter $\beta_\text{M}$ and the environmental covariate effect $\beta_\text{Ecov}$ are estimated.}\label{F_bias_M_estimated_beta_estimated}
+Estimating models also assume both the mean natural mortality parameter $\beta_M$ and the environmental covariate effect $\beta_E$ are estimated.}\label{F_bias_M_estimated_beta_estimated}
 \begin{center}

### 2025-05-21 Ecov_study/mortality/manuscript/manuscript.tex
saving work on Ecov-M

@@
 Legault\textsuperscript{2} \and Brandon Muffley\textsuperscript{5} \and John Wiedenmann\textsuperscript{6}}
-\date{13 May, 2025}
+\date{21 May, 2025}
 
@@
 State-space stock assessment models, with non-linear functions of latent processes and numerous observation types with different probability distribution assumptions represent one of most complex classes of state-space models.
-The literature on the ways we make inferences and the effects of various factors on reliabilty of inferences from state-space assessment models is growing \citep{lietal24, milleretal_inreview1, cadiganetal}.
+The literature on the ways we make inferences and the effects of various factors on reliabilty of inferences from state-space assessment models is growing {[}\citet{lietal24};\citet{milleretal_inreview1};cadiganetal{]}.
 The importance of contrast in population size and fishing mortality and quality of data used to fit assessment models including the state-space variety is known \citep{magnussonhilborn07, milleretal_inreview1}.
-Furthermore, estimation of natural mortality, and even temporal variablity is possible in many scenarios {[}\citet{leeetal11};\href{mailto:johnsonetal15;@cadigan16}{\nolinkurl{johnsonetal15;@cadigan16}};\citet{millerhyun18};\citet{milleretal_inreview1}{]}.
+Furthermore, estimation of natural mortality, and even temporal variablity is possible in many scenarios \citep{leeetal11, johnsonetal16, cadigan16, millerhyun18, milleretal_inreview1}.
 
-The effects of temporal variation in recruitment via unspecified or specified environmental factors have been extensively investigated in both traditional assessment models and state space models \citep{myers98, milleretal16, haltuchpapers}.
+The effects of temporal variation in recruitment via unspecified or specified environmental factors have been extensively investigated in both traditional assessment models and state space models \citep[haltuchpapers]{myers98, milleretal16}.
 Reliability of estimating environmental and spawning biomass effects on recruitment requires a combination of strong effects, good age composition data quality, contrast in the environmental covariate and lower recruitment variability \citep{brittenetal_inreview, milleretal_inreview1}.
 
-Temporal and environmental effects on growth and weight at age are also more important for short term projections and the reliability of estimation of those effects \citet{correaetal23}; \citet{correaetalinreview}.
+Temporal and environmental effects on growth and weight at age are also more important for short term projections and the reliability of estimation of those effects \citep{correaetal23, correaetal_inreview}.
 
@@
 
-See \citet{millerretal_inreview1} fro relevance of project 0 results (estimability of natural mortality).
-Some difficulty in distinquishing variation in natural mortality or effects of explicit covariates when variation in M random effects or Ecov is low relative to observation error.
+See \citet{milleretal_inreview1} for relevance of project 0 results (estimability of natural mortality).
+Some difficulty in distinguishing variation in natural mortality or effects of explicit covariates when variation in M random effects or Ecov is low relative to observation error.
 
@@
 
-Incorrect treatment of population attributes as temporally varying \citep{trijouletetal20, liljestrandetal24} could lead to misidentification of stock status and biased population estimates, ultimately impacting fisheries management decisions \citep{legaultpalmer16, szuwalskietal18, croninpunt21}.
-Furthermore, biological, fishery, and observational processes are often confounded in catch-at-age data, which may adversely affect ability to distinguish between true process variability and observational error \citep{puntetal14, stewartmonnahan17, croninpunt21, fischetal23, lietalinreview_a}.
+Mis-specified temporal population process errors could lead to misidentification of stock status and biased population estimates and poor fisheries management decisions \citep{trijouletetal20, legaultpalmer16, szuwalskietal18, croninpunt21, liljestrandetal24}.
 
@@
 
-Estimation of natural mortality is known to be challenging in stock assessment models (Lee et al., others). However, estimation has been shown to be reliable in some situations \citet{millerhyun18})
+Estimation of natural mortality is known to be challenging in stock assessment models (\citet{leeetal11}, others).
+However, estimation has been shown to be reliable in some situations \citep{millerhyun18}
 
@@
 
-All of our analyses used the Woods Hole Assessment Model (WHAM) to construct both OMs and EMs \citep{millerstock20, stockmiller21, milleretal_inreview1}.
-We used \href{https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3}{version 1.0.6.9000, commit 77bbd94} for this simulation study.
-This package has been used extensively to configure OMs and EMs for several other simulation studies \citep[\citet{lietalinreview}]{legaultetal23, lietal24} and is used to assess many commercially important stocks in the Northeast US.
+All of our analyses used the Woods Hole Assessment Model (WHAM) to construct both OMs and EMs \citep{millerstock20, stockmiller21, milleretal_inreview}.
+The WHAM package has been used extensively to configure OMs and EMs for several other simulation studies \citep{legaultetal23, lietal24, brittenetal_inreview, lietal_inreview_a} and is used to assess many commercially important stocks in the Northeast U.S.
+\citep[e.g.,][]{nefsc22, nefsc22a, nefsc24}.
+We used \href{https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3}{version 1.0.6.9000, commit 77bbd94} for to generate all results.
 
--------------- need to edit this part
+We completed a simulation study with a 288 operating models.
+The factors defining the configuration of each operating model which are described in detail in subsequent sections include source of population process error (3 levels), index and catch observation uncertainty (2 levels), environmental covariate uncertainty (2 levels), latent environmental covariate process stochasticity (4 levels), and fishing history (2 levels).
+We simulated 100 data sets for each operating model that included simulations of process errors.
 
-There are many similarities to the configuration of OMs and EMs in simulation studies conducted by \citet{milletetelinreview}.
-We completed a simulation study with a number of OMs that can be categorized based on where process error random effects are assumed: abundance at age (R, R+S), natural mortality (R+M), fleet selectivity (R+Sel), or index catchability (R+q).
-For each OM assumption about variance of process errors and observations are required and the values we used were based on a review of the range of estimates from applications of WHAM in management of stocks of haddock, butterfish, and American plaice in the NE US.
+For each simulated data set we fit a set of 12 estimating models (EMs).
+The factors that distinguish the estimating models which are also described in detail below include source of population process error type (3 levels) whether (median) natural mortality rate was estimated or assumed known (2 levels), and whether the effect of the environmental covariate on natural mortality was estimated or not (2 levels).
 
-In total, we configured 72 OMs with alternative assumptions about the source and variability of process errors, level of observation error in indices and age composition data, and contrast in fishing pressure over time.
-We fitted 20 EMs to observations from each of 100 simulations where process errors were also simulated.
-For R+M, R+Sel, and R+q OMs, we used unique seeds for each simulation, but we inadvertently used the same 100 seeds for all R and R+S OMs.
-Each of the EMs made alternative assumptions about the source of process errors and whether natural mortality (or the median for models with process error in natural mortality) was estimated and whether a Beverton-Holt stock recruit relationship was estimated within the EM.
-Details of each of the operating and EMs are described below.
-
-We did not use the log-normal bias-correction feature for process errors or observations described by \citet{stockmiller21} for operating and EMs \citep{lietalinreview}.
-Simulations and model fitting were all carried out on the University of Massachusetts Green High-Performance Computing Cluster.
-All code we used to perform the simulation study and summarize results can be found at \url{https://github.com/timjmiller/SSRTWG/tree/main/ecov_study/mortality/code}.
-
-\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
-
---------------------------- From project\_0 paper
-
-We used the Woods Hole Assessment Model (WHAM) to configure OMs and EMs in our simulation study \citep{millerstock20, stockmiller21}.
-WHAM is an R package freely available via a github repository and is built on the Template Model Builder package \citep{kristensenetal16}.
-For this study we used \href{https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3}{version 1.0.6.9000, commit 77bbd94}.
-WHAM has also been used to configure OMs and EMs for closed loop simulations evaluating index-based assessment methods \citep{legaultetal23} and is currently used or accepted for use in management of numerous NEUS fish stocks \citep[e.g.,][]{nefsc22, nefsc22a, nefsc24}.
-
-We completed a simulation study with a number of OMs that can be categorized based on where process error random effects were assumed: recruitment (R, assumed present in all models), apparent survival (denoted R+S), natural mortality (R+M), fleet selectivity (R+Sel), or index catchability (R+q).
-We refer to the (R+S) OMs as modeling apparent survival because on logscale the random effects (\(\epsilon_{a,y}\)) are additive to the total mortality (F+M) between numbers at age, thus they modify the survival term.
-However, as \citet{stockmiller21} note, these random effects can be due to events other than mortality, such as immigration, emigration, missreported catch, and other sources of misspecification.
-For each OM, assumptions about the magnitude of the variance of process errors and observations are required and the values we used were based on a review of the range of estimates from Northeast Unite States (NEUS) assessments using WHAM.
-
-In total, we configured 72 OMs with alternative assumptions about the source and magnitude of process errors, magnitude of observation error in indices and age composition data, and contrast in fishing pressure over time.
-We fitted 20 EMs to observations generated from each of 100 simulations where process errors were also simulated.
-Each EM differed in assumptions about the source of process errors, whether natural mortality (or the median for models with process error in natural mortality) was estimated, and whether a Beverton-Holt SRR was estimated within the EM.
-Details of each of the OMs and EMs are described below.
-
-We did not use the log-normal bias-correction feature for process errors or observations described by \citep{stockmiller21} for OMs and EMs to simplify interpretation of the study results \citep{lietalinreview}.
-All code we used to perform the simulation study and summarize results can be found at \url{https://github.com/timjmiller/SSRTWG/tree/main/Project_0/code}.
-
-\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
-
-We used the Woods Hole Assessment Model (WHAM) to configure operating and estimation models in our simulation study \citep{millerstock20, stockmiller21}.
-WHAM is an R package freely available as a github repository.
-For this study we used \href{https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3}{version 1.0.6.9000, commit 77bbd94}.
-This packages has also been used to configure operating and estimating models for closed loop simulations evaluating index-based assessment methods \citep{legaultetal23} and is used for management of haddock, butterfish, American plaice, bluefish, Atlantic cod, and black sea bass in the Northeast US.
-
-We completed a simulation study with a number of operating models.
-Many of the characteristics of the operating models are the same as those used in the working paper by Miller et al.~(WP 4).
-The factors defining the configuration of each operating model which are described in detail in subsequent sections include
-
-\begin{itemize}
-\item two configurations of fishing histories, 
-\item two configurations of index and age composition observation error, 
-\item four configurations of latent environmental covariate process errors, 
-\item two configurations of environmental covariate observation uncertainty, 
-\item three levels of covariate effects on natural mortality, and 
-\item three configurations of sources of process errors for temporal changes in abundances at age. 
-\end{itemize}
-
-The full factorial of the levels of each factor results in 288 different operating models. We simulated 100 data sets for each operating model that included simulations of process errors.
-
-For each simulated data set we fit a set of 12 estimating models. The factors defining the configuration or each estimating model which are also described in detail below include
-
-\begin{itemize}
-\item three configurations of sources of process errors for temporal changes in abundances at age, 
-\item whether the (mean) natural mortality rate was estimated or assumed known at the true value, and
-\item whether the effect of the environmental covariate on natural mortality was estimated.
-\end{itemize}
-
-We did not use the bias-correction feature for process errors or observations described by \citep{stockmiller21} for operating and estimating models.
+The sources of population process error that were used in the OMs or assumed in the EMs were on recruitment only (R), recruitment and changes in cohort abundance over time (R+S), or recruitment and natural mortality (R+M).
+We did not use the log-normal bias-correction feature for process errors or observations described by \citet{stockmiller21} for operating and EMs \citep{lietal_inreview}.
 Simulations were all carried out on the University of Massachusetts Green High-Performance Computing Cluster.
@@
 
+\hypertarget{environmental-covariate}{%
+\subsubsection*{Environmental covariate}\label{environmental-covariate}}
+\addcontentsline{toc}{subsubsection}{Environmental covariate}
+
+In the WHAM model, environmental covariates are assumed to be described as state-space processes with annual observations of the true latent covariate {[}\citet{milleretal16};stockmiller21{]}.
+In our simulations, the latent covariate is assumed to be a stationary first order autoregressive (AR1) process
+\[
+X_y|X_{y-1} \sim \text{N}\left(\mu_E\left(1-\rho_E\right) + \rho_E E_{y-1}, \left(1-\rho_E^2\right)\sigma^2_E\right)
+\]
+with marginal mean \(\mu_E=0\) and variance \(\sigma^2_E\).
+The four configuration of the latent environmental covariate in the operating models assume the one of two values for the marginal standard deviation (\(\sigma_E \in \{0.1, 0.5\}\)) and for the autocorrelation parameter (\(\rho_E \in \{0, 0.5\}\)).
+
+The observations of the latent environmental covariate are assumed to be unbiased and Gaussian
+\[
+x_y|X_y \sim \text{N}\left(X_y,\sigma^2_e\right)
+\]
+The standard deviation of the environmental observations in the operating models is one of two values \(\sigma_E \in \{0.1, 0.5\}\).
+Figure \ref{om_ecov_example} provides example simulations of the latent process and observations under the alternative configurations.
+
 \hypertarget{population}{%
@@
 
-The population model tracks 10 age classes: ages 1 to 10+ and we assume spawning occurs 1/4 of the way through the year.
-The maturity at age was a logistic curve with \(a_{50}\) = 2.89 and slope = 0.88 and assumed known in all estimation models (Figure \ref{om_maturity}).
-
-Weight at age was generated with a LVB growth function
+Many of the characteristics of the population biology and structure are the same as those in \citet{milleretal_inreview1}.
+We configures the stock with 10 age classes (ages 1 to 10+) where the last is a ``plus'' group that accumulates any older individuals.
+We model the population for 40 years with annual catch and index observations.
+We assume spawning occurs annually 1/4 of the way through the year.
+The maturity at age was a logistic function of age with age at 50\% maturity (\(a_{50}\)) = 2.89 and slope = 0.88 (Figure \ref{om_inputs_fig}, top left).
+We assumed a von Beralanffy growth function for weight at age with length at age defined as
 \[
@@
 \]
-where \(t_0 = 0\), \(L_\infty = 85\), and \(k = 0.3\), and a L-W relationship such that
+where \(t_0 = 0\), \(L_\infty = 85\), and \(k = 0.3\), and a length-weight
 \[
@@
 \]
-where \(\theta_1 = e^{-12.1}\) and \(\theta_2 = 3.2\) (Figure \ref{om_waa}).
+where \(\theta_1 = e^{-12.1}\) and \(\theta_2 = 3.2\) (Figure \ref{om_inputs_fig}, top right).
 
-We assumed a Beverton-Holt stock recruit function with constant pre-recruit mortality parameters (i.e., \(\alpha\) and \(\beta\)) for all operating models.
-All post-recruit productivity components are constant in the NAA and survey selectivity process error operating models.
-Therefore steepness and unfished recruitment are also constant over the time period for those operating models \citep{millerbrooks21}.
-We specified unfished recruitment = \(R_0 = e^{10}\) and \(\Fmsy = F_{40\%} = 0.348\) equated to a steepness of 0.69 and \(\alpha=0.60\) and \(\beta = 2.4 \times 10^{-5}\) for the
+The general model for natural mortality in year \(y\) is a log-linear function of a mean(median) parameter \(\beta_{M}\), effects of an environmental covariate \(E_y\) and a possibly autocorrelated normally distributed annual process error \(\varepsilon_{M,y}\)
 \[
-N_{1,y} = \frac{\alpha \text{SSB}_{y-1}}{1 + \beta \text{SSB}_{y-1}} 
+\log M_y = \beta_M + \beta_{E} E_y + \varepsilon_{M,y}
 \]
-Beverton-Holt parameterization (Figure \ref{om_sr}).
-For OMs without process errors on natural mortality we assumed the rate was assumed 0.2.
-For OMs with process errors on natural mortality the mean rate was 0.2.
-
-Two alternative fishing histories were used for operating models.
-In the first scenario, the stock experiences overfishing for the first 20 years and fishing at \Fmsy for the last 20 years.
-In the second scenario, the stock is fished at \Fmsy for the entire time period.
-The magnitude of the overfishing assumptions is based on average estimates of overfishing for NE groundfish stocks from \citet{wiedenmannetal19}.
-\citet{legaultetal23} also used similar approaches to defining fishing mortality histories for operating models.
-
-Initial population was configured at the equilibrium distribution fishing at either \(F = 2.5\times \Fmsy\) or \(F = \Fmsy\) for the two alternative fishing histories.
-That is for a deterministic model, the age composition would not change over time when the fishing mortality was constant at the respective level.
-
-For operating models with time-varying random effects or covariate effects on M, steepness is not constant, but we used the same alpha and beta parameters as other operating models this equates to a steepness and \(R_0\) at the mean of the time series process for M.
-For operating models with time-varying random effects for fishery selectivity, \Fmsy is also not constant however we use the same F history as other operating models which corresponds to \Fmsy at the mean selectivity parameters.
-
-\hypertarget{environmental-covariate}{%
-\subsubsection*{Environmental covariate}\label{environmental-covariate}}
-\addcontentsline{toc}{subsubsection}{Environmental covariate}
-
-In the WHAM model, environmental covariates are assumed to be described as state-space processes with annual observations of the true latent covariate.
-The latent covariate is assumed to be Gaussian AR1
-\[
-E_y|E_{y-1} \sim \text{N}\left(\mu_\text{Ecov}\left(1-\rho_\text{Ecov}\right) + \rho_\text{Ecov} E_{y-1}, \left(1-\rho_\text{Ecov}^2\right)\sigma^2_\text{Ecov}\right)
-\]
-with marginal mean \(\mu_\text{Ecov}=0\) and variance \(\sigma^2_\text{Ecov}\).
-The configuration of the latent environmental covariate in the operating models have one of two alternate assumptions about the marginal standard deviation and autocorrelation parameter:
-
-\begin{itemize}
-\item marginal standard deviation $\sigma_\text{Ecov} \in \{0.1, 0.5\}$
-\item AR1 correlation $\rho_\text{Ecov} \in \{0, 0.5\}$
-\end{itemize}
-
-The observations of the latent environmental covariate are assumed to be unbiased and Gaussian
-\[
-\text{ecov}_y|E_y \sim \text{N}\left(E_y,\sigma^2_\text{ecov}\right)
-\]
-The standard deviation of the environmental observations in the operating models is one of two values:
-
-\begin{itemize}
-\item standard deviation $\in \{0.1, 0.5\}$
-\end{itemize}
-
-Figure \ref{om_ecov_example} provides example simulations of the latent process and observations under the alternative configurations.
-
-\hypertarget{random-effects-on-recruitment-survival-and-natural-mortality}{%
-\subsubsection*{Random effects on recruitment, survival, and natural mortality}\label{random-effects-on-recruitment-survival-and-natural-mortality}}
-\addcontentsline{toc}{subsubsection}{Random effects on recruitment, survival, and natural mortality}
-
-Internally, WHAM treats natural mortality as a log-transformed parameter
-\[
-\log M_y = \beta_M + \beta_{\text{Ecov}} E_y + \varepsilon_{M,y}
-\]
-that is a linear combination of a mean \(\beta_M\) and any annual covariate \(\beta_{\text{Ecov}}\) and random effects marginally distributed as \(\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)\) \citep{stockmiller21}.
-For operating models with natural mortality random effects, we assume the same standard deviation \(\sigma_M = 0.3\).
-This is similar to the intermediate value used for R+M OMs by \citet{milleretal_inreview1}.
-
-The covariate effect is one of 3 alternative values in the operating models:
-
-\begin{itemize}
-\item $\beta_\text{Ecov} \in \{0,0.25,0.5\}$
-\end{itemize}
-
+\(\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)\) \citep{stockmiller21}.
+We assume the median natural mortality rate \(e^{\beta_M} = 0.2\) is constant across ages.
+For R and R+S OMs and EMs, the process errors are set equal to 0.
+For all R+M OMs, we assume the same standard deviation \(\sigma_M = 0.3\) and is estimated in the R+M EMs.
+The covariate effect is one of 3 alternative values in the operating models (\(\beta_\text{E} \in \{0,0.25,0.5\}\).
 The parameters defining the simulated covariate time series, size of the covariate effect, and any natural mortality random effects result in a range of different levels of variation in annual natural mortality rates (Figure \ref{M_example}).
 
-We configured all OMs with uncorrelated random effects on recruitment with standard devaition on log(recruitment) \(\sigma_R = 0.5\) for R, R+S, and R+M OMs.
-This same assumption was used by \citet{milleretalinreview} for R+M OMs and other OMs with fishery selectivity and index catchability process errors.
-For R+S OMs, apparent survival process errors were uncorrelated with \(\sigma_{2+} = 0.3\) and for R+M OMs, process errors on natural mortality were uncorrelated with \(\sigma_M = 0.3\).
+We assume is a single fishing fleet harvesting the population with a logistic selectivity function with age at 50\% selection \(a_{50} = 5\) and slope = 1 (Figure \ref{om_inputs_fig}, bottom left).
 
-\hypertarget{fleets}{%
-\subsubsection*{Fleets}\label{fleets}}
-\addcontentsline{toc}{subsubsection}{Fleets}
+We assumed expected recruitment each year from a Beverton-Holt stock-recruit relationship (SRR)
+\[
+R_{y} = \frac{a \text{SSB}_{y-1}}{1 + b \text{SSB}_{y-1}}.
+\]
+All biological inputs to calculations of spawning biomass per recruit (i.e., weight, maturity, and natural mortality at age) are constant in the R and R+S OMs without covariate effects on natural mortality.
+Therefore, steepness and unfished recruitment are also constant over the time period for those OMs \citep{millerbrooks21} and with our assumed biological inputs and selectivity the fishing mortality resulting in 40\% reduction in spawning biomass per recruit is \(F_{40\%} = 0.348\).
+With an assumed unfished recruitment of \(R_0 = e^{10}\), setting \(\Fmsy = \Fspr[40]\) results in a steepness of 0.69 and \(a=0.60\) and \(b = 2.4 \times 10^{-5}\) (Figure \ref{om_inputs_fig}, bottom right).
 
-We assumed a single fleet operating year round for catch observations with logistic selectivity for the fleet with \(a_{50} = 5\) and slope = 1 (Figure \ref{om_mean_selectivity}).
+We used two fishing scenarios for OMs.
+In the first scenario, the stock experiences overfishing at 2.5\Fmsy for the first 20 years followed by fishing at \Fmsy for the last 20 years (denoted \(2.5\Fmsy \rightarrow \Fmsy\)).
+In the second scenario, the stock is fished at \Fmsy for the entire time period (40 years).
+The magnitude of the overfishing assumptions is based on average estimates of overfishing for NEUS groundfish stocks from \citet{wiedenmannetal19} and similar to the approach in \citet{legaultetal23}.
+
+We specified initial population abundance at age at the equilibrium distribution that corresponds to fishing at either \(F = 2.5\times \Fmsy\) or \(F = \Fmsy\).
+This implies that, for a deterministic model, the abundance at age would not change from the first year to the next.
+
+For R+M OMs and all OMs with environmental covariate effects on natural mortality, steepness is not constant, but we used the same \(a\) and \(b\) parameters as other operating models which equates to a steepness and \(R_0\) at the median of the time series process for M.
+For operating models with time-varying random effects for fishery selectivity, \Fmsy is also not constant however we use the same fishing history as other operating models which corresponds to \Fmsy at the mean selectivity parameters.
+
+We configured all R, R+S, and R+M OMs with uncorrelated random effects on recruitment with standard deviation on log(recruitment) \(\sigma_R = 0.5\).
+This same assumption was used by \citet{milleretal_inreview1} for R+M OMs and other OMs with fishery selectivity and index catchability process errors.
+For R+S OMs, cohort temporal transition process errors were uncorrelated with \(\sigma_{2+} = 0.3\)
+
+\hypertarget{catch-and-index-observations}{%
+\subsubsection*{Catch and index observations}\label{catch-and-index-observations}}
+\addcontentsline{toc}{subsubsection}{Catch and index observations}
+
+We assumed a single fleet operating year round for catch observations with logistic selectivity for the fleet with \(a_{50} = 5\) and slope = 1 (Figure \ref{om_inputs_fig}).
 This selectivity was used to define \Fmsy, \Fspr[40], and steepness for the Beverton-Holt stock recruitment parameters above.
@@
 
-\hypertarget{indices}{%
-\subsubsection*{Indices}\label{indices}}
-\addcontentsline{toc}{subsubsection}{Indices}
-
 Two time series of surveys are assumed and observed in numbers rather than biomass for the entire 40 year period with one occurring in the spring (0.25 way through the year) and one in the fall (0.75 way through the year).
@@
 We assumed the same selectivity and age composition distribution as the fleet for both indices.
-
-\hypertarget{observation-uncertainty}{%
-\subsubsection*{Observation Uncertainty}\label{observation-uncertainty}}
-\addcontentsline{toc}{subsubsection}{Observation Uncertainty}
 
@@
 
-\hypertarget{measures-of-reliability}{%
-\subsection*{Measures of reliability}\label{measures-of-reliability}}
-\addcontentsline{toc}{subsection}{Measures of reliability}
+\hypertarget{performance-measures}{%
+\subsection*{Performance measures}\label{performance-measures}}
+\addcontentsline{toc}{subsection}{Performance measures}
 
@@
 
-\hypertarget{convergence-performance}{%
-\subsection*{Convergence performance}\label{convergence-performance}}
-\addcontentsline{toc}{subsection}{Convergence performance}
+Because \citet{milleretal_inreview1} found inferences are most reliable for scenarios with low observation error for indices and age composition data and with temporal contrast in fishing pressure, we restrict our attention to scenarios with these characteristics in the main paper (temporal contrast in fishing and low observation error), but include corresponding results for the other scenarios in the Supplementary Materials.
 
-When the (mean) log-natural mortality rate was assumed at the true value, corresponding to usual practice in application of assessment models, good convergence for all EMs was observed for R+S operating models with \Fmsy fishing history, more variation in the latent environmental covariate (\(\sigma_\text{Ecov} = 0.5\)), and lower error in the associated observations \(\sigma_\text{ecov} = 0.1\) (Figure \ref{convergence_M_fixed}).
+\hypertarget{convergence-1}{%
+\subsection*{Convergence}\label{convergence-1}}
+\addcontentsline{toc}{subsection}{Convergence}
+
+When the median natural mortality rate was assumed known, corresponding to usual practice in application of assessment models, good convergence for all EMs was observed for R+S operating models with \Fmsy fishing history, more variation in the latent environmental covariate (\(\sigma_E = 0.5\)), and lower error in the associated observations \(\sigma_e = 0.1\) (Figure \ref{convergence}).
 Good convergence for all EMs was also observed for R+S OMs with the step change in fishing history, but with low error in indices and age comp observations.
 
-EMs that assumed random effects just on recruitment (R), nearly always converged across all operating models that assumed more variation in the latent environmental covariate (\(\sigma_\text{Ecov} = 0.5\)) and lower error in the associated observations \(\sigma_\text{ecov} = 0.1\).
+EMs that assumed random effects just on recruitment (R), nearly always converged across all operating models that assumed more variation in the latent environmental covariate (\(\sigma_E = 0.5\)) and lower error in the associated observations \(\sigma_E = 0.1\).
 EMs that assumed recruitment and survival random effects (R+S), had poor converge probability when the OMs had alternative process errors (R or R+M).
-The R+S EMs showed best convergence for the R+S OMs where there was lower error in the environmental observations \(\sigma_\text{ecov} = 0.1\) or higher error in environmental observations of latent true environmental covariates with greater temporal variation.
+The R+S EMs showed best convergence for the R+S OMs where there was lower error in the environmental observations \(\sigma_E = 0.1\) or higher error in environmental observations of latent true environmental covariates with greater temporal variation.
 EMs that assumed recruitment and natural mortality random effects (R+M), had poor convergence probability for all OMs that had a step change in fishing history and higher uncertainty in indices and age composition.
@@
 
+When fitting the EMs to R and R+S OMs, AIC was generally accurate in determining the correct source of process errors and, when no effect of the covariate was simulated in the OMs, AIC was also accurate for the correct treatment of the environmental covariate \(\beta_E = 0\) (Figs.
+\ref{aic}, \ref{aic_Rom}, \ref{aic_RSom}).
+However, when OMs included an effect of the covariate AIC accuracy for estimating an effect in the EM was only found when the OMs also included lower covariate observation error and higher effect size and higher temporal variability in the covariate.
+On the other hand, for R+M Oms, AIC generally was not able to accurately determine the correct source of process errors, but was able to accurately determine the correct treatment of the environmental covariate (Figs.
+\ref{aic} and \ref{aic_RMom}).
+
 When estimating models assumed the mean natural mortality rate was known, the best accuracy of AIC for model selection occurred for models with R+S process errors.
-R+S estimating models ranked best with very low frequency for R or R+M operating models and with very high frequency for R+S operating models (Figure \ref{aic_rank_M_fixed}).
+R+S estimating models ranked best with very low frequency for R or R+M operating models and with very high frequency for R+S operating models (Figure \ref{aic}).
 R estimating models were determined best with high frequency for R operating models, but also for R+M operating models.
@@
 When the OMs had R+S process errors with low temporal variation in the true environmental covariate and lower uncertainty in the indices age age composition, estimated covariate effects were highly variable.
-In most cases the relative error of \(\beta_\text{Ecov}\) did not depend on the source of process error assumed in the EM.
+In most cases the relative error of \(\beta_E\) did not depend on the source of process error assumed in the EM.
 When there was an effect of the EM process error assumption it was when OMs had R+S process errors.
@@
 
-We present results for error in annual natural mortality rate conditional on three alternative EM configurations for the natural mortality parameters: 1) mean natural mortality rate parameter is fixed at the true value (\(\beta_M = log(0.2)\)) and no covariate effect is assumed (\(\beta_\text{Ecov} = 0\)), 2) \(\beta_M = log(0.2)\) and \(\beta_\text{Ecov}\) is estimated, and 3) both \(\beta_M\) and \(\beta_\text{Ecov}\) are estimated.
+We present results for error in annual natural mortality rate conditional on three alternative EM configurations for the natural mortality parameters: 1) mean natural mortality rate parameter is fixed at the true value (\(\beta_M = log(0.2)\)) and no covariate effect is assumed (\(\beta_E = 0\)), 2) \(\beta_M = log(0.2)\) and \(\beta_E\) is estimated, and 3) both \(\beta_M\) and \(\beta_E\) are estimated.
 
-When OMs and EMs assume \(\beta_M = log(0.2)\) and \(\beta_\text{Ecov} = 0\), there is no annual variation in natural mortality for OMs with R or R+S process errors simulated or assumed in the EM and, therefore, estimation bias is not possible (Figure \ref{annual_M_bias_M_fixed_beta_fixed}).
-We also observe little or no evidence of bias (confidence intervals include 0) for R or R+S EMs for any OMs even when the OMs included an effect of the covariate on natural mortality (\(\beta_\text{Ecov} > 0\)).
+When OMs and EMs assume \(\beta_M = log(0.2)\) and \(\beta_E = 0\), there is no annual variation in natural mortality for OMs with R or R+S process errors simulated or assumed in the EM and, therefore, estimation bias is not possible (Figure \ref{annual_M_bias_M_fixed_beta_fixed}).
+We also observe little or no evidence of bias (confidence intervals include 0) for R or R+S EMs for any OMs even when the OMs included an effect of the covariate on natural mortality (\(\beta_E > 0\)).
 Including process error on M (R+M) produces more variability in errors for R and R+S EMs than including increased level of effect of the covariate on M.
@@
 
-When \(\beta_\text{Ecov}\) was estimated, R EMs performed worse for R+S OMs with evidence of negative bias for OMs with constant fishing mortality, lower uncertainty in index and age composition observations, the largest covariate effect size, lower variability in the latent covariate, and larger uncertainty in the covariate observations (Figure \ref{annual_M_bias_M_fixed_beta_estimated}).
-There was little difference in the results for R+S EMs whether \(\beta_\text{Ecov}\) was estimated or not.
-The results for R+M EMs were generally similar to those when \(\beta_\text{Ecov}=0\) was assumed, except that the differences between the estimates at the beginning of the time series and later years for certain OM configurations did not occur.
+When \(\beta_E\) was estimated, R EMs performed worse for R+S OMs with evidence of negative bias for OMs with constant fishing mortality, lower uncertainty in index and age composition observations, the largest covariate effect size, lower variability in the latent covariate, and larger uncertainty in the covariate observations (Figure \ref{annual_M_bias_M_fixed_beta_estimated}).
+There was little difference in the results for R+S EMs whether \(\beta_E\) was estimated or not.
+The results for R+M EMs were generally similar to those when \(\beta_E=0\) was assumed, except that the differences between the estimates at the beginning of the time series and later years for certain OM configurations did not occur.
 
@@
 
-Like natural mortality, we present results for error in spawning biomass conditional on three alternative EM configurations for the natural mortality parameters: 1) mean natural mortality rate parameter is fixed at the true value (\(\beta_M = log(0.2)\)) and no covariate effect is assmed (\(\beta_\text{Ecov} = 0\)), 2) \(\beta_M = log(0.2)\) and \(\beta_\text{Ecov}\) is estimated, and 3) both \(\beta_M\) and \(\beta_\text{Ecov}\) are estimated.
+Like natural mortality, we present results for error in spawning biomass conditional on three alternative EM configurations for the natural mortality parameters: 1) mean natural mortality rate parameter is fixed at the true value (\(\beta_M = log(0.2)\)) and no covariate effect is assmed (\(\beta_E = 0\)), 2) \(\beta_M = log(0.2)\) and \(\beta_E\) is estimated, and 3) both \(\beta_M\) and \(\beta_E\) are estimated.
 
-When EMs assume \(\beta_M = log(0.2)\) and \(\beta_\text{Ecov} = 0\), there is little evidence of bias for R and R+M OMs except when there is a step change in fishing mortality and higher uncertainty in index and age composition observations, particularly at the end of the time series (Figure \ref{SSB_bias_M_fixed_beta_fixed}).
+When EMs assume \(\beta_M = log(0.2)\) and \(\beta_E = 0\), there is little evidence of bias for R and R+M OMs except when there is a step change in fishing mortality and higher uncertainty in index and age composition observations, particularly at the end of the time series (Figure \ref{SSB_bias_M_fixed_beta_fixed}).
 We observed a similar trend in median relative error for R+S OMs, but there was more evidence of positive bias at the beginning of the time series whereas the confidence intervals for the negative median errors at the end of the time series often included 0.
@@
 
-When EMs assume \(\beta_M = log(0.2)\), but estimate \(\beta_\text{Ecov}\), the median errors for SSB for R and and R+M OMs are similar to those when \(\beta_\text{Ecov} = 0\) (Figure \ref{SSB_bias_M_fixed_beta_estimated}).
+When EMs assume \(\beta_M = log(0.2)\), but estimate \(\beta_E\), the median errors for SSB for R and and R+M OMs are similar to those when \(\beta_E = 0\) (Figure \ref{SSB_bias_M_fixed_beta_estimated}).
 For R+S OMs, the median errors for EMs that also assume R+S process errors are also similar to those with no covariate effect assumed.
@@
 
-When EMs estimated both \(\beta_M\) and \(\beta_\text{Ecov}\), we observed large variation in the errors in SSB under the same OM configurations where we observed large variation in errors for annual natural mortality rates (Figure \ref{SSB_bias_M_estimated_beta_estimated}).
+When EMs estimated both \(\beta_M\) and \(\beta_E\), we observed large variation in the errors in SSB under the same OM configurations where we observed large variation in errors for annual natural mortality rates (Figure \ref{SSB_bias_M_estimated_beta_estimated}).
 Similarly, we observed little or no bias in SSB estimation for R+S EMs across all OM process error assumptions as long as there was a step change in fishing mortality and lower uncertainty in indices and age composition.
@@
 \citet{derisoetal08} was the first to model natural mortality as a function of explicit covariate and residual random annual variation. Any relevant points?
+
+Poor quality of biological, fishery, and observation may adversely affect ability to distinguish between true process and observational error \citep{puntetal14, stewartmonnahan17, croninpunt21, fischetal23, lietal_inreview_a}.
 

### 2025-05-21 Ecov_study/mortality/code/manuscript/aic_ecov_effect_results_new.R

@@ -1,184 +1,0 @@
-library(here)
-library(ggplot2)
-library(dplyr)
-library(ggh4x)
-library(tidyr)
-library(patchwork)
-library(scales)
-library(ggpattern)
-
-df.ems = readRDS(file.path(here(),"Ecov_study","mortality","inputs", "df.ems.RDS"))
-df.oms = readRDS(file.path(here(),"Ecov_study","mortality","inputs", "df.oms.RDS"))
-aic_res <- readRDS(file.path(here(),"Ecov_study","mortality", "results", "aic_results.RDS"))
-
-aic_fn <- function(all, est_ind, om_ind = NULL){
-  if(!is.null(om_ind)) all <- all[om_ind]
-  out <- sapply(all, function(res){
-    tmp <- sapply(res[est_ind], function(x) return(x))
-    tmp = apply(tmp,1, function(x) {
-      if(any(!is.na(x))) {
-        return(x == min(x,na.rm=T))
-      } else return(rep(NA, length(x)))
-    })
-    return(apply(tmp,1,sum,na.rm=T))
-  })  
-  return(out)
-}
-
-make_df_fn <- function(M_est = TRUE){
-  #all EM PE assumptions
-  for(i in 1:3) {
-    re_mod <- c("rec", "rec+1", "rec+M")[i]
-    #EM:  M estimated
-    em_ind <- which(df.ems$M_est == M_est) #all EM PE assumptions
-    om_ind <- which(df.oms$NAA_M_re == re_mod) #om and em match
-    Mfixed_rec <- aic_fn(aic_res, em_ind, om_ind)
-    res <- cbind(df.oms[rep(om_ind, each = length(em_ind)),], df.ems[rep(em_ind, length(om_ind)),], n = c(Mfixed_rec))
-    if(i == 1) {
-      df <- res
-    } else {
-      df <- rbind(df, res)
-    }
-  }
-  facs <- c("Ecov_obs_sig", "Fhist", "Ecov_re_sig","Ecov_re_cor","obs_error", "NAA_M_re", "re_config", "Ecov_est")
-  df[facs] <- lapply(df[facs], factor)
-  df <- df %>%
-    mutate(Ecov_obs_sig = recode(Ecov_obs_sig,
-      "0.1" = "sigma[italic(e)] == 0.1",
-      "0.5" = "sigma[italic(e)] == 0.5"
-    ))
-  df <- df %>%
-    mutate(obs_error = recode(obs_error,
[…]

### 2025-05-21 Ecov_study/mortality/code/manuscript/aic_ecov_effect_results_orig.R

@@ -1,206 +1,0 @@
-library(here)
-library(ggplot2)
-library(dplyr)
-df.ems = readRDS(file.path(here(),"Ecov_study","mortality","inputs", "df.ems.RDS"))
-df.oms = readRDS(file.path(here(),"Ecov_study","mortality","inputs", "df.oms.RDS"))
-aic_res <- readRDS(file.path(here(),"Ecov_study","mortality", "results", "aic_results.RDS"))
-
-aic_fn <- function(all, est_ind, om_ind = NULL){
-  if(!is.null(om_ind)) all <- all[om_ind]
-  out <- sapply(all, function(res){
-    tmp <- sapply(res[est_ind], function(x) return(x))
-    #print(est_ind)
-    #print(length(res))
-    #print(tmp)
-    tmp = apply(tmp,1, function(x) {
-      if(any(!is.na(x))) {
-        return(x == min(x,na.rm=T))
-      } else return(rep(NA, length(x)))
-    })
-    return(apply(tmp,1,sum,na.rm=T))
-  })  
-  return(out)
-}
-
-make_df_fn <- function(M_est = TRUE){
-  #all EM PE assumptions
-  for(i in 1:3) {
-    re_mod <- c("rec", "rec+1", "rec+M")[i]
-    #EM:  M estimated
-    em_ind <- which(df.ems$M_est == M_est) #all EM PE assumptions
-    print(length(em_ind))
-    om_ind <- which(df.oms$NAA_M_re == re_mod) #om PE
-    print(length(om_ind))
-    res <- aic_fn(aic_res, em_ind, om_ind) #n_ems x n_oms
-    print(res[1,])
-    print(res[,1])
-    res <- cbind(df.oms[rep(om_ind, each = length(em_ind)),], df.ems[rep(em_ind, length(om_ind)),], n = c(res))
-    if(i == 1) {
-      all_res <- res
-    } else {
-      all_res <- rbind(all_res, res)
-    }
-  }
-  facs <- c("Ecov_obs_sig", "Fhist", "Ecov_re_sig","Ecov_re_cor","obs_error", "NAA_M_re", "re_config", "Ecov_est")
-  all_res[facs] <- lapply(all_res[facs], factor)
-  all_res_mod <- all_res %>%
-    mutate(Ecov_obs_sig = recode(Ecov_obs_sig,
[…]

### 2025-05-21 Ecov_study/mortality/code/manuscript/aic_results.R

@@ -1,17 +1,21 @@
 library(here)
 library(ggplot2)
-library(ggh4x)
 library(dplyr)
+library(ggh4x)
 library(tidyr)
+library(patchwork)
+library(scales)
 library(ggpattern)
-df.ems = readRDS(file.path(here(),"Project_0","inputs", "df.ems.RDS"))
 
+df.ems = readRDS(here::here("Ecov_study","mortality","inputs", "df.ems.RDS"))
+df.oms = readRDS(here::here("Ecov_study","mortality","inputs", "df.oms.RDS"))
+aic_res <- readRDS(here::here("Ecov_study","mortality", "results", "aic_results.RDS"))
 
-all_naa_aic <- readRDS(file = file.path(here(),"Project_0","results", "all_naa_aic_results.RDS"))
 aic_fn <- function(all, est_ind, om_ind = NULL){
   if(!is.null(om_ind)) all <- all[om_ind]
   out <- sapply(all, function(res){
-    tmp = apply(res[est_ind,],2, function(x) {
+    tmp <- sapply(res[est_ind], function(x) return(x))
+    tmp = apply(tmp,1, function(x) {
       if(any(!is.na(x))) {
         return(x == min(x,na.rm=T))
       } else return(rep(NA, length(x)))
@@ -21,437 +25,144 @@ aic_fn <- function(all, est_ind, om_ind = NULL){
   return(out)
 }
 
-make_plot_df <- function(aic_res, df.ems, df.oms, em_ind = 1:20, SR_est =FALSE, M_est = FALSE){
-  use.df.ems <- df.ems[em_ind,]
-  SR_M_ind = which(use.df.ems$SR_model == ifelse(SR_est,3,2) & use.df.ems$M_est == M_est)
-  aic_res <- t(aic_fn(aic_res, SR_M_ind))
-  df <- cbind(df.oms[-1], aic_res)
-  print(df)
-  df <- df %>% pivot_longer(cols = as.character(1:5), names_to = "EM", values_to = "n") %>% as.data.frame
-  this.df.ems <- use.df.ems[SR_M_ind,]
-  print(SR_M_ind)
-  print(this.df.ems)
-  print(df$EM)
-  df <- cbind(this.df.ems[as.integer(df$EM),],df)
-  print(head(df))
-  df$correct_EM_PE <- "No"
-  if(max(em_ind)==20) {
[…]

### 2025-05-21 Ecov_study/mortality/code/manuscript/convergence_results.R

@@ -74,6 +74,13 @@ conv_res_plotting_fn <- function(conv_res, M_est = TRUE, Ecov_est = TRUE){
       "H" = "High observation error"
     ))
   df <- df %>%
+    mutate(oe = recode(obs_error,
+        "Low observation error" = "Low OE",
+        "High observation error" = "High OE"
+    ))
+  df$oe  <- factor(df$oe, levels = c("Low OE", "High OE"))
+
+  df <- df %>%
     mutate(M_assumption = recode(as.character(M_est),
       "TRUE" = "Estimated",
       "FALSE" = "Known"
@@ -163,3 +170,24 @@ cairo_pdf(here("Ecov_study","mortality","manuscript", "convergence.pdf"), width 
 print(plt)
 dev.off()
 
+
+OMs <- levels(all_res_mod$OM_process_error)
+OMs_lab <- c("Rom","RSom","RMom")
+for(i in 1:length(OMs)){
+  temp <- subset(df, Type == 3 & OM_process_error == OMs[i])
+  plt <- ggplot(temp, aes(x = Ecov_effect, y = p_pass, colour = EM_process_error, fill = EM_process_error, shape = M_assumption)) + 
+      facet_nested(Ecov_obs_sig+Ecov_re_sig+Ecov_re_cor ~ Fhist + oe + Ecov_assumption,
+        labeller = labeller(Ecov_obs_sig = label_parsed, Ecov_re_sig = label_parsed, Ecov_re_cor = label_parsed,  Fhist = label_parsed, Ecov_assumption = label_parsed)) +
+      scale_colour_viridis_d(begin = 0.2, end = 0.8, option = "turbo", drop = FALSE) +
+      scale_fill_viridis_d(begin = 0.2, end = 0.8, option = "turbo", drop = FALSE) +
+      coord_cartesian(ylim = c(0, 1)) + ylab("Probability of convergence") + xlab(expression("True "*beta[Ecov])) +
+      scale_x_continuous(breaks = c(0,0.25, 0.5)) + 
+      geom_line(position = position_dodge(0.1), linewidth = 1) + 
+      geom_point(position = position_dodge(0.1), size = 4) + 
+      geom_errorbar(aes(ymin = ci_lo, ymax = ci_hi, colour = EM_process_error), width = 0, position = position_dodge(0.1), linewidth = 1) +
+      labs(colour = "EM process error", fill = "EM process error", shape = "M assumption") +
[…]

### 2025-05-21 Ecov_study/mortality/code/manuscript/methods_figures.R

@@ -1,8 +1,8 @@
 library(here)
 library(dplyr)
 library(wham, lib.loc = "c:/work/wham/old_packages/77bbd94")
-df.oms <- readRDS(file.path(here(),"Project_0","inputs", "df.oms.RDS"))
-om_inputs <- readRDS(file.path(here(),"Project_0","inputs", "NAA_om_inputs.RDS"))
+df.oms <- readRDS(here::here("Ecov_study","mortality","inputs", "df.oms.RDS"))
+om_inputs <- readRDS(here::here("Ecov_study","mortality","inputs", "om_inputs.RDS"))
 x <- fit_wham(om_inputs[[1]], do.fit =F)
 library(ggplot2)
 library(patchwork)
@@ -22,14 +22,12 @@ mat_plt
 wgt_plt <- ggplot(df, aes(x = age, y = wgt)) + geom_line() + ylab("Mass (kg)") + xlab("Age") + scale_x_continuous(breaks=seq(1,10,1),labels=df$age_lab)
 wgt_plt
 
-png(here("Project_0","paper","om_sr.png"), width=7, height=7, units='in', res = 300)
 sr.fn <- function(ssb) exp(x$rep$log_SR_a[1])*ssb/(1+exp(x$rep$log_SR_b[1])*ssb)
-
 ssb <- seq(0,10e5,10)
 df <- cbind.data.frame(ssb = ssb/1000, R = sr.fn(ssb)/1000)
 sr_plt <- ggplot(df, aes(x = ssb, y = R)) + geom_line() + ylab(expression(Recruitment~(10^6))) + xlab("Spawning biomass (kmt)")
 sr_plt
 
-cairo_pdf(here("manuscript", "om_input_plots_figure.pdf"), width = 20, height = 20)
+cairo_pdf(here::here("Ecov_study","mortality","manuscript", "om_input_plots_figure.pdf"), width = 20, height = 20)
 mat_plt+wgt_plt+sel_plt+sr_plt
 dev.off()

### 2025-05-29 Ecov_study/mortality/manuscript/manuscript.Rmd
saving work on ecov-m

@@
 All of our analyses used the Woods Hole Assessment Model (WHAM) to construct both OMs and EMs [@millerstock20; @stockmiller21; @milleretal_inreview].
-The WHAM package has been used extensively to configure OMs and EMs for several other simulation studies [@legaultetal23; @lietal24; @brittenetal_inreview; @lietal_inreview_a] and is used to assess many commercially important stocks in the Northeast U.S.
+The WHAM package has been used extensively to configure OMs and EMs for several other simulation studies [@stocketal21;@legaultetal23; @lietal24; @brittenetal_inreview; @lietal_inreview_a] and is used to assess many commercially important stocks in the Northeast U.S.
 [e.g., @nefsc22;@nefsc22a;@nefsc24].
@@
 
-In the WHAM model, environmental covariates are assumed to be described as state-space processes with annual observations of the true latent covariate [@milleretal16;stockmiller21].
+In the WHAM model, environmental covariates are assumed to be described as state-space processes with annual observations of the true latent covariate [@milleretal16;@stockmiller21].
 In our simulations, the latent covariate is assumed to be a stationary first order autoregressive (AR1) process
 $$
-X_y|X_{y-1} \sim \text{N}\left(\mu_E\left(1-\rho_E\right) + \rho_E E_{y-1}, \left(1-\rho_E^2\right)\sigma^2_E\right)
+X_y|X_{y-1} \sim \text{N}\left(\mu_E\left(1-\rho_E\right) + \rho_E X_{y-1}, \left(1-\rho_E^2\right)\sigma^2_E\right)
 $$
 with marginal mean $\mu_E=0$ and variance $\sigma^2_E$.
-The four configuration of the latent environmental covariate in the operating models assume the one of two values for the marginal standard deviation ($\sigma_E \in \{0.1, 0.5\}$) and for the autocorrelation parameter ($\rho_E \in \{0, 0.5\}$).
+The four configuration of the latent environmental covariate in the operating models assume the one of two values for the marginal standard deviation $\sigma_E \in \{0.1, 0.5\}$ and for the autocorrelation parameter $\rho_E \in \{0, 0.5\}$.
 
@@
 $$
-The standard deviation of the environmental observations in the operating models is one of two values $\sigma_E \in \{0.1, 0.5\}$.
+The standard deviation of the environmental observations in the operating models is one of two values $\sigma_e \in \{0.1, 0.5\}$.
 Figure \ref{om_ecov_example} provides example simulations of the latent process and observations under the alternative configurations.
@@
 
-Many of the characteristics of the population biology and structure are the same as those in @milleretal_inreview1.
-We configures the stock with 10 age classes (ages 1 to 10+) where the last is a "plus" group that accumulates any older individuals.
-We model the population for 40 years with annual catch and index observations.
-We assume spawning occurs annually 1/4 of the way through the year.
-The maturity at age was a logistic function of age with age at 50\% maturity ($a_{50}$) = 2.89 and slope = 0.88 (Figure \ref{om_inputs_fig}, top left).
-We assumed a von Beralanffy growth function for weight at age  with length at age defined as
-$$
-L_a = L_{\infty}\left(1 - e^{-k(a - t_0)}\right)
-$$
-where $t_0 = 0$, $L_\infty = 85$, and $k = 0.3$, and a length-weight 
+Many of the characteristics of the population biology and structure including the age classes (10 age classes (ages 1 to 10+)), time span (40 years), maturity (Figure \ref{om_inputs_fig}, top left), growth  (Figure \ref{om_inputs_fig}, top right), time of spawning (1/4 of the year), and recruitment (Figure \ref{om_inputs_fig}, bottom right) are identical to @milleretal_inreview1.
+The maturity at age is a logistic function of age with age at 50\% maturity ($a_{50}$) = 2.89 and slope = 0.88 and weight at age is derived from a von Beralanffy growth function where $t_0 = 0$, $L_\infty = 85$, and $k = 0.3$, and a length-weight relationship
 $$
@@
 $$
-where $\theta_1 = e^{-12.1}$ and $\theta_2 = 3.2$ (Figure \ref{om_inputs_fig}, top right).
+where $\theta_1 = e^{-12.1}$ and $\theta_2 = 3.2$.
 
-The general model for natural mortality in year $y$ is a log-linear function of a mean(median) parameter $\beta_{M}$, effects of an environmental covariate $E_y$ and a possibly autocorrelated normally distributed annual process error $\varepsilon_{M,y}$
+Here we consider both covariate effects and process errors for natural mortality and the general model for natural mortality in year $y$ is a log-linear function of a  parameter $\beta_{M}$ that defines median natural mortality and possibly effects of the environmental covariate $X_y$ described above and autocorrelated normally distributed annual process error $\varepsilon_{M,y}$
 $$
-\log M_y = \beta_M + \beta_{E} E_y + \varepsilon_{M,y}
+\log M_y = \beta_M + \beta_{E} X_y + \varepsilon_{M,y}
 $$
-$\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)$ [@stockmiller21].
+$\epsilon_{M,y}|\epsilon_{M,y-1} \sim \text{N}\left(\epsilon_{M,y-1},(1-\rho_M^2)\sigma_M^2\right)$ [@stockmiller21].
 We assume the median natural mortality rate $e^{\beta_M} = 0.2$ is constant across ages.
-For R and R+S OMs and EMs, the process errors are set equal to 0.
+For R and R+S OMs and EMs, $\epsilon_{M,y} = 0$.
 For all R+M OMs, we assume the same standard deviation $\sigma_M = 0.3$ and is estimated in the R+M EMs.
@@
 The parameters defining the simulated covariate time series, size of the covariate effect, and any natural mortality random effects result in a range of different levels of variation in annual natural mortality rates (Figure \ref{M_example}).
-
-We assume is a single fishing fleet harvesting the population with a logistic selectivity function with age at 50\% selection $a_{50} = 5$ and slope = 1 (Figure \ref{om_inputs_fig}, bottom left).
 
@@
 All biological inputs to calculations of spawning biomass per recruit (i.e., weight, maturity, and natural mortality at age) are constant in the R and R+S OMs without covariate effects on natural mortality.
-Therefore, steepness and unfished recruitment are also constant over the time period for those OMs [@millerbrooks21] and with our assumed biological inputs and selectivity the fishing mortality resulting in 40\% reduction in spawning biomass per recruit is $F_{40\%} = 0.348$.
-With an assumed unfished recruitment of $R_0 = e^{10}$, setting $\Fmsy = \Fspr[40]$ results in a steepness of 0.69 and $a=0.60$ and $b = 2.4 \times 10^{-5}$ (Figure \ref{om_inputs_fig}, bottom right).
+Therefore, steepness and equilibrium unfished recruitment are also constant over the time period for those OMs [@millerbrooks21].
+As in @milleretal_inreview1, our assumed biological inputs and selectivity (defined below) with constant natural mortality result in equilibrium fishing mortality that reduces spawning biomass per recruit to 40\% of the unfished level is $\Fspr[40] = 0.348$.
+With an assumed unfished recruitment of $R_0 = e^{10}$, setting $\Fmsy = \Fspr[40]$ results in a steepness of 0.69 and $a=0.60$ and $b = 2.4 \times 10^{-5}$.
 
-We used two fishing scenarios for OMs.
+For R+M OMs and all OMs with covariate effects on natural mortality, steepness is not constant, but we used the same $a$ and $b$ parameters as other operating models which equates to a steepness and $R_0$ at the median of the time series process for M.
+For operating models with time-varying random effects for fishery selectivity, \Fmsy is also not constant however we use the same fishing history as other operating models which corresponds to \Fmsy at the mean selectivity parameters.
+
+We also used the same two fishing scenarios as @milleretal_inreview1 for OMs.
 In the first scenario, the stock experiences overfishing at 2.5\Fmsy for the first 20 years followed by fishing at \Fmsy for the last 20 years (denoted $2.5\Fmsy \rightarrow \Fmsy$).
 In the second scenario, the stock is fished at \Fmsy for the entire time period (40 years).
-The magnitude of the overfishing assumptions is based on average estimates of overfishing for NEUS groundfish stocks from @wiedenmannetal19 and similar to the approach in @legaultetal23.
-
-We specified initial population abundance at age at the equilibrium distribution that corresponds to fishing at either $F = 2.5\times \Fmsy$ or $F = \Fmsy$.
-This implies that, for a deterministic model, the abundance at age would not change from the first year to the next.
-
-For R+M OMs and all OMs with environmental covariate effects on natural mortality, steepness is not constant, but we used the same $a$ and $b$ parameters as other operating models which equates to a steepness and $R_0$ at the median of the time series process for M.
-For operating models with time-varying random effects for fishery selectivity, \Fmsy is also not constant however we use the same fishing history as other operating models which corresponds to \Fmsy at the mean selectivity parameters.
 
@@
 
-We assumed a single fleet operating year round for catch observations with logistic selectivity for the fleet with $a_{50} = 5$ and slope = 1 (Figure \ref{om_inputs_fig}).
-This selectivity was used to define \Fmsy, \Fspr[40], and steepness for the Beverton-Holt stock recruitment parameters above.
-We assumed a logistic-normal distribution for the age-composition observations associated with the fleet where errors in the multivariate normal transformation are independent.
+We define the generation of observations of total catch, aggregate indices, and corresponding age composition identical to @milleretal_inreview1.
+There is a single fleet operating year round for catch observations with logistic selectivity for the fleet with $a_{50} = 5$ and slope = 1 (Figure \ref{om_inputs_fig}, bottom left).
+Observations are generated for all 40 years of the model.
+There are two index time series intended to represent fishery-independent surveys occurring in the spring (0.25 way through the year) and the fall (0.75 way through the year).
+Catchability of both surveys are assumed to be 0.1.
+We assumed catch and index age composition observations aer generated from a logistic-normal distribution where errors on the multivariate normal scale are independent.
 The standard deviation parameter is also constant across ages.
-
-Two time series of surveys are assumed and observed in numbers rather than biomass for the entire 40 year period with one occurring in the spring (0.25 way through the year) and one in the fall (0.75 way through the year).
-Catchability of both surveys are assumed to be 0.1.
-We assumed the same selectivity and age composition distribution as the fleet for both indices.
 
@@
 
-For each data set simulated from an operating model 12 estimating models were fit. There were three factors defining the configuration of each estimating model
+For each data set simulated from an operating model 12 estimating models were fit.
+There were three factors defining the configuration of each estimating model: 1) whether the mean natural mortality $\beta_M$ was estimated or assumed known, 2) whether an environmental effect $\beta_E$ was estimated or not, and 3) whether the process errors were assumed on recruitment only (R), recruitment and survival (R+S), or recruitment and natural mortality (R+M).
 
-\begin{itemize}
-\item whether the mean natural mortality $\beta_M$ was estimated or assumed known ($\log 0.2$)
-\item whether an environmental effect $\beta_E$ was estimated or not (fixed at 0)
-\item whether the process errors were assumed on recruitment only (R), recruitment and survival (R+S), or recruitment and natural mortality (R+M)
-\end{itemize}
 The configuration of the process errors in the estimating models generally matched the corresponding options in the operating models.
-
 For example, uncorrelated R+S was assumed for both the estimating and operating model.
-However, R+M EMs did not assume M random effects were uncorrelated (parameter was estimated).
-The environmental covariate observations were included in all estimation models to ensure comparability of information criteria.
+However, R+M EMs did not assume M random effects were uncorrelated ($\rho_M$ was estimated).
+The environmental covariate observations were included in all estimation models to ensure comparability of AIC.
 All fixed effects parameters for selectivity, catchability, fully-selected fishing mortality, mean recruitment, initial abundance at age, and variances for logistic-normal age composition distributions were estimated.
@@
 
-### Convergence {-}
+### EM convergence {-}
 
@@
 $$
+All model fits that successfully completed the optimization were used in these results.
+We used all of these fits because some lack of convergence would be expected for the correct behavior of more complicated models that include process errors that did not exist in the operating model.
+For example R+M EMs fit to R OMs would be expected to estimate no variance in the natural mortality  random effects and the estimated variance parameter going to zero would cause poor convergence.
 
-### Bias {-}
+### Parameter inference reliability {-}
 
-We calculated median errors of
-\begin{itemize}
-\item $\beta_E$, the effect of environmental covariates on natural mortality, 
-\item $\beta_M$, the mean log-natural mortality rate,
-\end{itemize}
-and the median relative errors of 
-\begin{itemize}
-\item annual natural mortality rate,
-\item annual spawning stock biomass, and
-\item annual fully-selected fishing mortality rate.
-\end{itemize}
+All results here use OM simulations with fits that satisfied the convergence criterion described above.
+We used this conditioning, to reflect how practitioners would proceed in analyses of model fits with real assessment data.
+That is, practitioners would ensure models converged such that hessian-based standard errors were available for all model parameter estimates.
 
-Note that the exponentiating the median error of $\beta_M$ would be equivalent to the median relative error of the median natural mortality rate because the median of the exponential is the same as the exponential of the median.
+We calculated median errors (ME) of $\beta_E$, the effect of environmental covariates on natural mortality, and the median natural mortality rate parameter $\beta_M$ (the mean log-natural mortality rate).
+We calculated the median relative errors (MRE) of terminal year estimates of natural mortality rate, spawning stock biomass, and fully-selected fishing mortality rate.
+In preliminary analyses we inspected MRE of these parameters over the whole time series, but we found no appreciable differences in patterns across the various factors defining the OMs and EMs.
+Furthermore, results for terminal year fishing mortality were generally inversely related to those for spawning stock biomass, and are provided in the Supplementary Materials and not discussed further.
 
-Results for fishing mortality rate are provided in the Supplementary Materials. For the $i$th simulated data the relative error for a parameter $\theta$ provided from the fitted estimation model is
+We also calculated the root mean square error (RMSE) and estimated probability of coverage of constructed 95\% confidence intervals for $\beta_E$ and $\beta_M$ for EMs that estimated these parameters.
+Confidence intervals for probabilities of coverage are constructed using the same methods as those for probabilities of convergence.
+
+Results for fishing mortality rate For the $i$th simulated data the relative error for a parameter $\theta$ provided from the fitted estimation model is
 $$
@@
 $$
-and measured bias as the median relative error and constructed 95\% confidence intervals using the binomial distribution approach as in @stockmiller21.
-If the confidence interval contains zero we do not have evidence of bias from the simulation study.
-We used estimates from all simulations that satisfied the first type of convergence described above.
-We were not concerned about more restrictive types of convergence because some estimation models would not be expected to satisfy these criteria because of mis-specified structures.
-For example, an estimation model that includes survival random effects fitted to a set of observations simulated without survival random effects should have trouble estimating variance of the survival random effects, but the estimates of the parameters and derived output described above, might be reliable.
+and  constructed 95\% confidence intervals using the binomial distribution approach as in @stockmiller21 and @milleretal_inreview1.
+Because of the inverse relationship of the the confidence interval and the null hypothesis significance test, when the confidence interval contains zero we do not have evidence of bias from the simulation study.
 
-For natural mortality rate, SSB, and fully-selected fishing mortality rate, we summarized results for each of the annual values, but we present results only for estimates from the first year (start), year 21 (middle), and the last year (end), because there were no appreciable differences between the results for these three years and those for the other years.
+For natural mortality rate, SSB, and fully-selected fishing mortality rate, we summarized results for each of the annual values, but we present results only for estimates from the terminal year, because there were no appreciable differences between the results among all years.
 
@@
 
-## Convergence {-}
+## EM Convergence {-}
 
@@
 ## AIC performance {-}
+
+We only present results for EMs where the median natural mortality rate parameter was estimated.
+Results differed little between results where this parameter was estimated or assumed known, but all of these results can be found in the Supplementary materials.
 
@@
 
-When the EMs assumed the mean natural mortality rate was known, we observed generally accurate estimation of environmental effects across all EM and OM process error assumptions and all true covariate effect sizes, when ther was low uncertainty in environmental observations and larger temporal contrast in the simulated true environmental covariate (Figure \ref{Ecov_beta_bias_M_fixed}).
+When the EMs assumed the median natural mortality rate was known, we observed generally accurate estimation of environmental effects across all EM and OM process error assumptions and all true covariate effect sizes, when ther was low uncertainty in environmental observations and larger temporal contrast in the simulated true environmental covariate (Figure \ref{beta_E_bias}).
 We observed a negative trend in bias of the environmental effect with increased effect size when temporal variation in the covariate was lower and/or uncertainty in the covariate observations was higher.
@@
 
-When the mean natural mortality rate was estimated in the EM, results were similar except estimated effects were even more variable for data simulated with R+S process errors (Figure \ref{Ecov_beta_bias_M_estimated}).
+When the median natural mortality rate was estimated in the EM, results were similar except estimated effects were even more variable for data simulated with R+S process errors (Figure \ref{beta_E_bias}).
 There was also more separation of reliability of the estimation of the effect among EMs with different process error assumptions.
@@
 
-### Mean natural mortality rate {-}
+### Median natural mortality rate {-}
 
-We found high accuracy for estimation of the mean natural mortality rate parameter ($\beta_M$) for all EM process errors assumptions when OMs had step changes in fishing mortality, lower uncertainty in index and age composition observations, and either R or R+M process errors (Figure \ref{mean_M_bias_Ecov_beta_fixed}).
+We found high accuracy for estimation of the median natural mortality rate parameter ($\beta_M$) for all EM process errors assumptions when OMs had step changes in fishing mortality, lower uncertainty in index and age composition observations, and either R or R+M process errors (Figure \ref{beta_M_bias}).
 The most variation in estimates occurred when fishing mortality was constant and there was higher uncertainty in index and age composition observations.
@@
 
-When OMs and EMs assume $\beta_M = log(0.2)$ and $\beta_E = 0$, there is no annual variation in natural mortality for OMs with R or R+S process errors simulated or assumed in the EM and, therefore, estimation bias is not possible (Figure \ref{annual_M_bias_M_fixed_beta_fixed}).
+When OMs and EMs assume $\beta_M = log(0.2)$ and $\beta_E = 0$, there is no annual variation in natural mortality for OMs with R or R+S process errors simulated or assumed in the EM and, therefore, estimation bias is not possible (Figure \ref{terminal_M_bias}).
 We also observe little or no evidence of bias (confidence intervals include 0) for R or R+S EMs for any OMs even when the OMs included an effect of the covariate on natural mortality ($\beta_E > 0$).
@@
 
-When $\beta_E$ was estimated, R EMs performed worse for R+S OMs with evidence of negative bias for OMs with constant fishing mortality, lower uncertainty in index and age composition observations, the largest covariate effect size, lower variability in the latent covariate, and larger uncertainty in the covariate observations (Figure \ref{annual_M_bias_M_fixed_beta_estimated}).
+When $\beta_E$ was estimated, R EMs performed worse for R+S OMs with evidence of negative bias for OMs with constant fishing mortality, lower uncertainty in index and age composition observations, the largest covariate effect size, lower variability in the latent covariate, and larger uncertainty in the covariate observations (Figure \ref{terminal_M_bias}).
 There was little difference in the results for R+S EMs whether $\beta_E$ was estimated or not.
@@
 
-Allowing the EMs to estimate $\beta_M$ resulted in very large variability in estimates for all EMs for R and R+M OMs with constant fishing mortality and higher uncertainty in index and age composition observations (Figure \ref{annual_M_bias_M_estimated_beta_estimated}).
+Allowing the EMs to estimate $\beta_M$ resulted in very large variability in estimates for all EMs for R and R+M OMs with constant fishing mortality and higher uncertainty in index and age composition observations (Figure \ref{terminal_M_bias}).
 The same variability occurred for R+S OMs, but strong bias was estimated for R and R+M EMs.
@@
 
-When EMs assume $\beta_M = log(0.2)$ and $\beta_E = 0$, there is little evidence of bias for R and R+M OMs except when there is a step change in fishing mortality and higher uncertainty in index and age composition observations, particularly at the end of the time series (Figure \ref{SSB_bias_M_fixed_beta_fixed}).
+When EMs assume $\beta_M = log(0.2)$ and $\beta_E = 0$, there is little evidence of bias for R and R+M OMs except when there is a step change in fishing mortality and higher uncertainty in index and age composition observations, particularly at the end of the time series (Figure \ref{terminal_ssb_bias}).
 We observed a similar trend in median relative error for R+S OMs, but there was more evidence of positive bias at the beginning of the time series whereas the confidence intervals for the negative median errors at the end of the time series often included 0.
@@
 
-When EMs assume $\beta_M = log(0.2)$, but estimate $\beta_E$, the median errors for SSB for R and and R+M OMs are similar to those when $\beta_E = 0$ (Figure \ref{SSB_bias_M_fixed_beta_estimated}).
+When EMs assume $\beta_M = log(0.2)$, but estimate $\beta_E$, the median errors for SSB for R and and R+M OMs are similar to those when $\beta_E = 0$ (Figure \ref{terminal_ssb_bias}).
 For R+S OMs, the median errors for EMs that also assume R+S process errors are also similar to those with no covariate effect assumed.
@@
 
-When EMs estimated both $\beta_M$ and $\beta_E$, we observed large variation in the errors in SSB under the same OM configurations where we observed large variation in errors for annual natural mortality rates (Figure \ref{SSB_bias_M_estimated_beta_estimated}).
+When EMs estimated both $\beta_M$ and $\beta_E$, we observed large variation in the errors in SSB under the same OM configurations where we observed large variation in errors for annual natural mortality rates (Figure \ref{terminal_ssb_bias}).
 Similarly, we observed little or no bias in SSB estimation for R+S EMs across all OM process error assumptions as long as there was a step change in fishing mortality and lower uncertainty in indices and age composition.
@@
 # Discussion {-}
+
+We found estimation of environmental effects on M was possible and reliable in certain scenarios even when the process error was mis-specified (R+S EM and R+M OM) also R EMs were not reliable for Ecov effects when PE was mis-specified.
+Given that milleretal in review found AIC was accurate for R+S OMs and applications of wham to real assessment data typically finds R+S preferrable to R only, it suggests inferences on covariate effects on M without just R process errors (e.g., deriso et al 2008) should be reevaluated.
 
@@
 @derisoetal08 was the first to model natural mortality as a function of explicit covariate and residual random annual variation. Any relevant points?
+
+We conditioned our investigations of reliability of parameter estimation on simulations with fits that converged satisfactorily.
+However, these simulations could exclude many realizations of covariate, population, and observation time series that are not sufficiently informative for all the parameters that we were estimating.
+Why might this matter?
 
@@
 
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{convergence_main}
+\end{center}
+\caption{Estimated probability of fits providing hessian-based standard errors for EMs with alternative process error assumptions, treatment of median natural mortality ($e^\beta_M$ known or estimated), and treatment of covariate effect ($\beta_E = 0$ or estimated).
+The OMs have R (left) and R+S (middle), or R+M (right) process error structures, alternative configurations of covariate time series structure and levels of observation uncertainty (rows), and three levels of true covariate effect on median natural mortality (x axis).
+All OMs had low observation error for fish population observations and temporal contrast in fishing pressure.
+Vertical lines represent 95\% confidence intervals.}\label{convergence}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{aic_main}
+\end{center}
+\caption{For each OM, the proportion of simulated data sets where the EM type (treatment of environmental covariate effect and  assumed process error type) had the lowest AIC.
+All OMs had low observation error for fish population observations and temporal contrast in fishing pressure.
+All EMs estimated median natural mortality rate.}\label{aic}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_E_bias_main}
+\end{center}
+\caption{Median error (ME) of estimates of environmental effect on natural mortality $\beta_E$ from fitting EMs with alternative process error assumptions and treatment of median natural mortality ($e^\beta_M$ known or estimated).
+All OMs had low observation error and contrast in fishing mortality.
+Vertical lines represent 95\% confidence intervals.}\label{beta_E_bias}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_E_CI_coverage_main}
+\end{center}
+\caption{Probability of 95\% confidence interval for $\beta_E$ containing the true value for EMs with alternative process error assumptions and treatment of median natural mortality ($e^\beta_M$ known or estimated).
+All OMs had low observation error and contrast in fishing mortality.
+Vertical lines represent 95\% confidence intervals.}\label{beta_E_CI_coverage}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_M_bias_main}
+\end{center}
+\caption{Median error (ME) of estimates of $\beta_M$ from fitting EMs with alternative process error assumptions and treatment of covariate effect ($\beta_E = 0$ or estimated).
+All OMs had low observation error and contrast in fishing mortality.
+Vertical lines represent 95\% confidence intervals.}\label{beta_M_bias}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_M_CI_coverage_main}
+\end{center}
+\caption{Probability of 95\% confidence interval for $\beta_M$ containing the true value for EMs with alternative process error assumptions and treatment of covariate effect ($\beta_E = 0$ or estimated).
+All OMs had low observation error and contrast in fishing mortality.
+Vertical lines represent 95\% confidence intervals.}\label{beta_M_CI_coverage}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_M_bias_main}
+\end{center}
+\caption{Median relative error (MRE) of estimates of natural mortality rate in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).
+All OMs had low observation error and contrast in fishing mortality.}\label{terminal_M_bias}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_ssb_bias_main}
+\end{center}
+\caption{Median relative error (MRE) of estimates of spawning stock biomass (SSB) in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).
+All OMs had low observation error and contrast in fishing mortality.}\label{terminal_ssb_bias}
+\end{figure}
+\end{landscape}
+
+# Supplemental Materials {-}
+
+\setcounter{figure}{0}
+\renewcommand\thefigure{S\arabic{figure}}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width = \textwidth]{om_input_plots_figure}
+\end{center}
+\caption{The proportion mature at age, weight at age, fleet and index selectivity at age, and Beverton-Holt stock-recruit relationship assumed for the population in all operating models.
+For operating models with random effects on fleet selectivity, this represents the selectivity at the mean of the random effects.}\label{om_inputs_fig}
+\end{figure}
+
 \begin{figure}
@@
 \end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{convergence}
-\end{center}
-\caption{Estimated probability of fits providing hessian-based standard errors for EMs assuming alternative process error, that estimate or assume known median natural mortality, and that estimate or assume no covariate effect on median natural mortality when fitted to operating models that have R (left) and R+S (middle) , or R+M (right) process error structures and three levels of true covariate effect on median natural mortality (x axis).
-Vertical lines represent 95\% confidence intervals.}\label{convergence}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics[height = \textheight]{aic_main}
-\end{center}
-\caption{Proportion of simulated data sets for each operating model where the estimation model type (treatment of environmental covariate effect and  assumed process error type) had the lowest AIC.
-All OMs had low observation error for fish population observations and temporal contrast in fishing pressure.
-All EMs estimated median natural mortality rate.}\label{aic}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-%\includegraphics[height = \textheight]{Ecov_beta_bias}
-\end{center}
-\caption{Relative error of estimates of environmental effect on natural mortality $\beta_E$ from fitting EMs with alternative process error assumptions.
-All OMs had low observation error and contrast in fishing mortality.
-Vertical lines represent 95\% confidence intervals.}\label{ecov_beta_bias}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-%\includegraphics[height = \textheight]{median_log_M_bias}
-\end{center}
-\caption{Relative error of the estimated mean log-natural mortality parameter $\beta_{M}$ from fitting EMs with alternative process error assumptions.
-All OMs had low observation error and contrast in fishing mortality.
-Vertical lines represent 95\% confidence intervals.}\label{median_log_M_bias}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of annual natural mortality rate in terminal year from fitting simulated observations from each OM with alternative process error assumptions in the EM.
-Mean log-natural mortality parameter ($\beta_M$) and environmental covariate effect ($\beta_E$) are both estimated in the EMs.
-All OMs had low observation error and contrast in fishing mortality.}\label{terminal_M_bias}
-\begin{center}
-%\includegraphics[height = \textheight]{terminal_year_M_bias.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of annual SSB in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
-Estimating models also assume the mean natural mortality parameter $\beta_M = \log 0.2$ and no environmental covariate effect $\beta_E = 0$.}\label{SSB_bias_M_fixed_beta_fixed}
-\begin{center}
-%\includegraphics[height = \textheight]{SSB_bias_all_PE_effect_M_fixed_beta_fixed.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of annual SSB in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
-Estimating models also assume the mean natural mortality parameter $\beta_M = \log 0.2$ and the environmental covariate effect $\beta_E$ is estimated.}\label{SSB_bias_M_fixed_beta_estimated}
-\begin{center}
-%\includegraphics[height = \textheight]{SSB_bias_all_PE_effect_M_fixed_beta_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Median relative error of annual SSB in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
-Estimating models also assume both the mean natural mortality parameter $\beta_M$ and the environmental covariate effect $\beta_E$ are estimated.}\label{SSB_bias_M_estimated_beta_estimated}
-\begin{center}
-%\includegraphics[height = \textheight]{SSB_bias_all_PE_effect_M_estimated_beta_estimated.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
-# Supplemental Materials {-}
-
-\setcounter{figure}{0}
-\renewcommand\thefigure{S\arabic{figure}}
-
-\begin{figure}
-\begin{center}
-\includegraphics[width = \textwidth]{om_input_plots_figure}
-\end{center}
-\caption{The proportion mature at age, weight at age, fleet and index selectivity at age, and Beverton-Holt stock-recruit relationship assumed for the population in all operating models.
-For operating models with random effects on fleet selectivity, this represents the selectivity at the mean of the random effects.}\label{om_inputs_fig}
-\end{figure}
 
@@
 \begin{figure}
-\caption{Median relative error of annual natural mortality rate in terminal year from fitting simulated observations from each OM with alternative process error assumptions in the EM.
-Mean log-natural mortality parameter ($\beta_M$) is estimated, but EMs assume no environmental covariate effect ($\beta_E = 0 $) are both estimated in the EMs.
-All OMs had low observation error and contrast in fishing mortality.}\label{terminal_M_bias_beta_0}
 \begin{center}
-%\includegraphics[height = \textheight]{terminal_year_M_bias_ecov_beta_0.png}
+\includegraphics[height = \textheight]{beta_E_bias_Rom}
 \end{center}
+\caption{For R OMs, median error (ME) of estimates of environmental effect on natural mortality $\beta_E$ from fitting EMs with alternative process error assumptions and treatment of median natural mortality ($e^\beta_M$ known or estimated).
+Vertical lines represent 95\% confidence intervals.}\label{beta_E_bias_Rom}
 \end{figure}
@@
 \begin{figure}
-\caption{Median relative error of annual fully-selected fishing mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
-Estimating models also assume the mean natural mortality parameter $\beta_M = \log 0.2$ and no environmental covariate effect $\beta_E = 0$.}\label{F_bias_M_fixed_beta_fixed}
 \begin{center}
-%\includegraphics[height = \textheight]{F_bias_all_PE_effect_M_fixed_beta_fixed.png}
+\includegraphics[height = \textheight]{beta_E_bias_RSom}
 \end{center}
+\caption{For R+S OMs, median error (ME) of estimates of environmental effect on natural mortality $\beta_E$ from fitting EMs with alternative process error assumptions and treatment of median natural mortality ($e^\beta_M$ known or estimated).
+Vertical lines represent 95\% confidence intervals.}\label{beta_E_bias_RSom}
 \end{figure}
@@
 \begin{figure}
-\caption{Median relative error of annual fully-selected fishing mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
-Estimating models also assume the mean natural mortality parameter $\beta_M = \log 0.2$ and the environmental covariate effect $\beta_E$ is estimated.}\label{F_bias_M_fixed_beta_estimated}
 \begin{center}
-%\includegraphics[height = \textheight]{F_bias_all_PE_effect_M_fixed_beta_estimated.png}
+\includegraphics[height = \textheight]{beta_E_bias_RMom}
 \end{center}
+\caption{For R+M OMs, median error (ME) of estimates of environmental effect on natural mortality $\beta_E$ from fitting EMs with alternative process error assumptions and treatment of median natural mortality ($e^\beta_M$ known or estimated).
+Vertical lines represent 95\% confidence intervals.}\label{beta_E_bias_RMom}
 \end{figure}
@@
 \begin{figure}
-\caption{Median relative error of annual fully-selected fishing mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
-Estimating models also assume both the mean natural mortality parameter $\beta_M$ and the environmental covariate effect $\beta_E$ are estimated.}\label{F_bias_M_estimated_beta_estimated}
 \begin{center}
-%\includegraphics[height = \textheight]{F_bias_all_PE_effect_M_estimated_beta_estimated.png}
+\includegraphics[height = \textheight]{beta_E_CI_coverage_Rom}
 \end{center}
+\caption{For R OMs, probability of 95\% confidence interval for $\beta_E$ containing the true value for EMs with alternative process error assumptions and treatment of median natural mortality ($e^\beta_M$ known or estimated).
+Vertical lines represent 95\% confidence intervals.}\label{beta_E_CI_coverage_Rom}
 \end{figure}
 \end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_E_CI_coverage_RSom}
+\end{center}
+\caption{For R+S OMs, probability of 95\% confidence interval for $\beta_E$ containing the true value for EMs with alternative process error assumptions and treatment of median natural mortality ($e^\beta_M$ known or estimated).
+Vertical lines represent 95\% confidence intervals.}\label{beta_E_CI_coverage_RSom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_E_CI_coverage_RMom}
+\end{center}
+\caption{For R+M OMs, probability of 95\% confidence interval for $\beta_E$ containing the true value for EMs with alternative process error assumptions and treatment of median natural mortality ($e^\beta_M$ known or estimated).
+Vertical lines represent 95\% confidence intervals.}\label{beta_E_CI_coverage_RMom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_E_rmse_main}
+\end{center}
+\caption{Root mean square error (RMSE) of estimates of covariate effect on natural mortality $\beta_E$ from fitting EMs with alternative process error assumptions and treatment of median natural mortality ($e^\beta_M$ known or estimated).
+All OMs had low observation error and contrast in fishing mortality.}\label{beta_E_rmse}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_E_rmse_Rom}
+\end{center}
+\caption{For R OMs, root mean square error (RMSE) of estimates of covariate effect on natural mortality $\beta_E$ from fitting EMs with alternative process error assumptions and treatment of median natural mortality ($e^\beta_M$ known or estimated).
+}\label{beta_E_rmse_Rom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_E_rmse_RSom}
+\end{center}
+\caption{For R+S OMs, root mean square error (RMSE) of estimates of covariate effect on natural mortality $\beta_E$ from fitting EMs with alternative process error assumptions and treatment of median natural mortality ($e^\beta_M$ known or estimated).
+}\label{beta_E_rmse_RSom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_E_rmse_RMom}
+\end{center}
+\caption{For R+M OMs, root mean square error (RMSE) of estimates of covariate effect on natural mortality $\beta_E$ from fitting EMs with alternative process error assumptions and treatment of median natural mortality ($e^\beta_M$ known or estimated).
+}\label{beta_E_rmse_RMom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_M_bias_Rom}
+\end{center}
+\caption{For R OMs, median error (ME) of estimates of $\beta_M$ from fitting EMs with alternative process error assumptions and treatment of covariate effect ($\beta_E = 0$ or estimated).
+Vertical lines represent 95\% confidence intervals.}\label{beta_M_bias_Rom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_M_bias_RSom}
+\end{center}
+\caption{For R+S OMs, median error (ME) of estimates of $\beta_M$ from fitting EMs with alternative process error assumptions and treatment of covariate effect ($\beta_E = 0$ or estimated).
+Vertical lines represent 95\% confidence intervals.}\label{beta_M_bias_RSom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_M_bias_RMom}
+\end{center}
+\caption{For R+M OMs, median error (ME) of estimates of $\beta_M$ from fitting EMs with alternative process error assumptions and treatment of covariate effect ($\beta_E = 0$ or estimated).
+Vertical lines represent 95\% confidence intervals.}\label{beta_M_bias_RMom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_M_CI_coverage_Rom}
+\end{center}
+\caption{For R OMs, probability of 95\% confidence interval for $\beta_M$ containing the true value for EMs with alternative process error assumptions and treatment of covariate effect ($\beta_E = 0$ or estimated).
+Vertical lines represent 95\% confidence intervals.}\label{beta_M_CI_coverage_Rom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_M_CI_coverage_RSom}
+\end{center}
+\caption{For R+S OMs, probability of 95\% confidence interval for $\beta_M$ containing the true value for EMs with alternative process error assumptions and treatment of covariate effect ($\beta_E = 0$ or estimated).
+Vertical lines represent 95\% confidence intervals.}\label{beta_M_CI_coverage_RSom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_M_CI_coverage_RMom}
+\end{center}
+\caption{For R+M OMs, probability of 95\% confidence interval for $\beta_M$ containing the true value for EMs with alternative process error assumptions and treatment of covariate effect ($\beta_E = 0$ or estimated).
+Vertical lines represent 95\% confidence intervals.}\label{beta_M_CI_coverage_RMom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_M_rmse_main}
+\end{center}
+\caption{Root mean square error  (RMSE) of estimates of  $\beta_M$ from fitting EMs with alternative process error assumptions and treatment of covariate effect ($\beta_E = 0$ or estimated).
+All OMs had low observation error and contrast in fishing mortality.}\label{beta_M_rmse}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_M_rmse_Rom}
+\end{center}
+\caption{For R OMs, root mean square error (RMSE) of estimates of  $\beta_M$ from fitting EMs with alternative process error assumptions and treatment of covariate effect ($\beta_E = 0$ or estimated).}\label{beta_M_rmse_Rom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_M_rmse_RSom}
+\end{center}
+\caption{For R+S OMs, root mean square error (RMSE) of estimates of  $\beta_M$ from fitting EMs with alternative process error assumptions and treatment of covariate effect ($\beta_E = 0$ or estimated).}\label{beta_M_rmse_RSom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_M_rmse_RMom}
+\end{center}
+\caption{For R+M OMs, root mean square error (RMSE) of estimates of  $\beta_M$ from fitting EMs with alternative process error assumptions and treatment of covariate effect ($\beta_E = 0$ or estimated).}\label{beta_M_rmse_RMom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_M_bias_Rom}
+\end{center}
+\caption{For R OMs, median relative error (MRE) of estimates of natural mortality rate in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).}\label{terminal_M_bias_Rom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_M_bias_RSom}
+\end{center}
+\caption{For R+S OMs, median relative error (MRE) of estimates of natural mortality rate in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).}\label{terminal_M_bias_RSom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_M_bias_RMom}
+\end{center}
+\caption{For R+M OMs, median relative error (MRE) of estimates of natural mortality rate in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).}\label{terminal_M_bias_RMom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_ssb_bias_Rom}
+\end{center}
+\caption{For R OMs, median relative error (MRE) of estimates of spawning stock biomass (SSB) in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).}\label{terminal_ssb_bias_Rom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_ssb_bias_RSom}
+\end{center}
+\caption{For R+S OMs, median relative error (MRE) of estimates of spawning stock biomass (SSB) in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).}\label{terminal_ssb_bias_RSom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_ssb_bias_RMom}
+\end{center}
+\caption{For R+M OMs, median relative error (MRE) of estimates of spawning stock biomass (SSB) in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).}\label{terminal_ssb_bias_RMom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_F_bias_main}
+\end{center}
+\caption{Median relative error (MRE) of estimates of fully-selected fishing mortality ($F$) in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).
+All OMs had low observation error and contrast in fishing mortality.}\label{terminal_F_bias}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_F_bias_Rom}
+\end{center}
+\caption{For R OMs, median relative error (MRE) of estimates of fully-selected fishing mortality ($F$) in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).
+All OMs had low observation error and contrast in fishing mortality.}\label{terminal_F_bias_Rom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_F_bias_RSom}
+\end{center}
+\caption{For R+S OMs, median relative error (MRE) of estimates of fully-selected fishing mortality ($F$) in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).
+All OMs had low observation error and contrast in fishing mortality.}\label{terminal_F_bias_RSom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_F_bias_RMom}
+\end{center}
+\caption{For R+M OMs, median relative error (MRE) of estimates of fully-selected fishing mortality ($F$) in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).
+All OMs had low observation error and contrast in fishing mortality.}\label{terminal_F_bias_RMom}
+\end{figure}
+\end{landscape}

### 2025-05-29 Ecov_study/mortality/manuscript/manuscript.tex
saving work on ecov-m

@@
 Legault\textsuperscript{2} \and Brandon Muffley\textsuperscript{5} \and John Wiedenmann\textsuperscript{6}}
-\date{21 May, 2025}
+\date{29 May, 2025}
 
@@
 All of our analyses used the Woods Hole Assessment Model (WHAM) to construct both OMs and EMs \citep{millerstock20, stockmiller21, milleretal_inreview}.
-The WHAM package has been used extensively to configure OMs and EMs for several other simulation studies \citep{legaultetal23, lietal24, brittenetal_inreview, lietal_inreview_a} and is used to assess many commercially important stocks in the Northeast U.S.
+The WHAM package has been used extensively to configure OMs and EMs for several other simulation studies \citep{stocketal21, legaultetal23, lietal24, brittenetal_inreview, lietal_inreview_a} and is used to assess many commercially important stocks in the Northeast U.S.
 \citep[e.g.,][]{nefsc22, nefsc22a, nefsc24}.
@@
 
-In the WHAM model, environmental covariates are assumed to be described as state-space processes with annual observations of the true latent covariate {[}\citet{milleretal16};stockmiller21{]}.
+In the WHAM model, environmental covariates are assumed to be described as state-space processes with annual observations of the true latent covariate \citep{milleretal16, stockmiller21}.
 In our simulations, the latent covariate is assumed to be a stationary first order autoregressive (AR1) process
 \[
-X_y|X_{y-1} \sim \text{N}\left(\mu_E\left(1-\rho_E\right) + \rho_E E_{y-1}, \left(1-\rho_E^2\right)\sigma^2_E\right)
+X_y|X_{y-1} \sim \text{N}\left(\mu_E\left(1-\rho_E\right) + \rho_E X_{y-1}, \left(1-\rho_E^2\right)\sigma^2_E\right)
 \]
 with marginal mean \(\mu_E=0\) and variance \(\sigma^2_E\).
-The four configuration of the latent environmental covariate in the operating models assume the one of two values for the marginal standard deviation (\(\sigma_E \in \{0.1, 0.5\}\)) and for the autocorrelation parameter (\(\rho_E \in \{0, 0.5\}\)).
+The four configuration of the latent environmental covariate in the operating models assume the one of two values for the marginal standard deviation \(\sigma_E \in \{0.1, 0.5\}\) and for the autocorrelation parameter \(\rho_E \in \{0, 0.5\}\).
 
@@
 \]
-The standard deviation of the environmental observations in the operating models is one of two values \(\sigma_E \in \{0.1, 0.5\}\).
+The standard deviation of the environmental observations in the operating models is one of two values \(\sigma_e \in \{0.1, 0.5\}\).
 Figure \ref{om_ecov_example} provides example simulations of the latent process and observations under the alternative configurations.
@@
 
-Many of the characteristics of the population biology and structure are the same as those in \citet{milleretal_inreview1}.
-We configures the stock with 10 age classes (ages 1 to 10+) where the last is a ``plus'' group that accumulates any older individuals.
-We model the population for 40 years with annual catch and index observations.
-We assume spawning occurs annually 1/4 of the way through the year.
-The maturity at age was a logistic function of age with age at 50\% maturity (\(a_{50}\)) = 2.89 and slope = 0.88 (Figure \ref{om_inputs_fig}, top left).
-We assumed a von Beralanffy growth function for weight at age with length at age defined as
-\[
-L_a = L_{\infty}\left(1 - e^{-k(a - t_0)}\right)
-\]
-where \(t_0 = 0\), \(L_\infty = 85\), and \(k = 0.3\), and a length-weight
+Many of the characteristics of the population biology and structure including the age classes (10 age classes (ages 1 to 10+)), time span (40 years), maturity (Figure \ref{om_inputs_fig}, top left), growth (Figure \ref{om_inputs_fig}, top right), time of spawning (1/4 of the year), and recruitment (Figure \ref{om_inputs_fig}, bottom right) are identical to \citet{milleretal_inreview1}.
+The maturity at age is a logistic function of age with age at 50\% maturity (\(a_{50}\)) = 2.89 and slope = 0.88 and weight at age is derived from a von Beralanffy growth function where \(t_0 = 0\), \(L_\infty = 85\), and \(k = 0.3\), and a length-weight relationship
 \[
@@
 \]
-where \(\theta_1 = e^{-12.1}\) and \(\theta_2 = 3.2\) (Figure \ref{om_inputs_fig}, top right).
+where \(\theta_1 = e^{-12.1}\) and \(\theta_2 = 3.2\).
 
-The general model for natural mortality in year \(y\) is a log-linear function of a mean(median) parameter \(\beta_{M}\), effects of an environmental covariate \(E_y\) and a possibly autocorrelated normally distributed annual process error \(\varepsilon_{M,y}\)
+Here we consider both covariate effects and process errors for natural mortality and the general model for natural mortality in year \(y\) is a log-linear function of a parameter \(\beta_{M}\) that defines median natural mortality and possibly effects of the environmental covariate \(X_y\) described above and autocorrelated normally distributed annual process error \(\varepsilon_{M,y}\)
 \[
-\log M_y = \beta_M + \beta_{E} E_y + \varepsilon_{M,y}
+\log M_y = \beta_M + \beta_{E} X_y + \varepsilon_{M,y}
 \]
-\(\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)\) \citep{stockmiller21}.
+\(\epsilon_{M,y}|\epsilon_{M,y-1} \sim \text{N}\left(\epsilon_{M,y-1},(1-\rho_M^2)\sigma_M^2\right)\) \citep{stockmiller21}.
 We assume the median natural mortality rate \(e^{\beta_M} = 0.2\) is constant across ages.
-For R and R+S OMs and EMs, the process errors are set equal to 0.
+For R and R+S OMs and EMs, \(\epsilon_{M,y} = 0\).
 For all R+M OMs, we assume the same standard deviation \(\sigma_M = 0.3\) and is estimated in the R+M EMs.
@@
 The parameters defining the simulated covariate time series, size of the covariate effect, and any natural mortality random effects result in a range of different levels of variation in annual natural mortality rates (Figure \ref{M_example}).
-
-We assume is a single fishing fleet harvesting the population with a logistic selectivity function with age at 50\% selection \(a_{50} = 5\) and slope = 1 (Figure \ref{om_inputs_fig}, bottom left).
 
@@
 All biological inputs to calculations of spawning biomass per recruit (i.e., weight, maturity, and natural mortality at age) are constant in the R and R+S OMs without covariate effects on natural mortality.
-Therefore, steepness and unfished recruitment are also constant over the time period for those OMs \citep{millerbrooks21} and with our assumed biological inputs and selectivity the fishing mortality resulting in 40\% reduction in spawning biomass per recruit is \(F_{40\%} = 0.348\).
-With an assumed unfished recruitment of \(R_0 = e^{10}\), setting \(\Fmsy = \Fspr[40]\) results in a steepness of 0.69 and \(a=0.60\) and \(b = 2.4 \times 10^{-5}\) (Figure \ref{om_inputs_fig}, bottom right).
+Therefore, steepness and equilibrium unfished recruitment are also constant over the time period for those OMs \citep{millerbrooks21}.
+As in \citet{milleretal_inreview1}, our assumed biological inputs and selectivity (defined below) with constant natural mortality result in equilibrium fishing mortality that reduces spawning biomass per recruit to 40\% of the unfished level is \(\Fspr[40] = 0.348\).
+With an assumed unfished recruitment of \(R_0 = e^{10}\), setting \(\Fmsy = \Fspr[40]\) results in a steepness of 0.69 and \(a=0.60\) and \(b = 2.4 \times 10^{-5}\).
 
-We used two fishing scenarios for OMs.
+For R+M OMs and all OMs with covariate effects on natural mortality, steepness is not constant, but we used the same \(a\) and \(b\) parameters as other operating models which equates to a steepness and \(R_0\) at the median of the time series process for M.
+For operating models with time-varying random effects for fishery selectivity, \Fmsy is also not constant however we use the same fishing history as other operating models which corresponds to \Fmsy at the mean selectivity parameters.
+
+We also used the same two fishing scenarios as \citet{milleretal_inreview1} for OMs.
 In the first scenario, the stock experiences overfishing at 2.5\Fmsy for the first 20 years followed by fishing at \Fmsy for the last 20 years (denoted \(2.5\Fmsy \rightarrow \Fmsy\)).
 In the second scenario, the stock is fished at \Fmsy for the entire time period (40 years).
-The magnitude of the overfishing assumptions is based on average estimates of overfishing for NEUS groundfish stocks from \citet{wiedenmannetal19} and similar to the approach in \citet{legaultetal23}.
-
-We specified initial population abundance at age at the equilibrium distribution that corresponds to fishing at either \(F = 2.5\times \Fmsy\) or \(F = \Fmsy\).
-This implies that, for a deterministic model, the abundance at age would not change from the first year to the next.
-
-For R+M OMs and all OMs with environmental covariate effects on natural mortality, steepness is not constant, but we used the same \(a\) and \(b\) parameters as other operating models which equates to a steepness and \(R_0\) at the median of the time series process for M.
-For operating models with time-varying random effects for fishery selectivity, \Fmsy is also not constant however we use the same fishing history as other operating models which corresponds to \Fmsy at the mean selectivity parameters.
 
@@
 
-We assumed a single fleet operating year round for catch observations with logistic selectivity for the fleet with \(a_{50} = 5\) and slope = 1 (Figure \ref{om_inputs_fig}).
-This selectivity was used to define \Fmsy, \Fspr[40], and steepness for the Beverton-Holt stock recruitment parameters above.
-We assumed a logistic-normal distribution for the age-composition observations associated with the fleet where errors in the multivariate normal transformation are independent.
+We define the generation of observations of total catch, aggregate indices, and corresponding age composition identical to \citet{milleretal_inreview1}.
+There is a single fleet operating year round for catch observations with logistic selectivity for the fleet with \(a_{50} = 5\) and slope = 1 (Figure \ref{om_inputs_fig}, bottom left).
+Observations are generated for all 40 years of the model.
+There are two index time series intended to represent fishery-independent surveys occurring in the spring (0.25 way through the year) and the fall (0.75 way through the year).
+Catchability of both surveys are assumed to be 0.1.
+We assumed catch and index age composition observations aer generated from a logistic-normal distribution where errors on the multivariate normal scale are independent.
 The standard deviation parameter is also constant across ages.
-
-Two time series of surveys are assumed and observed in numbers rather than biomass for the entire 40 year period with one occurring in the spring (0.25 way through the year) and one in the fall (0.75 way through the year).
-Catchability of both surveys are assumed to be 0.1.
-We assumed the same selectivity and age composition distribution as the fleet for both indices.
 
@@
 
-For each data set simulated from an operating model 12 estimating models were fit. There were three factors defining the configuration of each estimating model
-
-\begin{itemize}
-\item whether the mean natural mortality $\beta_M$ was estimated or assumed known ($\log 0.2$)
-\item whether an environmental effect $\beta_E$ was estimated or not (fixed at 0)
-\item whether the process errors were assumed on recruitment only (R), recruitment and survival (R+S), or recruitment and natural mortality (R+M)
-\end{itemize}
+For each data set simulated from an operating model 12 estimating models were fit.
+There were three factors defining the configuration of each estimating model: 1) whether the mean natural mortality \(\beta_M\) was estimated or assumed known, 2) whether an environmental effect \(\beta_E\) was estimated or not, and 3) whether the process errors were assumed on recruitment only (R), recruitment and survival (R+S), or recruitment and natural mortality (R+M).
 
 The configuration of the process errors in the estimating models generally matched the corresponding options in the operating models.
-
 For example, uncorrelated R+S was assumed for both the estimating and operating model.
-However, R+M EMs did not assume M random effects were uncorrelated (parameter was estimated).
-The environmental covariate observations were included in all estimation models to ensure comparability of information criteria.
+However, R+M EMs did not assume M random effects were uncorrelated (\(\rho_M\) was estimated).
+The environmental covariate observations were included in all estimation models to ensure comparability of AIC.
 All fixed effects parameters for selectivity, catchability, fully-selected fishing mortality, mean recruitment, initial abundance at age, and variances for logistic-normal age composition distributions were estimated.
@@
 
-\hypertarget{convergence}{%
-\subsubsection*{Convergence}\label{convergence}}
-\addcontentsline{toc}{subsubsection}{Convergence}
+\hypertarget{em-convergence}{%
+\subsubsection*{EM convergence}\label{em-convergence}}
+\addcontentsline{toc}{subsubsection}{EM convergence}
 
@@
 \]
+All model fits that successfully completed the optimization were used in these results.
+We used all of these fits because some lack of convergence would be expected for the correct behavior of more complicated models that include process errors that did not exist in the operating model.
+For example R+M EMs fit to R OMs would be expected to estimate no variance in the natural mortality random effects and the estimated variance parameter going to zero would cause poor convergence.
 
-\hypertarget{bias}{%
-\subsubsection*{Bias}\label{bias}}
-\addcontentsline{toc}{subsubsection}{Bias}
+\hypertarget{parameter-inference-reliability}{%
+\subsubsection*{Parameter inference reliability}\label{parameter-inference-reliability}}
+\addcontentsline{toc}{subsubsection}{Parameter inference reliability}
 
-We calculated median errors of
+All results here use OM simulations with fits that satisfied the convergence criterion described above.
+We used this conditioning, to reflect how practitioners would proceed in analyses of model fits with real assessment data.
+That is, practitioners would ensure models converged such that hessian-based standard errors were available for all model parameter estimates.
+
+We calculated median errors (ME) of
 
@@
 
-and the median relative errors of
+and the median relative errors (MRE) of terminal year
 
 \begin{itemize}
-\item annual natural mortality rate,
-\item annual spawning stock biomass, and
-\item annual fully-selected fishing mortality rate.
+\item natural mortality rate ,
+\item spawning stock biomass, and
+\item fully-selected fishing mortality rate.
 \end{itemize}
 
-Note that the exponentiating the median error of \(\beta_M\) would be equivalent to the median relative error of the median natural mortality rate because the median of the exponential is the same as the exponential of the median.
+In preliminary analyses we inspected MRE of these parameters over the whole time series, but we found no appreciable differences in patterns across the various factors defining the OMs and EMs.
+
+We also calculated the root mean square error (RMSE) and estimated probability of coverage of constructed 95\% confidence intervals for \(\beta_E\) and \(\beta_M\) for EMs that estimated these parameters.
 
@@
 \]
-and measured bias as the median relative error and constructed 95\% confidence intervals using the binomial distribution approach as in \citet{stockmiller21}.
-If the confidence interval contains zero we do not have evidence of bias from the simulation study.
-We used estimates from all simulations that satisfied the first type of convergence described above.
-We were not concerned about more restrictive types of convergence because some estimation models would not be expected to satisfy these criteria because of mis-specified structures.
-For example, an estimation model that includes survival random effects fitted to a set of observations simulated without survival random effects should have trouble estimating variance of the survival random effects, but the estimates of the parameters and derived output described above, might be reliable.
+and constructed 95\% confidence intervals using the binomial distribution approach as in \citet{stockmiller21} and \citet{milleretal_inreview1}.
+Because of the inverse relationship of the the confidence interval and the null hypothesis significance test, when the confidence interval contains zero we do not have evidence of bias from the simulation study.
 
-For natural mortality rate, SSB, and fully-selected fishing mortality rate, we summarized results for each of the annual values, but we present results only for estimates from the first year (start), year 21 (middle), and the last year (end), because there were no appreciable differences between the results for these three years and those for the other years.
+For natural mortality rate, SSB, and fully-selected fishing mortality rate, we summarized results for each of the annual values, but we present results only for estimates from the terminal year, because there were no appreciable differences between the results among all years.
 
@@
 
-\hypertarget{convergence-1}{%
-\subsection*{Convergence}\label{convergence-1}}
-\addcontentsline{toc}{subsection}{Convergence}
+\hypertarget{em-convergence-1}{%
+\subsection*{EM Convergence}\label{em-convergence-1}}
+\addcontentsline{toc}{subsection}{EM Convergence}
 
@@
 
+We only present results for EMs where the median natural mortality rate parameter was estimated.
+Results differed little between results where this parameter was estimated or assumed known, but all of these results can be found in the Supplementary materials.
+
 When fitting the EMs to R and R+S OMs, AIC was generally accurate in determining the correct source of process errors and, when no effect of the covariate was simulated in the OMs, AIC was also accurate for the correct treatment of the environmental covariate \(\beta_E = 0\) (Figs.
@@
 
-\hypertarget{bias-1}{%
-\subsection*{Bias}\label{bias-1}}
+\hypertarget{bias}{%
+\subsection*{Bias}\label{bias}}
 \addcontentsline{toc}{subsection}{Bias}
@@
 
-\hypertarget{mean-natural-mortality-rate}{%
-\subsubsection*{Mean natural mortality rate}\label{mean-natural-mortality-rate}}
-\addcontentsline{toc}{subsubsection}{Mean natural mortality rate}
+\hypertarget{median-natural-mortality-rate}{%
+\subsubsection*{Median natural mortality rate}\label{median-natural-mortality-rate}}
+\addcontentsline{toc}{subsubsection}{Median natural mortality rate}
 
-We found high accuracy for estimation of the mean natural mortality rate parameter (\(\beta_M\)) for all EM process errors assumptions when OMs had step changes in fishing mortality, lower uncertainty in index and age composition observations, and either R or R+M process errors (Figure \ref{mean_M_bias_Ecov_beta_fixed}).
+We found high accuracy for estimation of the median natural mortality rate parameter (\(\beta_M\)) for all EM process errors assumptions when OMs had step changes in fishing mortality, lower uncertainty in index and age composition observations, and either R or R+M process errors (Figure \ref{beta_M_bias}).
 The most variation in estimates occurred when fishing mortality was constant and there was higher uncertainty in index and age composition observations.
@@
 
-When OMs and EMs assume \(\beta_M = log(0.2)\) and \(\beta_E = 0\), there is no annual variation in natural mortality for OMs with R or R+S process errors simulated or assumed in the EM and, therefore, estimation bias is not possible (Figure \ref{annual_M_bias_M_fixed_beta_fixed}).
+When OMs and EMs assume \(\beta_M = log(0.2)\) and \(\beta_E = 0\), there is no annual variation in natural mortality for OMs with R or R+S process errors simulated or assumed in the EM and, therefore, estimation bias is not possible (Figure \ref{terminal_M_bias}).
 We also observe little or no evidence of bias (confidence intervals include 0) for R or R+S EMs for any OMs even when the OMs included an effect of the covariate on natural mortality (\(\beta_E > 0\)).
@@
 
-When \(\beta_E\) was estimated, R EMs performed worse for R+S OMs with evidence of negative bias for OMs with constant fishing mortality, lower uncertainty in index and age composition observations, the largest covariate effect size, lower variability in the latent covariate, and larger uncertainty in the covariate observations (Figure \ref{annual_M_bias_M_fixed_beta_estimated}).
+When \(\beta_E\) was estimated, R EMs performed worse for R+S OMs with evidence of negative bias for OMs with constant fishing mortality, lower uncertainty in index and age composition observations, the largest covariate effect size, lower variability in the latent covariate, and larger uncertainty in the covariate observations (Figure \ref{terminal_M_bias}).
 There was little difference in the results for R+S EMs whether \(\beta_E\) was estimated or not.
@@
 
-Allowing the EMs to estimate \(\beta_M\) resulted in very large variability in estimates for all EMs for R and R+M OMs with constant fishing mortality and higher uncertainty in index and age composition observations (Figure \ref{annual_M_bias_M_estimated_beta_estimated}).
+Allowing the EMs to estimate \(\beta_M\) resulted in very large variability in estimates for all EMs for R and R+M OMs with constant fishing mortality and higher uncertainty in index and age composition observations (Figure \ref{terminal_M_bias}).
 The same variability occurred for R+S OMs, but strong bias was estimated for R and R+M EMs.
@@
 
-When EMs assume \(\beta_M = log(0.2)\) and \(\beta_E = 0\), there is little evidence of bias for R and R+M OMs except when there is a step change in fishing mortality and higher uncertainty in index and age composition observations, particularly at the end of the time series (Figure \ref{SSB_bias_M_fixed_beta_fixed}).
+When EMs assume \(\beta_M = log(0.2)\) and \(\beta_E = 0\), there is little evidence of bias for R and R+M OMs except when there is a step change in fishing mortality and higher uncertainty in index and age composition observations, particularly at the end of the time series (Figure \ref{terminal_ssb_bias}).
 We observed a similar trend in median relative error for R+S OMs, but there was more evidence of positive bias at the beginning of the time series whereas the confidence intervals for the negative median errors at the end of the time series often included 0.
@@
 
-When EMs assume \(\beta_M = log(0.2)\), but estimate \(\beta_E\), the median errors for SSB for R and and R+M OMs are similar to those when \(\beta_E = 0\) (Figure \ref{SSB_bias_M_fixed_beta_estimated}).
+When EMs assume \(\beta_M = log(0.2)\), but estimate \(\beta_E\), the median errors for SSB for R and and R+M OMs are similar to those when \(\beta_E = 0\) (Figure \ref{terminal_ssb_bias}).
 For R+S OMs, the median errors for EMs that also assume R+S process errors are also similar to those with no covariate effect assumed.
@@
 
-When EMs estimated both \(\beta_M\) and \(\beta_E\), we observed large variation in the errors in SSB under the same OM configurations where we observed large variation in errors for annual natural mortality rates (Figure \ref{SSB_bias_M_estimated_beta_estimated}).
+When EMs estimated both \(\beta_M\) and \(\beta_E\), we observed large variation in the errors in SSB under the same OM configurations where we observed large variation in errors for annual natural mortality rates (Figure \ref{terminal_ssb_bias}).
 Similarly, we observed little or no bias in SSB estimation for R+S EMs across all OM process error assumptions as long as there was a step change in fishing mortality and lower uncertainty in indices and age composition.
@@
 \addcontentsline{toc}{section}{Discussion}
+
+We found estimation of environmental effects on M was possible and reliable in certain scenarios even when the process error was mis-specified (R+S EM and R+M OM) also R EMs were not reliable for Ecov effects when PE was mis-specified.
+Given that milleretal in review found AIC was accurate for R+S OMs and applications of wham to real assessment data typically finds R+S preferrable to R only, it suggests inferences on covariate effects on M without just R process errors (e.g., deriso et al 2008) should be reevaluated.
 
@@
 \citet{derisoetal08} was the first to model natural mortality as a function of explicit covariate and residual random annual variation. Any relevant points?
+
+We conditioned our investigations of reliability of parameter estimation on simulations with fits that converged satisfactorily.
+However, these simulations could exclude many realizations of covariate, population, and observation time series that are not sufficiently informative for all the parameters that we were estimating.
+Why might this matter?
 
@@
 
-\begin{figure}
-\caption{Example simulations of environmental covariate latent processes and observations with different levels of observation error, and different assumptions about variability of the latent process.}\label{om_ecov_example}
-\begin{center}
-%\includegraphics[width = \textwidth]{Ecov_true_obs_example.png}
-\end{center}
-\end{figure}
-
-\begin{landscape}
-\begin{figure}
-\caption{Example simulations of annual natural mortality rates that may be a function of a temporally varying environmental covariate and autoregressive random effects.}\label{M_example}
-\begin{center}
-%\includegraphics[width = \textwidth]{M_example.png}
-\end{center}
-\end{figure}
-\end{landscape}
-
 \begin{landscape}
@@
 \begin{center}
-\includegraphics{convergence}
+\includegraphics{convergence_main}
 \end{center}
-\caption{Estimated probability of fits providing hessian-based standard errors for EMs assuming alternative process error, that estimate or assume known median natural mortality, and that estimate or assume no covariate effect on median natural mortality when fitted to operating models that have R (left) and R+S (middle) , or R+M (right) process error structures and three levels of true covariate effect on median natural mortality (x axis).
+\caption{Estimated probability of fits providing hessian-based standard errors for EMs with alternative process error assumptions, treatment of median natural mortality ($e^\beta_M$ known or estimated), and treatment of covariate effect ($\beta_E = 0$ or estimated).
+The OMs have R (left) and R+S (middle), or R+M (right) process error structures, alternative configurations of covariate time series structure and levels of observation uncertainty (rows), and three levels of true covariate effect on median natural mortality (x axis).
+All OMs had low observation error for fish population observations and temporal contrast in fishing pressure.
 Vertical lines represent 95\% confidence intervals.}\label{convergence}
@@
 \end{center}
-\caption{Proportion of simulated data sets for each operating model where the estimation model type (treatment of environmental covariate effect and  assumed process error type) had the lowest AIC.
+\caption{For each OM, the proportion of simulated data sets where the EM type (treatment of environmental covariate effect and  assumed process error type) had the lowest AIC.
 All OMs had low observation error for fish population observations and temporal contrast in fishing pressure.
@@
 \begin{center}
-%\includegraphics[height = \textheight]{Ecov_beta_bias}
+\includegraphics[height = \textheight]{beta_E_bias_main}
 \end{center}
-\caption{Relative error of estimates of environmental effect on natural mortality $\beta_E$ from fitting EMs with alternative process error assumptions.
+\caption{Median error (ME) of estimates of environmental effect on natural mortality $\beta_E$ from fitting EMs with alternative process error assumptions and treatment of median natural mortality ($e^\beta_M$ known or estimated).
 All OMs had low observation error and contrast in fishing mortality.
-Vertical lines represent 95\% confidence intervals.}\label{ecov_beta_bias}
+Vertical lines represent 95\% confidence intervals.}\label{beta_E_bias}
 \end{figure}
@@
 \begin{center}
-%\includegraphics[height = \textheight]{median_log_M_bias}
+\includegraphics[height = \textheight]{beta_E_CI_coverage_main}
 \end{center}
-\caption{Relative error of the estimated mean log-natural mortality parameter $\beta_{M}$ from fitting EMs with alternative process error assumptions.
+\caption{Probability of 95\% confidence interval for $\beta_E$ containing the true value for EMs with alternative process error assumptions and treatment of median natural mortality ($e^\beta_M$ known or estimated).
 All OMs had low observation error and contrast in fishing mortality.
-Vertical lines represent 95\% confidence intervals.}\label{median_log_M_bias}
+Vertical lines represent 95\% confidence intervals.}\label{beta_E_CI_coverage}
 \end{figure}
@@
 \begin{figure}
-\caption{Median relative error of annual natural mortality rate in terminal year from fitting simulated observations from each OM with alternative process error assumptions in the EM.
-Mean log-natural mortality parameter ($\beta_M$) and environmental covariate effect ($\beta_E$) are both estimated in the EMs.
-All OMs had low observation error and contrast in fishing mortality.}\label{terminal_M_bias}
 \begin{center}
-%\includegraphics[height = \textheight]{terminal_year_M_bias.png}
+\includegraphics[height = \textheight]{beta_M_bias_main}
 \end{center}
+\caption{Median error (ME) of estimates of $\beta_M$ from fitting EMs with alternative process error assumptions and treatment of covariate effect ($\beta_E = 0$ or estimated).
+All OMs had low observation error and contrast in fishing mortality.
+Vertical lines represent 95\% confidence intervals.}\label{beta_M_bias}
 \end{figure}
@@
 \begin{figure}
-\caption{Median relative error of annual SSB in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
-Estimating models also assume the mean natural mortality parameter $\beta_M = \log 0.2$ and no environmental covariate effect $\beta_E = 0$.}\label{SSB_bias_M_fixed_beta_fixed}
 \begin{center}
-%\includegraphics[height = \textheight]{SSB_bias_all_PE_effect_M_fixed_beta_fixed.png}
+\includegraphics[height = \textheight]{beta_M_CI_coverage_main}
 \end{center}
+\caption{Probability of 95\% confidence interval for $\beta_M$ containing the true value for EMs with alternative process error assumptions and treatment of covariate effect ($\beta_E = 0$ or estimated).
+All OMs had low observation error and contrast in fishing mortality.
+Vertical lines represent 95\% confidence intervals.}\label{beta_M_CI_coverage}
 \end{figure}
@@
 \begin{figure}
-\caption{Median relative error of annual SSB in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
-Estimating models also assume the mean natural mortality parameter $\beta_M = \log 0.2$ and the environmental covariate effect $\beta_E$ is estimated.}\label{SSB_bias_M_fixed_beta_estimated}
 \begin{center}
-%\includegraphics[height = \textheight]{SSB_bias_all_PE_effect_M_fixed_beta_estimated.png}
+\includegraphics[height = \textheight]{terminal_year_M_bias_main}
 \end{center}
+\caption{Median relative error (MRE) of estimates of natural mortality rate in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).
+All OMs had low observation error and contrast in fishing mortality.}\label{terminal_M_bias}
 \end{figure}
@@
 \begin{figure}
-\caption{Median relative error of annual SSB in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
-Estimating models also assume both the mean natural mortality parameter $\beta_M$ and the environmental covariate effect $\beta_E$ are estimated.}\label{SSB_bias_M_estimated_beta_estimated}
 \begin{center}
-%\includegraphics[height = \textheight]{SSB_bias_all_PE_effect_M_estimated_beta_estimated.png}
+\includegraphics[height = \textheight]{terminal_year_ssb_bias_main}
 \end{center}
+\caption{Median relative error (MRE) of estimates of spawning stock biomass (SSB) in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).
+All OMs had low observation error and contrast in fishing mortality.}\label{terminal_ssb_bias}
 \end{figure}
@@
 \end{figure}
+
+\begin{figure}
+\caption{Example simulations of environmental covariate latent processes and observations with different levels of observation error, and different assumptions about variability of the latent process.}\label{om_ecov_example}
+\begin{center}
+%\includegraphics[width = \textwidth]{Ecov_true_obs_example.png}
+\end{center}
+\end{figure}
+
+\begin{landscape}
+\begin{figure}
+\caption{Example simulations of annual natural mortality rates that may be a function of a temporally varying environmental covariate and autoregressive random effects.}\label{M_example}
+\begin{center}
+%\includegraphics[width = \textwidth]{M_example.png}
+\end{center}
+\end{figure}
+\end{landscape}
 
@@
 \begin{figure}
-\caption{Median relative error of annual fully-selected fishing mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model.
-Estimating models also assume both the mean natural mortality parameter $\beta_M$ and the environmental covariate effect $\beta_E$ are estimated.}\label{F_bias_M_estimated_beta_estimated}
 \begin{center}
-%\includegraphics[height = \textheight]{F_bias_all_PE_effect_M_estimated_beta_estimated.png}
+\includegraphics[height = \textheight]{beta_E_CI_coverage_Rom}
 \end{center}
+\caption{For R OMs, probability of 95\% confidence interval for $\beta_E$ containing the true value for EMs with alternative process error assumptions and treatment of median natural mortality ($e^\beta_M$ known or estimated).
+Vertical lines represent 95\% confidence intervals.}\label{beta_E_CI_coverage_Rom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_E_CI_coverage_RSom}
+\end{center}
+\caption{For R+S OMs, probability of 95\% confidence interval for $\beta_E$ containing the true value for EMs with alternative process error assumptions and treatment of median natural mortality ($e^\beta_M$ known or estimated).
+Vertical lines represent 95\% confidence intervals.}\label{beta_E_CI_coverage_RSom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_E_CI_coverage_RMom}
+\end{center}
+\caption{For R+M OMs, probability of 95\% confidence interval for $\beta_E$ containing the true value for EMs with alternative process error assumptions and treatment of median natural mortality ($e^\beta_M$ known or estimated).
+Vertical lines represent 95\% confidence intervals.}\label{beta_E_CI_coverage_RMom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_E_rmse_main}
+\end{center}
+\caption{Root mean square error (RMSE) of estimates of covariate effect on natural mortality $\beta_E$ from fitting EMs with alternative process error assumptions and treatment of median natural mortality ($e^\beta_M$ known or estimated).
+All OMs had low observation error and contrast in fishing mortality.}\label{beta_E_rmse}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_E_rmse_Rom}
+\end{center}
+\caption{For R OMs, root mean square error (RMSE) of estimates of covariate effect on natural mortality $\beta_E$ from fitting EMs with alternative process error assumptions and treatment of median natural mortality ($e^\beta_M$ known or estimated).
+}\label{beta_E_rmse_Rom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_E_rmse_RSom}
+\end{center}
+\caption{For R+S OMs, root mean square error (RMSE) of estimates of covariate effect on natural mortality $\beta_E$ from fitting EMs with alternative process error assumptions and treatment of median natural mortality ($e^\beta_M$ known or estimated).
+}\label{beta_E_rmse_RSom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_E_rmse_RMom}
+\end{center}
+\caption{For R+M OMs, root mean square error (RMSE) of estimates of covariate effect on natural mortality $\beta_E$ from fitting EMs with alternative process error assumptions and treatment of median natural mortality ($e^\beta_M$ known or estimated).
+}\label{beta_E_rmse_RMom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_M_bias_Rom}
+\end{center}
+\caption{For R OMs, median error (ME) of estimates of $\beta_M$ from fitting EMs with alternative process error assumptions and treatment of covariate effect ($\beta_E = 0$ or estimated).
+Vertical lines represent 95\% confidence intervals.}\label{beta_M_bias_Rom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_M_bias_RSom}
+\end{center}
+\caption{For R+S OMs, median error (ME) of estimates of $\beta_M$ from fitting EMs with alternative process error assumptions and treatment of covariate effect ($\beta_E = 0$ or estimated).
+Vertical lines represent 95\% confidence intervals.}\label{beta_M_bias_RSom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_M_bias_RMom}
+\end{center}
+\caption{For R+M OMs, median error (ME) of estimates of $\beta_M$ from fitting EMs with alternative process error assumptions and treatment of covariate effect ($\beta_E = 0$ or estimated).
+Vertical lines represent 95\% confidence intervals.}\label{beta_M_bias_RMom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_M_CI_coverage_Rom}
+\end{center}
+\caption{For R OMs, probability of 95\% confidence interval for $\beta_M$ containing the true value for EMs with alternative process error assumptions and treatment of covariate effect ($\beta_E = 0$ or estimated).
+Vertical lines represent 95\% confidence intervals.}\label{beta_M_CI_coverage_Rom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_M_CI_coverage_RSom}
+\end{center}
+\caption{For R+S OMs, probability of 95\% confidence interval for $\beta_M$ containing the true value for EMs with alternative process error assumptions and treatment of covariate effect ($\beta_E = 0$ or estimated).
+Vertical lines represent 95\% confidence intervals.}\label{beta_M_CI_coverage_RSom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_M_CI_coverage_RMom}
+\end{center}
+\caption{For R+M OMs, probability of 95\% confidence interval for $\beta_M$ containing the true value for EMs with alternative process error assumptions and treatment of covariate effect ($\beta_E = 0$ or estimated).
+Vertical lines represent 95\% confidence intervals.}\label{beta_M_CI_coverage_RMom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_M_rmse_main}
+\end{center}
+\caption{Root mean square error  (RMSE) of estimates of  $\beta_M$ from fitting EMs with alternative process error assumptions and treatment of covariate effect ($\beta_E = 0$ or estimated).
+All OMs had low observation error and contrast in fishing mortality.}\label{beta_M_rmse}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_M_rmse_Rom}
+\end{center}
+\caption{For R OMs, root mean square error (RMSE) of estimates of  $\beta_M$ from fitting EMs with alternative process error assumptions and treatment of covariate effect ($\beta_E = 0$ or estimated).}\label{beta_M_rmse_Rom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_M_rmse_RSom}
+\end{center}
+\caption{For R+S OMs, root mean square error (RMSE) of estimates of  $\beta_M$ from fitting EMs with alternative process error assumptions and treatment of covariate effect ($\beta_E = 0$ or estimated).}\label{beta_M_rmse_RSom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{beta_M_rmse_RMom}
+\end{center}
+\caption{For R+M OMs, root mean square error (RMSE) of estimates of  $\beta_M$ from fitting EMs with alternative process error assumptions and treatment of covariate effect ($\beta_E = 0$ or estimated).}\label{beta_M_rmse_RMom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_M_bias_Rom}
+\end{center}
+\caption{For R OMs, median relative error (MRE) of estimates of natural mortality rate in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).}\label{terminal_M_bias_Rom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_M_bias_RSom}
+\end{center}
+\caption{For R+S OMs, median relative error (MRE) of estimates of natural mortality rate in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).}\label{terminal_M_bias_RSom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_M_bias_RMom}
+\end{center}
+\caption{For R+M OMs, median relative error (MRE) of estimates of natural mortality rate in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).}\label{terminal_M_bias_RMom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_ssb_bias_Rom}
+\end{center}
+\caption{For R OMs, median relative error (MRE) of estimates of spawning stock biomass (SSB) in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).}\label{terminal_ssb_bias_Rom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_ssb_bias_RSom}
+\end{center}
+\caption{For R+S OMs, median relative error (MRE) of estimates of spawning stock biomass (SSB) in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).}\label{terminal_ssb_bias_RSom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_ssb_bias_RMom}
+\end{center}
+\caption{For R+M OMs, median relative error (MRE) of estimates of spawning stock biomass (SSB) in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).}\label{terminal_ssb_bias_RMom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_F_bias_main}
+\end{center}
+\caption{Median relative error (MRE) of estimates of fully-selected fishing mortality ($F$) in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).
+All OMs had low observation error and contrast in fishing mortality.}\label{terminal_F_bias}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_F_bias_Rom}
+\end{center}
+\caption{For R OMs, median relative error (MRE) of estimates of fully-selected fishing mortality ($F$) in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).
+All OMs had low observation error and contrast in fishing mortality.}\label{terminal_F_bias_Rom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_F_bias_RSom}
+\end{center}
+\caption{For R+S OMs, median relative error (MRE) of estimates of fully-selected fishing mortality ($F$) in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).
+All OMs had low observation error and contrast in fishing mortality.}\label{terminal_F_bias_RSom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_F_bias_RMom}
+\end{center}
+\caption{For R+M OMs, median relative error (MRE) of estimates of fully-selected fishing mortality ($F$) in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).
+All OMs had low observation error and contrast in fishing mortality.}\label{terminal_F_bias_RMom}
 \end{figure}

### 2025-05-29 Ecov_study/mortality/code/manuscript/M_results.R

@@ -1,0 +1,465 @@
+library(here)
+library(ggplot2)
+library(dplyr)
+library(ggh4x)
+library(tidyr)
+library(patchwork)
+library(scales)
+library(ggpattern)
+df.ems = readRDS(file.path(here(),"Ecov_study","mortality","inputs", "df.ems.RDS"))
+df.oms = readRDS(file.path(here(),"Ecov_study","mortality","inputs", "df.oms.RDS"))
+M_bias <- readRDS(file.path(here::here(),"Ecov_study","mortality", "results", "M_results.RDS"))
+conv_res <- readRDS(here("Ecov_study","mortality", "results", "convergence_results.RDS"))
+library(reshape2)
+#res <- melt(ssb_bias)
+
+#see convergence_results.R
+conv_fn <- function(om, em, conv_res, Type = 3){
+  x <- conv_res[[om]][[em]]
+  if(Type == 1) ind <- which(!is.na(x[,1]))
+  if(Type == 2) ind <- which(!is.na(x[,2]) & x[,2] == 0)
+  if(Type == 3) ind <- which(!is.na(x[,3]) & x[,3] == 0)
+  if(Type == 4) ind <- which(!is.na(x[,4]) & x[,4] < 1e-6)
+  if(Type == 5) ind <- which(!is.na(x[,5]) & x[,3] == 0 & x[,5] < 10)
+  return(ind)
+}
+# types of convergence
+#convergence information:
+#1: 0/1 model finished optimizing
+#2: nlminb convergence flag
+#3: number of NaNs in SEs for parameters, 0 = good invertible hessian
+#4: max gradient value < 1e-6
+#5: maximum non-NaN SE estimate < 10
+
+#all EM PE assumptions
+plot_df_fn <- function(df.ems, df.oms, Ecov_est = FALSE, M_est = FALSE, conv_type = 3) {
+  custom_boxplot_stat <- function(x){#, n.yrs = 1, n.sim = 100) {
+    x <- x[which(!is.na(x))]
+    n <- length(x)
+    bnds95 <- qbinom(c(0.025,0.975), n, 0.5)/n # 95% CI bounds for median (sims x years)
+    bnds80 <- qbinom(c(0.1,0.9), n, 0.5)/n # 80% CI bounds for median (sims x years)
+    r <- quantile(x, probs = c(bnds95[1], bnds80[1], 0.5, bnds80[2], bnds95[2]))
+    names(r) <- c("ymin", "lower", "middle", "upper", "ymax")
[…]

### 2025-05-29 Ecov_study/mortality/code/manuscript/annual_M_bias.R

@@ -1,373 +1,0 @@
-library(here)
-library(ggplot2)
-library(dplyr)
-library(ggh4x)
-library(tidyr)
-library(patchwork)
-library(scales)
-library(ggpattern)
-df.ems = readRDS(file.path(here(),"Ecov_study","mortality","inputs", "df.ems.RDS"))
-df.oms = readRDS(file.path(here(),"Ecov_study","mortality","inputs", "df.oms.RDS"))
-M_bias <- readRDS(file.path(here::here(),"Ecov_study","mortality", "results", "M_results.RDS"))
-library(reshape2)
-#res <- melt(ssb_bias)
-
-
-#all EM PE assumptions
-plot_df_fn <- function(df.ems, df.oms, Ecov_est = FALSE, M_est = FALSE) {
-  custom_boxplot_stat <- function(x){#, n.yrs = 1, n.sim = 100) {
-    x <- x[which(!is.na(x))]
-    n <- length(x)
-    bnds95 <- qbinom(c(0.025,0.975), n, 0.5)/n # 95% CI bounds for median (sims x years)
-    bnds80 <- qbinom(c(0.1,0.9), n, 0.5)/n # 80% CI bounds for median (sims x years)
-    r <- quantile(x, probs = c(bnds95[1], bnds80[1], 0.5, bnds80[2], bnds95[2]))
-    names(r) <- c("ymin", "lower", "middle", "upper", "ymax")
-    r
-  }
-  for(h in 1:3) {
-    re_mod <- c("rec", "rec+1", "rec+M")[h]
-    #EM:  M fixed, mean_M estimated
-    em_ind <- which(df.ems$Ecov_est== Ecov_est & df.ems$M_est == M_est)
-    print("em_ind")
-    print(em_ind)
-    om_ind <- which(df.oms$NAA_M_re == re_mod) #om and em match
-    res <- lapply(om_ind, function(x) {
-      M_res <- lapply(1:40, function(y) {
-        out <- matrix(NA,length(em_ind), 8)
-        for(j in em_ind){
-          r_e <- M_bias[[x]][,j,y,2]/M_bias[[x]][,j,y,1]-1
-          log_M_e <- log(M_bias[[x]][,j,y,2]) - log(M_bias[[x]][,j,y,1])
-          rmse <- sqrt(mean(log_M_e^2, na.rm = TRUE))
[…]

### 2025-05-29 Ecov_study/mortality/code/manuscript/ecov_beta_bias.R

@@ -1,143 +1,0 @@
-library(here)
-library(ggplot2)
-library(dplyr)
-library(ggh4x)
-library(tidyr)
-library(patchwork)
-library(scales)
-library(ggpattern)
-df.ems = readRDS(file.path(here(),"Ecov_study","mortality","inputs", "df.ems.RDS"))
-df.oms = readRDS(file.path(here(),"Ecov_study","mortality","inputs", "df.oms.RDS"))
-all_beta_bias <- readRDS(file.path(here::here(),"Ecov_study","mortality", "results", "ecov_beta_bias_results.RDS"))
-
-custom_boxplot_stat <- function(x, alpha = 0.05){#, n.yrs = 1, n.sim = 100) {
-  x <- x[which(!is.na(x))]
-  n <- length(x)
-  prob_ci <- qbinom(c(alpha/2,1-alpha/2), n, 0.5)/n # 95% CI bounds for median (sims x years)
-  r <- quantile(x, probs = c(prob_ci[1], 0.5, prob_ci[2]))
-  names(r) <- c("lower", "middle", "upper")
-  r
-}
-
-#all EM PE assumptions
-plot_df_fn <- function(df.ems, df.oms, M_est = FALSE, is_sdrep = TRUE) {
-  for(h in 1:3) {
-    re_mod <- c("rec", "rec+1", "rec+M")[h]
-    #EM:  M fixed, Ecov_beta estimated, OM and EM RE assumption match
-    df.ems. <- df.ems[df.ems$Ecov_est,]
-    em_ind <- which(df.ems.$M_est == M_est) # all 3 EM re_mods  
-    om_ind <- which(df.oms$NAA_M_re == re_mod) #om and em match
-    for(i in om_ind){
-      out <- matrix(NA,length(em_ind), 6)
-      for(j in em_ind){
-        x <- all_beta_bias[[i]][[j]]
-        if(is_sdrep) x <- x[which(!is.na(x[,2])),1] #remove fits without standard error estimates
-        out[which(em_ind==j),] <- c(median(x, na.rm = TRUE), sd(x, na.rm=T)/sqrt(sum(!is.na(x))), custom_boxplot_stat(x), sum(!is.na(x)))#quantile(x, probs = c(0.025,0.975)))
-      }
-      colnames(out) <- c("bias_est", "bias_se", "lower", "middle", "upper", "nfit")
[…]

### 2025-05-29 Ecov_study/mortality/code/manuscript/median_M_results.R

@@ -1,0 +1,225 @@
+library(here)
+library(ggplot2)
+library(dplyr)
+library(ggh4x)
+library(tidyr)
+library(patchwork)
+library(scales)
+library(ggpattern)
+df.ems = readRDS(file.path(here(),"Ecov_study","mortality","inputs", "df.ems.RDS"))
+df.oms = readRDS(file.path(here(),"Ecov_study","mortality","inputs", "df.oms.RDS"))
+median_M_bias <- readRDS(file.path(here::here(),"Ecov_study","mortality", "results", "mean_M_bias_results.RDS"))
+conv_res <- readRDS(here("Ecov_study","mortality", "results", "convergence_results.RDS"))
+
+custom_boxplot_stat <- function(x, alpha = 0.05){#, n.yrs = 1, n.sim = 100) {
+  x <- x[which(!is.na(x))]
+  n <- length(x)
+  prob_ci <- qbinom(c(alpha/2,1-alpha/2), n, 0.5)/n # 95% CI bounds for median (sims x years)
+  r <- quantile(x, probs = c(prob_ci[1], 0.5, prob_ci[2]))
+  names(r) <- c("lower", "middle", "upper")
+  r
+}
+
+#see convergence_results.R
+conv_fn <- function(om, em, conv_res, Type = 3){
+  x <- conv_res[[om]][[em]]
+  if(Type == 1) ind <- which(!is.na(x[,1]))
+  if(Type == 2) ind <- which(!is.na(x[,2]) & x[,2] == 0)
+  if(Type == 3) ind <- which(!is.na(x[,3]) & x[,3] == 0)
+  if(Type == 4) ind <- which(!is.na(x[,4]) & x[,4] < 1e-6)
+  if(Type == 5) ind <- which(!is.na(x[,5]) & x[,3] == 0 & x[,5] < 10)
+  return(ind)
+}
+# types of convergence
+#convergence information:
+#1: 0/1 model finished optimizing
+#2: nlminb convergence flag
+#3: number of NaNs in SEs for parameters, 0 = good invertible hessian
+#4: max gradient value < 1e-6
+#5: maximum non-NaN SE estimate < 10
+
+#all EM PE assumptions
+plot_df_fn <- function(df.ems, df.oms, Ecov_est = FALSE, conv_type = 3) {
[…]

### 2025-05-29 Ecov_study/mortality/code/manuscript/ssb_results.R

@@ -1,0 +1,243 @@
+library(here)
+library(ggplot2)
+library(dplyr)
+library(ggh4x)
+library(tidyr)
+library(patchwork)
+library(scales)
+library(ggpattern)
+df.ems = readRDS(file.path(here(),"Ecov_study","mortality","inputs", "df.ems.RDS"))
+df.oms = readRDS(file.path(here(),"Ecov_study","mortality","inputs", "df.oms.RDS"))
+ssb_bias <- readRDS(file.path(here::here(),"Ecov_study","mortality", "results", "ssb_results.RDS"))
+conv_res <- readRDS(here("Ecov_study","mortality", "results", "convergence_results.RDS"))
+library(reshape2)
+#res <- melt(ssb_bias)
+
+#see convergence_results.R
+conv_fn <- function(om, em, conv_res, Type = 3){
+  x <- conv_res[[om]][[em]]
+  if(Type == 1) ind <- which(!is.na(x[,1]))
+  if(Type == 2) ind <- which(!is.na(x[,2]) & x[,2] == 0)
+  if(Type == 3) ind <- which(!is.na(x[,3]) & x[,3] == 0)
+  if(Type == 4) ind <- which(!is.na(x[,4]) & x[,4] < 1e-6)
+  if(Type == 5) ind <- which(!is.na(x[,5]) & x[,3] == 0 & x[,5] < 10)
+  return(ind)
+}
+# types of convergence
+#convergence information:
+#1: 0/1 model finished optimizing
+#2: nlminb convergence flag
+#3: number of NaNs in SEs for parameters, 0 = good invertible hessian
+#4: max gradient value < 1e-6
+#5: maximum non-NaN SE estimate < 10
+
+custom_boxplot_stat <- function(x){#, n.yrs = 1, n.sim = 100) {
+  x <- x[which(!is.na(x))]
+  n <- length(x)
+  bnds95 <- qbinom(c(0.025,0.975), n, 0.5)/n # 95% CI bounds for median (sims x years)
+  bnds80 <- qbinom(c(0.1,0.9), n, 0.5)/n # 80% CI bounds for median (sims x years)
+  r <- quantile(x, probs = c(bnds95[1], bnds80[1], 0.5, bnds80[2], bnds95[2]))
+  names(r) <- c("ymin", "lower", "middle", "upper", "ymax")
+  r
+}
+plot_df_fn <- function(df.ems, df.oms, Ecov_est = FALSE, M_est = FALSE, conv_type = 3) {
[…]

### 2025-06-09 Ecov_study/mortality/manuscript/manuscript.Rmd
saving work on Ecov-M manuscript

@@
 
-We completed a large scale simulation study with 288 operating models, each with 100 simulated data sets, and 12 estimated models fit to each simulated data set.
-The factors defining operating model configuration included the source of process error on the population (recruitment, recruitment and survival, recruitment and natural mortality), the degree of temporal variation and autocorrelation of the environmental covariate, the uncertainty in the observation of the covariate, the uncertainty in indices and age composition observations, fishing history, and the magnitude of the effect of the covariate on natural mortality.
-The estimating models make alternative assumptions on whether to include the environmental effect, whether the mean/intercept log natural mortality (log(0.2)) is estimated or known, and whether process errors are on just recruitment, recruitment and survival, or recruitment and natural mortality.
-
-We found convergence of all estimation models was generally best when operating models assumed process errors in recruitment and survival, constant fishing rate, greater contrast in the true environmental covariate, and lower uncertainty in corresponding observations.
-Reliable convergence of all estimating models also occurred with the same process errors in the operating model and a step-change in fishing, but this also required lower uncertainty in index and age composition observations.
-Estimating models with process errors on recruitment and survival were unlikely to converge when the process errors in the operating model did not match whereas estimating models with process errors in recruitment and natural mortality converged for operating models without this match in certain cases.
-Probability of convergence generally decreased when the mean/intercept log-natural mortality rate parameter was estimated.
-
-Whether the mean log-natural mortality was estimated or not, the best accuracy of AIC for model selection occurred for models with process errors on recruitment and survival.
-AIC accuracy was poor for models with process errors on recruitment and natural mortality.
-estimating the mean natural mortality rate had small effects on the accuracy of AIC in selecting the appropriate process error.
-Estimating the mean log-natural mortality resulted in a small decrease in AIC accuracy.
-AIC was conservative for determining whether the environmental covariate affected natural mortality.
-AIC was very accurate in determining no effect when there was no effect in the operating model, but AIC often ranked the null model best when there was an effect.
-Accuracy of AIC for covariate effects improved with increased effect size, increased temporal contrast in the covariate, and lower uncertainty in observations.
-
-We found no evidence of bias in estimation of environmental effects regardless of process error assumptions when there was low uncertainty in the environmental observations and large constrast in the environmental covariate.
-In most cases the relative error of the estimated environmental effect did not depend on the source of process error assumed in the estimating model.
-The worst bias was observed when OMs assumed R+S process errors, high uncertainty in covariate observations, low variability in the covariate, and low uncertainty in index and age composition observations.
-Simultaneously estimating the mean/intercept log natural mortality resulted in larger variation in the relative errors of the estimated environmental effect.
-Estimation of the intercept was reliable for all EM process error assumptions when the operating models assumed process errors on recruitment and natural mortality, contrast in fishing pressure over time, and lower observation error.
-Estimating the mean/intercept for log natural mortality generally resulted in highly variable estimates of annual natural mortality and spawning biomass and evidence of bias for some operating and estimation model assumptions about process error source.
-Again reliability of annual natural mortality estimates was generally improved with lower observation error uncertainty and contrast in fishing pressure.
-
-Reliable detection of covariate effects requires informative data.
-AIC preferred simpler models than the true model when information content in data and contrast in covariates and abundance were low.
-The null model for environmental covariate effects (no covariate effect) was selected when contrast in the time series was low and/or uncertainty in observations was high.
-The selection of the null model by AIC also likely decreases with strength of the effect of the covariate on M.
-Similarly, when there was process error in recruitment and natural mortality, estimation models with process error only in recruitment were preferred presumably due to low variation in simulated natural mortality process errors.
-Covariate effect estimation can be robust to process error assumptions with high contrast in covariate and low observation error.
+Treatment of natural mortality is a major consideration in assessment models and state-space approaches allow estimation of temporal variation in this mortality rate as well as effects of specific covariates.
+However, there has been no investigation of the reliability of inferences made regarding natural mortality, associated covariate effects and important assessment output from state-space assessment models exists.
+We conducted a large-scale simulation study that considers models fit to data simulated from operating models with alternative assumptions defined by several factors, but we focus on scenarios where there is temporal contrast in fishing pressure and lower uncertainty in population observations (age composition and indices of abundance).
+The estimating models we fit to simulated observations had alternative assumptions on whether to include the environmental effect, whether the median natural mortality rate was estimated or known, and the source of temporal variability in the population demography.
+Lower uncertainty in covariate observations, higher covariate temporal variability and the source of demographic temporal variability were important for better inferences regarding the covariate effect on natural mortality.
+
+**keywords:** state-space assessment models, time-varying natural mortality, bias, AIC
 
@@
 State-space stock assessment models, with non-linear functions of latent processes and numerous observation types with different probability distribution assumptions represent one of most complex classes of state-space models.
-The literature on the ways we make inferences and the effects of various factors on reliabilty of inferences from state-space assessment models is growing [@lietal24;@milleretal_inreview1;cadiganetal].
+The literature on the effects of various factors on reliabilty of inferences from state-space assessment models is growing [@lietal24;@milleretal_inreview1].
 The importance of contrast in population size and fishing mortality and quality of data used to fit assessment models including the state-space variety is known [@magnussonhilborn07;@milleretal_inreview1].
-Furthermore, estimation of natural mortality, and even temporal variablity is possible in many scenarios [@leeetal11;@johnsonetal16;@cadigan16;@millerhyun18;@milleretal_inreview1].
-
-The effects of temporal variation in recruitment via unspecified or specified environmental factors have been extensively investigated in both traditional assessment models and state space models [@myers98;@milleretal16, haltuchpapers].
-Reliability of estimating environmental and spawning biomass effects on recruitment requires a combination of strong effects, good age composition data quality, contrast in the environmental covariate and lower recruitment variability [@brittenetal_inreview;@milleretal_inreview1].
-
-Temporal and environmental effects on growth and weight at age are also more important for short term projections and the reliability of estimation of those effects [@correaetal23; @correaetal_inreview].
-
-However, temporal variation in natural mortality and covariate effects on natural mortality are less studied @cadigan16.
-State-space assessment models currently used can treat the changes in cohort abundance over time as random effects and or  despite the importance of natural mortality in inferences for the size of fish populations, their productivity and projections necessary for making catch advice.
-In fact, natural mortality plays a more significant role in short term projections than recruitment due to immediate effects on older age classes that constitute spawning biomass and catch.
-
-See @milleretal_inreview1 for relevance of project 0 results (estimability of natural mortality).
-Some difficulty in distinguishing variation in natural mortality or effects of explicit covariates when variation in M random effects or Ecov is low relative to observation error.
-
-@derisoetal08 formulated the same natural mortality model as a function of covariates and random effects as we use in WHAM.
-
- Mis-specified temporal population process errors  could lead to misidentification of stock status and biased population estimates and poor fisheries management decisions [@trijouletetal20; @legaultpalmer16; @szuwalskietal18; @croninpunt21;@liljestrandetal24].
-
-In the present study, we conduct a simulation study with operating models (OMs) varying by degree of observation error, source and variability of process error, and fishing history.
-The simulations from these OMs are fitted with estimation models (EMs) that make alternative assumptions for sources of process error, whether a SRR was estimated, and whether natural mortality is estimated.
-Given the confounding nature of process errors, developing diagnostic tools to detect model misspecification is of great scientific interest and could aid the next generation of stock assessments [@augeretal21].
-We evaluate whether convergence and Akaike Information Criterion (AIC) can correctly determine the source of process error and the existence of a SRR.
-We also evaluate when retrospective patterns occur and the degree of bias in the outputs of the assessment model that are important for management.
-
-<!--
-Application of state-space models in fisheries stock assessment and management has expanded dramatically within ICES, Canada and the Northeast US [@nielsenberg14;@cadigan16;@stockmiller21].
-State-space approaches that use random effects to parameterize process errors is considered best practice and a requirement for the next generation of stock assessment models.
-
-Much is known about the reliability of state-space models that are linear or Gaussian [@aeberhardetal18], but applications in fisheries management are nonlinear and typically include multiple types of observations with varying distributional assumptions.
-Furthermore, there is a wide range of potential random effects structures and model parameters that can be treated as random effects in assessment models.
-We know relatively little about the statistical reliability of such models or the ability of information criteria to distinguish among such alternative structures.
-
- But those studies focus primarily on Gaussian linear models.
-review literature on reliability of hidden/latent process models. Primarily in other fields.
-
-Convergence results can be useful for understanding how bad convergence in applications to real data might direct the practitioner to which alternative random effects configurations is be more appropriate.
-The type of convergence that we might use as a diagnostic is important.
-And what rule of thumb we use for certain criteria e.g.
-1e-6?
-large SEs/CVs?
--->
-
-Estimation of natural mortality is known to be challenging in stock assessment models (@leeetal11, others). However, estimation has been shown to be reliable in some situations [@millerhyun18]
-
-Miller et al in review (project 0) found estimation of natural mortality was feasible when data are good and contrast in fishing pressure etc.
-
-Let's focus results here on those situations so that we can reduce the plots in figs.
-
-Variation in natural mortality has an immediate impact on projections unlike recruitment.
-Understanding this varation in M leads to better understanding of post-recruit productivity and therefore management.
-
-Here we conduct a simulation study with OMs varying by degree of observation error uncertainty, sources of process error, fishing history, temporal variation in environmental covariates, and magnitude of the effect of the covariate on natural mortality.
-The simulations from these operating models are fitted with estimating models that make alternative assumptions for sources of process error, and whether (mean) M is estimated.
-We evaluate effects of these factors on convergence of fitted models, whether AIC can correctly determine the correct source of process error and correct assumption about covariate effects on natural mortality, and the degree of bias in relevant parameters and outputs of the assessment model.
+Furthermore, estimation of natural mortality, and even temporal variablity is possible in many scenarios [@leeetal11;@cadigan16;@millerhyun18;@milleretal_inreview1].
+
+The effects of temporal variation in recruitment via undefined or explicit environmental factors have been extensively investigated in both traditional assessment models and state space models [@myers98; @haltuchpunt11; @johnsonetal16; @milleretal16].
+Reliability of estimating environmental and spawning biomass effects on recruitment in state-space assessment models requires a combination of strong effects, good age composition data quality, contrast in the environmental covariate and lower recruitment variability [@brittenetal_inreview;@milleretal_inreview1].
+
+A critical aspect of fisheries assessment models and their use in management is short-term projections that are used to determine catch advice.
+While understanding drivers of recruitment is important particularly for subsequent effects on reference points, recruitment in short-term projections typically has little impact on the exploitable biomass in the first few projection years.
+However, assumptions for natural mortality ($M$) have immediate and larger effects on projected biomass because they affect the abundances at older age classes at the end of the data time series that constitute spawning biomass and catch [@brodziaketal08;@stocketal21].
+
+Because of the effects of $M$ on both biological reference points and short term projections, better understanding sources of variation in $M$ would provide more accurate estimation of abundance and productivity and therefore improved management.
+Temporal variation in $M$ is less studied than recruitment, but its importance for explaining variability in observations has been demonstrated in state-space assessment models for Atlantic cod and yellowtail flounder [@cadigan16;@stocketal21].
+@derisoetal08 also demonstrated the importance of several factors affecting Pacific herring natural mortality.
+
+Assessment models could include temporal variation in many aspects of population dynamics or how observations are related to the population.
+For example the Woods Hole Assessment Model (WHAM) can include process errors treated as random effects for transition in cohorts over time, catchability for indices of abundance, selectivity of fishing fleets or indices, movement between regions, or in $M$ [@stockmiller21;@milleretal_inreview].
+However, misspecified temporal population process errors could lead to biased population and stock status estimation, and, therefore, poor fisheries management decisions [ @legaultpalmer16; @szuwalskietal18].
+Studies of the reliability of inferences we might make regarding the presence of temporal variability in $M$ are limited.
+@milleretal_inreview1 found AIC could accurately distinguish process errors in the changes in abundance for cohorts, but not for those specifically due to $M$ except when uncertainty in population observations (indices, catch and age composition) was low and there was greater temporal variation in natural mortality.
+In their simulation studies looking a models with multiple sources of process error, @lietal24 found including more sources of process error than existed in the operating model was a better model-building approach than excluding them a priori.
+
+Here we conduct a simulation study with operating models (OMs) varying by degree of observation error uncertainty, sources of process error, fishing history, temporal variation in environmental covariates, and magnitude of the effect of the covariate on natural mortality.
+The simulated observations from these OMs are fitted with estimating models that make alternative assumptions for sources of process error, and whether median M and covariate effects are estimated.
+We evaluate the effects of these factors on convergence of fitted models, whether Akaike's information criterion (AIC) can correctly determine the correct source of process error and correct assumption about covariate effects on natural mortality, and the degree of bias in inferences for relevant parameters and outputs of the assessment model.
 
@@
 For each simulated data set we fit a set of 12 estimating models (EMs).
-The factors that distinguish the estimating models which are also described in detail below include source of population process error type (3 levels) whether (median) natural mortality rate was estimated or assumed known (2 levels), and whether the effect of the environmental covariate on natural mortality was estimated or not (2 levels).
-
-The sources of population process error that were used in the OMs or assumed in the EMs were on recruitment only (R), recruitment and changes in cohort abundance over time (R+S), or recruitment and natural mortality (R+M).
+The factors that distinguish the estimating models which are also described in detail below include source of population process error type (3 levels) whether (median) $M$ was estimated or assumed known (2 levels), and whether the effect of the environmental covariate on $M$ was estimated or not (2 levels).
+
+The sources of population process error that were used in the OMs or assumed in the EMs were on recruitment only (R), recruitment and changes in cohort abundance over time (R+S), or recruitment and $M$ (R+M).
 We did not use the log-normal bias-correction feature for process errors or observations described by @stockmiller21 for operating and EMs [@lietal_inreview].
@@
 
-Here we consider both covariate effects and process errors for natural mortality and the general model for natural mortality in year $y$ is a log-linear function of a  parameter $\beta_{M}$ that defines median natural mortality and possibly effects of the environmental covariate $X_y$ described above and autocorrelated normally distributed annual process error $\varepsilon_{M,y}$
+The general model for $M$ in year $y$ is a log-linear function of both covariate effects $X_y$ and process errors $\varepsilon_{M,y}$ and a  parameter $\beta_{M}$ that defines median $M$ 
 $$
@@
 $$
-$\epsilon_{M,y}|\epsilon_{M,y-1} \sim \text{N}\left(\epsilon_{M,y-1},(1-\rho_M^2)\sigma_M^2\right)$ [@stockmiller21].
-We assume the median natural mortality rate $e^{\beta_M} = 0.2$ is constant across ages.
-For R and R+S OMs and EMs, $\epsilon_{M,y} = 0$.
+where the process errors are modeled as random effects that may, in general, be autocorrelated normal random variables 
+$$
+\varepsilon_{M,y}|\varepsilon_{M,y-1} \sim \text{N}\left(\varepsilon_{M,y-1},(1-\rho_M^2)\sigma_M^2\right)
+$$ 
+[@stockmiller21], but we assume $\rho_M = 0$ in our R+M OMs.
+We assume the median $M$ rate $e^{\beta_M} = 0.2$ is constant across ages.
+For R and R+S OMs and EMs, $\varepsilon_{M,y} = 0$.
 For all R+M OMs, we assume the same standard deviation $\sigma_M = 0.3$ and is estimated in the R+M EMs.
-The covariate effect is one of 3 alternative values in the operating models ($\beta_\text{E} \in \{0,0.25,0.5\}$.
-The parameters defining the simulated covariate time series, size of the covariate effect, and any natural mortality random effects result in a range of different levels of variation in annual natural mortality rates (Figure \ref{M_example}).
-
-We assumed expected recruitment each year from a Beverton-Holt stock-recruit relationship (SRR)
+The covariate effect is one of 3 alternative values in the operating models, $\beta_\text{E} \in \{0,0.25,0.5\}$.
+The parameters defining the simulated covariate time series, size of the covariate effect, and any $M$ random effects result in a range of different levels of variation in annual values (Figure \ref{M_example}).
+
+We assumed expected recruitment each year is from a Beverton-Holt stock-recruit relationship (SRR)
 \[
@@
 \]
-All biological inputs to calculations of spawning biomass per recruit (i.e., weight, maturity, and natural mortality at age) are constant in the R and R+S OMs without covariate effects on natural mortality.
+All biological inputs to calculations of spawning biomass per recruit (i.e., weight, maturity, and $M$ at age) are constant in the R and R+S OMs without covariate effects on $M$.
 Therefore, steepness and equilibrium unfished recruitment are also constant over the time period for those OMs [@millerbrooks21].
-As in @milleretal_inreview1, our assumed biological inputs and selectivity (defined below) with constant natural mortality result in equilibrium fishing mortality that reduces spawning biomass per recruit to 40\% of the unfished level is $\Fspr[40] = 0.348$.
-With an assumed unfished recruitment of $R_0 = e^{10}$, setting $\Fmsy = \Fspr[40]$ results in a steepness of 0.69 and $a=0.60$ and $b = 2.4 \times 10^{-5}$.
-
-For R+M OMs and all OMs with covariate effects on natural mortality, steepness is not constant, but we used the same $a$ and $b$ parameters as other operating models which equates to a steepness and $R_0$ at the median of the time series process for M.
-For operating models with time-varying random effects for fishery selectivity, \Fmsy is also not constant however we use the same fishing history as other operating models which corresponds to \Fmsy at the mean selectivity parameters.
+As in @milleretal_inreview1, our assumed biological inputs and selectivity (defined below) with constant $M$ result in equilibrium fishing mortality that reduces spawning biomass per recruit to 40\% of the unfished level is $F_{40\%} = 0.348$.
+With an assumed unfished recruitment of $R_0 = e^{10}$, setting $\Fmsy = F_{40\%}$ results in a steepness of 0.69 and $a=0.60$ and $b = 2.4 \times 10^{-5}$.
+For R+M OMs and all OMs with covariate effects on $M$, steepness is not constant, but we used the same $a$ and $b$ parameters as other operating models which equates to a steepness and $R_0$ at the median of the time series processes for $M$ and the covariate.
 
@@
 For each data set simulated from an operating model 12 estimating models were fit.
-There were three factors defining the configuration of each estimating model: 1) whether the mean natural mortality $\beta_M$ was estimated or assumed known, 2) whether an environmental effect $\beta_E$ was estimated or not, and 3) whether the process errors were assumed on recruitment only (R), recruitment and survival (R+S), or recruitment and natural mortality (R+M).
+There were three factors defining the configuration of each estimating model: 1) whether $\beta_M$ was estimated or assumed known, 2) whether an environmental effect $\beta_E$ was estimated or not, and 3) whether the process errors were assumed on recruitment only (R), recruitment and survival (R+S), or recruitment and $M$ (R+M).
 
@@
 All fixed effects parameters for selectivity, catchability, fully-selected fishing mortality, mean recruitment, initial abundance at age, and variances for logistic-normal age composition distributions were estimated.
-Any process error variance parameters for recruitment, survival, and natural mortality were also estimated.
+Any process error variance parameters for recruitment, survival, and $M$ were also estimated.
 The observation error variance of the environmental observations and aggregate catch and indices were all assumed known at the true values.
@@
 
-The first measure of reliability we investigated was frequency of convergence when fitting each estimating model to the simulated data sets.
-There are various ways to assess convergence of the fit, but we defined successful convergence as the hessian of the marginal log-likelihood being invertible and providing variance estimates for the fixed effects parameters.
+We measured the frequency of convergence when fitting each EM to the simulated data sets for each OM.
+There are various ways to assess convergence of the fit [e.g., @carvalhoetal21], but we defined successful convergence as the Hessian of the marginal log-likelihood being invertible and providing variance estimates for the fixed effects parameters as recommended by @milleretal_inreview1.
+We calculated 95\% confidence intervals for probability of convergence using the Clopper-Pearson exact method [@clopperpearson34;@thulin14].
 
@@
 We measured the frequency of correct model selection using marginal AIC.
-For a given operating model the set of models that were considered all made the same assumptions on whether or not to estimate (mean) natural mortality rate or it is assumed at the true value $\beta_M = log(0.2)$.
+For a given operating model the set of models that were considered all made the same assumptions on whether or not to estimate $\beta_M$ or it is assumed at the true value.
 For model $m$, the marginal AIC is a function of the marginal log-likelihood maximized with respect to the fixed effects in the model $\boldsymbol{\theta}$ and the number of fixed effects $n\left(\boldsymbol{\theta}\right)$ estimated,
@@
 We used all of these fits because some lack of convergence would be expected for the correct behavior of more complicated models that include process errors that did not exist in the operating model.
-For example R+M EMs fit to R OMs would be expected to estimate no variance in the natural mortality  random effects and the estimated variance parameter going to zero would cause poor convergence.
-
-### Parameter inference reliability {-}
+For example R+M EMs fit to R OMs would be expected to estimate no variance in the $M$  random effects and the estimated variance parameter going to zero would cause poor convergence.
+
+### Parameter estimation {-}
 
@@
 We used this conditioning, to reflect how practitioners would proceed in analyses of model fits with real assessment data.
-That is, practitioners would ensure models converged such that hessian-based standard errors were available for all model parameter estimates.
-
-We calculated median errors (ME) of $\beta_E$, the effect of environmental covariates on natural mortality, and the median natural mortality rate parameter $\beta_M$ (the mean log-natural mortality rate).
-We calculated the median relative errors (MRE) of terminal year estimates of natural mortality rate, spawning stock biomass, and fully-selected fishing mortality rate.
-In preliminary analyses we inspected MRE of these parameters over the whole time series, but we found no appreciable differences in patterns across the various factors defining the OMs and EMs.
+That is, practitioners would ensure models converged such that Hessian-based standard errors were available for all model parameter estimates.
+
+We focused on statistical behavior of estimators of the covariate effect on $M$ ($\widehat \beta_E$), the estimator of the median $M$ parameter ($\widehat \beta_M$), and estimators of terminal year natural mortality $\widehat M$, spawning stock biomass $\widehat{\text{SSB}}$, and fully-selected fishing mortality rate $\widehat F$.
+In preliminary analyses we results of the estimators of all the annual values for $M$, SSB and fishing mortality over the whole time series, but we found no appreciable differences in patterns across the various factors defining the OMs and EMs.
 Furthermore, results for terminal year fishing mortality were generally inversely related to those for spawning stock biomass, and are provided in the Supplementary Materials and not discussed further.
 
-We also calculated the root mean square error (RMSE) and estimated probability of coverage of constructed 95\% confidence intervals for $\beta_E$ and $\beta_M$ for EMs that estimated these parameters.
-Confidence intervals for probabilities of coverage are constructed using the same methods as those for probabilities of convergence.
-
-Results for fishing mortality rate For the $i$th simulated data the relative error for a parameter $\theta$ provided from the fitted estimation model is
-$$
-{\text{RE}_i}\left(\theta\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
-$$
-and  constructed 95\% confidence intervals using the binomial distribution approach as in @stockmiller21 and @milleretal_inreview1.
+We calculated median errors (ME) of $\widehat \beta_E$, $\widehat \beta_M$), and the Hessian-based standard error estimators of these parameters ($\widehat{SE}\left(\widehat \beta_E\right)$ and $\widehat{SE}\left(\widehat \beta_M\right)$).
+We calculated the median relative errors (MRE) of terminal year $\widehat M$, $\widehat{\text{SSB}}$, and $\widehat F$.
+We also calculated the root mean square error (RMSE) and estimated probability of coverage of constructed 95\% confidence intervals for $\widehat \beta_E$ and $\widehat \beta_M$ for EMs that estimated these parameters.
+We constructed the confidence intervals for probabilities of CI coverage using the same methods as those for probabilities of convergence.
+
+The true values for terminal year SSB and fishing mortality rate vary among simulations.
+For the $i$th simulated data the relative error for terminal year value $\theta_i$ provided from the fitted estimation model is
+$$
+\text{RE}_i\left(\theta\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
+$$
+We calculated RMSE as
+$$
+\text{RMSE}\left(\widehat \theta\right) = \sqrt{\frac{1}{n}\sum^n\left(\widehat \theta_i - \theta_i\right)^2}
+$$
+For ME and RME of estimators, we constructed 95\% confidence intervals using the binomial distribution approach as in @stockmiller21 and @milleretal_inreview1.
 Because of the inverse relationship of the the confidence interval and the null hypothesis significance test, when the confidence interval contains zero we do not have evidence of bias from the simulation study.
 
-For natural mortality rate, SSB, and fully-selected fishing mortality rate, we summarized results for each of the annual values, but we present results only for estimates from the terminal year, because there were no appreciable differences between the results among all years.
-
 # Results {-}
 
-Because @milleretal_inreview1 found inferences are most reliable for scenarios with low observation error for indices and age composition data and with temporal contrast in fishing pressure, we restrict our attention to scenarios with these characteristics in the main paper (temporal contrast in fishing and low observation error), but include corresponding results for the other scenarios in the Supplementary Materials.
+@milleretal_inreview1 found inferences are most reliable for distinguishing process error sources and stock-recruit relationships in scenarios with lower observation error for indices and age composition data and with temporal contrast in fishing pressure.
+Our expectation is that inferences for covariate effects on $M$ would require data to be at least that informative and inspection of the comprehensive set of results across all OMs generally confirms this (See supplemental Materials).
+Therefore, we restrict our attention to these OM scenarios with more informative data in the main paper (temporal contrast in fishing and lower population observation error).
 
@@
 
-When the median natural mortality rate was assumed known, corresponding to usual practice in application of assessment models, good convergence for all EMs was observed for R+S operating models with \Fmsy fishing history, more variation in the latent environmental covariate ($\sigma_E = 0.5$), and lower error in the associated observations $\sigma_e = 0.1$ (Figure \ref{convergence}).
-Good convergence for all EMs was also observed for R+S OMs with the step change in fishing history, but with low error in indices and age comp observations.
-
-EMs that assumed random effects just on recruitment (R), nearly always converged across all operating models that assumed more variation in the latent environmental covariate ($\sigma_E = 0.5$) and lower error in the associated observations $\sigma_E = 0.1$.
-EMs that assumed recruitment and survival random effects (R+S), had poor converge probability when the OMs had alternative process errors (R or R+M).
-The R+S EMs showed best convergence for the R+S OMs where there was lower error in the environmental observations $\sigma_E = 0.1$ or higher error in environmental observations of latent true environmental covariates with greater temporal variation.
-EMs that assumed recruitment and natural mortality random effects (R+M), had poor convergence probability for all OMs that had a step change in fishing history and higher uncertainty in indices and age composition.
-
-When the OMs and EMs both assume process errors only on recruitment (R) or on recruitment and survival (R+S), convergence was worst with less variation in the true environmental covariate and larger uncertainty in associated observations.
-When the OMs and EMs both assume process errors on recruitment and natural mortality (R+M) convergence was problematic for all OMs with the step-change in fishing history.
-The best convergence was observed with this match between OMs and EMs was when the fishing history was constant and there was low uncertainty in environmental observations.
-
-As might be expected, there was an overall drop in the probability of convergence when the mean natural mortality rate was estimated rather than assumed at the true value (Figure \ref{convergence}).
-Otherwise, general trends described above with mean natural mortality fixed, apply when estimated.
+In OMs with contrast in fishing pressure and lower uncertainty in population observations (catch, indices and age composition), R EMs generally converged with high frequency except when EMs estimated covariate effects on $M$ and the covariate uncertainty ($\sigma_e$) was high (Figure \ref{convergence}).
+R+S EMs generally converged with low frequency when the OM process error did not match the EM configuration.
+R+M EMs generally did not converge with high frequency except for R+S OMs with low covariate uncertainty and where the EMs assumed median $M$ as known.
+There was no strong effect of the treatment of median M (known or estimated) on convergence for any of the other EMs.
+The largest effect of the size of the true covariate effect ($\beta_E$) on convergence was observed for R and R+M EMs that estimated covariate effects in R and R+M OMs with high uncertainty in covariate observations ($\sigma_e$) and high variability in covariate process errors ($\sigma_E$).
+There was an increase in convergence frequency with increased variability in covariate process errors for many EMs in OMs with large covariate uncertainty.
 
@@
 
-We only present results for EMs where the median natural mortality rate parameter was estimated.
-Results differed little between results where this parameter was estimated or assumed known, but all of these results can be found in the Supplementary materials.
-
-When fitting the EMs to R and R+S OMs, AIC was generally accurate in determining the correct source of process errors and, when no effect of the covariate was simulated in the OMs, AIC was also accurate for the correct treatment of the environmental covariate $\beta_E = 0$ (Figs.
-\ref{aic}, \ref{aic_Rom}, \ref{aic_RSom}).
-However, when OMs included an effect of the covariate AIC accuracy for estimating an effect in the EM was only found when the OMs also included lower covariate observation error and higher effect size and higher temporal variability in the covariate.
-On the other hand, for R+M Oms, AIC generally was not able to accurately determine the correct source of process errors, but was able to accurately determine the correct treatment of the environmental covariate (Figs.
-\ref{aic} and \ref{aic_RMom}).
-
-When estimating models assumed the mean natural mortality rate was known, the best accuracy of AIC for model selection occurred for models with R+S process errors.
-R+S estimating models ranked best with very low frequency for R or R+M operating models and with very high frequency for R+S operating models (Figure \ref{aic}).
-R estimating models were determined best with high frequency for R operating models, but also for R+M operating models.
-R+M estimating models were rarely determined best for any operating models including those where the process errors matched.
-
-AIC was conservative for determining whether the environmental covariate affected natural mortality.
-AIC was highly accurate in determining no effect when there was no effect in the operating model, but AIC ranked the null model best with high frequency even when there was an effect in the operating model in many cases.
-However the accuracy of AIC improved in certain operating models.
-Increased effect size, increased temporal contrast in the covariate, and lower uncertainty in all observations types lead to increased accuracy of determining covariate effects.
-
-Relative to the assumption that the mean natural mortality rate was known, estimating the mean natural mortality rate had small effects on the accuracy of AIC in selecting the appropriate process error and whether the covariate affect natural mortality (Figure \ref{aic}).
-Where there were differences there were small decreases in accuracy for determining the appropriate process error and determining a covariate effect when there was one.
-
-## Bias {-}
-
-### Environmental effect {-}
-
-When the EMs assumed the median natural mortality rate was known, we observed generally accurate estimation of environmental effects across all EM and OM process error assumptions and all true covariate effect sizes, when ther was low uncertainty in environmental observations and larger temporal contrast in the simulated true environmental covariate (Figure \ref{beta_E_bias}).
-We observed a negative trend in bias of the environmental effect with increased effect size when temporal variation in the covariate was lower and/or uncertainty in the covariate observations was higher.
-When the OMs had R+S process errors with low temporal variation in the true environmental covariate and lower uncertainty in the indices age age composition, estimated covariate effects were highly variable.
-In most cases the relative error of $\beta_E$ did not depend on the source of process error assumed in the EM.
-When there was an effect of the EM process error assumption it was when OMs had R+S process errors.
-The worst bias was observed when OMs assumed R+S process errors, high uncertainty in covariate observations, low variability in the covariate, and low uncertainty in index and age composition observations.
-
-When the median natural mortality rate was estimated in the EM, results were similar except estimated effects were even more variable for data simulated with R+S process errors (Figure \ref{beta_E_bias}).
-There was also more separation of reliability of the estimation of the effect among EMs with different process error assumptions.
-The separation was most apparent when OMs simulated R+S process errors and larger variability in the environmental covariate wehere EMs with process errors other than R+S showing more bias than when the mean natural mortality rate was assumed known.
-
-### Median natural mortality rate {-}
-
-We found high accuracy for estimation of the median natural mortality rate parameter ($\beta_M$) for all EM process errors assumptions when OMs had step changes in fishing mortality, lower uncertainty in index and age composition observations, and either R or R+M process errors (Figure \ref{beta_M_bias}).
-The most variation in estimates occurred when fishing mortality was constant and there was higher uncertainty in index and age composition observations.
-For OMs with R+S process errors, the most reliable estimation of $\beta_M$ was obtained when the EM also assumed R+S process errors across all other factors.
-For OMs with R+M process errors, the matching assumption for the EM only showed the best reliability when there was low uncertainty in index and age composition observations and fishing mortality was constant.
-
-### Annual natural morality rate {-}
-
-We present results for error in annual natural mortality rate conditional on three alternative EM configurations for the natural mortality parameters: 1) mean natural mortality rate parameter is fixed at the true value ($\beta_M = log(0.2)$) and no covariate effect is assumed ($\beta_E = 0$),  2) $\beta_M = log(0.2)$ and $\beta_E$ is estimated, and 3) both $\beta_M$ and $\beta_E$ are estimated.
-
-When OMs and EMs assume $\beta_M = log(0.2)$ and $\beta_E = 0$, there is no annual variation in natural mortality for OMs with R or R+S process errors simulated or assumed in the EM and, therefore, estimation bias is not possible (Figure \ref{terminal_M_bias}).
-We also observe little or no evidence of bias (confidence intervals include 0) for R or R+S EMs for any OMs even when the OMs included an effect of the covariate on natural mortality ($\beta_E > 0$).
-Including process error on M (R+M) produces more variability in errors for R and R+S EMs than including increased level of effect of the covariate on M.
-For R and R+S OMs, errors in annual M for R+S EMs were less variable than those for R EMs.
-
-R+M EMs fit to R OMs with constant fishing mortality, or a step change in fishing mortality and lower uncertainty in index and age composition observations exhibited no bias in annual estimation of M indicating that the variance of the estimated random effects for M in these EMs collapsed to 0.
-R+M EMs were the only EMs to exhibit differences in the sign of the median errors across the time series.
-When OMs had a step change in fishing mortality and higher uncertainty in index and age composition observations, we observed positive median errors at the beginning of the time series and lower median errors in later years.
-However, whether there was evidence of bias, depended on the uncertainty in covariate observations and the degree of variability in the latent covariate.
-We observed negative median relative errors in annual M estimates for R+M EMs fit to R+S OMs with lower uncertainty in index and age composition observations across all levels of simulated effects of the covariate, but evidence of bias was strongest at the beginning of the time series.
-
-When $\beta_E$ was estimated, R EMs performed worse for R+S OMs with evidence of negative bias for OMs with constant fishing mortality, lower uncertainty in index and age composition observations, the largest covariate effect size, lower variability in the latent covariate, and larger uncertainty in the covariate observations (Figure \ref{terminal_M_bias}).
-There was little difference in the results for R+S EMs whether $\beta_E$ was estimated or not.
-The results for R+M EMs were generally similar to those when $\beta_E=0$ was assumed, except that the differences between the estimates at the beginning of the time series and later years for certain OM configurations did not occur.
-
-Allowing the EMs to estimate $\beta_M$ resulted in very large variability in estimates for all EMs for R and R+M OMs with constant fishing mortality and higher uncertainty in index and age composition observations (Figure \ref{terminal_M_bias}).
-The same variability occurred for R+S OMs, but strong bias was estimated for R and R+M EMs.
-Annual M estimation was most reliable when OMs had step changes in fishing mortality and lower uncertainty in index and age composition observations.
-For R and R+M OMs with those characteristics, all EMs generally provided accurate estimation of annual M, but only R+S EMs provided accurate estimation for R+S OMs.
-Little or no bias in annual M estimation was observed for R+S EMs across all OM process error assumptions as long as there was a step change in fishing mortality and lower uncertainty in indices and age composition.
-
-### Spawning stock biomass {-}
-
-Like natural mortality, we present results for error in spawning biomass conditional on three alternative EM configurations for the natural mortality parameters: 1) mean natural mortality rate parameter is fixed at the true value ($\beta_M = log(0.2)$) and no covariate effect is assmed ($\beta_E = 0$),  2) $\beta_M = log(0.2)$ and $\beta_E$ is estimated, and 3) both $\beta_M$ and $\beta_E$ are estimated.
-
-When EMs assume $\beta_M = log(0.2)$ and $\beta_E = 0$, there is little evidence of bias for R and R+M OMs except when there is a step change in fishing mortality and higher uncertainty in index and age composition observations, particularly at the end of the time series (Figure \ref{terminal_ssb_bias}).
-We observed a similar trend in median relative error for R+S OMs, but there was more evidence of positive bias at the beginning of the time series whereas the confidence intervals for the negative median errors at the end of the time series often included 0.
-However, there was more indication of positive bias of the incorrect process error assumption of the EMs for the R+S OMs.
-For R+S EMs fit to R+S OMs, there was indication of small positive bias at the beginning of the time series when there was constant fishing mortality and higher uncertainty in index and age composition observations.
-
-When EMs assume $\beta_M = log(0.2)$, but estimate $\beta_E$, the median errors for SSB for R and and R+M OMs are similar to those when $\beta_E = 0$ (Figure \ref{terminal_ssb_bias}).
-For R+S OMs, the median errors for EMs that also assume R+S process errors are also similar to those with no covariate effect assumed.
-However, for R and R+M EMs fit to R+S OMs with constant fishing mortality rate and lower uncertainty in indices and age composition observations, we observed evidences of negative bias when higher covariate observation uncertainty and lower variation in the latent covariate was simulated and positive bias for other configurations of the covariate and corresponding observation uncertainty.
-When R+S OMs had constant fishing mortality rate and higher uncertainty in indices and age composition observations or step changes in fishing mortality and lower uncertainty in indices and age composition, R and R+M EMs often provided positively biased SSB estimates when there was lower uncertainty in covariate observations.
-
-When EMs estimated both $\beta_M$ and $\beta_E$, we observed large variation in the errors in SSB under the same OM configurations where we observed large variation in errors for annual natural mortality rates (Figure \ref{terminal_ssb_bias}).
-Similarly, we observed little or no bias in SSB estimation for R+S EMs across all OM process error assumptions as long as there was a step change in fishing mortality and lower uncertainty in indices and age composition.
+We only present results for EMs where the median $M$ rate parameter was estimated because results differed little whether this parameter was estimated or assumed known (see Figures \ref{aic_Rom} to \ref{aic_RMom}).
+When there is temporal contrast in fishing pressure and lower uncertainty in population observations, the only OM scenarios where AIC was consistently accurate for both covariate effect and process error configurations were R OMs with high variability in the covariate and low uncertainty in observations of the covariate (Figure \ref{aic}).
+R+M OMs with those same covariate attributes were only accurate for the covariate effect (low Type I and II errors) and the incorrect process error almost always chosen was R rather than R+S.
+We also observed high accuracy for both covariate effect and process error configurations for R+S OMs where either no effect of the covariate or the strongest effect of the covariate was simulated.
+All R and R+S OMs were accurate for process error configuration and across all OMs, AIC accurately selects models without covariate effects when there are none (low Type I error).
+See Figures \ref{aic_Rom} to \ref{aic_RMom} for full results.
+
+## Covariate effect estimation {-}
+
+When there is temporal contrast in fishing pressure, lower uncertainty in population observations, low uncertainty in environmental observations, and larger temporal contrast in the simulated true environmental covariate, we observed generally little or no bias in estimation of covariate effects ($\widehat \beta_E$) across all EM and OM process error assumptions, both EM configurations of median $M$, and all true covariate effect sizes (Figure \ref{beta_E_bias}, rows 3 and 4).
+The only exception to this was when R EMs with median $M$ estimated were fit to R+S OMs with those configurations.
+Decreasing trends in bias of $\widehat \beta_E$ for EMs fit to many configurations of R and R+M OMs indicate that $\widehat \beta_E$ values were closer to zero on average even when the true effect increased.
+For R+S OMs, EMs with the matching process error configuration showed little evidence of bias in estimating $\beta_E$ across a range of covariate and covariate observation configurations.
+
+Bias of Hessian-based standard error estimation for $\widehat \beta_E$ was also close to zero in the same OM scenarios where bias in estimation of $\beta_E$ itself was close to zero (Figure \ref{se_beta_E_bias}, rows 3 and 4).
+However, in these high information OMs bias of standard error estimation was observed in for several EMs that did not have the matching R+S and R+M process error configuration.
+Estimation of standard errors was essentially unreliable for most EMs when OMs had higher uncertainty in covariate observations (Figure \ref{se_beta_E_bias}, rows 5 to 8).
+
+Despite the reliable estimation of covariate effects and standard errors with low uncertainty in environmental observations, and larger temporal contrast in the simulated true environmental covariate, we found poor confidence interval coverage for many EMs fitted to R+S and R+M OMs (Figure \ref{beta_E_CI_coverage}, rows 3 and 4).
+For example, R+M EMs fit to R+M OMs showed little bias in estimation of the covariate effect and standard errors, but confidence interval coverage was negatively biased (rows 3 and 4 in right two columns of Figures \ref{beta_E_bias}, \ref{se_beta_E_bias}, and \ref{beta_E_CI_coverage}.
+However, there appears to be a positive correlation of the covariate and standard error estimates such that estimates less than the true value had negatively biased standard error estimates which would make confidence intervals too small (Figure \ref{ex_lm_beta_E_SE_beta_E}).
+Confidence interval coverage was most reliable for R+S OMs when the EMs had the matching process error configuration (Figure \ref{beta_E_CI_coverage}, middle 2 columns).
+
+## Median natural mortality rate estimation {-}
+
+We observed negligible bias in estimating the median $M$ parameter ($\beta_M$) over a much wider set of operating models than that for estimating the covariate effect on $M$.
+When there is temporal contrast in fishing pressure, lower uncertainty in population observations, all EMs fit to R and R+M OMs showed little evidence of bias for $\widehat \beta_M$  (Figure \ref{beta_M_bias}, left and right sets of columns).
+For R+S OMs with those conditions, EMs with the matching process error assumption showed little evidence of bias.
+
+We also found little or no bias of Hessian-based standard error estimation for $\beta_M$ in the many of the same OM scenarios where bias of $\widehat \beta_M$ was close to zero except for some EMs with the process error source mis-specified or where the OM simulated the strongest covariate effect (Figure \ref{se_beta_M_bias}).
+Like $\widehat \beta_M$, reliability of standard error estimation in R+S OMs required the EMs to have the matching process error assumption.
+
+However, we found confidence interval coverage to be generally unreliable even for OM and EM combinations where bias in $\widehat \beta_M$ and its standard error estimators was negligible, much like that for $\widehat \beta_E$ (Figure \ref{beta_M_CI_coverage}).
+In the same EM OM combination we investigated above for $\widehat \beta_E$, we observed an opposite negative correlation of $\widehat \beta_M$ and its standard error estimates (Figure \ref{ex_lm_beta_M_SE_beta_M}), which would result in confidence intervals being too narrow when $\widehat \beta_M$ values are larger than average.
+
+## Terminal year natural morality rate {-}
+
+The EMs with R and R+S process error configurations with median M assumed known and covariate effects not estimated will have terminal year M correctly specified when fitted to R and R+S OMs with no covariate effect because $M$ is constant in both the EM and OM at the same value.
+We exclude those OM-EM combinations from our results here.
+
+In many of the OMs with temporal contrast in fishing pressure and lower uncertainty in population observations, there was little difference in bias for terminal year M among EMs whether the median $M$ parameter ($\beta_M$) was estimated (Figure \ref{terminal_M_bias}).
+However, when there were differences, bias in terminal year M was closer to zero when the EM assumed $\beta_M$ known, As would be expected.
+For R OMs, all estimating models exhibited little evidence of bias except when OMs simulated the largest covariate effect ($\beta_E = 0.5$).
+However, bias estimates were at worst around 10\% (Figure \ref{terminal_M_bias}, left columns).
+For R+S OMs, the R+S EMs generally exhibited the least evidence for bias and R+M EMs generally estimated terminal M lower than the true value, particularly when observation uncertainty in the covariate was lower.
+For R+M OMs, many of the EM configurations provided little bias in terminal $\widehat M$, but the biases were generally further from zero than the results for the R OMs (Figure \ref{terminal_M_bias}, right columns).
+Accuracy of terminal year $\widehat M$ was similar for all of the EMs when fitted to R+M OMs (Figure \ref{terminal_M_rmse}).
+For R+S OMs, accuracy was a bit better for R+S EM than others.
+As would be expected accuracy of estimate where median $M$ was assumed known was better than when it was estimated, when there were differences.
+
+## Terminal year spawning stock biomass {-}
+
+In OMs with temporal contrast in fishing pressure and lower uncertainty in population observations, we found little bias in estimates of terminal year SSB for all EM configurations when fitted to R or R+M OMs (Figure \ref{terminal_ssb_bias}).
+We also found little bias for R+S OMs when the EMs assumed the correct process error configuration.
+The worst bias in terminal year SSB estimation we found was in R+S OMs for R EMs with $\beta_M$ estimated.
+Like the bias results, accuracy as measured by RMSE was similar for many of EMs when fit to R and R+M OMs.
+Accuracy was generally worse when EMs were fit to R+S OMs (Figure \ref{terminal_ssb_rmse}).
 
@@
 
-We found estimation of environmental effects on M was possible and reliable in certain scenarios even when the process error was mis-specified (R+S EM and R+M OM) also R EMs were not reliable for Ecov effects when PE was mis-specified.
-Given that milleretal in review found AIC was accurate for R+S OMs and applications of wham to real assessment data typically finds R+S preferrable to R only, it suggests inferences on covariate effects on M without just R process errors (e.g., deriso et al 2008) should be reevaluated.
-
-The estimating models assumed variances of aggregate catch and index observations was known.
-This approximation may be appropriate for indices where we have a reliable estimate of uncertainty based on the survey design (), but there may be better approaches for the aggregate catch such as an informed prior on the standard errors with realistic bounds.
-
-We found EMs with R+M process errors were rarely determined as the appropriate model when OMs simulated R+M process errors.
-This was unexpected but is likely due to the size of the variance assumed for those process errors ($\sigma_M = 0.3$) relative to the variances assumed for index and age composition observations.
-This is related to the difficulty in separately estimating observation and process error variances when the ratio of process to observation uncertainty is low.
-
-Note that the results for bias of covariate effect all assume ecov beta is estimated.
-The lack of bias in certain situations might suggest including the effect even if AIC says it isn't better at least when contrast in Ecov is high.
-
-See project 0 paper for relevance of @lietal24 and @liljestrandetal24. See @milleretal_inreview1 fro relevance of project 0 results (estimability of natural mortality)
-
-@derisoetal08 was the first to model natural mortality as a function of explicit covariate and residual random annual variation. Any relevant points?
-
-We conditioned our investigations of reliability of parameter estimation on simulations with fits that converged satisfactorily.
+Our simulation study demonstrated that estimation of environmental effects on $M$ is possible and reliable in certain scenarios even when the process error was misspecified (e.g., R and R+S EMs fit to R+M OM).
+In many of these same OMs frequency of convergence of model fits did not appears to suffer when covariate effects on $M$ were estimated even when there was no effect simulated.
+However, these scenarios are information rich in that there was contrast in population size (due to changes in fishing pressure) and the covariate affecting the population, and low uncertainty in population and covariate observations.
+Previous research has shown that estimation of a constant $M$ parameter requires contrast in time series and informative data [@leeetal11], so it is no surprise that estimation of these effects also requires relatively good information via more precise observations and higher contrast in the covariate time series.
+
+On the other hand, AIC could only reliably detect covariate effects for R and R+M OMs with contrast in covariates and low covariate uncertainty and confidence interval coverage for the covariate effect was often biased even when bias in the estimators of the effect and the standard error were negligible.
+@cadiganetal24 found confidence interval coverage to be biased for SSB and fishing mortality estimation in a state-space model in some scenarios, but they attributed the poor coverage to bias in Hessian-based standard error estimation and their simulations held any process error random effects constant.
+The coverage bias we observed, at least for some OM-EM combinations, may be a related to correlation of the estimators of the covariate effect and the corresponding standard error and therefore consideration of other methods of calculating confidence intervals may be warranted (e.g., those based on profile likelihood and/or MCMC sampling of log-likelihood surface).
+
+@milleretal_inreview1 investigated R+M OMs with two levels of $M$ process error variability ($\sigma_M \in {0.1,0.5}$ and only found AIC able to accurately distinguish R+M process errors with the higher level of process error variability ($\sigma_M = 0.5$).
+We assumed $\sigma_M = 0.3$, intermediate to the values investigated by @milleretal_inreview1, and found process error inferences unreliable for the source of process error, indicating that the level of variability required for detecting $M$ process errors must be greater than $\sigma_M = 0.3$, but may still be less than 0.5.
+In our results and those by @milleretal_inreview1, AIC typically chose R EMs which would indicate the fitted R+M EMs would estimate no variability in the $M$ process errors.
+Future studies like ours where R+M OMs are simulated with greater variability in $M$ process errors would better inform reliability of covariate effect inferences under such scenarios.
+
+@derisoetal08 attempted to estimate process errors as well as covariate effects with $M$ for Pacific herring, but similarly found no variability in $M$, suggesting a there was too much uncertainty in the available observations relative to the true temporal variability in $M$.
+Given that we found covariate effect inferences using R EMs was reliable in R+M OMs with apparently little variability in $M$ process errors, the findings of @derisoetal08 on covariate effects for Pacific herring would presumably also be robust to true low variability in $M$.
+However, they did not account for uncertainty in covariate observations, some of which would probably have substantial uncertainty (e.g., competition and predation covariates).
+We found higher covariate observation uncertainty to cause true covariate effects to be less detectable using AIC, but we did not investigate the implications for incorrectly assuming no covariate uncertainty for covariate inferences.
+
+Any bias or poor accuracy for annual SSB estimation was primarily a function of whether or not the median $M$ parameter was estimated or known and the types of process errors than the treatment of the covariate effects on $M$.
+For example,  we found estimation of SSB was better when the EM had the process error correctly specified for R+S OMs.
+Fortunately, our results and those by @milleretal_inreview1 demonstrate that marginal AIC seems to be a good tool for determining whether this source of process error should be included in the model.
+However, the reliability of the estimation of SSB does break down in the less ideal scenarios when there is higher population observation error, and lack of contrast in fishing pressure (e.g., Figures \ref{terminal_ssb_bias_Rom} to \ref{terminal_ssb_bias_RMom}).
+
+The ability to accurately infer covariate effects on $M$ in some realistic situations indicates that such investigations may be fruitful.
+Ability to make inferences could improve further when WHAM is extended to incorporate tagging data as expected [@milleretal_inreview].
+Tagging data can greatly inform natural mortality estimation [@pollocketal91;@hampton00], and this impact on $M$ estimation should also apply to estimation of covariate effects or unexplained temporal variation in $M$.
+Given our findings coupled with future WHAM development, we expect investigations of and accounting for covariate effects on $M$ to become more common within the fisheries stock assessment process.
+
+<!-- We conditioned our investigations of reliability of parameter estimation on simulations with fits that converged satisfactorily.
 However, these simulations could exclude many realizations of covariate, population, and observation time series that are not sufficiently informative for all the parameters that we were estimating.
 Why might this matter?
-
-Poor quality of biological, fishery, and observation may adversely affect ability to distinguish between true process and observational error [@puntetal14; @stewartmonnahan17; @croninpunt21; @fischetal23; @lietal_inreview_a].
-
-Uncertainty in estimated population attributes such as spawnign biomass can be increased considerably when natural mortality is estimated.
-This is also a consideration for estimating covariate effects on natural mortality even if the median natural mortality is assumed known.
-
-The bias in natural mortality estimation resulted in biased estimation of stock size and likely harvest rates, but it will also propagate into biological reference points and possibly stock status.
-
-The lack of bias observed for annual M, using R+S EMs when no effect was assumed must be due to the annual M being equal to the assumed value (0.2) on average across simulations because the simulated environmental covariate has mean 0.
+-->
+
+<!-- Note that the results for bias of covariate effect all assume ecov beta is estimated.
+The lack of bias in certain situations might suggest including the effect even if AIC says it isn't better at least when contrast in Ecov is high.
+-->
+
+<!-- See project 0 paper for relevance of @lietal24 and @liljestrandetal24. See @milleretal_inreview1 fro relevance of project 0 results (estimability of $M$) -->
+
+<!-- @stocketal21 showed temporal variation in M to be important for yellowtail flounder whereas @milleretal_inreview found models without it (at age 1) (only on apparent survival) performed better for bsb.
+-->
+
+<!-- @derisoetal08 was the first to model $M$ as a function of explicit covariates and residual random annual variation, but they did not use the random effects for $M$ because they found them inestimable for their application.
+They also did not account for any uncertainty in the covariates they investigated for effects on $M$ although many of them would presumably have substantial uncertainty (e.g., predator consumption).
+Our study found the degree of uncertainty in the covariate observation can lead to incorrect conclusions about their effects.
+-->
+
+<!-- Poor quality of biological, fishery, and observation may adversely affect ability to distinguish between true process and observational error [@puntetal14; @stewartmonnahan17; @croninpunt21; @fischetal23; @lietal_inreview_a].
+-->
+
+<!-- Uncertainty in estimated population attributes such as spawnign biomass can be increased considerably when $M$ is estimated.
+This is also a consideration for estimating covariate effects on $M$ even if the median $M$ parameter is assumed known.
+-->
+
+<!-- The bias in $M$ estimation resulted in biased estimation of stock size and likely harvest rates, but it will also propagate into biological reference points and possibly stock status. -->
+
+<!-- The lack of bias observed for annual M, using R+S EMs when no effect was assumed must be due to the annual M being equal to the assumed value (0.2) on average across simulations because the simulated environmental covariate has mean 0.
 However, if one were to condition on the covariate time series we would expect biased estimation of annual M when there was a covariate effect.
-
-# Conclusions {-}
-
-As we would expect, reliable detection of covariate effects requires informative data. 
-AIC preferred simpler models than the true model when information content in data and contrast in covariates and abundance were low.
+-->
+
+<!-- # Conclusions {-} -->
+
+<!-- As we would expect, reliable detection of covariate effects requires informative data.  -->
+<!-- AIC preferred simpler models than the true model when information content in data and contrast in covariates and abundance were low.
 Null model for environmental covariate effect (no covariate effect) was selected when contrast in the time series was low  and/or uncertainty in observations was high.
@@
 Reliable estimation of environmental effect on M regardless of the process error assumed by the EM as long as the contrast in the covariate is sufficient and the uncertainty in the observations is low.
+-->
 
@@
 \end{center}
-\caption{Estimated probability of fits providing hessian-based standard errors for EMs with alternative process error assumptions, treatment of median natural mortality ($e^\beta_M$ known or estimated), and treatment of covariate effect ($\beta_E = 0$ or estimated).
+\caption{Estimated probability of fits providing Hessian-based standard errors for EMs with alternative process error assumptions, treatment of median natural mortality ($e^\beta_M$ known or estimated), and treatment of covariate effect ($\beta_E = 0$ or estimated).
 The OMs have R (left) and R+S (middle), or R+M (right) process error structures, alternative configurations of covariate time series structure and levels of observation uncertainty (rows), and three levels of true covariate effect on median natural mortality (x axis).
@@
 
+\begin{landscape}
 \begin{figure}
@@
 \begin{center}
-%\includegraphics[width = \textwidth]{Ecov_true_obs_example.png}
-\end{center}
-\end{figure}
+\includegraphics[height = \textheight]{Ecov_true_obs_example}
+\end{center}
+\end{figure}
+\end{landscape}
 
@@
 \begin{center}
-%\includegraphics[width = \textwidth]{M_example.png}
-\end{center}
-\end{figure}
-\end{landscape}
+\includegraphics[height = \textheight]{M_example}
+\end{center}
+\end{figure}
+\end{landscape}
+
+## Convergence results {-}
+
+For R OMs, R EMs converged frequently except when there was contrast in fishing and covariate effects ($\beta_E$) were estimated and covariate observation uncertainty ($\sigma_e$) was high (Figure \ref{convergence_Rom}).
+There was little difference in convergence for R EMs whether median M was estimated or not.
+When R OMs had contrast in fishing, convergence of R+M EMs was generally better when the EM estimated median M, whereas in OMs with constant fishing, convergence was better when treating median M as known.
+Convergence of R+S EMs was generally worse than other process error configurations for R OMs, and there was no difference between EMs that estimated median M or not except when fishing pressure was constant where better convergence occurred when M was treated as known.
+Effects of true covariate effect size on convergence were apparent in some R OMs, but there did not appear to be any consistent patterns.
+
+For R+S OMs, R+M EMs generally showed the worse probability of convergence that EMs with R and R+S configurations (Figure \ref{convergence_RSom}).
+Convergence probability was high for all R and R+S EMs when covariate observation uncertainty was low.
+Although probability of convergence was generally greater when covariate effects were not estimated, there were no strong trends in convergence for any of the EMS with increased true effect size.
+
+For R+M OMs, R+S EMs generally converged less frequently than those with R and R+M configurations (Figure \ref{convergence_RMom}).
+Convergence probability for R OMs was generally the greatest of the three process error configurations.
+As with R+S OMs, there were no strong trends in convergence for any of the EMS with increased true covariate effect size.
+However, there was also not strong effects of estimating the covariate effect on convergence except for some OMs with higher covariate observation uncertainty.
+Like R OMs, when R+M OMs had contrast in fishing, convergence of R+M EMs was generally better when the EM estimated median M and the convergence was better when treating median M as known when the OM had constant fishing.
 
@@
 \end{center}
-\caption{Estimated probability of fits providing hessian-based standard errors for EMs assuming alternative process error, that estimate or assume known median natural mortality, and that estimate or assume no covariate effect on median natural mortality when fitted to R OMs and three levels of true covariate effect on median natural mortality (x axis).
+\caption{Estimated probability of fits providing Hessian-based standard errors for EMs assuming alternative process error, that estimate or assume known median natural mortality, and that estimate or assume no covariate effect on median natural mortality when fitted to R OMs and three levels of true covariate effect on median natural mortality (x axis).
 Vertical lines represent 95\% confidence intervals.}\label{convergence_Rom}
@@
 \end{center}
-\caption{Estimated probability of fits providing hessian-based standard errors for EMs assuming alternative process error, that estimate or assume known median natural mortality, and that estimate or assume no covariate effect on median natural mortality when fitted to R+S OMs and three levels of true covariate effect on median natural mortality (x axis).
+\caption{Estimated probability of fits providing Hessian-based standard errors for EMs assuming alternative process error, that estimate or assume known median natural mortality, and that estimate or assume no covariate effect on median natural mortality when fitted to R+S OMs and three levels of true covariate effect on median natural mortality (x axis).
 Vertical lines represent 95\% confidence intervals.}\label{convergence_RSom}
@@
 \end{center}
-\caption{Estimated probability of fits providing hessian-based standard errors for EMs assuming alternative process error, that estimate or assume known median natural mortality, and that estimate or assume no covariate effect on median natural mortality when fitted to R+M OMs and three levels of true covariate effect on median natural mortality (x axis).
+\caption{Estimated probability of fits providing Hessian-based standard errors for EMs assuming alternative process error, that estimate or assume known median natural mortality, and that estimate or assume no covariate effect on median natural mortality when fitted to R+M OMs and three levels of true covariate effect on median natural mortality (x axis).
 Vertical lines represent 95\% confidence intervals.}\label{convergence_RMom}
@@
 
+## AIC results {-}
+
+<!-- When fitting the EMs to R and R+S OMs, AIC was generally accurate in determining the correct source of process errors and, when no effect of the covariate was simulated in the OMs, AIC was also accurate for the correct treatment of the environmental covariate $\beta_E = 0$ (Figs.
+\ref{aic}, \ref{aic_Rom}, \ref{aic_RSom}).
+However, when OMs included an effect of the covariate AIC accuracy for estimating an effect in the EM was only found when the OMs also included lower covariate observation error and higher effect size and higher temporal variability in the covariate.
+On the other hand, for R+M Oms, AIC generally was not able to accurately determine the correct source of process errors, but was able to accurately determine the correct treatment of the environmental covariate (Figs.
+\ref{aic} and \ref{aic_RMom}).
+-->
+
+<!-- When estimating models assumed the mean natural mortality rate was known, the best accuracy of AIC for model selection occurred for models with R+S process errors.
+R+S estimating models ranked best with very low frequency for R or R+M operating models and with very high frequency for R+S operating models (Figure \ref{aic}).
+R estimating models were determined best with high frequency for R operating models, but also for R+M operating models.
+R+M estimating models were rarely determined best for any operating models including those where the process errors matched.
+-->
+
+<!-- AIC was conservative for determining whether the environmental covariate affected natural mortality.
+AIC was highly accurate in determining no effect when there was no effect in the operating model, but AIC ranked the null model best with high frequency even when there was an effect in the operating model in many cases.
+However the accuracy of AIC improved in certain operating models.
+Increased effect size, increased temporal contrast in the covariate, and lower uncertainty in all observations types lead to increased accuracy of determining covariate effects.
+-->
+
+<!-- Relative to the assumption that the mean natural mortality rate was known, estimating the mean natural mortality rate had small effects on the accuracy of AIC in selecting the appropriate process error and whether the covariate affect natural mortality (Figure \ref{aic}).
+Where there were differences there were small decreases in accuracy for determining the appropriate process error and determining a covariate effect when there was one.
+-->
+
 \begin{landscape}
@@
 \end{landscape}
+
+## Covariate effect bias{-}
+
+<!-- When the EMs assumed the median natural mortality rate was known, we observed generally accurate estimation of environmental effects across all EM and OM process error assumptions and all true covariate effect sizes, when ther was low uncertainty in environmental observations and larger temporal contrast in the simulated true environmental covariate (Figure \ref{beta_E_bias}).
+We observed a negative trend in bias of the environmental effect with increased effect size when temporal variation in the covariate was lower and/or uncertainty in the covariate observations was higher.
+When the OMs had R+S process errors with low temporal variation in the true environmental covariate and lower uncertainty in the indices age age composition, estimated covariate effects were highly variable.
+In most cases the relative error of $\beta_E$ did not depend on the source of process error assumed in the EM.
+When there was an effect of the EM process error assumption it was when OMs had R+S process errors.
+The worst bias was observed when OMs assumed R+S process errors, high uncertainty in covariate observations, low variability in the covariate, and low uncertainty in index and age composition observations.
+-->
+
+<!-- When the median natural mortality rate was estimated in the EM, results were similar except estimated effects were even more variable for data simulated with R+S process errors (Figure \ref{beta_E_bias}).
+There was also more separation of reliability of the estimation of the effect among EMs with different process error assumptions.
+The separation was most apparent when OMs simulated R+S process errors and larger variability in the environmental covariate wehere EMs with process errors other than R+S showing more bias than when the mean natural mortality rate was assumed known.
+-->
 
@@
 
+## Covariate effect standard error estimation bias {-}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{se_beta_E_bias_main}
+\end{center}
+\caption{Median error (ME) of Hessian-based estimates of standard error for covariate effect on natural mortality $\beta_E$ from fitting EMs with alternative process error assumptions and treatment of median natural mortality ($e^\beta_M$ known or estimated).
+All OMs had low observation error and contrast in fishing mortality.
+True standard error is defined as the mean of the standard error estimates accross converged fits to simulated data sets for a given OM scenario.}\label{se_beta_E_bias}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{se_beta_E_bias_Rom}
+\end{center}
+\caption{For R OMs, median error (ME) of Hessian-based estimates of standard error for covariate effect on natural mortality $\beta_E$ from fitting EMs with alternative process error assumptions and treatment of median natural mortality ($e^\beta_M$ known or estimated).
+True standard error is defined as the mean of the standard error estimates accross converged fits to simulated data sets for a given OM scenario.}\label{se_beta_E_bias_Rom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{se_beta_E_bias_RSom}
+\end{center}
+\caption{For R+S OMs, median error (ME) of Hessian-based estimates of standard error for covariate effect on natural mortality $\beta_E$ from fitting EMs with alternative process error assumptions and treatment of median natural mortality ($e^\beta_M$ known or estimated).
+True standard error is defined as the mean of the standard error estimates accross converged fits to simulated data sets for a given OM scenario.}\label{se_beta_E_bias_RSom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{se_beta_E_bias_RMom}
+\end{center}
+\caption{For R+M OMs, median error (ME) of Hessian-based estimates of standard error for covariate effect on natural mortality $\beta_E$ from fitting EMs with alternative process error assumptions and treatment of median natural mortality ($e^\beta_M$ known or estimated).
+True standard error is defined as the mean of the standard error estimates accross converged fits to simulated data sets for a given OM scenario.}\label{se_beta_E_bias_RMom}
+\end{figure}
+\end{landscape}
+
+## Covariate effect confidence interval coverage {-}
+
 \begin{landscape}
@@
 
+## Covariate effect RMSE {-}
+
 \begin{landscape}
@@
 
+## Covariate effect estimate and standard error example  {-}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{om_69_em_5_beta_E_se_beta_E_lm_plot}
+\end{center}
+\caption{Positive correlation of covariate effect estimates and Hessian-based standard error estimates for EM that also estimates the median natural mortality parameter and has correct R+M process error assumption fitted to simulated data from the OM with R+M process errors, temporal contrast in fishing pressure, low observation uncertainty for both population ($Low OE$) and covariate observations ($\sigma_e = 0.1$), high and uncorrelated temporal variability in the true covariate ($\sigma_E = 0.5$ and $\rho_E = 0$), and the strongest covariate effect on natural mortality ($\beta_E = 0.5$).}\label{ex_lm_beta_E_SE_beta_E}
+\end{figure}
+\end{landscape}
+
+## Median Natural mortality parameter bias {-}
+
+<!-- We found high accuracy for estimation of the median natural mortality rate parameter ($\beta_M$) for all EM process errors assumptions when OMs had step changes in fishing mortality, lower uncertainty in index and age composition observations, and either R or R+M process errors (Figure \ref{beta_M_bias}).
+The most variation in estimates occurred when fishing mortality was constant and there was higher uncertainty in index and age composition observations.
+For OMs with R+S process errors, the most reliable estimation of $\beta_M$ was obtained when the EM also assumed R+S process errors across all other factors.
+For OMs with R+M process errors, the matching assumption for the EM only showed the best reliability when there was low uncertainty in index and age composition observations and fishing mortality was constant.
+-->
+
 \begin{landscape}
@@
 
+## Median natural mortality parameter standard error estimation bias {-}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{se_beta_M_bias_main}
+\end{center}
+\caption{Median error (ME) of Hessian-based estimates of standard error for median natural mortality parameter $\beta_M$ from fitting EMs with alternative process error assumptions and treatment of the covariate effect ($\beta_ E= 0$ or estimated).
+All OMs had low observation error and contrast in fishing mortality.
+True standard error is defined as the mean of the standard error estimates accross converged fits to simulated data sets for a given OM scenario.}\label{se_beta_M_bias}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{se_beta_M_bias_Rom}
+\end{center}
+\caption{For R OMs, median error (ME) of Hessian-based estimates of standard error for median natural mortality parameter $\beta_M$ from fitting EMs with alternative process error assumptions and treatment of the covariate effect ($\beta_ E= 0$ or estimated).
+True standard error is defined as the mean of the standard error estimates accross converged fits to simulated data sets for a given OM scenario.}\label{se_beta_M_bias_Rom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{se_beta_M_bias_RSom}
+\end{center}
+\caption{For R+S OMs, median error (ME) of Hessian-based estimates of standard error for median natural mortality parameter $\beta_M$ from fitting EMs with alternative process error assumptions and treatment of the covariate effect ($\beta_ E= 0$ or estimated).
+True standard error is defined as the mean of the standard error estimates accross converged fits to simulated data sets for a given OM scenario.}\label{se_beta_M_bias_RSom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{se_beta_M_bias_RMom}
+\end{center}
+\caption{For R+M OMs, median error (ME) of Hessian-based estimates of standard error for median natural mortality parameter $\beta_M$ from fitting EMs with alternative process error assumptions and treatment of the covariate effect ($\beta_ E= 0$ or estimated).
+True standard error is defined as the mean of the standard error estimates accross converged fits to simulated data sets for a given OM scenario.}\label{se_beta_M_bias_RMom}
+\end{figure}
+\end{landscape}
+
+## Median Natural mortality parameter confidence interval coverage {-}
+
 \begin{landscape}
@@
 
+## Median natural mortality parameter estimate and standard error example  {-}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{om_69_em_5_beta_M_se_beta_M_lm_plot}
+\end{center}
+\caption{Negative correlation of $\beta_M$ estimates and Hessian-based standard error estimates for EM that also estimates the covariate effect and has correct R+M process error assumption fitted to simulated data from the OM with R+M process errors, temporal contrast in fishing pressure, low observation uncertainty for both population ($Low OE$) and covariate observations ($\sigma_e = 0.1$), high and uncorrelated temporal variability in the true covariate ($\sigma_E = 0.5$ and $\rho_E = 0$), and the strongest covariate effect on natural mortality ($\beta_E = 0.5$).}\label{ex_lm_beta_M_SE_beta_M}
+\end{figure}
+\end{landscape}
+
+## Median Natural mortality parameter RMSE {-}
+
 \begin{landscape}
@@
 \caption{Root mean square error  (RMSE) of estimates of  $\beta_M$ from fitting EMs with alternative process error assumptions and treatment of covariate effect ($\beta_E = 0$ or estimated).
-All OMs had low observation error and contrast in fishing mortality.}\label{beta_M_rmse}
+All OMs had low observation error for population observations and contrast in fishing mortality.}\label{beta_M_rmse}
 \end{figure}
@@
 
+## Terminal year natural mortality bias {-}
+
+<!-- We present results for error in annual natural mortality rate conditional on three alternative EM configurations for the natural mortality parameters: 1) mean natural mortality rate parameter is fixed at the true value ($\beta_M = log(0.2)$) and no covariate effect is assumed ($\beta_E = 0$),  2) $\beta_M = log(0.2)$ and $\beta_E$ is estimated, and 3) both $\beta_M$ and $\beta_E$ are estimated.
+-->
+
+<!-- When OMs and EMs assume $\beta_M = log(0.2)$ and $\beta_E = 0$, there is no annual variation in natural mortality for OMs with R or R+S process errors simulated or assumed in the EM and, therefore, estimation bias is not possible (Figure \ref{terminal_M_bias}).
+We also observe little or no evidence of bias (confidence intervals include 0) for R or R+S EMs for any OMs even when the OMs included an effect of the covariate on natural mortality ($\beta_E > 0$).
+Including process error on M (R+M) produces more variability in errors for R and R+S EMs than including increased level of effect of the covariate on M.
+For R and R+S OMs, errors in annual M for R+S EMs were less variable than those for R EMs.
+-->
+
+<!-- R+M EMs fit to R OMs with constant fishing mortality, or a step change in fishing mortality and lower uncertainty in index and age composition observations exhibited no bias in annual estimation of M indicating that the variance of the estimated random effects for M in these EMs collapsed to 0.
+R+M EMs were the only EMs to exhibit differences in the sign of the median errors across the time series.
+When OMs had a step change in fishing mortality and higher uncertainty in index and age composition observations, we observed positive median errors at the beginning of the time series and lower median errors in later years.
+However, whether there was evidence of bias, depended on the uncertainty in covariate observations and the degree of variability in the latent covariate.
+We observed negative median relative errors in annual M estimates for R+M EMs fit to R+S OMs with lower uncertainty in index and age composition observations across all levels of simulated effects of the covariate, but evidence of bias was strongest at the beginning of the time series.
+-->
+
+<!-- When $\beta_E$ was estimated, R EMs performed worse for R+S OMs with evidence of negative bias for OMs with constant fishing mortality, lower uncertainty in index and age composition observations, the largest covariate effect size, lower variability in the latent covariate, and larger uncertainty in the covariate observations (Figure \ref{terminal_M_bias}).
+There was little difference in the results for R+S EMs whether $\beta_E$ was estimated or not.
+The results for R+M EMs were generally similar to those when $\beta_E=0$ was assumed, except that the differences between the estimates at the beginning of the time series and later years for certain OM configurations did not occur.
+-->
+
+<!-- Allowing the EMs to estimate $\beta_M$ resulted in very large variability in estimates for all EMs for R and R+M OMs with constant fishing mortality and higher uncertainty in index and age composition observations (Figure \ref{terminal_M_bias}).
+The same variability occurred for R+S OMs, but strong bias was estimated for R and R+M EMs.
+Annual M estimation was most reliable when OMs had step changes in fishing mortality and lower uncertainty in index and age composition observations.
+For R and R+M OMs with those characteristics, all EMs generally provided accurate estimation of annual M, but only R+S EMs provided accurate estimation for R+S OMs.
+Little or no bias in annual M estimation was observed for R+S EMs across all OM process error assumptions as long as there was a step change in fishing mortality and lower uncertainty in indices and age composition.
+-->
+
 \begin{landscape}
@@
 
+## Terminal year natural mortality RMSE {-}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_M_rmse_main}
+\end{center}
+\caption{Root mean square error (RMSE) of estimates of natural mortality rate in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).
+All OMs had low population observation error and contrast in fishing mortality.}\label{terminal_M_rmse}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_M_rmse_Rom}
+\end{center}
+\caption{For R OMs, root mean square error (RMSE) of estimates of natural mortality rate in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).}\label{terminal_M_rmse_Rom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_M_rmse_RSom}
+\end{center}
+\caption{For R+S OMs, root mean square error (RMSE) of estimates of natural mortality rate in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).}\label{terminal_M_rmse_RSom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_M_rmse_RMom}
+\end{center}
+\caption{For R+M OMs, root mean square error (RMSE) of estimates of natural mortality rate in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).}\label{terminal_M_rmse_RMom}
+\end{figure}
+\end{landscape}
+
+## Terminal year spawning stock biomass bias {-}
+
+<!-- Like natural mortality, we present results for error in spawning biomass conditional on three alternative EM configurations for the natural mortality parameters: 1) mean natural mortality rate parameter is fixed at the true value ($\beta_M = log(0.2)$) and no covariate effect is assmed ($\beta_E = 0$),  2) $\beta_M = log(0.2)$ and $\beta_E$ is estimated, and 3) both $\beta_M$ and $\beta_E$ are estimated.
+-->
+
+<!-- When EMs assume $\beta_M = log(0.2)$ and $\beta_E = 0$, there is little evidence of bias for R and R+M OMs except when there is a step change in fishing mortality and higher uncertainty in index and age composition observations, particularly at the end of the time series (Figure \ref{terminal_ssb_bias}).
+We observed a similar trend in median relative error for R+S OMs, but there was more evidence of positive bias at the beginning of the time series whereas the confidence intervals for the negative median errors at the end of the time series often included 0.
+However, there was more indication of positive bias of the incorrect process error assumption of the EMs for the R+S OMs.
+For R+S EMs fit to R+S OMs, there was indication of small positive bias at the beginning of the time series when there was constant fishing mortality and higher uncertainty in index and age composition observations.
+-->
+
+<!-- When EMs assume $\beta_M = log(0.2)$, but estimate $\beta_E$, the median errors for SSB for R and and R+M OMs are similar to those when $\beta_E = 0$ (Figure \ref{terminal_ssb_bias}).
+For R+S OMs, the median errors for EMs that also assume R+S process errors are also similar to those with no covariate effect assumed.
+However, for R and R+M EMs fit to R+S OMs with constant fishing mortality rate and lower uncertainty in indices and age composition observations, we observed evidences of negative bias when higher covariate observation uncertainty and lower variation in the latent covariate was simulated and positive bias for other configurations of the covariate and corresponding observation uncertainty.
+When R+S OMs had constant fishing mortality rate and higher uncertainty in indices and age composition observations or step changes in fishing mortality and lower uncertainty in indices and age composition, R and R+M EMs often provided positively biased SSB estimates when there was lower uncertainty in covariate observations.
+-->
+
+<!-- When EMs estimated both $\beta_M$ and $\beta_E$, we observed large variation in the errors in SSB under the same OM configurations where we observed large variation in errors for annual natural mortality rates (Figure \ref{terminal_ssb_bias}).
+Similarly, we observed little or no bias in SSB estimation for R+S EMs across all OM process error assumptions as long as there was a step change in fishing mortality and lower uncertainty in indices and age composition.
+-->
+
 \begin{landscape}
@@
 
+## Terminal year spawning stock biomass RMSE {-}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_ssb_rmse_main}
+\end{center}
+\caption{Root mean square error (RMSE) of estimates of spawning stock biomass in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).
+All OMs had low population observation error and contrast in fishing mortality.}\label{terminal_ssb_rmse}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_ssb_rmse_Rom}
+\end{center}
+\caption{For R OMs, root mean square error (RMSE) of estimates of spawning stock biomass in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).}\label{terminal_ssb_rmse_Rom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_ssb_rmse_RSom}
+\end{center}
+\caption{For R+S OMs, root mean square error (RMSE) of estimates of spawning stock biomass in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).}\label{terminal_ssb_rmse_RSom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_ssb_rmse_RMom}
+\end{center}
+\caption{For R+M OMs, root mean square error (RMSE) of estimates of spawning stock biomass in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).}\label{terminal_ssb_rmse_RMom}
+\end{figure}
+\end{landscape}
+
+## Terminal year fishing mortality bias {-}
+
 \begin{landscape}
@@
 \caption{Median relative error (MRE) of estimates of fully-selected fishing mortality ($F$) in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).
-All OMs had low observation error and contrast in fishing mortality.}\label{terminal_F_bias}
+All OMs had low population observation error and contrast in fishing mortality.}\label{terminal_F_bias}
 \end{figure}
@@
 \end{landscape}
+
+## Terminal year fishing mortality RMSE {-}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_F_rmse_main}
+\end{center}
+\caption{Root mean square error (RMSE) of estimates of fully-selected fishing mortality ($F$)  in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).
+All OMs had low population observation error and contrast in fishing mortality.}\label{terminal_F_rmse}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_F_rmse_Rom}
+\end{center}
+\caption{For R OMs, root mean square error (RMSE) of estimates of fully-selected fishing mortality ($F$)  in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).}\label{terminal_F_rmse_Rom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_F_rmse_RSom}
+\end{center}
+\caption{For R+S OMs, root mean square error (RMSE) of estimates of fully-selected fishing mortality ($F$)  in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).}\label{terminal_F_rmse_RSom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_F_rmse_RMom}
+\end{center}
+\caption{For R+M OMs, root mean square error (RMSE) of estimates of fully-selected fishing mortality ($F$)  in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).}\label{terminal_F_rmse_RMom}
+\end{figure}
+\end{landscape}

### 2025-06-09 Ecov_study/mortality/manuscript/manuscript.tex
saving work on Ecov-M manuscript

@@
 Legault\textsuperscript{2} \and Brandon Muffley\textsuperscript{5} \and John Wiedenmann\textsuperscript{6}}
-\date{29 May, 2025}
+\date{09 June, 2025}
 
@@
 
-We completed a large scale simulation study with 288 operating models, each with 100 simulated data sets, and 12 estimated models fit to each simulated data set.
-The factors defining operating model configuration included the source of process error on the population (recruitment, recruitment and survival, recruitment and natural mortality), the degree of temporal variation and autocorrelation of the environmental covariate, the uncertainty in the observation of the covariate, the uncertainty in indices and age composition observations, fishing history, and the magnitude of the effect of the covariate on natural mortality.
-The estimating models make alternative assumptions on whether to include the environmental effect, whether the mean/intercept log natural mortality (log(0.2)) is estimated or known, and whether process errors are on just recruitment, recruitment and survival, or recruitment and natural mortality.
-
-We found convergence of all estimation models was generally best when operating models assumed process errors in recruitment and survival, constant fishing rate, greater contrast in the true environmental covariate, and lower uncertainty in corresponding observations.
-Reliable convergence of all estimating models also occurred with the same process errors in the operating model and a step-change in fishing, but this also required lower uncertainty in index and age composition observations.
-Estimating models with process errors on recruitment and survival were unlikely to converge when the process errors in the operating model did not match whereas estimating models with process errors in recruitment and natural mortality converged for operating models without this match in certain cases.
-Probability of convergence generally decreased when the mean/intercept log-natural mortality rate parameter was estimated.
-
-Whether the mean log-natural mortality was estimated or not, the best accuracy of AIC for model selection occurred for models with process errors on recruitment and survival.
-AIC accuracy was poor for models with process errors on recruitment and natural mortality.
-estimating the mean natural mortality rate had small effects on the accuracy of AIC in selecting the appropriate process error.
-Estimating the mean log-natural mortality resulted in a small decrease in AIC accuracy.
-AIC was conservative for determining whether the environmental covariate affected natural mortality.
-AIC was very accurate in determining no effect when there was no effect in the operating model, but AIC often ranked the null model best when there was an effect.
-Accuracy of AIC for covariate effects improved with increased effect size, increased temporal contrast in the covariate, and lower uncertainty in observations.
-
-We found no evidence of bias in estimation of environmental effects regardless of process error assumptions when there was low uncertainty in the environmental observations and large constrast in the environmental covariate.
-In most cases the relative error of the estimated environmental effect did not depend on the source of process error assumed in the estimating model.
-The worst bias was observed when OMs assumed R+S process errors, high uncertainty in covariate observations, low variability in the covariate, and low uncertainty in index and age composition observations.
-Simultaneously estimating the mean/intercept log natural mortality resulted in larger variation in the relative errors of the estimated environmental effect.
-Estimation of the intercept was reliable for all EM process error assumptions when the operating models assumed process errors on recruitment and natural mortality, contrast in fishing pressure over time, and lower observation error.
-Estimating the mean/intercept for log natural mortality generally resulted in highly variable estimates of annual natural mortality and spawning biomass and evidence of bias for some operating and estimation model assumptions about process error source.
-Again reliability of annual natural mortality estimates was generally improved with lower observation error uncertainty and contrast in fishing pressure.
-
-Reliable detection of covariate effects requires informative data.
-AIC preferred simpler models than the true model when information content in data and contrast in covariates and abundance were low.
-The null model for environmental covariate effects (no covariate effect) was selected when contrast in the time series was low and/or uncertainty in observations was high.
-The selection of the null model by AIC also likely decreases with strength of the effect of the covariate on M.
-Similarly, when there was process error in recruitment and natural mortality, estimation models with process error only in recruitment were preferred presumably due to low variation in simulated natural mortality process errors.
-Covariate effect estimation can be robust to process error assumptions with high contrast in covariate and low observation error.
+Treatment of natural mortality is a major consideration in assessment models and state-space approaches allow estimation of temporal variation in this mortality rate as well as effects of specific covariates.
+However, there has been no investigation of the reliability of inferences made regarding natural mortality, associated covariate effects and important assessment output from state-space assessment models exists.
+We conducted a large-scale simulation study that considers models fit to data simulated from operating models with alternative assumptions defined by several factors, but we focus on scenarios where there is temporal contrast in fishing pressure and lower uncertainty in population observations (age composition and indices of abundance).
+The estimating models we fit to simulated observations had alternative assumptions on whether to include the environmental effect, whether the median natural mortality rate was estimated or known, and the source of temporal variability in the population demography.
+Lower uncertainty in covariate observations, higher covariate temporal variability and the source of demographic temporal variability were important for better inferences regarding the covariate effect on natural mortality.
+
+\textbf{keywords:} state-space assessment models, time-varying natural mortality, bias, AIC
 
@@
 State-space stock assessment models, with non-linear functions of latent processes and numerous observation types with different probability distribution assumptions represent one of most complex classes of state-space models.
-The literature on the ways we make inferences and the effects of various factors on reliabilty of inferences from state-space assessment models is growing {[}\citet{lietal24};\citet{milleretal_inreview1};cadiganetal{]}.
+The literature on the effects of various factors on reliabilty of inferences from state-space assessment models is growing \citep{lietal24, milleretal_inreview1}.
 The importance of contrast in population size and fishing mortality and quality of data used to fit assessment models including the state-space variety is known \citep{magnussonhilborn07, milleretal_inreview1}.
-Furthermore, estimation of natural mortality, and even temporal variablity is possible in many scenarios \citep{leeetal11, johnsonetal16, cadigan16, millerhyun18, milleretal_inreview1}.
-
-The effects of temporal variation in recruitment via unspecified or specified environmental factors have been extensively investigated in both traditional assessment models and state space models \citep[haltuchpapers]{myers98, milleretal16}.
-Reliability of estimating environmental and spawning biomass effects on recruitment requires a combination of strong effects, good age composition data quality, contrast in the environmental covariate and lower recruitment variability \citep{brittenetal_inreview, milleretal_inreview1}.
-
-Temporal and environmental effects on growth and weight at age are also more important for short term projections and the reliability of estimation of those effects \citep{correaetal23, correaetal_inreview}.
-
-However, temporal variation in natural mortality and covariate effects on natural mortality are less studied \citet{cadigan16}.
-State-space assessment models currently used can treat the changes in cohort abundance over time as random effects and or despite the importance of natural mortality in inferences for the size of fish populations, their productivity and projections necessary for making catch advice.
-In fact, natural mortality plays a more significant role in short term projections than recruitment due to immediate effects on older age classes that constitute spawning biomass and catch.
-
-See \citet{milleretal_inreview1} for relevance of project 0 results (estimability of natural mortality).
-Some difficulty in distinguishing variation in natural mortality or effects of explicit covariates when variation in M random effects or Ecov is low relative to observation error.
-
-\citet{derisoetal08} formulated the same natural mortality model as a function of covariates and random effects as we use in WHAM.
-
-Mis-specified temporal population process errors could lead to misidentification of stock status and biased population estimates and poor fisheries management decisions \citep{trijouletetal20, legaultpalmer16, szuwalskietal18, croninpunt21, liljestrandetal24}.
-
-In the present study, we conduct a simulation study with operating models (OMs) varying by degree of observation error, source and variability of process error, and fishing history.
-The simulations from these OMs are fitted with estimation models (EMs) that make alternative assumptions for sources of process error, whether a SRR was estimated, and whether natural mortality is estimated.
-Given the confounding nature of process errors, developing diagnostic tools to detect model misspecification is of great scientific interest and could aid the next generation of stock assessments \citep{augeretal21}.
-We evaluate whether convergence and Akaike Information Criterion (AIC) can correctly determine the source of process error and the existence of a SRR.
-We also evaluate when retrospective patterns occur and the degree of bias in the outputs of the assessment model that are important for management.
-
-Estimation of natural mortality is known to be challenging in stock assessment models (\citet{leeetal11}, others).
-However, estimation has been shown to be reliable in some situations \citep{millerhyun18}
-
-Miller et al in review (project 0) found estimation of natural mortality was feasible when data are good and contrast in fishing pressure etc.
-
-Let's focus results here on those situations so that we can reduce the plots in figs.
-
-Variation in natural mortality has an immediate impact on projections unlike recruitment.
-Understanding this varation in M leads to better understanding of post-recruit productivity and therefore management.
-
-Here we conduct a simulation study with OMs varying by degree of observation error uncertainty, sources of process error, fishing history, temporal variation in environmental covariates, and magnitude of the effect of the covariate on natural mortality.
-The simulations from these operating models are fitted with estimating models that make alternative assumptions for sources of process error, and whether (mean) M is estimated.
-We evaluate effects of these factors on convergence of fitted models, whether AIC can correctly determine the correct source of process error and correct assumption about covariate effects on natural mortality, and the degree of bias in relevant parameters and outputs of the assessment model.
+Furthermore, estimation of natural mortality, and even temporal variablity is possible in many scenarios \citep{leeetal11, cadigan16, millerhyun18, milleretal_inreview1}.
+
+The effects of temporal variation in recruitment via undefined or explicit environmental factors have been extensively investigated in both traditional assessment models and state space models \citep{myers98, haltuchpunt11, johnsonetal16, milleretal16}.
+Reliability of estimating environmental and spawning biomass effects on recruitment in state-space assessment models requires a combination of strong effects, good age composition data quality, contrast in the environmental covariate and lower recruitment variability \citep{brittenetal_inreview, milleretal_inreview1}.
+
+A critical aspect of fisheries assessment models and their use in management is short-term projections that are used to determine catch advice.
+While understanding drivers of recruitment is important particularly for subsequent effects on reference points, recruitment in short-term projections typically has little impact on the exploitable biomass in the first few projection years.
+However, assumptions for natural mortality (\(M\)) have immediate and larger effects on projected biomass because they affect the abundances at older age classes at the end of the data time series that constitute spawning biomass and catch \citep{brodziaketal08, stocketal21}.
+
+Because of the effects of \(M\) on both biological reference points and short term projections, better understanding sources of variation in \(M\) would provide more accurate estimation of abundance and productivity and therefore improved management.
+Temporal variation in \(M\) is less studied than recruitment, but its importance for explaining variability in observations has been demonstrated in state-space assessment models for Atlantic cod and yellowtail flounder \citep{cadigan16, stocketal21}.
+\citet{derisoetal08} also demonstrated the importance of several factors affecting Pacific herring natural mortality.
+
+Assessment models could include temporal variation in many aspects of population dynamics or how observations are related to the population.
+For example the Woods Hole Assessment Model (WHAM) can include process errors treated as random effects for transition in cohorts over time, catchability for indices of abundance, selectivity of fishing fleets or indices, movement between regions, or in \(M\) \citep{stockmiller21, milleretal_inreview}.
+However, misspecified temporal population process errors could lead to biased population and stock status estimation, and, therefore, poor fisheries management decisions \citep{legaultpalmer16, szuwalskietal18}.
+Studies of the reliability of inferences we might make regarding the presence of temporal variability in \(M\) are limited.
+\citet{milleretal_inreview1} found AIC could accurately distinguish process errors in the changes in abundance for cohorts, but not for those specifically due to \(M\) except when uncertainty in population observations (indices, catch and age composition) was low and there was greater temporal variation in natural mortality.
+In their simulation studies looking a models with multiple sources of process error, \citet{lietal24} found including more sources of process error than existed in the operating model was a better model-building approach than excluding them a priori.
+
+Here we conduct a simulation study with operating models (OMs) varying by degree of observation error uncertainty, sources of process error, fishing history, temporal variation in environmental covariates, and magnitude of the effect of the covariate on natural mortality.
+The simulated observations from these OMs are fitted with estimating models that make alternative assumptions for sources of process error, and whether median M and covariate effects are estimated.
+We evaluate the effects of these factors on convergence of fitted models, whether Akaike's information criterion (AIC) can correctly determine the correct source of process error and correct assumption about covariate effects on natural mortality, and the degree of bias in inferences for relevant parameters and outputs of the assessment model.
 
@@
 For each simulated data set we fit a set of 12 estimating models (EMs).
-The factors that distinguish the estimating models which are also described in detail below include source of population process error type (3 levels) whether (median) natural mortality rate was estimated or assumed known (2 levels), and whether the effect of the environmental covariate on natural mortality was estimated or not (2 levels).
-
-The sources of population process error that were used in the OMs or assumed in the EMs were on recruitment only (R), recruitment and changes in cohort abundance over time (R+S), or recruitment and natural mortality (R+M).
+The factors that distinguish the estimating models which are also described in detail below include source of population process error type (3 levels) whether (median) \(M\) was estimated or assumed known (2 levels), and whether the effect of the environmental covariate on \(M\) was estimated or not (2 levels).
+
+The sources of population process error that were used in the OMs or assumed in the EMs were on recruitment only (R), recruitment and changes in cohort abundance over time (R+S), or recruitment and \(M\) (R+M).
 We did not use the log-normal bias-correction feature for process errors or observations described by \citet{stockmiller21} for operating and EMs \citep{lietal_inreview}.
@@
 
-Here we consider both covariate effects and process errors for natural mortality and the general model for natural mortality in year \(y\) is a log-linear function of a parameter \(\beta_{M}\) that defines median natural mortality and possibly effects of the environmental covariate \(X_y\) described above and autocorrelated normally distributed annual process error \(\varepsilon_{M,y}\)
+The general model for \(M\) in year \(y\) is a log-linear function of both covariate effects \(X_y\) and process errors \(\varepsilon_{M,y}\) and a parameter \(\beta_{M}\) that defines median \(M\)
 \[
@@
 \]
-\(\epsilon_{M,y}|\epsilon_{M,y-1} \sim \text{N}\left(\epsilon_{M,y-1},(1-\rho_M^2)\sigma_M^2\right)\) \citep{stockmiller21}.
-We assume the median natural mortality rate \(e^{\beta_M} = 0.2\) is constant across ages.
-For R and R+S OMs and EMs, \(\epsilon_{M,y} = 0\).
+where the process errors are modeled as random effects that may, in general, be autocorrelated normal random variables
+\[
+\varepsilon_{M,y}|\varepsilon_{M,y-1} \sim \text{N}\left(\varepsilon_{M,y-1},(1-\rho_M^2)\sigma_M^2\right)
+\]
+\citep{stockmiller21}, but we assume \(\rho_M = 0\) in our R+M OMs.
+We assume the median \(M\) rate \(e^{\beta_M} = 0.2\) is constant across ages.
+For R and R+S OMs and EMs, \(\varepsilon_{M,y} = 0\).
 For all R+M OMs, we assume the same standard deviation \(\sigma_M = 0.3\) and is estimated in the R+M EMs.
-The covariate effect is one of 3 alternative values in the operating models (\(\beta_\text{E} \in \{0,0.25,0.5\}\).
-The parameters defining the simulated covariate time series, size of the covariate effect, and any natural mortality random effects result in a range of different levels of variation in annual natural mortality rates (Figure \ref{M_example}).
-
-We assumed expected recruitment each year from a Beverton-Holt stock-recruit relationship (SRR)
+The covariate effect is one of 3 alternative values in the operating models, \(\beta_\text{E} \in \{0,0.25,0.5\}\).
+The parameters defining the simulated covariate time series, size of the covariate effect, and any \(M\) random effects result in a range of different levels of variation in annual values (Figure \ref{M_example}).
+
+We assumed expected recruitment each year is from a Beverton-Holt stock-recruit relationship (SRR)
 \[
@@
 \]
-All biological inputs to calculations of spawning biomass per recruit (i.e., weight, maturity, and natural mortality at age) are constant in the R and R+S OMs without covariate effects on natural mortality.
+All biological inputs to calculations of spawning biomass per recruit (i.e., weight, maturity, and \(M\) at age) are constant in the R and R+S OMs without covariate effects on \(M\).
 Therefore, steepness and equilibrium unfished recruitment are also constant over the time period for those OMs \citep{millerbrooks21}.
-As in \citet{milleretal_inreview1}, our assumed biological inputs and selectivity (defined below) with constant natural mortality result in equilibrium fishing mortality that reduces spawning biomass per recruit to 40\% of the unfished level is \(\Fspr[40] = 0.348\).
-With an assumed unfished recruitment of \(R_0 = e^{10}\), setting \(\Fmsy = \Fspr[40]\) results in a steepness of 0.69 and \(a=0.60\) and \(b = 2.4 \times 10^{-5}\).
-
-For R+M OMs and all OMs with covariate effects on natural mortality, steepness is not constant, but we used the same \(a\) and \(b\) parameters as other operating models which equates to a steepness and \(R_0\) at the median of the time series process for M.
-For operating models with time-varying random effects for fishery selectivity, \Fmsy is also not constant however we use the same fishing history as other operating models which corresponds to \Fmsy at the mean selectivity parameters.
+As in \citet{milleretal_inreview1}, our assumed biological inputs and selectivity (defined below) with constant \(M\) result in equilibrium fishing mortality that reduces spawning biomass per recruit to 40\% of the unfished level is \(F_{40\%} = 0.348\).
+With an assumed unfished recruitment of \(R_0 = e^{10}\), setting \(\Fmsy = F_{40\%}\) results in a steepness of 0.69 and \(a=0.60\) and \(b = 2.4 \times 10^{-5}\).
+For R+M OMs and all OMs with covariate effects on \(M\), steepness is not constant, but we used the same \(a\) and \(b\) parameters as other operating models which equates to a steepness and \(R_0\) at the median of the time series processes for \(M\) and the covariate.
 
@@
 For each data set simulated from an operating model 12 estimating models were fit.
-There were three factors defining the configuration of each estimating model: 1) whether the mean natural mortality \(\beta_M\) was estimated or assumed known, 2) whether an environmental effect \(\beta_E\) was estimated or not, and 3) whether the process errors were assumed on recruitment only (R), recruitment and survival (R+S), or recruitment and natural mortality (R+M).
+There were three factors defining the configuration of each estimating model: 1) whether \(\beta_M\) was estimated or assumed known, 2) whether an environmental effect \(\beta_E\) was estimated or not, and 3) whether the process errors were assumed on recruitment only (R), recruitment and survival (R+S), or recruitment and \(M\) (R+M).
 
@@
 All fixed effects parameters for selectivity, catchability, fully-selected fishing mortality, mean recruitment, initial abundance at age, and variances for logistic-normal age composition distributions were estimated.
-Any process error variance parameters for recruitment, survival, and natural mortality were also estimated.
+Any process error variance parameters for recruitment, survival, and \(M\) were also estimated.
 The observation error variance of the environmental observations and aggregate catch and indices were all assumed known at the true values.
@@
 
-The first measure of reliability we investigated was frequency of convergence when fitting each estimating model to the simulated data sets.
-There are various ways to assess convergence of the fit, but we defined successful convergence as the hessian of the marginal log-likelihood being invertible and providing variance estimates for the fixed effects parameters.
+We measured the frequency of convergence when fitting each EM to the simulated data sets for each OM.
+There are various ways to assess convergence of the fit \citep[e.g.,][]{carvalhoetal21}, but we defined successful convergence as the Hessian of the marginal log-likelihood being invertible and providing variance estimates for the fixed effects parameters as recommended by \citet{milleretal_inreview1}.
+We calculated 95\% confidence intervals for probability of convergence using the Clopper-Pearson exact method \citep{clopperpearson34, thulin14}.
 
@@
 We measured the frequency of correct model selection using marginal AIC.
-For a given operating model the set of models that were considered all made the same assumptions on whether or not to estimate (mean) natural mortality rate or it is assumed at the true value \(\beta_M = log(0.2)\).
+For a given operating model the set of models that were considered all made the same assumptions on whether or not to estimate \(\beta_M\) or it is assumed at the true value.
 For model \(m\), the marginal AIC is a function of the marginal log-likelihood maximized with respect to the fixed effects in the model \(\boldsymbol{\theta}\) and the number of fixed effects \(n\left(\boldsymbol{\theta}\right)\) estimated,
@@
 We used all of these fits because some lack of convergence would be expected for the correct behavior of more complicated models that include process errors that did not exist in the operating model.
-For example R+M EMs fit to R OMs would be expected to estimate no variance in the natural mortality random effects and the estimated variance parameter going to zero would cause poor convergence.
-
-\hypertarget{parameter-inference-reliability}{%
-\subsubsection*{Parameter inference reliability}\label{parameter-inference-reliability}}
-\addcontentsline{toc}{subsubsection}{Parameter inference reliability}
+For example R+M EMs fit to R OMs would be expected to estimate no variance in the \(M\) random effects and the estimated variance parameter going to zero would cause poor convergence.
+
+\hypertarget{parameter-estimation}{%
+\subsubsection*{Parameter estimation}\label{parameter-estimation}}
+\addcontentsline{toc}{subsubsection}{Parameter estimation}
 
@@
 We used this conditioning, to reflect how practitioners would proceed in analyses of model fits with real assessment data.
-That is, practitioners would ensure models converged such that hessian-based standard errors were available for all model parameter estimates.
-
-We calculated median errors (ME) of
-
-\begin{itemize}
-\item $\beta_E$, the effect of environmental covariates on natural mortality, 
-\item $\beta_M$, the mean log-natural mortality rate,
-\end{itemize}
-
-and the median relative errors (MRE) of terminal year
-
-\begin{itemize}
-\item natural mortality rate ,
-\item spawning stock biomass, and
-\item fully-selected fishing mortality rate.
-\end{itemize}
-
-In preliminary analyses we inspected MRE of these parameters over the whole time series, but we found no appreciable differences in patterns across the various factors defining the OMs and EMs.
-
-We also calculated the root mean square error (RMSE) and estimated probability of coverage of constructed 95\% confidence intervals for \(\beta_E\) and \(\beta_M\) for EMs that estimated these parameters.
-
-Results for fishing mortality rate are provided in the Supplementary Materials.
-For the \(i\)th simulated data the relative error for a parameter \(\theta\) provided from the fitted estimation model is
+That is, practitioners would ensure models converged such that Hessian-based standard errors were available for all model parameter estimates.
+
+We focused on statistical behavior of estimators of the covariate effect on \(M\) (\(\widehat \beta_E\)), the estimator of the median \(M\) parameter (\(\widehat \beta_M\)), and estimators of terminal year natural mortality \(\widehat M\), spawning stock biomass \(\widehat{\text{SSB}}\), and fully-selected fishing mortality rate \(\widehat F\).
+In preliminary analyses we results of the estimators of all the annual values for \(M\), SSB and fishing mortality over the whole time series, but we found no appreciable differences in patterns across the various factors defining the OMs and EMs.
+Furthermore, results for terminal year fishing mortality were generally inversely related to those for spawning stock biomass, and are provided in the Supplementary Materials and not discussed further.
+
+We calculated median errors (ME) of \(\widehat \beta_E\), \(\widehat \beta_M\)), and the Hessian-based standard error estimators of these parameters (\(\widehat{SE}\left(\widehat \beta_E\right)\) and \(\widehat{SE}\left(\widehat \beta_M\right)\)).
+We calculated the median relative errors (MRE) of terminal year \(\widehat M\), \(\widehat{\text{SSB}}\), and \(\widehat F\).
+We also calculated the root mean square error (RMSE) and estimated probability of coverage of constructed 95\% confidence intervals for \(\widehat \beta_E\) and \(\widehat \beta_M\) for EMs that estimated these parameters.
+We constructed the confidence intervals for probabilities of CI coverage using the same methods as those for probabilities of convergence.
+
+The true values for terminal year SSB and fishing mortality rate vary among simulations.
+For the \(i\)th simulated data the relative error for terminal year value \(\theta_i\) provided from the fitted estimation model is
 \[
-{\text{RE}_i}\left(\theta\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
+\text{RE}_i\left(\theta\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
 \]
-and constructed 95\% confidence intervals using the binomial distribution approach as in \citet{stockmiller21} and \citet{milleretal_inreview1}.
+We calculated RMSE as
+\[
+\text{RMSE}\left(\widehat \theta\right) = \sqrt{\frac{1}{n}\sum^n\left(\widehat \theta_i - \theta_i\right)^2}
+\]
+For ME and RME of estimators, we constructed 95\% confidence intervals using the binomial distribution approach as in \citet{stockmiller21} and \citet{milleretal_inreview1}.
 Because of the inverse relationship of the the confidence interval and the null hypothesis significance test, when the confidence interval contains zero we do not have evidence of bias from the simulation study.
-
-For natural mortality rate, SSB, and fully-selected fishing mortality rate, we summarized results for each of the annual values, but we present results only for estimates from the terminal year, because there were no appreciable differences between the results among all years.
 
@@
 
-Because \citet{milleretal_inreview1} found inferences are most reliable for scenarios with low observation error for indices and age composition data and with temporal contrast in fishing pressure, we restrict our attention to scenarios with these characteristics in the main paper (temporal contrast in fishing and low observation error), but include corresponding results for the other scenarios in the Supplementary Materials.
+\citet{milleretal_inreview1} found inferences are most reliable for distinguishing process error sources and stock-recruit relationships in scenarios with lower observation error for indices and age composition data and with temporal contrast in fishing pressure.
+Our expectation is that inferences for covariate effects on \(M\) would require data to be at least that informative and inspection of the comprehensive set of results across all OMs generally confirms this (See supplemental Materials).
+Therefore, we restrict our attention to these OM scenarios with more informative data in the main paper (temporal contrast in fishing and lower population observation error).
 
@@
 
-When the median natural mortality rate was assumed known, corresponding to usual practice in application of assessment models, good convergence for all EMs was observed for R+S operating models with \Fmsy fishing history, more variation in the latent environmental covariate (\(\sigma_E = 0.5\)), and lower error in the associated observations \(\sigma_e = 0.1\) (Figure \ref{convergence}).
-Good convergence for all EMs was also observed for R+S OMs with the step change in fishing history, but with low error in indices and age comp observations.
-
-EMs that assumed random effects just on recruitment (R), nearly always converged across all operating models that assumed more variation in the latent environmental covariate (\(\sigma_E = 0.5\)) and lower error in the associated observations \(\sigma_E = 0.1\).
-EMs that assumed recruitment and survival random effects (R+S), had poor converge probability when the OMs had alternative process errors (R or R+M).
-The R+S EMs showed best convergence for the R+S OMs where there was lower error in the environmental observations \(\sigma_E = 0.1\) or higher error in environmental observations of latent true environmental covariates with greater temporal variation.
-EMs that assumed recruitment and natural mortality random effects (R+M), had poor convergence probability for all OMs that had a step change in fishing history and higher uncertainty in indices and age composition.
-
-When the OMs and EMs both assume process errors only on recruitment (R) or on recruitment and survival (R+S), convergence was worst with less variation in the true environmental covariate and larger uncertainty in associated observations.
-When the OMs and EMs both assume process errors on recruitment and natural mortality (R+M) convergence was problematic for all OMs with the step-change in fishing history.
-The best convergence was observed with this match between OMs and EMs was when the fishing history was constant and there was low uncertainty in environmental observations.
-
-As might be expected, there was an overall drop in the probability of convergence when the mean natural mortality rate was estimated rather than assumed at the true value (Figure \ref{convergence}).
-Otherwise, general trends described above with mean natural mortality fixed, apply when estimated.
+In OMs with contrast in fishing pressure and lower uncertainty in population observations (catch, indices and age composition), R EMs generally converged with high frequency except when EMs estimated covariate effects on \(M\) and the covariate uncertainty (\(\sigma_e\)) was high (Figure \ref{convergence}).
+R+S EMs generally converged with low frequency when the OM process error did not match the EM configuration.
+R+M EMs generally did not converge with high frequency except for R+S OMs with low covariate uncertainty and where the EMs assumed median \(M\) as known.
+There was no strong effect of the treatment of median M (known or estimated) on convergence for any of the other EMs.
+The largest effect of the size of the true covariate effect (\(\beta_E\)) on convergence was observed for R and R+M EMs that estimated covariate effects in R and R+M OMs with high uncertainty in covariate observations (\(\sigma_e\)) and high variability in covariate process errors (\(\sigma_E\)).
+There was an increase in convergence frequency with increased variability in covariate process errors for many EMs in OMs with large covariate uncertainty.
 
@@
 
-We only present results for EMs where the median natural mortality rate parameter was estimated.
-Results differed little between results where this parameter was estimated or assumed known, but all of these results can be found in the Supplementary materials.
-
-When fitting the EMs to R and R+S OMs, AIC was generally accurate in determining the correct source of process errors and, when no effect of the covariate was simulated in the OMs, AIC was also accurate for the correct treatment of the environmental covariate \(\beta_E = 0\) (Figs.
-\ref{aic}, \ref{aic_Rom}, \ref{aic_RSom}).
-However, when OMs included an effect of the covariate AIC accuracy for estimating an effect in the EM was only found when the OMs also included lower covariate observation error and higher effect size and higher temporal variability in the covariate.
-On the other hand, for R+M Oms, AIC generally was not able to accurately determine the correct source of process errors, but was able to accurately determine the correct treatment of the environmental covariate (Figs.
-\ref{aic} and \ref{aic_RMom}).
-
-When estimating models assumed the mean natural mortality rate was known, the best accuracy of AIC for model selection occurred for models with R+S process errors.
-R+S estimating models ranked best with very low frequency for R or R+M operating models and with very high frequency for R+S operating models (Figure \ref{aic}).
-R estimating models were determined best with high frequency for R operating models, but also for R+M operating models.
-R+M estimating models were rarely determined best for any operating models including those where the process errors matched.
-
-AIC was conservative for determining whether the environmental covariate affected natural mortality.
-AIC was highly accurate in determining no effect when there was no effect in the operating model, but AIC ranked the null model best with high frequency even when there was an effect in the operating model in many cases.
-However the accuracy of AIC improved in certain operating models.
-Increased effect size, increased temporal contrast in the covariate, and lower uncertainty in all observations types lead to increased accuracy of determining covariate effects.
-
-Relative to the assumption that the mean natural mortality rate was known, estimating the mean natural mortality rate had small effects on the accuracy of AIC in selecting the appropriate process error and whether the covariate affect natural mortality (Figure \ref{aic}).
-Where there were differences there were small decreases in accuracy for determining the appropriate process error and determining a covariate effect when there was one.
-
-\hypertarget{bias}{%
-\subsection*{Bias}\label{bias}}
-\addcontentsline{toc}{subsection}{Bias}
-
-\hypertarget{environmental-effect}{%
-\subsubsection*{Environmental effect}\label{environmental-effect}}
-\addcontentsline{toc}{subsubsection}{Environmental effect}
-
-When the EMs assumed the median natural mortality rate was known, we observed generally accurate estimation of environmental effects across all EM and OM process error assumptions and all true covariate effect sizes, when ther was low uncertainty in environmental observations and larger temporal contrast in the simulated true environmental covariate (Figure \ref{beta_E_bias}).
-We observed a negative trend in bias of the environmental effect with increased effect size when temporal variation in the covariate was lower and/or uncertainty in the covariate observations was higher.
-When the OMs had R+S process errors with low temporal variation in the true environmental covariate and lower uncertainty in the indices age age composition, estimated covariate effects were highly variable.
-In most cases the relative error of \(\beta_E\) did not depend on the source of process error assumed in the EM.
-When there was an effect of the EM process error assumption it was when OMs had R+S process errors.
-The worst bias was observed when OMs assumed R+S process errors, high uncertainty in covariate observations, low variability in the covariate, and low uncertainty in index and age composition observations.
-
-When the median natural mortality rate was estimated in the EM, results were similar except estimated effects were even more variable for data simulated with R+S process errors (Figure \ref{beta_E_bias}).
-There was also more separation of reliability of the estimation of the effect among EMs with different process error assumptions.
-The separation was most apparent when OMs simulated R+S process errors and larger variability in the environmental covariate wehere EMs with process errors other than R+S showing more bias than when the mean natural mortality rate was assumed known.
-
-\hypertarget{median-natural-mortality-rate}{%
-\subsubsection*{Median natural mortality rate}\label{median-natural-mortality-rate}}
-\addcontentsline{toc}{subsubsection}{Median natural mortality rate}
-
-We found high accuracy for estimation of the median natural mortality rate parameter (\(\beta_M\)) for all EM process errors assumptions when OMs had step changes in fishing mortality, lower uncertainty in index and age composition observations, and either R or R+M process errors (Figure \ref{beta_M_bias}).
-The most variation in estimates occurred when fishing mortality was constant and there was higher uncertainty in index and age composition observations.
-For OMs with R+S process errors, the most reliable estimation of \(\beta_M\) was obtained when the EM also assumed R+S process errors across all other factors.
-For OMs with R+M process errors, the matching assumption for the EM only showed the best reliability when there was low uncertainty in index and age composition observations and fishing mortality was constant.
-
-\hypertarget{annual-natural-morality-rate}{%
-\subsubsection*{Annual natural morality rate}\label{annual-natural-morality-rate}}
-\addcontentsline{toc}{subsubsection}{Annual natural morality rate}
-
-We present results for error in annual natural mortality rate conditional on three alternative EM configurations for the natural mortality parameters: 1) mean natural mortality rate parameter is fixed at the true value (\(\beta_M = log(0.2)\)) and no covariate effect is assumed (\(\beta_E = 0\)), 2) \(\beta_M = log(0.2)\) and \(\beta_E\) is estimated, and 3) both \(\beta_M\) and \(\beta_E\) are estimated.
-
-When OMs and EMs assume \(\beta_M = log(0.2)\) and \(\beta_E = 0\), there is no annual variation in natural mortality for OMs with R or R+S process errors simulated or assumed in the EM and, therefore, estimation bias is not possible (Figure \ref{terminal_M_bias}).
-We also observe little or no evidence of bias (confidence intervals include 0) for R or R+S EMs for any OMs even when the OMs included an effect of the covariate on natural mortality (\(\beta_E > 0\)).
-Including process error on M (R+M) produces more variability in errors for R and R+S EMs than including increased level of effect of the covariate on M.
-For R and R+S OMs, errors in annual M for R+S EMs were less variable than those for R EMs.
-
-R+M EMs fit to R OMs with constant fishing mortality, or a step change in fishing mortality and lower uncertainty in index and age composition observations exhibited no bias in annual estimation of M indicating that the variance of the estimated random effects for M in these EMs collapsed to 0.
-R+M EMs were the only EMs to exhibit differences in the sign of the median errors across the time series.
-When OMs had a step change in fishing mortality and higher uncertainty in index and age composition observations, we observed positive median errors at the beginning of the time series and lower median errors in later years.
-However, whether there was evidence of bias, depended on the uncertainty in covariate observations and the degree of variability in the latent covariate.
-We observed negative median relative errors in annual M estimates for R+M EMs fit to R+S OMs with lower uncertainty in index and age composition observations across all levels of simulated effects of the covariate, but evidence of bias was strongest at the beginning of the time series.
-
-When \(\beta_E\) was estimated, R EMs performed worse for R+S OMs with evidence of negative bias for OMs with constant fishing mortality, lower uncertainty in index and age composition observations, the largest covariate effect size, lower variability in the latent covariate, and larger uncertainty in the covariate observations (Figure \ref{terminal_M_bias}).
-There was little difference in the results for R+S EMs whether \(\beta_E\) was estimated or not.
-The results for R+M EMs were generally similar to those when \(\beta_E=0\) was assumed, except that the differences between the estimates at the beginning of the time series and later years for certain OM configurations did not occur.
-
-Allowing the EMs to estimate \(\beta_M\) resulted in very large variability in estimates for all EMs for R and R+M OMs with constant fishing mortality and higher uncertainty in index and age composition observations (Figure \ref{terminal_M_bias}).
-The same variability occurred for R+S OMs, but strong bias was estimated for R and R+M EMs.
-Annual M estimation was most reliable when OMs had step changes in fishing mortality and lower uncertainty in index and age composition observations.
-For R and R+M OMs with those characteristics, all EMs generally provided accurate estimation of annual M, but only R+S EMs provided accurate estimation for R+S OMs.
-Little or no bias in annual M estimation was observed for R+S EMs across all OM process error assumptions as long as there was a step change in fishing mortality and lower uncertainty in indices and age composition.
-
-\hypertarget{spawning-stock-biomass}{%
-\subsubsection*{Spawning stock biomass}\label{spawning-stock-biomass}}
-\addcontentsline{toc}{subsubsection}{Spawning stock biomass}
-
-Like natural mortality, we present results for error in spawning biomass conditional on three alternative EM configurations for the natural mortality parameters: 1) mean natural mortality rate parameter is fixed at the true value (\(\beta_M = log(0.2)\)) and no covariate effect is assmed (\(\beta_E = 0\)), 2) \(\beta_M = log(0.2)\) and \(\beta_E\) is estimated, and 3) both \(\beta_M\) and \(\beta_E\) are estimated.
-
-When EMs assume \(\beta_M = log(0.2)\) and \(\beta_E = 0\), there is little evidence of bias for R and R+M OMs except when there is a step change in fishing mortality and higher uncertainty in index and age composition observations, particularly at the end of the time series (Figure \ref{terminal_ssb_bias}).
-We observed a similar trend in median relative error for R+S OMs, but there was more evidence of positive bias at the beginning of the time series whereas the confidence intervals for the negative median errors at the end of the time series often included 0.
-However, there was more indication of positive bias of the incorrect process error assumption of the EMs for the R+S OMs.
-For R+S EMs fit to R+S OMs, there was indication of small positive bias at the beginning of the time series when there was constant fishing mortality and higher uncertainty in index and age composition observations.
-
-When EMs assume \(\beta_M = log(0.2)\), but estimate \(\beta_E\), the median errors for SSB for R and and R+M OMs are similar to those when \(\beta_E = 0\) (Figure \ref{terminal_ssb_bias}).
-For R+S OMs, the median errors for EMs that also assume R+S process errors are also similar to those with no covariate effect assumed.
-However, for R and R+M EMs fit to R+S OMs with constant fishing mortality rate and lower uncertainty in indices and age composition observations, we observed evidences of negative bias when higher covariate observation uncertainty and lower variation in the latent covariate was simulated and positive bias for other configurations of the covariate and corresponding observation uncertainty.
-When R+S OMs had constant fishing mortality rate and higher uncertainty in indices and age composition observations or step changes in fishing mortality and lower uncertainty in indices and age composition, R and R+M EMs often provided positively biased SSB estimates when there was lower uncertainty in covariate observations.
-
-When EMs estimated both \(\beta_M\) and \(\beta_E\), we observed large variation in the errors in SSB under the same OM configurations where we observed large variation in errors for annual natural mortality rates (Figure \ref{terminal_ssb_bias}).
-Similarly, we observed little or no bias in SSB estimation for R+S EMs across all OM process error assumptions as long as there was a step change in fishing mortality and lower uncertainty in indices and age composition.
+We only present results for EMs where the median \(M\) rate parameter was estimated because results differed little whether this parameter was estimated or assumed known (see Figures \ref{aic_Rom} to \ref{aic_RMom}).
+When there is temporal contrast in fishing pressure and lower uncertainty in population observations, the only OM scenarios where AIC was consistently accurate for both covariate effect and process error configurations were R OMs with high variability in the covariate and low uncertainty in observations of the covariate (Figure \ref{aic}).
+R+M OMs with those same covariate attributes were only accurate for the covariate effect (low Type I and II errors) and the incorrect process error almost always chosen was R rather than R+S.
+We also observed high accuracy for both covariate effect and process error configurations for R+S OMs where either no effect of the covariate or the strongest effect of the covariate was simulated.
+All R and R+S OMs were accurate for process error configuration and across all OMs, AIC accurately selects models without covariate effects when there are none (low Type I error).
+See Figures \ref{aic_Rom} to \ref{aic_RMom} for full results.
+
+\hypertarget{covariate-effect-estimation}{%
+\subsection*{Covariate effect estimation}\label{covariate-effect-estimation}}
+\addcontentsline{toc}{subsection}{Covariate effect estimation}
+
+When there is temporal contrast in fishing pressure, lower uncertainty in population observations, low uncertainty in environmental observations, and larger temporal contrast in the simulated true environmental covariate, we observed generally little or no bias in estimation of covariate effects (\(\widehat \beta_E\)) across all EM and OM process error assumptions, both EM configurations of median \(M\), and all true covariate effect sizes (Figure \ref{beta_E_bias}, rows 3 and 4).
+The only exception to this was when R EMs with median \(M\) estimated were fit to R+S OMs with those configurations.
+Decreasing trends in bias of \(\widehat \beta_E\) for EMs fit to many configurations of R and R+M OMs indicate that \(\widehat \beta_E\) values were closer to zero on average even when the true effect increased.
+For R+S OMs, EMs with the matching process error configuration showed little evidence of bias in estimating \(\beta_E\) across a range of covariate and covariate observation configurations.
+
+Bias of Hessian-based standard error estimation for \(\widehat \beta_E\) was also close to zero in the same OM scenarios where bias in estimation of \(\beta_E\) itself was close to zero (Figure \ref{se_beta_E_bias}, rows 3 and 4).
+However, in these high information OMs bias of standard error estimation was observed in for several EMs that did not have the matching R+S and R+M process error configuration.
+Estimation of standard errors was essentially unreliable for most EMs when OMs had higher uncertainty in covariate observations (Figure \ref{se_beta_E_bias}, rows 5 to 8).
+
+Despite the reliable estimation of covariate effects and standard errors with low uncertainty in environmental observations, and larger temporal contrast in the simulated true environmental covariate, we found poor confidence interval coverage for many EMs fitted to R+S and R+M OMs (Figure \ref{beta_E_CI_coverage}, rows 3 and 4).
+For example, R+M EMs fit to R+M OMs showed little bias in estimation of the covariate effect and standard errors, but confidence interval coverage was negatively biased (rows 3 and 4 in right two columns of Figures \ref{beta_E_bias}, \ref{se_beta_E_bias}, and \ref{beta_E_CI_coverage}.
+However, there appears to be a positive correlation of the covariate and standard error estimates such that estimates less than the true value had negatively biased standard error estimates which would make confidence intervals too small (Figure \ref{ex_lm_beta_E_SE_beta_E}).
+Confidence interval coverage was most reliable for R+S OMs when the EMs had the matching process error configuration (Figure \ref{beta_E_CI_coverage}, middle 2 columns).
+
+\hypertarget{median-natural-mortality-rate-estimation}{%
+\subsection*{Median natural mortality rate estimation}\label{median-natural-mortality-rate-estimation}}
+\addcontentsline{toc}{subsection}{Median natural mortality rate estimation}
+
+We observed negligible bias in estimating the median \(M\) parameter (\(\beta_M\)) over a much wider set of operating models than that for estimating the covariate effect on \(M\).
+When there is temporal contrast in fishing pressure, lower uncertainty in population observations, all EMs fit to R and R+M OMs showed little evidence of bias for \(\widehat \beta_M\) (Figure \ref{beta_M_bias}, left and right sets of columns).
+For R+S OMs with those conditions, EMs with the matching process error assumption showed little evidence of bias.
+
+We also found little or no bias of Hessian-based standard error estimation for \(\beta_M\) in the many of the same OM scenarios where bias of \(\widehat \beta_M\) was close to zero except for some EMs with the process error source mis-specified or where the OM simulated the strongest covariate effect (Figure \ref{se_beta_M_bias}).
+Like \(\widehat \beta_M\), reliability of standard error estimation in R+S OMs required the EMs to have the matching process error assumption.
+
+However, we found confidence interval coverage to be generally unreliable even for OM and EM combinations where bias in \(\widehat \beta_M\) and its standard error estimators was negligible, much like that for \(\widehat \beta_E\) (Figure \ref{beta_M_CI_coverage}).
+In the same EM OM combination we investigated above for \(\widehat \beta_E\), we observed an opposite negative correlation of \(\widehat \beta_M\) and its standard error estimates (Figure \ref{ex_lm_beta_M_SE_beta_M}), which would result in confidence intervals being too narrow when \(\widehat \beta_M\) values are larger than average.
+
+\hypertarget{terminal-year-natural-morality-rate}{%
+\subsection*{Terminal year natural morality rate}\label{terminal-year-natural-morality-rate}}
+\addcontentsline{toc}{subsection}{Terminal year natural morality rate}
+
+The EMs with R and R+S process error configurations with median M assumed known and covariate effects not estimated will have terminal year M correctly specified when fitted to R and R+S OMs with no covariate effect because \(M\) is constant in both the EM and OM at the same value.
+We exclude those OM-EM combinations from our results here.
+
+In many of the OMs with temporal contrast in fishing pressure and lower uncertainty in population observations, there was little difference in bias for terminal year M among EMs whether the median \(M\) parameter (\(\beta_M\)) was estimated (Figure \ref{terminal_M_bias}).
+However, when there were differences, bias in terminal year M was closer to zero when the EM assumed \(\beta_M\) known, As would be expected.
+For R OMs, all estimating models exhibited little evidence of bias except when OMs simulated the largest covariate effect (\(\beta_E = 0.5\)).
+However, bias estimates were at worst around 10\% (Figure \ref{terminal_M_bias}, left columns).
+For R+S OMs, the R+S EMs generally exhibited the least evidence for bias and R+M EMs generally estimated terminal M lower than the true value, particularly when observation uncertainty in the covariate was lower.
+For R+M OMs, many of the EM configurations provided little bias in terminal \(\widehat M\), but the biases were generally further from zero than the results for the R OMs (Figure \ref{terminal_M_bias}, right columns).
+Accuracy of terminal year \(\widehat M\) was similar for all of the EMs when fitted to R+M OMs (Figure \ref{terminal_M_rmse}).
+For R+S OMs, accuracy was a bit better for R+S EM than others.
+As would be expected accuracy of estimate where median \(M\) was assumed known was better than when it was estimated, when there were differences.
+
+\hypertarget{terminal-year-spawning-stock-biomass}{%
+\subsection*{Terminal year spawning stock biomass}\label{terminal-year-spawning-stock-biomass}}
+\addcontentsline{toc}{subsection}{Terminal year spawning stock biomass}
+
+In OMs with temporal contrast in fishing pressure and lower uncertainty in population observations, we found little bias in estimates of terminal year SSB for all EM configurations when fitted to R or R+M OMs (Figure \ref{terminal_ssb_bias}).
+We also found little bias for R+S OMs when the EMs assumed the correct process error configuration.
+The worst bias in terminal year SSB estimation we found was in R+S OMs for R EMs with \(\beta_M\) estimated.
+Like the bias results, accuracy as measured by RMSE was similar for many of EMs when fit to R and R+M OMs.
+Accuracy was generally worse when EMs were fit to R+S OMs (Figure \ref{terminal_ssb_rmse}).
 
@@
 
-We found estimation of environmental effects on M was possible and reliable in certain scenarios even when the process error was mis-specified (R+S EM and R+M OM) also R EMs were not reliable for Ecov effects when PE was mis-specified.
-Given that milleretal in review found AIC was accurate for R+S OMs and applications of wham to real assessment data typically finds R+S preferrable to R only, it suggests inferences on covariate effects on M without just R process errors (e.g., deriso et al 2008) should be reevaluated.
-
-The estimating models assumed variances of aggregate catch and index observations was known.
-This approximation may be appropriate for indices where we have a reliable estimate of uncertainty based on the survey design (), but there may be better approaches for the aggregate catch such as an informed prior on the standard errors with realistic bounds.
-
-We found EMs with R+M process errors were rarely determined as the appropriate model when OMs simulated R+M process errors.
-This was unexpected but is likely due to the size of the variance assumed for those process errors (\(\sigma_M = 0.3\)) relative to the variances assumed for index and age composition observations.
-This is related to the difficulty in separately estimating observation and process error variances when the ratio of process to observation uncertainty is low.
-
-Note that the results for bias of covariate effect all assume ecov beta is estimated.
-The lack of bias in certain situations might suggest including the effect even if AIC says it isn't better at least when contrast in Ecov is high.
-
-See project 0 paper for relevance of \citet{lietal24} and \citet{liljestrandetal24}. See \citet{milleretal_inreview1} fro relevance of project 0 results (estimability of natural mortality)
-
-\citet{derisoetal08} was the first to model natural mortality as a function of explicit covariate and residual random annual variation. Any relevant points?
-
-We conditioned our investigations of reliability of parameter estimation on simulations with fits that converged satisfactorily.
-However, these simulations could exclude many realizations of covariate, population, and observation time series that are not sufficiently informative for all the parameters that we were estimating.
-Why might this matter?
-
-Poor quality of biological, fishery, and observation may adversely affect ability to distinguish between true process and observational error \citep{puntetal14, stewartmonnahan17, croninpunt21, fischetal23, lietal_inreview_a}.
-
-Uncertainty in estimated population attributes such as spawnign biomass can be increased considerably when natural mortality is estimated.
-This is also a consideration for estimating covariate effects on natural mortality even if the median natural mortality is assumed known.
-
-The bias in natural mortality estimation resulted in biased estimation of stock size and likely harvest rates, but it will also propagate into biological reference points and possibly stock status.
-
-The lack of bias observed for annual M, using R+S EMs when no effect was assumed must be due to the annual M being equal to the assumed value (0.2) on average across simulations because the simulated environmental covariate has mean 0.
-However, if one were to condition on the covariate time series we would expect biased estimation of annual M when there was a covariate effect.
-
-\hypertarget{conclusions}{%
-\section*{Conclusions}\label{conclusions}}
-\addcontentsline{toc}{section}{Conclusions}
-
-As we would expect, reliable detection of covariate effects requires informative data.
-AIC preferred simpler models than the true model when information content in data and contrast in covariates and abundance were low.
-Null model for environmental covariate effect (no covariate effect) was selected when contrast in the time series was low and/or uncertainty in observations was high.
-Null selection likely decreases with strength of the effect on M.
-We only examined two non-zero effect sizes: 0.25, and 0.5, but our results suggest larger effect sizes, with the same observation error and contrast in time series, would allow better AIC performance for determining covariate effects.
-When there was process error in recruitment and M (R+M), models with process error only in recruitment were preferred.
-This could be because the variation in the M random effects was small relative to observation variability.
-The marginal variance for log M random effects was 0.3.
-Reliable estimation of environmental effect on M regardless of the process error assumed by the EM as long as the contrast in the covariate is sufficient and the uncertainty in the observations is low.
+Our simulation study demonstrated that estimation of environmental effects on \(M\) is possible and reliable in certain scenarios even when the process error was misspecified (e.g., R and R+S EMs fit to R+M OM).
+In many of these same OMs frequency of convergence of model fits did not appears to suffer when covariate effects on \(M\) were estimated even when there was no effect simulated.
+However, these scenarios are information rich in that there was contrast in population size (due to changes in fishing pressure) and the covariate affecting the population, and low uncertainty in population and covariate observations.
+Previous research has shown that estimation of a constant \(M\) parameter requires contrast in time series and informative data \citep{leeetal11}, so it is no surprise that estimation of these effects also requires relatively good information via more precise observations and higher contrast in the covariate time series.
+
+On the other hand, AIC could only reliably detect covariate effects for R and R+M OMs with contrast in covariates and low covariate uncertainty and confidence interval coverage for the covariate effect was often biased even when bias in the estimators of the effect and the standard error were negligible.
+\citet{cadiganetal24} found confidence interval coverage to be biased for SSB and fishing mortality estimation in a state-space model in some scenarios, but they attributed the poor coverage to bias in Hessian-based standard error estimation and their simulations held any process error random effects constant.
+The coverage bias we observed, at least for some OM-EM combinations, may be a related to correlation of the estimators of the covariate effect and the corresponding standard error and therefore consideration of other methods of calculating confidence intervals may be warranted (e.g., those based on profile likelihood and/or MCMC sampling of log-likelihood surface).
+
+\citet{milleretal_inreview1} investigated R+M OMs with two levels of \(M\) process error variability (\(\sigma_M \in {0.1,0.5}\) and only found AIC able to accurately distinguish R+M process errors with the higher level of process error variability (\(\sigma_M = 0.5\)).
+We assumed \(\sigma_M = 0.3\), intermediate to the values investigated by \citet{milleretal_inreview1}, and found process error inferences unreliable for the source of process error, indicating that the level of variability required for detecting \(M\) process errors must be greater than \(\sigma_M = 0.3\), but may still be less than 0.5.
+In our results and those by \citet{milleretal_inreview1}, AIC typically chose R EMs which would indicate the fitted R+M EMs would estimate no variability in the \(M\) process errors.
+Future studies like ours where R+M OMs are simulated with greater variability in \(M\) process errors would better inform reliability of covariate effect inferences under such scenarios.
+
+\citet{derisoetal08} attempted to estimate process errors as well as covariate effects with \(M\) for Pacific herring, but similarly found no variability in \(M\), suggesting a there was too much uncertainty in the available observations relative to the true temporal variability in \(M\).
+Given that we found covariate effect inferences using R EMs was reliable in R+M OMs with apparently little variability in \(M\) process errors, the findings of \citet{derisoetal08} on covariate effects for Pacific herring would presumably also be robust to true low variability in \(M\).
+However, they did not account for uncertainty in covariate observations, some of which would probably have substantial uncertainty (e.g., competition and predation covariates).
+We found higher covariate observation uncertainty to cause true covariate effects to be less detectable using AIC, but we did not investigate the implications for incorrectly assuming no covariate uncertainty for covariate inferences.
+
+Any bias or poor accuracy for annual SSB estimation was primarily a function of whether or not the median \(M\) parameter was estimated or known and the types of process errors than the treatment of the covariate effects on \(M\).
+For example, we found estimation of SSB was better when the EM had the process error correctly specified for R+S OMs.
+Fortunately, our results and those by \citet{milleretal_inreview1} demonstrate that marginal AIC seems to be a good tool for determining whether this source of process error should be included in the model.
+However, the reliability of the estimation of SSB does break down in the less ideal scenarios when there is higher population observation error, and lack of contrast in fishing pressure (e.g., Figures \ref{terminal_ssb_bias_Rom} to \ref{terminal_ssb_bias_RMom}).
+
+The ability to accurately infer covariate effects on \(M\) in some realistic situations indicates that such investigations may be fruitful.
+Ability to make inferences could improve further when WHAM is extended to incorporate tagging data as expected \citep{milleretal_inreview}.
+Tagging data can greatly inform natural mortality estimation \citep{pollocketal91, hampton00}, and this impact on \(M\) estimation should also apply to estimation of covariate effects or unexplained temporal variation in \(M\).
+Given our findings coupled with future WHAM development, we expect investigations of and accounting for covariate effects on \(M\) to become more common within the fisheries stock assessment process.
 
@@
 \begin{center}
-%\includegraphics[width = \textwidth]{M_example.png}
-\end{center}
-\end{figure}
-\end{landscape}
+\includegraphics[height = \textheight]{M_example}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\hypertarget{convergence-results}{%
+\subsection*{Convergence results}\label{convergence-results}}
+\addcontentsline{toc}{subsection}{Convergence results}
+
+For R OMs, R EMs converged frequently except when there was contrast in fishing and covariate effects (\(\beta_E\)) were estimated and covariate observation uncertainty (\(\sigma_e\)) was high (Figure \ref{convergence_Rom}).
+There was little difference in convergence for R EMs whether median M was estimated or not.
+When R OMs had contrast in fishing, convergence of R+M EMs was generally better when the EM estimated median M, whereas in OMs with constant fishing, convergence was better when treating median M as known.
+Convergence of R+S EMs was generally worse than other process error configurations for R OMs, and there was no difference between EMs that estimated median M or not except when fishing pressure was constant where better convergence occurred when M was treated as known.
+Effects of true covariate effect size on convergence were apparent in some R OMs, but there did not appear to be any consistent patterns.
+
+For R+S OMs, R+M EMs generally showed the worse probability of convergence that EMs with R and R+S configurations (Figure \ref{convergence_RSom}).
+Convergence probability was high for all R and R+S EMs when covariate observation uncertainty was low.
+Although probability of convergence was generally greater when covariate effects were not estimated, there were no strong trends in convergence for any of the EMS with increased true effect size.
+
+For R+M OMs, R+S EMs generally converged less frequently than those with R and R+M configurations (Figure \ref{convergence_RMom}).
+Convergence probability for R OMs was generally the greatest of the three process error configurations.
+As with R+S OMs, there were no strong trends in convergence for any of the EMS with increased true covariate effect size.
+However, there was also not strong effects of estimating the covariate effect on convergence except for some OMs with higher covariate observation uncertainty.
+Like R OMs, when R+M OMs had contrast in fishing, convergence of R+M EMs was generally better when the EM estimated median M and the convergence was better when treating median M as known when the OM had constant fishing.
 
@@
 
+\hypertarget{aic-results}{%
+\subsection*{AIC results}\label{aic-results}}
+\addcontentsline{toc}{subsection}{AIC results}
+
 \begin{landscape}
@@
 \end{landscape}
+
+\hypertarget{covariate-effect-bias}{%
+\subsection*{Covariate effect bias}\label{covariate-effect-bias}}
+\addcontentsline{toc}{subsection}{Covariate effect bias}
 
@@
 
+\hypertarget{covariate-effect-standard-error-estimation-bias}{%
+\subsection*{Covariate effect standard error estimation bias}\label{covariate-effect-standard-error-estimation-bias}}
+\addcontentsline{toc}{subsection}{Covariate effect standard error estimation bias}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{se_beta_E_bias_main}
+\end{center}
+\caption{Median error (ME) of Hessian-based estimates of standard error for covariate effect on natural mortality $\beta_E$ from fitting EMs with alternative process error assumptions and treatment of median natural mortality ($e^\beta_M$ known or estimated).
+All OMs had low observation error and contrast in fishing mortality.
+True standard error is defined as the mean of the standard error estimates accross converged fits to simulated data sets for a given OM scenario.}\label{se_beta_E_bias}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{se_beta_E_bias_Rom}
+\end{center}
+\caption{For R OMs, median error (ME) of Hessian-based estimates of standard error for covariate effect on natural mortality $\beta_E$ from fitting EMs with alternative process error assumptions and treatment of median natural mortality ($e^\beta_M$ known or estimated).
+True standard error is defined as the mean of the standard error estimates accross converged fits to simulated data sets for a given OM scenario.}\label{se_beta_E_bias_Rom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{se_beta_E_bias_RSom}
+\end{center}
+\caption{For R+S OMs, median error (ME) of Hessian-based estimates of standard error for covariate effect on natural mortality $\beta_E$ from fitting EMs with alternative process error assumptions and treatment of median natural mortality ($e^\beta_M$ known or estimated).
+True standard error is defined as the mean of the standard error estimates accross converged fits to simulated data sets for a given OM scenario.}\label{se_beta_E_bias_RSom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{se_beta_E_bias_RMom}
+\end{center}
+\caption{For R+M OMs, median error (ME) of Hessian-based estimates of standard error for covariate effect on natural mortality $\beta_E$ from fitting EMs with alternative process error assumptions and treatment of median natural mortality ($e^\beta_M$ known or estimated).
+True standard error is defined as the mean of the standard error estimates accross converged fits to simulated data sets for a given OM scenario.}\label{se_beta_E_bias_RMom}
+\end{figure}
+\end{landscape}
+
+\hypertarget{covariate-effect-confidence-interval-coverage}{%
+\subsection*{Covariate effect confidence interval coverage}\label{covariate-effect-confidence-interval-coverage}}
+\addcontentsline{toc}{subsection}{Covariate effect confidence interval coverage}
+
 \begin{landscape}
@@
 
+\hypertarget{covariate-effect-rmse}{%
+\subsection*{Covariate effect RMSE}\label{covariate-effect-rmse}}
+\addcontentsline{toc}{subsection}{Covariate effect RMSE}
+
 \begin{landscape}
@@
 
+\hypertarget{covariate-effect-estimate-and-standard-error-example}{%
+\subsection*{Covariate effect estimate and standard error example}\label{covariate-effect-estimate-and-standard-error-example}}
+\addcontentsline{toc}{subsection}{Covariate effect estimate and standard error example}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{om_69_em_5_beta_E_se_beta_E_lm_plot}
+\end{center}
+\caption{Positive correlation of covariate effect estimates and Hessian-based standard error estimates for EM that also estimates the median natural mortality parameter and has correct R+M process error assumption fitted to simulated data from the OM with R+M process errors, temporal contrast in fishing pressure, low observation uncertainty for both population ($Low OE$) and covariate observations ($\sigma_e = 0.1$), high and uncorrelated temporal variability in the true covariate ($\sigma_E = 0.5$ and $\rho_E = 0$), and the strongest covariate effect on natural mortality ($\beta_E = 0.5$).}\label{ex_lm_beta_E_SE_beta_E}
+\end{figure}
+\end{landscape}
+
+\hypertarget{median-natural-mortality-parameter-bias}{%
+\subsection*{Median Natural mortality parameter bias}\label{median-natural-mortality-parameter-bias}}
+\addcontentsline{toc}{subsection}{Median Natural mortality parameter bias}
+
 \begin{landscape}
@@
 
+\hypertarget{median-natural-mortality-parameter-standard-error-estimation-bias}{%
+\subsection*{Median natural mortality parameter standard error estimation bias}\label{median-natural-mortality-parameter-standard-error-estimation-bias}}
+\addcontentsline{toc}{subsection}{Median natural mortality parameter standard error estimation bias}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{se_beta_M_bias_main}
+\end{center}
+\caption{Median error (ME) of Hessian-based estimates of standard error for median natural mortality parameter $\beta_M$ from fitting EMs with alternative process error assumptions and treatment of the covariate effect ($\beta_ E= 0$ or estimated).
+All OMs had low observation error and contrast in fishing mortality.
+True standard error is defined as the mean of the standard error estimates accross converged fits to simulated data sets for a given OM scenario.}\label{se_beta_M_bias}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{se_beta_M_bias_Rom}
+\end{center}
+\caption{For R OMs, median error (ME) of Hessian-based estimates of standard error for median natural mortality parameter $\beta_M$ from fitting EMs with alternative process error assumptions and treatment of the covariate effect ($\beta_ E= 0$ or estimated).
+True standard error is defined as the mean of the standard error estimates accross converged fits to simulated data sets for a given OM scenario.}\label{se_beta_M_bias_Rom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{se_beta_M_bias_RSom}
+\end{center}
+\caption{For R+S OMs, median error (ME) of Hessian-based estimates of standard error for median natural mortality parameter $\beta_M$ from fitting EMs with alternative process error assumptions and treatment of the covariate effect ($\beta_ E= 0$ or estimated).
+True standard error is defined as the mean of the standard error estimates accross converged fits to simulated data sets for a given OM scenario.}\label{se_beta_M_bias_RSom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{se_beta_M_bias_RMom}
+\end{center}
+\caption{For R+M OMs, median error (ME) of Hessian-based estimates of standard error for median natural mortality parameter $\beta_M$ from fitting EMs with alternative process error assumptions and treatment of the covariate effect ($\beta_ E= 0$ or estimated).
+True standard error is defined as the mean of the standard error estimates accross converged fits to simulated data sets for a given OM scenario.}\label{se_beta_M_bias_RMom}
+\end{figure}
+\end{landscape}
+
+\hypertarget{median-natural-mortality-parameter-confidence-interval-coverage}{%
+\subsection*{Median Natural mortality parameter confidence interval coverage}\label{median-natural-mortality-parameter-confidence-interval-coverage}}
+\addcontentsline{toc}{subsection}{Median Natural mortality parameter confidence interval coverage}
+
 \begin{landscape}
@@
 
+\hypertarget{median-natural-mortality-parameter-estimate-and-standard-error-example}{%
+\subsection*{Median natural mortality parameter estimate and standard error example}\label{median-natural-mortality-parameter-estimate-and-standard-error-example}}
+\addcontentsline{toc}{subsection}{Median natural mortality parameter estimate and standard error example}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{om_69_em_5_beta_M_se_beta_M_lm_plot}
+\end{center}
+\caption{Negative correlation of $\beta_M$ estimates and Hessian-based standard error estimates for EM that also estimates the covariate effect and has correct R+M process error assumption fitted to simulated data from the OM with R+M process errors, temporal contrast in fishing pressure, low observation uncertainty for both population ($Low OE$) and covariate observations ($\sigma_e = 0.1$), high and uncorrelated temporal variability in the true covariate ($\sigma_E = 0.5$ and $\rho_E = 0$), and the strongest covariate effect on natural mortality ($\beta_E = 0.5$).}\label{ex_lm_beta_M_SE_beta_M}
+\end{figure}
+\end{landscape}
+
+\hypertarget{median-natural-mortality-parameter-rmse}{%
+\subsection*{Median Natural mortality parameter RMSE}\label{median-natural-mortality-parameter-rmse}}
+\addcontentsline{toc}{subsection}{Median Natural mortality parameter RMSE}
+
 \begin{landscape}
@@
 
+\hypertarget{terminal-year-natural-mortality-bias}{%
+\subsection*{Terminal year natural mortality bias}\label{terminal-year-natural-mortality-bias}}
+\addcontentsline{toc}{subsection}{Terminal year natural mortality bias}
+
 \begin{landscape}
@@
 
+\hypertarget{terminal-year-natural-mortality-rmse}{%
+\subsection*{Terminal year natural mortality RMSE}\label{terminal-year-natural-mortality-rmse}}
+\addcontentsline{toc}{subsection}{Terminal year natural mortality RMSE}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_M_rmse_main}
+\end{center}
+\caption{Root mean square error (RMSE) of estimates of natural mortality rate in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).
+All OMs had low population observation error and contrast in fishing mortality.}\label{terminal_M_rmse}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_M_rmse_Rom}
+\end{center}
+\caption{For R OMs, root mean square error (RMSE) of estimates of natural mortality rate in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).}\label{terminal_M_rmse_Rom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_M_rmse_RSom}
+\end{center}
+\caption{For R+S OMs, root mean square error (RMSE) of estimates of natural mortality rate in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).}\label{terminal_M_rmse_RSom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_M_rmse_RMom}
+\end{center}
+\caption{For R+M OMs, root mean square error (RMSE) of estimates of natural mortality rate in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).}\label{terminal_M_rmse_RMom}
+\end{figure}
+\end{landscape}
+
+\hypertarget{terminal-year-spawning-stock-biomass-bias}{%
+\subsection*{Terminal year spawning stock biomass bias}\label{terminal-year-spawning-stock-biomass-bias}}
+\addcontentsline{toc}{subsection}{Terminal year spawning stock biomass bias}
+
 \begin{landscape}
@@
 
+\hypertarget{terminal-year-spawning-stock-biomass-rmse}{%
+\subsection*{Terminal year spawning stock biomass RMSE}\label{terminal-year-spawning-stock-biomass-rmse}}
+\addcontentsline{toc}{subsection}{Terminal year spawning stock biomass RMSE}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_ssb_rmse_main}
+\end{center}
+\caption{Root mean square error (RMSE) of estimates of spawning stock biomass in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).
+All OMs had low population observation error and contrast in fishing mortality.}\label{terminal_ssb_rmse}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_ssb_rmse_Rom}
+\end{center}
+\caption{For R OMs, root mean square error (RMSE) of estimates of spawning stock biomass in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).}\label{terminal_ssb_rmse_Rom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_ssb_rmse_RSom}
+\end{center}
+\caption{For R+S OMs, root mean square error (RMSE) of estimates of spawning stock biomass in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).}\label{terminal_ssb_rmse_RSom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_ssb_rmse_RMom}
+\end{center}
+\caption{For R+M OMs, root mean square error (RMSE) of estimates of spawning stock biomass in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).}\label{terminal_ssb_rmse_RMom}
+\end{figure}
+\end{landscape}
+
+\hypertarget{terminal-year-fishing-mortality-bias}{%
+\subsection*{Terminal year fishing mortality bias}\label{terminal-year-fishing-mortality-bias}}
+\addcontentsline{toc}{subsection}{Terminal year fishing mortality bias}
+
 \begin{landscape}
@@
 
+\hypertarget{terminal-year-fishing-mortality-rmse}{%
+\subsection*{Terminal year fishing mortality RMSE}\label{terminal-year-fishing-mortality-rmse}}
+\addcontentsline{toc}{subsection}{Terminal year fishing mortality RMSE}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_F_rmse_main}
+\end{center}
+\caption{Root mean square error (RMSE) of estimates of fully-selected fishing mortality ($F$)  in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).
+All OMs had low population observation error and contrast in fishing mortality.}\label{terminal_F_rmse}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_F_rmse_Rom}
+\end{center}
+\caption{For R OMs, root mean square error (RMSE) of estimates of fully-selected fishing mortality ($F$)  in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).}\label{terminal_F_rmse_Rom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_F_rmse_RSom}
+\end{center}
+\caption{For R+S OMs, root mean square error (RMSE) of estimates of fully-selected fishing mortality ($F$)  in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).}\label{terminal_F_rmse_RSom}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = \textheight]{terminal_year_F_rmse_RMom}
+\end{center}
+\caption{For R+M OMs, root mean square error (RMSE) of estimates of fully-selected fishing mortality ($F$)  in the terminal year for EMs with alternative process error assumptions, treatment of covariate effect ($\beta_E = 0$ or estimated), and treatment of median natural mortality parameter ($\beta_M$ estimated or known).}\label{terminal_F_rmse_RMom}
+\end{figure}
+\end{landscape}
+
 \end{document}

### 2025-06-09 Ecov_study/mortality/code/manuscript/M_results.R

@@ -32,7 +32,7 @@ conv_fn <- function(om, em, conv_res, Type = 3){
 #5: maximum non-NaN SE estimate < 10
 
 #all EM PE assumptions
-plot_df_fn <- function(df.ems, df.oms, Ecov_est = FALSE, M_est = FALSE, conv_type = 3) {
+plot_df_fn <- function(df.ems, df.oms, Ecov_est = FALSE, M_est = FALSE, conv_type = 3, years = 40) {
   custom_boxplot_stat <- function(x){#, n.yrs = 1, n.sim = 100) {
     x <- x[which(!is.na(x))]
     n <- length(x)
@@ -43,32 +43,34 @@ plot_df_fn <- function(df.ems, df.oms, Ecov_est = FALSE, M_est = FALSE, conv_typ
     r
   }
   for(h in 1:3) {
-    re_mod <- c("rec", "rec+1", "rec+M")[h]
+    re_mod <- c("rec", "rec+1", "rec+M")[h] #OM re config
     #EM:  M fixed, mean_M estimated
-    em_ind <- which(df.ems$Ecov_est== Ecov_est & df.ems$M_est == M_est)
+    em_ind <- which(df.ems$Ecov_est== Ecov_est & df.ems$M_est == M_est) #all three PE configs for the ems
     print("em_ind")
     print(em_ind)
     om_ind <- which(df.oms$NAA_M_re == re_mod) #om and em match
     res <- lapply(om_ind, function(x) {
-      M_res <- lapply(1:40, function(y) {
+      if(is.null(years)) years <- 1:40
+      M_res <- lapply(years, function(y) {
         out <- matrix(NA,length(em_ind), 8)
         for(j in em_ind){
           conv_ind <- 1:100 #all of them
           if(!is.null(conv_type)) conv_ind <- conv_fn(x,j,conv_res,Type = conv_type) #subset consistent with convergence results
           r_e <- M_bias[[x]][conv_ind,j,y,2]/M_bias[[x]][conv_ind,j,y,1]-1
-          log_M_e <- log(M_bias[[x]][conv_ind,j,y,2]) - log(M_bias[[x]][conv_ind,j,y,1])
-          # rmse <- sqrt(mean(log_M_e^2, na.rm = TRUE))
           rmse <- sqrt(mean((M_bias[[x]][conv_ind,j,y,2]-M_bias[[x]][conv_ind,j,y,1])^2, na.rm = TRUE))
[…]

### 2025-06-09 Ecov_study/mortality/code/manuscript/convergence_results.R

@@ -85,7 +85,7 @@ conv_res_plotting_fn <- function(conv_res, M_est = TRUE, Ecov_est = TRUE){
       "TRUE" = "Estimated",
       "FALSE" = "Known"
     ))
-  df$M_assumption <- factor(df$M_assumption)
+  # df$M_assumption <- factor(df$M_assumption)
   df <- df %>%
     mutate(Ecov_assumption = recode(as.character(Ecov_est),
       "TRUE" = "beta[italic(E)]*' Estimated'",
@@ -145,9 +145,9 @@ df <- rbind(
   conv_res_plotting_fn(conv_res, M_est = TRUE, Ecov_est = FALSE))
 
 theme_set(theme_bw())
-theme_update(strip.text.x = element_text(size = rel(1.3)), strip.text.y = element_text(size = rel(1.5)), strip.placement = "outside", strip.background = element_rect(), #fill = "transparent"), 
-       axis.title = element_text(size = rel(1.5)), axis.text = element_text(size = rel(1.25)), legend.text = element_text(size = rel(1.5)), #text = element_text(size = rel(2)), 
-       legend.title = element_text(size = rel(1.5)))
+theme_update(strip.text = element_text(size = rel(1.5)), strip.placement = "outside", strip.background = element_rect(), #fill = "transparent"), 
+      axis.title = element_text(size = rel(2)), axis.text = element_text(size = rel(1.5)), legend.text = element_text(size = rel(2)), #text = element_text(size = rel(2)), 
+      legend.title = element_text(size = rel(2)), legend.title.align=0.5)
 
 levels(df$M_assumption)
 

### 2025-06-09 Ecov_study/mortality/code/manuscript/ssb_results.R

@@ -31,42 +31,46 @@ conv_fn <- function(om, em, conv_res, Type = 3){
 #4: max gradient value < 1e-6
 #5: maximum non-NaN SE estimate < 10
 
-custom_boxplot_stat <- function(x){#, n.yrs = 1, n.sim = 100) {
-  x <- x[which(!is.na(x))]
-  n <- length(x)
-  bnds95 <- qbinom(c(0.025,0.975), n, 0.5)/n # 95% CI bounds for median (sims x years)
-  bnds80 <- qbinom(c(0.1,0.9), n, 0.5)/n # 80% CI bounds for median (sims x years)
-  r <- quantile(x, probs = c(bnds95[1], bnds80[1], 0.5, bnds80[2], bnds95[2]))
-  names(r) <- c("ymin", "lower", "middle", "upper", "ymax")
-  r
-}
-plot_df_fn <- function(df.ems, df.oms, Ecov_est = FALSE, M_est = FALSE, conv_type = 3) {
-  print(which(df.oms$NAA_M_re == "rec+1"))
+#all EM PE assumptions
+plot_df_fn <- function(df.ems, df.oms, Ecov_est = FALSE, M_est = FALSE, conv_type = 3, years = 40) {
+  custom_boxplot_stat <- function(x){#, n.yrs = 1, n.sim = 100) {
+    x <- x[which(!is.na(x))]
+    n <- length(x)
+    bnds95 <- qbinom(c(0.025,0.975), n, 0.5)/n # 95% CI bounds for median (sims x years)
+    bnds80 <- qbinom(c(0.1,0.9), n, 0.5)/n # 80% CI bounds for median (sims x years)
+    r <- quantile(x, probs = c(bnds95[1], bnds80[1], 0.5, bnds80[2], bnds95[2]))
+    names(r) <- c("ymin", "lower", "middle", "upper", "ymax")
+    r
+  }
   for(i in 1:3) {
-    re_mod <- c("rec", "rec+1", "rec+M")[i]
+    re_mod <- c("rec", "rec+1", "rec+M")[i] #OM re config
     #EM:  M fixed, mean_M estimated
-    em_ind <- which(df.ems$Ecov_est== Ecov_est & df.ems$M_est == M_est)
+    em_ind <- which(df.ems$Ecov_est== Ecov_est & df.ems$M_est == M_est) #all three PE configs for the ems
     print("em_ind")
     print(em_ind)
[…]

### 2025-08-08 Ecov_study/mortality/manuscript/manuscript.Rmd
saving work on project_0 revisions

@@
   - Timothy J. Miller^1,2^
-  - Greg Britten^3^
+  - Gregory L. Britten^3^
   - Elizabeth N. Brooks^2^
   - Gavin Fay^4^
-  - Alex Hansell^2^
+  - Alexander C. Hansell^2^
   - Christopher M. Legault^2^
@@
 $^3$Biology Department, Woods Hole Oceanographic Institution, 266 Woods Hole Rd. Woods Hole, MA, USA\
-$^4$Department of Fisheries Oceanography, School for Marine Science and Technology, University of Massachusetts Dartmouth, 836 S Rodney French Boulevard, New Bedford, MA 02740, USA\
+$^4$Department of Fisheries Oceanography, School for Marine Science and Technology, University of Massachusetts Dartmouth, 836 S Rodney French Boulevard, New Bedford, MA 02744, USA\
 $^5$Mid-Atlantic Fishery Management Council, 800 North State Street, Suite 201, Dover, DE 19901 USA\
@@
 Treatment of natural mortality is a major consideration in assessment models and state-space approaches allow estimation of temporal variation in this mortality rate as well as effects of specific covariates.
-However, there has been no investigation of the reliability of inferences made regarding natural mortality, associated covariate effects and important assessment output from state-space assessment models exists.
+However, there has been no investigation of the reliability of inferences made regarding natural mortality, associated covariate effects, and important assessment output from state-space assessment models.
 We conducted a large-scale simulation study that considers models fit to data simulated from operating models with alternative assumptions defined by several factors, but we focus on scenarios where there is temporal contrast in fishing pressure and lower uncertainty in population observations (age composition and indices of abundance).
-The estimating models we fit to simulated observations had alternative assumptions on whether to include the environmental effect, whether the median natural mortality rate was estimated or known, and the source of temporal variability in the population demography.
-Lower uncertainty in covariate observations, higher covariate temporal variability and the source of demographic temporal variability were important for better inferences regarding the covariate effect on natural mortality.
+We fit estimating models to simulated observations with alternative assumptions on inclusion of environmental effects, estimation of the median natural mortality rate, and the source of temporal variability in the population demography.
+Our results suggest that estimation of environmental effects on natural mortality is possible and reliable even when the population process error source was misspecified in some scenarios with lower uncertainty in covariate observations, and higher covariate temporal variability.
 
@@
 Because application of these methods are considered best practice and recommended for the next generation of stock assessment models [@hoyleetal22; @punt23], it is expected their use will only grow globally.
-An appeal of state-space models lies in their formulation treating latent population characteristics as statistical time series with periodic observations that also may have error due to sampling or other sources of measurement error and therefore separating these sources of biological and measurement variability.
+An appeal of state-space models lies in their separation of sources of biological and measurement variability by treating latent population characteristics as statistical time series with periodic observations measured with error.
 Through advances in computational capacity, we can use sophisticated numerical approaches to estimate model parameters as mixed effects [@thorsonminto15;@kristensenetal16].
@@
 State-space stock assessment models, with non-linear functions of latent processes and numerous observation types with different probability distribution assumptions represent one of most complex classes of state-space models.
-The literature on the effects of various factors on reliabilty of inferences from state-space assessment models is growing [@lietal24;@milleretal_inreview1].
-The importance of contrast in population size and fishing mortality and quality of data used to fit assessment models including the state-space variety is known [@magnussonhilborn07;@milleretal_inreview1].
-Furthermore, estimation of natural mortality, and even temporal variablity is possible in many scenarios [@leeetal11;@cadigan16;@millerhyun18;@milleretal_inreview1].
+The literature on the effects of various factors on reliability of inferences from state-space assessment models is growing [@lietal24;@milleretal_inreview1].
+The importance of contrast in population size and fishing mortality ($F$) and quality of data used to fit assessment models including the state-space variety is known [@magnussonhilborn07;@milleretal_inreview1].
+Furthermore, estimation of natural mortality ($M$), and even temporal variability in $M$ is possible in many scenarios [@leeetal11;@cadigan16;@millerhyun18;@milleretal_inreview1].
 
@@
 While understanding drivers of recruitment is important particularly for subsequent effects on reference points, recruitment in short-term projections typically has little impact on the exploitable biomass in the first few projection years.
-However, assumptions for natural mortality ($M$) have immediate and larger effects on projected biomass because they affect the abundances at older age classes at the end of the data time series that constitute spawning biomass and catch [@brodziaketal08;@stocketal21].
+However, assumptions for $M$ have immediate and larger effects on projected biomass because they affect the abundances at older age classes at the end of the data time series that constitute spawning biomass and catch [@brodziaketal08;@stocketal21].
 
@@
 Temporal variation in $M$ is less studied than recruitment, but its importance for explaining variability in observations has been demonstrated in state-space assessment models for Atlantic cod and yellowtail flounder [@cadigan16;@stocketal21].
-@derisoetal08 also demonstrated the importance of several factors affecting Pacific herring natural mortality.
+@derisoetal08 also demonstrated the importance of several factors affecting $M$ for Pacific herring.
 
 Assessment models could include temporal variation in many aspects of population dynamics or how observations are related to the population.
-For example the Woods Hole Assessment Model (WHAM) can include process errors treated as random effects for transition in cohorts over time, catchability for indices of abundance, selectivity of fishing fleets or indices, movement between regions, or in $M$ [@stockmiller21;@milleretal_inreview].
+For example the Woods Hole Assessment Model (WHAM) can include process errors treated as random effects for transition in cohorts over time (hereafter referred to as apparent survival), catchability for indices of abundance, selectivity of fishing fleets or indices, movement between regions, or in $M$ [@stockmiller21;@milleretal_inreview].
 However, misspecified temporal population process errors could lead to biased population and stock status estimation, and, therefore, poor fisheries management decisions [ @legaultpalmer16; @szuwalskietal18].
-Studies of the reliability of inferences we might make regarding the presence of temporal variability in $M$ are limited.
-@milleretal_inreview1 found AIC could accurately distinguish process errors in the changes in abundance for cohorts, but not for those specifically due to $M$ except when uncertainty in population observations (indices, catch and age composition) was low and there was greater temporal variation in natural mortality.
-In their simulation studies looking a models with multiple sources of process error, @lietal24 found including more sources of process error than existed in the operating model was a better model-building approach than excluding them a priori.
-
-Here we conduct a simulation study with operating models (OMs) varying by degree of observation error uncertainty, sources of process error, fishing history, temporal variation in environmental covariates, and magnitude of the effect of the covariate on natural mortality.
-The simulated observations from these OMs are fitted with estimating models that make alternative assumptions for sources of process error, and whether median M and covariate effects are estimated.
-We evaluate the effects of these factors on convergence of fitted models, whether Akaike's information criterion (AIC) can correctly determine the correct source of process error and correct assumption about covariate effects on natural mortality, and the degree of bias in inferences for relevant parameters and outputs of the assessment model.
+Studies of the reliability of inferences regarding the presence of temporal variability in $M$ are limited.
+@milleretal_inreview1 found AIC could accurately distinguish process errors in apparent survival, but not for those specifically due to $M$ except when uncertainty in population observations (indices, catch and age composition) was low and there was greater temporal variation in $M$.
+In their simulation studies looking at models with multiple sources of process error, @lietal24 found including more sources of process error than existed in the operating model was a better model-building approach than excluding them a priori.
+
+Here we conduct a simulation study with operating models (OMs) varying by degree of observation error uncertainty, sources of process error, fishing history, temporal variation in environmental covariates, and magnitude of the effect of the covariate on $M$.
+The simulated observations from these OMs are fitted with estimating models that make alternative assumptions for sources of process error, and whether median $M$ and covariate effects are estimated.
+We evaluate the effects of these factors on convergence of fitted models, whether Akaike's information criterion (AIC) can determine the correct source of process error and correct assumption about covariate effects on $M$, and the accuracy for estimators of relevant parameters and stock size and harvest rates derived from the assessment model.
 
@@
 
-All of our analyses used the Woods Hole Assessment Model (WHAM) to construct both OMs and EMs [@millerstock20; @stockmiller21; @milleretal_inreview].
+Our analyses used the Woods Hole Assessment Model (WHAM) to construct both OMs and EMs [@millerstock20; @stockmiller21; @milleretal_inreview].
 The WHAM package has been used extensively to configure OMs and EMs for several other simulation studies [@stocketal21;@legaultetal23; @lietal24; @brittenetal_inreview; @lietal_inreview_a] and is used to assess many commercially important stocks in the Northeast U.S.
 [e.g., @nefsc22;@nefsc22a;@nefsc24].
-We used [version 1.0.6.9000, commit 77bbd94](https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3) for to generate all results.
-
-We completed a simulation study with a 288 operating models.
-The factors defining the configuration of each operating model which are described in detail in subsequent sections include source of population process error (3 levels), index and catch observation uncertainty (2 levels), environmental covariate uncertainty (2 levels), latent environmental covariate process stochasticity (4 levels), and fishing history (2 levels).
+We used [version 1.0.6.9000, commit 77bbd94](https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3) to generate all results.
+
+We completed a simulation study with 288 operating models.
+The factors defining the configuration of each operating model, described in detail in subsequent sections, include source of population process error (3 levels), index and catch observation uncertainty (2 levels), environmental covariate uncertainty (2 levels), temporal variation in the latent environmental covariate (4 levels), and fishing history (2 levels).
 We simulated 100 data sets for each operating model that included simulations of process errors.
@@
 For each simulated data set we fit a set of 12 estimating models (EMs).
-The factors that distinguish the estimating models which are also described in detail below include source of population process error type (3 levels) whether (median) $M$ was estimated or assumed known (2 levels), and whether the effect of the environmental covariate on $M$ was estimated or not (2 levels).
-
-The sources of population process error that were used in the OMs or assumed in the EMs were on recruitment only (R), recruitment and changes in cohort abundance over time (R+S), or recruitment and $M$ (R+M).
-We did not use the log-normal bias-correction feature for process errors or observations described by @stockmiller21 for operating and EMs [@lietal_inreview].
+The factors that distinguish the estimating models, also described in detail below, include source of population process error type (3 levels) whether (median) $M$ was estimated or assumed known (2 levels), and whether the effect of the environmental covariate on $M$ was estimated or not (2 levels).
+
+The sources of population process error that were used in the OMs or assumed in the EMs were on recruitment only (R), recruitment and apparent survival (R+S), or recruitment and $M$ (R+M).
+We did not use the log-normal bias-correction feature for process errors or observations described by @stockmiller21 for OMs and EMs [@lietal_inreview].
 Simulations were all carried out on the University of Massachusetts Green High-Performance Computing Cluster.
@@
 with marginal mean $\mu_E=0$ and variance $\sigma^2_E$.
-The four configuration of the latent environmental covariate in the operating models assume the one of two values for the marginal standard deviation $\sigma_E \in \{0.1, 0.5\}$ and for the autocorrelation parameter $\rho_E \in \{0, 0.5\}$.
+The four configurations of the latent environmental covariate in the operating models assume one of two values for the marginal standard deviation $\sigma_E \in \{0.1, 0.5\}$ and for the autocorrelation parameter $\rho_E \in \{0, 0.5\}$.
 
@@
 The standard deviation of the environmental observations in the operating models is one of two values $\sigma_e \in \{0.1, 0.5\}$.
-Figure \ref{om_ecov_example} provides example simulations of the latent process and observations under the alternative configurations.
+Figure \ref{om_ecov_example} provides example simulations of the latent covariate and observations under the alternative configurations.
 
@@
 Many of the characteristics of the population biology and structure including the age classes (10 age classes (ages 1 to 10+)), time span (40 years), maturity (Figure \ref{om_inputs_fig}, top left), growth  (Figure \ref{om_inputs_fig}, top right), time of spawning (1/4 of the year), and recruitment (Figure \ref{om_inputs_fig}, bottom right) are identical to @milleretal_inreview1.
-The maturity at age is a logistic function of age with age at 50\% maturity ($a_{50}$) = 2.89 and slope = 0.88 and weight at age is derived from a von Beralanffy growth function where $t_0 = 0$, $L_\infty = 85$, and $k = 0.3$, and a length-weight relationship
+The maturity at age is a logistic function with age at 50\% maturity ($a_{50}$) = 2.89 and slope = 0.88 and weight at age is derived from a von Beralanffy growth function where $t_0 = 0$, $L_\infty = 85$, and $k = 0.3$, and a length-weight relationship
 $$
@@
 For R and R+S OMs and EMs, $\varepsilon_{M,y} = 0$.
-For all R+M OMs, we assume the same standard deviation $\sigma_M = 0.3$ and is estimated in the R+M EMs.
+For all R+M OMs, we assume the same standard deviation $\sigma_M = 0.3$, which is estimated in the R+M EMs.
 The covariate effect is one of 3 alternative values in the operating models, $\beta_\text{E} \in \{0,0.25,0.5\}$.
@@
 Therefore, steepness and equilibrium unfished recruitment are also constant over the time period for those OMs [@millerbrooks21].
-As in @milleretal_inreview1, our assumed biological inputs and selectivity (defined below) with constant $M$ result in equilibrium fishing mortality that reduces spawning biomass per recruit to 40\% of the unfished level is $F_{40\%} = 0.348$.
+As in @milleretal_inreview1, our assumed biological inputs and selectivity (defined below) with constant $M$ result in equilibrium $F$ that reduces spawning biomass per recruit to 40\% of the unfished level is $F_{40\%} = 0.348$.
 With an assumed unfished recruitment of $R_0 = e^{10}$, setting $\Fmsy = F_{40\%}$ results in a steepness of 0.69 and $a=0.60$ and $b = 2.4 \times 10^{-5}$.
-For R+M OMs and all OMs with covariate effects on $M$, steepness is not constant, but we used the same $a$ and $b$ parameters as other operating models which equates to a steepness and $R_0$ at the median of the time series processes for $M$ and the covariate.
+For R+M OMs and all OMs with covariate effects on $M$, steepness is not constant, but we used the same $a$ and $b$ parameters as other operating models which equates to a steepness and $R_0$ at the median of the time series models for $M$ and the covariate.
 
@@
 In the second scenario, the stock is fished at \Fmsy for the entire time period (40 years).
+The magnitude of the overfishing assumptions is intended to reflect estimates of overfishing for Northeast US groundfish stocks from @wiedenmannetal19.
 
@@
 This same assumption was used by @milleretal_inreview1 for R+M OMs and other OMs with fishery selectivity and index catchability process errors.
-For R+S OMs, cohort temporal transition process errors were uncorrelated with $\sigma_{2+} = 0.3$
+For R+S OMs, apparent survival process errors were uncorrelated with $\sigma_{2+} = 0.3$.
 
@@
 
-We define the generation of observations of total catch, aggregate indices, and corresponding age composition identical to @milleretal_inreview1.
+We define the generation of observations of aggregate (total combined across ages) catch and indices, and corresponding age composition identical to @milleretal_inreview1.
 There is a single fleet operating year round for catch observations with logistic selectivity for the fleet with $a_{50} = 5$ and slope = 1 (Figure \ref{om_inputs_fig}, bottom left).
@@
 Catchability of both surveys are assumed to be 0.1.
-We assumed catch and index age composition observations aer generated from a logistic-normal distribution where errors on the multivariate normal scale are independent.
+We assumed catch and index age composition observations are generated from a logistic-normal distribution where errors on the multivariate normal scale are independent.
 The standard deviation parameter is also constant across ages.
@@
 Standard deviation for log-aggregate catch was 0.1.
-There were two levels of observation error variance for indices and age composition for both indices and fleet catch.
+There were two levels of observation error variance for aggregate indices and age composition for both indices and fleet catch.
 A low uncertainty specification assumed standard deviation of both series of log-aggregate index observations was 0.1 and the standard deviation of the logistic-normal for age composition observations was 0.3.
 In the high uncertainty specification the standard deviation for log-aggregate indices was 0.4 and that for the age composition observations was 1.5.
-For all estimating models, standard deviation for log-aggregate observations was assumed known whereas that for the logistic-normal age composition observations was estimated.
+For all estimating models, the standard deviation for log-aggregate observations was assumed known at the true value whereas that for the logistic-normal age composition observations was estimated.
 
@@
 
-For each data set simulated from an operating model 12 estimating models were fit.
+Estimating models were fit to each of 100 simulated data sets from each operating model.
 There were three factors defining the configuration of each estimating model: 1) whether $\beta_M$ was estimated or assumed known, 2) whether an environmental effect $\beta_E$ was estimated or not, and 3) whether the process errors were assumed on recruitment only (R), recruitment and survival (R+S), or recruitment and $M$ (R+M).
@@
 For example, uncorrelated R+S was assumed for both the estimating and operating model.
-However, R+M EMs did not assume M random effects were uncorrelated ($\rho_M$ was estimated).
-The environmental covariate observations were included in all estimation models to ensure comparability of AIC.
-All fixed effects parameters for selectivity, catchability, fully-selected fishing mortality, mean recruitment, initial abundance at age, and variances for logistic-normal age composition distributions were estimated.
+However, R+M EMs did not assume $M$ random effects were uncorrelated ($\rho_M$ was estimated).
+The environmental covariate observations were included in all estimation models, whether effects on $M$ were estimated or not, to ensure comparability of AIC.
+All fixed effects parameters for selectivity, catchability, fully-selected $F$, mean recruitment, initial abundance at age, and variances for logistic-normal age composition distributions were estimated.
 Any process error variance parameters for recruitment, survival, and $M$ were also estimated.
@@
 There are various ways to assess convergence of the fit [e.g., @carvalhoetal21], but we defined successful convergence as the Hessian of the marginal log-likelihood being invertible and providing variance estimates for the fixed effects parameters as recommended by @milleretal_inreview1.
-We calculated 95\% confidence intervals for probability of convergence using the Clopper-Pearson exact method [@clopperpearson34;@thulin14].
+We calculated 95\% confidence intervals (CIs) for probability of convergence using the Clopper-Pearson exact method [@clopperpearson34;@thulin14].
 
@@
 $$
-All model fits that successfully completed the optimization were used in these results.
+All model fits that successfully completed the optimization were used for this set of analyses.
 We used all of these fits because some lack of convergence would be expected for the correct behavior of more complicated models that include process errors that did not exist in the operating model.
@@
 All results here use OM simulations with fits that satisfied the convergence criterion described above.
-We used this conditioning, to reflect how practitioners would proceed in analyses of model fits with real assessment data.
+We used this conditioning to reflect how practitioners would proceed in analyses of model fits with real assessment data.
 That is, practitioners would ensure models converged such that Hessian-based standard errors were available for all model parameter estimates.
 
-We focused on statistical behavior of estimators of the covariate effect on $M$ ($\widehat \beta_E$), the estimator of the median $M$ parameter ($\widehat \beta_M$), and estimators of terminal year natural mortality $\widehat M$, spawning stock biomass $\widehat{\text{SSB}}$, and fully-selected fishing mortality rate $\widehat F$.
-In preliminary analyses we results of the estimators of all the annual values for $M$, SSB and fishing mortality over the whole time series, but we found no appreciable differences in patterns across the various factors defining the OMs and EMs.
-Furthermore, results for terminal year fishing mortality were generally inversely related to those for spawning stock biomass, and are provided in the Supplementary Materials and not discussed further.
-
-We calculated median errors (ME) of $\widehat \beta_E$, $\widehat \beta_M$), and the Hessian-based standard error estimators of these parameters ($\widehat{SE}\left(\widehat \beta_E\right)$ and $\widehat{SE}\left(\widehat \beta_M\right)$).
+We focused on statistical behavior of estimators of the covariate effect on $M$ ($\widehat \beta_E$), the estimator of the median $M$ parameter ($\widehat \beta_M$), and estimators of terminal year $M$ ($\widehat M$), spawning stock biomass ($\widehat{\text{SSB}}$), and fully-selected $F$ ($\widehat F$).
+In preliminary analyses we examined results for the estimators of all the annual values for $M$, SSB and $F$ over the whole time series, but we found no appreciable differences in patterns across the various factors defining the OMs and EMs.
+Furthermore, results for terminal year $F$ were generally inversely related to those for spawning stock biomass, and are provided in the Supplementary Materials and not discussed further.
+
+We calculated median errors (ME) of $\widehat \beta_E$ and $\widehat \beta_M$, and the Hessian-based standard error estimators of these parameters ($\widehat{SE}\left(\widehat \beta_E\right)$ and $\widehat{SE}\left(\widehat \beta_M\right)$).
 We calculated the median relative errors (MRE) of terminal year $\widehat M$, $\widehat{\text{SSB}}$, and $\widehat F$.
-We also calculated the root mean square error (RMSE) and estimated probability of coverage of constructed 95\% confidence intervals for $\widehat \beta_E$ and $\widehat \beta_M$ for EMs that estimated these parameters.
-We constructed the confidence intervals for probabilities of CI coverage using the same methods as those for probabilities of convergence.
-
-The true values for terminal year SSB and fishing mortality rate vary among simulations.
-For the $i$th simulated data the relative error for terminal year value $\theta_i$ provided from the fitted estimation model is
+We also calculated the root mean square error (RMSE) and estimated probability of coverage of constructed 95\% CIs for $\widehat \beta_E$ and $\widehat \beta_M$ for EMs that estimated these parameters.
+We constructed the CIs for probabilities of CI coverage using the same methods as those for probabilities of convergence.
+
+The true values for terminal year SSB and $F$ vary among simulations.  For the $i$th simulated data the relative error for terminal year value $\theta_i$ provided from the fitted estimation model is
 $$
@@
 $$
-For ME and RME of estimators, we constructed 95\% confidence intervals using the binomial distribution approach as in @stockmiller21 and @milleretal_inreview1.
-Because of the inverse relationship of the the confidence interval and the null hypothesis significance test, when the confidence interval contains zero we do not have evidence of bias from the simulation study.
+For ME and MRE of estimators, we constructed 95\% CIs using the binomial distribution approach as in @stockmiller21 and @milleretal_inreview1.
+Because of the inverse relationship of the CI and the null hypothesis significance test, when the CI contains zero we do not have evidence of bias from the simulation study.
 
@@
 @milleretal_inreview1 found inferences are most reliable for distinguishing process error sources and stock-recruit relationships in scenarios with lower observation error for indices and age composition data and with temporal contrast in fishing pressure.
-Our expectation is that inferences for covariate effects on $M$ would require data to be at least that informative and inspection of the comprehensive set of results across all OMs generally confirms this (See supplemental Materials).
-Therefore, we restrict our attention to these OM scenarios with more informative data in the main paper (temporal contrast in fishing and lower population observation error).
+Our expectation is that inferences for covariate effects on $M$ would require data to be at least that informative and inspection of the comprehensive set of results across all OMs generally confirms this (See Supplemental Materials).
+Therefore, we restrict our attention to these OM scenarios with more informative data (temporal contrast in fishing and lower population observation error).
 
@@
 R+M EMs generally did not converge with high frequency except for R+S OMs with low covariate uncertainty and where the EMs assumed median $M$ as known.
-There was no strong effect of the treatment of median M (known or estimated) on convergence for any of the other EMs.
-The largest effect of the size of the true covariate effect ($\beta_E$) on convergence was observed for R and R+M EMs that estimated covariate effects in R and R+M OMs with high uncertainty in covariate observations ($\sigma_e$) and high variability in covariate process errors ($\sigma_E$).
-There was an increase in convergence frequency with increased variability in covariate process errors for many EMs in OMs with large covariate uncertainty.
+There was generally little effect of the treatment of median $M$ (known or estimated) on convergence for any of the other EMs.
+The largest impact of the size of the true covariate effect ($\beta_E$) on convergence was observed for R and R+M EMs that estimated covariate effects in R and R+M OMs with high uncertainty in covariate observations ($\sigma_e$) and high temporal variability in the latent covariate ($\sigma_E$).
+There was an increase in convergence frequency with increased temporal variability in the covariate for many EMs in OMs with large covariate uncertainty.
 
@@
 We only present results for EMs where the median $M$ rate parameter was estimated because results differed little whether this parameter was estimated or assumed known (see Figures \ref{aic_Rom} to \ref{aic_RMom}).
-When there is temporal contrast in fishing pressure and lower uncertainty in population observations, the only OM scenarios where AIC was consistently accurate for both covariate effect and process error configurations were R OMs with high variability in the covariate and low uncertainty in observations of the covariate (Figure \ref{aic}).
+When there was temporal contrast in fishing pressure and lower uncertainty in population observations, the only OM scenarios where AIC was consistently accurate for both covariate effect and process error configurations were R OMs with high variability in the covariate and low uncertainty in observations of the covariate (Figure \ref{aic}).
 R+M OMs with those same covariate attributes were only accurate for the covariate effect (low Type I and II errors) and the incorrect process error almost always chosen was R rather than R+S.
@@
 
-When there is temporal contrast in fishing pressure, lower uncertainty in population observations, low uncertainty in environmental observations, and larger temporal contrast in the simulated true environmental covariate, we observed generally little or no bias in estimation of covariate effects ($\widehat \beta_E$) across all EM and OM process error assumptions, both EM configurations of median $M$, and all true covariate effect sizes (Figure \ref{beta_E_bias}, rows 3 and 4).
+When there was temporal contrast in fishing pressure, lower uncertainty in population observations, low uncertainty in environmental observations, and larger temporal contrast in the simulated true environmental covariate, we observed little or no bias in estimation of covariate effects ($\widehat \beta_E$) across all EM and OM process error assumptions, both EM configurations of median $M$, and all true covariate effect sizes (Figure \ref{beta_E_bias}, rows 3 and 4).
 The only exception to this was when R EMs with median $M$ estimated were fit to R+S OMs with those configurations.
@@
 For R+S OMs, EMs with the matching process error configuration showed little evidence of bias in estimating $\beta_E$ across a range of covariate and covariate observation configurations.
+R+S and R+M EMs generally performed similarly when fitted to either R+S or R+M OMs except R+S OMs with high covariate uncertainty and low covariate variability (Figure \ref{beta_E_bias}, rows 5 and 6 of middle columns).
 
 Bias of Hessian-based standard error estimation for $\widehat \beta_E$ was also close to zero in the same OM scenarios where bias in estimation of $\beta_E$ itself was close to zero (Figure \ref{se_beta_E_bias}, rows 3 and 4).
-However, in these high information OMs bias of standard error estimation was observed in for several EMs that did not have the matching R+S and R+M process error configuration.
+However, in these high information OMs, bias of standard error estimation was observed for several EMs that did not have the matching R+S and R+M process error configuration.
 Estimation of standard errors was essentially unreliable for most EMs when OMs had higher uncertainty in covariate observations (Figure \ref{se_beta_E_bias}, rows 5 to 8).
 
-Despite the reliable estimation of covariate effects and standard errors with low uncertainty in environmental observations, and larger temporal contrast in the simulated true environmental covariate, we found poor confidence interval coverage for many EMs fitted to R+S and R+M OMs (Figure \ref{beta_E_CI_coverage}, rows 3 and 4).
-For example, R+M EMs fit to R+M OMs showed little bias in estimation of the covariate effect and standard errors, but confidence interval coverage was negatively biased (rows 3 and 4 in right two columns of Figures \ref{beta_E_bias}, \ref{se_beta_E_bias}, and \ref{beta_E_CI_coverage}.
-However, there appears to be a positive correlation of the covariate and standard error estimates such that estimates less than the true value had negatively biased standard error estimates which would make confidence intervals too small (Figure \ref{ex_lm_beta_E_SE_beta_E}).
-Confidence interval coverage was most reliable for R+S OMs when the EMs had the matching process error configuration (Figure \ref{beta_E_CI_coverage}, middle 2 columns).
+Despite the reliable estimation of covariate effects and standard errors with low uncertainty in environmental observations, and larger temporal contrast in the simulated true environmental covariate, we found poor CI coverage for many EMs fitted to R+S and R+M OMs (Figure \ref{beta_E_CI_coverage}, rows 3 and 4).
+For example, R+M EMs fit to R+M OMs showed little bias in estimation of the covariate effect and standard errors, but CI coverage was negatively biased (rows 3 and 4 in right two columns of Figures \ref{beta_E_bias}, \ref{beta_E_CI_coverage}, and \ref{se_beta_E_bias}).
+However, there appears to be a positive correlation of the covariate and standard error estimates such that estimates less than the true value had negatively biased standard error estimates which would make CIs too small (Figure \ref{ex_lm_beta_E_SE_beta_E}).
+CI coverage was most reliable for R+S OMs when the EMs had the matching process error configuration (Figure \ref{beta_E_CI_coverage}, middle 2 columns).
 
@@
 When there is temporal contrast in fishing pressure, lower uncertainty in population observations, all EMs fit to R and R+M OMs showed little evidence of bias for $\widehat \beta_M$  (Figure \ref{beta_M_bias}, left and right sets of columns).
-For R+S OMs with those conditions, EMs with the matching process error assumption showed little evidence of bias.
-
-We also found little or no bias of Hessian-based standard error estimation for $\beta_M$ in the many of the same OM scenarios where bias of $\widehat \beta_M$ was close to zero except for some EMs with the process error source mis-specified or where the OM simulated the strongest covariate effect (Figure \ref{se_beta_M_bias}).
+For R+S OMs with those conditions, only EMs with the matching process error assumption showed little evidence of bias.
+
+We also found little or no bias of Hessian-based standard error estimation for $\beta_M$ in many of the same OM scenarios where bias of $\widehat \beta_M$ was close to zero except for some EMs with the process error source mis-specified or where the OM simulated the strongest covariate effect (Figure \ref{se_beta_M_bias}).
 Like $\widehat \beta_M$, reliability of standard error estimation in R+S OMs required the EMs to have the matching process error assumption.
 
-However, we found confidence interval coverage to be generally unreliable even for OM and EM combinations where bias in $\widehat \beta_M$ and its standard error estimators was negligible, much like that for $\widehat \beta_E$ (Figure \ref{beta_M_CI_coverage}).
-In the same EM OM combination we investigated above for $\widehat \beta_E$, we observed an opposite negative correlation of $\widehat \beta_M$ and its standard error estimates (Figure \ref{ex_lm_beta_M_SE_beta_M}), which would result in confidence intervals being too narrow when $\widehat \beta_M$ values are larger than average.
+we found CI coverage to be unreliable for many EM-OM combinations even for OM and EM combinations where bias in $\widehat \beta_M$ and its standard error estimators was negligible, much like that for $\widehat \beta_E$ (Figure \ref{beta_M_CI_coverage}).
+In the same EM OM combination we investigated above for $\widehat \beta_E$, we observed an opposite negative correlation of $\widehat \beta_M$ and its standard error estimates (Figure \ref{ex_lm_beta_M_SE_beta_M}), which would result in CIs being too narrow when $\widehat \beta_M$ values are larger than average.
+The exceptions to the general poor CI coverage were for all EMs fit to several R OMs with low covariate uncertainty and covariate variability (Figure \ref{beta_M_CI_coverage}, first two columns) and for R+S EMs in many R+S OMs (Figure \ref{beta_M_CI_coverage}, middle two columns) .
 
@@
 
-The EMs with R and R+S process error configurations with median M assumed known and covariate effects not estimated will have terminal year M correctly specified when fitted to R and R+S OMs with no covariate effect because $M$ is constant in both the EM and OM at the same value.
+The EMs with either R or R+S process error configurations with median $M$ assumed known and covariate effects not estimated will have terminal year $M$ correctly specified when fitted to either R or R+S OMs with no covariate effect because $M$ is constant in both the EM and OM at the same value.
 We exclude those OM-EM combinations from our results here.
 
-In many of the OMs with temporal contrast in fishing pressure and lower uncertainty in population observations, there was little difference in bias for terminal year M among EMs whether the median $M$ parameter ($\beta_M$) was estimated (Figure \ref{terminal_M_bias}).
-However, when there were differences, bias in terminal year M was closer to zero when the EM assumed $\beta_M$ known, As would be expected.
+In many of the OMs with temporal contrast in fishing pressure and lower uncertainty in population observations, there was little difference in bias for terminal year $M$ among EMs whether the median $M$ parameter ($\beta_M$) was estimated (Figure \ref{terminal_M_bias}).
+However, when there were differences, bias in terminal year $M$ was closer to zero when the EM assumed $\beta_M$ known, as would be expected.
 For R OMs, all estimating models exhibited little evidence of bias except when OMs simulated the largest covariate effect ($\beta_E = 0.5$).
 However, bias estimates were at worst around 10\% (Figure \ref{terminal_M_bias}, left columns).
-For R+S OMs, the R+S EMs generally exhibited the least evidence for bias and R+M EMs generally estimated terminal M lower than the true value, particularly when observation uncertainty in the covariate was lower.
+For R+S OMs, the R+S EMs generally exhibited the least evidence for bias and R+M EMs generally estimated terminal $M$ lower than the true value, particularly when observation uncertainty in the covariate was lower.
 For R+M OMs, many of the EM configurations provided little bias in terminal $\widehat M$, but the biases were generally further from zero than the results for the R OMs (Figure \ref{terminal_M_bias}, right columns).
-Accuracy of terminal year $\widehat M$ was similar for all of the EMs when fitted to R+M OMs (Figure \ref{terminal_M_rmse}).
-For R+S OMs, accuracy was a bit better for R+S EM than others.
-As would be expected accuracy of estimate where median $M$ was assumed known was better than when it was estimated, when there were differences.
+
+Accuracy (lower RMSE) of terminal year $\widehat M$ was similar for all of the EMs when fitted to R+M OMs (Figure \ref{terminal_M_rmse}).
+For R+S OMs, accuracy was generally better for R+S EM than other process error assumptions, particularly when $\beta_E$ was also estimated.
+Accuracy for R+S EMs was also generally as good as other EM process error assumptions even when the OM process error type was different.
+As would be expected, the accuracy of terminal year $\widehat M$ where median $M$ was assumed known was better than when it was estimated, when there were differences.
 
@@
 In OMs with temporal contrast in fishing pressure and lower uncertainty in population observations, we found little bias in estimates of terminal year SSB for all EM configurations when fitted to R or R+M OMs (Figure \ref{terminal_ssb_bias}).
-We also found little bias for R+S OMs when the EMs assumed the correct process error configuration.
-The worst bias in terminal year SSB estimation we found was in R+S OMs for R EMs with $\beta_M$ estimated.
-Like the bias results, accuracy as measured by RMSE was similar for many of EMs when fit to R and R+M OMs.
-Accuracy was generally worse when EMs were fit to R+S OMs (Figure \ref{terminal_ssb_rmse}).
+We also found little or no evidence of bias for R+S OMs when the EMs assumed the correct process error configuration.
+We found the worst bias in terminal year SSB estimation for R EMs with $\beta_M$ estimated when fitted to R+S OMs, and R+M EMs also had more bias than R+S EMs in these OMs.
+Like the bias results, accuracy, as measured by RMSE, was similar for many of EMs when fit to R and R+M OMs.
+Accuracy was generally worse (higher RMSE) when EMs were fit to R+S OMs, and the worst accuracy occurred when the EM process error assumption did not match that of the OM (Figure \ref{terminal_ssb_rmse}).
 
@@
 Our simulation study demonstrated that estimation of environmental effects on $M$ is possible and reliable in certain scenarios even when the process error was misspecified (e.g., R and R+S EMs fit to R+M OM).
-In many of these same OMs frequency of convergence of model fits did not appears to suffer when covariate effects on $M$ were estimated even when there was no effect simulated.
-However, these scenarios are information rich in that there was contrast in population size (due to changes in fishing pressure) and the covariate affecting the population, and low uncertainty in population and covariate observations.
+In many of these same OMs, frequency of convergence of fitted models did not appear to suffer when covariate effects on $M$ were estimated even when there was no effect simulated.
+However, these scenarios are information rich in that there was contrast in population size (due to changes in fishing pressure) and the covariate affecting the population, and there was low uncertainty in population and covariate observations.
 Previous research has shown that estimation of a constant $M$ parameter requires contrast in time series and informative data [@leeetal11], so it is no surprise that estimation of these effects also requires relatively good information via more precise observations and higher contrast in the covariate time series.
 
-On the other hand, AIC could only reliably detect covariate effects for R and R+M OMs with contrast in covariates and low covariate uncertainty and confidence interval coverage for the covariate effect was often biased even when bias in the estimators of the effect and the standard error were negligible.
-@cadiganetal24 found confidence interval coverage to be biased for SSB and fishing mortality estimation in a state-space model in some scenarios, but they attributed the poor coverage to bias in Hessian-based standard error estimation and their simulations held any process error random effects constant.
-The coverage bias we observed, at least for some OM-EM combinations, may be a related to correlation of the estimators of the covariate effect and the corresponding standard error and therefore consideration of other methods of calculating confidence intervals may be warranted (e.g., those based on profile likelihood and/or MCMC sampling of log-likelihood surface).
-
-@milleretal_inreview1 investigated R+M OMs with two levels of $M$ process error variability ($\sigma_M \in {0.1,0.5}$ and only found AIC able to accurately distinguish R+M process errors with the higher level of process error variability ($\sigma_M = 0.5$).
+Even though estimation of covariate effects was unbiased in many scenarios, AIC could only reliably detect covariate effects for R and R+M OMs with contrast in covariates and low covariate uncertainty.
+In those scenarios where the covariate effect could be detected, CI coverage for the covariate effect was often biased even when there was little or no bias in the estimators of the effect and its standard error.
+@cadiganetal24 found CI coverage to be biased for SSB and $F$ estimation in a state-space model in some scenarios, but they attributed the poor coverage to bias in Hessian-based standard error estimation, and their simulations held any process error random effects constant.
+The coverage bias we observed, at least for some OM-EM combinations, may be related to correlation of the estimators of the covariate effect and the corresponding standard error and therefore consideration of other methods of calculating CIs may be warranted (e.g., those based on profile likelihood and/or Monte Carlo sampling of the log-likelihood surface).
+
+@milleretal_inreview1 investigated R+M OMs with two levels of $M$ process error variability ($\sigma_M \in \{0.1,0.5\}$) and only found AIC able to accurately distinguish R+M process errors with the higher level of process error variability ($\sigma_M = 0.5$).
 We assumed $\sigma_M = 0.3$, intermediate to the values investigated by @milleretal_inreview1, and found process error inferences unreliable for the source of process error, indicating that the level of variability required for detecting $M$ process errors must be greater than $\sigma_M = 0.3$, but may still be less than 0.5.
@@
 
-@derisoetal08 attempted to estimate process errors as well as covariate effects with $M$ for Pacific herring, but similarly found no variability in $M$, suggesting a there was too much uncertainty in the available observations relative to the true temporal variability in $M$.
+@derisoetal08 attempted to estimate process errors as well as covariate effects with $M$ for Pacific herring, but similarly found no variability in $M$, suggesting there was too much uncertainty in the available observations relative to the true temporal variability in $M$.
 Given that we found covariate effect inferences using R EMs was reliable in R+M OMs with apparently little variability in $M$ process errors, the findings of @derisoetal08 on covariate effects for Pacific herring would presumably also be robust to true low variability in $M$.
@@
 
-Any bias or poor accuracy for annual SSB estimation was primarily a function of whether or not the median $M$ parameter was estimated or known and the types of process errors than the treatment of the covariate effects on $M$.
+Any bias or poor accuracy for annual SSB estimation was primarily a function of whether or not the median $M$ parameter was estimated or known and the types of process errors, rather than the treatment of the covariate effects on $M$.
 For example,  we found estimation of SSB was better when the EM had the process error correctly specified for R+S OMs.
@@
 
+The R+S and R+M EMs both include process error for the survival of cohorts and would be expected to perform similarly, and they did in our simulations when the OM and EM matched.
+However, we found that the biases using R+M EMs for R+S OMs were generally worse than using R+S EMs for R+M OMs.
+Additionally, there are implications for biological reference points for R+M EMs [e.g., @legaultpalmer16] that are not present with R+S EMs.
+So, we recommend following the suggestion from @lietal24 to prefer R+S EMs over R+M OMs unless strong biological evidence is present to support a particular R+M OM.
+Such support could be found through both biological understanding [e.g., @trijouletetal20] as well as statistical properties such as a large delta AIC for R+M compared to R+S associated with greater temporal variability in natural mortality [@milleretal_inreview1].
+
 The ability to accurately infer covariate effects on $M$ in some realistic situations indicates that such investigations may be fruitful.
-Ability to make inferences could improve further when WHAM is extended to incorporate tagging data as expected [@milleretal_inreview].
+Ability to make inferences could improve further when WHAM is extended to incorporate tagging data [@milleretal_inreview].
 Tagging data can greatly inform natural mortality estimation [@pollocketal91;@hampton00], and this impact on $M$ estimation should also apply to estimation of covariate effects or unexplained temporal variation in $M$.
-Given our findings coupled with future WHAM development, we expect investigations of and accounting for covariate effects on $M$ to become more common within the fisheries stock assessment process.
+Given our findings and planned future WHAM development, we expect investigations of and accounting for covariate effects on $M$ to become more common within the fisheries stock assessment process.
+At the same time, it will be equally important to conduct research that will improve our understanding of how best to measure the depletion of stocks and determine catch advice for these stocks with covariate effects on $M$.
 
@@
 \end{center}
-\caption{Estimated probability of fits providing Hessian-based standard errors for EMs with alternative process error assumptions, treatment of median natural mortality ($e^\beta_M$ known or estimated), and treatment of covariate effect ($\beta_E = 0$ or estimated).
+\caption{Estimated probability of fits providing Hessian-based standard errors for EMs with alternative process error assumptions, treatment of median natural mortality ($\beta_M$ known or estimated), and treatment of covariate effect ($\beta_E = 0$ or estimated).
 The OMs have R (left) and R+S (middle), or R+M (right) process error structures, alternative configurations of covariate time series structure and levels of observation uncertainty (rows), and three levels of true covariate effect on median natural mortality (x axis).
@@
 \end{center}
-\caption{Median error (ME) of estimates of environmental effect on natural mortality $\beta_E$ from fitting EMs with alternative process error assumptions and treatment of median natural mortality ($e^\beta_M$ known or estimated).
+\caption{Median error (ME) of estimates of environmental effect on natural mortality $\beta_E$ from fitting EMs with alternative process error assumptions and treatment of median natural mortality ($\beta_M$ known or estimated).
 All OMs had low observation error and contrast in fishing mortality.
@@
 \end{center}
-\caption{Probability of 95\% confidence interval for $\beta_E$ containing the true value for EMs with alternative process error assumptions and treatment of median natural mortality ($e^\beta_M$ known or estimated).
+\caption{Probability of 95\% confidence interval for $\beta_E$ containing the true value for EMs with alternative process error assumptions and treatment of median natural mortality ($\beta_M$ known or estimated).
 All OMs had low observation error and contrast in fishing mortality.
@@
 
-For R OMs, R EMs converged frequently except when there was contrast in fishing and covariate effects ($\beta_E$) were estimated and covariate observation uncertainty ($\sigma_e$) was high (Figure \ref{convergence_Rom}).
-There was little difference in convergence for R EMs whether median M was estimated or not.
-When R OMs had contrast in fishing, convergence of R+M EMs was generally better when the EM estimated median M, whereas in OMs with constant fishing, convergence was better when treating median M as known.
-Convergence of R+S EMs was generally worse than other process error configurations for R OMs, and there was no difference between EMs that estimated median M or not except when fishing pressure was constant where better convergence occurred when M was treated as known.
+<!-- For R OMs, R EMs converged frequently except when there was contrast in fishing and covariate effects ($\beta_E$) were estimated and covariate observation uncertainty ($\sigma_e$) was high (Figure \ref{convergence_Rom}).
+There was little difference in convergence for R EMs whether median $M$ was estimated or not.
+When R OMs had contrast in fishing, convergence of R+M EMs was generally better when the EM estimated median $M$, whereas in OMs with constant fishing, convergence was better when treating median $M$ as known.
+Convergence of R+S EMs was generally worse than other process error configurations for R OMs, and there was no difference between EMs that estimated median $M$ or not except when fishing pressure was constant where better convergence occurred when $M$ was treated as known.
 Effects of true covariate effect size on convergence were apparent in some R OMs, but there did not appear to be any consistent patterns.
-
-For R+S OMs, R+M EMs generally showed the worse probability of convergence that EMs with R and R+S configurations (Figure \ref{convergence_RSom}).
+-->
+
+<!-- For R+S OMs, R+M EMs generally showed the worse probability of convergence that EMs with R and R+S configurations (Figure \ref{convergence_RSom}).
 Convergence probability was high for all R and R+S EMs when covariate observation uncertainty was low.
 Although probability of convergence was generally greater when covariate effects were not estimated, there were no strong trends in convergence for any of the EMS with increased true effect size.
-
-For R+M OMs, R+S EMs generally converged less frequently than those with R and R+M configurations (Figure \ref{convergence_RMom}).
+-->
+
+<!-- For R+M OMs, R+S EMs generally converged less frequently than those with R and R+M configurations (Figure \ref{convergence_RMom}).
 Convergence probability for R OMs was generally the greatest of the three process error configurations.
@@
 However, there was also not strong effects of estimating the covariate effect on convergence except for some OMs with higher covariate observation uncertainty.
-Like R OMs, when R+M OMs had contrast in fishing, convergence of R+M EMs was generally better when the EM estimated median M and the convergence was better when treating median M as known when the OM had constant fishing.
+Like R OMs, when R+M OMs had contrast in fishing, convergence of R+M EMs was generally better when the EM estimated median $M$ and the convergence was better when treating median $M$ as known when the OM had constant fishing.
+-->
 
@@
 We also observe little or no evidence of bias (confidence intervals include 0) for R or R+S EMs for any OMs even when the OMs included an effect of the covariate on natural mortality ($\beta_E > 0$).
-Including process error on M (R+M) produces more variability in errors for R and R+S EMs than including increased level of effect of the covariate on M.
-For R and R+S OMs, errors in annual M for R+S EMs were less variable than those for R EMs.
--->
-
-<!-- R+M EMs fit to R OMs with constant fishing mortality, or a step change in fishing mortality and lower uncertainty in index and age composition observations exhibited no bias in annual estimation of M indicating that the variance of the estimated random effects for M in these EMs collapsed to 0.
+Including process error on $M$ (R+M) produces more variability in errors for R and R+S EMs than including increased level of effect of the covariate on $M$.
+For R and R+S OMs, errors in annual $M$ for R+S EMs were less variable than those for R EMs.
+-->
+
+<!-- R+M EMs fit to R OMs with constant fishing mortality, or a step change in fishing mortality and lower uncertainty in index and age composition observations exhibited no bias in annual estimation of $M$ indicating that the variance of the estimated random effects for $M$ in these EMs collapsed to 0.
 R+M EMs were the only EMs to exhibit differences in the sign of the median errors across the time series.
@@
 However, whether there was evidence of bias, depended on the uncertainty in covariate observations and the degree of variability in the latent covariate.
-We observed negative median relative errors in annual M estimates for R+M EMs fit to R+S OMs with lower uncertainty in index and age composition observations across all levels of simulated effects of the covariate, but evidence of bias was strongest at the beginning of the time series.
+We observed negative median relative errors in annual $M$ estimates for R+M EMs fit to R+S OMs with lower uncertainty in index and age composition observations across all levels of simulated effects of the covariate, but evidence of bias was strongest at the beginning of the time series.
 -->
@@
 The same variability occurred for R+S OMs, but strong bias was estimated for R and R+M EMs.
-Annual M estimation was most reliable when OMs had step changes in fishing mortality and lower uncertainty in index and age composition observations.
-For R and R+M OMs with those characteristics, all EMs generally provided accurate estimation of annual M, but only R+S EMs provided accurate estimation for R+S OMs.
-Little or no bias in annual M estimation was observed for R+S EMs across all OM process error assumptions as long as there was a step change in fishing mortality and lower uncertainty in indices and age composition.
+Annual $M$ estimation was most reliable when OMs had step changes in fishing mortality and lower uncertainty in index and age composition observations.
+For R and R+M OMs with those characteristics, all EMs generally provided accurate estimation of annual $M$, but only R+S EMs provided accurate estimation for R+S OMs.
+Little or no bias in annual $M$ estimation was observed for R+S EMs across all OM process error assumptions as long as there was a step change in fishing mortality and lower uncertainty in indices and age composition.
 -->

### 2025-08-08 Ecov_study/mortality/manuscript/manuscript.tex
saving work on project_0 revisions

@@
   pdftitle={Factors affecting inferences on natural mortality and associated environmental effects in state-space age-structured assessment models},
-  pdfauthor={Timothy J. Miller1,2; Greg Britten3; Elizabeth N. Brooks2; Gavin Fay4; Alex Hansell2; Christopher M. Legault2; Brandon Muffley5; John Wiedenmann6},
+  pdfauthor={Timothy J. Miller1,2; Gregory L. Britten3; Elizabeth N. Brooks2; Gavin Fay4; Alexander C. Hansell2; Christopher M. Legault2; Brandon Muffley5; John Wiedenmann6},
   hidelinks,
@@
 \author{Timothy J.
-Miller\textsuperscript{1,2} \and Greg Britten\textsuperscript{3} \and Elizabeth N.
-Brooks\textsuperscript{2} \and Gavin Fay\textsuperscript{4} \and Alex Hansell\textsuperscript{2} \and Christopher M.
+Miller\textsuperscript{1,2} \and Gregory L.
+Britten\textsuperscript{3} \and Elizabeth N.
+Brooks\textsuperscript{2} \and Gavin Fay\textsuperscript{4} \and Alexander C.
+Hansell\textsuperscript{2} \and Christopher M.
 Legault\textsuperscript{2} \and Brandon Muffley\textsuperscript{5} \and John Wiedenmann\textsuperscript{6}}
-\date{09 June, 2025}
+\date{18 June, 2025}
 
@@
 \(^3\)Biology Department, Woods Hole Oceanographic Institution, 266 Woods Hole Rd. Woods Hole, MA, USA\\
-\(^4\)Department of Fisheries Oceanography, School for Marine Science and Technology, University of Massachusetts Dartmouth, 836 S Rodney French Boulevard, New Bedford, MA 02740, USA\\
+\(^4\)Department of Fisheries Oceanography, School for Marine Science and Technology, University of Massachusetts Dartmouth, 836 S Rodney French Boulevard, New Bedford, MA 02744, USA\\
 \(^5\)Mid-Atlantic Fishery Management Council, 800 North State Street, Suite 201, Dover, DE 19901 USA\\
@@
 Because application of these methods are considered best practice and recommended for the next generation of stock assessment models \citep{hoyleetal22, punt23}, it is expected their use will only grow globally.
-An appeal of state-space models lies in their formulation treating latent population characteristics as statistical time series with periodic observations that also may have error due to sampling or other sources of measurement error and therefore separating these sources of biological and measurement variability.
+An appeal of state-space models lies in their separation of sources of biological and measurement variability by treating latent population characteristics as statistical time series with periodic observations measured with error.
 Through advances in computational capacity, we can use sophisticated numerical approaches to estimate model parameters as mixed effects \citep{thorsonminto15, kristensenetal16}.
@@
 State-space stock assessment models, with non-linear functions of latent processes and numerous observation types with different probability distribution assumptions represent one of most complex classes of state-space models.
-The literature on the effects of various factors on reliabilty of inferences from state-space assessment models is growing \citep{lietal24, milleretal_inreview1}.
-The importance of contrast in population size and fishing mortality and quality of data used to fit assessment models including the state-space variety is known \citep{magnussonhilborn07, milleretal_inreview1}.
-Furthermore, estimation of natural mortality, and even temporal variablity is possible in many scenarios \citep{leeetal11, cadigan16, millerhyun18, milleretal_inreview1}.
+The literature on the effects of various factors on reliability of inferences from state-space assessment models is growing \citep{lietal24, milleretal_inreview1}.
+The importance of contrast in population size and fishing mortality (\(F\)) and quality of data used to fit assessment models including the state-space variety is known \citep{magnussonhilborn07, milleretal_inreview1}.
+Furthermore, estimation of natural mortality (\(M\)), and even temporal variability in \(M\) is possible in many scenarios \citep{leeetal11, cadigan16, millerhyun18, milleretal_inreview1}.
 
@@
 While understanding drivers of recruitment is important particularly for subsequent effects on reference points, recruitment in short-term projections typically has little impact on the exploitable biomass in the first few projection years.
-However, assumptions for natural mortality (\(M\)) have immediate and larger effects on projected biomass because they affect the abundances at older age classes at the end of the data time series that constitute spawning biomass and catch \citep{brodziaketal08, stocketal21}.
+However, assumptions for \(M\) have immediate and larger effects on projected biomass because they affect the abundances at older age classes at the end of the data time series that constitute spawning biomass and catch \citep{brodziaketal08, stocketal21}.
 
@@
 Temporal variation in \(M\) is less studied than recruitment, but its importance for explaining variability in observations has been demonstrated in state-space assessment models for Atlantic cod and yellowtail flounder \citep{cadigan16, stocketal21}.
-\citet{derisoetal08} also demonstrated the importance of several factors affecting Pacific herring natural mortality.
+\citet{derisoetal08} also demonstrated the importance of several factors affecting \(M\) for Pacific herring.
 
 Assessment models could include temporal variation in many aspects of population dynamics or how observations are related to the population.
-For example the Woods Hole Assessment Model (WHAM) can include process errors treated as random effects for transition in cohorts over time, catchability for indices of abundance, selectivity of fishing fleets or indices, movement between regions, or in \(M\) \citep{stockmiller21, milleretal_inreview}.
+For example the Woods Hole Assessment Model (WHAM) can include process errors treated as random effects for transition in cohorts over time (hereafter referred to as apparent survival), catchability for indices of abundance, selectivity of fishing fleets or indices, movement between regions, or in \(M\) \citep{stockmiller21, milleretal_inreview}.
 However, misspecified temporal population process errors could lead to biased population and stock status estimation, and, therefore, poor fisheries management decisions \citep{legaultpalmer16, szuwalskietal18}.
-Studies of the reliability of inferences we might make regarding the presence of temporal variability in \(M\) are limited.
-\citet{milleretal_inreview1} found AIC could accurately distinguish process errors in the changes in abundance for cohorts, but not for those specifically due to \(M\) except when uncertainty in population observations (indices, catch and age composition) was low and there was greater temporal variation in natural mortality.
-In their simulation studies looking a models with multiple sources of process error, \citet{lietal24} found including more sources of process error than existed in the operating model was a better model-building approach than excluding them a priori.
-
-Here we conduct a simulation study with operating models (OMs) varying by degree of observation error uncertainty, sources of process error, fishing history, temporal variation in environmental covariates, and magnitude of the effect of the covariate on natural mortality.
-The simulated observations from these OMs are fitted with estimating models that make alternative assumptions for sources of process error, and whether median M and covariate effects are estimated.
-We evaluate the effects of these factors on convergence of fitted models, whether Akaike's information criterion (AIC) can correctly determine the correct source of process error and correct assumption about covariate effects on natural mortality, and the degree of bias in inferences for relevant parameters and outputs of the assessment model.
+Studies of the reliability of inferences regarding the presence of temporal variability in \(M\) are limited.
+\citet{milleretal_inreview1} found AIC could accurately distinguish process errors in apparent survival, but not for those specifically due to \(M\) except when uncertainty in population observations (indices, catch and age composition) was low and there was greater temporal variation in \(M\).
+In their simulation studies looking at models with multiple sources of process error, \citet{lietal24} found including more sources of process error than existed in the operating model was a better model-building approach than excluding them a priori.
+
+Here we conduct a simulation study with operating models (OMs) varying by degree of observation error uncertainty, sources of process error, fishing history, temporal variation in environmental covariates, and magnitude of the effect of the covariate on \(M\).
+The simulated observations from these OMs are fitted with estimating models that make alternative assumptions for sources of process error, and whether median \(M\) and covariate effects are estimated.
+We evaluate the effects of these factors on convergence of fitted models, whether Akaike's information criterion (AIC) can determine the correct source of process error and correct assumption about covariate effects on \(M\), and the accuracy for estimators of relevant parameters and stock size and harvest rates derived from the assessment model.
 
@@
 
-All of our analyses used the Woods Hole Assessment Model (WHAM) to construct both OMs and EMs \citep{millerstock20, stockmiller21, milleretal_inreview}.
+Our analyses used the Woods Hole Assessment Model (WHAM) to construct both OMs and EMs \citep{millerstock20, stockmiller21, milleretal_inreview}.
 The WHAM package has been used extensively to configure OMs and EMs for several other simulation studies \citep{stocketal21, legaultetal23, lietal24, brittenetal_inreview, lietal_inreview_a} and is used to assess many commercially important stocks in the Northeast U.S.
 \citep[e.g.,][]{nefsc22, nefsc22a, nefsc24}.
-We used \href{https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3}{version 1.0.6.9000, commit 77bbd94} for to generate all results.
-
-We completed a simulation study with a 288 operating models.
-The factors defining the configuration of each operating model which are described in detail in subsequent sections include source of population process error (3 levels), index and catch observation uncertainty (2 levels), environmental covariate uncertainty (2 levels), latent environmental covariate process stochasticity (4 levels), and fishing history (2 levels).
+We used \href{https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3}{version 1.0.6.9000, commit 77bbd94} to generate all results.
+
+We completed a simulation study with 288 operating models.
+The factors defining the configuration of each operating model, described in detail in subsequent sections, include source of population process error (3 levels), index and catch observation uncertainty (2 levels), environmental covariate uncertainty (2 levels), temporal variation in the latent environmental covariate (4 levels), and fishing history (2 levels).
 We simulated 100 data sets for each operating model that included simulations of process errors.
@@
 For each simulated data set we fit a set of 12 estimating models (EMs).
-The factors that distinguish the estimating models which are also described in detail below include source of population process error type (3 levels) whether (median) \(M\) was estimated or assumed known (2 levels), and whether the effect of the environmental covariate on \(M\) was estimated or not (2 levels).
-
-The sources of population process error that were used in the OMs or assumed in the EMs were on recruitment only (R), recruitment and changes in cohort abundance over time (R+S), or recruitment and \(M\) (R+M).
-We did not use the log-normal bias-correction feature for process errors or observations described by \citet{stockmiller21} for operating and EMs \citep{lietal_inreview}.
+The factors that distinguish the estimating models, also described in detail below, include source of population process error type (3 levels) whether (median) \(M\) was estimated or assumed known (2 levels), and whether the effect of the environmental covariate on \(M\) was estimated or not (2 levels).
+
+The sources of population process error that were used in the OMs or assumed in the EMs were on recruitment only (R), recruitment and apparent survival (R+S), or recruitment and \(M\) (R+M).
+We did not use the log-normal bias-correction feature for process errors or observations described by \citet{stockmiller21} for OMs and EMs \citep{lietal_inreview}.
 Simulations were all carried out on the University of Massachusetts Green High-Performance Computing Cluster.
@@
 with marginal mean \(\mu_E=0\) and variance \(\sigma^2_E\).
-The four configuration of the latent environmental covariate in the operating models assume the one of two values for the marginal standard deviation \(\sigma_E \in \{0.1, 0.5\}\) and for the autocorrelation parameter \(\rho_E \in \{0, 0.5\}\).
+The four configurations of the latent environmental covariate in the operating models assume one of two values for the marginal standard deviation \(\sigma_E \in \{0.1, 0.5\}\) and for the autocorrelation parameter \(\rho_E \in \{0, 0.5\}\).
 
@@
 The standard deviation of the environmental observations in the operating models is one of two values \(\sigma_e \in \{0.1, 0.5\}\).
-Figure \ref{om_ecov_example} provides example simulations of the latent process and observations under the alternative configurations.
+Figure \ref{om_ecov_example} provides example simulations of the latent covariate and observations under the alternative configurations.
 
@@
 Many of the characteristics of the population biology and structure including the age classes (10 age classes (ages 1 to 10+)), time span (40 years), maturity (Figure \ref{om_inputs_fig}, top left), growth (Figure \ref{om_inputs_fig}, top right), time of spawning (1/4 of the year), and recruitment (Figure \ref{om_inputs_fig}, bottom right) are identical to \citet{milleretal_inreview1}.
-The maturity at age is a logistic function of age with age at 50\% maturity (\(a_{50}\)) = 2.89 and slope = 0.88 and weight at age is derived from a von Beralanffy growth function where \(t_0 = 0\), \(L_\infty = 85\), and \(k = 0.3\), and a length-weight relationship
+The maturity at age is a logistic function with age at 50\% maturity (\(a_{50}\)) = 2.89 and slope = 0.88 and weight at age is derived from a von Beralanffy growth function where \(t_0 = 0\), \(L_\infty = 85\), and \(k = 0.3\), and a length-weight relationship
 \[
@@
 For R and R+S OMs and EMs, \(\varepsilon_{M,y} = 0\).
-For all R+M OMs, we assume the same standard deviation \(\sigma_M = 0.3\) and is estimated in the R+M EMs.
+For all R+M OMs, we assume the same standard deviation \(\sigma_M = 0.3\), which is estimated in the R+M EMs.
 The covariate effect is one of 3 alternative values in the operating models, \(\beta_\text{E} \in \{0,0.25,0.5\}\).
@@
 Therefore, steepness and equilibrium unfished recruitment are also constant over the time period for those OMs \citep{millerbrooks21}.
-As in \citet{milleretal_inreview1}, our assumed biological inputs and selectivity (defined below) with constant \(M\) result in equilibrium fishing mortality that reduces spawning biomass per recruit to 40\% of the unfished level is \(F_{40\%} = 0.348\).
+As in \citet{milleretal_inreview1}, our assumed biological inputs and selectivity (defined below) with constant \(M\) result in equilibrium \(F\) that reduces spawning biomass per recruit to 40\% of the unfished level is \(F_{40\%} = 0.348\).
 With an assumed unfished recruitment of \(R_0 = e^{10}\), setting \(\Fmsy = F_{40\%}\) results in a steepness of 0.69 and \(a=0.60\) and \(b = 2.4 \times 10^{-5}\).
-For R+M OMs and all OMs with covariate effects on \(M\), steepness is not constant, but we used the same \(a\) and \(b\) parameters as other operating models which equates to a steepness and \(R_0\) at the median of the time series processes for \(M\) and the covariate.
+For R+M OMs and all OMs with covariate effects on \(M\), steepness is not constant, but we used the same \(a\) and \(b\) parameters as other operating models which equates to a steepness and \(R_0\) at the median of the time series models for \(M\) and the covariate.
 
@@
 In the second scenario, the stock is fished at \Fmsy for the entire time period (40 years).
+The magnitude of the overfishing assumptions is intended to reflect estimates of overfishing for Northeast US groundfish stocks from \citet{wiedenmannetal19}.
 
@@
 This same assumption was used by \citet{milleretal_inreview1} for R+M OMs and other OMs with fishery selectivity and index catchability process errors.
-For R+S OMs, cohort temporal transition process errors were uncorrelated with \(\sigma_{2+} = 0.3\)
+For R+S OMs, apparent survival process errors were uncorrelated with \(\sigma_{2+} = 0.3\).
 
@@
 
-We define the generation of observations of total catch, aggregate indices, and corresponding age composition identical to \citet{milleretal_inreview1}.
+We define the generation of observations of aggregate (total combined across ages) catch and indices, and corresponding age composition identical to \citet{milleretal_inreview1}.
 There is a single fleet operating year round for catch observations with logistic selectivity for the fleet with \(a_{50} = 5\) and slope = 1 (Figure \ref{om_inputs_fig}, bottom left).
@@
 
-For each data set simulated from an operating model 12 estimating models were fit.
+Estimating models were fit to each of 100 simulated data sets from each operating model.
 There were three factors defining the configuration of each estimating model: 1) whether \(\beta_M\) was estimated or assumed known, 2) whether an environmental effect \(\beta_E\) was estimated or not, and 3) whether the process errors were assumed on recruitment only (R), recruitment and survival (R+S), or recruitment and \(M\) (R+M).
@@
 For example, uncorrelated R+S was assumed for both the estimating and operating model.
-However, R+M EMs did not assume M random effects were uncorrelated (\(\rho_M\) was estimated).
-The environmental covariate observations were included in all estimation models to ensure comparability of AIC.
-All fixed effects parameters for selectivity, catchability, fully-selected fishing mortality, mean recruitment, initial abundance at age, and variances for logistic-normal age composition distributions were estimated.
+However, R+M EMs did not assume \(M\) random effects were uncorrelated (\(\rho_M\) was estimated).
+The environmental covariate observations were included in all estimation models, whether effects on \(M\) were estimated or not, to ensure comparability of AIC.
+All fixed effects parameters for selectivity, catchability, fully-selected \(F\), mean recruitment, initial abundance at age, and variances for logistic-normal age composition distributions were estimated.
 Any process error variance parameters for recruitment, survival, and \(M\) were also estimated.
@@
 There are various ways to assess convergence of the fit \citep[e.g.,][]{carvalhoetal21}, but we defined successful convergence as the Hessian of the marginal log-likelihood being invertible and providing variance estimates for the fixed effects parameters as recommended by \citet{milleretal_inreview1}.
-We calculated 95\% confidence intervals for probability of convergence using the Clopper-Pearson exact method \citep{clopperpearson34, thulin14}.
+We calculated 95\% confidence intervals (CIs) for probability of convergence using the Clopper-Pearson exact method \citep{clopperpearson34, thulin14}.
 
@@
 \]
-All model fits that successfully completed the optimization were used in these results.
+All model fits that successfully completed the optimization were used for this set of analyses.
 We used all of these fits because some lack of convergence would be expected for the correct behavior of more complicated models that include process errors that did not exist in the operating model.
@@
 All results here use OM simulations with fits that satisfied the convergence criterion described above.
-We used this conditioning, to reflect how practitioners would proceed in analyses of model fits with real assessment data.
+We used this conditioning to reflect how practitioners would proceed in analyses of model fits with real assessment data.
 That is, practitioners would ensure models converged such that Hessian-based standard errors were available for all model parameter estimates.
 
-We focused on statistical behavior of estimators of the covariate effect on \(M\) (\(\widehat \beta_E\)), the estimator of the median \(M\) parameter (\(\widehat \beta_M\)), and estimators of terminal year natural mortality \(\widehat M\), spawning stock biomass \(\widehat{\text{SSB}}\), and fully-selected fishing mortality rate \(\widehat F\).
-In preliminary analyses we results of the estimators of all the annual values for \(M\), SSB and fishing mortality over the whole time series, but we found no appreciable differences in patterns across the various factors defining the OMs and EMs.
-Furthermore, results for terminal year fishing mortality were generally inversely related to those for spawning stock biomass, and are provided in the Supplementary Materials and not discussed further.
-
-We calculated median errors (ME) of \(\widehat \beta_E\), \(\widehat \beta_M\)), and the Hessian-based standard error estimators of these parameters (\(\widehat{SE}\left(\widehat \beta_E\right)\) and \(\widehat{SE}\left(\widehat \beta_M\right)\)).
+We focused on statistical behavior of estimators of the covariate effect on \(M\) (\(\widehat \beta_E\)), the estimator of the median \(M\) parameter (\(\widehat \beta_M\)), and estimators of terminal year \(M\) (\(\widehat M\)), spawning stock biomass (\(\widehat{\text{SSB}}\)), and fully-selected \(F\) (\(\widehat F\)).
+In preliminary analyses we examined results for the estimators of all the annual values for \(M\), SSB and \(F\) over the whole time series, but we found no appreciable differences in patterns across the various factors defining the OMs and EMs.
+Furthermore, results for terminal year \(F\) were generally inversely related to those for spawning stock biomass, and are provided in the Supplementary Materials and not discussed further.
+
+We calculated median errors (ME) of \(\widehat \beta_E\) and \(\widehat \beta_M\), and the Hessian-based standard error estimators of these parameters (\(\widehat{SE}\left(\widehat \beta_E\right)\) and \(\widehat{SE}\left(\widehat \beta_M\right)\)).
 We calculated the median relative errors (MRE) of terminal year \(\widehat M\), \(\widehat{\text{SSB}}\), and \(\widehat F\).
-We also calculated the root mean square error (RMSE) and estimated probability of coverage of constructed 95\% confidence intervals for \(\widehat \beta_E\) and \(\widehat \beta_M\) for EMs that estimated these parameters.
-We constructed the confidence intervals for probabilities of CI coverage using the same methods as those for probabilities of convergence.
-
-The true values for terminal year SSB and fishing mortality rate vary among simulations.
+We also calculated the root mean square error (RMSE) and estimated probability of coverage of constructed 95\% CIs for \(\widehat \beta_E\) and \(\widehat \beta_M\) for EMs that estimated these parameters.
+We constructed the CIs for probabilities of CI coverage using the same methods as those for probabilities of convergence.
+
+The true values for terminal year SSB and \(F\) vary among simulations.
 For the \(i\)th simulated data the relative error for terminal year value \(\theta_i\) provided from the fitted estimation model is
@@
 \]
-For ME and RME of estimators, we constructed 95\% confidence intervals using the binomial distribution approach as in \citet{stockmiller21} and \citet{milleretal_inreview1}.
-Because of the inverse relationship of the the confidence interval and the null hypothesis significance test, when the confidence interval contains zero we do not have evidence of bias from the simulation study.
+For ME and MRE of estimators, we constructed 95\% CIs using the binomial distribution approach as in \citet{stockmiller21} and \citet{milleretal_inreview1}.
+Because of the inverse relationship of the CI and the null hypothesis significance test, when the CI contains zero we do not have evidence of bias from the simulation study.
 
@@
 \citet{milleretal_inreview1} found inferences are most reliable for distinguishing process error sources and stock-recruit relationships in scenarios with lower observation error for indices and age composition data and with temporal contrast in fishing pressure.
-Our expectation is that inferences for covariate effects on \(M\) would require data to be at least that informative and inspection of the comprehensive set of results across all OMs generally confirms this (See supplemental Materials).
-Therefore, we restrict our attention to these OM scenarios with more informative data in the main paper (temporal contrast in fishing and lower population observation error).
+Our expectation is that inferences for covariate effects on \(M\) would require data to be at least that informative and inspection of the comprehensive set of results across all OMs generally confirms this (See Supplemental Materials).
+Therefore, we restrict our attention to these OM scenarios with more informative data (temporal contrast in fishing and lower population observation error).
 
@@
 R+M EMs generally did not converge with high frequency except for R+S OMs with low covariate uncertainty and where the EMs assumed median \(M\) as known.
-There was no strong effect of the treatment of median M (known or estimated) on convergence for any of the other EMs.
-The largest effect of the size of the true covariate effect (\(\beta_E\)) on convergence was observed for R and R+M EMs that estimated covariate effects in R and R+M OMs with high uncertainty in covariate observations (\(\sigma_e\)) and high variability in covariate process errors (\(\sigma_E\)).
-There was an increase in convergence frequency with increased variability in covariate process errors for many EMs in OMs with large covariate uncertainty.
+There was generally little effect of the treatment of median \(M\) (known or estimated) on convergence for any of the other EMs.
+The largest impact of the size of the true covariate effect (\(\beta_E\)) on convergence was observed for R and R+M EMs that estimated covariate effects in R and R+M OMs with high uncertainty in covariate observations (\(\sigma_e\)) and high temporal variability in the latent covariate (\(\sigma_E\)).
+There was an increase in convergence frequency with increased temporal variability in the covariate for many EMs in OMs with large covariate uncertainty.
 
@@
 We only present results for EMs where the median \(M\) rate parameter was estimated because results differed little whether this parameter was estimated or assumed known (see Figures \ref{aic_Rom} to \ref{aic_RMom}).
-When there is temporal contrast in fishing pressure and lower uncertainty in population observations, the only OM scenarios where AIC was consistently accurate for both covariate effect and process error configurations were R OMs with high variability in the covariate and low uncertainty in observations of the covariate (Figure \ref{aic}).
+When there was temporal contrast in fishing pressure and lower uncertainty in population observations, the only OM scenarios where AIC was consistently accurate for both covariate effect and process error configurations were R OMs with high variability in the covariate and low uncertainty in observations of the covariate (Figure \ref{aic}).
 R+M OMs with those same covariate attributes were only accurate for the covariate effect (low Type I and II errors) and the incorrect process error almost always chosen was R rather than R+S.
@@
 
-When there is temporal contrast in fishing pressure, lower uncertainty in population observations, low uncertainty in environmental observations, and larger temporal contrast in the simulated true environmental covariate, we observed generally little or no bias in estimation of covariate effects (\(\widehat \beta_E\)) across all EM and OM process error assumptions, both EM configurations of median \(M\), and all true covariate effect sizes (Figure \ref{beta_E_bias}, rows 3 and 4).
+When there was temporal contrast in fishing pressure, lower uncertainty in population observations, low uncertainty in environmental observations, and larger temporal contrast in the simulated true environmental covariate, we observed little or no bias in estimation of covariate effects (\(\widehat \beta_E\)) across all EM and OM process error assumptions, both EM configurations of median \(M\), and all true covariate effect sizes (Figure \ref{beta_E_bias}, rows 3 and 4).
 The only exception to this was when R EMs with median \(M\) estimated were fit to R+S OMs with those configurations.
@@
 For R+S OMs, EMs with the matching process error configuration showed little evidence of bias in estimating \(\beta_E\) across a range of covariate and covariate observation configurations.
+R+S and R+M EMs generally performed similarly when fitted to either R+S or R+M OMs except R+S OMs with high covariate uncertainty and low covariate variability (Figure \ref{beta_E_bias}, rows 5 and 6 of middle columns).
 
 Bias of Hessian-based standard error estimation for \(\widehat \beta_E\) was also close to zero in the same OM scenarios where bias in estimation of \(\beta_E\) itself was close to zero (Figure \ref{se_beta_E_bias}, rows 3 and 4).
-However, in these high information OMs bias of standard error estimation was observed in for several EMs that did not have the matching R+S and R+M process error configuration.
+However, in these high information OMs, bias of standard error estimation was observed for several EMs that did not have the matching R+S and R+M process error configuration.
 Estimation of standard errors was essentially unreliable for most EMs when OMs had higher uncertainty in covariate observations (Figure \ref{se_beta_E_bias}, rows 5 to 8).
 
-Despite the reliable estimation of covariate effects and standard errors with low uncertainty in environmental observations, and larger temporal contrast in the simulated true environmental covariate, we found poor confidence interval coverage for many EMs fitted to R+S and R+M OMs (Figure \ref{beta_E_CI_coverage}, rows 3 and 4).
-For example, R+M EMs fit to R+M OMs showed little bias in estimation of the covariate effect and standard errors, but confidence interval coverage was negatively biased (rows 3 and 4 in right two columns of Figures \ref{beta_E_bias}, \ref{se_beta_E_bias}, and \ref{beta_E_CI_coverage}.
-However, there appears to be a positive correlation of the covariate and standard error estimates such that estimates less than the true value had negatively biased standard error estimates which would make confidence intervals too small (Figure \ref{ex_lm_beta_E_SE_beta_E}).
-Confidence interval coverage was most reliable for R+S OMs when the EMs had the matching process error configuration (Figure \ref{beta_E_CI_coverage}, middle 2 columns).
+Despite the reliable estimation of covariate effects and standard errors with low uncertainty in environmental observations, and larger temporal contrast in the simulated true environmental covariate, we found poor CI coverage for many EMs fitted to R+S and R+M OMs (Figure \ref{beta_E_CI_coverage}, rows 3 and 4).
+For example, R+M EMs fit to R+M OMs showed little bias in estimation of the covariate effect and standard errors, but CI coverage was negatively biased (rows 3 and 4 in right two columns of Figures \ref{beta_E_bias}, \ref{beta_E_CI_coverage}, and \ref{se_beta_E_bias}).
+However, there appears to be a positive correlation of the covariate and standard error estimates such that estimates less than the true value had negatively biased standard error estimates which would make CIs too small (Figure \ref{ex_lm_beta_E_SE_beta_E}).
+CI coverage was most reliable for R+S OMs when the EMs had the matching process error configuration (Figure \ref{beta_E_CI_coverage}, middle 2 columns).
 
@@
 When there is temporal contrast in fishing pressure, lower uncertainty in population observations, all EMs fit to R and R+M OMs showed little evidence of bias for \(\widehat \beta_M\) (Figure \ref{beta_M_bias}, left and right sets of columns).
-For R+S OMs with those conditions, EMs with the matching process error assumption showed little evidence of bias.
-
-We also found little or no bias of Hessian-based standard error estimation for \(\beta_M\) in the many of the same OM scenarios where bias of \(\widehat \beta_M\) was close to zero except for some EMs with the process error source mis-specified or where the OM simulated the strongest covariate effect (Figure \ref{se_beta_M_bias}).
+For R+S OMs with those conditions, only EMs with the matching process error assumption showed little evidence of bias.
+
+We also found little or no bias of Hessian-based standard error estimation for \(\beta_M\) in many of the same OM scenarios where bias of \(\widehat \beta_M\) was close to zero except for some EMs with the process error source mis-specified or where the OM simulated the strongest covariate effect (Figure \ref{se_beta_M_bias}).
 Like \(\widehat \beta_M\), reliability of standard error estimation in R+S OMs required the EMs to have the matching process error assumption.
 
-However, we found confidence interval coverage to be generally unreliable even for OM and EM combinations where bias in \(\widehat \beta_M\) and its standard error estimators was negligible, much like that for \(\widehat \beta_E\) (Figure \ref{beta_M_CI_coverage}).
-In the same EM OM combination we investigated above for \(\widehat \beta_E\), we observed an opposite negative correlation of \(\widehat \beta_M\) and its standard error estimates (Figure \ref{ex_lm_beta_M_SE_beta_M}), which would result in confidence intervals being too narrow when \(\widehat \beta_M\) values are larger than average.
+we found CI coverage to be unreliable for many EM-OM combinations even for OM and EM combinations where bias in \(\widehat \beta_M\) and its standard error estimators was negligible, much like that for \(\widehat \beta_E\) (Figure \ref{beta_M_CI_coverage}).
+In the same EM OM combination we investigated above for \(\widehat \beta_E\), we observed an opposite negative correlation of \(\widehat \beta_M\) and its standard error estimates (Figure \ref{ex_lm_beta_M_SE_beta_M}), which would result in CIs being too narrow when \(\widehat \beta_M\) values are larger than average.
+The exceptions to the general poor CI coverage were for all EMs fit to several R OMs with low covariate uncertainty and covariate variability (Figure \ref{beta_M_CI_coverage}, first two columns) and for R+S EMs in many R+S OMs (Figure \ref{beta_M_CI_coverage}, middle two columns) .
 
@@
 
-The EMs with R and R+S process error configurations with median M assumed known and covariate effects not estimated will have terminal year M correctly specified when fitted to R and R+S OMs with no covariate effect because \(M\) is constant in both the EM and OM at the same value.
+The EMs with either R or R+S process error configurations with median \(M\) assumed known and covariate effects not estimated will have terminal year \(M\) correctly specified when fitted to either R or R+S OMs with no covariate effect because \(M\) is constant in both the EM and OM at the same value.
 We exclude those OM-EM combinations from our results here.
 
-In many of the OMs with temporal contrast in fishing pressure and lower uncertainty in population observations, there was little difference in bias for terminal year M among EMs whether the median \(M\) parameter (\(\beta_M\)) was estimated (Figure \ref{terminal_M_bias}).
-However, when there were differences, bias in terminal year M was closer to zero when the EM assumed \(\beta_M\) known, As would be expected.
+In many of the OMs with temporal contrast in fishing pressure and lower uncertainty in population observations, there was little difference in bias for terminal year \(M\) among EMs whether the median \(M\) parameter (\(\beta_M\)) was estimated (Figure \ref{terminal_M_bias}).
+However, when there were differences, bias in terminal year \(M\) was closer to zero when the EM assumed \(\beta_M\) known, as would be expected.
 For R OMs, all estimating models exhibited little evidence of bias except when OMs simulated the largest covariate effect (\(\beta_E = 0.5\)).
 However, bias estimates were at worst around 10\% (Figure \ref{terminal_M_bias}, left columns).
-For R+S OMs, the R+S EMs generally exhibited the least evidence for bias and R+M EMs generally estimated terminal M lower than the true value, particularly when observation uncertainty in the covariate was lower.
+For R+S OMs, the R+S EMs generally exhibited the least evidence for bias and R+M EMs generally estimated terminal \(M\) lower than the true value, particularly when observation uncertainty in the covariate was lower.
 For R+M OMs, many of the EM configurations provided little bias in terminal \(\widehat M\), but the biases were generally further from zero than the results for the R OMs (Figure \ref{terminal_M_bias}, right columns).
-Accuracy of terminal year \(\widehat M\) was similar for all of the EMs when fitted to R+M OMs (Figure \ref{terminal_M_rmse}).
-For R+S OMs, accuracy was a bit better for R+S EM than others.
-As would be expected accuracy of estimate where median \(M\) was assumed known was better than when it was estimated, when there were differences.
+
+Accuracy (lower RMSE) of terminal year \(\widehat M\) was similar for all of the EMs when fitted to R+M OMs (Figure \ref{terminal_M_rmse}).
+For R+S OMs, accuracy was generally better for R+S EM than other process error assumptions, particularly when \(\beta_E\) was also estimated.
+Accuracy for R+S EMs was also generally as good as other EM process error assumptions even when the OM process error type was different.
+As would be expected, the accuracy of terminal year \(\widehat M\) where median \(M\) was assumed known was better than when it was estimated, when there were differences.
 
@@
 In OMs with temporal contrast in fishing pressure and lower uncertainty in population observations, we found little bias in estimates of terminal year SSB for all EM configurations when fitted to R or R+M OMs (Figure \ref{terminal_ssb_bias}).
-We also found little bias for R+S OMs when the EMs assumed the correct process error configuration.
-The worst bias in terminal year SSB estimation we found was in R+S OMs for R EMs with \(\beta_M\) estimated.
-Like the bias results, accuracy as measured by RMSE was similar for many of EMs when fit to R and R+M OMs.
-Accuracy was generally worse when EMs were fit to R+S OMs (Figure \ref{terminal_ssb_rmse}).
+We also found little or no evidence of bias for R+S OMs when the EMs assumed the correct process error configuration.
+We found the worst bias in terminal year SSB estimation for R EMs with \(\beta_M\) estimated when fitted to R+S OMs, and R+M EMs also had more bias than R+S EMs in these OMs.
+Like the bias results, accuracy, as measured by RMSE, was similar for many of EMs when fit to R and R+M OMs.
+Accuracy was generally worse (higher RMSE) when EMs were fit to R+S OMs, and the worst accuracy occurred when the EM process error assumption did not match that of the OM (Figure \ref{terminal_ssb_rmse}).
 
@@
 Our simulation study demonstrated that estimation of environmental effects on \(M\) is possible and reliable in certain scenarios even when the process error was misspecified (e.g., R and R+S EMs fit to R+M OM).
-In many of these same OMs frequency of convergence of model fits did not appears to suffer when covariate effects on \(M\) were estimated even when there was no effect simulated.
-However, these scenarios are information rich in that there was contrast in population size (due to changes in fishing pressure) and the covariate affecting the population, and low uncertainty in population and covariate observations.
+In many of these same OMs, frequency of convergence of fitted models did not appear to suffer when covariate effects on \(M\) were estimated even when there was no effect simulated.
+However, these scenarios are information rich in that there was contrast in population size (due to changes in fishing pressure) and the covariate affecting the population, and there was low uncertainty in population and covariate observations.
 Previous research has shown that estimation of a constant \(M\) parameter requires contrast in time series and informative data \citep{leeetal11}, so it is no surprise that estimation of these effects also requires relatively good information via more precise observations and higher contrast in the covariate time series.
 
-On the other hand, AIC could only reliably detect covariate effects for R and R+M OMs with contrast in covariates and low covariate uncertainty and confidence interval coverage for the covariate effect was often biased even when bias in the estimators of the effect and the standard error were negligible.
-\citet{cadiganetal24} found confidence interval coverage to be biased for SSB and fishing mortality estimation in a state-space model in some scenarios, but they attributed the poor coverage to bias in Hessian-based standard error estimation and their simulations held any process error random effects constant.
-The coverage bias we observed, at least for some OM-EM combinations, may be a related to correlation of the estimators of the covariate effect and the corresponding standard error and therefore consideration of other methods of calculating confidence intervals may be warranted (e.g., those based on profile likelihood and/or MCMC sampling of log-likelihood surface).
-
-\citet{milleretal_inreview1} investigated R+M OMs with two levels of \(M\) process error variability (\(\sigma_M \in {0.1,0.5}\) and only found AIC able to accurately distinguish R+M process errors with the higher level of process error variability (\(\sigma_M = 0.5\)).
+Even though estimation of covariate effects was unbiased in many scenarios, AIC could only reliably detect covariate effects for R and R+M OMs with contrast in covariates and low covariate uncertainty.
+In those scenarios where the covariate effect could be detected, CI coverage for the covariate effect was often biased even when there was little or no bias in the estimators of the effect and its standard error.
+\citet{cadiganetal24} found CI coverage to be biased for SSB and \(F\) estimation in a state-space model in some scenarios, but they attributed the poor coverage to bias in Hessian-based standard error estimation, and their simulations held any process error random effects constant.
+The coverage bias we observed, at least for some OM-EM combinations, may be related to correlation of the estimators of the covariate effect and the corresponding standard error and therefore consideration of other methods of calculating CIs may be warranted (e.g., those based on profile likelihood and/or Monte Carlo sampling of the log-likelihood surface).
+
+\citet{milleretal_inreview1} investigated R+M OMs with two levels of \(M\) process error variability (\(\sigma_M \in \{0.1,0.5\}\)) and only found AIC able to accurately distinguish R+M process errors with the higher level of process error variability (\(\sigma_M = 0.5\)).
 We assumed \(\sigma_M = 0.3\), intermediate to the values investigated by \citet{milleretal_inreview1}, and found process error inferences unreliable for the source of process error, indicating that the level of variability required for detecting \(M\) process errors must be greater than \(\sigma_M = 0.3\), but may still be less than 0.5.
@@
 
-\citet{derisoetal08} attempted to estimate process errors as well as covariate effects with \(M\) for Pacific herring, but similarly found no variability in \(M\), suggesting a there was too much uncertainty in the available observations relative to the true temporal variability in \(M\).
+\citet{derisoetal08} attempted to estimate process errors as well as covariate effects with \(M\) for Pacific herring, but similarly found no variability in \(M\), suggesting there was too much uncertainty in the available observations relative to the true temporal variability in \(M\).
 Given that we found covariate effect inferences using R EMs was reliable in R+M OMs with apparently little variability in \(M\) process errors, the findings of \citet{derisoetal08} on covariate effects for Pacific herring would presumably also be robust to true low variability in \(M\).
@@
 
-Any bias or poor accuracy for annual SSB estimation was primarily a function of whether or not the median \(M\) parameter was estimated or known and the types of process errors than the treatment of the covariate effects on \(M\).
+Any bias or poor accuracy for annual SSB estimation was primarily a function of whether or not the median \(M\) parameter was estimated or known and the types of process errors, rather than the treatment of the covariate effects on \(M\).
 For example, we found estimation of SSB was better when the EM had the process error correctly specified for R+S OMs.
@@
 
+The R+S and R+M EMs both include process error for the survival of cohorts and would be expected to perform similarly, and they did in our simulations when the OM and EM matched.
+However, we found that the biases using R+M EMs for R+S OMs were generally worse than using R+S EMs for R+M OMs.
+Additionally, there are implications for biological reference points for R+M EMs \citep[e.g.,][]{legaultpalmer16} that are not present with R+S EMs.
+So, we recommend following the suggestion from \citet{lietal24} to prefer R+S EMs over R+M OMs unless strong biological evidence is present to support a particular R+M OM.
+Such support could be found through both biological understanding \citep[e.g.,][]{trijouletetal20} as well as statistical properties such as a large delta AIC for R+M compared to R+S associated with greater temporal variability in natural mortality \citep{milleretal_inreview1}.
+
 The ability to accurately infer covariate effects on \(M\) in some realistic situations indicates that such investigations may be fruitful.
-Ability to make inferences could improve further when WHAM is extended to incorporate tagging data as expected \citep{milleretal_inreview}.
+Ability to make inferences could improve further when WHAM is extended to incorporate tagging data \citep{milleretal_inreview}.
 Tagging data can greatly inform natural mortality estimation \citep{pollocketal91, hampton00}, and this impact on \(M\) estimation should also apply to estimation of covariate effects or unexplained temporal variation in \(M\).
-Given our findings coupled with future WHAM development, we expect investigations of and accounting for covariate effects on \(M\) to become more common within the fisheries stock assessment process.
+Given our findings and planned future WHAM development, we expect investigations of and accounting for covariate effects on \(M\) to become more common within the fisheries stock assessment process.
+At the same time, it will be equally important to conduct research that will improve our understanding of how best to measure the depletion of stocks and determine catch advice for these stocks with covariate effects on \(M\).
 
@@
 
-For R OMs, R EMs converged frequently except when there was contrast in fishing and covariate effects (\(\beta_E\)) were estimated and covariate observation uncertainty (\(\sigma_e\)) was high (Figure \ref{convergence_Rom}).
-There was little difference in convergence for R EMs whether median M was estimated or not.
-When R OMs had contrast in fishing, convergence of R+M EMs was generally better when the EM estimated median M, whereas in OMs with constant fishing, convergence was better when treating median M as known.
-Convergence of R+S EMs was generally worse than other process error configurations for R OMs, and there was no difference between EMs that estimated median M or not except when fishing pressure was constant where better convergence occurred when M was treated as known.
-Effects of true covariate effect size on convergence were apparent in some R OMs, but there did not appear to be any consistent patterns.
-
-For R+S OMs, R+M EMs generally showed the worse probability of convergence that EMs with R and R+S configurations (Figure \ref{convergence_RSom}).
-Convergence probability was high for all R and R+S EMs when covariate observation uncertainty was low.
-Although probability of convergence was generally greater when covariate effects were not estimated, there were no strong trends in convergence for any of the EMS with increased true effect size.
-
-For R+M OMs, R+S EMs generally converged less frequently than those with R and R+M configurations (Figure \ref{convergence_RMom}).
-Convergence probability for R OMs was generally the greatest of the three process error configurations.
-As with R+S OMs, there were no strong trends in convergence for any of the EMS with increased true covariate effect size.
-However, there was also not strong effects of estimating the covariate effect on convergence except for some OMs with higher covariate observation uncertainty.
-Like R OMs, when R+M OMs had contrast in fishing, convergence of R+M EMs was generally better when the EM estimated median M and the convergence was better when treating median M as known when the OM had constant fishing.
-
 \begin{landscape}

### 2025-08-08 Project_0/manuscript/manuscript_R1.Rmd
saving work on project_0 revisions

--- 2025-05-09 Project_0/manuscript/manuscript.Rmd
@@
 
-All model selection results condition on whether all of the compared estimating models completed the optimization process without failure.
+All model selection results condition only on whether all of the compared estimating models completed the optimization process without failure.
 We did not condition on convergence as defined by a gradient threshold or invertibity of the hessian because optimization could correctly determine an inappropriate process error assumption by estimating variance parameters at the lower bound of zero.
@@
 
+### Summarizing results across OMs {-}
+
+Because of the numerous factors investigated for each of the ways we assessed reliability of the state-space assessment models, we used  two methods to summarize the most important factors that explain differences for each measure of reliability.
+The first method was fitting regression models with the response being each of the measures of reliability described above and predictor variables were defined based on OM and EM characteristics.
+For convergence results we performed logistic regressions and for AIC results we performed multinomial regressions for the type of process error that provided the lowest AIC.
+For other measures of reliability we fit standard linear regression models.
+Because relative bias and Mohn's $\rho$ for the various parameters are bounded below at -1, we used as a response the log of the absolute log-scale errors
+\begin{equation}\label{bias_regression_response}
+y_i = \log\left|\log\left(\widehat \theta_i\right) - \log(\left(\theta_i\right)\right|,
+\end{equation}
+so that lower values measure less bias whatever the sign.
+For relative bias we also restricted the fits to simulations where the estimated CV of the parameter or model output was less than 100% because values greater than this would typically lead the assessment scientist away from the particular configuration of the estimating model.
+For all regressions we fit models with factors included individually and with all factors and with all second and third order interactions.
+For the multinomial regression, we used the \verb|vglm| function from the VGAM package [@yee08;@yee15].
+We tabulated percent reduction in residual deviance for each of regression fits.
+
+The second method involved fitting classification and regression trees [@breimanetal84] to show how the primary factors that split the values for each measure of reliability and subsets of the measures that were alloted to each branch of the splits.
+Classification trees were fit to categorical measures (convergence and AIC) and regression trees were fit to the other measures with continuous scales.
+The response variables were the same as the regressions for the deviance reduction analyses.
+We used the \verb|rpart| function in the rpart package (https://cran.r-project.org/package=rpart) to fit trees and we pruned them to show only the primary branches for clarity.
+Generally, the factors defining primary splits for the trees corresponded to the factors with the largest reductions in deviance fitting the regression models.
+
 # Results {-}
@@
 ## Convergence performance {-}
+
+### Deviance reduction and classification trees {-}
+
+### Full results {-}
 
@@
 
+### Deviance reduction and classification trees {-}
+
+### Full results {-}
+
 Marginal AIC accurately determined the correct process error assumptions in EMs when data were generated from R and R+S OMs, regardless of whether median natural mortality or a SRR was estimated (Figure \ref{pe_aic}, A).
@@
 ## AIC performance for the stock-recruit relationship {-}
+
+### Deviance reduction and classification trees {-}
+
+### Full results {-}
 
@@
 
-### Spawning stock biomass and recruitment {-}
+### Deviance reduction and regression trees {-}
+
+We observed different effects of factors on deviance reduction for errors in estimation of the two Beverton-Holt stock-recruit parameters.
+When analyses were For bias of the density-dependent parameter $b$ as measured by the log absolute value of log-scale errors, deviance was not reduced by any of the individual OM or EM factors except for variation in recruitment $\sigma_R$ for R and R+S OMs.
+However, second and(or) third order interactions of the OM and EM factors also provided large reductions in deviance for all OM process error types which indicated that some combinations of factors could reduce bias for the $b$ parameter (Table \ref{}).
+Despite the explanation variation in the $b$ estimation erorrs, bias was still significant regardless of the combination of OM and EM factors (Figures regtrees and full results).
+
+### Full results {-}
+
+#### Spawning stock biomass {-}
 
@@
 ## Mohn's $\rho$ {-}
+
+### Deviance reduction and regression trees {-}
+
+### Full results {-}
 
@@
 \begin{center}
+\includegraphics{convergence_classification_plots}
+\end{center}
+\caption{Classification trees indicating primary factors determining convergence as defined by providing hessian-based standard errors for R, R+S, R+M, R+Sel and R+q OMs.
+Lower and higher convergence rates are indicated by red and green, respectively}\label{conv_class}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{AIC_PE_classification_plots}
+\end{center}
+\caption{Classification trees indicating primary factors determining which EM process error assumption provides the lowest AIC for R+S, R+M, R+Sel and R+q OMs.
+Lower and higher accuracy of the process error assumption are indicated by red and green, respectively}\label{AIC_PE_class}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{SR_a_bias_regtree_plots}
+\end{center}
+\caption{Regression trees indicating primary factors determining reductions in sums of squares of errors in estimation measured by Eq.
+\ref{bias_regression_response} for the Beverton-Holt stock-recruit parameter $a$ for R+S, R+M, R+Sel and R+q OMs.
+Lower and higher median absolute errors of the process error assumption are indicated by green and red, respectively.}\label{SR_a_bias_regtree}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{SR_b_bias_regtree_plots}
+\end{center}
+\caption{Regression trees indicating primary factors determining reductions in sums of squares of errors in estimation measured by Eq.
+\ref{bias_regression_response} for the Beverton-Holt stock-recruit parameter $b$ for R+S, R+M, R+Sel and R+q OMs.
+Lower and higher median absolute errors of the process error assumption are indicated by green and red, respectively.}\label{SR_b_bias_regtree}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{med_M_bias_regtree_plots}
+\end{center}
+\caption{Regression trees indicating primary factors determining reductions in sums of squares of errors in estimation measured by Eq.
+\ref{bias_regression_response} for the median natural mortality rate for R+S, R+M, R+Sel and R+q OMs.
+Lower and higher median absolute errors of the process error assumption are indicated by green and red, respectively.}\label{med_MA_bias_regtree}
+\end{figure}
+\end{landscape}
+
+\pagebreak
+
+\begin{table}
+\caption{For each OM process error type (columns), percent reduction in deviance for logistic regression models fit to indicators of convergence with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{convergence_PRD_table}
+{\input{convergence_PRD_table}}
+\end{table}
+
+\begin{table}
+\caption{For each OM process error type (columns), percent reduction in deviance for multinomial logistic regression models fit to indicators of EM process error assumption with lowest AIC with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{AIC_PE_PRD_table}
+{\input{AIC_PE_PRD_table}}
+\end{table}
+
+\begin{landscape}
+\begin{table}
+\caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
+\ref{bias_regression_response} for the Beverton-Holt stock recruit relation ship parameters with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_SR_pars_PRD_table}
+{\input{bias_SR_pars_PRD_table}}
+\end{table}
+\end{landscape}
+
+\begin{table}
+\caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
+\ref{bias_regression_response} for the median natural mortality rate parameter with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_median_M_PRD_table}
+{\input{bias_median_M_PRD_table}}
+\end{table}
+
+\begin{table}
+\caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
+\ref{bias_regression_response} for the terminal year spawning biomass with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_SSB_PRD_table}
+{\input{bias_SSB_PRD_table}}
+\end{table}
+
+\pagebreak
+
+\setcounter{figure}{0}
+\renewcommand\thefigure{S\arabic{figure}}
+
+\setcounter{table}{0}
+\renewcommand\thetable{S\arabic{table}}
+
+\pagebreak
+
+# Supplementary Materials {-}
+
+\pagebreak
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
 \includegraphics{type_4_convergence_plots}
@@
 \end{landscape}
-
-\pagebreak
-
-\setcounter{figure}{0}
-\renewcommand\thefigure{S\arabic{figure}}
-
-\setcounter{table}{0}
-\renewcommand\thetable{S\arabic{table}}
-
-# Supplementary Materials {-}
 

### 2025-08-08 Project_0/manuscript/manuscript_R1.tex
saving work on project_0 revisions

--- 2025-05-02 Project_0/manuscript/manuscript.tex
@@
 C. Stock\textsuperscript{6} \and John Wiedenmann\textsuperscript{7}}
-\date{02 May, 2025}
+\date{08 August, 2025}
 
@@
 
-All model selection results condition on whether all of the compared
-estimating models completed the optimization process without failure. We
-did not condition on convergence as defined by a gradient threshold or
-invertibity of the hessian because optimization could correctly
-determine an inappropriate process error assumption by estimating
-variance parameters at the lower bound of zero. Such an optimization
-could indicate poor convergence but the likelihood would be equivalent
-to that without the mis-specified random effects and the AIC would be
-appropriately higher because more (variance) parameters were estimated.
-All other measures of reliability described below (bias and Mohn's
-\(\rho\)) use these same criteria for inclusion of EM fits in the
+All model selection results condition only on whether all of the
+compared estimating models completed the optimization process without
+failure. We did not condition on convergence as defined by a gradient
+threshold or invertibity of the hessian because optimization could
+correctly determine an inappropriate process error assumption by
+estimating variance parameters at the lower bound of zero. Such an
+optimization could indicate poor convergence but the likelihood would be
+equivalent to that without the mis-specified random effects and the AIC
+would be appropriately higher because more (variance) parameters were
+estimated. All other measures of reliability described below (bias and
+Mohn's \(\rho\)) use these same criteria for inclusion of EM fits in the
 summarized results.
@@
 
+\hypertarget{summarizing-results-across-oms}{%
+\subsubsection*{Summarizing results across
+OMs}\label{summarizing-results-across-oms}}
+\addcontentsline{toc}{subsubsection}{Summarizing results across OMs}
+
+Because of the numerous factors investigated for each of the ways we
+assessed reliability of the state-space assessment models, we used two
+methods to summarize the most important factors that explain differences
+for each measure of reliability. The first method was fitting regression
+models with the response being each of the measures of reliability
+described above and predictor variables were defined based on OM and EM
+characteristics. For convergence results we performed logistic
+regressions and for AIC results we performed multinomial regressions for
+the type of process error that provided the lowest AIC. For other
+measures of reliability we fit standard linear regression models.
+Because relative bias and Mohn's \(\rho\) for the various parameters are
+bounded below at -1, we used as a response the log of the absolute
+log-scale errors \begin{equation}\label{bias_regression_response}
+y_i = \log\left|\log\left(\widehat \theta_i\right) - \log(\left(\theta_i\right)\right|,
+\end{equation} so that lower values measure less bias whatever the sign.
+For relative bias we also restricted the fits to simulations where the
+estimated CV of the parameter or model output was less than 100\%
+because values greater than this would typically lead the assessment
+scientist away from the particular configuration of the estimating
+model. For all regressions we fit models with factors included
+individually and with all factors and with all second and third order
+interactions. For the multinomial regression, we used the \verb|vglm|
+function from the VGAM package (Yee 2008; Yee 2015). We tabulated
+percent reduction in residual deviance for each of regression fits.
+
+The second method involved fitting classification and regression trees
+(Breiman et al. 1984) to show how the primary factors that split the
+values for each measure of reliability and subsets of the measures that
+were alloted to each branch of the splits. Classification trees were fit
+to categorical measures (convergence and AIC) and regression trees were
+fit to the other measures with continuous scales. The response variables
+were the same as the regressions for the deviance reduction analyses. We
+used the \verb|rpart| function in the rpart package
+(\url{https://cran.r-project.org/package=rpart}) to fit trees and we
+pruned them to show only the primary branches for clarity. Generally,
+the factors defining primary splits for the trees corresponded to the
+factors with the largest reductions in deviance fitting the regression
+models.
+
 \hypertarget{results}{%
@@
 \addcontentsline{toc}{subsection}{Convergence performance}
+
+\hypertarget{deviance-reduction-and-classification-trees}{%
+\subsubsection*{Deviance reduction and classification
+trees}\label{deviance-reduction-and-classification-trees}}
+\addcontentsline{toc}{subsubsection}{Deviance reduction and
+classification trees}
+
+\hypertarget{full-results}{%
+\subsubsection*{Full results}\label{full-results}}
+\addcontentsline{toc}{subsubsection}{Full results}
 
@@
 
+\hypertarget{deviance-reduction-and-classification-trees-1}{%
+\subsubsection*{Deviance reduction and classification
+trees}\label{deviance-reduction-and-classification-trees-1}}
+\addcontentsline{toc}{subsubsection}{Deviance reduction and
+classification trees}
+
+\hypertarget{full-results-1}{%
+\subsubsection*{Full results}\label{full-results-1}}
+\addcontentsline{toc}{subsubsection}{Full results}
+
 Marginal AIC accurately determined the correct process error assumptions
@@
 relationship}
+
+\hypertarget{deviance-reduction-and-classification-trees-2}{%
+\subsubsection*{Deviance reduction and classification
+trees}\label{deviance-reduction-and-classification-trees-2}}
+\addcontentsline{toc}{subsubsection}{Deviance reduction and
+classification trees}
+
+\hypertarget{full-results-2}{%
+\subsubsection*{Full results}\label{full-results-2}}
+\addcontentsline{toc}{subsubsection}{Full results}
 
@@
 
-\hypertarget{spawning-stock-biomass-and-recruitment}{%
-\subsubsection*{Spawning stock biomass and
-recruitment}\label{spawning-stock-biomass-and-recruitment}}
-\addcontentsline{toc}{subsubsection}{Spawning stock biomass and
-recruitment}
+\hypertarget{deviance-reduction-and-regression-trees}{%
+\subsubsection*{Deviance reduction and regression
+trees}\label{deviance-reduction-and-regression-trees}}
+\addcontentsline{toc}{subsubsection}{Deviance reduction and regression
+trees}
+
+We observed different effects of factors on deviance reduction for
+errors in estimation of the two Beverton-Holt stock-recruit parameters.
+When analyses were For bias of the density-dependent parameter \(b\) as
+measured by the log absolute value of log-scale errors, deviance was not
+reduced by any of the individual OM or EM factors except for variation
+in recruitment \(\sigma_R\) for R and R+S OMs. However, second and(or)
+third order interactions of the OM and EM factors also provided large
+reductions in deviance for all OM process error types which indicated
+that some combinations of factors could reduce bias for the \(b\)
+parameter (Table \ref{}). Despite the explanation variation in the \(b\)
+estimation erorrs, bias was still significant regardless of the
+combination of OM and EM factors (Figures regtrees and full results).
+
+\hypertarget{full-results-3}{%
+\subsubsection*{Full results}\label{full-results-3}}
+\addcontentsline{toc}{subsubsection}{Full results}
+
+\hypertarget{spawning-stock-biomass}{%
+\paragraph*{Spawning stock biomass}\label{spawning-stock-biomass}}
+\addcontentsline{toc}{paragraph}{Spawning stock biomass}
 
@@
 \addcontentsline{toc}{subsection}{Mohn's \(\rho\)}
+
+\hypertarget{deviance-reduction-and-regression-trees-1}{%
+\subsubsection*{Deviance reduction and regression
+trees}\label{deviance-reduction-and-regression-trees-1}}
+\addcontentsline{toc}{subsubsection}{Deviance reduction and regression
+trees}
+
+\hypertarget{full-results-4}{%
+\subsubsection*{Full results}\label{full-results-4}}
+\addcontentsline{toc}{subsubsection}{Full results}
 
@@
 This work was funded by NOAA Fisheries Northeast Fisheries Science
-Center.
+Center. We thank Jon Deroba for helpful comments on an earlier version
+of this manuscript.
 
@@
 doi:\href{https://doi.org/10.1002/ecm.1470}{10.1002/ecm.1470}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-breimanetal84}{}}%
+Breiman, L., Friedman, J.H., Olshen, R.A., and Stone, C.J. 1984.
+Classification and regression trees. Chapman; Hall/CRC, New York, NY
+USA.
+doi:\href{https://doi.org/10.1201/9781315139470}{10.1201/9781315139470}.
 
@@
 \textbf{158}: 124--134.
+doi:\href{https://doi.org/10.1016/j.fishres.2013.06.003}{10.1016/j.fishres.2013.06.003}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-yee08}{}}%
+Yee, T.W. 2008. The {VGAM} package. R News \textbf{8}(2): 28--39.
+Available from
+\url{https://journal.r-project.org/articles/RN-2008-014/}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-yee15}{}}%
+Yee, T.W. 2015. Vector generalized linear and additive models: With an
+implementation in {R}. Springer, New York, NY USA.
+doi:\href{https://doi.org/10.1007/978-1-4939-2818-7}{10.1007/978-1-4939-2818-7}.
+
 \end{CSLReferences}
@@
 \clearpage
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{convergence_classification_plots}
+\end{center}
+\caption{Classification trees indicating primary factors determining convergence as defined by providing hessian-based standard errors for R, R+S, R+M, R+Sel and R+q OMs.
+Lower and higher convergence rates are indicated by red and green, respectively}\label{conv_class}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{AIC_PE_classification_plots}
+\end{center}
+\caption{Classification trees indicating primary factors determining which EM process error assumption provides the lowest AIC for R+S, R+M, R+Sel and R+q OMs.
+Lower and higher accuracy of the process error assumption are indicated by red and green, respectively}\label{AIC_PE_class}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{SR_a_bias_regtree_plots}
+\end{center}
+\caption{Regression trees indicating primary factors determining reductions in sums of squares of errors in estimation measured by Eq.
+\ref{bias_regression_response} for the Beverton-Holt stock-recruit parameter $a$ for R+S, R+M, R+Sel and R+q OMs.
+Lower and higher median absolute errors of the process error assumption are indicated by green and red, respectively.}\label{SR_a_bias_regtree}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{SR_b_bias_regtree_plots}
+\end{center}
+\caption{Regression trees indicating primary factors determining reductions in sums of squares of errors in estimation measured by Eq.
+\ref{bias_regression_response} for the Beverton-Holt stock-recruit parameter $b$ for R+S, R+M, R+Sel and R+q OMs.
+Lower and higher median absolute errors of the process error assumption are indicated by green and red, respectively.}\label{SR_b_bias_regtree}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{med_M_bias_regtree_plots}
+\end{center}
+\caption{Regression trees indicating primary factors determining reductions in sums of squares of errors in estimation measured by Eq.
+\ref{bias_regression_response} for the median natural mortality rate for R+S, R+M, R+Sel and R+q OMs.
+Lower and higher median absolute errors of the process error assumption are indicated by green and red, respectively.}\label{med_MA_bias_regtree}
+\end{figure}
+\end{landscape}
+
+\pagebreak
+
+\begin{table}
+\caption{For each OM process error type (columns), percent reduction in deviance for logistic regression models fit to indicators of convergence with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{convergence_PRD_table}
+{\input{convergence_PRD_table}}
+\end{table}
+
+\begin{table}
+\caption{For each OM process error type (columns), percent reduction in deviance for multinomial logistic regression models fit to indicators of EM process error assumption with lowest AIC with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{AIC_PE_PRD_table}
+{\input{AIC_PE_PRD_table}}
+\end{table}
+
+\begin{landscape}
+\begin{table}
+\caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
+\ref{bias_regression_response} for the Beverton-Holt stock recruit relation ship parameters with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_SR_pars_PRD_table}
+{\input{bias_SR_pars_PRD_table}}
+\end{table}
+\end{landscape}
+
+\begin{table}
+\caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
+\ref{bias_regression_response} for the median natural mortality rate parameter with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_median_M_PRD_table}
+{\input{bias_median_M_PRD_table}}
+\end{table}
+
+\begin{table}
+\caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
+\ref{bias_regression_response} for the terminal year spawning biomass with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_SSB_PRD_table}
+{\input{bias_SSB_PRD_table}}
+\end{table}
+
+\pagebreak
+
+\setcounter{figure}{0}
+\renewcommand\thefigure{S\arabic{figure}}
+
+\setcounter{table}{0}
+\renewcommand\thetable{S\arabic{table}}
+
+\pagebreak
+
+\hypertarget{supplementary-materials}{%
+\section*{Supplementary Materials}\label{supplementary-materials}}
+\addcontentsline{toc}{section}{Supplementary Materials}
+
+\pagebreak
 
@@
 
-\pagebreak
-
-\setcounter{figure}{0}
-\renewcommand\thefigure{S\arabic{figure}}
-
-\setcounter{table}{0}
-\renewcommand\thetable{S\arabic{table}}
-
-\hypertarget{supplementary-materials}{%
-\section*{Supplementary Materials}\label{supplementary-materials}}
-\addcontentsline{toc}{section}{Supplementary Materials}
-
 \begin{figure}[!ht]

### 2025-08-08 Project_0/code/manuscript/aic_SRR_summarize.R

@@ -1,0 +1,295 @@
+library(here)
+library(dplyr)
+library(tidyr)
+library(Hmisc)
+
+aic_fn <- function(x, all_aic, df.oms, df.ems, rec_mod, M_est) {
+  if(!is.null(df.oms$NAA_sig)){
+    if(is.na(df.oms$NAA_sig[x])) re_config <- "rec"
+    else re_config <- "rec+1"  
+    em_ind <- which(df.ems$SR_model == rec_mod & df.ems$M_est == M_est & df.ems$re_config == re_config)
+  }
+  if(!is.null(df.oms$M_sig)){
+    if(df.oms$M_cor[x] == 0) re_config <- "iid"
+    else re_config <- "ar1_y"  
+    em_ind <- which(df.ems$SR_model == rec_mod & df.ems$M_est == M_est & df.ems$M_re_cor == re_config)
+  }
+  if(!is.null(df.oms$Sel_sig)){
+    if(df.oms$Sel_cor[x] == 0) re_config <- "iid"
+    else re_config <- "ar1_y"  
+    em_ind <- which(df.ems$SR_model == rec_mod & df.ems$M_est == M_est & df.ems$sel_re_cor == re_config)
+  }
+  if(!is.null(df.oms$q_sig)){
+    if(df.oms$q_cor[x] == 0) re_config <- "iid"
+    else re_config <- "ar1"  
+    em_ind <- which(df.ems$SR_model == rec_mod & df.ems$M_est == M_est & df.ems$q_re_cor == re_config)
+  }
+  return(all_aic[[x]][em_ind,]) #single EM/row
+}
+
+df.ems = readRDS(file.path(here(),"Project_0","inputs", "df.ems.RDS"))
+# em_ind <- c(5:20,25:28)
+# use.df.ems <- df.ems[em_ind,]
+types <- c("naa", "M", "Sel", "q")
+types.plt <- c("R, R+S","R+M", "R+Sel", "R+q")
+est_conds <- cbind(c(FALSE,TRUE))
+em_inds <- cbind(1:20, 5:24, c(5:20,25:28), c(5:20, 29:32))
+pred_dfs <- list()
+obs_dfs <- list()
+for(i in 1:4) {
+  all_aic <- readRDS(file = here("Project_0","results", paste0("all_", types[i], "_aic_results.RDS")))
[…]

### 2025-08-08 Project_0/code/manuscript/bias_results_summarize.R

@@ -1,0 +1,1103 @@
+library(here)
+library(dplyr)
+library(tidyr)
+library(Hmisc)
+library(reshape2)
+library(rpart)
+library(viridis)
+#modified rpart.plot package to use plotmath in split labels...
+pkgload::load_all("c:/work/rpart.plot")
+
+
+make_plot_df <- function(om_type = "naa", res = naa_relSR_results, is_SE = FALSE, M_or_SR = "SR", year = NULL) {
+  df.ems <- readRDS(here::here("Project_0","inputs", "df.ems.RDS"))
+  if(om_type == "naa") {
+    em_ind <- 1:20
+  }
+  if(om_type == "M") {
+    em_ind <- 5:24
+  }
+  if(om_type == "Sel") {
+    em_ind <- c(5:20,25:28)
+  }
+  if(om_type == "q") {
+    em_ind <- c(5:20,29:32)
+  }
+  df.ems <- df.ems[em_ind,]
+  if(!is.null(M_or_SR)) {
+    if(M_or_SR == "SR") em_ind <- which(df.ems$SR_model == 3)
+    if(M_or_SR == "M") em_ind <- which(df.ems$M_est)
+    df.ems <- df.ems[em_ind,]
+  }
+  print("df.ems")
+  print(df.ems)
+
+  res <- melt(res)
+  names(res) <- c("par", "column", "value", "em", "sim","om") #em = 1: M fixed, em = 2: M estimated
+  if(!is.null(year)) res <- filter(res, par == year)
+
+  print("unique(res$em)")
+  print(unique(res$em))
+  print(all(unique(res$em) %in% 1:NROW(df.ems)))
+  print(length(unique(res$em)))
+  print(length(unique(res$em)))
+  print(NROW(df.ems))
+  if(!(all(unique(res$em) %in% 1:NROW(df.ems)) & length(unique(res$em)) == NROW(df.ems))) stop("df.ems does not seem to be difined correctly for the number of ems in res.")
+  res <- cbind(df.ems[res$em,], res)
+  res <- res %>% mutate(column = recode(column,
+      "1" = "relerror",
+      "2" = "cv",
+      "3" = "in_ci"
+    ))
+
+
+  df <- res %>% pivot_wider(names_from = column, values_from = value) %>% as.data.frame
+  if(is_SE) df <- filter(df, !is.na(cv))
+
+  df$relerror = df$relerror - 1
+
[…]

### 2025-09-10 Project_0/manuscript/manuscript_R1.Rmd
saving work on revision

@@
 We evaluate whether convergence and Akaike Information Criterion (AIC) can correctly determine the source of process error and the existence of a SRR.
-We also evaluate when retrospective patterns occur and the degree of bias in the outputs of the assessment model that are important for management.
+We also evaluate when retrospective patterns occur and the degree of bias in outputs of the assessment model that are important for management.
 
@@
 For a given model attribute we calculated the relative error 
-$$
+\begin{equation}\label{relerror}
 \text{RE}\left(\theta_i\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
-$$
+\end{equation}
 from fitting a given estimating model to simulated data set $i$ configured for a given OM where $\widehat \theta_i$ and $\theta_i$ are the estimated and true values for simulation $i$.
@@
 
-We calculated Mohn's $\rho$ for SSB, fully-selected fishing mortality, and recruitment for each EM [@mohn99].
-We fit 7 peels for each EM and calculated median 95\% confidence intervals for Mohn's $\rho$ using the same methods as that for relative bias.
+We calculated Mohn's $\rho$ for SSB, fishing mortality (averaged over all age classes), and recruitment for each EM [@mohn99].
+We fit $P =7$ peels to each simulated data set and calculated Mohn's $\rho$ for a given attribute $\theta$ as
+\begin{equation}\label{mohns_rho}
+\rho\left(\theta\right) = \frac{1}{P}\sum^P_{p=1} \frac{\widehat \theta_{Y-p,Y-p} - \widehat\theta_{Y-p,Y}}{\widehat\theta_{Y-p,Y}}
+\end{equation}
+where $\widehat \theta_{i,j}$ is the estimate for attribute $\theta$ in year $i$ from a model fit using data up to year $j$.
+For each EM we calculated median and 95\% confidence intervals using the same methods as that for relative bias.
 
-### Summarizing results across OMs {-}
+### Summarizing results across OM and EM attributes {-}
 
-Because of the numerous factors investigated for each of the ways we assessed reliability of the state-space assessment models, we used  two methods to summarize the most important factors that explain differences for each measure of reliability.
-The first method was fitting regression models with the response being each of the measures of reliability described above and predictor variables were defined based on OM and EM characteristics.
-For convergence results we performed logistic regressions and for AIC results we performed multinomial regressions for the type of process error that provided the lowest AIC.
+The measures of central tendency and variability of of observed values for specific OM and EM attributes (e.g., low or high observation error) that we described above can indicate scenarios that provide better or poorer reliability.
+However, the OM and EM attributes that we investigated are numerous, so we used two methods to summarize the most important factors for differences in results.
+The first method was fitting regression models with the response being each of the measures of reliability described above and predictor variables were defined based on OM and EM characteristics [e.g., @mackinnonetal95; @wangetal17; @harwelletal18].
+For the binary results for convergence and AIC-based selection of a stock-recruit relationship, we performed logistic regressions.
+For results for AIC-based selection of EM process error type (multiple categories) we performed multinomial regressions.
 For other measures of reliability we fit standard linear regression models.
-Because relative bias and Mohn's $\rho$ for the various parameters are bounded below at -1, we used as a response the log of the absolute log-scale errors
+Because relative errors (Eq.
+\ref{relerror}) and Mohn's $\rho$ for the various parameters are bounded below at -1, we used a transformation of these values
 \begin{equation}\label{bias_regression_response}
-y_i = \log\left|\log\left(\widehat \theta_i\right) - \log(\left(\theta_i\right)\right|,
+y_i = \log\left[f\left(\widehat \theta_i,\theta_i\right)+1\right]
 \end{equation}
-so that lower values measure less bias whatever the sign.
-For relative bias we also restricted the fits to simulations where the estimated CV of the parameter or model output was less than 100% because values greater than this would typically lead the assessment scientist away from the particular configuration of the estimating model.
-For all regressions we fit models with factors included individually and with all factors and with all second and third order interactions.
+where $f$ is either the relative error (Eq.
+\ref{relerror}) or Mohn's $\rho$ (Eq.
+\ref{mohns_rho}) for simulation $i$, so that values are unbounded.
+For relative errors, $y_i$ is the log-scale error.
+We omitted simulation estimates equal to zero (RE = -1).
+For all regressions we fit separate models with individual factors included, with all factors combined, with including all second order interactions, and including all third order interactions.
 For the multinomial regression, we used the \verb|vglm| function from the VGAM package [@yee08;@yee15].
 We tabulated percent reduction in residual deviance for each of regression fits.
+We did not perform formal statistical analyses of effects of OM and EM attributes on results (e.g., ANOVA) because of the lack of independence of the "observations" that results from fitting multiple EMs to each simulated data set.
 
-The second method involved fitting classification and regression trees [@breimanetal84] to show how the primary factors that split the values for each measure of reliability and subsets of the measures that were alloted to each branch of the splits.
-Classification trees were fit to categorical measures (convergence and AIC) and regression trees were fit to the other measures with continuous scales.
+The second method involved fitting classification and regression trees [@breimanetal84] to show how the OM and EM attributes, and their interactions, split the values for each measure of reliability [e.g., @gonzalezetal18;@collieretal22].
+We used classification trees for categorical measures (convergence and AIC) and regression trees for the other measures with continuous scales (relative error and Mohn's $\rho$).
 The response variables were the same as the regressions for the deviance reduction analyses.
-We used the \verb|rpart| function in the rpart package (https://cran.r-project.org/package=rpart) to fit trees and we pruned them to show only the primary branches for clarity.
-Generally, the factors defining primary splits for the trees corresponded to the factors with the largest reductions in deviance fitting the regression models.
+We used the \verb|rpart| function in the rpart package (https://cran.r-project.org/package=rpart) to fit trees and, for clarity we pruned them to show the primary branches.
 
@@
 ### Deviance reduction and classification trees {-}
+
+EM process error assumption was the single attribute that resulted in the largest percent reduction in deviance for all OM process error types other than R+S OMs where the EM M assumption explained the most residual deviance (Table \ref{convergence_PRD_table}).
+However, also including interactions of OM and EM factors also provided large reductions in residual deviance, suggesting successful convergence depended on a combination OM and EM attributes.
+Classification trees for each OM process error type, all had the primary branch defined using the same attribute that provided the largest reduction in deviance (Figure \ref{conv_class}).
+
+Generally, the factors defining primary splits for the trees corresponded to those with the largest reductions in deviance fitting the regression models.
 
@@
 
+Across all OM process error types, fishing history was the most important single factor in reducing residual deviation in log-scale errors for the Beverton-Holt density-independent parameter $a$, and also for the density-independent parameter $b$ for R+M, R+Sel, and R+q OMs (Table \ref{bias_SR_pars_PRD_table}).
+For R and R+S OMs the EM process error assumption was equally important for estimating $b$, but the the reductions were small (mostly less than 4%) for any factor and any OM process error type.
+There was also no substantial reduction in deviance with second or third order interactions of factors for either parameter in any OM process error types.
+
+The regression trees indicate that having contrast in fishing pressure reduced the median error for both SR parameters for all OM types
+
 We observed different effects of factors on deviance reduction for errors in estimation of the two Beverton-Holt stock-recruit parameters.
 When analyses were For bias of the density-dependent parameter $b$ as measured by the log absolute value of log-scale errors, deviance was not reduced by any of the individual OM or EM factors except for variation in recruitment $\sigma_R$ for R and R+S OMs.
-However, second and(or) third order interactions of the OM and EM factors also provided large reductions in deviance for all OM process error types which indicated that some combinations of factors could reduce bias for the $b$ parameter (Table \ref{}).
+However, second and(or) third order interactions of the OM and EM factors also provided large reductions in deviance for all OM process error types which indicated that some combinations of factors could reduce bias for the $b$ parameter.
 Despite the explanation variation in the $b$ estimation erorrs, bias was still significant regardless of the combination of OM and EM factors (Figures regtrees and full results).
+
+Median natural mortality rate showed reductions in deviance for R, R+S, and R+M OMs with second order interactions and R+M also showed reductions with third order interactions.
 
@@
 ### Deviance reduction and regression trees {-}
+
+Fits of glms showed little explanatory power of any of the OM an EM attributes for Mohn's rho for SSB, F, and R.
+Note that the factors that showed the largest reductions in deviance (which were very small) often do not comport with factors that show differences in median Mohn's rho results.
+This is because the glms and regression trees are reducing residual variance and there could be large variability in values, but the medians are less sensitive to extreme values.
+For example we can see that there is large Mohn's rho for R in R OMs with larger variability assumed in recruitment and higher observation error (Figure X), but these factors suggest little reduction in residual deviance (Table X).
 
@@
 \caption{Classification trees indicating primary factors determining convergence as defined by providing hessian-based standard errors for R, R+S, R+M, R+Sel and R+q OMs.
-Lower and higher convergence rates are indicated by red and green, respectively}\label{conv_class}
+Lower or higher convergence rates are indicated by red or green polygons, respectively}\label{conv_class}
 \end{figure}
@@
 \caption{Classification trees indicating primary factors determining which EM process error assumption provides the lowest AIC for R+S, R+M, R+Sel and R+q OMs.
-Lower and higher accuracy of the process error assumption are indicated by red and green, respectively}\label{AIC_PE_class}
+Each node shows the proportion of EM process error models with lowest AIC (top) and number of observations (bottom) for the corresponding subset.
+Lower or higher accuracy of the process error assumption are indicated by red or green polygons, respectively.}\label{AIC_PE_class}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{AIC_SRR_classification_plots}
+\end{center}
+\caption{Classification trees indicating primary factors determining which EM stock recruitment assumption (none or Beverton-Holt) provides the lowest AIC for R, R+S, R+M, R+Sel and R+q OMs.
+Each node shows the proportion of EMs that assume the stock-recruit relationship with lowest AIC (top) and number of observations (bottom) for the corresponding subset.
+Lower or higher accuracy of the process error assumption are indicated by more red or green polygons, respectively.}\label{AIC_SRR_class}
 \end{figure}
@@
 \ref{bias_regression_response} for the Beverton-Holt stock-recruit parameter $a$ for R+S, R+M, R+Sel and R+q OMs.
-Lower and higher median absolute errors of the process error assumption are indicated by green and red, respectively.}\label{SR_a_bias_regtree}
+Each node shows the median error (top) and number of observations (bottom) for the corresponding subset.
+Lower or higher median absolute errors of the process error assumption are indicated by more green or red polygons, respectively.}\label{SR_a_bias_regtree}
 \end{figure}
@@
 \ref{bias_regression_response} for the Beverton-Holt stock-recruit parameter $b$ for R+S, R+M, R+Sel and R+q OMs.
-Lower and higher median absolute errors of the process error assumption are indicated by green and red, respectively.}\label{SR_b_bias_regtree}
+Each node shows the median error (top) and number of observations (bottom) for the corresponding subset.
+Lower or higher median absolute errors of the process error assumption are indicated by more green or red polygons, respectively.}\label{SR_b_bias_regtree}
 \end{figure}
@@
 \ref{bias_regression_response} for the median natural mortality rate for R+S, R+M, R+Sel and R+q OMs.
-Lower and higher median absolute errors of the process error assumption are indicated by green and red, respectively.}\label{med_MA_bias_regtree}
+Each node shows the median error (top) and number of observations (bottom) for the corresponding subset.
+Lower or higher median absolute errors of the process error assumption are indicated by more green or red polygons, respectively.}\label{med_M_bias_regtree}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{SSB_bias_regtree_plots}
+\end{center}
+\caption{Regression trees indicating primary factors determining reductions in sums of squares of errors in estimation measured by Eq.
+\ref{bias_regression_response} for terminal year spawning stock biomass for R+S, R+M, R+Sel and R+q OMs.
+Each node shows the median error (top) and number of observations (bottom) for the corresponding subset.
+Median errors closer to or further from zero are indicated by more green or red polygons, respectively.}\label{SSB_bias_regtree}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{SSB_mohns_rho_regtree_plots}
+\end{center}
+\caption{Regression trees indicating primary factors determining reductions in sums of squares of errors in transformed Mohn's $\rho$ (Eq.
+\ref{bias_regression_response}) for spawning stock biomass for R+S, R+M, R+Sel and R+q OMs.
+Each node shows the median Mohn's $\rho$ (top) and number of observations (bottom) for the corresponding subset.
+Median Mohn's $\rho$ closer to or further from zero are indicated by more green or red polygons, respectively.}\label{SSB_mohns_rho_regtree}
 \end{figure}
@@
 {\input{AIC_PE_PRD_table}}
+\end{table}
+
+\begin{table}
+\caption{For each OM process error type (columns), percent reduction in deviance for logistic regression models fit to indicators of EM stock-recruit assumption (none or Beverton-Holt) with lowest AIC with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{AIC_SRR_PRD_table}
+{\input{AIC_SRR_PRD_table}}
 \end{table}
@@
 
+\begin{table}
+\caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to transformed Mohn's $\rho$ values for each simulation (Eq.
+\ref{bias_regression_response}) for spawning biomass with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{mohns_rho_SSB_PRD_table}
+{\input{mohns_rho_SSB_PRD_table}}
+\end{table}
+
 \pagebreak
@@
 
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width = \textwidth]{om_input_plots_figure}
+\end{center}
+\caption{The proportion mature at age, weight at age, fleet and index selectivity at age, and Beverton-Holt stock-recruit relationship assumed for the population in all operating models.
+For operating models with random effects on fleet selectivity, this represents the selectivity at the mean of the random effects.}\label{om_inputs_fig}
+\end{figure}
+
 \begin{landscape}
@@
 \clearpage
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{type_3_convergence_plots}
+\end{center}
+\caption{Probability of estimating models providing maximum absolute values of gradients less than $10^{-6}$ assuming alternative process error (colored points and lines), and median natural mortality (estimated or known) and Beverton-Holt stock-recruit relationships (estimated or not; along x-axis) when fitted to operating models that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{gradient_convergence}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{hess_grad_convergence_plots}
+\end{center}
+\caption{The maximum of the absolute values of all gradient values for all fits that provided hessian-based standard errors across all simuated data sets of a given OM configuration (A: R and R+S, B: R+M, C: R+Sel, or D: R+q).
+Results are conditional on EM fits with alternative process error type (colored points and lines), median natural mortality (estimated or known) and recruitment assumptions (Beverton-Holt stock-recruit relationship or not).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{hess_grad}
+\end{figure}
+\end{landscape}
 
@@
 \begin{center}
+\includegraphics{sr_aic_plots}
+\end{center}
+\caption{Estimated probability of lowest AIC from logistic regression on the log-standard deviation of the true log(SSB) in each simulation for estimating model with Beverton-Holt stock-recruit relationships, rather than the otherwise equivalent EM without the stock-recruit relationship.
+Results are conditional on alternative assumptions for median natural mortality (estimated or known) and on EMs having the correct process error structure: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Rug along x-axis denotes $SD(\log(SSB))$ values for each simulation and polygons represent 95\% confidence intervals.}\label{sr_aic_supp}
+\end{figure}
+\end{landscape}
+
+\clearpage
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
 \includegraphics{term_SSB_bias_plots}
@@
 \begin{center}
+\includegraphics{term_F_bias_plots}
+\end{center}
+\caption{Median relative error of terminal year fully-selected fishing mortality for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{F_rel_error}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{term_R_bias_plots}
+\end{center}
+\caption{Median relative error of terminal year recruitment for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{R_rel_error}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{sr_bias_plots}
+\end{center}
+\caption{Median relative error of Beverton-Holt stock-recruit parameters ($a$ and $b$) for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{SR_rel_error}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{M_bias_plots}
+\end{center}
+\caption{Median relative error of median natural mortality for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{M_rel_error}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
 \includegraphics{mohns_rho_ssb_plots}
 \end{center}
-\caption{Median Mohn's rho for SSB for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+\caption{Median Mohn's $\rho$ for SSB for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
 Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_ssb}
@@
 
-\begin{figure}[!ht]
+\begin{landscape}
+\begin{figure}
 \begin{center}
-\includegraphics[width = \textwidth]{om_input_plots_figure}
+\includegraphics{mohns_rho_F_plots}
 \end{center}
-\caption{The proportion mature at age, weight at age, fleet and index selectivity at age, and Beverton-Holt stock-recruit relationship assumed for the population in all operating models.
-For operating models with random effects on fleet selectivity, this represents the selectivity at the mean of the random effects.}\label{om_inputs_fig}
+\caption{Median Mohn's $\rho$ of fishing mortality averaged over all age classes for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_F}
 \end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics{mohns_rho_R_plots}
+\end{center}
+\caption{Median Mohn's $\rho$ of recruitment for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_R}
+\end{figure}
+\end{landscape}
 
@@
 
-\clearpage
+\begin{table}
+\caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to transformed Mohn's $\rho$ values for each simulation (Eq.
+\ref{bias_regression_response}) for fishing mortality averaged over all age classes with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{mohns_rho_F_PRD_table}
+{\input{mohns_rho_F_PRD_table}}
+\end{table}
 
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{hess_grad_convergence_plots}
-\end{center}
-\caption{The maximum of the absolute values of all gradient values for all fits that provided hessian-based standard errors across all simuated data sets of a given OM configuration (A: R and R+S, B: R+M, C: R+Sel, or D: R+q).
-Results are conditional on EM fits with alternative process error type (colored points and lines), median natural mortality (estimated or known) and recruitment assumptions (Beverton-Holt stock-recruit relationship or not).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{hess_grad}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{type_3_convergence_plots}
-\end{center}
-\caption{Probability of estimating models providing maximum absolute values of gradients less than $10^{-6}$ assuming alternative process error (colored points and lines), and median natural mortality (estimated or known) and Beverton-Holt stock-recruit relationships (estimated or not; along x-axis) when fitted to operating models that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{gradient_convergence}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{sr_aic_plots}
-\end{center}
-\caption{Estimated probability of lowest AIC from logistic regression on the log-standard deviation of the true log(SSB) in each simulation for estimating model with Beverton-Holt stock-recruit relationships, rather than the otherwise equivalent EM without the stock-recruit relationship.
-Results are conditional on alternative assumptions for median natural mortality (estimated or known) and on EMs having the correct process error structure: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Rug along x-axis denotes $SD(\log(SSB))$ values for each simulation and polygons represent 95\% confidence intervals.}\label{sr_aic_supp}
-\end{figure}
-\end{landscape}
-
-\clearpage
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{term_F_bias_plots}
-\end{center}
-\caption{Median relative error of terminal year fully-selected fishing mortality for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{F_rel_error}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{term_R_bias_plots}
-\end{center}
-\caption{Median relative error of terminal year recruitment for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{R_rel_error}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{sr_bias_plots}
-\end{center}
-\caption{Median relative error of Beverton-Holt stock-recruit parameters ($a$ and $b$) for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{SR_rel_error}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{M_bias_plots}
-\end{center}
-\caption{Median relative error of median natural mortality for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{M_rel_error}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{mohns_rho_F_plots}
-\end{center}
-\caption{Median Mohn's $\rho$ of fishing mortality averaged over all age classes for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_F}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{mohns_rho_R_plots}
-\end{center}
-\caption{Median Mohn's $\rho$ of recruitment for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_R}
-\end{figure}
-\end{landscape}
+\begin{table}
+\caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to transformed Mohn's $\rho$ values for each simulation (Eq.
+\ref{bias_regression_response}) for recruitment with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{mohns_rho_R_PRD_table}
+{\input{mohns_rho_R_PRD_table}}
+\end{table}

### 2025-09-10 Project_0/manuscript/manuscript_R1.tex
saving work on revision

@@
 \usepackage{amsmath,amssymb}
-\usepackage{lmodern}
 \usepackage{iftex}
@@
 \else % if luatex or xetex
-  \usepackage{unicode-math}
+  \usepackage{unicode-math} % this also loads fontspec
   \defaultfontfeatures{Scale=MatchLowercase}
   \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+\usepackage{lmodern}
+\ifPDFTeX\else
+  % xetex/luatex font selection
 \fi
@@
 \usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={An investigation of factors affecting inferences from and reliability of state-space age-structured assessment models},
-  pdfauthor={Timothy J. Miller1,2; Greg Britten3; Elizabeth N. Brooks2; Gavin Fay4; Alex Hansell2; Christopher M. Legault2; Chengxue Li2; Brandon Muffley5; Brian C. Stock6; John Wiedenmann7},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
 \usepackage[margin=1in]{geometry}
@@
 \setcounter{secnumdepth}{5}
+% definitions for citeproc citations
+\NewDocumentCommand\citeproctext{}{}
+\NewDocumentCommand\citeproc{mm}{%
+  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
+\makeatletter
+ % allow citations to break across lines
+ \let\@cite@ofmt\@firstofone
+ % avoid brackets around text for \cite:
+ \def\@biblabel#1{}
+ \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
+\makeatother
 \newlength{\cslhangindent}
@@
 \setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
+\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
+ {\begin{list}{}{%
+  \setlength{\itemindent}{0pt}
+  \setlength{\leftmargin}{0pt}
+  \setlength{\parsep}{0pt}
   % turn on hanging indent if param 1 is 1
   \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
+   \setlength{\leftmargin}{\cslhangindent}
+   \setlength{\itemindent}{-1\cslhangindent}
   \fi
   % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
+  \setlength{\itemsep}{#2\baselineskip}}}
+ {\end{list}}
 \usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
@@
 \fi
+\usepackage{bookmark}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same}
+\hypersetup{
+  pdftitle={An investigation of factors affecting inferences from and reliability of state-space age-structured assessment models},
+  pdfauthor={Timothy J. Miller1,2; Greg Britten3; Elizabeth N. Brooks2; Gavin Fay4; Alex Hansell2; Christopher M. Legault2; Chengxue Li2; Brandon Muffley5; Brian C. Stock6; John Wiedenmann7},
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
 
@@
 C. Stock\textsuperscript{6} \and John Wiedenmann\textsuperscript{7}}
-\date{08 August, 2025}
+\date{10 September, 2025}
 
@@
 
-\hypertarget{abstract}{%
-\subsection*{Abstract}\label{abstract}}
+\subsection*{Abstract}\label{abstract}
 \addcontentsline{toc}{subsection}{Abstract}
@@
 
-\hypertarget{introduction}{%
-\section*{Introduction}\label{introduction}}
+\section*{Introduction}\label{introduction}
 \addcontentsline{toc}{section}{Introduction}
@@
 also evaluate when retrospective patterns occur and the degree of bias
-in the outputs of the assessment model that are important for
-management.
-
-\hypertarget{methods}{%
-\section*{Methods}\label{methods}}
+in outputs of the assessment model that are important for management.
+
+\section*{Methods}\label{methods}
 \addcontentsline{toc}{section}{Methods}
@@
 
-\hypertarget{operating-models}{%
-\subsection*{Operating models}\label{operating-models}}
+\subsection*{Operating models}\label{operating-models}
 \addcontentsline{toc}{subsection}{Operating models}
 
-\hypertarget{population}{%
-\subsubsection*{Population}\label{population}}
+\subsubsection*{Population}\label{population}
 \addcontentsline{toc}{subsubsection}{Population}
@@
 
-\hypertarget{fleets}{%
-\subsubsection*{Fleets}\label{fleets}}
+\subsubsection*{Fleets}\label{fleets}
 \addcontentsline{toc}{subsubsection}{Fleets}
@@
 
-\hypertarget{indices}{%
-\subsubsection*{Indices}\label{indices}}
+\subsubsection*{Indices}\label{indices}
 \addcontentsline{toc}{subsubsection}{Indices}
@@
 
-\hypertarget{observation-uncertainty}{%
-\subsubsection*{Observation Uncertainty}\label{observation-uncertainty}}
+\subsubsection*{Observation Uncertainty}\label{observation-uncertainty}
 \addcontentsline{toc}{subsubsection}{Observation Uncertainty}
@@
 
-\hypertarget{operating-models-with-random-effects-on-numbers-at-age}{%
 \subsubsection*{Operating models with random effects on numbers at
-age}\label{operating-models-with-random-effects-on-numbers-at-age}}
+age}\label{operating-models-with-random-effects-on-numbers-at-age}
 \addcontentsline{toc}{subsubsection}{Operating models with random
@@
 
-\hypertarget{operating-models-with-random-effects-on-natural-mortality}{%
 \subsubsection*{Operating models with random effects on natural
-mortality}\label{operating-models-with-random-effects-on-natural-mortality}}
+mortality}\label{operating-models-with-random-effects-on-natural-mortality}
 \addcontentsline{toc}{subsubsection}{Operating models with random
@@
 
-\hypertarget{operating-models-with-random-effects-on-fleet-selectivity}{%
 \subsubsection*{Operating models with random effects on fleet
-selectivity}\label{operating-models-with-random-effects-on-fleet-selectivity}}
+selectivity}\label{operating-models-with-random-effects-on-fleet-selectivity}
 \addcontentsline{toc}{subsubsection}{Operating models with random
@@
 
-\hypertarget{operating-models-with-random-effects-on-index-catchability}{%
 \subsubsection*{Operating models with random effects on index
-catchability}\label{operating-models-with-random-effects-on-index-catchability}}
+catchability}\label{operating-models-with-random-effects-on-index-catchability}
 \addcontentsline{toc}{subsubsection}{Operating models with random
@@
 
-\hypertarget{estimation-models}{%
-\subsection*{Estimation models}\label{estimation-models}}
+\subsection*{Estimation models}\label{estimation-models}
 \addcontentsline{toc}{subsection}{Estimation models}
@@
 
-\hypertarget{measures-of-reliability}{%
-\subsection*{Measures of reliability}\label{measures-of-reliability}}
+\subsection*{Measures of reliability}\label{measures-of-reliability}
 \addcontentsline{toc}{subsection}{Measures of reliability}
 
-\hypertarget{convergence}{%
-\subsubsection*{Convergence}\label{convergence}}
+\subsubsection*{Convergence}\label{convergence}
 \addcontentsline{toc}{subsubsection}{Convergence}
@@
 
-\hypertarget{aic-for-model-selection}{%
-\subsubsection*{AIC for model selection}\label{aic-for-model-selection}}
+\subsubsection*{AIC for model selection}\label{aic-for-model-selection}
 \addcontentsline{toc}{subsubsection}{AIC for model selection}
@@
 
-\hypertarget{bias}{%
-\subsubsection*{Bias}\label{bias}}
+\subsubsection*{Bias}\label{bias}
 \addcontentsline{toc}{subsubsection}{Bias}
 
-For a given model attribute we calculated the relative error \[
+For a given model attribute we calculated the relative error
+\begin{equation}\label{relerror}
 \text{RE}\left(\theta_i\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
-\] from fitting a given estimating model to simulated data set \(i\)
-configured for a given OM where \(\widehat \theta_i\) and \(\theta_i\)
-are the estimated and true values for simulation \(i\). We estimated
-bias as the median of the relative errors across all simulations for a
-given OM and EM combination. We constructed 95\% confidence intervals
-for the median relative bias using the binomial distribution approach as
-in Miller and Hyun (2018) and Stock and Miller (2021). We present
-results for bias in terminal year estimates of SSB and recruitment,
-Beverton-Holt stock recruit parameters (\(a\) and \(b\)), and median
-natural mortality rate. Results for terminal year fishing mortality were
-strongly negatively correlated with those for SSB and are provided in
-the Supplementary Materials.
-
-\hypertarget{mohns-rho}{%
+\end{equation} from fitting a given estimating model to simulated data
+set \(i\) configured for a given OM where \(\widehat \theta_i\) and
+\(\theta_i\) are the estimated and true values for simulation \(i\). We
+estimated bias as the median of the relative errors across all
+simulations for a given OM and EM combination. We constructed 95\%
+confidence intervals for the median relative bias using the binomial
+distribution approach as in Miller and Hyun (2018) and Stock and Miller
+(2021). We present results for bias in terminal year estimates of SSB
+and recruitment, Beverton-Holt stock recruit parameters (\(a\) and
+\(b\)), and median natural mortality rate. Results for terminal year
+fishing mortality were strongly negatively correlated with those for SSB
+and are provided in the Supplementary Materials.
+
 \subsubsection*{\texorpdfstring{Mohn's
-\(\rho\)}{Mohn's \textbackslash rho}}\label{mohns-rho}}
+\(\rho\)}{Mohn's \textbackslash rho}}\label{mohns-rho}
 \addcontentsline{toc}{subsubsection}{Mohn's \(\rho\)}
 
-We calculated Mohn's \(\rho\) for SSB, fully-selected fishing mortality,
-and recruitment for each EM (Mohn 1999). We fit 7 peels for each EM and
-calculated median 95\% confidence intervals for Mohn's \(\rho\) using
-the same methods as that for relative bias.
-
-\hypertarget{summarizing-results-across-oms}{%
-\subsubsection*{Summarizing results across
-OMs}\label{summarizing-results-across-oms}}
-\addcontentsline{toc}{subsubsection}{Summarizing results across OMs}
-
-Because of the numerous factors investigated for each of the ways we
-assessed reliability of the state-space assessment models, we used two
-methods to summarize the most important factors that explain differences
-for each measure of reliability. The first method was fitting regression
-models with the response being each of the measures of reliability
-described above and predictor variables were defined based on OM and EM
-characteristics. For convergence results we performed logistic
-regressions and for AIC results we performed multinomial regressions for
-the type of process error that provided the lowest AIC. For other
-measures of reliability we fit standard linear regression models.
-Because relative bias and Mohn's \(\rho\) for the various parameters are
-bounded below at -1, we used as a response the log of the absolute
-log-scale errors \begin{equation}\label{bias_regression_response}
-y_i = \log\left|\log\left(\widehat \theta_i\right) - \log(\left(\theta_i\right)\right|,
-\end{equation} so that lower values measure less bias whatever the sign.
-For relative bias we also restricted the fits to simulations where the
-estimated CV of the parameter or model output was less than 100\%
-because values greater than this would typically lead the assessment
-scientist away from the particular configuration of the estimating
-model. For all regressions we fit models with factors included
-individually and with all factors and with all second and third order
-interactions. For the multinomial regression, we used the \verb|vglm|
-function from the VGAM package (Yee 2008; Yee 2015). We tabulated
-percent reduction in residual deviance for each of regression fits.
+We calculated Mohn's \(\rho\) for SSB, fishing mortality (averaged over
+all age classes), and recruitment for each EM (Mohn 1999). We fit
+\(P =7\) peels to each simulated data set and calculated Mohn's \(\rho\)
+for a given attribute \(\theta\) as \begin{equation}\label{mohns_rho}
+\rho\left(\theta\right) = \frac{1}{P}\sum^P_{p=1} \frac{\widehat \theta_{Y-p,Y-p} - \widehat\theta_{Y-p,Y}}{\widehat\theta_{Y-p,Y}}
+\end{equation} where \(\widehat \theta_{i,j}\) is the estimate for
+attribute \(\theta\) in year \(i\) from a model fit using data up to
+year \(j\). For each EM we calculated median and 95\% confidence
+intervals using the same methods as that for relative bias.
+
+\subsubsection*{Summarizing results across OM and EM
+attributes}\label{summarizing-results-across-om-and-em-attributes}
+\addcontentsline{toc}{subsubsection}{Summarizing results across OM and
+EM attributes}
+
+The measures of central tendency and variability of of observed values
+for specific OM and EM attributes (e.g., low or high observation error)
+that we described above can indicate scenarios that provide better or
+poorer reliability. However, the OM and EM attributes that we
+investigated are numerous, so we used two methods to summarize the most
+important factors for differences in results. The first method was
+fitting regression models with the response being each of the measures
+of reliability described above and predictor variables were defined
+based on OM and EM characteristics (e.g., MacKinnon et al. 1995; Wang et
+al. 2017; Harwell et al. 2018). For the binary results for convergence
+and AIC-based selection of a stock-recruit relationship, we performed
+logistic regressions. For results for AIC-based selection of EM process
+error type (multiple categories) we performed multinomial regressions.
+For other measures of reliability we fit standard linear regression
+models. Because relative errors (Eq. \ref{relerror}) and Mohn's \(\rho\)
+for the various parameters are bounded below at -1, we used a
+transformation of these values
+\begin{equation}\label{bias_regression_response}
+y_i = \log\left[f\left(\widehat \theta_i,\theta_i\right)+1\right]
+\end{equation} where \(f\) is either the relative error (Eq.
+\ref{relerror}) or Mohn's \(\rho\) (Eq. \ref{mohns_rho}) for simulation
+\(i\), so that values are unbounded. For relative errors, \(y_i\) is the
+log-scale error. We omitted simulation estimates equal to zero (RE =
+-1). For all regressions we fit separate models with individual factors
+included, with all factors combined, with including all second order
+interactions, and including all third order interactions. For the
+multinomial regression, we used the \verb|vglm| function from the VGAM
+package (Yee 2008; Yee 2015). We tabulated percent reduction in residual
+deviance for each of regression fits. We did not perform formal
+statistical analyses of effects of OM and EM attributes on results
+(e.g., ANOVA) because of the lack of independence of the
+``observations'' that results from fitting multiple EMs to each
+simulated data set.
 
 The second method involved fitting classification and regression trees
-(Breiman et al. 1984) to show how the primary factors that split the
-values for each measure of reliability and subsets of the measures that
-were alloted to each branch of the splits. Classification trees were fit
-to categorical measures (convergence and AIC) and regression trees were
-fit to the other measures with continuous scales. The response variables
-were the same as the regressions for the deviance reduction analyses. We
-used the \verb|rpart| function in the rpart package
-(\url{https://cran.r-project.org/package=rpart}) to fit trees and we
-pruned them to show only the primary branches for clarity. Generally,
-the factors defining primary splits for the trees corresponded to the
-factors with the largest reductions in deviance fitting the regression
-models.
-
-\hypertarget{results}{%
-\section*{Results}\label{results}}
+(Breiman et al. 1984) to show how the OM and EM attributes, and their
+interactions, split the values for each measure of reliability (e.g.,
+Gonzalez et al. 2018; Collier et al. 2022). We used classification trees
+for categorical measures (convergence and AIC) and regression trees for
+the other measures with continuous scales (relative error and Mohn's
+\(\rho\)). The response variables were the same as the regressions for
+the deviance reduction analyses. We used the \verb|rpart| function in
+the rpart package (\url{https://cran.r-project.org/package=rpart}) to
+fit trees and, for clarity we pruned them to show the primary branches.
+
+\section*{Results}\label{results}
 \addcontentsline{toc}{section}{Results}
 
-\hypertarget{convergence-performance}{%
-\subsection*{Convergence performance}\label{convergence-performance}}
+\subsection*{Convergence performance}\label{convergence-performance}
 \addcontentsline{toc}{subsection}{Convergence performance}
 
-\hypertarget{deviance-reduction-and-classification-trees}{%
 \subsubsection*{Deviance reduction and classification
-trees}\label{deviance-reduction-and-classification-trees}}
+trees}\label{deviance-reduction-and-classification-trees}
 \addcontentsline{toc}{subsubsection}{Deviance reduction and
@@
 
-\hypertarget{full-results}{%
-\subsubsection*{Full results}\label{full-results}}
+EM process error assumption was the single attribute that resulted in
+the largest percent reduction in deviance for all OM process error types
+other than R+S OMs where the EM M assumption explained the most residual
+deviance (Table \ref{convergence_PRD_table}). However, also including
+interactions of OM and EM factors also provided large reductions in
+residual deviance, suggesting successful convergence depended on a
+combination OM and EM attributes. Classification trees for each OM
+process error type, all had the primary branch defined using the same
+attribute that provided the largest reduction in deviance (Figure
+\ref{conv_class}).
+
+Generally, the factors defining primary splits for the trees
+corresponded to those with the largest reductions in deviance fitting
+the regression models.
+
+\subsubsection*{Full results}\label{full-results}
 \addcontentsline{toc}{subsubsection}{Full results}
@@
 
-\hypertarget{aic-performance-for-process-error-structure}{%
 \subsection*{AIC performance for process error
-structure}\label{aic-performance-for-process-error-structure}}
+structure}\label{aic-performance-for-process-error-structure}
 \addcontentsline{toc}{subsection}{AIC performance for process error
@@
 
-\hypertarget{deviance-reduction-and-classification-trees-1}{%
 \subsubsection*{Deviance reduction and classification
-trees}\label{deviance-reduction-and-classification-trees-1}}
+trees}\label{deviance-reduction-and-classification-trees-1}
 \addcontentsline{toc}{subsubsection}{Deviance reduction and
@@
 
-\hypertarget{full-results-1}{%
-\subsubsection*{Full results}\label{full-results-1}}
+\subsubsection*{Full results}\label{full-results-1}
 \addcontentsline{toc}{subsubsection}{Full results}
@@
 
-\hypertarget{aic-performance-for-the-stock-recruit-relationship}{%
 \subsection*{AIC performance for the stock-recruit
-relationship}\label{aic-performance-for-the-stock-recruit-relationship}}
+relationship}\label{aic-performance-for-the-stock-recruit-relationship}
 \addcontentsline{toc}{subsection}{AIC performance for the stock-recruit
@@
 
-\hypertarget{deviance-reduction-and-classification-trees-2}{%
 \subsubsection*{Deviance reduction and classification
-trees}\label{deviance-reduction-and-classification-trees-2}}
+trees}\label{deviance-reduction-and-classification-trees-2}
 \addcontentsline{toc}{subsubsection}{Deviance reduction and
@@
 
-\hypertarget{full-results-2}{%
-\subsubsection*{Full results}\label{full-results-2}}
+\subsubsection*{Full results}\label{full-results-2}
 \addcontentsline{toc}{subsubsection}{Full results}
@@
 
-\hypertarget{bias-1}{%
-\subsection*{Bias}\label{bias-1}}
+\subsection*{Bias}\label{bias-1}
 \addcontentsline{toc}{subsection}{Bias}
 
-\hypertarget{deviance-reduction-and-regression-trees}{%
 \subsubsection*{Deviance reduction and regression
-trees}\label{deviance-reduction-and-regression-trees}}
+trees}\label{deviance-reduction-and-regression-trees}
 \addcontentsline{toc}{subsubsection}{Deviance reduction and regression
 trees}
+
+Across all OM process error types, fishing history was the most
+important single factor in reducing residual deviation in log-scale
+errors for the Beverton-Holt density-independent parameter \(a\), and
+also for the density-independent parameter \(b\) for R+M, R+Sel, and R+q
+OMs (Table \ref{bias_SR_pars_PRD_table}). For R and R+S OMs the EM
+process error assumption was equally important for estimating \(b\), but
+the the reductions were small (mostly less than 4\%) for any factor and
+any OM process error type. There was also no substantial reduction in
+deviance with second or third order interactions of factors for either
+parameter in any OM process error types.
+
+The regression trees indicate that having contrast in fishing pressure
+reduced the median error for both SR parameters for all OM types
 
@@
 that some combinations of factors could reduce bias for the \(b\)
-parameter (Table \ref{}). Despite the explanation variation in the \(b\)
-estimation erorrs, bias was still significant regardless of the
-combination of OM and EM factors (Figures regtrees and full results).
-
-\hypertarget{full-results-3}{%
-\subsubsection*{Full results}\label{full-results-3}}
+parameter. Despite the explanation variation in the \(b\) estimation
+erorrs, bias was still significant regardless of the combination of OM
+and EM factors (Figures regtrees and full results).
+
+Median natural mortality rate showed reductions in deviance for R, R+S,
+and R+M OMs with second order interactions and R+M also showed
+reductions with third order interactions.
+
+\subsubsection*{Full results}\label{full-results-3}
 \addcontentsline{toc}{subsubsection}{Full results}
 
-\hypertarget{spawning-stock-biomass}{%
-\paragraph*{Spawning stock biomass}\label{spawning-stock-biomass}}
+\paragraph*{Spawning stock biomass}\label{spawning-stock-biomass}
 \addcontentsline{toc}{paragraph}{Spawning stock biomass}
@@
 
-\hypertarget{beverton-holt-parameters}{%
 \subsubsection*{Beverton-Holt
-parameters}\label{beverton-holt-parameters}}
+parameters}\label{beverton-holt-parameters}
 \addcontentsline{toc}{subsubsection}{Beverton-Holt parameters}
@@
 
-\hypertarget{median-natural-mortality-rate}{%
 \subsubsection*{Median natural mortality
-rate}\label{median-natural-mortality-rate}}
+rate}\label{median-natural-mortality-rate}
 \addcontentsline{toc}{subsubsection}{Median natural mortality rate}
@@
 
-\hypertarget{mohns-rho-1}{%
 \subsection*{\texorpdfstring{Mohn's
-\(\rho\)}{Mohn's \textbackslash rho}}\label{mohns-rho-1}}
+\(\rho\)}{Mohn's \textbackslash rho}}\label{mohns-rho-1}
 \addcontentsline{toc}{subsection}{Mohn's \(\rho\)}
 
-\hypertarget{deviance-reduction-and-regression-trees-1}{%
 \subsubsection*{Deviance reduction and regression
-trees}\label{deviance-reduction-and-regression-trees-1}}
+trees}\label{deviance-reduction-and-regression-trees-1}
 \addcontentsline{toc}{subsubsection}{Deviance reduction and regression
@@
 
-\hypertarget{full-results-4}{%
-\subsubsection*{Full results}\label{full-results-4}}
+Fits of glms showed little explanatory power of any of the OM an EM
+attributes for Mohn's rho for SSB, F, and R. Note that the factors that
+showed the largest reductions in deviance (which were very small) often
+do not comport with factors that show differences in median Mohn's rho
+results. This is because the glms and regression trees are reducing
+residual variance and there could be large variability in values, but
+the medians are less sensitive to extreme values. For example we can see
+that there is large Mohn's rho for R in R OMs with larger variability
+assumed in recruitment and higher observation error (Figure X), but
+these factors suggest little reduction in residual deviance (Table X).
+
+\subsubsection*{Full results}\label{full-results-4}
 \addcontentsline{toc}{subsubsection}{Full results}
@@
 
-\hypertarget{discussion}{%
-\section*{Discussion}\label{discussion}}
+\section*{Discussion}\label{discussion}
 \addcontentsline{toc}{section}{Discussion}
 
-\hypertarget{convergence-1}{%
-\subsection*{Convergence}\label{convergence-1}}
+\subsection*{Convergence}\label{convergence-1}
 \addcontentsline{toc}{subsection}{Convergence}
@@
 
-\hypertarget{aic}{%
-\subsection*{AIC}\label{aic}}
+\subsection*{AIC}\label{aic}
 \addcontentsline{toc}{subsection}{AIC}
@@
 
-\hypertarget{bias-2}{%
-\subsection*{Bias}\label{bias-2}}
+\subsection*{Bias}\label{bias-2}
 \addcontentsline{toc}{subsection}{Bias}
@@
 
-\hypertarget{retrospective-patterns}{%
-\subsection*{Retrospective patterns}\label{retrospective-patterns}}
+\subsection*{Retrospective patterns}\label{retrospective-patterns}
 \addcontentsline{toc}{subsection}{Retrospective patterns}
@@
 
-\hypertarget{conclusions}{%
-\subsection*{Conclusions}\label{conclusions}}
+\subsection*{Conclusions}\label{conclusions}
 \addcontentsline{toc}{subsection}{Conclusions}
@@
 
-\hypertarget{acknowledgements}{%
-\section*{Acknowledgements}\label{acknowledgements}}
+\section*{Acknowledgements}\label{acknowledgements}
 \addcontentsline{toc}{section}{Acknowledgements}
@@
 
-\hypertarget{references}{%
-\section*{References}\label{references}}
+\section*{References}\label{references}
 \addcontentsline{toc}{section}{References}
 
-\hypertarget{refs}{}
+\phantomsection\label{refs}
 \begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-aeberhardetal18}{}}%
+\bibitem[\citeproctext]{ref-aeberhardetal18}
 Aeberhard, W.H., Flemming, J.M., and Nielsen, A. 2018. Review of
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-augeretal16}{}}%
+\bibitem[\citeproctext]{ref-augeretal16}
 Auger-Méthé, M., Field, C., Albertsen, C.M., Derocher, A.E., Lewis,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-augeretal21}{}}%
+\bibitem[\citeproctext]{ref-augeretal21}
 Auger-Méthé, M., Newman, K., Cole, D., Empacher, F., Gryba, R., King,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-breimanetal84}{}}%
+\bibitem[\citeproctext]{ref-breimanetal84}
 Breiman, L., Friedman, J.H., Olshen, R.A., and Stone, C.J. 1984.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-breiviketal23}{}}%
+\bibitem[\citeproctext]{ref-breiviketal23}
 Breivik, O.N., Aldrin, M., Fuglebakk, E., and Nielsen, A. 2023.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-brittenetalinreview}{}}%
+\bibitem[\citeproctext]{ref-brittenetalinreview}
 Britten, G., Brooks, E.N., and Miller, T.J. In review. Identification
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-brookslegault16}{}}%
+\bibitem[\citeproctext]{ref-brookslegault16}
 Brooks, E.N., and Legault, C.M. 2016. Retrospective forecasting --
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-cadigan16}{}}%
+\bibitem[\citeproctext]{ref-cadigan16}
 Cadigan, N.G. 2016. A state-space stock assessment model for northern
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-carvalhoetal21}{}}%
+\bibitem[\citeproctext]{ref-carvalhoetal21}
 Carvalho, F., Winker, H., Courtney, D., Kapur, M., Kell, L., Cardinale,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-clopperpearson34}{}}%
+\bibitem[\citeproctext]{ref-clopperpearson34}
 Clopper, C.J., and Pearson, E.S. 1934. The use of confidence or fiducial
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-connetal10}{}}%
+\bibitem[\citeproctext]{ref-collieretal22}
+Collier, Z.K., Zhang, H., and Soyoye, O. 2022. Alternative methods for
+interpreting {M}onte {C}arlo experiments. Communications in Statistics -
+Simulation and Computation: 1--16.
+doi:\href{https://doi.org/10.1080/03610918.2022.2082474}{10.1080/03610918.2022.2082474}.
+
+\bibitem[\citeproctext]{ref-connetal10}
 Conn, P.B., Williams, E.H., and Shertzer, K.W. 2010. When can we
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-croninpunt21}{}}%
+\bibitem[\citeproctext]{ref-croninpunt21}
 Cronin-Fine, L., and Punt, A.E. 2021. Modeling time-varying selectivity
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-devalpinehastings02}{}}%
+\bibitem[\citeproctext]{ref-devalpinehastings02}
 de Valpine, P., and Hastings, A. 2002. Fitting population models
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-fischetal23}{}}%
+\bibitem[\citeproctext]{ref-fischetal23}
 Fisch, N., Shertzer, K., Camp, E., Maunder, M., and Ahrens, R. 2023.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-fleischmanetal13}{}}%
+\bibitem[\citeproctext]{ref-fleischmanetal13}
 Fleischman, S.J., Catalano, M.J., Clark, R.A., and Bernard, D.R. 2013.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-hoenigetal25}{}}%
+\bibitem[\citeproctext]{ref-gonzalezetal18}
+Gonzalez, O., O'Rourke, H.P., Wurpts, I.C., and Grimm, K.J. 2018.
+Analyzing {M}onte {C}arlo simulation studies with classification and
+regression trees. Structural Equation Modeling: A Multidisciplinary
+Journal \textbf{25}(3): 403--413.
+doi:\href{https://doi.org/10.1080/10705511.2017.1369353}{10.1080/10705511.2017.1369353}.
+
+\bibitem[\citeproctext]{ref-harwelletal18}
+Harwell, M., Kohli, N., and Peralta-Torres, Y. 2018. A survey of
+reporting practices of computer simulation studies in statistical
+research. The American Statistician \textbf{72}(4): 321--327.
+doi:\href{https://doi.org/10.1080/00031305.2017.1342692}{10.1080/00031305.2017.1342692}.
+
+\bibitem[\citeproctext]{ref-hoenigetal25}
 Hoenig, J.M., Hearn, W.S., Leigh, G.M., and Latour, R.J. 2025.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-hoyleetal22}{}}%
+\bibitem[\citeproctext]{ref-hoyleetal22}
 Hoyle, S.D., Maunder, M.N., Punt, A.E., Mace, P.M., Devine, J.A., and
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-hurtadoferroetal15}{}}%
+\bibitem[\citeproctext]{ref-hurtadoferroetal15}
 Hurtado-Ferro, F., Szuwalski, C.S., Valero, J.L., Anderson, S.C.,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-huynhetal22}{}}%
+\bibitem[\citeproctext]{ref-huynhetal22}
 Huynh, Q.C., Legault, C.M., Hordyk, A.R., and Carruthers, T.R. 2022. A
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-johnsonetal16}{}}%
+\bibitem[\citeproctext]{ref-johnsonetal16}
 Johnson, K.F., Councill, E., Thorson, J.T., Brooks, E., Methot, R.D.,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-kapuretal24}{}}%
+\bibitem[\citeproctext]{ref-kapuretal24}
 Kapur, M.S., Ducharme-Barth, N., Oshima, M., and Carvalho, F. 2025. Good
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-kassraftery95}{}}%
+\bibitem[\citeproctext]{ref-kassraftery95}
 Kass, R.E., and Raftery, A.E. 1995. Bayes factors. Journal of the
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-katz81}{}}%
+\bibitem[\citeproctext]{ref-katz81}
 Katz, R.W. 1981. On some criteria for estimating the order of a {M}arkov
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-knape08}{}}%
+\bibitem[\citeproctext]{ref-knape08}
 Knape, J. 2008. Estimability of density dependence in models of time
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-kristensenetal16}{}}%
+\bibitem[\citeproctext]{ref-kristensenetal16}
 Kristensen, K., Nielsen, A., Berg, C.W., Skaug, H., and Bell, B.M. 2016.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-leeetal11}{}}%
+\bibitem[\citeproctext]{ref-leeetal11}
 Lee, H.-H., Maunder, M.N., Piner, K.R., and Methot, R.D. 2011.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-legault09}{}}%
+\bibitem[\citeproctext]{ref-legault09}
 Legault, C.M. 2009. Report of the retrospective working group, 14-16
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-legaultpalmer16}{}}%
+\bibitem[\citeproctext]{ref-legaultpalmer16}
 Legault, C.M., and Palmer, M.C. 2016. In what direction should the
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-legaultrestrepo99}{}}%
+\bibitem[\citeproctext]{ref-legaultrestrepo99}
 Legault, C.M., and Restrepo, V.R. 1999. A flexible forward
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-legaultetal23}{}}%
+\bibitem[\citeproctext]{ref-legaultetal23}
 Legault, C.M., Wiedenmann, J., Deroba, J.J., Fay, G., Miller, T.J.,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-lietalinreview_a}{}}%
+\bibitem[\citeproctext]{ref-lietalinreview_a}
 Li, C., Deroba, J.J., Berger, A.M., Goethel, D.R., Langseth, B.J.,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-lietalinreview}{}}%
+\bibitem[\citeproctext]{ref-lietalinreview}
 Li, C., Deroba, J.J., Miller, T.J., Legault, C.M., and Perretti, C. In
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-lietal24}{}}%
+\bibitem[\citeproctext]{ref-lietal24}
 Li, C., Deroba, J.J., Miller, T.J., Legault, C.M., and Perretti, C.T.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-liljestrandetal24}{}}%
+\bibitem[\citeproctext]{ref-liljestrandetal24}
 Liljestrand, E.M., Bence, J.R., and Deroba, J.J. 2024. The effect of
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-magnussonhilborn07}{}}%
+\bibitem[\citeproctext]{ref-mackinnonetal95}
+MacKinnon, D.P., Warsi, G., and Dwyer, J.H. 1995. A simulation study of
+mediated effect measures. Multivariate Behavioral Research
+\textbf{30}(1): 41--62.
+doi:\href{https://doi.org/10.1207/s15327906mbr3001/_3}{10.1207/s15327906mbr3001\textbackslash\_3}.
+
+\bibitem[\citeproctext]{ref-magnussonhilborn07}
 Magnusson, A., and Hilborn, R. 2007. What makes fisheries data
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-methotwetzel13}{}}%
+\bibitem[\citeproctext]{ref-methotwetzel13}
 Methot, R.D., and Wetzel, C.R. 2013. Stock synthesis: A biological and
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-millerbrooks21}{}}%
+\bibitem[\citeproctext]{ref-millerbrooks21}
 Miller, T.J., and Brooks, E.N. 2021. Steepness is a slippery slope. Fish
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-milleretal16}{}}%
+\bibitem[\citeproctext]{ref-milleretal16}
 Miller, T.J., Hare, J.A., and Alade, L. 2016. A state-space approach to
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-millerhyun18}{}}%
+\bibitem[\citeproctext]{ref-millerhyun18}
 Miller, T.J., and Hyun, S.-Y. 2018. Evaluating evidence for alternative
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-millerlegault17}{}}%
+\bibitem[\citeproctext]{ref-millerlegault17}
 Miller, T.J., and Legault, C.M. 2017. Statistical behavior of
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-milleretal18}{}}%
+\bibitem[\citeproctext]{ref-milleretal18}
 Miller, T.J., O'Brien, L., and Fratantoni, P.S. 2018. Temporal and
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-millerstock20}{}}%
+\bibitem[\citeproctext]{ref-millerstock20}
 Miller, T.J., and Stock, B.C. 2020. The {Woods Hole Assessment Model}
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-mohn99}{}}%
+\bibitem[\citeproctext]{ref-mohn99}
 Mohn, R. 1999. The retrospective problem in sequential population
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-nefsc22}{}}%
+\bibitem[\citeproctext]{ref-nefsc22}
 NEFSC. 2022a. Final report of the haddock research track assessment
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-nefsc22a}{}}%
+\bibitem[\citeproctext]{ref-nefsc22a}
 NEFSC. 2022b. Report of the {A}merican plaice research track working
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-nefsc24}{}}%
+\bibitem[\citeproctext]{ref-nefsc24}
 NEFSC. 2024. Butterfish research track assessment report. US Dept Commer
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-nefsc25}{}}%
+\bibitem[\citeproctext]{ref-nefsc25}
 NEFSC. 2025. Yellowttail flounder research track working group report.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-nielsenberg14}{}}%
+\bibitem[\citeproctext]{ref-nielsenberg14}
 Nielsen, A., and Berg, C.W. 2014. Estimation of time-varying selectivity
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-pedersenberg17}{}}%
+\bibitem[\citeproctext]{ref-pedersenberg17}
 Pedersen, M.W., and Berg, C.W. 2017. A stochastic surplus production
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-perrettietal20}{}}%
+\bibitem[\citeproctext]{ref-perrettietal20}
 Perretti, C.T., Deroba, J.J., and Legault, C.M. 2020. Simulation testing
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-polanskyetal09}{}}%
+\bibitem[\citeproctext]{ref-polanskyetal09}
 Polansky, L., De Valpine, P., Lloyd-Smith, J.O., and Getz, W.M. 2009.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-dupontaviceetal22}{}}%
+\bibitem[\citeproctext]{ref-dupontaviceetal22}
 Pontavice, H. du, Miller, T.J., Stock, B.C., Chen, Z., and Saba, V.S.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-punt23}{}}%
+\bibitem[\citeproctext]{ref-punt23}
 Punt, A.E. 2023. Those who fail to learn from history are condemned to
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-puntetal14}{}}%
+\bibitem[\citeproctext]{ref-puntetal14}
 Punt, A.E., Hurtado-Ferro, F., and Whitten, A.R. 2014. Model selection
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-ruddthorson18}{}}%
+\bibitem[\citeproctext]{ref-ruddthorson18}
 Rudd, M.B., and Thorson, J.T. 2018. Accounting for variable recruitment
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-shibata76}{}}%
+\bibitem[\citeproctext]{ref-shibata76}
 Shibata, R. 1976. Selection of the order of an autoregressive model by
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-stewartmonnahan17}{}}%
+\bibitem[\citeproctext]{ref-stewartmonnahan17}
 Stewart, I.J., and Monnahan, C.C. 2017. Implications of process error in
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-stockmiller21}{}}%
+\bibitem[\citeproctext]{ref-stockmiller21}
 Stock, B.C., and Miller, T.J. 2021. The {W}oods {H}ole {A}ssessment
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-stocketal21}{}}%
+\bibitem[\citeproctext]{ref-stocketal21}
 Stock, B.C., Xu, H., Miller, T.J., Thorson, J.T., and Nye, J.A. 2021.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-szuwalskietal18}{}}%
+\bibitem[\citeproctext]{ref-szuwalskietal18}
 Szuwalski, C.S., Ianelli, J.N., and Punt, A.E. 2018. Reducing
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-thorsonminto15}{}}%
+\bibitem[\citeproctext]{ref-thorsonminto15}
 Thorson, J.T., and Minto, C. 2015. Mixed effects: A unifying framework
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-thulin14}{}}%
+\bibitem[\citeproctext]{ref-thulin14}
 Thulin, M. 2014. {The cost of using exact confidence intervals for a
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-trijouletetal20}{}}%
+\bibitem[\citeproctext]{ref-trijouletetal20}
 Trijoulet, V., Fay, G., and Miller, T.J. 2020. Performance of a
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-wiedenmannetal19}{}}%
+\bibitem[\citeproctext]{ref-wangetal17}
+Wang, S., Cadigan, N.G., and Benoît, H.P. 2017. Inference about
+regression parameters using highly stratified survey count data with
+over-dispersion and repeated measurements. Journal of Applied Statistics
+\textbf{44}(6): 1013--1030.
+doi:\href{https://doi.org/10.1080/02664763.2016.1191622}{10.1080/02664763.2016.1191622}.
+
+\bibitem[\citeproctext]{ref-wiedenmannetal19}
 Wiedenmann, J., Free, C.M., and Jensen, O.P. 2019. Evaluating the
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-xuetal19}{}}%
+\bibitem[\citeproctext]{ref-xuetal19}
 Xu, H., Thorson, J.T., Methot, R.D., and Taylor, I.G. 2019. A new
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-yee08}{}}%
+\bibitem[\citeproctext]{ref-yee08}
 Yee, T.W. 2008. The {VGAM} package. R News \textbf{8}(2): 28--39.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-yee15}{}}%
+\bibitem[\citeproctext]{ref-yee15}
 Yee, T.W. 2015. Vector generalized linear and additive models: With an
@@
 
-\hypertarget{supplementary-materials}{%
-\section*{Supplementary Materials}\label{supplementary-materials}}
+\section*{Supplementary Materials}\label{supplementary-materials}
 \addcontentsline{toc}{section}{Supplementary Materials}
@@
 \pagebreak
+
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width = \textwidth]{om_input_plots_figure}
+\end{center}
+\caption{The proportion mature at age, weight at age, fleet and index selectivity at age, and Beverton-Holt stock-recruit relationship assumed for the population in all operating models.
+For operating models with random effects on fleet selectivity, this represents the selectivity at the mean of the random effects.}\label{om_inputs_fig}
+\end{figure}
 
@@
 
-\clearpage
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{hess_grad_convergence_plots}
-\end{center}
-\caption{The maximum of the absolute values of all gradient values for all fits that provided hessian-based standard errors across all simuated data sets of a given OM configuration (A: R and R+S, B: R+M, C: R+Sel, or D: R+q).
-Results are conditional on EM fits with alternative process error type (colored points and lines), median natural mortality (estimated or known) and recruitment assumptions (Beverton-Holt stock-recruit relationship or not).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{hess_grad}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{type_3_convergence_plots}
-\end{center}
-\caption{Probability of estimating models providing maximum absolute values of gradients less than $10^{-6}$ assuming alternative process error (colored points and lines), and median natural mortality (estimated or known) and Beverton-Holt stock-recruit relationships (estimated or not; along x-axis) when fitted to operating models that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{gradient_convergence}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{sr_aic_plots}
-\end{center}
-\caption{Estimated probability of lowest AIC from logistic regression on the log-standard deviation of the true log(SSB) in each simulation for estimating model with Beverton-Holt stock-recruit relationships, rather than the otherwise equivalent EM without the stock-recruit relationship.
-Results are conditional on alternative assumptions for median natural mortality (estimated or known) and on EMs having the correct process error structure: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Rug along x-axis denotes $SD(\log(SSB))$ values for each simulation and polygons represent 95\% confidence intervals.}\label{sr_aic_supp}
-\end{figure}
-\end{landscape}
-
-\clearpage
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{term_F_bias_plots}
-\end{center}
-\caption{Median relative error of terminal year fully-selected fishing mortality for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{F_rel_error}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{term_R_bias_plots}
-\end{center}
-\caption{Median relative error of terminal year recruitment for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{R_rel_error}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{sr_bias_plots}
-\end{center}
-\caption{Median relative error of Beverton-Holt stock-recruit parameters ($a$ and $b$) for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{SR_rel_error}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{M_bias_plots}
-\end{center}
-\caption{Median relative error of median natural mortality for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{M_rel_error}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{mohns_rho_F_plots}
-\end{center}
-\caption{Median Mohn's $\rho$ of fishing mortality averaged over all age classes for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_F}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{mohns_rho_R_plots}
-\end{center}
-\caption{Median Mohn's $\rho$ of recruitment for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_R}
-\end{figure}
-\end{landscape}
+\begin{table}
+\caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to transformed Mohn's $\rho$ values for each simulation (Eq.
+\ref{bias_regression_response}) for fishing mortality averaged over all age classes with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{mohns_rho_F_PRD_table}
+{\input{mohns_rho_F_PRD_table}}
+\end{table}
+
+\begin{table}
+\caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to transformed Mohn's $\rho$ values for each simulation (Eq.
+\ref{bias_regression_response}) for recruitment with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{mohns_rho_R_PRD_table}
+{\input{mohns_rho_R_PRD_table}}
+\end{table}
 

### 2025-09-10 Project_0/code/manuscript/aic_SRR_summarize.R

@@ -2,6 +2,9 @@ library(here)
 library(dplyr)
 library(tidyr)
 library(Hmisc)
+#modified rpart.plot package to use plotmath in split labels...
+pkgload::load_all("/home/tmiller/FSAM_research/aug_backup/work/rpart.plot")
+# pkgload::load_all("c:/work/rpart.plot")
 
 aic_fn <- function(x, all_aic, df.oms, df.ems, rec_mod, M_est) {
   if(!is.null(df.oms$NAA_sig)){
@@ -129,7 +132,7 @@ colnames(PRD.table) <- c("R","R+S","R+M","R+Sel","R+q")
 rnames <- gsub("_", " ", All.facs, fixed = TRUE)
 rnames <- gsub("F ", "$F$ ", rnames, fixed = TRUE)
 rnames <- gsub("EM M", "EM $M$ Assumption", rnames, fixed = TRUE)
-rnames <- gsub("log sd log SSB", "$\\log\\left(SD_\\text{SSB}]\\right)$", rnames, fixed = TRUE)
+rnames <- gsub("log sd log SSB", "$\\log\\left(\\text{SD}_\\text{SSB}\\right)$", rnames, fixed = TRUE)
 rnames <- gsub("NAA sigma", "$\\sigma_{2+}$ ", rnames, fixed = TRUE)
 rnames <- gsub("R sigma", "$\\sigma_R$", rnames, fixed = TRUE)
 rnames <- gsub("q sigma", "$\\sigma_q$", rnames, fixed = TRUE)
@@ -145,7 +148,7 @@ y <- interactions.dev.table
 rownames(y) <- c("All factors", "+ All Two Way", "+ All Three Way")
 PRD.table <- rbind(PRD.table,y)
 
-x[] <- format(round(100*PRD.table,2), nsmall = 2)
+x <- format(round(100*PRD.table,2), nsmall = 2)
 dim(x)
 x[which(is.na(as.numeric(x)))] <- "--"
 x[which(as.numeric(x) == 0)] <- "< 0.01"
@@ -159,7 +162,7 @@ x <- latex(x, file = here("Project_0","manuscript","AIC_SRR_PRD_table.tex"),
 library(rpart)
 #modified rpart.plot package to use plotmath in split labels...
 #library(rpart.plot)
-pkgload::load_all("c:/work/rpart.plot")
+# pkgload::load_all("c:/work/rpart.plot")
 split.fun <- function(type = "R") {
[…]

### 2025-09-10 Project_0/code/manuscript/bias_results_summarize.R

@@ -6,7 +6,8 @@ library(reshape2)
 library(rpart)
 library(viridis)
 #modified rpart.plot package to use plotmath in split labels...
-pkgload::load_all("c:/work/rpart.plot")
+pkgload::load_all("/home/tmiller/FSAM_research/aug_backup/work/rpart.plot")
+# pkgload::load_all("c:/work/rpart.plot")
 
 
 make_plot_df <- function(om_type = "naa", res = naa_relSR_results, is_SE = FALSE, M_or_SR = "SR", year = NULL) {
@@ -170,7 +171,7 @@ split.fun <- function(type = "R") {
       labs <- gsub("q rho", "rho['q']", labs, fixed = TRUE) #sigma_q
       labs <- gsub(" M rho", " rho['M']", labs, fixed = TRUE) #sigma_M
       labs <- gsub("R sigma", "sigma[R]", labs, fixed = TRUE) #sigma_R
-      labs <- gsub("NAA SD", "sigma['2+']", labs, fixed = TRUE) #sigma_2+
+      labs <- gsub("NAA sigma", "sigma['2+']", labs, fixed = TRUE) #sigma_2+
       labs <- gsub(" M", "*phantom(0)*italic(M)", labs, fixed = TRUE) #italic M
       labs <- gsub("process error", "Process*phantom(0)*Error", labs, fixed = TRUE)
       labs <- gsub("F ", "italic(F)*phantom(0)*", labs, fixed = TRUE)
@@ -191,7 +192,7 @@ split.fun <- function(type = "R") {
       labs <- gsub("F ", "italic(F)*phantom(0)*", labs, fixed = TRUE)
       labs <- gsub("EM SR", "EM*phantom(0)*SR*phantom(0)*Assumption", labs, fixed = TRUE)
       labs <- gsub("R sigma", "sigma[R]", labs, fixed = TRUE) #sigma_2+
-      labs <- gsub("NAA SD", "sigma['2+']", labs, fixed = TRUE) #sigma_2+
+      labs <- gsub("NAA sigma", "sigma['2+']", labs, fixed = TRUE) #sigma_2+
       labs <- gsub(" ", "*phantom(0)*", labs, fixed = TRUE)
       labs <- gsub("*%", "%", labs, fixed = TRUE)
       labs <- gsub("%*", "%", labs, fixed = TRUE)
@@ -264,16 +265,18 @@ add_to_frame <- function(obj, data){
   origx$frame$median_abs_RE <- NA
[…]

### 2025-09-10 Project_0/code/manuscript/convergence_results_summarize.R

@@ -4,6 +4,10 @@ library(tidyr)
 library(mgcv)
 library(gratia)
 library(Hmisc)
+#modified rpart.plot package to use plotmath in split labels...
+pkgload::load_all("/home/tmiller/FSAM_research/aug_backup/work/rpart.plot")
+# pkgload::load_all("c:/work/rpart.plot")
+
 
 df.oms = list(naa = readRDS(here("Project_0", "inputs", "df.oms.RDS")))
 df.oms$M = readRDS(here("Project_0", "inputs", "df.M.oms.RDS"))

### 2025-09-10 Project_0/code/manuscript/mohns_rho_results_summarize.R

@@ -1,0 +1,468 @@
+library(here)
+library(dplyr)
+library(tidyr)
+library(Hmisc)
+library(reshape2)
+library(rpart)
+library(viridis)
+library(gratia)
+#modified rpart.plot package to use plotmath in split labels...
+pkgload::load_all("/home/tmiller/FSAM_research/aug_backup/work/rpart.plot")
+
+
+make_plot_df <- function(om_type = "naa", res = naa_relSR_results) {
+  df.ems <- readRDS(here::here("Project_0","inputs", "df.ems.RDS"))
+  if(om_type == "naa") {
+    em_ind <- 1:20
+  }
+  if(om_type == "M") {
+    em_ind <- 5:24
+  }
+  if(om_type == "Sel") {
+    em_ind <- c(5:20,25:28)
+  }
+  if(om_type == "q") {
+    em_ind <- c(5:20,29:32)
+  }
+  df.ems <- df.ems[em_ind,]
+  print("df.ems")
+  print(df.ems)
+  
+  res <- melt(res)
+  names(res) <- c("rho", "em", "sim","om")
+  res$Type = c("SSB","F","R")
+  print(res[1:20,])
+
+  print("unique(res$em)")
+  print(unique(res$em))
+  print(all(unique(res$em) %in% 1:NROW(df.ems)))
+  print(length(unique(res$em)))
+  print(length(unique(res$em)))
+  print(NROW(df.ems))
+  if(!(all(unique(res$em) %in% 1:NROW(df.ems)) & length(unique(res$em)) == NROW(df.ems))) stop("df.ems does not seem to be difined correctly for the number of ems in res.")
+  res <- cbind(df.ems[res$em,], res)
+
+  df <- res
+
+  if(om_type == "naa") {
+    df.oms <- readRDS(here::here("Project_0", "inputs", "df.oms.RDS"))
+  } else {
+    df.oms <- readRDS(here::here("Project_0", "inputs", paste0("df.", om_type, ".oms.RDS")))
+  }
+  df <- cbind(df, df.oms[df$om,])
+  
+  df <- df %>%
+    mutate(EM_process_error = recode(re_config,
+                                     "rec" = "R",
+                                     "rec+1" = "R+S",
+                                     "M_re" = "R+M",
+                                     "q_re" = "R+q",
[…]

### 2025-10-01 Project_0/manuscript/bib_text_marked.tex
saving work on project 0 revisons

--- 2025-09-10 Project_0/manuscript/manuscript_R1.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[
-  12pt,
-]{article}
-\usepackage{amsmath,amssymb}
-\usepackage{iftex}
-\ifPDFTeX
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math} % this also loads fontspec
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-\usepackage{lmodern}
-\ifPDFTeX\else
-  % xetex/luatex font selection
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\usepackage[margin=1in]{geometry}
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{5}
-% definitions for citeproc citations
-\NewDocumentCommand\citeproctext{}{}
-\NewDocumentCommand\citeproc{mm}{%
-  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
-\makeatletter
- % allow citations to break across lines
- \let\@cite@ofmt\@firstofone
- % avoid brackets around text for \cite:
- \def\@biblabel#1{}
- \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
-\makeatother
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
- {\begin{list}{}{%
-  \setlength{\itemindent}{0pt}
-  \setlength{\leftmargin}{0pt}
-  \setlength{\parsep}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-   \setlength{\leftmargin}{\cslhangindent}
-   \setlength{\itemindent}{-1\cslhangindent}
-  \fi
-  % set entry spacing
-  \setlength{\itemsep}{#2\baselineskip}}}
- {\end{list}}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-\usepackage{url}
-\usepackage{setspace}
-%\singlespacing
-%\onehalfspacing
-\doublespacing
-\usepackage{lineno}
-\linenumbers
-\usepackage[belowskip=0pt,aboveskip=0pt]{caption}
-\usepackage{relsize}
-\usepackage{float}
-\usepackage{lscape}
-\usepackage{longtable}
-\usepackage{amsmath,rotating}
-\usepackage[scanall]{psfrag}
-\usepackage{bm}
-\usepackage{caption,graphics}
-\usepackage{graphicx}
-\usepackage{sectsty}
-\usepackage{color}
-\usepackage{fancyhdr}
-\usepackage{xspace}
-\usepackage{textcomp}
-\usepackage{upgreek}
-\renewcommand\figurename{Fig.}
-\captionsetup{labelsep=period, singlelinecheck=false}
-\newcommand{\changesize}[1]{\fontsize{#1pt}{#1pt}\selectfont}
-\renewcommand{\arraystretch}{1.5}
-%\renewcommand\theadfont{}
-
-\newcommand{\Fmsy}{\ensuremath{F_{\text{MSY}}}\xspace}
-\newcommand{\Fspr}[1]{\ensuremath{F_{\text{{#1}\%}}}\xspace}
-\newcommand{\afrb}{Alaska Fishery Research Bulletin\xspace}
-\newcommand{\ajms}{African Journal of Marine Science\xspace}
-\newcommand{\amb}{Advances in Marine Biology\xspace}
-\newcommand{\bms}{Bulletin of Marine Science\xspace}
-\newcommand{\bjssf}{Bulletin of the Japanese Society of Scientific Fisheries\xspace}
-\newcommand{\cb}{Conservation Biology\xspace}
-\newcommand{\cjfas}{Canadian Journal of Fisheries and Aquatic Sciences\xspace}
-\newcommand{\ea}{Ecological Applications\xspace}
-\newcommand{\eer}{Evolutionary Ecology Research\xspace}
-\newcommand{\elet}{Ecology Letters\xspace}
-\newcommand{\emod}{Ecological Modelling\xspace}
-\newcommand{\ebf}{Environmental Biology of Fishes\xspace}
-\newcommand{\ff}{Fish and Fisheries\xspace}
-\newcommand{\fo}{Fisheries Oceanography\xspace}
-\newcommand{\fr}{Fisheries Research\xspace}
-\newcommand{\fb}{Fishery Bulletin\xspace}
-\newcommand{\ijms}{ICES Journal of Marine Science\xspace}
-\newcommand{\iccat}{Collective Volume of Scientific Papers ICCAT\xspace}
-\newcommand{\jae}{Journal of Animal Ecology\xspace}
-\newcommand{\jai}{Journal of Applied Ichthyology\xspace}
-\newcommand{\jdc}{Journal Du Conseil International Pour L'exploration De La Mer\xspace}
-\newcommand{\jdcp}{Journal Du Conseil Permanent International Pour L'exploration De La Mer\xspace}
-\newcommand{\jembe}{Journal of Experimental Marine Biology and Ecology\xspace}
-\newcommand{\jfb}{Journal of Fish Biology\xspace}
-\newcommand{\jsr}{Journal of Sea Research\xspace}
-\newcommand{\jtb}{Journal of Theoretical Biology\xspace}
-\newcommand{\jfrbc}{Journal of the Fisheries Research Board of Canada\xspace}
-\newcommand{\jnwafs}{Journal of Northwest Atlantic Fisheries Science\xspace}
-\newcommand{\mcf}{Marine and Coastal Fisheries: Dynamics, Management, and Ecosystem Science\xspace}
-\newcommand{\mb}{Marine Biology\xspace}
-\newcommand{\meps}{Marine Ecology Progress Series\xspace}
-\newcommand{\mfr}{Marine Fisheries Review\xspace}
-\newcommand{\mpb}{Marine Pollution Bulletin\xspace}
-\newcommand{\najfm}{North American Journal of Fisheries Management\xspace}
-\newcommand{\nzjmfr}{New Zealand Journal of Marine and Freshwater Research\xspace}
-\newcommand{\pnas}{Proceedings of the National Academy of Sciences USA\xspace}
-\newcommand{\rpvrciemm}{Rapports et Proc\`es-Verbaux des R\'eunions. Conseil Internationale pour l'Exploration de la Mer\xspace}
-\newcommand{\rpvrcpiemm}{Rapports et Proc\`es-Verbaux des R\'eunions. Conseil Permanent Internationale pour l'Exploration de la Mer\xspace}
-\newcommand{\rfbf}{Reviews in Fish Biology and Fisheries\xspace}
-\newcommand{\sajms}{South African Journal of Marine Science\xspace}
-\newcommand{\tafs}{Transactions of the American Fisheries Society\xspace}
-
-\newcommand{\anzjs}{Australian \& New Zealand Journal of Statistics\xspace}
-\newcommand{\as}{Applied Statistics\xspace}
-\newcommand{\csda}{Computational Statistics \& Data Analysis\xspace}
-\newcommand{\ees}{Environmental and Ecological Statistics\xspace}
-\newcommand{\jas}{Journal of Applied Statistics\xspace}
-\newcommand{\jabes}{Journal of Agricultural, Biological, and Environmental Statistics\xspace}
-\newcommand{\jasa}{Journal of the American Statistical Association\xspace}
-\newcommand{\jrssb}{Journal of the Royal Statistical Society. Series B\xspace}
-\newcommand{\sm}{Statistics in Medicine}
-
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{wrapfig}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{pdflscape}
-\usepackage{tabu}
-\usepackage{threeparttable}
-\usepackage{threeparttablex}
-\usepackage[normalem]{ulem}
-\usepackage{makecell}
-\usepackage{xcolor}
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\usepackage{bookmark}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same}
-\hypersetup{
-  pdftitle={An investigation of factors affecting inferences from and reliability of state-space age-structured assessment models},
-  pdfauthor={Timothy J. Miller1,2; Greg Britten3; Elizabeth N. Brooks2; Gavin Fay4; Alex Hansell2; Christopher M. Legault2; Chengxue Li2; Brandon Muffley5; Brian C. Stock6; John Wiedenmann7},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-
-\title{An investigation of factors affecting inferences from and
-reliability of state-space age-structured assessment models}
-\author{Timothy J. Miller\textsuperscript{1,2} \and Greg
-Britten\textsuperscript{3} \and Elizabeth N.
-Brooks\textsuperscript{2} \and Gavin Fay\textsuperscript{4} \and Alex
-Hansell\textsuperscript{2} \and Christopher M.
-Legault\textsuperscript{2} \and Chengxue
-Li\textsuperscript{2} \and Brandon Muffley\textsuperscript{5} \and Brian
-C. Stock\textsuperscript{6} \and John Wiedenmann\textsuperscript{7}}
-\date{10 September, 2025}
-
-\begin{document}
-\maketitle
-
-\(^1\)corresponding author:
-\href{mailto:timothy.j.miller@noaa.gov}{\nolinkurl{timothy.j.miller@noaa.gov}}\\
-\(^2\)Northeast Fisheries Science Center, Woods Hole Laboratory, 166
-Water Street, Woods Hole, MA 02543 USA\\
-\(^3\)Biology Department, Woods Hole Oceanographic Institution, 266
-Woods Hole Rd. Woods Hole, MA, USA\\
-\(^4\)Department of Fisheries Oceanography, School for Marine Science
-and Technology, University of Massachusetts Dartmouth, 836 S Rodney
-French Boulevard, New Bedford, MA 02740, USA\\
-\(^5\)Mid-Atlantic Fishery Management Council, 800 North State Street,
-Suite 201, Dover, DE 19901 USA\\
-\(^6\)Institute of Marine Research, Nye Flødevigveien 20, 4817 His,
-Norway\\
-\(^7\)Department of Ecology, Evolution, and Natural Resources. Rutgers
-University\\
-
-\pagebreak
-
-\textbf{keywords:} stock assessment, state-space, model selection, bias,
-convergence, retrospective patterns
-
-\subsection*{Abstract}\label{abstract}
-\addcontentsline{toc}{subsection}{Abstract}
-
-State-space models are increasingly used for stock assessment, and
-evaluations of their statistical reliability and best practices for
-selecting among process error configurations are needed. We simulated 72
-operating models that varied fishing pressure and observation error
-across process errors in recruitment, survival, selectivity,
-catchability, and/or natural mortality. We fit estimating models with
-different assumptions on the process error source and whether median
-natural mortality or a stock-recruit relationship were estimated.
-Estimating models without a stock-recruit relationship that assumed the
-correct process error source and median natural mortality had high
-convergence rates and low bias. Bias was also low under many incorrect
-process error assumptions when there was contrast in fishing pressure
-and low observation error. Marginal AIC most accurately distinguished
-process errors on recruitment, survival, and selectivity, and other
-process error sources when variability was greater. Retrospective
-patterns were generally small but were sizable for recruitment when
-observation error was high. These results help establish the statistical
-reliability of state space assessment models and pave the way for the
-next-generation of fisheries stock assessment.
-
-\pagebreak
-
-\section*{Introduction}\label{introduction}
-\addcontentsline{toc}{section}{Introduction}
-
-Application of state-space models in fisheries stock assessment and
-management has expanded dramatically within International Council for
-the Exploration of the Sea (ICES), Canada, and the Northeast US (Nielsen
-and Berg 2014; Cadigan 2016; Pedersen and Berg 2017; Stock and Miller
-2021). State-space models latent population characteristics as
-statistical time series with periodic observations that also may have
-error due to sampling or other sources of measurement error. Traditional
-assessment models may use state-space approaches to account for temporal
-variability in population characteristics (Legault and Restrepo 1999;
-Methot and Wetzel 2013), but these models treat the annual parameters as
-penalized fixed effects parameters where the variance parameters
-controlling the penalties are assumed known (Thorson and Minto 2015).
-Modern state-space models can estimate the annually varying parameters
-as random effects with variance parameters estimated using maximum
-marginal likelihood or corresponding Bayesian approaches. These latter
-approaches are considered best practice and a recommended for the next
-generation of stock assessment models (Hoyle et al. 2022; Punt 2023).
-
-State-space stock assessment models, with nonlinear functions of latent
-parameters and multiple types of observations with varying
-distributional assumptions, are one of the most complex examples of this
-analytical approach. Statistical aspects of state-space models and their
-application within fisheries have been studied extensively, but previous
-work has focused primarily on linear and Gaussian state-space models
-(Aeberhard et al. 2018; Auger-Méthé et al. 2021). Therefore, current
-understanding of the reliability of state-space models does not extend
-to usage for stock assessment.
-
-As state-space models provide greater flexibility by allowing multiple
-processes to vary as random effects (Nielsen and Berg 2014; Aeberhard et
-al. 2018; Stock et al. 2021), one of the most immediate questions
-regards the implications of mis-specification among alternative sources
-of process error. Incorrect treatment of population attributes as
-temporally varying (Trijoulet et al. 2020; Liljestrand et al. 2024)
-could lead to misidentification of stock status and biased population
-estimates, ultimately impacting fisheries management decisions (Legault
-and Palmer 2016; Szuwalski et al. 2018; Cronin-Fine and Punt 2021).
-Furthermore, biological, fishery, and observational processes are often
-confounded in catch-at-age data, which may adversely affect ability to
-distinguish between true process variability and observational error (Li
-et al. In review; Punt et al. 2014; Stewart and Monnahan 2017;
-Cronin-Fine and Punt 2021; Fisch et al. 2023).
-
-Li et al. (2024) conducted a full-factorial simulation-estimation study
-to assess model reliability when confounding random-effects processes
-(numbers-at-age, fishery selectivity, and natural mortality) were
-included. Their results suggest that while state-space models can
-generally identify sources of process error, overly complex models, even
-when misspecified (i.e., incorporating process error that did not exist
-in reality), often performed similarly to correctly specified models,
-with little to no bias in key management quantities. Similarly,
-Liljestrand et al. (2024) found little downside in assuming process
-error in recruitment or selectivity, even when it was absent.
-
-Despite mounting efforts, several limitations remain. First, confounding
-processes that can be treated as random effects in the model were not
-thoroughly examined or tested within a simulation-estimation framework.
-Second, previous studies relied on operating models conditioned on
-specific fisheries, limiting their generalizability (Li et al. In
-review; Liljestrand et al. 2024). In particular, the effects of
-observation error and underlying fishing history have not been fully
-isolated in simulation study designs, making it challenging to
-disentangle the interplay between process and observation error
-magnitudes, as demonstrated in Fisch et al. (2023). Third, explicitly
-modeling stock-recruit relationships (SRRs) as mechanistic drivers of
-population dynamics is promising (Fleischman et al. 2013; Pontavice et
-al. 2022), but reliability of inferences within integrated state-space
-age-structured models has not been evaluated. Evidence from other
-studies suggests that when both process and observation errors are
-unknown, estimating density dependence parameters becomes highly
-uncertain (Knape 2008; Polansky et al. 2009). In particular, Knape
-(2008) demonstrated that stronger density dependence becomes
-increasingly difficult to estimate in the presence of observation error.
-Therefore, it is crucial to assess whether density dependence mechanisms
-can be estimated with sufficient precision for use in fisheries
-management (Auger-Méthé et al. 2016). Finally, although the importance
-of autocorrelation in process errors is recognized, investigations of
-the ability to distinguish state-space assessment models with and
-without autocorrelation and whether such misspecification is detrimental
-to estimation of important population metrics are lacking (Johnson et
-al. 2016; Xu et al. 2019).
-
-In the present study, we conduct a simulation study with operating
-models (OMs) varying by degree of observation error, source and
-variability of process error, and fishing history. The simulations from
-these OMs are fitted with estimation models (EMs) that make alternative
-assumptions for sources of process error, whether a SRR was estimated,
-and whether natural mortality is estimated. Given the confounding nature
-of process errors, developing diagnostic tools to detect model
-misspecification is of great scientific interest and could aid the next
-generation of stock assessments (Auger-Méthé et al. 2021). We evaluate
-whether convergence and Akaike Information Criterion (AIC) can correctly
-determine the source of process error and the existence of a SRR. We
-also evaluate when retrospective patterns occur and the degree of bias
-in outputs of the assessment model that are important for management.
-
-\section*{Methods}\label{methods}
-\addcontentsline{toc}{section}{Methods}
-
-We used the Woods Hole Assessment Model (WHAM) to configure OMs and EMs
-in our simulation study (Miller and Stock 2020; Stock and Miller 2021).
-WHAM is an R package freely available via a github repository and is
-built on the Template Model Builder package (Kristensen et al. 2016).
-For this study we used
-\href{https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3}{version
-1.0.6.9000, commit 77bbd94}. WHAM has also been used to configure OMs
-and EMs for closed loop simulations evaluating index-based assessment
-methods (Legault et al. 2023) and is currently used or accepted for use
-in management of numerous NEUS fish stocks (e.g., NEFSC 2022a, 2022b;
-NEFSC 2024).
-
-We completed a simulation study with a number of OMs that can be
-categorized based on where process error random effects were assumed:
-recruitment (R, assumed present in all models), apparent survival
-(denoted R+S), natural mortality (R+M), fleet selectivity (R+Sel), or
-index catchability (R+q). We refer to the (R+S) OMs as modeling apparent
-survival because on logscale the random effects (\(\epsilon_{a,y}\)) are
-additive to the total mortality (F+M) between numbers at age, thus they
-modify the survival term. However, as Stock and Miller (2021) note,
-these random effects can be due to events other than mortality, such as
-immigration, emigration, missreported catch, and other sources of
-misspecification. For each OM, assumptions about the magnitude of the
-variance of process errors and observations are required and the values
-we used were based on a review of the range of estimates from Northeast
-Unite States (NEUS) assessments using WHAM.
-
-In total, we configured 72 OMs with alternative assumptions about the
-source and magnitude of process errors, magnitude of observation error
-in indices and age composition data, and contrast in fishing pressure
-over time. We fitted 20 EMs to observations generated from each of 100
-simulations where process errors were also simulated. Each EM differed
-in assumptions about the source of process errors, whether natural
-mortality (or the median for models with process error in natural
-mortality) was estimated, and whether a Beverton-Holt SRR was estimated
-within the EM. Details of each of the OMs and EMs are described below.
-
-We did not use the log-normal bias-correction feature for process errors
-or observations described by (Stock and Miller 2021) for OMs and EMs to
-simplify interpretation of the study results (Li et al. In review). All
-code we used to perform the simulation study and summarize results can
-be found at
-\url{https://github.com/timjmiller/SSRTWG/tree/main/Project_0/code}.
-
-\subsection*{Operating models}\label{operating-models}
-\addcontentsline{toc}{subsection}{Operating models}
-
-\subsubsection*{Population}\label{population}
-\addcontentsline{toc}{subsubsection}{Population}
-
-The population consists of 10 age classes, ages 1 to 10+, with the last
-being a plus group that accumulates ages 10 and older. We assume
-spawning occurs annually 1/4 of the way through the year. The maturity
-at age was a logistic curve with \(a_{50}\) = 2.89 and slope = 0.88
-(Figure \ref{om_inputs_fig}, top left).
-
-Weight at age was generated with a von Bertalanffy growth function \[
-L_a = L_{\infty}\left(1 - e^{-k(a - t_0)}\right)
-\] where \(t_0 = 0\), \(L_\infty = 85\), and \(k = 0.3\), and a L-W
-relationship such that \[
-W_a = \theta_1 L_a^{\theta_2}
-\] where \(\theta_1 = e^{-12.1}\) and \(\theta_2 = 3.2\) (Figure
-\ref{om_inputs_fig}, top right).
-
-We assumed a Beverton-Holt SRR with constant pre-recruit mortality
-parameters for all OMs. All biological inputs to calculations of
-spawning biomass per recruit (i.e., weight, maturity, and natural
-mortality at age) are constant in the apparent survival (R+S)
-selectivity (R+Sel), and survey catchability (R+q) process error OMs.
-Therefore, steepness and unfished recruitment are also constant over the
-time period for those OMs (Miller and Brooks 2021). We specified
-unfished recruitment equal to \(e^{10}\) and
-\(\Fmsy = F_{40\%} = 0.348\), which equates to a steepness of 0.69 and
-\(a=0.60\) and \(b = 2.4 \times 10^{-5}\) for the Beverton-Holt
-parameterization \[
-N_{1,y} = \frac{a \text{SSB}_{y-1}}{1 + b \text{SSB}_{y-1}} 
-\] (Figure \ref{om_inputs_fig}, bottom right). We assumed a value of 0.2
-for the natural mortality rate in OMs without process errors on natural
-mortaliyt and for the median rate for OMs with process errors on natural
-mortality.
-
-We used two fishing scenarios for OMs. In the first scenario, the stock
-experiences overfishing at 2.5\Fmsy for the first 20 years followed by
-fishing at \Fmsy for the last 20 years (denoted
-\(2.5\Fmsy \rightarrow \Fmsy\)). In the second scenario, the stock is
-fished at \Fmsy for the entire time period (40 years). The magnitude of
-the overfishing assumptions is based on average estimates of overfishing
-for NEUS groundfish stocks from Wiedenmann et al. (2019) and similar to
-the approach in Legault et al. (2023).
-
-We specified initial population abundance at age at the equilibrium
-distribution that corresponds to fishing at either
-\(F = 2.5\times \Fmsy\) or \(F = \Fmsy\). This implies that, for a
-deterministic model, the abundance at age would not change from the
-first year to the next.
-
-For OMs with time-varying random effects for M , steepness is not
-constant. However, we used the same \(a\) and \(b\) parameters as other
-OMs, which equates to a steepness and R0 at the median of the time
-series process for M. For OMs with time-varying random effects for
-fishery selectivity, \Fmsy is also not constant, but since we use the
-same F history as other OMs, this corresponds to \Fmsy at the mean
-selectivity parameters.
-
-\subsubsection*{Fleets}\label{fleets}
-\addcontentsline{toc}{subsubsection}{Fleets}
-
-We assumed a single fleet operating year round for catch observations
-with logistic selectivity for the fleet (\(a_{50} = 5\) and slope = 1;
-Figure \ref{om_inputs_fig}, bottom left). This selectivity was used to
-define \Fmsy for the Beverton-Holt SRR parameters above. We assumed a
-logistic-normal distribution with no correlation on the multivariate
-normal scale for the age-composition observations for the fleet.
-
-\subsubsection*{Indices}\label{indices}
-\addcontentsline{toc}{subsubsection}{Indices}
-
-Two time series of fishery-independent surveys in numbers are generated
-for the entire 40 year period with one occurring in the spring (0.25 of
-each year) and one in the fall (0.75 of each year). Catchability of both
-surveys are assumed to be 0.1. Like the fishing fleet, we assumed
-logistic selectivity for both indices (\(a_{50} = 5\) and slope = 1) and
-a logistic-normal distribution with no correlation on the multivariate
-normal scale for the age-composition observations.
-
-\subsubsection*{Observation Uncertainty}\label{observation-uncertainty}
-\addcontentsline{toc}{subsubsection}{Observation Uncertainty}
-
-The standard deviation for log-aggregate catch was 0.1. Two levels of
-observation error variance (high and low) were specified for indices and
-all age composition observations (both indices and catch). The low
-uncertainty specification assumed a standard deviation of 0.1 for both
-series of log-aggregate index observations, and the standard deviation
-of the logistic-normal for age composition observations was 0.3. In the
-high uncertainty specification, the standard deviation for log-aggregate
-indices was 0.4 and that for the age composition observations was 1.5.
-For all EMs, the standard deviation for log-aggregate observations was
-assumed known whereas that for the logistic-normal age composition
-observations was estimated.
-
-\subsubsection*{Operating models with random effects on numbers at
-age}\label{operating-models-with-random-effects-on-numbers-at-age}
-\addcontentsline{toc}{subsubsection}{Operating models with random
-effects on numbers at age}
-
-For operating models with random effects on recruitment and(or) apparent
-survival (R, R+S), we assumed marginal standard deviations for
-recruitment of \(\sigma_R \in \{0.5,1.5\}\) and marginal standard
-deviations for older age classes of \(\sigma_{2+} \in \{0,0.25, 0.5\}\).
-The full factorial combination of these process error assumptions (2x3
-levels) and scenarios for fishing history (2 levels) and observation
-error (2 levels) scenarios described above results in 24 different R
-(\(\sigma_{2+} = 0\)) and R+S operating models (Table
-\ref{naa_om_table}).
-
-\subsubsection*{Operating models with random effects on natural
-mortality}\label{operating-models-with-random-effects-on-natural-mortality}
-\addcontentsline{toc}{subsubsection}{Operating models with random
-effects on natural mortality}
-
-All R+M OMs treat natural mortality as constant across age, but with
-annually varying random effects. WHAM treats natural mortality as a
-log-transformed parameter \[
-\log M_{y,a} = \mu_{M} + \epsilon_{M,y}
-\] that is a linear combination of a mean log-natural mortality
-parameter that is constant across ages (\(\mu_{M} = \log(0.2)\)) and any
-annual random effects are marginally distributed as
-\(\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)\). The marginal
-standard deviations we assumed for log natural mortality random effects
-were \(\sigma_M \in \{0.1, 0.5\}\) and the random effects were either
-uncorrelated or first-order autoregressive (AR1,
-\(\rho_M \in \{0,0.9\}\)). Uncorrelated random effects were also
-included on recruitment with \(\sigma_R = 0.5\) (hence, we denote these
-OMs as R+M). The full factorial combination of these process error
-assumptions and fishing history (2 levels) and observation error (2
-levels) scenarios described above results in 16 different R+M OMs (Table
-\ref{M_om_table}).
-
-\subsubsection*{Operating models with random effects on fleet
-selectivity}\label{operating-models-with-random-effects-on-fleet-selectivity}
-\addcontentsline{toc}{subsubsection}{Operating models with random
-effects on fleet selectivity}
-
-WHAM treats selectivity parameter \(s\) as a logit-transformed parameter
-\[
-\log\left(\frac{p_{s,y}-l_{s}}{u_{s}-p_{s,y}}\right) = \mu_s + \epsilon_{s,y}
-\] that is a linear combination of a mean \(\mu_s\) and any annual
-random effects marginally distributed as
-\(\epsilon_{s,y} \sim \text{N}\left(0,\sigma_s^2\right)\), where the
-lower and upper bounds of the parameter (\(l_s\) and \(u_s\)) can be
-specified by the user. All selectivity parameters (\(a_{50}\) and slope
-parameters) were bounded by 0 and 10 for all OMs and EMs. The marginal
-standard deviations we assumed for logit scale random effects were
-\(\sigma_s \in \{0.1, 0.5\}\) and AR1 autocorrelation parameters of
-\(\rho_s \in \{0,0.9\}\). Like R+M OMs, the full factorial combination
-of these process error assumptions (2x2 levels) and scenarios described
-above for fishing history (2 levels) and observation error (2 levels)
-results in 16 different R+Sel OMs (Table \ref{sel_om_table}).
-
-\subsubsection*{Operating models with random effects on index
-catchability}\label{operating-models-with-random-effects-on-index-catchability}
-\addcontentsline{toc}{subsubsection}{Operating models with random
-effects on index catchability}
-
-Like selectivity parameters, WHAM treats catchability for an index \(i\)
-as a logit-transformed parameter \[
-\log\left(\frac{q_{i,y}-l_{i}}{u_{i}-q_{i,y}}\right) = \mu_i + \epsilon_{i,y}
-\] that is a linear combination of a mean \(\mu_i\) and any annual
-random effects marginally distributed as
-\(\epsilon_{i,y} \sim \text{N}\left(0,\sigma_i^2\right)\) where the
-lower and upper bounds of the catchability (\(l_i\) and \(u_i\)) can be
-specified by the user. We assumed bounds of 0 and 1000 for all OMs and
-EMs. For all OMs and EMs with process errors on catchability, the
-temporal variation only applies to the first index, which could be
-interpreted as capturing some unmeasured seasonal process that affects
-availability to the survey. The marginal standard deviations we assumed
-for logit scale random effects were \(\sigma_i \in \{0.1, 0.5\}\) and
-AR1 autocorrelation parameters of \(\rho_i \in \{0,0.9\}\). Like R+M and
-R+Sel OMs, the full factorial combination of these process error
-assumptions and fishing history (2 levels) and observation error (2
-levels) scenarios described above results in 16 different R+q OMs (Table
-\ref{q_om_table}).
-
-\subsection*{Estimation models}\label{estimation-models}
-\addcontentsline{toc}{subsection}{Estimation models}
-
-For each of the data sets simulated from an OM, 20 EMs were fit. A total
-of 32 different EMs were fit across OMs where the subset of 20 depended
-on the source of process error in the OM (Table \ref{em_table}). The EMs
-have different assumptions about the source of process error (R+S, R+M,
-R+Sel, R+q) and whether or not 1) there is temporal autocorrelation, 2)
-a Beverton-Holt SRR is estimated, and 3) the natural mortality rate
-(\(\mu_M\), the constant or mean on log scale for R+M EMs) is estimated.
-For simplicity we refer to the derived estimate \(e^{\mu_M}\) as the
-median natural mortality rate regardless of whether natural mortality
-random effects are estimated in the EM.
-
-Subsets of 20 EMs in Table \ref{em_table} were fit to simulate data sets
-from each of the OM process error categories. For R and R+S OMs, fitted
-EMs had matching process error assumptions as well as R+Sel, R+M, and
-R+q assumptions without autocorrelation. Similarly, For other OM process
-error categories, we fit EMs with matching process error assumptions as
-well as other process error types without autocorrelation. The maturity
-at age, weight at age for catch and SSB, and observation error variance
-of aggregate catch and indices were all assumed known at the true
-values. However, the variance parameters for the logistic-normal
-distributions for age composition observations were estimated in the
-EMs. As such, EMs would either be configured completely correctly for
-the OM, or there could be mis-specification in assumptions of process
-error autocorrelation, the type of process error, or the SRR
-(Beverton-Holt or none).
-
-\subsection*{Measures of reliability}\label{measures-of-reliability}
-\addcontentsline{toc}{subsection}{Measures of reliability}
-
-\subsubsection*{Convergence}\label{convergence}
-\addcontentsline{toc}{subsubsection}{Convergence}
-
-The first measure of reliability we investigated was frequency of
-convergence when fitting each EM to the simulated data sets. There are
-various ways to assess convergence of the fit (e.g., Carvalho et al.
-2021; Kapur et al. 2025), but given the importance of estimates of
-uncertainty when using assessment models in management, we estimated
-probablity of convergence as measured by occurrence of a
-positive-definite hessian matrix at the optimized negative
-log-likelihood that could be inverted. We also provide results in the
-Supplementary Materials for the maximum of the absolute values among all
-gradients for all fits of a given EM to all simulated data sets from a
-given OM that produced hessian-based standard errors for all estimated
-fixed effects. This provides an indication of how poor the calculated
-gradients can be, but still presumably converged adequately enough for
-parameter inferences. We used the Clopper-Pearson exact method for
-constructing 95\% confidence intervals of the probabilities of
-convergence (Clopper and Pearson 1934; Thulin 2014).
-
-\subsubsection*{AIC for model selection}\label{aic-for-model-selection}
-\addcontentsline{toc}{subsubsection}{AIC for model selection}
-
-We estimated the probability of selection of each process error model
-structure (R, R+S, R+M, R+Sel, R+q) using marginal AIC. For a given
-operating model, we compared AIC for EMs that were all configured the
-same for median natural mortality (known or estimated) and the SRR
-(Beverton-Holt or none).
-
-We also estimated the probability of correctly selecting between EMs
-with Beverton-Holt SRR assumed and models without the SRR (null model).
-We made these comparisons between models that otherwise assumed the same
-process error structure as the operating model and both of the compared
-models either estimate median natural mortality or assume it is known.
-Contrast in fishing pressure and time series with recruitment at low
-stock size has been shown to improve estimation of SRR parameters
-(Magnusson and Hilborn 2007; Conn et al. 2010). Our preliminary
-inspections of the proportions of simulations where the correct
-recruitment model was chosen for a given set of OM factors (including
-contrast in fishing pressure) indicated generally poor performance of
-AIC. Therefore, we fit logistic regression models to the indicator of
-Beverton-Holt models having lower AIC as a function of the log-standard
-deviation of the true log(SSB) (similar to the log of the coefficient of
-variation for SSB) since simulations with realized SSB producing low and
-high recruitments would have larger variation in realized SSB.
-
-All model selection results condition only on whether all of the
-compared estimating models completed the optimization process without
-failure. We did not condition on convergence as defined by a gradient
-threshold or invertibity of the hessian because optimization could
-correctly determine an inappropriate process error assumption by
-estimating variance parameters at the lower bound of zero. Such an
-optimization could indicate poor convergence but the likelihood would be
-equivalent to that without the mis-specified random effects and the AIC
-would be appropriately higher because more (variance) parameters were
-estimated. All other measures of reliability described below (bias and
-Mohn's \(\rho\)) use these same criteria for inclusion of EM fits in the
-summarized results.
-
-\subsubsection*{Bias}\label{bias}
-\addcontentsline{toc}{subsubsection}{Bias}
-
-For a given model attribute we calculated the relative error
-\begin{equation}\label{relerror}
-\text{RE}\left(\theta_i\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
-\end{equation} from fitting a given estimating model to simulated data
-set \(i\) configured for a given OM where \(\widehat \theta_i\) and
-\(\theta_i\) are the estimated and true values for simulation \(i\). We
-estimated bias as the median of the relative errors across all
-simulations for a given OM and EM combination. We constructed 95\%
-confidence intervals for the median relative bias using the binomial
-distribution approach as in Miller and Hyun (2018) and Stock and Miller
-(2021). We present results for bias in terminal year estimates of SSB
-and recruitment, Beverton-Holt stock recruit parameters (\(a\) and
-\(b\)), and median natural mortality rate. Results for terminal year
-fishing mortality were strongly negatively correlated with those for SSB
-and are provided in the Supplementary Materials.
-
-\subsubsection*{\texorpdfstring{Mohn's
-\(\rho\)}{Mohn's \textbackslash rho}}\label{mohns-rho}
-\addcontentsline{toc}{subsubsection}{Mohn's \(\rho\)}
-
-We calculated Mohn's \(\rho\) for SSB, fishing mortality (averaged over
-all age classes), and recruitment for each EM (Mohn 1999). We fit
-\(P =7\) peels to each simulated data set and calculated Mohn's \(\rho\)
-for a given attribute \(\theta\) as \begin{equation}\label{mohns_rho}
-\rho\left(\theta\right) = \frac{1}{P}\sum^P_{p=1} \frac{\widehat \theta_{Y-p,Y-p} - \widehat\theta_{Y-p,Y}}{\widehat\theta_{Y-p,Y}}
-\end{equation} where \(\widehat \theta_{i,j}\) is the estimate for
-attribute \(\theta\) in year \(i\) from a model fit using data up to
-year \(j\). For each EM we calculated median and 95\% confidence
-intervals using the same methods as that for relative bias.
-
-\subsubsection*{Summarizing results across OM and EM
-attributes}\label{summarizing-results-across-om-and-em-attributes}
-\addcontentsline{toc}{subsubsection}{Summarizing results across OM and
-EM attributes}
-
-The measures of central tendency and variability of of observed values
-for specific OM and EM attributes (e.g., low or high observation error)
-that we described above can indicate scenarios that provide better or
-poorer reliability. However, the OM and EM attributes that we
-investigated are numerous, so we used two methods to summarize the most
-important factors for differences in results. The first method was
-fitting regression models with the response being each of the measures
-of reliability described above and predictor variables were defined
-based on OM and EM characteristics (e.g., MacKinnon et al. 1995; Wang et
-al. 2017; Harwell et al. 2018). For the binary results for convergence
-and AIC-based selection of a stock-recruit relationship, we performed
-logistic regressions. For results for AIC-based selection of EM process
-error type (multiple categories) we performed multinomial regressions.
-For other measures of reliability we fit standard linear regression
-models. Because relative errors (Eq. \ref{relerror}) and Mohn's \(\rho\)
-for the various parameters are bounded below at -1, we used a
-transformation of these values
-\begin{equation}\label{bias_regression_response}
-y_i = \log\left[f\left(\widehat \theta_i,\theta_i\right)+1\right]
-\end{equation} where \(f\) is either the relative error (Eq.
-\ref{relerror}) or Mohn's \(\rho\) (Eq. \ref{mohns_rho}) for simulation
-\(i\), so that values are unbounded. For relative errors, \(y_i\) is the
-log-scale error. We omitted simulation estimates equal to zero (RE =
--1). For all regressions we fit separate models with individual factors
-included, with all factors combined, with including all second order
-interactions, and including all third order interactions. For the
-multinomial regression, we used the \verb|vglm| function from the VGAM
-package (Yee 2008; Yee 2015). We tabulated percent reduction in residual
-deviance for each of regression fits. We did not perform formal
-statistical analyses of effects of OM and EM attributes on results
-(e.g., ANOVA) because of the lack of independence of the
-``observations'' that results from fitting multiple EMs to each
-simulated data set.
-
-The second method involved fitting classification and regression trees
-(Breiman et al. 1984) to show how the OM and EM attributes, and their
-interactions, split the values for each measure of reliability (e.g.,
-Gonzalez et al. 2018; Collier et al. 2022). We used classification trees
-for categorical measures (convergence and AIC) and regression trees for
-the other measures with continuous scales (relative error and Mohn's
-\(\rho\)). The response variables were the same as the regressions for
-the deviance reduction analyses. We used the \verb|rpart| function in
-the rpart package (\url{https://cran.r-project.org/package=rpart}) to
-fit trees and, for clarity we pruned them to show the primary branches.
-
-\section*{Results}\label{results}
-\addcontentsline{toc}{section}{Results}
-
-\subsection*{Convergence performance}\label{convergence-performance}
-\addcontentsline{toc}{subsection}{Convergence performance}
-
-\subsubsection*{Deviance reduction and classification
-trees}\label{deviance-reduction-and-classification-trees}
-\addcontentsline{toc}{subsubsection}{Deviance reduction and
-classification trees}
-
-EM process error assumption was the single attribute that resulted in
-the largest percent reduction in deviance for all OM process error types
-other than R+S OMs where the EM M assumption explained the most residual
-deviance (Table \ref{convergence_PRD_table}). However, also including
-interactions of OM and EM factors also provided large reductions in
-residual deviance, suggesting successful convergence depended on a
-combination OM and EM attributes. Classification trees for each OM
-process error type, all had the primary branch defined using the same
-attribute that provided the largest reduction in deviance (Figure
-\ref{conv_class}).
-
-Generally, the factors defining primary splits for the trees
-corresponded to those with the largest reductions in deviance fitting
-the regression models.
-
-\subsubsection*{Full results}\label{full-results}
-\addcontentsline{toc}{subsubsection}{Full results}
-
-For many R and R+S OMs, convergence rate declined when either the median
-natural mortality rate or the Beverton-Holt SRR was estimated even when
-the process error assumptions of the EMs and OMs matched (Figure
-\ref{hessian_SE_convergence}, A). When there was high observation error
-and constant fishing pressure (\(F = F_{\text{MSY}}\) for all 40 years),
-convergence was poor for of all EM process error configurations other
-than R EMs when fitted to R OMs (\(\sigma_{2+} = 0\)) regardless of
-whether median natural mortality and SRRs were estimated. Convergence of
-R EMs was high for all R and R+S OMs except when there was high
-observation error and constant fishing pressure, and when median natural
-mortality and SRRs were estimated. R+S EMs fit to R OMs exhibited poor
-convergence regardless of whether natural mortality or a SRR was
-estimated. R+S EMs fit to R+S OMs had highest convergence rates when
-there was contrast in fishing pressure and low observation error.
-Convergence rates were high for all EMs when fit to data from R+S OMs
-with lower observation error except those where median natural mortality
-and/or SRRs were estimated.
-
-Convergence of all EMs fitted to R+M OMs was highest when the OMs had
-higher natural mortality process error variability, low observation
-error, and contrast in fishing pressure (Figure
-\ref{hessian_SE_convergence}, B). R+M EMs that estimated autocorrelation
-of process errors had poor convergence for R+M OMs when there was low
-natural mortality process error variability regardless of
-autocorrelation of the simulated process errors. R+S EMs fitted to data
-generated from R+M OMs always converged poorly whether or not median
-natural mortality and the Beverton-Holt SRR were estimated.
-
-The R+S EMs, in particular, had poor convergence when fit to data
-generated from R+Sel OMs with lower selectivity process error
-variability or higher observation error (Figure
-\ref{hessian_SE_convergence}, C). R+Sel EMs generally converged better
-than other EMs for R+Sel OMs with higher process error variability,
-lower observation error, and contrast in fishing pressure regardless of
-whether median natural mortality or a SRR was estimated.
-
-For R+q OMs, convergence of R+q EMs was generally better than that of
-other EMs when there was contrast in fishing pressure (Figure
-(\ref{hessian_SE_convergence}, D). Convergence of R+S EMs was generally
-worse than that of all other EMs across all OMs whether or not median
-natural mortality or a SRR was estimated. Again, convergence probability
-generally declined for all EMs when median natural mortality or a SRR
-was estimated.
-
-We found a wide range of maximum absolute values of gradients for models
-that converged (Figure \ref{hess_grad}). The largest value observed for
-a given EM and OM combination was typically \(<10^{-3}\), but many
-converged models had values greater than 1. For many OMs, EMs that
-assumed the correct process error type and did not estimate median
-natural mortality or the Beverton-Holt SRR produced the lowest gradient
-values.
-
-\subsection*{AIC performance for process error
-structure}\label{aic-performance-for-process-error-structure}
-\addcontentsline{toc}{subsection}{AIC performance for process error
-structure}
-
-\subsubsection*{Deviance reduction and classification
-trees}\label{deviance-reduction-and-classification-trees-1}
-\addcontentsline{toc}{subsubsection}{Deviance reduction and
-classification trees}
-
-\subsubsection*{Full results}\label{full-results-1}
-\addcontentsline{toc}{subsubsection}{Full results}
-
-Marginal AIC accurately determined the correct process error assumptions
-in EMs when data were generated from R and R+S OMs, regardless of
-whether median natural mortality or a SRR was estimated (Figure
-\ref{pe_aic}, A). Attempting to estimate median natural mortality or a
-SRR separately had a negligible effect on the accuracy of determining
-the correct process error assumption. When both were estimated, there
-was a noticeable reduction in accuracy when OMs had a constant fishing
-pressure, low observation error, and larger variability in recruitment
-process errors.
-
-For R+M OMs, marginal AIC only accurately determined the correct process
-error model and correlation structure when observation error was low and
-variability in natural mortality process errors was high (Figures
-\ref{pe_aic}, B). Of these OMs, estimating the median natural mortality
-rate only reduced the accuracy of AIC when natural mortality process
-errors were independent and fishing pressure was constant. For OMs with
-poor model selection accuracy, AIC most frequently selected EMs with
-process errors in catchability (R+q) or selectivity (R+Sel). Selection
-of R+S EMs was generally unlikely.
-
-Marginal AIC most accurately determined the correct source of process
-error and correlation structure for R+Sel OMs with low observation error
-(Figures \ref{pe_aic}, C). When there was low variability in selectivity
-process errors and high observation error, R+q or R+S EMs were more
-likely to have the best AIC. Whether median natural morality or SRRs
-were estimated appeared to have little effect on the performance of AIC.
-
-Marginal AIC most accurately determined the correct source of process
-error and correlation structure for R+q OMs with high variability in
-catchability process errors (Figures \ref{pe_aic},D). The R+q OMs with
-low variability in catchability process errors and high observation
-error had the least model selection accuracy. However, for these OMs,
-the marginal AIC accurately determined the correct source of process
-error (but not correlation structure) except when OMs assumed a constant
-fishing pressure and EMs estimated both median natural morality and the
-SRR.
-
-\subsection*{AIC performance for the stock-recruit
-relationship}\label{aic-performance-for-the-stock-recruit-relationship}
-\addcontentsline{toc}{subsection}{AIC performance for the stock-recruit
-relationship}
-
-\subsubsection*{Deviance reduction and classification
-trees}\label{deviance-reduction-and-classification-trees-2}
-\addcontentsline{toc}{subsubsection}{Deviance reduction and
-classification trees}
-
-\subsubsection*{Full results}\label{full-results-2}
-\addcontentsline{toc}{subsubsection}{Full results}
-
-Our comparisons of model performance conditioned on assuming the true
-process error configuration is known (EM and OM process error types
-match) and we focus on results where the EMs assume median natural
-mortality is known because there was little difference in results when
-the EMs estimated this parameter. Broadly, we found generally poor
-accuracy of AIC in selecting models assuming a Beverton-Holt SRR over
-the null model without an SRR for all OMs. However, we also found
-increased accuracy of AIC in determining the Beverton-Holt SRR when the
-simulated population exhibited greater variation in spawning biomass for
-nearly every OM (Figure \ref{sr_aic}).
-
-With R and R+S process error assumptions, probability of lowest AIC for
-the B-H SRR as a function of SSB variability were greatest for OMs with
-contrast in fishing pressure and lower process variability in
-recruitment (Figure \ref{sr_aic}, A). The largest variation in SSB
-occurred in OMs with larger recruitment variability (\(\sigma_R = 1.5\);
-Figure \ref{sr_aic}, A, right column group), but the same high AIC
-accuracy was achieved for OMs with lower recruitment variability at
-lower levels of SSB variation. The level of observation error had little
-effect on AIC accuracy.
-
-For R+M OMs, probability of lowest AIC for the Beverton-Holt SRR
-increased steeply with variation in SSB whether it was induced by
-contrast in fishing or variation in natural mortality process error.
-(Figure \ref{sr_aic}, B). There was little difference in AIC accuracy
-whether the natural mortality process errors were correleted and,
-similar to R+S OMs, there was also little effect due to level of
-observation error.
-
-For R+Sel OMs, contrast in fishing pressure over time was the primary
-source of variation in SSB and these are the OMs where AIC accuracy for
-the Beverton-Holt SRR was greatest (Figure \ref{sr_aic}, C). There was
-little effect of variability or correlation of selectivity process
-errors or the level of observation error on AIC accuracy.
-
-Like the R+Sel OMs, the greatest accuracy for AIC in selecting the
-Beverton-Holt SRR occurred for R+q OMs where there was contrast in
-fishing pressure over time which is also where there was the greatest
-variation in SSB (Figure \ref{sr_aic}, D). There was also little effect
-of variability or correlation of catchability process errors or the
-level of observation error on AIC accuracy.
-
-\subsection*{Bias}\label{bias-1}
-\addcontentsline{toc}{subsection}{Bias}
-
-\subsubsection*{Deviance reduction and regression
-trees}\label{deviance-reduction-and-regression-trees}
-\addcontentsline{toc}{subsubsection}{Deviance reduction and regression
-trees}
-
-Across all OM process error types, fishing history was the most
-important single factor in reducing residual deviation in log-scale
-errors for the Beverton-Holt density-independent parameter \(a\), and
-also for the density-independent parameter \(b\) for R+M, R+Sel, and R+q
-OMs (Table \ref{bias_SR_pars_PRD_table}). For R and R+S OMs the EM
-process error assumption was equally important for estimating \(b\), but
-the the reductions were small (mostly less than 4\%) for any factor and
-any OM process error type. There was also no substantial reduction in
-deviance with second or third order interactions of factors for either
-parameter in any OM process error types.
-
-The regression trees indicate that having contrast in fishing pressure
-reduced the median error for both SR parameters for all OM types
-
-We observed different effects of factors on deviance reduction for
-errors in estimation of the two Beverton-Holt stock-recruit parameters.
-When analyses were For bias of the density-dependent parameter \(b\) as
-measured by the log absolute value of log-scale errors, deviance was not
-reduced by any of the individual OM or EM factors except for variation
-in recruitment \(\sigma_R\) for R and R+S OMs. However, second and(or)
-third order interactions of the OM and EM factors also provided large
-reductions in deviance for all OM process error types which indicated
-that some combinations of factors could reduce bias for the \(b\)
-parameter. Despite the explanation variation in the \(b\) estimation
-erorrs, bias was still significant regardless of the combination of OM
-and EM factors (Figures regtrees and full results).
-
-Median natural mortality rate showed reductions in deviance for R, R+S,
-and R+M OMs with second order interactions and R+M also showed
-reductions with third order interactions.
-
-\subsubsection*{Full results}\label{full-results-3}
-\addcontentsline{toc}{subsubsection}{Full results}
-
-\paragraph*{Spawning stock biomass}\label{spawning-stock-biomass}
-\addcontentsline{toc}{paragraph}{Spawning stock biomass}
-
-For R OMs (\(\sigma_{2+} = 0\)), there was no indication of bias (95\%
-confidence intervals included 0) in terminal year SSB for any of the
-estimating models regardless of process error assumptions, except when
-no SR assumption was made, recruitment variability was low, and there
-was contrast in fishing mortality and high observation error (Figure
-\ref{SSB_rel_error}, A). However, errors in terminal SSB estimates were
-highly variable when median natural mortality was estimated and there
-was constant fishing pressure and high observation error (Figure
-\ref{SSB_rel_error}, A, second row).
-
-For R+S OMs, the EMs with matching process error assumptions generally
-produced unbiased estimation of terminal SSB except when median natural
-mortality was estimated and there was high observation error. In R+S OMs
-with low observation error, EMs with incorrect process error assumptions
-typically provided biased estimation of terminal year SSB. Estimating
-the Beverton-Holt SRR had little discernible effect on bias of terminal
-year SSB estimation whereas estimating median M tended to produce more
-variability in errors in terminal SSB estimation similar to R OMs.
-
-For R+ M OMs with low variability in natural mortality process errors,
-low observation error and contrast in fishing motality over time all EMs
-produced low variability in SSB estimation error that indicated
-unbiasedness (Figure \ref{SSB_rel_error}, B, third row). However, larger
-variability in natural mortality process errors increased bias of EMs
-without the correct process error type. Estimating median natural
-mortality increased variability of SSB estimation error particularly for
-OMs with high observation error and constant fishing pressure over time.
-It also increased bias in SSB estimation for many R+M OMs. Like R and
-R+S OMs, estimating a SRR had little discernible effect on SSB bias.
-
-For R+Sel OMs, there was no evidence of bias for any EMs when
-variability in selectivity process error and observation error was low,
-and with contrast in fishing mortality (Figure \ref{SSB_rel_error}, C).
-The largest bias occurred for any EMs that estimated median natural
-mortality when the OMs had high observation error, constant fishing
-pressure, and greater variability in selectivity process errors
-(\(\sigma_{\text{Sel}} = 0.5\)) or low selectivity process errors
-(\(\sigma_{\text{Sel}} = 0.1\)) and low observation error. However,
-there was no evidence of SSB bias for correctly specified R+Sel EMs when
-observation error was low and variation in selectivity process errors
-was larger, whether median natural mortality was estimated or not
-(Figure \ref{SSB_rel_error}, C, third row). We only observed an effect
-of estimating the Beverton-Holt SRR for R+Sel OMs that had high
-observation error and contrast in fishing pressure where estimating the
-SRR produced less biased SSB estimation for many EMs (Figure
-\ref{SSB_rel_error}, C, top row).
-
-All EMs fit to data from R+q OMs with low observation error and contrast
-in fishing pressure exhibited little evidence of bias in terminal SSB
-estimation except for R+M EMs when there was no AR1 correlation in
-catchability process errors (Figure \ref{SSB_rel_error}, D). Many EMs
-also performed well in R+q OMs with low observation error, but no
-contrast in fishing pressure. For R+q OMs with high observation error
-and contrast in fishing pressure, EMs that estimated the Beverton-Holt
-SRR exhibited less SSB bias than those that did not. Estimating median
-natural mortality in the EMs only resulted in much more variable SSB
-estimation errors when there was no contrast in fishing pressure (Figure
-\ref{SSB_rel_error}, D, first and third rows).
-
-For all OM process error types, relative errors in terminal year
-recruitment were generally more variable than SSB, but effects of R and
-R+S OM and EM attributes on bias (i.e, negative or positive or none)
-were similar (Figure \ref{R_rel_error}, A). Furthermore, for EM
-configurations where bias in terminal SSB was evident, median relative
-errors in recruitment often indicated stronger bias in recruitment of
-the same sign.
-
-\subsubsection*{Beverton-Holt
-parameters}\label{beverton-holt-parameters}
-\addcontentsline{toc}{subsubsection}{Beverton-Holt parameters}
-
-Across all OMs, there was generally less bias and(or) lower variability
-in estimation of the Beverton Holt \(a\) parameter than the \(b\)
-parameter. In R and R+S OMs, EMs with the correct assumptions about
-process errors provided the least biased estimation of Beverton-Holt SRR
-parameters when there was a change in fishing pressure over time and
-lower variability of recruitment process errors, but there was little
-effect of estimating median natural mortality and a small increase in
-bias for those OMs that had high observation error (Figure
-\ref{SR_rel_error}, A). For other R and R+S OMs, estimating natural
-mortality often resulted in less biased estimation of SRR parameters.
-There was generally large variability in relative errors of the SRR
-parameter estimates, but the lowest variability occurred with low
-variability in recruitment and little or no variability in survival
-process errors (\(\sigma_{2+} \in \{0,0.25\}\)), and contrast in fishing
-pressure.
-
-In R+M OMs, the most accurate estimation of SRR parameters for all EM
-process error assumptions occurred when there was a change in fishing
-pressure, greater variability in natural mortality process errors, and
-lower observation error (Figure \ref{SR_rel_error}, B). Relative to the
-R, and R+S OMs, there was even less effect of estimating median natural
-mortality on estimation bias for the SRR parameters.
-
-Bias for SRR parameters was large and variability in relative errors was
-greatest for most EMs fit to R+Sel OMs with constant fishing pressure
-(Figure \ref{SR_rel_error}, C). Less bias in parameter estimation
-occurred for OMs with a change in fishing pressure and the best accuracy
-occurred for those OMs that had low observation error and more variable
-and uncorrelated selectivity process errors, and when the EMs had with
-the correct process error assumption. There was little effect of
-estimating natural mortality on relative errors for SRR parameters.
-
-Like R+Sel OMs, relative errors in SRR parameters for R+q OMs were more
-accurate for most EM process error types when OMs had contrast in
-fishing pressure and lower observation error (Figure \ref{SR_rel_error},
-D). However, the best accuracy occurred for those OMs that had lower
-variability in catchability process errors. The worst accuracy of SRR
-parameter estimation regardless of EM type occurred when R+q OMs had low
-observation error and constant fishing pressure (Figure
-\ref{SR_rel_error}, D, fourth row).
-
-\subsubsection*{Median natural mortality
-rate}\label{median-natural-mortality-rate}
-\addcontentsline{toc}{subsubsection}{Median natural mortality rate}
-
-Across all OMs and EMs there was little effect of estimating SRRs on the
-bias in estimation of median natural mortality (Figure
-\ref{M_rel_error}). Median natural mortality rate was estimated
-accurately by all EM process error types for all R OMs except those with
-high observation error and constant fishing pressure, in which case
-relative errors were high (Figure \ref{M_rel_error}, A,
-\(\sigma_{2+} = 0\)). For R+S OMs estimation of median natural mortality
-rate was most accurate when observation error was low and there was
-contrast in fishing pressure and the EM process error type was correct.
-
-For R+M OMs, median natural mortality was estimated most accurately,
-regardless of EM process error type, when OMs had a change in fishing
-pressure and low observation error (Figure \ref{M_rel_error}, B).
-However, those R+M OMs that also had greatest variability in AR1
-correlated natural mortality process errors only had unbiased estimation
-when the EM process error type was correct.
-
-All EM process error types accurately estimated median natural mortality
-rate for R+Sel OMs that had contrast in fishing pressure, low
-observation error, and low selectivity process error variability (Figure
-\ref{M_rel_error}, C). When selectivity process error variability
-increased, the incorrect EM process errors produce more biased
-estimation of median natural mortality rate. The least accurate
-estimation occurred for all EM process error types when observation
-error was high and fishing pressure was constant.
-
-Like R+Sel OMs, all EM process error types produced accurate estimation
-of median natural mortality rate when fit to R+q OMs with contrast in
-fishing pressure, low observation error and low catchability process
-error variability (Figure \ref{M_rel_error}, D). Most EM process error
-types produced biased estimation of median natural mortality when R+q
-OMs had high observaiton error and constant fishing pressure.
-
-\subsection*{\texorpdfstring{Mohn's
-\(\rho\)}{Mohn's \textbackslash rho}}\label{mohns-rho-1}
-\addcontentsline{toc}{subsection}{Mohn's \(\rho\)}
-
-\subsubsection*{Deviance reduction and regression
-trees}\label{deviance-reduction-and-regression-trees-1}
-\addcontentsline{toc}{subsubsection}{Deviance reduction and regression
-trees}
-
-Fits of glms showed little explanatory power of any of the OM an EM
-attributes for Mohn's rho for SSB, F, and R. Note that the factors that
-showed the largest reductions in deviance (which were very small) often
-do not comport with factors that show differences in median Mohn's rho
-results. This is because the glms and regression trees are reducing
-residual variance and there could be large variability in values, but
-the medians are less sensitive to extreme values. For example we can see
-that there is large Mohn's rho for R in R OMs with larger variability
-assumed in recruitment and higher observation error (Figure X), but
-these factors suggest little reduction in residual deviance (Table X).
-
-\subsubsection*{Full results}\label{full-results-4}
-\addcontentsline{toc}{subsubsection}{Full results}
-
-Mohn's \(\rho\) for SSB was small in absolute value for all R and R+S
-OMs, regardless of EM process error types, and whether median natural
-mortality rate or SRRs were estimated (Figure \ref{mohns_rho_ssb}, A).
-The strongest retrospective patterns (highest absolute Mohn's \(\rho\)
-values) occurred in OMs with the largest apparent survival process error
-variability, high observation error, and contrast in fishing pressure,
-but only for EMs with the incorrect process error type and where median
-natural mortality rate was assumed known (median \(\rho\) was
-approximately -0.15). For R+M, R+Sel, and R+q OMs, Mohn's \(\rho\) was
-also small in absolute value, but median values were all closer to 0
-than the largest values in the R and R+S OMs (Figure
-\ref{mohns_rho_ssb},B-D). For these OMs, there was no noticeable effect
-of estimation of median natural mortality rate or SRRs on Mohn's
-\(\rho\) for any EM process error types.
-
-Mohn's \(\rho\) for recruitment was small in absolute value for all R
-OMs with low variability in recruitment process errors, regardless of EM
-process error type, and whether median natural mortality rate or SRRs
-were estimated (Figure \ref{mohns_rho_R}, A). However, R and R+S OMs
-with greater recruitment process variability and higher observation
-error had median Mohn's \(\rho\) for recruitment greater than zero for
-most EMs even when the EM process error type was correct. In R+S OMs
-with lower observation error, EMs with the correct process error type
-exhibited better median Mohn's \(\rho\) close to 0 than EMs with the
-incorrect process error type. For R+M, R+Sel, and R+q OMs, results for
-Mohn's \(\rho\) for recruitment are similar to those for SSB, but the
-range in median values and variation in Mohn's \(\rho\) values for a
-given OM are generally larger for recruitment (Figure \ref{mohns_rho_R},
-B-D).
-
-\section*{Discussion}\label{discussion}
-\addcontentsline{toc}{section}{Discussion}
-
-\subsection*{Convergence}\label{convergence-1}
-\addcontentsline{toc}{subsection}{Convergence}
-
-Analyses of model convergence across simulations can be useful for
-understanding the utility of alternative convergence criteria used in
-applications to real data for directing the practitioner to more
-appropriate random effects configurations. It is common during the
-assessment model fitting process to check that the maximum absolute
-gradient component is less than some threshold prior to inspecting the
-Hessian of the optimized likelihood for invertibility (Carvalho et al.
-2021). However, there is no accepted standard for the gradient threshold
-(e.g., Lee et al. 2011; Hurtado-Ferro et al. 2014; Rudd and Thorson
-2018) and some thresholds would exclude models that in fact have an
-invertible Hessian. We found the Hessian at the optimized log-likelihood
-can often be invertible when the maximum absolute gradient was much
-larger than what would perceived to be a sensible threshold.
-
-Li et al. (2024) found that convergence rate could be a useful
-diagnostic especially for separating the correct model from overly
-complex models. However, the criteria for convergence used in their
-study may also lead to limited ability to distinguish the correct model
-from overly simplistic models, a pattern that was also noted by
-Liljestrand et al. (2024) in which one process error may absorb all
-sources of process error when the magnitude of other process errors are
-low.
-
-Often poor convergence result when parameter estimates are at their
-bounds (Carvalho et al. 2021), and this also applies to variance
-parameters for random effects with state-space assessment models. Even
-when the Hessian is invertible, parameters that are poorly informed will
-have extremely large variance estimates. This further inspection can
-lead to a more appropriate and often more parsimonious model
-configuration where the problematic parameters are not estimated. For
-example, process error variance parameters that are estimated close to 0
-indicates that the random effects are estimated to have little or no
-variability and removing these process errors is warranted. Generally,
-our results suggest we can expect lower probability of convergence of
-state-space assessment models when estimating natural mortality or SRRs
-because of the difficulty distinguishing these parameters from others
-being estimated in assessment model with data that are typically
-available. Our experiments did not aim to emulate the practitioner
-decision process in developing model configurations (e.g.~removing a
-source of process error and refitting the model when process error
-variance parameters were estimated close to 0). Evaluating the efficacy
-of such a decision process when applying EMs might be important in
-closed loop simulations (e.g.~MSE) aimed at quantifying management
-performance.
-
-A factor affecting the convergence criteria, particularly for maximum
-likelihood estimation of models with random effects, is numerical
-accuracy. All optimizations performed in these simulations are of the
-Laplace approximation of the marginal likelihood and, therefore,
-gradients and Hessians are also with respect to this approximation (see
-TMB::sdreport in the Template Model Builder package). Functionality
-within the Template Model Builder package exists (i.e.,
-TMB::checkConsistency) to check the validity of the Laplace
-approximation and the utility of this as a diagnostic for state-space
-assessment models should be explored further. Furthermore, numerical
-methods are used to calculate and invert the Hessian for variance
-estimation for models with random effects. Along with our results, the
-potential lack of accuracy imposed by these approximations, suggests at
-least investigating whether the Hessian is positive definite when the
-calculated absolute gradients are not terribly large (e.g, \(< 1\)).
-
-\subsection*{AIC}\label{aic}
-\addcontentsline{toc}{subsection}{AIC}
-
-Of the OM process error configurations we considered, we found AIC to be
-accurate for selecting models with process errors on recruitment and
-apparent survival (R and R+S). Fitting models to other OMs rarely
-preferred R+S EMs, and R and R+S EMs were nearly always selected for the
-matching OMs; a similar result was reported by Liljestrand et al.
-(2024). For other sources of process error, accuracy of AIC was improved
-when there was larger variability in the process errors and/or lower
-observation error.
-
-Across all OM process error configurations, AIC performed poorly in
-identifying that the presence of the Beverton-Holt SRR in the OM unless
-there was contrast in fishing pressure possibly in combination with
-other factors such as lower variability in recruitment process errors
-(in R and R+S models) or greater variation in natural mortality process
-errors (for R+M OMs, Fig. \ref{sr_aic}). As such, properly accounting
-for process error in natural mortality could be important (Li et al.
-2024) when evaluating SRRs in state-space models. Curiously, we did not
-find a marked effect of the level of observation error on ability to
-detect the SRR, but it is possible that AIC would perform better if
-observations have even lower uncertainty than we considered.
-
-Although we did not compare models with alternative SRRs (e.g., Ricker
-and Beverton-Holt), we do not expect AIC to perform any better
-distinguishing between relationships. Our finding that AIC tended to
-choose simpler recruitment models in most cases contrasts with the noted
-bias in AIC for more complex models (Shibata 1976; Katz 1981; Kass and
-Raftery 1995), but, whereas those findings apply to the much more common
-comparison of models that are fit to raw and independent observations,
-here we are comparing state-space models which account for observation
-error and estimate process errors in latent variables.
-
-Our results comport with those of de Valpine and Hastings (2002) who
-found AIC could not distinguish among state-space SRRs that were fit
-just to SSB and recruitment observations (i.e., not an assessment
-model). Similarly, Britten et al. (In review) found AIC could not
-reliably distinguish alternative environmental effects on SRR
-parameters. However, Miller et al. (2016) did find AIC to prefer a SRR
-with environmental effects when applied to data for the SNEMA yellowtail
-flounder stock and AIC also selected an environmental covariate on a SRR
-for the most recent stock assessment of Georges Bank yellowtail flounder
-(NEFSC 2025). Both of these yellowtail flounder stocks have large
-changes in stock size and the values of environmental covariates over
-time. Additionally, this species is well-observed by the bottom trawl
-survey that is used for an index in assessment models.
-
-\subsection*{Bias}\label{bias-2}
-\addcontentsline{toc}{subsection}{Bias}
-
-As expected, bias in all parameters and assessment output was generally
-improved with lower observation error. Estimation of SRR parameters was
-reliable in ideal scenarios of low observation error and contrast in
-fishing for some R+Sel and R+M OMs, but generally estimation was biased
-and(or) highly variable. We found substantial bias in estimated SRR
-parameters in R and R+S OMs particularly with high variability in
-recruitment and apparent survival process errors, suggesting that
-practitioners should be cautious when fitted assessment models have
-these properties.
-
-On the other hand, estimation of median natural mortality was reliable
-in many OM scenarios with contrast in fishing pressure, consistent with
-Hoenig et al. (2025). In some OMs, when EMs estimated the SRR parameters
-and median natural mortality, bias for those parameters was improved.
-Conversely, for some R+Sel and R+q OMs where there was bias in natural
-mortality due to high observation error, estimating the SRR reduced the
-bias in median natural mortality rate. However, estimating median
-natural mortality did cause poor accuracy in SSB estimation in many OMs
-without contrast in fishing pressure over time and with higher
-observation error. Thus, estimating median natural mortality should be
-approached with caution in state-space assessment models, particularly
-given its significant impact on determination of reference point and
-stock status (Li et al. 2024).
-
-\subsection*{Retrospective patterns}\label{retrospective-patterns}
-\addcontentsline{toc}{subsection}{Retrospective patterns}
-
-Incorrect EM process error assumptions did not produce strong
-retrospective patterns for SSB for any OMs regardless of whether median
-natural mortality or a SRR was estimated, but some weak retrospective
-patterns occur when observation error was high and there was contrast in
-fishing pressure. However, retrospective patterns tended to be more
-variable for recruitment and were sometimes large even when the EM was
-correct. Therefore, we recommend emphasis on inspection of retrospective
-patterns primarily for SSB and \(F\), but further research on
-retrospective patterns in other assessment model parameters, management
-quantities such as biological reference points, and projections may be
-beneficial (Brooks and Legault 2016).
-
-The general lack of retrospective patterns with mis-specified process
-errors is perhaps to be expected. Retrospective patterns are often
-induced in simulation studies by rapid changes in a quantity such as
-index catchability, natural mortality, or perceived catch during years
-toward the end of the time series (Legault 2009; Miller and Legault
-2017; Huynh et al. 2022; Breivik et al. 2023). In our simulations, the
-process errors changing over time may have trends in particular
-simulations, particularly when strong autocorrelation is imposed, but
-the random effects have no trend on average across simulations.
-Szuwalski et al. (2018) and lietal24 also found relatively small
-retrospective patterns when the source of mis-specification was temporal
-variation in demography attributes. Indeed, it is common for the
-flexibility provided by temporal random effects to reduce retrospective
-patterns (Miller et al. 2018; Stock et al. 2021; Stock and Miller 2021),
-though it does not necessarily indicate a more accurate assessment model
-(Perretti et al. 2020; Li et al. 2024; Liljestrand et al. 2024). Our
-results together with the existing literature seem to suggest that when
-a strong retrospective pattern is observed in an assessment it is more
-likely to be due to a mis-specification of a rapid shift in some model
-attribute rather than whether a particular process is assumed to be
-randomly varying temporally.
-
-\subsection*{Conclusions}\label{conclusions}
-\addcontentsline{toc}{subsection}{Conclusions}
-
-Our simulation study examined the importance of several factors for
-reliable inferences from state-space age-structured assessment models.
-Contrast in fishing pressure was consistently an important factor across
-all measures of reliability we examined. AIC accurately distinguished
-models with process errors on recruitment only (R) or on recruitment and
-apparent survival (R+S). Accuracy for other process error types required
-a strong signal (high process variability) with low noise (low
-observation uncertainty). Therefore, we expect practitioners will find
-R+S configurations to provide satisfactory diagnostics across a range of
-life history and data quality scenarios. AIC generally performed poorly
-for selecting the SRR, but performance was improved with low recruitment
-variability and contrast in fishing pressure. Some bias in estimation in
-at least one of the SRR parameters existed in nearly all OM-EM
-combinations. Because bias in terminal SSB and retrospective patterns
-were indifferent to whether or not the SRR was estimated, and
-convergence was slightly better without the SRR, a sensible default
-would be to fit models without an assumed SRR.
-
-\section*{Acknowledgements}\label{acknowledgements}
-\addcontentsline{toc}{section}{Acknowledgements}
-
-This work was funded by NOAA Fisheries Northeast Fisheries Science
-Center. We thank Jon Deroba for helpful comments on an earlier version
-of this manuscript.
-
-\pagebreak
-
 \section*{References}\label{references}
@@
 
-\phantomsection\label{refs}
+\protect\phantomsection\label{refs}
 \begin{CSLReferences}{1}{0}
@@
 \bibitem[\citeproctext]{ref-augeretal21}
-Auger-Méthé, M., Newman, K., Cole, D., Empacher, F., Gryba, R., King,
+{Auger-Méthé, M., Newman, K., Cole, D., Empacher, F., Gryba, R., King,
 A.A., Leos-Barajas, V., Mills Flemming, J., Nielsen, A., Petris, G., and
-others. 2021. A guide to state--space modeling of ecological time
+others}. 2021. A guide to state--space modeling of ecological time
 series. Ecological Monographs \textbf{91}(4): e01470.
@@
 \bibitem[\citeproctext]{ref-devalpinehastings02}
-de Valpine, P., and Hastings, A. 2002. Fitting population models
+{de Valpine, P., and Hastings, A.} 2002. Fitting population models
 incorporating process noise and observation error. Ecological Monographs
@@
 \end{CSLReferences}
-
-\pagebreak
-
-\clearpage
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{convergence_classification_plots}
-\end{center}
-\caption{Classification trees indicating primary factors determining convergence as defined by providing hessian-based standard errors for R, R+S, R+M, R+Sel and R+q OMs.
-Lower or higher convergence rates are indicated by red or green polygons, respectively}\label{conv_class}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{AIC_PE_classification_plots}
-\end{center}
-\caption{Classification trees indicating primary factors determining which EM process error assumption provides the lowest AIC for R+S, R+M, R+Sel and R+q OMs.
-Each node shows the proportion of EM process error models with lowest AIC (top) and number of observations (bottom) for the corresponding subset.
-Lower or higher accuracy of the process error assumption are indicated by red or green polygons, respectively.}\label{AIC_PE_class}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{AIC_SRR_classification_plots}
-\end{center}
-\caption{Classification trees indicating primary factors determining which EM stock recruitment assumption (none or Beverton-Holt) provides the lowest AIC for R, R+S, R+M, R+Sel and R+q OMs.
-Each node shows the proportion of EMs that assume the stock-recruit relationship with lowest AIC (top) and number of observations (bottom) for the corresponding subset.
-Lower or higher accuracy of the process error assumption are indicated by more red or green polygons, respectively.}\label{AIC_SRR_class}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{SR_a_bias_regtree_plots}
-\end{center}
-\caption{Regression trees indicating primary factors determining reductions in sums of squares of errors in estimation measured by Eq.
-\ref{bias_regression_response} for the Beverton-Holt stock-recruit parameter $a$ for R+S, R+M, R+Sel and R+q OMs.
-Each node shows the median error (top) and number of observations (bottom) for the corresponding subset.
-Lower or higher median absolute errors of the process error assumption are indicated by more green or red polygons, respectively.}\label{SR_a_bias_regtree}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{SR_b_bias_regtree_plots}
-\end{center}
-\caption{Regression trees indicating primary factors determining reductions in sums of squares of errors in estimation measured by Eq.
-\ref{bias_regression_response} for the Beverton-Holt stock-recruit parameter $b$ for R+S, R+M, R+Sel and R+q OMs.
-Each node shows the median error (top) and number of observations (bottom) for the corresponding subset.
-Lower or higher median absolute errors of the process error assumption are indicated by more green or red polygons, respectively.}\label{SR_b_bias_regtree}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{med_M_bias_regtree_plots}
-\end{center}
-\caption{Regression trees indicating primary factors determining reductions in sums of squares of errors in estimation measured by Eq.
-\ref{bias_regression_response} for the median natural mortality rate for R+S, R+M, R+Sel and R+q OMs.
-Each node shows the median error (top) and number of observations (bottom) for the corresponding subset.
-Lower or higher median absolute errors of the process error assumption are indicated by more green or red polygons, respectively.}\label{med_M_bias_regtree}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{SSB_bias_regtree_plots}
-\end{center}
-\caption{Regression trees indicating primary factors determining reductions in sums of squares of errors in estimation measured by Eq.
-\ref{bias_regression_response} for terminal year spawning stock biomass for R+S, R+M, R+Sel and R+q OMs.
-Each node shows the median error (top) and number of observations (bottom) for the corresponding subset.
-Median errors closer to or further from zero are indicated by more green or red polygons, respectively.}\label{SSB_bias_regtree}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{SSB_mohns_rho_regtree_plots}
-\end{center}
-\caption{Regression trees indicating primary factors determining reductions in sums of squares of errors in transformed Mohn's $\rho$ (Eq.
-\ref{bias_regression_response}) for spawning stock biomass for R+S, R+M, R+Sel and R+q OMs.
-Each node shows the median Mohn's $\rho$ (top) and number of observations (bottom) for the corresponding subset.
-Median Mohn's $\rho$ closer to or further from zero are indicated by more green or red polygons, respectively.}\label{SSB_mohns_rho_regtree}
-\end{figure}
-\end{landscape}
-
-\pagebreak
-
-\begin{table}
-\caption{For each OM process error type (columns), percent reduction in deviance for logistic regression models fit to indicators of convergence with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{convergence_PRD_table}
-{\input{convergence_PRD_table}}
-\end{table}
-
-\begin{table}
-\caption{For each OM process error type (columns), percent reduction in deviance for multinomial logistic regression models fit to indicators of EM process error assumption with lowest AIC with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{AIC_PE_PRD_table}
-{\input{AIC_PE_PRD_table}}
-\end{table}
-
-\begin{table}
-\caption{For each OM process error type (columns), percent reduction in deviance for logistic regression models fit to indicators of EM stock-recruit assumption (none or Beverton-Holt) with lowest AIC with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{AIC_SRR_PRD_table}
-{\input{AIC_SRR_PRD_table}}
-\end{table}
-
-\begin{landscape}
-\begin{table}
-\caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
-\ref{bias_regression_response} for the Beverton-Holt stock recruit relation ship parameters with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_SR_pars_PRD_table}
-{\input{bias_SR_pars_PRD_table}}
-\end{table}
-\end{landscape}
-
-\begin{table}
-\caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
-\ref{bias_regression_response} for the median natural mortality rate parameter with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_median_M_PRD_table}
-{\input{bias_median_M_PRD_table}}
-\end{table}
-
-\begin{table}
-\caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
-\ref{bias_regression_response} for the terminal year spawning biomass with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_SSB_PRD_table}
-{\input{bias_SSB_PRD_table}}
-\end{table}
-
-\begin{table}
-\caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to transformed Mohn's $\rho$ values for each simulation (Eq.
-\ref{bias_regression_response}) for spawning biomass with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{mohns_rho_SSB_PRD_table}
-{\input{mohns_rho_SSB_PRD_table}}
-\end{table}
-
-\pagebreak
-
-\setcounter{figure}{0}
-\renewcommand\thefigure{S\arabic{figure}}
-
-\setcounter{table}{0}
-\renewcommand\thetable{S\arabic{table}}
-
-\pagebreak
-
-\section*{Supplementary Materials}\label{supplementary-materials}
-\addcontentsline{toc}{section}{Supplementary Materials}
-
-\pagebreak
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width = \textwidth]{om_input_plots_figure}
-\end{center}
-\caption{The proportion mature at age, weight at age, fleet and index selectivity at age, and Beverton-Holt stock-recruit relationship assumed for the population in all operating models.
-For operating models with random effects on fleet selectivity, this represents the selectivity at the mean of the random effects.}\label{om_inputs_fig}
-\end{figure}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{type_4_convergence_plots}
-\end{center}
-\caption{Estimated probability of fits providing hessian-based standard errors for EMs assuming alternative process error (colored points and lines), and median natural mortality (estimated or known) and Beverton-Holt stock-recruit relationships (estimated or not; along x-axis) when fitted to operating models that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{hessian_SE_convergence}
-\end{figure}
-\end{landscape}
-
-\clearpage
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{type_3_convergence_plots}
-\end{center}
-\caption{Probability of estimating models providing maximum absolute values of gradients less than $10^{-6}$ assuming alternative process error (colored points and lines), and median natural mortality (estimated or known) and Beverton-Holt stock-recruit relationships (estimated or not; along x-axis) when fitted to operating models that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{gradient_convergence}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{hess_grad_convergence_plots}
-\end{center}
-\caption{The maximum of the absolute values of all gradient values for all fits that provided hessian-based standard errors across all simuated data sets of a given OM configuration (A: R and R+S, B: R+M, C: R+Sel, or D: R+q).
-Results are conditional on EM fits with alternative process error type (colored points and lines), median natural mortality (estimated or known) and recruitment assumptions (Beverton-Holt stock-recruit relationship or not).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{hess_grad}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{pe_aic_plots}
-\end{center}
-\caption{Estimated probability of lowest AIC for EMs assuming alternative process error structures (colored bars) conditional on alternative assumptions for median natural mortality (estimated or known) and Beverton-Holt stock-recruit relationships (estimated or not; along x-axis) when fitted to operating models that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
-Striped bars indicate results where the EM process error structure matches that of the operating model.}\label{pe_aic}
-\end{figure}
-\end{landscape}
-
-\clearpage
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics[height = 0.8\textheight]{sr_aic_plots_rev}
-\end{center}
-\caption{Estimated probability of lowest AIC from logistic regression on the log-standard deviation of the true log(SSB) in each simulation for estimating model with Beverton-Holt stock-recruit relationships, rather than the otherwise equivalent EM without the stock-recruit relationship.
-Results are conditional on median M is known in the EM and alternative assumptions EMs having the correct process error structure: R and R+S (A), R+Sel (B), R+M (C), or R+q (D), and median M is assumed known in the EM.
-Solid and dashed lines are for OMs with and without temporal contrast in fishing pressure, respectively, and polygons represent 95\% confidence intervals.
-Range of results indicates the range of log-standard deviation of log(SSB) for simulations of the particular OM.}\label{sr_aic}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{sr_aic_plots}
-\end{center}
-\caption{Estimated probability of lowest AIC from logistic regression on the log-standard deviation of the true log(SSB) in each simulation for estimating model with Beverton-Holt stock-recruit relationships, rather than the otherwise equivalent EM without the stock-recruit relationship.
-Results are conditional on alternative assumptions for median natural mortality (estimated or known) and on EMs having the correct process error structure: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Rug along x-axis denotes $SD(\log(SSB))$ values for each simulation and polygons represent 95\% confidence intervals.}\label{sr_aic_supp}
-\end{figure}
-\end{landscape}
-
-\clearpage
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{term_SSB_bias_plots}
-\end{center}
-\caption{Median relative error of terminal year SSB for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{SSB_rel_error}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{term_F_bias_plots}
-\end{center}
-\caption{Median relative error of terminal year fully-selected fishing mortality for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{F_rel_error}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{term_R_bias_plots}
-\end{center}
-\caption{Median relative error of terminal year recruitment for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{R_rel_error}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{sr_bias_plots}
-\end{center}
-\caption{Median relative error of Beverton-Holt stock-recruit parameters ($a$ and $b$) for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{SR_rel_error}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{M_bias_plots}
-\end{center}
-\caption{Median relative error of median natural mortality for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{M_rel_error}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{mohns_rho_ssb_plots}
-\end{center}
-\caption{Median Mohn's $\rho$ for SSB for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_ssb}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{mohns_rho_F_plots}
-\end{center}
-\caption{Median Mohn's $\rho$ of fishing mortality averaged over all age classes for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_F}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{mohns_rho_R_plots}
-\end{center}
-\caption{Median Mohn's $\rho$ of recruitment for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_R}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{table}
-\caption{Distinguishing characteristics of the operating models with random effects on recruitment and apparent survival (R.R+S).
-Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
-Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.}\label{naa_om_table}
-{\footnotesize \input{naa_om_tab}}
-\end{table}
-\end{landscape}
-
-\begin{landscape}
-\begin{table}
-\caption{Distinguishing characteristics of the operating models with random effects on recruitment and natural mortality (R+M).
-Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
-Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
-For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}\label{M_om_table}
-{\input{M_om_tab}}
-\end{table}
-\end{landscape}
-
-\begin{landscape}
-\begin{table}
-\caption{Distinguishing characteristics of the operating models with random effects on recruitment and selectivity (R+Sel).
-Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
-Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
-For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}\label{sel_om_table}
-{\input{Sel_om_tab}}
-\end{table}
-\end{landscape}
-
-\begin{landscape}
-\begin{table}
-\caption{Distinguishing characteristics of the operating models with random effects on recruitment and catchability (R+q).
-Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
-Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
-For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}\label{q_om_table}
-{\input{q_om_tab}}
-\end{table}
-\end{landscape}
-
-\begin{table}
-\caption{Distinguishing characteristics of the estimating models and operating model process error categories (R, R+S, R+M, R+Sel, R+q) where used.}\label{em_table}
-{\scriptsize \input{em_tab}}
-\end{table}
-
-\begin{table}
-\caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to transformed Mohn's $\rho$ values for each simulation (Eq.
-\ref{bias_regression_response}) for fishing mortality averaged over all age classes with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{mohns_rho_F_PRD_table}
-{\input{mohns_rho_F_PRD_table}}
-\end{table}
-
-\begin{table}
-\caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to transformed Mohn's $\rho$ values for each simulation (Eq.
-\ref{bias_regression_response}) for recruitment with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{mohns_rho_R_PRD_table}
-{\input{mohns_rho_R_PRD_table}}
-\end{table}
-
-\end{document}

### 2025-10-01 Project_0/manuscript/manuscript_R1.Rmd
saving work on project 0 revisons

@@
 
-State-space models are increasingly used for stock assessment, and evaluations of their statistical reliability and best practices for selecting among process error configurations are needed.
-We simulated 72 operating models that varied fishing pressure and observation error across process errors in recruitment, survival, selectivity, catchability, and/or natural mortality.
-We fit estimating models with different assumptions on the process error source and whether median natural mortality or a stock-recruit relationship were estimated.
-Estimating models without a stock-recruit relationship that assumed the correct process error source and median natural mortality had high convergence rates and low bias.
-Bias was also low under many incorrect process error assumptions when there was contrast in fishing pressure and low observation error.
-Marginal AIC most accurately distinguished process errors on recruitment, survival, and selectivity, and other process error sources when variability was greater.
-Retrospective patterns were generally small but were sizable for recruitment when observation error was high.
-These results help establish the statistical reliability of state space assessment models and pave the way for the next-generation of fisheries stock assessment.
+State-space models are increasingly used for stock assessment, and evaluations of statistical inferences made from them is needed.
+We simulated 72 operating models that varied fishing pressure and observation error across process errors in recruitment, apparent survival, selectivity, catchability, and/or natural mortality.
+We fit estimating models with different assumptions on the process error source and whether natural mortality and stock-recruit relationship were estimated.
+We measured reliability of estimating models by probability of convergence, accuracy of AIC for correct process error configuration and stock-recruit function, estimation bias, and magnitude of retrospective patterns.
+Across all measures reliability was generally better with lower observation error, contrast in fishing pressure over time, and median natural mortality rate known, but using marginal AIC to distinguishing the stock recruit relationship further required large contrast in spawning biomass and low residual variation in recruitment.
+Marginal AIC accurately distinguished process errors on recruitment, apparent survival, and selectivity.
+These results improve our understanding of when results from state space assessment models, the next-generation of fisheries stock assessment, can be expected to be reliable.
 
@@
 Similarly, For other OM process error categories, we fit EMs with matching process error assumptions as well as other process error types without autocorrelation.
-The maturity at age, weight at age for catch and SSB, and observation error variance of aggregate catch and indices were all assumed known at the true values.
+The maturity at age, weight at age for catch and spawning stock biomass (SSB), and observation error variance of aggregate catch and indices were all assumed known at the true values.
 However, the variance parameters for the logistic-normal distributions for age composition observations were estimated in the EMs.
@@
 The first measure of reliability we investigated was frequency of convergence when fitting each EM to the simulated data sets.
-There are various ways to assess convergence of the fit [e.g., @carvalhoetal21; @kapuretal24], but given the importance of estimates of uncertainty when using assessment models in management, we estimated probablity of convergence as measured by occurrence of a positive-definite hessian matrix at the optimized negative log-likelihood that could be inverted.
-We also provide results in the Supplementary Materials for the maximum of the absolute values among all gradients for all fits of a given EM to all simulated data sets from a given OM that produced hessian-based standard errors for all estimated fixed effects.
+There are various ways to assess convergence of the fit [e.g., @carvalhoetal21; @kapuretal24], but given the importance of estimates of uncertainty when using assessment models in management, we estimated probability of convergence as measured by occurrence of a positive-definite hessian matrix at the optimized negative log-likelihood that could be inverted (i.e., providing hessian-based standard error estimates).
+We also provide results in the Supplementary Materials for convergence defined by the maximum absolute gradient < $1^{-6}$ and the maximum of the absolute gradient values for all fits of a given EM to all simulated data sets from a given OM that produced hessian-based standard errors for all estimated fixed effects.
 This provides an indication of how poor the calculated gradients can be, but still presumably converged adequately enough for parameter inferences.
-We used the Clopper-Pearson exact method for constructing 95\% confidence intervals of the probabilities of convergence [@clopperpearson34;@thulin14].
 
@@
 
-We estimated the probability of selection of each process error model structure (R, R+S, R+M, R+Sel, R+q) using marginal AIC.
-For a given operating model, we compared AIC for EMs that were all configured the same for median natural mortality (known or estimated) and the SRR (Beverton-Holt or none).
+We investigated the reliability of AIC-based model selection for two purposes.
+First, we analyzed selection of each process error model structure (R, R+S, R+M, R+Sel, R+q) using marginal AIC.
+For a given OM simulated data set, we compared AIC for EMs with different process error assumption conditional on whether median natural mortality rate and the Beverton-Holt SRR were estimated.
+We tabulated the models providing the lowest AIC across simulated data sets.
+Second, we analyzed AIC-based selection between EMs with and without the Beverton-Holt SRR assumed.
+Contrast in fishing pressure and time series with recruitment at low stock size has been shown to improve estimation of SRR parameters [@magnussonhilborn07; @connetal10].
+Our preliminary inspections indicated generally poor performance of AIC in determining the Beverton-Holt model for a given set of OM factors (including contrast in fishing pressure).
+Therefore, we also considered the effect of of the log-standard deviation of the true log(SSB) (similar to the log of the coefficient of variation for SSB) on model selection since simulations with realized SSB producing low and high recruitment would have larger variation in realized SSB.
 
-We also estimated the probability of correctly selecting between EMs with Beverton-Holt SRR assumed and models without the SRR (null model).
-We made these comparisons between models that otherwise assumed the same process error structure as the operating model and both of the compared models either estimate median natural mortality or assume it is known.
-Contrast in fishing pressure and time series with recruitment at low stock size has been shown to improve estimation of SRR parameters [@magnussonhilborn07; @connetal10].
-Our preliminary inspections of the proportions of simulations where the correct recruitment model was chosen for a given set of OM factors (including contrast in fishing pressure) indicated generally poor performance of AIC.
-Therefore, we fit logistic regression models to the indicator of Beverton-Holt models having lower AIC as a function of the log-standard deviation of the true log(SSB) (similar to the log of the coefficient of variation for SSB) since simulations with realized SSB producing low and high recruitments would have larger variation in realized SSB.
-
-All model selection results condition only on whether all of the compared estimating models completed the optimization process without failure.
-We did not condition on convergence as defined by a gradient threshold or invertibity of the hessian because optimization could correctly determine an inappropriate process error assumption by estimating variance parameters at the lower bound of zero.
+All model selection results condition only on completion of the optimization process without failure for all of the compared EMs.
+We did not condition on convergence as defined above because optimization could correctly determine an inappropriate process error assumption by estimating variance parameters at the lower bound of zero.
 Such an optimization could indicate poor convergence but the likelihood would be equivalent to that without the mis-specified random effects and the AIC would be appropriately higher because more (variance) parameters were estimated.
@@
 
-For a given model attribute we calculated the relative error 
+We also investigated bias in estimation of various model attributes as a measure of reliability. For a given model attribute we calculated the relative error 
 \begin{equation}\label{relerror}
@@
 from fitting a given estimating model to simulated data set $i$ configured for a given OM where $\widehat \theta_i$ and $\theta_i$ are the estimated and true values for simulation $i$.
-We estimated bias as the median of the relative errors across all simulations for a given OM and EM combination.
-We constructed 95\% confidence intervals for the median relative bias using the binomial distribution approach as in @millerhyun18 and @stockmiller21.
-We present results for bias in terminal year estimates of SSB and recruitment, Beverton-Holt stock recruit parameters ($a$ and $b$), and median natural mortality rate.
-Results for terminal year fishing mortality were strongly negatively correlated with those for SSB and are provided in the Supplementary Materials.
+We analyzed simulation results for estimates of terminal year estimates of SSB and recruitment, Beverton-Holt SRR parameters ($a$ and $b$), and median natural mortality rate.
 
@@
 
-We calculated Mohn's $\rho$ for SSB, fishing mortality (averaged over all age classes), and recruitment for each EM [@mohn99].
-We fit $P =7$ peels to each simulated data set and calculated Mohn's $\rho$ for a given attribute $\theta$ as
+Finally, we investigated presence of retrospective patterns in fitted models as a measure of reliability.
+We calculated Mohn's $\rho$ for SSB, fishing mortality (averaged over all age classes), and recruitment for each EM fit to each OM simulated data set [@mohn99].
+We fit $P = 7$ peels to each simulated data set and calculated Mohn's $\rho$ for a given attribute $\theta$ as
 \begin{equation}\label{mohns_rho}
@@
 where $\widehat \theta_{i,j}$ is the estimate for attribute $\theta$ in year $i$ from a model fit using data up to year $j$.
-For each EM we calculated median and 95\% confidence intervals using the same methods as that for relative bias.
 
@@
 
-The measures of central tendency and variability of of observed values for specific OM and EM attributes (e.g., low or high observation error) that we described above can indicate scenarios that provide better or poorer reliability.
+The measures of central tendency and variability of observed values for specific OM and EM attributes (e.g., low or high observation error) that we described above can indicate scenarios that provide better or poorer reliability.
 However, the OM and EM attributes that we investigated are numerous, so we used two methods to summarize the most important factors for differences in results.
 The first method was fitting regression models with the response being each of the measures of reliability described above and predictor variables were defined based on OM and EM characteristics [e.g., @mackinnonetal95; @wangetal17; @harwelletal18].
-For the binary results for convergence and AIC-based selection of a stock-recruit relationship, we performed logistic regressions.
-For results for AIC-based selection of EM process error type (multiple categories) we performed multinomial regressions.
-For other measures of reliability we fit standard linear regression models.
+For the binary indicators of convergence and AIC-based selection of a SRR, we performed logistic regressions.
+For indicators of AIC-based selection of EM process error type (multiple categories) we performed multinomial regressions.
+For other measures of reliability we fit linear regression models to transformed responses.
 Because relative errors (Eq.
@@
 The response variables were the same as the regressions for the deviance reduction analyses.
-We used the \verb|rpart| function in the rpart package (https://cran.r-project.org/package=rpart) to fit trees and, for clarity we pruned them to show the primary branches.
+We used the \verb|rpart| function in the rpart package (https://cran.r-project.org/package=rpart) to fit trees.
+Full trees were determined using default settings except that we increased the number of cross-validations to 100.
+For clarity, we pruned the full trees to show just the primary branches.
+
+We also provide detailed results for all measures of reliability at each combination of OM and EM attributes in the Supplementary Materials.
+For confidence intervals of probability of convergence, we used the Clopper-Pearson exact method [@clopperpearson34;@thulin14].
+For AIC selection of process error configuration we provide estimates of the proportions of simulations where each EM type was selected.
+For AIC selection of the SRR, we provide predicted probabilities from logistic regressions as a function of SSB variability for each OM and EM type.
+We estimated bias as the median of the relative errors across all simulations for a given OM and EM combination.
+We constructed 95\% confidence intervals for the median relative bias, and Mohn's $\rho$ using the binomial distribution approach as in @millerhyun18 and @stockmiller21.
+For each EM we calculated median and 95\% confidence intervals using the same methods as that for relative bias.
 
@@
 
-### Deviance reduction and classification trees {-}
+For probability of convergence, EM process error assumption was the single attribute that resulted in the largest percent reduction in deviance (14-28\%) for all OM process error types other than R+S OMs where the EM M assumption explained the most residual deviance (>11\%; Table \ref{convergence_PRD_table}).
+However, including interactions of OM and EM factors also provided large reductions in residual deviance (35-47\%), suggesting successful convergence depended on a combination OM and EM attributes.
 
-EM process error assumption was the single attribute that resulted in the largest percent reduction in deviance for all OM process error types other than R+S OMs where the EM M assumption explained the most residual deviance (Table \ref{convergence_PRD_table}).
-However, also including interactions of OM and EM factors also provided large reductions in residual deviance, suggesting successful convergence depended on a combination OM and EM attributes.
 Classification trees for each OM process error type, all had the primary branch defined using the same attribute that provided the largest reduction in deviance (Figure \ref{conv_class}).
+EMs that assumed R+S process errors converged poorly for all OMs that were simulated with the alternative process error assumptions (R, R+M, R+Sel, an R+q OMs).
+For all trees, branches based on the OM fishing mortality history showed better convergence when the OM included a change in fishing pressure.
+Branches based on whether the Beverton-Holt SRR was assumed or not, showed better convergence when it was not estimated and branches based on whether the EM estimated the median natural mortality rate showed better convergence when it was treated as known.
+For certain R+M and R+Sel OMs, better convergence was also observed when there was lower observation uncertainty.
 
-Generally, the factors defining primary splits for the trees corresponded to those with the largest reductions in deviance fitting the regression models.
-
-### Full results {-}
-
-For many R and R+S OMs, convergence rate declined when either the median natural mortality rate or the Beverton-Holt SRR was estimated even when the process error assumptions of the EMs and OMs matched (Figure \ref{hessian_SE_convergence}, A).
-When there was high observation error and constant fishing pressure ($F = F_{\text{MSY}}$ for all 40 years), convergence was poor for of all EM process error configurations other than R EMs when fitted to R OMs ($\sigma_{2+} = 0$) regardless of whether median natural mortality and SRRs were estimated.
-Convergence of R EMs was high for all R and R+S OMs except when there was high observation error and constant fishing pressure, and when median natural mortality and SRRs were estimated.
-R+S EMs fit to R OMs exhibited poor convergence regardless of whether natural mortality or a SRR was estimated.
-R+S EMs fit to R+S OMs had highest convergence rates when there was contrast in fishing pressure and low observation error.
-Convergence rates were high for all EMs when fit to data from R+S OMs with lower observation error except those where median natural mortality and/or SRRs were estimated.
-
-Convergence of all EMs fitted to R+M OMs was highest when the OMs had higher natural mortality process error variability, low observation error, and contrast in fishing pressure  (Figure \ref{hessian_SE_convergence}, B).
-R+M EMs that estimated autocorrelation of process errors had poor convergence for R+M OMs when there was low natural mortality process error variability regardless of autocorrelation of the simulated process errors.
-R+S EMs fitted to data generated from R+M OMs always converged poorly whether or not median natural mortality and the Beverton-Holt SRR were estimated.
-
-The R+S EMs, in particular, had poor convergence when fit to data generated from R+Sel OMs with lower selectivity process error variability or higher observation error (Figure \ref{hessian_SE_convergence}, C).
-R+Sel EMs generally converged better than other EMs for R+Sel OMs with higher process error variability, lower observation error, and contrast in fishing pressure regardless of whether median natural mortality or a SRR was estimated.
-
-For R+q OMs, convergence of R+q EMs was generally better than that of other EMs when there was contrast in fishing pressure (Figure (\ref{hessian_SE_convergence}, D).
-Convergence of R+S EMs was generally worse than that of all other EMs across all OMs whether or not median natural mortality or a SRR was estimated.
-Again, convergence probability generally declined for all EMs when median natural mortality or a SRR was estimated.
-
-We found a wide range of maximum absolute values of gradients for models that converged (Figure \ref{hess_grad}).
+When convergence is defined by a gradient threshold, the primary factor explaining deviance reduction is the same for all OM process error types, but there are some differences in deviance reduction for secondary factors (Table \ref{convergence_gradient_PRD_table}), and probability of convergence, overall, was lower (Figure \ref{conv_gradient_class}).
+We found a wide range of maximum absolute values of gradients for models that had invertible Hessians (Figure \ref{hess_grad}).
 The largest value observed for a given EM and OM combination was typically $<10^{-3}$, but many converged models had values greater than 1.
@@
 
-## AIC performance for process error structure {-}
+## AIC performance {-}
 
-### Deviance reduction and classification trees {-}
+### Process error structure {-}
 
-### Full results {-}
+For AIC selection of the correct process error configuration, the magnitude of observation and process error were the attributes that resulted in the largest percent reductions in deviance across OM process error types other than R OMs (Table \ref{AIC_PE_PRD_table}).
+Both variance of apparent survival random effects ($\sigma_{2+}$) and degree of observation error explained the largest reductions for R+S (17-22\%) and R+Sel (8-26\%) OMs, whereas variance of process errors provided the largest reductions in R+M (>9\%) and R+q (>13\%) OMs.
+Comparatively, none of the OM or EM attributes explained particularly large reductions in deviance for R OMs, but fishing history, whether a SRR was estimated, and whether median natural mortality was known or estimated provided similar and the largest reductions (approximately 5-6\%).
+Inclusion of second and third order interactions, did not provide large reductions in deviance for any of the OM process error types.
 
-Marginal AIC accurately determined the correct process error assumptions in EMs when data were generated from R and R+S OMs, regardless of whether median natural mortality or a SRR was estimated (Figure \ref{pe_aic}, A).
-Attempting to estimate median natural mortality or a SRR separately had a negligible effect on the accuracy of determining the correct process error assumption.
-When both were estimated, there was a noticeable reduction in accuracy when OMs had a constant fishing pressure, low observation error, and larger variability in recruitment process errors.
+For all OM process error types other than R OMs, the attributes defining the primary branches of classification trees matched those that provided the largest reductions in deviance (Figure \ref{AIC_PE_class}).
+Across all OMs, AIC was more accurate for the process error type when process error variability was greater and when observation error was lower.
+No branches were estimated for classification trees fit to the R OMs, likely because accuracy was high across all simulations (0.94), although inspection of the fine-scale results shows there is some degradation in AIC selection when a SRR and median natural mortality rate are estimated for R OMs with constant fishing pressure and high observation error (Figure \ref{pe_aic}, top left).
 
-For R+M OMs, marginal AIC only accurately determined the correct process error model and correlation structure when observation error was low and variability in natural mortality process errors was high (Figures \ref{pe_aic}, B).
-Of these OMs, estimating the median natural mortality rate only reduced the accuracy of AIC when natural mortality process errors were independent and fishing pressure was constant.
-For OMs with poor model selection accuracy, AIC most frequently selected EMs with process errors in catchability (R+q) or selectivity (R+Sel).
-Selection of R+S EMs was generally unlikely.
+### Stock-recruit relationship {-}
 
-Marginal AIC most accurately determined the correct source of process error and correlation structure for R+Sel OMs with low observation error (Figures \ref{pe_aic}, C).
-When there was low variability in selectivity process errors and high observation error, R+q or R+S EMs were more likely to have the best AIC.
-Whether median natural morality or SRRs were estimated appeared to have little effect on the performance of AIC.
+Logistic regressions for AIC selection of the Beverton-Holt SRR, showed OM fishing history and variation in SSB ($\log \text{SD}_{\text{SSB}}$) provided substantial reductions in deviance for R+M (>13\%), R+Sel (>26\%), and R+q (>24\%) OMs (Table \ref{AIC_SRR_PRD_table}).
+For R OMs, fishing history provided the largest reduction in deviance (>9\%) whereas none of the attributes individually provided large reductions in deviance for R+S OMs (all <5\%).
+However, inclusion of all attributes provided larger reductions in deviance than the sum of individual contributions for both R (>30\%) and R+S (>18\%) OMs.
+For all OM process error types, inclusion of interaction terms provided relatively little reduction in residual deviance.
 
-Marginal AIC most accurately determined the correct source of process error and correlation structure for R+q OMs with high variability in catchability process errors (Figures \ref{pe_aic},D).
-The R+q OMs with low variability in catchability process errors and high observation error had the least model selection accuracy.
-However, for these OMs, the marginal AIC accurately determined the correct source of process error (but not correlation structure) except when OMs assumed a constant fishing pressure and EMs estimated both median natural morality and the SRR.
-
-## AIC performance for the stock-recruit relationship {-}
-
-### Deviance reduction and classification trees {-}
-
-### Full results {-}
-
-Our comparisons of model performance conditioned on assuming the true process error configuration is known (EM and OM process error types match) and we focus on results where the EMs assume median natural mortality is known because there was little difference in results when the EMs estimated this parameter.
-Broadly, we found generally poor accuracy of AIC in selecting models assuming a Beverton-Holt SRR over the null model without an SRR for all OMs.
-However, we also found increased accuracy of AIC in determining the Beverton-Holt SRR when the simulated population exhibited greater variation in spawning biomass for nearly every OM (Figure \ref{sr_aic}).
-
-With R and R+S process error assumptions, probability of lowest AIC for the B-H SRR as a function of SSB variability were greatest for OMs with contrast in fishing pressure and lower process variability in recruitment (Figure \ref{sr_aic}, A).
-The largest variation in SSB occurred in OMs with larger recruitment variability ($\sigma_R = 1.5$; Figure \ref{sr_aic}, A, right column group), but the same high AIC accuracy was achieved for OMs with lower recruitment variability at lower levels of SSB variation.
-The level of observation error had little effect on AIC accuracy.
-
-For R+M OMs, probability of lowest AIC for the Beverton-Holt SRR increased steeply with variation in SSB whether it was induced by contrast in fishing or variation in natural mortality process error.
-(Figure \ref{sr_aic}, B).
-There was little difference in AIC accuracy whether the natural mortality process errors were correleted and, similar to R+S OMs, there was also little effect due to level of observation error.
-
-For R+Sel OMs, contrast in fishing pressure over time was the primary source of variation in SSB and these are the OMs where AIC accuracy for the Beverton-Holt SRR was greatest (Figure \ref{sr_aic}, C).
-There was little effect of variability or correlation of selectivity process errors or the level of observation error on AIC accuracy.
-
-Like the R+Sel OMs, the greatest accuracy for AIC in selecting the Beverton-Holt SRR occurred for R+q OMs where there was contrast in fishing pressure over time which is also where there was the greatest variation in SSB (Figure \ref{sr_aic}, D).
-There was also little effect of variability or correlation of catchability process errors or the level of observation error on AIC accuracy.
+Attributes defining the primary branches of classification trees for AIC selection of the SRR assumption were the same as those explaining the largest reductions in deviance for the logistic regression models (Figure \ref{AIC_SRR_class}).
+All branches based on variation in SSB showed better accuracy with larger variability in SSB and all branches based on fishing history showed better accuracy when there was contrast in fishing pressure.
+Branched based on OM observation error or recruitment variability showed better accuracy when they were lower.
+For R OMs, a combination of lower recruitment variability, contrast in fishing pressure, and higher SSB variability produced AIC accuracy over 0.8.
+For R+S OMs, lower recruitment variability and observation error and higher SSB variability produced AIC accuracy of 0.79.
+For R+M, R+Sel, and R+q OM,s accuracy of 0.87 to 0.94 was observed just with higher SSB variability.
 
@@
 
-### Deviance reduction and regression trees {-}
+### Terminal year spawning stock biomass, fishing mortality, and recruitment {-}
 
-Across all OM process error types, fishing history was the most important single factor in reducing residual deviation in log-scale errors for the Beverton-Holt density-independent parameter $a$, and also for the density-independent parameter $b$ for R+M, R+Sel, and R+q OMs (Table \ref{bias_SR_pars_PRD_table}).
-For R and R+S OMs the EM process error assumption was equally important for estimating $b$, but the the reductions were small (mostly less than 4%) for any factor and any OM process error type.
-There was also no substantial reduction in deviance with second or third order interactions of factors for either parameter in any OM process error types.
+Regression models for log-scale errors in SSB that included the various OM and EM factors showed little reduction in deviance (<5\%) for any of the factors across all OM process error types (Table \ref{bias_SSB_PRD_table}).
+The attributes producing the largest reductions were the EM assumption for median natural mortality (known or estimated) for R, R+M, R+Sel, and R+q OMs (1-3\%), EM process error type for R+S OMs (4\%) and fishing history for all OM types (1-5\%).
+Including second interactions provided largest reductions in residual deviance (10- 26\%).
+Including third order interactions also provided large further reductions for R, R+S, and R+q OMs (between 5 and 11\%).
+In all regression trees, branches based on fishing history and level of observation error generally showed lower bias in SSB with contrast in fishing and lower observation error (Figure \ref{SSB_bias_regtree}).
+For branches based on treatment of median natural mortality rate, bias was generally lower when it was known rather than estimated.
+For some R+Sel and R+q OMs, less bias in SSB was shown when the EM process error configuration was correct.
 
-The regression trees indicate that having contrast in fishing pressure reduced the median error for both SR parameters for all OM types
+Results for bias in fishing mortality and recruitment generally matched those for SSB, except that directions of bias for fishing mortality were opposite to those for SSB and recruitment.
+Effects of individual OM and EM factors on regression models were similarly small as measured by reduction in deviance (Tables \ref{bias_F_PRD_table} and \ref{bias_R_PRD_table}).
+Primary branches of regression trees were in most cases identical to those for SSB  (Figures \ref{F_bias_regtree} and \ref{R_bias_regtree}).
 
-We observed different effects of factors on deviance reduction for errors in estimation of the two Beverton-Holt stock-recruit parameters.
-When analyses were For bias of the density-dependent parameter $b$ as measured by the log absolute value of log-scale errors, deviance was not reduced by any of the individual OM or EM factors except for variation in recruitment $\sigma_R$ for R and R+S OMs.
-However, second and(or) third order interactions of the OM and EM factors also provided large reductions in deviance for all OM process error types which indicated that some combinations of factors could reduce bias for the $b$ parameter.
-Despite the explanation variation in the $b$ estimation erorrs, bias was still significant regardless of the combination of OM and EM factors (Figures regtrees and full results).
+### Stock-recruit parameters {-}
 
-Median natural mortality rate showed reductions in deviance for R, R+S, and R+M OMs with second order interactions and R+M also showed reductions with third order interactions.
+Regression models for transformed relative errors of estimates of both the Beverton-Holt $a$ and $b$ parameters showed none of the factors explained large percent reductions in deviance (Table \ref{bias_SR_pars_PRD_table}).
+The OM fishing history provided the largest reduction for most OM process error types for both parameters, but percent reductions were less than 5.6\% except for R+Sel OMs where the percent reductions were 11.37\% and 7.97\% for the $a$ and $b$ parameters, respectively and for just the $b$ parameter for R+q OMs (10\%).
+The EM process error assumption provided similar reductions in deviance for both parameters for R OMs.
+Including interactions also did not produce important reductions in deviance (increase of .
 
-### Full results {-}
-
-#### Spawning stock biomass {-}
-
-For R OMs ($\sigma_{2+} = 0$), there was no indication of bias (95\% confidence intervals included 0) in terminal year SSB for any of the estimating models regardless of process error assumptions, except when no SR assumption was made, recruitment variability was low, and there was contrast in fishing mortality and high observation error (Figure \ref{SSB_rel_error}, A).
-However, errors in terminal SSB estimates were highly variable when median natural mortality was estimated and there was constant fishing pressure and high observation error (Figure \ref{SSB_rel_error}, A, second row).
-
-For R+S OMs, the EMs with matching process error assumptions generally produced unbiased estimation of terminal SSB except when median natural mortality was estimated and there was high observation error.
-In R+S OMs with low observation error, EMs with incorrect process error assumptions typically provided biased estimation of terminal year SSB.
-Estimating the Beverton-Holt SRR had little discernible effect on bias of terminal year SSB estimation whereas estimating median M tended to produce more variability in errors in terminal SSB estimation similar to R OMs.
-
-For R+ M OMs with low variability in natural mortality process errors, low observation error and contrast in fishing motality over time all EMs produced low variability in SSB estimation error that indicated unbiasedness (Figure \ref{SSB_rel_error}, B, third row).
-However, larger variability in natural mortality process errors increased bias of EMs without the correct process error type.
-Estimating median natural mortality increased variability of SSB estimation error particularly for OMs with high observation error and constant fishing pressure over time.
-It also increased bias in SSB estimation for many R+M OMs.
-Like R and R+S OMs, estimating a SRR had little discernible effect on SSB bias.
-
-For R+Sel OMs, there was no evidence of bias for any EMs when variability in selectivity process error and observation error was low, and with contrast in fishing mortality (Figure \ref{SSB_rel_error}, C).
-The largest bias occurred for any EMs that estimated median natural mortality when the OMs had high observation error, constant fishing pressure, and greater variability in selectivity process errors ($\sigma_{\text{Sel}} = 0.5$) or low selectivity process errors ($\sigma_{\text{Sel}} = 0.1$) and low observation error.
-However, there was no evidence of SSB bias for correctly specified R+Sel EMs when observation error was low and variation in selectivity process errors was larger, whether median natural mortality was estimated or not (Figure \ref{SSB_rel_error}, C, third row).
-We only observed an effect of estimating the Beverton-Holt SRR for R+Sel OMs that had high observation error and contrast in fishing pressure where estimating the SRR produced less biased SSB estimation for many EMs (Figure \ref{SSB_rel_error}, C, top row).
-
-All EMs fit to data from R+q OMs with low observation error and  contrast in fishing pressure exhibited little evidence of bias in terminal SSB estimation except for R+M EMs when there was no AR1 correlation in catchability process errors (Figure \ref{SSB_rel_error}, D).
-Many EMs also performed well in R+q OMs with low observation error, but no contrast in fishing pressure.
-For R+q OMs with high observation error and contrast in fishing pressure, EMs that estimated the Beverton-Holt SRR exhibited less SSB bias than those that did not.
-Estimating median natural mortality in the EMs only resulted in much more variable SSB estimation errors when there was no contrast in fishing pressure (Figure \ref{SSB_rel_error}, D, first and third rows).
-
-For all OM process error types, relative errors in terminal year recruitment were generally more variable than SSB, but effects of R and R+S OM and EM attributes on bias (i.e, negative or positive or none) were similar (Figure \ref{R_rel_error}, A).
-Furthermore, for EM configurations where bias in terminal SSB was evident, median relative errors in recruitment often indicated stronger bias in recruitment of the same sign.
-
-### Beverton-Holt parameters{-}
-
-Across all OMs, there was generally less bias and(or) lower variability in estimation of the Beverton Holt $a$ parameter than the $b$ parameter.
-In R and R+S OMs, EMs with the correct assumptions about process errors provided the least biased estimation of Beverton-Holt SRR parameters when there was a change in fishing pressure over time and lower variability of recruitment process errors, but there was little effect of estimating median natural mortality and a small increase in bias for those OMs that had high observation error (Figure \ref{SR_rel_error}, A).
-For other R and R+S OMs, estimating natural mortality often resulted in less biased estimation of SRR parameters.
-There was generally large variability in relative errors of the SRR parameter estimates, but the lowest variability occurred with low variability in recruitment and little or no variability in survival process errors ($\sigma_{2+} \in \{0,0.25\}$), and contrast in fishing pressure.
-
-In R+M OMs, the most accurate estimation of SRR parameters for all EM process error assumptions occurred when there was a change in fishing pressure, greater variability in natural mortality process errors, and lower observation error (Figure \ref{SR_rel_error}, B).
-Relative to the R, and R+S OMs, there was even less effect of estimating median natural mortality on estimation bias for the SRR parameters.
-
-Bias for SRR parameters was large and variability in relative errors was greatest for most EMs fit to R+Sel OMs with constant fishing pressure (Figure \ref{SR_rel_error}, C).
-Less bias in parameter estimation occurred for OMs with a change in fishing pressure and the best accuracy occurred for those OMs that had low observation error and more variable and uncorrelated selectivity process errors, and when the EMs had with the correct process error assumption.
-There was little effect of estimating natural mortality on relative errors for SRR parameters.
-
-Like R+Sel OMs, relative errors in SRR parameters for R+q OMs were more accurate for most EM process error types when OMs had contrast in fishing pressure and lower observation error (Figure \ref{SR_rel_error}, D).
-However, the best accuracy occurred for those OMs that had lower variability in catchability process errors.
-The worst accuracy of SRR parameter estimation regardless of EM type occurred when R+q OMs had low observation error and constant fishing pressure (Figure \ref{SR_rel_error}, D, fourth row).
+For regression trees of log-scale error in Beverton-Holt $a$ and $b$ parameters, lower bias was indicated with contrast in OM fishing pressure for all branches in trees for each OM process error type (Figures \ref{SR_a_bias_regtree} and \ref{SR_b_bias_regtree}).
+For all branches based on recruitment variability in trees for R and R+S OMs, lower bias in both $a$ and $b$ was observed with less recruitment variability.
+For R OMs with contrast in fishing pressure and greater recruitment variability EMs that assumed the incorrect R+M process errors produced lower bias in both $a$ and $b$ than other process error assumptions.
+Across all OMs, there was generally less bias and(or) lower variability in estimation of the $a$ parameter than the $b$ parameter (Figure \ref{SR_rel_error}).
 
@@
 
-Across all OMs and EMs there was little effect of estimating SRRs on the bias in estimation of median natural mortality (Figure \ref{M_rel_error}).
-Median natural mortality rate was estimated accurately by all EM process error types for all R OMs except those with high observation error and constant fishing pressure, in which case relative errors were high (Figure \ref{M_rel_error}, A, $\sigma_{2+} = 0$).
-For R+S OMs estimation of median natural mortality rate was most accurate when observation error was low and there was contrast in fishing pressure and the EM process error type was correct.
+Fitted regression models for log-scale errors in median natural mortality rate showed largest percent reductions in residual deviance for R+S and R_M models (Table \ref{bias_median_M_PRD_table}).
+The largest reductions for a single attribute was the EM process error assumption (>20\%) and fishing history (>15\%) for R+S OMs.
+Fishing history also provided >10\% reduction for R+M OMs, but reductions for all factors in R, R+Sel, and R+q OMs were relatively low (<6\%).
+Interactions of OM and EM factors also provided substantial further reductions for R+S and R+M OMs (between 8 and 15 \% for second order interactions).
 
-For R+M OMs, median natural mortality was estimated most accurately, regardless of EM process error type, when OMs had a change in fishing pressure and low observation error (Figure \ref{M_rel_error}, B).
-However, those R+M OMs that also had greatest variability in AR1 correlated natural mortality process errors only had unbiased estimation when the EM process error type was correct.
-
-All EM process error types accurately estimated median natural mortality rate for R+Sel OMs that had contrast in fishing pressure, low observation error, and low selectivity process error variability (Figure \ref{M_rel_error}, C).
-When selectivity process error variability increased, the incorrect EM process errors produce more biased estimation of median natural mortality rate.
-The least accurate estimation occurred for all EM process error types when observation error was high and fishing pressure was constant.
-
-Like R+Sel OMs, all EM process error types produced accurate estimation of median natural mortality rate when fit to R+q OMs with contrast in fishing pressure, low observation error and low catchability process error variability (Figure \ref{M_rel_error}, D).
-Most EM process error types produced biased estimation of median natural  mortality when R+q OMs had high observaiton error and constant fishing pressure.
+Regression trees with branches based on fishing history showed lower bias in median natural mortality rate with contrast in fishing pressure and branches based on level of observation error showed lower bias with more precise observations (Figure \ref{med_M_bias_regtree}).
+For R OMs, branches based on EM process error assumption showed lower bias with EMs assuming the correct R and the incorrect R+S assumption.
+For R+S and R+M OMs, branches based on EM process error showed only the correct EM process error assumption with less bias.
 
@@
 
-### Deviance reduction and regression trees {-}
+Regression models for Mohn's $\rho$ of SSB showed little reduction in deviance for any of the OM an EM attributes (<2\%; Table \ref{mohns_rho_SSB_PRD_table}).
+The lack of explanatory power is also reflected in the regression trees where median Mohn's $\rho$ values are near zero unless a large combinations of OM and EM conditions occur (Figure \ref{SSB_mohns_rho_regtree}).
+For example, In R+S OMs, with constant fishing pressure, high observation error, and higher apparent survival process error, EMs that assume R+M process errors have a median Mohn's $\rho = -0.068$.
 
-Fits of glms showed little explanatory power of any of the OM an EM attributes for Mohn's rho for SSB, F, and R.
-Note that the factors that showed the largest reductions in deviance (which were very small) often do not comport with factors that show differences in median Mohn's rho results.
-This is because the glms and regression trees are reducing residual variance and there could be large variability in values, but the medians are less sensitive to extreme values.
-For example we can see that there is large Mohn's rho for R in R OMs with larger variability assumed in recruitment and higher observation error (Figure X), but these factors suggest little reduction in residual deviance (Table X).
-
-### Full results {-}
-
-Mohn's $\rho$ for SSB was small in absolute value for all R and R+S OMs, regardless of EM process error types, and whether median natural mortality rate or SRRs were estimated (Figure \ref{mohns_rho_ssb}, A).
-The strongest retrospective patterns (highest absolute Mohn’s $\rho$ values) occurred in OMs with the largest apparent survival process error variability, high observation error, and contrast in fishing pressure, but only for EMs with the incorrect process error type and where median natural mortality rate was assumed known (median $\rho$ was approximately -0.15).
-For R+M, R+Sel, and R+q OMs, Mohn's $\rho$ was also small in absolute value, but median values were all closer to 0 than the largest values in the R and R+S OMs (Figure \ref{mohns_rho_ssb},B-D).
-For these OMs, there was no noticeable effect of estimation of median natural mortality rate or SRRs on Mohn's $\rho$ for any EM process error types.
-
-Mohn's $\rho$ for recruitment was small in absolute value for all R OMs with low variability in recruitment process errors, regardless of EM process error type, and whether median natural mortality rate or SRRs were estimated (Figure \ref{mohns_rho_R}, A).
-However, R and R+S OMs with greater recruitment process variability and higher observation error had median Mohn's $\rho$ for recruitment greater than zero for most EMs even when the EM process error type was correct.
-In R+S OMs with lower observation error, EMs with the correct process error type exhibited better median Mohn's $\rho$ close to 0 than EMs with the incorrect process error type.
-For R+M, R+Sel, and R+q OMs, results for Mohn's $\rho$ for recruitment are similar to those for SSB, but the range in median values and variation in Mohn's $\rho$ values for a given OM are generally larger for recruitment (Figure \ref{mohns_rho_R}, B-D).
+Similarly poor explanatory power of the OM and EM attributes occured when we fit regression models for Mohn's $\rho$ of fishing mortality and recruitment (Tables \ref{mohns_rho_F_PRD_table} and \ref{mohns_rho_R_PRD_table}).
+Regression trees for Mohn's $\rho$ of fishing mortality were similar to those for SSB in that median values of Mohn's $\rho$ were close to zero for most combinations of OM and EM attributes (Figure \ref{F_mohns_rho_regtree}).
+However, we observed median Mohn's $\rho$ for recruitment greater than 0.1 at  branches much closer to the base of the trees with fewer interactions of the OM and EM attributes (Figure \ref{R_mohns_rho_regtree}).
+These branches with consistently large retrospective patterns were typically defined by larger OM observation error, OM constant fishing pressure, or incorrect EM process error configuration.
+Comparing regression model and regression tree fits, attributes defining the primary branches for all regression trees of all Mohn's $\rho$ values (SSB, fishing mortality, and recruitment) generally matched those that explained the largest reductions in deviance.
 
@@
 However, there is no accepted standard for the gradient threshold [e.g., @leeetal11; @hurtadoferroetal15;@ruddthorson18] and some thresholds would exclude models that in fact have an invertible Hessian.
-We found the Hessian at the optimized log-likelihood can often be invertible when the maximum absolute gradient was much larger than what would perceived to be a sensible threshold.
+We found the Hessian at the optimized log-likelihood can often be invertible when the maximum absolute gradient was much larger than what might be perceived to be a sensible threshold.
 
@@
 Furthermore, numerical methods are used to calculate and invert the Hessian for variance estimation for models with random effects.
-Along with our results, the potential lack of accuracy imposed by these approximations, suggests at least investigating whether the Hessian is positive definite when the calculated absolute gradients are not terribly large (e.g, $< 1$).
+Our results, along with the potential lack of accuracy imposed by these approximations, suggest at least investigating whether the Hessian is positive definite when the calculated absolute gradients are not terribly large (e.g, $< 1$).
 
@@
 In our simulations, the process errors changing over time may have trends in particular simulations, particularly when strong autocorrelation is imposed, but the random effects have no trend on average across simulations.
-@szuwalskietal18 and lietal24 also found relatively small retrospective patterns when the source of mis-specification was temporal variation in demography attributes.
+@szuwalskietal18 and @lietal24 also found relatively small retrospective patterns when the source of mis-specification was temporal variation in demography attributes.
 Indeed, it is common for the flexibility provided by temporal random effects to reduce retrospective patterns [@milleretal18; @stockmiller21; @stocketal21], though it does not necessarily indicate a more accurate assessment model [@perrettietal20; @lietal24; @liljestrandetal24].
 Our results together with the existing literature seem to suggest that when a strong retrospective pattern is observed in an assessment it is more likely to be due to a mis-specification of a rapid shift in some model attribute rather than whether a particular process is assumed to be randomly varying temporally.
+
+## Analytical approach {-}
+
+We found the use of regression models and classification and regression trees extremely useful in understanding the most important OM and EM attributes explaining variation in the measures of reliability we examined across all simulations.
+The classification and regression trees are generally a good tool for determining the values of the OM and EM attributes that produce better or worse measures of reliability.
+However, determining the combination of attributes that produce the best or worst measures of reliability can be challenging using the trees.
+For example, in the regression tree for median natural mortality rate estimates in R OMs (Figure \ref{median_M_bias_regtree}),  the both of the first branches imply bias is low regardless of OM fishing history, but when OM fishing pressure is constant, results are much better when OM observation error is low (median RE about -6\%) than when OM observation error is high (median RE about 40\%).
+Therefore, inspection of results by all combinations of OM and EM factors (as provided in the supplementary materials) can be important.
 
@@
 Our simulation study examined the importance of several factors for reliable inferences from state-space age-structured assessment models.
-Contrast in fishing pressure was consistently an important factor across all measures of reliability we examined.
+Contrast in fishing pressure and lower observation error consistently improved all measures of reliability we examined.
 AIC accurately distinguished models with process errors on recruitment only (R) or on recruitment and apparent survival (R+S).
-Accuracy for other process error types required a strong signal (high process variability) with low noise (low observation uncertainty).
 Therefore, we expect practitioners will find R+S configurations to provide satisfactory diagnostics across a range of life history and data quality scenarios.
-AIC generally performed poorly for selecting the SRR, but performance was improved with low recruitment variability and contrast in fishing pressure.
-Some bias in estimation in at least one of the SRR parameters existed in nearly all OM-EM combinations.
-Because bias in terminal SSB and retrospective patterns were indifferent to whether or not the SRR was estimated, and convergence was slightly better without the SRR, a sensible default would be to fit models without an assumed SRR.
+Accurate AIC selection of the Beverton-Holt SRR generally required a combination of low recruitment variability, contrast in fishing pressure and large variation in SSB over time.
+However, when the Beverton-Holt was correctly assumed, some bias in estimation in at least one of the parameters existed regardless of any of the other OM and EM configurations.
+Because bias in terminal SSB and retrospective patterns were indifferent to whether or not the SRR was estimated, the prevalence of bias in SRR parameter estimation, and often better convergence without the SRR, a sensible default would be to not assume a SRR when fitting assessment models.
 
@@
 
-This work was funded by NOAA Fisheries Northeast Fisheries Science Center. We thank Jon Deroba for helpful comments on an earlier version of this manuscript.
+This work was funded by NOAA Fisheries Northeast Fisheries Science Center.
+We thank Jon Deroba, 2 anonymous reviewers, and the associate editor for helpful comments on earlier versions of this manuscript that markedly improved its clarity.
 
@@
 \begin{center}
-\includegraphics{convergence_classification_plots}
+\includegraphics[width = 1.4\textwidth]{convergence_classification_plots}
 \end{center}
 \caption{Classification trees indicating primary factors determining convergence as defined by providing hessian-based standard errors for R, R+S, R+M, R+Sel and R+q OMs.
-Lower or higher convergence rates are indicated by red or green polygons, respectively}\label{conv_class}
+Lower or higher convergence rates are indicated by more red or green polygons, respectively}\label{conv_class}
 \end{figure}
@@
 \begin{center}
-\includegraphics{AIC_PE_classification_plots}
+\includegraphics[width = 1.4\textwidth]{AIC_PE_classification_plots}
 \end{center}
@@
 Each node shows the proportion of EM process error models with lowest AIC (top) and number of observations (bottom) for the corresponding subset.
-Lower or higher accuracy of the process error assumption are indicated by red or green polygons, respectively.}\label{AIC_PE_class}
+Lower or higher accuracy of the process error assumption are indicated by more red or green polygons, respectively.}\label{AIC_PE_class}
 \end{figure}
@@
 \begin{center}
-\includegraphics{AIC_SRR_classification_plots}
+\includegraphics[width = 1.4\textwidth]{AIC_SRR_classification_plots}
 \end{center}
-\caption{Classification trees indicating primary factors determining which EM stock recruitment assumption (none or Beverton-Holt) provides the lowest AIC for R, R+S, R+M, R+Sel and R+q OMs.
-Each node shows the proportion of EMs that assume the stock-recruit relationship with lowest AIC (top) and number of observations (bottom) for the corresponding subset.
+\caption{Classification trees indicating primary factors determining which EM SRR assumption (none or Beverton-Holt) provides the lowest AIC for R, R+S, R+M, R+Sel and R+q OMs.
+Each node shows the proportion of EMs that assume the SRR with lowest AIC (top) and number of observations (bottom) for the corresponding subset.
 Lower or higher accuracy of the process error assumption are indicated by more red or green polygons, respectively.}\label{AIC_SRR_class}
@@
 \begin{center}
-\includegraphics{SR_a_bias_regtree_plots}
+\includegraphics[width = 1.4\textwidth]{SSB_bias_regtree_plots}
 \end{center}
 \caption{Regression trees indicating primary factors determining reductions in sums of squares of errors in estimation measured by Eq.
-\ref{bias_regression_response} for the Beverton-Holt stock-recruit parameter $a$ for R+S, R+M, R+Sel and R+q OMs.
+\ref{bias_regression_response} for terminal year SSB for R+S, R+M, R+Sel and R+q OMs.
+Each node shows the median error (top) and number of observations (bottom) for the corresponding subset.
+Median errors closer to or further from zero are indicated by more green or red polygons, respectively.}\label{SSB_bias_regtree}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[width = 1.4\textwidth]{SR_a_bias_regtree_plots}
+\end{center}
+\caption{Regression trees indicating primary factors determining reductions in sums of squares of errors in estimation measured by Eq.
+\ref{bias_regression_response} for the Beverton-Holt SRR parameter $a$ for R+S, R+M, R+Sel and R+q OMs.
 Each node shows the median error (top) and number of observations (bottom) for the corresponding subset.
@@
 \begin{center}
-\includegraphics{SR_b_bias_regtree_plots}
+\includegraphics[width = 1.4\textwidth]{SR_b_bias_regtree_plots}
 \end{center}
 \caption{Regression trees indicating primary factors determining reductions in sums of squares of errors in estimation measured by Eq.
-\ref{bias_regression_response} for the Beverton-Holt stock-recruit parameter $b$ for R+S, R+M, R+Sel and R+q OMs.
+\ref{bias_regression_response} for the Beverton-Holt SRR parameter $b$ for R+S, R+M, R+Sel and R+q OMs.
 Each node shows the median error (top) and number of observations (bottom) for the corresponding subset.
@@
 \begin{center}
-\includegraphics{med_M_bias_regtree_plots}
+\includegraphics[width = 1.4\textwidth]{med_M_bias_regtree_plots}
 \end{center}
@@
 \begin{center}
-\includegraphics{SSB_bias_regtree_plots}
-\end{center}
-\caption{Regression trees indicating primary factors determining reductions in sums of squares of errors in estimation measured by Eq.
-\ref{bias_regression_response} for terminal year spawning stock biomass for R+S, R+M, R+Sel and R+q OMs.
-Each node shows the median error (top) and number of observations (bottom) for the corresponding subset.
-Median errors closer to or further from zero are indicated by more green or red polygons, respectively.}\label{SSB_bias_regtree}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{SSB_mohns_rho_regtree_plots}
+\includegraphics[width = 1.4\textwidth]{SSB_mohns_rho_regtree_plots}
 \end{center}
 \caption{Regression trees indicating primary factors determining reductions in sums of squares of errors in transformed Mohn's $\rho$ (Eq.
-\ref{bias_regression_response}) for spawning stock biomass for R+S, R+M, R+Sel and R+q OMs.
+\ref{bias_regression_response}) for SSB for R+S, R+M, R+Sel and R+q OMs.
 Each node shows the median Mohn's $\rho$ (top) and number of observations (bottom) for the corresponding subset.
@@
 \begin{table}
-\caption{For each OM process error type (columns), percent reduction in deviance for logistic regression models fit to indicators of convergence with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{convergence_PRD_table}
+\caption{For each OM process error type (columns), percent reduction in deviance for logistic regression models fit to indicators of convergence (providing hessian-based standard errors) with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{convergence_PRD_table}
 {\input{convergence_PRD_table}}
@@
 \begin{table}
-\caption{For each OM process error type (columns), percent reduction in deviance for logistic regression models fit to indicators of EM stock-recruit assumption (none or Beverton-Holt) with lowest AIC with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{AIC_SRR_PRD_table}
+\caption{For each OM process error type (columns), percent reduction in deviance for logistic regression models fit to indicators of EM SRR assumption (none or Beverton-Holt) with lowest AIC with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{AIC_SRR_PRD_table}
 {\input{AIC_SRR_PRD_table}}
+\end{table}
+
+\begin{table}
+\caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
+\ref{bias_regression_response} for the terminal year SSB with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_SSB_PRD_table}
+{\input{bias_SSB_PRD_table}}
 \end{table}
@@
 \caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
-\ref{bias_regression_response} for the Beverton-Holt stock recruit relation ship parameters with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_SR_pars_PRD_table}
+\ref{bias_regression_response} for the Beverton-Holt SRR parameters with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_SR_pars_PRD_table}
 {\input{bias_SR_pars_PRD_table}}
@@
 \begin{table}
-\caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
-\ref{bias_regression_response} for the terminal year spawning biomass with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_SSB_PRD_table}
-{\input{bias_SSB_PRD_table}}
-\end{table}
-
-\begin{table}
 \caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to transformed Mohn's $\rho$ values for each simulation (Eq.
-\ref{bias_regression_response}) for spawning biomass with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{mohns_rho_SSB_PRD_table}
+\ref{bias_regression_response}) for SSB with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{mohns_rho_SSB_PRD_table}
 {\input{mohns_rho_SSB_PRD_table}}
@@
 
+\begin{landscape}
+\end{landscape}
 \pagebreak
@@
 
+## Referenced Figures {-}
+
 \begin{figure}[!ht]
@@
 \end{center}
-\caption{The proportion mature at age, weight at age, fleet and index selectivity at age, and Beverton-Holt stock-recruit relationship assumed for the population in all operating models.
+\caption{The proportion mature at age, weight at age, fleet and index selectivity at age, and Beverton-Holt SRR assumed for the population in all operating models.
 For operating models with random effects on fleet selectivity, this represents the selectivity at the mean of the random effects.}\label{om_inputs_fig}
@@
 \begin{center}
-\includegraphics{type_4_convergence_plots}
+\includegraphics[width = 1.4\textwidth]{convergence_gradient_classification_plots}
 \end{center}
-\caption{Estimated probability of fits providing hessian-based standard errors for EMs assuming alternative process error (colored points and lines), and median natural mortality (estimated or known) and Beverton-Holt stock-recruit relationships (estimated or not; along x-axis) when fitted to operating models that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{hessian_SE_convergence}
-\end{figure}
-\end{landscape}
-
-\clearpage
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{type_3_convergence_plots}
-\end{center}
-\caption{Probability of estimating models providing maximum absolute values of gradients less than $10^{-6}$ assuming alternative process error (colored points and lines), and median natural mortality (estimated or known) and Beverton-Holt stock-recruit relationships (estimated or not; along x-axis) when fitted to operating models that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{gradient_convergence}
+\caption{Classification trees indicating primary factors determining convergence as defined by a maximum absolute gradient < $10^{-6}$ for R, R+S, R+M, R+Sel and R+q OMs.
+Lower or higher convergence rates are indicated by more red or green polygons, respectively}\label{conv_gradient_class}
 \end{figure}
@@
 \begin{center}
-\includegraphics{hess_grad_convergence_plots}
+\includegraphics[width = 1.4\textwidth]{hess_grad_convergence_plots}
 \end{center}
 \caption{The maximum of the absolute values of all gradient values for all fits that provided hessian-based standard errors across all simuated data sets of a given OM configuration (A: R and R+S, B: R+M, C: R+Sel, or D: R+q).
-Results are conditional on EM fits with alternative process error type (colored points and lines), median natural mortality (estimated or known) and recruitment assumptions (Beverton-Holt stock-recruit relationship or not).
+Results are conditional on EM fits with alternative process error type (colored points and lines), median natural mortality (estimated or known) and recruitment assumptions (Beverton-Holt SRR or not).
 Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{hess_grad}
@@
 \begin{center}
-\includegraphics{pe_aic_plots}
+\includegraphics[width = 1.4\textwidth]{pe_aic_plots}
 \end{center}
-\caption{Estimated probability of lowest AIC for EMs assuming alternative process error structures (colored bars) conditional on alternative assumptions for median natural mortality (estimated or known) and Beverton-Holt stock-recruit relationships (estimated or not; along x-axis) when fitted to operating models that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
+\caption{Estimated probability of lowest AIC for EMs assuming alternative process error structures (colored bars) conditional on alternative assumptions for median natural mortality (estimated or known) and Beverton-Holt SRR (estimated or not; along x-axis) when fitted to operating models that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
 Striped bars indicate results where the EM process error structure matches that of the operating model.}\label{pe_aic}
-\end{figure}
-\end{landscape}
-
-\clearpage
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics[height = 0.8\textheight]{sr_aic_plots_rev}
-\end{center}
-\caption{Estimated probability of lowest AIC from logistic regression on the log-standard deviation of the true log(SSB) in each simulation for estimating model with Beverton-Holt stock-recruit relationships, rather than the otherwise equivalent EM without the stock-recruit relationship.
-Results are conditional on median M is known in the EM and alternative assumptions EMs having the correct process error structure: R and R+S (A), R+Sel (B), R+M (C), or R+q (D), and median M is assumed known in the EM.
-Solid and dashed lines are for OMs with and without temporal contrast in fishing pressure, respectively, and polygons represent 95\% confidence intervals.
-Range of results indicates the range of log-standard deviation of log(SSB) for simulations of the particular OM.}\label{sr_aic}
 \end{figure}
@@
 \begin{center}
-\includegraphics{sr_aic_plots}
+\includegraphics[width = 1.4\textwidth]{F_bias_regtree_plots}
 \end{center}
-\caption{Estimated probability of lowest AIC from logistic regression on the log-standard deviation of the true log(SSB) in each simulation for estimating model with Beverton-Holt stock-recruit relationships, rather than the otherwise equivalent EM without the stock-recruit relationship.
-Results are conditional on alternative assumptions for median natural mortality (estimated or known) and on EMs having the correct process error structure: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Rug along x-axis denotes $SD(\log(SSB))$ values for each simulation and polygons represent 95\% confidence intervals.}\label{sr_aic_supp}
-\end{figure}
-\end{landscape}
-
-\clearpage
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{term_SSB_bias_plots}
-\end{center}
-\caption{Median relative error of terminal year SSB for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{SSB_rel_error}
+\caption{Regression trees indicating primary factors determining reductions in sums of squares of errors in estimation measured by Eq.
+\ref{bias_regression_response} for terminal year fully-selected fishing mortality for R+S, R+M, R+Sel and R+q OMs.
+Each node shows the median error (top) and number of observations (bottom) for the corresponding subset.
+Median errors closer to or further from zero are indicated by more green or red polygons, respectively.}\label{F_bias_regtree}
 \end{figure}
@@
 \begin{center}
-\includegraphics{term_F_bias_plots}
+\includegraphics[width = 1.4\textwidth]{R_bias_regtree_plots}
 \end{center}
-\caption{Median relative error of terminal year fully-selected fishing mortality for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{F_rel_error}
+\caption{Regression trees indicating primary factors determining reductions in sums of squares of errors in estimation measured by Eq.
+\ref{bias_regression_response} for terminal year recruitment for R+S, R+M, R+Sel and R+q OMs.
+Each node shows the median error (top) and number of observations (bottom) for the corresponding subset.
+Median errors closer to or further from zero are indicated by more green or red polygons, respectively.}\label{R_bias_regtree}
 \end{figure}
@@
 \begin{center}
-\includegraphics{term_R_bias_plots}
+\includegraphics[width = 1.4\textwidth]{sr_bias_plots}
 \end{center}
-\caption{Median relative error of terminal year recruitment for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{R_rel_error}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{sr_bias_plots}
-\end{center}
-\caption{Median relative error of Beverton-Holt stock-recruit parameters ($a$ and $b$) for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+\caption{Median relative error of Beverton-Holt SRR parameters ($a$ and $b$) for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
 Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{SR_rel_error}
@@
 \begin{center}
-\includegraphics{M_bias_plots}
+\includegraphics[width = 1.4\textwidth]{F_mohns_rho_regtree_plots}
 \end{center}
-\caption{Median relative error of median natural mortality for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{M_rel_error}
+\caption{Regression trees indicating primary factors determining reductions in sums of squares of errors in transformed Mohn's $\rho$ (Eq.
+\ref{bias_regression_response}) for fishing mortality averaged over all age classes for R+S, R+M, R+Sel and R+q OMs.
+Each node shows the median Mohn's $\rho$ (top) and number of observations (bottom) for the corresponding subset.
+Median Mohn's $\rho$ closer to or further from zero are indicated by more green or red polygons, respectively.}\label{F_mohns_rho_regtree}
 \end{figure}
@@
 \begin{center}
-\includegraphics{mohns_rho_ssb_plots}
+\includegraphics[width = 1.4\textwidth]{R_mohns_rho_regtree_plots}
 \end{center}
-\caption{Median Mohn's $\rho$ for SSB for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_ssb}
+\caption{Regression trees indicating primary factors determining reductions in sums of squares of errors in transformed Mohn's $\rho$ (Eq.
+\ref{bias_regression_response}) for recruitment for R+S, R+M, R+Sel and R+q OMs.
+Each node shows the median Mohn's $\rho$ (top) and number of observations (bottom) for the corresponding subset.
+Median Mohn's $\rho$ closer to or further from zero are indicated by more green or red polygons, respectively.}\label{R_mohns_rho_regtree}
 \end{figure}
@@
 
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{mohns_rho_F_plots}
-\end{center}
-\caption{Median Mohn's $\rho$ of fishing mortality averaged over all age classes for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_F}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{mohns_rho_R_plots}
-\end{center}
-\caption{Median Mohn's $\rho$ of recruitment for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_R}
-\end{figure}
-\end{landscape}
+## Referenced Tables {-}
 
@@
 \begin{table}
+\caption{For each OM process error type (columns), percent reduction in deviance for logistic regression models fit to indicators of convergence (maximum absolute gradient < $10^{-6}$) with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{convergence_gradient_PRD_table}
+{\input{convergence_gradient_PRD_table}}
+\end{table}
+
+\begin{table}
+\caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
+\ref{bias_regression_response} for the terminal year fully-selected fishing mortality with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_F_PRD_table}
+{\input{bias_F_PRD_table}}
+\end{table}
+
+\begin{table}
+\caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
+\ref{bias_regression_response} for the terminal year recruitment with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_R_PRD_table}
+{\input{bias_R_PRD_table}}
+\end{table}
+
+\clearpage
+
+## Further Detailed Results {-}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[width = 1.4\textwidth]{term_SSB_bias_plots}
+\end{center}
+\caption{Median relative error of terminal year SSB for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{SSB_rel_error}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[width = 1.4\textwidth]{term_F_bias_plots}
+\end{center}
+\caption{Median relative error of terminal year fully-selected fishing mortality for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{F_rel_error}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[width = 1.4\textwidth]{term_R_bias_plots}
+\end{center}
+\caption{Median relative error of terminal year recruitment for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{R_rel_error}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[width = 1.4\textwidth]{M_bias_plots}
+\end{center}
+\caption{Median relative error of median natural mortality for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{M_rel_error}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[width = 1.4\textwidth]{mohns_rho_ssb_plots}
+\end{center}
+\caption{Median Mohn's $\rho$ for SSB for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_ssb}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[width = 1.4\textwidth]{mohns_rho_F_plots}
+\end{center}
+\caption{Median Mohn's $\rho$ of fishing mortality averaged over all age classes for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_F}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[width = 1.4\textwidth]{mohns_rho_R_plots}
+\end{center}
+\caption{Median Mohn's $\rho$ of recruitment for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_R}
+\end{figure}
+\end{landscape}
+
+\begin{table}
 \caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to transformed Mohn's $\rho$ values for each simulation (Eq.
@@
 \end{table}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[width = 1.4\textwidth]{type_4_convergence_plots}
+\end{center}
+\caption{Probability of EMs providing hessian-based standard errors with alternative process error (colored points and lines), and median natural mortality (estimated or known) and Beverton-Holt SRR (estimated or not; along x-axis) assumptions when fitted to OMs that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{hessian_SE_convergence}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[width = 1.4\textwidth]{type_3_convergence_plots}
+\end{center}
+\caption{Probability of EMs providing maximum absolute values of gradients less than $10^{-6}$ with alternative process error (colored points and lines), and median natural mortality (estimated or known) and Beverton-Holt SRR (estimated or not; along x-axis) assumptions when fitted to OMs that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{gradient_convergence}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = 0.8\textheight]{sr_aic_plots_rev}
+\end{center}
+\caption{Probability of lowest AIC from logistic regression on the log-standard deviation of the true log(SSB) in each simulation for estimating model with Beverton-Holt SRRs, rather than the otherwise equivalent EM without the SRR.
+Results are conditional on median M is known in the EM and alternative assumptions EMs having the correct process error structure: R and R+S (A), R+Sel (B), R+M (C), or R+q (D), and median M is assumed known in the EM.
+Solid and dashed lines are for OMs with and without temporal contrast in fishing pressure, respectively, and polygons represent 95\% confidence intervals.
+Range of results indicates the range of log-standard deviation of log(SSB) for simulations of the particular OM.}\label{sr_aic}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[width = 1.4\textwidth]{sr_aic_plots}
+\end{center}
+\caption{Estimated probability of lowest AIC from logistic regression on the log-standard deviation of the true log(SSB) in each simulation for estimating model with Beverton-Holt SRRs, rather than the otherwise equivalent EM without the SRR.
+Results are conditional on alternative assumptions for median natural mortality (estimated or known) and on EMs having the correct process error structure: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Rug along x-axis denotes $SD(\log(SSB))$ values for each simulation and polygons represent 95\% confidence intervals.}\label{sr_aic_supp}
+\end{figure}
+\end{landscape}

### 2025-10-01 Project_0/manuscript/manuscript_R1.tex
saving work on project 0 revisons

@@
 \PassOptionsToPackage{hyphens}{url}
-%
 \documentclass[
@@
 ]{article}
+\usepackage{xcolor}
+\usepackage[margin=1in]{geometry}
 \usepackage{amsmath,amssymb}
+\setcounter{secnumdepth}{5}
 \usepackage{iftex}
@@
 \makeatother
-\usepackage{xcolor}
-\usepackage[margin=1in]{geometry}
 \usepackage{graphicx}
 \makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\newsavebox\pandoc@box
+\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
+  \sbox\pandoc@box{#1}%
+  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
+  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
+  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
+  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
+  \else\usebox{\pandoc@box}%
+  \fi%
+}
 % Set default figure placement to htbp
-\makeatletter
 \def\fps@figure{htbp}
 \makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{5}
 % definitions for citeproc citations
@@
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 \usepackage{url}
@@
 \usepackage{xcolor}
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
 \usepackage{bookmark}
@@
 C. Stock\textsuperscript{6} \and John Wiedenmann\textsuperscript{7}}
-\date{10 September, 2025}
+\date{30 September, 2025}
 
@@
 State-space models are increasingly used for stock assessment, and
-evaluations of their statistical reliability and best practices for
-selecting among process error configurations are needed. We simulated 72
-operating models that varied fishing pressure and observation error
-across process errors in recruitment, survival, selectivity,
-catchability, and/or natural mortality. We fit estimating models with
-different assumptions on the process error source and whether median
-natural mortality or a stock-recruit relationship were estimated.
-Estimating models without a stock-recruit relationship that assumed the
-correct process error source and median natural mortality had high
-convergence rates and low bias. Bias was also low under many incorrect
-process error assumptions when there was contrast in fishing pressure
-and low observation error. Marginal AIC most accurately distinguished
-process errors on recruitment, survival, and selectivity, and other
-process error sources when variability was greater. Retrospective
-patterns were generally small but were sizable for recruitment when
-observation error was high. These results help establish the statistical
-reliability of state space assessment models and pave the way for the
-next-generation of fisheries stock assessment.
+evaluations of statistical inferences made from them is needed. We
+simulated 72 operating models that varied fishing pressure and
+observation error across process errors in recruitment, apparent
+survival, selectivity, catchability, and/or natural mortality. We fit
+estimating models with different assumptions on the process error source
+and whether natural mortality and stock-recruit relationship were
+estimated. We measured reliability of estimating models by probability
+of convergence, accuracy of AIC for correct process error configuration
+and stock-recruit function, estimation bias, and magnitude of
+retrospective patterns. Across all measures reliability was generally
+better with lower observation error, contrast in fishing pressure over
+time, and median natural mortality rate known, but using marginal AIC to
+distinguishing the stock recruit relationship further required large
+contrast in spawning biomass and low residual variation in recruitment.
+Marginal AIC accurately distinguished process errors on recruitment,
+apparent survival, and selectivity. These results improve our
+understanding of when results from state space assessment models, the
+next-generation of fisheries stock assessment, can be expected to be
+reliable.
 
@@
 work has focused primarily on linear and Gaussian state-space models
-(Aeberhard et al. 2018; Auger-Méthé et al. 2021). Therefore, current
+(Aeberhard et al. 2018; {Auger-Méthé et al.} 2021). Therefore, current
 understanding of the reliability of state-space models does not extend
@@
 misspecification is of great scientific interest and could aid the next
-generation of stock assessments (Auger-Méthé et al. 2021). We evaluate
+generation of stock assessments ({Auger-Méthé et al.} 2021). We evaluate
 whether convergence and Akaike Information Criterion (AIC) can correctly
@@
 well as other process error types without autocorrelation. The maturity
-at age, weight at age for catch and SSB, and observation error variance
-of aggregate catch and indices were all assumed known at the true
-values. However, the variance parameters for the logistic-normal
-distributions for age composition observations were estimated in the
-EMs. As such, EMs would either be configured completely correctly for
-the OM, or there could be mis-specification in assumptions of process
-error autocorrelation, the type of process error, or the SRR
+at age, weight at age for catch and spawning stock biomass (SSB), and
+observation error variance of aggregate catch and indices were all
+assumed known at the true values. However, the variance parameters for
+the logistic-normal distributions for age composition observations were
+estimated in the EMs. As such, EMs would either be configured completely
+correctly for the OM, or there could be mis-specification in assumptions
+of process error autocorrelation, the type of process error, or the SRR
 (Beverton-Holt or none).
@@
 uncertainty when using assessment models in management, we estimated
-probablity of convergence as measured by occurrence of a
+probability of convergence as measured by occurrence of a
 positive-definite hessian matrix at the optimized negative
-log-likelihood that could be inverted. We also provide results in the
-Supplementary Materials for the maximum of the absolute values among all
-gradients for all fits of a given EM to all simulated data sets from a
-given OM that produced hessian-based standard errors for all estimated
-fixed effects. This provides an indication of how poor the calculated
+log-likelihood that could be inverted (i.e., providing hessian-based
+standard error estimates). We also provide results in the Supplementary
+Materials for convergence defined by the maximum absolute gradient
+\textless{} \(1^{-6}\) and the maximum of the absolute gradient values
+for all fits of a given EM to all simulated data sets from a given OM
+that produced hessian-based standard errors for all estimated fixed
+effects. This provides an indication of how poor the calculated
 gradients can be, but still presumably converged adequately enough for
-parameter inferences. We used the Clopper-Pearson exact method for
-constructing 95\% confidence intervals of the probabilities of
-convergence (Clopper and Pearson 1934; Thulin 2014).
+parameter inferences.
 
@@
 
-We estimated the probability of selection of each process error model
-structure (R, R+S, R+M, R+Sel, R+q) using marginal AIC. For a given
-operating model, we compared AIC for EMs that were all configured the
-same for median natural mortality (known or estimated) and the SRR
-(Beverton-Holt or none).
-
-We also estimated the probability of correctly selecting between EMs
-with Beverton-Holt SRR assumed and models without the SRR (null model).
-We made these comparisons between models that otherwise assumed the same
-process error structure as the operating model and both of the compared
-models either estimate median natural mortality or assume it is known.
+We investigated the reliability of AIC-based model selection for two
+purposes. First, we analyzed selection of each process error model
+structure (R, R+S, R+M, R+Sel, R+q) using marginal AIC. For a given OM
+simulated data set, we compared AIC for EMs with different process error
+assumption conditional on whether median natural mortality rate and the
+Beverton-Holt SRR were estimated. We tabulated the models providing the
+lowest AIC across simulated data sets. Second, we analyzed AIC-based
+selection between EMs with and without the Beverton-Holt SRR assumed.
 Contrast in fishing pressure and time series with recruitment at low
@@
 (Magnusson and Hilborn 2007; Conn et al. 2010). Our preliminary
-inspections of the proportions of simulations where the correct
-recruitment model was chosen for a given set of OM factors (including
-contrast in fishing pressure) indicated generally poor performance of
-AIC. Therefore, we fit logistic regression models to the indicator of
-Beverton-Holt models having lower AIC as a function of the log-standard
-deviation of the true log(SSB) (similar to the log of the coefficient of
-variation for SSB) since simulations with realized SSB producing low and
-high recruitments would have larger variation in realized SSB.
-
-All model selection results condition only on whether all of the
-compared estimating models completed the optimization process without
-failure. We did not condition on convergence as defined by a gradient
-threshold or invertibity of the hessian because optimization could
+inspections indicated generally poor performance of AIC in determining
+the Beverton-Holt model for a given set of OM factors (including
+contrast in fishing pressure). Therefore, we also considered the effect
+of of the log-standard deviation of the true log(SSB) (similar to the
+log of the coefficient of variation for SSB) on model selection since
+simulations with realized SSB producing low and high recruitment would
+have larger variation in realized SSB.
+
+All model selection results condition only on completion of the
+optimization process without failure for all of the compared EMs. We did
+not condition on convergence as defined above because optimization could
 correctly determine an inappropriate process error assumption by
@@
 
-For a given model attribute we calculated the relative error
-\begin{equation}\label{relerror}
+We also investigated bias in estimation of various model attributes as a
+measure of reliability. For a given model attribute we calculated the
+relative error \begin{equation}\label{relerror}
 \text{RE}\left(\theta_i\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
@@
 \(\theta_i\) are the estimated and true values for simulation \(i\). We
-estimated bias as the median of the relative errors across all
-simulations for a given OM and EM combination. We constructed 95\%
-confidence intervals for the median relative bias using the binomial
-distribution approach as in Miller and Hyun (2018) and Stock and Miller
-(2021). We present results for bias in terminal year estimates of SSB
-and recruitment, Beverton-Holt stock recruit parameters (\(a\) and
-\(b\)), and median natural mortality rate. Results for terminal year
-fishing mortality were strongly negatively correlated with those for SSB
-and are provided in the Supplementary Materials.
+analyzed simulation results for estimates of terminal year estimates of
+SSB and recruitment, Beverton-Holt SRR parameters (\(a\) and \(b\)), and
+median natural mortality rate.
 
@@
 
-We calculated Mohn's \(\rho\) for SSB, fishing mortality (averaged over
-all age classes), and recruitment for each EM (Mohn 1999). We fit
-\(P =7\) peels to each simulated data set and calculated Mohn's \(\rho\)
-for a given attribute \(\theta\) as \begin{equation}\label{mohns_rho}
+Finally, we investigated presence of retrospective patterns in fitted
+models as a measure of reliability. We calculated Mohn's \(\rho\) for
+SSB, fishing mortality (averaged over all age classes), and recruitment
+for each EM fit to each OM simulated data set (Mohn 1999). We fit
+\(P = 7\) peels to each simulated data set and calculated Mohn's
+\(\rho\) for a given attribute \(\theta\) as
+\begin{equation}\label{mohns_rho}
 \rho\left(\theta\right) = \frac{1}{P}\sum^P_{p=1} \frac{\widehat \theta_{Y-p,Y-p} - \widehat\theta_{Y-p,Y}}{\widehat\theta_{Y-p,Y}}
@@
 attribute \(\theta\) in year \(i\) from a model fit using data up to
-year \(j\). For each EM we calculated median and 95\% confidence
-intervals using the same methods as that for relative bias.
+year \(j\).
 
@@
 
-The measures of central tendency and variability of of observed values
-for specific OM and EM attributes (e.g., low or high observation error)
-that we described above can indicate scenarios that provide better or
-poorer reliability. However, the OM and EM attributes that we
-investigated are numerous, so we used two methods to summarize the most
-important factors for differences in results. The first method was
-fitting regression models with the response being each of the measures
-of reliability described above and predictor variables were defined
-based on OM and EM characteristics (e.g., MacKinnon et al. 1995; Wang et
-al. 2017; Harwell et al. 2018). For the binary results for convergence
-and AIC-based selection of a stock-recruit relationship, we performed
-logistic regressions. For results for AIC-based selection of EM process
-error type (multiple categories) we performed multinomial regressions.
-For other measures of reliability we fit standard linear regression
-models. Because relative errors (Eq. \ref{relerror}) and Mohn's \(\rho\)
-for the various parameters are bounded below at -1, we used a
-transformation of these values
-\begin{equation}\label{bias_regression_response}
+The measures of central tendency and variability of observed values for
+specific OM and EM attributes (e.g., low or high observation error) that
+we described above can indicate scenarios that provide better or poorer
+reliability. However, the OM and EM attributes that we investigated are
+numerous, so we used two methods to summarize the most important factors
+for differences in results. The first method was fitting regression
+models with the response being each of the measures of reliability
+described above and predictor variables were defined based on OM and EM
+characteristics (e.g., MacKinnon et al. 1995; Wang et al. 2017; Harwell
+et al. 2018). For the binary indicators of convergence and AIC-based
+selection of a SRR, we performed logistic regressions. For indicators of
+AIC-based selection of EM process error type (multiple categories) we
+performed multinomial regressions. For other measures of reliability we
+fit linear regression models to transformed responses. Because relative
+errors (Eq. \ref{relerror}) and Mohn's \(\rho\) for the various
+parameters are bounded below at -1, we used a transformation of these
+values \begin{equation}\label{bias_regression_response}
 y_i = \log\left[f\left(\widehat \theta_i,\theta_i\right)+1\right]
@@
 the rpart package (\url{https://cran.r-project.org/package=rpart}) to
-fit trees and, for clarity we pruned them to show the primary branches.
+fit trees. Full trees were determined using default settings except that
+we increased the number of cross-validations to 100. For clarity, we
+pruned the full trees to show just the primary branches.
+
+We also provide detailed results for all measures of reliability at each
+combination of OM and EM attributes in the Supplementary Materials. For
+confidence intervals of probability of convergence, we used the
+Clopper-Pearson exact method (Clopper and Pearson 1934; Thulin 2014).
+For AIC selection of process error configuration we provide estimates of
+the proportions of simulations where each EM type was selected. For AIC
+selection of the SRR, we provide predicted probabilities from logistic
+regressions as a function of SSB variability for each OM and EM type. We
+estimated bias as the median of the relative errors across all
+simulations for a given OM and EM combination. We constructed 95\%
+confidence intervals for the median relative bias, and Mohn's \(\rho\)
+using the binomial distribution approach as in Miller and Hyun (2018)
+and Stock and Miller (2021). For each EM we calculated median and 95\%
+confidence intervals using the same methods as that for relative bias.
 
@@
 
-\subsubsection*{Deviance reduction and classification
-trees}\label{deviance-reduction-and-classification-trees}
-\addcontentsline{toc}{subsubsection}{Deviance reduction and
-classification trees}
-
-EM process error assumption was the single attribute that resulted in
-the largest percent reduction in deviance for all OM process error types
-other than R+S OMs where the EM M assumption explained the most residual
-deviance (Table \ref{convergence_PRD_table}). However, also including
-interactions of OM and EM factors also provided large reductions in
-residual deviance, suggesting successful convergence depended on a
-combination OM and EM attributes. Classification trees for each OM
-process error type, all had the primary branch defined using the same
-attribute that provided the largest reduction in deviance (Figure
-\ref{conv_class}).
-
-Generally, the factors defining primary splits for the trees
-corresponded to those with the largest reductions in deviance fitting
-the regression models.
-
-\subsubsection*{Full results}\label{full-results}
-\addcontentsline{toc}{subsubsection}{Full results}
-
-For many R and R+S OMs, convergence rate declined when either the median
-natural mortality rate or the Beverton-Holt SRR was estimated even when
-the process error assumptions of the EMs and OMs matched (Figure
-\ref{hessian_SE_convergence}, A). When there was high observation error
-and constant fishing pressure (\(F = F_{\text{MSY}}\) for all 40 years),
-convergence was poor for of all EM process error configurations other
-than R EMs when fitted to R OMs (\(\sigma_{2+} = 0\)) regardless of
-whether median natural mortality and SRRs were estimated. Convergence of
-R EMs was high for all R and R+S OMs except when there was high
-observation error and constant fishing pressure, and when median natural
-mortality and SRRs were estimated. R+S EMs fit to R OMs exhibited poor
-convergence regardless of whether natural mortality or a SRR was
-estimated. R+S EMs fit to R+S OMs had highest convergence rates when
-there was contrast in fishing pressure and low observation error.
-Convergence rates were high for all EMs when fit to data from R+S OMs
-with lower observation error except those where median natural mortality
-and/or SRRs were estimated.
-
-Convergence of all EMs fitted to R+M OMs was highest when the OMs had
-higher natural mortality process error variability, low observation
-error, and contrast in fishing pressure (Figure
-\ref{hessian_SE_convergence}, B). R+M EMs that estimated autocorrelation
-of process errors had poor convergence for R+M OMs when there was low
-natural mortality process error variability regardless of
-autocorrelation of the simulated process errors. R+S EMs fitted to data
-generated from R+M OMs always converged poorly whether or not median
-natural mortality and the Beverton-Holt SRR were estimated.
-
-The R+S EMs, in particular, had poor convergence when fit to data
-generated from R+Sel OMs with lower selectivity process error
-variability or higher observation error (Figure
-\ref{hessian_SE_convergence}, C). R+Sel EMs generally converged better
-than other EMs for R+Sel OMs with higher process error variability,
-lower observation error, and contrast in fishing pressure regardless of
-whether median natural mortality or a SRR was estimated.
-
-For R+q OMs, convergence of R+q EMs was generally better than that of
-other EMs when there was contrast in fishing pressure (Figure
-(\ref{hessian_SE_convergence}, D). Convergence of R+S EMs was generally
-worse than that of all other EMs across all OMs whether or not median
-natural mortality or a SRR was estimated. Again, convergence probability
-generally declined for all EMs when median natural mortality or a SRR
-was estimated.
-
-We found a wide range of maximum absolute values of gradients for models
-that converged (Figure \ref{hess_grad}). The largest value observed for
-a given EM and OM combination was typically \(<10^{-3}\), but many
-converged models had values greater than 1. For many OMs, EMs that
-assumed the correct process error type and did not estimate median
-natural mortality or the Beverton-Holt SRR produced the lowest gradient
-values.
-
-\subsection*{AIC performance for process error
-structure}\label{aic-performance-for-process-error-structure}
-\addcontentsline{toc}{subsection}{AIC performance for process error
-structure}
-
-\subsubsection*{Deviance reduction and classification
-trees}\label{deviance-reduction-and-classification-trees-1}
-\addcontentsline{toc}{subsubsection}{Deviance reduction and
-classification trees}
-
-\subsubsection*{Full results}\label{full-results-1}
-\addcontentsline{toc}{subsubsection}{Full results}
-
-Marginal AIC accurately determined the correct process error assumptions
-in EMs when data were generated from R and R+S OMs, regardless of
-whether median natural mortality or a SRR was estimated (Figure
-\ref{pe_aic}, A). Attempting to estimate median natural mortality or a
-SRR separately had a negligible effect on the accuracy of determining
-the correct process error assumption. When both were estimated, there
-was a noticeable reduction in accuracy when OMs had a constant fishing
-pressure, low observation error, and larger variability in recruitment
-process errors.
-
-For R+M OMs, marginal AIC only accurately determined the correct process
-error model and correlation structure when observation error was low and
-variability in natural mortality process errors was high (Figures
-\ref{pe_aic}, B). Of these OMs, estimating the median natural mortality
-rate only reduced the accuracy of AIC when natural mortality process
-errors were independent and fishing pressure was constant. For OMs with
-poor model selection accuracy, AIC most frequently selected EMs with
-process errors in catchability (R+q) or selectivity (R+Sel). Selection
-of R+S EMs was generally unlikely.
-
-Marginal AIC most accurately determined the correct source of process
-error and correlation structure for R+Sel OMs with low observation error
-(Figures \ref{pe_aic}, C). When there was low variability in selectivity
-process errors and high observation error, R+q or R+S EMs were more
-likely to have the best AIC. Whether median natural morality or SRRs
-were estimated appeared to have little effect on the performance of AIC.
-
-Marginal AIC most accurately determined the correct source of process
-error and correlation structure for R+q OMs with high variability in
-catchability process errors (Figures \ref{pe_aic},D). The R+q OMs with
-low variability in catchability process errors and high observation
-error had the least model selection accuracy. However, for these OMs,
-the marginal AIC accurately determined the correct source of process
-error (but not correlation structure) except when OMs assumed a constant
-fishing pressure and EMs estimated both median natural morality and the
-SRR.
-
-\subsection*{AIC performance for the stock-recruit
-relationship}\label{aic-performance-for-the-stock-recruit-relationship}
-\addcontentsline{toc}{subsection}{AIC performance for the stock-recruit
-relationship}
-
-\subsubsection*{Deviance reduction and classification
-trees}\label{deviance-reduction-and-classification-trees-2}
-\addcontentsline{toc}{subsubsection}{Deviance reduction and
-classification trees}
-
-\subsubsection*{Full results}\label{full-results-2}
-\addcontentsline{toc}{subsubsection}{Full results}
-
-Our comparisons of model performance conditioned on assuming the true
-process error configuration is known (EM and OM process error types
-match) and we focus on results where the EMs assume median natural
-mortality is known because there was little difference in results when
-the EMs estimated this parameter. Broadly, we found generally poor
-accuracy of AIC in selecting models assuming a Beverton-Holt SRR over
-the null model without an SRR for all OMs. However, we also found
-increased accuracy of AIC in determining the Beverton-Holt SRR when the
-simulated population exhibited greater variation in spawning biomass for
-nearly every OM (Figure \ref{sr_aic}).
-
-With R and R+S process error assumptions, probability of lowest AIC for
-the B-H SRR as a function of SSB variability were greatest for OMs with
-contrast in fishing pressure and lower process variability in
-recruitment (Figure \ref{sr_aic}, A). The largest variation in SSB
-occurred in OMs with larger recruitment variability (\(\sigma_R = 1.5\);
-Figure \ref{sr_aic}, A, right column group), but the same high AIC
-accuracy was achieved for OMs with lower recruitment variability at
-lower levels of SSB variation. The level of observation error had little
-effect on AIC accuracy.
-
-For R+M OMs, probability of lowest AIC for the Beverton-Holt SRR
-increased steeply with variation in SSB whether it was induced by
-contrast in fishing or variation in natural mortality process error.
-(Figure \ref{sr_aic}, B). There was little difference in AIC accuracy
-whether the natural mortality process errors were correleted and,
-similar to R+S OMs, there was also little effect due to level of
-observation error.
-
-For R+Sel OMs, contrast in fishing pressure over time was the primary
-source of variation in SSB and these are the OMs where AIC accuracy for
-the Beverton-Holt SRR was greatest (Figure \ref{sr_aic}, C). There was
-little effect of variability or correlation of selectivity process
-errors or the level of observation error on AIC accuracy.
-
-Like the R+Sel OMs, the greatest accuracy for AIC in selecting the
-Beverton-Holt SRR occurred for R+q OMs where there was contrast in
-fishing pressure over time which is also where there was the greatest
-variation in SSB (Figure \ref{sr_aic}, D). There was also little effect
-of variability or correlation of catchability process errors or the
-level of observation error on AIC accuracy.
+For probability of convergence, EM process error assumption was the
+single attribute that resulted in the largest percent reduction in
+deviance (14-28\%) for all OM process error types other than R+S OMs
+where the EM M assumption explained the most residual deviance
+(\textgreater11\%; Table \ref{convergence_PRD_table}). However,
+including interactions of OM and EM factors also provided large
+reductions in residual deviance (35-47\%), suggesting successful
+convergence depended on a combination OM and EM attributes.
+
+Classification trees for each OM process error type, all had the primary
+branch defined using the same attribute that provided the largest
+reduction in deviance (Figure \ref{conv_class}). EMs that assumed R+S
+process errors converged poorly for all OMs that were simulated with the
+alternative process error assumptions (R, R+M, R+Sel, an R+q OMs). For
+all trees, branches based on the OM fishing mortality history showed
+better convergence when the OM included a change in fishing pressure.
+Branches based on whether the Beverton-Holt SRR was assumed or not,
+showed better convergence when it was not estimated and branches based
+on whether the EM estimated the median natural mortality rate showed
+better convergence when it was treated as known. For certain R+M and
+R+Sel OMs, better convergence was also observed when there was lower
+observation uncertainty.
+
+When convergence is defined by a gradient threshold, the primary factor
+explaining deviance reduction is the same for all OM process error
+types, but there are some differences in deviance reduction for
+secondary factors (Table \ref{convergence_gradient_PRD_table}), and
+probability of convergence, overall, was lower (Figure
+\ref{conv_gradient_class}). We found a wide range of maximum absolute
+values of gradients for models that had invertible Hessians (Figure
+\ref{hess_grad}). The largest value observed for a given EM and OM
+combination was typically \(<10^{-3}\), but many converged models had
+values greater than 1. For many OMs, EMs that assumed the correct
+process error type and did not estimate median natural mortality or the
+Beverton-Holt SRR produced the lowest gradient values.
+
+\subsection*{AIC performance}\label{aic-performance}
+\addcontentsline{toc}{subsection}{AIC performance}
+
+\subsubsection*{Process error structure}\label{process-error-structure}
+\addcontentsline{toc}{subsubsection}{Process error structure}
+
+For AIC selection of the correct process error configuration, the
+magnitude of observation and process error were the attributes that
+resulted in the largest percent reductions in deviance across OM process
+error types other than R OMs (Table \ref{AIC_PE_PRD_table}). Both
+variance of apparent survival random effects (\(\sigma_{2+}\)) and
+degree of observation error explained the largest reductions for R+S
+(17-22\%) and R+Sel (8-26\%) OMs, whereas variance of process errors
+provided the largest reductions in R+M (\textgreater9\%) and R+q
+(\textgreater13\%) OMs. Comparatively, none of the OM or EM attributes
+explained particularly large reductions in deviance for R OMs, but
+fishing history, whether a SRR was estimated, and whether median natural
+mortality was known or estimated provided similar and the largest
+reductions (approximately 5-6\%). Inclusion of second and third order
+interactions, did not provide large reductions in deviance for any of
+the OM process error types.
+
+For all OM process error types other than R OMs, the attributes defining
+the primary branches of classification trees matched those that provided
+the largest reductions in deviance (Figure \ref{AIC_PE_class}). Across
+all OMs, AIC was more accurate for the process error type when process
+error variability was greater and when observation error was lower. No
+branches were estimated for classification trees fit to the R OMs,
+likely because accuracy was high across all simulations (0.94), although
+inspection of the fine-scale results shows there is some degradation in
+AIC selection when a SRR and median natural mortality rate are estimated
+for R OMs with constant fishing pressure and high observation error
+(Figure \ref{pe_aic}, top left).
+
+\subsubsection*{Stock-recruit
+relationship}\label{stock-recruit-relationship}
+\addcontentsline{toc}{subsubsection}{Stock-recruit relationship}
+
+Logistic regressions for AIC selection of the Beverton-Holt SRR, showed
+OM fishing history and variation in SSB
+(\(\log \text{SD}_{\text{SSB}}\)) provided substantial reductions in
+deviance for R+M (\textgreater13\%), R+Sel (\textgreater26\%), and R+q
+(\textgreater24\%) OMs (Table \ref{AIC_SRR_PRD_table}). For R OMs,
+fishing history provided the largest reduction in deviance
+(\textgreater9\%) whereas none of the attributes individually provided
+large reductions in deviance for R+S OMs (all \textless5\%). However,
+inclusion of all attributes provided larger reductions in deviance than
+the sum of individual contributions for both R (\textgreater30\%) and
+R+S (\textgreater18\%) OMs. For all OM process error types, inclusion of
+interaction terms provided relatively little reduction in residual
+deviance.
+
+Attributes defining the primary branches of classification trees for AIC
+selection of the SRR assumption were the same as those explaining the
+largest reductions in deviance for the logistic regression models
+(Figure \ref{AIC_SRR_class}). All branches based on variation in SSB
+showed better accuracy with larger variability in SSB and all branches
+based on fishing history showed better accuracy when there was contrast
+in fishing pressure. Branched based on OM observation error or
+recruitment variability showed better accuracy when they were lower. For
+R OMs, a combination of lower recruitment variability, contrast in
+fishing pressure, and higher SSB variability produced AIC accuracy over
+0.8. For R+S OMs, lower recruitment variability and observation error
+and higher SSB variability produced AIC accuracy of 0.79. For R+M,
+R+Sel, and R+q OM,s accuracy of 0.87 to 0.94 was observed just with
+higher SSB variability.
 
@@
 
-\subsubsection*{Deviance reduction and regression
-trees}\label{deviance-reduction-and-regression-trees}
-\addcontentsline{toc}{subsubsection}{Deviance reduction and regression
-trees}
-
-Across all OM process error types, fishing history was the most
-important single factor in reducing residual deviation in log-scale
-errors for the Beverton-Holt density-independent parameter \(a\), and
-also for the density-independent parameter \(b\) for R+M, R+Sel, and R+q
-OMs (Table \ref{bias_SR_pars_PRD_table}). For R and R+S OMs the EM
-process error assumption was equally important for estimating \(b\), but
-the the reductions were small (mostly less than 4\%) for any factor and
-any OM process error type. There was also no substantial reduction in
-deviance with second or third order interactions of factors for either
-parameter in any OM process error types.
-
-The regression trees indicate that having contrast in fishing pressure
-reduced the median error for both SR parameters for all OM types
-
-We observed different effects of factors on deviance reduction for
-errors in estimation of the two Beverton-Holt stock-recruit parameters.
-When analyses were For bias of the density-dependent parameter \(b\) as
-measured by the log absolute value of log-scale errors, deviance was not
-reduced by any of the individual OM or EM factors except for variation
-in recruitment \(\sigma_R\) for R and R+S OMs. However, second and(or)
-third order interactions of the OM and EM factors also provided large
-reductions in deviance for all OM process error types which indicated
-that some combinations of factors could reduce bias for the \(b\)
-parameter. Despite the explanation variation in the \(b\) estimation
-erorrs, bias was still significant regardless of the combination of OM
-and EM factors (Figures regtrees and full results).
-
-Median natural mortality rate showed reductions in deviance for R, R+S,
-and R+M OMs with second order interactions and R+M also showed
-reductions with third order interactions.
-
-\subsubsection*{Full results}\label{full-results-3}
-\addcontentsline{toc}{subsubsection}{Full results}
-
-\paragraph*{Spawning stock biomass}\label{spawning-stock-biomass}
-\addcontentsline{toc}{paragraph}{Spawning stock biomass}
-
-For R OMs (\(\sigma_{2+} = 0\)), there was no indication of bias (95\%
-confidence intervals included 0) in terminal year SSB for any of the
-estimating models regardless of process error assumptions, except when
-no SR assumption was made, recruitment variability was low, and there
-was contrast in fishing mortality and high observation error (Figure
-\ref{SSB_rel_error}, A). However, errors in terminal SSB estimates were
-highly variable when median natural mortality was estimated and there
-was constant fishing pressure and high observation error (Figure
-\ref{SSB_rel_error}, A, second row).
-
-For R+S OMs, the EMs with matching process error assumptions generally
-produced unbiased estimation of terminal SSB except when median natural
-mortality was estimated and there was high observation error. In R+S OMs
-with low observation error, EMs with incorrect process error assumptions
-typically provided biased estimation of terminal year SSB. Estimating
-the Beverton-Holt SRR had little discernible effect on bias of terminal
-year SSB estimation whereas estimating median M tended to produce more
-variability in errors in terminal SSB estimation similar to R OMs.
-
-For R+ M OMs with low variability in natural mortality process errors,
-low observation error and contrast in fishing motality over time all EMs
-produced low variability in SSB estimation error that indicated
-unbiasedness (Figure \ref{SSB_rel_error}, B, third row). However, larger
-variability in natural mortality process errors increased bias of EMs
-without the correct process error type. Estimating median natural
-mortality increased variability of SSB estimation error particularly for
-OMs with high observation error and constant fishing pressure over time.
-It also increased bias in SSB estimation for many R+M OMs. Like R and
-R+S OMs, estimating a SRR had little discernible effect on SSB bias.
-
-For R+Sel OMs, there was no evidence of bias for any EMs when
-variability in selectivity process error and observation error was low,
-and with contrast in fishing mortality (Figure \ref{SSB_rel_error}, C).
-The largest bias occurred for any EMs that estimated median natural
-mortality when the OMs had high observation error, constant fishing
-pressure, and greater variability in selectivity process errors
-(\(\sigma_{\text{Sel}} = 0.5\)) or low selectivity process errors
-(\(\sigma_{\text{Sel}} = 0.1\)) and low observation error. However,
-there was no evidence of SSB bias for correctly specified R+Sel EMs when
-observation error was low and variation in selectivity process errors
-was larger, whether median natural mortality was estimated or not
-(Figure \ref{SSB_rel_error}, C, third row). We only observed an effect
-of estimating the Beverton-Holt SRR for R+Sel OMs that had high
-observation error and contrast in fishing pressure where estimating the
-SRR produced less biased SSB estimation for many EMs (Figure
-\ref{SSB_rel_error}, C, top row).
-
-All EMs fit to data from R+q OMs with low observation error and contrast
-in fishing pressure exhibited little evidence of bias in terminal SSB
-estimation except for R+M EMs when there was no AR1 correlation in
-catchability process errors (Figure \ref{SSB_rel_error}, D). Many EMs
-also performed well in R+q OMs with low observation error, but no
-contrast in fishing pressure. For R+q OMs with high observation error
-and contrast in fishing pressure, EMs that estimated the Beverton-Holt
-SRR exhibited less SSB bias than those that did not. Estimating median
-natural mortality in the EMs only resulted in much more variable SSB
-estimation errors when there was no contrast in fishing pressure (Figure
-\ref{SSB_rel_error}, D, first and third rows).
-
-For all OM process error types, relative errors in terminal year
-recruitment were generally more variable than SSB, but effects of R and
-R+S OM and EM attributes on bias (i.e, negative or positive or none)
-were similar (Figure \ref{R_rel_error}, A). Furthermore, for EM
-configurations where bias in terminal SSB was evident, median relative
-errors in recruitment often indicated stronger bias in recruitment of
-the same sign.
-
-\subsubsection*{Beverton-Holt
-parameters}\label{beverton-holt-parameters}
-\addcontentsline{toc}{subsubsection}{Beverton-Holt parameters}
-
-Across all OMs, there was generally less bias and(or) lower variability
-in estimation of the Beverton Holt \(a\) parameter than the \(b\)
-parameter. In R and R+S OMs, EMs with the correct assumptions about
-process errors provided the least biased estimation of Beverton-Holt SRR
-parameters when there was a change in fishing pressure over time and
-lower variability of recruitment process errors, but there was little
-effect of estimating median natural mortality and a small increase in
-bias for those OMs that had high observation error (Figure
-\ref{SR_rel_error}, A). For other R and R+S OMs, estimating natural
-mortality often resulted in less biased estimation of SRR parameters.
-There was generally large variability in relative errors of the SRR
-parameter estimates, but the lowest variability occurred with low
-variability in recruitment and little or no variability in survival
-process errors (\(\sigma_{2+} \in \{0,0.25\}\)), and contrast in fishing
-pressure.
-
-In R+M OMs, the most accurate estimation of SRR parameters for all EM
-process error assumptions occurred when there was a change in fishing
-pressure, greater variability in natural mortality process errors, and
-lower observation error (Figure \ref{SR_rel_error}, B). Relative to the
-R, and R+S OMs, there was even less effect of estimating median natural
-mortality on estimation bias for the SRR parameters.
-
-Bias for SRR parameters was large and variability in relative errors was
-greatest for most EMs fit to R+Sel OMs with constant fishing pressure
-(Figure \ref{SR_rel_error}, C). Less bias in parameter estimation
-occurred for OMs with a change in fishing pressure and the best accuracy
-occurred for those OMs that had low observation error and more variable
-and uncorrelated selectivity process errors, and when the EMs had with
-the correct process error assumption. There was little effect of
-estimating natural mortality on relative errors for SRR parameters.
-
-Like R+Sel OMs, relative errors in SRR parameters for R+q OMs were more
-accurate for most EM process error types when OMs had contrast in
-fishing pressure and lower observation error (Figure \ref{SR_rel_error},
-D). However, the best accuracy occurred for those OMs that had lower
-variability in catchability process errors. The worst accuracy of SRR
-parameter estimation regardless of EM type occurred when R+q OMs had low
-observation error and constant fishing pressure (Figure
-\ref{SR_rel_error}, D, fourth row).
+\subsubsection*{Terminal year spawning stock biomass, fishing mortality,
+and
+recruitment}\label{terminal-year-spawning-stock-biomass-fishing-mortality-and-recruitment}
+\addcontentsline{toc}{subsubsection}{Terminal year spawning stock
+biomass, fishing mortality, and recruitment}
+
+Regression models for log-scale errors in SSB that included the various
+OM and EM factors showed little reduction in deviance (\textless5\%) for
+any of the factors across all OM process error types (Table
+\ref{bias_SSB_PRD_table}). The attributes producing the largest
+reductions were the EM assumption for median natural mortality (known or
+estimated) for R, R+M, R+Sel, and R+q OMs (1-3\%), EM process error type
+for R+S OMs (4\%) and fishing history for all OM types (1-5\%).
+Including second interactions provided largest reductions in residual
+deviance (10- 26\%). Including third order interactions also provided
+large further reductions for R, R+S, and R+q OMs (between 5 and 11\%).
+In all regression trees, branches based on fishing history and level of
+observation error generally showed lower bias in SSB with contrast in
+fishing and lower observation error (Figure \ref{SSB_bias_regtree}). For
+branches based on treatment of median natural mortality rate, bias was
+generally lower when it was known rather than estimated. For some R+Sel
+and R+q OMs, less bias in SSB was shown when the EM process error
+configuration was correct.
+
+Results for bias in fishing mortality and recruitment generally matched
+those for SSB, except that directions of bias for fishing mortality were
+opposite to those for SSB and recruitment. Effects of individual OM and
+EM factors on regression models were similarly small as measured by
+reduction in deviance (Tables \ref{bias_F_PRD_table} and
+\ref{bias_R_PRD_table}). Primary branches of regression trees were in
+most cases identical to those for SSB (Figures \ref{F_bias_regtree} and
+\ref{R_bias_regtree}).
+
+\subsubsection*{Stock-recruit
+parameters}\label{stock-recruit-parameters}
+\addcontentsline{toc}{subsubsection}{Stock-recruit parameters}
+
+Regression models for transformed relative errors of estimates of both
+the Beverton-Holt \(a\) and \(b\) parameters showed none of the factors
+explained large percent reductions in deviance (Table
+\ref{bias_SR_pars_PRD_table}). The OM fishing history provided the
+largest reduction for most OM process error types for both parameters,
+but percent reductions were less than 5.6\% except for R+Sel OMs where
+the percent reductions were 11.37\% and 7.97\% for the \(a\) and \(b\)
+parameters, respectively and for just the \(b\) parameter for R+q OMs
+(10\%). The EM process error assumption provided similar reductions in
+deviance for both parameters for R OMs. Including interactions also did
+not produce important reductions in deviance (increase of .
+
+For regression trees of log-scale error in Beverton-Holt \(a\) and \(b\)
+parameters, lower bias was indicated with contrast in OM fishing
+pressure for all branches in trees for each OM process error type
+(Figures \ref{SR_a_bias_regtree} and \ref{SR_b_bias_regtree}). For all
+branches based on recruitment variability in trees for R and R+S OMs,
+lower bias in both \(a\) and \(b\) was observed with less recruitment
+variability. For R OMs with contrast in fishing pressure and greater
+recruitment variability EMs that assumed the incorrect R+M process
+errors produced lower bias in both \(a\) and \(b\) than other process
+error assumptions. Across all OMs, there was generally less bias and(or)
+lower variability in estimation of the \(a\) parameter than the \(b\)
+parameter (Figure \ref{SR_rel_error}).
 
@@
 
-Across all OMs and EMs there was little effect of estimating SRRs on the
-bias in estimation of median natural mortality (Figure
-\ref{M_rel_error}). Median natural mortality rate was estimated
-accurately by all EM process error types for all R OMs except those with
-high observation error and constant fishing pressure, in which case
-relative errors were high (Figure \ref{M_rel_error}, A,
-\(\sigma_{2+} = 0\)). For R+S OMs estimation of median natural mortality
-rate was most accurate when observation error was low and there was
-contrast in fishing pressure and the EM process error type was correct.
-
-For R+M OMs, median natural mortality was estimated most accurately,
-regardless of EM process error type, when OMs had a change in fishing
-pressure and low observation error (Figure \ref{M_rel_error}, B).
-However, those R+M OMs that also had greatest variability in AR1
-correlated natural mortality process errors only had unbiased estimation
-when the EM process error type was correct.
-
-All EM process error types accurately estimated median natural mortality
-rate for R+Sel OMs that had contrast in fishing pressure, low
-observation error, and low selectivity process error variability (Figure
-\ref{M_rel_error}, C). When selectivity process error variability
-increased, the incorrect EM process errors produce more biased
-estimation of median natural mortality rate. The least accurate
-estimation occurred for all EM process error types when observation
-error was high and fishing pressure was constant.
-
-Like R+Sel OMs, all EM process error types produced accurate estimation
-of median natural mortality rate when fit to R+q OMs with contrast in
-fishing pressure, low observation error and low catchability process
-error variability (Figure \ref{M_rel_error}, D). Most EM process error
-types produced biased estimation of median natural mortality when R+q
-OMs had high observaiton error and constant fishing pressure.
+Fitted regression models for log-scale errors in median natural
+mortality rate showed largest percent reductions in residual deviance
+for R+S and R\_M models (Table \ref{bias_median_M_PRD_table}). The
+largest reductions for a single attribute was the EM process error
+assumption (\textgreater20\%) and fishing history (\textgreater15\%) for
+R+S OMs. Fishing history also provided \textgreater10\% reduction for
+R+M OMs, but reductions for all factors in R, R+Sel, and R+q OMs were
+relatively low (\textless6\%). Interactions of OM and EM factors also
+provided substantial further reductions for R+S and R+M OMs (between 8
+and 15 \% for second order interactions).
+
+Regression trees with branches based on fishing history showed lower
+bias in median natural mortality rate with contrast in fishing pressure
+and branches based on level of observation error showed lower bias with
+more precise observations (Figure \ref{med_M_bias_regtree}). For R OMs,
+branches based on EM process error assumption showed lower bias with EMs
+assuming the correct R and the incorrect R+S assumption. For R+S and R+M
+OMs, branches based on EM process error showed only the correct EM
+process error assumption with less bias.
 
@@
 
-\subsubsection*{Deviance reduction and regression
-trees}\label{deviance-reduction-and-regression-trees-1}
-\addcontentsline{toc}{subsubsection}{Deviance reduction and regression
-trees}
-
-Fits of glms showed little explanatory power of any of the OM an EM
-attributes for Mohn's rho for SSB, F, and R. Note that the factors that
-showed the largest reductions in deviance (which were very small) often
-do not comport with factors that show differences in median Mohn's rho
-results. This is because the glms and regression trees are reducing
-residual variance and there could be large variability in values, but
-the medians are less sensitive to extreme values. For example we can see
-that there is large Mohn's rho for R in R OMs with larger variability
-assumed in recruitment and higher observation error (Figure X), but
-these factors suggest little reduction in residual deviance (Table X).
-
-\subsubsection*{Full results}\label{full-results-4}
-\addcontentsline{toc}{subsubsection}{Full results}
-
-Mohn's \(\rho\) for SSB was small in absolute value for all R and R+S
-OMs, regardless of EM process error types, and whether median natural
-mortality rate or SRRs were estimated (Figure \ref{mohns_rho_ssb}, A).
-The strongest retrospective patterns (highest absolute Mohn's \(\rho\)
-values) occurred in OMs with the largest apparent survival process error
-variability, high observation error, and contrast in fishing pressure,
-but only for EMs with the incorrect process error type and where median
-natural mortality rate was assumed known (median \(\rho\) was
-approximately -0.15). For R+M, R+Sel, and R+q OMs, Mohn's \(\rho\) was
-also small in absolute value, but median values were all closer to 0
-than the largest values in the R and R+S OMs (Figure
-\ref{mohns_rho_ssb},B-D). For these OMs, there was no noticeable effect
-of estimation of median natural mortality rate or SRRs on Mohn's
-\(\rho\) for any EM process error types.
-
-Mohn's \(\rho\) for recruitment was small in absolute value for all R
-OMs with low variability in recruitment process errors, regardless of EM
-process error type, and whether median natural mortality rate or SRRs
-were estimated (Figure \ref{mohns_rho_R}, A). However, R and R+S OMs
-with greater recruitment process variability and higher observation
-error had median Mohn's \(\rho\) for recruitment greater than zero for
-most EMs even when the EM process error type was correct. In R+S OMs
-with lower observation error, EMs with the correct process error type
-exhibited better median Mohn's \(\rho\) close to 0 than EMs with the
-incorrect process error type. For R+M, R+Sel, and R+q OMs, results for
-Mohn's \(\rho\) for recruitment are similar to those for SSB, but the
-range in median values and variation in Mohn's \(\rho\) values for a
-given OM are generally larger for recruitment (Figure \ref{mohns_rho_R},
-B-D).
+Regression models for Mohn's \(\rho\) of SSB showed little reduction in
+deviance for any of the OM an EM attributes (\textless2\%; Table
+\ref{mohns_rho_SSB_PRD_table}). The lack of explanatory power is also
+reflected in the regression trees where median Mohn's \(\rho\) values
+are near zero unless a large combinations of OM and EM conditions occur
+(Figure \ref{SSB_mohns_rho_regtree}). For example, In R+S OMs, with
+constant fishing pressure, high observation error, and higher apparent
+survival process error, EMs that assume R+M process errors have a median
+Mohn's \(\rho = -0.068\).
+
+Similarly poor explanatory power of the OM and EM attributes occured
+when we fit regression models for Mohn's \(\rho\) of fishing mortality
+and recruitment (Tables \ref{mohns_rho_F_PRD_table} and
+\ref{mohns_rho_R_PRD_table}). Regression trees for Mohn's \(\rho\) of
+fishing mortality were similar to those for SSB in that median values of
+Mohn's \(\rho\) were close to zero for most combinations of OM and EM
+attributes (Figure \ref{F_mohns_rho_regtree}). However, we observed
+median Mohn's \(\rho\) for recruitment greater than 0.1 at branches much
+closer to the base of the trees with fewer interactions of the OM and EM
+attributes (Figure \ref{R_mohns_rho_regtree}). These branches with
+consistently large retrospective patterns were typically defined by
+larger OM observation error, OM constant fishing pressure, or incorrect
+EM process error configuration. Comparing regression model and
+regression tree fits, attributes defining the primary branches for all
+regression trees of all Mohn's \(\rho\) values (SSB, fishing mortality,
+and recruitment) generally matched those that explained the largest
+reductions in deviance.
 
@@
 can often be invertible when the maximum absolute gradient was much
-larger than what would perceived to be a sensible threshold.
+larger than what might be perceived to be a sensible threshold.
 
@@
 methods are used to calculate and invert the Hessian for variance
-estimation for models with random effects. Along with our results, the
-potential lack of accuracy imposed by these approximations, suggests at
+estimation for models with random effects. Our results, along with the
+potential lack of accuracy imposed by these approximations, suggest at
 least investigating whether the Hessian is positive definite when the
@@
 
-Our results comport with those of de Valpine and Hastings (2002) who
+Our results comport with those of {de Valpine and Hastings} (2002) who
 found AIC could not distinguish among state-space SRRs that were fit
@@
 the random effects have no trend on average across simulations.
-Szuwalski et al. (2018) and lietal24 also found relatively small
+Szuwalski et al. (2018) and Li et al. (2024) also found relatively small
 retrospective patterns when the source of mis-specification was temporal
@@
 
+\subsection*{Analytical approach}\label{analytical-approach}
+\addcontentsline{toc}{subsection}{Analytical approach}
+
+We found the use of regression models and classification and regression
+trees extremely useful in understanding the most important OM and EM
+attributes explaining variation in the measures of reliability we
+examined across all simulations. The classification and regression trees
+are generally a good tool for determining the values of the OM and EM
+attributes that produce better or worse measures of reliability.
+However, determining the combination of attributes that produce the best
+or worst measures of reliability can be challenging using the trees. For
+example, in the regression tree for median natural mortality rate
+estimates in R OMs (Figure \ref{median_M_bias_regtree}), the both of the
+first branches imply bias is low regardless of OM fishing history, but
+when OM fishing pressure is constant, results are much better when OM
+observation error is low (median RE about -6\%) than when OM observation
+error is high (median RE about 40\%). Therefore, inspection of results
+by all combinations of OM and EM factors (as provided in the
+supplementary materials) can be important.
+
 \subsection*{Conclusions}\label{conclusions}
@@
 reliable inferences from state-space age-structured assessment models.
-Contrast in fishing pressure was consistently an important factor across
-all measures of reliability we examined. AIC accurately distinguished
-models with process errors on recruitment only (R) or on recruitment and
-apparent survival (R+S). Accuracy for other process error types required
-a strong signal (high process variability) with low noise (low
-observation uncertainty). Therefore, we expect practitioners will find
-R+S configurations to provide satisfactory diagnostics across a range of
-life history and data quality scenarios. AIC generally performed poorly
-for selecting the SRR, but performance was improved with low recruitment
-variability and contrast in fishing pressure. Some bias in estimation in
-at least one of the SRR parameters existed in nearly all OM-EM
-combinations. Because bias in terminal SSB and retrospective patterns
-were indifferent to whether or not the SRR was estimated, and
-convergence was slightly better without the SRR, a sensible default
-would be to fit models without an assumed SRR.
+Contrast in fishing pressure and lower observation error consistently
+improved all measures of reliability we examined. AIC accurately
+distinguished models with process errors on recruitment only (R) or on
+recruitment and apparent survival (R+S). Therefore, we expect
+practitioners will find R+S configurations to provide satisfactory
+diagnostics across a range of life history and data quality scenarios.
+Accurate AIC selection of the Beverton-Holt SRR generally required a
+combination of low recruitment variability, contrast in fishing pressure
+and large variation in SSB over time. However, when the Beverton-Holt
+was correctly assumed, some bias in estimation in at least one of the
+parameters existed regardless of any of the other OM and EM
+configurations. Because bias in terminal SSB and retrospective patterns
+were indifferent to whether or not the SRR was estimated, the prevalence
+of bias in SRR parameter estimation, and often better convergence
+without the SRR, a sensible default would be to not assume a SRR when
+fitting assessment models.
 
@@
 This work was funded by NOAA Fisheries Northeast Fisheries Science
-Center. We thank Jon Deroba for helpful comments on an earlier version
-of this manuscript.
+Center. We thank Jon Deroba, 2 anonymous reviewers, and the associate
+editor for helpful comments on earlier versions of this manuscript that
+markedly improved its clarity.
 
@@
 Hoyle, S.D., Maunder, M.N., Punt, A.E., Mace, P.M., Devine, J.A., and
-A???mar, Z.T. 2022. Preface: Developing the next generation of stock
+A'mar, Z.T. 2022. Preface: Developing the next generation of stock
 assessment software. Fisheries Research \textbf{246}: 106176.
@@
 \begin{center}
-\includegraphics{convergence_classification_plots}
+\includegraphics[width = 1.4\textwidth]{convergence_classification_plots}
 \end{center}
 \caption{Classification trees indicating primary factors determining convergence as defined by providing hessian-based standard errors for R, R+S, R+M, R+Sel and R+q OMs.
-Lower or higher convergence rates are indicated by red or green polygons, respectively}\label{conv_class}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{AIC_PE_classification_plots}
+Lower or higher convergence rates are indicated by more red or green polygons, respectively}\label{conv_class}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[width = 1.4\textwidth]{AIC_PE_classification_plots}
 \end{center}
@@
 Each node shows the proportion of EM process error models with lowest AIC (top) and number of observations (bottom) for the corresponding subset.
-Lower or higher accuracy of the process error assumption are indicated by red or green polygons, respectively.}\label{AIC_PE_class}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{AIC_SRR_classification_plots}
-\end{center}
-\caption{Classification trees indicating primary factors determining which EM stock recruitment assumption (none or Beverton-Holt) provides the lowest AIC for R, R+S, R+M, R+Sel and R+q OMs.
-Each node shows the proportion of EMs that assume the stock-recruit relationship with lowest AIC (top) and number of observations (bottom) for the corresponding subset.
+Lower or higher accuracy of the process error assumption are indicated by more red or green polygons, respectively.}\label{AIC_PE_class}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[width = 1.4\textwidth]{AIC_SRR_classification_plots}
+\end{center}
+\caption{Classification trees indicating primary factors determining which EM SRR assumption (none or Beverton-Holt) provides the lowest AIC for R, R+S, R+M, R+Sel and R+q OMs.
+Each node shows the proportion of EMs that assume the SRR with lowest AIC (top) and number of observations (bottom) for the corresponding subset.
 Lower or higher accuracy of the process error assumption are indicated by more red or green polygons, respectively.}\label{AIC_SRR_class}
@@
 \begin{table}
-\caption{For each OM process error type (columns), percent reduction in deviance for logistic regression models fit to indicators of EM stock-recruit assumption (none or Beverton-Holt) with lowest AIC with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{AIC_SRR_PRD_table}
+\caption{For each OM process error type (columns), percent reduction in deviance for logistic regression models fit to indicators of EM SRR assumption (none or Beverton-Holt) with lowest AIC with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{AIC_SRR_PRD_table}
 {\input{AIC_SRR_PRD_table}}
@@
 
-\begin{landscape}
 \begin{table}
 \caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
-\ref{bias_regression_response} for the Beverton-Holt stock recruit relation ship parameters with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_SR_pars_PRD_table}
+\ref{bias_regression_response} for the terminal year SSB with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_SSB_PRD_table}
+{\input{bias_SSB_PRD_table}}
+\end{table}
+
+\begin{landscape}
+\begin{table}
+\caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
+\ref{bias_regression_response} for the Beverton-Holt SRR parameters with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_SR_pars_PRD_table}
 {\input{bias_SR_pars_PRD_table}}
@@
 
+\subsection*{Referenced Figures}\label{referenced-figures}
+\addcontentsline{toc}{subsection}{Referenced Figures}
+
 \begin{figure}[!ht]
@@
 \begin{center}
-\includegraphics{type_4_convergence_plots}
-\end{center}
-\caption{Estimated probability of fits providing hessian-based standard errors for EMs assuming alternative process error (colored points and lines), and median natural mortality (estimated or known) and Beverton-Holt stock-recruit relationships (estimated or not; along x-axis) when fitted to operating models that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{hessian_SE_convergence}
-\end{figure}
-\end{landscape}
-
-\clearpage
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{type_3_convergence_plots}
-\end{center}
-\caption{Probability of estimating models providing maximum absolute values of gradients less than $10^{-6}$ assuming alternative process error (colored points and lines), and median natural mortality (estimated or known) and Beverton-Holt stock-recruit relationships (estimated or not; along x-axis) when fitted to operating models that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{gradient_convergence}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{hess_grad_convergence_plots}
+\includegraphics[width = 1.4\textwidth]{convergence_gradient_classification_plots}
+\end{center}
+\caption{Classification trees indicating primary factors determining convergence as defined by a maximum absolute gradient < $10^{-6}$ for R, R+S, R+M, R+Sel and R+q OMs.
+Lower or higher convergence rates are indicated by more red or green polygons, respectively}\label{conv_gradient_class}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[width = 1.4\textwidth]{hess_grad_convergence_plots}
 \end{center}
 \caption{The maximum of the absolute values of all gradient values for all fits that provided hessian-based standard errors across all simuated data sets of a given OM configuration (A: R and R+S, B: R+M, C: R+Sel, or D: R+q).
-Results are conditional on EM fits with alternative process error type (colored points and lines), median natural mortality (estimated or known) and recruitment assumptions (Beverton-Holt stock-recruit relationship or not).
+Results are conditional on EM fits with alternative process error type (colored points and lines), median natural mortality (estimated or known) and recruitment assumptions (Beverton-Holt SRR or not).
 Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{hess_grad}
@@
 \begin{center}
-\includegraphics{pe_aic_plots}
-\end{center}
-\caption{Estimated probability of lowest AIC for EMs assuming alternative process error structures (colored bars) conditional on alternative assumptions for median natural mortality (estimated or known) and Beverton-Holt stock-recruit relationships (estimated or not; along x-axis) when fitted to operating models that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
+\includegraphics[width = 1.4\textwidth]{pe_aic_plots}
+\end{center}
+\caption{Estimated probability of lowest AIC for EMs assuming alternative process error structures (colored bars) conditional on alternative assumptions for median natural mortality (estimated or known) and Beverton-Holt SRR (estimated or not; along x-axis) when fitted to operating models that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
 Striped bars indicate results where the EM process error structure matches that of the operating model.}\label{pe_aic}
@@
 
-\clearpage
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics[height = 0.8\textheight]{sr_aic_plots_rev}
-\end{center}
-\caption{Estimated probability of lowest AIC from logistic regression on the log-standard deviation of the true log(SSB) in each simulation for estimating model with Beverton-Holt stock-recruit relationships, rather than the otherwise equivalent EM without the stock-recruit relationship.
-Results are conditional on median M is known in the EM and alternative assumptions EMs having the correct process error structure: R and R+S (A), R+Sel (B), R+M (C), or R+q (D), and median M is assumed known in the EM.
-Solid and dashed lines are for OMs with and without temporal contrast in fishing pressure, respectively, and polygons represent 95\% confidence intervals.
-Range of results indicates the range of log-standard deviation of log(SSB) for simulations of the particular OM.}\label{sr_aic}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{sr_aic_plots}
-\end{center}
-\caption{Estimated probability of lowest AIC from logistic regression on the log-standard deviation of the true log(SSB) in each simulation for estimating model with Beverton-Holt stock-recruit relationships, rather than the otherwise equivalent EM without the stock-recruit relationship.
-Results are conditional on alternative assumptions for median natural mortality (estimated or known) and on EMs having the correct process error structure: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Rug along x-axis denotes $SD(\log(SSB))$ values for each simulation and polygons represent 95\% confidence intervals.}\label{sr_aic_supp}
-\end{figure}
-\end{landscape}
-
-\clearpage
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{term_SSB_bias_plots}
-\end{center}
-\caption{Median relative error of terminal year SSB for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{SSB_rel_error}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{term_F_bias_plots}
-\end{center}
-\caption{Median relative error of terminal year fully-selected fishing mortality for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{F_rel_error}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{term_R_bias_plots}
-\end{center}
-\caption{Median relative error of terminal year recruitment for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{R_rel_error}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{sr_bias_plots}
-\end{center}
-\caption{Median relative error of Beverton-Holt stock-recruit parameters ($a$ and $b$) for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[width = 1.4\textwidth]{F_bias_regtree_plots}
+\end{center}
+\caption{Regression trees indicating primary factors determining reductions in sums of squares of errors in estimation measured by Eq.
+\ref{bias_regression_response} for terminal year fully-selected fishing mortality for R+S, R+M, R+Sel and R+q OMs.
+Each node shows the median error (top) and number of observations (bottom) for the corresponding subset.
+Median errors closer to or further from zero are indicated by more green or red polygons, respectively.}\label{F_bias_regtree}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[width = 1.4\textwidth]{R_bias_regtree_plots}
+\end{center}
+\caption{Regression trees indicating primary factors determining reductions in sums of squares of errors in estimation measured by Eq.
+\ref{bias_regression_response} for terminal year recruitment for R+S, R+M, R+Sel and R+q OMs.
+Each node shows the median error (top) and number of observations (bottom) for the corresponding subset.
+Median errors closer to or further from zero are indicated by more green or red polygons, respectively.}\label{R_bias_regtree}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[width = 1.4\textwidth]{sr_bias_plots}
+\end{center}
+\caption{Median relative error of Beverton-Holt SRR parameters ($a$ and $b$) for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
 Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{SR_rel_error}
@@
 \begin{center}
-\includegraphics{M_bias_plots}
-\end{center}
-\caption{Median relative error of median natural mortality for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{M_rel_error}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{mohns_rho_ssb_plots}
-\end{center}
-\caption{Median Mohn's $\rho$ for SSB for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_ssb}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{mohns_rho_F_plots}
-\end{center}
-\caption{Median Mohn's $\rho$ of fishing mortality averaged over all age classes for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_F}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\begin{center}
-\includegraphics{mohns_rho_R_plots}
-\end{center}
-\caption{Median Mohn's $\rho$ of recruitment for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_R}
-\end{figure}
-\end{landscape}
+\includegraphics[width = 1.4\textwidth]{F_mohns_rho_regtree_plots}
+\end{center}
+\caption{Regression trees indicating primary factors determining reductions in sums of squares of errors in transformed Mohn's $\rho$ (Eq.
+\ref{bias_regression_response}) for fishing mortality averaged over all age classes for R+S, R+M, R+Sel and R+q OMs.
+Each node shows the median Mohn's $\rho$ (top) and number of observations (bottom) for the corresponding subset.
+Median Mohn's $\rho$ closer to or further from zero are indicated by more green or red polygons, respectively.}\label{F_mohns_rho_regtree}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[width = 1.4\textwidth]{R_mohns_rho_regtree_plots}
+\end{center}
+\caption{Regression trees indicating primary factors determining reductions in sums of squares of errors in transformed Mohn's $\rho$ (Eq.
+\ref{bias_regression_response}) for recruitment for R+S, R+M, R+Sel and R+q OMs.
+Each node shows the median Mohn's $\rho$ (top) and number of observations (bottom) for the corresponding subset.
+Median Mohn's $\rho$ closer to or further from zero are indicated by more green or red polygons, respectively.}\label{R_mohns_rho_regtree}
+\end{figure}
+\end{landscape}
+
+\subsection*{Referenced Tables}\label{referenced-tables}
+\addcontentsline{toc}{subsection}{Referenced Tables}
 
@@
 \begin{table}
+\caption{For each OM process error type (columns), percent reduction in deviance for logistic regression models fit to indicators of convergence (maximum absolute gradient < $10^{-6}$) with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{convergence_gradient_PRD_table}
+{\input{convergence_gradient_PRD_table}}
+\end{table}
+
+\begin{table}
+\caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
+\ref{bias_regression_response} for the terminal year fully-selected fishing mortality with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_F_PRD_table}
+{\input{bias_F_PRD_table}}
+\end{table}
+
+\begin{table}
+\caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
+\ref{bias_regression_response} for the terminal year recruitment with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_R_PRD_table}
+{\input{bias_R_PRD_table}}
+\end{table}
+
+\clearpage
+
+\subsection*{Further Detailed Results}\label{further-detailed-results}
+\addcontentsline{toc}{subsection}{Further Detailed Results}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[width = 1.4\textwidth]{term_SSB_bias_plots}
+\end{center}
+\caption{Median relative error of terminal year SSB for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{SSB_rel_error}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[width = 1.4\textwidth]{term_F_bias_plots}
+\end{center}
+\caption{Median relative error of terminal year fully-selected fishing mortality for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{F_rel_error}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[width = 1.4\textwidth]{term_R_bias_plots}
+\end{center}
+\caption{Median relative error of terminal year recruitment for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{R_rel_error}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[width = 1.4\textwidth]{M_bias_plots}
+\end{center}
+\caption{Median relative error of median natural mortality for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{M_rel_error}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[width = 1.4\textwidth]{mohns_rho_ssb_plots}
+\end{center}
+\caption{Median Mohn's $\rho$ for SSB for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_ssb}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[width = 1.4\textwidth]{mohns_rho_F_plots}
+\end{center}
+\caption{Median Mohn's $\rho$ of fishing mortality averaged over all age classes for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_F}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[width = 1.4\textwidth]{mohns_rho_R_plots}
+\end{center}
+\caption{Median Mohn's $\rho$ of recruitment for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_R}
+\end{figure}
+\end{landscape}
+
+\begin{table}
 \caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to transformed Mohn's $\rho$ values for each simulation (Eq.
@@
 
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[width = 1.4\textwidth]{type_4_convergence_plots}
+\end{center}
+\caption{Probability of EMs providing hessian-based standard errors with alternative process error (colored points and lines), and median natural mortality (estimated or known) and Beverton-Holt SRR (estimated or not; along x-axis) assumptions when fitted to OMs that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{hessian_SE_convergence}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[width = 1.4\textwidth]{type_3_convergence_plots}
+\end{center}
+\caption{Probability of EMs providing maximum absolute values of gradients less than $10^{-6}$ with alternative process error (colored points and lines), and median natural mortality (estimated or known) and Beverton-Holt SRR (estimated or not; along x-axis) assumptions when fitted to OMs that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
+Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{gradient_convergence}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[height = 0.8\textheight]{sr_aic_plots_rev}
+\end{center}
+\caption{Probability of lowest AIC from logistic regression on the log-standard deviation of the true log(SSB) in each simulation for estimating model with Beverton-Holt SRRs, rather than the otherwise equivalent EM without the SRR.
+Results are conditional on median M is known in the EM and alternative assumptions EMs having the correct process error structure: R and R+S (A), R+Sel (B), R+M (C), or R+q (D), and median M is assumed known in the EM.
+Solid and dashed lines are for OMs with and without temporal contrast in fishing pressure, respectively, and polygons represent 95\% confidence intervals.
+Range of results indicates the range of log-standard deviation of log(SSB) for simulations of the particular OM.}\label{sr_aic}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[width = 1.4\textwidth]{sr_aic_plots}
+\end{center}
+\caption{Estimated probability of lowest AIC from logistic regression on the log-standard deviation of the true log(SSB) in each simulation for estimating model with Beverton-Holt SRRs, rather than the otherwise equivalent EM without the SRR.
+Results are conditional on alternative assumptions for median natural mortality (estimated or known) and on EMs having the correct process error structure: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Rug along x-axis denotes $SD(\log(SSB))$ values for each simulation and polygons represent 95\% confidence intervals.}\label{sr_aic_supp}
+\end{figure}
+\end{landscape}
+
 \end{document}

### 2025-10-01 Project_0/code/manuscript/aic_results_summarize.R

@@ -6,7 +6,8 @@ library(gratia)
 library(Hmisc)
 library(VGAM)
 #modified rpart.plot package to use plotmath in split labels...
-pkgload::load_all("/home/tmiller/FSAM_research/aug_backup/work/rpart.plot")
+pkgload::load_all(file.path(here(),"../../rpart.plot"))
+# pkgload::load_all("/home/tmiller/FSAM_research/aug_backup/work/rpart.plot")
 # pkgload::load_all("c:/work/rpart.plot")
 
 df.oms = list(naa = readRDS(here("Project_0", "inputs", "df.oms.RDS")))
@@ -137,28 +138,42 @@ for(i in names(om.em.rows)) {
 
   
   aic_df[[i]] <- aic_df[[i]] %>%
-    mutate(Obs._Error = recode(obs_error,
+    mutate(OM_Obs._Error = recode(obs_error,
       "L" = "Low",
       "H" = "High"
     ))
+  
   aic_df[[i]] <- aic_df[[i]] %>%
-    mutate(F_History = recode(as.character(Fhist),
-      "H-MSY" = "H->MSY",
-      "MSY" = "MSY"
-    ))
-
+    mutate(OM_F_History = recode(Fhist,
+      "H-MSY"  = "2.5*italic(F)[MSY] %->% italic(F)[MSY]",
+      "MSY" = "italic(F)[MSY]"))
+  
+  aic_df[[i]] <- aic_df[[i]] %>%
+  mutate(EM_SR = recode(as.character(SR_model),
+                        "2" = "None",
+                        "3" = "Estimated"
+  ))
+  
+  aic_df[[i]] <- aic_df[[i]] %>%
+    mutate(EM_M = recode(as.character(M_est),
+                       "TRUE" = "Estimated",
+                       "FALSE" = "Known"
+  ))
+  
   aic_df[[i]] <- aic_df[[i]][sort(names(aic_df[[i]]))]
 
 }
 aic_df <- do.call(rbind.data.frame, aic_df)
 aic_df$EM_assumption <- as.factor(aic_df$EM_assumption)
-aic_df$OM_F_History <- as.factor(aic_df$F_History)
[…]

### 2025-10-01 Project_0/code/manuscript/convergence_results_summarize.R

@@ -5,8 +5,7 @@ library(mgcv)
 library(gratia)
 library(Hmisc)
 #modified rpart.plot package to use plotmath in split labels...
-pkgload::load_all("/home/tmiller/FSAM_research/aug_backup/work/rpart.plot")
-# pkgload::load_all("c:/work/rpart.plot")
+pkgload::load_all(file.path(here(),"../../rpart.plot"))
 
 
 df.oms = list(naa = readRDS(here("Project_0", "inputs", "df.oms.RDS")))
@@ -119,8 +118,8 @@ for(i in names(om.em.rows)) {
     ))
   conv_df[[i]] <- conv_df[[i]] %>%
     mutate(EM_SR = recode(as.character(SR_model),
-      "2" = "Estimated",
-      "3" = "None"
+      "2" = "None",
+      "3" = "Estimated"
     ))
   conv_df[[i]] <- conv_df[[i]] %>%
     mutate(F_History = recode(as.character(Fhist),
@@ -229,7 +228,8 @@ x <- latex(x, file = here("Project_0","manuscript","convergence_PRD_table.tex"),
 library(rpart)
 #modified rpart.plot package to use plotmath in split labels...
 #library(rpart.plot)
-pkgload::load_all("c:/work/rpart.plot")
+pkgload::load_all(file.path(here(),"../../rpart.plot"))
+#pkgload::load_all("c:/work/rpart.plot")
 split.fun <- function(type = "R") {
 	# replace commas with spaces (needed for strwrap)
 	if(!type %in% c("R","R+S")){
@@ -330,11 +330,12 @@ for(type in names(factors)){
   form <- as.formula(paste("conv_txt ~", paste(factors[[type]], collapse = "+")))# (EM_PE + OM_R_SD + EM_M + EM_SR + OM_Obs._Error + OM_F_History)
 	full.trees[[type]] <- rpart(form, data=temp_df, method = "class", control=rpart.control(cp=0, xval = 100))
 }
+anova.palette.sd <- 0.25
 
 cairo_pdf(here("Project_0","manuscript", paste0("convergence_classification_plots.pdf")), width = 30*2/3, height = 20*2/3)
 x <- matrix(c(1,1,2,2,3,3,0,4,4,5,5,0), 2, 6, byrow = TRUE)
 layout.x <- layout(x) 
-par(mar = c(0,0,5,0), oma = c(0,0,0,0))
[…]

### 2025-10-01 Project_0/code/manuscript/mohns_rho_results_summarize.R

@@ -7,7 +7,8 @@ library(rpart)
 library(viridis)
 library(gratia)
 #modified rpart.plot package to use plotmath in split labels...
-pkgload::load_all("/home/tmiller/FSAM_research/aug_backup/work/rpart.plot")
+pkgload::load_all(file.path(here(),"../../rpart.plot"))
+# pkgload::load_all("/home/tmiller/FSAM_research/aug_backup/work/rpart.plot")
 
 
 make_plot_df <- function(om_type = "naa", res = naa_relSR_results) {
@@ -61,8 +62,8 @@ make_plot_df <- function(om_type = "naa", res = naa_relSR_results) {
   ))
   
   df <- df %>% mutate(SR_model = recode(SR_model,
-                                        "2" = "No SR",
-                                        "3" = "SR"
+                                        "2" = "None",
+                                        "3" = "Estimated"
   ))
   ind <- which(df$M_re_cor == "iid" | df$sel_re_cor == "iid" | df$q_re_cor == "iid")
   df$EM_process_error[ind] <- paste0(df$EM_process_error[ind], "(iid)")
@@ -451,11 +452,15 @@ plot.prune(prune(full.trees[["F"]][["R+q"]],c(12,15)), cp = 0.0005, type = "R+q"
 mtext("R+q OMs", side = 3, line = 0, cex = 2)
 dev.off()
 
+par(mfrow = c(1,2))
+plot.prune(full.trees[["R"]][["R+S"]], cp = 0.0001, type = "R", roundint = FALSE, extra = 1, mar = c(0,0,5,0), tweak = 1.2)
+ 
 cairo_pdf(here("Project_0","manuscript", paste0("R_mohns_rho_regtree_plots.pdf")), width = 30*2/3, height = 15*2/3)
 x <- matrix(c(1,0,1,4,2,4,2,5,3,5,3,0), 2, 6)
 layout.x <- layout(x) 
 par(mar = c(0,0,5,0), oma = c(0,4,0,1))
-plot.prune(prune(full.trees[["R"]][["R"]], c(5,14,15)), cp = 0.0001, type = "R", roundint = FALSE, extra = 1, mar = c(0,0,5,0), tweak = 1.2)
[…]

### 2025-12-04 Project_0/manuscript/manuscript_R1.Rmd
saving revisions for project_0 manuscript

@@
 
-State-space models are increasingly used for stock assessment, and evaluations of statistical inferences made from them is needed.
-We simulated 72 operating models that varied fishing pressure and observation error across process errors in recruitment, apparent survival, selectivity, catchability, and/or natural mortality.
-We fit estimating models with different assumptions on the process error source and whether natural mortality and stock-recruit relationship were estimated.
-We measured reliability of estimating models by probability of convergence, accuracy of AIC for correct process error configuration and stock-recruit function, estimation bias, and magnitude of retrospective patterns.
-Across all measures reliability was generally better with lower observation error, contrast in fishing pressure over time, and median natural mortality rate known, but using marginal AIC to distinguishing the stock recruit relationship further required large contrast in spawning biomass and low residual variation in recruitment.
-Marginal AIC accurately distinguished process errors on recruitment, apparent survival, and selectivity.
-These results improve our understanding of when results from state space assessment models, the next-generation of fisheries stock assessment, can be expected to be reliable.
+State-space models have been promoted as the next-generation of fisheries stock assessment and evaluations of their reliability is needed.
+We simulated operating models that varied fishing pressure, magnitude of observation error, and sources of process error.
+For each operating model, we fit a range estimating models with a range of correct and incorrect assumptions.
+We measured reliability of estimating models by convergence rate, accuracy of marginal AIC, estimation bias, and magnitude of retrospective patterns.
+All reliability measures were generally better with lower observation error, contrast in fishing pressure over time, and when median natural mortality rate is known.
+Magnitude of the log-likelihood gradients was not a reliable indicator of convergence.
+AIC can generally distinguish process error type with lower observation error and higher true process error variability.
+Distinguishing the stock recruit relationship with AIC required large contrast in spawning biomass and low recruitment variation, but bias in stock-recruit parameter estimation was prevalent.
+Retrospective patterns were not large for mis-specified models.
+These findings improve our understanding of when results from state space models will be reliable.
 
@@
 Application of state-space models in fisheries stock assessment and management has expanded dramatically within International Council for the Exploration of the Sea (ICES), Canada, and the Northeast US [@nielsenberg14; @cadigan16; @pedersenberg17; @stockmiller21].
-State-space models latent population characteristics as statistical time series with periodic observations that also may have error due to sampling or other sources of measurement error.
-Traditional assessment models may use state-space approaches to account for temporal variability in population characteristics [@methotwetzel13;@legaultrestrepo99], but these models treat the annual parameters as penalized fixed effects parameters where the variance parameters controlling the penalties are assumed known [@thorsonminto15].
+State-space models treat latent population characteristics as statistical time series with periodic observations that also may have error due to sampling or other other measurement properties.
+Traditional assessment models may use state-space approaches to account for temporal variability in population characteristics [@methotwetzel13;@legaultrestrepo99], but these models treat the annual parameters as penalized fixed effect parameters where the variance parameters controlling the penalties are assumed known [@thorsonminto15].
 Modern state-space models can estimate the annually varying parameters as random effects with variance parameters estimated using maximum marginal likelihood or corresponding Bayesian approaches.
-These latter approaches are considered best practice and a recommended for the next generation of stock assessment models [@hoyleetal22; @punt23].
+These latter approaches are considered best practice and are recommended for the next generation of stock assessment models [@hoyleetal22; @punt23].
 
@@
 Incorrect treatment of population attributes as temporally varying [@trijouletetal20; @liljestrandetal24] could lead to misidentification of stock status and biased population estimates, ultimately impacting fisheries management decisions [@legaultpalmer16; @szuwalskietal18; @croninpunt21].
-Furthermore, biological, fishery, and observational processes are often confounded in catch-at-age data, which may adversely affect ability to distinguish between true process variability and observational error [@puntetal14; @stewartmonnahan17; @croninpunt21; @fischetal23; @lietalinreview_a].
+Furthermore, biological, fishery, and observational processes are often confounded in catch-at-age data, which may adversely affect the ability to distinguish between true process variability and observational error [@puntetal14; @stewartmonnahan17; @croninpunt21; @fischetal23; @lietalinreview_a].
 
@@
 Despite mounting efforts, several limitations remain.
-First, confounding processes that can be treated as random effects in the model were not thoroughly examined or tested within a simulation-estimation framework.
+First, confounding processes that can be treated as random effects in the model have not been thoroughly examined or tested within a simulation-estimation framework.
 Second, previous studies relied on operating models conditioned on specific fisheries, limiting their generalizability [@lietalinreview_a; @liljestrandetal24].
@@
 Given the confounding nature of process errors, developing diagnostic tools to detect model misspecification is of great scientific interest and could aid the next generation of stock assessments [@augeretal21].
-We evaluate whether convergence and Akaike Information Criterion (AIC) can correctly determine the source of process error and the existence of a SRR.
+We evaluate whether OM and EM attributes affect rates of convergence and the ability of Akaike Information Criterion (AIC) to correctly determine the source of process error or the existence of a SRR.
 We also evaluate when retrospective patterns occur and the degree of bias in outputs of the assessment model that are important for management.
@@
 
-We used the Woods Hole Assessment Model (WHAM) to configure OMs and EMs in our simulation study [@millerstock20;@stockmiller21].
-WHAM is an R package freely available via a github repository and is built on the Template Model Builder package [@kristensenetal16].
+We used the Woods Hole Assessment Model (WHAM) to configure OMs and EMs in our simulation study [@stockmiller21;@milleretal25].
+WHAM is an R package freely available via a Github repository and is built on the Template Model Builder package [@kristensenetal16].
 For this study we used [version 1.0.6.9000, commit 77bbd94](https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3).
@@
 We completed a simulation study with a number of OMs that can be categorized based on where process error random effects were assumed: recruitment (R, assumed present in all models), apparent survival (denoted R+S), natural mortality (R+M), fleet selectivity (R+Sel), or index catchability (R+q).
-We refer to the (R+S) OMs as modeling apparent survival because on logscale the random effects ($\epsilon_{a,y}$) are additive to the total mortality (F+M) between numbers at age, thus they modify the survival term.
+We refer to the (R+S) OMs as modeling apparent survival because on log-scale the random effects are additive to the total mortality (F+M) between numbers at age, thus they modify the survival term.
 However, as @stockmiller21 note, these random effects can be due to events other than mortality, such as immigration, emigration, missreported catch, and other sources of misspecification.
@@
 In total, we configured 72 OMs with alternative assumptions about the source and magnitude of process errors, magnitude of observation error in indices and age composition data, and contrast in fishing pressure over time.
-We fitted 20 EMs to observations generated from each of 100 simulations where process errors were also simulated.
-Each EM differed in assumptions about the source of process errors, whether natural mortality (or the median for models with process error in natural mortality) was estimated, and whether a Beverton-Holt SRR was estimated within the EM.
+For each OM, we simulated 100 population time series with process errors and, for each time series, simulated observation data sets.
+For each data set, we fitted a number of EMs that differed in assumptions about the source of process errors, whether natural mortality (or the median for models with process error in natural mortality) was estimated, and whether a Beverton-Holt SRR was estimated within the EM.
 Details of each of the OMs and EMs are described below.
 
-We did not use the log-normal bias-correction feature for process errors or observations described by [@stockmiller21] for OMs and EMs to simplify interpretation of the study results [@lietalinreview].
+We did not use the log-normal bias-correction feature for process errors or observations described by [@stockmiller21] for OMs and EMs to simplify interpretation of the study results [@lietal25].
 All code we used to perform the simulation study and summarize results can be found at https://github.com/timjmiller/SSRTWG/tree/main/Project_0/code.
@@
 
+We intended the population demographics and observation types to represent a general NEUS groundfish stock.
 The population consists of 10 age classes, ages 1 to 10+, with the last being a plus group that accumulates ages 10 and older.
@@
 Therefore, steepness and unfished recruitment are also constant over the time period for those OMs [@millerbrooks21].
+We assumed a value of 0.2 for the natural mortality rate in OMs without process errors on natural mortality.
 We specified unfished recruitment equal to $e^{10}$ and $\Fmsy = F_{40\%} = 0.348$, which equates to a steepness of 0.69 and $a=0.60$ and $b = 2.4 \times 10^{-5}$ for the Beverton-Holt parameterization
@@
 (Figure \ref{om_inputs_fig}, bottom right).
-We assumed a value of 0.2 for the natural mortality rate in OMs without process errors on natural mortaliyt and for the median rate for OMs with process errors on natural mortality.
+For OMs with time-varying random effects for natural mortality, steepness is not constant.
+However, we used the same $a$  and $b$ parameters as other OMs, which equates to a steepness and R0 at the median of the time series process for natural mortality.
+Similarly, For OMs with time-varying random effects for fishery selectivity, \Fmsy also varies temporally, so equilibrium conditions for these OMs are defined for mean selectivity parameters.
 
@@
 The magnitude of the overfishing assumptions is based on average estimates of overfishing for NEUS groundfish stocks from @wiedenmannetal19 and similar to the approach in @legaultetal23.
-
-We specified initial population abundance at age at the equilibrium distribution that corresponds to fishing at either $F = 2.5\times \Fmsy$ or $F = \Fmsy$.
+The second scenario represents the ideal situation where the stock is fished at an optimal level, but provides less contrast in stock sizes over time.
+We specified initial population abundance at age at the equilibrium distribution that corresponds to fishing at either $F = 2.5\Fmsy$ or $F = \Fmsy$.
 This implies that, for a deterministic model, the abundance at age would not change from the first year to the next.
-
-For OMs with time-varying random effects for M , steepness is not constant.
-However, we used the same $a$  and $b$ parameters as other OMs, which equates to a steepness and R0 at the median of the time series process for M.
-For OMs with time-varying random effects for fishery selectivity, \Fmsy is also not constant, but since we use the same F history as other OMs, this corresponds to \Fmsy at the mean selectivity parameters.
 
@@
 
-We assumed a single fleet operating year round for catch observations with logistic selectivity for the fleet ($a_{50} = 5$ and slope = 1; Figure \ref{om_inputs_fig}, bottom left).
+We assumed a single fleet operating year round for catch observations with logistic selectivity ($a_{50} = 5$ and slope = 1; Figure \ref{om_inputs_fig}, bottom left).
 This selectivity was used to define \Fmsy for the Beverton-Holt SRR parameters above.
-We assumed a logistic-normal distribution with no correlation on the multivariate normal scale for the age-composition observations for the fleet.
+We assumed a logistic-normal distribution with no correlation on the multivariate normal scale for the corresponding annual age-composition observations.
 
@@
 
-Two time series of fishery-independent surveys in numbers are generated for the entire 40 year period with one occurring in the spring (0.25 of each year) and one in the fall (0.75 of each year).
+Two time series of fishery-independent surveys measured in numbers are generated for the entire 40 year period with one occurring in the spring (0.25 of each year) and one in the fall (0.75 of each year), representing current bottom trawl surveys conducted in the NEUS.
 Catchability of both surveys are assumed to be 0.1.
-Like the fishing fleet, we assumed logistic selectivity for both indices ($a_{50} = 5$ and slope = 1) and a logistic-normal distribution with no correlation on the multivariate normal scale for the age-composition observations.
+Like the fishing fleet, we assumed logistic selectivity for both indices ($a_{50} = 5$ and slope = 1) and a logistic-normal distribution with no correlation on the multivariate normal scale for the annual age-composition observations.
 
@@
 
-The standard deviation for log-aggregate catch was 0.1.
+The standard deviation for log-aggregate catch was 0.1 for all OMs, a common assumption for NEUS stock assessments.
 Two levels of observation error variance (high and low) were specified for indices and all age composition observations (both indices and catch).
@@
 In the high uncertainty specification, the standard deviation for log-aggregate indices was 0.4 and that for the age composition observations was 1.5.
+The low standard deviation for index observations is typical for fish stocks that are consistently sampled across survey stations whereas the high value is typical for more sporadically sampled stocks.
+The standard deviations for the age composition observations were determined from the range of values estimated from WHAM fits to NEUS stocks that assumed the logistic-normal model.
 For all EMs, the standard deviation for log-aggregate observations was assumed known whereas that for the logistic-normal age composition observations was estimated.
@@
 
-For operating models with random effects on recruitment and(or) apparent survival (R, R+S), we assumed marginal standard deviations for recruitment of $\sigma_R \in \{0.5,1.5\}$ and marginal standard deviations for older age classes of $\sigma_{2+} \in \{0,0.25, 0.5\}$.
-The full factorial combination of these process error assumptions (2x3 levels) and scenarios for fishing history (2 levels) and observation error (2 levels) scenarios described above results in 24 different R ($\sigma_{2+} = 0$) and R+S operating models (Table \ref{naa_om_table}).
+For operating models with random effects on recruitment and(or) apparent survival (R, R+S), we assumed marginal standard deviations for recruitment of $\sigma_R \in \{0.5,1.5\}$.
+The marginal standard deviations for apparent survival random effects at older age classes were $\sigma_{2+} \in \{0,0.25, 0.5\}$.
+The full factorial combination of these process error assumptions ($2\times 3$ levels) and scenarios for fishing history (2 levels) and observation error (2 levels) scenarios described above results in 24 different R ($\sigma_{2+} = 0$) and R+S operating models (Table \ref{naa_om_table}).
 
@@
 For R and R+S OMs, fitted EMs had matching process error assumptions as well as R+Sel, R+M, and R+q assumptions without autocorrelation.
-Similarly, For other OM process error categories, we fit EMs with matching process error assumptions as well as other process error types without autocorrelation.
-The maturity at age, weight at age for catch and spawning stock biomass (SSB), and observation error variance of aggregate catch and indices were all assumed known at the true values.
+For other OM process error categories, we fit EMs with correct process error assumptions, the correct process error type but incorrect correlation assumption, and the incorrect process error types without autocorrelation.
+As such, EMs were configured completely correctly for the OM, or they had mis-specification in assumptions of process error autocorrelation, the type of process error, and(or) the SRR (Beverton-Holt or none).
+
+The maturity at age, weight at age for catch and spawning stock biomass (SSB), and observation error standard deviations for aggregate catch and indices were all assumed known at the true values.
 However, the variance parameters for the logistic-normal distributions for age composition observations were estimated in the EMs.
-As such, EMs would either be configured completely correctly for the OM, or there could be mis-specification in assumptions of process error autocorrelation, the type of process error, or the SRR (Beverton-Holt or none).
 
@@
 The first measure of reliability we investigated was frequency of convergence when fitting each EM to the simulated data sets.
-There are various ways to assess convergence of the fit [e.g., @carvalhoetal21; @kapuretal24], but given the importance of estimates of uncertainty when using assessment models in management, we estimated probability of convergence as measured by occurrence of a positive-definite hessian matrix at the optimized negative log-likelihood that could be inverted (i.e., providing hessian-based standard error estimates).
-We also provide results in the Supplementary Materials for convergence defined by the maximum absolute gradient < $1^{-6}$ and the maximum of the absolute gradient values for all fits of a given EM to all simulated data sets from a given OM that produced hessian-based standard errors for all estimated fixed effects.
+There are various ways to assess convergence of the fit [e.g., @carvalhoetal21; @kapuretal24], but given the importance of estimates of uncertainty when using assessment models in management, we estimated probability of convergence as measured by occurrence of a positive-definite Hessian matrix at the optimized negative log-likelihood that could be inverted (i.e., providing Hessian-based standard error estimates).
+We also provide results in the Supplementary Materials for convergence defined by the maximum absolute gradient < $1^{-6}$ and the maximum of the absolute gradient values for all fits of a given EM to all simulated data sets from a given OM that produced Hessian-based standard errors for all estimated fixed effects.
 This provides an indication of how poor the calculated gradients can be, but still presumably converged adequately enough for parameter inferences.
@@
 For a given OM simulated data set, we compared AIC for EMs with different process error assumption conditional on whether median natural mortality rate and the Beverton-Holt SRR were estimated.
-We tabulated the models providing the lowest AIC across simulated data sets.
 Second, we analyzed AIC-based selection between EMs with and without the Beverton-Holt SRR assumed.
 Contrast in fishing pressure and time series with recruitment at low stock size has been shown to improve estimation of SRR parameters [@magnussonhilborn07; @connetal10].
-Our preliminary inspections indicated generally poor performance of AIC in determining the Beverton-Holt model for a given set of OM factors (including contrast in fishing pressure).
-Therefore, we also considered the effect of of the log-standard deviation of the true log(SSB) (similar to the log of the coefficient of variation for SSB) on model selection since simulations with realized SSB producing low and high recruitment would have larger variation in realized SSB.
+Our preliminary inspections indicated generally poor performance of AIC in determining the Beverton-Holt model for a given set of OM factors (including contrast in fishing pressure), even when the EM was configured with the correct process error type.
+Therefore, we conditioned on the EMs having the correct process error assumption and also considered the effect of the log-standard deviation of the true log(SSB) ($\log \text{SD}_{\text{SSB}}$; similar to the log of the coefficient of variation for SSB) on model selection since simulations with realized SSB producing low and high recruitment would have larger variation in realized SSB.
 
@@
 \begin{equation}\label{relerror}
-\text{RE}\left(\theta_i\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
+\text{RE}\left(\theta_j\right) = \frac{\widehat \theta_j - \theta_j}{\theta_j}
 \end{equation}
-from fitting a given estimating model to simulated data set $i$ configured for a given OM where $\widehat \theta_i$ and $\theta_i$ are the estimated and true values for simulation $i$.
+from fitting a given estimating model to simulated data set $j$ configured for a given OM where $\widehat \theta_j$ and $\theta_j$ are the estimated and true values for simulation $j$.
 We analyzed simulation results for estimates of terminal year estimates of SSB and recruitment, Beverton-Holt SRR parameters ($a$ and $b$), and median natural mortality rate.
@@
 \end{equation}
-where $\widehat \theta_{i,j}$ is the estimate for attribute $\theta$ in year $i$ from a model fit using data up to year $j$.
+where $Y$ is last year of the full set of observations and $\widehat \theta_{y,y'}$ is the estimate for attribute $\theta$ in year $y$ from a model fit using data up to year $y'\geq y$.
 
@@
 
-The measures of central tendency and variability of observed values for specific OM and EM attributes (e.g., low or high observation error) that we described above can indicate scenarios that provide better or poorer reliability.
-However, the OM and EM attributes that we investigated are numerous, so we used two methods to summarize the most important factors for differences in results.
+<!-- The measures of central tendency and variability of observed values for specific OM and EM attributes (e.g., low or high observation error) that we described above can indicate scenarios that provide better or poorer reliability.
+-->
+
+Because the OM and EM attributes that we investigated are numerous, we used two methods to summarize the most important factors for differences in results within a given OM process error type.
 The first method was fitting regression models with the response being each of the measures of reliability described above and predictor variables were defined based on OM and EM characteristics [e.g., @mackinnonetal95; @wangetal17; @harwelletal18].
@@
 \begin{equation}\label{bias_regression_response}
-y_i = \log\left[f\left(\widehat \theta_i,\theta_i\right)+1\right]
+y_j = \log\left[f\left(\widehat \theta_j,\theta_j\right)+1\right]
 \end{equation}
@@
 \ref{relerror}) or Mohn's $\rho$ (Eq.
-\ref{mohns_rho}) for simulation $i$, so that values are unbounded.
-For relative errors, $y_i$ is the log-scale error.
-We omitted simulation estimates equal to zero (RE = -1).
-For all regressions we fit separate models with individual factors included, with all factors combined, with including all second order interactions, and including all third order interactions.
+\ref{mohns_rho}) for simulation $j$, so that values are unbounded.
+For relative errors, $y_j$ is the log-scale error.
+We omitted simulations where estimated attributes equal to zero (RE = -1).
+For all regressions we fit separate models with just individual OM and EM factors included, with all factors included, with  all second order interactions, and with all third order interactions.
 For the multinomial regression, we used the \verb|vglm| function from the VGAM package [@yee08;@yee15].
@@
 For AIC selection of process error configuration we provide estimates of the proportions of simulations where each EM type was selected.
-For AIC selection of the SRR, we provide predicted probabilities from logistic regressions as a function of SSB variability for each OM and EM type.
+For AIC selection of the SRR (a binary indicator for each simulated data set), we fit logistic regressions and present resulting predicted probabilities of correctly selecting the SRR as a function of SSB variability ($\log \text{SD}_{\text{SSB}}$ described above).
 We estimated bias as the median of the relative errors across all simulations for a given OM and EM combination.
-We constructed 95\% confidence intervals for the median relative bias, and Mohn's $\rho$ using the binomial distribution approach as in @millerhyun18 and @stockmiller21.
-For each EM we calculated median and 95\% confidence intervals using the same methods as that for relative bias.
+We constructed 95\% confidence intervals for the median relative bias, and Mohn's $\rho$ using the binomial distribution approach [@thompson36] as in @millerhyun18 and @stockmiller21.
 
@@
 
-For probability of convergence, EM process error assumption was the single attribute that resulted in the largest percent reduction in deviance (14-28\%) for all OM process error types other than R+S OMs where the EM M assumption explained the most residual deviance (>11\%; Table \ref{convergence_PRD_table}).
+For probability of convergence, EM process error assumption was the single attribute that resulted in the largest percent reduction in deviance (14-28\%) for all OM process error types other than R+S OMs where the EM median natural mortality rate assumption (estimated or known) explained the most residual deviance (>11\%; Table \ref{convergence_PRD_table}).
 However, including interactions of OM and EM factors also provided large reductions in residual deviance (35-47\%), suggesting successful convergence depended on a combination OM and EM attributes.
@@
 For all trees, branches based on the OM fishing mortality history showed better convergence when the OM included a change in fishing pressure.
-Branches based on whether the Beverton-Holt SRR was assumed or not, showed better convergence when it was not estimated and branches based on whether the EM estimated the median natural mortality rate showed better convergence when it was treated as known.
-For certain R+M and R+Sel OMs, better convergence was also observed when there was lower observation uncertainty.
+Branches based on whether the Beverton-Holt SRR was assumed or not, showed better convergence when it was not estimated and branches based on the median natural mortality rate assumption showed better convergence when it was treated as known.
+For some R+M and R+Sel OMs, better convergence was also observed when there was lower observation uncertainty.
 
-When convergence is defined by a gradient threshold, the primary factor explaining deviance reduction is the same for all OM process error types, but there are some differences in deviance reduction for secondary factors (Table \ref{convergence_gradient_PRD_table}), and probability of convergence, overall, was lower (Figure \ref{conv_gradient_class}).
+When convergence is defined by a gradient threshold, the primary factor explaining deviance reduction is the same that for Hessian-based convergence for all OM process error types, but there are some differences in deviance reduction for secondary factors (Table \ref{convergence_gradient_PRD_table}), and probability of convergence, overall, was lower (Figure \ref{conv_gradient_class}).
 We found a wide range of maximum absolute values of gradients for models that had invertible Hessians (Figure \ref{hess_grad}).
@@
 
-For AIC selection of the correct process error configuration, the magnitude of observation and process error were the attributes that resulted in the largest percent reductions in deviance across OM process error types other than R OMs (Table \ref{AIC_PE_PRD_table}).
-Both variance of apparent survival random effects ($\sigma_{2+}$) and degree of observation error explained the largest reductions for R+S (17-22\%) and R+Sel (8-26\%) OMs, whereas variance of process errors provided the largest reductions in R+M (>9\%) and R+q (>13\%) OMs.
+For AIC selection of the correct process error configuration, the magnitude of observation and process error variation were the attributes that resulted in the largest percent reductions in deviance across OM process error types other than R OMs (Table \ref{AIC_PE_PRD_table}).
+Both sources of variation explained large reductions in deviance for R+S (17-22\%) and R+Sel (8-26\%) OMs, whereas variance of process errors provided the major reductions for R+M (>9\%) and R+q (>13\%) OMs.
 Comparatively, none of the OM or EM attributes explained particularly large reductions in deviance for R OMs, but fishing history, whether a SRR was estimated, and whether median natural mortality was known or estimated provided similar and the largest reductions (approximately 5-6\%).
@@
 Across all OMs, AIC was more accurate for the process error type when process error variability was greater and when observation error was lower.
+For R+S OMs, there was a tendency to select R OMs when observation error was higher and apparent survival variation was lower ($\sigma_{2+} = 0.25$), but accuracy for the process error was otherwise highly accurate.
+Similarly, low accuracy for process error type occurred for R+Sel and R+q OMs only with high observation error and lower process error variability, but high accuracy for R+M OMs required both low observation error and high process error variance.
+This was also required for accurate identification of the correct correlation structure for R+M, R+q, and R+M OMs (Figure \ref{pe_aic}).
 No branches were estimated for classification trees fit to the R OMs, likely because accuracy was high across all simulations (0.94), although inspection of the fine-scale results shows there is some degradation in AIC selection when a SRR and median natural mortality rate are estimated for R OMs with constant fishing pressure and high observation error (Figure \ref{pe_aic}, top left).
@@
 
-Logistic regressions for AIC selection of the Beverton-Holt SRR, showed OM fishing history and variation in SSB ($\log \text{SD}_{\text{SSB}}$) provided substantial reductions in deviance for R+M (>13\%), R+Sel (>26\%), and R+q (>24\%) OMs (Table \ref{AIC_SRR_PRD_table}).
-For R OMs, fishing history provided the largest reduction in deviance (>9\%) whereas none of the attributes individually provided large reductions in deviance for R+S OMs (all <5\%).
-However, inclusion of all attributes provided larger reductions in deviance than the sum of individual contributions for both R (>30\%) and R+S (>18\%) OMs.
+Logistic regressions for AIC selection of the Beverton-Holt SRR, showed OM fishing history and $\log \text{SD}_{\text{SSB}}$ provided substantial reductions in deviance for R+M (>13\%), R+Sel (>26\%), and R+q (>24\%) OMs (Table \ref{AIC_SRR_PRD_table}).
+For R OMs, fishing history provided the largest reduction in deviance (>9\%), whereas none of the attributes individually provided large reductions in deviance for R+S OMs (all <5\%).
+However, inclusion of all attributes provided larger reductions in deviance than the sum of individual contributions for both R (>30\%) and R+S (~19\%) OMs.
+Further fits for R and R+S OMs including all combinations two factors showed those with $\log \text{SD}_{\text{SSB}}$ and recruitment variation provided essentially the same reduction in deviance as the models with all factors.
 For all OM process error types, inclusion of interaction terms provided relatively little reduction in residual deviance.
@@
 Attributes defining the primary branches of classification trees for AIC selection of the SRR assumption were the same as those explaining the largest reductions in deviance for the logistic regression models (Figure \ref{AIC_SRR_class}).
-All branches based on variation in SSB showed better accuracy with larger variability in SSB and all branches based on fishing history showed better accuracy when there was contrast in fishing pressure.
-Branched based on OM observation error or recruitment variability showed better accuracy when they were lower.
+All branches based on $\log \text{SD}_{\text{SSB}}$ showed better accuracy with larger variability in SSB and all branches based on fishing history showed better accuracy when there was contrast in fishing pressure.
+Branched based on OM observation error or recruitment variability (R and R+S OMs) showed better accuracy when they were lower.
 For R OMs, a combination of lower recruitment variability, contrast in fishing pressure, and higher SSB variability produced AIC accuracy over 0.8.
 For R+S OMs, lower recruitment variability and observation error and higher SSB variability produced AIC accuracy of 0.79.
-For R+M, R+Sel, and R+q OM,s accuracy of 0.87 to 0.94 was observed just with higher SSB variability.
+For R+M, R+Sel, and R+q OM,s accuracy of 0.87 to 0.94 was observed with just increased SSB variability.
 
@@
 The attributes producing the largest reductions were the EM assumption for median natural mortality (known or estimated) for R, R+M, R+Sel, and R+q OMs (1-3\%), EM process error type for R+S OMs (4\%) and fishing history for all OM types (1-5\%).
-Including second interactions provided largest reductions in residual deviance (10- 26\%).
-Including third order interactions also provided large further reductions for R, R+S, and R+q OMs (between 5 and 11\%).
-In all regression trees, branches based on fishing history and level of observation error generally showed lower bias in SSB with contrast in fishing and lower observation error (Figure \ref{SSB_bias_regtree}).
-For branches based on treatment of median natural mortality rate, bias was generally lower when it was known rather than estimated.
+Including second order interactions provided the largest reductions in residual deviance (10- 26\%).
+Including third order interactions also provided further reductions for R, R+S, and R+q OMs between 5 and 11\%.
+
+In all regression trees, branches based on fishing history and level of observation error generally showed less bias in SSB with contrast in fishing and lower observation error (Figure \ref{SSB_bias_regtree}).
+For scenarios where there was bias, it was generally positive (over-estimation).
+For branches based on treatment of median natural mortality rate, bias was generally less when it was known rather than estimated.
 For some R+Sel and R+q OMs, less bias in SSB was shown when the EM process error configuration was correct.
@@
 Effects of individual OM and EM factors on regression models were similarly small as measured by reduction in deviance (Tables \ref{bias_F_PRD_table} and \ref{bias_R_PRD_table}).
-Primary branches of regression trees were in most cases identical to those for SSB  (Figures \ref{F_bias_regtree} and \ref{R_bias_regtree}).
+Factors defining the primary branches of regression trees were in most cases identical to those for SSB  (Figures \ref{F_bias_regtree} and \ref{R_bias_regtree}).
 
@@
 
-Regression models for transformed relative errors of estimates of both the Beverton-Holt $a$ and $b$ parameters showed none of the factors explained large percent reductions in deviance (Table \ref{bias_SR_pars_PRD_table}).
-The OM fishing history provided the largest reduction for most OM process error types for both parameters, but percent reductions were less than 5.6\% except for R+Sel OMs where the percent reductions were 11.37\% and 7.97\% for the $a$ and $b$ parameters, respectively and for just the $b$ parameter for R+q OMs (10\%).
+Regression models for log-scale errors of estimates of both the Beverton-Holt $a$ and $b$ parameters showed none of the factors explained large percent reductions in deviance (Table \ref{bias_SR_pars_PRD_table}).
+The OM fishing history provided the largest deviance reduction for most OM process error types for both parameters, but reductions were less than 5.6\% except for R+Sel OMs where the reductions were 11.37\% and 7.97\% for the $a$ and $b$ parameters, respectively and for just the $b$ parameter for R+q OMs (10\%).
 The EM process error assumption provided similar reductions in deviance for both parameters for R OMs.
-Including interactions also did not produce important reductions in deviance (increase of .
+Including interactions also did not produce important reductions in deviance.
 
-For regression trees of log-scale error in Beverton-Holt $a$ and $b$ parameters, lower bias was indicated with contrast in OM fishing pressure for all branches in trees for each OM process error type (Figures \ref{SR_a_bias_regtree} and \ref{SR_b_bias_regtree}).
-For all branches based on recruitment variability in trees for R and R+S OMs, lower bias in both $a$ and $b$ was observed with less recruitment variability.
-For R OMs with contrast in fishing pressure and greater recruitment variability EMs that assumed the incorrect R+M process errors produced lower bias in both $a$ and $b$ than other process error assumptions.
-Across all OMs, there was generally less bias and(or) lower variability in estimation of the $a$ parameter than the $b$ parameter (Figure \ref{SR_rel_error}).
+For regression trees of log-scale errors in Beverton-Holt $a$ and $b$ parameter estimates, less bias was indicated with contrast in OM fishing pressure for all branches in trees for each OM process error type (Figures \ref{SR_a_bias_regtree} and \ref{SR_b_bias_regtree}).
+For all branches based on recruitment variability in trees for R and R+S OMs, less bias in both $a$ and $b$ was observed with less recruitment variability.
+For R OMs with contrast in fishing pressure and greater recruitment variability EMs that assumed the incorrect R+M process errors produced less bias in both $a$ and $b$ than other process error assumptions.
+Across all combinations of OM and EM attributes, some bias was observed for both parameters, but there was generally less bias and(or) lower variability in estimation of the $a$ parameter than the $b$ parameter (Figure \ref{SR_rel_error}).
 
@@
 
-Fitted regression models for log-scale errors in median natural mortality rate showed largest percent reductions in residual deviance for R+S and R_M models (Table \ref{bias_median_M_PRD_table}).
+Fitted regression models for log-scale errors in median natural mortality rate showed largest percent reductions in residual deviance for R+S and R+M models (Table \ref{bias_median_M_PRD_table}).
 The largest reductions for a single attribute was the EM process error assumption (>20\%) and fishing history (>15\%) for R+S OMs.
 Fishing history also provided >10\% reduction for R+M OMs, but reductions for all factors in R, R+Sel, and R+q OMs were relatively low (<6\%).
-Interactions of OM and EM factors also provided substantial further reductions for R+S and R+M OMs (between 8 and 15 \% for second order interactions).
+Interactions of OM and EM factors also provided substantial further reductions for R+S and R+M OMs (between 8 and 15\% for second order interactions).
 
-Regression trees with branches based on fishing history showed lower bias in median natural mortality rate with contrast in fishing pressure and branches based on level of observation error showed lower bias with more precise observations (Figure \ref{med_M_bias_regtree}).
-For R OMs, branches based on EM process error assumption showed lower bias with EMs assuming the correct R and the incorrect R+S assumption.
+Regression trees with branches based on fishing history showed less bias in median natural mortality rate with contrast in fishing pressure and branches based on level of observation error showed less bias with more precise observations (Figure \ref{med_M_bias_regtree}).
+For R OMs, branches based on EM process error assumption showed less bias with EMs assuming the correct R and the incorrect R+S assumption.
 For R+S and R+M OMs, branches based on EM process error showed only the correct EM process error assumption with less bias.
@@
 
-Similarly poor explanatory power of the OM and EM attributes occured when we fit regression models for Mohn's $\rho$ of fishing mortality and recruitment (Tables \ref{mohns_rho_F_PRD_table} and \ref{mohns_rho_R_PRD_table}).
+Similarly, poor explanatory power of the OM and EM attributes occurred when we fit regression models for Mohn's $\rho$ of fishing mortality and recruitment (Tables \ref{mohns_rho_F_PRD_table} and \ref{mohns_rho_R_PRD_table}).
 Regression trees for Mohn's $\rho$ of fishing mortality were similar to those for SSB in that median values of Mohn's $\rho$ were close to zero for most combinations of OM and EM attributes (Figure \ref{F_mohns_rho_regtree}).
@@
 
-## Convergence {-}
+## Assessing convergence {-}
 
@@
 
-Often poor convergence result when parameter estimates are at their bounds [@carvalhoetal21], and this also applies to variance parameters for random effects with state-space assessment models.
+Often poor convergence occurs when parameter estimates are at their bounds [@carvalhoetal21], and this also applies to variance parameters for random effects with state-space assessment models.
 Even when the Hessian is invertible, parameters that are poorly informed will have extremely large variance estimates.
@@
 For example, process error variance parameters that are estimated close to 0 indicates that the random effects are estimated to have little or no variability and removing these process errors is warranted.
-Generally, our results suggest we can expect lower probability of convergence of state-space assessment models when estimating natural mortality or SRRs because of the difficulty distinguishing these parameters from others being estimated in assessment model with data that are typically available.
+Generally, our results suggest we can expect lower probability of convergence of state-space assessment models when estimating natural mortality or SRRs because of the difficulty distinguishing these parameters from others being estimated in assessment model in common scenarios where data quality is less than ideal.
 Our experiments did not aim to emulate the practitioner decision process in developing model configurations (e.g.
@@
 
-## AIC {-}
+## Configuring process error {-}
 
-Of the OM process error configurations we considered, we found AIC to be accurate for selecting models with process errors on recruitment and apparent survival (R and R+S).
-Fitting models to other OMs rarely preferred R+S EMs, and R and R+S EMs were nearly always selected for the matching OMs; a similar result was reported by @liljestrandetal24.
-For other sources of process error, accuracy of AIC was improved when there was larger variability in the process errors and/or lower observation error.
+We found accuracy of marginal AIC for process error type required only low observation error for R, R+S, R+Sel, and R+q OMs.
+R+M OMs further required higher process error variability, but this also improved accuracy for the other OM process errors types when there was higher observation error.
 
-Across all OM process error configurations, AIC performed poorly in identifying that the presence of the Beverton-Holt SRR in the OM unless there was contrast in fishing pressure possibly in combination with other factors such as lower variability in recruitment process errors (in R and R+S models) or greater variation in natural mortality process errors (for R+M OMs, Fig.
-\ref{sr_aic}).
-As such, properly accounting for process error in natural mortality could be important [@lietal24] when evaluating SRRs in state-space models.
-Curiously, we did not find a marked effect of the level of observation error on ability to detect the SRR, but it is possible that AIC would perform better if observations have even lower uncertainty than we considered.
+## Stock recruitment relationships {-}
 
-Although we did not compare models with alternative SRRs (e.g., Ricker and Beverton-Holt), we do not expect AIC to perform any better distinguishing between relationships.
-Our finding that AIC tended to choose simpler recruitment models in most cases contrasts with the noted bias in AIC for more complex models [@shibata76;@katz81;@kassraftery95], but, whereas those findings apply to the much more common comparison of models that are fit to raw and independent observations, here we are comparing state-space models which account for observation error and estimate process errors in latent variables.
+Variation in SSB was the most important factor for using marginal AIC to distinguish the the Beverton-Holt SRR.
+For R+M, R+Sel, and R+q OMs, the SRR was accurately detected when the CV of SSB over the time series was at least 40 to 50\% ($\log \text{SD}_{\text{SSB}}$ = -0.9 to -0.7) regardless of any other OM or EM attributes.
+Detection of the SRR for R and R+S OMs required lower recruitment variability, but this lower level ($\sigma_R = 0.5$) was assumed for all of the other OM process error types.
+Our results assumed that the EM process error configuration was correct, but this may be a strong limation given the ability of AIC to distinguish the process error type in many scenarios.
+
+Although we did not compare models with alternative SRRs (e.g., Ricker and Beverton-Holt), we do not expect AIC to perform any better distinguishing between relationships and may be more difficult than distinguishing from the null model even with larger variability in SSB.
+Our finding that AIC tended to choose simpler recruitment models in many cases contrasts with the noted bias in AIC for more complex models [@shibata76;@katz81;@kassraftery95], but, whereas those findings apply to the much more common comparison of models that are fit to raw and independent observations, here we are comparing state-space models which account for observation error and estimate process errors in latent variables.
 
@@
 
-## Bias {-}
+However, estimation of SRR parameters was only moderately reliable in ideal scenarios of low observation error and contrast in fishing for R+Sel and R+M OMs with large temporal variability in process errors.
+Otherwise, SRR parameter estimation was biased and(or) highly variable.
+We found substantial bias in estimated SRR parameters in R and R+S OMs particularly with high variability in recruitment and apparent survival process errors, suggesting that practitioners should be cautious SRR inferences when fitted assessment models have these properties.
+We only evaluated effects of SSB variability on accuracy of AIC in identifying the SRR, but those results suggests we might find less bias for the SRR parameters in such cases as well.
+Similarly, restricting results to fits that converged may also yield better accuracy of SRR parameter estimation.
+<!-- A proper regression analysis analogous to those we performed may require using an alternative transformation of the estimation errors to measure proximity to 0 irrespective of sign as the response variable (e.g., log of absolute value of relative errors).
+-->
 
-As expected, bias in all parameters and assessment output was generally improved with lower observation error.
-Estimation of SRR parameters was reliable in ideal scenarios of low observation error and contrast in fishing for some R+Sel and R+M OMs, but generally estimation was biased and(or) highly variable.
-We found substantial bias in estimated SRR parameters in R and R+S OMs particularly with high variability in recruitment and apparent survival process errors, suggesting that practitioners should be cautious when fitted assessment models have these properties.
+<!-- However, AIC performed poorly in identifying that the presence of the Beverton-Holt SRR in the OM unless there was contrast in fishing pressure possibly in combination with other factors such as lower variability in recruitment process errors (in R and R+S models) or greater variation in natural mortality process errors (for R+M OMs, Fig.
+\ref{sr_aic}).
+As such, properly accounting for process error in natural mortality could be important [@lietal24] when evaluating SRRs in state-space models.
+Curiously, we did not find a marked effect of the level of observation error on ability to detect the SRR, but it is possible that AIC would perform better if observations have even lower uncertainty than we considered.
+-->
 
-On the other hand, estimation of median natural mortality was reliable in many OM scenarios with contrast in fishing pressure, consistent with @hoenigetal25.
-In some OMs, when EMs estimated the SRR parameters and median natural mortality, bias for those parameters was improved.
-Conversely, for some R+Sel and R+q OMs where there was bias in natural mortality due to high observation error, estimating the SRR reduced the bias in median natural mortality rate.
-However, estimating median natural mortality did cause poor accuracy in SSB estimation in many OMs without contrast in fishing pressure over time and with higher observation error.
-Thus, estimating median natural mortality should be approached with caution in state-space assessment models, particularly given its significant impact on determination of reference point and stock status [@lietal24].
+## Estimating assessment model quantities {-}
 
-## Retrospective patterns {-}
+As expected, bias in parameters, SSB, and other assessment output was generally improved with lower observation error.
+Estimation of median natural mortality was reliable in many OM scenarios with contrast in fishing pressure, consistent with @hoenigetal25.
+However, we found poor accuracy in terminal SSB estimation when estimating median natural mortality in many OMs when there was no contrast in fishing pressure over time and higher observation error.
+Therefore, estimating median natural mortality should be approached with caution in state-space assessment models, particularly given its significant impact on determination of reference point and stock status [@lietal24].
 
-Incorrect EM process error assumptions did not produce strong retrospective patterns for SSB for any OMs regardless of whether median natural mortality or a SRR was estimated, but some weak retrospective patterns occur when observation error was high and there was contrast in fishing pressure.
+## Negligible retrospective patterns {-}
+
+Incorrect EM process error assumptions did not produce strong retrospective patterns for SSB for any OMs regardless of whether median natural mortality or a SRR was estimated although some weak patterns occurred when observation error was high and there was contrast in fishing pressure.
 However, retrospective patterns tended to be more variable for recruitment and were sometimes large even when the EM was correct.
-Therefore, we recommend emphasis on inspection of retrospective patterns primarily for SSB and $F$, but further research on retrospective patterns in other assessment model parameters, management quantities such as biological reference points, and projections may be beneficial [@brookslegault16].
+Therefore, we recommend de-emphasis on inspection of patterns for recruitment, but further research on retrospective patterns in other assessment model parameters, management quantities such as biological reference points, and projections may be beneficial [@brookslegault16].
 
@@
 
-## Analytical approach {-}
+## Summarization approach {-}
 
@@
 The classification and regression trees are generally a good tool for determining the values of the OM and EM attributes that produce better or worse measures of reliability.
-However, determining the combination of attributes that produce the best or worst measures of reliability can be challenging using the trees.
-For example, in the regression tree for median natural mortality rate estimates in R OMs (Figure \ref{median_M_bias_regtree}),  the both of the first branches imply bias is low regardless of OM fishing history, but when OM fishing pressure is constant, results are much better when OM observation error is low (median RE about -6\%) than when OM observation error is high (median RE about 40\%).
-Therefore, inspection of results by all combinations of OM and EM factors (as provided in the supplementary materials) can be important.
+However, determining the combination of attributes that produce the best or worst measures of reliability can be challenging using the trees alone.
+For example, in the regression tree for median natural mortality rate estimates in R OMs (Figure \ref{med_M_bias_regtree}), both of the first branches imply bias is low regardless of OM fishing history, but when OM fishing pressure is constant, results are much better when OM observation error is low (median RE about -6\%) than when OM observation error is high (median RE about 40\%).
+The default pruning of the trees can exclude these lower branches.
+However, inspection of deviance explained by various regression models shows the ~9\% reduction in residual deviance by including second order interaction of all OM and EM factors (Table \ref{bias_median_M_PRD_table}), indicating that the interaction of factors may be important, thereby complimenting the regression tree analysis.
+Higher order interactions of some factors could also provide reductions in deviance and, therefore, inspection of results for each combinations of OM and EM factors, as provided in the Supplementary Materials, can also be important.
 
-## Conclusions {-}
+## Recommendations and conclusions {-}
 
-Our simulation study examined the importance of several factors for reliable inferences from state-space age-structured assessment models.
+Our findings regarding model convergence suggests practitioners using state-space models and maximum marginal likelihood for estimation should not heavily weight the magnitude of the gradient values in determining convergence as long as the maximum absolute values is around 1 or lower.
+Instead, positive-definiteness of the Hessian of the minimized negative log-likelihood should be evaluated.
+
+<!-- @lietal24 recommends exploring all potential sources of process error when building a state-space assessment model and removing those that are estimated to be degenerate (no variation in the random effects).
+-->
+
+Unfortunately, whether the practitioner includes a Beverton-Holt SRR will often depend on biological plausibility of this particular SRR because using AIC to determine its validity required a combination of low recruitment variability, contrast in fishing pressure, large variation in SSB over time, and lower observation error, which applies to a limited number of managed stocks.
+Furthermore, some bias in estimation of the SRR parameters should be expected, which presumably also applies to MSY-based reference points.
+Because bias in terminal SSB and retrospective patterns were indifferent to whether or not the SRR was estimated, the prevalence of bias in SRR parameter estimation, and often better convergence without the SRR, we recommend a sensible default is to exclude a SRR when fitting assessment models, as also suggested by @brooks24.
+
+We found marginal AIC can, in many cases, accurately distinguished models with process errors.
+We saw the best accuracy for models with process errors on recruitment only (R), recruitment and apparent survival (R+S), and recruitment and selectivity (R+Sel), especially with lower observation error.
+However, AIC could also distinguish R+M and R+q process errors when variability of those processes was greater.
+The R+S assumption for process errors is common in applications of WHAM in the NEUS and the SAM assessment framework [@nielsenberg14] in ICES, and we can have some confidence that practitioners are correctly arriving at this assumption over other sources of process error using marginal AIC.
+
+<!-- Our simulation study examined the importance of several factors for reliable inferences from state-space age-structured assessment models.
 Contrast in fishing pressure and lower observation error consistently improved all measures of reliability we examined.
-AIC accurately distinguished models with process errors on recruitment only (R) or on recruitment and apparent survival (R+S).
-Therefore, we expect practitioners will find R+S configurations to provide satisfactory diagnostics across a range of life history and data quality scenarios.
-Accurate AIC selection of the Beverton-Holt SRR generally required a combination of low recruitment variability, contrast in fishing pressure and large variation in SSB over time.
-However, when the Beverton-Holt was correctly assumed, some bias in estimation in at least one of the parameters existed regardless of any of the other OM and EM configurations.
-Because bias in terminal SSB and retrospective patterns were indifferent to whether or not the SRR was estimated, the prevalence of bias in SRR parameter estimation, and often better convergence without the SRR, a sensible default would be to not assume a SRR when fitting assessment models.
+These findings are consistent with other simulation studies of traditional statistical catch at age assessment models and state-space assessment models [@magnussonhilborn07;@fischetal23].
+-->
 
@@
 This work was funded by NOAA Fisheries Northeast Fisheries Science Center.
-We thank Jon Deroba, 2 anonymous reviewers, and the associate editor for helpful comments on earlier versions of this manuscript that markedly improved its clarity.
+We thank Jon Deroba, two anonymous reviewers, and the associate editor for helpful comments on earlier versions of this manuscript that markedly improved its clarity.
 
@@
 \end{center}
-\caption{Classification trees indicating primary factors determining convergence as defined by providing hessian-based standard errors for R, R+S, R+M, R+Sel and R+q OMs.
+\caption{Classification trees indicating primary factors determining convergence as defined by providing Hessian-based standard errors for R, R+S, R+M, R+Sel and R+q OMs.
 Lower or higher convergence rates are indicated by more red or green polygons, respectively}\label{conv_class}
@@
 \begin{table}
-\caption{For each OM process error type (columns), percent reduction in deviance for logistic regression models fit to indicators of convergence (providing hessian-based standard errors) with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{convergence_PRD_table}
+\caption{For each OM process error type (columns), percent reduction in deviance for logistic regression models fit to indicators of convergence (providing Hessian-based standard errors) with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{convergence_PRD_table}
 {\input{convergence_PRD_table}}
@@
 \end{center}
-\caption{The proportion mature at age, weight at age, fleet and index selectivity at age, and Beverton-Holt SRR assumed for the population in all operating models.
-For operating models with random effects on fleet selectivity, this represents the selectivity at the mean of the random effects.}\label{om_inputs_fig}
+\caption{The proportion mature at age, weight at age, fleet and index selectivity at age, and Beverton-Holt SRR assumed for the population in all OMs.
+For OMs with random effects on fleet selectivity, this represents the selectivity at the mean of the random effects.}\label{om_inputs_fig}
 \end{figure}
@@
 \end{center}
-\caption{The maximum of the absolute values of all gradient values for all fits that provided hessian-based standard errors across all simuated data sets of a given OM configuration (A: R and R+S, B: R+M, C: R+Sel, or D: R+q).
+\caption{The maximum of the absolute values of all gradient values for all fits that provided Hessian-based standard errors across all simuated data sets of a given OM configuration (A: R and R+S, B: R+M, C: R+Sel, or D: R+q).
 Results are conditional on EM fits with alternative process error type (colored points and lines), median natural mortality (estimated or known) and recruitment assumptions (Beverton-Holt SRR or not).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{hess_grad}
+Circled values indicate results where the EM process error structure matches that of the OM and vertical lines represent 95\% confidence intervals.}\label{hess_grad}
 \end{figure}
@@
 \end{center}
-\caption{Estimated probability of lowest AIC for EMs assuming alternative process error structures (colored bars) conditional on alternative assumptions for median natural mortality (estimated or known) and Beverton-Holt SRR (estimated or not; along x-axis) when fitted to operating models that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
-Striped bars indicate results where the EM process error structure matches that of the operating model.}\label{pe_aic}
+\caption{Estimated probability of lowest AIC for EMs assuming alternative process error structures (colored bars) conditional on alternative assumptions for median natural mortality (estimated or known) and Beverton-Holt SRR (estimated or not; along x-axis) when fitted to OMs that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
+Striped bars indicate results where the EM process error structure matches that of the OM.}\label{pe_aic}
 \end{figure}
@@
 \end{center}
-\caption{Median relative error of Beverton-Holt SRR parameters ($a$ and $b$) for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{SR_rel_error}
+\caption{Median relative error of Beverton-Holt SRR parameters ($a$ and $b$) for EMs fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the OM and vertical lines represent 95\% confidence intervals.}\label{SR_rel_error}
 \end{figure}
@@
 \begin{table}
-\caption{Distinguishing characteristics of the operating models with random effects on recruitment and apparent survival (R.R+S).
-Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
-Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.}\label{naa_om_table}
+\caption{Distinguishing characteristics of the OMs with random effects on recruitment and apparent survival (R, R+S).
+When observation uncertainty is low, standard deviations for log-normal distributed indices and logistic normal distributed age composition observations are 0.1 and 0.3, respectively, and when it is high, standard deviations are 0.4 and 1.5, respectively.
+Fishing mortality either changes from 2.5$F_{\text{MSY}}$ to $F_{\text{MSY}}$ after year 20 (of 40) or is constant at $F_{\text{MSY}}$ over all years.}\label{naa_om_table}
 {\footnotesize \input{naa_om_tab}}
@@
 \begin{table}
-\caption{Distinguishing characteristics of the operating models with random effects on recruitment and natural mortality (R+M).
-Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
-Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
-For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}\label{M_om_table}
+\caption{Distinguishing characteristics of the OMs with random effects on recruitment and natural mortality (R+M).
+When observation uncertainty is low, standard deviations for log-normal distributed indices and logistic normal distributed age composition observations are 0.1 and 0.3, respectively, and when it is high, standard deviations are 0.4 and 1.5, respectively.
+Fishing mortality either changes from 2.5$F_{\text{MSY}}$ to $F_{\text{MSY}}$ after year 20 (of 40) or is constant at $F_{\text{MSY}}$ over all years.
+For AR1 process errors, $\sigma_M$ is defined for the marginal distribution of the processes.}\label{M_om_table}
 {\input{M_om_tab}}
@@
 \begin{table}
-\caption{Distinguishing characteristics of the operating models with random effects on recruitment and selectivity (R+Sel).
-Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
-Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
-For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}\label{sel_om_table}
+\caption{Distinguishing characteristics of the OMs with random effects on recruitment and selectivity (R+Sel).
+When observation uncertainty is low, standard deviations for log-normal distributed indices and logistic normal distributed age composition observations are 0.1 and 0.3, respectively, and when it is high, standard deviations are 0.4 and 1.5, respectively.
+Fishing mortality either changes from 2.5$F_{\text{MSY}}$ to $F_{\text{MSY}}$ after year 20 (of 40) or is constant at $F_{\text{MSY}}$ over all years.
+For AR1 process errors, $\sigma_{\text{Sel}}$ is defined for the marginal distribution of the processes.}\label{sel_om_table}
 {\input{Sel_om_tab}}
@@
 \begin{table}
-\caption{Distinguishing characteristics of the operating models with random effects on recruitment and catchability (R+q).
-Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices).
-Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.
-For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}\label{q_om_table}
+\caption{Distinguishing characteristics of the OMs with random effects on recruitment and catchability (R+q).
+When observation uncertainty is low, standard deviations for log-normal distributed indices and logistic normal distributed age composition observations are 0.1 and 0.3, respectively, and when it is high, standard deviations are 0.4 and 1.5, respectively.
+Fishing mortality either changes from 2.5$F_{\text{MSY}}$ to $F_{\text{MSY}}$ after year 20 (of 40) or is constant at $F_{\text{MSY}}$ over all years.
+For AR1 process errors, $\sigma_q$ is defined for the marginal distribution of the processes.}\label{q_om_table}
 {\input{q_om_tab}}
@@
 \begin{table}
-\caption{Distinguishing characteristics of the estimating models and operating model process error categories (R, R+S, R+M, R+Sel, R+q) where used.}\label{em_table}
+\caption{Distinguishing characteristics of the EMs and indication (+) of which OM process error categories (R, R+S, R+M, R+Sel, R+q) each EM configuration was fit.}\label{em_table}
 {\scriptsize \input{em_tab}}
@@
 \end{center}
-\caption{Median relative error of terminal year SSB for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{SSB_rel_error}
+\caption{Median relative error of terminal year SSB for EMs fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the OM and vertical lines represent 95\% confidence intervals.}\label{SSB_rel_error}
 \end{figure}
@@
 \end{center}
-\caption{Median relative error of terminal year fully-selected fishing mortality for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{F_rel_error}
+\caption{Median relative error of terminal year fully-selected fishing mortality for EMs fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the OM and vertical lines represent 95\% confidence intervals.}\label{F_rel_error}
 \end{figure}
@@
 \end{center}
-\caption{Median relative error of terminal year recruitment for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{R_rel_error}
+\caption{Median relative error of terminal year recruitment for EMs fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the OM and vertical lines represent 95\% confidence intervals.}\label{R_rel_error}
 \end{figure}
@@
 \end{center}
-\caption{Median relative error of median natural mortality for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{M_rel_error}
+\caption{Median relative error of median natural mortality for EMs fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the OM and vertical lines represent 95\% confidence intervals.}\label{M_rel_error}
 \end{figure}
@@
 \end{center}
-\caption{Median Mohn's $\rho$ for SSB for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_ssb}
+\caption{Median Mohn's $\rho$ for SSB for EMs fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the OM and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_ssb}
 \end{figure}
@@
 \end{center}
-\caption{Median Mohn's $\rho$ of fishing mortality averaged over all age classes for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_F}
+\caption{Median Mohn's $\rho$ of fishing mortality averaged over all age classes for EMs fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the OM and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_F}
 \end{figure}
@@
 \end{center}
-\caption{Median Mohn's $\rho$ of recruitment for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_R}
+\caption{Median Mohn's $\rho$ of recruitment for EMs fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+Circled values indicate results where the EM process error structure matches that of the OM and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_R}
 \end{figure}
@@
 \end{center}
-\caption{Probability of EMs providing hessian-based standard errors with alternative process error (colored points and lines), and median natural mortality (estimated or known) and Beverton-Holt SRR (estimated or not; along x-axis) assumptions when fitted to OMs that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{hessian_SE_convergence}
+\caption{Probability of EMs providing Hessian-based standard errors with alternative process error (colored points and lines), and median natural mortality (estimated or known) and Beverton-Holt SRR (estimated or not; along x-axis) assumptions when fitted to OMs that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
+Circled values indicate results where the EM process error structure matches that of the OM and vertical lines represent 95\% confidence intervals.}\label{hessian_SE_convergence}
 \end{figure}
@@
 \caption{Probability of EMs providing maximum absolute values of gradients less than $10^{-6}$ with alternative process error (colored points and lines), and median natural mortality (estimated or known) and Beverton-Holt SRR (estimated or not; along x-axis) assumptions when fitted to OMs that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
-Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{gradient_convergence}
+Circled values indicate results where the EM process error structure matches that of the OM and vertical lines represent 95\% confidence intervals.}\label{gradient_convergence}
 \end{figure}
@@
 \end{center}
-\caption{Probability of lowest AIC from logistic regression on the log-standard deviation of the true log(SSB) in each simulation for estimating model with Beverton-Holt SRRs, rather than the otherwise equivalent EM without the SRR.
+\caption{Probability of lowest AIC from logistic regression on the log-standard deviation of the true log(SSB) in each simulation for EM with Beverton-Holt SRRs, rather than the otherwise equivalent EM without the SRR.
 Results are conditional on median M is known in the EM and alternative assumptions EMs having the correct process error structure: R and R+S (A), R+Sel (B), R+M (C), or R+q (D), and median M is assumed known in the EM.
@@
 \end{center}
-\caption{Estimated probability of lowest AIC from logistic regression on the log-standard deviation of the true log(SSB) in each simulation for estimating model with Beverton-Holt SRRs, rather than the otherwise equivalent EM without the SRR.
+\caption{Estimated probability of lowest AIC from logistic regression on the log-standard deviation of the true log(SSB) in each simulation for EM with Beverton-Holt SRRs, rather than the otherwise equivalent EM without the SRR.
 Results are conditional on alternative assumptions for median natural mortality (estimated or known) and on EMs having the correct process error structure: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).

### 2025-12-04 Project_0/manuscript/manuscript_R1.tex
saving revisions for project_0 manuscript

@@
 C. Stock\textsuperscript{6} \and John Wiedenmann\textsuperscript{7}}
-\date{30 September, 2025}
+\date{04 December, 2025}
 
@@
 
-State-space models are increasingly used for stock assessment, and
-evaluations of statistical inferences made from them is needed. We
-simulated 72 operating models that varied fishing pressure and
-observation error across process errors in recruitment, apparent
-survival, selectivity, catchability, and/or natural mortality. We fit
-estimating models with different assumptions on the process error source
-and whether natural mortality and stock-recruit relationship were
-estimated. We measured reliability of estimating models by probability
-of convergence, accuracy of AIC for correct process error configuration
-and stock-recruit function, estimation bias, and magnitude of
-retrospective patterns. Across all measures reliability was generally
-better with lower observation error, contrast in fishing pressure over
-time, and median natural mortality rate known, but using marginal AIC to
-distinguishing the stock recruit relationship further required large
-contrast in spawning biomass and low residual variation in recruitment.
-Marginal AIC accurately distinguished process errors on recruitment,
-apparent survival, and selectivity. These results improve our
-understanding of when results from state space assessment models, the
-next-generation of fisheries stock assessment, can be expected to be
-reliable.
+State-space models have been promoted as the next-generation of
+fisheries stock assessment and evaluations of their reliability is
+needed. We simulated operating models that varied fishing pressure,
+magnitude of observation error, and sources of process error. For each
+operating model, we fit a range estimating models with a range of
+correct and incorrect assumptions. We measured reliability of estimating
+models by convergence rate, accuracy of marginal AIC, estimation bias,
+and magnitude of retrospective patterns. All reliability measures were
+generally better with lower observation error, contrast in fishing
+pressure over time, and when median natural mortality rate is known.
+Magnitude of the log-likelihood gradients was not a reliable indicator
+of convergence. AIC can generally distinguish process error type with
+lower observation error and higher true process error variability.
+Distinguishing the stock recruit relationship with AIC required large
+contrast in spawning biomass and low recruitment variation, but bias in
+stock-recruit parameter estimation was prevalent. Retrospective patterns
+were not large for mis-specified models. These findings improve our
+understanding of when results from state space models will be reliable.
 
@@
 and Berg 2014; Cadigan 2016; Pedersen and Berg 2017; Stock and Miller
-2021). State-space models latent population characteristics as
+2021). State-space models treat latent population characteristics as
 statistical time series with periodic observations that also may have
-error due to sampling or other sources of measurement error. Traditional
+error due to sampling or other other measurement properties. Traditional
 assessment models may use state-space approaches to account for temporal
@@
 Methot and Wetzel 2013), but these models treat the annual parameters as
-penalized fixed effects parameters where the variance parameters
+penalized fixed effect parameters where the variance parameters
 controlling the penalties are assumed known (Thorson and Minto 2015).
@@
 marginal likelihood or corresponding Bayesian approaches. These latter
-approaches are considered best practice and a recommended for the next
+approaches are considered best practice and are recommended for the next
 generation of stock assessment models (Hoyle et al. 2022; Punt 2023).
@@
 Furthermore, biological, fishery, and observational processes are often
-confounded in catch-at-age data, which may adversely affect ability to
-distinguish between true process variability and observational error (Li
-et al. In review; Punt et al. 2014; Stewart and Monnahan 2017;
+confounded in catch-at-age data, which may adversely affect the ability
+to distinguish between true process variability and observational error
+(Li et al. In press; Punt et al. 2014; Stewart and Monnahan 2017;
 Cronin-Fine and Punt 2021; Fisch et al. 2023).
@@
 Despite mounting efforts, several limitations remain. First, confounding
-processes that can be treated as random effects in the model were not
-thoroughly examined or tested within a simulation-estimation framework.
-Second, previous studies relied on operating models conditioned on
-specific fisheries, limiting their generalizability (Li et al. In
-review; Liljestrand et al. 2024). In particular, the effects of
+processes that can be treated as random effects in the model have not
+been thoroughly examined or tested within a simulation-estimation
+framework. Second, previous studies relied on operating models
+conditioned on specific fisheries, limiting their generalizability (Li
+et al. In press; Liljestrand et al. 2024). In particular, the effects of
 observation error and underlying fishing history have not been fully
@@
 generation of stock assessments ({Auger-Méthé et al.} 2021). We evaluate
-whether convergence and Akaike Information Criterion (AIC) can correctly
-determine the source of process error and the existence of a SRR. We
-also evaluate when retrospective patterns occur and the degree of bias
-in outputs of the assessment model that are important for management.
+whether OM and EM attributes affect rates of convergence and the ability
+of Akaike Information Criterion (AIC) to correctly determine the source
+of process error or the existence of a SRR. We also evaluate when
+retrospective patterns occur and the degree of bias in outputs of the
+assessment model that are important for management.
 
@@
 We used the Woods Hole Assessment Model (WHAM) to configure OMs and EMs
-in our simulation study (Miller and Stock 2020; Stock and Miller 2021).
-WHAM is an R package freely available via a github repository and is
+in our simulation study (Stock and Miller 2021; Miller et al. 2025).
+WHAM is an R package freely available via a Github repository and is
 built on the Template Model Builder package (Kristensen et al. 2016).
@@
 index catchability (R+q). We refer to the (R+S) OMs as modeling apparent
-survival because on logscale the random effects (\(\epsilon_{a,y}\)) are
-additive to the total mortality (F+M) between numbers at age, thus they
-modify the survival term. However, as Stock and Miller (2021) note,
-these random effects can be due to events other than mortality, such as
-immigration, emigration, missreported catch, and other sources of
-misspecification. For each OM, assumptions about the magnitude of the
-variance of process errors and observations are required and the values
-we used were based on a review of the range of estimates from Northeast
-Unite States (NEUS) assessments using WHAM.
+survival because on log-scale the random effects are additive to the
+total mortality (F+M) between numbers at age, thus they modify the
+survival term. However, as Stock and Miller (2021) note, these random
+effects can be due to events other than mortality, such as immigration,
+emigration, missreported catch, and other sources of misspecification.
+For each OM, assumptions about the magnitude of the variance of process
+errors and observations are required and the values we used were based
+on a review of the range of estimates from Northeast Unite States (NEUS)
+assessments using WHAM.
 
@@
 in indices and age composition data, and contrast in fishing pressure
-over time. We fitted 20 EMs to observations generated from each of 100
-simulations where process errors were also simulated. Each EM differed
-in assumptions about the source of process errors, whether natural
+over time. For each OM, we simulated 100 population time series with
+process errors and, for each time series, simulated observation data
+sets. For each data set, we fitted a number of EMs that differed in
+assumptions about the source of process errors, whether natural
 mortality (or the median for models with process error in natural
@@
 or observations described by (Stock and Miller 2021) for OMs and EMs to
-simplify interpretation of the study results (Li et al. In review). All
-code we used to perform the simulation study and summarize results can
-be found at
+simplify interpretation of the study results (Li et al. 2025). All code
+we used to perform the simulation study and summarize results can be
+found at
 \url{https://github.com/timjmiller/SSRTWG/tree/main/Project_0/code}.
@@
 
-The population consists of 10 age classes, ages 1 to 10+, with the last
-being a plus group that accumulates ages 10 and older. We assume
-spawning occurs annually 1/4 of the way through the year. The maturity
-at age was a logistic curve with \(a_{50}\) = 2.89 and slope = 0.88
-(Figure \ref{om_inputs_fig}, top left).
+We intended the population demographics and observation types to
+represent a general NEUS groundfish stock. The population consists of 10
+age classes, ages 1 to 10+, with the last being a plus group that
+accumulates ages 10 and older. We assume spawning occurs annually 1/4 of
+the way through the year. The maturity at age was a logistic curve with
+\(a_{50}\) = 2.89 and slope = 0.88 (Figure \ref{om_inputs_fig}, top
+left).
 
@@
 Therefore, steepness and unfished recruitment are also constant over the
-time period for those OMs (Miller and Brooks 2021). We specified
-unfished recruitment equal to \(e^{10}\) and
-\(\Fmsy = F_{40\%} = 0.348\), which equates to a steepness of 0.69 and
-\(a=0.60\) and \(b = 2.4 \times 10^{-5}\) for the Beverton-Holt
+time period for those OMs (Miller and Brooks 2021). We assumed a value
+of 0.2 for the natural mortality rate in OMs without process errors on
+natural mortality. We specified unfished recruitment equal to \(e^{10}\)
+and \(\Fmsy = F_{40\%} = 0.348\), which equates to a steepness of 0.69
+and \(a=0.60\) and \(b = 2.4 \times 10^{-5}\) for the Beverton-Holt
 parameterization \[
 N_{1,y} = \frac{a \text{SSB}_{y-1}}{1 + b \text{SSB}_{y-1}} 
-\] (Figure \ref{om_inputs_fig}, bottom right). We assumed a value of 0.2
-for the natural mortality rate in OMs without process errors on natural
-mortaliyt and for the median rate for OMs with process errors on natural
-mortality.
+\] (Figure \ref{om_inputs_fig}, bottom right). For OMs with time-varying
+random effects for natural mortality, steepness is not constant.
+However, we used the same \(a\) and \(b\) parameters as other OMs, which
+equates to a steepness and R0 at the median of the time series process
+for natural mortality. Similarly, For OMs with time-varying random
+effects for fishery selectivity, \Fmsy also varies temporally, so
+equilibrium conditions for these OMs are defined for mean selectivity
+parameters.
 
@@
 for NEUS groundfish stocks from Wiedenmann et al. (2019) and similar to
-the approach in Legault et al. (2023).
-
-We specified initial population abundance at age at the equilibrium
-distribution that corresponds to fishing at either
-\(F = 2.5\times \Fmsy\) or \(F = \Fmsy\). This implies that, for a
-deterministic model, the abundance at age would not change from the
-first year to the next.
-
-For OMs with time-varying random effects for M , steepness is not
-constant. However, we used the same \(a\) and \(b\) parameters as other
-OMs, which equates to a steepness and R0 at the median of the time
-series process for M. For OMs with time-varying random effects for
-fishery selectivity, \Fmsy is also not constant, but since we use the
-same F history as other OMs, this corresponds to \Fmsy at the mean
-selectivity parameters.
+the approach in Legault et al. (2023). The second scenario represents
+the ideal situation where the stock is fished at an optimal level, but
+provides less contrast in stock sizes over time. We specified initial
+population abundance at age at the equilibrium distribution that
+corresponds to fishing at either \(F = 2.5\Fmsy\) or \(F = \Fmsy\). This
+implies that, for a deterministic model, the abundance at age would not
+change from the first year to the next.
 
@@
 We assumed a single fleet operating year round for catch observations
-with logistic selectivity for the fleet (\(a_{50} = 5\) and slope = 1;
-Figure \ref{om_inputs_fig}, bottom left). This selectivity was used to
-define \Fmsy for the Beverton-Holt SRR parameters above. We assumed a
+with logistic selectivity (\(a_{50} = 5\) and slope = 1; Figure
+\ref{om_inputs_fig}, bottom left). This selectivity was used to define
+\Fmsy for the Beverton-Holt SRR parameters above. We assumed a
 logistic-normal distribution with no correlation on the multivariate
-normal scale for the age-composition observations for the fleet.
+normal scale for the corresponding annual age-composition observations.
 
@@
 
-Two time series of fishery-independent surveys in numbers are generated
-for the entire 40 year period with one occurring in the spring (0.25 of
-each year) and one in the fall (0.75 of each year). Catchability of both
-surveys are assumed to be 0.1. Like the fishing fleet, we assumed
-logistic selectivity for both indices (\(a_{50} = 5\) and slope = 1) and
-a logistic-normal distribution with no correlation on the multivariate
-normal scale for the age-composition observations.
+Two time series of fishery-independent surveys measured in numbers are
+generated for the entire 40 year period with one occurring in the spring
+(0.25 of each year) and one in the fall (0.75 of each year),
+representing current bottom trawl surveys conducted in the NEUS.
+Catchability of both surveys are assumed to be 0.1. Like the fishing
+fleet, we assumed logistic selectivity for both indices (\(a_{50} = 5\)
+and slope = 1) and a logistic-normal distribution with no correlation on
+the multivariate normal scale for the annual age-composition
+observations.
 
@@
 
-The standard deviation for log-aggregate catch was 0.1. Two levels of
-observation error variance (high and low) were specified for indices and
-all age composition observations (both indices and catch). The low
-uncertainty specification assumed a standard deviation of 0.1 for both
-series of log-aggregate index observations, and the standard deviation
-of the logistic-normal for age composition observations was 0.3. In the
-high uncertainty specification, the standard deviation for log-aggregate
+The standard deviation for log-aggregate catch was 0.1 for all OMs, a
+common assumption for NEUS stock assessments. Two levels of observation
+error variance (high and low) were specified for indices and all age
+composition observations (both indices and catch). The low uncertainty
+specification assumed a standard deviation of 0.1 for both series of
+log-aggregate index observations, and the standard deviation of the
+logistic-normal for age composition observations was 0.3. In the high
+uncertainty specification, the standard deviation for log-aggregate
 indices was 0.4 and that for the age composition observations was 1.5.
-For all EMs, the standard deviation for log-aggregate observations was
-assumed known whereas that for the logistic-normal age composition
-observations was estimated.
+The low standard deviation for index observations is typical for fish
+stocks that are consistently sampled across survey stations whereas the
+high value is typical for more sporadically sampled stocks. The standard
+deviations for the age composition observations were determined from the
+range of values estimated from WHAM fits to NEUS stocks that assumed the
+logistic-normal model. For all EMs, the standard deviation for
+log-aggregate observations was assumed known whereas that for the
+logistic-normal age composition observations was estimated.
 
@@
 survival (R, R+S), we assumed marginal standard deviations for
-recruitment of \(\sigma_R \in \{0.5,1.5\}\) and marginal standard
-deviations for older age classes of \(\sigma_{2+} \in \{0,0.25, 0.5\}\).
-The full factorial combination of these process error assumptions (2x3
-levels) and scenarios for fishing history (2 levels) and observation
-error (2 levels) scenarios described above results in 24 different R
+recruitment of \(\sigma_R \in \{0.5,1.5\}\). The marginal standard
+deviations for apparent survival random effects at older age classes
+were \(\sigma_{2+} \in \{0,0.25, 0.5\}\). The full factorial combination
+of these process error assumptions (\(2\times 3\) levels) and scenarios
+for fishing history (2 levels) and observation error (2 levels)
+scenarios described above results in 24 different R
 (\(\sigma_{2+} = 0\)) and R+S operating models (Table
@@
 EMs had matching process error assumptions as well as R+Sel, R+M, and
-R+q assumptions without autocorrelation. Similarly, For other OM process
-error categories, we fit EMs with matching process error assumptions as
-well as other process error types without autocorrelation. The maturity
-at age, weight at age for catch and spawning stock biomass (SSB), and
-observation error variance of aggregate catch and indices were all
-assumed known at the true values. However, the variance parameters for
-the logistic-normal distributions for age composition observations were
-estimated in the EMs. As such, EMs would either be configured completely
-correctly for the OM, or there could be mis-specification in assumptions
-of process error autocorrelation, the type of process error, or the SRR
-(Beverton-Holt or none).
+R+q assumptions without autocorrelation. For other OM process error
+categories, we fit EMs with correct process error assumptions, the
+correct process error type but incorrect correlation assumption, and the
+incorrect process error types without autocorrelation. As such, EMs were
+configured completely correctly for the OM, or they had
+mis-specification in assumptions of process error autocorrelation, the
+type of process error, and(or) the SRR (Beverton-Holt or none).
+
+The maturity at age, weight at age for catch and spawning stock biomass
+(SSB), and observation error standard deviations for aggregate catch and
+indices were all assumed known at the true values. However, the variance
+parameters for the logistic-normal distributions for age composition
+observations were estimated in the EMs.
 
@@
 probability of convergence as measured by occurrence of a
-positive-definite hessian matrix at the optimized negative
-log-likelihood that could be inverted (i.e., providing hessian-based
+positive-definite Hessian matrix at the optimized negative
+log-likelihood that could be inverted (i.e., providing Hessian-based
 standard error estimates). We also provide results in the Supplementary
@@
 for all fits of a given EM to all simulated data sets from a given OM
-that produced hessian-based standard errors for all estimated fixed
+that produced Hessian-based standard errors for all estimated fixed
 effects. This provides an indication of how poor the calculated
@@
 assumption conditional on whether median natural mortality rate and the
-Beverton-Holt SRR were estimated. We tabulated the models providing the
-lowest AIC across simulated data sets. Second, we analyzed AIC-based
+Beverton-Holt SRR were estimated. Second, we analyzed AIC-based
 selection between EMs with and without the Beverton-Holt SRR assumed.
@@
 the Beverton-Holt model for a given set of OM factors (including
-contrast in fishing pressure). Therefore, we also considered the effect
-of of the log-standard deviation of the true log(SSB) (similar to the
-log of the coefficient of variation for SSB) on model selection since
-simulations with realized SSB producing low and high recruitment would
-have larger variation in realized SSB.
+contrast in fishing pressure), even when the EM was configured with the
+correct process error type. Therefore, we conditioned on the EMs having
+the correct process error assumption and also considered the effect of
+the log-standard deviation of the true log(SSB)
+(\(\log \text{SD}_{\text{SSB}}\); similar to the log of the coefficient
+of variation for SSB) on model selection since simulations with realized
+SSB producing low and high recruitment would have larger variation in
+realized SSB.
 
@@
 relative error \begin{equation}\label{relerror}
-\text{RE}\left(\theta_i\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
+\text{RE}\left(\theta_j\right) = \frac{\widehat \theta_j - \theta_j}{\theta_j}
 \end{equation} from fitting a given estimating model to simulated data
-set \(i\) configured for a given OM where \(\widehat \theta_i\) and
-\(\theta_i\) are the estimated and true values for simulation \(i\). We
+set \(j\) configured for a given OM where \(\widehat \theta_j\) and
+\(\theta_j\) are the estimated and true values for simulation \(j\). We
 analyzed simulation results for estimates of terminal year estimates of
@@
 \rho\left(\theta\right) = \frac{1}{P}\sum^P_{p=1} \frac{\widehat \theta_{Y-p,Y-p} - \widehat\theta_{Y-p,Y}}{\widehat\theta_{Y-p,Y}}
-\end{equation} where \(\widehat \theta_{i,j}\) is the estimate for
-attribute \(\theta\) in year \(i\) from a model fit using data up to
-year \(j\).
+\end{equation} where \(Y\) is last year of the full set of observations
+and \(\widehat \theta_{y,y'}\) is the estimate for attribute \(\theta\)
+in year \(y\) from a model fit using data up to year \(y'\geq y\).
 
@@
 
-The measures of central tendency and variability of observed values for
-specific OM and EM attributes (e.g., low or high observation error) that
-we described above can indicate scenarios that provide better or poorer
-reliability. However, the OM and EM attributes that we investigated are
-numerous, so we used two methods to summarize the most important factors
-for differences in results. The first method was fitting regression
-models with the response being each of the measures of reliability
-described above and predictor variables were defined based on OM and EM
-characteristics (e.g., MacKinnon et al. 1995; Wang et al. 2017; Harwell
-et al. 2018). For the binary indicators of convergence and AIC-based
-selection of a SRR, we performed logistic regressions. For indicators of
-AIC-based selection of EM process error type (multiple categories) we
-performed multinomial regressions. For other measures of reliability we
-fit linear regression models to transformed responses. Because relative
-errors (Eq. \ref{relerror}) and Mohn's \(\rho\) for the various
-parameters are bounded below at -1, we used a transformation of these
-values \begin{equation}\label{bias_regression_response}
-y_i = \log\left[f\left(\widehat \theta_i,\theta_i\right)+1\right]
+Because the OM and EM attributes that we investigated are numerous, we
+used two methods to summarize the most important factors for differences
+in results within a given OM process error type. The first method was
+fitting regression models with the response being each of the measures
+of reliability described above and predictor variables were defined
+based on OM and EM characteristics (e.g., MacKinnon et al. 1995; Wang et
+al. 2017; Harwell et al. 2018). For the binary indicators of convergence
+and AIC-based selection of a SRR, we performed logistic regressions. For
+indicators of AIC-based selection of EM process error type (multiple
+categories) we performed multinomial regressions. For other measures of
+reliability we fit linear regression models to transformed responses.
+Because relative errors (Eq. \ref{relerror}) and Mohn's \(\rho\) for the
+various parameters are bounded below at -1, we used a transformation of
+these values \begin{equation}\label{bias_regression_response}
+y_j = \log\left[f\left(\widehat \theta_j,\theta_j\right)+1\right]
 \end{equation} where \(f\) is either the relative error (Eq.
 \ref{relerror}) or Mohn's \(\rho\) (Eq. \ref{mohns_rho}) for simulation
-\(i\), so that values are unbounded. For relative errors, \(y_i\) is the
-log-scale error. We omitted simulation estimates equal to zero (RE =
--1). For all regressions we fit separate models with individual factors
-included, with all factors combined, with including all second order
-interactions, and including all third order interactions. For the
-multinomial regression, we used the \verb|vglm| function from the VGAM
-package (Yee 2008; Yee 2015). We tabulated percent reduction in residual
-deviance for each of regression fits. We did not perform formal
+\(j\), so that values are unbounded. For relative errors, \(y_j\) is the
+log-scale error. We omitted simulations where estimated attributes equal
+to zero (RE = -1). For all regressions we fit separate models with just
+individual OM and EM factors included, with all factors included, with
+all second order interactions, and with all third order interactions.
+For the multinomial regression, we used the \verb|vglm| function from
+the VGAM package (Yee 2008; Yee 2015). We tabulated percent reduction in
+residual deviance for each of regression fits. We did not perform formal
 statistical analyses of effects of OM and EM attributes on results
@@
 the proportions of simulations where each EM type was selected. For AIC
-selection of the SRR, we provide predicted probabilities from logistic
-regressions as a function of SSB variability for each OM and EM type. We
+selection of the SRR (a binary indicator for each simulated data set),
+we fit logistic regressions and present resulting predicted
+probabilities of correctly selecting the SRR as a function of SSB
+variability (\(\log \text{SD}_{\text{SSB}}\) described above). We
 estimated bias as the median of the relative errors across all
@@
 confidence intervals for the median relative bias, and Mohn's \(\rho\)
-using the binomial distribution approach as in Miller and Hyun (2018)
-and Stock and Miller (2021). For each EM we calculated median and 95\%
-confidence intervals using the same methods as that for relative bias.
+using the binomial distribution approach (Thompson 1936) as in Miller
+and Hyun (2018) and Stock and Miller (2021).
 
@@
 deviance (14-28\%) for all OM process error types other than R+S OMs
-where the EM M assumption explained the most residual deviance
-(\textgreater11\%; Table \ref{convergence_PRD_table}). However,
-including interactions of OM and EM factors also provided large
-reductions in residual deviance (35-47\%), suggesting successful
-convergence depended on a combination OM and EM attributes.
+where the EM median natural mortality rate assumption (estimated or
+known) explained the most residual deviance (\textgreater11\%; Table
+\ref{convergence_PRD_table}). However, including interactions of OM and
+EM factors also provided large reductions in residual deviance
+(35-47\%), suggesting successful convergence depended on a combination
+OM and EM attributes.
 
@@
 showed better convergence when it was not estimated and branches based
-on whether the EM estimated the median natural mortality rate showed
-better convergence when it was treated as known. For certain R+M and
-R+Sel OMs, better convergence was also observed when there was lower
-observation uncertainty.
+on the median natural mortality rate assumption showed better
+convergence when it was treated as known. For some R+M and R+Sel OMs,
+better convergence was also observed when there was lower observation
+uncertainty.
 
 When convergence is defined by a gradient threshold, the primary factor
-explaining deviance reduction is the same for all OM process error
-types, but there are some differences in deviance reduction for
-secondary factors (Table \ref{convergence_gradient_PRD_table}), and
-probability of convergence, overall, was lower (Figure
-\ref{conv_gradient_class}). We found a wide range of maximum absolute
-values of gradients for models that had invertible Hessians (Figure
-\ref{hess_grad}). The largest value observed for a given EM and OM
-combination was typically \(<10^{-3}\), but many converged models had
-values greater than 1. For many OMs, EMs that assumed the correct
-process error type and did not estimate median natural mortality or the
-Beverton-Holt SRR produced the lowest gradient values.
+explaining deviance reduction is the same that for Hessian-based
+convergence for all OM process error types, but there are some
+differences in deviance reduction for secondary factors (Table
+\ref{convergence_gradient_PRD_table}), and probability of convergence,
+overall, was lower (Figure \ref{conv_gradient_class}). We found a wide
+range of maximum absolute values of gradients for models that had
+invertible Hessians (Figure \ref{hess_grad}). The largest value observed
+for a given EM and OM combination was typically \(<10^{-3}\), but many
+converged models had values greater than 1. For many OMs, EMs that
+assumed the correct process error type and did not estimate median
+natural mortality or the Beverton-Holt SRR produced the lowest gradient
+values.
 
@@
 For AIC selection of the correct process error configuration, the
-magnitude of observation and process error were the attributes that
-resulted in the largest percent reductions in deviance across OM process
-error types other than R OMs (Table \ref{AIC_PE_PRD_table}). Both
-variance of apparent survival random effects (\(\sigma_{2+}\)) and
-degree of observation error explained the largest reductions for R+S
+magnitude of observation and process error variation were the attributes
+that resulted in the largest percent reductions in deviance across OM
+process error types other than R OMs (Table \ref{AIC_PE_PRD_table}).
+Both sources of variation explained large reductions in deviance for R+S
 (17-22\%) and R+Sel (8-26\%) OMs, whereas variance of process errors
-provided the largest reductions in R+M (\textgreater9\%) and R+q
+provided the major reductions for R+M (\textgreater9\%) and R+q
 (\textgreater13\%) OMs. Comparatively, none of the OM or EM attributes
@@
 all OMs, AIC was more accurate for the process error type when process
-error variability was greater and when observation error was lower. No
-branches were estimated for classification trees fit to the R OMs,
-likely because accuracy was high across all simulations (0.94), although
-inspection of the fine-scale results shows there is some degradation in
-AIC selection when a SRR and median natural mortality rate are estimated
-for R OMs with constant fishing pressure and high observation error
-(Figure \ref{pe_aic}, top left).
+error variability was greater and when observation error was lower. For
+R+S OMs, there was a tendency to select R OMs when observation error was
+higher and apparent survival variation was lower
+(\(\sigma_{2+} = 0.25\)), but accuracy for the process error was
+otherwise highly accurate. Similarly, low accuracy for process error
+type occurred for R+Sel and R+q OMs only with high observation error and
+lower process error variability, but high accuracy for R+M OMs required
+both low observation error and high process error variance. This was
+also required for accurate identification of the correct correlation
+structure for R+M, R+q, and R+M OMs (Figure \ref{pe_aic}). No branches
+were estimated for classification trees fit to the R OMs, likely because
+accuracy was high across all simulations (0.94), although inspection of
+the fine-scale results shows there is some degradation in AIC selection
+when a SRR and median natural mortality rate are estimated for R OMs
+with constant fishing pressure and high observation error (Figure
+\ref{pe_aic}, top left).
 
@@
 Logistic regressions for AIC selection of the Beverton-Holt SRR, showed
-OM fishing history and variation in SSB
-(\(\log \text{SD}_{\text{SSB}}\)) provided substantial reductions in
-deviance for R+M (\textgreater13\%), R+Sel (\textgreater26\%), and R+q
-(\textgreater24\%) OMs (Table \ref{AIC_SRR_PRD_table}). For R OMs,
-fishing history provided the largest reduction in deviance
-(\textgreater9\%) whereas none of the attributes individually provided
-large reductions in deviance for R+S OMs (all \textless5\%). However,
-inclusion of all attributes provided larger reductions in deviance than
-the sum of individual contributions for both R (\textgreater30\%) and
-R+S (\textgreater18\%) OMs. For all OM process error types, inclusion of
-interaction terms provided relatively little reduction in residual
-deviance.
+OM fishing history and \(\log \text{SD}_{\text{SSB}}\) provided
+substantial reductions in deviance for R+M (\textgreater13\%), R+Sel
+(\textgreater26\%), and R+q (\textgreater24\%) OMs (Table
+\ref{AIC_SRR_PRD_table}). For R OMs, fishing history provided the
+largest reduction in deviance (\textgreater9\%), whereas none of the
+attributes individually provided large reductions in deviance for R+S
+OMs (all \textless5\%). However, inclusion of all attributes provided
+larger reductions in deviance than the sum of individual contributions
+for both R (\textgreater30\%) and R+S (\textasciitilde19\%) OMs. Further
+fits for R and R+S OMs including all combinations two factors showed
+those with \(\log \text{SD}_{\text{SSB}}\) and recruitment variation
+provided essentially the same reduction in deviance as the models with
+all factors. For all OM process error types, inclusion of interaction
+terms provided relatively little reduction in residual deviance.
 
@@
 largest reductions in deviance for the logistic regression models
-(Figure \ref{AIC_SRR_class}). All branches based on variation in SSB
-showed better accuracy with larger variability in SSB and all branches
-based on fishing history showed better accuracy when there was contrast
-in fishing pressure. Branched based on OM observation error or
-recruitment variability showed better accuracy when they were lower. For
-R OMs, a combination of lower recruitment variability, contrast in
-fishing pressure, and higher SSB variability produced AIC accuracy over
-0.8. For R+S OMs, lower recruitment variability and observation error
-and higher SSB variability produced AIC accuracy of 0.79. For R+M,
-R+Sel, and R+q OM,s accuracy of 0.87 to 0.94 was observed just with
-higher SSB variability.
+(Figure \ref{AIC_SRR_class}). All branches based on
+\(\log \text{SD}_{\text{SSB}}\) showed better accuracy with larger
+variability in SSB and all branches based on fishing history showed
+better accuracy when there was contrast in fishing pressure. Branched
+based on OM observation error or recruitment variability (R and R+S OMs)
+showed better accuracy when they were lower. For R OMs, a combination of
+lower recruitment variability, contrast in fishing pressure, and higher
+SSB variability produced AIC accuracy over 0.8. For R+S OMs, lower
+recruitment variability and observation error and higher SSB variability
+produced AIC accuracy of 0.79. For R+M, R+Sel, and R+q OM,s accuracy of
+0.87 to 0.94 was observed with just increased SSB variability.
 
@@
 for R+S OMs (4\%) and fishing history for all OM types (1-5\%).
-Including second interactions provided largest reductions in residual
-deviance (10- 26\%). Including third order interactions also provided
-large further reductions for R, R+S, and R+q OMs (between 5 and 11\%).
+Including second order interactions provided the largest reductions in
+residual deviance (10- 26\%). Including third order interactions also
+provided further reductions for R, R+S, and R+q OMs between 5 and 11\%.
+
 In all regression trees, branches based on fishing history and level of
-observation error generally showed lower bias in SSB with contrast in
+observation error generally showed less bias in SSB with contrast in
 fishing and lower observation error (Figure \ref{SSB_bias_regtree}). For
-branches based on treatment of median natural mortality rate, bias was
-generally lower when it was known rather than estimated. For some R+Sel
-and R+q OMs, less bias in SSB was shown when the EM process error
-configuration was correct.
+scenarios where there was bias, it was generally positive
+(over-estimation). For branches based on treatment of median natural
+mortality rate, bias was generally less when it was known rather than
+estimated. For some R+Sel and R+q OMs, less bias in SSB was shown when
+the EM process error configuration was correct.
 
@@
 reduction in deviance (Tables \ref{bias_F_PRD_table} and
-\ref{bias_R_PRD_table}). Primary branches of regression trees were in
-most cases identical to those for SSB (Figures \ref{F_bias_regtree} and
-\ref{R_bias_regtree}).
+\ref{bias_R_PRD_table}). Factors defining the primary branches of
+regression trees were in most cases identical to those for SSB (Figures
+\ref{F_bias_regtree} and \ref{R_bias_regtree}).
 
@@
 
-Regression models for transformed relative errors of estimates of both
-the Beverton-Holt \(a\) and \(b\) parameters showed none of the factors
+Regression models for log-scale errors of estimates of both the
+Beverton-Holt \(a\) and \(b\) parameters showed none of the factors
 explained large percent reductions in deviance (Table
 \ref{bias_SR_pars_PRD_table}). The OM fishing history provided the
-largest reduction for most OM process error types for both parameters,
-but percent reductions were less than 5.6\% except for R+Sel OMs where
-the percent reductions were 11.37\% and 7.97\% for the \(a\) and \(b\)
+largest deviance reduction for most OM process error types for both
+parameters, but reductions were less than 5.6\% except for R+Sel OMs
+where the reductions were 11.37\% and 7.97\% for the \(a\) and \(b\)
 parameters, respectively and for just the \(b\) parameter for R+q OMs
@@
 deviance for both parameters for R OMs. Including interactions also did
-not produce important reductions in deviance (increase of .
-
-For regression trees of log-scale error in Beverton-Holt \(a\) and \(b\)
-parameters, lower bias was indicated with contrast in OM fishing
-pressure for all branches in trees for each OM process error type
-(Figures \ref{SR_a_bias_regtree} and \ref{SR_b_bias_regtree}). For all
-branches based on recruitment variability in trees for R and R+S OMs,
-lower bias in both \(a\) and \(b\) was observed with less recruitment
-variability. For R OMs with contrast in fishing pressure and greater
-recruitment variability EMs that assumed the incorrect R+M process
-errors produced lower bias in both \(a\) and \(b\) than other process
-error assumptions. Across all OMs, there was generally less bias and(or)
-lower variability in estimation of the \(a\) parameter than the \(b\)
-parameter (Figure \ref{SR_rel_error}).
+not produce important reductions in deviance.
+
+For regression trees of log-scale errors in Beverton-Holt \(a\) and
+\(b\) parameter estimates, less bias was indicated with contrast in OM
+fishing pressure for all branches in trees for each OM process error
+type (Figures \ref{SR_a_bias_regtree} and \ref{SR_b_bias_regtree}). For
+all branches based on recruitment variability in trees for R and R+S
+OMs, less bias in both \(a\) and \(b\) was observed with less
+recruitment variability. For R OMs with contrast in fishing pressure and
+greater recruitment variability EMs that assumed the incorrect R+M
+process errors produced less bias in both \(a\) and \(b\) than other
+process error assumptions. Across all combinations of OM and EM
+attributes, some bias was observed for both parameters, but there was
+generally less bias and(or) lower variability in estimation of the \(a\)
+parameter than the \(b\) parameter (Figure \ref{SR_rel_error}).
 
@@
 mortality rate showed largest percent reductions in residual deviance
-for R+S and R\_M models (Table \ref{bias_median_M_PRD_table}). The
+for R+S and R+M models (Table \ref{bias_median_M_PRD_table}). The
 largest reductions for a single attribute was the EM process error
@@
 provided substantial further reductions for R+S and R+M OMs (between 8
-and 15 \% for second order interactions).
-
-Regression trees with branches based on fishing history showed lower
-bias in median natural mortality rate with contrast in fishing pressure
-and branches based on level of observation error showed lower bias with
-more precise observations (Figure \ref{med_M_bias_regtree}). For R OMs,
-branches based on EM process error assumption showed lower bias with EMs
+and 15\% for second order interactions).
+
+Regression trees with branches based on fishing history showed less bias
+in median natural mortality rate with contrast in fishing pressure and
+branches based on level of observation error showed less bias with more
+precise observations (Figure \ref{med_M_bias_regtree}). For R OMs,
+branches based on EM process error assumption showed less bias with EMs
 assuming the correct R and the incorrect R+S assumption. For R+S and R+M
@@
 
-Similarly poor explanatory power of the OM and EM attributes occured
+Similarly, poor explanatory power of the OM and EM attributes occurred
 when we fit regression models for Mohn's \(\rho\) of fishing mortality
@@
 
-\subsection*{Convergence}\label{convergence-1}
-\addcontentsline{toc}{subsection}{Convergence}
+\subsection*{Assessing convergence}\label{assessing-convergence}
+\addcontentsline{toc}{subsection}{Assessing convergence}
 
@@
 
-Often poor convergence result when parameter estimates are at their
+Often poor convergence occurs when parameter estimates are at their
 bounds (Carvalho et al. 2021), and this also applies to variance
@@
 because of the difficulty distinguishing these parameters from others
-being estimated in assessment model with data that are typically
-available. Our experiments did not aim to emulate the practitioner
-decision process in developing model configurations (e.g.~removing a
-source of process error and refitting the model when process error
-variance parameters were estimated close to 0). Evaluating the efficacy
-of such a decision process when applying EMs might be important in
-closed loop simulations (e.g.~MSE) aimed at quantifying management
-performance.
+being estimated in assessment model in common scenarios where data
+quality is less than ideal. Our experiments did not aim to emulate the
+practitioner decision process in developing model configurations
+(e.g.~removing a source of process error and refitting the model when
+process error variance parameters were estimated close to 0). Evaluating
+the efficacy of such a decision process when applying EMs might be
+important in closed loop simulations (e.g.~MSE) aimed at quantifying
+management performance.
 
@@
 
-\subsection*{AIC}\label{aic}
-\addcontentsline{toc}{subsection}{AIC}
-
-Of the OM process error configurations we considered, we found AIC to be
-accurate for selecting models with process errors on recruitment and
-apparent survival (R and R+S). Fitting models to other OMs rarely
-preferred R+S EMs, and R and R+S EMs were nearly always selected for the
-matching OMs; a similar result was reported by Liljestrand et al.
-(2024). For other sources of process error, accuracy of AIC was improved
-when there was larger variability in the process errors and/or lower
+\subsection*{Configuring process error}\label{configuring-process-error}
+\addcontentsline{toc}{subsection}{Configuring process error}
+
+We found accuracy of marginal AIC for process error type required only
+low observation error for R, R+S, R+Sel, and R+q OMs. R+M OMs further
+required higher process error variability, but this also improved
+accuracy for the other OM process errors types when there was higher
 observation error.
 
-Across all OM process error configurations, AIC performed poorly in
-identifying that the presence of the Beverton-Holt SRR in the OM unless
-there was contrast in fishing pressure possibly in combination with
-other factors such as lower variability in recruitment process errors
-(in R and R+S models) or greater variation in natural mortality process
-errors (for R+M OMs, Fig. \ref{sr_aic}). As such, properly accounting
-for process error in natural mortality could be important (Li et al.
-2024) when evaluating SRRs in state-space models. Curiously, we did not
-find a marked effect of the level of observation error on ability to
-detect the SRR, but it is possible that AIC would perform better if
-observations have even lower uncertainty than we considered.
+\subsection*{Stock recruitment
+relationships}\label{stock-recruitment-relationships}
+\addcontentsline{toc}{subsection}{Stock recruitment relationships}
+
+Variation in SSB was the most important factor for using marginal AIC to
+distinguish the the Beverton-Holt SRR. For R+M, R+Sel, and R+q OMs, the
+SRR was accurately detected when the CV of SSB over the time series was
+at least 40 to 50\% (\(\log \text{SD}_{\text{SSB}}\) = -0.9 to -0.7)
+regardless of any other OM or EM attributes. Detection of the SRR for R
+and R+S OMs required lower recruitment variability, but this lower level
+(\(\sigma_R = 0.5\)) was assumed for all of the other OM process error
+types. Our results assumed that the EM process error configuration was
+correct, but this may be a strong limation given the ability of AIC to
+distinguish the process error type in many scenarios.
 
@@
 and Beverton-Holt), we do not expect AIC to perform any better
-distinguishing between relationships. Our finding that AIC tended to
-choose simpler recruitment models in most cases contrasts with the noted
-bias in AIC for more complex models (Shibata 1976; Katz 1981; Kass and
-Raftery 1995), but, whereas those findings apply to the much more common
-comparison of models that are fit to raw and independent observations,
-here we are comparing state-space models which account for observation
-error and estimate process errors in latent variables.
+distinguishing between relationships and may be more difficult than
+distinguishing from the null model even with larger variability in SSB.
+Our finding that AIC tended to choose simpler recruitment models in many
+cases contrasts with the noted bias in AIC for more complex models
+(Shibata 1976; Katz 1981; Kass and Raftery 1995), but, whereas those
+findings apply to the much more common comparison of models that are fit
+to raw and independent observations, here we are comparing state-space
+models which account for observation error and estimate process errors
+in latent variables.
 
@@
 
-\subsection*{Bias}\label{bias-2}
-\addcontentsline{toc}{subsection}{Bias}
-
-As expected, bias in all parameters and assessment output was generally
-improved with lower observation error. Estimation of SRR parameters was
-reliable in ideal scenarios of low observation error and contrast in
-fishing for some R+Sel and R+M OMs, but generally estimation was biased
-and(or) highly variable. We found substantial bias in estimated SRR
-parameters in R and R+S OMs particularly with high variability in
-recruitment and apparent survival process errors, suggesting that
-practitioners should be cautious when fitted assessment models have
-these properties.
-
-On the other hand, estimation of median natural mortality was reliable
-in many OM scenarios with contrast in fishing pressure, consistent with
-Hoenig et al. (2025). In some OMs, when EMs estimated the SRR parameters
-and median natural mortality, bias for those parameters was improved.
-Conversely, for some R+Sel and R+q OMs where there was bias in natural
-mortality due to high observation error, estimating the SRR reduced the
-bias in median natural mortality rate. However, estimating median
-natural mortality did cause poor accuracy in SSB estimation in many OMs
-without contrast in fishing pressure over time and with higher
-observation error. Thus, estimating median natural mortality should be
-approached with caution in state-space assessment models, particularly
-given its significant impact on determination of reference point and
-stock status (Li et al. 2024).
-
-\subsection*{Retrospective patterns}\label{retrospective-patterns}
-\addcontentsline{toc}{subsection}{Retrospective patterns}
+However, estimation of SRR parameters was only moderately reliable in
+ideal scenarios of low observation error and contrast in fishing for
+R+Sel and R+M OMs with large temporal variability in process errors.
+Otherwise, SRR parameter estimation was biased and(or) highly variable.
+We found substantial bias in estimated SRR parameters in R and R+S OMs
+particularly with high variability in recruitment and apparent survival
+process errors, suggesting that practitioners should be cautious SRR
+inferences when fitted assessment models have these properties. We only
+evaluated effects of SSB variability on accuracy of AIC in identifying
+the SRR, but those results suggests we might find less bias for the SRR
+parameters in such cases as well. Similarly, restricting results to fits
+that converged may also yield better accuracy of SRR parameter
+estimation.
+
+\subsection*{Estimating assessment model
+quantities}\label{estimating-assessment-model-quantities}
+\addcontentsline{toc}{subsection}{Estimating assessment model
+quantities}
+
+As expected, bias in parameters, SSB, and other assessment output was
+generally improved with lower observation error. Estimation of median
+natural mortality was reliable in many OM scenarios with contrast in
+fishing pressure, consistent with Hoenig et al. (2025). However, we
+found poor accuracy in terminal SSB estimation when estimating median
+natural mortality in many OMs when there was no contrast in fishing
+pressure over time and higher observation error. Therefore, estimating
+median natural mortality should be approached with caution in
+state-space assessment models, particularly given its significant impact
+on determination of reference point and stock status (Li et al. 2024).
+
+\subsection*{Negligible retrospective
+patterns}\label{negligible-retrospective-patterns}
+\addcontentsline{toc}{subsection}{Negligible retrospective patterns}
 
@@
 retrospective patterns for SSB for any OMs regardless of whether median
-natural mortality or a SRR was estimated, but some weak retrospective
-patterns occur when observation error was high and there was contrast in
+natural mortality or a SRR was estimated although some weak patterns
+occurred when observation error was high and there was contrast in
 fishing pressure. However, retrospective patterns tended to be more
 variable for recruitment and were sometimes large even when the EM was
-correct. Therefore, we recommend emphasis on inspection of retrospective
-patterns primarily for SSB and \(F\), but further research on
-retrospective patterns in other assessment model parameters, management
-quantities such as biological reference points, and projections may be
-beneficial (Brooks and Legault 2016).
+correct. Therefore, we recommend de-emphasis on inspection of patterns
+for recruitment, but further research on retrospective patterns in other
+assessment model parameters, management quantities such as biological
+reference points, and projections may be beneficial (Brooks and Legault
+2016).
 
@@
 
-\subsection*{Analytical approach}\label{analytical-approach}
-\addcontentsline{toc}{subsection}{Analytical approach}
+\subsection*{Summarization approach}\label{summarization-approach}
+\addcontentsline{toc}{subsection}{Summarization approach}
 
@@
 However, determining the combination of attributes that produce the best
-or worst measures of reliability can be challenging using the trees. For
-example, in the regression tree for median natural mortality rate
-estimates in R OMs (Figure \ref{median_M_bias_regtree}), the both of the
+or worst measures of reliability can be challenging using the trees
+alone. For example, in the regression tree for median natural mortality
+rate estimates in R OMs (Figure \ref{med_M_bias_regtree}), both of the
 first branches imply bias is low regardless of OM fishing history, but
@@
 observation error is low (median RE about -6\%) than when OM observation
-error is high (median RE about 40\%). Therefore, inspection of results
-by all combinations of OM and EM factors (as provided in the
-supplementary materials) can be important.
-
-\subsection*{Conclusions}\label{conclusions}
-\addcontentsline{toc}{subsection}{Conclusions}
-
-Our simulation study examined the importance of several factors for
-reliable inferences from state-space age-structured assessment models.
-Contrast in fishing pressure and lower observation error consistently
-improved all measures of reliability we examined. AIC accurately
-distinguished models with process errors on recruitment only (R) or on
-recruitment and apparent survival (R+S). Therefore, we expect
-practitioners will find R+S configurations to provide satisfactory
-diagnostics across a range of life history and data quality scenarios.
-Accurate AIC selection of the Beverton-Holt SRR generally required a
-combination of low recruitment variability, contrast in fishing pressure
-and large variation in SSB over time. However, when the Beverton-Holt
-was correctly assumed, some bias in estimation in at least one of the
-parameters existed regardless of any of the other OM and EM
-configurations. Because bias in terminal SSB and retrospective patterns
-were indifferent to whether or not the SRR was estimated, the prevalence
-of bias in SRR parameter estimation, and often better convergence
-without the SRR, a sensible default would be to not assume a SRR when
-fitting assessment models.
+error is high (median RE about 40\%). The default pruning of the trees
+can exclude these lower branches. However, inspection of deviance
+explained by various regression models shows the \textasciitilde9\%
+reduction in residual deviance by including second order interaction of
+all OM and EM factors (Table \ref{bias_median_M_PRD_table}), indicating
+that the interaction of factors may be important, thereby complimenting
+the regression tree analysis. Higher order interactions of some factors
+could also provide reductions in deviance and, therefore, inspection of
+results for each combinations of OM and EM factors, as provided in the
+Supplementary Materials, can also be important.
+
+\subsection*{Recommendations and
+conclusions}\label{recommendations-and-conclusions}
+\addcontentsline{toc}{subsection}{Recommendations and conclusions}
+
+Our findings regarding model convergence suggests practitioners using
+state-space models and maximum marginal likelihood for estimation should
+not heavily weight the magnitude of the gradient values in determining
+convergence as long as the maximum absolute values is around 1 or lower.
+Instead, positive-definiteness of the Hessian of the minimized negative
+log-likelihood should be evaluated.
+
+Unfortunately, whether the practitioner includes a Beverton-Holt SRR
+will often depend on biological plausibility of this particular SRR
+because using AIC to determine its validity required a combination of
+low recruitment variability, contrast in fishing pressure, large
+variation in SSB over time, and lower observation error, which applies
+to a limited number of managed stocks. Furthermore, some bias in
+estimation of the SRR parameters should be expected, which presumably
+also applies to MSY-based reference points. Because bias in terminal SSB
+and retrospective patterns were indifferent to whether or not the SRR
+was estimated, the prevalence of bias in SRR parameter estimation, and
+often better convergence without the SRR, we recommend a sensible
+default is to exclude a SRR when fitting assessment models, as also
+suggested by Brooks (2024).
+
+We found marginal AIC can, in many cases, accurately distinguished
+models with process errors. We saw the best accuracy for models with
+process errors on recruitment only (R), recruitment and apparent
+survival (R+S), and recruitment and selectivity (R+Sel), especially with
+lower observation error. However, AIC could also distinguish R+M and R+q
+process errors when variability of those processes was greater. The R+S
+assumption for process errors is common in applications of WHAM in the
+NEUS and the SAM assessment framework (Nielsen and Berg 2014) in ICES,
+and we can have some confidence that practitioners are correctly
+arriving at this assumption over other sources of process error using
+marginal AIC.
 
@@
 This work was funded by NOAA Fisheries Northeast Fisheries Science
-Center. We thank Jon Deroba, 2 anonymous reviewers, and the associate
+Center. We thank Jon Deroba, two anonymous reviewers, and the associate
 editor for helpful comments on earlier versions of this manuscript that
@@
 Fisheries and Aquatic Sciences.
+
+\bibitem[\citeproctext]{ref-brooks24}
+Brooks, E.N. 2024. Pragmatic approaches to modeling recruitment in
+fisheries stock assessment: A perspective. Fisheries Research
+\textbf{270}: 106896.
+doi:\href{https://doi.org/10.1016/j.fishres.2023.106896}{10.1016/j.fishres.2023.106896}.
 
@@
 Li, C., Deroba, J.J., Berger, A.M., Goethel, D.R., Langseth, B.J.,
-Schueller, A.M., and Miller, T.J. In review. Random effects on
+Schueller, A.M., and Miller, T.J. In press. Random effects on
 numbers-at-age transitions implictly account for movement dynamics and
@@
 
-\bibitem[\citeproctext]{ref-lietalinreview}
-Li, C., Deroba, J.J., Miller, T.J., Legault, C.M., and Perretti, C. In
-review. Guidance on bias-correction of log-normal random effects and
+\bibitem[\citeproctext]{ref-lietal25}
+Li, C., Deroba, J.J., Miller, T.J., Legault, C.M., and Perretti, C.
+2025. Guidance on bias-correction of log-normal random effects and
 observations in state-space assessment models. Canadian Journal of
 Fisheries and Aquatic Sciences.
+doi:\href{https://doi.org/10.1139/cjfas-2025-0093}{10.1139/cjfas-2025-0093}.
 
@@
 doi:\href{https://doi.org/10.1111/faf.12534}{10.1111/faf.12534}.
+
+\bibitem[\citeproctext]{ref-milleretal25}
+Miller, T.J., Curti, K.L., and Hansell, A.C. 2025. Space for WHAM: A
+multi-region, multi-stock generalization of the woods hole assessment
+model with an application to black sea bass. Canadian Journal of
+Fisheries and Aquatic Sciences \textbf{82}: 1--26.
+doi:\href{https://doi.org/10.1139/cjfas-2025-0097}{10.1139/cjfas-2025-0097}.
 
@@
 
-\bibitem[\citeproctext]{ref-millerstock20}
-Miller, T.J., and Stock, B.C. 2020. The {Woods Hole Assessment Model}
-({WHAM}). Available from \url{https://timjmiller.github.io/wham/}.
-
 \bibitem[\citeproctext]{ref-mohn99}
@@
 
+\bibitem[\citeproctext]{ref-thompson36}
+Thompson, W.R. 1936. On confidence ranges for the median and other
+expectation distributions for populations of unknown distribution form.
+Annals of Mathematical Statistics \textbf{7}(3): 122--128.
+doi:\href{https://doi.org/10.1214/aoms/1177732502}{10.1214/aoms/1177732502}.
+
 \bibitem[\citeproctext]{ref-thorsonminto15}

### 2025-12-04 Project_0/code/manuscript/aic_SRR_summarize.R

@@ -3,31 +3,42 @@ library(dplyr)
 library(tidyr)
 library(Hmisc)
 #modified rpart.plot package to use plotmath in split labels...
-pkgload::load_all("/home/tmiller/FSAM_research/aug_backup/work/rpart.plot")
+pkgload::load_all(file.path(here(),"../../rpart.plot"))
+#pkgload::load_all("/home/tmiller/FSAM_research/aug_backup/work/rpart.plot")
 # pkgload::load_all("c:/work/rpart.plot")
 
-aic_fn <- function(x, all_aic, df.oms, df.ems, rec_mod, M_est) {
+aic_fn <- function(x, all_aic, df.oms, df.ems, rec_mod, M_est, EM_PE_is_correct = TRUE) {
   if(!is.null(df.oms$NAA_sig)){
     if(is.na(df.oms$NAA_sig[x])) re_config <- "rec"
     else re_config <- "rec+1"  
-    em_ind <- which(df.ems$SR_model == rec_mod & df.ems$M_est == M_est & df.ems$re_config == re_config)
+    
+    if(EM_PE_is_correct) em_ind <- which(df.ems$SR_model == rec_mod & df.ems$M_est == M_est & df.ems$re_config == re_config)
+    else em_ind <- which(df.ems$SR_model == rec_mod & df.ems$M_est == M_est)
   }
   if(!is.null(df.oms$M_sig)){
     if(df.oms$M_cor[x] == 0) re_config <- "iid"
     else re_config <- "ar1_y"  
-    em_ind <- which(df.ems$SR_model == rec_mod & df.ems$M_est == M_est & df.ems$M_re_cor == re_config)
+    
+    if(EM_PE_is_correct) em_ind <- which(df.ems$SR_model == rec_mod & df.ems$M_est == M_est & df.ems$M_re_cor == re_config)
+    else em_ind <- which(df.ems$SR_model == rec_mod & df.ems$M_est == M_est)
   }
   if(!is.null(df.oms$Sel_sig)){
     if(df.oms$Sel_cor[x] == 0) re_config <- "iid"
     else re_config <- "ar1_y"  
-    em_ind <- which(df.ems$SR_model == rec_mod & df.ems$M_est == M_est & df.ems$sel_re_cor == re_config)
+    
[…]

### 2025-12-04 Project_0/code/manuscript/aic_results_summarize.R

@@ -227,9 +227,9 @@ rnames <- gsub("NAA SD", "$\\sigma_{2+}$ ", rnames, fixed = TRUE)
 rnames <- gsub("R SD", "$\\sigma_R$", rnames, fixed = TRUE)
 rnames <- gsub("q sig", "$\\sigma_q$", rnames, fixed = TRUE)
 rnames <- gsub("M sig", "$\\sigma_M$", rnames, fixed = TRUE)
-rnames <- gsub("Sel sig", "$\\sigma_{Sel}$", rnames, fixed = TRUE)
-rnames <- gsub("M cor", "$\\rho_R$", rnames, fixed = TRUE)
-rnames <- gsub("Sel cor", "$\\rho_{Sel}$", rnames, fixed = TRUE)
+rnames <- gsub("Sel sig", "$\\sigma_{\\text{Sel}}$", rnames, fixed = TRUE)
+rnames <- gsub("M cor", "$\\rho_M$", rnames, fixed = TRUE)
+rnames <- gsub("Sel cor", "$\\rho_{\\text{Sel}}$", rnames, fixed = TRUE)
 rnames <- gsub("q cor", "$\\rho_q$", rnames, fixed = TRUE)
 rownames(PRD.table) <- c(rnames, "All factors", "+ All Two Way", "+ All Three Way")
 for(i in c("R","R+S","R+M","R+Sel","R+q")) PRD.table[,i] <- c(dev.tables[["PRD"]][[i]][match(All.facs, names(dev.tables[["PRD"]][[i]]))], dev.tables[["PRD"]][[i]][paste0("all",c("",2:3))])

### 2025-12-04 Project_0/code/manuscript/om_em_description_tables.R

@@ -8,11 +8,12 @@ df.oms = readRDS(file.path(here(),"Project_0","inputs", "df.oms.RDS"))
 om_inputs = readRDS(file.path(here(),"Project_0","inputs", "NAA_om_inputs.RDS"))
 
 naa_om_tab <- df.oms
-naa_om_tab$Model <- paste0("$\\text{NAA}_{", 1:24, "}$")
+#naa_om_tab$Model <- paste0("$\\text{NAA}_{", 1:24, "}$")
+naa_om_tab$Model <- 1:24
 naa_om_tab <- naa_om_tab %>%
   mutate(obs_error = recode(obs_error,
-    "L" = "Index SD = 0.1, Age composition SD = 0.3",
-    "H" = "Index SD = 0.4, Age composition SD = 1.5"
+    "L" = "Low",
+    "H" = "High"
   ))
 naa_om_tab <- naa_om_tab %>%
   mutate(Fhist = recode(Fhist,
@@ -34,11 +35,12 @@ df.oms = readRDS(file.path(here(),"Project_0","inputs", "df.M.oms.RDS"))
 om_inputs = readRDS(file.path(here(),"Project_0","inputs", "M_om_inputs.RDS"))
 
 om_tab <- df.oms
-om_tab$Model <- paste0("$M_{", 1:16, "}$")
+#om_tab$Model <- paste0("$M_{", 1:16, "}$")
+om_tab$Model <- 1:16
 om_tab <- om_tab %>%
   mutate(obs_error = recode(obs_error,
-    "L" = "Index SD = 0.1, Age composition SD = 0.3",
-    "H" = "Index SD = 0.4, Age composition SD = 1.5"
+                            "L" = "Low",
+                            "H" = "High"
   ))
 om_tab <- om_tab %>%
   mutate(Fhist = recode(Fhist,
@@ -61,11 +63,12 @@ df.oms = readRDS(file.path(here(),"Project_0","inputs", "df.Sel.oms.RDS"))
 om_inputs = readRDS(file.path(here(),"Project_0","inputs", "Sel_om_inputs.RDS"))
 
 om_tab <- df.oms
-om_tab$Model <- paste0("$\\text{Sel}_{", 1:16, "}$")
+#om_tab$Model <- paste0("$\\text{Sel}_{", 1:16, "}$")
+om_tab$Model <- 1:16
 om_tab <- om_tab %>%
[…]

### 2025-12-19 Ecov_study/mortality/manuscript/manuscript_R1.Rmd
revision for project 0 complete, save work for M-Ecov revision

--- 2025-08-08 Ecov_study/mortality/manuscript/manuscript.Rmd
@@
 
-Our analyses used the Woods Hole Assessment Model (WHAM) to construct both OMs and EMs [@millerstock20; @stockmiller21; @milleretal_inreview].
+Our analyses used the Woods Hole Assessment Model (WHAM) to construct both OMs and estimation models (EMs) [@millerstock20; @stockmiller21; @milleretal_inreview].
 The WHAM package has been used extensively to configure OMs and EMs for several other simulation studies [@stocketal21;@legaultetal23; @lietal24; @brittenetal_inreview; @lietal_inreview_a] and is used to assess many commercially important stocks in the Northeast U.S.
@@
 
-For each simulated data set we fit a set of 12 estimating models (EMs).
+For each simulated data set we fit a set of 12 EMs.
 The factors that distinguish the estimating models, also described in detail below, include source of population process error type (3 levels) whether (median) $M$ was estimated or assumed known (2 levels), and whether the effect of the environmental covariate on $M$ was estimated or not (2 levels).
@@
 \begin{center}
+\includegraphics[width = 1.4\textwidth]{convergence_classification_plots}
+\end{center}
+\caption{Classification trees indicating primary factors determining convergence as defined by providing Hessian-based standard errors for R, R+S, and R+M OMs.
+Nodes denote percent convergence (top) and number of fits (bottom)  for the corresponding subset.
+Lower or higher convergence rates are indicated by more red or green polygons, respectively}\label{conv_class}
+\end{figure}
+\end{landscape}
+
+\pagebreak
+
+\begin{table}
+\caption{For each OM process error source (columns), percent reduction in deviance for logistic regression models fit to indicators of convergence (providing Hessian-based standard errors) with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{convergence_PRD_table}
+{\input{convergence_PRD_table}}
+\end{table}
+
+\begin{table}
+\caption{For each OM process error source (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
+\ref{bias_regression_response} for the median natural mortality rate parameter with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_median_M_converged_PRD_table}
+{\input{bias_mean_M_converged_PRD_table}}
+\end{table}
+
+\begin{table}
+\caption{For each OM process error source (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
+\ref{bias_regression_response} for the the standard error of the MLE for median natural mortality rate ($\widehat {SE}\left(\widehat \beta_M\right)$) with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_median_M_se_PRD_table}
+{\input{bias_mean_M_se_PRD_table}}
+\end{table}
+
+\begin{table}
+\caption{For each OM process error source (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
+\ref{bias_regression_response} for the covariate effect on natural mortality ($\beta_E$) with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_ecov_beta_converged_PRD_table}
+{\input{bias_ecov_beta_converged_PRD_table}}
+\end{table}
+
+\begin{table}
+\caption{For each OM process error source (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
+\ref{bias_regression_response} for the the standard error of the MLE for covariate effect on natural mortality  ($\widehat {SE}\left(\widehat \beta_E\right)$) with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_median_M_se_PRD_table}
+{\input{bias_ecov_beta_se_PRD_table}}
+\end{table}
+
+\begin{table}
+\caption{For each OM process error source (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
+\ref{bias_regression_response} for the terminal year SSB with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_SSB_PRD_table}
+{\input{bias_SSB_PRD_table}}
+\end{table}
+
+\pagebreak
+
+# Supplemental Materials {-}
+
+\setcounter{figure}{0}
+\renewcommand\thefigure{S\arabic{figure}}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width = \textwidth]{om_input_plots_figure}
+\end{center}
+\caption{The proportion mature at age, weight at age, fleet and index selectivity at age, and Beverton-Holt stock-recruit relationship assumed for the population in all operating models.
+For operating models with random effects on fleet selectivity, this represents the selectivity at the mean of the random effects.}\label{om_inputs_fig}
+\end{figure}
+
+\begin{landscape}
+\begin{figure}
+\caption{Example simulations of environmental covariate latent processes and observations with different levels of observation error, and different assumptions about variability of the latent process.}\label{om_ecov_example}
+\begin{center}
+\includegraphics[height = \textheight]{Ecov_true_obs_example}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Example simulations of annual natural mortality rates that may be a function of a temporally varying environmental covariate and autoregressive random effects.}\label{M_example}
+\begin{center}
+\includegraphics[height = \textheight]{M_example}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{table}
+\caption{For each OM process error source (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
+\ref{bias_regression_response} for the median natural mortality rate parameter with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_median_M_complete_PRD_table}
+{\input{bias_mean_M_complete_PRD_table}}
+\end{table}
+
+\begin{table}
+\caption{For each OM process error source (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
+\ref{bias_regression_response} for the covariate effect on natural mortality ($\beta_E$) with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_ecov_beta_complete_PRD_table}
+{\input{bias_ecov_beta_complete_PRD_table}}
+\end{table}
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
 \includegraphics{convergence_main}
@@
 All OMs had low observation error and contrast in fishing mortality.}\label{terminal_ssb_bias}
-\end{figure}
-\end{landscape}
-
-# Supplemental Materials {-}
-
-\setcounter{figure}{0}
-\renewcommand\thefigure{S\arabic{figure}}
-
-\begin{figure}
-\begin{center}
-\includegraphics[width = \textwidth]{om_input_plots_figure}
-\end{center}
-\caption{The proportion mature at age, weight at age, fleet and index selectivity at age, and Beverton-Holt stock-recruit relationship assumed for the population in all operating models.
-For operating models with random effects on fleet selectivity, this represents the selectivity at the mean of the random effects.}\label{om_inputs_fig}
-\end{figure}
-
-\begin{landscape}
-\begin{figure}
-\caption{Example simulations of environmental covariate latent processes and observations with different levels of observation error, and different assumptions about variability of the latent process.}\label{om_ecov_example}
-\begin{center}
-\includegraphics[height = \textheight]{Ecov_true_obs_example}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Example simulations of annual natural mortality rates that may be a function of a temporally varying environmental covariate and autoregressive random effects.}\label{M_example}
-\begin{center}
-\includegraphics[height = \textheight]{M_example}
-\end{center}
 \end{figure}

### 2025-12-19 Ecov_study/mortality/manuscript/manuscript_R1.tex
revision for project 0 complete, save work for M-Ecov revision

--- 2025-08-08 Ecov_study/mortality/manuscript/manuscript.tex
@@
 ]{article}
+\usepackage{xcolor}
+\usepackage[margin=1in]{geometry}
 \usepackage{amsmath,amssymb}
-\usepackage{lmodern}
+\setcounter{secnumdepth}{5}
 \usepackage{iftex}
@@
 \makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={Factors affecting inferences on natural mortality and associated environmental effects in state-space age-structured assessment models},
-  pdfauthor={Timothy J. Miller1,2; Gregory L. Britten3; Elizabeth N. Brooks2; Gavin Fay4; Alexander C. Hansell2; Christopher M. Legault2; Brandon Muffley5; John Wiedenmann6},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage[margin=1in]{geometry}
 \usepackage{longtable,booktabs,array}
@@
 \makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\newsavebox\pandoc@box
+\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
+  \sbox\pandoc@box{#1}%
+  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
+  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
+  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
+  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
+  \else\usebox{\pandoc@box}%
+  \fi%
+}
 % Set default figure placement to htbp
-\makeatletter
 \def\fps@figure{htbp}
@@
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{5}
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
-  \fi
-  % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+\usepackage[round]{natbib}
+\bibliographystyle{cjfas.bst}
 \usepackage{url}
@@
 \usepackage{xcolor}
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\usepackage[round]{natbib}
-\bibliographystyle{cjfas.bst}
+\usepackage{bookmark}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same}
+\hypersetup{
+  pdftitle={Factors affecting inferences on natural mortality and associated environmental effects in state-space age-structured assessment models},
+  pdfauthor={Timothy J. Miller1,2; Gregory L. Britten3; Elizabeth N. Brooks2; Gavin Fay4; Alexander C. Hansell2; Christopher M. Legault2; Brandon Muffley5; John Wiedenmann6},
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
 
@@
 Legault\textsuperscript{2} \and Brandon Muffley\textsuperscript{5} \and John Wiedenmann\textsuperscript{6}}
-\date{18 June, 2025}
+\date{18 December, 2025}
 
@@
 
-\hypertarget{methods}{%
-\section*{Methods}\label{methods}}
+\section*{Methods}\label{methods}
 \addcontentsline{toc}{section}{Methods}
 
-Our analyses used the Woods Hole Assessment Model (WHAM) to construct both OMs and EMs \citep{millerstock20, stockmiller21, milleretal_inreview}.
+Our analyses used the Woods Hole Assessment Model (WHAM) to construct both OMs and estimation models (EMs) \citep{millerstock20, stockmiller21, milleretal_inreview}.
 The WHAM package has been used extensively to configure OMs and EMs for several other simulation studies \citep{stocketal21, legaultetal23, lietal24, brittenetal_inreview, lietal_inreview_a} and is used to assess many commercially important stocks in the Northeast U.S.
@@
 
-For each simulated data set we fit a set of 12 estimating models (EMs).
+For each simulated data set we fit a set of 12 EMs.
 The factors that distinguish the estimating models, also described in detail below, include source of population process error type (3 levels) whether (median) \(M\) was estimated or assumed known (2 levels), and whether the effect of the environmental covariate on \(M\) was estimated or not (2 levels).
@@
 
-\hypertarget{operating-models}{%
-\subsection*{Operating models}\label{operating-models}}
+\subsection*{Operating models}\label{operating-models}
 \addcontentsline{toc}{subsection}{Operating models}
 
-\hypertarget{environmental-covariate}{%
-\subsubsection*{Environmental covariate}\label{environmental-covariate}}
+\subsubsection*{Environmental covariate}\label{environmental-covariate}
 \addcontentsline{toc}{subsubsection}{Environmental covariate}
@@
 
-\hypertarget{population}{%
-\subsubsection*{Population}\label{population}}
+\subsubsection*{Population}\label{population}
 \addcontentsline{toc}{subsubsection}{Population}
@@
 
-\hypertarget{catch-and-index-observations}{%
-\subsubsection*{Catch and index observations}\label{catch-and-index-observations}}
+\subsubsection*{Catch and index observations}\label{catch-and-index-observations}
 \addcontentsline{toc}{subsubsection}{Catch and index observations}
@@
 
-\hypertarget{estimating-models}{%
-\subsection*{Estimating models}\label{estimating-models}}
+\subsection*{Estimating models}\label{estimating-models}
 \addcontentsline{toc}{subsection}{Estimating models}
@@
 
-\hypertarget{performance-measures}{%
-\subsection*{Performance measures}\label{performance-measures}}
+\subsection*{Performance measures}\label{performance-measures}
 \addcontentsline{toc}{subsection}{Performance measures}
 
-\hypertarget{em-convergence}{%
-\subsubsection*{EM convergence}\label{em-convergence}}
+\subsubsection*{EM convergence}\label{em-convergence}
 \addcontentsline{toc}{subsubsection}{EM convergence}
@@
 
-\hypertarget{parameter-estimation}{%
-\subsubsection*{Parameter estimation}\label{parameter-estimation}}
+\subsubsection*{Parameter estimation}\label{parameter-estimation}
 \addcontentsline{toc}{subsubsection}{Parameter estimation}
@@
 
-\hypertarget{results}{%
-\section*{Results}\label{results}}
+\section*{Results}\label{results}
 \addcontentsline{toc}{section}{Results}
@@
 
-\hypertarget{em-convergence-1}{%
-\subsection*{EM Convergence}\label{em-convergence-1}}
+\subsection*{EM Convergence}\label{em-convergence-1}
 \addcontentsline{toc}{subsection}{EM Convergence}
@@
 
-\hypertarget{aic-performance}{%
-\subsection*{AIC performance}\label{aic-performance}}
+\subsection*{AIC performance}\label{aic-performance}
 \addcontentsline{toc}{subsection}{AIC performance}
@@
 
-\hypertarget{covariate-effect-estimation}{%
-\subsection*{Covariate effect estimation}\label{covariate-effect-estimation}}
+\subsection*{Covariate effect estimation}\label{covariate-effect-estimation}
 \addcontentsline{toc}{subsection}{Covariate effect estimation}
@@
 
-\hypertarget{median-natural-mortality-rate-estimation}{%
-\subsection*{Median natural mortality rate estimation}\label{median-natural-mortality-rate-estimation}}
+\subsection*{Median natural mortality rate estimation}\label{median-natural-mortality-rate-estimation}
 \addcontentsline{toc}{subsection}{Median natural mortality rate estimation}
@@
 
-\hypertarget{terminal-year-natural-morality-rate}{%
-\subsection*{Terminal year natural morality rate}\label{terminal-year-natural-morality-rate}}
+\subsection*{Terminal year natural morality rate}\label{terminal-year-natural-morality-rate}
 \addcontentsline{toc}{subsection}{Terminal year natural morality rate}
@@
 
-\hypertarget{terminal-year-spawning-stock-biomass}{%
-\subsection*{Terminal year spawning stock biomass}\label{terminal-year-spawning-stock-biomass}}
+\subsection*{Terminal year spawning stock biomass}\label{terminal-year-spawning-stock-biomass}
 \addcontentsline{toc}{subsection}{Terminal year spawning stock biomass}
@@
 
-\hypertarget{discussion}{%
-\section*{Discussion}\label{discussion}}
+\section*{Discussion}\label{discussion}
 \addcontentsline{toc}{section}{Discussion}
@@
 
-\hypertarget{refs}{}
-\begin{CSLReferences}{0}{0}
-\end{CSLReferences}
-
 \pagebreak
+
+\begin{landscape}
+\begin{figure}
+\begin{center}
+\includegraphics[width = 1.4\textwidth]{convergence_classification_plots}
+\end{center}
+\caption{Classification trees indicating primary factors determining convergence as defined by providing Hessian-based standard errors for R, R+S, and R+M OMs.
+Nodes denote percent convergence (top) and number of fits (bottom)  for the corresponding subset.
+Lower or higher convergence rates are indicated by more red or green polygons, respectively}\label{conv_class}
+\end{figure}
+\end{landscape}
+
+\pagebreak
+
+\begin{table}
+\caption{For each OM process error source (columns), percent reduction in deviance for logistic regression models fit to indicators of convergence (providing Hessian-based standard errors) with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{convergence_PRD_table}
+{\input{convergence_PRD_table}}
+\end{table}
+
+\begin{table}
+\caption{For each OM process error source (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
+\ref{bias_regression_response} for the median natural mortality rate parameter with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_median_M_converged_PRD_table}
+{\input{bias_mean_M_converged_PRD_table}}
+\end{table}
+
+\begin{table}
+\caption{For each OM process error source (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
+\ref{bias_regression_response} for the the standard error of the MLE for median natural mortality rate with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_median_M_se_PRD_table}
+{\input{bias_mean_M_se_PRD_table}}
+\end{table}
+
+\begin{table}
+\caption{For each OM process error source (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
+\ref{bias_regression_response} for the covariate effect on natural mortality ($\beta_E$) with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_ecov_beta_converged_PRD_table}
+{\input{bias_ecov_beta_converged_PRD_table}}
+\end{table}
+
+\begin{table}
+\caption{For each OM process error source (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
+\ref{bias_regression_response} for the terminal year SSB with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_SSB_PRD_table}
+{\input{bias_SSB_PRD_table}}
+\end{table}
+
+\begin{table}
+\caption{For each OM process error source (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
+\ref{bias_regression_response} for the median natural mortality rate parameter with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_median_M_complete_PRD_table}
+{\input{bias_mean_M_complete_PRD_table}}
+\end{table}
+
+\begin{table}
+\caption{For each OM process error source (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
+\ref{bias_regression_response} for the covariate effect on natural mortality ($\beta_E$) with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_ecov_beta_complete_PRD_table}
+{\input{bias_ecov_beta_complete_PRD_table}}
+\end{table}
+
+\pagebreak
+
+\section*{Supplemental Materials}\label{supplemental-materials}
+\addcontentsline{toc}{section}{Supplemental Materials}
+
+\setcounter{figure}{0}
+\renewcommand\thefigure{S\arabic{figure}}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width = \textwidth]{om_input_plots_figure}
+\end{center}
+\caption{The proportion mature at age, weight at age, fleet and index selectivity at age, and Beverton-Holt stock-recruit relationship assumed for the population in all operating models.
+For operating models with random effects on fleet selectivity, this represents the selectivity at the mean of the random effects.}\label{om_inputs_fig}
+\end{figure}
+
+\begin{landscape}
+\begin{figure}
+\caption{Example simulations of environmental covariate latent processes and observations with different levels of observation error, and different assumptions about variability of the latent process.}\label{om_ecov_example}
+\begin{center}
+\includegraphics[height = \textheight]{Ecov_true_obs_example}
+\end{center}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+\caption{Example simulations of annual natural mortality rates that may be a function of a temporally varying environmental covariate and autoregressive random effects.}\label{M_example}
+\begin{center}
+\includegraphics[height = \textheight]{M_example}
+\end{center}
+\end{figure}
+\end{landscape}
 
@@
 
-\hypertarget{supplemental-materials}{%
-\section*{Supplemental Materials}\label{supplemental-materials}}
-\addcontentsline{toc}{section}{Supplemental Materials}
-
-\setcounter{figure}{0}
-\renewcommand\thefigure{S\arabic{figure}}
-
-\begin{figure}
-\begin{center}
-\includegraphics[width = \textwidth]{om_input_plots_figure}
-\end{center}
-\caption{The proportion mature at age, weight at age, fleet and index selectivity at age, and Beverton-Holt stock-recruit relationship assumed for the population in all operating models.
-For operating models with random effects on fleet selectivity, this represents the selectivity at the mean of the random effects.}\label{om_inputs_fig}
-\end{figure}
-
-\begin{landscape}
-\begin{figure}
-\caption{Example simulations of environmental covariate latent processes and observations with different levels of observation error, and different assumptions about variability of the latent process.}\label{om_ecov_example}
-\begin{center}
-\includegraphics[height = \textheight]{Ecov_true_obs_example}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-\caption{Example simulations of annual natural mortality rates that may be a function of a temporally varying environmental covariate and autoregressive random effects.}\label{M_example}
-\begin{center}
-\includegraphics[height = \textheight]{M_example}
-\end{center}
-\end{figure}
-\end{landscape}
-
-\hypertarget{convergence-results}{%
-\subsection*{Convergence results}\label{convergence-results}}
+\subsection*{Convergence results}\label{convergence-results}
 \addcontentsline{toc}{subsection}{Convergence results}
@@
 
-\hypertarget{aic-results}{%
-\subsection*{AIC results}\label{aic-results}}
+\subsection*{AIC results}\label{aic-results}
 \addcontentsline{toc}{subsection}{AIC results}
@@
 
-\hypertarget{covariate-effect-bias}{%
-\subsection*{Covariate effect bias}\label{covariate-effect-bias}}
+\subsection*{Covariate effect bias}\label{covariate-effect-bias}
 \addcontentsline{toc}{subsection}{Covariate effect bias}
@@
 
-\hypertarget{covariate-effect-standard-error-estimation-bias}{%
-\subsection*{Covariate effect standard error estimation bias}\label{covariate-effect-standard-error-estimation-bias}}
+\subsection*{Covariate effect standard error estimation bias}\label{covariate-effect-standard-error-estimation-bias}
 \addcontentsline{toc}{subsection}{Covariate effect standard error estimation bias}
@@
 
-\hypertarget{covariate-effect-confidence-interval-coverage}{%
-\subsection*{Covariate effect confidence interval coverage}\label{covariate-effect-confidence-interval-coverage}}
+\subsection*{Covariate effect confidence interval coverage}\label{covariate-effect-confidence-interval-coverage}
 \addcontentsline{toc}{subsection}{Covariate effect confidence interval coverage}
@@
 
-\hypertarget{covariate-effect-rmse}{%
-\subsection*{Covariate effect RMSE}\label{covariate-effect-rmse}}
+\subsection*{Covariate effect RMSE}\label{covariate-effect-rmse}
 \addcontentsline{toc}{subsection}{Covariate effect RMSE}
@@
 
-\hypertarget{covariate-effect-estimate-and-standard-error-example}{%
-\subsection*{Covariate effect estimate and standard error example}\label{covariate-effect-estimate-and-standard-error-example}}
+\subsection*{Covariate effect estimate and standard error example}\label{covariate-effect-estimate-and-standard-error-example}
 \addcontentsline{toc}{subsection}{Covariate effect estimate and standard error example}
@@
 
-\hypertarget{median-natural-mortality-parameter-bias}{%
-\subsection*{Median Natural mortality parameter bias}\label{median-natural-mortality-parameter-bias}}
+\subsection*{Median Natural mortality parameter bias}\label{median-natural-mortality-parameter-bias}
 \addcontentsline{toc}{subsection}{Median Natural mortality parameter bias}
@@
 
-\hypertarget{median-natural-mortality-parameter-standard-error-estimation-bias}{%
-\subsection*{Median natural mortality parameter standard error estimation bias}\label{median-natural-mortality-parameter-standard-error-estimation-bias}}
+\subsection*{Median natural mortality parameter standard error estimation bias}\label{median-natural-mortality-parameter-standard-error-estimation-bias}
 \addcontentsline{toc}{subsection}{Median natural mortality parameter standard error estimation bias}
@@
 
-\hypertarget{median-natural-mortality-parameter-confidence-interval-coverage}{%
-\subsection*{Median Natural mortality parameter confidence interval coverage}\label{median-natural-mortality-parameter-confidence-interval-coverage}}
+\subsection*{Median Natural mortality parameter confidence interval coverage}\label{median-natural-mortality-parameter-confidence-interval-coverage}
 \addcontentsline{toc}{subsection}{Median Natural mortality parameter confidence interval coverage}
@@
 
-\hypertarget{median-natural-mortality-parameter-estimate-and-standard-error-example}{%
-\subsection*{Median natural mortality parameter estimate and standard error example}\label{median-natural-mortality-parameter-estimate-and-standard-error-example}}
+\subsection*{Median natural mortality parameter estimate and standard error example}\label{median-natural-mortality-parameter-estimate-and-standard-error-example}
 \addcontentsline{toc}{subsection}{Median natural mortality parameter estimate and standard error example}
@@
 
-\hypertarget{median-natural-mortality-parameter-rmse}{%
-\subsection*{Median Natural mortality parameter RMSE}\label{median-natural-mortality-parameter-rmse}}
+\subsection*{Median Natural mortality parameter RMSE}\label{median-natural-mortality-parameter-rmse}
 \addcontentsline{toc}{subsection}{Median Natural mortality parameter RMSE}
@@
 
-\hypertarget{terminal-year-natural-mortality-bias}{%
-\subsection*{Terminal year natural mortality bias}\label{terminal-year-natural-mortality-bias}}
+\subsection*{Terminal year natural mortality bias}\label{terminal-year-natural-mortality-bias}
 \addcontentsline{toc}{subsection}{Terminal year natural mortality bias}
@@
 
-\hypertarget{terminal-year-natural-mortality-rmse}{%
-\subsection*{Terminal year natural mortality RMSE}\label{terminal-year-natural-mortality-rmse}}
+\subsection*{Terminal year natural mortality RMSE}\label{terminal-year-natural-mortality-rmse}
 \addcontentsline{toc}{subsection}{Terminal year natural mortality RMSE}
@@
 
-\hypertarget{terminal-year-spawning-stock-biomass-bias}{%
-\subsection*{Terminal year spawning stock biomass bias}\label{terminal-year-spawning-stock-biomass-bias}}
+\subsection*{Terminal year spawning stock biomass bias}\label{terminal-year-spawning-stock-biomass-bias}
 \addcontentsline{toc}{subsection}{Terminal year spawning stock biomass bias}
@@
 
-\hypertarget{terminal-year-spawning-stock-biomass-rmse}{%
-\subsection*{Terminal year spawning stock biomass RMSE}\label{terminal-year-spawning-stock-biomass-rmse}}
+\subsection*{Terminal year spawning stock biomass RMSE}\label{terminal-year-spawning-stock-biomass-rmse}
 \addcontentsline{toc}{subsection}{Terminal year spawning stock biomass RMSE}
@@
 
-\hypertarget{terminal-year-fishing-mortality-bias}{%
-\subsection*{Terminal year fishing mortality bias}\label{terminal-year-fishing-mortality-bias}}
+\subsection*{Terminal year fishing mortality bias}\label{terminal-year-fishing-mortality-bias}
 \addcontentsline{toc}{subsection}{Terminal year fishing mortality bias}
@@
 
-\hypertarget{terminal-year-fishing-mortality-rmse}{%
-\subsection*{Terminal year fishing mortality RMSE}\label{terminal-year-fishing-mortality-rmse}}
+\subsection*{Terminal year fishing mortality RMSE}\label{terminal-year-fishing-mortality-rmse}
 \addcontentsline{toc}{subsection}{Terminal year fishing mortality RMSE}

### 2025-12-19 Project_0/manuscript/manuscript_R1.Rmd
revision for project 0 complete, save work for M-Ecov revision

@@
 
-State-space models have been promoted as the next-generation of fisheries stock assessment and evaluations of their reliability is needed.
+State-space models have been promoted as the next-generation of fisheries stock assessment and evaluation of their reliability is needed.
 We simulated operating models that varied fishing pressure, magnitude of observation error, and sources of process error.
-For each operating model, we fit a range estimating models with a range of correct and incorrect assumptions.
-We measured reliability of estimating models by convergence rate, accuracy of marginal AIC, estimation bias, and magnitude of retrospective patterns.
+For each operating model, we fit a range of estimating models with correct and incorrect configurations.
+We measured reliability of estimating models by convergence rate, accuracy of AIC-based model selection, estimation bias, and magnitude of retrospective patterns.
 All reliability measures were generally better with lower observation error, contrast in fishing pressure over time, and when median natural mortality rate is known.
-Magnitude of the log-likelihood gradients was not a reliable indicator of convergence.
-AIC can generally distinguish process error type with lower observation error and higher true process error variability.
+The magnitude of the log-likelihood gradients was not a reliable indicator of convergence.
+AIC can generally distinguish process error source with lower observation error and higher true process error variability.
 Distinguishing the stock recruit relationship with AIC required large contrast in spawning biomass and low recruitment variation, but bias in stock-recruit parameter estimation was prevalent.
@@
 
-Application of state-space models in fisheries stock assessment and management has expanded dramatically within International Council for the Exploration of the Sea (ICES), Canada, and the Northeast US [@nielsenberg14; @cadigan16; @pedersenberg17; @stockmiller21].
-State-space models treat latent population characteristics as statistical time series with periodic observations that also may have error due to sampling or other other measurement properties.
+Application of state-space models in fisheries stock assessment and management has expanded dramatically within the International Council for the Exploration of the Sea (ICES), Canada, and the Northeast US [@nielsenberg14; @cadigan16; @pedersenberg17; @stockmiller21].
+State-space models treat latent population characteristics as statistical time series with periodic observations that also may have error due to sampling or other measurement properties.
 Traditional assessment models may use state-space approaches to account for temporal variability in population characteristics [@methotwetzel13;@legaultrestrepo99], but these models treat the annual parameters as penalized fixed effect parameters where the variance parameters controlling the penalties are assumed known [@thorsonminto15].
 Modern state-space models can estimate the annually varying parameters as random effects with variance parameters estimated using maximum marginal likelihood or corresponding Bayesian approaches.
-These latter approaches are considered best practice and are recommended for the next generation of stock assessment models [@hoyleetal22; @punt23].
+These random-effects approaches are considered best practice and are recommended for the next generation of stock assessment models [@hoyleetal22; @punt23].
 
@@
 Incorrect treatment of population attributes as temporally varying [@trijouletetal20; @liljestrandetal24] could lead to misidentification of stock status and biased population estimates, ultimately impacting fisheries management decisions [@legaultpalmer16; @szuwalskietal18; @croninpunt21].
-Furthermore, biological, fishery, and observational processes are often confounded in catch-at-age data, which may adversely affect the ability to distinguish between true process variability and observational error [@puntetal14; @stewartmonnahan17; @croninpunt21; @fischetal23; @lietalinreview_a].
+Furthermore, biological, fishery, and observational processes are often confounded in catch-at-age data, which may adversely affect the ability to distinguish between true process variability and observational error [@puntetal14; @stewartmonnahan17; @croninpunt21; @fischetal23; @lietal25a].
 
@@
 
-Despite mounting efforts, several limitations remain.
+Despite increasing research on state space assessment models, several uncertainties in state space assessment modeling remain.
 First, confounding processes that can be treated as random effects in the model have not been thoroughly examined or tested within a simulation-estimation framework.
-Second, previous studies relied on operating models conditioned on specific fisheries, limiting their generalizability [@lietalinreview_a; @liljestrandetal24].
+Second, previous studies relied on operating models conditioned on specific fisheries, limiting their generalizability [@lietal25a; @liljestrandetal24].
 In particular, the effects of observation error and underlying fishing history have not been fully isolated in simulation study designs, making it challenging to disentangle the interplay between process and observation error magnitudes, as demonstrated in @fischetal23.
@@
 In particular, @knape08 demonstrated that stronger density dependence becomes increasingly difficult to estimate in the presence of observation error.
-Therefore, it is crucial to assess whether density dependence mechanisms can be estimated with sufficient precision for use in fisheries management [@augeretal16].
+Therefore, it is crucial to assess whether density-dependent mechanisms can be estimated with sufficient precision for use in fisheries management [@augeretal16].
 Finally, although the importance of autocorrelation in process errors is recognized, investigations of the ability to distinguish state-space assessment models with and without autocorrelation and whether such misspecification is detrimental to estimation of important population metrics are lacking [@johnsonetal16;@xuetal19].
@@
 In the present study, we conduct a simulation study with operating models (OMs) varying by degree of observation error, source and variability of process error, and fishing history.
-The simulations from these OMs are fitted with estimation models (EMs) that make alternative assumptions for sources of process error, whether a SRR was estimated, and whether natural mortality is estimated.
+The simulations from these OMs are fitted with estimation models (EMs) that make alternative assumptions for sources of process error, whether an SRR was estimated, and whether natural mortality is estimated.
 Given the confounding nature of process errors, developing diagnostic tools to detect model misspecification is of great scientific interest and could aid the next generation of stock assessments [@augeretal21].
-We evaluate whether OM and EM attributes affect rates of convergence and the ability of Akaike Information Criterion (AIC) to correctly determine the source of process error or the existence of a SRR.
-We also evaluate when retrospective patterns occur and the degree of bias in outputs of the assessment model that are important for management.
+We evaluate whether OM and EM attributes affect rates of convergence and the ability of Akaike Information Criterion (AIC) to correctly determine the source of process error or the existence of an SRR.
+We also evaluate effects of OM and EM attributes on magnitude of retrospective patterns and bias in estimation of parameters and other model outputs important for management.
 
@@
 For this study we used [version 1.0.6.9000, commit 77bbd94](https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3).
-WHAM has also been used to configure OMs and EMs for closed loop simulations evaluating index-based assessment methods [@legaultetal23] and is currently used or accepted for use in management of numerous NEUS fish stocks [e.g., @nefsc22;@nefsc22a;@nefsc24].
-
-We completed a simulation study with a number of OMs that can be categorized based on where process error random effects were assumed: recruitment (R, assumed present in all models), apparent survival (denoted R+S), natural mortality (R+M), fleet selectivity (R+Sel), or index catchability (R+q).
-We refer to the (R+S) OMs as modeling apparent survival because on log-scale the random effects are additive to the total mortality (F+M) between numbers at age, thus they modify the survival term.
-However, as @stockmiller21 note, these random effects can be due to events other than mortality, such as immigration, emigration, missreported catch, and other sources of misspecification.
-For each OM, assumptions about the magnitude of the variance of process errors and observations are required and the values we used were based on a review of the range of estimates from Northeast Unite States (NEUS) assessments using WHAM.
+WHAM has also been used to configure OMs and EMs for closed loop simulations evaluating index-based assessment methods [@legaultetal23] and is currently used or accepted for use in management of numerous Northeast United States (NEUS) fish stocks [e.g., @nefsc22;@nefsc22a;@nefsc24].
+
+We completed a simulation study with a number of OMs that can be categorized based on where process error random effects were assumed.
+R OMs assume process error for recruitment only.
+Other OM categories assume recruitment process errors along with process errors for apparent survival (R+S), natural mortality (R+M), fleet selectivity (R+Sel), or index catchability (R+q).
+We refer to the R+S OMs as modeling apparent survival because on log-scale the random effects are additive to the total mortality (fishing and natural mortality) between numbers at age, thus they modify the survival term.
+For each OM, assumptions about the magnitude of the variance of process errors and observations are required and the values we used were based on a review of the range of estimates from NEUS assessments using WHAM.
+
+<!-- However, as @stockmiller21 note, these random effects can be due to events other than mortality, such as immigration, emigration, missreported catch, and other sources of misspecification.   -->
 
 In total, we configured 72 OMs with alternative assumptions about the source and magnitude of process errors, magnitude of observation error in indices and age composition data, and contrast in fishing pressure over time.
-For each OM, we simulated 100 population time series with process errors and, for each time series, simulated observation data sets.
+For each OM, we simulated 100 time series of abundance at age with process errors, and for each realized time series, we simulated observation data sets.
 For each data set, we fitted a number of EMs that differed in assumptions about the source of process errors, whether natural mortality (or the median for models with process error in natural mortality) was estimated, and whether a Beverton-Holt SRR was estimated within the EM.
 Details of each of the OMs and EMs are described below.
-
-We did not use the log-normal bias-correction feature for process errors or observations described by [@stockmiller21] for OMs and EMs to simplify interpretation of the study results [@lietal25].
+We did not use the log-normal bias-correction feature for process errors or observations described by @stockmiller21 for OMs and EMs to simplify interpretation of the study results [@lietal25].
 All code we used to perform the simulation study and summarize results can be found at https://github.com/timjmiller/SSRTWG/tree/main/Project_0/code.
@@
 The population consists of 10 age classes, ages 1 to 10+, with the last being a plus group that accumulates ages 10 and older.
-We assume spawning occurs annually 1/4 of the way through the year.
 The maturity at age was a logistic curve with $a_{50}$ = 2.89 and slope = 0.88 (Figure \ref{om_inputs_fig}, top left).
-
-Weight at age was generated with a von Bertalanffy growth function
-$$
-L_a = L_{\infty}\left(1 - e^{-k(a - t_0)}\right)
-$$
-where $t_0 = 0$, $L_\infty = 85$, and $k = 0.3$, and a L-W relationship such that 
-$$
-W_a = \theta_1 L_a^{\theta_2}
+Weight at age ($W_a$) was generated with a von Bertalanffy growth function defining length at age:
+$$
+L_a = L_{\infty}\left(1 - e^{-k(a - t_0)}\right),
+$$
+where $t_0 = 0$, $L_\infty = 85$, and $k = 0.3$, and a length-weight relationship such that 
+$$
+W_a = \theta_1 L_a^{\theta_2},
 $$
@@
 We assumed a Beverton-Holt SRR with constant pre-recruit mortality parameters for all OMs.
-All biological inputs to calculations of spawning biomass per recruit (i.e., weight, maturity, and natural mortality at age) are constant in the apparent survival (R+S) selectivity (R+Sel), and survey catchability (R+q) process error OMs.
+We assume spawning occurs annually 0.25 of each year and recruitment at age 1 ($N_{1,y}$).
+All biological inputs to calculations of spawning stock biomass (SSB) per recruit (i.e., weight, maturity, and natural mortality at age) are constant in the R+S, R+Sel, and R+q process error OMs.
 Therefore, steepness and unfished recruitment are also constant over the time period for those OMs [@millerbrooks21].
@@
 We specified unfished recruitment equal to $e^{10}$ and $\Fmsy = F_{40\%} = 0.348$, which equates to a steepness of 0.69 and $a=0.60$ and $b = 2.4 \times 10^{-5}$ for the Beverton-Holt parameterization
-\[
+$$
 N_{1,y} = \frac{a \text{SSB}_{y-1}}{1 + b \text{SSB}_{y-1}} 
-\]
+$$
 (Figure \ref{om_inputs_fig}, bottom right).
@@
 However, we used the same $a$  and $b$ parameters as other OMs, which equates to a steepness and R0 at the median of the time series process for natural mortality.
-Similarly, For OMs with time-varying random effects for fishery selectivity, \Fmsy also varies temporally, so equilibrium conditions for these OMs are defined for mean selectivity parameters.
+Similarly, for OMs with time-varying random effects for fishery selectivity, \Fmsy also varies temporally, so equilibrium conditions for these OMs are defined for mean selectivity parameters.
 
@@
 We specified initial population abundance at age at the equilibrium distribution that corresponds to fishing at either $F = 2.5\Fmsy$ or $F = \Fmsy$.
-This implies that, for a deterministic model, the abundance at age would not change from the first year to the next.
+This implies that, for a deterministic model, the abundance at age would not change from year to year at the beginning of the time series.
 
@@
 
-The standard deviation for log-aggregate catch was 0.1 for all OMs, a common assumption for NEUS stock assessments.
+The standard deviation for log-aggregate catch was 0.1 for all OMs, a common assumption for commercial removals in NEUS stock assessments.
 Two levels of observation error variance (high and low) were specified for indices and all age composition observations (both indices and catch).
@@
 
-For operating models with random effects on recruitment and(or) apparent survival (R, R+S), we assumed marginal standard deviations for recruitment of $\sigma_R \in \{0.5,1.5\}$.
+For operating models with random effects on recruitment only and also on apparent survival (R, R+S), we assumed marginal standard deviations for recruitment of $\sigma_R \in \{0.5,1.5\}$.
 The marginal standard deviations for apparent survival random effects at older age classes were $\sigma_{2+} \in \{0,0.25, 0.5\}$.
@@
 
-WHAM treats selectivity parameter $s$ as a logit-transformed parameter 
+WHAM treats each selectivity parameter $s$ as a logit-transformed parameter 
 $$
@@
 that is a linear combination of a mean $\mu_s$ and any annual random effects marginally distributed as $\epsilon_{s,y} \sim \text{N}\left(0,\sigma_s^2\right)$, where the lower and upper bounds of the parameter ($l_s$ and $u_s$) can be specified by the user.
-All selectivity parameters ($a_{50}$ and slope parameters) were bounded by 0 and 10 for all OMs and EMs.
+All selectivity parameters ($a_{50}$ and slope parameters) were bounded by $s_l = 0$ and $s_u = 10$ for all OMs and EMs.
 The marginal standard deviations we assumed for logit scale random effects were $\sigma_s \in \{0.1, 0.5\}$ and AR1 autocorrelation parameters of $\rho_s \in \{0,0.9\}$.
@@
 
-Subsets of 20 EMs in Table \ref{em_table} were fit to simulate data sets from each of the OM process error categories.
+Subsets of 20 EMs in Table \ref{em_table} were fit to simulated data sets from each of the OM process error sources.
 For R and R+S OMs, fitted EMs had matching process error assumptions as well as R+Sel, R+M, and R+q assumptions without autocorrelation.
-For other OM process error categories, we fit EMs with correct process error assumptions, the correct process error type but incorrect correlation assumption, and the incorrect process error types without autocorrelation.
-As such, EMs were configured completely correctly for the OM, or they had mis-specification in assumptions of process error autocorrelation, the type of process error, and(or) the SRR (Beverton-Holt or none).
-
-The maturity at age, weight at age for catch and spawning stock biomass (SSB), and observation error standard deviations for aggregate catch and indices were all assumed known at the true values.
+For other OM process error sources, we fit EMs with correct process error assumptions, the correct process error source but incorrect correlation assumption, and the incorrect process error source without autocorrelation.
+As such, EMs were configured correctly for the OM, or they had mis-specification in assumptions of process error autocorrelation, the source of process error, and(or) the SRR (Beverton-Holt or none).
+
+The maturity at age, weight at age for catch and SSB, and observation error standard deviations for aggregate catch and indices were all assumed known at the true values.
 However, the variance parameters for the logistic-normal distributions for age composition observations were estimated in the EMs.
@@
 We investigated the reliability of AIC-based model selection for two purposes.
-First, we analyzed selection of each process error model structure (R, R+S, R+M, R+Sel, R+q) using marginal AIC.
+First, we analyzed selection of each process error model source (R, R+S, R+M, R+Sel, R+q) using marginal AIC.
 For a given OM simulated data set, we compared AIC for EMs with different process error assumption conditional on whether median natural mortality rate and the Beverton-Holt SRR were estimated.
 Second, we analyzed AIC-based selection between EMs with and without the Beverton-Holt SRR assumed.
-Contrast in fishing pressure and time series with recruitment at low stock size has been shown to improve estimation of SRR parameters [@magnussonhilborn07; @connetal10].
-Our preliminary inspections indicated generally poor performance of AIC in determining the Beverton-Holt model for a given set of OM factors (including contrast in fishing pressure), even when the EM was configured with the correct process error type.
+Contrast in fishing pressure and time series with recruitment at low stock size have been shown to improve estimation of SRR parameters [@magnussonhilborn07; @connetal10].
+Our preliminary inspections indicated generally poor performance of AIC in determining the Beverton-Holt SRR model for a given set of OM factors (including contrast in fishing pressure), even when the EM was configured with the correct process error source.
 Therefore, we conditioned on the EMs having the correct process error assumption and also considered the effect of the log-standard deviation of the true log(SSB) ($\log \text{SD}_{\text{SSB}}$; similar to the log of the coefficient of variation for SSB) on model selection since simulations with realized SSB producing low and high recruitment would have larger variation in realized SSB.
@@
 \end{equation}
-from fitting a given estimating model to simulated data set $j$ configured for a given OM where $\widehat \theta_j$ and $\theta_j$ are the estimated and true values for simulation $j$.
-We analyzed simulation results for estimates of terminal year estimates of SSB and recruitment, Beverton-Holt SRR parameters ($a$ and $b$), and median natural mortality rate.
+from fitting a given EM to simulated data set $j$ configured for a given OM where $\widehat \theta_j$ and $\theta_j$ are the estimated and true values for simulation $j$.
+We analyzed simulation results for estimates of terminal year SSB and recruitment, Beverton-Holt SRR parameters ($a$ and $b$), and median natural mortality rate.
 
@@
 where $Y$ is last year of the full set of observations and $\widehat \theta_{y,y'}$ is the estimate for attribute $\theta$ in year $y$ from a model fit using data up to year $y'\geq y$.
+Thus, $\theta$ terms where $y'=Y$ refer to estimates from the fit to all years of data.
 
@@
 
-Because the OM and EM attributes that we investigated are numerous, we used two methods to summarize the most important factors for differences in results within a given OM process error type.
+Because the OM and EM attributes that we investigated are numerous, we used two methods to summarize the most important factors for differences in results within a given OM process error source.
 The first method was fitting regression models with the response being each of the measures of reliability described above and predictor variables were defined based on OM and EM characteristics [e.g., @mackinnonetal95; @wangetal17; @harwelletal18].
-For the binary indicators of convergence and AIC-based selection of a SRR, we performed logistic regressions.
-For indicators of AIC-based selection of EM process error type (multiple categories) we performed multinomial regressions.
+For the binary indicators of convergence and AIC-based selection of an SRR, we performed logistic regressions.
+For indicators of AIC-based selection of EM process error source (multiple categories) we performed multinomial regressions.
 For other measures of reliability we fit linear regression models to transformed responses.
@@
 For the multinomial regression, we used the \verb|vglm| function from the VGAM package [@yee08;@yee15].
-We tabulated percent reduction in residual deviance for each of regression fits.
+We tabulated percent reduction in residual deviance for each of the regression fits.
 We did not perform formal statistical analyses of effects of OM and EM attributes on results (e.g., ANOVA) because of the lack of independence of the "observations" that results from fitting multiple EMs to each simulated data set.
 
-The second method involved fitting classification and regression trees [@breimanetal84] to show how the OM and EM attributes, and their interactions, split the values for each measure of reliability [e.g., @gonzalezetal18;@collieretal22].
+The second method involved fitting classification and regression trees [@breimanetal84] to show how the OM and EM attributes, and their interactions, partition the values for each measure of reliability [e.g., @gonzalezetal18;@collieretal22].
 We used classification trees for categorical measures (convergence and AIC) and regression trees for the other measures with continuous scales (relative error and Mohn's $\rho$).
 The response variables were the same as the regressions for the deviance reduction analyses.
-We used the \verb|rpart| function in the rpart package (https://cran.r-project.org/package=rpart) to fit trees.
+We used the \verb|rpart| function in the rpart package [@rpart] to fit trees.
 Full trees were determined using default settings except that we increased the number of cross-validations to 100.
-For clarity, we pruned the full trees to show just the primary branches.
+For clarity, we manually pruned the full trees to show just the primary branches.
 
@@
 For confidence intervals of probability of convergence, we used the Clopper-Pearson exact method [@clopperpearson34;@thulin14].
-For AIC selection of process error configuration we provide estimates of the proportions of simulations where each EM type was selected.
+For AIC selection of process error source we provide estimates of the proportions of simulations where each EM type was selected.
 For AIC selection of the SRR (a binary indicator for each simulated data set), we fit logistic regressions and present resulting predicted probabilities of correctly selecting the SRR as a function of SSB variability ($\log \text{SD}_{\text{SSB}}$ described above).
@@
 
-For probability of convergence, EM process error assumption was the single attribute that resulted in the largest percent reduction in deviance (14-28\%) for all OM process error types other than R+S OMs where the EM median natural mortality rate assumption (estimated or known) explained the most residual deviance (>11\%; Table \ref{convergence_PRD_table}).
+For probability of convergence, the EM process error assumption was the single attribute that resulted in the largest percent reduction in deviance (14-28\%) for all OM process error sources other than R+S OMs where the EM median natural mortality rate assumption (estimated or known) explained the most residual deviance (>11\%; Table \ref{convergence_PRD_table}).
 However, including interactions of OM and EM factors also provided large reductions in residual deviance (35-47\%), suggesting successful convergence depended on a combination OM and EM attributes.
 
-Classification trees for each OM process error type, all had the primary branch defined using the same attribute that provided the largest reduction in deviance (Figure \ref{conv_class}).
-EMs that assumed R+S process errors converged poorly for all OMs that were simulated with the alternative process error assumptions (R, R+M, R+Sel, an R+q OMs).
+Classification trees for each OM process error source all had the primary branch defined using the same attribute that provided the largest reduction in deviance (Figure \ref{conv_class}).
+EMs that assumed R+S process errors converged poorly for all OMs that were simulated with the alternative process error assumptions (R, R+M, R+Sel, and R+q OMs).
 For all trees, branches based on the OM fishing mortality history showed better convergence when the OM included a change in fishing pressure.
@@
 
-When convergence is defined by a gradient threshold, the primary factor explaining deviance reduction is the same that for Hessian-based convergence for all OM process error types, but there are some differences in deviance reduction for secondary factors (Table \ref{convergence_gradient_PRD_table}), and probability of convergence, overall, was lower (Figure \ref{conv_gradient_class}).
+When convergence is defined by a gradient threshold, the primary factor explaining deviance reduction is the same that for Hessian-based convergence for all OM process error sources, but there are some differences in deviance reduction for secondary factors (Table \ref{convergence_gradient_PRD_table}), and probability of convergence, overall, was lower (Figure \ref{conv_gradient_class}).
 We found a wide range of maximum absolute values of gradients for models that had invertible Hessians (Figure \ref{hess_grad}).
 The largest value observed for a given EM and OM combination was typically $<10^{-3}$, but many converged models had values greater than 1.
-For many OMs, EMs that assumed the correct process error type and did not estimate median natural mortality or the Beverton-Holt SRR produced the lowest gradient values.
+For many OMs, EMs that assumed the correct process error source and did not estimate median natural mortality or the Beverton-Holt SRR produced the lowest gradient values.
 
@@
 
-### Process error structure {-}
-
-For AIC selection of the correct process error configuration, the magnitude of observation and process error variation were the attributes that resulted in the largest percent reductions in deviance across OM process error types other than R OMs (Table \ref{AIC_PE_PRD_table}).
+### Process error source {-}
+
+For AIC selection of the correct process error configuration, the magnitude of observation and process error variation were the attributes that resulted in the largest percent reductions in deviance across OM process error sources other than R OMs (Table \ref{AIC_PE_PRD_table}).
 Both sources of variation explained large reductions in deviance for R+S (17-22\%) and R+Sel (8-26\%) OMs, whereas variance of process errors provided the major reductions for R+M (>9\%) and R+q (>13\%) OMs.
 Comparatively, none of the OM or EM attributes explained particularly large reductions in deviance for R OMs, but fishing history, whether a SRR was estimated, and whether median natural mortality was known or estimated provided similar and the largest reductions (approximately 5-6\%).
-Inclusion of second and third order interactions, did not provide large reductions in deviance for any of the OM process error types.
-
-For all OM process error types other than R OMs, the attributes defining the primary branches of classification trees matched those that provided the largest reductions in deviance (Figure \ref{AIC_PE_class}).
-Across all OMs, AIC was more accurate for the process error type when process error variability was greater and when observation error was lower.
-For R+S OMs, there was a tendency to select R OMs when observation error was higher and apparent survival variation was lower ($\sigma_{2+} = 0.25$), but accuracy for the process error was otherwise highly accurate.
-Similarly, low accuracy for process error type occurred for R+Sel and R+q OMs only with high observation error and lower process error variability, but high accuracy for R+M OMs required both low observation error and high process error variance.
-This was also required for accurate identification of the correct correlation structure for R+M, R+q, and R+M OMs (Figure \ref{pe_aic}).
-No branches were estimated for classification trees fit to the R OMs, likely because accuracy was high across all simulations (0.94), although inspection of the fine-scale results shows there is some degradation in AIC selection when a SRR and median natural mortality rate are estimated for R OMs with constant fishing pressure and high observation error (Figure \ref{pe_aic}, top left).
+Inclusion of second and third order interactions, did not provide large reductions in deviance for any of the OM process error sources.
+
+For all OM process error sources other than R OMs, the attributes defining the primary branches of classification trees matched those that provided the largest reductions in deviance (Figure \ref{AIC_PE_class}).
+Across all OMs, AIC was more accurate for the process error source when process error variability was greater and when observation error was lower.
+For R+S OMs, there was a tendency to select R OMs when observation error was higher and apparent survival variation was lower ($\sigma_{2+} = 0.25$), but accuracy for the process error source was otherwise highly accurate.
+Larger variability of process error relative to observation error was also required for accurate identification of the correct correlation structure for R+M, R+q, and R+Sel OMs (Figure \ref{pe_aic}).
+No branches were estimated for classification trees fit to the R OMs, likely because accuracy was high across all simulations (0.94), although inspection of the fine-scale results shows there is some degradation in AIC selection when an SRR and median natural mortality rate are estimated for R OMs with constant fishing pressure and high observation error (Figure \ref{pe_aic}, top left).
+
+<!-- Similarly, low accuracy for process error source occurred for R+Sel and R+q OMs only with high observation error and lower process error variability, but high accuracy for R+M OMs required both low observation error and high process error variance.
+-->
 
@@
 However, inclusion of all attributes provided larger reductions in deviance than the sum of individual contributions for both R (>30\%) and R+S (~19\%) OMs.
-Further fits for R and R+S OMs including all combinations two factors showed those with $\log \text{SD}_{\text{SSB}}$ and recruitment variation provided essentially the same reduction in deviance as the models with all factors.
-For all OM process error types, inclusion of interaction terms provided relatively little reduction in residual deviance.
+Further fits for R and R+S OMs that including different combinations of two factors additively showed fits that included $\log \text{SD}_{\text{SSB}}$ and recruitment variation only provided essentially the same reduction in deviance as the models with all factors.
+For all OM process error sources, inclusion of interaction terms provided relatively little reduction in residual deviance.
 
@@
 All branches based on $\log \text{SD}_{\text{SSB}}$ showed better accuracy with larger variability in SSB and all branches based on fishing history showed better accuracy when there was contrast in fishing pressure.
-Branched based on OM observation error or recruitment variability (R and R+S OMs) showed better accuracy when they were lower.
+Branches based on OM observation error or recruitment variability (R and R+S OMs) showed better accuracy when they were lower.
 For R OMs, a combination of lower recruitment variability, contrast in fishing pressure, and higher SSB variability produced AIC accuracy over 0.8.
 For R+S OMs, lower recruitment variability and observation error and higher SSB variability produced AIC accuracy of 0.79.
-For R+M, R+Sel, and R+q OM,s accuracy of 0.87 to 0.94 was observed with just increased SSB variability.
+For R+M, R+Sel, and R+q OMs, accuracy of 0.87 to 0.94 was observed with just increased SSB variability.
 
@@
 
-Regression models for log-scale errors in SSB that included the various OM and EM factors showed little reduction in deviance (<5\%) for any of the factors across all OM process error types (Table \ref{bias_SSB_PRD_table}).
-The attributes producing the largest reductions were the EM assumption for median natural mortality (known or estimated) for R, R+M, R+Sel, and R+q OMs (1-3\%), EM process error type for R+S OMs (4\%) and fishing history for all OM types (1-5\%).
+Regression models for log-scale errors in SSB that included the various OM and EM factors showed little reduction in deviance (<5\%) for any of the factors across all OM process error sources (Table \ref{bias_SSB_PRD_table}).
+The attributes producing the largest reductions were the EM assumption for median natural mortality (known or estimated) for R, R+M, R+Sel, and R+q OMs (1-3\%), EM process error assumption for R+S OMs (4\%) and fishing history for all OM process error sources (1-5\%).
 Including second order interactions provided the largest reductions in residual deviance (10- 26\%).
@@
 For branches based on treatment of median natural mortality rate, bias was generally less when it was known rather than estimated.
-For some R+Sel and R+q OMs, less bias in SSB was shown when the EM process error configuration was correct.
+For some R+Sel and R+q OMs, less bias in SSB was shown when the EM process error assumption was correct.
 
@@
 Regression models for log-scale errors of estimates of both the Beverton-Holt $a$ and $b$ parameters showed none of the factors explained large percent reductions in deviance (Table \ref{bias_SR_pars_PRD_table}).
-The OM fishing history provided the largest deviance reduction for most OM process error types for both parameters, but reductions were less than 5.6\% except for R+Sel OMs where the reductions were 11.37\% and 7.97\% for the $a$ and $b$ parameters, respectively and for just the $b$ parameter for R+q OMs (10\%).
+The OM fishing history provided the largest deviance reduction for most OM process error sources for both parameters, but reductions were generally less than 6\%.
+Exceptions were the R+Sel OMs where $a$ and $b$ were reduced by approximately 11\% and 8\%, respectively, and the R+q OMs where $b$ was reduced by 10\%.
 The EM process error assumption provided similar reductions in deviance for both parameters for R OMs.
@@
 
-For regression trees of log-scale errors in Beverton-Holt $a$ and $b$ parameter estimates, less bias was indicated with contrast in OM fishing pressure for all branches in trees for each OM process error type (Figures \ref{SR_a_bias_regtree} and \ref{SR_b_bias_regtree}).
+For regression trees of log-scale errors in Beverton-Holt $a$ and $b$ parameter estimates, less bias was indicated with contrast in OM fishing pressure for all branches in trees for each OM process error source (Figures \ref{SR_a_bias_regtree} and \ref{SR_b_bias_regtree}).
 For all branches based on recruitment variability in trees for R and R+S OMs, less bias in both $a$ and $b$ was observed with less recruitment variability.
@@
 The lack of explanatory power is also reflected in the regression trees where median Mohn's $\rho$ values are near zero unless a large combinations of OM and EM conditions occur (Figure \ref{SSB_mohns_rho_regtree}).
-For example, In R+S OMs, with constant fishing pressure, high observation error, and higher apparent survival process error, EMs that assume R+M process errors have a median Mohn's $\rho = -0.068$.
+For example, in R+S OMs, with constant fishing pressure, high observation error, and higher apparent survival process error, EMs that assume R+M process errors have a median Mohn's $\rho = -0.068$.
 
@@
 
-Analyses of model convergence across simulations can be useful for understanding the utility of alternative convergence criteria used in applications to real data for directing the practitioner to more appropriate random effects configurations.
+Poor convergence was common in our results when the incorrect process error source was assumed.
+@lietal24 found that convergence could be a useful diagnostic especially for separating the correct simpler process error assumption from overly complex models.
+Poor convergence often occurs when parameter estimates are at their bounds [@carvalhoetal21].
+However, even when the Hessian is invertible for a converged model, parameters that are poorly informed will have extremely large variance estimates.
+This further inspection can lead to a more appropriate and often more parsimonious model configuration where the problematic parameters are not estimated.
+For example, process error variance parameters in state-space models that are estimated close to 0 indicates that the random effects are estimated to have little or no variability and removing these process errors is warranted.
+Our experiments did not aim to emulate the practitioner decision process in determining an appropriate model configurations, but evaluating the efficacy of such a decision process when applying EMs might be important in closed loop simulations aimed at quantifying management performance (e.g., a management strategy evaluation).
+
+<!-- In the common scenario where index and age composition data do not measure the population well, our results also suggest we can expect lack of convergence of state-space assessment models when estimating natural mortality and(or) SRRs.
+-->
+
+It is common during the assessment model fitting process to check that the maximum absolute gradient component is less than some threshold prior to inspecting the Hessian of the optimized likelihood for invertibility [@carvalhoetal21], but we found reliance on magnitude of the gradient values for fitted models as a convergence criterion questionable.
+There is no accepted standard for the gradient threshold [e.g., @leeetal11; @hurtadoferroetal15;@ruddthorson18], but the Hessian at the optimized log-likelihood was often invertible when the maximum absolute gradient was much larger than what might be perceived to be a sensible threshold in some of our simulations.
+Therefore, the gradient criterion could exclude models that in fact have an invertible Hessian.
+
+<!-- However, the invertible-Hessian convergence criterion used in their study may also lead to limited ability to distinguish the correct model from overly simplistic models, a pattern that was also noted by @liljestrandetal24 when one assumed process error source may absorb all sources of process error when the magnitude of other process errors are low.
+-->
+
+<!-- A factor affecting the convergence criteria, particularly for maximum likelihood estimation of models with random effects, is numerical accuracy.
+All optimizations performed in these simulations are of the Laplace approximation of the marginal likelihood and, therefore, gradients and Hessians are also with respect to this approximation (see TMB::sdreport in the Template Model Builder package).
+Functionality within the Template Model Builder package exists (i.e., TMB::checkConsistency) to check the validity of the Laplace approximation and the utility of this as a diagnostic for state-space assessment models should be explored further.
+Furthermore, numerical methods are used to calculate and invert the Hessian for variance estimation for models with random effects.
+Our results, along with the potential lack of accuracy imposed by these approximations, suggest at least investigating whether the Hessian is positive definite when the calculated absolute gradients are not terribly large (e.g, $< 1$).
+-->
+
+<!-- Our analyses of model convergence improves our understanding of the utility of alternative convergence criteria used in applications to real data for directing the practitioner to more appropriate random effects configurations.
+-->
+
+<!-- Analyses of model convergence across simulations can be useful for understanding the utility of alternative convergence criteria used in applications to real data for directing the practitioner to more appropriate random effects configurations.
 It is common during the assessment model fitting process to check that the maximum absolute gradient component is less than some threshold prior to inspecting the Hessian of the optimized likelihood for invertibility [@carvalhoetal21].
@@
 We found the Hessian at the optimized log-likelihood can often be invertible when the maximum absolute gradient was much larger than what might be perceived to be a sensible threshold.
-
-@lietal24 found that convergence rate could be a useful diagnostic especially for separating the correct model from overly complex models.
+-->
+
+<!-- @lietal24 found that convergence rate could be a useful diagnostic especially for separating the correct model from overly complex models.
 However, the criteria for convergence used in their study may also lead to limited ability to distinguish the correct model from overly simplistic models, a pattern that was also noted by @liljestrandetal24 in which one process error may absorb all sources of process error when the magnitude of other process errors are low.
-
-Often poor convergence occurs when parameter estimates are at their bounds [@carvalhoetal21], and this also applies to variance parameters for random effects with state-space assessment models.
+-->
+
+<!-- Often poor convergence occurs when parameter estimates are at their bounds [@carvalhoetal21], and this also applies to variance parameters for random effects with state-space assessment models.
 Even when the Hessian is invertible, parameters that are poorly informed will have extremely large variance estimates.
@@
 MSE) aimed at quantifying management performance.
+-->
 
@@
 
-We found accuracy of marginal AIC for process error type required only low observation error for R, R+S, R+Sel, and R+q OMs.
-R+M OMs further required higher process error variability, but this also improved accuracy for the other OM process errors types when there was higher observation error.
+We found accuracy of marginal AIC-based selection for the correct process error source required only low observation error for R, R+S, R+Sel, and R+q OMs.
+R+M OMs further required higher process error variability, but this also improved accuracy for the other OM process errors sources when there was higher observation error.
+These results seem consistent with @lietal24.
+Their simulation studies investigated models with multiple process error sources and found good accuracy of AIC in detecting correct process error assumptions for simulations based on two stocks (Gulf of Maine cod and Southern New England-Mid-Atlantic yellowtail founder) that are well sampled by NEUS bottom trawl surveys used as indices in the respective assessments and poor accuracy for Atlantic mackerel, a semi-pelagic species that is observed relatively poorly.
+
+<!-- The assumptions of OMs with the low observation error reflects scenarios where the population is well-sampled with fishery-independent surveys and extensive age composition data. -->
 
@@
 
-Variation in SSB was the most important factor for using marginal AIC to distinguish the the Beverton-Holt SRR.
+Variation in SSB was the most important factor for using marginal AIC to correctly distinguish the Beverton-Holt SRR from the null model without an SRR.
 For R+M, R+Sel, and R+q OMs, the SRR was accurately detected when the CV of SSB over the time series was at least 40 to 50\% ($\log \text{SD}_{\text{SSB}}$ = -0.9 to -0.7) regardless of any other OM or EM attributes.
-Detection of the SRR for R and R+S OMs required lower recruitment variability, but this lower level ($\sigma_R = 0.5$) was assumed for all of the other OM process error types.
-Our results assumed that the EM process error configuration was correct, but this may be a strong limation given the ability of AIC to distinguish the process error type in many scenarios.
-
-Although we did not compare models with alternative SRRs (e.g., Ricker and Beverton-Holt), we do not expect AIC to perform any better distinguishing between relationships and may be more difficult than distinguishing from the null model even with larger variability in SSB.
-Our finding that AIC tended to choose simpler recruitment models in many cases contrasts with the noted bias in AIC for more complex models [@shibata76;@katz81;@kassraftery95], but, whereas those findings apply to the much more common comparison of models that are fit to raw and independent observations, here we are comparing state-space models which account for observation error and estimate process errors in latent variables.
-
-Our results comport with those of @devalpinehastings02 who found AIC could not distinguish among state-space SRRs that were fit just to SSB and recruitment observations (i.e., not an assessment model).
-Similarly,  @brittenetalinreview found AIC could not reliably distinguish alternative environmental effects on SRR parameters.
-However, @milleretal16 did find AIC to prefer a SRR with environmental effects when applied to data for the SNEMA yellowtail flounder stock and  AIC also selected an environmental covariate on a SRR for the most recent stock assessment of Georges Bank yellowtail flounder [@nefsc25].
+Detection of the SRR for R and R+S OMs required lower recruitment variability, but this lower level ($\sigma_R = 0.5$) was assumed for all of the other OM process error source and represents the lower range of estimates from recent NEUS stock assessments.
+Our results assumed that the EM process error configuration was correct, but this may not be a strong limitation given the ability of AIC to distinguish the process error source in many scenarios.
+
+Although we did not compare models with alternative SRRs (e.g., Ricker vs.
+Beverton-Holt), we do not expect AIC to perform any better distinguishing between relationships and may be more difficult than distinguishing from the null model even with larger variability in SSB.
+Our finding that AIC tended to choose simpler recruitment models in many cases contrasts with the noted bias in AIC for more complex models [@shibata76;@katz81;@kassraftery95].
+However, these earlier findings apply to the much more common comparison of models that are fit to raw and independent observations, whereas our comparisons of state-space models account for observation error and separately estimate process errors in latent variables.
+
+Our results comport with those of @devalpinehastings02 who found AIC could not distinguish among state-space SRRs that were fit just to SSB and recruitment observations (i.e., not within an assessment model).
+Similarly,  @brittenetalinreview found AIC could not reliably distinguish the Beverton-Holt SRR from no SRR, nor identify alternative environmental effects on SRR parameters.
+However, @milleretal16 did find AIC to prefer an SRR with environmental effects when applied to data for the Southern New England-Mid-Atlantic yellowtail flounder stock and AIC also selected an environmental covariate on an SRR for the most recent stock assessment of Georges Bank yellowtail flounder [@nefsc25].
 Both of these yellowtail flounder stocks have large changes in stock size and the values of environmental covariates over time.
@@
 
-However, estimation of SRR parameters was only moderately reliable in ideal scenarios of low observation error and contrast in fishing for R+Sel and R+M OMs with large temporal variability in process errors.
+Estimation of SRR parameters was only moderately reliable in ideal scenarios of low observation error and contrast in fishing for R+Sel and R+M OMs with large temporal variability in process errors.
 Otherwise, SRR parameter estimation was biased and(or) highly variable.
-We found substantial bias in estimated SRR parameters in R and R+S OMs particularly with high variability in recruitment and apparent survival process errors, suggesting that practitioners should be cautious SRR inferences when fitted assessment models have these properties.
+We found substantial bias in estimated SRR parameters in R and R+S OMs particularly with high variability in recruitment and apparent survival process errors, suggesting that practitioners should be cautious with SRR inferences when fitted assessment models have these properties.
 We only evaluated effects of SSB variability on accuracy of AIC in identifying the SRR, but those results suggests we might find less bias for the SRR parameters in such cases as well.
-Similarly, restricting results to fits that converged may also yield better accuracy of SRR parameter estimation.
+Another condition that could improve perception of bias in our simulation studies is restricting results to fits that converged with Hessian-based standard errors for all parameters, but @brittenetalinreview did not find less SRR parameter bias when restricting estimates using a gradient-based criterion.
+A simulation study by @stockmiller21 examining configurations of environmental covariate effects on a Beverton-Holt SRR for the previously mentioned, well-observed, Southern New England-Mid-Atlantic yellowtail flounder stock found little or no bias for the density-independent mortality parameter $a$, but still biased estimation of the density-dependent parameter $b$.
+
 <!-- A proper regression analysis analogous to those we performed may require using an alternative transformation of the estimation errors to measure proximity to 0 irrespective of sign as the response variable (e.g., log of absolute value of relative errors).
@@
 
-Incorrect EM process error assumptions did not produce strong retrospective patterns for SSB for any OMs regardless of whether median natural mortality or a SRR was estimated although some weak patterns occurred when observation error was high and there was contrast in fishing pressure.
+Incorrect EM process error assumptions did not produce strong retrospective patterns for SSB for any OMs regardless of whether median natural mortality or an SRR was estimated, although some weak patterns occurred when observation error was high and there was contrast in fishing pressure.
 However, retrospective patterns tended to be more variable for recruitment and were sometimes large even when the EM was correct.
@@
 Retrospective patterns are often induced in simulation studies by rapid changes in a quantity such as index catchability, natural mortality, or perceived catch during years toward the end of the time series [@legault09; @millerlegault17; @huynhetal22; @breiviketal23].
-In our simulations, the process errors changing over time may have trends in particular simulations, particularly when strong autocorrelation is imposed, but the random effects have no trend on average across simulations.
-@szuwalskietal18 and @lietal24 also found relatively small retrospective patterns when the source of mis-specification was temporal variation in demography attributes.
+In our simulations, the process errors changing over time may have trends in certain simulations, particularly when strong autocorrelation is imposed, but the random effects have no trend on average across simulations.
+@szuwalskietal18 and @lietal24 also found relatively small retrospective patterns when the source of mis-specification was temporal variation in demographic attributes.
 Indeed, it is common for the flexibility provided by temporal random effects to reduce retrospective patterns [@milleretal18; @stockmiller21; @stocketal21], though it does not necessarily indicate a more accurate assessment model [@perrettietal20; @lietal24; @liljestrandetal24].
@@
 We found the use of regression models and classification and regression trees extremely useful in understanding the most important OM and EM attributes explaining variation in the measures of reliability we examined across all simulations.
-The classification and regression trees are generally a good tool for determining the values of the OM and EM attributes that produce better or worse measures of reliability.
+The classification and regression trees are generally a good tool for determining the OM and EM attributes that produce better or worse measures of reliability.
 However, determining the combination of attributes that produce the best or worst measures of reliability can be challenging using the trees alone.
@@
 
-Our findings regarding model convergence suggests practitioners using state-space models and maximum marginal likelihood for estimation should not heavily weight the magnitude of the gradient values in determining convergence as long as the maximum absolute values is around 1 or lower.
+Our findings regarding model convergence suggests practitioners using state-space models and maximum marginal likelihood for estimation should not heavily weight the magnitude of the gradient values in determining convergence as long as the maximum absolute value is around 1 or lower.
 Instead, positive-definiteness of the Hessian of the minimized negative log-likelihood should be evaluated.
@@
 Unfortunately, whether the practitioner includes a Beverton-Holt SRR will often depend on biological plausibility of this particular SRR because using AIC to determine its validity required a combination of low recruitment variability, contrast in fishing pressure, large variation in SSB over time, and lower observation error, which applies to a limited number of managed stocks.
-Furthermore, some bias in estimation of the SRR parameters should be expected, which presumably also applies to MSY-based reference points.
-Because bias in terminal SSB and retrospective patterns were indifferent to whether or not the SRR was estimated, the prevalence of bias in SRR parameter estimation, and often better convergence without the SRR, we recommend a sensible default is to exclude a SRR when fitting assessment models, as also suggested by @brooks24.
+Furthermore, some bias in estimation of the SRR parameters (and MSY-based reference points should be expected.
+Because bias in terminal SSB and retrospective patterns were indifferent to whether or not the SRR was estimated, the prevalence of bias in SRR parameter estimation, and often better convergence without the SRR, we recommend a sensible default is to exclude an SRR when fitting assessment models, as also suggested by @brooks24.
 
@@
 \caption{Classification trees indicating primary factors determining convergence as defined by providing Hessian-based standard errors for R, R+S, R+M, R+Sel and R+q OMs.
+Nodes denote percent convergence (top) and number of fits (bottom)  for the corresponding subset.
 Lower or higher convergence rates are indicated by more red or green polygons, respectively}\label{conv_class}
@@
 \caption{Classification trees indicating primary factors determining which EM SRR assumption (none or Beverton-Holt) provides the lowest AIC for R, R+S, R+M, R+Sel and R+q OMs.
-Each node shows the proportion of EMs that assume the SRR with lowest AIC (top) and number of observations (bottom) for the corresponding subset.
+All EMs assume the correct process error source.
+Nodes denote the percentage of EMs that assume the SRR with lowest AIC (top) and number of observations (bottom) for the corresponding subset.
 Lower or higher accuracy of the process error assumption are indicated by more red or green polygons, respectively.}\label{AIC_SRR_class}
@@
 \begin{table}
-\caption{For each OM process error type (columns), percent reduction in deviance for logistic regression models fit to indicators of convergence (providing Hessian-based standard errors) with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{convergence_PRD_table}
+\caption{For each OM process error source (columns), percent reduction in deviance for logistic regression models fit to indicators of convergence (providing Hessian-based standard errors) with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{convergence_PRD_table}
 {\input{convergence_PRD_table}}
@@
 \begin{table}
-\caption{For each OM process error type (columns), percent reduction in deviance for multinomial logistic regression models fit to indicators of EM process error assumption with lowest AIC with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{AIC_PE_PRD_table}
+\caption{For each OM process error source (columns), percent reduction in deviance for multinomial logistic regression models fit to indicators of EM process error assumption with lowest AIC with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{AIC_PE_PRD_table}
 {\input{AIC_PE_PRD_table}}
@@
 \begin{table}
-\caption{For each OM process error type (columns), percent reduction in deviance for logistic regression models fit to indicators of EM SRR assumption (none or Beverton-Holt) with lowest AIC with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{AIC_SRR_PRD_table}
+\caption{For each OM process error source (columns), percent reduction in deviance for logistic regression models fit to indicators of EM SRR assumption (none or Beverton-Holt) with lowest AIC with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{AIC_SRR_PRD_table}
 {\input{AIC_SRR_PRD_table}}
@@
 \begin{table}
-\caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
+\caption{For each OM process error source (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
 \ref{bias_regression_response} for the terminal year SSB with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_SSB_PRD_table}
@@
 \begin{table}
-\caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
+\caption{For each OM process error source (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
 \ref{bias_regression_response} for the Beverton-Holt SRR parameters with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_SR_pars_PRD_table}
@@
 \begin{table}
-\caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
+\caption{For each OM process error source (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
 \ref{bias_regression_response} for the median natural mortality rate parameter with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_median_M_PRD_table}
@@
 \begin{table}
-\caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to transformed Mohn's $\rho$ values for each simulation (Eq.
+\caption{For each OM process error source (columns), percent reduction in deviance for linear regression models fit to transformed Mohn's $\rho$ values for each simulation (Eq.
 \ref{bias_regression_response}) for SSB with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{mohns_rho_SSB_PRD_table}
@@
 \caption{Classification trees indicating primary factors determining convergence as defined by a maximum absolute gradient < $10^{-6}$ for R, R+S, R+M, R+Sel and R+q OMs.
+Nodes denote percent convergence (top) and number of fits (bottom)  for the corresponding subset.
 Lower or higher convergence rates are indicated by more red or green polygons, respectively}\label{conv_gradient_class}
@@
 \caption{The maximum of the absolute values of all gradient values for all fits that provided Hessian-based standard errors across all simuated data sets of a given OM configuration (A: R and R+S, B: R+M, C: R+Sel, or D: R+q).
-Results are conditional on EM fits with alternative process error type (colored points and lines), median natural mortality (estimated or known) and recruitment assumptions (Beverton-Holt SRR or not).
+Results are conditional on EM fits with alternative process error assumptions (colored points and lines), median natural mortality (estimated or known) and recruitment assumptions (Beverton-Holt SRR or not).
 Circled values indicate results where the EM process error structure matches that of the OM and vertical lines represent 95\% confidence intervals.}\label{hess_grad}
@@
 \end{center}
-\caption{Estimated probability of lowest AIC for EMs assuming alternative process error structures (colored bars) conditional on alternative assumptions for median natural mortality (estimated or known) and Beverton-Holt SRR (estimated or not; along x-axis) when fitted to OMs that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
+\caption{Estimated probability of lowest AIC for EMs assuming alternative process error assumptions (colored bars) conditional on alternative assumptions for median natural mortality (estimated or known) and Beverton-Holt SRR (estimated or not; along x-axis) when fitted to OMs that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error sources.
 Striped bars indicate results where the EM process error structure matches that of the OM.}\label{pe_aic}
@@
 
-\begin{landscape}
 \begin{table}
@@
 \end{table}
-\end{landscape}
-
-\begin{landscape}
+
 \begin{table}
@@
 \end{table}
-\end{landscape}
-
-\begin{table}
-\caption{Distinguishing characteristics of the EMs and indication (+) of which OM process error categories (R, R+S, R+M, R+Sel, R+q) each EM configuration was fit.}\label{em_table}
+
+\begin{table}
+\caption{Distinguishing characteristics of the EMs and indication (+) of which OM process error sources (R, R+S, R+M, R+Sel, R+q) each EM configuration was fit.}\label{em_table}
 {\scriptsize \input{em_tab}}
@@
 \begin{table}
-\caption{For each OM process error type (columns), percent reduction in deviance for logistic regression models fit to indicators of convergence (maximum absolute gradient < $10^{-6}$) with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{convergence_gradient_PRD_table}
+\caption{For each OM process error source (columns), percent reduction in deviance for logistic regression models fit to indicators of convergence (maximum absolute gradient < $10^{-6}$) with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{convergence_gradient_PRD_table}
 {\input{convergence_gradient_PRD_table}}
@@
 \begin{table}
-\caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
+\caption{For each OM process error source (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
 \ref{bias_regression_response} for the terminal year fully-selected fishing mortality with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_F_PRD_table}
@@
 \begin{table}
-\caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
+\caption{For each OM process error source (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq.
 \ref{bias_regression_response} for the terminal year recruitment with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_R_PRD_table}
@@
 
-## Further Detailed Results {-}
+## Further Results {-}
 
@@
 \end{center}
-\caption{Median relative error of terminal year SSB for EMs fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+\caption{Median relative error of terminal year SSB for EMs fitted to data sets simulated with alternative process error sources: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
 Circled values indicate results where the EM process error structure matches that of the OM and vertical lines represent 95\% confidence intervals.}\label{SSB_rel_error}
@@
 \end{center}
-\caption{Median relative error of terminal year fully-selected fishing mortality for EMs fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+\caption{Median relative error of terminal year fully-selected fishing mortality for EMs fitted to data sets simulated with alternative process error sources: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
 Circled values indicate results where the EM process error structure matches that of the OM and vertical lines represent 95\% confidence intervals.}\label{F_rel_error}
@@
 \end{center}
-\caption{Median relative error of terminal year recruitment for EMs fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+\caption{Median relative error of terminal year recruitment for EMs fitted to data sets simulated with alternative process error sources: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
 Circled values indicate results where the EM process error structure matches that of the OM and vertical lines represent 95\% confidence intervals.}\label{R_rel_error}
@@
 \end{center}
-\caption{Median relative error of median natural mortality for EMs fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+\caption{Median relative error of median natural mortality for EMs fitted to data sets simulated with alternative process error sources: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
 Circled values indicate results where the EM process error structure matches that of the OM and vertical lines represent 95\% confidence intervals.}\label{M_rel_error}
@@
 \end{center}
-\caption{Median Mohn's $\rho$ for SSB for EMs fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+\caption{Median Mohn's $\rho$ for SSB for EMs fitted to data sets simulated with alternative process error sources: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
 Circled values indicate results where the EM process error structure matches that of the OM and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_ssb}
@@
 \end{center}
-\caption{Median Mohn's $\rho$ of fishing mortality averaged over all age classes for EMs fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+\caption{Median Mohn's $\rho$ of fishing mortality averaged over all age classes for EMs fitted to data sets simulated with alternative process error sources: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
 Circled values indicate results where the EM process error structure matches that of the OM and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_F}
@@
 \end{center}
-\caption{Median Mohn's $\rho$ of recruitment for EMs fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
+\caption{Median Mohn's $\rho$ of recruitment for EMs fitted to data sets simulated with alternative process error sources: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).
 Circled values indicate results where the EM process error structure matches that of the OM and vertical lines represent 95\% confidence intervals.}\label{mohns_rho_R}
@@
 
-\begin{table}
-\caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to transformed Mohn's $\rho$ values for each simulation (Eq.
-\ref{bias_regression_response}) for fishing mortality averaged over all age classes with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{mohns_rho_F_PRD_table}
-{\input{mohns_rho_F_PRD_table}}
-\end{table}
-
-\begin{table}
-\caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to transformed Mohn's $\rho$ values for each simulation (Eq.
-\ref{bias_regression_response}) for recruitment with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{mohns_rho_R_PRD_table}
-{\input{mohns_rho_R_PRD_table}}
-\end{table}
-
 \begin{landscape}
@@
 \end{center}
-\caption{Probability of EMs providing Hessian-based standard errors with alternative process error (colored points and lines), and median natural mortality (estimated or known) and Beverton-Holt SRR (estimated or not; along x-axis) assumptions when fitted to OMs that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
+\caption{Probability of EMs providing Hessian-based standard errors with alternative process error (colored points and lines), and median natural mortality (estimated or known) and Beverton-Holt SRR (estimated or not; along x-axis) assumptions when fitted to OMs that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error sources.
 Circled values indicate results where the EM process error structure matches that of the OM and vertical lines represent 95\% confidence intervals.}\label{hessian_SE_convergence}
@@
 \end{center}
-\caption{Probability of EMs providing maximum absolute values of gradients less than $10^{-6}$ with alternative process error (colored points and lines), and median natural mortality (estimated or known) and Beverton-Holt SRR (estimated or not; along x-axis) assumptions when fitted to OMs that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error structures.
+\caption{Probability of EMs providing maximum absolute values of gradients less than $10^{-6}$ with alternative process error (colored points and lines), and median natural mortality (estimated or known) and Beverton-Holt SRR (estimated or not; along x-axis) assumptions when fitted to OMs that have R and R+S (A), R+Sel (B), R+M (C), or R+q (D) process error sources.
 Circled values indicate results where the EM process error structure matches that of the OM and vertical lines represent 95\% confidence intervals.}\label{gradient_convergence}
@@
 \end{landscape}
+
+\begin{table}
+\caption{For each OM process error source (columns), percent reduction in deviance for linear regression models fit to transformed Mohn's $\rho$ values for each simulation (Eq.
+\ref{bias_regression_response}) for fishing mortality averaged over all age classes with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{mohns_rho_F_PRD_table}
+{\input{mohns_rho_F_PRD_table}}
+\end{table}
+
+\begin{table}
+\caption{For each OM process error source (columns), percent reduction in deviance for linear regression models fit to transformed Mohn's $\rho$ values for each simulation (Eq.
+\ref{bias_regression_response}) for recruitment with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{mohns_rho_R_PRD_table}
+{\input{mohns_rho_R_PRD_table}}
+\end{table}

### 2025-12-19 Project_0/manuscript/manuscript_R1.tex
revision for project 0 complete, save work for M-Ecov revision

@@
   pdftitle={An investigation of factors affecting inferences from and reliability of state-space age-structured assessment models},
-  pdfauthor={Timothy J. Miller1,2; Greg Britten3; Elizabeth N. Brooks2; Gavin Fay4; Alex Hansell2; Christopher M. Legault2; Chengxue Li2; Brandon Muffley5; Brian C. Stock6; John Wiedenmann7},
+  pdfauthor={Timothy J.
+Miller1,2; Gregory L.
+Britten3; Elizabeth N.
+Brooks2; Gavin Fay4; Alexander C.
+Hansell2; Christopher M.
+Legault2; Chengxue Li2; Brandon Muffley5; Brian C.
+Stock6; John Wiedenmann7},
   hidelinks,
@@
 reliability of state-space age-structured assessment models}
-\author{Timothy J. Miller\textsuperscript{1,2} \and Greg
+\author{Timothy J. Miller\textsuperscript{1,2} \and Gregory L.
 Britten\textsuperscript{3} \and Elizabeth N.
-Brooks\textsuperscript{2} \and Gavin Fay\textsuperscript{4} \and Alex
+Brooks\textsuperscript{2} \and Gavin
+Fay\textsuperscript{4} \and Alexander C.
 Hansell\textsuperscript{2} \and Christopher M.
@@
 C. Stock\textsuperscript{6} \and John Wiedenmann\textsuperscript{7}}
-\date{04 December, 2025}
+\date{19 December, 2025}
 
@@
 State-space models have been promoted as the next-generation of
-fisheries stock assessment and evaluations of their reliability is
+fisheries stock assessment and evaluation of their reliability is
 needed. We simulated operating models that varied fishing pressure,
 magnitude of observation error, and sources of process error. For each
-operating model, we fit a range estimating models with a range of
-correct and incorrect assumptions. We measured reliability of estimating
-models by convergence rate, accuracy of marginal AIC, estimation bias,
-and magnitude of retrospective patterns. All reliability measures were
-generally better with lower observation error, contrast in fishing
-pressure over time, and when median natural mortality rate is known.
-Magnitude of the log-likelihood gradients was not a reliable indicator
-of convergence. AIC can generally distinguish process error type with
+operating model, we fit a range of estimating models with correct and
+incorrect configurations. We measured reliability of estimating models
+by convergence rate, accuracy of AIC-based model selection, estimation
+bias, and magnitude of retrospective patterns. All reliability measures
+were generally better with lower observation error, contrast in fishing
+pressure over time, and when median natural mortality rate is known. The
+magnitude of the log-likelihood gradients was not a reliable indicator
+of convergence. AIC can generally distinguish process error source with
 lower observation error and higher true process error variability.
@@
 Application of state-space models in fisheries stock assessment and
-management has expanded dramatically within International Council for
-the Exploration of the Sea (ICES), Canada, and the Northeast US (Nielsen
-and Berg 2014; Cadigan 2016; Pedersen and Berg 2017; Stock and Miller
-2021). State-space models treat latent population characteristics as
-statistical time series with periodic observations that also may have
-error due to sampling or other other measurement properties. Traditional
+management has expanded dramatically within the International Council
+for the Exploration of the Sea (ICES), Canada, and the Northeast US
+(Nielsen and Berg 2014; Cadigan 2016; Pedersen and Berg 2017; Stock and
+Miller 2021). State-space models treat latent population characteristics
+as statistical time series with periodic observations that also may have
+error due to sampling or other measurement properties. Traditional
 assessment models may use state-space approaches to account for temporal
@@
 as random effects with variance parameters estimated using maximum
-marginal likelihood or corresponding Bayesian approaches. These latter
-approaches are considered best practice and are recommended for the next
-generation of stock assessment models (Hoyle et al. 2022; Punt 2023).
+marginal likelihood or corresponding Bayesian approaches. These
+random-effects approaches are considered best practice and are
+recommended for the next generation of stock assessment models (Hoyle et
+al. 2022; Punt 2023).
 
@@
 to distinguish between true process variability and observational error
-(Li et al. In press; Punt et al. 2014; Stewart and Monnahan 2017;
-Cronin-Fine and Punt 2021; Fisch et al. 2023).
+(Punt et al. 2014; Stewart and Monnahan 2017; Cronin-Fine and Punt 2021;
+Fisch et al. 2023; Li et al. 2025a).
 
@@
 
-Despite mounting efforts, several limitations remain. First, confounding
-processes that can be treated as random effects in the model have not
-been thoroughly examined or tested within a simulation-estimation
-framework. Second, previous studies relied on operating models
-conditioned on specific fisheries, limiting their generalizability (Li
-et al. In press; Liljestrand et al. 2024). In particular, the effects of
-observation error and underlying fishing history have not been fully
-isolated in simulation study designs, making it challenging to
-disentangle the interplay between process and observation error
-magnitudes, as demonstrated in Fisch et al. (2023). Third, explicitly
-modeling stock-recruit relationships (SRRs) as mechanistic drivers of
-population dynamics is promising (Fleischman et al. 2013; Pontavice et
-al. 2022), but reliability of inferences within integrated state-space
-age-structured models has not been evaluated. Evidence from other
-studies suggests that when both process and observation errors are
-unknown, estimating density dependence parameters becomes highly
-uncertain (Knape 2008; Polansky et al. 2009). In particular, Knape
-(2008) demonstrated that stronger density dependence becomes
-increasingly difficult to estimate in the presence of observation error.
-Therefore, it is crucial to assess whether density dependence mechanisms
-can be estimated with sufficient precision for use in fisheries
-management (Auger-Méthé et al. 2016). Finally, although the importance
-of autocorrelation in process errors is recognized, investigations of
-the ability to distinguish state-space assessment models with and
-without autocorrelation and whether such misspecification is detrimental
-to estimation of important population metrics are lacking (Johnson et
-al. 2016; Xu et al. 2019).
+Despite increasing research on state space assessment models, several
+uncertainties in state space assessment modeling remain. First,
+confounding processes that can be treated as random effects in the model
+have not been thoroughly examined or tested within a
+simulation-estimation framework. Second, previous studies relied on
+operating models conditioned on specific fisheries, limiting their
+generalizability (Liljestrand et al. 2024; Li et al. 2025a). In
+particular, the effects of observation error and underlying fishing
+history have not been fully isolated in simulation study designs, making
+it challenging to disentangle the interplay between process and
+observation error magnitudes, as demonstrated in Fisch et al. (2023).
+Third, explicitly modeling stock-recruit relationships (SRRs) as
+mechanistic drivers of population dynamics is promising (Fleischman et
+al. 2013; Pontavice et al. 2022), but reliability of inferences within
+integrated state-space age-structured models has not been evaluated.
+Evidence from other studies suggests that when both process and
+observation errors are unknown, estimating density dependence parameters
+becomes highly uncertain (Knape 2008; Polansky et al. 2009). In
+particular, Knape (2008) demonstrated that stronger density dependence
+becomes increasingly difficult to estimate in the presence of
+observation error. Therefore, it is crucial to assess whether
+density-dependent mechanisms can be estimated with sufficient precision
+for use in fisheries management (Auger-Méthé et al. 2016). Finally,
+although the importance of autocorrelation in process errors is
+recognized, investigations of the ability to distinguish state-space
+assessment models with and without autocorrelation and whether such
+misspecification is detrimental to estimation of important population
+metrics are lacking (Johnson et al. 2016; Xu et al. 2019).
 
@@
 these OMs are fitted with estimation models (EMs) that make alternative
-assumptions for sources of process error, whether a SRR was estimated,
+assumptions for sources of process error, whether an SRR was estimated,
 and whether natural mortality is estimated. Given the confounding nature
@@
 of Akaike Information Criterion (AIC) to correctly determine the source
-of process error or the existence of a SRR. We also evaluate when
-retrospective patterns occur and the degree of bias in outputs of the
-assessment model that are important for management.
+of process error or the existence of an SRR. We also evaluate effects of
+OM and EM attributes on magnitude of retrospective patterns and bias in
+estimation of parameters and other model outputs important for
+management.
 
@@
 methods (Legault et al. 2023) and is currently used or accepted for use
-in management of numerous NEUS fish stocks (e.g., NEFSC 2022a, 2022b;
-NEFSC 2024).
+in management of numerous Northeast United States (NEUS) fish stocks
+(e.g., NEFSC 2022a, 2022b; NEFSC 2024).
 
 We completed a simulation study with a number of OMs that can be
-categorized based on where process error random effects were assumed:
-recruitment (R, assumed present in all models), apparent survival
-(denoted R+S), natural mortality (R+M), fleet selectivity (R+Sel), or
-index catchability (R+q). We refer to the (R+S) OMs as modeling apparent
+categorized based on where process error random effects were assumed. R
+OMs assume process error for recruitment only. Other OM categories
+assume recruitment process errors along with process errors for apparent
+survival (R+S), natural mortality (R+M), fleet selectivity (R+Sel), or
+index catchability (R+q). We refer to the R+S OMs as modeling apparent
 survival because on log-scale the random effects are additive to the
-total mortality (F+M) between numbers at age, thus they modify the
-survival term. However, as Stock and Miller (2021) note, these random
-effects can be due to events other than mortality, such as immigration,
-emigration, missreported catch, and other sources of misspecification.
-For each OM, assumptions about the magnitude of the variance of process
-errors and observations are required and the values we used were based
-on a review of the range of estimates from Northeast Unite States (NEUS)
-assessments using WHAM.
+total mortality (fishing and natural mortality) between numbers at age,
+thus they modify the survival term. For each OM, assumptions about the
+magnitude of the variance of process errors and observations are
+required and the values we used were based on a review of the range of
+estimates from NEUS assessments using WHAM.
 
@@
 in indices and age composition data, and contrast in fishing pressure
-over time. For each OM, we simulated 100 population time series with
-process errors and, for each time series, simulated observation data
-sets. For each data set, we fitted a number of EMs that differed in
-assumptions about the source of process errors, whether natural
-mortality (or the median for models with process error in natural
-mortality) was estimated, and whether a Beverton-Holt SRR was estimated
-within the EM. Details of each of the OMs and EMs are described below.
-
-We did not use the log-normal bias-correction feature for process errors
-or observations described by (Stock and Miller 2021) for OMs and EMs to
-simplify interpretation of the study results (Li et al. 2025). All code
-we used to perform the simulation study and summarize results can be
-found at
+over time. For each OM, we simulated 100 time series of abundance at age
+with process errors, and for each realized time series, we simulated
+observation data sets. For each data set, we fitted a number of EMs that
+differed in assumptions about the source of process errors, whether
+natural mortality (or the median for models with process error in
+natural mortality) was estimated, and whether a Beverton-Holt SRR was
+estimated within the EM. Details of each of the OMs and EMs are
+described below. We did not use the log-normal bias-correction feature
+for process errors or observations described by Stock and Miller (2021)
+for OMs and EMs to simplify interpretation of the study results (Li et
+al. 2025b). All code we used to perform the simulation study and
+summarize results can be found at
 \url{https://github.com/timjmiller/SSRTWG/tree/main/Project_0/code}.
@@
 age classes, ages 1 to 10+, with the last being a plus group that
-accumulates ages 10 and older. We assume spawning occurs annually 1/4 of
-the way through the year. The maturity at age was a logistic curve with
-\(a_{50}\) = 2.89 and slope = 0.88 (Figure \ref{om_inputs_fig}, top
-left).
-
-Weight at age was generated with a von Bertalanffy growth function \[
-L_a = L_{\infty}\left(1 - e^{-k(a - t_0)}\right)
-\] where \(t_0 = 0\), \(L_\infty = 85\), and \(k = 0.3\), and a L-W
-relationship such that \[
-W_a = \theta_1 L_a^{\theta_2}
+accumulates ages 10 and older. The maturity at age was a logistic curve
+with \(a_{50}\) = 2.89 and slope = 0.88 (Figure \ref{om_inputs_fig}, top
+left). Weight at age (\(W_a\)) was generated with a von Bertalanffy
+growth function defining length at age: \[
+L_a = L_{\infty}\left(1 - e^{-k(a - t_0)}\right),
+\] where \(t_0 = 0\), \(L_\infty = 85\), and \(k = 0.3\), and a
+length-weight relationship such that \[
+W_a = \theta_1 L_a^{\theta_2},
 \] where \(\theta_1 = e^{-12.1}\) and \(\theta_2 = 3.2\) (Figure
@@
 We assumed a Beverton-Holt SRR with constant pre-recruit mortality
-parameters for all OMs. All biological inputs to calculations of
-spawning biomass per recruit (i.e., weight, maturity, and natural
-mortality at age) are constant in the apparent survival (R+S)
-selectivity (R+Sel), and survey catchability (R+q) process error OMs.
-Therefore, steepness and unfished recruitment are also constant over the
-time period for those OMs (Miller and Brooks 2021). We assumed a value
-of 0.2 for the natural mortality rate in OMs without process errors on
-natural mortality. We specified unfished recruitment equal to \(e^{10}\)
-and \(\Fmsy = F_{40\%} = 0.348\), which equates to a steepness of 0.69
-and \(a=0.60\) and \(b = 2.4 \times 10^{-5}\) for the Beverton-Holt
-parameterization \[
+parameters for all OMs. We assume spawning occurs annually 0.25 of each
+year and recruitment at age 1 (\(N_{1,y}\)). All biological inputs to
+calculations of spawning stock biomass (SSB) per recruit (i.e., weight,
+maturity, and natural mortality at age) are constant in the R+S, R+Sel,
+and R+q process error OMs. Therefore, steepness and unfished recruitment
+are also constant over the time period for those OMs (Miller and Brooks
+2021). We assumed a value of 0.2 for the natural mortality rate in OMs
+without process errors on natural mortality. We specified unfished
+recruitment equal to \(e^{10}\) and \(\Fmsy = F_{40\%} = 0.348\), which
+equates to a steepness of 0.69 and \(a=0.60\) and
+\(b = 2.4 \times 10^{-5}\) for the Beverton-Holt parameterization \[
 N_{1,y} = \frac{a \text{SSB}_{y-1}}{1 + b \text{SSB}_{y-1}} 
@@
 equates to a steepness and R0 at the median of the time series process
-for natural mortality. Similarly, For OMs with time-varying random
+for natural mortality. Similarly, for OMs with time-varying random
 effects for fishery selectivity, \Fmsy also varies temporally, so
@@
 implies that, for a deterministic model, the abundance at age would not
-change from the first year to the next.
+change from year to year at the beginning of the time series.
 
@@
 The standard deviation for log-aggregate catch was 0.1 for all OMs, a
-common assumption for NEUS stock assessments. Two levels of observation
-error variance (high and low) were specified for indices and all age
-composition observations (both indices and catch). The low uncertainty
-specification assumed a standard deviation of 0.1 for both series of
-log-aggregate index observations, and the standard deviation of the
-logistic-normal for age composition observations was 0.3. In the high
-uncertainty specification, the standard deviation for log-aggregate
-indices was 0.4 and that for the age composition observations was 1.5.
-The low standard deviation for index observations is typical for fish
-stocks that are consistently sampled across survey stations whereas the
-high value is typical for more sporadically sampled stocks. The standard
-deviations for the age composition observations were determined from the
-range of values estimated from WHAM fits to NEUS stocks that assumed the
-logistic-normal model. For all EMs, the standard deviation for
-log-aggregate observations was assumed known whereas that for the
-logistic-normal age composition observations was estimated.
+common assumption for commercial removals in NEUS stock assessments. Two
+levels of observation error variance (high and low) were specified for
+indices and all age composition observations (both indices and catch).
+The low uncertainty specification assumed a standard deviation of 0.1
+for both series of log-aggregate index observations, and the standard
+deviation of the logistic-normal for age composition observations was
+0.3. In the high uncertainty specification, the standard deviation for
+log-aggregate indices was 0.4 and that for the age composition
+observations was 1.5. The low standard deviation for index observations
+is typical for fish stocks that are consistently sampled across survey
+stations whereas the high value is typical for more sporadically sampled
+stocks. The standard deviations for the age composition observations
+were determined from the range of values estimated from WHAM fits to
+NEUS stocks that assumed the logistic-normal model. For all EMs, the
+standard deviation for log-aggregate observations was assumed known
+whereas that for the logistic-normal age composition observations was
+estimated.
 
@@
 
-For operating models with random effects on recruitment and(or) apparent
-survival (R, R+S), we assumed marginal standard deviations for
+For operating models with random effects on recruitment only and also on
+apparent survival (R, R+S), we assumed marginal standard deviations for
 recruitment of \(\sigma_R \in \{0.5,1.5\}\). The marginal standard
@@
 
-WHAM treats selectivity parameter \(s\) as a logit-transformed parameter
-\[
+WHAM treats each selectivity parameter \(s\) as a logit-transformed
+parameter \[
 \log\left(\frac{p_{s,y}-l_{s}}{u_{s}-p_{s,y}}\right) = \mu_s + \epsilon_{s,y}
@@
 specified by the user. All selectivity parameters (\(a_{50}\) and slope
-parameters) were bounded by 0 and 10 for all OMs and EMs. The marginal
-standard deviations we assumed for logit scale random effects were
-\(\sigma_s \in \{0.1, 0.5\}\) and AR1 autocorrelation parameters of
-\(\rho_s \in \{0,0.9\}\). Like R+M OMs, the full factorial combination
-of these process error assumptions (2x2 levels) and scenarios described
-above for fishing history (2 levels) and observation error (2 levels)
-results in 16 different R+Sel OMs (Table \ref{sel_om_table}).
+parameters) were bounded by \(s_l = 0\) and \(s_u = 10\) for all OMs and
+EMs. The marginal standard deviations we assumed for logit scale random
+effects were \(\sigma_s \in \{0.1, 0.5\}\) and AR1 autocorrelation
+parameters of \(\rho_s \in \{0,0.9\}\). Like R+M OMs, the full factorial
+combination of these process error assumptions (2x2 levels) and
+scenarios described above for fishing history (2 levels) and observation
+error (2 levels) results in 16 different R+Sel OMs (Table
+\ref{sel_om_table}).
 
@@
 
-Subsets of 20 EMs in Table \ref{em_table} were fit to simulate data sets
-from each of the OM process error categories. For R and R+S OMs, fitted
-EMs had matching process error assumptions as well as R+Sel, R+M, and
-R+q assumptions without autocorrelation. For other OM process error
-categories, we fit EMs with correct process error assumptions, the
-correct process error type but incorrect correlation assumption, and the
-incorrect process error types without autocorrelation. As such, EMs were
-configured completely correctly for the OM, or they had
-mis-specification in assumptions of process error autocorrelation, the
-type of process error, and(or) the SRR (Beverton-Holt or none).
-
-The maturity at age, weight at age for catch and spawning stock biomass
-(SSB), and observation error standard deviations for aggregate catch and
-indices were all assumed known at the true values. However, the variance
-parameters for the logistic-normal distributions for age composition
-observations were estimated in the EMs.
+Subsets of 20 EMs in Table \ref{em_table} were fit to simulated data
+sets from each of the OM process error sources. For R and R+S OMs,
+fitted EMs had matching process error assumptions as well as R+Sel, R+M,
+and R+q assumptions without autocorrelation. For other OM process error
+sources, we fit EMs with correct process error assumptions, the correct
+process error source but incorrect correlation assumption, and the
+incorrect process error source without autocorrelation. As such, EMs
+were configured correctly for the OM, or they had mis-specification in
+assumptions of process error autocorrelation, the source of process
+error, and(or) the SRR (Beverton-Holt or none).
+
+The maturity at age, weight at age for catch and SSB, and observation
+error standard deviations for aggregate catch and indices were all
+assumed known at the true values. However, the variance parameters for
+the logistic-normal distributions for age composition observations were
+estimated in the EMs.
 
@@
 purposes. First, we analyzed selection of each process error model
-structure (R, R+S, R+M, R+Sel, R+q) using marginal AIC. For a given OM
+source (R, R+S, R+M, R+Sel, R+q) using marginal AIC. For a given OM
 simulated data set, we compared AIC for EMs with different process error
@@
 Contrast in fishing pressure and time series with recruitment at low
-stock size has been shown to improve estimation of SRR parameters
+stock size have been shown to improve estimation of SRR parameters
 (Magnusson and Hilborn 2007; Conn et al. 2010). Our preliminary
 inspections indicated generally poor performance of AIC in determining
-the Beverton-Holt model for a given set of OM factors (including
+the Beverton-Holt SRR model for a given set of OM factors (including
 contrast in fishing pressure), even when the EM was configured with the
-correct process error type. Therefore, we conditioned on the EMs having
-the correct process error assumption and also considered the effect of
-the log-standard deviation of the true log(SSB)
+correct process error source. Therefore, we conditioned on the EMs
+having the correct process error assumption and also considered the
+effect of the log-standard deviation of the true log(SSB)
 (\(\log \text{SD}_{\text{SSB}}\); similar to the log of the coefficient
@@
 \text{RE}\left(\theta_j\right) = \frac{\widehat \theta_j - \theta_j}{\theta_j}
-\end{equation} from fitting a given estimating model to simulated data
-set \(j\) configured for a given OM where \(\widehat \theta_j\) and
-\(\theta_j\) are the estimated and true values for simulation \(j\). We
-analyzed simulation results for estimates of terminal year estimates of
-SSB and recruitment, Beverton-Holt SRR parameters (\(a\) and \(b\)), and
-median natural mortality rate.
+\end{equation} from fitting a given EM to simulated data set \(j\)
+configured for a given OM where \(\widehat \theta_j\) and \(\theta_j\)
+are the estimated and true values for simulation \(j\). We analyzed
+simulation results for estimates of terminal year SSB and recruitment,
+Beverton-Holt SRR parameters (\(a\) and \(b\)), and median natural
+mortality rate.
 
@@
 and \(\widehat \theta_{y,y'}\) is the estimate for attribute \(\theta\)
-in year \(y\) from a model fit using data up to year \(y'\geq y\).
+in year \(y\) from a model fit using data up to year \(y'\geq y\). Thus,
+\(\theta\) terms where \(y'=Y\) refer to estimates from the fit to all
+years of data.
 
@@
 used two methods to summarize the most important factors for differences
-in results within a given OM process error type. The first method was
+in results within a given OM process error source. The first method was
 fitting regression models with the response being each of the measures
@@
 al. 2017; Harwell et al. 2018). For the binary indicators of convergence
-and AIC-based selection of a SRR, we performed logistic regressions. For
-indicators of AIC-based selection of EM process error type (multiple
-categories) we performed multinomial regressions. For other measures of
-reliability we fit linear regression models to transformed responses.
-Because relative errors (Eq. \ref{relerror}) and Mohn's \(\rho\) for the
-various parameters are bounded below at -1, we used a transformation of
-these values \begin{equation}\label{bias_regression_response}
+and AIC-based selection of an SRR, we performed logistic regressions.
+For indicators of AIC-based selection of EM process error source
+(multiple categories) we performed multinomial regressions. For other
+measures of reliability we fit linear regression models to transformed
+responses. Because relative errors (Eq. \ref{relerror}) and Mohn's
+\(\rho\) for the various parameters are bounded below at -1, we used a
+transformation of these values
+\begin{equation}\label{bias_regression_response}
 y_j = \log\left[f\left(\widehat \theta_j,\theta_j\right)+1\right]
@@
 the VGAM package (Yee 2008; Yee 2015). We tabulated percent reduction in
-residual deviance for each of regression fits. We did not perform formal
-statistical analyses of effects of OM and EM attributes on results
-(e.g., ANOVA) because of the lack of independence of the
+residual deviance for each of the regression fits. We did not perform
+formal statistical analyses of effects of OM and EM attributes on
+results (e.g., ANOVA) because of the lack of independence of the
 ``observations'' that results from fitting multiple EMs to each
@@
 (Breiman et al. 1984) to show how the OM and EM attributes, and their
-interactions, split the values for each measure of reliability (e.g.,
-Gonzalez et al. 2018; Collier et al. 2022). We used classification trees
-for categorical measures (convergence and AIC) and regression trees for
-the other measures with continuous scales (relative error and Mohn's
-\(\rho\)). The response variables were the same as the regressions for
-the deviance reduction analyses. We used the \verb|rpart| function in
-the rpart package (\url{https://cran.r-project.org/package=rpart}) to
-fit trees. Full trees were determined using default settings except that
-we increased the number of cross-validations to 100. For clarity, we
-pruned the full trees to show just the primary branches.
+interactions, partition the values for each measure of reliability
+(e.g., Gonzalez et al. 2018; Collier et al. 2022). We used
+classification trees for categorical measures (convergence and AIC) and
+regression trees for the other measures with continuous scales (relative
+error and Mohn's \(\rho\)). The response variables were the same as the
+regressions for the deviance reduction analyses. We used the
+\verb|rpart| function in the rpart package (Therneau and Atkinson 2025)
+to fit trees. Full trees were determined using default settings except
+that we increased the number of cross-validations to 100. For clarity,
+we manually pruned the full trees to show just the primary branches.
 
@@
 Clopper-Pearson exact method (Clopper and Pearson 1934; Thulin 2014).
-For AIC selection of process error configuration we provide estimates of
-the proportions of simulations where each EM type was selected. For AIC
+For AIC selection of process error source we provide estimates of the
+proportions of simulations where each EM type was selected. For AIC
 selection of the SRR (a binary indicator for each simulated data set),
@@
 
-For probability of convergence, EM process error assumption was the
+For probability of convergence, the EM process error assumption was the
 single attribute that resulted in the largest percent reduction in
-deviance (14-28\%) for all OM process error types other than R+S OMs
+deviance (14-28\%) for all OM process error sources other than R+S OMs
 where the EM median natural mortality rate assumption (estimated or
@@
 
-Classification trees for each OM process error type, all had the primary
-branch defined using the same attribute that provided the largest
-reduction in deviance (Figure \ref{conv_class}). EMs that assumed R+S
-process errors converged poorly for all OMs that were simulated with the
-alternative process error assumptions (R, R+M, R+Sel, an R+q OMs). For
-all trees, branches based on the OM fishing mortality history showed
-better convergence when the OM included a change in fishing pressure.
-Branches based on whether the Beverton-Holt SRR was assumed or not,
-showed better convergence when it was not estimated and branches based
-on the median natural mortality rate assumption showed better
-convergence when it was treated as known. For some R+M and R+Sel OMs,
-better convergence was also observed when there was lower observation
-uncertainty.
+Classification trees for each OM process error source all had the
+primary branch defined using the same attribute that provided the
+largest reduction in deviance (Figure \ref{conv_class}). EMs that
+assumed R+S process errors converged poorly for all OMs that were
+simulated with the alternative process error assumptions (R, R+M, R+Sel,
+and R+q OMs). For all trees, branches based on the OM fishing mortality
+history showed better convergence when the OM included a change in
+fishing pressure. Branches based on whether the Beverton-Holt SRR was
+assumed or not, showed better convergence when it was not estimated and
+branches based on the median natural mortality rate assumption showed
+better convergence when it was treated as known. For some R+M and R+Sel
+OMs, better convergence was also observed when there was lower
+observation uncertainty.
 
@@
 explaining deviance reduction is the same that for Hessian-based
-convergence for all OM process error types, but there are some
+convergence for all OM process error sources, but there are some
 differences in deviance reduction for secondary factors (Table
@@
 converged models had values greater than 1. For many OMs, EMs that
-assumed the correct process error type and did not estimate median
+assumed the correct process error source and did not estimate median
 natural mortality or the Beverton-Holt SRR produced the lowest gradient
@@
 
-\subsubsection*{Process error structure}\label{process-error-structure}
-\addcontentsline{toc}{subsubsection}{Process error structure}
+\subsubsection*{Process error source}\label{process-error-source}
+\addcontentsline{toc}{subsubsection}{Process error source}
 
@@
 that resulted in the largest percent reductions in deviance across OM
-process error types other than R OMs (Table \ref{AIC_PE_PRD_table}).
+process error sources other than R OMs (Table \ref{AIC_PE_PRD_table}).
 Both sources of variation explained large reductions in deviance for R+S
@@
 interactions, did not provide large reductions in deviance for any of
-the OM process error types.
-
-For all OM process error types other than R OMs, the attributes defining
-the primary branches of classification trees matched those that provided
-the largest reductions in deviance (Figure \ref{AIC_PE_class}). Across
-all OMs, AIC was more accurate for the process error type when process
-error variability was greater and when observation error was lower. For
-R+S OMs, there was a tendency to select R OMs when observation error was
-higher and apparent survival variation was lower
-(\(\sigma_{2+} = 0.25\)), but accuracy for the process error was
-otherwise highly accurate. Similarly, low accuracy for process error
-type occurred for R+Sel and R+q OMs only with high observation error and
-lower process error variability, but high accuracy for R+M OMs required
-both low observation error and high process error variance. This was
-also required for accurate identification of the correct correlation
-structure for R+M, R+q, and R+M OMs (Figure \ref{pe_aic}). No branches
-were estimated for classification trees fit to the R OMs, likely because
-accuracy was high across all simulations (0.94), although inspection of
-the fine-scale results shows there is some degradation in AIC selection
-when a SRR and median natural mortality rate are estimated for R OMs
-with constant fishing pressure and high observation error (Figure
-\ref{pe_aic}, top left).
+the OM process error sources.
+
+For all OM process error sources other than R OMs, the attributes
+defining the primary branches of classification trees matched those that
+provided the largest reductions in deviance (Figure \ref{AIC_PE_class}).
+Across all OMs, AIC was more accurate for the process error source when
+process error variability was greater and when observation error was
+lower. For R+S OMs, there was a tendency to select R OMs when
+observation error was higher and apparent survival variation was lower
+(\(\sigma_{2+} = 0.25\)), but accuracy for the process error source was
+otherwise highly accurate. Larger variability of process error relative
+to observation error was also required for accurate identification of
+the correct correlation structure for R+M, R+q, and R+Sel OMs (Figure
+\ref{pe_aic}). No branches were estimated for classification trees fit
+to the R OMs, likely because accuracy was high across all simulations
+(0.94), although inspection of the fine-scale results shows there is
+some degradation in AIC selection when an SRR and median natural
+mortality rate are estimated for R OMs with constant fishing pressure
+and high observation error (Figure \ref{pe_aic}, top left).
 
@@
 for both R (\textgreater30\%) and R+S (\textasciitilde19\%) OMs. Further
-fits for R and R+S OMs including all combinations two factors showed
-those with \(\log \text{SD}_{\text{SSB}}\) and recruitment variation
-provided essentially the same reduction in deviance as the models with
-all factors. For all OM process error types, inclusion of interaction
+fits for R and R+S OMs that including different combinations of two
+factors additively showed fits that included
+\(\log \text{SD}_{\text{SSB}}\) and recruitment variation only provided
+essentially the same reduction in deviance as the models with all
+factors. For all OM process error sources, inclusion of interaction
 terms provided relatively little reduction in residual deviance.
@@
 variability in SSB and all branches based on fishing history showed
-better accuracy when there was contrast in fishing pressure. Branched
+better accuracy when there was contrast in fishing pressure. Branches
 based on OM observation error or recruitment variability (R and R+S OMs)
@@
 recruitment variability and observation error and higher SSB variability
-produced AIC accuracy of 0.79. For R+M, R+Sel, and R+q OM,s accuracy of
+produced AIC accuracy of 0.79. For R+M, R+Sel, and R+q OMs, accuracy of
 0.87 to 0.94 was observed with just increased SSB variability.
@@
 OM and EM factors showed little reduction in deviance (\textless5\%) for
-any of the factors across all OM process error types (Table
+any of the factors across all OM process error sources (Table
 \ref{bias_SSB_PRD_table}). The attributes producing the largest
 reductions were the EM assumption for median natural mortality (known or
-estimated) for R, R+M, R+Sel, and R+q OMs (1-3\%), EM process error type
-for R+S OMs (4\%) and fishing history for all OM types (1-5\%).
-Including second order interactions provided the largest reductions in
-residual deviance (10- 26\%). Including third order interactions also
-provided further reductions for R, R+S, and R+q OMs between 5 and 11\%.
+estimated) for R, R+M, R+Sel, and R+q OMs (1-3\%), EM process error
+assumption for R+S OMs (4\%) and fishing history for all OM process
+error sources (1-5\%). Including second order interactions provided the
+largest reductions in residual deviance (10- 26\%). Including third
+order interactions also provided further reductions for R, R+S, and R+q
+OMs between 5 and 11\%.
 
@@
 estimated. For some R+Sel and R+q OMs, less bias in SSB was shown when
-the EM process error configuration was correct.
+the EM process error assumption was correct.
 
@@
 \ref{bias_SR_pars_PRD_table}). The OM fishing history provided the
-largest deviance reduction for most OM process error types for both
-parameters, but reductions were less than 5.6\% except for R+Sel OMs
-where the reductions were 11.37\% and 7.97\% for the \(a\) and \(b\)
-parameters, respectively and for just the \(b\) parameter for R+q OMs
-(10\%). The EM process error assumption provided similar reductions in
-deviance for both parameters for R OMs. Including interactions also did
-not produce important reductions in deviance.
+largest deviance reduction for most OM process error sources for both
+parameters, but reductions were generally less than 6\%. Exceptions were
+the R+Sel OMs where \(a\) and \(b\) were reduced by approximately 11\%
+and 8\%, respectively, and the R+q OMs where \(b\) was reduced by 10\%.
+The EM process error assumption provided similar reductions in deviance
+for both parameters for R OMs. Including interactions also did not
+produce important reductions in deviance.
 
@@
 fishing pressure for all branches in trees for each OM process error
-type (Figures \ref{SR_a_bias_regtree} and \ref{SR_b_bias_regtree}). For
-all branches based on recruitment variability in trees for R and R+S
+source (Figures \ref{SR_a_bias_regtree} and \ref{SR_b_bias_regtree}).
+For all branches based on recruitment variability in trees for R and R+S
 OMs, less bias in both \(a\) and \(b\) was observed with less
@@
 are near zero unless a large combinations of OM and EM conditions occur
-(Figure \ref{SSB_mohns_rho_regtree}). For example, In R+S OMs, with
+(Figure \ref{SSB_mohns_rho_regtree}). For example, in R+S OMs, with
 constant fishing pressure, high observation error, and higher apparent
@@
 
-Analyses of model convergence across simulations can be useful for
-understanding the utility of alternative convergence criteria used in
-applications to real data for directing the practitioner to more
-appropriate random effects configurations. It is common during the
-assessment model fitting process to check that the maximum absolute
-gradient component is less than some threshold prior to inspecting the
-Hessian of the optimized likelihood for invertibility (Carvalho et al.
-2021). However, there is no accepted standard for the gradient threshold
+Poor convergence was common in our results when the incorrect process
+error source was assumed. Li et al. (2024) found that convergence could
+be a useful diagnostic especially for separating the correct simpler
+process error assumption from overly complex models. Poor convergence
+often occurs when parameter estimates are at their bounds (Carvalho et
+al. 2021). However, even when the Hessian is invertible for a converged
+model, parameters that are poorly informed will have extremely large
+variance estimates. This further inspection can lead to a more
+appropriate and often more parsimonious model configuration where the
+problematic parameters are not estimated. For example, process error
+variance parameters in state-space models that are estimated close to 0
+indicates that the random effects are estimated to have little or no
+variability and removing these process errors is warranted. Our
+experiments did not aim to emulate the practitioner decision process in
+determining an appropriate model configurations, but evaluating the
+efficacy of such a decision process when applying EMs might be important
+in closed loop simulations aimed at quantifying management performance
+(e.g., a management strategy evaluation).
+
+It is common during the assessment model fitting process to check that
+the maximum absolute gradient component is less than some threshold
+prior to inspecting the Hessian of the optimized likelihood for
+invertibility (Carvalho et al. 2021), but we found reliance on magnitude
+of the gradient values for fitted models as a convergence criterion
+questionable. There is no accepted standard for the gradient threshold
 (e.g., Lee et al. 2011; Hurtado-Ferro et al. 2014; Rudd and Thorson
-2018) and some thresholds would exclude models that in fact have an
-invertible Hessian. We found the Hessian at the optimized log-likelihood
-can often be invertible when the maximum absolute gradient was much
-larger than what might be perceived to be a sensible threshold.
-
-Li et al. (2024) found that convergence rate could be a useful
-diagnostic especially for separating the correct model from overly
-complex models. However, the criteria for convergence used in their
-study may also lead to limited ability to distinguish the correct model
-from overly simplistic models, a pattern that was also noted by
-Liljestrand et al. (2024) in which one process error may absorb all
-sources of process error when the magnitude of other process errors are
-low.
-
-Often poor convergence occurs when parameter estimates are at their
-bounds (Carvalho et al. 2021), and this also applies to variance
-parameters for random effects with state-space assessment models. Even
-when the Hessian is invertible, parameters that are poorly informed will
-have extremely large variance estimates. This further inspection can
-lead to a more appropriate and often more parsimonious model
-configuration where the problematic parameters are not estimated. For
-example, process error variance parameters that are estimated close to 0
-indicates that the random effects are estimated to have little or no
-variability and removing these process errors is warranted. Generally,
-our results suggest we can expect lower probability of convergence of
-state-space assessment models when estimating natural mortality or SRRs
-because of the difficulty distinguishing these parameters from others
-being estimated in assessment model in common scenarios where data
-quality is less than ideal. Our experiments did not aim to emulate the
-practitioner decision process in developing model configurations
-(e.g.~removing a source of process error and refitting the model when
-process error variance parameters were estimated close to 0). Evaluating
-the efficacy of such a decision process when applying EMs might be
-important in closed loop simulations (e.g.~MSE) aimed at quantifying
-management performance.
+2018), but the Hessian at the optimized log-likelihood was often
+invertible when the maximum absolute gradient was much larger than what
+might be perceived to be a sensible threshold in some of our
+simulations. Therefore, the gradient criterion could exclude models that
+in fact have an invertible Hessian.
 
@@
 
-We found accuracy of marginal AIC for process error type required only
-low observation error for R, R+S, R+Sel, and R+q OMs. R+M OMs further
-required higher process error variability, but this also improved
-accuracy for the other OM process errors types when there was higher
-observation error.
+We found accuracy of marginal AIC-based selection for the correct
+process error source required only low observation error for R, R+S,
+R+Sel, and R+q OMs. R+M OMs further required higher process error
+variability, but this also improved accuracy for the other OM process
+errors sources when there was higher observation error. These results
+seem consistent with Li et al. (2024). Their simulation studies
+investigated models with multiple process error sources and found good
+accuracy of AIC in detecting correct process error assumptions for
+simulations based on two stocks (Gulf of Maine cod and Southern New
+England-Mid-Atlantic yellowtail founder) that are well sampled by NEUS
+bottom trawl surveys used as indices in the respective assessments and
+poor accuracy for Atlantic mackerel, a semi-pelagic species that is
+observed relatively poorly.
 
@@
 Variation in SSB was the most important factor for using marginal AIC to
-distinguish the the Beverton-Holt SRR. For R+M, R+Sel, and R+q OMs, the
-SRR was accurately detected when the CV of SSB over the time series was
-at least 40 to 50\% (\(\log \text{SD}_{\text{SSB}}\) = -0.9 to -0.7)
-regardless of any other OM or EM attributes. Detection of the SRR for R
-and R+S OMs required lower recruitment variability, but this lower level
-(\(\sigma_R = 0.5\)) was assumed for all of the other OM process error
-types. Our results assumed that the EM process error configuration was
-correct, but this may be a strong limation given the ability of AIC to
-distinguish the process error type in many scenarios.
+correctly distinguish the Beverton-Holt SRR from the null model without
+an SRR. For R+M, R+Sel, and R+q OMs, the SRR was accurately detected
+when the CV of SSB over the time series was at least 40 to 50\%
+(\(\log \text{SD}_{\text{SSB}}\) = -0.9 to -0.7) regardless of any other
+OM or EM attributes. Detection of the SRR for R and R+S OMs required
+lower recruitment variability, but this lower level (\(\sigma_R = 0.5\))
+was assumed for all of the other OM process error source and represents
+the lower range of estimates from recent NEUS stock assessments. Our
+results assumed that the EM process error configuration was correct, but
+this may not be a strong limitation given the ability of AIC to
+distinguish the process error source in many scenarios.
 
 Although we did not compare models with alternative SRRs (e.g., Ricker
-and Beverton-Holt), we do not expect AIC to perform any better
+vs.~Beverton-Holt), we do not expect AIC to perform any better
 distinguishing between relationships and may be more difficult than
@@
 cases contrasts with the noted bias in AIC for more complex models
-(Shibata 1976; Katz 1981; Kass and Raftery 1995), but, whereas those
+(Shibata 1976; Katz 1981; Kass and Raftery 1995). However, these earlier
 findings apply to the much more common comparison of models that are fit
-to raw and independent observations, here we are comparing state-space
-models which account for observation error and estimate process errors
-in latent variables.
+to raw and independent observations, whereas our comparisons of
+state-space models account for observation error and separately estimate
+process errors in latent variables.
 
@@
 found AIC could not distinguish among state-space SRRs that were fit
-just to SSB and recruitment observations (i.e., not an assessment
+just to SSB and recruitment observations (i.e., not within an assessment
 model). Similarly, Britten et al. (In review) found AIC could not
-reliably distinguish alternative environmental effects on SRR
-parameters. However, Miller et al. (2016) did find AIC to prefer a SRR
-with environmental effects when applied to data for the SNEMA yellowtail
-flounder stock and AIC also selected an environmental covariate on a SRR
-for the most recent stock assessment of Georges Bank yellowtail flounder
-(NEFSC 2025). Both of these yellowtail flounder stocks have large
-changes in stock size and the values of environmental covariates over
-time. Additionally, this species is well-observed by the bottom trawl
-survey that is used for an index in assessment models.
-
-However, estimation of SRR parameters was only moderately reliable in
-ideal scenarios of low observation error and contrast in fishing for
-R+Sel and R+M OMs with large temporal variability in process errors.
-Otherwise, SRR parameter estimation was biased and(or) highly variable.
-We found substantial bias in estimated SRR parameters in R and R+S OMs
+reliably distinguish the Beverton-Holt SRR from no SRR, nor identify
+alternative environmental effects on SRR parameters. However, Miller et
+al. (2016) did find AIC to prefer an SRR with environmental effects when
+applied to data for the Southern New England-Mid-Atlantic yellowtail
+flounder stock and AIC also selected an environmental covariate on an
+SRR for the most recent stock assessment of Georges Bank yellowtail
+flounder (NEFSC 2025). Both of these yellowtail flounder stocks have
+large changes in stock size and the values of environmental covariates
+over time. Additionally, this species is well-observed by the bottom
+trawl survey that is used for an index in assessment models.
+
+Estimation of SRR parameters was only moderately reliable in ideal
+scenarios of low observation error and contrast in fishing for R+Sel and
+R+M OMs with large temporal variability in process errors. Otherwise,
+SRR parameter estimation was biased and(or) highly variable. We found
+substantial bias in estimated SRR parameters in R and R+S OMs
 particularly with high variability in recruitment and apparent survival
-process errors, suggesting that practitioners should be cautious SRR
-inferences when fitted assessment models have these properties. We only
-evaluated effects of SSB variability on accuracy of AIC in identifying
-the SRR, but those results suggests we might find less bias for the SRR
-parameters in such cases as well. Similarly, restricting results to fits
-that converged may also yield better accuracy of SRR parameter
-estimation.
+process errors, suggesting that practitioners should be cautious with
+SRR inferences when fitted assessment models have these properties. We
+only evaluated effects of SSB variability on accuracy of AIC in
+identifying the SRR, but those results suggests we might find less bias
+for the SRR parameters in such cases as well. Another condition that
+could improve perception of bias in our simulation studies is
+restricting results to fits that converged with Hessian-based standard
+errors for all parameters, but Britten et al. (In review) did not find
+less SRR parameter bias when restricting estimates using a
+gradient-based criterion. A simulation study by Stock and Miller (2021)
+examining configurations of environmental covariate effects on a
+Beverton-Holt SRR for the previously mentioned, well-observed, Southern
+New England-Mid-Atlantic yellowtail flounder stock found little or no
+bias for the density-independent mortality parameter \(a\), but still
+biased estimation of the density-dependent parameter \(b\).
 
@@
 retrospective patterns for SSB for any OMs regardless of whether median
-natural mortality or a SRR was estimated although some weak patterns
+natural mortality or an SRR was estimated, although some weak patterns
 occurred when observation error was high and there was contrast in
@@
 2017; Huynh et al. 2022; Breivik et al. 2023). In our simulations, the
-process errors changing over time may have trends in particular
+process errors changing over time may have trends in certain
 simulations, particularly when strong autocorrelation is imposed, but
@@
 retrospective patterns when the source of mis-specification was temporal
-variation in demography attributes. Indeed, it is common for the
+variation in demographic attributes. Indeed, it is common for the
 flexibility provided by temporal random effects to reduce retrospective
@@
 examined across all simulations. The classification and regression trees
-are generally a good tool for determining the values of the OM and EM
-attributes that produce better or worse measures of reliability.
-However, determining the combination of attributes that produce the best
-or worst measures of reliability can be challenging using the trees
-alone. For example, in the regression tree for median natural mortality
-rate estimates in R OMs (Figure \ref{med_M_bias_regtree}), both of the
-first branches imply bias is low regardless of OM fishing history, but
-when OM fishing pressure is constant, results are much better when OM
-observation error is low (median RE about -6\%) than when OM observation
-error is high (median RE about 40\%). The default pruning of the trees
-can exclude these lower branches. However, inspection of deviance
-explained by various regression models shows the \textasciitilde9\%
-reduction in residual deviance by including second order interaction of
-all OM and EM factors (Table \ref{bias_median_M_PRD_table}), indicating
-that the interaction of factors may be important, thereby complimenting
-the regression tree analysis. Higher order interactions of some factors
-could also provide reductions in deviance and, therefore, inspection of
-results for each combinations of OM and EM factors, as provided in the
-Supplementary Materials, can also be important.
+are generally a good tool for determining the OM and EM attributes that
+produce better or worse measures of reliability. However, determining
+the combination of attributes that produce the best or worst measures of
+reliability can be challenging using the trees alone. For example, in
+the regression tree for median natural mortality rate estimates in R OMs
+(Figure \ref{med_M_bias_regtree}), both of the first branches imply bias
+is low regardless of OM fishing history, but when OM fishing pressure is
+constant, results are much better when OM observation error is low
+(median RE about -6\%) than when OM observation error is high (median RE
+about 40\%). The default pruning of the trees can exclude these lower
+branches. However, inspection of deviance explained by various
+regression models shows the \textasciitilde9\% reduction in residual
+deviance by including second order interaction of all OM and EM factors
+(Table \ref{bias_median_M_PRD_table}), indicating that the interaction
+of factors may be important, thereby complimenting the regression tree
+analysis. Higher order interactions of some factors could also provide
+reductions in deviance and, therefore, inspection of results for each
+combinations of OM and EM factors, as provided in the Supplementary
+Materials, can also be important.
 
@@
 not heavily weight the magnitude of the gradient values in determining
-convergence as long as the maximum absolute values is around 1 or lower.
+convergence as long as the maximum absolute value is around 1 or lower.
 Instead, positive-definiteness of the Hessian of the minimized negative
@@
 to a limited number of managed stocks. Furthermore, some bias in
-estimation of the SRR parameters should be expected, which presumably
-also applies to MSY-based reference points. Because bias in terminal SSB
-and retrospective patterns were indifferent to whether or not the SRR
-was estimated, the prevalence of bias in SRR parameter estimation, and
-often better convergence without the SRR, we recommend a sensible
-default is to exclude a SRR when fitting assessment models, as also
-suggested by Brooks (2024).
+estimation of the SRR parameters (and MSY-based reference points should
+be expected. Because bias in terminal SSB and retrospective patterns
+were indifferent to whether or not the SRR was estimated, the prevalence
+of bias in SRR parameter estimation, and often better convergence
+without the SRR, we recommend a sensible default is to exclude an SRR
+when fitting assessment models, as also suggested by Brooks (2024).
 
@@
 
-\bibitem[\citeproctext]{ref-lietalinreview_a}
+\bibitem[\citeproctext]{ref-lietal25a}
 Li, C., Deroba, J.J., Berger, A.M., Goethel, D.R., Langseth, B.J.,
-Schueller, A.M., and Miller, T.J. In press. Random effects on
+Schueller, A.M., and Miller, T.J. 2025a. Random effects on
 numbers-at-age transitions implictly account for movement dynamics and
 improve performance within a state-space stock assessment. Canadian
-Journal of Fisheries and Aquatic Sciences.
+Journal of Fisheries and Aquatic Sciences \textbf{82}.
+doi:\href{https://doi.org/10.1139/cjfas-2025-0092}{10.1139/cjfas-2025-0092}.
 
@@
 Li, C., Deroba, J.J., Miller, T.J., Legault, C.M., and Perretti, C.
-2025. Guidance on bias-correction of log-normal random effects and
+2025b. Guidance on bias-correction of log-normal random effects and
 observations in state-space assessment models. Canadian Journal of
@@
 
+\bibitem[\citeproctext]{ref-rpart}
+Therneau, T., and Atkinson, B. 2025. Rpart: {R}ecursive {P}artitioning
+and {R}egression {T}rees. Available from
+\url{https://github.com/bethatkinson/rpart}.
+
 \bibitem[\citeproctext]{ref-thompson36}
@@
 \end{table}
-\end{landscape}
 
 \begin{table}
-\caption{Distinguishing characteristics of the EMs and indication (+) of which OM process error categories (R, R+S, R+M, R+Sel, R+q) each EM configuration was fit.}\label{em_table}
+\caption{Distinguishing characteristics of the EMs and indication (+) of which OM process error sources (R, R+S, R+M, R+Sel, R+q) each EM configuration was fit.}\label{em_table}
 {\scriptsize \input{em_tab}}
@@
 
-\subsection*{Further Detailed Results}\label{further-detailed-results}
-\addcontentsline{toc}{subsection}{Further Detailed Results}
+\subsection*{Further Results}\label{further-results}
+\addcontentsline{toc}{subsection}{Further Results}
 
@@
 
+\begin{table}
+\caption{For each OM process error source (columns), percent reduction in deviance for linear regression models fit to transformed Mohn's $\rho$ values for each simulation (Eq.
+\ref{bias_regression_response}) for fishing mortality averaged over all age classes with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{mohns_rho_F_PRD_table}
+{\input{mohns_rho_F_PRD_table}}
+\end{table}
+
+\begin{table}
+\caption{For each OM process error source (columns), percent reduction in deviance for linear regression models fit to transformed Mohn's $\rho$ values for each simulation (Eq.
+\ref{bias_regression_response}) for recruitment with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{mohns_rho_R_PRD_table}
+{\input{mohns_rho_R_PRD_table}}
+\end{table}
+
 \end{document}

### 2025-12-19 Project_0/code/manuscript/convergence_results__type_3_summarize.R

@@ -1,349 +1,0 @@
-library(here)
-library(dplyr)
-library(tidyr)
-library(mgcv)
-library(gratia)
-library(Hmisc)
-#modified rpart.plot package to use plotmath in split labels...
-pkgload::load_all(file.path(here(),"../../rpart.plot"))
-
-
-df.oms = list(naa = readRDS(here("Project_0", "inputs", "df.oms.RDS")))
-df.oms$M = readRDS(here("Project_0", "inputs", "df.M.oms.RDS"))
-df.oms$Sel = readRDS(here("Project_0", "inputs", "df.Sel.oms.RDS"))
-df.oms$q = readRDS(here("Project_0", "inputs", "df.q.oms.RDS"))
-om.em.rows <- list(naa = 1:20, M = 5:24, Sel = c(5:20, 25:28), q = c(5:20, 29:32))
-df.ems = readRDS(here("Project_0", "inputs", "df.ems.RDS"))
-
-#convergence information:
-#1: 0/1 model finished optimizing
-#2: nlminb convergence flag
-#3: max gradient value
-#4: number of NaNs in SEs for parameters, 0 = good invertible hessian
-#5: maximum non-NaN SE estimate
-
-om_conv_res <- list(naa = readRDS(here("Project_0", "results", "naa_om_convergence_results.RDS")))
-# naa_om_conv_res <- readRDS(here("Project_0", "results", "naa_om_convergence_results.RDS"))
-om_conv_res$M <- readRDS(here("Project_0", "results", "M_om_convergence_results.RDS"))
-om_conv_res$Sel <- readRDS(here("Project_0", "results", "Sel_om_convergence_results.RDS"))
-om_conv_res$q <- readRDS(here("Project_0", "results", "q_om_convergence_results.RDS"))
-est_config <- c("M known, No SR", "M estimated, No SR", "M known, SR", "M estimated, SR")
-EM_process_error <- c("R","R+S","R+M (iid)","R+M (AR1)","R+Sel (iid)","R+Sel (AR1)","R+q (iid)","R+q (AR1)")
-
-conv_fn <- function(all, em_rows, om_type, df.ems, om_ind = NULL, est_ind = NULL, type = 3){
-  
-  #df.ems. is the subset of df.ems for the current om_type
[…]
